\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233780}
\author{Daniel Van Olmen\affiliation{Lancaster University} and Adri Breed\affiliation{North-West University Potchefstroom} and Ernst Kotzé\affiliation{Nelson Mandela University Port Elizabeth}}
\title{Negation in Afrikaans}
\abstract{This chapter is a description of negation in Afrikaans. We cover its standard negation construction (\textit{nie}\textsubscript{1} \textit{… nie}\textsubscript{2} ‘not’), which is used for subordinate clause negation and interrogative negation among other things too, as well as its dedicated strategies for imperative negation (\textit{moenie}\textsubscript{1} \textit{… nie}\textsubscript{2} ‘don’t!’) and certain types of negative stative predication (\textit{geen … nie}\textsubscript{2} ‘no’). Attention is also paid to phenomena such as negative indefinites (e.g. \textit{nooit … nie}\textsubscript{2} ‘never’), negative polarity items (e.g. \textit{veel} ‘much’), response particles (e.g. the use of \textit{nee} ‘no’ to deny the questioner’s presumed concerns), negative derivation (e.g. \textit{{}-loos} ‘-less’), the reinforcement of negation (e.g. \textit{glad nie}\textsubscript{1} \textit{… nie}\textsubscript{2} ‘not at all) and expletive negation (e.g. in clauses headed by \textit{tensy} ‘unless’).
\keywords{Afrikaans, two-part negation, negative imperative, negative stative predication, negative indefinite, expletive negation}}

\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document} 
\maketitle 

\section{The language} 
\label{sec:afr:1}
Afrikaans (ISO: \href{https://iso639-3.sil.org/code/afr}{afr}, Glottocode: \href{https://glottolog.org/resource/languoid/id/afri1274}{afri1274}) is primarily spoken in South Africa and Namibia. It is also the language of small minorities in many other Southern African countries, including Botswana and Zambia, and of immigrant communities in, for instance, Australia, Canada and even Argentina (see \citealt{Toit1995}). In South Africa, where it is one of the eleven official languages, it is spoken as a first language by approximately 6,740,000 people, making up 12\% of the population (\citealt{StatisticsSouthAfrica2016}), and as a second language by an estimated further ten million \citep[78]{Webb2002}. Afrikaans has a strong literary tradition as well as its own news and media outlets, is widely used in education (although it has been losing ground to English, particularly in the tertiary sector) and has arguably witnessed something of a revival in movies and music in recent times. In Namibia, where it is recognized as one of eight national languages alongside the official language of English, it is the mother tongue of circa 180,000 individuals or 8\% of the entire population (\cite{UN2014}). It also often serves as the only lingua franca between speakers of other languages.

In genealogical terms, Afrikaans may be described as a sister of Present-day Dutch or, in other words, as Indo-European and (West) Germanic. It is this genus’s “contact language par excellence” \citep[72]{McWhorter2007}, though. Arriving in the Cape in 1652, Early Modern Dutch began interacting with the indigenous Khoisan languages (e.g. \textit{dagga} ‘cannabis’) and with languages like Malay and Creole Portuguese (e.g. \textit{baie} ‘a lot, very’ and \textit{mielie} ‘corn’), spoken by the Asian and African slaves brought to the region (as well as with the languages of other early European migrants, such as French and German). It came into contact with the Bantu languages in the rest of Southern Africa later on (e.g. \textit{babelas} ‘hangover’) and has been in a relationship of mutual influence with English since the advent of the British in 1795 (e.g. \textit{worrie} ‘worry’). 

The role that these interactions, especially the initial ones, played in the development of Afrikaans (grammar) has been and remains a topic of much debate (see \citealt[249, 257--258]{AuweraOlmen2017}), mainly as to whether the language is a creole or not. A plausible compromise is that Afrikaans, whose standardization and codification started some sixty years before it became an official language of South Africa in 1925, is the product of two opposing forces. The Dutch spoken and passed on by the people of European descent and serving as the Cape’s superstrate formed a more conservative force. A more progressive one, involving morphological simplification and substrate influence, was the Dutch that the Khoi and the slaves learned “imperfectly” and most of them eventually switched to. Actually, the varying impact of Settler Dutch, Khoisan and Malay still manifests itself in the three dialects that are typically distinguished for Present-day Afrikaans – i.e., respectively, Eastern Border Afrikaans, Gariep Afrikaans (also known as Orange River Afrikaans) and Cape Afrikaans (for a detailed history of the language, see \citealt{Rensburg2012}). The focus here will be on General Afrikaans (see \citealt{Huyssteen2020}) and the way that it is used both formally and informally.\footnote{There is evidence of dialectal variation in negation. \citet[165]{Roberge2000}, for instance, observes that the second element of the normal two-part negation of the general language “is facultative in Orange River Afrikaans” (see also \citealt{Stell2009}).}

The rest of the present chapter will take the basic facts of Afrikaans grammar for granted (see \citealt{Donaldson1993} and \citeauthor{LandsbergenEtAl2014}'s \citeyear{LandsbergenEtAl2014}  \textit{Taalportaal} website for a comprehensive description). One of these is that the language is V2 in main clauses (and complementizer-less subordinate clauses). This fact means that: first, the finite verb occupies second position, like \textit{het} ‘has/have’ in \REF{ex:afr:1a} or \REF{ex:afr:1b}; second, the subject usually precedes it, as in \REF{ex:afr:1a}, or immediately follows it if another constituent is in initial place, as in \REF{ex:afr:1b}; third, any non-finite verb (or any separable verb particle) appears in clause-final position, like \textit{gesien} ‘seen’ in \REF{ex:afr:1a} or \REF{ex:afr:1b}; and, fourth, the object normally occurs between the finite and non-finite slots, like \textit{die kar} ‘the car’ in \REF{ex:afr:1a} or \REF{ex:afr:1b}. In subordinate clauses, Afrikaans is SOV, as \REF{ex:afr:1c} shows.

\ea\label{ex:afr:1}
    
    \ea\label{ex:afr:1a}
        \gll Sy het die kar gesien. \\ 
             \textsc{3sg.f.sbj} have.\textsc{prs} \textsc{def} car see.\textsc{pst.ptcp} \\
        \glt ‘She saw the car.’
    
    \ex\label{ex:afr:1b}
        \gll Gister het sy die kar gesien. \\
             yesterday  have.\textsc{prs} \textsc{3sg.f.sbj} \textsc{def} car  see.\textsc{pst.ptcp} \\
        \glt ‘Yesterday, she saw the car.’
    \ex\label{ex:afr:1c}
        \gll … dat sy die kar gesien het. \\
             {} \textsc{comp} \textsc{3sg.f.sbj} \textsc{def} car  see.\textsc{pst.ptcp} have.\textsc{prs}\\
        \glt ‘… that she saw the car.’ 
    \z
\z 


Another essential fact concerns the general lack of inflection. The language does not indicate case or gender on nouns and, in its pronominal system, such distinctions are restricted to the singular (e.g. \textit{hy/hom} ‘he/him’ and \textit{sy/haar} ‘she/her’ versus \textit{hulle} ‘they, them’). There is differential object marking, though, with the preposition \textit{vir} ‘for’, which seems preferred with proper nouns, optional with human or animate nouns and with pronouns and unacceptable with inanimate nouns. Compare \REF{ex:afr:1a} and \REF{ex:afr:2}.

\ea\label{ex:afr:2}
    \gll Sy het vir Henk gesien. \\ 
         3\textsc{sg.f.sbj} have.\textsc{prs} for Henk see.\textsc{pst.ptcp} \\
    \glt ‘She saw Henk.’
\z 


Verb inflection is limited too. They show no agreement and tend to have just a base form and a past participle form (e.g. \textit{gaan} ‘go’, \textit{gegaan} ‘gone’). Only the modal auxiliaries and ‘know’, ‘be’ and ‘have’ (still) have a preterite (e.g. \textit{is} ‘am/are/is’, \textit{was} ‘was/were’). The final two verbs are also the sole ones in the language with a distinctive infinitive form (e.g. \textit{wees} ‘be’, \textit{hê} ‘have’).

Note that all examples in this chapter are based on earlier research, internet queries (for which we used the Google search engine and relied on factors such as context, authors’ names and dates of publication to assess the relevance of the results) and on actual corpus attestations (from the numerous corpora available and searchable on the \citeauthor{VirtueleInstituutvirAfrikaans2019}'s \citeyear{VirtueleInstituutvirAfrikaans2019} corpus portal – with recent data from blogs, books, newspapers and plays, among other things). For reasons of space and presentation, they have, however, typically been shortened and/or slightly altered (though native speakers have always been consulted to check their grammaticality). So most examples are constructed to at least some extent and no reference can/will be provided for them. For examples taken from secondary sources or reproduced verbatim from other sources, we will obviously give precise references. The description of negation in Afrikaans in this chapter is based on the questionnaire provided by the editors (\citetv{chapters/questionnaire}) and the structure of the chapter follows the model given in the questionnaire.

\section{Clausal negation}
\subsection{Standard negation}
\label{sec:afr:2.1}
Negating a declarative verbal main clause is typically achieved through two identical particles, as in \REF{ex:afr:3}. The first \textit{nie}, glossed as \textsc{neg}\textsubscript{1}, is clause-medial and comes after the subject, the finite verb slot and, if present, sentential adverbs like \textit{waarskynlik} ‘probably’ in \REF{ex:afr:3a}. In focus-neutral circumstances, it precedes the object, as in \REF{ex:afr:3b}, unless this object is a pronoun without \textit{vir}, as in \REF{ex:afr:3c}. The second \textit{nie} or \textsc{neg}\textsubscript{2} is clause-final and normally even follows subordinate clauses, like in \REF{ex:afr:3d}, and extraposed constituents after the non-finite verb slot.

\ea\label{ex:afr:3}
    
    \ea\label{ex:afr:3a}
        \gll Die bestuurder-s verkoop waarskynlik nie kaart-jie-s nie. \\ 
             \textsc{def} driver-\textsc{pl} sell probably \textsc{neg}\textsubscript{1} card-\textsc{dim}{}-\textsc{pl}  \textsc{neg}\textsubscript{2} \\
        \glt ‘The drivers probably do not sell tickets.’
    
    \ex\label{ex:afr:3b}
        \gll Jy het nie die kar gesien nie. \\
             \textsc{2sg.sbj} have\textsc{.prs} \textsc{neg}\textsubscript{1} \textsc{def} car  see.\textsc{pst.ptcp} \textsc{neg}\textsubscript{2}\\
        \glt ‘You did not see the car.’
    
    \ex\label{ex:afr:3c}
        \gll Jy het hom nie gesien nie. \\ 
             2\textsc{sg.sbj} have.\textsc{prs} \textsc{3sg.m.nsbj} \textsc{neg}\textsubscript{1} see.\textsc{pst.ptcp}  \textsc{neg}\textsubscript{2} \\
        \glt ‘You did not see him.’
    
    \ex\label{ex:afr:3d}
        \gll Ek het nie gesê (*nie) dat die kar stukkend was nie. \\
             \textsc{1sg.sbj} have.\textsc{prs} \textsc{neg}\textsubscript{1} say.\textsc{pst.ptcp}  \textsc{neg}\textsubscript{2} \textsc{comp}  \textsc{def} car  broken be.\textsc{pst} \textsc{neg}\textsubscript{2}\\
        \glt ‘I did not say that the car was broken.’
    \z
\z 


The two particles differ from each other in several respects (see \citealt[3--4]{BiberauerCyrino2009}). Only \textit{nie}\textsubscript{1} can be modified, for instance, as in \REF{ex:afr:4a}. It is also \textit{nie}\textsubscript{1} through which negation is primarily expressed: \REF{ex:afr:4a} without \textit{nie}\textsubscript{2}, though ill-formed, will still be interpreted as a negative clause. In the same vein, it is \textit{nie}\textsubscript{1} that is replaced when a more emphatic negator is required or a negative indefinite is included. \textit{Nie}\textsubscript{2} stays put in either case, as \REF{ex:afr:4b} shows.\footnote{Because of these differences, Afrikaans linguistics tends to analyze \textit{nie}\textsubscript{1} as a negative adverb, i.e. the true negator, and \textit{nie}\textsubscript{2} as a negative particle (e.g. the Taalportaal grammar). A similar distinction can be found in much generative work on the phenomenon (e.g. \citealt{Oosthuizen1998}).} What the markers have in common is the ability to cliticize onto the preceding word in very colloquial language. Example \REF{ex:afr:4c}, coming from the hip hop group \textit{Die Antwoord}’s lyrics, is a case in point.

\ea\label{ex:afr:4}
    
    \ea\label{ex:afr:4a}
        \gll Sy wil (glad) nie eet (*glad) nie. \\ 
             3\textsc{sg.f.sbj} want\textsc{.prs} altogether \textsc{neg}\textsubscript{1} eat altogether  \textsc{neg}\textsubscript{2} \\
        \glt ‘She does not(/NOT) want to eat at all.’
    
    \ex\label{ex:afr:4b}
        \gll Hy wil niks/geensins eet nie. \\
             3\textsc{sg.m.sbj} want\textsc{.prs} nothing/in.no.way eat \textsc{neg}\textsubscript{2}\\
        \glt ‘He does not want to eat anything/at all.’

    \ex\label{ex:afr:4c}
        \gll Ek weet=ie waar ek iss=ie. \\
             \textsc{1sg.sbj} know=\textsc{neg}\textsubscript{1} where  \textsc{1sg.sbj} be.\textsc{prs=neg}\textsubscript{2}\\
        \glt ‘I do not know where I am.’\footnote{\href{https://l-hit.com/en/433108}{https://l-hit.com/en/433108}, accessed 2021-11-02.}
    
    \z
\z 


As to (a)symmetry à la \citet{Miestamo2005}, this discontinuous two-part negation can be described as fully symmetric. Deleting \textit{nie}\textsubscript{1} and \textit{nie}\textsubscript{2} from the sentences in \REF{ex:afr:3}, for example, produces the positive ones in \REF{ex:afr:5}.

\ea\label{ex:afr:5}
    \ea\label{ex:afr:5a}
        \gll Die bestuurder-s verkoop waarskynlik kaart-jie-s. \\ 
             \textsc{def} driver-\textsc{pl} sell probably card-\textsc{dim-pl} \\
        \glt ‘The drivers probably sell tickets.’
    
    \ex\label{ex:afr:5b}
        \gll Jy het haar gesien. \\
             \textsc{2sg.sbj} have.\textsc{prs} \textsc{3sg.f.nsjb} see.\textsc{pst.ptcp}\\
        \glt ‘You saw her.’

    \ex\label{ex:afr:5c}
        \gll Jy het die kar gesien. \\ 
             2\textsc{sg.sbj} have.\textsc{prs} \textsc{def} car see\textsc{.pst.ptcp} \\
        \glt ‘You saw the car.’
    
    \ex\label{ex:afr:5d}
        \gll Ek het gesê dat die kar stukkend was. \\
             \textsc{1sg.sbj} have.\textsc{prs} say\textsc{.pst.ptcp} \textsc{comp} \textsc{def} car broken be.\textsc{pst}\\
        \glt ‘I said that the car was broken.’
    \z
\z 

Afrikaans standard negation thus exhibits constructional symmetry. There also exist no differences between positive and negative declarative verbal main clauses in, inter alia, finiteness or tense-aspect-mood marking, making \textit{nie}\textsubscript{1} … \textit{nie}\textsubscript{2} paradigmatically symmetric as well.

The above account of the two particles’ distribution needs to be qualified in a number of ways. First, \textit{nie}\textsubscript{2} is ungrammatical when all slots after \textit{nie}\textsubscript{1} are empty and \textit{nie}\textsubscript{1} is in final position (see \citealt[31]{Biberauer2007}  on a syntactic haplology mechanism in Afrikaans). This situation arises in clauses consisting only of a subject and a finite verb, as in \REF{ex:afr:6a} (which would be interpreted as ‘they don’t not sing’ with \textit{nie nie}). It also occurs in clauses with an additional pronominal object preceding \textit{nie}\textsubscript{1}, as in \REF{ex:afr:6b}, where the modifier \textit{glad} ‘altogether’ indicates that the single negator is indeed \textit{nie}\textsubscript{1} rather than \textit{nie}\textsubscript{2}. This aversion to consecutive cases of \textit{nie} extends to clauses that might have been expected to end in \textit{nie}\textsubscript{2} \textit{nie}\textsubscript{2}. In \REF{ex:afr:6c}, for example, one could have envisaged the occurrence of a first \textit{nie}\textsubscript{2} for the main clause negation and a second \textit{nie}\textsubscript{2} for the subordinate clause negation but only one \textit{nie}\textsubscript{2} appears. If \textit{nie}\textsubscript{1} in structures like \REF{ex:afr:6a} and \REF{ex:afr:6b} is substituted by a negative indefinite, as in \REF{ex:afr:6d}, \textit{nie}\textsubscript{2} is optional (just like when negative indefinites are used as short replies to questions).

\ea\label{ex:afr:6}
    \ea\label{ex:afr:6a}
        \gll Hulle sing nie (*nie). \\ 
             \textsc{3pl} sing \textsc{neg}\textsubscript{1} \textsc{neg}\textsubscript{2} \\
        \glt ‘They do not sing.’
    
    \ex\label{ex:afr:6b}
        \gll Die kind-ers verstaan my glad nie (*nie). \\
             \textsc{def} child-\textsc{pl} understand \textsc{1sg.nsbj} altogether \textsc{neg}\textsubscript{1} \textsc{neg}\textsubscript{2}\\
        \glt ‘The children do not understand me at all.’
    
    \ex\label{ex:afr:6c}
        \gll Hulle het nie gesê dat jy nie werk nie (*nie). \\ 
              3\textsc{pl} have.\textsc{prs} \textsc{neg} \textsubscript{1}say.\textsc{pst.ptcp} \textsc{comp} \textsc{2sg.sbj} \textsc{neg}\textsubscript{1} work \textsc{neg}\textsubscript{2} \textsc{neg}\textsubscript{2} \\
        \glt ‘They did not say that you do not work.’
    
    \ex\label{ex:afr:6d}
        \gll Hy eet niks (nie). \\
             3\textsc{sg.m.sbj} eat nothing \textsc{neg}\textsubscript{2}\\
        \glt ‘He does not eat anything.’
    \z
\z 


Second, \textit{nie}\textsubscript{2} can be omitted in certain genres. These include poetry, signs and newspaper headlines (see \citealt{Deumert2017}: 105). The headline from the national news publication \textit{Beeld} in \REF{ex:afr:7} can serve as an example.

\ea\label{ex:afr:7}
    \gll AfriForum mag nie appelleer oor ou vlag. \\ 
         AfriForum may.\textsc{prs} \textsc{neg}\textsubscript{1} appeal about old flag \\
    \glt ‘AfriForum may not appeal about old flag.’\footnote{https://www.netwerk24.com/netwerk24/nuus/hof/afriforum-mag-nie-appelleer-oor-ou-vlag-20190920, accessed 2021-11-02} 
\z 

Third, as \REF{ex:afr:8a} shows, \textit{nie}\textsubscript{1} can be left out too, with a very specific effect. According to \citet[5]{Biberauer2007}, the difference from the two-particle equivalent \REF{ex:afr:8b} is that, in the former, the object “is necessarily interpreted as old information” while, in the latter, it is “part of the information focus, either independently as new/focused information or as part of an all-rhematic structure”.

\ea\label{ex:afr:8}
    \ea\label{ex:afr:8a}
        \gll Ek ken hier-die boek nie. \\ 
             \textsc{1sg.sbj} know  here-\textsc{def} book \textsc{neg}\textsubscript{2} \\
        \glt ‘I do not know this book.’
    
    \ex\label{ex:afr:8b}
        \gll Ek ken  nie  hier-die    boek  nie. \\
             \textsc{1sg.sbj} know \textsc{neg}\textsubscript{1} here-\textsc{def} book  \textsc{neg}\textsubscript{2} \\
        \glt ‘I do not know this book.’
    
    \z
\z 

Fourth, and finally, in informal language, \textit{nie}\textsubscript{2} sometimes appears ahead of extraposed constituents, especially when they are quite heavy (probably for reasons of processing; see \citealt[117]{Molnárfi2001}). In such cases, there would not normally be a second \textit{nie}\textsubscript{2} at the end. Compare \REF{ex:afr:9} to \REF{ex:afr:3d}.

\ea\label{ex:afr:9}
    \gll Sy het nooit gesê nie dat mens net so gelukkig is soos jou ongelukkig-ste kind. \\ 
    \textsc{3sg.f.sbj} have.\textsc{prs} never say.\textsc{pst.ptcp} \textsc{neg}\textsubscript{2} \textsc{comp} human just so happy be.\textsc{prs} as \textsc{2sg.nsbj} unhappy-\textsc{supl} child\\
    \glt ‘She never said that one is only as happy as one’s unhappiest child.’
\z 


\subsection{Negation in non-declaratives}
\label{sec:afr:2.2}
The negation of interrogatives, which are formed by inverting the subject and the finite verb, is governed by the same principles as standard negation. Consider the respective counterparts of \REF{ex:afr:3b}, \REF{ex:afr:3d}, \REF{ex:afr:4b} and \REF{ex:afr:6a} in \REF{ex:afr:10} and notice, inter alia, the two-part negation in \REF{ex:afr:10a}, the truly clause-final position of \textit{nie}\textsubscript{2} in \REF{ex:afr:10b}, the replacement of \textit{nie}\textsubscript{1} by a negative indefinite in \REF{ex:afr:10c} and the ungrammaticality of \textit{nie nie} in \REF{ex:afr:10d}.

\ea\label{ex:afr:10}
    \ea\label{ex:afr:10a}
        \gll Het jy hom nie gesien nie? \\ 
             have.\textsc{prs} \textsc{2sg.sbj} \textsc{3sg.m.nsbj} \textsc{neg}\textsubscript{1} see.\textsc{pst.ptcp} \textsc{neg}\textsubscript{2} \\
        \glt ‘Did you not see him?’
    
    \ex\label{ex:afr:10b}
        \gll Het jy nie gesê dat die kar stukkend was nie? \\
             have.\textsc{prs} \textsc{2sg.sbj} \textsc{neg}\textsubscript{1} say\textsc{.pst.ptcp} \textsc{comp} \textsc{def} car  broken be.\textsc{pst} \textsc{neg}\textsubscript{2} \\
        \glt ‘Did you not say that the car was broken?’
 
    \ex\label{ex:afr:10c}
        \gll Hoekom wil hy niks eet nie? \\ 
             why want.\textsc{prs} \textsc{3sg.m.sbj} nothing eat \textsc{neg}\textsubscript{2} \\
        \glt ‘Why does he not want to eat anything?’
    
    \ex\label{ex:afr:10d}
        \gll Sing hulle nie (*nie)? \\
             sing  3\textsc{pl}  \textsc{neg}\textsubscript{1} \textsc{neg}\textsubscript{2}\\
        \glt ‘Do they not sing?’
    \z
\z 


It should therefore not come as a surprise that interrogative negation is also entirely symmetric. Deleting the two negators in \REF{ex:afr:10a}, for instance, results in a sentence with the meaning ‘did you see him?’. As to the type of answer expected, the negative polar questions in \REF{ex:afr:10a}, \REF{ex:afr:10b} and \REF{ex:afr:10d} may function in a neutral manner but they are often biased toward a positive answer (see \citealt[40]{Holmberg2016}).

Imperative negation is more complex. As \REF{ex:afr:11a} and \REF{ex:afr:11b} make clear, the imperative cannot be negated in the standard way. The construction in \REF{ex:afr:11c} is employed instead. It features \textit{moenie} ‘don’t!’ rather than \textit{nie}\textsubscript{1} and has the main verb in the non-finite rather than the finite slot (thus belonging to  \citeauthor{AuweraEtAl2013}'s \citeyear{AuweraEtAl2013}  [\textemdash imperative verb] [\textemdash standard negation] type). This loss of finiteness manifests itself here in the behavior of the separable verb particle \textit{weg} ‘away’. In the imperative in \REF{ex:afr:11a}, the stem \textit{gaan} ‘go’ is in the finite slot and, hence, \textit{weg} occupies the non-finite slot, just like in declarative main clauses such as \REF{ex:afr:11d}. In the negative imperative in \REF{ex:afr:11c}, by contrast, \textit{weg} is attached to \textit{gaan}, which the declarative main clause in \REF{ex:afr:11e} indicates is characteristic of verbs in the non-finite slot.

\ea\label{ex:afr:11}
    
    \ea\label{ex:afr:11a}
        \gll Gaan weg! \\ 
             go away \\
        \glt ‘Go away!’
    
    \ex\label{ex:afr:11b}
        \gll *Gaan nie weg nie! \\
             go \textsc{neg}\textsubscript{1} away  \textsc{neg}\textsubscript{2} \\
        \glt ‘Do not go away!’
   
    \ex\label{ex:afr:11c}
        \gll Moenie weg-gaan nie! \\ 
             must.not.\textsc{prs} away-go    \textsc{neg}\textsubscript{2} \\
        \glt ‘Do not go away!’
    
    \ex\label{ex:afr:11d}
        \gll Ons gaan ’n bietjie weg. \\
             1\textsc{pl} go \textsc{indf} little.bit away \\
        \glt ‘We are going away for a little bit.’
    
    \ex\label{ex:afr:11e}
        \gll Ons wil nie weg-gaan nie. \\
             \textsc{1pl}  want.\textsc{prs} \textsc{neg}\textsubscript{1} away-go    \textsc{neg}\textsubscript{2}\\
        \glt ‘We do not want to go away.’ 
    \z
\z 


The constructional asymmetry in finiteness (see \citealt[73--96]{Miestamo2005}) between \REF{ex:afr:11a} and \REF{ex:afr:11c} can be attributed to the auxiliary origin of \textit{moenie}. This marker is a combination of \textit{moet} ‘must’ and \textit{nie}\textsubscript{1}. In fact, the two components can be separated, for example, by a pronoun without \textit{vir} taking up its ordinary place between the finite slot and \textit{nie}\textsubscript{1}, as in \REF{ex:afr:12a}. \textit{Moet} can occur on its own as well, when a negative indefinite or an emphatic negator substitutes for \textit{nie}\textsubscript{1}, as in \REF{ex:afr:12b}. Still, object pronouns without \textit{vir}, which are ordinarily found before the first negator in other clause types (cf. \REF{ex:afr:3b}), actually tend to follow rather than precede it in the negative imperative, facilitating the fusion of \textit{moet} and \textit{nie} in \REF{ex:afr:12c}. This fact is suggestive of some degree of univerbation and grammaticalization (for more details, see \citealt[500--501]{Olmen2010}).\footnote{The fused form is not actually limited to negative imperatives. It can also be used for non-performative expressions of negative obligation such as \textit{Piet moenie worrie nie} ‘Pete should not worry’.\label{footnote:afr:3}}

\ea\label{ex:afr:12}
    \ea\label{ex:afr:12a}
        \gll Moet jou nie ontstel nie! \\ 
             must.\textsc{prs} \textsc{2sg.nsubj} \textsc{neg}\textsubscript{1} trouble  \textsc{neg}\textsubscript{2}  \\
        \glt ‘Do not trouble yourself!’
    
    \ex\label{ex:afr:12b}
        \gll Moet niks eet nie! \\
             must.\textsc{prs} nothing eat    \textsc{neg}\textsubscript{2} \\
        \glt ‘Do not eat anything!’
        
    \ex\label{ex:afr:12c}
        \gll Moenie hom dood-maak nie! \\
             must.not.\textsc{prs} \textsc{3sg.m.nsbj} dead-make  \textsc{neg}\textsubscript{2}\\
        \glt ‘Do not kill him!’
    \z
\z 

\textit{Moenie} can also have a more specific negative imperative sense: when it directly introduces a complementizer clause as in \REF{ex:afr:13a}, it expresses the meaning ‘don’t let’. A function that it does not fulfill is the negation of non-addressee-oriented or “non-canonical” \citep[47]{Aikhenvald2010} imperatives. They require the auxiliary \textit{laat} ‘let’ and adhere to the standard way of negating, as \REF{ex:afr:13b} illustrates.

\ea\label{ex:afr:13}
    \ea\label{ex:afr:13a}
        \gll Moenie dat hulle jou bang-maak nie! \\ 
             must.not.\textsc{prs} \textsc{comp} \textsc{3pl} \textsc{2sg.nsbj} scared-make \textsc{neg}\textsubscript{2} \\
        \glt ‘Do not let them scare you!’
    
    \ex\label{ex:afr:13b}
        \gll Laat ons (nie) vergeet (nie)! \\
             let \textsc{1pl} \textsc{neg}\textsubscript{1} forget \textsc{nie}\textsubscript{2}\\
        \glt ‘Let’s (not) forget!’
    \z
\z 

\subsection{Negation in stative predications}
\label{sec:afr:2.3}
Some types of stative predication (see \citealt[1386--1389]{Veselinova2014}) can only be negated in the standard, symmetric manner. These are equation, attribution and locative predication, as in \REF{ex:afr:14a} to \REF{ex:afr:14c} respectively.

\ea\label{ex:afr:14}
    \ea\label{ex:afr:14a}
        \gll Hy is (nie) my pa (nie). \\ 
             \textsc{3sg.m.sbj}  be.\textsc{prs} \textsc{neg}\textsubscript{1} \textsc{1sg.nsbj} dad  \textsc{neg}\textsubscript{2} \\
        \glt ‘He is (not) my dad.’
    
    \ex\label{ex:afr:14b}
        \gll Hulle is (nie) braaf (nie). \\
             \textsc{3pl} be.\textsc{prs} \textsc{neg}\textsubscript{1} brave    \textsc{neg}\textsubscript{2}\\
        \glt ‘They are (not) brave.’

    \ex\label{ex:afr:14c}
        \gll Die kar is (nie) in die motorhuis (nie). \\
             \textsc{def} car  be.\textsc{prs} \textsc{neg}\textsubscript{1} in  \textsc{def} garage \textsc{neg}\textsubscript{2}\\
        \glt ‘The car is (not) in the garage.’
    \z
\z 


The negation of the other types may be similar, as possessive predication in \REF{ex:afr:15a} shows, but it is also possible to use the negative determiner \textit{geen} ‘no’ (or its informal variant \textit{g’n}) instead of \textit{nie}\textsubscript{1}. If it precedes an indefinite singular noun phrase, as in \REF{ex:afr:15b}, the article \textit{’n} ‘a(n)’ from the corresponding positive clause is lost. \textit{Nie}\textsubscript{2} is optional here.

\ea\label{ex:afr:15}
    
    \ea\label{ex:afr:15a}
        \gll Ek het (nie) ’n foon (nie). \\ 
             \textsc{1sg.sbj}  have.\textsc{prs} \textsc{neg}\textsubscript{1} \textsc{indf} phone  \textsc{neg}\textsubscript{2} \\
        \glt ‘I (do not) have a phone.’
    
    \ex\label{ex:afr:15b}
        \gll Ek het geen foon (nie). \\
             \textsc{1sg.sbj} have.\textsc{prs} no phone  \textsc{neg}\textsubscript{2} \\
        \glt ‘I have no phone.’
    \z
\z 
 

This variation is found with proper inclusion too, as in \REF{ex:afr:16}, and with locative-existential and purely existential predication, as in \REF{ex:afr:17} and \REF{ex:afr:18} respectively, both of which employ the construction \textit{daar is} ‘there is/are’. Existential predication can also be expressed by \textit{bestaan} ‘exist’, however, and this verb is negated in the standard way, as in \REF{ex:afr:18c}.

\ea\label{ex:afr:16}
    \ea\label{ex:afr:16a}
        \gll Henk is (nie) ’n onderwyser (nie). \\ 
             Henk be.\textsc{prs} \textsc{neg}\textsubscript{1} \textsc{indf} teacher      \textsc{neg}\textsubscript{2} \\
        \glt ‘Henk is (not) a teacher.’
    
    \ex\label{ex:afr:16b}
        \gll Henk is geen onderwyser nie. \\
             Henk be.\textsc{prs} no teacher \textsc{neg}\textsubscript{2} \\
        \glt  ‘Henk is no teacher.’
    \z
\z 

\ea\label{ex:afr:17}
    \ea\label{ex:afr:17a}
        \gll Daar is (nie) tran-e in die hemel (nie). \\ 
             there  be.\textsc{prs}  \textsc{neg}\textsubscript{1} tear-\textsc{pl} in  \textsc{def} heaven \textsc{neg}\textsubscript{2} \\
        \glt ‘There are (no) tears in heaven.’
    
    \ex\label{ex:afr:17b}
        \gll Daar is geen tran-e in die hemel nie. \\
             there be.\textsc{prs} no tear-\textsc{pl} in  \textsc{def} heaven  \textsc{neg}\textsubscript{2} \\
        \glt ‘There are no tears in heaven.’
    \z
\z 

\ea\label{ex:afr:18}
    \ea\label{ex:afr:18a}
        \gll Daar is (nie) spok-e (nie). \\ 
             there  be.\textsc{prs}  \textsc{neg}\textsubscript{1} ghost-\textsc{pl} \textsc{neg}\textsubscript{2} \\
        \glt ‘There are (no) ghosts.’
    
    \ex\label{ex:afr:18b}
        \gll Daar is geen spok-e nie. \\
             there be.\textsc{prs} no ghost-\textsc{pl} \textsc{neg}\textsubscript{2}\\
        \glt ‘There are no ghosts.’

    \ex\label{ex:afr:18c}
        \gll Spok-e bestaan (nie). \\
             ghost-\textsc{pl} exist    \textsc{neg}\textsubscript{1} \\
        \glt  ‘Ghosts (do not) exist.’   
    \z
\z 

According to \citet[413]{Donaldson1993}, the two variants are interchangeable but it is likely that, as in other Germanic languages, which exhibit considerable variation in this area, the choice of one over the other is determined by a variety of factors. They include but are not limited to the type of stative predication, the complexity of the verb phrase, the scope of negation, the old versus new information status of the proposition and the conventionalized character of phrases (e.g. \citealt{Peters2008, Bordal2017, BurnettEtAl2018, Wallage2020}). An in-depth study of their potential impact falls outside the scope of the present chapter, though. Let it suffice to say that the variant with \textit{nie}\textsubscript{1} seems to occur much more often than that with \textit{geen} in Present-day Afrikaans.\footnote{This claim is based on Google searches and an exploratory examination of the Taalkommissiekorpus (\citealt{VirtueleInstituutvirAfrikaans2019}), a large reference corpus of present-day written language.} This fact is interesting from the perspective of Dutch, where only \textit{geen} is grammatical in possessive predication, proper inclusion and (locative-)existential predication.\footnote{As one reviewer rightly points out, Dutch \textit{niet} ‘not’ may still appear with these predication types in contrastive contexts. Taking \REF{ex:afr:16b} as a starting point, if one wanted to communicate ‘Henk is not a teacher but a professor’, one could say \textit{Henk is niet onderwijzer maar professor.} See also \sectref{sec:afr:4.1}.} The suggestion therefore is that Afrikaans has extended the standard negation pattern to all types of stative predication. Note, finally, that \REF{ex:afr:14} to \REF{ex:afr:18} also demonstrate that no grammatical differences exist between generic and specific predicates or between temporary and permanent property assignment.

\subsection{Negation in non-main clauses}
\label{sec:afr:2.4}
Negating subordinate clauses happens in much the same way as standard negation and the negation of stative predication. There are, in other words, no special negators or negative non-finite forms. The subordinate finite equivalents of the standard negation in \REF{ex:afr:3a} and the negated stative predicates in \REF{ex:afr:18a} and \REF{ex:afr:18b} are given in \REF{ex:afr:19}. They show that negation is, essentially, completely symmetric here too.

\ea\label{ex:afr:19}
    \ea\label{ex:afr:19a}
        \gll Jy beweer dat hulle (nie) kaart-jie-s verkoop (nie). \\ 
             \textsc{2sg.sbj} claim \textsc{comp} \textsc{3pl} \textsc{neg}\textsubscript{1}  card-\textsc{dim-pl} sell    \textsc{neg}\textsubscript{2} \\
        \glt ‘You claim that they (do not) sell tickets.’
    
    \ex\label{ex:afr:19b}
        \gll Hulle sê dat daar (nie) spok-e is (nie). \\
             3\textsc{pl} say \textsc{comp} there  \textsc{neg}\textsubscript{1} ghost-\textsc{pl} be.\textsc{prs} \textsc{neg}\textsubscript{2}\\
        \glt ‘They say that there are (no) ghosts.’

    \ex\label{ex:afr:19c}
        \gll Hulle sê dat daar (geen) spok-e is (nie). \\
             3\textsc{pl} say \textsc{comp} there no ghost-\textsc{pl} be.\textsc{prs} \textsc{neg}\textsubscript{2}\\
        \glt ‘They say that there are (no) ghosts.’
    \z
\z 


The word order in finite subordinate clauses is normally SOV, with \textit{nie}\textsubscript{1} (or some negative indefinite or emphatic negator) appearing before the verb and \textit{nie}\textsubscript{2} after the verb. This syntax has as a result that consecutive occurrences of \textit{nie} are impossible. Main clauses with only \textit{nie}\textsubscript{1} thus acquire a second negator when they are turned into subordinate clauses. The comparison of \REF{ex:afr:6b} to \REF{ex:afr:20a} can serve as an example. The complementizer \textit{dat} ‘that’ is often omitted, though (see \citealt{RooyKruger2016}), in which case the subordinate clause has V2 word order and \textit{nie}\textsubscript{2} is again blocked, as in \REF{ex:afr:20b}. 

\ea\label{ex:afr:20}
    
    \ea\label{ex:afr:20a}
        \gll Ek dink dat die kind-ers my glad nie verstaan nie.\\
             1\textsc{sg.sbj} think  \textsc{comp} \textsc{def} child-\textsc{pl} \textsc{1sg.nsbj} altogether \textsc{neg}\textsubscript{1} understand  \textsc{neg}\textsubscript{2}\\
        \glt ‘I think that the children do not understand me at all.’
    
    \ex\label{ex:afr:20b}
        \gll Ek dink die kind-ers verstaan my glad nie (*nie). \\
             1\textsc{sg.sbj} think  \textsc{def} child-\textsc{pl} understand  \textsc{1sg.nsbj} altogether \textsc{neg}\textsubscript{1} \textsc{neg}\textsubscript{2}\\
        \glt ‘I think that the children do not understand me at all.’
    \z
\z 


The general requirement for two negators extends to cases where, in the main clause, only \textit{nie}\textsubscript{2} is found because the object is old information, as in \REF{ex:afr:8a}, repeated here as \REF{ex:afr:21a}. Its subordinate counterpart in \REF{ex:afr:21b} has \textit{nie}\textsubscript{1} coming after the object. When the object is part of the information focus, as in \REF{ex:afr:21c}, \textit{nie}\textsubscript{1} precedes it (see \citealt[106--107]{Biberauer2008}).

\ea\label{ex:afr:21}
    \ea\label{ex:afr:21a}
        \gll Ek ken hier-die boek nie. \\ 
             \textsc{1sg.sbj} know  here-\textsc{def} book  \textsc{neg}\textsubscript{2} \\
        \glt ‘I do not know this book.’
    
    \ex\label{ex:afr:21b}
        \gll …  dat ek hier-die boek nie ken nie. \\
             {} \textsc{comp} \textsc{1sg.sbj} here-\textsc{def} book  \textsc{neg}\textsubscript{1} know  \textsc{neg}\textsubscript{2}\\
        \glt ‘… that I do not know this book.’
    \ex\label{ex:afr:21c}
        \gll … dat ek nie  hier-die boek ken nie. \\
             {} \textsc{comp} \textsc{1sg.sbj} \textsc{neg}\textsubscript{1} here-\textsc{def} book know \textsc{neg}\textsubscript{2}\\
        \glt ‘… that I do not know this book.’ 
    \z
\z 

The negation of non-finite subordinate clauses, like that of finite ones, tends to behave in fairly straightforward ways, as \REF{ex:afr:22} makes clear. Of note in \REF{ex:afr:22a} is that \textit{nie}\textsubscript{2} is placed not at the end of the entire sentence but directly after the preposed purpose clause. In \REF{ex:afr:22b}, it is the position of \textit{nie}\textsubscript{1} after \textit{om} ‘for’ that makes the subordinate clause negative. If it precedes \textit{om}, it negates the main verb, as in \REF{ex:afr:22c}.

\ea\label{ex:afr:22}
    \ea\label{ex:afr:22a}
        \gll Ten einde sak-e (nie) te kompliseer (nie), volg ons die advies. \\ 
             in order thing-\textsc{pl} \textsc{neg}\textsubscript{1} to  complicate  \textsc{neg}\textsubscript{2} follow \textsc{1pl} \textsc{def} advice \\
        \glt ‘In order (not) to complicate things, we are following the advice.’
    
    \ex\label{ex:afr:22b}
        \gll Sy vra julle om (nie) te sing (nie). \\
             \textsc{3sg.f.sbj} ask \textsc{2pl} for    \textsc{neg}\textsubscript{1} to  sing  \textsc{neg}\textsubscript{2} \\
        \glt ‘She is asking you (not) to sing.’
        
    \ex\label{ex:afr:22c}
        \gll Sy vra julle nie om te sing nie. \\
             \textsc{3sg.f.sbj} ask  \textsc{2pl} \textsc{neg}\textsubscript{1} for to sing  \textsc{neg}\textsubscript{2}\\
        \glt ‘She is not asking you to sing.’
    \z
\z

For non-finite complement clauses with bare infinitives, it seems necessary to make a distinction between verbs of perception like \textit{sien} ‘see’ and \textit{hoor} ‘hear’ and verbs like \textit{help} ‘help’ and \textit{leer} ‘teach’. The former can be negated in just one structural way, as in \REF{ex:afr:23a}, and it is normally understood as applying to the main verb. To get the (admittedly awkward) interpretation where only the subordinate clause is negated, the negation has to be stressed, as in \REF{ex:afr:23b}.\footnote{The generative literature (e.g. \citealt[4]{BiberauerCyrino2009}) indicates that only \textit{nie}\textsubscript{1} is stressable. The native speakers consulted for the present chapter disagree and maintain that, in particular contexts, like \REF{ex:afr:23b}, \textit{nie}\textsubscript{2} is able/needs to be emphasized as well.} With the latter verbs, as in \REF{ex:afr:23c}, the simple addition of \textit{nie}\textsubscript{1} and \textit{nie}\textsubscript{2} produces the same meaning as in \REF{ex:afr:23a}, namely the negation of the main verb.\footnote{In line with the description in \sectref{sec:afr:2.1}, \textit{nie}\textsubscript{1} follows the pronoun (without \textit{vir}) in \REF{ex:afr:23a} and \REF{ex:afr:23c}. \textit{Nie}\textsubscript{1} before \textit{haar} and \textit{jou} would be a marked order. In keeping with the earlier discussion too, \textit{nie}\textsubscript{1} would precede a lexical replacement of these pronouns in \REF{ex:afr:23a} and \REF{ex:afr:23c} and the reversed order would be marked.} To solely negate the complement clause, the bare infinitive needs to be turned into an \textit{om te} ‘for to’ infinitive, as in \REF{ex:afr:23d}.

\ea\label{ex:afr:23}
    \ea\label{ex:afr:23a}
        \gll Ek sien haar (nie) kom (nie). \\ 
             \textsc{1sg.sbj} see  \textsc{3sg.f.nsbj} \textsc{neg}\textsubscript{1} come  \textsc{neg}\textsubscript{2} \\
        \glt ‘I (do not) see her come.’
    
    \ex\label{ex:afr:23b}
        \gll Ek sien haar NIE kom NIE. \\
             \textsc{1sg.sbj} see  \textsc{3sg.f.nsbj} \textsc{neg}\textsubscript{1} come  \textsc{neg}\textsubscript{2} \\
        \glt ‘I see her not come.’

    \ex\label{ex:afr:23c}
        \gll Dit leer jou nie kop-hou nie. \\ 
             \textsc{3sg.n} teach  2\textsc{sg.nsbj} \textsc{neg}\textsubscript{1} head-hold  \textsc{neg}\textsubscript{2} \\
        \glt ‘This does not teach you to stay calm.’
    
    \ex\label{ex:afr:23d}
        \gll Dit leer jou om nie kwaad te word nie. \\
             \textsc{3sg.n} teach 2\textsc{sg.nsbj} for    \textsc{neg}\textsubscript{1} angry to become \textsc{neg}\textsubscript{2}\\
        \glt ‘This teaches you not to get angry.’
    \z
\z 

See also \sectref{sec:afr:3.3} about (non-)finite subordinate clauses introduced by \textit{sonder} ‘without’ and \sectref{sec:afr:4.1} about the scope of negation in complex clauses. 

\subsection{Negative lexicalizations}

There are no negative lexicalizations in Afrikaans, with the possible exception of \textit{moenie} ‘must not’, which is discussed in \sectref{sec:afr:2.2} (see footnote \ref{footnote:afr:3} too).
A potential test for the presence of inherently negative verbs such as ‘not exist’, ‘not want’ and ‘not know’ (see \citealt{Veselinova2013lexicalization}) would be the compulsory use of \textit{nie}\textsubscript{2} in clauses in which they occur. To our knowledge, no verb exhibits this behavior.\footnote{\label{footnote:afr:8}\citet[79]{Oosthuizen1998} and \citet[116]{Biberauer2008} do mention the ability of verbs with a contradictory meaning such as \textit{weier} ‘refuse’, as well as of intrinsically negative adverbs such as \textit{nouliks} ‘barely’, to be followed by \textit{nie}\textsubscript{2} in colloquial language. Examples like \REF{ex:afr:i} and \REF{ex:afr:ii} are, however, almost consistently judged to be unacceptable by native speakers, as \citegen[58--60]{Potgieter2018} experimental data shows. 

\ea\label{ex:afr:i}
    \gll Ek weier om saam te kom nie.  \\
    \textsc{1sg.sbj} refuse  for  together  to  come \textsc{neg}\textsubscript{2} \\
    \glt ‘I refuse to come along.’ \citep[116]{Biberauer2008} \\
\ex\label{ex:afr:ii}
    \gll Ek kan my nouliks in-hou nie. \\
    \textsc{sg.sbj} can.\textsc{prs} \textsc{1sg.nsbj} barely in-hold \textsc{neg}\textsubscript{2} \\
    \glt ‘I can barely contain myself.’  \citep[116]{Biberauer2008}
\z
} %this closes the footnote

\section{Non-clausal negation}

\subsection{Negative replies}

To disagree with the polarity of a positive polar question like \REF{ex:afr:24}, Afrikaans uses the particle \textit{nee} ‘no’. To agree with its polarity, the particle \textit{ja} ‘yes’ is employed. The replies to a negative polar question like \REF{ex:afr:25} are more complex. \textit{Nee} can serve to express agreement with the interrogative’s negative polarity on its own. By contrast, \textit{ja} in isolation will, as \citegen[143]{Holmberg2016} informant for Afrikaans points out, “almost certainly result in confusion if it isn’t accompanied by some form of disambiguating intonation or other paratactic device which the hearer can draw on to expand ‘yes’ in the required way\textit{”.} \textit{Ja} as a reply to the negative polar question in \REF{ex:afr:25} could, in other words, be interpreted as either ‘the dog is indeed not barking’ or ‘but the dog \textit{is} barking’.\footnote{To express the latter meaning, Dutch has a dedicated particle \textit{jawel} ‘but yes’. There exists no equivalent of this particle in Afrikaans.} Adding \textit{hy blaf (nie)} ‘he is (not) barking’ after \textit{ja} can help to make the reply clearer, though.\footnote{Following a reviewer’s suggestion, we also want to mention \textit{(h)aikona} ‘no way!’ here. It is a relatively common borrowing from the Nguni languages in colloquial Afrikaans (and South African English) and serves, among other things, to convey shock or disbelief and to express emphatic disagreement.}

\ea\label{ex:afr:24}
    \begin{xlist}[A2:]
    \exi{Q: }\label{ex:afr:24a}
        \gll Blaf die hond?  \\ 
             bark \textsc{def} dog \\ 
        \glt ‘Is the dog barking?’    
    \exi{A1: }\label{ex:afr:24b}
        \gll Nee. \\ 
             no  \\
        \glt ‘No(, the dog isn’t barking).’ 
    \exi{A2: }\label{ex:afr:24c}
        \gll  Ja. \\ 
              yes \\
        \glt  ‘Yes(, the dog is barking).’
    \end{xlist}
\z 


\ea\label{ex:afr:25}
    \begin{xlist}[A2:]
    \exi{Q: }\label{ex:afr:25a}
        \gll Blaf die hond nie?  \\ 
             bark \textsc{def} dog \textsc{neg}\textsubscript{1} \\ 
        \glt ‘Isn't the dog barking?’    
    \exi{A1: }\label{ex:afr:25b}
        \gll Nee. \\ 
             no  \\
        \glt ‘No(, the dog isn’t barking).’ 
    \exi{A2: }\label{ex:afr:25c}
        \gll  ?Ja. \\ 
              yes \\
        \glt  ‘Yes(, the dog is(n’t) barking).’
    \end{xlist}
\z 

Interestingly, \textit{nee} does occasionally occur in replies expressing agreement with the polarity of positive polar interrogatives, as in \REF{ex:afr:26a}. In such cases, it fulfills not its regular function but the intersubjective one of denying the concerns that the addressee is presumed to have. It can even be used in this way in replies to content questions, as \REF{ex:afr:26b} shows.

\ea\label{ex:afr:26a}
    \begin{xlist}
    \exi{Q: }\label{ex:afr:26a.A}
        \gll Is jy okay? \\ 
             be.\textsc{prs} \textsc{2sg.sbj} okay\\
        \glt ‘Are you okay?’ 
    \exi{A: }\label{ex:afr:26a.B}
        \gll Nee, ek is reg. \\ 
             no 1\textsc{sg.sbj} be.\textsc{prs} alright \\
        \glt ‘No, (do not worry,) I am alright.’  (\citealt{BiberauerEtAl2017}: 79) 
    \end{xlist}
\z 
\ea\label{ex:afr:26b}
    \begin{xlist}
    \exi{Q: }\label{ex:afr:26b.A}
        \gll Hoe gaan dit? \\
             how go \textsc{3sg.n} \\
        \glt  ‘How are you doing?’ 
    \exi{A: }\label{ex:afr:26b.B}
        \gll Nee, goed. \\
             no good\\
        \glt ‘No, (do not worry,) well.’ (\citealt{BiberauerEtAl2017}: 79) 
    \end{xlist}
\z 
% \ea\label{ex:afr:26}%HC I have made 2x2 examples out if this (just above), but I am not sure if it is ok because it is cited from a published work, so I keep this outcommented until I know the authors agree with the AB version above
%     \ea\label{ex:afr:26a}
%         \gll Is jy okay? -- Nee, ek is reg. \\ 
%              be.\textsc{prs} \textsc{2sg.sbj} okay -- no 1\textsc{sg.sbj} be.\textsc{prs} alright \\
%         \glt ‘Are you okay?’ ‘No, do not worry, I am alright.’  (\citealt{BiberauerEtAl2017}: 79)
    
%     \ex\label{ex:afr:26b}
%         \gll Hoe gaan dit? -- Nee, goed. \\
%              how go \textsc{3sg.n} -- no good\\
%         \glt  ‘How are you doing?’ ‘No, do not worry, well.’ (\citealt{BiberauerEtAl2017}: 79) 
%     \z
% \z 

Another peculiarity of \textit{nee} is its frequent combination with \textit{ja}, as in \REF{ex:afr:27}. \textit{Ja-nee} has essentially the same meaning as \textit{ja} in \REF{ex:afr:24} but the sense of agreement is always qualified in some manner (e.g. reluctance, enthusiasm, resignation; see \citealt[76--77]{BiberauerEtAl2017}). It seems to be used more for confirming statements than for replying to questions, though.

\ea\label{ex:afr:27}
    \begin{xlist}
    \exi{Q: }\label{ex:afr:27A}
        \gll Dink jy hier-die boek is belangrik? \\
         think 2\textsc{sg.sbj} here-\textsc{def} book  be.\textsc{prs} important \\
    \glt ‘Do you think that this book is important?’ 
    \exi{A: }\label{ex:afr:27B}
        \gll Ja-nee, ek dink so. \\
         yes-no 1\textsc{sg.sbj} think so\\
    \glt ‘Yeah, well, I think so.’
    \end{xlist}
\z 

\subsection{Negative indefinites}
\label{sec:afr:3.2}

The list of negative indefinites in the language includes \textit{niemand} ‘nobody’, \textit{niks} ‘nothing’, \textit{nooit} ‘never’, \textit{nêrens} ‘nowhere’ and \textit{geen/g’n} ‘no’ (see \citealt{Huddlestone2010}). As discussed in \sectref{sec:afr:2.1}, they usually take the place of \textit{nie}\textsubscript{1}, as in \REF{ex:afr:28a}, although they can appear before the finite verb as well. When they do, as in \REF{ex:afr:28b}, the clause does not feature \textit{nie}\textsubscript{1} but still calls for the presence of \textit{nie}\textsubscript{2} (see \citealt{OlmenBreed2015}: 237).\footnote{Note, for the sake of completeness, that the slot before the finite verb is not limited to subjects in a V2 language like Afrikaans (see \sectref{sec:afr:1}). In \textit{ons sal hulle nooit vergeet nie} ‘we will never forget them’ (cf. \REF{ex:afr:28d}), \textit{nooit} ‘never’ appears in its “normal” position and the clause conforms to the pattern in \REF{ex:afr:28a}. With \textit{nooit} in initial position, as in \textit{nooit sal ons hulle vergeet nie}‘never will we forget them’, the clause has the same pattern as \REF{ex:afr:28b}.} This requirement is relaxed only when the negative indefinite occupies the clause-final position or serves as a short reply to a question. In these circumstances, \textit{nie}\textsubscript{2} is optional, as in \REF{ex:afr:28c}. Note also that, in informal language, negative indefinites can actually co-occur with \textit{nie}\textsubscript{1}. Sentences \REF{ex:afr:28d} and \REF{ex:afr:28e} are cases in point (see also \citealt[102--103]{Biberauer2009}).


\ea\label{ex:afr:28}
    \ea\label{ex:afr:28a}
        \gll Hy wil niks eet nie. \\ 
             3\textsc{sg.m.sbj} want\textsc{.prs} nothing eat \textsc{neg}\textsubscript{2} \\
        \glt ‘He does not want to eat anything.’
    
    \ex\label{ex:afr:28b}
        \gll Niemand het haar gesien nie. \\
             nobody have.\textsc{prs} \textsc{3sg.f.nsbj} see.\textsc{pst.ptcp} \textsc{neg}\textsubscript{2}\\
        \glt ‘Nobody saw her.’

    \ex\label{ex:afr:28c}
        \gll Hy eet niks (nie). \\ 
             3\textsc{sg.m.sbj} eat nothing \textsc{neg}\textsubscript{2} \\
        \glt ‘He does not eat anything.’
    
    \ex\label{ex:afr:28d}
        \gll Ons sal hulle nooit nie vergeet nie. \\
             \textsc{1pl} shall \textsc{3pl} never    \textsc{neg}\textsubscript{1} forget  \textsc{neg}\textsubscript{2} \\
        \glt ‘We will never forget them.’
        
    \ex\label{ex:afr:28e}
        \gll Ons sal niemand nie vergeet nie. \\
             1\textsc{pl} shall nobody \textsc{neg}\textsubscript{1} forget  \textsc{neg}\textsubscript{2}\\
        \glt ‘We will forget nobody.’
    \z
\z 


When a negative clause contains multiple indefinites, the first one is normally negative and the subsequent ones non-negative, like \textit{nooit} and \textit{êrens} ‘somewhere’ or \textit{iemand} ‘someone’ in \REF{ex:afr:29a} and \REF{ex:afr:29b}. Two negative indefinites in the same sentence cancel each other out, in formal language and according to prescriptive tradition. Example \REF{ex:afr:29c} thus receives the interpretation of ‘she hurt someone at some point’. Yet, in informal language, it is not uncommon for negation to spread to all indefinites in the clause, without resulting in any mutual cancelation of negative meanings or implying a more emphatic negation. So, colloquially, \REF{ex:afr:29c} signifies the same as \REF{ex:afr:29b}. In fact, experimental research by \citet{HuddlestoneSwart2014} indicates that clauses such as \REF{ex:afr:29c} are almost always understood as plainly negative by native speakers.

\ea\label{ex:afr:29}
    \ea\label{ex:afr:29a}
        \gll Hulle gaan nooit êrens kom nie. \\ 
             3\textsc{pl} go never somewhere come \textsc{neg}\textsubscript{2} \\
        \glt ‘They are never going to get anywhere.’
    
    \ex\label{ex:afr:29b}
        \gll Sy het nooit iemand seer-gemaak nie. \\
             \textsc{3sg.f.sbj} have.\textsc{prs} never someone pain-make.\textsc{pst.ptcp} \textsc{neg}\textsubscript{2} \\
        \glt ‘She never hurt anyone.’
    
    \ex\label{ex:afr:29c}
        \gll Sy het nooit niemand seer-gemaak nie. \\
             \textsc{3sg.f.sbj} have.\textsc{prs} never nobody pain-make.\textsc{pst.ptcp} \textsc{neg}\textsubscript{2}\\
        \glt ‘She hurt someone at some point.’ or ‘She never hurt anyone.’
    \z
\z


The pattern in \REF{ex:afr:29b} occurs much more often than that in \REF{ex:afr:29c}, though, at least in writing (see \citealt[139]{Huddlestone2010}). The only exception to this trend is when \textit{g’n} precedes another indefinite. Combinations of \textit{g’n} and a negative indefinite, as in \REF{ex:afr:30a}, are more frequent than those of \textit{g’n} and a non-negative indefinite, as in \REF{ex:afr:30b} (see \citealt[137]{Huddlestone2010}).\footnote{As pointed out in \sectref{sec:afr:4.4}, \textit{g’n} may also serve as a colloquial equivalent of \textit{nie}\textsubscript{1}. With emphasis on \textit{niemand} \REF{ex:afr:30a} can therefore be interpreted as ‘we didn’t NOT give anybody a lift’ as well. Similarly, in the right context, \REF{ex:afr:30b} can function as a strong denial too, meaning ‘we certainly did not give anyone a lift’.} Now consider \REF{ex:afr:30c} and \REF{ex:afr:30d}.

\ea\label{ex:afr:30}
    \ea\label{ex:afr:30a}
        \gll Ons het g’n niemand op-getel nie. \\ 
             \textsc{1pl} have.\textsc{prs} no nobody up-count.\textsc{pst.ptcp} \textsc{neg}\textsubscript{2} \\
        \glt ‘We did not give anyone a lift.’
    
    \ex\label{ex:afr:30b}
        \gll Ons het g’n iemand op-getel nie. \\
             \textsc{1pl} have.\textsc{prs} no someone up-count.\textsc{pst.ptcp} \textsc{neg}\textsubscript{2}  \\
        \glt ‘We did not give anyone a lift.’
    
    \ex\label{ex:afr:30c}
        \gll Hulle het geen/g’n kind-ers nie. \\ 
             \textsc{3pl} have.\textsc{prs} no child-\textsc{pl} \textsc{neg}\textsubscript{2} \\
        \glt ‘They do not have any children.’
    
    \ex\label{ex:afr:30d}
        \gll Ons het geen niemand op-getel nie. \\
             \textsc{1pl} have.\textsc{prs} no nobody up-count.\textsc{pst.ptcp} \textsc{neg}\textsubscript{2}\\
        \glt ‘We did not give anyone a lift.’ 
    
    \z
\z 


As simple negative determiners, like in \REF{ex:afr:30c}, \textit{geen} and \textit{g’n} are interchangeable, with the former being the more formal or written variant. In combinations with negative indefinites, as in \REF{ex:afr:30d}, however, \textit{geen} is generally considered much less acceptable than \textit{g’n} (see \citealt[167]{Huddlestone2010}, as well as \sectref{sec:afr:4.4}).

The negative indefinites’ non-negative counterparts – i.e. \textit{iemand}, \textit{iets} ‘something’, \textit{ooit} ‘ever’ and \textit{êrens} ‘somewhere’ – can clearly occur in negative clauses when a negative indefinite is present. On their own, they are essentially restricted to what \citet[31--52]{Haspelmath1997}, for his semantic map of indefinites, calls specific-known, specific-unknown, irrealis-non-specific, conditional, question and indirect negation contexts, which are illustrated in \REF{ex:afr:31a} to \REF{ex:afr:31f} respectively.

\ea\label{ex:afr:31}
    \ea\label{ex:afr:31a}
        \gll Ek het iemand ontmoet; sy is ’n navorser. \\ 
             1\textsc{sg.sbj} have.\textsc{prs} someone meet \textsc{3sg.f.sbj} be.\textsc{prs} \textsc{indf} researcher \\
        \glt ‘I met someone; she is a researcher.’
    
    \ex\label{ex:afr:31b}
        \gll Hy het iemand ontmoet maar ek weet nie wie nie. \\
             \textsc{3sg.m.sbj} have.\textsc{prs} someone meet but \textsc{1sg.sbj} know  \textsc{neg}\textsubscript{1} who  \textsc{neg}\textsubscript{2}\\
        \glt ‘He met someone but I do not know who.’
 
    \ex\label{ex:afr:31c}
        \gll Gaan iets lekker-s doen! \\ 
             go something nice-\textsc{gen} do \\
        \glt ‘Go and do something nice!’
    
    \ex\label{ex:afr:31d}
        \gll As jy ooit Potch toe kom, … \\ 
             if \textsc{2sg.sbj} ever Potch to come\\
        \glt ‘If you ever come to Potch, …’

    \ex\label{ex:afr:31e}
        \gll Gaan julle êrens heen? \\ 
             go \textsc{2pl} somewhere from.here \\
        \glt ‘Are you going somewhere?’
    
    \ex\label{ex:afr:31f}
        \gll Ek dink nie dat hulle iets gesien het nie. \\
             \textsc{1sg.sbj} think \textsc{neg}\textsubscript{1} \textsc{comp} \textsc{3pl} something  see.\textsc{pst.ptcp} have.\textsc{prs} \textsc{neg}\textsubscript{2} \\
        \glt ‘I do not think that they saw anything.’
    
    \z
\z 

For \citegen{Haspelmath1997} comparative and free-choice uses, the non-negative indefinites need to combine with \textit{enige} ‘any’, as in \REF{ex:afr:32a} and \REF{ex:afr:32b} respectively (see \sectref{sec:afr:4.2} for more about the negative polarity item \textit{enige}).\footnote{For the contexts in \REF{ex:afr:32}, Afrikaans can actually also use the more grammaticalized combinations \textit{enigiets} ‘anything at all’ and \textit{enigiemand} ‘anybody at all’.}

\ea\label{ex:afr:32}
    \ea\label{ex:afr:32a}
        \gll Die seun hardloop vinnig-er as enige iemand anders. \\ 
             \textsc{def} boy run fast-\textsc{cmpr} as any someone  else \\
        \glt ‘The boy runs faster than anyone else.’
    
    \ex\label{ex:afr:32b}
        \gll Hier-die probleem kan deur enige iemand opgelos word. \\
             here-\textsc{def} problem can by any someone solve.\textsc{pst.ptcp} become \\
        \glt ‘This problem can be solved by anyone.’
    \z
\z 


\subsection{Negative derivation and case marking}
\label{sec:afr:3.3}
It will come as no surprise that Afrikaans, as a language that makes case distinctions in singular personal pronouns only, does not have any “negative case marking”. Abessive meaning at the phrasal level is expressed by the preposition \textit{sonder} ‘without’, as in \REF{ex:afr:33a}.\footnote{In non-native Afrikaans, \textit{nie}\textsubscript{2}is sometimes added to sentences like \REF{ex:afr:33a}.} This marker can also introduce a finite subordinate clause, when followed by \textit{dat}, or a non-finite one, as in \REF{ex:afr:33b} and \REF{ex:afr:33c} respectively.

\ea\label{ex:afr:33}
    \ea\label{ex:afr:33a}
        \gll Ons was sonder kos en water. \\ 
             1\textsc{pl} be.\textsc{prs} without food and water \\
        \glt ‘We were without food and water.’
    
    \ex\label{ex:afr:33b}
        \gll Julle het dit gedoen sonder dat julle ons geraadpleeg het. \\
             2\textsc{pl} have.\textsc{prs} \textsc{3sg.n} do.\textsc{pst.ptcp} without \textsc{comp} \textsc{2pl} \textsc{1pl} consult.\textsc{pst.ptcp} have.\textsc{prs} \\
        \glt ‘You did it without consulting us.’
        
    \ex\label{ex:afr:33c}
        \gll Julle  het      dit     gedoen  sonder  om   ons   te  raadpleeg. \\
             2\textsc{pl} have.\textsc{prs} \textsc{3sg.n} do.\textsc{pst.ptcp} without for \textsc{1pl} to consult\\
        \glt ‘You did it without consulting us.’   
    \z
\z 

Negative derivation is somewhat under-researched in Afrikaans and an exhaustive study of the phenomenon goes beyond the scope of the present chapter in any case. Let the following comments therefore suffice (but see also \citealt{Huyssteen2019} and the Afrikaans dictionaries available via the  \citealt{VirtueleInstituutvirAfrikaans2019}).

The language seems to have a range of negative prefixes at its disposal but few negative suffixes. The affix -\textit{loos} ‘-less’, which turns nouns such as \textit{bodem} ‘bottom’ into adjectives such as \textit{bodemloos} ‘bottomless’, may be the only one – although \textit{vry} ‘free’, which can combine with a noun like \textit{vet} ‘fat’ to make an adjective like \textit{vetvry} ‘fat-free’, could potentially be considered here too. The list of prefixes includes \textit{on-} ‘un-’ and \textit{nie-} ‘not-’, which can attach to both nouns and adjectives without changing their word classes. The former signifies lack of the quality conveyed by the stem. \textit{(On)trou} ‘(un)faithfulness’ and \textit{(on)gesond} ‘(un)healthy’ are cases in point. The latter expresses exclusion from what the stem refers to, like in \textit{(nie-)roker} ‘(non-)smoker’ and \textit{(nie-)christelik} ‘(non-)Christian’. Another prefix, for which it is hard to provide a single translation, is \textit{wan-}. It signifies that what the stem conveys is inadequate or wrong, as in \textit{(wan)besteding} ‘(inappropriate) spending’ and \textit{(wan)smaak} ‘(bad) taste’. 

There also exist a number of prefixes that come from other languages. \textit{In-} ‘un-’, for instance, is found with adjectives (and nouns derived from them), like \textit{(in)formeel} ‘(in)formal’, and frequently varies with \textit{on-}, as \textit{(in)stabiel} ‘(in)stable’ and \textit{(on)stabiel} ‘(un)stable’ show. \textit{Non-} ‘non-’ is attested occasionally too, as an alternative to \textit{nie-}, as in \textit{(non)figuratief} ‘(non-)figurative’. Other examples are \textit{kontra-} ‘contra-’ and \textit{anti-} ‘anti-’, in nouns like \textit{(Kontra)reformasie} ‘(Counter\nobreakdash-)\-Reformation’ and adjectives like \textit{(anti)fascisties} ‘(anti-)fascist’, where they have the meaning of ‘opposed to’. This last marker can express the sense of ‘reverse of’ as well, as in (\textit{anti)klimaks} ‘(anti-)climax’. Two more mainly nominal or adjectival prefixes that are borrowed are \textit{eks-} ‘ex-’ and \textit{a-} ‘a-’. The first one, as in \textit{(eks-)burgemeester} ‘(ex-)mayor’, means ‘former’ whereas the second one, as in \textit{(a)moreel} ‘(a)moral’, signifies unrelatedness to what the stem conveys. 

Affixes that seem to operate primarily – though not exclusively – in the verbal domain, finally, are \textit{dis}{}- ‘dis-’ and \textit{ont-} ‘un-’. The former turns the meaning of a verb into its opposite, as in \textit{(dis)kwalifiseer} ‘(dis)qualify’, but is not very productive. The latter has a variety of functions (which will not all be discussed here). It can, for instance, combine with verbs to convey the reverse of their original semantics, as in \textit{(ont)laai} ‘(un)load’ and \textit{(ont)koppel} ‘(dis)connect’. The prefix may also be attached to a noun, with the resulting verb expressing the act of removing what the stem denotes, as in \textit{ontkurk} ‘uncork’ and \textit{onthoof} ‘decapitate’. It can be employed to derive verbs from adjectives as well. The meaning produced this way is that of undoing the quality conveyed by the stem. \textit{Ontrond} ‘unround’ (e.g. when pronouncing a sound) is a case in point and so is \textit{ontgroen} ‘haze’ (e.g. during an initiation ceremony), i.e. making someone not or less green in the metaphorical sense.

\section{Other aspects of negation}
\subsection{The scope of negation}
\label{sec:afr:4.1}
If negation has scope over only one constituent, for instance, for contrastive purposes, \textit{nie}\textsubscript{1} and \textit{nie}\textsubscript{2} can just surround the constituent in question, as in \REF{ex:afr:34a}. To express such a contrast in a coordination construction, \textit{nie}\textsubscript{1} can move from its ordinary place in \REF{ex:afr:34b} and directly precede the constituent in focus, as in \REF{ex:afr:34c} (see \citealt[231--232]{Robbers1992}). This change in position has a similar effect in \REF{ex:afr:34d} and \REF{ex:afr:34e}.


\ea\label{ex:afr:34}
    \ea\label{ex:afr:34a}
        \gll Adri, nie Daniel nie, is       verantwoordelik. \\ 
             Adri  \textsc{neg}\textsubscript{1} Daniel \textsc{neg}\textsubscript{2} be.\textsc{prs} responsible \\
        \glt ‘Adri, not Daniel, is responsible.’
    
    \ex\label{ex:afr:34b}
        \gll Daniel is nie verantwoordelik nie, maar Adri. \\
             Daniel be.\textsc{prs} \textsc{neg}\textsubscript{1} responsible      \textsc{neg}\textsubscript{2} but Adri\\
        \glt ‘Daniel is not responsible but Adri.’

    \ex\label{ex:afr:34c}
        \gll Nie Daniel is verantwoordelik nie, maar Adri. \\ 
             \textsc{neg}\textsubscript{1} Daniel be.\textsc{prs} responsible \textsc{neg}\textsubscript{2} but Adri \\
        \glt ‘Not Daniel is responsible but Adri.’
    
    \ex\label{ex:afr:34d}
        \gll Daniel mag dit nie sê nie. \\
             Daniel  may.\textsc{prs} \textsc{3sg.n} \textsc{neg}\textsubscript{1} say  \textsc{neg}\textsubscript{2}\\
        \glt ‘Daniel is not allowed to say that.’

    \ex\label{ex:afr:34e}
        \gll Daniel mag nie dit sê nie. \\
             Daniel may.\textsc{prs} \textsc{neg}\textsubscript{1} \textsc{3sg.n} say  \textsc{neg}\textsubscript{2}\\
        \glt ‘Daniel is not allowed to say \textit{that}.’
    \z
\z 

Another area of interest for issues of scope is adverbs (see \citealt[122--124]{Biberauer2008}). If negation ranges over them, \textit{nie}\textsubscript{1} comes before them, as in \REF{ex:afr:35a}. If an adverb has scope over negation, however, \textit{nie}\textsubscript{1} follows it, for which \REF{ex:afr:35b} can serve as a (somewhat marked) example. Not surprisingly therefore, sentential adverbs like \textit{waarskynlik} normally always precede \textit{nie}\textsubscript{1}, as in \REF{ex:afr:35c} (\textit{nie waarskynlik} ‘not probably’ would be acceptable, though, when contrasted to \textit{maar beslis} ‘but definitely’).


\ea\label{ex:afr:35}
    \ea\label{ex:afr:35a}
        \gll Hy lag nie altyd vir my grap-pie-s nie. \\ 
             \textsc{3sg.m.sbj} laugh \textsc{neg}\textsubscript{1} always for \textsc{1sg.nsbj} joke-\textsc{dim-pl} \textsc{neg}\textsubscript{2}
 \\
        \glt ‘He does not always laugh at my jokes.’
    
    \ex\label{ex:afr:35b}
        \gll Hy lag altyd nie vir my grap-pie-s nie. \\
             \textsc{3sg.m.sbj} laugh always \textsc{neg}\textsubscript{1} for \textsc{1sg.nsbj} joke-\textsc{dim-pl} \textsc{neg}\textsubscript{2} \\
        \glt ‘He always does not laugh at my jokes.’
   
    \ex\label{ex:afr:35c}
        \gll  Hulle kom waarskynlik nie Potch toe nie. \\ 
             \textsc{3pl} come probably \textsc{neg}\textsubscript{1}  Potch to \textsc{neg}\textsubscript{2} \\
        \glt ‘They are probably not coming to Potch.’
    \z
\z 

A final phenomenon relevant to the scope of negation is the position of \textit{nie}\textsubscript{2} in complex clauses. As mentioned in \sectref{sec:afr:2.1}, it is usually found at the end of the entire sentence, including any subordinate clause that the negation ranges over – such as the complement clause in \REF{ex:afr:3c}, repeated here as \REF{ex:afr:36a}, and the relative clause in \REF{ex:afr:36b}.

\ea\label{ex:afr:36}
    \ea\label{ex:afr:36a}
        \gll Ek het nie gesê dat die kar stukkend was nie. \\ 
             \textsc{1sg.sbj} have.\textsc{prs} \textsc{neg}\textsubscript{1} say.\textsc{pst.ptcp} \textsc{comp}  \textsc{def} car broken be.\textsc{pst} \textsc{neg}\textsubscript{2} \\
        \glt ‘I did not say that the car was broken.’
    
    \ex\label{ex:afr:36b}
        \gll Dit is nie die plek waar sy begrawe is nie. \\
             \textsc{3sg.n} be.\textsc{prs} \textsc{neg}\textsubscript{1}  \textsc{def} place  where  \textsc{3sg.f.sbj} bury.\textsc{pst.ptcp} be.\textsc{prs}  \textsc{neg}\textsubscript{2}\\
        \glt ‘This is not the place where she is buried.’
    \z
\z 

Still, when a subordinate clause falls outside the scope of negation, \textit{nie}\textsubscript{2} appears at the end of the main clause and not at the end of the whole sentence. In \REF{ex:afr:37a}, for instance, the subordinate clause provides a “positive” reason for the “negative” state of affairs in the main clause and \textit{nie}\textsubscript{2} occurs before the conjunction \textit{aangesien} ‘because’.\footnote{The issue of scope with conjunctions meaning ‘because’ is interesting.\textit{Want} ‘because’ is a coordinator and the clause that it introduces therefore always falls outside the scope of a preceding clause’s negation. Only the pattern in \REF{ex:afr:37a} – and \REF{ex:afr:37c} – is possible. By contrast, like \textit{aangesien}, \textit{omdat} ‘because’ is a subordinator and the clause headed by it may be inside the scope of a foregoing clause’s negation in particular contexts. For instance, if one wanted to say ‘he does not want to eat because he is ill but because …’, the reason would fall within the scope of the main clause negation. In that case,\textit{nie}\textsubscript{2} would follow the first subordinate clause:\textit{hy wil nie eet omdat hy siek is nie maar omdat …}} Conversely, when the negation applies only to the subordinate clause, \textit{nie}\textsubscript{2} is placed at the end of that clause and not at the end of the complete sentence. In \REF{ex:afr:37b}, for example, the main clause makes a “positive” claim, about people who are defined “negatively” in the relative clause, and \textit{nie}\textsubscript{2} occurs after \textit{drink} ‘drink’ rather than \textit{vreemd} ‘weird’. Similarly, when a negative clause is followed by a coordinated positive clause, as in \REF{ex:afr:37c}, \textit{nie}\textsubscript{2} is found at the end of the first clause and not at that of the complex clause. 

\ea\label{ex:afr:37}
    \ea\label{ex:afr:37a}
        \gll Hy wil nie eet nie aangesien hy siek is. \\ 
             \textsc{3sg.m.sbj} want.\textsc{prs} \textsc{neg}\textsubscript{1} eat \textsc{neg}\textsubscript{2} because \textsc{3sg.m.sbj} ill be.\textsc{prs} \\
        \glt ‘He does not want to eat, because he is ill.’
    
    \ex\label{ex:afr:37b}
        \gll Mens-e wat nie drink nie, is vreemd. \\
             human-\textsc{pl} what \textsc{neg}\textsc\textsubscript{1} drink \textsc{neg}\textsubscript{2} be.\textsc{prs} weird \\
        \glt ‘People who do not drink are weird.’
    \ex\label{ex:afr:37c}
        \gll Sy lees nie boek-e nie en hy is ’n skrywer. \\
             \textsc{3sg.f.sbj} read  \textsc{neg}\textsubscript{1} book-\textsc{pl} \textsc{neg}\textsubscript{2} and  \textsc{3sg.m.sbj} be.\textsc{prs} \textsc{indf} writer \\
        \glt ‘She does not read books and he is a writer.’    
    \z
\z 

See also \sectref{sec:afr:2.4} about issues of scope in non-finite subordinate clauses.

\subsection{Negative polarity}
\label{sec:afr:4.2}
The inventory of negative polarity items includes the modal auxiliaries \textit{hoef} ‘need’ and \textit{durf} ‘dare’. The former expresses absence of necessity in negative clauses, as in \REF{ex:afr:38a}, but can  be employed in other negative polarity contexts too, such as polar questions and clauses with the restrictive adverb \textit{net} ‘only’, as \REF{ex:afr:38b} and \REF{ex:afr:38c} respectively show. Its use in a plainly positive clause like \REF{ex:afr:38d} is not grammatical, however.

\ea\label{ex:afr:38}
    \ea\label{ex:afr:38a}
        \gll Ons hoef nie die wet-te te gehoorsaam nie. \\ 
             \textsc{1pl} need \textsc{neg}\textsubscript{1}  \textsc{def} law-\textsc{pl} to obey \textsc{neg}\textsubscript{2} \\
        \glt ‘We do not need to obey the laws.’
    
    \ex\label{ex:afr:38b}
        \gll Hoef ons die wet-te te gehoorsaam? \\
             need \textsc{1pl} \textsc{def} law-\textsc{pl} to obey\\
        \glt  ‘Do we need to obey the laws?’
    
    \ex\label{ex:afr:38c}
        \gll Ons hoef net die wet-te te gehoorsaam. \\ 
             \textsc{1pl} need only \textsc{def} law-\textsc{pl} to obey \\
        \glt ‘We only need to obey the laws.’
    
    \ex\label{ex:afr:38d}
        \gll ?Ons hoef die wet-te te gehoorsaam. \\
             \textsc{1pl} need \textsc{def} law-\textsc{pl} to obey \\
        \glt ‘We need to obey the laws.’
    \z
\z 

\textit{Durf} exhibits the same behavior, as \REF{ex:afr:39} aims to suggest: \REF{ex:afr:39a} would be unacceptable without the negators and the conditional in \REF{ex:afr:39b} is yet another type of negative polarity context.

\ea\label{ex:afr:39}
    \ea\label{ex:afr:39a}
        \gll Ek durf dit nie sê nie. \\ 
             \textsc{1sg.sbj} dare \textsc{3sg.n} \textsc{neg}\textsubscript{1} say \textsc{neg}\textsubscript{2} \\
        \glt ‘I dare not say it.’
    
    \ex\label{ex:afr:39b}
        \gll As jy durf lag, moer ek jou. \\
             if \textsc{2sg.sbj} dare laugh hit \textsc{1sg.sbj} \textsc{2sg.nsbj} \\
        \glt ‘If you dare laugh, I will hit you.’
    \z
\z 

Another negative polarity item is \textit{veel} ‘much’ (see also \citealt{KotzéBreed2019}). It can appear in the negative clause in \REF{ex:afr:40a} but has to be replaced by \textit{baie} ‘much’ (a borrowing from Malay) in its positive equivalent in \REF{ex:afr:40b} (see \citealt{KotzéBreed2018}). \textit{Veel} serves as a(n often rather marked) negative polar alternative to \textit{baie}, which is always possible too, in a variety of contexts – such as with count and mass nouns like \REF{ex:afr:40a}, with eventive verbs like \REF{ex:afr:40c} and in comparatives like \REF{ex:afr:40d} (see \citealt[23--24]{Berghoff2016}, the inspiration for the examples). With gradable adjectives, as in \REF{ex:afr:40e}, \textit{veel} is simply not acceptable, though.

\ea\label{ex:afr:40}
    \ea\label{ex:afr:40a}
        \gll Daar is nie veel/baie skade nie. \\ 
             There be.\textsc{prs} \textsc{neg}\textsubscript{1} much damage \textsc{neg}\textsubscript{2} \\
        \glt ‘There is not much damage.’
    
    \ex\label{ex:afr:40b}
        \gll Daar is baie/*veel skade. \\
             there be.\textsc{prs} much damage \\
        \glt ‘There is much damage.’

    \ex\label{ex:afr:40c}
        \gll Sy reis nie veel/baie nie. \\ 
             \textsc{3sg.f.sbj} travel \textsc{neg}\textsubscript{1} much \textsc{neg}\textsubscript{2} \\
        \glt ‘She does not travel a lot.’
    
    \ex\label{ex:afr:40d}
        \gll Gys is (nie) veel/baie slim-mer as Piet (nie). \\
            Gys be.\textsc{prs} \textsc{neg}\textsubscript{1} much smart-\textsc{cmpr} as Piet \textsc{neg}\textsubscript{2} \\
        \glt  ‘Gys is (not) much smarter than Piet.’

    \ex\label{ex:afr:40e}
        \gll Jan is (nie) baie/*veel snaaks (nie). \\
             Jan be.\textsc{prs} \textsc{neg}\textsubscript{1} much funny \textsc{neg}\textsubscript{2}\\
        \glt ‘Jan is (not) very funny.’
    \z
\z 

Importantly, in \REF{ex:afr:40d}, \textit{veel} can even occur without negation. Other positive contexts in which it can be found are the excessive and equative constructions in \REF{ex:afr:41a}, where \textit{baie} would be very marked, and questions like \REF{ex:afr:41b} as \textit{hoeveel} ‘how much’, where \textit{baie} is impossible. The likely reason why \textit{veel} is found in these positive environments is that phrases like \textit{te veel} ‘too much’ and \textit{ewe veel} ‘as much as’, which are occasionally joined up in writing, conventionalized as units before it became a negative polarity item (\textit{veel} is not negatively polar in Dutch).

\ea\label{ex:afr:41}
    \ea\label{ex:afr:41a}
        \gll Louise het te(/ewe) veel/?baie geld (as Henk). \\ 
             Louise have.\textsc{prs} too/as much money as Henk \\
        \glt  ‘Louise has too(/as) much money (as Henk).’
    
    \ex\label{ex:afr:41b}
        \gll Hoe-veel/*-baie geld het jy? \\
             how-much money have.\textsc{prs} \textsc{2sg.sbj} \\
        \glt ‘How much money do you have?’
    \z
\z  

The last item to be considered here is \textit{enige} ‘any’ (see \citealt[159--160]{Donaldson1993} too). As mentioned in \sectref{sec:afr:3.2}, it can appear in \citegen{Haspelmath1997} comparative and free-choice uses, of which \REF{ex:afr:42a} is another example. It can also occur in conditional, interrogative and indirect negation contexts, as in \REF{ex:afr:42b} to \REF{ex:afr:42d} respectively, but not in specific-known, specific-unknown and irrealis-non-specific ones. Sentence \REF{ex:afr:42e} is a case in point.

\ea\label{ex:afr:42}
    \ea\label{ex:afr:42a}
        \gll Enige student kan hier-die klas-se by-woon. \\ 
             any student can.\textsc{prs} here-\textsc{def} class-\textsc{pl} at-reside \\
        \glt ‘Any student can attend these classes.’
    
    \ex\label{ex:afr:42b}
        \gll As jy enige vra-e het, is nou jou kans. \\
             if \textsc{2sg.sbj} any question-\textsc{pl} have.\textsc{prs} be.\textsc{prs} now \textsc{2sg.nsbj} chance \\
        \glt ‘If you have any questions, now is your chance.’

    \ex\label{ex:afr:42c}
        \gll Lees jy op die oomblik enige boek? \\ 
             read \textsc{2sg.sbj} on \textsc{def} moment any book \\
        \glt ‘Are you reading any book at the moment?’
    
    \ex\label{ex:afr:42d}
        \gll Ek dink nie dat julle enige moeite hier-mee sal hê nie.\\
             \textsc{1sg.sbj} think \textsc{neg}\textsubscript{1} \textsc{comp} \textsc{2pl} any difficulty here-with shall have.\textsc{inf} \textsc{neg}\textsubscript{2} \\
        \glt ‘I do not think that you will have any difficulty with this.’
    
    \ex\label{ex:afr:42e}
        \gll *Ek het enige boek-e. \\
             \textsc{1sg.sbj} have.\textsc{prs} any book-\textsc{pl} \\
        \glt ‘I have some books.’
    \z
\z 

\textit{Enige} can be attested under direct negation as well. In this context, it acquires the meaning of ‘any … at all’, though, as \REF{ex:afr:43} shows.

\ea\label{ex:afr:43}
    \gll Ek het nie enige foon nie. \\ 
         \textsc{1sg.sbj} have.\textsc{prs} \textsc{neg}\textsubscript{1} any phone \textsc{neg}\textsubscript{2} \\
    \glt ‘I do not have any phone at all.’
 
\z 

\subsection{Marking of noun phrases in the scope of negation}

To our knowledge, there are no changes in, for instance, case or focus marking for noun phrases within the scope of negation in Afrikaans. The only exception seems to be when \textit{geen} heads a singular indefinite noun phrase and the article \textit{’n} disappears (see \sectref{sec:afr:2.3}, as well as \citealt[124]{Ponelis1979}).

\subsection{Reinforcing negation}
\label{sec:afr:4.4}
Afrikaans has various ways of reinforcing negation (see also \sectref{sec:afr:4.2} about \textit{enige} under direct negation). One can, for instance, insert adverbs such as \textit{glad} ‘altogether’, \textit{geheel} ‘wholly’ and \textit{hoegenaamd} ‘whatsoever’ before \textit{nie}\textsubscript{1}, as in \REF{ex:afr:44a} (see \citealt[26]{Ponelis1968}). Another strategy is the use of \textit{geensins} ‘in no way’ instead of \textit{nie}\textsubscript{1}, as in \REF{ex:afr:44b}, or that of \textit{niks} and \textit{nooit}, with their literal meanings still somewhat present, when \textit{nie}\textsubscript{1} would technically have sufficed, as in \REF{ex:afr:44c} and \REF{ex:afr:44d} respectively (see \citealt[109--110]{Biberauer2008}).

\ea\label{ex:afr:44}
    \ea\label{ex:afr:44a}
        \gll Hy wil glad/geheel/hoegenaamd nie eet nie. \\ 
             \textsc{3sg.m.sbj} want.\textsc{prs} altogether/wholly/whatsoever \textsc{neg}\textsubscript{1} eat \textsc{neg}\textsubscript{2} \\
        \glt ‘He does not want to eat at all.’
    
    \ex\label{ex:afr:44b}
        \gll Sy wil geensins eet nie. \\
             3\textsc{sg.f.sbj} want.\textsc{prs} in.no.way  eat \textsc{neg}\textsubscript{2} \\
        \glt ‘She does not want to eat at all.’
        
    \ex\label{ex:afr:44c}
        \gll Hulle het niks verander nie. \\ 
             3\textsc{pl} have.\textsc{prs} nothing  change \textsc{neg}\textsubscript{2} \\
        \glt ‘They have not changed a thing.’ (i.e. ‘They have not changed at all.’)
    
    \ex\label{ex:afr:44d}
        \gll Hulle is vegetaries; hulle eet nooit vleis nie. \\
             3\textsc{pl} be.\textsc{prs} vegetarian \textsc{3pl} eat never meat  \textsc{neg}\textsubscript{2} \\
        \glt ‘They are vegetarian; they never eat meat.’ (i.e. ‘They do not eat meat at all.’)
    \z
\z 

In informal language, it is also possible to use \textit{niks} \textit{(nie)} to reinforce negation in circumstances where the meaning of ‘nothing’ is entirely bleached, as in \REF{ex:afr:45a}. Another option in this register is \textit{g’n (nie)} – though not \textit{geen (nie)} – but only to strongly deny “an activated, salient, discourse-old proposition” \citep[3]{Huddlestone2010}. The dialogue in \REF{ex:afr:45b} can serve as an illustration. One more informal strategy involves the collocation \textit{g’n niks (nie)},\footnote{Note, for the sake of completeness, that this combination can be employed for other purposes than strong standard negation. As \citet[265, 269]{Huddlestone2010} points out, \textit{g’n} can also function as a reinforcer of \textit{niks} as a genuine negative indefinite, resulting in the meaning ‘nothing at all’, and as a negator canceling out the negation of indefinite \textit{niks}, producing the meaning ‘not nothing’.} as in \REF{ex:afr:45c}.\footnote{As one of the reviewers points out, negation may also be mitigated, of course. One way in which Afrikaans can do this is by putting an adverb like \textit{eintlik} ‘actually’ or \textit{rêrig} ‘really’ after \textit{nie}\textsubscript{1}. The mitigated equivalent of \REF{ex:afr:44a} would thus be: \textit{hy wil nie eintlik/rêrig eet nie} ‘he does not really want to eat’.}

\ea\label{ex:afr:45a}
    \gll Sy is niks (nie) tevrede nie. \\ 
         \textsc{3sg.f.sbj} be.\textsc{prs} nothing \textsc{neg}\textsubscript{1} satisfied \textsc{neg}\textsubscript{2} \\
    \glt ‘She is not remotely satisfied.’ (\citealt[11]{BiberauerCyrino2009}) 
\z 

\ea\label{ex:afr:45b}
    \begin{xlist}
    \exi{A: }\label{ex:afr:45b.A}
        \gll Jy is my vriend. \\
         2\textsc{sg.sbj} be.\textsc{prs} \textsc{1sg.nsbj} friend \\
    \glt ‘You are my friend.’ 
    \exi{B: }\label{ex:afr:45b.B}
        \gll Ek is g’n jou vriend nie. \\
         \textsc{1sg.sbj} be.\textsc{prs} no \textsc{2sg.nsbj} friend  \textsc{neg}\textsubscript{2}\\
    \glt ‘I am NOT your friend.’ (\citealt[9]{BiberauerCyrino2009})
    \end{xlist} 
\z
        
\ea\label{ex:afr:45c}
    \gll Dit is g’n niks (nie) so moeilik nie. \\
         3\textsc{sg.n} be.\textsc{prs} no nothing \textsc{neg}\textsubscript{1} so difficult \textsc{neg}\textsubscript{2} \\
    \glt ‘It is NOT so difficult.’ (\citealt[12]{BiberauerCyrino2009})
\z


\subsection{Negation, coordination and complex clauses}

The way that negation behaves in complex clauses, both coordinate and subordinate ones, has been discussed at some length (see \sectref{sec:afr:2.4} and \ref{sec:afr:4.1}). Important things to add here are that Afrikaans does not have any special negative subordinators such as ‘lest’ and that the language possesses no dedicated constructions for contrastive negation (see \citealt{Silvennoinen2019}), to our knowledge. Examples of this last type of negation have been given in \REF{ex:afr:34a} to \REF{ex:afr:34c}. Two more comments are in order. First, the special negative coordinator in Afrikaans is \textit{nóg} ‘nor’. As \REF{ex:afr:46a} shows, it does not allow \textit{nie}\textsubscript{2} and, as \REF{ex:afr:46b} makes clear, it can also link phrases and occur only before the second conjunct.

\ea\label{ex:afr:46}
    \ea\label{ex:afr:46a}
        \gll Nóg sy het ons gehelp, nóg haar ma het iets gedoen. \\  
             nor  \textsc{3sg.f.sbj} have.\textsc{prs} \textsc{1pl} help.\textsc{pst.ptcp} nor \textsc{3sg.f.nsbj}  mum  have.\textsc{prs} something do.\textsc{pst.ptcp}\\
        \glt ‘Neither did she help us nor did her mum do anything.’
    
    \ex\label{ex:afr:46b}
        \gll (Nóg) jy, nóg ek moet gaan. \\
             nor \textsc{2sg.sbj} nor \textsc{1sg.sbj} must.\textsc{prs} go\\
        \glt ‘(Neither) you nor I should go.’
    \z
\z  


Second, the language has an intriguing construction where a negative clause is coordinated by \textit{of} ‘or’ with a positive clause and, semantically, the former serves as a restrictive positive subordinate clause for the latter. The literal translation of \REF{ex:afr:47} is ‘you cannot move or they gossip’ but its meaning is ‘as soon as you move, they gossip’ (see also \citealt[29]{Ponelis1968}, for further discussion).

\ea\label{ex:afr:47}
    \gll Jy kan nie jou roer nie of hulle skinder. \\ 
         \textsc{2sg.sbj} can.\textsc{prs} \textsc{neg}\textsubscript{1} \textsc{2sg.nsbj} stir \textsc{neg}\textsubscript{2} or 3\textsc{pl} gossip \\
    \glt ‘As soon as you move, they gossip.’ \citep[29]{Ponelis1968}
\z 

\subsection{Miscellaneous aspects of negation}
\subsubsection{Negative transport}

The phenomenon also known as negative raising is attested in the language, though it does not appear to have been studied in any detail yet. Compare the “non-raised” case in \REF{ex:afr:48a} featuring the negation in the finite complement clause, with its “raised” counterpart in \REF{ex:afr:48b}, where the main verb is negated instead but the meaning is more or less the same. Examples with a non-finite subordinate clause are given in \REF{ex:afr:48c} and \REF{ex:afr:48d}.

\ea\label{ex:afr:48}
    \ea\label{ex:afr:48a}
        \gll Ek dink dat ons nie ’n keuse het nie. \\ 
             \textsc{1sg.sbj} think \textsc{comp} \textsc{1pl} \textsc{neg}\textsubscript{1} \textsc{indf} choice  have.\textsc{prs} \textsc{neg}\textsubscript{2} \\
        \glt ‘I think that we do not have a choice.’
    
    \ex\label{ex:afr:48b}
        \gll Ek dink nie dat ons ’n keuse het nie. \\
             \textsc{1sg.sbj} think \textsc{neg}\textsubscript{1} \textsc{comp} \textsc{1pl} \textsc{indf} choice  have.\textsc{prs} \textsc{neg}\textsubscript{2} \\
        \glt ‘I do not think that we have a choice.’
  
    \ex\label{ex:afr:48c}
        \gll Ek beplan om nie lank hier te wees nie. \\ 
             \textsc{1sg.sbj} plan for \textsc{neg}\textsubscript{1} long here to be.\textsc{inf} \textsc{neg}\textsubscript{2} \\
        \glt ‘I plan not to be here long.’
    
    \ex\label{ex:afr:48d}
        \gll Ek beplan nie om lank hier te wees nie. \\
             \textsc{1sg.sbj} plan    \textsc{neg}\textsubscript{1} for long here to be.\textsc{inf} \textsc{neg}\textsubscript{2}\\
        \glt ‘I do not plan to be here long.’
    \z
\z 

An exploration based on \citegen[323]{Horn1989} and \citegen[51]{Popp2016} lists of predicates suggests that negative raising can occur with at least: the attitude predicates \textit{die indruk kry} ‘feel, get the impression’, \textit{dink} ‘think’, \textit{glo} ‘believe’ and \textit{hoop} ‘hope’; the perception predicates \textit{klink asof} ‘sound like’ and \textit{lyk asof} ‘look like’; the volition predicates \textit{beplan} ‘plan’ and \textit{wil} ‘want’; and the judgment predicates \textit{aanraai} ‘advise’, \textit{dit is gewens} ‘it is desirable’ and \textit{voorstel} ‘suggest’. More in-depth (corpus) research is needed, though, to get the full picture.

\subsubsection{Expletive negation}

The literature discusses a number of situations in which negation is overtly expressed but does not actually contribute to the meaning. Verbs with contradictory semantics like \textit{weier} ‘refuse’ and inherently negative adverbs like \textit{nouliks} ‘barely’ have been mentioned in this context. The expletive use of \textit{nie}\textsubscript{2} with such items is, however, almost consistently seen as unacceptable by native speakers (see footnote \ref{footnote:afr:8}). Expletive negation in conditional clauses with \textit{tensy} ‘unless’ and \textit{voordat} ‘before’ receives a more – though still not generally – positive reception (see \citealt[60--63]{Potgieter2018}). In informal language, the subordinate clauses in \REF{ex:afr:49} can (but need not) be negated in spite of the fact that their meaning remains positive.\footnote{\citet[102]{Biberauer2009} adds that \textit{nie}\textsubscript{1} may not be compulsory in the types of expletive negation discussed so far but that it is yet unclear what the difference is between the cases with \textit{nie}\textsubscript{1} and \textit{nie}\textsubscript{2} and those with just \textit{nie}\textsubscript{2}.}

\ea\label{ex:afr:49}
    \ea\label{ex:afr:49a}
        \gll Jy kan nie gaan voordat jou werk nie klaar is nie. \\ 
             \textsc{2sg.sbj} can.\textsc{prs} \textsc{neg}\textsubscript{1} go    before  2\textsc{sg.nsbj} work \textsc{neg}\textsubscript{1} ready  be.\textsc{prs} \textsc{neg}\textsubscript{2} \\
        \glt ‘You cannot go before your work is done.’ \citep[405]{Donaldson1993}
    
    \ex\label{ex:afr:49b}
        \gll Tensy jou werk nie klaar is nie, kan jy nie gaan nie. \\
             unless  2\textsc{sg.nsbj} work \textsc{neg}\textsubscript{1} ready  be.\textsc{prs} \textsc{neg}\textsubscript{2} can.\textsc{prs} \textsc{2sg.sbj} \textsc{neg}\textsubscript{1} go \textsc{neg}\textsubscript{2} \\
        \glt ‘Unless your work is done, you cannot go.’\citep[405]{Donaldson1993}
    \z
\z 

Exclamatives are another context where the input of negation, whose occurrence there is widely accepted (see \citealt[63--64]{Potgieter2018}), is not entirely clear: the cases in \REF{ex:afr:50} are essentially both expressions of, for instance, surprise at or emotion toward a positive state of affairs.

\ea\label{ex:afr:50}
    \ea\label{ex:afr:50a}
        \gll Hoe vinnig gaan die lewe (nie) verby (nie)! \\ 
             how fast go \textsc{def} life \textsc{neg}\textsubscript{1} past    \textsc{neg}\textsubscript{2} \\
        \glt ‘How fast life passes us by!’ (\citealt{BiberauerPotgieter2017}: 3)
    
    \ex\label{ex:afr:50b}
        \gll Is dit nie alte fraai nie! \\
             be.\textsc{prs} \textsc{3sg.n} \textsc{neg}\textsubscript{1} all.too  beautiful \textsc{neg}\textsubscript{2}\\
        \glt ‘Isn’t that beautiful!’ (\citealt{BiberauerPotgieter2017}: 3)
    \z
\z 

In fact, in so-called wh-exclamatives, like \REF{ex:afr:50a}, the negation can quite easily be dropped with little change in meaning. The difference has to do with the discourse status of the information. A positive wh-exclamative conveys new information whereas a negative one expresses a reaction to known information. In what are termed polar exclamatives, like \REF{ex:afr:50b}, by contrast, the negation is not optional. Deleting it would result in something that looks like an interrogative with an interpretation of ‘is it all too beautiful?’.

In a request such as \REF{ex:afr:51}, finally, the addition of negation also has little impact on the act performed with the sentence. It functions as a call on the addressee to close the door with or without \textit{nie}\textsubscript{1} … \textit{nie}\textsubscript{2}. The information’s discourse status is again different, however. With the negation, the information is presented as known, which may make the request more insistent – implying, for instance, that the door should have been closed already. Still, in other situations, it may also have a mitigating effect.

\ea\label{ex:afr:51}
    \gll Kan jy (nie) asseblief die deur toemaak (nie)? \\
         can.\textsc{prs} \textsc{2sg.sbj} \textsc{neg}\textsubscript{1} please \textsc{def} door close      \textsc{neg}\textsubscript{2}\\
    \glt ‘Could(n’t) you close the door please?’
\z 

\subsubsection{Diachronic notes and observations}

The origin of the two-part negation is a matter of much debate in Afrikaans linguistics, to which the present section cannot really do justice. It will therefore only provide a brief description of this and other diachronic issues and point to some relevant literature. As to the two-part negation (see \sectref{sec:afr:2.1}), some scholars have argued that it comes directly from particular Dutch dialects with a similar phenomenon while others maintain that it is an innovation of Afrikaans (e.g. \textit{nie}\textsubscript{2} as deriving from a clause-final resumptive negator). Among the latter, there is disagreement about whether its development was autonomous or whether contact with, for example, the indigenous Khoisan languages in the Cape played a role. The evidence for any single one of these positions seems to be quite scarce, however. A fairly comprehensive overview of the entire debate is given in \citet{Roberge2000}. Other important references are \citet{Besten1986}, \citet{Bell2001}, \citet{Deumert2017}, \citet{Rensburg2018}.\footnote{An issue that, to our knowledge, has not been taken into account fully in the debate about the two-part negation’s origin is the syntactic position of\textit{nie}\textsubscript{1}. As one of the reviewers rightly points out, its unmarked position in a clause meaning ‘I have not seen the house’ is before the lexical object \textit{die huis} ‘the house’: \textit{ek het nie}\textsubscript{1}\textrm{ die huis gesien nie}\textsubscript{2}. In (Present-day) Dutch,\textit{niet} ‘not’ comes after \textit{het huis} ‘the house’:\textit{ik heb het huis niet gezien}. This difference (possibly of a diachronic nature) needs to be explained in any scenario linking Afrikaans negation to Dutch negation.} 

There is also some discussion on the origin of \textit{moenie} in negative imperatives (see \sectref{sec:afr:2.2}). It has, for instance, been argued to be a calque of some eighteenth-century Portuguese Pidgin or Creole construction. It could also be the product of an independent process of grammaticalization for which the politeness strategy of deontic declaratives (i.e. \textit{jy moet nie …} ‘you should not …’) was the source. Relevant research includes \citet{Besten2004} and \citet{Olmen2010}. 

As regards the negation of stative predication (see \sectref{sec:afr:2.3}), the comparison with Dutch and German, where the negative determiner construction is the only acceptable option for certain types of stative predicate, suggests that the grammaticality in Afrikaans of \textit{nie}\textsubscript{1} … \textit{nie}\textsubscript{2} in all types results from an extension of the standard negation construction. 

Finally, contrasting the language with Dutch also indirectly supports \citegen[232]{Donaldson1988} and \citegen[843]{Hoeksema2010} claim that the use of \textit{enige} in comparative and free-choice contexts (see \sectref{sec:afr:3.2} and \sectref{sec:afr:4.2}) is due to English influence. Unlike \textit{any} and Afrikaans \textit{enige}, Dutch \textit{enig} cannot be used in this way.

\section{Summary}
% \hc{LSP will deal with that table later}
\tabref{tab:afr:1} gives an overview of the most prominent negative constructions in Afrikaans, together with their functions. For details, we refer to the relevant sections in this chapter. 
%
% \textbf{\textit{nie}\textsubscript{1} … \textit{nie}\textsubscript{2} ‘not … not’}
%
% \begin{tabularx}{\textwidth}{QQ}
% \textbf{Functions} & \textbf{Notes}\\
% standard negation
% \newline interrogative negation
% \newline non-canonical imperative negation
% \newline negation of stative predication
% \newline subordinate clause negation
% \newline contrastive negation
% &
% \textit{nie}\textsubscript{1} is clause-medial
% \newline \textit{nie}\textsubscript{2} clause-final
% \newline \textit{nie}\textsubscript{2} can be omitted in certain, very specific contexts
% \newline the negation is entirely symmetrical
% \end{tabularx}
%
% \textbf{\textit{moenie}\textsubscript{1} … \textit{nie}\textsubscript{2} ‘must.not … not’}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
%  canonical imperative negation & the negation exhibits constructional asymmetry in finiteness\\
% \end{tabularx}
%
% \textbf{\textit{moenie}\textsubscript{1} \textit{dat … nie}\textsubscript{2} ‘must.not that … not’}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% ‘don’t let (someone do something)!’ & -
% \end{tabularx}
%
%
% \textbf{\textit{geen/g’n} … (\textit{nie}\textsubscript{2}) ‘no … not’}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
%  negation of possessive predication
% \newline negation of proper inclusion
% \newline (locative-)existential negation
% &
% standard negation is more common in these types of stative predication
% \newline \textit{g’n} is the informal variant \\
% \end{tabularx}
%
%
%
% \textbf{\textit{nee} ‘no’ }
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% disagreement with positive polar questions
% \newline agreement with negative polar questions
% \newline denying a questioner’s presumed concerns & - \\
% \end{tabularx}
%
%
% \textbf{\textit{ja} ‘yes’}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% agreement with positive polar questions & as a reply to negative polar questions, it is vague vis-à-vis (dis)agreement \\
% \end{tabularx}
%
%
% \textbf{\textit{ja-nee} ‘yes-no’}\\
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% subjectively qualified agreement & - \\
% \end{tabularx}
%
%
% \textbf{\textit{niemand … nie}\textsubscript{2} ‘nobody … not’; \textit{niks … nie}\textsubscript{2} ‘nothing … not’; \textit{nooit … nie}\textsubscript{2} ‘never … not’; \textit{nêrens … nie}\textsubscript{2} ‘nowhere … not’}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
%  negative indefinites & \textit{nie}\textsubscript{2} is optional in certain, very specific contexts \textit{nie}\textsubscript{1} can be added in colloquial language with more than one indefinite, only the first one is normally negative in informal language, multiple negative indefinites are possible without negations canceling each other out \textit{niks} and \textit{nooit} may be used simply to reinforce a standard negation in particular contexts \\
% \end{tabularx}
%
%
% \textbf{\textit{iemand} ‘someone’; \textit{iets} ‘something’; \textit{ooit} ‘ever’; \textit{êrens} ‘somewhere’}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
%  Indefinites
% & they require \textit{enige} ‘any’ in comparative and free-choice contexts \\
% \midrule
% \end{tabularx}
%
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}
%
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}
%
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}
%
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}
%
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}
%
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}
%
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}
%
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}
%
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}
%
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}
%
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}
%
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}
%
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}
%
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}
%
%
% \textbf{}
%
% \begin{tabularx}{\textwidth}{QQ}
%  \textbf{Functions} & \textbf{Notes}\\
% \end{tabularx}


\begin{table} \footnotesize
\begin{tabularx}{\textwidth}{p{2cm}QQ}
\lsptoprule
{Construction} & {Functions} & {Notes} \\
\midrule
\textit{nie}\textsubscript{1} … \textit{nie}\textsubscript{2} ‘not … not’
&
standard negation
\newline interrogative negation
\newline non-canonical imperative negation
\newline negation of stative predication
\newline subordinate clause negation
\newline contrastive negation
&
\textit{nie}\textsubscript{1} is clause-medial
\newline \textit{nie}\textsubscript{2} clause-final
\newline \textit{nie}\textsubscript{2} can be omitted in certain, very specific contexts
\newline the negation is entirely symmetrical \\
\midrule

\textit{moenie}\textsubscript{1} … \textit{nie}\textsubscript{2} ‘must.not … not’ & canonical imperative negation & the negation exhibits constructional asymmetry in finiteness \\
\midrule

\textit{moenie}\textsubscript{1} \textit{dat … nie}\textsubscript{2} ‘must.not that … not’ & ‘don’t let (someone do something)!’ & \\
\midrule

\textit{geen/g’n} … (\textit{nie}\textsubscript{2}) ‘no … not’
&
negation of possessive predication
\newline negation of proper inclusion
\newline (locative-)existential negation
&
standard negation is more common in these types of stative predication
\newline \textit{g’n} is the informal variant \\
\midrule

\textit{nee} ‘no’
&
disagreement with positive polar questions
\newline agreement with negative polar questions
\newline denying a questioner’s presumed concerns & \\
\midrule

\textit{ja} ‘yes’ & agreement with positive polar questions & as a reply to negative polar questions, it is vague vis-à-vis (dis)agreement \\
\midrule

\textit{ja-nee} ‘yes-no’ & subjectively qualified agreement & \\
\midrule

\textit{niemand … nie}\textsubscript{2} ‘nobody … not’
\newline \textit{niks … nie}\textsubscript{2} ‘nothing … not’
\newline \textit{nooit … nie}\textsubscript{2} ‘never … not’
\newline \textit{nêrens … nie}\textsubscript{2} ‘nowhere … not’
&
{negative indefinites}
&
\textit{nie}\textsubscript{2} is optional in certain, very specific contexts
\newline \textit{nie}\textsubscript{1} can be added in colloquial language
\newline with more than one indefinite, only the first one is normally negative
\newline in informal language, multiple negative indefinites are possible without negations canceling each other out
\newline \textit{niks} and \textit{nooit} may be used simply to reinforce a standard negation in particular contexts \\
\midrule
\end{tabularx}
\caption{Negation in Afrikaans}
\label{tab:afr:1}
\end{table}

\begin{table} \footnotesize
\begin{tabularx}{\textwidth}{p{3cm}QQ}
\midrule
{Construction} & {Functions} & {Notes} \\
\midrule
\textit{iemand} ‘someone’
\newline \textit{iets} ‘something’
\newline \textit{ooit} ‘ever’
\newline \textit{êrens} ‘somewhere’
&
Indefinites
& they require \textit{enige} ‘any’ in comparative and free-choice contexts \\
\midrule
\textit{sonder} ‘without’ & abessive meaning & it can be followed by a noun phrase and a finite or non-finite subordinate clause \\
\midrule
\textit{-loos} ‘-less’ & forming adjectives from nouns, with abessive meaning & \\
\midrule

\textit{on-} / \textit{in-} ‘un-’ & lack of quality denoted by stem & \textit{on-} is more frequent \\
\midrule

\textit{nie-} / \textit{non-} ‘not- / non-’ & exclusion from what stem denotes & \textit{nie-} is more frequent \\
\midrule

\textit{anti-} ‘anti-’ & ‘opposed to’, ‘reverse of’ & \\
\midrule

\textit{kontra-} ‘contra-‘ & ‘opposed to’ & \\
\midrule

\textit{eks-} ‘ex-’ & ‘former’ & \\
\midrule

\textit{a-} ‘a-’ & unrelatedness to what stem denotes & \\
\midrule

\textit{ont-} ‘un-’ & reverse of action denoted by verbal stem
\newline removal of referent of nominal stem
\newline undoing of quality denoted by adjectival stem & \\
\midrule

\textit{hoef} ‘need’
\newline \textit{durf} ‘dare’ & modal verbs & they can only occur in negative polarity contexts \\
% \midrule
% \end{tabularx}
% \end{table}
%
% \begin{table} \footnotesize
% \begin{tabularx}{\textwidth}{>{\raggedright}p{2cm}QQ}
% \lsptoprule
{Construction} & {Functions} & {Notes} \\
\midrule
\textit{veel} ‘much’ & negatively polar alternative to \textit{baie} ‘much’ & it can also be found in interrogative, excessive and equative constructions \\
\midrule

\textit{enige} ‘any’ & negative polarity item & it can occur in comparative, free-choice, conditional, interrogative and (in)direct negation contexts \\
\midrule
\end{tabularx}
\end{table}

\begin{table} \footnotesize
\begin{tabularx}{\textwidth}{p{4cm}QQ}
\midrule
{Construction} & {Functions} & {Notes} \\
\midrule


\textit{hoegenaamd} / \textit{glad} /\textit{geheel} / \textit{nie}\textsubscript{1} … \textit{nie}\textsubscript{2} ‘whatsoever / altogether / wholly not … not’ & {reinforcing negation} & \\
\midrule

\textit{geensins … nie}\textsubscript{2} ‘in.no.way … not’ & {reinforcing negation} & \\
\midrule

\textit{(nóg) … nóg …} ‘neither … nor …’ & {negative coordination} & \\
\midrule

\textit{tensy} ‘unless’
\newline \textit{voordat} ‘before’ & conjunctions with a (potentially) conditional meaning  & the subordinate clauses that they introduce may feature expletive negation in informal language \\

\lspbottomrule
\end{tabularx}
\end{table}

\section*{Abbreviations}
\begin{tabularx}{.45\textwidth}{lQ}
 \scshape 1 & first person \\
 \scshape 2 & second person \\
 \scshape 3 & third person \\
 \scshape cmpr & comparative \\
 \scshape comp & complementizer \\
 \scshape def & definite \\
 \scshape dim & diminutive \\
 \scshape f & feminine \\
 \scshape gen & genitive \\
 \scshape indf & indefinite \\
 \scshape inf & infinitive \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
 \scshape m & masculine \\
 \scshape n & neuter \\
 \scshape neg & negation \\
 \scshape nsbj & non-subject \\
 \scshape pl & plural \\
 \scshape prs & present \\
 \scshape pst & past \\
 \scshape ptcp & participle \\
 \scshape sbj & subject \\
 \scshape sg & singular \\
 \scshape supl & superlative \\
\end{tabularx}


\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
\end{document} 
