\documentclass[output=paper,colorlinks,citecolor=brown,draft]{langscibook}
\ChapterDOI{10.5281/zenodo.18234101}
\author{Simon E. Overall\affiliation{University of Otago}}
\title{Negation in Aguaruna}
\abstract{The chapter describes negation in Aguaruna (Chicham). Verbal negation, including standard negation, makes use of inflectional verbal suffixes, part of a rich system of verbal morphology in the language. Two suffixed negators are selected according to aspect and finiteness of the verb, and a third  suffix appears only in some nondeclarative clause types. The verbal paradigms are largely symmetrical, and the only category that is completely neutralized under negation is a distinction between plain and familiar imperative. Nominals can be derivationally negated, and this strategy is the basis for negating clauses with nominal predicates. When marked on the predicate, negation has clausal scope, and this is true both of verbal and nominal predicates. The chapter concludes with a discussion of the diachronic relation between the negative existential verb and one of the verbal negative suffixes, and suggests that the former is the source of the latter.
\keywords{Chicham, Amazonian, grammaticalization, aspect, tense}
}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 

\section{The language}\label{sec:agu:1}

This chapter describes the grammatical means for expressing negation in Agua\-runa (ISO: \href{https://iso639-3.sil.org/code/agr}{agr}, Glottocode: \href{https://glottolog.org/resource/languoid/id/agua1253}{agua1253}), a Chicham language of north Peru. The Chicham family (formerly known as Jivaroan) consists of five closely related languages all spoken in the south of Ecuador and north of Peru, in an area mostly within the Santiago, Pastaza and Marañón River basins. Aguaruna speakers refer to the language as Iiniá Chicham or Awajún, and the latter is the currently preferred label in Peruvian Spanish.
Over 56,000 people identified as first language speakers of Aguaruna in the 2017 census \citep[64]{MinisteriodeEducacióndelPerú2018}. Most speakers live in areas of native title in Peru’s Amazonas region, with communities also in San Martín, Loreto, and Cajamarca regions. Bilingual schools have operated in Aguaruna communities since 1953, and there is a high degree of Aguaruna–Spanish bilingualism.


Much of the data in this chapter comes from \citet{Overall2017a}, a full descriptive grammar of Aguaruna. This is supplemented with the author’s field data, personal communication with native speakers, and written sources including a translation of the New Testament (\citealt{LigaBíblica2008}) and public social media posts.
Examples labelled SLCK were provided in personal communication by Segundo Lucio Cungumas Kujancham. 

Aguaruna typically shows SV/APV constituent order, although with considerable freedom of movement for information structural purposes, and NP arguments need not be overt. Nominative-accusative alignment is manifested in case marking of NPs (zero-marked nominative, accusative enclitic \textit{=na}) and verbal agreement, as well as grammatical processes such as nominalization and switch-reference, which distinguish subject (S or A) from non-subjects.

The morphology is both head and dependent marking: at the clause level, subjects and speech-act participant objects are marked with verbal suffixes, and NP arguments are marked for case; within the possessive NP, possessed nouns are morphologically marked as possessed, along with person and number of the possessor, and possessors are marked with genitive case, or accusative case if they are pronominal.

Morphological structure is basically agglutinating, with some fusion. Most allomorphy is the result of regular processes of syncope and apocope that elide the vowels of some CV syllables. The full underlying forms of morphs are cited in the text, while numbered examples show both surface forms and full morphemic forms. 
In examples, the first line gives the surface form after the application of vowel elision, and the second line gives the underlying morphological forms. Loanwords from Spanish are italicised in the second line. Examples are given in IPA, with the exception of those taken from written sources (such as example (\ref{ex:agu:3})), in which the first line (between angle brackets <…>) is in the Spanish-based vernacular orthography, which differs from IPA in the following graphs: <ch> = [tʃ]; <e> = [ɨ]; <g> = [ɰ] (between vowels) or [ŋ]; <h> = [ʔ]; <j> = [h]; <sh> = [ʃ]; <y> = [j]. In  the glosses, affixes are separated with hyphens and clitics with equal signs. The analysis of a given form as an affix or a clitic is based on whether the form is associated with a particular word class or a phrasal constituent: affixes are positioned by morphological principles, while clitics are positioned by syntactic principles. See \citet[9--10]{Overall2017a} for a fuller discussion. Square brackets are used in examples to delimit potentially unclear multiword constituents, such as the complement clauses in example (\ref{ex:agu:3}). 

Finite verbs in Aguaruna are marked for the following verbal grammatical categories: aspect, tense, person/number and mood/modality. These categories, along with derivational morphology and polarity, are marked in the verbal word, which is entirely suffixing apart from an unproductive causative prefix. The ordering of the categories is largely as predicted by \citet{Bybee1985}, except that the inflectional category of aspect precedes negation in Aguaruna. The productive suffixing morphology can be usefully viewed in terms of morphological slots, as in the schematic overview in \figref{fig:agu:1}. The valency-changing derivational morphology of slot A is not obligatory, but the rest of the categories are obligatorily specified, assuming that the paradigms include zero-marked members that contrast meaningfully with overt markers. Consequently, a finite verb can be defined as one that includes specification for all of the grammatical categories in slots B to G.

\begin{figure}
\begin{tabular}{llllllll}
%\begin{tabularx}{\textwidth}{XXXXXXXX}
\lsptoprule
& {A} & {B} & {C} & {D} & {E} & {F} & {G}\\
%\midrule
{\scshape root} & {Valency} & {Object} & {Aspect} & {Negation} & {Tense} & {Subject} & {Mood}\\
\lspbottomrule
%\end{tabularx}
\end{tabular}
\caption{Productive morphological slots in the verbal word}
\label{fig:agu:1}
\end{figure}

Aspect marking in slot C forms a set of stems which can then combine with tense (slot E) and other morphology. The major aspectual stems are \textsc{imperfective} and \textsc{perfective}. Imperfective is marked with a suffix \textit{{}-a}, or \textit{{}-ina} if the subject is plural, while perfective is marked with one of a set of suffixes that also give some further specification of the manner of the action. The selection of perfective suffix is largely lexically specified \citep[328]{Overall2017a}. Less frequently encountered aspectual stems are \textsc{potential} and \textsc{durative} (which combines only with imperative mood and gives the sense of ‘keep on doing’ or ‘do for a while’). The \textsc{unmarked} verb root also forms a stem to which nominalizers and some finite verbal forms can be added. \tabref{tab:agu:1} shows the full range of possibilities for marking in the aspect morphological slot.

\begin{table}
\begin{tabular}{lll}
\lsptoprule
{Stem}                          & \multicolumn{2}{l}{Marker(s)}\\
\midrule 
\multirow{2}{*}{imperfective} & \textit{{}-a}  & singular subject \\
                                & \textit{{}-ina}       & plural subject\\
{perfective}           & \multicolumn{2}{l}{(one of a paradigm of suffixes)}\\
{potential}            & \textit{-mai}     & \\
{durative}             & \textit{{}-ma}       & only combines with imperative \\
{unmarked}             & \multicolumn{2}{l}{Ø}\\
\lspbottomrule
\end{tabular}
\caption{Aspect marked/unmarked stems}
\label{tab:agu:1}
\end{table}

Aguaruna has a rich system of tense marking (slot E) with “four synthetic past tenses, plus a past tense nominalizer, and two synthetic future tenses, plus a future nominalizer” \citep[16]{Overall2017a} in addition to zero-marked present/immediate past tense. There are also various auxiliary constructions and nominalized forms; it is likely that some past tense forms are relatively recently grammaticalized from auxiliary constructions. Note in particular the use of verbs nominalized with the suffix \textit{{}-u} (see \sectref{sec:agu:3.3}), which may function in lieu of a finite verb, giving a non-firsthand evidentiality reading \citep{Overall2014}. This construction is very frequent in traditional stories, and can be seen combined with the copula enclitic in \REF{ex:agu:36} \textit{wɨ\nobreakdash-u=ai} (go.\textsc{pfv-nmlz=cop.3.decl}) ‘he went’ and without the copula in \REF{ex:agu:40} \textit{dakita-u} (refuse.\textsc{pfv-nmlz}) ‘he refused’.

Although they form morphologically distinct paradigms, tense and aspect markers typically combine in predictable ways. Most notably, non-present tenses almost always combine with perfective aspect. The morphologically unmarked present tense is formed on the imperfective stem, while the perfective stem with no overt tense marker is treated as an immediate past tense (\citealt{Overall2017a}: 339–340). The remote past tense is unusual in that it can combine with imperfective, perfective, or the unmarked stem (as in \REF{ex:agu:6} below). The unmarked verb stem appears in a few finite verb forms in addition to remote past tense, but these constructions appear to be lexically restricted and unproductive \citep[253]{Overall2017a}.

Aguaruna is a clause-chaining language and makes heavy use of dependent verb forms and nominalizations, with relatively few finite verbs in narrative texts. Dependent verb forms lack tense and mood marking. They are marked for their dependent status, for person of the subject (using a different paradigm from that used in finite verbs) and for switch-reference. 

The remainder of this chapter is structured according to the questionnaire provided by the editors (\citetv{chapters/questionnaire}). Section \sectref{sec:agu:2} describes clausal negation, beginning with standard negation then moving on to non-declarative clause types, stative predications, non-main clauses and lexicalized negation. Section \sectref{sec:agu:3} describes non-clausal negation, including negative replies, negative indefinites, and negative derivation. Section \sectref{sec:agu:4} addresses some other aspects of negation, including questions of scope and marking of NPs under negation. In \sectref{sec:agu:4.6} some diachronic observations are presented. Finally \sectref{sec:agu:5} offers some concluding remarks.

\section{Clausal negation}\label{sec:agu:2}
\subsection{Standard negation}\label{sec:agu:2.1}

Standard negation (SN) is defined as applying to a pragmatically unmarked, finite, declarative clause with a verbal predicate \citep{Payne1985,Miestamo2005,Miestamo2017}. The definition of verbal finiteness provided above makes it relatively easy to identify SN in Aguaruna. SN makes use of two verbal suffixes, \nobreakdash-\textit{tʃa} and \textit{\nobreakdash-tsu,} which appear in morphological slot D (\figref{fig:agu:1}). Examples (\ref{ex:agu:1}--\ref{ex:agu:3}) illustrate the use of these suffixes. Note that in both suffixes the affricate is typically changed to a fricative when syncope places it in syllable-final position, following a general allophonic rule (\citealt{Overall2017a}: 45–46). Also note that the negative suffix \textit{{}-tʃa} triggers accent shift to the verb root, as in \REF{ex:agu:2a} where the corresponding affirmative form has the accent falling on a suffix \REF{ex:agu:2b} \citep[113]{Overall2017a}.

\ea\label{ex:agu:1}
    \ea\label{ex:agu:1a}
        \glll dakástʃattahai \\ 
              daka-sa-tʃa-tata-ha-i \\
              wait-\textsc{pfv-neg-fut-1sg-decl} \\
        \glt ‘I will not wait.’ \citep[403]{Overall2017a}
    \ex\label{ex:agu:1b}
        \glll dakástathai \\
              daka-sa-tata-ha-i \\
              wait-\textsc{pfv-fut-1sg-decl} \\
        \glt ‘I will wait.’ (SLCK)
    \z
\z 

\ea\label{ex:agu:2}
    \ea\label{ex:agu:2a}
        \glll waínkaʃmahai \\ 
              waina-ka-tʃa-ma-ha-i \\
              see-\textsc{pfv-neg-pst-1sg-decl} \\
        \glt ‘I didn’t see it.’ (SLCK)
    \ex\label{ex:agu:2b}
        \glll wainkámhai \\
              waina-ka-ma-ha-i \\
              see-\textsc{pfv-pst-1sg-decl}\\
        \glt ‘I saw it.’ (SLCK)
    \z
\z  

\ea\label{ex:agu:3}
    \ea\label{ex:agu:3a}
        \glll <Wika dekatsjai ame tame nunak> \\ 
              wi=ka  dɨka-tsu-ha-i  [amɨ  ta-mɨ nu=na=ka]\\
              \textsc{1sg}=\textsc{top}  know.\textsc{ipfv-neg-1sg-decl}  \textsc{2sg}  say.\textsc{ipfv-2sg} \textsc{ana=acc=top} \\
        \glt ‘I don’t know what you’re talking about.’ (\citealt{LigaBíblica2008}: 65; Matthew 26:70)
    \ex\label{ex:agu:3b}
        \glll <Mesías Kristu daagtin taatna nunak dekajai> \\
              [\emph{mesías} Kristu  naa-hu-tinu  ta-a-tina=a nu=na=ka]  dɨka-ha-i \\ 
              Messiah  Christ  name-\textsc{pssd-attr}  arrive-\textsc{pfv-fut.nmlz=cop.3} \textsc{ana=acc=top}  know.\textsc{ipfv-1sg-decl}\\
        \glt ‘I know that Messiah (called Christ) is coming.’ (\citealt{LigaBíblica2008}: 176; John 4:25)
    \z
\z

The use of verbal suffixes means that Aguaruna conforms to Dryer's Type 6 (SO[V-Neg]) in his \citeyear{Dryer2013-wals144}  typology. The SN construction is symmetric \citep{Miestamo2005}, as the negative clause differs from the positive clause only in the presence of the negative marker itself (but note that negative clauses apparently favour the use of topic marking, see \sectref{sec:agu:4.3}). SN is also symmetric at the level of the paradigm, as all finite declarative clause types can be negated with the same SN constructions (in fact, negative paradigms are consistently symmetric throughout the grammar, while the constructions are asymmetric for imperative/prohibitive, see \sectref{sec:agu:2.2.1}).

The primary factor in the selection of negative suffix \textit{{}-tʃa}  or \textit{\nobreakdash-tsu} appears to be aspect (and see \citealt{Croft1991}: 10–11 for some parallel cases of aspect-conditioned alternation in SN markers), but because tense and aspect markers rarely vary independently (despite forming distinct paradigms) there are few examples that allow this to be demonstrated robustly.

Within the most frequently encountered finite verb forms, the selection of negator is predictable based on the combinations of aspect and tense: perfective aspect combined with non-present tenses triggers the marker \textit{{}-tʃa} (this was illustrated with future tense in example \REF{ex:agu:1a}, and past tense in \REF{ex:agu:2a}), while imperfective aspect combined with present tense triggers \textit{\nobreakdash-tsu} \REF{ex:agu:3a}.

The few tensed forms that involve neither perfective nor imperfective aspect show that the selection of negator is only partly predictable from the present/non-present tense distinction. Finite verb forms based on the potential stem are negated with \textit{\nobreakdash-tʃa} in past tense \REF{ex:agu:4a} and \textit{\nobreakdash-tsu} in present tense \REF{ex:agu:5}. These constructions are infrequent, however \citep[336]{Overall2017a}, as the potential stem is typically followed by a nominalizer, and then uses the nominal negation strategy involving the negating/nominalizing suffix \textit{\nobreakdash-tʃau} (for example, the form in \REF{ex:agu:4b} was first offered to translate Spanish \textit{no pude cruzar} ‘I couldn’t cross’, and compare \REF{ex:agu:6}; see \sectref{sec:agu:2.3} for details of the nominal negation strategy).

\ea\label{ex:agu:4}
    \ea\label{ex:agu:4a}
        \glll katiímaitʃamɨ \\ 
              katii-mai-tʃa-mɨ \\
              cross-\textsc{pot-neg-pst.3.decl} \\
        \glt ‘One could not cross.’ (for example, a flooded river) (SLCK)
    \ex\label{ex:agu:4b}
        \glll katiímaintʃau dɨkapsámhai \\
              katii-mai-inu-tʃau  dɨkapɨ-sa-ma-ha-i \\
              cross-\textsc{pot-nmlz-neg.nmlz}  feel-\textsc{pfv-pst-1sg-decl} \\
        \glt \textsc{‘}I didn’t feel able to cross.’ (SLCK)
    \z
\z 

\ea\label{ex:agu:5}
    \glll wíka hiínmaitsuhai, húna itípka nahánkun hɨ̃ɰ̃ã́  puhúttahai \\ 
           wi=ka  hiina-mai-tsu-ha-i  hu=na  itipaka nahana-ku-nu  hɨɰã  puhu-tata-ha-i\\
          \textsc{1sg=top}  go.out-\textsc{pot-neg-1sg-decl}  \textsc{dem.prox=acc}  kilt.\textsc{acc} make.\textsc{ipfv-sim-1sg.ss}  house.\textsc{loc}  live-\textsc{fut-1sg-decl}\\
    \glt ‘I can’t go out, I’ll stay at home making this kilt.’ (agr040719\_01)\footnote{Unpublished Aguaruna examples are cited with the code of the recording, currently being prepared for archiving.}
\z 

\ea\label{ex:agu:6}
    \glll tʃitʃasán inímtanaʃ inimáintʃau dɨkapɨ́jahai \\ 
          tʃitʃa-sa-nu  inima-ta=na=ʃa  inima-mai-inu-tʃau dɨkapɨ-ja-ha-i\\
          speak-\textsc{sub-1sg.ss}  ask-\textsc{nmlz=acc=add}  ask-\textsc{pot-nmlz-neg.nmlz} feel-\textsc{rem.pst-1sg-decl}\\
    \glt ‘I felt unable to even ask a question by speaking (because I didn’t know Spanish).’ \citep[609]{Overall2017a}
\z 

In contrast to the potential forms, remote past tense combined with the unmarked stem is negated with \textit{{}-tsa} (which \citealt{Overall2017a} assumes is an allomorph of \textit{\nobreakdash-tsu}), and not with \nobreakdash-\textit{tʃa} as would be expected if non-present tense were the relevant factor in the selection of negator \REF{ex:agu:7}.

\ea\label{ex:agu:7}
    \glll wahúk háaha núna tumáin dɨkáptsajahai\\ 
          [wahuka  ha-a-ha  nu=na]  tu-mai-inu dɨkapɨ-tsa-ja-ha-i \\
          how  be.sick-\textsc{ipfv-1sg}  \textsc{ana=acc}  say-\textsc{pot-nmlz} feel-\textsc{neg-rem.pst-1sg-decl} \\
    \glt ‘I didn’t feel able to say in what way I was sick.’ \citep[402]{Overall2017a}
\z 

The ‘normative’ form, also combining with unmarked stem, gives a gnomic reading of “the way things are done”, and this form is negated with \textit{{}-tʃa} \REF{ex:agu:8}.

\ea\label{ex:agu:8}
    \glll <uvashkam jagkinmayag juushtayame> \\ 
          uva=ʃakama  haŋki=numa=ja=ka  huu-tʃa-tajamɨ \\
          grape=\textsc{add}  thorn=\textsc{loc=abl=top}  take-\textsc{neg-norm} \\
    \glt ‘We don’t pick grapes from amongst thorns.’ (\citealt{LigaBíblica2008}: 19; Matthew 7:16)
\z 

Although part of the tense paradigm in Aguaruna, the normative form is synchronically a nominalizer in the other Chicham languages, and the use of \textit{{}-tʃa} is consistent with its origin in a nominalizer (see \sectref{sec:agu:3.3}). Similarly, the ‘narrative past’ tense marker \textit{{}-haku} appears to have its etymological source in a nominalizer, and this means that it is incompatible with SN suffixes. Instead, a periphrastic negation strategy is used, which involves negating the verb root with the negative nominalizer (see \sectref{sec:agu:2.3} and \sectref{sec:agu:3.3}), then adding the tense marker to an auxiliary verb \REF{ex:agu:9}.

\ea\label{ex:agu:9}
    \glll dɨ́katʃu ahakúi utʃíŋmitkatnak \\ 
          dɨka-tʃau  a-haku=i utʃiha-mitika-ta=na=ka\\
          know-\textsc{neg.nmlz}  \textsc{cop-narr.pst=cop.3.decl} give.birth-\textsc{caus-act.nmlz=acc=top} \\
    \glt ‘(They) didn’t know how to make her give birth.’ \citep[300]{Overall2017a}
\z 

In sum, finite tensed forms with perfective or imperfective aspect show a clear distribution of negators according to the aspect marking. Combinations of negators with other stems (potential or unmarked) are not readily predictable, and simply depend on different constructions.\footnote{The durative stem appears only in combination with positive imperative and therefore does not combine with negative marking.} Such forms have low token frequency and may be paraphrased with constructions that are not strictly verbal, such as the nominalization in \REF{ex:agu:4b} above. In non-finite verb forms, including nominalizations, the marker \textit{\nobreakdash-tʃa} is the default (\sectref{sec:agu:2.4}, \sectref{sec:agu:3.3}).

The mood marking paradigm in morphological slot G (\figref{fig:agu:1}) includes traditional sentence-type markers (declarative, interrogative) and some epistemic markers (e.g. ‘counter-expectation’, ‘speculative’ see \REF{ex:agu:48}) as well as a dedicated marker of traditional narrative genre. These other declarative clause types seem to use SN strategies, although there are few examples. Negation in interrogative clauses also uses SN, as described in \sectref{sec:agu:2.2.2}. Imperative mood is marked in the same morphological slot as tense, and its negation is described in \sectref{sec:agu:2.2.1}.

\subsection{Negation in non-declaratives}\label{sec:agu:2.2}
\subsubsection{Negation of commands}\label{sec:agu:2.2.1}

There are three distinct morphological forms for the expression of positive and negative commands in Aguaruna:

(i)  Canonical imperative is based on the suffix \textit{{}-ta} (\textit{{}-ti} in third person) related to future tense markers and the purpose clause construction;

(ii)  The suffix \textit{{}-mi} marks hortative;

(iii) The apprehensive marker \textit{{}-i} forms the basis of the prohibitive form – functionally the negative counterpart of imperative.

\tabref{tab:agu:2} shows an overview of imperative forms and their negated counterparts in Aguaruna.\footnote{The ‘plain’ versus ‘familiar’ distinction in second person positive imperative is neutralized in prohibitive, which has only a single set of forms.}

\begin{table}
\begin{tabular}{llll}
\lsptoprule
{person} & \multicolumn{2}{c}{{positive/command}} & {negative/prohibition}\\
               & {plain}  & {{familiar}} & \\
               \midrule 
{\scshape 2sg} & \textit{-ta}       & \textit{-kia}     & \textit{-i-pa}\\
{\scshape 2pl} & \textit{-ta-hum}   & \textit{-khua}    & {\textit{-i-hupa} or \textit{-i-pa-humɨ}}\\
{\scshape 3sg} & \multicolumn{2}{c}{\textit{-ti}}       & \textit{-ĩ-ka}\\
{\scshape 3pl} & \multicolumn{2}{c}{\textit{-ti-numɨ}}  & \textit{-i-numɨ-ka}\\
{\scshape 1pl} & \multicolumn{2}{c}{\textit{-mi}}       & \textit{-tʃa-mi}\\
\lspbottomrule
\end{tabular}
\caption{Aguaruna imperative and prohibitive verb forms \citep[68]{Overall2017b}}
\label{tab:agu:2}
\end{table}

Two important points to note are that (i) the \textit{{}-ta} forms for positive imperative do not extend to \textsc{1pl}, which only uses the dedicated hortative marker -\textit{mi}; and (ii) only the hortative \textit{{}-mi} form is negated with the SN suffix -\textit{tʃa} (compare positive and negative forms \REF{ex:agu:10}), while the \textit{{}-ta} forms all have distinct prohibitive forms based on apprehensive marking, described below. (Note, however, that the etymological source of imperative \textit{{}-ta}, the immediate future tense marker, can be negated using SN as in \REF{ex:agu:18} below.)

\ea\label{ex:agu:10}
    \glll <Iina namakji aidau kuitamkami, tsuwapakchami.> \\ 
          ii=na  namaka-hĩ  a-ina-u  kuitama-ka-mi tsuwapa-ka-tʃa-mi \\
          \textsc{1pl}=\textsc{acc}  river-\textsc{pssd.1pl}  exist-\textsc{pl.ipfv-nmlz}  care.for-\textsc{pfv-hort} make.dirty-\textsc{pfv-neg-hort}\\
    \glt ‘Let’s look after our rivers, let’s not make them dirty.’ (\textit{Cultura Awajun}, Facebook post, 9 August 2019; the original post includes the Spanish translation ‘{No tiremos la basura a nuestros ríos, cuidémosla}.’)
\z 

Prohibitive (negative imperative) forms use the apprehensive suffix \textit{\nobreakdash-i} combined with suffixes that indicate person and number. This suffix normally forms a negative counterpart of the purpose clause (\sectref{sec:agu:2.4.3}), but uses a different set of person suffixes when forming a prohibitive clause, as shown in the rightmost column of \tabref{tab:agu:2} (compare these to the apprehensive set in \tabref{tab:agu:3}). Note that there are two forms for \textsc{2pl}: \textit{\nobreakdash-hupa} or a form with \textsc{2sg} \textit{{}-pa} followed by the \textsc{2pl} marker from \tabref{tab:agu:3}, i.e. \textit{\nobreakdash-pa-humɨ}. There does not appear to be any semantic distinction between the two forms for second person plural (\citealt{Overall2017b}: 74, also see \citealt{Peña2015}: 678 on a similar free variation in Wampis).

The prohibitive construction conforms to the fourth type described by \citet{AuweraEtAl2013}, as it involves a negator that is distinct from that used in SN, and is based on a form that is distinct from the positive imperative.\footnote{As noted above, negation of the \textsc{1pl} hortative form is performed with exactly the same marking as SN, added directly to the positive construction. It is unclear how this compares to crosslinguistic patterns, as there is little typological work directly addressing first person plural prohibitive forms – for example, \citet{AuweraEtAl2013} restrict themselves to second person prohibitives, and \citet{AuweraKrasnoukhova2020} explicitly exclude \textsc{1pl} forms for reasons of space.}

Examples (\ref{ex:agu:11}--\ref{ex:agu:12}) illustrate singular and plural imperative contrasted with prohibitive, \REF{ex:agu:13} shows the alternative plural prohibitive form, and \REF{ex:agu:14} gives a prohibitive in context, from a direct speech report in a recorded narrative. Example \REF{ex:agu:15} illustrates jussive, the third person imperative, and \REF{ex:agu:16} shows its negative counterpart, third person prohibitive, formed with the suffix \textit{{}-ka}, which is unique to this form.

\ea\label{ex:agu:11}
    \ea\label{ex:agu:11a}
        \glll atʃiktá \\ 
              atʃi-ka-ta \\
              grab-\textsc{pfv-imp} \\
        \glt ‘Grab (it)!’
    \ex\label{ex:agu:11b}
        \glll atʃikáipa \\
              atʃi-ka-i-pa \\
              grab-\textsc{pfv-appr-2.proh} \\
        \glt ‘Don’t grab (it)!’ \citep[74]{Overall2017b}
    \z
\z 

\ea\label{ex:agu:12}
    \ea\label{ex:agu:12a}
        \glll atʃiktáhum \\ 
              atʃi-ka-ta-humɨ \\
              grab-\textsc{pfv-imp-2pl} \\
        \glt ‘Grab (it)!’
    \ex\label{ex:agu:12b}
        \glll atʃikáiŋpa! \\
              atʃi-ka-i-hupa \\
              grab-\textsc{pfv{}-appr-2pl.proh} \\
        \glt ‘Don’t grab (it)!’ \citep[74]{Overall2017b}
    \z
\z 

\ea\label{ex:agu:13}
    \glll wáinkam ihuípahum!\\ 
          wainka\nobreakdash-mɨ  ihu-i-pa-humɨ\\
          in.vain-2  stab-\textsc{appr-2.proh-2pl} \\
    \glt ‘Don’t stab it in vain!’ \citep[75]{Overall2017b}
\z 

\ea\label{ex:agu:14}
    \glll atsá, ámɨk miníipa, wíka iʃámainnum wɨ́ɰahai, ámɨk titú puhustá \\ 
          atsa  amɨ=ka  mini-i-pa  wi=ka iʃama-mai-inu=numa  wɨ{}-a-ha-i  amɨ=ka  titu puhu-sa-ta \\
          no  \textsc{2sg}=\textsc{top}  come.\textsc{pfv-appr-2.proh}  \textsc{1sg}=\textsc{top} fear-\textsc{pot-nmlz=loc}  go\textsc{-ipfv-1sg-decl} \textsc{2sg=top} still live-\textsc{pfv-imp}\\
    \glt ‘No, don’t you come, I’m going into danger, you stay still!’ \citep[583]{Overall2017a}
\z           

\ea\label{ex:agu:15}
    \glll <Yutái susatajum yuwati>\\ 
          ju-taĩ  su-sa-ta-humɨ  ju-a-ti \\
          eat-\textsc{nmlz}  give-\textsc{pfv-imp-2pl}  eat-\textsc{pfv-juss} \\
    \glt ‘Give her food and let her eat!’ (\citealt{LigaBíblica2008}: 79; Mark 5:43)
\z 

\ea\label{ex:agu:16}
    \glll útʃi húna saɨ́ptʃihin ijáunakɨʃ juwáwaiŋka \\ 
          utʃi  hu=na  saɨpɨ-utʃi-hĩ=na  ija-u=na=kɨ=ʃa ju-aw-aĩ-ka \\
          child  \textsc{dem}.\textsc{prox}=\textsc{acc}  rind-\textsc{dim-pssd.3=acc}  fall.\textsc{ipfv-nmlz=acc=restr=add} eat-\textsc{pfv-appr.3sg-proh} \\
    \glt ‘Don’t let the child eat this rind that is falling!’ \citep[75]{Overall2017b}
\z 

\subsubsection{Negation of questions}\label{sec:agu:2.2.2}

Interrogative clauses use the same SN strategies as declarative clauses. Negated content questions (\ref{ex:agu:17}--\ref{ex:agu:18}) are rare in published material. 

\ea\label{ex:agu:17}
    \glll maʔ, “antúkta” táha! wáŋka ántasmɨ? \\ 
          [maʔ  antu-ka-ta  ta-ha]  [waŋka  anta-tsu-mɨ] \\
          hey  listen-\textsc{pfv-imp}  say.\textsc{ipfv-1sg.exclam}  why  listen\textsc{.ipfv-neg-2sg} \\
    \glt ‘Hey, “listen” I say! Why don’t you listen?’ \citep[563]{Overall2017a}
\z 

\ea\label{ex:agu:18}
    \glll wáŋka ahápaʃtamɨ? \\ 
          waŋka  ahapa-tʃa-ta-mɨ \\
          why  give.birth.\textsc{pfv-neg-imm.fut-2sg} \\
    \glt ‘Why aren’t you going to give birth?’ \citep[354]{Overall2017a}
\z           

\citet[394]{Overall2017a} notes that polar questions often include negative marking, apparently pleonastically (\REF{ex:agu:19}, also \REF{ex:agu:26} and \REF{ex:agu:41} below). Example \REF{ex:agu:19} is a rhetorical question: the speaker knows that the addressee has killed the dog, and the question is a rebuke. A more detailed study would be necessary to better understand the pragmatics involved in negated polar questions.

\ea\label{ex:agu:19}
    \glll j̃ãw̃ã́ã íinu maáʃmakum? \\ 
          [jawaã  ii-nau]  maa-tʃa-ma=ka-umɨ \\
          dog  \textsc{1pl-poss}  kill.\textsc{pfv-neg-pst=q-2sg} \\
    \glt ‘Have you killed our dog?’ (lit. ‘haven’t you killed our dog?’) \citep[155]{Overall2017a}
\z

\subsection{Negation in stative predications}\label{sec:agu:2.3}

Non-verbal predication in Aguaruna may be encoded by means of an enclitic copula, a verbal copula, or juxtaposition of subject and predicate. Clauses expressing equation, proper inclusion and attribution generally make use of a partially verblike copula =\textit{aita} (with various reduced allomorphs), which is enclitic to a nominal or adjectival predicate and is inflected with finite verbal morphology. The adjective word class is fairly small \citep[130]{Overall2017a}, and many property concepts are encoded with intransitive lexical verbs. 

The distribution of the enclitic copula depends on grammatical factors including finiteness and tense. Where the enclitic cannot be used, a separate verbal copula \textit{a-} appears instead. In present tense declarative clauses, the enclitic copula may be omitted, resulting in simple juxtaposition of subject and predicate.

Neither the enclitic copula nor the copula verb is compatible with SN, instead making use of a suffix \nobreakdash-\textit{tʃau} (reduced to \textit{{}-tʃu} following a vowel) affixed directly to the non-verbal predicate. This suffix also nominalizes and negates verbal roots, and etymologically consists of the negative suffix \textit{{}-tʃa} and nominalizer \textit{\nobreakdash-u}, hence \citet{Overall2017a} labels it a ‘negative nominalizer’, and shows that synchronically it is a suffix in its own right \citep[513–514, 527]{Overall2017a}. Examples in this chapter follow \citet{Overall2017a} in glossing this suffix \textsc{neg.nmlz} ‘negative nominalizer’.

Example \REF{ex:agu:20} shows the enclitic copula combining with a negated nominal in the first clause, and a non-negated pronoun in the second clause. Example \REF{ex:agu:21} shows a negated nominal predicate combined with the copula verb, as the enclitic copula cannot appear in this type of subordinate clause. Example \REF{ex:agu:22} shows a negated nominal predicate juxtaposed to the subject in a present tense, declarative clause with third person singular subject.

\ea\label{ex:agu:20}
    \glll pákitʃauwaithi, íijaithi \\ 
          paki-tʃau=aita-hi  ii=aita-hi\\
          white.lipped.peccary-\textsc{neg.nmlz=cop-1pl.decl}  \textsc{1pl=cop-1pl.decl}  \\
    \glt ‘We’re not peccaries, we’re us (i.e. people).’ \citep[406]{Overall2017a}
\z           

\ea\label{ex:agu:21}
     \glll <Atumek judiuchu asajum…> \\ 
           atumɨ=ka  hudiu-tʃau  asa-humɨ \\
           2\textsc{pl=top}  Jew-\textsc{neg}  \textsc{cop.sub-2pl.ss} \\
    \glt ‘As you are not Jews…’ (\citealt{LigaBíblica2008}: 291; Romans 11:24)
\z           

\ea\label{ex:agu:22}
    \glll húka mídautʃu \\ 
          hu=ka  mi-nau-tʃau \\
          \textsc{dem.prox=top}  \textsc{1sg-poss-neg.nmlz} \\
    \glt ‘This is not mine.’ \citep[406]{Overall2017a}
\z
          
Although the copula verb is not compatible with SN, it does combine with the apprehensive marker in prohibitive clauses, as in \REF{ex:agu:23} (see \sectref{sec:agu:2.2.1} above for details of the prohibitive construction).

\ea\label{ex:agu:23}
    \glll kása áipa\\ 
          kasa  a-i-pa\\
          thief  \textsc{cop-appr-2.proh}\\
    \glt ‘Don’t be a thief!’ \citep[78]{Overall2017b}
\z           

Existential predications are formed with a verb \textit{a-}, which is morphologically distinct from the homophonous copula verb \citep{Overall2018,OverallForthcoming}. Negative existential is encoded with a verb \textit{atsu-}. The two forms are illustrated in \REF{ex:agu:24}).

\ea\label{ex:agu:24}
    \ea\label{ex:agu:24a}
        \glll ã\~{w}\~\i{} awai {Western Union} \\ 
              au=\~\i{}  a-wa-i  \textit{Western Union} \\
              \textsc{dem.dist=loc} exist.\textsc{ipfv}{}-3-\textsc{decl}  {Western Union} \\
        \glt ‘There’s a {Western Union} there.’
    \ex\label{ex:agu:24b}
        \glll h\~u\~{w}\~{i}  Nieva atsawai \\
              hu=\~\i{} \textit{Nieva}  atsa-wa-i \\
              \textsc{dem.prox=loc}  Nieva  exist\textsc{.neg.ipfv-3-decl}\\
        \glt ‘Here in Nieva there isn’t one.’ \citep[158]{Overall2017a}
    \z
\z     

It is tempting to analyse \textit{atsu-} as consisting of the existential verb \textit{a-} and SN suffix \textit{\nobreakdash-tsu} (\citealt{Peña2015} and \citealt{Kohlberger2020} adopt this analysis for cognate forms in Wampis and Shiwiar, respectively), but \citet[407]{Overall2017a} shows that this is unsatisfactory, as both valency-changing morphology in slot (B) and imperfective aspect in slot (C) follow the root \textit{atsu-} (as can be seen in example \REF{ex:agu:29} below), whereas both would be expected to intervene between the root and an SN suffix in slot (D) (see \sectref{sec:agu:1}, \figref{fig:agu:1}). The diachrony of the negative existential is discussed further in \sectref{sec:agu:4.6} below. 

Locative predication makes use of the lexical verb \textit{puhu-} ‘live’. This verb has the sense of “be alive” or “dwell in a place” in its lexical use, but encodes temporary location when used as a copula with a locational predicate, as in example \REF{ex:agu:25}. This verb takes the full range of verbal morphology, including SN. Example \REF{ex:agu:26} illustrates the negated form in a question (note that the use of negation here is pleonastic, cf. \sectref{sec:agu:2.2.2}). Example \REF{ex:agu:27} shows that in a declarative sentence the use of the expected negated form \textit{puha-tsu-wa-i} (live.\textsc{ipfv-neg-3-decl}) is dispreferred with a human subject, apparently due to ambiguity with the lexical meaning ‘he is not alive’. Instead, the negative existential verb \textit{atsu-} is used.

\ea\label{ex:agu:25}
    \glll <Juwi pujawai> \\ 
          hu=ĩ  puha-wa-i \\
          \textsc{dem.prox=loc}  live.\textsc{ipfv-3-decl}\\
    \glt ‘He is here.’ (\citealt{LigaBíblica2008}: 151; Luke 17:23)
\z 

\ea\label{ex:agu:26}
    \glll <¿Ju jega juwi Simón daagtin, tikich daaji Pedro tutaya nu pujatsuak?>\\ 
          hu  hɨɰã  hu=ĩ  \textit{Simón} naa-hĩ-tinu  tikitʃi naa-hĩ  \textit{Pedro}  tu-tajã  nu  puha-tsu-wa=ka\\
          \textsc{dem.prox}  house  \textsc{dem.prox=loc}  Simon  name-\textsc{pssd.3-attr}  other name-\textsc{pssd.3}  Peter  say-\textsc{norm}  \textsc{ana}  live.\textsc{ipfv-neg-3=q}\\
    \glt ‘Is Simon, also known as Peter, (not) in this house?’ (\citealt{LigaBíblica2008}: 232; Acts 10:18)
\z 

\ea\label{ex:agu:27}
     \glll <juwig atsawai>\\ 
           hu=ĩ=ka  atsa-wa-i \\
           \textsc{dem.prox=loc=top}  exist.\textsc{neg.ipfv-3-decl} \\
    \glt ‘He isn’t here.’ (\citealt{LigaBíblica2008}: 104; Mark 16:6)
\z 
          
Finally, possessive predication is based on the existential verb, with the possessum encoded as subject and possessor encoded as an object added by the applicative derivation that typically encodes beneficiary or maleficiary, shown in \REF{ex:agu:28}. This construction is negated using the applicativized negative existential, as in \REF{ex:agu:29}.

\ea\label{ex:agu:28}
    \glll <Siete pag, nuigtú wajumchik namak ajutjamui>\\ 
          \textit{siete}  paŋka  nuiŋtu  wahuma-tʃi=ki  namaka a-hu-tuhama-wa-i \\
          seven  bread  furthermore  few-\textsc{dim=restr}  fish   exist-\textsc{appl-1pl.obj.ipfv-3-decl} \\
    \glt ‘We have seven loaves of bread and just a few fish.’ (\citealt{LigaBíblica2008}: 39; Matthew 15:34)
\z 

\ea\label{ex:agu:29}
    \glll <minak tudauk atsugtawai> \\ 
          mi=na=ka  tudau=ka  atsu-hu-ta-wa-i \\
          \textsc{1sg=acc=top}  sin=\textsc{top}  exist.\textsc{neg-appl-1sg.obj.ipfv-3-decl} \\
    \glt ‘I have no sin.’ (\citealt{LigaBíblica2008}: 434; 1 John 1:8)
\z 

In summary, stative predications in Aguaruna use two strategies distinct from SN: nominal and adjectival predications mark negation on the non-verbal predicate using a derivational suffix, while existential and possessive predications make use of a distinct negative existential verb. Locative predication may use SN, but may also make use of the negative existential verb. 

\subsection{Negation in non-main clauses}\label{sec:agu:2.4}

\subsubsection{Preliminaries}
Aguaruna makes heavy use of non-finite clause types in clause-chaining constructions, especially in narrative texts. These clauses are morphosyntactically dependent in that they can only appear in a construction with an associated finite predicate: the verbs of dependent clauses are marked for some of the same categories as finite verbs, but lack tense and mood marking. Subject agreement in dependent verbs is marked with a slightly different paradigm from that of finite verbs, and the agreement markers follow suffixes that mark the verb’s dependent status and to some extent its semantic relation to the larger construction. Dependent verbs are also marked for switch-reference, indicating whether the subject of the dependent verb is the same as or different from that of a controlling verb (\citealt{HaimanMunro1983}).

Non-finite verbs forming temporal clauses are negated with the verbal negator \nobreakdash-\textit{tʃa}, and do not show the aspect-based alternation of \nobreakdash-\textit{tʃa} and \textit{\nobreakdash-tsu} observed in finite verbs, as shown in \sectref{sec:agu:2.4.1}. The same negator is involved in negating nominalizations, as described in \sectref{sec:agu:3.3}, apart from one construction that appears to combine \textit{\nobreakdash-tsu} with a nominalizer, which is described in \sectref{sec:agu:2.4.2}. Purpose clauses have a negative counterpart formed with the ‘apprehensive’ marker \nobreakdash-\textit{i}, described in \sectref{sec:agu:2.4.3}. The same marker forms the basis of the prohibitive forms described above (\sectref{sec:agu:2.2.1}). Finally, relativization and complementation are mostly achieved through the use of nominalized clauses, as described in \sectref{sec:agu:2.4.4}.

\subsubsection{Temporal subordinate clauses}\label{sec:agu:2.4.1}

The most frequently encountered subordinate clauses are sequential clauses, which are formed  on the perfective stem, and simultaneous clauses, formed on the imperfective stem plus the ‘simultaneous’ marker \nobreakdash-\textit{ku}, as in \REF{ex:agu:30}. Although the imperfective stem is associated with the \textit{\nobreakdash-tsu} negator in finite clauses, simultaneous subordinate clauses take the \textit{{}-tʃa} negator (\ref{ex:agu:31}--\ref{ex:agu:32}). I have not encountered any examples of negated subordinate clauses based on the perfective stem. The addition of the topic enclitic \textit{=ka} to any of these clause types forms a conditional clause (as in \REF{ex:agu:32}; also see \REF{ex:agu:49} below).

\ea\label{ex:agu:30}
    \glll tíkish wɨ́nakũĩ \\ 
          tikichi  wɨ-ina-ku-\~\i\\
          other  go-\textsc{pl.ipfv-sim.3-ds}\\
    \glt ‘When others were going…’ \citep[608]{Overall2017a}
\z 

\ea\label{ex:agu:31}
     \glll ɨ́kɨ akihúinaʃkũĩ \\ 
           ɨkɨ  aki-hu-ina-tʃa-ku-\~\i \\
          yet  pay-\textsc{1sg.obj-pl.ipfv-neg-sim.1/3-ds}\\
    \glt ‘They’re not paying me yet…’ \citep[408]{Overall2017a}
\z 

\ea\label{ex:agu:32}
    \glll núna waʃín duʃíkiaʃkunuk tukúmain awakáhai \\ 
          [nu=na  waʃi=na  duʃiki-a-tʃa-ku-nu=ka] tuku-mai-inu  awakɨ{}-ka-ha-i \\
          \textsc{ana=acc}  spider.monkey=\textsc{acc}  laugh.at-\textsc{ipfv-neg-sim-1sg.ss=top} shoot-\textsc{pot-nmlz}  overcome\textsc{{}-pfv-1sg-decl}\\
    \glt ‘If I hadn’t laughed at that monkey, I would have been able to shoot it.’ (agr041102\_07)
\z 

\subsubsection{`Without doing' clauses}\label{sec:agu:2.4.2}

Nominalized clauses with the \textit{{}-tsu} negator and final /k/ (\ref{ex:agu:33}--\ref{ex:agu:35}) are analysed by \citet[404]{Overall2017a} as a special construction meaning `without doing something' (cf. \citealt{PellizaroNáwech2005}: 83; \citealt{Saad2014}: 64, 96 on a similar construction in Shuar). This is the only construction in which the \textit{{}-tsu} negator combines with a nominalizer, as nominalizations are normally negated with \textit{{}-tʃa} (\sectref{sec:agu:3.3}). There is no positive counterpart to this construction.

\ea\label{ex:agu:33}
    \glll áuhtsuk papínak puhújahai \\ 
          [auhu-tsu-u=ka  papi=na=ka]  puhu-ja-ha-i\\
          study\textsc{{}-neg-nmlz=top} book\textsc{=acc=top}  live\textsc{{}-rem.pst-1sg-decl} \\
    \glt ‘I was not studying books.’ \citep[263]{Overall2017a}
\z 

\ea\label{ex:agu:34}
    \glll mína saíŋnakɨʃ úhatsuk… \\ 
          [mi=na  sai-hu=na=kɨ=ʃa uha-tsu-u=ka]\\
          \textsc{1sg}=\textsc{acc}  brother.in.law-\textsc{pssd.1sg=acc=restr=add} tell-\textsc{neg-nmlz=top}\\
    \glt  ‘Without telling even my brother-in-law…’ \citep[192]{Overall2017a}
\z 

\ea\label{ex:agu:35}
    \glll titú puhustá áanuk tútsuk \\ 
          titu  puhu-sa-ta  aanu=ka  tu-tsu-u=ka \\
          quietly  live-\textsc{pfv-imp}  \textsc{dem}.\textsc{med=top}  say-\textsc{neg-nmlz=top} \\
    \glt ‘Keep calm and don’t say that!’ (agr041005\_22)
\z 

\subsubsection{Purpose and apprehensive clauses}\label{sec:agu:2.4.3}

The negative counterpart of a purpose clause is formed with the apprehensive marker. The apprehensive form is marked with a suffix \textit{\nobreakdash-i}, as shown in \tabref{tab:agu:3} (note that third person singular is marked only with nasalization).

\begin{table}
\begin{tabular}{lll}
\lsptoprule
{person} & {singular} & {plural}\\
\midrule 
{1} & \textit{-i-ha} & \textit{-i-hi}\\
& {\scshape -appr-1sg} & {\scshape -appr-1pl}\\
{2} & \textit{-i-mɨ} & \textit{-i-humɨ}\\
& {\scshape -appr-2sg} & {\scshape -appr-2pl}\\
{3} & \textit{-ĩ} & \textit{-i-numɨ̃}\\
& {\scshape -appr.3sg} & {\scshape -appr-3pl}\\
\lspbottomrule
\end{tabular}
\caption{Apprehensive marking in Aguaruna \citep[73]{Overall2017b}}
\label{tab:agu:3}
\end{table}


\citet[73--74]{Overall2017b} notes that the apprehensive verb form is morphologically finite (subject is marked as for declarative verbs apart from the third person forms, and there is no switch-reference morphology), but is generally used in constructions that link its clause conceptually to another finite verb, making it appear functionally like a dependent clause. This is perhaps to be expected, given that it is semantically the negative counterpart of the dependent purpose clause type illustrated in \REF{ex:agu:36}. In particular, apprehensive verbs often appear in dependent speech report constructions (\REF{ex:agu:37}; the underlined part is the speech report); but \REF{ex:agu:38}, consisting of two juxtaposed finite clauses, shows that this is not necessarily the case.\footnote{Note that the purpose clause marker itself comes from a combination of the future marker \textit{{}-ta} with a dependent form of the speech verb: compare the third person purpose clause marker \textit{\nobreakdash-tatus} \REF{ex:agu:36} with the third person dependent speech verb \textit{tus} in \REF{ex:agu:37}.}

\ea\label{ex:agu:36}
    \glll wɨkaɨɰák wɨ́uwai kuntínun mantumaátatus \\ 
          wɨkaɨɰa-k\~u  wɨ-u=ai  [kuntinu=na mantu-ma-a-tatus]\\
          walk\textsc{.pfv-sim.3.ss}  go.\textsc{pfv-nmlz=cop.3.decl}  animal=\textsc{acc} kill.\textsc{appl-refl-pfv-intl.3.ss}\\
    \glt ‘He went walking to kill animals for himself.’ (i.e. ‘he went hunting’) (\citealt{Overall2017a}: 492–493)
\z 

\ea\label{ex:agu:37}
    \glll áimak ɨ́mamkɨmas intáhaiŋ tus \\ 
          aima-kũ  ɨmamkɨma-sã  [\underline{inta-ha-i-ha}  tus] \\
          fill.\textsc{ipfv}{}-\textsc{sim.3.ss}  be.careful-\textsc{sub.3.ss}  break-\textsc{pfv-appr-1sg}  say.\textsc{sub.3.ss}\\
    \glt ‘Filling them carefully, lest he should break them…’ (lit. ‘…saying “lest I break them”’) \citep[363]{Overall2017a}
\z 

 \ea\label{ex:agu:38}
    \glll wakítkita ámɨk, mantámawainum! \\ 
          wakita-ki-ta  amɨ=ka  mã-tama-aw-ai-numɨ \\
          go.back\textsc{{}-pfv-imp}  \textsc{2sg=top}\textsubscript{} kill-2\textsc{.obj-pfv-appr-3pl} \\
    \glt ‘You go back, lest they kill you!’ \citep[491]{Overall2017a}
\z 

The apprehensive forms are the basis of the prohibitive paradigm, the negative counterpart of the imperative, as described in \sectref{sec:agu:2.2.1} above.

\subsubsection{Relative and complement clauses}\label{sec:agu:2.4.4}
\largerpage
Relative clauses are formed with nominalizations, see for example the NP [\textit{hu=na saɨpɨ{}-utʃi-hĩ=na} [\textit{ija-u=na=kɨ=ʃa}]\textsubscript{RC}]\textsubscript{NP} (\textsc{dem.prox=acc} rind-\textsc{dim-3=acc} fall.\textsc{ipfv-nmlz=acc=restr=add)} \textsc{‘}this rind that is falling’ in example \REF{ex:agu:16}. Negative counterparts can be formed using the negative nominalizer \nobreakdash-\textit{tʃau}, for example [\textit{n\~\i} [\textit{waina-tʃau=ka}]\textsubscript{RC}]\textsubscript{NP} (\textsc{3sg} see-\textsc{neg.nmlz=top}) ‘one who doesn’t know’ in example \REF{ex:agu:48} below (and see \sectref{sec:agu:3.3} for details of the negative nominalization).

Most complement clauses are formed with nominalizations (a complementation strategy, in \citegen{Dixon2006} terminology), but the verb \textit{dɨka-} can take a juxtaposed finite clause as complement. There do not appear to be any grammatical restrictions on negating the predicate either of the complement or the main clause. Example \REF{ex:agu:39} includes a negated predicate of the complement clause (‘they did not enter’) while the main predicate (‘we know’) is positive. Compare this with example \REF{ex:agu:9} above, where the same verb is negated (‘(they) didn’t know [how to make her give birth]’), and note that when the main clause predicate is negated, the complement clause takes the form of a nominalization.

\ea\label{ex:agu:39}
    \glll <Yamaik shiig dekainaji, ditak Apajuin dekaskeapi tuinachu asag, Apajuí ditan nugká anagkuauwa nuwig wayachajui.> \\ 
          yamai=ka  ʃiiha  dɨka-ina-hi  [dita=ka  apahui=na dɨkaskɨ-api  tu-ina-tʃau  asa-ahã  apahui  dita=na nuŋka  anaŋku-a-u=a  nu=ĩ=ka wajã-tʃa-aha-u=i] \\
          now=\textsc{top}  well  know-\textsc{pl.ipfv-1pl.decl}  \textsc{3pl=top}  God=\textsc{acc} true-\textsc{tag}  say-\textsc{pl.ipfv-neg.nmlz}  \textsc{cop.sub-pl.3.ss}  God  \textsc{3pl=acc} land.\textsc{acc}  promise-\textsc{pfv-nmlz=cop.3}  \textsc{ana=loc=top} enter.\textsc{pfv}{}-\textsc{neg-pl-nmlz=cop.3.decl} \\
    \glt ‘Now we know well, that because they don’t believe in God (lit. ‘they don’t say of God “he must be true”’), they did not enter into the land that God promised them.’ (\citealt{LigaBíblica2008}: 398; Hebrews 3:19)
\z 
  
The verb \textit{dakitu}{}- ‘refuse’, illustrated in \REF{ex:agu:40}, is an inherently negative counterpart to \textit{wakɨɰa-} ‘want’, and is discussed further in \sectref{sec:agu:2.5}.

\ea\label{ex:agu:40}
    \glll dakítau apahı̃́hãĩ wɨtán\\ 
          dakita-a-u  [apa-h\~\i=ha\~\i{}  wɨ-ta=na] \\
          refuse-\textsc{ipfv-nmlz}  father-\textsc{pssd.3=com}  go-\textsc{act.nmlz=acc} \\
    \glt ‘He refused to go with his father.’ \citep[530]{Overall2017a}
\z 

\subsection{Negative lexicalizations}\label{sec:agu:2.5}

The negative existential verb \textit{atsu-} is clearly related to the existential \textit{a-} and the negative suffix \nobreakdash-\textit{tsu}, but is best analysed synchronically as a morphologically simple verb root, as shown above (\sectref{sec:agu:2.3}), and discussed further in \sectref{sec:agu:4.6} below. A few other verb roots are semantically negative, at least when translated into Spanish or English, for example \textit{dakitu-} ‘not want’; \textit{tuhintu-} ‘be unable’;  \textit{kanina-} ‘not fit’. Note, however, that these verbs do not preclude negated forms of the corresponding positive verbs. For example, \REF{ex:agu:41} shows that \textit{dakitu-} ‘refuse, not want’ is not formally the negation of \textit{wakɨɰa-} ‘want’, since this verb can be negated with SN suffixes.\footnote{Note that the context is a polar question, where the negative marker could be considered pleonastic – see \sectref{sec:agu:2.2.2}.} In other words, the meaning of \textit{dakitu-} is not simply the negation of \textit{wakɨɰa-}, but must add some other sense too.

\ea\label{ex:agu:41}
    \glll amɨ́ʃ Lima wɨ́tasamɨʃ wakɨ́ɰasmɨk? \\ 
          amɨ=ʃa  [Lima  wɨ-tasa-mɨ=ʃa]  wakɨɰa-a-tsu-mɨ=ka \\
          \textsc{2sg}=\textsc{qtop}  Lima  go.\textsc{pfv-intl-2=qtop}  want\textsc{{}-ipfv-neg-2sg=q} \\
    \glt ‘Do you want to go to Lima?’ (lit. ‘Don’t you want to go to Lima?’) \citep[480]{Overall2017a}
\z 

Two particles function as pro-clauses and are semantically negative: \textit{wããs} ‘I don’t know why’ introduces rhetorical questions \REF{ex:agu:42}; and an answer word \textit{atʃá} ‘I don’t know’ \citep[409]{Overall2017a}.

\ea\label{ex:agu:42}
    \glll w̃ã́ãs dúʃa núnikabia muúntaʃ\\ 
          wããs  nu=ʃa  nuni-ka-amaja  muunta=ʃa \\
          I.don’t.know.why  \textsc{ana=qtop}  do\textsc{{}-pfv-dist.pst.3} adult\textsc{=qtop} \\
    \glt ‘I don’t know why the man did that.’ \citep[401]{Overall2017a}
\z 

\section{Non-clausal negation}\label{sec:agu:3}

Aguaruna has a pro-sentence negative particle \textit{atsa} ‘no’, used for answering questions (\sectref{sec:agu:3.1}). There are no dedicated indefinite expressions, and strategies for expressing negative indefinites are discussed in \sectref{sec:agu:3.2}. Negative derivation and negation of nominalizations are described in \sectref{sec:agu:3.3}.

\subsection{Negative replies}\label{sec:agu:3.1}

The word for ‘no’ is \textit{atsá}, clearly related to the negative existential \textit{atsu-}. The corresponding affirmative particle is \textit{ɨ̃ʔɨ̃́ \textasciitilde{} hɨ̃ʔɨ̃́ \textasciitilde{} hɨ̃ʔá } ‘yes’. Both particles are frequently accompanied by repetition of the predicate, and both relate to the polarity of the predicate in the answer, regardless of that in the question. That is, there is no grammatical or lexical indication of whether or not the answer contradicts the question. A third particle \textit{atʃá} ‘I don’t know’, mentioned in \sectref{sec:agu:2.5} above, is not normally accompanied by a repetition of the questioned predicate.

Recorded Aguaruna data is largely narrative texts, and there are few examples of question and answer exchanges. The examples below are taken from a Bible translation, and illustrate the four possible combinations of polarity in the question and answer.

Example \REF{ex:agu:43} shows the affirmative particle contradicting the negation in the polar question. Compare with \REF{ex:agu:44}, in which a positive question is answered in the affirmative.
% \hc{@LSP: there is a long space before the translation of \ref{ex:agu:43}, could you fix this?}
\ea\label{ex:agu:43}
    \glll <¿Amina jintinjamnuk akikmakchattawak jega Apajuí ememattainum kuichik itaata tibauwa nunak? tuidau. Tama ni ayaak: Ehé akikmaktatui>\\
          ~~~~amina  hintintu-hama-inu=ka  akikma-ka-tʃa-tata-wa=ka  [hɨɰã  apahui ɨmɨma-tu-taĩ=numa  kuitʃiki  i-ta-a-ta ti-mau=a  nu=na=ka]  tu-ina-u  ta-ma nĩ  aya-a-kũ  ɨʔɨ  akikma-ka-tata-wa-i\\
          ~~~~\textsc{2sg}.\textsc{acc}  teach-\textsc{2.obj-nmlz=top}  pay-\textsc{pfv-neg-fut-3-q}  house  God feel.pride-\textsc{appl-nmlz=loc}  money  \textsc{caus-}arrive-\textsc{pfv-imp} say.\textsc{pfv-nmlz=cop.3}  \textsc{ana=acc=top}  say-\textsc{pl.ipfv-nmlz}  say.\textsc{ipfv-nsbj/sbj.coref} \textsc{3sg}  reply-\textsc{ipfv-sim.3.ss}  yes  pay-\textsc{pfv-fut-3-decl}\\
    \glt ‘They said “won’t your teacher pay the temple tax?”, when they said that to him he replied “yes, he will pay it”.’ (Note the translation of \textit{temple tax} is literally ‘that of which is said “bring money to the house where one praises God”’.) (\citealt{LigaBíblica2008}: 43; Matthew 17:24–25)
\z 

\ea\label{ex:agu:44}
    \glll <¿Antamek jujú imataina dusha? Tama Jisus ayaak: Ehé, antajai.> \\ 
          ~~~~anta-mɨ=ka  huhu  imata-ina  nu=ʃa ta-ma  Jisus  aya-a-kũ  ɨʔɨ  anta-ha-i \\
          ~~~~hear.\textsc{ipfv-2sg=q}  \textsc{dem.prox}  shout-\textsc{pl.ipfv}  \textsc{ana=top} say.\textsc{ipfv-nsbj/sbj.coref}  Jesus  reply-\textsc{ipfv-sim.3.ss}  yes  hear.\textsc{ipfv-1sg-decl} \\
    \glt ‘(The priests and teachers asked) do you hear what these (children) are shouting? When they said that, Jesus replied “yes, I hear it”’. (\citealt{LigaBíblica2008}: 49–50; Matthew 21:16)
\z 

Turning to negative replies, in \REF{ex:agu:45} a question with positive polarity is answered in the negative, and in \REF{ex:agu:46} a question with negative polarity is answered with negative polarity.

\ea\label{ex:agu:45}
    \glll <¿Apajuí etsegtin taatna dukaitam? Nu tamashkam: Atsa, duchauwaitjai, tau.> \\ 
           ~~~~apahui  ɨtsɨhu-tu-inu  ta-a-tin=a  nu=ka=ita-mɨ  nu  ta-ma=ʃakama  atsa  nu-tʃau=aita-ha-i ta-u \\
           ~~~~God  preach-\textsc{appl-nmlz}  arrive-\textsc{pfv-fut.nmlz=cop.3}  \textsc{ana=q=cop-2sg} \textsc{ana}  say.\textsc{ipfv-nsbj/sbj.coref=add}  no  \textsc{ana-neg.nmlz=cop-1sg-decl} say.\textsc{ipfv-nmlz}\\
    \glt ‘Are you the one who will arrive to preach on God’s behalf? He, when they said that to him, said “no, I’m not that one.”’ (\citealt{LigaBíblica2008}: 170; John 1:21)
\z 

\ea\label{ex:agu:46}
    \glll <¿Amesh auna aentsun achikaja nuna jintintaijichukaitam? tusa iniau. Tama Pedro ayaak: Atsa, wika auna jintintaijichuitjai, tau.> \\ 
          ~~~~amɨ=ʃa  au=na  aɨntsu=na  atʃi-ka-aha  nu=na hintintu-taĩ-hĩ=tʃau=ka=ita-mɨ  tu-sã  ini-a-u ta-ma  Pedro  aja-a-kũ atsa  wi=ka  au=na  hintintu-taĩ-hĩ-tʃau=ita-ha-i ta-u \\
          ~~~~2\textsc{sg=top}  \textsc{dem.dist=acc}  person=\textsc{acc}  grab-\textsc{pfv-pl}  \textsc{ana=acc} teach-\textsc{nmlz-pssd.3=neg.nmlz=q=cop-2sg}  say-\textsc{sub.3.ss}  ask-\textsc{ipfv-nmlz} say.\textsc{ipfv-nsbj/sbj.coref}  Peter  reply-\textsc{ipfv-sim.3.ss} no  \textsc{1sg=top}  \textsc{dem.dist=acc}  teach-\textsc{nmlz-pssd.3-neg.nmlz=cop-1sg-decl} say.\textsc{ipfv-nmlz}\\
   \glt ‘“Are you not a disciple of that person they’re holding?” she asked, when she said that Peter replied “no, I’m not his disciple”.’ (\citealt{LigaBíblica2008}: 207; John 18:17) 
\z 

In all of the above examples, answers to polar questions consist of the particle and repetition of the predicate. An answer may also consist of only the relevant particle, or only the predicate. \tabref{tab:agu:4} summarizes the combinations of polarity in the question-answer pairs illustrated above.

\begin{table}
\begin{tabular}{llll}
\lsptoprule
{example} & {polarity of Q} & {polarity of A} & {particle used}\\
\midrule 
{\REF{ex:agu:43}} & {Affirmative} & {Affirmative} & {\textit{ɨʔ\'{ɨ} } ‘yes’}\\
{\REF{ex:agu:44}} & {Negative} & {Affirmative} & {\textit{ɨʔ\'{ɨ} } ‘yes’}\\
{\REF{ex:agu:45}} & {Affirmative} & {Negative} & {\textit{atsá} ‘no’}\\
{\REF{ex:agu:46}} & {Negative} & {Negative} & {\textit{atsá} ‘no’}\\
\lspbottomrule
\end{tabular}
\caption{Summary of question and answer pairs in examples}
\label{tab:agu:4}
\end{table}
%\lv{To Sebastian and Héloïse: table \ref{tab:agu:4} is misplaced in the pdf file as it appears in the section on negative indefinites. Would it be possible move it to its own section?}
\subsection{Negative indefinites and quantifiers}\label{sec:agu:3.2}

Aguaruna lacks indefinite expressions, both positive (\textit{something, somebody}) and negative (\textit{nothing, nobody}). In contexts where positive indefinites appear in translations, nominalizations (often “headless relatives”, see \citealt{Overall2017a}: 546) are used, whether in positive \REF{ex:agu:47} or negative \REF{ex:agu:48} polarity contexts. Note that example \REF{ex:agu:48} comes from a discussion about the origin of a particular place name. The verb \textit{waina}{}- ‘see’ is used with the sense of ‘know’ when referring to knowing a person or place.

\ea\label{ex:agu:47}
    \glll kánu hapikbáuwa íman tahák akaikíuwai \\ 
          [kanu  hapi-ki-mau=a  iman] taha-k\~u  akai-ki-u=ai \\
          canoe  pull-\textsc{pfv-nsbj.nmlz=cop.3}  \textsc{ints.nmlz} clear.\textsc{ipfv-sim.3.ss}  go.down\textsc{{}-pfv-nmlz=cop.3.decl}\\
    \glt ‘(Something) as big as a canoe being dragged had come down clearing a path through the undergrowth.’ \citep[551]{Overall2017a}
\z 

\ea\label{ex:agu:48}
    \glll túhã nı̃́ĩ wáintʃauk waináinatsuastai \\ 
          tuhã  [n\~\i{}  waina-tʃau=ka]  waina-ina-tsu-u=a-tsa-tai \\
          but  \textsc{3sg}  see-\textsc{neg.nmlz=top}  see-\textsc{pl.ipfv-neg-nmlz=cop.3-specl-specl} \\
    \glt ‘But one who doesn’t know (the place) will probably never know.’ \citep[378]{Overall2017a}
\z 

Example \REF{ex:agu:49} is from a text about hunting techniques; the speaker tells how people build hides in places where game animals come to eat fruits, and kill the animals when they arrive. An implicit object (an animal) is simply unmarked in this example, which could just as well be translated ‘if we don’t kill anything'.

\ea\label{ex:agu:49}
    \glll núniaʃkuik, núu dútika máaʃkuik, káʃiʃ wɨtájamɨ̃ \\ 
          nuni-a-tʃa-ku-i=ka  [nu  dutika] [ma-a-tʃa-ku-i=ka]  kaʃi=ʃa  wɨ-tajamɨ \\
          do\textsc{{}-ipfv-neg-sim-1pl.ss=top}  \textsc{ana} do.\textsc{pfv.1pl.ss} kill\textsc{{}-ipfv-neg-sim-1pl.ss=top}  night=\textsc{add}  go-\textsc{norm} \\
    \glt ‘If we don’t do that, if having done all that we don’t kill (an animal), we even go at night.’ \citep[619]{Overall2017a}
\z 

A strategy consisting of the numeral ‘one’ plus the restrictive and additive enclitics is used to encode the sense of ‘nobody at all’ or ‘nothing at all’ (i.e. ‘not even one’) in \REF{ex:agu:50}; and example \REF{ex:agu:51} shows a similar construction based on the quantifier \textit{matʃik} ‘a little bit’ with the positive reading ‘anything at all’.

\ea\label{ex:agu:50}
    \glll <tujash wika makichik aentsnakesh etegkeatsjai.>\\ 
          tuhã=ʃa  wi=ka  makitʃiki  aɨntsu=na=kɨ=ʃa ɨtɨŋkɨ{}-a-tsu-ha-i \\
          but=\textsc{add}  \textsc{1sg=top}  one  person=\textsc{acc=restr=add} judge-\textsc{ipfv-neg-1sg-decl} \\
    \glt ‘But I pass judgement on no-one.’ (\citealt{LigaBíblica2008}: 186; John 8:15)
\z 

\ea\label{ex:agu:51}
    \glll <Ame machikiuchikesh dutikmainaitkumek> \\ 
          amɨ  matʃiki-utʃi=kɨ=ʃa  dutika-mai-inu=aita-ku-mɨ=ka \\
          \textsc{2sg}  bit-\textsc{dim-restr=add}  do-\textsc{pot-nmlz=cop-sim-2sg=top}\\ 
    \glt ‘If you can do anything at all…’ (\citealt{LigaBíblica2008}: 87; Mark 9:22)
\z 

\subsection{Negative derivation and nominalization}\label{sec:agu:3.3}

Nominalizations are widely used in Aguaruna, forming relative and complement clauses (see \sectref{sec:agu:2.4.4}) and also functioning in lieu of finite verbs in some constructions \citep{Overall2014}. The nominalizing suffixes are shown in \tabref{tab:agu:5}.

\begin{table}
\begin{tabularx}{\textwidth}{llQ}
\lsptoprule
{suffix} & {referent} & {stem(s) selected}\\
\midrule
\textit{-ta} & {action} & {unmarked}\\
\textit{-inu} & {subject participant} & {unmarked or potential}\\
\textit{-ta\~\i} & {non-subject participant} & {unmarked}\\
\textit{-u} & {subject participant} & {perfective or imperfective}\\
\textit{-mau} & {non-subject participant or action} & {perfective or imperfective}\\
\textit{-tinu} & {participant + future} & {perfective}\\
\textit{-tʃau} & {participant + negation} & {perfective, imperfective or unmarked}\\
\textit{-haku} & {subject participant + remote past} & {unmarked}\\
\lspbottomrule
\end{tabularx}
\caption{Nominalizing suffixes in Aguaruna \citep[504]{Overall2017a}}
\label{tab:agu:5}
\end{table}

Some nominalizers can combine with the SN negator \textit{{}-tʃa} \REF{ex:agu:52}, others have their negative counterpart in the nominalizer \textit{{}-tʃau} \REF{ex:agu:53}.

\ea\label{ex:agu:52}
    \ea\label{ex:agu:52a}
        \glll jutã́ĩ \\
              ju-taĩ \\ 
              eat-\textsc{nmlz} \\
        \glt ‘food’ (‘what is eaten’) \citep[66]{Overall2017a}
    \ex\label{ex:agu:52b}
       \glll jútʃatãĩ \\
             ju-tʃa-taĩ \\
             eat-\textsc{neg-nmlz} \\
        \glt ‘inedible’ (‘what is not eaten’) (SLCK)
    \z
\z 

\ea\label{ex:agu:53}
    \ea\label{ex:agu:53a}
        \glll <nemajin> \\ 
              nɨma-hu-inu \\
              follow-\textsc{appl-nmlz} \\
        \glt ‘follower’ (\citealt{LigaBíblica2008}: 308)
    \ex\label{ex:agu:53b}
        \glll <nemagchau> \\
              nɨma-hu-tʃau \\
              follow-\textsc{appl-neg.nmlz} \\
        \glt ‘non-follower’ (\citealt{LigaBíblica2008}: 308)
    \z
\z 

The negative nominalizer \textit{{}-tʃau} is etymologically composed of the negator \textit{\nobreakdash-tʃa} and nominalizer \textit{{}-u}, but \citet[513--514]{Overall2017a} presents arguments for treating it as synchronically non-compositional. In addition to nominalizing and negating verb roots, \textit{{}-tʃau} also negates nominals, including nouns, adjectives and already nominalized verbs. In the case of adjectives, the negator has a derivational function in forming antonyms e.g. \textit{pɨŋkɨŋ-tʃau} (good-\textsc{neg.nmlz}) ‘bad’; and see \sectref{sec:agu:2.3} for its role in negating stative predications.

The negative nominalizer also functions to negate the derivational suffix \textit{{}-tin} ‘attributive’ \REF{ex:agu:54a} giving a privative sense \REF{ex:agu:54b}.

\ea\label{ex:agu:54}
    \ea\label{ex:agu:54a}
        \glll núwɨntin \\ 
              nuwɨ-tinu \\
              woman.\textsc{pssd.3-attr} \\
        \glt ‘married man’ (i.e. ‘possessed of a wife’) \citep[88]{UwaraiYagkugRegan1998}
    \ex\label{ex:agu:54b}
        \glll núwɨntʃau \\
              nuwɨ\nobreakdash-tu\nobreakdash-tʃau \\
              woman.\textsc{pssd.3-attr-neg.nmlz} \\
        \glt ‘bachelor’ \citep[88]{UwaraiYagkugRegan1998}
    \z
\z 

\section{Other aspects of negation}\label{sec:agu:4}
\subsection{The scope of negation}\label{sec:agu:4.1}

Verbal negation has clausal scope. In its derivational function, the nominal negator \textit{{}-tʃau} has scope over the negated nominal: note in \REF{ex:agu:55} the translation implies ‘they do [non-good]' rather than ‘they don’t [do good]'.

\ea\label{ex:agu:55}
    \glll <tuja aents pegkegchaushkam pegkegchaunak takaawai.> \\ 
          ~~tuha  aɨntsu  pɨŋkɨha-tʃau=ʃakama  pɨŋkɨha-tʃau=na=ka  taka-a-wa-i\\
          ~~but  person  good-\textsc{neg=add}  good-\textsc{neg=acc=top}  work-\textsc{ipfv-3-decl}\\
    \glt ‘But bad people do bad things.’ (\citealt{LigaBíblica2008}: 19; Matthew 7:16)
\z 

When the nominal negation strategy is used to negate a copula clause, however, it appears to scope over the whole clause including any associated dependent clauses. Note the translation of \REF{ex:agu:56}: ‘I am not [a dog who will be beaten]', as opposed to ‘I am [a non-dog] who will be beaten.'

\ea\label{ex:agu:56}
    \glll wíka j̃ãw̃ã́ãtʃuithai, ihuntáŋ suwimám átinun \\ 
          wi=ka  jawaã-tʃau=ita-ha-i  ih\~u-tu-hã suwima-ma  a-tinu-nu\\
          \textsc{1sg}=\textsc{top}  dog-\textsc{neg.nmlz=cop-1sg-decl}  gang.up-\textsc{appl-pfv.3.ss} beat.\textsc{pfv-nsbj/sbj.coref}  \textsc{cop-fut.nmlz-1sg.ss} \\
    \glt ‘I am not a dog, to be beaten by gangs of people.’ \citep[406]{Overall2017a}
\z 

\subsection{Negative polarity}\label{sec:agu:4.2}

An adverb \textit{ɨkɨ} ‘yet’ is favoured in negative contexts, as in \REF{ex:agu:57}, although it does appear in positive contexts too (as in \REF{ex:agu:58}). In conversation, however, \textit{ɨkɨ} can be used alone with the sense ‘not yet’ (for example, answering a question about whether something has been done).

\ea\label{ex:agu:57}
    \glll ɨ́kɨ kánashai \\ 
          ɨkɨ  kana-a-tsu-ha-i \\
           yet  sleep-\textsc{ipfv-neg-1sg-decl}\\
    \glt ‘I am not sleeping yet.’ \citep[409]{Overall2017a}
\z 

\ea\label{ex:agu:58}
    \glll <Tusa Jisus eke chichaak wajai> \\ 
          ~~tu-sã  Jisus  ɨkɨ  tʃitʃaa-kũ  waha-ĩ \\
          ~~say-\textsc{sub.3.ss}  Jesus  yet  speak.\textsc{ipfv-sim.3.ss}  stand.\textsc{ipfv-ds} \\
    \glt ‘As he said that, while Jesus was still speaking…’ (\citealt{LigaBíblica2008}: 63; Matthew 26:47)
\z 

\subsection{Marking of NPs in the scope of negation.}\label{sec:agu:4.3}

There is no obligatory marking of NPs in the scope of negation, but \citet[584--585]{Overall2017a} notes that negated clauses favour the appearance of the topic enclitic \textit{=ka} with one or more constituents of the clause (compare \REF{ex:agu:25} and \REF{ex:agu:27} above, among others). Outside of negative contexts, the topic enclitic is favoured when (re)introducing characters in narrative; for the subjects of verbless and copula clauses; marking contrast; strong assertion; and when adding editorial commentary in narratives \citep[581]{Overall2017a}.

\subsection{Reinforcing negation}\label{sec:agu:4.4}

The ‘not even one’ construction described in \sectref{sec:agu:3.2}, example \REF{ex:agu:50}, gives an intensified negation. I am not aware of any other strategy for reinforcing negation specifically, outside of the general means for expressing speaker’s attitude in clauses.

\subsection{Negation, coordination and complex clauses}\label{sec:agu:4.5}

Aguaruna has a highly hypotactic syntax, and there is no formally marked paratactic coordination. Negation of subordinate clauses, including various temporal clause types, is described in \sectref{sec:agu:2.4}, and negation of nominalizations (which may form relative or complement clauses) is described in \sectref{sec:agu:3.3}. The apprehensive clause type, which translates as ‘lest’, is not a subordinate clause syntactically but frequently appears in a speech report construction that links it to another clause – this is described in \sectref{sec:agu:2.4.3}.

\subsection{Miscellaneous aspects of negation}\label{sec:agu:4.6}
\largerpage
The bulk of the grammar is identical for all of the Chicham languages, which are very closely related, but some differences are valuable indicators of the progression of grammatical change. These include different paradigms of tense markers and different arrangements of the imperative/prohibitive paradigms.

The negative existential is formed with a verb \textit{atsu-} which cannot be analysed as morphologically compositional, as shown above (\sectref{sec:agu:2.3}). Although this form could reflect an etymological fusion of the positive existential and the negator (cf. \citealt{Veselinova2013neg.exist}: 136–138), a more attractive analysis is that the negative existential verb itself is the diachronic source of the negative suffix \textit{{}-tsu} through renalysis of an auxiliation construction. As noted above (\sectref{sec:agu:1}), there is evidence for some verbal tense markers having arisen from phonological fusion of non-finite stems with a following finite auxiliary. This analysis would explain the aspect-based distribution of the two SN negators, by assuming that the \textit{{}-tsu} suffix entered the system via some type of progressive construction, with the negative existential as auxiliary (cf. \citealt{Croft1991}). The distinct strategy for negating perfective stems with the suffix \textit{{}-tʃa} would then be older, and this is consistent with the fact that this marker is also the default in non-finite verb forms. In the typology developed by \citet{Croft1991} and refined by \citet{Veselinova2016} this is Type C: “the negative existential is used as SN but the constructions where it is used differ for the negation of verbs and for the negation of existential predications” \citep[144]{Veselinova2016}. In Aguaruna the difference is morphological, as the separate negative existential verb \textit{atsu-} has fused with the stem to become the suffix \textit{{}-tsu}, displacing the earlier negator -\textit{tʃa} only in imperfective finite verbs.

\section{Summary}\label{sec:agu:5}

In Aguaruna, predicates and nominals can be negated. Verbal predicates make use of two inflectional suffixes, distributed according to aspect and finiteness of the verb, while nominal predicates must use the derivational negator. Negation marked in the predicate has clausal scope, and this is true both of verbal and nominal predicates. A distinct marker \textit{{}-i} appears in some non-declarative clause types. The verbal paradigms are largely symmetrical, and the only category that is completely neutralized under negation is the plain/familiar imperative distinction (\sectref{sec:agu:2.2.1}).

There are no dedicated negative indefinites; negative transport or negative raising is not in evidence; and there is no evidence of negation used in non-negative contexts, apart from the possibility of pleonastic negation in polar questions (see \sectref{sec:agu:2.2.2} above).

\tabref{tab:agu:6} gives an overview of the negative markers of Aguaruna, the word class(es) that can host each marker, and their function(s).

\begin{table}
\small
    \begin{tabularx}{\textwidth}{llXl}
\lsptoprule 
marker & word class  & function(s) & section\\
& of host &  \\
\midrule 
-\textit{tsu} & V 
&Negation of finite clause types, associated with imperfective aspect and present tense & \sectref{sec:agu:2.1}  \\
&&Marks ‘without doing’ construction & \sectref{sec:agu:2.4.2}  \\

-\textit{tʃa} & V 
&Negation of finite clause types, associated with perfective aspect and non-present tenses & \sectref{sec:agu:2.1}  \\
&&Negation of subordinate clause types & \sectref{sec:agu:2.4}  \\
&&Negation of hortative clause type & \sectref{sec:agu:2.2.1}  \\

-\textit{i} & V
&Formation of apprehensive clauses (negative equivalent of purpose clause type) & \sectref{sec:agu:2.4.3}  \\
&&Negation of imperative clause types, except hortative & \sectref{sec:agu:2.2.1}  \\

-\textit{tʃau} & V, N, Adj 
&Derivation of negated nominal from verb & \sectref{sec:agu:3.3}  \\
&& Derivation of negated form of adjective or noun & \sectref{sec:agu:3.3}  \\
&& Clausal negation with nonverbal predicate & \sectref{sec:agu:2.3}  \\

\textit{atsu}- &– 
&Negative existential verb & \sectref{sec:agu:2.3}  \\

\textit{wããs}  &– 
&Negative pro-clause ‘I don’t know why’ & \sectref{sec:agu:2.5}  \\

\textit{atʃa}  &– 
&Negative pro-sentence, in response to question ‘I don’t know’ & \sectref{sec:agu:2.5}  \\

\textit{atsa}  &– 
&Negative pro-sentence, in response to polar question & \sectref{sec:agu:3.1} \\
\lspbottomrule
    \end{tabularx}
    \caption{Properties of negative markers}
    \label{tab:agu:6}
\end{table}
    

Future research will undoubtedly shed light on the subtleties of negation in Aguaruna, and also contribute to a better understanding of the diachronic development of the constructions described here.

\section*{Acknowledgments}
\largerpage[2]
I gratefully acknowledge Segundo Lucio Cungumas Kujancham’s contributions to this paper, not only in providing examples but also through his insightful discussions of semantic distinctions. I am also grateful to the editors of the volume and an anonymous reviewer for very helpful comments on earlier drafts. Needless to say, I take full responsibility for the analysis presented here, and for any errors that may remain.

\section*{Abbreviations}

\begin{tabularx}{.45\textwidth}{lQ}
1 & first person \\
2 & second person \\
3 & third person \\
\scshape a & most agent-like argument of a transitive clause \\
\scshape abl & ablative \\
\scshape acc & accusative \\
\scshape acn & action  \\
\scshape add & additive \\
\scshape ana & anaphoric pronoun \\
\scshape appl & applicative \\
\scshape appr & apprehensive \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape attr & attributive \\
\scshape caus & causative \\
\scshape cop & copula \\
\scshape decl & declarative \\
\scshape dem & demonstrative \\
\scshape dim & diminutive \\
\scshape dist & distal \\
\scshape ds & different subject \\
\scshape exclam & exclamative \\
\scshape fut & future \\
\scshape hort & hortative \\
\scshape imm & immediate \\
\scshape imp & imperative \\
\end{tabularx}


\begin{tabularx}{.45\textwidth}{lQ}
\scshape ints & intensifier \\
\scshape intl & intentional \\
\scshape ipfv & imperfective \\
\scshape juss & jussive \\
\scshape loc & locative \\
\scshape med & medial \\
\scshape narr & narrative \\
\scshape neg & negative \\
\scshape nmlz & nominalizer \\
\scshape norm & normative \\
\scshape NP & noun phrase \\
\scshape nsbj & non-subject \\
\scshape nsbj/sbj.coref & non-subject in marked clause is subject of controlling clause \\
\scshape obj & object \\
\scshape p & most patient-like argument of a transitive clause \\
\scshape pfv & perfective \\
\scshape pl & plural \\
\end{tabularx}%
\begin{tabularx}{.54\textwidth}{lQ}%larger than normal
\scshape pot & potential \\
\scshape proh & prohibitive \\
\scshape prox & proximal \\
\scshape poss & possessive \\
\scshape pssd & possessed \\
\scshape pst & past \\
\scshape q & polar question marker \\
\scshape qtop & topic marker in interrogative clause \\
\scshape RC & relative clause \\
\scshape refl & reflexive \\
\scshape rem & remote  \\
\scshape restr & restrictive \\
\scshape s & single argument of an intransitive clause \\
\scshape sub & subordinate \\
\scshape sg & singular \\
\scshape sim & simultaneous \\
\scshape sn & standard negation \\
\scshape specl & speculative \\
\scshape ss & same subject \\
\scshape tag & tag question marker \\
\scshape top & topic \\
\scshape v & verb \\
\\
\end{tabularx}

\largerpage
\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
\end{document} 
