\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233972}
\author{Nailya Philippova\affiliation{}}
\title{Negation in Amri Karbi}
\abstract{This paper provides a description of the negation system in Amri Karbi. The description is based mainly on data collected by the author during fieldwork. The outline of the description follows the structure of the \citetitle{chapters/questionnaire} \parencitetv{chapters/questionnaire}. The paper starts with a short introduction, then it proceeds with description of standard negation in Amri Karbi that is marked with the negative suffix -C\textit{e}. Following this is a discussion of non-clausal negation and the final section of the paper provides description of other aspects of negation, including negative polarity and negation in complex clauses among other topics.

\keywords{Amri Karbi, negation, Tibeto-Burman, duplifix}
}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 

\section{The language}\label{sec:amr:1}

Amri Karbi (ISO: \href{https://iso639-3.sil.org/code/ajz}{ajz}, Glottocode: \href{https://glottolog.org/resource/languoid/id/amri1238}{amri1238}) is a Tibeto-Burman language spoken in the Kamrup district of Assam and the Ri-Bhoi district of Meghalaya in North East India. 
The Amri Karbis, along with a closely related group Karbis, are an indigenous tribe of North East India. They are mentioned as the Mikir in the Constitution Order of the Government of India. The Amri Karbis and the Karbis are closely related linguistically and ethnically. The Karbis live mainly in the Karbi Anglong district of Assam. 

A full description of Amri Karbi grammar is provided in \citet{Philippova2021}. The latest description of the closest related language Karbi can be found in \citet{Konnerth2014}.

The Amri Karbi live in scattered villages located either in the plains or the hills; they live among an Assamese-speaking population and other tribes that speak either related languages like Tiwa or genealogically unrelated languages like Khasi (Austro-Asiatic). The \textit{Ethnologue} (\citealt{EberhardEtAl2019}) mentions 125,000 native speakers of Amri Karbi. The Census of India from 2011 states that there are 511,732 Karbi/Mikir native speakers in Assam and 14,380 Karbi/Mikir speakers in Meghalaya, and a total of 528,503 Karbi/Mikir speakers in all of India.\footnote{\href{http://censusindia.gov.in/}{http://censusindia.gov.in/}} The number of speakers for Assam includes both Amri Karbi speakers and Karbi speakers of Karbi Anglong. Therefore, the exact number of Amri Karbi speakers is uncertain.

Amri Karbi is a verb-final language and TAM distinctions are expressed through verb morphology. Amri Karbi is a synthetic language, with some degree of variation in the form of morphemes and fusion at morpheme boundaries.

\tabref{tab:amr:1} gives a summary of verbal affixes and their assigned slots. 

\begin{sidewaystable}
\begin{tabularx}{\textwidth}{l@{}l|l|l|l|l|l|l|l}
% \lsptoprule
{1} & {2} & {3} & {4} & {5} & {6} & {7} & {8} & {9}\\
\midrule 
{ \textit{nang}=} & \textsc{nmlz} \textit{ki-} & \textsc{rr} \textit{chi-} & \bfseries\scshape root & \textsc{pred}.\textsc{deriv} & \textsc{red} & {\textbf{\textsc{neg} \textbf{-C}}\textbf{\textit{e}}} & {\textsc{r} \textit{{}-lò}} & {\textsc{tag} \textit{=dik}}\\

{1 \textit{ne}=} & \textsc{pst} \textit{ki-} & \textsc{caus} \textit{pa-} & & \scshape (red) & & \textsc{imp} \textit{{}-not} & {\textsc{hab} \textit{{}-man}} & {\textsc{assum} =hole/=hele}\\

{2 \textit{nang}=} & & & & \textsc{(neg)} & & \textsc{imp}.\textsc{pol} \textit{–tha} & {\textsc{pot} \textit{{}-bò}} & {\textsc{conf}.\textsc{q}\textit{=bō}}\\

{\textsc{1pl}.\textsc{incl} \textit{i=}} & & & & & &\textsc{proh} {}-\textit{nek} & {\textsc{fut} {}-\textit{ye}} & {or =\textit{ba}}\\

{\textsc{1pl}.\textsc{excl} \textit{hali}=} & & & & & &\textsc{hort} \textsc{{}-}\textit{nang} & {\textsc{irr1} \textit{{}-bang}} & {\textsc{rep} \textsc{=}\textit{tàng/=tànghò}}\\

&&&&&&& {\textsc{nfnl}.\textsc{r} \textit{{}-si}} & {\textsc{q} =\textit{ma}}\\
&&&&&&& {\textsc{nfnl}.\textsc{irr} \textit{{}-ne}}& {\textsc{q2} =\textit{te}}\\
&&&&&&& {yet –\textit{lang}}& {\textsc{add} =\textit{ta}}\\
&&&&&&& {\textsc{def} {}-\textit{po}} & {\textsc{uncert} =\textit{ne}}\\
&&&&&&&& {may.be =\textit{mine}}\\
&&&&&&&& {\textsc{repl} \textit{=to}}\\
&&&&&&&& {\textsc{interact} =\textit{hō}}\\
&&&&&&&& {\textsc{emph} =\textit{he}}\\
&&&&&&&& {\textsc{supp} =\textit{tema}}\\
&&&&&&&& {\textsc{decl} =\textit{ro}}\\
&&&&&&&& {\textsc{indf} \textit{=bak}}\\
% \lspbottomrule
\end{tabularx}
\caption{Template of Amri Karbi verb structure}
\label{tab:amr:1}
\end{sidewaystable}

Amri Karbi is a tonal language, with three tones, low, medium and high, which exhibit low functional load. In the text examples tone is marked only on those items where the tone category was identified. Since there is a high variability and fluctuation of tone, only underlying tone is marked in the text examples. Tone in the text examples is marked by diacritics, e.g. \textit{á} ‘High’, \textit{à} ‘Low’, \textit{ā} ‘Mid’. 

Most examples provided in this paper come from natural data collected by the author during fieldtrips (in the period 2013--2017) as a part of the Amri Karbi grammatical description project (see \citealt{Philippova2021}). There are two hours and thirty-four minutes of transcribed natural data; there are also several hours of recorded audio and video data that were not transcribed but were used for this paper to supplement the transcribed data. The transcribed data consists of data recorded from thirty speakers, about 5 minutes per speaker. The paper contains some elicited examples that were recorded during the fieldtrips or remotely from language consultants. When an example in this paper was elicited, this is indicated after it. There are some examples that come from the New Testament \citet{AmriNT} that was translated in 2013; these examples are marked as (NT). The description of negation in Amri Karbi in this chapter is based on the questionnaire provided by the editors (\citetv{chapters/questionnaire}) and the structure of the chapter follows the model given in the questionnaire.

\section{Clausal negation}\label{sec:amr:2}
\subsection{Standard negation}\label{sec:amr:2.1}
Standard negation (\citealt{Payne1985,Miestamo2005}) in Amri Karbi is expressed by the negative marker on a verb which is an onset reduplication suffix or a duplifix -C\textit{e}. A duplifix is a mixture of an affix and reduplication where an element attached to the base consists of both copied segments and fixed segments \citep[24]{Haspelmath2002}. In Amri Karbi, the copied segment is the onset of the last syllable of the verb stem. The fixed segment is the vowel \textit{e}. For example, the verb \textit{cho} ‘eat’ has an onset \textit{ch}, so the negative suffix for \textit{cho} is \textit{che} and the negative form of the verb is \textit{choche} ‘not eat’. If the last syllable of the verb stem does not have an onset, then the negative suffix starts with the glottal stop; for example, \textit{ùn} ‘be able’ does not have an onset consonant, so the negative suffix for \textit{ùn} is \textit{e} and the negative form of the verb is produced as \textit{ùn’e} ‘not able’.\footnote{Negative suffixes starting with a glottal stop are written with an apostrophe.} In the text examples, the negative suffix is written as -C\textit{e} on the morpheme line, where C stands for the onset consonant and \textit{e} is the fixed vowel of the negative suffix.

The negative suffix attaches to a verb stem (that may consist of a verb and a predicate derivation suffix) \REF{ex:amr:1}, thus appearing between the stem and TAM suffix.

\ea\label{ex:amr:1}
    \ea\label{ex:amr:1a} Ne dame.  \\ 
        \gll 
              ne dam-ye \\
              1 go-\textsc{fut}\\
        \glt ‘I will go.’
    \ex\label{ex:amr:1b}Ne damdeye. \\
        \gll ne dam-Ce-ye \\
              1 go-\textsc{neg}{}-\textsc{fut}\\
        \glt ‘I will not go.’ (elicited)
    \z 
\z

Amri Karbi standard negation forms are different from their affirmative counterparts with regard to TAM marking, where the basic negated verb stem serves as a general negative form. As is shown in this section, there is a paradigmatic asymmetry (the term adapted from the typology suggested by \citealt{Miestamo2005} between Amri Karbi’s affirmative and negative structures with some TAM categories). This asymmetry is exhibited by the tense-aspect distinction in the negative being neutralized. For example, in the habitual there is a constructional asymmetry as the habitual \textit{{}-man} disappears in the negative \REF{ex:amr:2}; in other words, there is a cooccurrence restriction of the negative -C\textit{e} and the habitual \textit{{}-man}. There is also a variation in the negative of the future \textit{{}-ye} \REF{ex:amr:3}, with either keeping the future \textit{{}-ye} in the negative or not. As a result, negative forms of the habitual \textit{{}-man} and the future -\textit{ye} are sometimes identical, and the formal distinction between these forms is neutralized in the negative; their senses have to be inferred from context. 

\ea\label{ex:amr:2} 
    \ea\label{ex:amr:2a}Ne an choman.  \\ 
        \gll ne an cho-man \\
              1 rice eat-\textsc{hab} \\
        \glt  ‘I eat rice.’ 
    \ex\label{ex:amr:2b}Ne an choche. \\
        \gll ne an cho-Ce\\
             1 rice eat-\textsc{neg} \\
        \glt ‘I do not eat rice.’ (elicited)
    \z
\z 

\ea\label{ex:amr:3} 
    \ea\label{ex:amr:3a}Ne an choye.  \\
        \gll  ne an cho-ye \\
              1 rice eat-\textsc{fut} \\
        \glt ‘I will eat rice.’ 
    \ex\label{ex:amr:3b}Ne an chocheye. \\
        \gll  ne an cho-Ce-ye \\
              1 rice eat-\textsc{neg}{}-\textsc{fut} \\
        \glt ‘I will not eat rice.’ 
    \ex\label{ex:amr:3c} Ne an choche. \\
        \gll ne an cho-Ce\\
              1 rice eat-\textsc{neg} \\
        \glt ‘I will not eat rice.’ (elicited)
    \z
\z 

In the database, the negative and the future -\textit{ye} occurs primarily along with the conditional clauses \REF{ex:amr:4}. The general negative form of the verb stem and the negative duplifix are found in contexts referring to future events \REF{ex:amr:5}. 

{\ea\label{ex:amr:4} Aro labang hakpo kiphok phoknge putang putene nang eku hlonghleye. \\
    \gll aro labang hak-po ki-phok phok-ye \{putang putene\} nang eku \textbf{hlong-Ce-ye}\\
         and \textsc{dem} bucket-\textsc{def} \textsc{nmlz}{}-open open-\textsc{fut} then {} 2 nothing get-\textsc{neg}{}-\textsc{fut} \\
    \glt ‘And this bucket, if you open it, then you will not get anything.’ (natural data)
\z}

\ea\label{ex:amr:5} Tene kaiyokta thanek neta thanthe.\\
    \gll tene kai=yok=ta than-ek ne-ta \textbf{than-Ce} \\
          then who=\textsc{nsubj}=\textsc{add} say-\textsc{proh} 1=\textsc{add} say-\textsc{neg}\\
    \glt ‘So do not tell anyone and I will not tell (anyone).’ (natural data)
\z

There is a different form of paradigmatic asymmetry between negative and affirmative in the case of the potential -\textit{bo}. The potential \textit{{}-bo} marks events that are non-asserted or that the speaker considers not to be the case. There is a cooccurrence restriction between the potential -\textit{bo} and the negative marker -C\textit{e}. The elicited negative form of the potential \textit{{}-bo} appears as the realis -\textit{lo} or as a basic negative form with the stem and the negative marker \REF{ex:amr:6}, as is the case with the negative habitual and future that was shown above. However, the realis -\textit{lo} with the negative suffix yields an emphatic meaning of ‘no more’ or ‘never’, which suggests that it is not an exact negative counterpart of the potential -\textit{bo} which does not have this type of emphatic meaning.\footnote{Note that the construction verb-\textsc{neg}{}-\textit{lo} does not need to have the word ‘never’ to produce this meaning, which is different than the construction verb-\textsc{neg}{}-\textsc{negex} that is discussed in \sectref{sec:amr:3.2}, where the adverb \textit{jatsi} ‘never’ is required.} 
% \mm[inline,size=\tiny,color=green!40]{Is there a way to refer to the examples that are from the natural data? If so, it would be good to have it after each example (and explained in the introduction).}
% \aut[inline,size=\tiny,color=yellow!40]{The examples (6 a, b, c) have been elicited for this study. Below each of them the examples of these forms being used in natural data.
%hūr-bò monon=ke ne=ta ne=khoi-po wang-hleng-dun-bò=mine
%stand.up-POT now=TOP 1=ADD 1=friend-DEF come-reach-join-POT=may.be
%‘I will stand up and now my friend will reach me, may be’ (HI Turtle and rabbit)

% te inghongnema inghonghe pusi anangke pulo neke inghonhelo
% te inghong-ye=ma inghong-Ce pu-si anang=ke pu-lò ne=ke inghon-Ce-lò
% then wait-FUT=Q wait-NEG say-NF:RL 3-TOP say-RL 1=TOP wait-NEG-RL
% 275
% tene anangyok thanot nekuwang pune, aro neta dambo
% tene anang=yok than-not ne=ki-wang pu-ne aro ne=ta dam-bò
% then 3=NSUBJ say-IMP 1=NMLZ-come say-NF:IRR and 1=ADD go-POT
% ‘will you wait or not (for your son-in-law) she said, I will not wait any longer, then say to him, I came,
% and I will go’

% we arleng ayokke chiap’e tike ayoksi
% we arleng a-yok=ke chi-ap-Ce tike a-yok=si
% negex.COP2 man POSS-NSUBJ=TOP RR-shoot-NEG tiger POSS-NSUBJ=FOC:RL
% apmeta
% ap-ye=ta
% shoot-FUT=ADD
% ‘(we) do not shoot people, (we) will shoot tiger only’
% }
% \mm[inline,size=\tiny,color=green!40]{I didn't mean you need to replace the elicited examples by natural ones. What I meant was that it would be good to explicate the sources for natural data as well. In the best possible case, there would a specific reference to a location in your corpus. But I realize this may not be possible for various reasons. Can you at least give a name of a text/recording where each example comes from. Or at the very least, an indication that the message comes from your fieldwork data after each example just like you do for elicited examples.}
% \aut{You are right that I do not say what peace of data the natural data examples come from. Since I do not give all the data here, I am not sure how to reference their locations. In the introduction I do say that the data comes from the natural data collected during my fieldwork and elicited examples.}

\ea\label{ex:amr:6}
    \ea\label{ex:amr:6a} Ne an chobo.  \\ 
        \gll ne an cho-bo \\
               1 rice eat-\textsc{pot} \\
        \glt ‘I am going to/will eat rice.’ 
    \ex\label{ex:amr:6b} Ne an chochelo. \\
        \gll ne an cho-Ce-lo \\
              1 rice eat-\textsc{neg}{}-\textsc{r}\\
        \glt  ‘I will not eat rice (any more).’ 
    \ex\label{ex:amr:6c} Ne an choche. \\
        \gll ne an cho-Ce\\
              1 rice eat-\textsc{neg} \\
        \glt ‘I will not eat rice.’ (elicited)
    \z
\z 


Since the realis -\textit{lo} produces the future meaning in the negative, one would expect the negative realis to have a different form. The suffix -\textit{lo} has functions that fluctuate between the categories of the realis mood, the perfect and the perfective aspect. The suffix \textit{-lo} functions as both a realis mood marker and a temporal-aspectual marker in Amri Karbi. As a mood marker, it expresses the actuality of a state or event and typically appears with copular or adjectival predicates, which are generally stative in Aktionsart terms (cf. \citealt[143-160]{Vendler1957},  \citealt{VanValin1997}). As a temporal-aspectual marker, \textit{-lo} is used with eventive (dynamic) verbs to indicate either perfect aspect (completed events with relevance to the present) or perfective aspect (events viewed as a bounded whole). The negative counterpart of the realis \textit{{}-lo} has three forms. One is when the negated verb stem is marked with the realis -\textit{lo.} In this case the meaning of the verb has an aspectual meaning of ‘no more, never’, which is similar to the future meaning of -\textit{lo} ‘no more, never’ discussed above, but it does not have future meaning \REF{ex:amr:7}. Another form consists of the negated verb stem marked with the reduced form of the negative existential \textit{ingyong}, i.e. \textit{{}-yong} (\ref{ex:amr:8}). The third form involves the negated verb stem marked with \textit{-yong}, which is further suffixed by the realis marker \textit{-lo}. In this construction, the combination of \textit{-yong} and \textit{-lo} not only negates the event but also introduces an additional layer of interpretation: the speaker has to provide or imply a reason, explanation, or justification for why the event did not occur, see example \REF{ex:amr:8}. See also \tabref{tab:amr:2} that illustrates the different negative counterpart constructions of affirmative -\textit{lo}.

\ea\label{ex:amr:7} Taramta angpasiselo.\\
    \gll  taram=ta nang=pasi-Ce-lo\\
          call=\textsc{add} \textsc{cis}=respond-\textsc{neg}{}-\textsc{r}\\
    \glt ‘Even though (he) called (him), (he) never responded.’ (natural data)
\z

\ea\label{ex:amr:8}
    \ea\label{ex:amr:8a} Ne an cholo.  \\
        \gll  ne an cho-lo \\
              1 rice eat-\textsc{r} \\
        \glt ‘I have eaten rice.’ 
    \ex\label{ex:amr:8b} Ne an chocheyong/chocheyongo. \\
        \gll ne an cho-Ce-yong/cho-Ce-yong-lo \\
              1 rice eat-\textsc{neg}{}-\textsc{negex}/eat-\textsc{neg}{}-\textsc{negex}{}-\textsc{r} \\
        \glt ‘I have not eaten.’ / ‘I have not eaten (because I was sick).’ (elicited)
    \z
\z 

\begin{table}
\centering
\begin{tabularx}{.5\textwidth}{lX}
\lsptoprule
affirmative & negative \\
\midrule 
\itshape dam-lo & \itshape dam-de-yong \\
& \itshape dam-de-yong-o \\
& \textit{dam-de-lo} ‘future sense’\\
\lspbottomrule
\end{tabularx}
    \caption{The negative counterpart constructions of affirmative -\textit{lo}}
    \label{tab:amr:2}
\end{table}

The irrealis -\textit{bang} marks events that have not happened but are going to happen. There is a component of intention, desire and necessity in actions marked with -\textit{bang}. The irrealis -\textit{bang} may occur both in the affirmative and negative, but the senses of \textit{{}-bang} are not symmetric in the affirmative and negative. In the negative, the intentionality of -\textit{bang} is not negated, the action that it marks is negated and an inference of the expectation of the action is produced \REF{ex:amr:9}. In addition, similar to other TAM discussed above (habitual, future and potential), the general negative form with the verb root and the negative suffix may be used and the meaning of \textit{{}-bang} may be inferred through context.

\ea\label{ex:amr:9}
    \ea\label{ex:amr:9a} Ne hat dambang.  \\ 
        \gll ne hat dam-bang \\
              1 market go-\textsc{irr1} \\
        \glt ‘I am going to the market now.’\footnote{\textit{Ne hat dambang} ‘I am going to the market now’ can mean that a person intends to go to the market at the time of speaking or that a person is on their way to the market.} 
    \ex\label{ex:amr:9b}  Ne hat damdebang. \\
        \gll ne hat dam-Ce-bang\\
              1 market go-\textsc{neg}{}-\textsc{irr1}\\
        \glt  ‘I am not going to the market now/yet (but I will after some time).’
    \ex\label{ex:amr:9c} Ne hat damde. \\
        \gll ne hat dam-Ce\\
              1 market go-\textsc{neg}\\
        \glt ‘I am not going to the market.’ (elicited)
    \z
\z 

The elicited negated past tense forms appear symmetrical, where the negative suffix -C\textit{e} is added to the affirmative past tense e.g. \REF{ex:amr:10}. However, in the available natural data, the past tense \textit{ki-} never occurs in the negative. Negatives with past time references in the data have the form involving reduced negative existential copula -\textit{yong/-yong-o}. This is the same negative construction that is used for the negative realis, e.g. \REF{ex:amr:11}. 

{\ea\label{ex:amr:10}
    \ea\label{ex:amr:10a}Kiwang.  \\
        \gll  ki-wang \\
              \textsc{pst}{}-come \\
        \glt ‘(I) came.’ 
    \ex\label{ex:amr:10b}  Kiwangwe. \\
        \gll ki-wang-Ce\\
              \textsc{pst}{}-come-\textsc{neg} \\
        \glt  ‘(I) did not come.’ (elicited)
    \z
\z}

\ea\label{ex:amr:11} Chiklo jonni dordelengke anang jangchilangdunde\textbf{yongo} mengsurungpo ahem ayok.\\ 
    \gll chiklo jon-ni dor-Ce-leng=ke anang nang=chi-lang-dun-Ce-yong-lo mengsurung-po a-hem a-yok\\
    month \textsc{clf}.month-two complete-\textsc{neg}{}-reach=\textsc{top} 3 \textsc{cis=rr}-look-follow-\textsc{neg}{}-\textsc{negex}.\textsc{cop}{}-\textsc{r} fox-\textsc{def} \textsc{poss}{}-house \textsc{poss}{}-\textsc{dir}\\
    \glt ‘He (the crocodile) did not go to the fox’s house for two months.’  (natural data)
\z 

\subsection{Negation in non-declaratives}\label{sec:amr:2.2}
\subsubsection{Prohibitive}\label{sec:amr:2.2.1}
Second person imperatives or commands are expressed by a bare stem or by one of the imperative suffixes \textit{{}-not} or \textit{{}-tha}. The imperative -\textit{tha} is used for plural addressees or in a formal setting, e.g. talking to one’s in-laws. The imperative -\textit{not} is used for plural addressees or also in a formal setting, e.g. talking to strangers, but it is less formal than the imperative -\textit{tha}. The bare stem form is used in informal settings, e.g. talking to friends, siblings, etc. The negative form of an imperative has a dedicated construction, which is the prohibitive suffix -\textit{yek}. The prohibitive suffix -\textit{yek} is used for all three types of imperatives; as a result, the negative of imperative does not distinguish between formal and informal forms, e.g. \REF{ex:amr:12}, thus showing paradigmatic asymmetry.

\ea\label{ex:amr:12}
    \ea\label{ex:amr:12a} Cho! / Chonot! / Chotha!  \\ 
        \gll cho / cho-not / cho-tha \\
              eat / eat-\textsc{imp} / eat-\textsc{imp}.\textsc{pol} \\
        \glt  ‘Eat! / Eat! / Eat, please!’ 
    \ex\label{ex:amr:12b}Choyek! \\
        \gll cho-yek \\
              eat-\textsc{proh} \\
        \glt ‘Do not eat!’ (elicited)
    \z
\z 

The Amri Karbi prohibitive -\textit{yek} does not occur with the imperative suffixes, which is different from Karbi (see \citealt{Konnerth2014}: 536) where the Karbi prohibitive suffix \textit{{}-rī} occurs with the imperative suffixes. 

\subsubsection{Modals}\label{sec:amr:2.2.2}
Some negative non-declarative clause types are expressed with the modal predicate derivational suffixes -\textit{lag(e)} ‘want, should’ and \textit{{}-un} ‘be able’. In negation of the modal -\textit{lag(e)} there is an asymmetry as the negative suffix does not attach to -\textit{lag(e),} but to a different modal suffix -\textit{nang} e.g. \REF{ex:amr:13} (note that in the case of the predicate derivational suffix -\textit{nang}, the scope of negation is on the entire predicate). The suffix -\textit{nang} is an older modal that was used both in the affirmative and negative, but nowadays, Amri Karbi speakers prefer to use the Assamese borrowed modal -\textit{lag(e)} in the affirmative. However, the old form \textit{nang} is still used in the negative. Cross-linguistically, it is common for older elements to be sometimes preserved under negation, and Amri Karbi -\textit{lag(e),} and \textit{-nang} are good examples of this. Amri Karbis remember that they previously used the modal \textit{-nang} in both the affirmative and the negative and it was used in the New Testament translation \cite{AmriNT} in the affirmative. This form is still used in the closely related language Karbi \citep[356]{Konnerth2014}, where the modal -\textit{lag(e)} has not been adopted. 

\ea\label{ex:amr:13}
    \ea\label{ex:amr:13a} Chobellage.  \\ 
        \gll chobel-lage \\
              deceive-need \\
        \glt ‘You should deceive.’ 
    \ex\label{ex:amr:13b}Chobelnangne. \\
        \gll chobel-nang-Ce\\
              deceive-need-\textsc{neg} \\
        \glt ‘You should not deceive.’ (elicited)
    \z
\z

Non-declarative clauses constructed with the modal -\textit{un} are negated with the negative verbal duplifix -C\textit{e} \REF{ex:amr:14}. 

\ea\label{ex:amr:14} Nanghem kihlenghlelengke nang labang hakpo phoun’e.\\
    \gll nang-hēm ki-hleng-Ce=\textbf{lengke} nang la=bang hak-po pho-\`{u}n-Ce\\
         2-house \textsc{nmlz}{}-reach-\textsc{neg}=until 2 \textsc{dem}=\textsc{gnr} bucket-\textsc{def} open-be.able-\textsc{neg} \\
    \glt \textsc{‘}Before you reach home, you must not open the bucket.’ (PM Bilangthe)
\z

\subsubsection{Negative questions}\label{sec:amr:2.2.3} 
Polar questions are constructed with the question particle \textit{=ma} that marks the predicate. In negative polar questions, the verb has the negative suffix -C\textit{e} and the question particle is the same =\textit{ma} as in the standard negation questions, e.g. \REF{ex:amr:15}. However, in negative questions that involve the negative existential \textit{ingyong} the question particle is =\textit{te} instead of =\textit{ma}, e.g. \REF{ex:amr:16}. The question particle \textit{=te} is also a dedicated question particle for rhetorical questions and reprimands.

\ea\label{ex:amr:15}
    \ea\label{ex:amr:15a}  Mohan hat kidamma?  \\
        \gll Mohan hat ki-dam=ma \\
              Mohan market \textsc{pst}{}-go=\textsc{q} \\
        \glt ‘Did Mohan go to the market?’
    \ex\label{ex:amr:15b} Mohan hat damdema?  \\
        \gll Mohan hat dam-Ce=ma \\
              Mohan market go-\textsc{neg}=\textsc{q}\\
        \glt ‘Didn’t Mohan go to the market?’ (elicited)
    \z
\z 

\ea\label{ex:amr:16} 
    \ea\label{ex:amr:16a}Mohanpo hem doma?  \\ 
        \gll Mohan-po hem do=ma \\
              Mohan-\textsc{m} home \textsc{cop}=\textsc{q} \\
        \glt ‘Is Mohan at home?’ 
    \ex\label{ex:amr:16b} Mohan-po hem ingyongte?  \\
        \gll Mohan-po hem ingyong=te\\
              Mohan-\textsc{m} home \textsc{negex}=\textsc{q}\\
        \glt ‘Isn’t Mohan at home?’ (elicited)
    \z
\z 

\subsection{Negation in stative predications}\label{sec:amr:2.3}
Stative predications are classified according to \citet[111]{Payne1997} with the following types of stative predication clauses: equation, class inclusion, attribution, locative predication, existential predication and possessive predication.

\subsubsection{Equative and class inclusion clauses}\label{sec:amr:2.3.1}
In equative and class inclusion clauses, nouns can be used as predicates without any further elements in the present tense. The subject is topicalized and the nominal predicate usually carries the definite marker \textit{{}-po/-pe}. When the nominal clause is in the past, its predicate has to include a combination of a noun and the copula \textit{do}, and the copula is marked with the past tense \textit{ki-}. In the future tense, the equative and class inclusion clauses generally use the verb \textit{chiplang} ‘become’ instead of the copula \textit{do}. The negative equative is \textit{kalek}, which is placed at the end of the negative equative and class inclusion clauses. The negative \textit{kalek} does not take TAM markers, but it can be marked with the second syllable of the existential \textit{ingyong}, i.e. \textit{kalekyong}. The negative \textit{kalek} is used in negated present tense clauses, even though there is no affirmative copula in equative and class inclusion clauses of the present tense, e.g. (\ref{ex:amr:17}--\ref{ex:amr:18++}). The present and past tense negative equative and class inclusion clauses are identical in form and the past tense is inferred through context e.g. (\ref{ex:amr:17}--\ref{ex:amr:18++}). The fact that the past tense distinction is eliminated in the negative makes the pattern asymmetric in the past tense. The future negative equative and class inclusion clauses are marked on the verb \textit{chiplang} with the negative duplifix -C\textit{e} e.g. (\ref{ex:amr:17}--\ref{ex:amr:18++}).


\ea\label{ex:amr:17}
    \ea\label{ex:amr:17a}Anangke neaipe.  \\ 
        \gll anang=ke ne-ai-pe \\
              3=\textsc{top} 1-mother-\textsc{f} \\
        \glt ‘She is my mother.’ 
    \ex\label{ex:amr:17b}Anangke neaipe kalek. \\
        \gll anang=ke ne-ai-pe kalek\\
              3=\textsc{top} 1-mother-\textsc{f} \textsc{neg.eq}\\
        \glt ‘She is not/was not/will not be my mother.’ (elicited)
    \z 
\z 
\ea\label{ex:amr:17+}
    \ea\label{ex:amr:17c}Anangke neaipe kido.  \\ 
        \gll anang=ke ne=ai-pe ki-do \\
              3=\textsc{top} 1=mother-\textsc{f} \textsc{pst}{}-\textsc{cop} \\
        \glt ‘She was my mother.’ 
    \ex\label{ex:amr:17d}Anangke neaipe kalek.  \\
        \gll  anang=ke ne-ai-pe kalek \\
              3=\textsc{top} 1-mother-\textsc{f} \textsc{neg.eq} \\
        \glt ‘She is not/was not/will not be my mother.’ (elicited)
    \z
\z 

{\ea\label{ex:amr:17++}
    \ea\label{ex:amr:17e} Anangke neɑipe chiplangye.  \\ 
        \gll anang=ke ne=ɑi-pe chiplang-ye \\
              3=\textsc{top} 1=mother-\textsc{f} become-\textsc{fut} \\
        \glt ‘She will become my mother.’ 
    \ex\label{ex:amr:17f}Anangke neɑipe chiplangpleye.  \\
        \gll anang=ke ne=ɑi-pe chiplang-Ce-ye\\
             3=\textsc{top} 1=mother-\textsc{f} become-\textsc{neg}{}-\textsc{fut} \\
        \glt ‘She will not become my mother.’ (elicited)
    \z
\z }

\newpage
\ea\label{ex:amr:18} 
    \ea\label{ex:amr:18a} Mohanpoke sikaripo.  \\ 
        \gll Mohan-po=ke sikari-po \\
               Mohan-\textsc{def}=\textsc{top} hunter-\textsc{def} \\
        \glt ‘Mohan is a hunter.’ 
    \ex\label{ex:amr:18b}Mohanpoke sikaripo kalek.  \\
        \gll Mohan-po=ke sikari-po kalek\\
              Mohan-\textsc{def}=\textsc{top} hunter-\textsc{def} \textsc{neg}.\textsc{eq}\\
        \glt ‘Mohan is not a hunter.’ (elicited)
    \z
\z 
\ea\label{ex:amr:18+}
    \ea\label{ex:amr:18c}Mohanpoke sikaripo kido.  \\ 
        \gll Mohan-po=ke sikari-po ki-do \\
              Mohan-\textsc{def}=\textsc{top} hunter-\textsc{def} \textsc{pst}{}-\textsc{cop} \\
        \glt ‘Mohan was a hunter.’ 
    \ex\label{ex:amr:18d} Mohanpoke sikaripo kalek.  \\
        \gll Mohan-po=ke sikari-po kalek\\
              Mohan-\textsc{def}=\textsc{top} hunter-\textsc{def} \textsc{neg}.\textsc{eq}\\
        \glt ‘Mohan was not a hunter.’ (elicited)
    \z 
\z 
\ea\label{ex:amr:18++}
    \ea\label{ex:amr:18e} Mohanpoke sikaripo chiplangye.  \\ 
        \gll Mohan-po=ke sikari-po chiplang-ye \\
              Mohan-\textsc{def}=\textsc{top} hunter-\textsc{def} become\textsc{{}-fut} \\
        \glt ‘Mohan will become a hunter.’ 
    \ex\label{ex:amr:18f} Mohanpoke sikaripo chiplangpleye.  \\
        \gll  Mohan-po=ke sikari-po chiplang-Ce-ye \\
              Mohan-\textsc{def}=\textsc{top} hunter-\textsc{def} become-\textsc{neg-fut} \\
        \glt ‘Mohan will not become a hunter.’ (elicited)
    \z
\z 

\subsubsection{Attributive clauses}\label{sec:amr:2.3.2}
In Amri Karbi, the attributive function is usually fulfilled by adjectival verbs that are nominalized and usually carry the possessive prefix \textit{a-}. Since adjectival verbs are verbs, the verbal negative duplifix -C\textit{e} is used when attributive clauses are negated, e.g. \REF{ex:amr:19}. However, since adjectival verbs are nominalized, the nominal predicate negation strategy can also be used to negate adjectival verbs. So, to negate an attributive clause, the negative equational \textit{kalek} can be used, but the negative existential \textit{ingyong} can be used as well. Examples \REF{ex:amr:19} and \REF{ex:amr:20} show an adjectival verb \textit{akime} ‘be.good’ being negated either with the standard negator \REF{ex:amr:19b} or the negative equative \textit{kalek} or the negative existential \textit{ingyong} \REF{ex:amr:20b}, the meaning in both cases is the same. There is a symmetry between affirmative and negative attributive clauses in examples \REF{ex:amr:19} and \REF{ex:amr:20}. 

\ea\label{ex:amr:19} 
    \ea\label{ex:amr:19a} An akime.  \\ 
        \gll an a-ki-me \\
              rice \textsc{poss-nmlz}{}-be.good \\
        \glt ‘Rice (generally) is good.’ 
    \ex\label{ex:amr:19b}An akimeme.  \\
        \gll an a-ki-me-Ce \\
              3 \textsc{poss-nmlz}{}-be.good-\textsc{neg} \\
        \glt ‘Rice (generally) is not good.’ (elicited) 
    \z
\z 

\ea\label{ex:amr:20}
    \ea\label{ex:amr:20a}An akime.  \\ 
        \gll an a-ki-me \\
              rice \textsc{poss}{}-\textsc{nmlz}{}-be.good \\
        \glt  ‘Rice (generally) is good.’ 
    \ex\label{ex:amr:20b}An akime kalek/ingyong.  \\
        \gll an a-ki-me kalek/ingyong \\
              rice \textsc{poss}{}-\textsc{nmlz}{}-be.good \textsc{neg}.\textsc{eq}/\textsc{negex} \\
        \glt ‘Rice (generally) is not good.’ (elicited) 
    \z
\z 

When attributive clauses appear with different tenses, the constructions are partially symmetric. In the past tense the affirmative attributive clause involves the copula \textit{do} that inflects for past tense; in the negative, the negative suffix attaches to the copula \textit{do,} thus being symmetric. In the future tense, the adjectival verb itself functions as a verb and inflects for the future tense, while in the negative, the negative verbal suffix attaches to the adjectival verb, but the future tense is lost, thus being asymmetric, e.g. \REF{ex:amr:21}.

\ea\label{ex:amr:21}
    \ea\label{ex:amr:21a}An akime kido.  \\ 
        \gll rice a-ki-me ki-do \\
               rice \textsc{poss}{}-\textsc{nmlz}{}-be.good \textsc{pst}{}-\textsc{cop} \\
        \glt ‘(The cooked) rice was good.’ 
    \ex\label{ex:amr:21b}An akime kidode. \\
        \gll an a-ki-me ki-do-Ce\\
              rice \textsc{poss}{}-\textsc{nmlz}{}-be.good \textsc{nmlz}{}-\textsc{cop}{}-\textsc{neg}\\
        \glt ‘(The cooked) rice was not good.’
    \ex\label{ex:amr:21c}An akimeye/meye.  \\ 
        \gll an a-ki-me-ye/me-ye \\
              rice \textsc{poss}{}-\textsc{nmlz}{}-be.good-\textsc{fut}/be.good-\textsc{fut} \\
        \glt ‘(The cooked) rice will be good.’ 
    \ex\label{ex:amr:21d}An akimeme/meme. \\
        \gll an a-ki-me-Ce/me-Ce\\
              rice \textsc{poss}{}-\textsc{nmlz}{}-be.good-\textsc{neg}/be.good-\textsc{neg}\\
        \glt ‘(The cooked) rice will not be good.’ (elicited)
    \z
\z 


\subsubsection{Existential clauses} \label{sec:amr:2.3.3}
\largerpage
Existential clauses that usually perform a presentative function are expressed by a combination of a noun and the copula \textit{do} regardless of tense, where the copula \textit{do} occurs clause-finally. The copula \textit{do} may occur without TAM markers and has a present tense meaning and may inflect for TAM. Existential clauses are negated by the negative existentials \textit{ingyong} or \textit{we}, e.g. \REF{ex:amr:22}. The negative existentials \textit{ingyong} and \textit{we} may take some TAM markers; both \textit{ingyong} and \textit{we} may be marked with the realis \textit{{}-lo} for the aspectual meaning, e.g. ‘no more’, and both \textit{ingyong} and \textit{we} may be marked with the future -\textit{ye}, e.g. \textit{weye} ‘will not be’. The negative existential \textit{ingyong} also occurs with the aspect \textit{{}-lang} ‘yet’, the imperative -\textit{not} and the non-final verbal suffix \textit{{}-si}. The past tense is not marked on the existential and is inferred through the context, which makes this pattern asymetric. Subject arguments in negative existential clauses are usually definite. Examples (\ref{ex:amr:22}--\ref{ex:amr:24}) show an indefinite subject in the positive existential marked with the numeral \textit{isi} ‘one’. If we replace the positive copula \textit{do} with the negative \textit{ingyong} in (\ref{ex:amr:22}--\ref{ex:amr:24}) the meaning will be \textit{isi anijom ingyong} ‘one custom is missing’. So, the subject in the negative is identifiable and the numeral \textit{isi} ‘one’ may no longer function as an indefinite marker in (\ref{ex:amr:22}--\ref{ex:amr:24}), but as the numeral ‘one’. On the other hand, since the obligatory possessive prefix \textit{a-} is on the word \textit{anijom,} it can also have a third person possessive reading of the existential clause ‘they do not have a custom/ he or she does not have a custom’. If we look at a noun without the obligatory possessive prefix \textit{a}{}- (for example \textit{tike} ‘tiger’) in \REF{ex:amr:25}, in the negative numeral \textit{i-jon} ‘one-\textsc{clf}.\textsc{anim’} will again have a numeric meaning rather than indefinite and without the numeral, negation will induce the definite marker -\textit{po} or an information structure topic =\textit{ke}, focus =\textit{si}, additive =\textit{ta} or a plural \textit{{}-hel/-(po)mar} on the subject, depending on the context.\footnote{"The negative copular verbs *\textit{ingyong} and \textit{we} are generally in free variation. The copula \textit{kangwe}, likely formed by the lexicalization of the nominalized negative copulas \textit{ingyong} and \textit{we} (with the initial vowel of \textit{ingyong} dropped, as typical with the prefix \textit{ing}- in Amri Karbi), is primarily used in possessive clauses, where \textit{we} rarely occurs.}

\ea\label{ex:amr:22}
        \ea\label{ex:amr:22a}Isi anijom do.  \\ 
            \gll isi a-nijom do \\
                  one \textsc{poss}{}-custom \textsc{cop} \\
            \glt ‘There is one custom.’ 
        \ex\label{ex:amr:22b} Anijom(po) ingyong/we. \\
            \gll a-nijom(po) ingyong/we\\
                  \textsc{poss}{}-custom-(\textsc{def}) \textsc{negex} \\
            \glt ‘There is no/was no/will not be custom.’ / ‘They do not have (the) custom.’ (natural data)
       \z
\z
\ea\label{ex:amr:23}
        \ea\label{ex:amr:23a} Isi anijom ki-do.  \\ 
            \gll isi a-nijom ki-do \\
                  one \textsc{poss}{}-custom \textsc{pst}{}-\textsc{cop} \\
            \glt ‘There was one custom.’ 
        \ex\label{ex:amr:23b}Anijom(po) ingyong/we.  \\
            \gll a-nijom(po) ingyong/we \\
                  \textsc{poss}{}-custom-(\textsc{def}) \textsc{negex} \\
            \glt ‘There was no custom.’ (elicited)\\
        \z
    \z
\ea\label{ex:amr:24}
        \ea\label{ex:amr:24a} Isi anijom doye.  \\
            \gll isi a-nijom do-ye \\
                  one \textsc{poss}{}-custom \textsc{cop-fut} \\
            \glt  ‘There will be one custom.'
        \ex\label{ex:amr:24b} Anijom(po) weye.  \\
            \gll  a-nijom(po) we-ye \\
                  \textsc{poss}{}-custom-(\textsc{def}) \textsc{negex-fut} \\
            \glt ‘There will be no custom.’ (elicited)
        \z 
    \z 


\ea\label{ex:amr:25}
    \ea\label{ex:amr:25a} Tike ijon do.  \\ 
        \gll tike i-jon do \\
              tiger one-\textsc{clf}.\textsc{anim} \textsc{cop} \\
        \glt  ‘There is one tiger.’ 
    \ex\label{ex:amr:25b}Tike ijon ingyong.  \\
        \gll tike i-jon ingyong\\
              tiger one-\textsc{clf}.\textsc{anim} \textsc{negex}\\
        \glt ‘One tiger is missing.’
\newpage
    \ex\label{ex:amr:25c}Tikepo ingyong. \\
        \gll tike-po ingyong\\
              tiger-\textsc{def} \textsc{negex}\\
        \glt ‘There is no tiger.’ (elicited)
    \z
\z 
% \lv{Apparently, there are two different negative existentials, \textit{ingyong} and \textit{we}. Could you please discuss the difference between them? Are \textit{we} and \textit{kangwe} related?}
% \aut{the negative copular verbs inyong and we are in free variantion in most cases. The copula "kangwe" is a nominalized form that likely originated from the lexicalization of the two nominalized negative copular verbs "ingyong" and "we" combined together. In this process, the initial vowel of the copula ingyong" is dropped, as is typical with the prefix "ing-" in Amri Karbi. The copula "kangwe" is primarily used in possessive clauses, whereas "we" typically does not appear in possessive clauses.}
% \hc{Could you formulate a footnote about this?}
% \aut{Here is the footnote:
% "}


\subsubsection{Possessive clauses}\label{sec:amr:2.3.4}
Amri Karbi possessive clauses employ the existential structure, which is a combination of a noun and a copula regardless of tense and mood. Similarly to existential clauses, the copula in possessive clauses may inflect for tense and mood and appears clause-finally. Possessive clauses are negated by the negative existential \textit{ingyong} or the negative existential \textit{kangwe} ‘not.exist’, e.g. (\ref{ex:amr:26}--\ref{ex:amr:27}). The negative existential \textit{ingyong} may be marked with the realis \textit{{}-lo} (with the meaning ‘no more’), the future -\textit{ye} and \textit{{}-lang} ‘yet’; \textit{kangwe} is a nominalized form and does not occur marked with TAM. Negative possessive clauses that appear in past time contexts are also construed as having past time reference \REF{ex:amr:27}. Possessive clauses with future time reference, i.e. ‘I will have a house’ could not be produced.

\ea\label{ex:amr:26}
    \ea\label{ex:amr:26a}Nehem do.  \\ 
        \gll ne-hem do \\
               1-house \textsc{cop} \\
        \glt ‘I have a house.’ (lit. ‘My house exists.') 
    \ex\label{ex:amr:26b}Nehem ingyong/kangwe.  \\
        \gll ne-hem ingyong/kangwe\\
              1-house \textsc{negex}/not.exist \\
        \glt  ‘I do not have a house.’ (lit. ‘My house does not exist.')
    \ex\label{ex:amr:26c}Nehem kido.  \\ 
        \gll ne-hem ki-do \\
              1-house \textsc{pst}{}-\textsc{cop} \\
        \glt  ‘I had a house.’ (lit. ‘My house exists.') 
    \ex\label{ex:amr:26d}Nehem ingyong/kangwe.  \\
        \gll ne-hem ingyong/kangwe\\
              1-house \textsc{neg}.\textsc{ex}/not.exist\\
        \glt ‘I did not have a house.’ (lit. ‘My house did not exist.') (elicited)
    \z
\z 

\newpage
\ea\label{ex:amr:27}Atomo asarpe himepe ahem kangwe aret kangwe salkam katiki kaita kangwe. \\
    \gll a-tomo a-sar-pe hime-pe a-hem \textbf{kangwe} a-ret \textbf{kangwe} sal-kam ki-tiki kaita \textbf{kangwe} \\
         \textsc{poss}{}-story \textsc{poss}{}-be.old-\textsc{f} widow-\textsc{f} \textsc{poss}{}-house not.exist \textsc{poss}{}-field not.exist work-work \textsc{nmlz}{}-work nobody not.exist\\
    \glt ‘A story about an old widow, (she) had no home, no field, nobody to work (with her).’ (natural data)
\z

\subsubsection{Locative predicate}\label{sec:amr:2.3.5}
Locative predicate clauses also use existential construction and are constructed with the copular verb \textit{do} and one of the locative relator nouns, e.g. -\textit{long} ‘\textsc{loc’}. The copula \textit{do} inflects for TAM and appears clause-finally. Locative predicates are negated by negative existentials \textit{ingyong} or \textit{we}, that occur clause-finally and may inflect for TAM as in existential clauses, e.g. (\ref{ex:amr:28}--\ref{ex:amr:30}). Locative predicate clauses in past time contexts may also infer past tense through the context \REF{ex:amr:31}. 


\ea\label{ex:amr:28}
        \ea\label{ex:amr:28a} Hikpo hem along do.  \\ 
            \gll hik-po hem a-long do \\
                  dog-\textsc{def} house \textsc{poss}{}-\textsc{loc} \textsc{cop} \\
            \glt  ‘The dog is in the house.’ 
        \ex\label{ex:amr:28b} Hikpo hem along ingyong.  \\
            \gll hik-po hem a-long ingyong\\
                   dog-\textsc{def} house \textsc{poss}{}-\textsc{loc} \textsc{negex}\\
            \glt ‘The dog is not in the house.’ (elicited)\\
        \z
    \z
    \ea\label{ex:amr:29}
        \ea\label{ex:amr:29a}Hikpo hem along kido.  \\
            \gll hik-po hem a-long ki-do \\
                  dog-\textsc{def} house \textsc{poss}{}-\textsc{loc} \textsc{pst}{}-\textsc{cop} \\
            \glt ‘The dog was in the house.’ 
        \ex\label{ex:amr:29b}Hikpo hem along ingyongo.  \\
            \gll hik-po hem a-long ingyong-lo \\
                  dog-\textsc{def} house \textsc{poss}{}-\textsc{loc} \textsc{negex-r} \\
            \glt ‘There was no dog in the house (anymore).’ (elicited)\\
    \z
 \z
\ea\label{ex:amr:30}
        \ea\label{ex:amr:30a}Hikpo hem along doye. \\ 
            \gll hik-po hem a-long do-ye  \\
                  dog-\textsc{def} house \textsc{poss}{}-\textsc{loc} \textsc{cop}{}-\textsc{fut}  \\
            \glt  ‘The dog will be in the house.’ 
        \ex\label{ex:amr:30b}Hikpo hem along weye. \\
            \gll hik-po hem a-long we-ye \\
                  dog-\textsc{def} house \textsc{poss}{}-\textsc{loc} \textsc{negex}{}-\textsc{fut}\\
            \glt ‘The dog will not be in the house.’ (elicited)
        \z
    \z 


\ea\label{ex:amr:31} Anangai wang kilang ayokke ososo hemtak along ingyong.\\
    \gll anang-ai wang ki-lang a-yok=ke oso-so hemtak a-long ingyong\\
          3-mother come \textsc{nmlz}{}-look \textsc{poss}{}-\textsc{purp}=\textsc{top} child-\textsc{def} hut \textsc{poss}{}-\textsc{loc} \textsc{negex}\\
    \glt ‘When his mother came to look, the baby was not in the hut.’ (natural data)
\z

\subsubsection{Summary of negation in stative predications}\label{sec:amr:2.3.6}
There are three constructions for stative predications. One is the equative construction that is also used in class inclusion clauses; they are constructed with a nominal predicate in the present tense and involve a copular verb in the past and future. Other stative predicates - existential, possessive and locative - employ one existential construction that involves the copula \textit{do}, which inflects for tense. The negation of stative predications does not involve the negative suffix -C\textit{e} that is used in standard negation. The equative clauses are negated with the negative equative \textit{kalek} which does not inflect for tense. Negative existential clauses are constructed with the negative existentials \textit{ingyong} and \textit{we} for existential and locative clauses, \textit{ingyong} and \textit{kangwe} for possessive clauses. The negative existentials \textit{ingyong} and \textit{we} may inflect for TAM or receive their temporal reference through context. Thus the pattern is asymmetric and possessives and locatives inherit this asymmetry from existentials.

\subsection{Negation in non-main clauses} \label{sec:amr:2.4}
Subordinate clauses are constructed by the nominalizer \textit{ki-} marking the subordinate verb. Besides nominalization, subordination can also be generated by the realis non-final suffix \textit{–si} or the irrealis \textit{{}-ne.} The non-final suffix -\textit{si} is used for subordination in declarative clauses and in the realis mood, whereas the non-final suffix -\textit{ne} is used for subordination of non-declarative clauses and declarative clauses in the irrealis mood. Nominalized subordinate clauses \REF{ex:amr:32}, \REF{ex:amr:34}, as well as those that are marked with non-final suffixes -\textit{si/-ne} \REF{ex:amr:33}, are negated by the verbal negative suffix {}-C\textit{e}. Therefore, subordinate clauses use standard negation in a regular way without the asymmetry that was observed in the main clauses. The affirmative counterparts of the examples \REF{ex:amr:32}, \REF{ex:amr:33}, and \REF{ex:amr:34} could be constructed by leaving out the negative suffix -C\textit{e}. The main reason for the symmetry is that there is usually no TAM marking in the subordinate clauses. 

\ea\label{ex:amr:32} Parake anang bupeke kampihel kikatune ayok hem pademlo.  \\
    \gll parake anang bu-pe=ke kampi-hel \textbf{ki-kat-un-Ce} a-yok hem padem-lo\\
          then \textsc{3} grandmother-\textsc{fem=top} monkey-\textsc{pl} \textsc{nmlz}{}-run-be.able-\textsc{neg} \textsc{poss}{}-\textsc{purp} house repair-\textsc{r}\\
    \glt ‘And their grandma repaired the house so that monkeys could not run away.’ (natural data)
\z

\ea\label{ex:amr:33}Manta phurulpoke chiklangklesi arlengposi kimead arlengposi chiklangdungsi wango.\\
    \gll manta phurul-po=ke \textbf{chi-klang-Ce-si} arleng-po=si ki-me-ad arleng-po=si chi-klang-dung-si wang-lo\\
         but snake-\textsc{def}=\textsc{top} \textsc{rr}{}-show-\textsc{neg}{}-\textsc{nfnl}.\textsc{r} man-\textsc{def}=\textsc{foc.r} \textsc{nmlz}{}-be.good-be.much man-m=\textsc{foc}.\textsc{r} \textsc{rr}{}-show-near-\textsc{nfnl}.\textsc{r} come-\textsc{r}\\
    \glt ‘But the snake did not show himself (as a snake) he came out as a handsome man.’ (natural data)
\z

\ea\label{ex:amr:34}Lasi labadisi anang damman kihlonghle apor manta {aneng arong} doman.\\
    \gll lasi labadi=si anang dam-man \textbf{ki-hlong-Ce} a-por manta {\{aneng arong\}} do-man\\
         so like.this=\textsc{foc}.\textsc{r} 3 go-\textsc{hab} \textsc{nmlz}{}-get-\textsc{neg} \textsc{poss}{}-time but happy \textsc{cop}{}-\textsc{hab}\\
    \glt ‘So he goes hunting like this, even when he did not get (birds) he was happy.’ (natural data)
\z

The subordinator \textit{=lengke} `until' (which is a negative polarity item, see \sectref{sec:amr:4.2}) occurs with the negated main clause that marks the punctual ‘until’ for the event in the main clause to happen (\ref{ex:amr:53}--\ref{ex:amr:54}). The verb in the subordinate clause marked with =\textit{lengke} is also negated. This phenomenon is called "expanded negation" \citep[145]{Wälchli2018}, where negation in temporal clauses in translation equivalents in many other languages lack negation. An ‘until’ clause in Amri Karbi identifies the time when the event in the main clause may happen, happened or will happen. These negated adverbial clauses demonstrate an “extremely close relation between linguistic form and function” \citep[145--146]{Hetterle2015}; in Amri Karbi, if the event in the adverbial clause marked with =\textit{lengke} does not happen, the event in the main clause also does not happen. The ‘until’ clause marked with =\textit{lengke} may also occur with an affirmative main clause \REF{ex:amr:55}, where the adverbial clause identifies a point of time until which the event of the main clause may last. 
  


\ea\label{ex:amr:52}Lasi bekso niyok panghongtolo apor chokchelang apodsi.\\
    \gll lasi bek-so ne-yok pa-inghong-tol-lo apor chok-Ce-\textbf{lang} apod=si \\
         so little-\textsc{dim} 1-\textsc{n.subj} \textsc{caus}{}-wait-send-\textsc{r} time finish-\textsc{neg}{}-yet because=\textsc{foc}\\
    \glt ‘So I was asked to wait for sometime because time is not yet finished.’ (natural data)
\z

\ea\label{ex:amr:53}Chiklo jonni dordelengke anang jangchilangdundeyongo mengsurungpo ahem ayok. \\
    \gll chiklo jon-ni dor-Ce=\textbf{lengke} anang nang=chi-l\`{a}ng-dun\textbf{{}-}Ce-yong-lò mengsurung-po a-h\={e}m a-yok \\
         month \textsc{clf}:month-two complete-\textsc{neg}=until 3 \textsc{cis}=\textsc{rr}{}-look-follow-\textsc{neg}{}-\textsc{neg}.\textsc{ex}.\textsc{cop}{}-\textsc{r} fox-\textsc{def} \textsc{poss}{}-house \textsc{poss}{}-\textsc{dir} \\
    \glt ‘He (the crocodile) did not come to check (his children) at the fox’s house until two months had passed.’  (natural data)
\z

\ea\label{ex:amr:54}Nanghem kihlenghlelengke nang labang hakpo phoun’e. \\
    \gll nang-hēm ki-hleng-Ce=\textbf{lengke} nang la=bang hak-po pho-\`{u}n-Ce\\
          2-house \textsc{nmlz}{}-reach-\textsc{neg}=until 2 \textsc{dem}=\textsc{gnr} bucket-\textsc{def} open-be.able-\textsc{neg}\\
    \glt ‘Before you reach home, you cannot open the bucket.’ (natural data)
\z

\ea\label{ex:amr:55} aro Herodpo thithelengke lalongsi dolo.\\
    \gll aro Herod{}-po thi{}-Ce=\textbf{lengke} lalong=si do-lo\\
         and \textsc{pn}:Herod-\textsc{m} die-\textsc{neg}=until there=\textsc{foc}.\textsc{r} stay-\textsc{r}\\
    \glt ‘and was there until the death of Herod’ (NT)
\z

\subsection{Negative lexicalizations} \label{sec:amr:2.5}

Negative lexicalizations include ‘not yet’, ‘not exist’, ‘not be of identity', ‘not know', ‘cannot, not be able to', ‘not want', ‘not talk', ‘not need', ‘not enough', ‘not like', and ‘not see' (cf. \citealt{Veselinova2013lexicalization}). Some of these lexicalizations are treated in other sections, e.g. ‘not exist’ and ‘not be of identity’ are treated in \sectref{sec:amr:2.3}, ‘not yet’ in \sectref{sec:amr:4.2} and ‘not want/not should’ is treated in \sectref{sec:amr:2.2.2}. 

\subsection{Other clausal negation constructions}\label{sec:amr:2.6}
In Amri Karbi, negation can be also achieved by derivational means. There is at least one predicate derivation -\textit{kel} that yields a negative meaning without the verbal negative suffix. Example \REF{ex:amr:35} shows the verb \textit{arju} ‘hear’ marked with the predicate derivation -\textit{kel} that gives meaning that the action denoted by the verb did not happen. In the example \REF{ex:amr:36} \textit{{}-kel} occurs with the verb \textit{{}-hlong} ‘get’. The negative predicate derivation -\textit{kel} is not a productive suffix and occurs mainly with the verbs that denote senses and perceptions.

\ea\label{ex:amr:35}Anangaipeke arjukelbakngo sal katiki along.\\
    \gll anang-ai-pe=ke arju-\textbf{kel}{}-bak-lo sal ki-tiki a-long\\
         3-mother-\textsc{f=top} hear-\textsc{neg.drv}{}-\textsc{indf}{}-\textsc{r} work \textsc{nmlz}{}-work \textsc{poss}{}-\textsc{loc}\\
    \glt ‘Its (baby’s) mother did not hear because of hard work.’  (natural data)
\z

\ea\label{ex:amr:36}Arlengpo angsong para wangkisun ayokke bachik isi hlongkelo.\\
    \gll arleng-po angsong para wang-ki-sun a-yok=ke bachik isi hlong\textbf{{}-kel-}lo\\
          man-\textsc{def} underneath para come-\textsc{nmlz}{}-come.down \textsc{poss}{}-\textsc{nsubj}=\textsc{top} basket one get-\textsc{neg}.\textsc{drv}{}-\textsc{r}\\
    \glt ‘When the man came down from the tree (up to where the basket was) and did not get one basket (i.e. ‘one basket got missing’).’ (natural data)
\z

\section{Non-clausal negation}\label{sec:amr:3}
\subsection{Negative and positive replies}\label{sec:amr:3.1}
The positive answer to a nominal predicate question is \textit{oik} ‘yes’. The positive answer to a verbal predicate question can either be \textit{oik} ‘yes’ or the verb itself, i.e. the verb is echoed in the answer, e.g. \REF{ex:amr:37} or both \textit{oik} ‘yes’ or the echoed verb. 
More research is needed to understand the reasons for this variation. Social and pragmatic factors, such as the need for clarity, politeness, or emphasis in conversation, may influence the choice of a positive answer in some languages (see \citealt{EnfieldEtAl2019}).
The negative answer to a question always corresponds to the type of predication being used in the question. When a predicate in the question is a verb, the negative existentials \textit{ingyong} or \textit{we} are used; with stative questions, the negative equational \textit{kalek} is used.
This finding aligns with \citegen{Veselinova2013neg.exist} observation that languages employ specific negative existentials, such as \textit{ingyong} or \textit{we}, for verbal predicates, while using forms like \textit{kalek} to negate stative predicates. \citeauthor{Veselinova2013neg.exist}'s research illustrates how these constructions serve as effective standalone answers, relying on contextual clues to convey their full meaning. This approach underscores the importance of predicate-specific negation for clarity and precision, a pattern also evident in the Amri Karbi language.
The predicate in the question can also be negated with the standard negation verbal suffix -C\textit{e} \REF{ex:amr:38}. Positive and negative answers to questions are asymmetrical in a way that there are two negatives used in the answers depending on the type of predicate, and there is only one positive \textit{oik} ‘yes’ for all predicate types. In addition, in the answers with the echoed verb, the TAM distinction is preserved in the positive replies and is not preserved in the negative replies. 

\ea\label{ex:amr:37}
    \begin{xlist}
    \exi{A:}Anangke nangai ma? \\ 
        \gll anang=ke nang-ai ma \\ 
              3=\textsc{top} 2-mother \textsc{q}\\
        \glt ‘Is she your mother?'
    \exi{B:} Oik. \\ 
        \gll oik\\
               yes \\
        \glt ‘Yes.' (elicited)
    \end{xlist}
\z

\ea\label{ex:amr:38}
    \begin{xlist}
    \exi{Q:}Hik inguman ma?\\ 
        \gll hik ingu-man ma\\ 
             dog bark-\textsc{hab} \textsc{q}\\
        \glt ‘Does a dog bark?'
    \exi{A:}  Oik./Inguman.\\ 
        \gll oik/ingu-man\\
               yes/bark\textsc{{}-hab}\\
        \glt ‘Yes, (it) barks.’ (elicited)
    \end{xlist}
    \z


\ea\label{ex:amr:39}
    \begin{xlist}
    \exi{Q:} Anangke nangai ma?\\ 
        \gll anang=ke nang-ai ma\\ 
              3=\textsc{top} 2-mother \textsc{q}\\
        \glt ‘Is she your mother?'
\newpage
    \exi{A:} Kalek.\\
        \gll  kalek\\
              \textsc{neg}.\textsc{eq}\\
        \glt ‘No.' (elicited)
    \end{xlist}
\z
    \ea\label{ex:amr:40}
    \begin{xlist}
    \exi{Q:} Hik inguman ma?\\ 
        \gll hik ingu-man ma\\ 
              dog bark-\textsc{hab} \textsc{q} \\
        \glt ‘Does the dog bark?'
    \exi{A:} Ingyong./We./Inguge.\\
        \gll  ingyong/we/ingu-Ce\\
              \textsc{negex}/\textsc{negex}/bark-\textsc{neg}\\
        \glt ‘No, it does not bark.’ (elicited)
    \end{xlist}
    \z
 

Negative and positive replies to negative questions are similar to replies to positive questions and a negated predicate is usually repeated, e.g. (\ref{ex:amr:41}--\ref{ex:amr:44}).

\ea\label{ex:amr:41} 
    \begin{xlist}
    \exi{Q:} Anangke nangai kalek ma?\\
        \gll  anang=ke nang-ai kalek ma\\ 
            3=\textsc{top} 2-mother \textsc{neg}.\textsc{eq} \textsc{q}\\
        \glt ‘Isn’t she your mother?'
    \exi{A:} Oik.\\
        \gll oik \\
               {yes}\\
        \glt ‘Yes.' (elicited)
    \end{xlist}
    \z
\ea\label{ex:amr:42} 
    \begin{xlist}
    \exi{Q:}  Hik inguge ma?\\
        \gll  hik ingu-Ce ma\\ 
            dog bark-\textsc{neg} \textsc{q}\\
        \glt ‘Doesn’t the dog bark?'
    \exi{A:}  Oik./Inguman.\\ 
        \gll oik/ingu-man\\
                yes/bark\textsc{{}-hab}\\
        \glt ‘Yes/it barks.’ (elicited)
    \end{xlist}
    \z


\ea\label{ex:amr:43}
    \begin{xlist}
    \exi{Q:} Anangke nangai kalek ma?\\ 
        \gll  anang=ke nang-ai kalek ma\\ 
             3=\textsc{top} 2-mother \textsc{neg}.\textsc{eq} \textsc{q}\\
        \glt ‘Isn’t she your mother?'
    \exi{A:}  Kalek.\\ 
        \gll kalek\\
              \textsc{neg}.\textsc{eq} \\
        \glt ‘No.' (elicited)
    \end{xlist}
\z
    \ea\label{ex:amr:44}
    \begin{xlist}
    \exi{Q:} Hik inguge ma?\\ 
        \gll  hik ingu-Ce ma\\ 
             dog bark-\textsc{neg} \textsc{q}\\
        \glt ‘Is the dog barking?'
    \exi{A:} Ingyong./We./Inguge.\\
        \gll  ingyong/we/ingu-Ce\\
                \textsc{negex/}\textsc{negex/}bark\textsc{{}-neg}\\
        \glt ‘No, it is not barking.’ (elicited)
    \end{xlist}
\z

\subsection{Negative indefinites and quantifiers} \label{sec:amr:3.2} 

Amri Karbi has three negative indefinites, listed in \tabref{tab:amr:3} along with the corresponding affirmatives. Out of the three, only \textit{kaita} ‘anybody/nobody’ occurs in both positive and negative clauses, other indefinites, \textit{iku} ‘nothing’ and \textit{jat} ‘never’, occur only in negative clauses. The pronoun \textit{kaita} ‘nobody’ is the only one that is constructed with the interrogative pronoun \textit{kai} ‘who’, which is attached with the additive suffix \textit{=ta} to form a universal (or non-specific) pronoun ‘anybody’. The other pronouns \textit{iku} ‘nothing’ and \textit{jat} ‘never’ have Assamese origin. The pronoun \textit{jat} ‘never’ may occur with the additive =\textit{ta} sometimes. The pronoun \textit{iku} can cooccur with the negative existential \textit{ingyong}. Negative pronouns are used together with clausal negation, e.g. (\ref{ex:amr:45}--\ref{ex:amr:47}).

\begin{table}
\centering
\begin{tabularx}{.95\textwidth}{lll}
\lsptoprule
 & Negative & Affirmative\\
\midrule
Person  & \textit{kaita} ‘anybody’ / ‘nobody’             & \textit{kaibak/inghonebak} ‘somebody’\\
Thing   & \textit{iku(si)/eku(si)/eko(si)} ‘nothing’    & \textit{penebak} ‘something’\\
Time    & \textit{jat/jatsi/jatta} ‘never’              & \textit{daknebak/kanebak} ‘sometime’\\
\lspbottomrule
\end{tabularx}
\caption{Indefinites and quantifiers}
\label{tab:amr:3}
\end{table}

\ea\label{ex:amr:45}  Kaita thektheto kaita ingyong holek.\\
    \gll \textbf{kai=ta} thek-Ce-to kai=ta ingyong holek\\
          who=\textsc{add} see-\textsc{neg}{}-\textsc{emph} who=\textsc{add} \textsc{negex} here\\
    \glt ‘Nobody will see, nobody is here.’ (elicited)
\z

\ea\label{ex:amr:46} Ikusi tikikelangta. \\
    \gll \textbf{iku=si} tiki-\textbf{C}\textbf{e}{}-lang=ta \\
         nothing=\textsc{foc}:\textsc{r} work-\textsc{neg}{}-yet=\textsc{add} \\
    \glt ‘Nothing has been done yet.’ (Mrs \& Mrs gathering)
\z

\ea\label{ex:amr:47} Israyel adak labadi jatsi thekhlonghleyong. \\
    \gll  israyel a-dak labadi \textbf{jatsi} th\'{e}k-hlong-Ce-ingyong\\
         \textsc{pn}:Israel \textsc{poss}{}-among like.this never see-get-\textsc{neg}{}-\textsc{negex}\\
    \glt ‘It was never so seen in Israel.’\footnote{The construction \textit{jatsi} verb\textit{{}-}\textsc{neg-negex} is found to be used only in the realis mood.} (NT)
\z

\subsection{Negative derivation}\label{sec:amr:3.3}
There is no dedicated strategy to express the caritive in Amri Karbi. Meanings of ‘without’ / ‘-less’ / ‘un-’ on nouns are expressed with the negative existentials \textit{we} or \textit{kangwe} ‘not.exist’ that follow a noun, e.g. \textit{lang (kang)we} 'water \textsc{negex/}' ‘waterless’ or literally ‘no water’, \textit{perik (kang)we} 'clothes \textsc{negex/}' ‘naked, unclothed’ or literally ‘no clothes’. These nouns can modify other nouns e.g. \REF{ex:amr:48}.

On deverbal adjectives that are nominalized, meanings of ‘without’ / ‘-less’ / ‘un-’ are expressed with the negative suffix -C\textit{e} on a verb, e.g. \textit{ki-krol-kre} \textsc{nmlz}{}-believe-\textsc{neg} ‘faithless’, literally ‘not believing’. These deverbal adjectives also modify nouns, e.g. \REF{ex:amr:49}. There is no other way of constructing a caritive from deverbal adjectives, so meanings like ‘unfaithful’ will have a similar form to ‘faithless’, i.e. \textit{kikrolkre}. However, since deverbal adjectives are verbs, they can extend their meaning with the help of predicate derivational suffixes and, depending on the context, can express different meanings of the caritive by the same negative verbal suffix -C\textit{e} attaching to a predicate derivation suffix. For example, if a modal predicate derivation suffix attaches to a deverbal negative adjective or a caritive, it means ‘the lack of ability’, e.g. \textit{ki-krol-un-e} ‘\textsc{nmlz}{}-believe-be.able-\textsc{neg’} ‘unfaithful’ / ‘faithless’, literally ‘not being able to believe’. 

\ea\label{ex:amr:48} Lang kangwe athem.\\
    \gll lang kangwe a-them\\
          water \textsc{negex} \textsc{poss}{}-place\\
    \glt ‘waterless region/place’ (NT)
\z

\ea\label{ex:amr:49} Kikrolkre arleng.\\
    \gll ki-krol-Ce arleng\\
          \textsc{nmlz}{}-believe-\textsc{neg} man\\
    \glt ‘unfaithful/faithless man’ (NT)
\z

\section{Other aspects of negation} \label{sec:amr:4}
\subsection{The scope of negation}\label{sec:amr:4.1} 
Amri Karbi has a large class of semantically rich predicate derivation suffixes that attach to the right of the verbal root in slot 5 of the verbal template. This type of suffix is found in other Tibeto-Burman/Trans-Himalayan languages and has been labeled in a variety of ways (see \citealt{Post2010}). I follow \citet{Post2010} in choosing the term \textit{predicate derivation suffixes}. There are about fifty-four predicate derivation suffixes found so far; these can be divided according to their subsequent functions: manner, result, direction, argument structure emphasis, aspect/aktionsart and others. When a verb with a predicate derivation suffix is negated, the negation has scope over either the predicate derivation, e.g. \REF{ex:amr:50}, or over the entire predicate, as in, e.g. \REF{ex:amr:51} \textit{thekdunde} ‘I did not know’. Each predicate derivation allows only two options for scope of negation, i.e. either only on the predicate derivation or over the entire verb. 

\ea\label{ex:amr:50}Borketke.\\
    \gll  bor-ket-Ce\\
         wrap-tight-\textsc{neg}\\
    \glt‘Wrap but not tightly.’ (elicited)
\z

\ea\label{ex:amr:51}potpotbadi chiporhelone ingyongne neke thekdunde potsi chininelo.\\
    \gll …pot{\textasciitilde}pot-badi chi-porhe-lo=ne ingyong=ne ne=ke thek-dun-Ce potsi chini-Ce-lo\\
          what{\textasciitilde}what-as \textsc{rr}{}-learn-\textsc{r}=\textsc{indf} \textsc{negex}=\textsc{indf} 1=\textsc{top} know-follow-\textsc{neg} reason know-\textsc{neg}{}-\textsc{r} \\
    \glt ‘I did not know whether they studied or not, because I do not know (I am uneducated).’  (natural data)
\z

\subsection{Negative polarity}\label{sec:amr:4.2}
Besides the negative polarity items listed in \sectref{sec:amr:3.2} there are three more negative polarity items that can be marked on the verb. The first one is the modal predicate derivation suffix -\textit{nang} that occurs only in the negation discussed in \sectref{sec:amr:2.2.2}. The second item is the aspect -\textit{lang} ‘yet’ that indicates that something has not happened yet. This aspect occurs only with standard negation, e.g. \REF{ex:amr:52}. The third negative polarity item is the subordinator =\textit{lengke} ‘until’, which was discussed in \sectref{sec:amr:2.4}.



\subsection{Marking of NPs in the scope of negation}\label{sec:amr:4.3}
It was observed in \sectref{sec:amr:2.3.3} that negative existential clauses cannot have unidentifiable subjects and the forms of negative possessive and existential clauses appear to be identical with inalienable subjects. 

\subsection{Reinforcing negation}\label{sec:amr:4.4}
Negation can be reinforced with the conjunction \textit{tene} ‘then’. The conjunction \textit{tene} is usually used with conditional clauses in the consequent part. Example \REF{ex:amr:56} contains an equative clause, where the predicate is a nominalized verb \textit{kasuru} ‘crying’ and \textit{tene} is used to reinforce negation and occurs after the negative equational \textit{kalek}. 

\ea\label{ex:amr:56} Kasuru kalek tene khoi.\\
    \gll ki-suru kalek tene khoi\\
          \textsc{nmlz}{}-cry \textsc{neg}.\textsc{eq} then friend\\
    \glt ‘This is not crying, friend.’  (natural data)
\z

\subsection{Negation, coordination and complex clauses} \label{sec:amr:4.5}
Subordination is usually marked by the nominalizer \textit{ki-} or the non-final suffix -\textit{si}. Complex clauses that have a positive subordinate clause and a negative main clause involve either the additive clitic =\textit{ta} or the topic clitic =\textit{ke} (depending on the intended meaning), marking a subordinate verb instead of the usual nominalization or non-final verb strategy. There is no dedicated coordinator for negative independent clauses which are coordinated by juxtaposition.



Subordinate clauses that can be translated as ‘even though’ are constructed by the additive =\textit{ta}, when =\textit{ta} marks a positive subordinate clause that is followed by a negative clause e.g. \REF{ex:amr:57}. 


\ea\label{ex:amr:57} Aro khoi puta pasiselo.\\
    \gll aro khoi pu=\textbf{ta} pasi-Ce-lo \\
          and friend say=\textsc{add} reply-\textsc{neg}{}-\textsc{r}\\
    \glt ‘Even though he called out "friend!", he did not reply.’ (natural data)
\z

The topic =\textit{ke} marks the quotative \textit{pu} instead of the usual non-final suffix -\textit{si} when the quotative \textit{pu} cooccurs with the negation, e.g. \REF{ex:amr:58}.

\largerpage
{\ea\label{ex:amr:58}  Manta alang neyok sunnot puke nepupeyong.\\
    \gll manta alang ne=yok sun-not pu=\textbf{ke} ne=pu-Ce-ingyong\\
         but 3 1=\textsc{nsubj} get.out-\textsc{imp} say=\textsc{top} 1=say-\textsc{neg}{}-\textsc{negex} \\
    \glt ‘But he didn’t tell me get out.’ (natural data)
\z}

Two or more negative independent clauses are coordinated by juxtaposition. Example \REF{ex:amr:59} shows three independent clauses appearing next to each without any conjunction. 

\ea\label{ex:amr:59}Atomo asarpe himepe ahem kangwe aret kangwe salkam katiki kaita kangwe.\\
    \gll a-tomo a-sar-pe hime-pe a-hem kangwe a-ret kangwe salkam ki-tiki kaita kangwe \\
         \textsc{poss}{}-story \textsc{poss}{}-be.old-\textsc{f} widow-\textsc{f} \textsc{poss}{}-house not.exist \textsc{poss}{}-field not.exist work-work \textsc{nmlz}{}-work nobody not.exist\\
    \glt ‘A long time ago there was a widow, she had no home, no field, nobody to work (with her).’ (natural data)
\z

\subsection{Miscellaneous aspects of negation}\label{sec:amr:4.6}
This section deals with miscellaneous aspects of negation. Negative transport (neg-raising), which is when a higher-clause negative is interpreted as negating a lower-clause predicate, does not occur in Amri Karbi.

Metalinguistic negation has not been found in the current Amri Karbi database. Since this type of negation is difficult to elicit, more natural data are needed in order to learn how metalinguistic negation is expressed in Amri Karbi. 

Contrastive negation occurs several times in the database and is exemplified in \REF{ex:amr:60}. Example \REF{ex:amr:60} shows that the first clause objects to the previous utterance, it is negated and the second clause that conveys the correct information follows right after it. The objecting element of the proposition is the O argument of the verb \textit{ap} ‘shoot’, which is marked with topic =\textit{ke} in the objection part and the focus =\textit{si} in the correction part of the proposition. 

{\ea\label{ex:amr:60}We arleng ayokke chiap’e tike ayoksi apmeta. \\
    \gll we arleng a-yok=ke chi-ap-Ce tike a-yok=si ap-ye=ta \\
          \textsc{negex} man \textsc{poss}{}-\textsc{nsubj}=\textsc{top} \textsc{rr}{}-shoot-\textsc{neg} tiger \textsc{poss}{}-\textsc{nsubj}=\textsc{foc} shoot-\textsc{fut}=\textsc{add}\\
    \glt  ‘(We) do not shoot people, (we) will shoot tiger only.’ (natural data)
\z}

Non-negative uses of negatives and expletive negation have not been attested in Amri Karbi.

\section{Summary} \label{sec:amr:5}

The negative strategies in Amri Karbi are summarized in \tabref{tab:amr:new}.

\begin{table}
\fittable{
    \begin{tabular}{ll}
\lsptoprule
Negative markers & Functions \\
\midrule
\itshape -Ce  & general verbal negation suffix: habitual, potential, future, past \\
\textsc{pst}-\textit{V-Ce}  & past  \\
\itshape -Ce-lo  & ‘no more’  \\
\itshape -Ce-bang  & ‘not now, but after some time’  \\
\itshape -nang-Ce  & ‘should not’  \\
\itshape -Ce-lengke  & ‘not  until’  \\
\itshape -Ce-lang  & ‘not yet’  \\
\itshape -Ce-te & negative question particle \\
\itshape - Ce-yong & {realis negation}\\
\itshape - Ce-yong-lo  &  {realis negation} \\
\itshape {ingyong} & existential, locative  \\ 
\itshape {we}  & existential, locative  \\
\itshape ingyong-lo  & existential ‘no more’  \\
\itshape ingyong & possessive  \\
\itshape kangwe  & possessive  \\
\itshape kalek  & equational  \\
\itshape -yek  & prohibitive  \\
\itshape -kel  & negation of verbs with meaning of sense and perception  \\

\lspbottomrule
    \end{tabular}
    }
    \caption{Summary of the negation strategies in Amri Karbi  }
    \label{tab:amr:new}
\end{table}


The negative suffix -C\textit{e} can occur as the only verbal operator, and the stem with the negative suffix can function as a general negative verb form for most TAM. The exception is the realis -\textit{lo}, which requires the reduced negative existential -\textit{yong} to form its negative. Besides, the realis -\textit{lo} may have future meaning in the negative. The habitual -\textit{man} is different from other forms in a way that it has only one negative form option, which is a basic \textsc{v}{}-\textsc{neg} form.
As a summary, \tabref{tab:amr:4} provides an overview of the main negative forms with the verb \textit{dam} ‘go’ that has the negative duplifix {}-\textit{de}. 





\begin{table}[t]
\begin{tabularx}{\textwidth}{p{23mm}Qp{25mm}}
\lsptoprule
{Affirmative} & {Negative} & {Functions}\\
\midrule
{\itshape dam-bo}\newline {‘will go’} & {\itshape dam-de-lo} {‘will go no more’}\newline {\itshape dam-de} {‘not going’}  &{Potential \textit{{}-bo}}\\
\midrule
{\itshape dame dam-ye} {‘will go’} & {\itshape dam-de-ye} {‘will not go’ (cooccurrence with conditional clauses)} & {Future -\textit{ye}}\\
&{\itshape dam-de} {‘will not go’} & \\
\midrule
{\itshape dam-bang}\newline  {‘going/will go’} & {\itshape dam-de} {‘not going’}\newline  {\itshape dam-de-bang} {‘not going now, but will go after some time’}  & {Irrealis} {\itshape {}-bang}\\
\midrule
{\itshape dam-lo} & {\itshape dam-de-yong}& {Realis \textit{{}-lo}}\\
& {\itshape dam-de-yong-o}& \\
& {\textit{dam-de-lo} ‘future sense’} & \\
\midrule
{\itshape dam-man} & {\itshape dam-de} & {Habitual -\textit{man}}\\
\midrule
{\itshape ki-dam} & {\itshape ki-dam-de} & {Past \textit{ki-}}\\
&{\itshape dam-de} &\\
&{\itshape dam-de-yong} &\\
\midrule
{\itshape lag(e)} & {\itshape nang-ne} & {Modal \textit{nang}\newline  ‘want, should’;}\\
{\itshape un} & {\itshape un-e} & {\textit{un} ‘be.able’}\\
\midrule
{\itshape do} & {\itshape ingyong} & {Existential} \\
& {\textit{ingyongo} ‘no more’} &\\
& {\itshape we} & \\
\midrule
{\itshape Ø} & {\itshape kalek}& {Equation}\\
&{\itshape kaleyong} & \\
\midrule
{\itshape cho-not} & {\itshape cho-nek} & {Command}\\
\lspbottomrule
\end{tabularx}
    \caption{Affirmative and negative paradigm}
    \label{tab:amr:4}
\end{table}

The negative form of an imperative has a dedicated construction, the prohibitive suffix -\textit{yek}. The prohibitive suffix -\textit{yek} is used for all types of imperatives, as a result the negative of imperative does not distinguish between formal vs. informal forms. The modal suffix -\textit{lag}, which is a borrowing, does not occur in the negative, and the older form -\textit{nang} is used in the negative instead. In the negative polar questions, the verb has the negative suffix -C\textit{e} and the question particle is =\textit{ma} as is in the positive polar questions. However, negative questions that involve the negative existential \textit{ingyong} use the question particle =\textit{te} instead of =\textit{ma}.
\largerpage
Equations are negated with the negative equative \textit{kalek} that does not inflect for tense. Other stative predicates - existential, possessive and locative are negated with the negative existential \textit{ingyong} and \textit{we} for existential clauses, \textit{ingyong} and \textit{kangwe} for possessive clauses and \textit{ingyong} and \textit{we} for locative clauses. The negative existential \textit{ingyong} and \textit{we} may inflect for TAM.


\section*{Abbreviations}
\begin{tabularx}{.45\textwidth}{lQ}
1 & first person \\
2 & second person \\
3 & third person \\
\scshape add & additive \\
\scshape assum & assumption \\
\scshape aug & augmentative \\
\scshape auto & autobenefactive \\
\scshape ben & benefactive \\
\scshape caus & causative \\
\scshape cis & cislocative \\
\scshape clf & classifier \\
\scshape conf & confirmation \\
\scshape cop & copula \\
\scshape decl & declarative \\
\scshape def & definite \\
\scshape dem & demonstartive \\
\scshape dim & diminutive \\
\scshape dir & directive \\
\scshape dist & distal demonstrative \\
\scshape drv & derivation \\
\scshape ee & elaborate expression \\
\scshape emph & emphatic \\
\scshape eq & equative, equational \\
\scshape ex & existential \\
\scshape excl & exclusive \\
\scshape f & feminine \\
\scshape foc & focus \\
\scshape fut & future \\
\scshape gnr & generic \\
\scshape hort & hortative \\
\scshape imp & imperative \\
\scshape incl & inclusive \\
\end{tabularx}
\begin{tabularx}{.5\textwidth}{lQ}%HC broader than normal
\scshape indf & indefinite \\
%\scshape intens & intensifier \\ not found above
\scshape interact & interactive \\
\scshape irr & irrealis \\
\scshape loc & locative \\
\scshape m & masculine \\
\scshape mal & malefactive \\
\scshape neg & negation \\
\scshape negex & negative existential \\
\scshape nfnl & non-final \\
\scshape nmlz & nominalizer \\
\scshape nsubj & non-subject \\
\scshape nt & new testament\\
\scshape o & object\\
\scshape pl & plural\\
\scshape pol & polite \\
\scshape poss & possessive \\
\scshape pot & potential \\
\scshape pred.deriv & predicate derivation \\
\scshape proh & prohibitive \\
\scshape pst & past \\
\scshape q & question \\
\scshape quot & quotative \\
\scshape red & reduplication \\
\scshape rep & reportative \\
\scshape repl & reply \\
\scshape r & realis \\
\scshape rr & reflexive-reciprocal \\
\scshape sg & singular \\
\scshape supp & supposition \\
\scshape tag & tag question \\
\scshape TAM & tense, aspect, mood \\
\scshape vb & verb \\
\end{tabularx}

\sloppy\printbibliography[heading=subbibliography,notkeyword=this]

\end{document} 
