\documentclass[output=paper,colorlinks=true, citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233966}
\author{Timur Maisak\affiliation{}}
\title{Negation in Andi}
\abstract{This paper presents an overview of negation in Andi, a Nakh-Daghestanian (East Caucasian) language of the Avar-Andic branch. I proceed from clausal negation to non-clausal negation, also covering various additional aspects of negation (the scope of negation, reinforcing negation, etc.), and conclude with diachronic observations and a summary of findings. Negation in the verb paradigm of Andi is symmetric in the sense that all categories have affirmative and negative forms. A peculiarity of Andi among the closely related languages is the expansion of one and the same negative marker into several domains of negation: the suffix which is the marker of standard negation is also used with non-indicative and non-finite verb forms, and is homophonous with the negative copula. Another peculiarity is the phenomenon of endoclisis, which is only attested in negative verb forms. Besides, Andi possesses a series of counterexpectation verb forms which seem to go back to a periphrastic construction involving some “inherently negative” auxiliary verb; forms with such semantics are not common in the languages of the family.

\keywords{Nakh-Daghestanian, Andi, endoclitics, counterexpectation, negation, negative copula}
}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 
%\shorttitlerunninghead{}%%use this for an abridged title in the page headers


\section{The language} \label{sec:and:1}

Andi (ISO: \href{https://iso639-3.sil.org/code/ani}{ani}, Glottocode: \href{https://glottolog.org/resource/languoid/id/andi1255}{andi1255}) belongs to the Andic group of the Avar-Andic branch of the Nakh-Daghestanian (East Caucasian) family. The Avar-Andic branch is one of the main subdivisions of the family alongside the Nakh, Tsezic, Dargwa and Lezgic language groups, and the Lak and Khinalug languages, each constituting its own branch. Alternatively, Avar-Andic can be united with Tsezic in a single Avar-Andic-Tsezic branch; see \citet[30--31]{DobrushinaEtAl2020} and \citet[88]{GanenkovMaisak2020} for details. The other Andic languages are Akhvakh, Bagvalal, Botlikh, Chamalal, Godoberi, Karata, and Tindi, all of which are minority languages spoken in the north-western part of mountain Daghestan (Russia) by a few hundred to a few thousand speakers each. By and large, all these languages remain unwritten and are not taught at schools, but only function as means of everyday communication in the respective villages. Avar, the closest relative of the Andic languages, is, in contrast, one of the major languages of Daghestan, with more than 700,000 speakers, an elaborate dialectal system and a literary standard. 

Estimates of the number of Andi speakers vary, but it is most plausible that the language is spoken by about 25,000 people, living in a few mountain villages of the Botlikhsky District (Andi, Gagatli, Rikvani, Ashali, Chanko, Zilo, and a few others) as well as in the lowlands of the Republic of Daghestan. Traditionally, the Andi varieties are subdivided into Upper Andi and Lower Andi, the latter (spoken in the villages Muni and Kvankhidatli) being mutually unintelligible with the former. In this sketch, I only use the data from Upper Andi varieties, in particular concentrating on the subdialects of the Andi and Rikvani villages, with some parallels from the Gagatli and Zilo varieties where relevant. In the examples, the English translations of the examples taken from the Gospel of Luke are from the New International Version of the Bible in English. In some cases, where the formulation of the verse in English differs a lot from the formulation of the Andi verse in the example, an additional more literal translation is given.

Although a number of grammatical sketches of Andi have been produced since the early 20\textsuperscript{th} century, the language remains underdescribed. \citet{Dirr1906}, \citet{Sulejmanov1957}, \citet{Cercvadze1965}, and \citeauthor{Salimov2010} (\citeyear{Salimov2010}, based on his 1968 dissertation) provide general overviews of Andi phonology and morphology. For this paper, I rely both on these sources and my own fieldwork in the village of Rikvani. Besides, I use the data from the available text corpus which includes a number of texts appended to the four aforementioned grammars (presenting different dialects), and the two books published in the 2010s which represent the first serious examples of literary production in Andi. The first of these is a collection of a dozen fairy tales from around the world, the second is a translation of the Gospel of Luke; although both are based on the variety of the largest Andi village (Andi proper), the Gospel of Luke in some aspects represents a mixture of multiple dialects. For the Zilo Andi data, I am indebted to Samira Verhees and other participants of the Zilo Andi documentation project (see also \citet{KayeEtAlFC}).

Like the Nakh-Daghestanian languages in general, Andic languages are morphologically ergative and left-branching with SOV basic word order, and have rich case systems (including many locative forms) and pervasive gender agreement (comprising three to six genders depending on the language and dialect). Verb systems in Andic are also fairly elaborate, with many synthetic as well as periphrastic forms. Andi is unusual among the closely related languages of this branch in that its present{}-tense copula \textit{i} (or \textit{ži}, \textit{ǯi}, depending on the dialect) has become restricted almost entirely to existential, possessive and locative contexts, thus representing a morphologically defective existential verb rather than a true copula. It is also not used as an auxiliary, which makes the core tense and aspect paradigm of Andi entirely\textsf{} synthetic.

The description below is based on the questionnaire provided by the editors \parencitetv{chapters/questionnaire} and follows its structure. I proceed from clausal negation in \sectref{sec:and:2} to non-clausal negation in \sectref{sec:and:3}, while \sectref{sec:and:4} covers various additional aspects of negation (in particular, the scope of negation, reinforcing negation, etc.). In \sectref{sec:and:4} I also analyse the phenomenon of endoclisis, which is only attested in negative verb forms in Andi. I conclude with a few diachronic observations and a summary of findings.


\section{Clausal negation} \label{sec:and:2}

\subsection{Standard negation} \label{sec:and:2.1}

The marker of standard negation in Andi is \textit{{}-sːu}. The same marker \textit{{}-sːu} is used with non-indicative and non-finite verb forms; moreover, the negative copula in Andi also has the form \textit{sːu}, and \textit{sːu} again is employed in constituent negation. In verb forms, I treat \textit{{}-sːu} as a bound marker (suffix), as discussed in more detail in \sectref{sec:and:4.6.3}.

The core of the Andi verb paradigm is synthetic, and all the tense-aspect-mood inflection in synthetic verb forms is suffixal. In the indicative, there are two morphological subsystems based on two verb stems. There is a perfective subsystem comprising the two main perfective past tenses, namely the aorist and the perfect (the latter is syncretic with the perfective converb). The present progressive also belongs to this subsystem, being derived from the same verb stem. The other verb stem serves as the base for forms belonging to the imperfective and future/prospective domains, namely the present, the present habitual, the counterexpectation present, the prospective and the intentional, as well as the future proper (which is syncretic with the imperfective participle). \tabref{tab:and:1} lists all the finite and the main non-finite synthetic forms of the verb ‘say’ in the Rikvani dialect, together with their negative counterparts. Other Upper Andi dialects may have minor differences in the form of the tense-aspect suffixes (e.g. the per\-fect/perfective converb suffix, which is \textit{{}-d} in Rikvani, \textit{\nobreakdash-dːu} in the Andi and Gagatli dialects, and \textit{{}-j} in Zilo). Note that the ‘not{}-yet’ present tense, i.e. the counterexpectation present, is included in the group of affirmative forms in \tabref{tab:and:1}, both because its meaning is not purely negative and because it has a regular morphological counterpart bearing the negative suffix \textit{{}-sːu}; on the semantics and structure of this form, see further \sectref{sec:and:2.6}.


\begin{table}
\begin{tabularx}{\textwidth}{Xll} 
\lsptoprule
& {Affirmative} & {Negative}\\
\midrule
\textbf{Aorist and related forms} &  & \\
\midrule
Aorist & \textit{ruʟi} & \textit{ruʟi-\textbf{sːu}}\\
Perfect (= Perfective Converb) & \textit{ruʟi{}-d} & \textit{ruʟi{}-d{}-\textbf{sːu}} / \textit{ruʟi{}-\textbf{č’igu}} \\
Perfective Participle & \textit{ruʟi{}-b} & \textit{ruʟi{}-\textbf{sːu}{}-b} \\
Masdar (action nominal) & \textit{ruʟi{}-r} & \textit{ruʟi{}-\textbf{sːu}{}-r} \\
Progressive Present & \textit{ruʟ-rado} & \textit{ruʟ-rado-\textbf{sːu}} \\
Imperative & \textit{ruʟ-o} & –\\
\tablevspace
\textbf{Infinitive and related forms} &  & \\
\midrule 
Infinitive & \textit{ruʟ-du} & \textit{ruʟ-du-\textbf{sːu}} \\
Present & \textit{ruʟ{}-duq} & \textit{ruʟ{}-duq{}-\textbf{sːu}} \\
Prospective & \textit{ruʟ-dulq} & \textit{ruʟ-dulq-\textbf{sːu}} \\
Intentional & \textit{ruʟ{}-duʁojd} & \textit{ruʟ{}-duʁojd{}-\textbf{sːu}} \\
Habitual Present & \textit{ruʟ-do} & \textit{ruʟ-do-\textbf{sːu}} \\
Prohibitive & – & \textit{ruʟ-do-\textbf{sːub}} \\
Future (= Imperfective Participle) & \textit{ruʟ-dja} & \textit{ruʟ-do-\textbf{sːja}} \\
Counterexpectation Present & \textit{ruʟ{}-daʁid} & \textit{ruʟ{}-daʁid{}-\textbf{sːu}} \\
\lspbottomrule
\end{tabularx}
\caption{Affirmative and negative forms of the verb ‘say’ in the Rikvani dialect}
\label{tab:and:1}
\end{table}

\largerpage
In most verb forms, the marking of negation is straightforward: \textit{{}-sːu} is simply added to an affirmative verb form, cf. the aorist (perfective past) \textit{ruʟi} ‘said’ {\textasciitilde} \textit{ruʟi-}\textbf{\textit{sːu}} ‘didn’t say’, the perfect \textit{ruʟi-d} ‘has said’ {\textasciitilde} \textit{ruʟi-d-}\textbf{\textit{sːu}} ‘hasn’t said’, the progressive present \textit{ruʟ-rado} ‘is saying’ {\textasciitilde} \textit{ruʟ-rado-}\textbf{\textit{sːu}} ‘isn’t saying’ (R),\footnote{In example sources the first letter stands for the dialect represented: “A” for Andi, “G” for Gagatli, “R” for Rikvani and “Z” for Zilo. “\citetitle{Tales}” refers to the collection of world fairy tales \citep{Tales}, and “\citetitle{Luke}” refers to the translation of the Gospel of Luke. If a text/grammar source is not indicated, the example results from elicitation.} etc. The following examples illustrate the use of two indicative forms with regular negation marking, the aorist in (\ref{ex:and:1}–\ref{ex:and:2}) and the present habitual in (\ref{ex:and:3}–\ref{ex:and:4}).

\ea 
    \label{ex:and:1}
    \gll  hege{}-li{}-ʟi riʟ’i \textbf{ikʷon}...\\
          \textsc{dem-obl.pl.fn{}-gen} meat eat.\textsc{aor}\\
    \glt [I killed two deer.] ‘I ate their meat...’ (A; \citetitle{Tales})
\z 

\ea
    \label{ex:and:2}
    \gll \textbf{ikʷon{}-sːu} den du{}-j{}-il boš{}-ul\\
         eat.\textsc{aor-neg} \textsc{1sg} \textsc{2sg}{}-\textsc{gen.iii.pl-pl} offspring{}-\textsc{pl}\\
    \glt [The wolf says to the goat:] ‘I didn’t eat your kids.’ (A; \citetitle{Tales})
\z

\ea
    \label{ex:and:3}
    \gll di{}-‹w›o \textbf{c’in{}-no} men i‹m› hek’a=bolo!\\
          \textsc{1sg}-\textsc{afft<m>} know-\textsc{hab} \textsc{2sg} who<\textsc{h>} man=\textsc{q.indir}\\
    \glt ‘I know who you are!’ (A; \citetitle{Luke} 4:34)
\z 

\ea\label{ex:and:4} 
    \gll di-‹w›o he{}-w \textbf{c’in{}-no{}-su}, jecːi...  \\ 
          \textsc{1sg}-\textsc{afft<m>} \textsc{dem-m} know-\textsc{hab-neg} sister \\
    \glt ‘I don’t know him, sister...’ (A; \citetitle{Luke} 22:57)\footnote{The transliteration of intensive (geminate) consonants in the Gospel of Luke translation is not always consistent, so the negative marker \textit{{}-sːu} is sometimes written there as \textit{{}-su}. In examples, I have kept the transliteration as in the original.}  
\z 


Thus, negation in the Andi verb paradigm is overwhelmingly symmetric in \citegen{Miestamo2005} terms.  There are, however, negative verb forms which are not derived from the corresponding affirmative ones by simply adding \textit{{}-sːu} on the right edge, after the inflectional suffix of the corresponding tense-aspect-mood category. First, in a few non-finite forms \textit{{}-sːu} precedes the inflectional suffix rather than following it (see \sectref{sec:and:2.4}). Second, in two verb forms the negation marker \textit{{}-sːu} is historically present, but is fused with another marker. The forms in question are the negative future in \textit{-dosːja} and the prohibitive (negative imperative) in \textit{\nobreakdash-dosːub}, which are not derived from their affirmative counterparts. Historically, they both probably go back to the habitual in \textit{\nobreakdash-do}. The affirmative future in \textit{{}-dja} seems to be derived from the habitual by means of the suffix *\textit{{}-ja}, which ousts the final vowel of the habitual (\textit{{}-dja} < *\textit{{}-do-ja}). In the negative future, we see the same habitual marker \textit{\nobreakdash-do} before the negation marker \textit{{}-sːu} (reduced to \textit{{}-sː}) and the same suffix *\textit{{}-ja} follows (\textit{{}-dosːja} < *\textit{{}-do-sːu{}-ja}). On the structure of the prohibitive, see \sectref{sec:and:2.2}. Both the negative future and the prohibitive display asymmetry of the \mbox{A/Cat/TAM} type, i.e. affecting the marking of tense-aspect-mood \citep[116]{Miestamo2005}.

The third exception is different in nature, and is the only source of paradigmatic asymmetry in Andi negation, as the relation between the affirmative form and the negative form here is not one-to-one. In the perfect, which is syncretic with the perfective converb, alongside the expected form in \textit{{}-sːu} there exists a form involving a special negation marker \textit{\nobreakdash-č’igu}, which does not occur elsewhere in the language and is unrelated to \textit{{}-sːu} historically. This second, “irregular” variant of the negative perfect / perfective converb is produced by adding \textit{\nobreakdash-č’igu} directly to the perfective stem, without the perfect ending \textit{{}-d}, cf. \textit{ruʟi-d} ‘has said’ {\textasciitilde} \textit{ruʟi-}\textbf{\textit{č’igu}} ‘hasn’t said’. Thus, \textit{\nobreakdash-č’igu} simultaneously expresses negation and the corresponding tense-aspect-mood value. On the whole, the “irregular” negative form in \textit{\nobreakdash-č’igu} seems to be the neutral one for both the perfect and the perfective converb, while the “regular” one is very marginal. In the available written texts of Andi (“\citetitle{Tales}” and “\citetitle{Luke}”), the regular form in \textit{{}-dːu-sːu} is only found 5 times (as opposed to more than 170 occurrences of the form in \textit{\nobreakdash-č’igu}), and only in subordinate clauses with the adverbializer \textit{\nobreakdash-gu}, i.e. with a general meaning ‘without having done’ (see \sectref{sec:and:4.6.1}).\footnote{\citet{Kibrik1985} puts forward a hypothesis on the choice between \textrm{\textit{{}-sːu}} and \textrm{\textit{\nobreakdash-č’igu}} depending on the meaning of the (finite) perfect: he suggests that the regular negation is used with the perfect/resultative meaning of the form, while \textrm{\textit{\nobreakdash-č’igu}} appears only with the indirective evidential meaning. However, this conclusion is not supported by the data we have. See also \citet[201--209]{Verhees2019} on the structure and use of the perfective converb (in her terms, general converb) in Andi.}

Examples (\ref{ex:and:5}–\ref{ex:and:7}) illustrate the use of the affirmative and negative future and the affirmative and “irregular” negative perfect.

\ea
    \label{ex:and:5}
    \gll  den \textbf{w{}-uɢi{}-dja} men=nogu, mijarek’a!\\
          \textsc{1sg} \textsc{m}{}-come-\textsc{fut} \textsc{2sg}=\textsc{com} lord\\
    \glt ‘I will follow (lit. come with) you, Lord!’ (A; \citetitle{Luke} 9:61)
\z 
        
\ea 
    \label{ex:and:6} 
    \gll den \textbf{j{}-iɢi{}-dosːja} bisːi{}-ʟi{}-kːu se{}-w{}-gulo{}-šːu{}-j\\
          \textsc{1sg} \textsc{f}{}-come-\textsc{fut.neg} \textsc{2pl{}-inter-ela} one-\textsc{m-indf-obl.m-dat}\\
    \glt ‘I will not marry (lit. come to) any of you.’ (A; \citetitle{Tales})
\z 

\ea
    \label{ex:and:7}
    \gll mašenʲka{}-di hinc’ːu \textbf{t’ik’ešo{}-dːu}– se{}-w{}-gulo \textbf{w{}-ot’on{}-č’igu}\\
          Mashenka-\textsc{erg} door knock-\textsc{prf} one-\textsc{m-indf} \textsc{m}{}-talk-\textsc{prf.neg}\\
    \glt ‘Mashenka knocked on the door – no one answered.’ (A; \citetitle{Tales})
\z 

As already mentioned in \sectref{sec:and:1}, there are no periphrastic forms in Upper Andi with a present-tense copula as auxiliary. There are, however, a few periphrastic constructions employing a regular verb \textit{bik’udu} ‘be, become, happen’ or \textit{biʁidu} ‘be, stop, stand’. The latter verb is employed in the Rikvani dialect only with the meaning ‘come to a stop, stand’, while at least in the Andi, Gagatli and Zilo dialects it has evolved into a more general existential verb, widely used as an auxiliary (see \citealt{VerheesRochant2017} for details). As an auxiliary it is postposed and preferably occurs in one of the past tenses, e.g. the aorist or the perfect. Most common types of periphrastic constructions are those based on the perfect (perfective converb) or the progressive present form of a lexical verb: the combination of the former with a past tense auxiliary yields a pluperfect (‘had done’), while the combination of the latter with a past tense auxiliary is an imperfect (‘was doing, used to do’).

When a periphrastic form is negated, in the most neutral case the negation is found on the auxiliary verb, as in \REF{ex:and:8} with the periphrastic future perfect (perfective converb + ‘be’ in the future tense) or \REF{ex:and:9} with the periphrastic past habitual (imperfective participle + ‘be’ in the perfect tense). 

\ea
    \label{ex:and:8}
    \gll  \textbf{b{}-eχukːu{}-dːu} \textbf{b{}-ik’u{}-dosːja}=bu!\\
          \textsc{iv}{}-be.late-\textsc{cvb} \textsc{iv}{}-be-\textsc{fut.neg=ptcl}\\
    \glt ‘There’s no hurry!’ (lit. ‘It won’t be late') (A; \citetitle{Tales})
\z 

\ea\label{ex:and:9} 
    \gll  hege{}-li{}-di oχːodːoro{}-l{}-la{}-kːu heč’eɢido{}-l{}-lo=ɢilo se{}-b{}-gulo \textbf{i{}-dja} \textbf{b{}-ik’ʷo{}-č’igu}\\ 
          \textsc{dem-obl.pl.fn{}-erg} morning-\textsc{obl}{}-\textsc{sup{}-ela} at.night{}-\textsc{obl}{}-\textsc{sup.lat}=until  one-\textsc{iv-indf} do-\textsc{ipfv.ptcp} \textsc{iv}{}-be-\textsc{prf.neg}\\
    \glt ‘They just lounged all day long.’ (lit. ‘They were not doing anything from morning till night.’) (A; \citetitle{Tales})
\z 

At the same time, the marking of negation on the lexical verb is also attested, as can be seen in examples \REF{ex:and:10} and \REF{ex:and:11} with the same constructions: in \REF{ex:and:10}, it is the perfective converb which bears negation, and in \REF{ex:and:11} it is the imperfective participle.

\ea
    \label{ex:and:10} 
        \gll ama q’adi šu{}-b=gu zamana{}-l{}-di \textbf{razi{}-ɬi{}-č’igu} \textbf{w{}-uk’o{}-dːu}.\\
          but judge good-\textsc{iv=ints} time{}-\textsc{obl}{}-\textsc{erg} content-\textsc{vbz-cvb.neg} \textsc{m}{}-be-\textsc{prf}\\
         \glt ‘But for some time the judge did not agree’ (A; \citetitle{Luke} 18:4)
\z

\ea\label{ex:and:11} 
    \gll  amma en{}-šːu{}-b=gu ulka{}-l{}-ʟ{}-ol adam{}-i{}-‹w›o  hege{}-w \textbf{hilu{}-dosja} \textbf{w{}-uk’o{}-dːu}\\ 
           but self-\textsc{obl.m-gen.iv=ints} country{}-\textsc{obl}{}-\textsc{gen-pl} people{}-\textsc{obl}{}-\textsc{afft<m>}  \textsc{dem-m} love-\textsc{ipfv.ptcp.neg} \textsc{m}{}-be-\textsc{prf}\\
    \glt ‘But his subjects did not like him.’ lit. 'But people of his country did not like him' (A; \citetitle{Luke} 19:14)
\z 

\subsection{Negation in non-declaratives} \label{sec:and:2.2}

In the formation of the negative imperative (prohibitive) Andi follows a strategy common to most other Nakh-Daghestanian languages in that the prohibitive is not derived directly from the imperative. In Andi, as in Andic languages in general, transitive and intransitive verbs have different imperative forms, the former in \textit{{}-o} and the latter in \textit{{}-b}. Intransitive imperatives are derived from the perfective stem and are syncretic with perfective participles (e.g. the form \textit{bik’o-b} can be both the imperative ‘be, become!’ and the participle ‘which became, which happened’). The prohibitives of transitives and intransitives, meanwhile, are derived uniformly and belong morphologically to the future/imperfective subsystem. Most probably, the prohibitive marker \textit{{}-dosːub} is historically derived from the habitual in \textit{{}-do} followed by the negative \textit{{}-sːu} and the participial suffix \textit{{}-b} \mbox{(\textit{{}-dosːub}} < *\textit{{}-do-sːu-b}). Thus, the prohibitive of the verb ‘be, become’ has the form \textit{bik’u-dosːub}; see also examples in \REF{ex:and:12}.

\ea
    \label{ex:and:12}
    \gll hinc’ːu \textbf{arχin{}-nosːub}, se{}-w{}-gulo hinni=lo \textbf{eldːosːub}\\
          door open-\textsc{proh} one-\textsc{m-indf} inside=\textsc{add} let+\textsc{proh} \\
    \glt ‘Don’t open the door, don’t let anyone in.’ (A; \citetitle{Tales})
\z 

Both yes/no-questions and wh-questions in Andi employ pairs of question enclitics, one used with the predicate in the aorist and the other with the predicate in any other form \citep{Maisak2020}. In polar questions, the question markers are by default cliticized to the verb, but they can follow the focus of a question in case of narrow (constituent) focus. In wh-questions, the question markers are hosted by the wh-phrase. The use of question markers does not generally influence the marking of negation, cf. a wh-question in \REF{ex:and:13} with the negative aorist.

\ea
    \label{ex:and:13}
    \gll  e‹b›i=di bisːi{}-di \textbf{č’erʟ’i{}-sːu}?\\
          what<\textsc{iv>=wh} \textsc{2pl{}-erg} divide.\textsc{aor-neg} \\
    \glt ‘Why are you quarreling?’ (lit. ‘What haven’t you divided among yourselves?’) (A; \citetitle{Tales})
\z

At the same time, it is common for negative polar questions to be employed with purely rhetorical function, when the speaker is sure that the addressee in fact knows the answer well, and a reply is not expected, cf. (\ref{ex:and:14}–\ref{ex:and:16}). 

\ea
    \label{ex:and:14}
    \gll du{}-‹b›o den c’inni{}-r \textbf{sːu=le}?\\
          \textsc{2sg}{}-\textsc{afft<iii>} \textsc{1sg} know-\textsc{prog} \textsc{cop.neg=q}\\
    \glt [The mouse asks the lion:] ‘Don’t you recognize me?’ [I am the one whose life you spared.] (A; \citetitle{Tales}) 
\z
         
\ea
    \label{ex:and:15}
    \gll men mesihi \textbf{sːu=le}?\\
          \textsc{2sg} messiah \textsc{cop.neg=q}\\
    \glt [One of the criminals asked:] ‘Aren’t you the Messiah?’ [Save yourself and us!] (A; \citetitle{Luke} 23:39) 
\z

\ea
    \label{ex:and:16} 
    \gll du{}-‹b›o \textbf{c’in{}-no{}-su=le} hila=si{}-šːu{}-b{}-ul amr{}-odul\\
          \textsc{2sg}{}-\textsc{afft<iv>} know-\textsc{hab-neg=q} above=\textsc{attr-obl.m-gen.iv-pl} order-\textsc{pl}\\
    \glt ‘Don’t you know the commandments of the Lord (lit. the one above)?’ (A; \citetitle{Luke} 18:20)
\z 

This use of negative polar questions is frequent in the translation of the Gospel of Luke, where negative questions are used to translate many sentences which are rendered in the affirmative declarative in e.g. English or Russian versions of the Gospel, as in \REF{ex:and:17} and \REF{ex:and:18}.



\ea
    \label{ex:and:17} 
        \gll ja, χʷažain, hede{}-š{}-č’u ho{}-ʟ’ol=lo hoc’o{}-gu misidi \textbf{i=sːu=le}\\
    \textsc{voc} landlord \textsc{dem-obl.m-cnt} \textsc{dem-mnr=add} ten-\textsc{num}  gold \textsc{cop=cop.neg=q}\\
    \glt ‘Sir, he already has (lit. hasn’t he?) ten golden coins as it is.’ (A; \citetitle{Luke} 19:25)
\z 

\ea
    \label{ex:and:18}
    \gll talihi=logu=s{}-ol \textbf{sːu=le} moč’i ihi{}-su{}-b{}-ul, he-w-ul w{}-oχːonɬi{}-su{}-b{}-ul ješi{}-l!\\
          luck=\textsc{com=attr-pl} \textsc{cop.neg=q} child do-\textsc{neg-pfv.ptcp-pl}  \textsc{dem-m-pl} \textsc{m-}suck.\textsc{caus-neg-pfv.ptcp-pl} girl-\textsc{pl}\\
    \glt ‘Blessed are (lit. aren’t?) the women that have never bore children and never fed them!’ (A; \citetitle{Luke} 23:29)
\z 


\subsection{Negation in stative predications} \label{sec:and:2.3}

Affirmative stative predications do not behave in a uniform way. Although the affirmative present copula \textit{i} (dialectal variants \textit{ži}, \textit{ǯi}) occasionally appears in contexts of equation, proper inclusion or attribution, these types of stative predication normally do not include a copula (\ref{ex:and:19}–\ref{ex:and:21}). On the other hand, in existential, locative and possessive predications the copula is regularly present (\ref{ex:and:22}–\ref{ex:and:24}). 

\ea 
    \label{ex:and:19} 
    \gll hono-w \textbf{i} žuhut’-a-w pačːahi\\
          \textsc{dem-m} \textsc{cop} Jew-\textsc{obl.pl-gen.m} king\\
    \glt ‘This is the king of the Jews.’ (A; \citetitle{Luke} 23:38)
\z

\ea
    \label{ex:and:20}
    \gll hege-w iši-ʟi mijarek’a\\
          \textsc{dem-m} \textsc{1pl.excl-gen} landlord\\
    \glt [Domestic animals about Man:] ‘He is our owner.’ (A; \citetitle{Tales}) 
\z 

\ea
    \label{ex:and:21}
    \gll den=no ila=lo zolo miskin{}-ol\\
          \textsc{1sg}=\textsc{add} mother=\textsc{add} very poor{}-\textsc{pl}\\
    \glt ‘Me and my mother are very poor.’ (G; \citealt{Salimov2010}: 222) 
\z

\ea
    \label{ex:and:22}
    \gll sor{}-da{}-j reč’ː{}-obil \textbf{i}, šekm{}-a{}-j bešem{}-il \textbf{i}\\
          fox-\textsc{obl.pl-dat} cave-\textsc{pl} \textsc{cop} bird-\textsc{obl.pl-dat} nest-\textsc{pl} \textsc{cop}\\
    \glt ‘For foxes there are holes and for birds there are nests.’ (A; \citetitle{Luke} 9:58)
\z

\ea
    \label{ex:and:23} 
    \gll di-w wošo armi-l-la \textbf{i}\\
         \textsc{1sg}-\textsc{gen.m} son army-\textsc{obl-sup.ess} \textsc{cop}\\
    \glt ‘My son is in the army.’ (A; \citealt[58]{Alisultanova2010})
\z 

\ea
    \label{ex:and:24} 
    \gll di-b č’e-gu mašina \textbf{ži}\\
         \textsc{1sg}-\textsc{gen.iv} two-\textsc{num} car \textsc{cop} \\
    \glt ‘I have two cars.’ (R)
\z

Under negation, the situation is different, as all these types make use of the negative copula \textit{sːu} (identical in form with the standard verbal negation marker, as already mentioned above).\footnote{The situation when a standard negation marker and a negative existential are formally identical, but one is a free form and the other is bound, is not unique cross-linguistically, see \citet[113--114]{Veselinova2013neg.exist} on the Dravidian language Kannada. Cases of such morphological difference are not very numerous in the languages of the world, e.g. in \citegen[116]{Veselinova2013neg.exist} sample of 95 languages there are only 4 languages that demonstrate this type of syncretism; at the same time, they all belong to different genealogical stocks.} Thus \REF{ex:and:25} illustrates the context of equation, \REF{ex:and:26} illustrates attribution, \REF{ex:and:27} describes location, \REF{ex:and:28} describes possession and \REF{ex:and:29} describes existence. There does not seem to be any difference in the negation marking depending on temporary vs. permanent property assignment, or generic vs. specific predicates.

\ea\label{ex:and:25}
    \gll hajman boc’o \textbf{sːu}\\
          sheep wolf \textsc{cop.neg}\\
    \glt ‘A sheep is not a wolf.’ (A; \citetitle{Tales})
\z 

\ea\label{ex:and:26}
    \gll išːil he-w-ul=ɢʷa burt’-ol \textbf{sːu}\\
          \textsc{1pl.excl} \textsc{dem-m-pl=degr} rich-\textsc{pl} \textsc{cop.neg}\\
    \glt ‘We are not so rich as they are.’ (A; \citetitle{Tales})
\z 

\ea
    \label{ex:and:27}
    \gll ho{}-ɬːu hege{}-w \textbf{sːu}. hege{}-w w{}-uč’o{}-b{}-ɬu{}-kːu w{}-uč’ːi\\
          \textsc{dem-loc} \textsc{dem-m} \textsc{cop.neg} \textsc{dem-m} \textsc{m}{}-die-\textsc{pfv.ptcp-loc{}-ela} \textsc{m}-rise.\textsc{aor}\\
    \glt ‘He is not here. He has risen (lit. from where he died).’ (A; \citetitle{Luke} 24:6)
\z 

\ea\label{ex:and:28} 
    \gll hono{}-w moč’i{}-šːu{}-r χuča \textbf{sːu}\\
          \textsc{dem-m} child\textsc{{}-obl.m}{}-\textsc{gen.v} book \textsc{cop.neg}\\
    \glt ‘This boy doesn’t have a book.’ (R)
\z 

\ea\label{ex:and:29} 
    \gll insan{}-šː{}-č’u sːihir dan se{}-b{}-gulo \textbf{sːu}!\\
          person\textsc{{}-obl.m}{}-\textsc{cnt} sly thing one-\textsc{iv-indf} \textsc{cop.neg}\\
    \glt ‘There is no one as sly as Man!’ (A; \citetitle{Tales})
\z 

From the negative copula \textit{sːu} a participle \textit{sːu-b} is derived, see \REF{ex:and:30}. The corresponding converbial function can be fulfilled by the form \textit{sːu-gu}, featuring the regular adverbializer \textit{{}-gu}. The form \textit{sːu-gu} is also found in contexts of exclusion (‘except for’, ‘only’), see \sectref{sec:and:4.6.1}.

\ea\label{ex:and:30}
    \gll di{}-č’u soru \textbf{sːu{}-b} hek’a\\
          \textsc{1sg}-\textsc{cnt} together \textsc{cop.neg-ptcp} man \\
    \glt ‘Whoever is not with me [is against me].’ (lit. ‘The man who is not with me.’)  (A; \citetitle{Luke} 11:23)
\z 

\subsection{Negation in non-main clauses} \label{sec:and:2.4}

As in Nakh-Daghestanian languages in general, complementation and subordination in Andi involves the use of non-finite forms – converbs, participles, the infinitive and the action nominal traditionally labelled “masdar”. All of these forms employ the standard negator \textit{\nobreakdash-sːu}. The only exception is the perfective converb, which is syncretic with the finite perfect tense; although a standard negative form does exist, this generally uses the special negative morpheme \textit{\nobreakdash-č’igu} discussed in \sectref{sec:and:2.1} above, cf. \REF{ex:and:31} with a converbial dependent clause expressing temporal precedence.

\ea
    \label{ex:and:31} 
    \gll hila=si{}-šːu{}-b pačːahi{}-ɬir [adam{}-i{}-‹b›o \textbf{haɢo{}-č’igu}] b-iɢi-dja\\
         above=\textsc{attr-obl.m-gen.iv} king-\textsc{abstr} {\db}people\textsc{{}-obl}{}-\textsc{afft<iv>} see-\textsc{cvb.neg}  \textsc{iv}{}-come-\textsc{fut} \\
    \glt ‘The kingdom of God will come unseen by the people (lit. without the people having seen it).’ (A; \citetitle{Luke} 17:20)
\z 

There are two dedicated participle forms: the perfective participle in \textit{{}-b} (which happens to be syncretic with the imperative of intransitive verbs) and the imperfective participle in \textit{\nobreakdash-dja} (which is syncretic with the future tense). While the latter has an irregular negative form marker \textit{{}-dosːja} identical to that of the future (see \sectref{sec:and:2.1} above), in the perfective participle the negator \textit{\nobreakdash-sːu} precedes the participial suffix rather than following it, cf. \textit{haɢo-b} ‘(the one) that saw’ {\textasciitilde} \textit{haɢo-}\textbf{\textit{sːu}}\textit{{}-b} ‘(the one) that didn’t see’. Examples \REF{ex:and:32} and \REF{ex:and:33} show the use of participles in prenominal relative clauses.

\ea
    \label{ex:and:32}
    \gll [...] hit’on{}-dːu {\ob}onšːi{}-gulo b{}-it’i– b{}-it’i boc’o=gu \textbf{haɢo-sːu-b}{\cb} nuf-nuf{}-o{}-di=lo \\
          {} say-\textsc{prf} {\db}once-\textsc{indf} \textsc{iii}{}-alive \textsc{iii}{}-alive wolf=\textsc{ints} see-\textsc{neg-pfv.ptcp} Nuf-Nuf\textsc{{}-obl}{}-\textsc{erg=add} \\
    \glt ‘[...] said Nuf-Nuf, who had never seen a living wolf.’ (A; \citetitle{Tales})
\z 

\ea
    \label{ex:and:33}
    \gll {\ob}dunja{}-l{}-la q’ir=lo \textbf{b{}-ižun{}-nosːja}{\cb} rihi\\
          {\db}world\textsc{{}-obl}{}-\textsc{sup.ess} corn=\textsc{add} \textsc{iv}{}-grow-\textsc{ipfv.ptcp.neg} time \\
    \glt ‘In the times when rice didn’t yet grow on earth’ (A; \citetitle{Tales})
\z 

The conditional form in \textit{{}-bor}, which is historically derived from the perfective participle, has the same position for the negative suffix before the conditional mood suffix,\footnote{Conditional forms (the ones used in protases of conditional sentences) are traditionally described in Nakh-Daghestanian languages as belonging to the group of specialized converbs, as they are almost totally restricted to subordinate clauses.} cf. \textit{bik’o-bor} ‘if (it) is’ {\textasciitilde} \textit{bik’o-}\textbf{\textit{sːu}}\textit{{}-bor} ‘if (it) is not’, see also \REF{ex:and:34}.

\ea
    \label{ex:and:34} 
    \gll di{}-b t’alab \textbf{t’obannoɬi{}-sːu{}-bor}, du{}-č’u gal{}-la mijar=lo r{}-ik’u{}-dosːja \\
          \textsc{1sg}-\textsc{gen.iv} demand be.fulfilled.\textsc{caus-neg-cond} \textsc{2sg}{}-\textsc{cnt} shoulder-\textsc{sup.ess} head=\textsc{add} \textsc{v}{}-be-\textsc{fut.neg} \\
    \glt ‘If you do not satisfy my requirement, it will cost you dear.’ (lit. ‘your head won’t remain on your shoulders’) (A; \citetitle{Tales})
\z 

Another non-finite form with the same placement of a negator before the inflectional marker is the verbal noun (“masdar”) in \textit{{}-r}, cf. \textit{bik’o-r} ‘being, existence’ {\textasciitilde} \textit{bik’o-}\textbf{\textit{sːu}}\textit{{}-r} ‘not being, non-existence’. The use of the masdar, which inflects for case and number like a regular noun, can be seen in \REF{ex:and:35}.

\ea
    \label{ex:and:35} 
    \gll tawbu i{}-du b{}-ukːu{}-dosja hoč’oc’o=lo hoč’o{}-gu  muʔmin \textbf{w-uk’o-su-r-o-qi}, se-w q’osuni-w \textbf{kant’uni{}-r{}-o{}-č’u{}-kːu} \\
          repentance do{}-\textsc{inf} \textsc{iv}{}-need\textsc{{}-ipfv.ptcp.neg} ninety=\textsc{add} nine-\textsc{num}  faithful \textsc{m}{}-be\textsc{{}-neg-msd-obl-apud} one-\textsc{m} sinner{}-\textsc{m} repent-\textsc{msd{}-obl{}-cnt-ela} \\
    \glt ‘[There will be more rejoicing in heaven] over one sinner’s repentance than over the ninety-nine righteous persons who do not need to repent.’ (A; \citetitle{Luke} 15:7)
\z 

Unlike the perfective participle and the masdar, the negative infinitive is derived from its affirmative counterpart in the same way as negative finite forms are derived, i.e. by placing the negator on the right edge, cf. \textit{bik’u-du} ‘to be(come)’ {\textasciitilde} \textit{bik’u-du-}\textbf{\textit{sːu}} ‘not to be(come)’. The infinitive is employed both in purpose clauses, as in \REF{ex:and:36}, and in complements of complement-taking verbs like ‘can, be able’, ‘want’ or ‘need’.

\ea
    \label{ex:and:36} %KM ...
    \gll [...] he{}-lu{}-di men kodi w{}-uχi{}-dja, hinc’o{}-l{}-č’u=cin du{}-b čuk’a \textbf{tunkun{}-nu{}-su} \\
          {} \textsc{dem-obl.pl.m{}-erg} \textsc{2sg} in.hand.\textsc{lat} \textsc{m}{}-take-\textsc{fut} stone\textsc{{}-obl}{}-\textsc{cnt=ptcl} \textsc{2sg}{}-\textsc{gen.iv} leg strike-\textsc{inf-neg} \\
    \glt ‘[...] they will lift you up in their hands, so that you will not strike your foot against a stone.’ (A; \citetitle{Luke} 4:11)
\z 

Besides the perfective converb described above, there are a number of other forms which can head adverbial clauses, and which are derived from various finite verb forms by means of the adverbializer \textit{{}-gu}, often in combination with the emphatic clitic \textit{=ža}, i.e. \textit{{}-guža}. In all these cases, negative counterparts are derived as in finite verb forms, with \textit{{}-gu(ža)} following the negated finite form \citep{Maisak2022}. \REF{ex:and:37} illustrates the use of an adverbialized negative future heading a preposed subordinate clause.

\newpage
\ea
    \label{ex:and:37}
    \gll {\ob}he{}-ldu se{}-r t’iri i{}-du \textbf{beʁel-losja=gu=ža}{\cb}, sebu{\textasciitilde}se{}-b{}-ul=lo hok’or{}-ol dan{}-ol i{}-du w{}-uɢin{}-nu w{}-uk’u{}-dja=le bisːil? \\
    {\db}\textsc{dem-mnr} one-\textsc{v} a.little do-\textsc{inf} can-\textsc{fut.neg=advz=emph} \textsc{red}:one-\textsc{iv-pl=add}
     heavy-\textsc{pl} thing-\textsc{pl} do-\textsc{inf} \textsc{m}{}-\textsc{pl}.get.down{}-\textsc{inf} \textsc{m}{}-\textsc{pl}.need{}-\textsc{fut=q} \textsc{2pl} \\
    \glt ‘Since you cannot do this very little thing, why do you have to worry about the various bigger things?’ (A; \citetitle{Luke} 12:26)
\z 

\subsection{Negative lexicalizations} \label{sec:and:2.5} 

Andi does not seem to have any true negative lexicalizations. As a phenomenon partly resembling a negative lexicalization, one can mention the use of the verb \textit{ɬidu} ‘become; be possible’ in the negative future \textit{ɬidosːja} (or \textit{ɬidosːa}), with the meaning ‘(it is) impossible’, ‘(one) may not’. This verb can be employed with the inchoative meaning ‘become’, in particular as a light verb in intransitive deadjectival complex verbs such as \textit{razi{}-ɬi} ‘agreed’ (< \textit{razi} ‘content, accepting’), \textit{baʁu{}-ɬi} ‘became beautiful’ (< \textit{baʁu} ‘beautiful’) etc.; see \citet[116--125]{Rochant2018} for details. The same verb is used as a possibilitive modal, taking a complement clause headed by the infinitive. 
With the modal meaning, the verb mostly occurs in the future tense,\footnote{The use of the verb ‘become’ as a possibilitive modal with an infinitival complement, and the preference for the future tense in modal function, are not exclusive to Andi, but are common to many other Nakh-Daghestanian languages.} i.e. \textit{ɬidja} ‘can, may’ and \textit{ɬidosːja} ‘cannot, may not’ in the negative, cf. \REF{ex:and:38}. The negative form \textit{ɬidosːja} alone can be used as a semi-lexicalized modal word ‘(it is) impossible’, ‘(it is) not allowed’, as in the Bear’s answer to Mashenka’s question in \REF{ex:and:39}. \citet{Salimov2010} even includes both \textit{ɬidja} ‘(one) may’ and \textit{ɬidosːa} ‘(one) may not’ (Russian \textit{možno} and \textit{nel’zja}) in his Andi–Russian lexicon. Such impersonal use is, apparently, completely in line with the modal semantics of \textit{ɬidu}. Note also that the sense ‘cannot, be unable’ is a cross-linguistically common negative lexicalization (the fourth most common in \citegen{Veselinova2013lexicalization} variety sample of 105 languages).

\ea
    \label{ex:and:38} 
    \gll išːi{}-di imperator{}-šu{}-j naluki ičːi{}-du \textbf{ɬi{}-dja}=le jagi \textbf{ɬi{}-dosːa}=le? \\
          \textsc{1pl.excl-erg} emperor-\textsc{obl.m}{}-\textsc{dat} tax five-\textsc{inf} become-\textsc{fut=q} or become-\textsc{fut.neg=q} \\
    \glt ‘Are we to pay taxes to the Emperor or not?’ (A; \citetitle{Luke} 20:22)
\z 

\ea\label{ex:and:39}
    \begin{xlist}
    \exi{A: }\label{ex:and:39A}
        \glt [Oh Bear, let me go and see my grandparents.]
    \exi{B: }\label{ex:and:39B}
        \gll  \textbf{ɬi{}-dosːja}, men rešː{}-ʟi j{}-it’u{}-dja\\ 
               become-\textsc{fut.neg} \textsc{2sg} forest-\textsc{inter} \textsc{f-}get.lost-\textsc{fut}\\
        \glt ‘No (lit. may not), you will get lost in the woods [said Bear].’ (A; \citetitle{Tales}) 
    \end{xlist}
\z 

\subsection{Other clausal negation constructions: the ‘not-yet’ tense} \label{sec:and:2.6}

Andi possesses a remarkable case of a ‘not{}-yet’ grammaticalization in the verbal domain, namely the verb form we label “counterexpectation present” or ‘still not' present (see \citealt{Maisak2018} and \citealt{MaisakVerhees2024} for a more detailed account). In isolated usage, the form in \textit{\nobreakdash-daʁid} listed in \tabref{tab:and:1} seems to have a negative meaning by itself, without any negation marker: that is, forms like \textit{ruʟ-daʁid} (from the verb ‘say’) tend to be translated simply as ‘doesn’t say’. This form has a regular negative equivalent in {}-\textit{sːu}, whose meaning is (as might be expected) affirmative, e.g. \textit{ruʟ-daʁid}{}-\textit{sːu} ‘says’, or more precisely ‘it’s not true that s/he doesn’t say’, see also \REF{ex:and:40}.

\ea
    \label{ex:and:40} 
    \gll musa-d wocː{}-u-l’o kaʁar \textbf{qʷar-daʁid-sːu}\\
          Musa-\textsc{erg} brother-\textsc{obl}{}-\textsc{sup.lat} letter write-\textsc{cnte.prs-neg}\\
    \glt ‘Musa is indeed writing his brother a letter.’ (In reply to another speaker who doubts that this is the case.) (R) 
\z 


However, it becomes clear from other examples that the form \textit{{}-daʁid} does not simply express negative meaning, but signals counterexpectation on the part of the speaker – the correct translation of \textit{ruʟ-daʁid} would be not just ‘doesn’t say’, but rather ‘s/he still doesn’t say, although I expected him/her to’, see \REF{ex:and:41} and \REF{ex:and:42}. 

\ea 
    \label{ex:and:41}
    \gll anzi r-eʟ-daʁid\\
          snow \textsc{v}{}-go-\textsc{cnte.prs} \\
    \glt ‘It still isn’t snowing [even though it is already January].’ (R)
\z 

\ea 
    \label{ex:and:42}
    \gll djo hege-b c’in-naʁid\\
          \textsc{1sg.afft} \textsc{dem-iv} know-\textsc{cnte.prs} \\
    \glt  ‘I still don’t quite understand that [although you have explained it to me].’ (R)
\z 

Such a dedicated verb form with counterexpectation meaning is not typical of the languages of the family, and functionally resembles ‘not-yet’ tenses in Africa, also known under the labels \textit{tardative} \citep[319]{Plungian1999} or \textit{nondum} (\citealt{VeselinovaDevos2021}). Cross-dialectal comparison helps to see that the Andi counterexpectation present, which is morphologically bound in most dialects (cf. the Rikvani suffix \textit{{}-daʁid}, the Andi/Gagatli suffix \textit{{}-daʁidːu}), probably originated in a periphrastic construction with an auxiliary *\textit{b-aʁi}, with a gender agreement prefix. In the Zilo dialect, the form is less morphologized, and its underlying structure is “habitual in \textit{{}-e} + auxiliary verb \textit{b-aʁij}” \REF{ex:and:43}. In the other dialects, the original structure “habitual in \textit{{}-do} + auxiliary verb \textit{b\nobreakdash-aʁid(ːu)}” must have fused into a single morpheme \textit{{}-daʁid(ːu)}, dropping the gender marker.

\ea
    \label{ex:and:43}
    \gll he-w k’amm{}-e-b-aʁij\\
          \textsc{dem-m} eat-\textsc{hab-iv-cnte.prs} \\
    \glt ‘He does not [want to] eat.’ (Z)
\z 

The form of the auxiliary verb in the counterexpectation present, i.e. \nobreakdash-\textit{daʁid} / \textit{{}-daʁidːu} in most dialects, and \textit{{}-b-aʁij} in Zilo, is the perfect in \textit{{}-d} / \textit{{}-dːu} / \textit{{}-j} which in Andi also has resultative meaning expressing a current state (e.g. ‘is sitting’ or ‘is standing’ is expressed with the resultative form of the verbs ‘sit down’ and ‘come to a stop’, respectively). Although attested almost exclusively in the perfect/resultative form, the counterexpectation present auxiliary can also appear in the unmarked aorist (perfective past). The meaning of such forms is something like a “counterexpectation aorist”, i.e. ‘did not do (as expected)’. Thus, in \REF{ex:and:44} the “counterexpectation aorist” \REF{ex:and:44a} is close in meaning to the negative aorist \REF{ex:and:44b}.

\ea\label{ex:and:44}
    \ea\label{ex:and:44a} 
        \gll sun wošu-d ruʟi-b ži-daʁi \\
              yesterday boy-\textsc{erg} say-\textsc{pfv.ptcp} do-\textsc{cnte.aor}  \\
        \glt 
    \ex\label{ex:and:44b} 
        \gll sun wošu-d ruʟi-b žij-sːu \\
              yesterday boy-\textsc{erg} say-\textsc{pfv.ptcp} do.\textsc{aor-neg} \\
        \glt  ‘Yesterday, the boy did not do what (he) was told.’ (R)
    \z 
\z 

The original meaning of the auxiliary verb \textit{b\nobreakdash-aʁi} is not clear. In modern Andi, this stem is used as a lexical verb only with the meaning ‘get tired’. So, either ‘get tired’ was indeed the grammaticalization source, or else the verb has changed its original semantics, which remains unknown. Comparison with the other Andic languages does not help much, as cognate lexical verbs seem to be lacking, except in Bagvalal, where a rare verb \textit{b\nobreakdash-aʁi} is attested with the meaning ‘resist, refrain from doing’ \REF{ex:and:45}. For details on possible cognates among the negation markers of the Andic languages, see \citet{MaisakVerhees2024}.

\ea
    \label{ex:and:45}
    \gll k’ʷan-di haddiɬːir q’erʟ’eri din b-ihi-ra \textbf{b-aʁi-‹b›o} \\
          Kvanada-\textsc{pl} long.while struggle.\textsc{aor} religion \textsc{n}{}-take-\textsc{ipfv.inf} \textsc{n}{}-resist-\textsc{cvb.pfv<n}> \\
    \glt ‘The people from Kvanada struggled for a long time, refusing to accept the religion.’ (Bagvalal; \citealt{KibrikEtAl2001}: 732)
\z 

Although I am not aware of any Daghestanian grammaticalizations involving an inherently negative verb like ‘resist, refuse’ (or ‘be tired’), the development itself looks semantically plausible. According to \citet[382--383]{Givón2001}, “inherently negative modality verbs such as ‘fail’, ‘lack’, ‘refuse’, ‘decline’, or ‘avoid’ commonly grammaticalize as negation markers”, as “their more specific semantic features are bleached out, leaving only their negative inference”. For a detailed recent study of the functions and the origin of ‘not yet’ expressions in Bantu languages, see \citet{VeselinovaDevos2021}.

\section{Non-clausal negation} \label{sec:and:3}

\subsection{Negative replies} \label{sec:and:3.1}

The positive reply to a polar question is \textit{hun} ‘yes’;\footnote{Or rather \textrm{\textit{h\~u}}, with a nasalized vowel (see also \citealt{Dirr1906}: 130; \citealt{Salimov2010}: 12).} in the available written corpus this word is very rare and only attested as \textit{hu}. As a negative reply, the negative copula \textit{sːu} can be used \REF{ex:and:46}, which is a very common function of negative existentials worldwide (\citealt{Veselinova2013neg.exist}: 118, 139). Even more common is a negative reply \textit{hunsːu} ‘no’ which, quite remarkably, represents the fusion of both ‘yes’ and ‘no’, see \REF{ex:and:47}. Both negative replies negate the propositional content (rather than disagree with the polarity of the question).

\ea\label{ex:and:46}
    \begin{xlist}
    \exi{A: }\label{ex:and:46A}
        \glt [Do you think I came to bring peace on earth?]
    \exi{B: }\label{ex:and:46B}
        \gll  hit’o{}-mado bisːi{}-qi, \textbf{sːu}, den w{}-uɢo bisːil sonso{}-č’u{}-kːu bat’{}-ol{}-i{}-du\\ 
               say-\textsc{prog} \textsc{2pl{}-apud} \textsc{cop.neg} \textsc{1sg} \textsc{m}{}-come.\textsc{aor} \textsc{2pl} each.other-\textsc{cnt-ela} separate{}-\textsc{pl}{}-\textsc{fact{}-inf}\\
        \glt ‘No, I tell you, I came to divide you.’ (A; \citetitle{Luke} 12:51)
    \end{xlist}
\z 

\newpage
\ea\label{ex:and:47}
    \begin{xlist}
    \exi{A: }\label{ex:and:47A}
        \glt [We ask for your hand!]
    \exi{B: }\label{ex:and:47B}
        \gll  \textbf{hun=sːu}, he{}-b sːu=b{}-iʁi di{}-j haǯatib dan\\ 
               yes=\textsc{cop.neg} \textsc{dem-iv} \textsc{cop.neg=iv-aux.aor} \textsc{1sg}-\textsc{dat} necessary thing\\
        \glt ‘No, this was not\footnote{In the Andi dialect, \textrm{\textit{sːu}} can be combined with the past tense auxiliary verb \textrm{\textit{b\nobreakdash-iʁi}} to express past meaning (‘was not’).} what I needed.’ (A; \citetitle{Tales})
    \end{xlist}
\z 

In \citegen{Dirr1906} texts, a negative expression \textit{haj joχ} ‘no’ is used several times as a negative reply, which seems to include the Turkic negative reply \textit{joχ}. In modern texts, there is also a particle \textit{aj} attested as a form of negative reply, cf. \REF{ex:and:48}, which, however, remains an isolated example. \citet[245]{Salimov2010} describes the meaning of \textit{aj} as ‘not to do anything’.

\ea\label{ex:and:48}
    \begin{xlist}
    \exi{A: }\label{ex:and:48A} 
        \glt [Open the door, or I’ll break it! said the Wolf.]
    \exi{B: }\label{ex:and:48B}
        \gll  \textbf{aj}, den arχin{}-nosja!\\ 
               \textsc{neg.ptcl} \textsc{1sg} open-\textsc{fut.neg}\\
        \glt ‘No, I won’t open!’ [said the piglet.] (A; \citetitle{Tales})
    \end{xlist}
\z 

On the whole, one gets the impression that negative replies employing particles are rarely found in texts (even in dialogues), and that it is rather the negative verb form which is preferably used in such contexts, as in \REF{ex:and:49}. The same is true of the affirmative reply \textit{hu(n)}: according to my consultants, it is not very felicitous to use ‘yes’ as a reply to a question like ‘Did you do it?’ – normally, the reply will include just a verb form (e.g. ‘Did’). 

\ea\label{ex:and:49}
    \begin{xlist}
    \exi{A: }\label{ex:and:49A} 
        \glt [Do you think they were more guilty than all the others living in Jerusalem?]
    \exi{B: }\label{ex:and:49B}
        \gll  hit’o{}-mado bisːi{}-qi, \textbf{w{}-ok’o{}-su}=ʁodːu\\ 
               say-\textsc{prog} \textsc{2pl{}-apud} \textsc{m}{}-\textsc{pl.}be\textsc{.aor-neg=quot}\\
        \glt ‘I tell you, they were not!’ (A; \citetitle{Luke} 13:5)
    \end{xlist}
\z 

\subsection{Negative indefinites and quantifiers} \label{sec:and:3.2}

Negative indefinites are derived by means of a suffix \textit{{}-gulo}, which is a lexicalized combination of an intensifying clitic \textit{=gu} and an additive clitic \textit{=lo} ‘also, even’.\footnote{For a description of the range of functions typically associated with intensifying markers, see \citet{Forker2015} on the Avar enclitic \textit{go} which is a close equivalent, and a cognate, of the Andi \textit{gu}. See also \citet{Forker2016} on the semantic typology of additives.} Besides being a marker of negative indefinites, \textit{gulo} can be also used as an emphatic particle, see \sectref{sec:and:4.6.2}, example \REF{ex:and:86}.

The most frequent negative indefinite pronoun is \textit{se-w-gulo} (\textit{se-b-gulo}, etc.) ‘no one, nothing’ based on the numeral \textit{se-w} ‘one’, with a final gender agreement suffix. This pronoun can be used both attributively and as an NP head, in the latter case it inflects for case and number (the absolutive case is unmarked, and the other case forms are derived from an oblique stem). Case/number inflection follows \textit{{}-gulo}, which confirms its status as a suffix rather than a clitic. Negative indefinites in \textit{{}-gulo} occur together with clausal negation. In \REF{ex:and:50}, \textit{sewgulo} is a modifier (‘no man’), in \REF{ex:and:51} and also \REF{ex:and:7}, \REF{ex:and:9}, \REF{ex:and:12} and \REF{ex:and:29} above it occurs in the unmarked absolutive form (‘no one, nothing’), and in \REF{ex:and:6} above it occurs in the dative (‘to no one’) heading an NP in an argument position. 

\ea
    \label{ex:and:50}
    \gll \textbf{se{}-w{}-gulo} hek’a q’int’i{}-dosub bisːi{}-di\\
          one-\textsc{m-indf} man press-\textsc{proh} \textsc{2pl{}-erg} \\
    \glt ‘Don’t extort money from any man.’ (lit. 'Don't press money from any man') (A; \citetitle{Luke} 3:14)
\z 

\ea
    \label{ex:and:51} 
    \gll amma hege{}-lu{}-‹b›o \textbf{se{}-b{}-gulo} bič’ːuni{}-su=ɬodi\\
          but \textsc{dem-obl.pl.m{}-afft<iv>} one-\textsc{iv-indf} understand.\textsc{aor-neg=rep} \\
    \glt ‘But they did not understand anything.’ (A; \citetitle{Luke} 18:34)
\z 

Although in the overwhelming majority of examples \textit{sebgulo} is found in negative clauses, it is possible to use it outside negative contexts as well, e.g. in questions. In \REF{ex:and:52} the sentence includes several conjoined dependent clauses with negative converbs, but there is no negation in the main clause (‘did you lack anything?’). Thus, in \citegen[199]{Haspelmath1997} terms, \textit{sebgulo} would be a negative indefinite pronoun “in the deliberately vague sense ‘indefinite pronoun that has “direct negation” as an important function’”.

\ea
    \label{ex:and:52}
    \gll ja orsi b{}-ik’ol{}-dːu tarha b{}-iχi{}-č’igu, ja q’aj b{}-ik’ol{}-dːu q’ʷač’a r{}-iχi{}-č’igu, ja b{}-eršol{}-dːu b{}-oli{}-du nasil{}-ol b{}-oχi{}-č’igu, bisːil den{}-ni w{}-oʟelta{}-b rihi,  \textbf{se{}-b{}-gulo} kami b{}-ik’o=de bisːi{}-j? \\
          or money \textsc{iv}{}-be.\textsc{caus-cvb} wallet \textsc{iv}{}-take-\textsc{cvb.neg} or belongings \textsc{iv}{}-be.\textsc{caus-cvb} sack \textsc{v}{}-take-\textsc{cvb.neg} or \textsc{iv}{}-change.\textsc{caus-cvb} \textsc{iv}{}-put.on-\textsc{inf} shoe-\textsc{pl} \textsc{iv}{}-\textsc{pl}.take-\textsc{cvb.neg} \textsc{2pl} \textsc{1sg}-\textsc{erg} \textsc{m}{}-let-\textsc{pfv.ptcp} time one-\textsc{iv-indf} lacking \textsc{iv}{}-be.\textsc{aor=q} \textsc{2pl-dat} \\
    \glt ‘When I sent you without having taken a purse, without a bag with your things, without shoes to wear, did you lack anything?’ (A; \citetitle{Luke} 22:35)
\z 

It is not only the numeral ‘one’ from which a negative/indefinite pronoun can be derived by means of \textit{{}-gulo}, see \REF{ex:and:53}--\REF{ex:and:55}. Thus, \citet{Salimov2010} mentions in his dictionary such interrogative-based derivatives as \textit{inu-gulo} ‘anywhere’ (< \textit{inu} ‘where?’), \textit{i‹w›a{}-gulo} ‘to anywhere’ (< \textit{i‹w›a} ‘where to?’, with a gender agreement infix), \textit{inna-gulo} ‘always’ (< \textit{inna} ‘when?’), \textit{iʟ’ol-gulo} ‘because’ (< \textit{iʟ’ol} ‘why?’), cf. also \textit{čom-gulo} ‘some’ (< \textit{čom} ‘how many?’). \textit{Onš}\textbf{ː}\textit{i-gulo} ‘ever’ is derived from an adverb \textit{onš}\textbf{ː}\textit{i} ‘once’. 

\ea
    \label{ex:and:53} 
    \gll di{}-b burt’a{}-ɬir \textbf{inna{}-gulo} di{}-č’u soru\\
          \textsc{1sg}-\textsc{gen.iv} rich-\textsc{abstr} when-\textsc{indf} \textsc{1sg}-\textsc{cnt} together\\
    \glt ‘My wealth is always with me.’ (A; \citetitle{Tales})
\z 


\ea
    \label{ex:and:54} 
    \gll naf-naf ho{}-r \textbf{čom{}-gulo} zubu en{}-ʟi=gu milki{}-l{}-č’u b-eqi-du b-ik’ʷo-dːu \\
          Naf-Naf \textsc{dem-v} some-\textsc{indf} day self.\textsc{fn}{}-\textsc{gen=ints} house-\textsc{obl}{}-\textsc{cnt} \textsc{iii}{}-reach-\textsc{inf} \textsc{iii}{}-be-\textsc{prf} \\
    \glt ‘Naf-Naf had been busy building his house for a few days already.’ (A; \citetitle{Tales})
\z 

\ea
    \label{ex:and:55} 
    \gll hege{}-l{}-č’u reχudu min \textbf{onšːi{}-gulo} b{}-ič’i{}-dosːja!\\
          \textsc{dem{}-obl.fn-cnt} after \textsc{2sg} once-\textsc{indf} \textsc{iii}{}-die-\textsc{fut.neg} \\
    \glt ‘After that you will never die!’ (A; \citetitle{Tales})
\z 

Within \citeauthor{Haspelmath1997}'s (\citeyear[Ch. 8]{Haspelmath1997}, \citeyear{Haspelmath2013-wals115}) typology of indefinite and negative pronouns, the series of pronouns in \textit{{}-gulo} are “interrogative-based negative indefinites” which co-occur with predicate negation (both the interrogative-based structure and the co-occurence with predicate negation are the most cross-linguis\-tically common values). There is also another interrogative-based series of indefinites in Andi derived by means of the marker \textit{{}-bolo} (\textit{\nobreakdash-bollo}) which represents a concessive mood ending: the concessive is derived from the conditional in \textit{\nobreakdash-bor} by means of the additive \textit{=lo} ‘also, even’, which assimilates /r/ (i.e \textit{bolo} < \textit{bollo} < *\textit{bor=lo} [\textsc{cond=add}]). Indefinites in \textit{{}-bolo} generally correspond to the English \textit{some}{}-series pronouns, e.g. \textit{im-molo} ‘someone’, \textit{ib-bolo} ‘something’, \textit{čom-bolo} ‘some amount’, \textit{inna-bolo} ‘at some time’, etc. Concessive forms of auxiliary verbs ‘become’ and ‘be’, i.e. \textit{ɬibollo} and \textit{bik’obollo}, can also be used to form indefinite phrases. The series of indefinites in \textit{{}-bolo} (\textit{\nobreakdash-bollo}) represents what is usually known as “ordinary indefinites” (i.e. not negative, not free-choice and not specific indefinites), and is not used in direct negation contexts.

\subsection{Negative derivation and the caritive meaning} \label{sec:and:3.3}

The meaning ‘without’ can be expressed by a dedicated particle \textit{ti}, or by constructions involving the participle of the negative copula/existential verb \textit{sːu}, i.e. \textit{sːu-b}. Among the Andi particles, \citet[216]{Salimov2010} even lists \textit{sːub} and \textit{tigu} as “negative particles”, giving the following illustrations:

\ea
    \label{ex:and:56}
    \gll me=\textbf{sːu{}-b} dunjal, miɬi=\textbf{sːu{}-b} bičːin\\
          \textsc{2sg}=\textsc{cop.neg-ptcp} world sun=\textsc{cop.neg-ptcp} shadow.side \\
    \glt ‘A world without you [is] a mountainside without sunshine.’ (G; \citealt{Salimov2010}: 216)
\z 

\ea 
    \label{ex:and:57}
    \gll hede{}-w men=\textbf{ti{}-gu} w{}-uɢo\\
          \textsc{dem-m} \textsc{2sg}=\textsc{car-advz} \textsc{m}{}-come.\textsc{aor} \\
    \glt ‘He came without you.’ (G; \citealt{Salimov2010}: 216)
\z 

As for \REF{ex:and:56}, \textit{sːub} here is not a particle but a participial form of the copula, heading relative clauses, namely ‘(a world) that does not have you’ and ‘(a mountainside) that does not have the sun’, see also \REF{ex:and:30} above. The caritive meaning here is thus expressed by means of a standard syntactic construction, not by a dedicated marker. Note that, according to \citet[118]{Veselinova2013neg.exist}, the use “as a pre-/postposition meaning ‘without’ or as a privative marker” is one of the common cross-linguistic functions of negative existentials.

In contrast, \textit{ti} (in \REF{ex:and:57}, with the adverbializer) is a dedicated caritive enclitic or function word, which does not have any other uses.\footnote{It is unclear whether \textrm{\textit{ti}} is a clitic or an autonomous function word: it is written together with its host in the single example by \citet{Salimov2010}, but always separately in the Tales and the Gospel of Luke. Interestingly, \citet[361]{Salimov2010} also includes \textrm{\textit{tigu}} as a separate lexical entry in the dictionary appended to his grammar sketch. Note that the caritive \textrm{\textit{ti}} is unrelated to the comitative, which is \textrm{\textit{=logu}}, an enclitic which represents the fusion of the additive marker \textrm{\textit{=lo}} and the adverbializer \textrm{\textit{{}-gu}}.} It is hosted by noun phrases, and is used with the adverbializer when it heads an adverbial clause, as in \REF{ex:and:57} and also in \REF{ex:and:58}. The only example of \textit{ti} attested (in texts) without the adverbializer \REF{ex:and:59} seems to represent use in a secondary predication ‘more dead than alive’ (lit. ‘being neither dead nor alive’).

\ea
    \label{ex:and:58} 
    \gll išːi{}-w{}-ul=gu saldat{}-ol \textbf{ti{}-gu} išːi{}-‹b›o se{}-b{}-gulo  i-du bildːosːja\\
          \textsc{1pl.excl-gen.m-pl=ints} soldier-\textsc{pl} \textsc{car-advz} \textsc{1pl.excl-afft<iv}> one-\textsc{iv-indf}  do-\textsc{inf} can+\textsc{fut.neg} \\
    \glt ‘We cannot do anything without our soldiers.’ (A; \citetitle{Tales})
\z 

\ea
    \label{ex:and:59} 
    \gll nif-nif=lo nuf-nuf=lo hege{}-b zaman j{}-it’{}-ol{}-gu=lo \textbf{ti}, j{}-oč’o{}-dːu=lo \textbf{ti} naf-naf{}-o{}-ʟi milki{}-l{}-χo helli-dːu j-eqo-dːu \\
          Nif-Nif=\textsc{add} Nuf-Nuf=\textsc{add} \textsc{dem-iv} time \textsc{iii.pl}{}-alive-\textsc{pl-advz=add} \textsc{car} \textsc{iii.pl}{}-\textsc{pl}.die-\textsc{cvb=add} \textsc{car} Naf-Naf-\textsc{obl}{}-\textsc{gen} house-\textsc{obl}{}-\textsc{ad.lat} run-\textsc{cvb} \textsc{iii.pl}{}-reach-\textsc{prf} \\
    \glt ‘And Nif-Nif and Nuf-Nuf, who were then more dead than alive, ran to Naf-Naf’s house.’ (A; \citetitle{Tales})
\z 

Thus, the caritive meaning in Andi is expressed by means of syntactic constructions rather than morphologically: as an attribute ‘which does not have’, this is a participial clause headed by \textit{sːub}, and as an adverbial ‘without having’, this is a clause headed by \textit{ti(gu)}. There are no dedicated derivational affixes with caritive or related meanings. 

\section{Other aspects of negation} \label{sec:and:4}

\subsection{The scope of negation} \label{sec:and:4.1}

Specification of the scope of negation can be achieved by negating a particular constituent, and not the main predicate. Thus, while in \REF{ex:and:60a} we have a standard negative copular clause with a negative copula (‘shouldn’t’), in \REF{ex:and:60b} the focus of negation is on the dependent clause, so it is the infinitive as the head of a complement that bears negation (‘not to speak’).

\ea\label{ex:and:60} 
    \ea\label{ex:and:60a}
        \gll  ruʟ{}-du haǯat \textbf{sːu}\\ 
               say-\textsc{inf} necessary \textsc{cop.neg} \\
        \glt ‘One shouldn’t talk.’ (R)
    \ex\label{ex:and:60b}
        \gll  ruʟ{}-du{}-\textbf{sːu} haǯat\\
               say-\textsc{inf-neg} necessary \\
        \glt ‘One should keep silence (lit. not-talk).’ (R)
    \z
\z 

Constituent negation on a non-verbal constituent can be also marked with \textit{sːu}.\footnote{In Rikvani, the use of \textrm{\textit{sːub}} (originally, the participial copula form) is also attested in constituent focus marking, see \REF{ex:and:63}.} Thus, \REF{ex:and:61a} illustrates standard clausal negation, \REF{ex:and:61b} shows that the agent NP in the scope of negation hosts \textit{sːu}, and in \REF{ex:and:61c} it is the patient NP which is in the scope of negation and hosts \textit{sːu}.

\ea\label{ex:and:61} 
    \ea\label{ex:and:61a}
        \gll  imu-d riʟ’ b-iχi-\textbf{sːu}\\ 
               father-\textsc{erg} meat \textsc{iv-}take.\textsc{aor-neg} \\
        \glt ‘Father didn’t buy meat.’ (R)
    \ex\label{ex:and:61b} 
        \gll  [imu-d]=\textbf{sːu} riʟ’ b-iχi, ilu-ʟ wocːu-d\\
               father-\textsc{erg=neg} meat \textsc{iv-}take.\textsc{aor} mother-\textsc{gen} brother-\textsc{erg} \\
        \glt ‘It was not father who bought meat, but mother’s brother.’ (R)
    \ex\label{ex:and:61c}
        \gll  imu-d [riʟ’]=\textbf{sːu} b-iχi, ʟurʟa r-iχi\\
                 father-\textsc{erg} meat=\textsc{neg} \textsc{iv-}take.\textsc{aor} butter \textsc{v-}take.\textsc{aor} \\
        \glt ‘It was not meat that father bought, but butter.’ (R)
    \z
\z 

A particle \textit{=ža} can be optionally added in the assertive part to emphasize the correct alternative, as in (\ref{ex:and:62}–\ref{ex:and:63}). Judging from its typical uses in other contexts, \textit{=ža} can be tentatively labelled the marker of “uncontroversial information” (in the sense of \citealt{Panov2020}).

\ea
    \label{ex:and:62}
    \gll den χːuča qʷar-rado-sːu, r-aʟ’-rado(=\textbf{ža})\\
          \textsc{1sg} book write-\textsc{prog-neg} \textsc{v-}read-\textsc{prog=ptcl} \\
    \glt ‘I am not writing a book, I am reading one.’ (R)
\z 

\ea
    \label{ex:and:63}
    \gll den χːuča=sːub, kazjat(=\textbf{ža}) b-aʟ’-rado\\
          \textsc{1sg} book=\textsc{neg} newspaper=\textsc{ptcl} \textsc{iv}{}-read-\textsc{prog} \\
    \glt ‘I am not reading a book, (but) a newspaper.’ (R)
\z 

The question arises whether in \REF{ex:and:61b} and \REF{ex:and:61c} \textit{sːu} is just “the same” negator as occurs in \REF{ex:and:61a} on the verb, but simply “moved” to a focused constituent. This is a natural assumption given that focus marking by means of predicative elements, including copulas, is a common strategy in Nakh-Daghestanian languages, including those of the Andic branch (see \citealt{Kazenin2003}, \citealt{KalininaSumbatova2007}, \citealt{Semënova2014} and \citealt{Maisak2023} for details). Consider the two pairs of examples from Bagvalal, another Andic language. In \REF{ex:and:64a} and \REF{ex:and:65a} with the affirmative and the negative periphrastic perfect, the auxiliary verb (copula) is placed in the default position after the lexical verb. In \REF{ex:and:64b} and \REF{ex:and:65b}, the auxiliary appears after the subject NP, which is in focus. Thus, in the affirmative and the negative clauses the focus marking strategy is absolutely identical. 

\ea\label{ex:and:64} Bagvalal (Avar-Andic, Nakh-Daghestanian) \\
    \ea\label{ex:and:64a}
        \gll  maħammad w-a-‹w›o \textbf{ek’ʷa}\\ 
               Muhammad \textsc{m}{}-come-\textsc{cvb.pfv<m}> \textsc{cop} \\
        \glt ‘Muhammad came.’ (\citealt{KibrikEtAl2001}: 683)
    \ex\label{ex:and:64b} %KM moved brackets to the first line
        \gll   [maħammad] \textbf{ek’ʷa} w-a-‹w›o\\
               {\db}Muhammad \textsc{cop} \textsc{m}{}-come-\textsc{cvb.pfv<m}> \\
        \glt ‘It was Muhammad who came.’ (\citealt{KibrikEtAl2001}: 683)
    \z
\z 

\ea\label{ex:and:65} Bagvalal (Avar-Andic, Nakh-Daghestanian) \\
    \ea\label{ex:and:65a}
        \gll  maħammad w-a-‹w›o \textbf{weč’e}\\ 
               Muhammad \textsc{m}{}-come-\textsc{cvb.pfv<m}> \textsc{cop.neg} \\
        \glt ‘Muhammad didn’t come.’ (\citealt{KibrikEtAl2001}: 683)
    \ex\label{ex:and:65b} %KM moved brackets to the first line
        \gll  [maħammad] \textbf{weč’e} w-a-‹w›o\\
               {\db}Muhammad \textsc{cop.neg} \textsc{m}{}-come-\textsc{cvb.pfv<m}> \\
        \glt ‘It was not Muhammad who came.’ (\citealt{KibrikEtAl2001}: 683)
    \z
\z 


As mentioned above, in Upper Andi the copula does not occur in periphrastic forms, so focus constructions comparable to those in Bagvalal do not exist. It may be, however, that the mechanism of constituent focusing by means of copula placement is only applicable under negation. In this case, what we see in (\ref{ex:and:61b}--\ref{ex:and:61c}) could be interpreted as focus marking by means of a negative copula, as in \REF{ex:and:64b} and \REF{ex:and:65b} from Bagvalal. Still, it is not the case that in examples like (\ref{ex:and:61b}--\ref{ex:and:61c}) the Andi negative marker \textit{sːu} simply changes its linear position. It is possible to have \textit{\nobreakdash-sːu} both on the verb and on the focused constituent in those cases when, e.g., the speaker focuses the lack of involvement of a certain participant in the situation. Thus, in \REF{ex:and:66} the idea that the speaker does not talk is being rejected: what is being asserted is that it is the addressee who does not do this.

\ea
    \label{ex:and:66}
    \gll den=\textbf{sːu} ruʟ-rado-\textbf{sːu}, men\\
          \textsc{1sg}=\textsc{neg} say-\textsc{prog-neg} \textsc{2sg} \\
    \glt ‘It is not me who isn’t talking, (but) you.’
\z 

While the latter example was constructed by me and approved by a native speaker, not all speakers accept sentences with two \textit{sːu} markers, one as constituent negation and the other as verbal negation. At the same time, such examples can be produced during elicitation by Andi speakers themselves, cf. \REF{ex:and:67} suggested as the translation (from a Russian stimulus) for ‘It was not the father who did not come, but the brother’. Although I do not propose any final decision on the status of \textit{sːu} as a constituent negator, it seems highly dubious that \textit{sːu} on an NP can be analysed simply as the result of a “mobile” verb negator shifting to a new location. Note, for example, that in \REF{ex:and:67} the perfect verb form does not make use of the standard negator \textit{sːu} at all (see also \sectref{sec:and:4.6.3} for further discussion).

\ea
    \label{ex:and:67}
    \gll ima=\textbf{sːu} w-uɢo-č’igu, wocː\\
          father=\textsc{neg} \textsc{m}{}-come-\textsc{prf.neg} brother \\
    \glt ‘It was not the father who did not come, (but) the brother.’
\z 

\subsection{Negative polarity} \label{sec:and:4.2}

With the exception of negative indefinites in \textit{{}-gulo} (see \ref{ex:and:50}--\ref{ex:and:52} in \sectref{sec:and:3.2}) and probably also the reinforcing element \textit{t’iri} (see \ref{ex:and:73}--\ref{ex:and:74} in \sectref{sec:and:4.4}), no negative polarity items have been identified.

\subsection{Marking of NPs in the scope of negation} \label{sec:and:4.3}

There are no effects of negation on case marking, nor on any other types of marking of NPs.

\subsection{Reinforcing negation} \label{sec:and:4.4}

There seem to be no special grammatical means for reinforcing negation. In constituent negation constructions, contrastive negation can be expressed by changing the linear position of the negative marker \textit{sːu}, as shown in \sectref{sec:and:4.1}.

Negated sentences can be made emphatic by means of the intensifying clitic \textit{=gu}, as in \REF{ex:and:68} with the negative aorist and \REF{ex:and:69} with the negative future. As shown in \REF{ex:and:70} with the negative perfective converb, the emphatic use of \textit{=gu} is also possible on subordinate clauses. In general, however, the emphatic use of \textit{=gu} is not restricted to negative contexts, and it can occur in affirmative clauses as well. 

\ea
    \label{ex:and:68} 
    \gll di-b χoj b-ik’o=\textbf{gu}=sːu\\
          \textsc{1sg}-\textsc{gen.iii} dog \textsc{iii}{}-be.\textsc{aor=ints=neg} \\
    \glt ‘I had no dog at all.’ (R)
\z 

\ea
    \label{ex:and:69}
    \gll išː-l’o sːe-b{}-gulo ruʟ-do‹\textbf{gu}›sːja\\
          \textsc{1pl.excl-sup.lat} one-\textsc{iv}{}-\textsc{indf} say-\textsc{fut.neg<ints>} \\
    \glt ‘He won’t tell us anything at all.’ (R)
\z 
  
\ea
    \label{ex:and:70} 
    \gll  [...] halt’unni-r j-ik’ʷo-dːu lek’u šanni=\textbf{gu}=č’igu\\
          {} work-\textsc{prog} \textsc{f-}be-\textsc{prf} more take.rest=\textsc{ints=cvb.neg} \\
    \glt ‘[...] she was working hard, without even taking a rest.’ (A; \citetitle{Tales}) 
\z 
        
The additive clitic is also employed in contexts of emphatic negation, where it regularly expresses the scalar meaning ‘even’, see \REF{ex:and:71}, which is functionally similar to \REF{ex:and:70}. In \REF{ex:and:72}, the meaning ‘not a single one’ is rendered by using both the numeral ‘one’ and the additive clitic. In affirmative contexts, the meaning of the additive can be both additive proper (‘and, also’, see \sectref{sec:and:4.5}) and scalar emphatic (‘even’).

\ea
    \label{ex:and:71} 
    \gll [...] amma hede{}-b hege{}-li{}-lo=χːardːu harč’on=\textbf{no}=č’igu\\
          {} but \textsc{dem-iii} \textsc{dem-obl.pl.fn{}-sup.lat}=towards look\textsc{=add}=\textsc{prf.neg}\\
    \glt ‘[...] but he didn’t even look back at them.’ (A; \citetitle{Tales})
\z 

\ea
    \label{ex:and:72}
    \gll men{}-ni di{}-j se{}-b c’ek’ir=ɢa=\textbf{lo} ičːi{}-su...\\
          \textsc{2sg}{}-\textsc{erg} \textsc{1sg}-\textsc{dat} one-\textsc{iii} goat=\textsc{degr=add} give.\textsc{aor{}-neg} \\
    \glt ‘Yet you never gave me even a young goat...’ (A; \citetitle{Luke} 15:29)
\z 

Note that with negative verb forms, both \textit{=lo} and \textit{=gu} behave as endoclitics, being placed word-internally (see \sectref{sec:and:4.6.3} for discussion).

One more construction employed in contexts of reinforcing negation includes the word \textit{t’iri}, originally a noun meaning ‘spatter, drop’ according to \citegen[365]{Salimov2010} lexical materials. In all examples attested in texts, \textit{t’iri} in the reinforcing function is accompanied by the additive clitic, as in (\ref{ex:and:73}–\ref{ex:and:74}). The reinforcing use of \textit{t’iri} seems to be restricted to negative clauses.

\ea
    \label{ex:and:73} 
    \gll amma hege{}-b \textbf{t’iri}=lo sir{}-du dan b{}-ik’ʷo{}-č’igu en-ʟi=gu hinc’o-l-ʟi milki-l-č’u hinu \\
          but \textsc{dem-iv} drop=\textsc{add} fear-\textsc{inf} thing \textsc{iv}{}-be-\textsc{prf.neg} self.\textsc{fn}{}-\textsc{gen=ints} stone\textsc{{}-obl}{}-\textsc{gen} house\textsc{{}-obl}{}-\textsc{cnt} inside \\
    \glt ‘But they had nothing to fear (lit. not a thing to fear even a drop) in his stone house.’ (A; \citetitle{Tales})
\z      

\ea
    \label{ex:and:74} 
    \gll hinǯerʟja kunt’a \textbf{t’iri}=lo hege{}-l{}-la razi ɬi{}-č’igu\\
          this.time husband drop=\textsc{add} \textsc{dem-obl.fn-sup.ess} content become-\textsc{prf.neg} \\
    \glt ‘This time the husband was not content at all (lit. not a drop).’ (A; \citetitle{Tales})
\z

\subsection{Negation, coordination and complex clauses} \label{sec:and:4.5}

Finite coordination strategies, whether with affirmative or negative predicates, are rarely used in Andi, where non-finite constructions are much more prominent. Finite clauses, affirmative as well as negative, can be conjoined symmetrically by means of the additive clitic \textit{=lo} ‘and, also’, which is not a coordinating device as such, as it can occur with a single constituent as well. With negative verb forms, \textit{=lo} is placed word-internally as an endoclitic, before the negation marker, as in \REF{ex:and:75} (see \sectref{sec:and:4.6.3}).

\ea
    \label{ex:and:75}
    \gll c’ːa-do=\textbf{lo}=sːu, k’amm{}-o=\textbf{lo}=sːu\\
          drink-\textsc{hab=add=neg} eat-\textsc{hab=add=neg} \\
    \glt ‘He neither drinks nor eats.’ (R)
\z 


There is a double phrase-initial conjunction \textit{ja... ja} (an Arabic loan found in many Nakh-Daghestanian languages) which is used with the meaning ‘either... or’ in affirmative sentences, and ‘neither... nor’ in negative ones, as in \REF{ex:and:76}. Like the additive, the disjunctive \textit{ja... ja} is not exclusively a clausal conjunction, as it can combine with NPs as well. The two devices are also mutually compatible, cf. \REF{ex:and:77} where \textit{ja... ja} is employed clause-initially, and the additive is placed on the two NPs in the two contrasted clauses.

\ea
    \label{ex:and:76} 
    \gll \textbf{ja} dan{}-a{}-di b{}-ic’i{}-b ust’ol sːu, \textbf{ja} r-ikʷokʷo c’a sːu haq’u{}-la\\
          or thing{}-\textsc{obl}.\textsc{pl}{}-\textsc{erg} \textsc{iv}{}-be.filled-\textsc{pfv.ptcp} table \textsc{cop.neg} or \textsc{v}{}-burn.\textsc{ptcp} fire \textsc{cop.neg} house-\textsc{in.ess} \\
    \glt  ‘In the house there is neither a table set (for guests), nor a fire burning (in the fireplace).’ (A; \citetitle{Tales})
\z 

\ea
    \label{ex:and:77} 
    \gll \textbf{ja} hila=si{}-š{}-č’u{}-kːu=\textbf{lo} sir{}-do{}-su den, \textbf{ja} adam{}-č’u{}-kːu=\textbf{lo} sul{}-ol{}-i{}-do{}-su \\
          or above=\textsc{attr-obl.m-cnt-ela=add} fear-\textsc{hab-neg} \textsc{1sg} or people-\textsc{cnt-ela=add} shame{}-\textsc{pl}{}-\textsc{fact}{}-\textsc{hab-neg} \\
    \glt ‘I don’t fear God or care what people think.’ lit. (A; \citetitle{Luke} 18:4)
\z 

Although highly marginal, a negative particle \textit{aj} (which we saw in \sectref{sec:and:3.1}) is also attested in clause-final position in the ‘neither... nor’ context, see \REF{ex:and:78}.

\ea
    \label{ex:and:78} 
    \gll c’ːa{}-du=lo \textbf{aj}, ikun{}-nu=lo \textbf{aj}, he{}-l{}-lo harč’in{}-nu ɢošːi{}-dːu=logu \\
          drink-\textsc{inf=add} \textsc{neg.ptcl} eat-\textsc{inf=add} \textsc{neg.ptcl} \textsc{dem-obl.fn-sup.lat} look-\textsc{inf} get-\textsc{cvb=com} \\
    \glt [A girl so beautiful that people used to say looking at her:] ‘(One should) neither drink nor eat, only look at her.’ (A; \citetitle{Tales})
\z 


\subsection{Miscellaneous aspects of negation} \label{sec:and:4.6}

\subsubsection{Negative copula in the ‘only X’ construction} \label{sec:and:4.6.1}

There is a frequent construction involving an adverbial form of the copula \textit{sːu-gu} in the context of clausal negation (that is, there are two negation markers in a single clause), see (\ref{ex:and:79}--\ref{ex:and:82}). This construction is a standard means of expressing the meaning ‘only’ or ‘nobody/nothing but’.\footnote{The same construction type has been described for Bagvalal (\citealt{KibrikEtAl2001}: 182, 721–722), see also \citet{Tatevosov2021} for a semantic analysis.} Both noun phrases and verb phrases, as well as adverbial phrases, can host \textit{sːugu}, which in this construction should be probably regarded as a particle, not functioning as a copula or an existential verb.

\ea
    \label{ex:and:79} 
    \gll hoʟ’u-gu rešin=\textbf{su-gu} k’unt’{}-u-χa \textbf{r-uχi-č’igu} hege-l{}-dːi \\
          seven-\textsc{num} year=\textsc{cop.neg-advz} husband-\textsc{obl}{}-\textsc{ad.ess} \textsc{v}{}-take-\textsc{prf.neg} \textsc{dem-obl.fn{}-erg} \\
    \glt ‘She had lived with her husband for only seven years.’ (A; \citetitle{Luke} 2:36)
\z 


\ea
    \label{ex:and:80} 
    \gll  [...] sːurat-a-la=\textbf{sːu-gu} onšːi-gulo boc’o \textbf{haɢo-sːu-b} nif-nif-o-di \\
          {} picture-\textsc{obl.pl}{}-\textsc{sup.lat=cop.neg-advz} once-\textsc{indf} wolf see-\textsc{neg-pfv.ptcp} Nif-Nif-\textsc{obl}{}-\textsc{erg} \\
    \glt ‘[...] [said] Nif-Nif, who had only seen wolves in pictures.’ (A; \citetitle{Tales})
\z 

\ea
    \label{ex:and:81} 
    \gll hila=si-šː-di munah-ol mič’-ol-gu a‹b›čolja-šːu-‹b›o mič’i-gu=\textbf{sːu-gu} \textbf{hilu-dosja} \\
          above=\textsc{attr-obl.m-erg} sin-\textsc{pl} little-\textsc{pl-advz} clean<\textsc{iv>.caus+ipfv.ptcp-obl(m)-afft<iv>} little-\textsc{advz=cop.neg-advz} love-\textsc{fut.neg} \\
    \glt ‘Whoever has been forgiven little (by God) will only love little.’ (A; \citetitle{Luke} 7:47)
\z 

\ea
    \label{ex:and:82}
    \gll b-ihu azab-aq’uba=lo ičːi-dːu=\textbf{su-gu} \textbf{w-oʟelto=lo=sːu} \\
          \textsc{iv}{}-many punishment-suffering=\textsc{add} give-\textsc{cvb=cop.neg-advz} \textsc{m}{}-let.\textsc{hab=add=neg} \\
    \glt ‘[A spirit] doesn’t leave him without torturing him.’ (A; \citetitle{Luke} 9:39)
\z

\subsubsection{Negative forms in disjunction constructions} \label{sec:and:4.6.2}


It is common for Nakh-Daghestanian languages to express contrastive disjunction (‘or else’, ‘otherwise’) by means of a form involving negation, e.g. a negative conditional form of a copula/existential verb, literally ‘if it is not’. Andi conforms to this tendency. Contrastive disjunction can be expressed with the form \textit{sːuɬibor} ‘if not’, which is a combination of \textit{sːu} and the conditional form \textit{ɬibor} of the verb ‘become’, see \REF{ex:and:83} and \REF{ex:and:84}.\footnote{As a disjunction particle, \textrm{\textit{jaɬibor}}, a combination of \textrm{\textit{ja}} ‘or’ with \textrm{\textit{ɬibor}}, can also be used.}

\ea
    \label{ex:and:83} 
    \gll žeɬal \textbf{sːu=ɬi-bor} šːela w-uč’i-dja din\\
          today \textsc{cop.neg}=become-\textsc{cond} tomorrow \textsc{m-}die-\textsc{fut} \textsc{1sg} \\
    \glt [My dear son, you see that I am one foot in the grave,] ‘I will die any day now (lit. today or tomorrow).’ (A; \citetitle{Tales})
\z 

\ea
    \label{ex:and:84}
    \gll \textbf{su=ɬi-bor} e‹m›i haɢi-du=di w-olon-dːu w-ok’o?\\
          \textsc{cop.neg}=become-\textsc{cond} who<\textsc{h>} see-\textsc{inf=wh} \textsc{m}{}-\textsc{pl}.go-\textsc{cvb} \textsc{m-pl}.be.\textsc{aor} \\
    \glt [What did you go out into the desert to see? A reed swayed by the wind?] ‘If not, who did you go out to see?’ (A; \citetitle{Luke} 7:25)
\z 

Another way of expressing disjunction is the use of a negative reply word \textit{hunsːu} ‘no’ (see \sectref{sec:and:3.1}), either on its own or with an additional emphatic particle \textit{gulo}, cf. \REF{ex:and:85} and \REF{ex:and:86}. The negative copula \textit{sːu}, which is also used as a negative reply, can be found in the same function \REF{ex:and:87}.

\ea
    \label{ex:and:85}
    \gll ho-b e‹b›i=ʁil, χːunt’{}-u-ʟi boši{}-llo-j=sːi milki=le, \textbf{hun=sːu} qala=le? \\
          \textsc{dem-iv} what<\textsc{iv>=wh} pig\textsc{{}-obl}{}-\textsc{gen} offspring\textsc{{}-obl}{}-\textsc{dat=attr} house=\textsc{q} yes=\textsc{cop.neg} fortress=\textsc{q} \\
    \glt ‘What is this, a piglet’s house or a fortress?’ (A; \citetitle{Tales})
\z 

\ea
    \label{ex:and:86}
    \gll \textbf{hun=sːu=gulo} den ho-r r-iqon-nja!\\
          yes=\textsc{cop.neg=emph} \textsc{1sg} \textsc{dem-v} \textsc{v}{}-break.\textsc{caus-fut} \\ 
    \glt ‘[Open the door,] or I’ll break it!’ (A; \citetitle{Tales}) 
\z 

\ea
    \label{ex:and:87}
    \gll mijarek’a, se-w-ul=gu išːi-j=le hene-b misal men-ni boso-mado, \textbf{su}, ib-ul{}-olo-j=le? \\
          lord one-\textsc{m-pl=ints} \textsc{1pl.excl-dat=q} \textsc{dem-iv} parable \textsc{2sg}{}-\textsc{erg}  tell-\textsc{prog} \textsc{cop.neg} all-\textsc{pl{}-obl.pl{}-dat=q} \\
    \glt ‘Lord, are you telling this parable to us alone, or to everyone?’ (A; \citetitle{Luke} 12:41)
\z 

\subsubsection{Endoclisis in negative verb forms and the status of \textit{sːu}} \label{sec:and:4.6.3}

As already illustrated in \sectref{sec:and:4.4} and \sectref{sec:and:4.5} above, when the additive clitic \textit{=lo} or the intensifying marker \textit{=gu} are hosted by a negative verb form, they are placed word-internally, directly before the negation marker. That is, in forms with the standard negation marker \textit{{}-sːu}, these clitics precede it (see \REF{ex:and:68}, \REF{ex:and:75}, \REF{ex:and:82}), while in the irregular forms of the negative future and the prohibitive they appear inside the cumulative tense-aspect-mood-negation marker, after \textit{-do} and before the negator \textit{{}-sːu} fused with another former morpheme (see \REF{ex:and:69} and also additional examples (\ref{ex:and:88}--\ref{ex:and:89})).\footnote{In the examples to follow, as well as in \REF{ex:and:69} above, I adopt a convention of glossing \textit{=lo} and \textit{=gu} inside synchronically unanalysable suffixes by means of angle brackets.} In the perfect / perfective converb, the clitics occur before \textit{-č’igu} (see \REF{ex:and:70}). 

\ea 
    \label{ex:and:88}
    \gll men araq’ c’ːa-do‹\textbf{gu}›sːub!\\
          \textsc{2sg} vodka drink-\textsc{proh<ints>} \\
    \glt ‘Don’t you drink vodka at all!’ (R)
\z 

\ea
    \label{ex:and:89} 
    \gll  hege{}-ši-lo bužu-do‹\textbf{lo}›sːub\\
          \textsc{dem{}-obl.m-sup.lat} believe-\textsc{proh<add>} \\
    \glt ‘[Be careful] and also do not trust him.’ (A; \citetitle{Tales})
\z 

Both \textit{=lo} and \textit{=gu} are clitics, not affixes, as they display highly “promiscuous” behaviour and, besides verb forms, can be hosted by various other constituents (including noun phrases in various cases, adjectives, adverbs, etc.). On the other hand, negative verb forms are affixal, which is especially obvious in the case of forms with irregular negation. It can thus be argued that the placement of \textit{=lo} and \textit{=gu} in negative verb forms in Andi is an instance of endoclisis (for details of the argumentation, see \citealt{Maisak2021}).\footnote{Endoclitics, a.k.a. \textit{intraclitics} or \textit{mesoclitics}, are a rare type of clitics occurring inside words, either intermorphemically or intramorphemically. They are attested in a number of languages, such as European Portuguese, the Iranian languages Pashto and Sorani Kurdish, the Delta Edoid language Degema and some Southern Mande languages. Endoclisis of person agreement markers (and also negation markers and the additive clitic) has previously been described for a Nakh-Daghestanian language, namely Udi, a Lezgic language spoken in Northern Azerbaijan and Eastern Georgia \citep{Harris2002}.} To my knowledge, in Andi only \textit{=lo} and \textit{=gu} behave as endoclitics, and it is only the negative verb forms that can host endoclisis.

The placement of \textit{=lo} and \textit{=gu} in regular negative forms before \textit{sːu} (e.g. \textit{c’ːa-do=}\textbf{\textit{lo}}\textit{=sːu}, \textit{k’amm{}-o=}\textbf{\textit{lo}}\textit{=sːu} ‘neither drinks nor eats’ with the present habitual) does not by itself provide conclusive evidence in favour of the endoclitic analysis. The reason is that it is not immediately clear whether \textit{sːu} in verb forms should be regarded as a suffix or as a clitic. To summarize, the main Andi negator \textit{sːu} occurs in the following contexts:

\begin{enumerate}[i.]
    \item as an agglutinative right-periphery marker in (most) negative verb forms, 
    \item as a suffix preceding another suffixal inflection in a few non-finite forms, 
    \item as a former suffix now fused with other morphemes into a single negative complex (in the prohibitive and the negative future markers \textit{{}-dosːub} and \textit{{}-dosːja}),
    \item as a negative copula, 
    \item as a constituent negation marker.
\end{enumerate}


Although \textit{sːu} can be considered a clitic in functions (iv) and (v), a suffix in function (ii) and a “submorph” in function (iii), the status of \textit{sːu} in (i) is not so clear. Moreover, different approaches can be taken to decide how many different \textit{sːu} morphemes we are dealing with. Historically, of course, the negative copula and the constituent negation marker, as well as the verbal negator(s), all go back to one and the same morpheme, but on a synchronic level they may represent multiple distinct morphemes (at least two, the suffixal \textit{sːu} and the clitic \textit{sːu}).

As I have argued in \sectref{sec:and:4.1}, \textit{sːu} as a verbal negator and \textit{sːu} as a constituent negation marker are probably not one and the same “movable” morpheme synchronically, as the constituent negation marker can appear in predicates which either already contain the verbal negator (meaning that there are two instances of \textit{sːu} in the same clause), or cannot take it on at all (e.g. with the verb form in \textit{\nobreakdash-č’igu}). Likewise, one probably should not treat the verbal negator \textit{sːu} synchronically as the same morpheme as the negative copula \textit{sːu}, unless one wishes to say that all negative verb forms are periphrastic (as opposed to the affirmative forms which are synthetic). First, in the perfective participle and the masdar \textit{sːu} is definitely a suffix, and cannot be a copula. Likewise, although verb-final \textit{sːu} occurs with some non-finite forms (cf. the infinitive \textit{bik’udu-sːu} ‘not to be, not to happen’), it would be strange to claim that the negative infinitive is derived from its affirmative counterpart with the help of a negative copula. An alternative analysis might be that among the negative verb forms in \textit{sːu}, some are underlying periphrastic forms with a negative copula \textit{sːu}, while others (in particular, non-finite forms) are simply suffixal, but this analysis seems to be unreasonably complicated. I conclude that there are three \textit{sːu} morphemes in modern Andi, namely a suffix found on verb forms (except some irregular ones) and two clitics, a constituent negator and a negative copula.

\subsubsection{Diachronic notes and observations} \label{sec:and:4.6.4}

The wide distribution of a single marker \textit{sːu} which is found on both declarative and non-declarative, finite and non-finite verb forms is an unusual feature of Andi, as in Andic languages verb forms do not always have regular negative equivalents and\textsf{} different negation morphemes are usually used for the encoding of different domains of negation.

According to \citet[119]{Alekseev1988} the main negation marker at the Proto-Andic level was probably *\textit{\nobreakdash-č’i}, although in Andi \textit{\nobreakdash-č’igu} is the only remnant of this proto-language morpheme. One can possibly identify the element \textit{gu} in \textit{\nobreakdash-č’igu} as diachronically related to the intensifier, but in present-day Andi \textit{\nobreakdash-č’i} cannot be used without \textit{gu} in any context, nor can \textit{\nobreakdash-č’igu} as a whole be employed in any other way except as part of the negative perfect (perfective converb). It is also not impossible that the form or the very existence of the marker \textit{\nobreakdash-č’igu} in Andi has been influenced by Avar, where the negative perfective converb has the suffix \textit{\nobreakdash-č’ogo} (and \textit{go} is an intensifier clitic in Avar).

Furthermore, as pointed out by \citet[119]{Alekseev1988}, another Proto-Andic negation marker *\textit{sːi} is generally restricted to prohibitive forms in modern Andic languages, so its expansion in Andi may be a “secondary phenomenon”.\footnote{It is not impossible that some version of the Negative Existential Cycle has been in operation in Andi and has fostered the expansion of a new standard negation marker from a negative existential (copula). On the Negative Existential Cycle, see a seminal paper by \citet{Croft1991} and a recent overview by \citet{VeselinovaHamari2022intro}, among others.} The details of the historical evolution of negation in Andi are, however, obscure, as no records of the language are available from earlier centuries.

\section{Summary} \label{sec:and:5}

A peculiarity of Andi among the closely related languages of the Avar-Andic branch is the expansion of one and the same negative marker \textit{sːu} into several domains of negation: the suffix \textit{{}-sːu} is the marker of standard negation, but it is also used with non-indicative and non-finite verb forms. Besides, the negative copula has the same form \textit{sːu}, and \textit{sːu} is employed in constituent negation as well.

Negation in the verb paradigm is symmetric in the sense that all categories have affirmative and negative forms. In most verb forms, the suffixal marker \textit{sːu} either follows or precedes another inflectional suffix, depending on the particular form. Although most negative constructions are symmetric in that “the negative does not differ from the corresponding affirmative in any other meaningful way than by the presence of the negative marker(s)” \citep[61]{Miestamo2005}, three verb forms display constructional asymmetry of the A/Cat/TAM type.

There is a series of counterexpectation verb forms which seem to go back to a periphrastic construction involving some inherently negative auxiliary verb (*\textit{b-aʁi}); in general, forms with such semantics are not common in the languages of the family.

An interesting variant of a negative reply is \textit{hunsːu} which is a combination of a positive reply ‘yes’ with a negative copula (the latter can also function as a negative reply ‘no’ on its own).

The dedicated marker of negative indefinite pronouns \textit{{}-gulo} is a lexicalized combination of two clitics, the intensifier \textit{=gu} and the additive/scalar marker \textit{=lo} ‘also, even’. There is also a dedicated adverbial caritive marker \textit{ti(gu)} ‘without having’, while in the attributive construction, caritive phrases are headed by the participle of the negative copula.

As is common in Nakh-Daghestanian languages, contrastive disjunction (‘or else’, ‘otherwise’) can be expressed by means of a negative copula, a negative reply or a negative conditional form ‘if it is not’. A conventional means of expressing ‘only X’, ‘nobody/nothing except X’ is a combination of the adverbial form of the negative copula \textit{sːugu} with clausal negation (i.e. ‘saw only X’ will be expressed as ‘did not see not X’).

\tabref{tab:and:2} summarizes the negative morphemes of Andi and the main constructions associated with them. 

\begin{table}
\small
\begin{tabularx}{\textwidth}{p{22mm}@{}Q}
\lsptoprule
{Negative\newline morphemes}/ \newline {constructions} & {Functions}\\
\midrule
{}-sːu & standard negation (indicative, non-indicative, non-finite verb forms)\\
{}-dosːja & negation of the future tense / imperfective participle (portmanteau morpheme historically including \textit{{}-sːu}, < *\textit{{}-do-sːu-ja} [-\textsc{hab-neg-ipfv.ptcp}])\\
{}-dosːub & negation of the imperative (portmanteau morpheme historically including \textit{{}-sːu}, < *\textit{{}-do-sːu-b} [-\textsc{hab-neg-imp}])\\
{}-č’igu & negation of the perfect tense / perfective converb (portmanteau morpheme, not cognate with \textit{{}-sːu})\\
{}-daʁidːu/ {}-daʁid/b-aʁij  & counterexpectation present (going back to a combination involving an inherently negative auxiliary verb *\textit{b-aʁi})\\
sːu & negative copula; also negative reply ‘no’; also constituent negation marker\\
ɬidosːja/\newline ɬidosːa & ‘cannot, may not, is not allowed’ (< ‘become’ in the negative future)\\
hunsːu & negative reply ‘no’ (< \textit{hun} ‘yes’ + negative copula \textit{sːu})\\
haj joχ, aj & negative replies ‘no’ (marginal)\\
{}-gulo & marker of negative indefinites (< intensifier \textit{=gu} + additive \textit{=lo} ‘also, even’)\\
{}-{sːub} & head of an attributive caritive phrase (‘one that does not have’, < participle of the negative copula \textit{sːu})\\
{ti(gu)} & head of an adverbial caritive phrase (‘without having’, dedicated marker with or without the adverbializer \textit{gu})\\
{t’iri} & an element occasionally used to reinforce negation (< noun ‘spatter, drop’)\\
X={sːugu}\newline+ clausal\newline negation & conventional means of expressing ‘only’, ‘nobody/nothing except’\\
{sːu}, {sːuɬibor}  & contrastive disjunction marker (‘or else’, ‘otherwise’): negative copula \textit{sːu} or its combination with a conditional form \textit{ɬibor} ‘if it is, if there is’\\
{hunsːu}(={gulo}) & contrastive disjunction marker (‘or else’, ‘otherwise’): negative reply ‘no’ or its combination with an emphatic particle\\
{ja... ja...} & phrase-initial conjunction ‘neither... nor’ in negative sentences\\
{...aj ...aj} & phrase-final conjunction ‘neither... nor’ in negative sentences (marginal)\\
\lspbottomrule
\end{tabularx}
\caption{Summary of negative morphemes and constructions}
\label{tab:and:2}
\end{table}



\section*{Acknowledgements}

My thanks are due to Chopan Gaziev and Mesedu Alisultanova for their native speaker judgements on Andi, and to Steven Kaye for the help with the English of the paper and a number of helpful suggestions. I am also grateful to two anonymous reviewers and both editors for their insightful comments.


\section*{Abbreviations}

\begin{tabularx}{.45\textwidth}{lQ}
{\scshape 1} & first person\\
{\scshape 2} & second person\\
{\scshape iii} &  third (neutral) gender \\
{\scshape iv} & fourth (neutral) gender \\
{\scshape v} & fifth (neutral) gender \\
{\scshape abstr} & abstract noun\\
{\scshape ad} & localization ‘near’ \\
{\scshape add} & additive \\
{\scshape advz} & adverbializer \\
{\scshape afft} & affective (case) \\
{\scshape aor} & aorist \\
{\scshape apud} & localization ‘close to’ \\ 
{\scshape attr} & attributivizer \\
{\scshape aux} & auxiliary \\
{\scshape car} & caritive \\
{\scshape caus} & causative \\
{\scshape com} & comitative \\
{\scshape cond} & conditional \\
{\scshape cnt} & localization ‘in contact’ \\
{\scshape cop} & copula \\
{\scshape cvb} & converb \\
{\scshape dat} & dative \\
{\scshape degr} & degree  \\
{\scshape dem} & demonstrative \\
{\scshape ela} & elative \\
{\scshape emph} & emphatic \\
{\scshape erg} & ergative \\
{\scshape ess} & essive  \\
{\scshape excl} & exclusive  \\
{\scshape f} & feminine \\
{\scshape fn} & feminine or neutral \\
{\scshape fact} & factitive \\
{\scshape fut} & future \\
{\scshape gen} & genitive \\
{\scshape h} & human \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
{\scshape hab} & habitual \\
\scshape in & localization ‘inside’ \\
{\scshape indf} & indefiniteness \\
{\scshape indir} & indirect \\
{\scshape inf} & infinitive \\
{\scshape ints} & intensifier \\
{\scshape inter} & localization ‘in a mass’ \\
{\scshape ipfv} & imperfective \\
{\scshape lat} & lative \\
{\scshape loc} & locative \\
{\scshape m} & masculine \\
{\scshape mnr} & manner adverbial \\
{\scshape msd} & masdar (action nominal) \\
{\scshape n} & neutral gender \\
{\scshape neg} & negation \\
{\scshape num} & numeral  \\
{\scshape obl} & oblique   \\
{\scshape pfv} & perfective \\
{\scshape pl} & plural \\
{\scshape prf} & perfect \\
{\scshape prog} & progressive \\
{\scshape proh} & prohibitive  \\
{\scshape prs} & present \\
{\scshape ptcl} & particle \\
{\scshape ptcp} & participle \\
{\scshape q} & polar question \\
{\scshape quot} & quotative \\
{\scshape red} & reduplication \\
{\scshape rep} & reportative \\
{\scshape sg} & singular \\
{\scshape sup} & localization ‘on top’ \\
{\scshape cnte} & counterexpectation \\
{\scshape vbz} & verbalizer \\
{\scshape voc} & vocative \\
{\scshape wh} & content question \\
\end{tabularx}
\medskip

\noindent

Unmarked values (absolutive, singular) are not glossed.


\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
\end{document} 
