\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233752}
\author{Martine Vanhove\affiliation{CNRS-LLACAN}}
\title{Negation in Beja}
\abstract{This chapter presents the negative constructions of Beja, a North-Cushitic (Afroasiatic) language, which are basically asymmetrical. It discusses in turn 
(i) standard negation in declarative main clauses with A/Cat/TAM asymmetry; 
(ii) the two negative types in non-declaratives, one of which does not use a negator but stem alternation and specific inflection; 
(iii) negation in non-main clauses (even more asymmetrical than declaratives); 
(iv) privatives, which include one case of clause chaining instead of a negator; 
(v) phasal polarity expressions;
(vi) negation in stative predications; 
(vii) negative replies and their pragmatics. It also discusses the absence of negative transport, as well as of negative indefinites and the strategies used instead.

\keywords{declarative, non-declarative, clausal, non-clausal, grammaticalization}}

\IfFileExists{../localcommands.tex}{
 \addbibresource{../localbibliography.bib}
 \input{../localpackages}
 \input{../localcommands} 
 \input{../localhyphenation} 
 \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 

\section{The language}

Beja (ISO: \href{https://iso639-3.sil.org/code/bej}{bej}, Glottocode: \href{https://glottolog.org/resource/languoid/id/beja1238}{beja1238}), locally named \textit{biɖawijeːt}, is the sole language of the North-Cushitic branch of the Afroasiatic phylum. It is lexically and grammatically quite distant from its closest Lowland East and Central Cushitic relatives, Afar-Saho and Agaw, and is considered as peripheral within the Cushitic family. It is spoken in the northernmost part of the Cushitic-speaking area, mainly in Eastern Sudan between the Red Sea and the river Atbara by some 2,000,000 speakers, and in Northern Eritrea (approx. 60,000 speakers). It used to be spoken in Southern Egypt, but it seems speakers have now shifted to Arabic.

Beja dialectology is ill-known, but dialects do not seem much differentiated. Three main dialectal zones are identified \citep{Morin1995}: North, Centre and South, which have further local and tribal-based subdivisions \citep{Wedekind2012}. Dialects are differentiated on the basis of isoglosses based on vowel variation, pronominal sets, morphological use of pitch accent for plural formation, accommodations of Arabic borrowings and some lexical peculiarities.

Beja has had intensive contacts with Arabic (Central Semitic) and Tigre (North Ethiosemitic branch of South Semitic), probably also with Nubian (Nilo-Saharan). Today, in Sudan, where I did my fieldwork, bilingualism with Sudanese Arabic is widespread and expanding. However, this is less so for women who lead a cloistered life. Only a minority of the Bejas have knowledge of Tigre (mostly in Eritrea).

In Sudan, formal education is increasing among young villagers and city dwellers, but girls seldom go further than primary level. The sole language of education is classical Arabic, with a mixture of colloquial Sudanese Arabic, and more recently also with some oral explanations in Beja \citep{Onour2015}. In addition, Quranic Arabic is taught in Quranic schools to young boys and girls alike.\footnote{The sociolinguistic situation described here is that which prevailed before the current war in Sudan between rival armed groups, which broke out in April 2023.}

The typological profile that follows is a brief summary of \citet{Vanhove2017}. Beja has a rich and complex morphology, inflectional and derivational, both in the nominal and verbal domains. It is partly templatic for verb inflection and derivation, noun formation, verb-noun derivation, adjective and plural formation. Beja is also partly agglutinative-fusional. Suffixes and enclitics form the majority of the inflectional and derivational devices. Prefixes and proclitics are also used. In all cases, they are often portmanteau morphemes, as e.g. the definite articles.

Beja is a marked nominative language with four nominal cases, two for the verb's core arguments, nominative and accusative, and two for noun phrases, genitive and vocative. The case of core arguments is marked on determiners by vocalic ablaut (or a consonant in one marginal case); the genitive and the vocative are marked by vocalic suffixes on nouns. Pronouns have in addition dative and ablative sets, but no vocative form.

Verbs can be finite or non-finite. Finite forms are organized along a three-term aspectual system. It distinguishes, through inflectional morphemes and apophony in the stem, Perfective, Imperfective and Aorist, which index also person, number and gender (only in 2 and 3 \textsc{sg}) of the subject. These verb forms are enriched by auxiliaries or a copula to form other TAM forms, among them the Perfect. There are two morphological verb classes. V1 contains the majority of verbs (57\%), and has prefixed paradigms for monosyllabic stems, and an infix for disyllabic stems (plural indices are suffixes). This is historically the oldest class. V2 verbs have only suffixes, and represent a common Cushitic innovation. There is in addition a rich system of semantic and voice derivation involving ablaut, reduplication and affixal devices (pluractional, intensive, middle, passive, reciprocal, causative, and marginally double causative and double pluractional). The non-finite forms amount to four converbs labelled General, Sequential, Simultaneity and Manner. They are invariable, except for the Manner converb, which may vary for gender in specific syntactic environments. The converbs are used in deranked subordinate clauses and the Manner and General converbs also as auxiliated forms in complex predicates.

Syntactically, Beja is predominantly head-final; the canonical constituent order is (X)(S)(O)V, and dependent clause – main clause. Constituent order is not particularly rigid and may vary for pragmatic reasons.

The data used for this paper come from semi-spontaneous oral texts that I recorded in Sudan, mainly in Sinkat and Wagar, between 2000 and 2011. Part of them are available online at \url{http://corporan.huma-num.fr/Archives/}. The structure of the paper follows the questionnaire provided by the editors (\textcitetv{chapters/questionnaire}).

\section{Clausal negation}

\subsection{Standard negation}
\label{sec:beja:2.1}
In declarative verbal main clauses aspect is neutralized. The five indicative tenses of declarative utterances, Imperfective, Perfective, Aorist, Perfect and Future are reduced to three in the negative polarity: the Perfective, Aorist and Perfect use the paradigm of the Perfect for the core verb and a negated auxiliary \REF{ex:beja:2}. Beja thus exhibits a paradigmatic asymmetry which corresponds to \citegen{Miestamo2005} A/Cat/TAM type. In addition, there is also a constructional asymmetry: the Imperfective negative stem and paradigm are based on the Perfective \REF{ex:beja:1b}. This is because during the course of the general evolution of the verb system, a new Imperfective paradigm was introduced, which resulted in a dramatic change in the aspectual system: the former Perfective became an Aorist while the former Imperfective became a Perfective.\footnote{A detailed discussion of this evolution is found in \citet{Cohen1972, Cohen1973}.}

Although all three negative forms share the same proclitic negative particle \textit{ki=} (\textit{ka=} in 1\textsc{sg} because of vowel harmony), the position of the negator is different: The negator directly precedes the verb in the Imperfective \REF{ex:beja:1}, but it precedes an auxiliary in the Perfective form for the other four TAM forms: \textit{di} ‘say’ for the Future \REF{ex:beja:3}, \textit{ak} ‘become’ for the other three paradigms \REF{ex:beja:2}. In the affirmative, the Perfect goes back to a complex predicate with a Manner converb and a non-verbal copula. In the negative, the copula is replaced by the verb \textit{ak} ‘become; be’ which hosts the negator. The Future affirmative is also a complex predicate with an auxiliary: The core verb originates from a frozen form of the Aorist (which still varies for number in the central dialect), followed by the verb \textit{di} ‘say’ (whose plural forms have a short \textit{-a-} in the stem of the Imperfective instead of a long \textit{-aː-}). Compare examples \REF{ex:beja:1}-\REF{ex:beja:3}.\footnote{The transcription follows the IPA except for geminated consonants which are written as a double consonant. Africanists and Cushitists, who are used to other transcription systems, should be aware that \textit{j} transcribes the semi-consonant, not an affricate (noted as \textit{ʤ}).}

\ea\label{ex:beja:1}
Imperfective \\
    \ea\label{ex:beja:1a} Affirmative \\
        \gll handi wari=b ha<n>riːw\\
             tree other\textbackslash\textsc{pl=indf.m.acc} seek\textsc{<ipfv>}\textsc{[3sg.m]}\\
        \glt `He looks/is looking for other pieces of wood.' \\

    \ex\label{ex:beja:1b} Negative \\
        \gll handi wari=b ki=hariw\\
             tree other{\textbackslash}\textsc{pl=indf.m.acc} \textsc{neg.ipfv}=seek\textsc{{\textbackslash}pfv.\textsc[3sg]}\\
        \glt `He does not look/is not looking for other pieces of wood.' [\textsc{narr\_54\_aa}\_gazelle\_15–16]\footnote{Examples extracted from my online corpus are referred to by text number (54), genre (\textsc{narr}), short title (gazelle), and the number of the intonation unit(s) (15–16). Elicited examples have no specific mention.}
    \z
\z 

\ea\label{ex:beja:2}
Perfective, Perfect, Aorist \\
    \ea\label{ex:beja:2a} Perfective affirmative \\
        \gll lawwaːw-ani \\
          prowl\textsc{.ints}-\textsc{pfv.1sg} \\
        \glt `I prowled around him.'
    \ex\label{ex:beja:2b} Perfect affirmative\\
        \gll lawwaːw-aː-b-i\\
             prowl\textsc{.ints}-\textsc{prf-m-1sg}\\
        \glt `I have prowled around him.'
    \ex\label{ex:beja:2c} Aorist affirmative \\
        \gll lawwaːw-i\\
             prowl\textsc{.ints}-\textsc{aor.1sg}\\
        \glt `I used to prowl around him.'
    \ex\label{ex:beja:2d} Perfective/Perfect/Aorist negative\\
        \gll lawwaːw-aː-b ka=a-ki\\
             prowl\textsc{.ints}-\textsc{prf-m} \textsc{neg.ipfv=1sg-}become{\textbackslash}\textsc{pfv}\footnotemark{} \\
        \footnotetext{For the sake of glossing consistency, the enclitic negator \textit{ki=} / \textit{ka=} is systematically glossed as \textsc{neg.ipfv}, even when occurring in periphrastic tenses that do not express the Imperfective aspect.}
        \glt `I did not prowl around him.' / `I have not prowled around him.' / `I did not use to prowl around him.' [\textsc{narr}\_04\_djinn\_184]
    \z
\z 

\protectedex{
\ea\label{ex:beja:3} 
Future \\
    \ea\label{ex:beja:3a} Affirmative\\
        \gll jam gʷʔa-s-i=hoːk a-<n>di\\
             water drink-\textsc{caus-fut.sg=obj.2sg} \textsc{1sg-}say<\textsc{ipfv>}\\
        \glt `I will give you water.'
    \ex\label{ex:beja:3b} Negative \\
        \gll jam gʷʔa-s-i=hoːk ka=a-di\\
             water drink-\textsc{caus-fut.sg=obj.2sg} \textsc{neg.ipfv=1sg}{}-say{\textbackslash}\textsc{pfv}\\
        \glt `I won’t give you water.' [\textsc{narr}\_39\_Bafalib\_26]
    \z         
\z
}

Beja has no dedicated marker or paradigm for negation of phasal polarity expressions. It may use instead the standard negation, together with adverbs. None of them is obligatory, and they are indeed quite rare in my data.

Three adverbs, \textit{w=ʔagija} ‘anew’ \REF{ex:beja:4}, \textit{ʃaːwi} ‘again’ \REF{ex:beja:5}, and \textit{mhasi} ‘still, again, yet, ever’  may be used in the negative polarity. Only \textit{ʃaːwi} can be postposed to the verb. Note that \textit{w=ʔagija} behaves morphologically as a noun and hosts the definite article.

\ea\label{ex:beja:4}
Phasal polarity expression with \textit{w=ʔagija} \\
    \gll tʔa w=ʔagija iː-bɖan ka=a-di\\
         now \textsc{def.sg.m}=anew \textsc{fut.sg}{}-forget{\textbackslash}\textsc{ints} \textsc{neg.ipfv=1sg-}say{\textbackslash}\textsc{pfv}\\
    \glt `Now, I won’t forget him anymore!' [\textsc{narr}\_13\_grave\_080]
\z 

\ea\label{ex:beja:5}
{  Phasal polarity expression with \textit{ʃaːwi}}\\
    \gll ani=wa baruːk=wa ki=n-am-da{\textasciitilde}dʔar ʃaːwi\\
         \textsc{1sg.nom=coord} \textsc{2sg.m.nom=coord} \textsc{neg.ipfv=1pl-recp-plac}{\textasciitilde}marry{\textbackslash}\textsc{pfv} again\\
    \glt `Me and you, we are not husband and wife anymore.' [\textsc{narr}\_23\_lezard\_075–076]
\z


Another type of phasal polarity expression expressed with the standard negation is ‘not yet’, and Beja has no special ‘not yet’ tense. Such a meaning is very rarely overtly expressed in my data. Below \REF{ex:beja:6} is an example with the negative Imperfective followed by the adverb \textit{nuːn} ‘only’.

\protectedex{
\ea\label{ex:beja:6}
{  ‘Not yet’} \\
    \gll eːra-m-aː-b winneːt ki=i-ki nuːn\\
         white-\textsc{mid-cvb.mnr-indf.m} plenty \textsc{neg.ipfv=3sg.m}{}-become{\textbackslash}\textsc{pfv} only\\
    \glt `It is not broad day light yet.' (lit. `It is not becoming very white only.') [\textsc{narr}\_05\_eritrea\_387]
\z
}

\subsection{Negation in non-declaratives}

\subsubsection{Prohibitive}
\label{sec:beja:2.2}
Imperatives have a dedicated negative construction, different from standard negation. The Prohibitive is formed with the negative proclitic particle \textit{baː=} (\textsc{2sg.m, 2pl}) / \textit{biː=} (\textsc{2sg.f}). Inflectional morphemes are all suffixes, even for the verb class conjugated with prefixes in the indicative (V1). They differ slightly between V1 and V2, and are similar (with one minor difference for V2 \textsc{2sg.f}) to those of the Imperative, as shown in \tabref{tab:beja:1}.

\begin{table} 
\begin{tabularx}{.65\textwidth}{X ll ll}
\lsptoprule
& \multicolumn{2}{c}{{\bfseries V1}} & \multicolumn{2}{c}{{\bfseries V2}}\\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
& {\bfseries\scshape imp} & {\bfseries\scshape proh} & {\bfseries\scshape imp} & {\bfseries\scshape proh}\\
\midrule 
{\scshape sg.m} & {{}-\textit{a}} & {\textit{baː}=V1-\textit{a}} & {{}-\textit{a}} & {\textit{baː}=V2-\textit{a}}\\
{\scshape sg.f} & {{}-\textit{i}} & {\textit{biː}=V1-\textit{i}} & {{}-\textit{i}} & {\textit{biː}=V2-\textit{aj}}\\
{\scshape pl} & \textit{{}-na} & {\textit{baː}=V1-\textit{na}} & \textit{{}-aːna} & {\textit{baː}=V2-\textit{aːna}}\\
\lspbottomrule
\end{tabularx} 
\caption{Imperative vs Prohibitive}
\label{tab:beja:1}
\end{table}

One further difference between Imperative and Prohibitive concerns the stem of V1 (i.e. the verbal base without inflectional morphemes), which is similar to that of the Imperfective for monosyllabic stems, CiːC, and partly similar to the singular stem of the Imperfective for disyllabic stems, minus the aspectual inflectional infix \textit{<n>}: CaCiːC (\textit{vs} Ca<n>CiːC in the Imperfective). The resulting opposition between the Imperative and Prohibitive stems is a difference in vowel length in the stem of V1 verbs: Short in the Imperative, long in the Prohibitive \REF{ex:beja:7}. There is thus also a paradigmatic asymmetry between the Imperative and the Prohibitive for V1.

The Prohibitive of V2 belongs to the most common cross-linguistic strategy described by \citet{AuweraEtAl2013} as using “the verbal construction of the second singular imperative and a sentential negative strategy not found in (indicative) declaratives”. On the other hand, V1 Prohibitive belongs to the second next most common strategy where “the prohibitive uses a verbal construction other than the second singular imperative and a sentential negative strategy not found in (indicative) declaratives”. 

Below are examples of the Prohibitive for V1 \REF{ex:beja:7} and V2 \REF{ex:beja:8}.

\protectedex{\ea\label{ex:beja:7}
{V1 Prohibitive} \\
    \gll baː=hariːd-a\\
     \textsc{neg.proh}=slaughter-\textsc{imp.sg.m}\\
    \glt `Don’t slaughter it!' [\textsc{narr}\_50\_fox\_hunt\_235]
    \\ (cp. Imperative \textit{harid-a}! `slaughter it!’)
\z }

\ea\label{ex:beja:8}
V2 Prohibitive\\
    \gll ti-bari=jeː=na baː=raːt-a\\
     \textsc{2sg.m}{}-have{\textbackslash}\textsc{ipfv=rel}=thing \textsc{neg.proh}=ask-\textsc{imp.sg.m}\\
    \glt `Don’t ask for what you have!’ [\textsc{narr}\_04\_djinn\_156]
    \\ (cp. Imperative \textit{raːt-a} `ask!')
\z

\subsubsection{Negative Optative}

Beja has a second mood paradigm, that of the Optative. It is also asymmetrical in terms of polarity. The Optative, affirmative and negative, is marked by the proclitic particle \textit{bi=} (which harmonizes with the 1\textsc{sg} inflectional prefix \textit{a-}). For both V1 and V2, the paradigms are different. In the affirmative Optative \textit{bi=} cliticizes to the Aorist paradigms of V1 and V2. In the negative Optative, \textit{bi=} \textit{/ ba=} cliticizes to other inflectional morphemes. V1 verbs have in addition a different stem. Both V1 and V2 have prefixed personal indices, identical to those of the indicative Perfective of V1. For V2, the stem, which I recall is invariable, is in addition followed by specific suffixes, except 2\textsc{pl} and 3\textsc{pl}, which have the same plural suffixes as the Aorist. The stems of V1 are identical to the Prohibitive stems. Thus, what distinguishes polarities in the Optative is not a negative proclitic as with other TAM, but the stem difference for V1, and the inflectional affixes for V2.

Compare Tables \ref{tab:beja:2} and \ref{tab:beja:3}, which provide the full paradigms of the Optative, affirmative and negative respectively.

\begin{table}
\begin{tabularx}{\textwidth}{lXXl}
\lsptoprule 
& {\bfseries V2} & {\bfseries V1 monosyllabic} & {\bfseries V1 disyllabic} \\
& { ‘eat’} & { ‘leave’} & { ‘arrive’}\\
\midrule 
{\scshape 1sg} & \textit{bi=tam-i} & \textit{bi=iː-dif} & \textit{bi=iː-ktim}\\
{\scshape 2sg.m} & \textit{bi=tam-tija} & \textit{bi=tiː-dif-a} & \textit{bi=tiː-ktim-a}\\
{\scshape 2sg.f} & \textit{bi=tam-tiː} & \textit{bi=tiː-dif-i} & \textit{bi=tiː-ktim-i}\\
{\scshape 3sg.m} & \textit{bi=tam-i} & \textit{bi=iː-dif} & \textit{bi=iː-ktim}\\
{\scshape 3sg.f} & \textit{bi=tam-ti} & \textit{bi=tiː-dif} & \textit{bi=tiː-ktim}\\
{\scshape 1pl} & \textit{bi=tam-ni} & \textit{bi=niː-dif} & \textit{bi=niː-ktim}\\
{\scshape 2pl} & \textit{bi=tam-tiːn(a)} & \textit{bi=tiː-dif-na} & \textit{bi=tiː-ktim-na}\\
{\scshape 3pl} & \textit{bi=tam-iːn(a)} & \textit{bi=iː-dif-na} & \textit{bi=iː-ktim-na}\\
\lspbottomrule
\end{tabularx}
\caption{Affirmative Optative paradigms}
\label{tab:beja:2} 
\end{table}

\begin{table}
\begin{tabularx}{\textwidth}{lXXl}
\lsptoprule
& {\bfseries V2} & {\bfseries V1 monosyllabic} & {\bfseries V1 disyllabic}\\
& { `eat'} & { `leave'} & { `arrive'}\\
\midrule 
{\scshape 1sg} & \textit{ba=a-tam-aj} & \textit{ba=a-diːf} & \textit{ba=a-katiːm}\\
{\scshape 2sg.m} & \textit{bi=t-tam-aja} & \textit{bi=t-diːf-a} & \textit{bi=t-katiːm}\\
{\scshape 2sg.f} & \textit{bi=t-tam-aj} & \textit{bi=t-diːf-i} & \textit{bi=t-katiːm}\\
{\scshape 3sg.m} & \textit{bi=i-tam-aj} & \textit{bi=i-diːf} & \textit{bi=i-katiːm}\\
{\scshape 3sg.f} & \textit{bi=t-tam-aj} & \textit{bi=t-diːf} & \textit{bi=i-katiːm}\\
{\scshape 1pl} & \textit{bi-n-tam-aj} & \textit{bi=n-diːf} & \textit{bi=n-katiːm}\\
{\scshape 2pl} & \textit{bi=t-tam-iːna} & \textit{bi=t-diːf-na} & \textit{bi=t-katiːm-na}\\
{\scshape 3pl} & \textit{bi=i-tam-iːna} & \textit{bi=i-diːf-na} & \textit{bi=i-katiːm-na}\\
\lspbottomrule
\end{tabularx}
\itshape
\caption{Negative Optative paradigms}
\label{tab:beja:3}
\end{table}

Below are examples of the negative Optative (\ref{ex:beja:9}--\ref{ex:beja:10}). Note that the Optative also expresses deontic modalities with 3\textsuperscript{rd} persons, such as ability \REF{ex:beja:11} or obligation \REF{ex:beja:12}.

\protectedex{\ea\label{ex:beja:9}
V2 \\
    \gll ti=dirʔa han ba=a-ʃaga-am-aj \\
         \textsc{def.f}=field also \textsc{opt=1sg}-work-\textsc{mid-neg.opt} \\
    \glt `Let me not work in the field anymore!' [\textsc{narr}\_02\_farmer\_143]
    \\ (cp. affirmative Optative \textit{bi-ʃaga-am-i} `let me work!') 
\z}

\protectedex{
\ea\label{ex:beja:10}
V1\\
    \gll oːn i=kʷiːri bi=n-hariːd\\
         \textsc{prox.sg.m.acc} \textsc{def.m}=ostrich \textsc{opt=1pl}-slaughter{\textbackslash}\textsc{neg.opt}\\
    \glt `Let’s not slaughter this ostrich!' [\textsc{narr}\_38\_ostrich\_040–041]
    \\ (cp. affirmative Optative \textit{bi-niː-hrid} `let us slaughter!') 
\z }

\ea\label{ex:beja:11}
Deontic ability (V2)\\
    \gll bani ʔaːdam han bi=i-hass-aj\\
         son Adam also \textsc{opt=3sg.m}{}-pass-\textsc{neg.opt}\\
    \glt `A human being could not even pass!' [\textsc{narr}\_05\_eritrea\_178]
    \\ (cp. affirmative Optative \textit{bi-hass-i} `let him pass!') 
\z 

\ea\label{ex:beja:12} 
Deontic obligation (V2)\\
    \gll harʔiː=siːsi bi=t-jʔ-a=heːb\footnotemark{} \\
         after=\textsc{poss.3sg.abl} \textsc{opt=3sg.f}-come-\textsc{neg.opt=obj.1sg}\\
    \glt `She must not come to me from behind it.' [\textsc{narr}\_05\_eritrea\_329]
    \\ (cp. affirmative Optative \textit{bi-ʔeː-ti} `let her come!') 
    \footnotetext{The final \textit{j} of the negative Optative suffixes of V2 drops before object enclitics.}
\z

\subsubsection{Negative questions}

Negative questions have the same morphological and constructional properties as standard negation. They are used to force a positive answer in a polemical situation, which is not the case with positive questions. Like in English, they need not be answered positively. 

In \REF{ex:beja:13} a rich and proud man is angry because a young boy did not bow in front of him. In \REF{ex:beja:14} two persons are arguing about moral behaviour and rules of honour: One speaker tries to force the other to agree with him, which he does not.

\ea\label{ex:beja:13}
{Negative question with positive reply}\\
    \gll aneːb ki=t-kan=heːb han\\
         \textsc{1sg.acc} \textsc{neg.ipfv=2sg.m}-know{\textbackslash}\textsc{mid.pfv=obj.1sg} \textsc{q.plr}\\
    \glt `Don’t you know \textsc{me?}' [\textsc{narr}\_31\_king\_40]
\z 

\protectedex{
\ea\label{ex:beja:14}
{Negative question with negative reply}\\
    \gll w=hagg=oː hoːj i-jaj-n=eːt toː=na ki=t-ti han […]\\
         \textsc{def.sg.m}=right=\textsc{poss.3sg.acc} \textsc{loc} 3-take{\textbackslash}\textsc{ipfv-pl=rel.f} \textsc{def.sg.f.acc}=thing \textsc{neg.ipfv=3sg.f}-become{\textbackslash}\textsc{pfv} \textsc{q.plr}\\
    \glt `Should they not stand for their rights?' (lit. `Is it not that they take from him their rights?') [\textsc{conv}\_01\_rich\_\textsc{sp1}\_214–215]
\z}

\subsection{Negation in stative predications}

Stative predications have no special negators. They are the same as the ones in standard and non-declarative negations. Nevertheless the constructional entities and the type of asymmetry differ.

\subsubsection{Equational, proper inclusion and attribution clauses}

In the affirmative, equational, proper inclusion and attribution clauses are all expressed with a non-verbal copula, enclitic to its host, which varies for person, gender and number (with some conflation), but not for TAM, as shown in \tabref{tab:beja:4}.

\begin{table}
\begin{tabularx}{0.5\textwidth}{XXl}
\lsptoprule
& {\bfseries \textsc{sg}} & {\bfseries \textsc{pl}}\\
\midrule 
{ 1 \& 3} & \textit{=i / =u} & \textit{=a}\\
{ 2} & \textit{=wa / =uːj} & \textit{=aːna}\\
\lspbottomrule
\end{tabularx}
\itshape
\caption{Paradigm of the copula}
\label{tab:beja:4}
\end{table}

The negative polarity shows constructional asymmetry, since the copula is replaced by \textit{ak,} an irregular verb meaning ‘be, become’.\footnote{The verb \textit{ak} is also used in independent affirmative clauses, as in \REF{ex:beja:38} below.} The TAM used with \textit{ak} depends on the clause type: Indicative paradigms in main and independent clauses (\ref{ex:beja:16}--\ref{ex:beja:17}), but Optative in dependent clauses (\REF{ex:beja:15}; see \sectref{sec:beja:2.4}). Note that there is no overtly marked difference between permanent and temporary property assignment.

\protectedex{\ea\label{ex:beja:15}
{Equational}\\
    \ea\label{ex:beja:15a}
    {Affirmative}\\
        \gll t=hikuːma ʔaraw=eː=ja\\
             \textsc{def.f=}power friend{\textbackslash}\textsc{pl=poss.1pl.acc=cop.3pl}\\
        \glt `The policemen are my friends.' [\textsc{narr}\_08\_drunkard\_064]
    
    \ex\label{ex:beja:15b}
    {Negative}\\
        \gll w=ʔaraːw=oː ba=a-kaj=eːt toː=na ti-kan=heːb\\
             \textsc{def.sg.m}=friend=\textsc{poss.3sg.acc} \textsc{opt=1sg-}become{\textbackslash}\textsc{neg.opt=rel.f} \textsc{def.sg.f.acc}=thing \textsc{3sg.f-}know{\textbackslash}\textsc{mid.pfv=obj.1sg}\\
        \glt `She knew that I was not her boyfriend.' [\textsc{narr}\_01\_shelter\_133–137]
    \z
\z }    

\ea\label{ex:beja:16}
{Proper inclusion}\\
    \ea\label{ex:beja:16a}
    {Affirmative}\\
        \gll   t=ʔanoː=t=a\\
             \textsc{def.f}=sheep=\textsc{indf.f=cop.3pl}\\
        \glt `They are ewes.' [\textsc{narr}\_02\_farmer\_119]
    
    \ex\label{ex:beja:16b}
    {Negative}\\
        \gll  tak ka=a-ki\\
             man \textsc{neg.ipfv=1sg}{}-become{\textbackslash}\textsc{pfv}\\
        \glt `I am not a man.' [\textsc{narr}\_25\_orphan\_263]
    \z
\z 

\ea\label{ex:beja:17} 
{Attribution}\\
    \ea\label{ex:beja:17a}
    Affirmative\\
        \gll uː=tak amaːg=i\\
             \textsc{def.sg.m.nom=}man bad=\textsc{cop.3sg} \\
        \glt `The man is bad.' [\textsc{conv}\_01\_rich\_\textsc{sp2}\_049]
    
    \ex\label{ex:beja:17b}
    Negative\\
        \gll ʃawaːj ka=a-ki\\
             free \textsc{neg.ipfv=1sg}{}-become{\textbackslash}\textsc{pfv}\\
        \glt `I am not free.' [\textsc{narr}\_18\_Adam\_devil\_285]
    \z
\z

\subsubsection{Existential and locative predications}
\label{sec:beja:2.3.2}

Existential and locative predications are expressed with three locative verbs \textit{faj}, \textit{haj} and \textit{daː} ‘be there’, depending mainly on the dialect and on morphological constraints. They are marked for TAM with some restrictions: \textit{haj} has no affirmative Perfective, no Imperative and no 1\textsc{sg} Imperfective; \textit{faj} has no Perfective. These predications have only two functions: existential and locative. None can be used as a privative, or as a short answer.

Existential and locative negative predications belong to the second most frequent type of \citet{Veselinova2013lexicalization}, i.e. there is no special existential negator. However, \textit{faj} cannot be negated. Only the two other verbs can be negated, as in \REF{ex:beja:18} and \REF{ex:beja:18+}. Moreover, in my data, the use of \textit{daː} is restricted to abstract notions in existential and locative constructions (\ref{ex:beja:19}--\ref{ex:beja:20}). These two verbs, although borderline cases, nevertheless show that negation of existence is not completely symmetrical to the affirmative.

\ea\label{ex:beja:18}
\ea \textit{haj} existential affirmation \\
    \gll ʔeːtniː=t=uːk na t-ha=eːk\\
    come\_\textsc{ipfv.3sg.f=indf.f=poss.2sg.nom} thing \textsc{3sg.f}-be\_there\_\textsc{ipfv}=if\\
    \glt ‘If there is something that happens to you …’ \textsc{[conv}\_01\_rich\_sp2\_216]
    \ex {\textit{haj} existential negation}\\
    \gll mʔari=t han  oː=doːr beːb ki=t-haːj=it\\
         food=\textsc{indf.f} also  \textsc{def.sg.m.acc}=time \textsc{dist.sg.m.acc} \textsc{neg.ipfv=3sg.f-}be\_there{\textbackslash}\textsc{pfv=csl}\\
    \glt `Since there is no food at that time either…' \textsc{[narr}\_46\_tirik\_062]
\z \z

\ea\label{ex:beja:18+}
\ea {\textit{haj} locative affirmation}\\
    \gll amaːg	darab=iːb	t-haj-a \\
        bad	road=\textsc{loc.sg}	\textsc{2sg}-be\_there\_\textsc{ipfv-m} \\
    \glt ‘You are going astray.’ (lit. ‘You are in a bad road.’) [\textsc{narr}\_08\_drunkard\_029]
\ex {\textit{haj} locative negation}\\
    \gll suːr jiwaːʃ hoːj ki=haːj \\
        before dirt \textsc{loc} \textsc{neg.ipfv}=be\_there\_\textsc{pfv.[3sg.m]} \\
    \glt ‘Before, there was no dirt in it.’ [\textsc{narr}\_27\_goat\_172-173]
\z \z

\ea\label{ex:beja:19}
\ea {\textit{daː} existential affirmation}\\
    \gll w=ʔiʃa han daː-jiːni \\
        \textsc{def.sg.m}=evening\_prayer also be\_there-\textsc{ipfv.3sg.m} \\
    \glt ‘There is also the evening prayer.’ [\textsc{narr}\_05\_eritrea\_195]
\newpage
\ex {\textit{daː} existential negation}\\
    \gll ittifaːg bi=i-daː-aj=eːt toː=naː=t=i\\
         agreement \textsc{opt=3sg.m}{}-be\_there-\textsc{neg.opt=rel.f} \textsc{def.sg.f.acc}=thing=\textsc{indf.f=cop.3sg}\\
    \glt `There should not be an agreement.' (lit. `The thing is that there is no agreement.') [\textsc{conv}\_01\_rich\_\textsc{sp1}\_192]
\z  \z

\ea\label{ex:beja:20}
\ea {\textit{daː} locative affirmation}\\
    \gll khoːrʔarab=wa mariː daːj-iːni \\
        Khor\_Arab=\textsc{coord} from\_the\_direction be\_there-\textsc{ipfv.3sg.m} \\
    \glt ‘It is in the direction of Khor Arab.’ [\textsc{narr}\_03\_camel\_110]
\ex {\textit{daː} locative negation}\\
    \gll mi-rkʷaj hoː ka=da-ja nuːn i=ginʔ=iːb\\
        \textsc{acn}-fear \textsc{1sg.dat} \textsc{neg.ipfv}=be\_there-\textsc{pfv.3sg.m} only \textsc{def.m}=heart=\textsc{loc.sg} \\
    \glt `Nevertheless I am not afraid in my heart.' (lit. `fear is not in my heart to me' = `I am courageous.') [\textsc{narr}\_05\_eritrea\_220]
\z \z 

\subsubsection{Possessive predication}

Possessive predication is expressed with a ‘have’ verb, \textit{biri}, which is defective and, like a handful of verbs such as \textit{faj} in \sectref{sec:beja:2.3.2}, shows traces of the previous aspectual system: It has no Perfective, and the Aorist (historically the former Perfective) is used instead. Consequently, the Imperfective negative, with the same negators as in Sections \sectref{sec:beja:2.1} and \sectref{sec:beja:2.2}, is built on the Imperfective paradigm and stem (instead of the Perfective for regular V1 verbs), as shown in \REF{ex:beja:21} below.

\ea\label{ex:beja:21}
{Possession}\\

    \ea\label{ex:beja:21a}
    {Affirmative}\\
        \gll nʔaːj=t iː-bri tak harʔiː=isaj ʔeːja \\ 
             goat=\textsc{indf.f} \textsc{3sg.m}-have{\textbackslash}\textsc{aor} man  after=\textsc{poss.3sg.abl} come{\textbackslash}\textsc{pfv.3sg.m}\\
        \glt `A man who had a goat came from behind him.' [\textsc{narr}\_28\_pear1\_21-23]
    
    \ex\label{ex:beja:21b}
    {Negative}\\
        \gll ahamad i=ragad-a ki=i-bari\\
             Ahmed \textsc{def.m}=leg\textsc{-pl} \textsc{neg.ipfv=3sg.m-}have{\textbackslash}\textsc{ipfv}\\
        \glt `Ahmed has no legs.' [\textsc{narr}\_05\_eritrea\_022]
    \z
\z


\subsection{Negation in non-main clauses}
\label{sec:beja:2.4}

In all balanced subordinate clauses with finite verb forms in whatever TAM, be they relative, complement or adverbial clauses, only the non-declarative construction with the negative Optative paradigm can be used, one more case of asymmetry in the polarity system. In contrast, their positive counterparts also license an Indicative TAM. The TAM interpretation of the non-main negative clauses can only be deduced from the context, as shown in (\ref{ex:beja:22}--\ref{ex:beja:25}). 

\ea\label{ex:beja:22} 
{Relative clause}\\
    \gll naː=t bi=t-katiːm=i mhiːn\\
         thing=\textsc{indf.f} \textsc{opt=3sg.f}-arrive{\textbackslash}\textsc{neg.opt=rel} place\\
    \glt `[The donkey stopped in] a place where nothing (can in general) arrive.' [\textsc{narr}\_05\_eritrea\_183]
\z

\ea\label{ex:beja:23}
{Complement clause}\\
    \gll ɖaːb-i bi=i-diː=jeːb hiːs-an\\
         run-\textsc{fut.sg} \textsc{opt=3sg.m}-say{\textbackslash}\textsc{neg.opt=rel.m} think-\textsc{pfv.1sg}\\
    \glt `I thought it would not be able to run.' [\textsc{narr}\_03\_camel\_153]
\z 

\ea\label{ex:beja:24}
{Conditional clause}\\
    \gll naː=t hoːk bi=i-dʔiː-n=eːk\\
         thing=\textsc{indf.f} \textsc{2sg.dat} \textsc{opt=3}-do{\textbackslash}\textsc{neg.opt-pl}=if \\
    \glt `If they don’t do anything to you [we are going to complain to God.]' [\textsc{narr}\_08\_drunkard\_067]
\z 

\ea\label{ex:beja:25}
{Temporal clause}\\
    \gll bi=i-m-takʷaːkʷ-n=hoːb\\
         \textsc{opt=3-recp}-repair{\textbackslash}\textsc{ints.neg.opt-pl}=when\\
    \glt `[There was a man, and he had a wife, and they did not get along well.] As they were not getting along well…' [\textsc{narr}\_24\_chief\_010]
\z

\newpage
Negation may also occur in one type of deranked subordinate clause, that with the simultaneity converb.\footnote{Note that the other three converbs cannot be negated. Negation is formed with balanced subordinate clauses with a verb in the negative Optative, another type of asymmetry.} Used with the Prohibitive negator \textit{baː=}\footnote{The reasons for the use of the Prohibitive negator with the simultaneity converb are unclear.}, the construction with this converb marks the absence of an event, with the scope of negation on the predicate (\ref{ex:beja:26}--\ref{ex:beja:27}). The clause gets a privative meaning.

\ea\label{ex:beja:26}
{Privative with scope over predicate}\\
    \gll baː=sinaːkir-eː fidin-ti ɖaːb-eːtiːt\\
         \textsc{neg.proh}=listen-\textsc{cvb.sim} move\_away-\textsc{cvb.grl} run-\textsc{cvb.seq}\\
    \glt `Without paying attention, she runs away and...' [\textsc{narr}\_36\_hunchback\_443] 
\z 

\ea\label{ex:beja:27} 
{Privative with scope over predicate}\\
    \gll ʔalaːl=iːb simha-ti da<nn>ʔi baː=gʷab-eː\\
         duststorm=\textsc{loc.sg} get\_rid\_of-\textsc{cvb.grl} do<\textsc{ipfv}>.[\textsc{3sg.m}] \textsc{neg.proh}=overturn-\textsc{cvb.sim}\\
    \glt `It makes a whirl of dust without overturning.' [\textsc{narr}\_51\_camel\_stallion\_116–117]
\z

\subsection{Negative lexicalizations}

Among the nine semantic domains of negative lexicalizations attested crosslinguistically (\citealt{Zeshan2004}; \citealt{Veselinova2013neg.exist}), Beja has only one, namely cognition, represented by the cognition verb \textit{gam} ‘ignore, not know’ \REF{ex:beja:28a} which does not also mean ‘neglect, not pay attention to’ which is expressed by other lexical items in idiomatic constructions. Unlike other verbs with negative content, such as ‘forget’ or ‘die’, \textit{gam} cannot be negated. It is itself a negator of the positive verb \textit{kan} ‘know’. \textit{gam} can be used for related cognitive domains, namely absence of understanding \REF{ex:beja:28b} and recognition \REF{ex:beja:28c}, and, in a polysemous manner, also for the expression of the incapacity modality \REF{ex:beja:28d}.

Note that in \REF{ex:beja:28} \textit{gam} could be replaced by the regular negation of \textit{kan} ‘know’.\footnote{See \REF{ex:beja:13} for the negation of \textit{kan}.} But Beja has no capacity modal verb (apart from the recent and infrequent loan from Arabic \textit{agdar}) and the cognitive verb \textit{kan} could not be used in a modal sense of capacity instead of \textit{gam} in \REF{ex:beja:28d}.

\newpage
\ea\label{ex:beja:28}
\textit{gam} ‘not know’\\
    \ea\label{ex:beja:28a}
         \gll oːn i=mheːl=oː eː-t-giːm\\
             \textsc{prox.sg.m.acc} \textsc{def.m}=medicine{\textbackslash}\textsc{acn=poss.3sg.acc} \textsc{3sg.m-mid}{}-not\_know{\textbackslash}\textsc{ipfv}\\
        \glt `He does not know the appropriate treatment.' [\textsc{narr}\_53\_pregnant\_man\_14]
    
    \ex\label{ex:beja:28b}    
         \gll w=ʔoːr-i   bhali=jeːb  naː=t eː-t-giːm\\
             \textsc{def.sg.m}=child-\textsc{gen} word=\textsc{loc.pl} thing=\textsc{indf.f} \textsc{3sg.m}{}-\textsc{mid}{}-not\_know{\textbackslash}\textsc{ipfv}\\
        \glt `He does not understand anything about the boy’s words.' [\textsc{narr}\_31\_king\_74–75]

    \ex\label{ex:beja:28c}
         \gll i=mhaj ragad-a eː=raw a-gam\\
             \textsc{def.m}=three leg-\textsc{pl} \textsc{def.pl.m.acc}=other{\textbackslash}\textsc{pl} \textsc{1sg}-not\_know{\textbackslash}\textsc{mid.pfv}\\
        \glt `I did not recognize the other three legs.' [\textsc{narr}\_27\_goat\_159]
        
    \ex\label{ex:beja:28d}
         \gll ɖa{\textasciitilde}ɖib-ti=jeːk wari=t hariw-a naː=t teː-t-giːm=eːk\\
             \textsc{plac}{\textasciitilde}fall-\textsc{acn=poss.2pl.acc} other{\textbackslash}\textsc{pl=indf.f} seek-\textsc{imp.sg.m} thing=\textsc{indf.f} \textsc{2sg.f-mid}-not\_know{\textbackslash}\textsc{ipfv}=if\\
        \glt `Look for another job, if you can’t do anything else!' [\textsc{narr}\_04\_djinn\_162]
    \z
\z 

Note that non-utterance negation is not lexicalized in Beja, but in addition to several verbs of saying which are regularly negated, there is also an affirmative construction consisting of the noun \textit{his / hus} ‘voice’ in a light verb construction with \textit{ak} ‘become, be’ meaning ‘not talk, shut up’ \REF{ex:beja:29}.

\ea\label{ex:beja:29}
{Non utterance}\\
    \gll u=jaːs-i=d hus ak-a i-jad-na\\
         \textsc{def.sg.m.acc}=dog-\textsc{gen=dir} voice become-\textsc{imp.sg.m} \textsc{3}-say{\textbackslash}\textsc{ipfv-pl}\\
    \glt `They tell the dog: Shut up!' [\textsc{narr}\_18\_Adam\_devil\_221]
\z

\section{Non-clausal negation}

\subsection{Negative replies}

Beja has three ways to form negative replies. 

A frequent one in everyday interactions is a dental click for ‘no’. It can be used in replies to negative and positive questions, and rarely in reaction to affirmative or negative statements. Its affirmative counterpart is a lateral click. As far as I could observe, they are only used among close relatives and friends, in informal settings.

In narratives, as well as more formal conversations, the most common construction to express negative replies to positive and negative polar questions is a verbal strategy, using the full paradigm of the negative Imperfective of \textit{ak} ‘be, become’, which functions in this case as a dummy verb.

The negation with \textit{ak} in reply to a positive polar question disagrees with its contents \REF{ex:beja:30}.


\ea\label{ex:beja:30}
{Negative reply with \textit{ak} to a positive polar question}\\
    \ea\label{ex:beja:30a}
        \gll ti=jʔa naː=t  hoːj t-ʔiʃ-a kʷiːkʷʔaj\\
             \textsc{def.f}=milk thing=\textsc{indf.f} \textsc{loc} 2-let{\textbackslash}\textsc{pfv-sg.m} crow\\
        \glt `Did you leave some milk from it, Crow?' [\textsc{narr}\_16\_Prophet\_Fox\_332–333]
    \ex\label{ex:beja:30b}    
        \gll ka=a-ki\\
             \textsc{neg.ipfv=1sg}-become{\textbackslash}\textsc{pfv}\\
        \glt `No.' [\textsc{narr}\_16\_Prophet\_Fox\_334]
    \z
\z 

In replies to negative polar questions, the verbal negation with \textit{ak} agrees with the negative polarity of the question and means ‘no’, as in \REF{ex:beja:31} below. It does not function as a polarity-reversing particle; \textit{awoː} ‘yes’ is used instead.

\ea\label{ex:beja:31}
{Negative reply with \textit{ak} to a negative polar question}\\
    \ea\label{ex:beja:31a}
         \gll Ahmed nʔir-aː-b ki=i-ki han\\
             Ahmed cure-\textsc{prf}-\textsc{m} \textsc{neg.ipfv=3sg.m}-become{\textbackslash}\textsc{pfv} \textsc{q.plr}\\
        \glt `Has not Ahmed healed?'
        
     \ex\label{ex:beja:31b}   
         \gll ki=i-ki\\
             \textsc{neg.ipfv=3sg.m}-become{\textbackslash}\textsc{pfv}\\
        \glt `No, he has not.'
    \z 
\z

The third possibility is to use the negative particle \textit{aʔaː} ‘no’. The only examples I have are answers to negative questions, as a sign of disagreement, not as a polarity-reversing particle. In \REF{ex:beja:32}, \textit{aʔaː} occurs in a strong polemical context, reinforced by the Arabic loan \textit{abadan} ‘never’.

\newpage
\ea\label{ex:beja:32}
{Negative question with negative reply}\\
    \ea\label{ex:beja:32a}
        \gll w=hagg=oː hoːj i-jaj-n=eːt toː=na ki=t-ti han […]\\
             \textsc{def.sg.m}=right=\textsc{poss.3sg.acc} \textsc{loc} 3-take{\textbackslash}\textsc{ipfv-pl=rel.f} \textsc{def.sg.f.acc}=thing \textsc{neg.ipfv=3sg.f}-become{\textbackslash}\textsc{pfv} \textsc{q.plr}\\
        \glt `Should they not stand for their rights?' (lit. `Is it not that they take from him their rights?') [\textsc{conv}\_01\_rich\_\textsc{sp1}\_214–215]
        
    \ex\label{ex:beja:32b}
        \gll abadan aʔaː\\
             never no\\
        \glt `Never! No!' [\textsc{conv}\_01\_rich\_\textsc{sp2}\_249–250]
    \z
\z 

\subsection{Negative indefinites}
\label{sec:beja:3.2}

Beja has no category of indefinite pronouns. For human entities one may use the numeral \textit{gaːl} ‘one’, and its plural form \textit{gaːli} ‘ones’ \REF{ex:beja:33a}, in an affirmative utterance. But \textit{gaːl} cannot be used in negative utterances, and the generic noun \textit{naː} ‘thing’ \REF{ex:beja:33c} or \textit{tak} ‘man’ \REF{ex:beja:33b} are used instead, together with a negative predicate.

\ea\label{ex:beja:33}
{Animate indefinite}\\
    \ea\label{ex:beja:33a}
    {Affirmative}\\
        \gll gaːl dhaːj jʔ-i\\
             one \textsc{dir} come-\textsc{aor.3sg.m}\\
        \glt `Someone was coming towards him.' [\textsc{narr}\_11\_coffee\_03]
    
    \ex\label{ex:beja:33b}
    {Negative}\\
        \gll tak ba=a-sakʷhiː\\
             man \textsc{opt}=\textsc{1sg}{}-follow{\textbackslash}\textsc{neg.opt}\\
        \glt `Let me not be followed by anyone!' [\textsc{narr}\_55\_tanduuy\_114]
    
    \ex\label{ex:beja:33c}
    {Negative}\\
        \gll w=haːʃ=iːb han naː=t ki=i-keː-n\\
             \textsc{def.sg.m}=land=\textsc{loc.sg} also thing=\textsc{indf.f} \textsc{neg.ipfv=3}-become{\textbackslash}\textsc{pfv-pl}\\
        \glt [The people in this country flew away from her.] `There is even no one in the country.' (lit. `They are not a thing.') [\textsc{narr}\_33\_teeth\_31–32]
    \z
\z 

Note that the use of \textit{gaːl} or \textit{tak} is not obligatory, and a finite verb form can simply be used instead, without an independent subject pronoun as in \REF{ex:beja:34} which is the very beginning of an anecdote.

\ea\label{ex:beja:34}
{No overt indefinite}\\
    \gll tam-iːni\\
         eat-\textsc{ipfv.3sg.m}\\
    \glt `Someone is eating.' [\textsc{narr}\_07\_cold\_01]
\z

For indefinite inanimates, and non-human animates as in \REF{ex:beja:35}, the generic noun \textit{naː} ‘thing’ is used for the expression of ‘something’ in affirmative utterances and ‘nothing’ with a negative predicate. In accordance with morphological rules of indefiniteness marking, the generic noun bears no indefinite article in subject position \REF{ex:beja:35} – except in relative clauses, see \REF{ex:beja:22} – but does in object position \REF{ex:beja:36}. 

\ea\label{ex:beja:35} 
{Inanimate subject}\\
    \ea\label{ex:beja:35a}
    {Affirmative}\\
        \gll ti-ʃik naː tiː-fi\\
             \textsc{3sg.f}-hunt{\textbackslash}\textsc{pfv} thing \textsc{3sg.f-}be\_there{\textbackslash}\textsc{aor}\\
        \glt `There was something that hunted it.' [\textsc{narr}\_05\_eritrea\_261]
    
    \ex\label{ex:beja:35b}
    {Negative}\\
        \gll naː dhaːj ka=ʔeːta\\
             thing \textsc{dir} \textsc{neg.ipfv=}come{\textbackslash}\textsc{pfv.3sg.f}\\
        \glt `Nothing came to it.' [\textsc{narr}\_16\_Prophet\_Fox\_071]
    \z    
\z 
        
\ea\label{ex:beja:36}
    {Inanimate object}\\
    
    \ea\label{ex:beja:36a}
    {Affirmative}\\
        \gll naː=t heː=hoːn\\
             thing\textsc{=indf.f} give{\textbackslash}\textsc{imp.[sg.m]=obj.1pl}\\
        \glt `Give us something!' [\textsc{narr}\_50\_fox\_hunt\_171]
    
    \ex\label{ex:beja:36b}
    {Negative}\\
        \gll naː=t hoːk bi=i-dʔiː-n=eːk\\
             thing=\textsc{indf.f} \textsc{2sg.dat} \textsc{opt=3}-do{\textbackslash}\textsc{neg.opt-pl}=if\\
        \glt `If they don’t do anything to you…' [\textsc{narr}\_08\_drunkard\_067]
    \z
\z 

Apart from the Arabic loan \textit{abadan} ‘never’ used as a strong refutation or disapproval to a statement (see \REF{ex:beja:32b} above), Beja uses standard negation or non-declarative negation to express an event that does not occur at any moment.

As for ‘nowhere’, one simply uses the noun \textit{mhiːn} ‘place’ in its indefinite form in a negative utterance \REF{ex:beja:37}.

\ea\label{ex:beja:37}
{Nowhere}\\
    \gll lhiːt mhiːn baː-baj-a\\
         tomorrow place \textsc{neg.proh}-go-\textsc{imp.2sg.m}\\
    \glt `Tomorrow, you go nowhere!'
\z

\subsection{Privative}

The absence of an entity is marked by three types of non-clausal constructions. Two of them are different from the ones used for the privative of events (cp. (\ref{ex:beja:26}--\ref{ex:beja:27}) \sectref{sec:beja:2.4}).

The first construction contains the dedicated postposition \textit{anu} ‘without’. It has scope over a noun or an overt pronoun as in \REF{ex:beja:38}.

\ea\label{ex:beja:38}
{Scope on (pro)noun with postposition \textit{anu}}\\
    \gll kibir anu aːdi tak i-ka\\
         pride without normal man \textsc{3sg.m-}become{\textbackslash}\textsc{pfv}\\
    \glt `He became a normal man without pride.' [\textsc{narr}\_31\_king\_94]
\z

As for the privative of events, it is also possible to use the negative simultaneity converb. For the privative of entities, only the simultaneity converb of \textit{ak} ‘become’ is licensed, placed after a noun or an overt pronoun \REF{ex:beja:39}. This construction is rare in my data.


\ea\label{ex:beja:39}
{Scope over noun with negative simultaneity converb of \textit{ak}} \\
    \gll i=nawa=wwa ti=miːtat=wa baː=kaːj\\
         \textsc{def.m}=sinew=\textsc{coord} \textsc{def.f}=bone{\textbackslash}\textsc{pl=coord} \textsc{neg.proh}=become{\textbackslash}\textsc{cvb.sim}\\
    \glt `Without the sinews and the bones.' [\textsc{narr}\_41\_Emirab\_leopard\_085] 
\z

If the scope is a covert 3\textsuperscript{rd} person pronoun referring to animate or inanimate entities, and if the predicate of the utterance is a motion verb, absence is marked with the general converb of \textit{ʔiʃ} ‘let’ \REF{ex:beja:40}.

\ea\label{ex:beja:40}
{Scope over covert pronoun with motion verb}\\

    \gll ʔiʃ-ti giːg-iːni\\
         let-\textsc{cvb.grl} leave-\textsc{ipfv.3sg.m}\\
    \glt `He leaves without (them).' [\textsc{narr}\_18\_Adam\_devil\_289]
\z

\section{Other aspects of negation}

\subsection{The scope of negation}

At this stage of research, generalizations about the scope of negation cannot be made.

\subsection{Negative polarity}

There are no negative polarity item per se in Beja (see \sectref{sec:beja:3.2}), apart from the recent loan from Arabic \textit{abadan} ‘never’. Prosody does not seem to play a role, but further research is needed.

\subsection{Marking of NPs in the scope of negation}

Case marking on NPs is not affected by the scope of negation.

\subsection{Reinforcing negation}

Ordinary clausal negations are often used rhetorically as understatements (see below \sectref{sec:beja:4.6.3}). To make clear that a statement is truly negative, and contrasts with an expected event or state, speakers use a complex predicate, only grammaticalized in the Perfective. It is built with a verbo-nominal form of the core verb with the suffix \textit{{-at / \nobreakdash-it}}\footnote{This verbo-nominal form only occurs in some complex predicates. See \citet[141 and 148]{Vanhove2017} for further details.} and the auxiliary verb \textit{rib} ‘refuse’ \REF{ex:beja:41} (for details, see \citet{HamidVanhove2002}).

\protectedex{\ea\label{ex:beja:41}
{\textit{rib} as reinforcing negation}\\
    \gll t=ʔaba=t=i dh=i tʔi-it ti-rib\\
         \textsc{def.f}=wadi=\textsc{indf.f=poss.1sg.nom} \textsc{dir=poss.1sg.acc} resemble-\textsc{vn} \textsc{3sg.f-}refuse{\textbackslash}\textsc{pfv}\\
    \glt `To me, my wadi did not look like it at all.' [\textsc{narr}\_05\_eritrea\_309]
\z}

\subsection{Negation, coordination and complex clauses}
Beja has no special negative coordinators, no special constructions for contrastive negation of NPs, and no negative conjunction. 

\subsection{Miscellaneous aspects of negation}

\subsubsection{Negative transport}

Negative transport does not occur in Beja. In complement clauses, whatever the semantic category of the predicate of the main clause, negation only has scope over its host. This means that if the main predicate is negated, the complement clause in the affirmative has an affirmative interpretation \REF{ex:beja:42}, and that negative complement clauses are always marked with a negator \REF{ex:beja:43}.

\ea\label{ex:beja:42}
{Scope over main predicate}\\
    \gll t=hoːj=t=eːb naːjloː=b hiːs-i=jeːk ka=a-kan\\
         \textsc{def.f}=belly=\textsc{indf.f=loc.pl} plastic=\textsc{indf.m.acc} think-\textsc{aor.1sg}=if \textsc{neg.ipfv=1sg}-know{\textbackslash}\textsc{mid.pfv}\\
    \glt `I don’t know if I had thought of a plastic bag in its belly.' [\textsc{narr}\_27\_goat\_176]
\z

\ea\label{ex:beja:43}
{Scope over complement clause}\\
    \gll ɖaːb-i bi=i-diː=jeːb hiːs-an\\
         run-\textsc{fut.sg} \textsc{opt=3sg.m}-say{\textbackslash}\textsc{neg.opt=rel.m} think-\textsc{pfv.1sg}\\
    \glt `I thought it would not be able to run.' [\textsc{narr}\_03\_camel\_153]
\z

 \subsubsection{Metalinguistic negation}

At this stage of data collection, I don’t know how metalinguistic negation works in Beja, or even if it is at all possible.

 \subsubsection{Non-negative uses of negatives and expletive negation}
\label{sec:beja:4.6.3}
The data did not provide a single example of expletive negation similar to French \textit{je crains qu’elle ne vienne} ‘I am afraid she will come’.

However Beja makes extensive non-negative uses of negative constructions. This is associated with highly positive values, in compliance with Beja social rules of politeness and honour, in order to soften statements and their social or psychological impact \REF{ex:beja:44}. 

\ea\label{ex:beja:44}
{Negation used as understatement}\\
    \ea\label{ex:beja:44a}
         \gll winneːt hoːj ɖib-a-b\\
             plenty \textsc{loc} fall-\textsc{prf-m}\\
        \gll ki=i-ki\\
             \textsc{neg.ipfv=3sg.m}{}-become{\textbackslash}\textsc{pfv}\\
        \glt `He is really angry at him.' (lit. `He has not fallen a lot in him.') [\textsc{narr}\_38\_ostrich\_143–144]
    \ex\label{ex:beja:44b}    
        \gll dijar-an=eːk ka=a-kan\\
             be\_tired-\textsc{pfv.1sg=}if \textsc{neg.ipfv=1sg}-know{\textbackslash}\textsc{mid.pfv}\\
        \glt `I am really exhausted.' (lit. `I don’t know if I am tired.') [\textsc{narr}\_18\_Adam\_devil\_165]
    \z 
\z

\subsubsection{Diachronic notes and observations}

Beja shows no trace of the Jespersen’s \& Meillet’s Cycle, nor of the Negative Existential Cycle as described by \citet{Croft1991} and \citet{Veselinova2014}.

\section{Summary}
\largerpage[2] %HC I don't know who wrote this line (same a bit below). if it's not Sebastian/Felix, we can delete it I guess %MM not me
To sum up, the negative system of Beja clearly distinguishes constructions in declarative and non-declarative utterances, and within declaratives and non-decl\-aratives, it has a reduced asymmetrical TAM system and uses different negators. TAM is even more reduced and asymmetrical in non-main clauses, where only the negative Optative or the negative Simultaneity converb occurs, whereas all affirmative declarative verb forms and converbs are possible in balanced and deranked clauses respectively.

What is especially noteworthy is the fact that, unlike other tenses, the negative Optative does not use an enclitic negator, but has recourse instead to stem alternation (V1) and specific inflectional morphemes (V2). 

Regarding stative predications, Beja shows different polarity systems depending on the type of predication. Possessive predication, expressed with a dedicated possessive verb, has the same negator and construction as standard negation in main and independent clauses, and as non-standard negation in non-main clauses. Negation is asymmetrical for equational, proper inclusion and attributive predications with the replacement of the copula by a negative ‘be/become’ verb, but there is no difference between permanent and temporary property assignment. For existential and locative predications, which cannot be used as privatives nor as short answers, the asymmetry concerns the restriction on the use of one of the three verbs.

In addition to the basic negative system, Beja has grammaticalized an auxiliated construction to reinforce negation, but only in the Perfective aspect, which is where TAM reduction occurs in ordinary negation. It is possible that social reasons are behind this restriction, since the Perfective aspect is less committing for speakers, and since it is considered as impolite to state loud and clear one’s opinion.
\newpage
Privatives are marked with four constructions, one clausal and three non-clausal. One of the non-clausal constructions, that with motion verbs, is historically based on clause chaining with a converb denoting abandonment (‘let’). It does not use a negator.

Phasal polarity expressions are rarely overtly expressed and are usually left to contextual interpretation. Nevertheless three adverbs denoting addition or repetition (‘also’, ‘anew’, ‘again’) can be added to negative utterances. One of these adverbials (‘only’) may be used for the expression of ‘not yet’; however, it is rarely overtly expressed.

Beja has no indefinite pronouns, but may use either the numeral \textit{gaːl} ‘one’, the generic noun \textit{naː} ‘thing’, the noun referring to a male \textit{tak} ‘man’. Only the latter two occur in negative utterances.

The language has only one negative lexicalization in the domain of cognition (polysemous with modality) competing with an affirmative verb, and very marginally a non-utterance one based on a light verb construction. 

As regards other aspects of negation, Beja does not show negative transport and does not seem to have recourse to expletive negation. On the other hand, it makes extensive non-negative uses of negative constructions, up to the point that a reinforcing strategy with an auxiliary verb has been grammaticalized.

Finally, negative replies use different strategies, from a simple dental click, to an adverb meaning ‘no’, or a dummy negative verb. Their use largely depends on pragmatics.

\tabref{tab:beja:5} summarizes the various functions of the negators and negative constructions.
% \clearpage

\begin{table} 
\small 
\begin{tabularx}{\textwidth}{p{3cm}p{2cm}Q}

\lsptoprule

{\bfseries Type} &   & {\bfseries\scshape form}\\
\midrule
\textbf{ Clausal negation}\\
\midrule
{Standard} &  & {\textit{ki= (/ka=)} + verb (\textsc{ipfv})}\newline {verb + \textit{ki= (/ka=)} + \textit{ak} ‘become’ (\textsc{pfv, prf, aor})}\newline  {verb + \textit{ki= (/ka=)} + \textit{di} ‘say’ (\textsc{fut})}\\
                     {Non-declarative} & { Prohibitive} & {\textit{baː= / biː=} + verb}\\
                      & {Optative} & {\textsc{png} of \textsc{pfv} + V1 stem of \textsc{proh}}\newline {\textsc{png} of \textsc{pfv} of V1 + V2 + suff. of \textsc{aor}}\\
                      & {Question} & {\textit{ki= (/ka=)} + verb (\textsc{ipfv})\newline verb + \textit{ki= (/ka=)} + \textit{ak} ‘become’ (\textsc{pfv,} \textsc{prf,} \textsc{aor})}\newline {verb + \textit{ki= (/ka=)} + \textit{di} ‘say’ (\textsc{fut})}\\
                     {Stative predication} & {Possession} & {\textit{ki=} + \textit{biri} ‘have’}\\
                      & {Equational, proper\newline inclusion, attributive}  & {verb + \textit{ki=} + \textit{ak} ‘become’ or verb + \textit{baː= / biː=} + \textit{ak} ‘become’}\\
                      & { Existential, locative} & {\textit{ki=} or \textit{baː= / biː=} + \textit{haj} or \textit{daː} ‘be there’}\\
                     {Non-main clause} &  & {Negative Optative in balanced clause; \textit{baː=} + \textsc{cvb.sim} in deranked clause}\\
                     {Negative\newline lexicalization} &  & {\textit{gam} ‘not know’\newline non-utterance: \textit{his} ‘voice’ + \textit{ak} ‘become’}\\
                     {Privative} &  & {\textit{baː=} + \textsc{cvb.sim}}\\
% \tablevspace
\midrule
\textbf{Non-clausal}\\
\midrule
{Reply} &  & {dental click, \textit{ki=} + \textit{ak} ‘become’, \textit{aʔaː}}\\
                     {Indefinite} &  & {\textit{tak} ‘man’ + negative verb, \textit{naː(t)} ‘thing’ + negative verb, \textit{mhiːn} ‘place’ + negative verb}\\
                     {Privative} &  & {NP + \textit{anu} ‘without’, NP + \textit{baː=} + \textsc{cvb.sim}, \textit{ʔiʃ-ti} \textsc{cvb.grl} of ‘let’ + motion verb}\\

% \tablevspace
\midrule
\textbf{Other}\\
\midrule
{Reinforcing} &  & {verb + \textit{-at/ -it}  + \textit{rib} ‘refuse’}\\
\lspbottomrule
\end{tabularx}
\caption{Negative strategies in Beja}
\label{tab:beja:5}
\end{table}

\tabref{tab:beja:6} provides an overview of the various roles played by the verb \textit{ak} ‘be, become’ (together with a negator) in negative constructions.

\begin{table}
\begin{tabularx}{\textwidth}{lX}
\lsptoprule
Standard neg. & \textsc{pfv},  \textsc{prf},  \textsc{aor} \\
Interrogative neg. &\textsc{pfv},  \textsc{prf},  \textsc{aor} \\
Stative neg. & equational,  proper inclusion, attributive existential locative  \\
Non-clausal & privative \\
Neg. lexicalization &non-utterance \\
Negative reply & ‘no’ \\
\lspbottomrule
\end{tabularx}

\caption{\textit{ak} ‘be, become’ in negative constructions in Beja}
\label{tab:beja:6}
\end{table}

\section*{Acknowledgements}

I gratefully acknowledge the financial support of the LLACAN, the ANR projects CorpAfroAs and CorTypo, and the excellence cluster Labex EFL (Empirical Foundations of Linguistics). My gratitude also goes to my colleagues and consultants in Sudan, in particular Mohamed-Tahir Hamid Ahmed, Ahmed Abdallah Mo\-hamed-Tahir and his family at Sinkat, as well as Yacine Ahmed Hamid and his family in Khartoum. Last but not least, many thanks also to my colleague Yvonne Treis for her very helpful comments and careful reading, to the editors of this volume and to the anonymous reviewers. The author is affiliated to the research centre Langage, langues et cultures d’Afrique (LLACAN, UMR 8135 du CNRS/Inalco/EPHE).

\newpage 
\section*{Abbreviations}
\begin{tabularx}{.45\textwidth}{lQ}
\textsc{1}      & first person \\ 
\textsc{2}      & second person \\ 
\textsc{3}      & third person \\ 
\textsc{abl}    & ablative \\ 
\textsc{acc}    &  accusative \\ 
\textsc{acn}   & action noun  \\ 
\textsc{aor}    &  Aorist \\ 
\textsc{caus}   &  causative \\ 
\textsc{coord}  &  coordination \\ 
\textsc{cop}    &  copula \\ 
\textsc{csl}    &  causal \\ 
\textsc{cvb}  &  converb  \\ 
\textsc{dat}    &  dative  \\ 
\textsc{def}    &  definite \\ 
\textsc{dir}    &  directional \\ 
\textsc{dist}   &  distal \\ 
\textsc{f}      &  feminine  \\ 
\textsc{fut}    &  Future \\ 
\textsc{gen}    &  genitive \\ 
\textsc{grl}  &  general   \\ 
\textsc{imp}    & imperative \\ 
\textsc{indf}   & indefinite \\ 
\textsc{ints}    & intensive  \\ 
\textsc{ipfv}   & Imperfective \\ 
\textsc{loc}    & locative \\ 
\textsc{m}      & masculine \\ 
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\textsc{mid}    & middle \\ 
\textsc{mnr}  &  manner  \\ 
\textsc{neg}    & negation \\ 
\textsc{nom}    & nominative \\ 
NP               & noun phrase \\ 
\textsc{obj}    & object \\ 
\textsc{opt}    & Optative  \\ 
\textsc{pfv}    & Perfective \\ 
\textsc{pl}     & plural \\ 
\textsc{plac}   & pluractional \\ 
PNG             & person, number, gender \\ 
\textsc{poss}   & possessive \\ 
\textsc{prf}    & Perfect  \\ 
\textsc{proh}   & Prohibitive \\ 
\textsc{prox}   & proximal \\ 
\textsc{q.plr}  & polar question \\ 
\textsc{recp}   & reciprocal \\ 
\textsc{rel}    & relative marker \\ 
\textsc{seq}   &  sequential  \\ 
\textsc{sg}     & singular \\ 
\textsc{sim}  &  simultaneity  \\ 
TAM             & tense, aspect, mood \\ 
V1              & verb class 1 \\ 
V2              & verb class 2 \\ 
\textsc{vn}     & verbo-nominal\\
\\
\end{tabularx}

\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
% \mm{@LSP: is there a way of reducing the spaces in the URL of Veselinova 2013a?}
% LSP@mm: No.
%HC: I asked about punctuation and capitalization of example sentences, but we won't change that since it is phonetic and not orthographic.

\end{document}
