\documentclass[output=paper,colorlinks=true, citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233954}

\author{Ljuba Veselinova\affiliation{Stockholm University}}
\title{Negation in Bulgarian}
\abstract{This article offers an overview of negation strategies in Bulgarian. The language follows several wide-spread cross-linguistic trends: its most general Standard Negation marker is a pre-verbal particle; there are several dedicated encodings of the prohibitive; negative indefinite pronouns are abundantly used with predicate negation. In terms of symmetric or asymmetric relations between affirmative and negative statements, we observe symmetric negation in  predications set in present and past tense as well as in hortatives. However, there is a constructional asymmetry between affirmative and negated predications in the future tense. In this connection, we should highlight the fact that negation encoded by a complex compound predicate is, in fact, commonly used in Bulgarian. Such encoding includes indicative predications set in the future as well as several variants of the prohibitive. The negative lexicalizations identified in the language fall into known cross-linguistic tendencies, namely lexicalizing of expressions for nonexistence, lack of ability and lack of knowledge. Bulgarian features special emphatic negators \textit{nito...nito} or \textit{ni...ni} which also function as coordinative negators; this trait has been shown to be typologically less common and also contingent on areal distribution. Another characteristic that is perhaps typologically less common is the presence of a special negative question particle \textit{nima}.

\keywords{symmetric and asymmetric negation, negation by particle, negation by complex compound predicate, coordinative negators, negative question particle}
}

% %move the following commands to the "local..." files of the master project when integrating this chapter
% \usepackage{tabularx}
% \usepackage{langsci-basic}
% \usepackage{langsci-optional}
% \usepackage{langsci-gb4e}
% \bibliography{localbibliography}
% \addbibresource{localbibliographyLV.bib}

% \usepackage{color, colortbl}
% \definecolor{Gray}{gray}{0.9}
% %\usepackage{caption}
% %\usepackage{multirow,array}
% %\usepackage[gen]{eurosym}
% %\usepackage[english]{babel}
% \usepackage{hyperref}
% \usepackage{tipx}


\IfFileExists{../localcommands.tex}{%hack to check whether this is being compiled as part of a collection or standalone
   \input{../localpackages}
   \input{../localcommands}
   \input{../localhyphenation}
%    \bibliography{localbibliography}
    \togglepaper[23]
}{}

% \captionsetup{labelfont=bf,
%               justification=raggedright,
%               singlelinecheck=false}
              
\begin{document}
\maketitle
\section{The language}
Bulgarian (ISO: \href{https://iso639-3.sil.org/code/bul}{bul}, Glottocode: \href{https://glottolog.org/resource/languoid/id/bulg1262}{bulg1262}) is a South Slavonic language currently spoken by about 9 million people, out of whom close to 7 million live in Bulgaria. Modern Bulgarian is a direct descendant of Old Church Slavonic and has a long written history, with the Glagolithic script being spread in Bulgaria in the 9th century and with the later creation and maintenance of the Cyrillic alphabet. \citet[188]{Scatton1993} summarizes the main periods of the history of the language as follows. (i) \textit{Prehistoric period}, which coincides with Proto-Slavic, from the time of the Slavonic invasion of the Eastern Balkans to the Cyrillo-Methodian mission in Moravia in the 9th century. (ii) \textit{Old Bulgarian}, 9th to 11th centuries, recorded in Old Church Slavonic manuscripts. (iii) \textit{Middle Bulgarian}, 11th through the beginning of the 15th century. This period includes the Second Bulgarian Empire (1185--1393/96) during which its territory expanded on the Balkans but also its collapse and  occupation by the Ottomans at the end of the 14th century. As one of the reviewers points out, literary activity was vibrant and continued some decades after the collapse of the Empire. Many structural innovations in the language can be traced back to that time. (iv) \textit{Modern Bulgarian}, in \citegen{Scatton1993} view from the 15th century until 1878. This covers the time of the Ottoman domination (1394-1878). 

It should be  noted that many of the innovations that occurred during the Middle and Early Modern Bulgarian periods are not systematically reflected in written records. As \citet[188]{Scatton1993} states, there are a number of reasons for this, one of the most important ones being conservative scribal attitude which prevented vernacular features from entering the written language. A prescriptive attitude dominated Bulgarian grammar writing well into the 20th century. Consequently, the formal and written varieties have been the focus of many descriptions where examples are commonly taken from literary works. In contrast, the \textbf{\href{https://dcl.bas.bg/bulnc/}{Bulgarian National Corpus}} (\citealt{BNC2023}), developed since the late 1990's, includes a rich variety of samples from both written and spoken registers. 

Bulgarian dialects are traditionally divided into Western and Eastern, based on the realization of Proto-Slavonic \textit{*ě} /\textit{jat}/. Thus in Western dialects \textit{*ě} becomes /e/ as in \textit{млеко/mleko} `milk', \textit{нема/nema} `not.have, not.exist' etc. while in Eastern dialects \textit{*e} is typically, though not always, realized as /ja/ as in \textit{мляко/mljako} `milk', \textit{няма/njama} `not.have, not.exist'. In addition to this feature, there are also several other phonological and grammatical characteristics that support the Eastern-Western divide, see Scatton (\citeyear[244--245]{Scatton1993}) for a concise outline. The Eastern dialects, specifically those from Veliko Tərnovo, the capital of Medieval Bulgaria and one of the main centers of the national renaissance, were the basis for the written as well as the formal and prestigious language for a long time. Once Sofia, which lies in the Western part of the country, became capital in 1878, some western features penetrated formal registers and the literary language. A detailed overview of Bulgarian dialects is considered outside the scope of this article; for a comprehensive analysis and data, see \citet{StoykovMladenov1993}, \citet{AntonovaEtAl2016}. 

Modern Bulgarian has 6 vowels and 22 consonant phonemes.\footnote{The number of consonant phonemes is reported here following \citet{Maddieson2013-wals-1}. Scholars who consider palatalization to be a distinctive feature in the language count 35 consonant phonemes, cf. \citet[191]{Scatton1993}} The language has free stress, rich inflectional and derivational morphology, relatively well defined morpho-syntactic categories of open classes such as nouns, verbs, adjectives and adverbs. The most neutral word order is SVO. Bulgarian has a number of features which set it apart from other Slavonic languages. The most important ones include the loss of the case system, the development of a definite article suffixed to the noun and the loss of the infinitive form. On the other hand, Bulgarian has retained and further complicated the Proto-Slavic tense-aspect system. As a result, currently the verb system covers a complex interplay of tense-aspect categories expressed mostly by prefixes on the verb stem but sometimes also by suffixes and infixes. As a consequence of the long contact with Turkish, Bulgarian has evidential as a grammatical category, which encodes the source of information, witnessed or inferred, or reported speech.

The language has been in intensive contact with other languages throughout its history. It is obviously part of the Balkan Sprachbund and shares a number of features with its neighbors. Some of them were already mentioned above: a vowel system consisting of six cardinal vowels, the development of a suffixed definite article and the loss of infinitives. Other features typically associated with the Balkan Sprachbund and observed in Bulgarian as well include the development of a periphrastic future formed by a particle developed from the verb `want', cf. (\ref{ex:bul:prs-3a}) in \sectref{sec:bul:sn}; distinctions between accomplishments and ongoing activities in the verb system; the use of one and the same construction to indicate subordination and subjunctive functions. A variant of this construction is also used for the encoding of specific subdomains of negation such as negation of predications set in the future as well as several variants of the prohibitive (see \sectref{sec:bul:sn} as well as \sectref{sec:bul:prohibitives}).
Borrowing of lexicon has been extensive throughout the history of Bulgarian. Greek borrowings date from the time of Old Bulgarian. They became especially frequent during the Ottoman rule. Borrowings from Turkish are numerous during the time of the Ottoman domination. As noted by \citet[240]{Scatton1993} a major change that occurred after 1878, the end of the Ottoman domination, included the replacement of Turkish loan words with Bulgarian neologisms or borrowings from Russian and other European languages such as Latin, Italian, French and German. In the 20th century and to this day, we observe a constant influx of English loans.

Most of the data in this chapter have been produced by the author who is a native speaker of the language. The \citet{BNC2023} has been amply consulted as well together with reference sources such as grammars, dictionaries and earlier analyses of negation in Bulgarian. The description of negation in Bulgarian in this chapter is based on the questionnaire provided by the editors (\citetv{chapters/questionnaire}) and the structure of the chapter follows the model given in the questionnaire.

% when using Word Count from the Menu on the upper left, subtract 210. This is what I calcutlated that the headers take up.
%Nice table to use later
%\begin{table}
%\caption{Frequencies of word classes}
%\label{tab:bul:1:frequencies}
 %\begin{tabular}{l rrrr}
  %\lsptoprule
   %         & nouns & verbs  & adjectives & adverbs\\
  %\midrule
  %absolute  &   12  &    34  &    23      & 13\\
  %relative  &   3.1 &   8.9  &    5.7     & 3.2\\
  %\lspbottomrule
 %\end{tabular}
%\end{table}


\is{aspect} %add "aspect" to subject index for this page
\il{Bulgarian} %add "Bulgarian" to language index for this page %HC why here and not right at the beginning of the chapter, on chapter title page?

\section{Clausal negation}
\subsection{Standard Negation} \label{sec:bul:sn}
Simple declarative sentences with an overt lexical verb as a predicate are negated by a pre-verbal particle \textit{ne} when the verb is in present or in past tense, cf. examples (\ref{ex:bul:prs-1}) and (\ref{ex:bul:prs-2}) below.
 
\ea Negation in predications with present time reference \label{ex:bul:prs-1}
  \ea С̇тефан тича.\label{ex:bul:prs-1a} \\ 
  	\gll Stefan tič-a. \\
     Stefan runs.\textsc{ipfv}-\textsc{3sg}.\textsc{prs}\\
    \glt `Stefan runs.' (constructed example)
  \ex С̇тефан не тича.\label{ex:bul:prs-1b}\\
  	\gll Stefan ne  tič-a.\\
  	Stefan \textsc{neg}    run.\textsc{ipfv}-\textsc{3sg}.\textsc{prs} \\
    \glt `Stefan does not run.' (constructed example)
  \z 
\z

\ea Negation in predications with past time reference\label{ex:bul:prs-2}
  \ea С̇тефан тичаше.\label{ex:bul:prs-2a} \\ 
  	\gll Stefan tič-aše. \\
     Stefan run.\textsc{ipfv}-\textsc{3sg}.\textsc{impf}\\
    \glt `Stefan was running.' (constructed example)
  \ex С̇тефан не тичаше.\label{ex:bul:prs-2b}\\
  	\gll Stefan ne  tič-aše.\\
  	Stefan \textsc{neg}    run.\textsc{ipfv}-\textsc{3sg}.\textsc{impf} \\
    \glt `Stefan was not running.' (constructed example)
  \z 
\z

In very broad lines, we can describe negation in predications with non-future time reference as symmetric, following \citet{Miestamo2005}. However, it has to be said that the preference for aspect in affirmative versus negative predications is yet to be studied. For instance, the verbs in examples (\ref{ex:bul:prs-1b}) and (\ref{ex:bul:prs-2b}) above are imperfective. Setting them in perfective aspect would be grammatical but it is my impression that imperfective is generally preferred in negated predications. More research is necessary to fully substantiate this statement. The preference for aspect will become even more important in the discussion on negation in non-declaratives, in \sectref{sec:2.2}.

The future tense is formed by a particle \textit{šte} and an inflected form of the lexical verb, cf. (\ref{ex:bul:prs-3a}). The construction can be traced back to the Old Church Slavonic construction which involved the verb \textit{xotěti} `want' plus the infinitive of a lexical verb. It was one among several alternative ways to indicate futurity, cf. \citet[418]{Duridanov1991}, \citet[288-296]{Slavova217}. Verbal predications in the future tense are negated by a complex compound predicate that includes the negative existential \textit{njama} together with the lexical verb introduced by the conjunction \textit{da}\footnote{I thank Björn Wiemer for the discussion on this construction. The views expressed here are entirely my own.}, see example in (\ref{ex:bul:prs-3b}) and \sectref{sec:bul:diachrony} for a discussion on diachrony.

\ea Negation in predications with future time reference\label{ex:bul:prs-3}
  \ea С̇тефан ще тича.\label{ex:bul:prs-3a} \\ 
  	\gll Stefan šte tiča. \\
     Stefan \textsc{fut} run.\textsc{ipfv}.\textsc{3sg}.\textsc{prs}\\
    \glt `Stefan will run.' (constructed example)
  \ex С̇тефан няма да тича.\label{ex:bul:prs-3b}\\
  	\gll Stefan njama da  tič-a.\\
  	Stefan \textsc{negex} \textsc{conj} run.\textsc{ipfv}-\textsc{3sg}.\textsc{prs} \\
    \glt `Stefan will not run.' (constructed example)
  \z 
\z

It should be noted that compound complex predicates that include an auxiliary and a lexical verb preceded by \textit{da} are seemingly identical to complex sentences that consist of a main clause and a subordinate one. In this sense, the complex compound predicates are strongly reminiscent of the Balkan subordinate construction. However, in a complex sentence, the subject of the main clause may be different from the one in the subordinate clause as in (\ref{ex:bul:complex-clause-1}). A complex compound predicate has one single subject. The particle \textit{da} is a multifunctional conjunction used in a variety of contexts. It is beyond the scope of this article to offer a detailed description of its uses. The ones most relevant here are collocating with present tense of verbs as in (\ref{ex:bul:prs-3b})\footnote{This is the modern version of an infinitive, see \citet[378]{Scatton1980}.}, indicating a complement or a subordinated clause as in (\ref{ex:bul:complex-clause-1}) or being used with subjunctive-like functions in hortative constructions, see (\ref{ex:bul:hort-1b}). Throughout this text, \textit{da} is referred to as either a conjunction or a subordinator depending on which function is most relevant.

\ea  
  Искам да дойдат тук.\\ \label{ex:bul:complex-clause-1}
  	\gll Iska-m da dojdat tuk.\\
     want-1\textsc{sg.prs} \textsc{sub.conj} come.\textsc{pfv}-\textsc{3pl.prs} here\\
\glt `I want them to come here.' (constructed example)

\z


Following \citegen{Miestamo2005} typology, we can say that in Bulgarian negation in predications with future time reference shows constructional asymmetry and is in the type (A\slash Fin). Affirmative and negative verbal clauses with future time reference differ in terms of constructional complexity of the predicate: affirmatives are encoded by a simple predicate and negatives by a complex one. 

To conclude, SN in Bulgarian can be described as symmetric for present or past tense, and asymmetric with regard to future tense predications. The symmetric construction falls in line with well known cross-linguistic tendencies to encode verbal negation by a particle that precedes the verb (cf. \citealt{Dahl1979,Dahl2010,Dryer2013-wals112,Dryer2013-wals143}). The asymmetric construction can be said to abide by general cross-linguistic tendencies as well since the negative auxiliary precedes the lexical verb which is to be negated.

\subsection{Negation in non-declaratives} \label{sec:2.2}
\subsubsection{Negative commands} \label{sec:bul:prohibitives}
Negative commands can be formed in several ways: 
(i) by adding the particle \textit{ne} to a positive imperative as in (\ref{ex:bul:imp-1b}), and (ii) by a compound complex predicate  which includes an auxiliary and a lexical verb or a proposition to be negated introduced by the conjuction \textit{da}. In fact, this construction is exactly the same as the one used to negate verbal clauses in the future. There are several different verbs or specific verb forms that can be used as prohibitive auxiliaries:  \textit{nedej} `do not\slash don't', \textit{nemoj} `not.be.able.to.\textsc{2sg.imp}', \textit{njama} `negative existential' and also \textit{stiga} `enough', see (\ref{ex:bul:imp-1c}), (\ref{ex:bul:imp-1d}), (\ref{ex:bul:imp-stiga}) for some illustrations. This prohibitive formed by a complex predicate is hereafter referred to as the \textit{complex predicate prohibitive}; whenever a specific sub-construction is discussed it is referred to by citing auxiliary in question, for instance the \textit{nedej}-prohibitive. Discussion on the diachronic origins of the prohibitive auxiliaries is offered in \sectref{sec:bul:diachrony}.

\newpage
\ea Imperative vs. prohibitives\label{ex:bul:imp-1}
  \ea Тичай!\label{ex:bul:imp-1a}\\ 
  	\gll Tič-aj!\\
        run.\textsc{ipfv}-\textsc{2sg.imp}\\
    \glt `Run!' (constructed example)
  \ex Не тичай!\label{ex:bul:imp-1b}\\
  	\gll Ne tič-aj!\\
  	 \textsc{neg} run.\textsc{ipfv}-\textsc{2sg.imp}\\
    \glt `Do not run!' (constructed example)
  \ex Недей да тичаш!\label{ex:bul:imp-1c}\\
  	\gll Nedej da tič-aš!\\
  	 do\_not.\textsc{2sg.imp} \textsc{conj} run.\textsc{ipfv}-\textsc{2sg.prs}\\ 
    \glt `Do not run!' (constructed example)
  \ex Няма да тичаш с празни ръце.\label{ex:bul:imp-1d}\\
    \gll Njama da tič-aš s prazni rətse\\
        \textsc{negex} \textsc{conj} run.\textsc{ipfv}-2\textsc{sg.prs} with empty.\textsc{pl} hand.\textsc{pl}\\
    \glt `Don't run with empty hands' (literal translation `You won't run with empty hands.') (Bulgarian National Corpus)
  \ex Стига си тичал!\\\label{ex:bul:imp-stiga}
    \gll Stiga si tič-al\\
        enough be.\textsc{2sg.prs} run.\textsc{ipfv}-\textsc{pst.ptcp.m.sg}\\
    \glt `Stop running/enough running/do not run!' (constructed example).  
  \z 
\z

Counts the Bulgarian National Corpus indicate that the regularly negated imperative as in (\ref{ex:bul:imp-1b}) is definitely more frequent than the \textit{complex predicate prohibitives} shown in (\ref{ex:bul:imp-1c}, \ref{ex:bul:imp-1d}, \ref{ex:bul:imp-stiga}). However, the existing variation in the encoding of the prohibitive should still be noted.

A comment on the form \textit{Стига\slash Stiga} shown in (\ref{ex:bul:imp-stiga}) is in order. In Bulgarian dictionaries, see \citet[925--926]{PopovEtAl1994}, the form \textit{stiga} is classified as a particle used in prohibitive commands. Unlike other admonitive particles such as \textit{haide} `come on' and \textit{neka} `let's', that have no obvious relation to any lexical items,\footnote{The word \textit{hajde} is a borrowing from Turkish.} the form \textit{stiga} clearly comes from the verb \textit{stigam} `arrive, reach, be enough'; see also \citet[11]{Auwera2006} citing \citet[166]{Greenberg1996}. 
It is a polysemous word with a number of functions. With some of them, such as being a prohibitive auxiliary or part of a complex subjunction \textit{стига да\slash stiga da} `as long as, if', \textit{stiga} can be described as a word with more general grammatical functions. Since it is also used as a motion verb meaning `arrive, reach', both literally and metaphorically, it clearly belongs to the lexicon as well. Morphologically, \textit{stiga} is the third person singular present form of \textit{stigam}. However, this description is relevant only when it is used with its lexical senses; when used with its grammatical functions, it is best described as an invariant form. For this reason, it is probably justified to classify it as a particle when used as a grammatical item. Seeing as it is clearly still used with a referential sense as well, its classification becomes a complex task. See further discussion and data on this issue in \sectref{sec:bul:diachrony}. 

The WALS classification \citep{wals-71} of the Bulgarian prohibitive takes into account only the \textit{nedej}-prohibitive; thus Bulgarian is classified in the type \textit{Special imperative + special negative} which is cross-linguistically the most frequent type. In an updated classification Bulgarian should be also classified in the type \textit{Normal imperative + normal negative}, as in (\ref{ex:bul:imp-1a}-\ref{ex:bul:imp-1b}) which comes third in cross-linguistic frequency among the four possible types of the prohibitive. Generally, it is safe to say that Bulgarian follows well-established cross-linguistic tendencies for the formation of its prohibitive.

The following can be said as regards aspect in prohibitives. \citet[46]{Scatton1980} notes that negative imperatives are not formed from perfective stems. This statement is probably best illustrated by one of the most common commands `Come!' (\ref{ex:bul:imp-7a}) which is a suppletive perfective form and one of the most frequently used direct commands. It cannot be negated but the corresponding suppletive imperfective has to be used in the prohibitive, cf. (\ref{ex:bul:imp-7b}). \citet[46]{Scatton1980} also notes that “positive imperatives from perfective stems are stylistically unmarked; positive imperatives from imperfectives carry a sense of urgency", see (\ref{ex:bul:imp-7c}). 

\ea \label{ex:bul:imp-7}
    \ea Ела!\label{ex:bul:imp-7a}\\ 
      	\gll Ela!\\
            come.\textsc{pfv.2sg.imp}\\
        \glt `Come!' (constructed example)
    \ex Идвай!\label{ex:bul:imp-7c}\\
        \gll Id-v-aj!\\
            come-\textsc{ipfv-2sg.imp}\\
        \glt `Come!' [with an implication of `Come right now\slash immediately!] (constructed example)
    \ex Не идвай!\label{ex:bul:imp-7b}\\
      	\gll Ne id-v-aj!\\
      	 \textsc{neg} come-\textsc{ipfv}-\textsc{2sg.imp}\\
        \glt `Don't come!' (constructed example)
    \z 
\z 

An example of a regular verb such as \textit{сядам\slash sjadam} `sit' is shown in (\ref{ex:bul:imp-sit}). The perfective form, cited in (\ref{ex:bul:imp-sit-a}), is a neutral invitation to sit. The imperfective form in (\ref{ex:bul:imp-sit-b}) is more marked. It may indicate irritation or urgency. Only the imperfective form can be negated, see (\ref{ex:bul:imp-sit-c1}); the use of the negator with the perfective \textit{sedni} in (\ref{ex:bul:imp-sit-c2}) is ungrammatical. Following Miestamo (\citeyear{Miestamo2005}), this should be classified as a paradigmatic asymmetry. 
\ea \label{ex:bul:imp-sit}
    \ea[]{Седни!\label{ex:bul:imp-sit-a}\\
      	\gll Sed-n-i!\\
            sit-\textsc{pfv-2sg.imp}\\
        \glt `Sit (down)!' (constructed example)}
    \ex[]{Сядай!\label{ex:bul:imp-sit-b}\\
        \gll Sjad-aj!\\
            sit.\textsc{ipfv-2sg.imp}\\
        \glt `Sit (down) (now)!' (constructed example)}
    \ex[]{Не сядай!\label{ex:bul:imp-sit-c1}\\ 
      	\gll Ne sjad-aj!\\
      	 \textsc{neg} sit.\textsc{ipfv}-\textsc{2sg.imp}\\
        \glt `Don't sit!' (constructed example)}
    \ex[*]{Не седни\label{ex:bul:imp-sit-c2}\\ 
      	\gll Ne sed-n-i\\
      	 \textsc{neg} sit-\textsc{pfv-2sg.imp}\\
        \glt Intended reading`Don't sit!' (constructed example)}
    \z 
\z 

It appears that it is much more common to use the imperfective aspect in the prohibitives and generally under negation. However, as already mentioned in \sectref{sec:bul:sn}, the preference for specific aspect categories under negation has to be subject to a more focused study.\footnote{If it is indeed valid that imperfective aspect is the preferred one under negation, then we could say that Bulgarian falls in line with \citegen{Schmidt1980} suggestion that such a preference is cross-linguistically common. Unlike \citet{Schmidt1980}, who does not rely on a stratified sample, \citet{MiestamoAuwera2011} use data from a balanced sample with world coverage. These authors argue that the available cross-linguistic data do not provide any support for a clear preference of either perfective or imperfective aspect under negation. As one of the reviewers points out, while both standard and non-standard Bulgarian seem to conform to generalization that imperfective is preferred under negation, data from a closely related language, Serbian, has constructions similar to (\ref{ex:bul:imp-6a}) where a perfective verb is used e.g. \textit{Nemoj da se naljutiš} `Don't be cross'.}

There is variation in the form of the \textit{complex predicate prohibitive} see \href{https://ibl.bas.bg/lib/bda4/#page/171/mode/1up}{Map 145} in \citet{Map145}.\footnote{The following disclaimer is offered as suggested by one of the reviewers: The value of the Bulgarian Dialect Atlas is in that it displays the entire Eastern South Slavic dialectal continuum. Note, however, that many of the speakers outside the Bulgarian state borders do not consider themselves Bulgarian.} The observed variation is along several parameters: (i) the choice of the prohibitive auxiliary, (ii) the form of the negated lexical verb, and (iii) the presence or absence of the conjunction \textit{da}. These parameters are discussed below.%HC check all these map links, depending on Ljuba's answer below

As already stated, the auxiliary in the \textit{complex predicate prohibitive} can be \textit{nedej} `don't do', \textit{nemoj} `not.be.able.to.\textsc{2sg.imp}', \textit{njama} `not.have', \textit{stiga} `enough'. The forms \textit{nedej} and \textit{nemoj} are restricted to the negative domain without any positive counterparts, see \sectref{sec:bul:diachrony} for further discussion of their origins as well as for reasoning on the grammaticalization of \textit{stiga} as a prohibitive auxiliary. 

The form of the negated/prohibited lexical verb can be: (i)  modern citation form or a short\slash truncated infinitive as in (\ref{ex:bul:imp-5a}); (ii) regular imperative, as in (\ref{ex:bul:imp-5b}); (iii) an older infinitive as in (\ref{ex:bul:imp-6a}); or (iv) an active past participle as in (\ref{ex:bul:imp-6b}). 

The forms identified in the Bulgarian Dialect Atlas, \href{https://ibl.bas.bg/lib/bda4/#page/171/mode/1up}{Map 145} \citep{Map145} are shown in (\ref{ex:bul:imp-5}) below. The punctuation in the examples follows the punctuation in the source. 

\ea\label{ex:bul:imp-5} 
  \ea Недей/Немой плака.\label{ex:bul:imp-5a}\\
  	\gll Nedej/Nemoj plaka.\\
        do.not.\textsc{2sg.imp} cry.\textsc{ipfv}.\textsc{short.infinitive}\\
    \glt `Do not cry.' (Central Eastern dialects, red-pink area on \href{https://ibl.bas.bg/lib/bda4/#page/171/mode/1up}{Map 145})
  \ex Недей/Немой плачи.\label{ex:bul:imp-5b}\\
	\gll Nedej/nemoj plači.\\
        do.not.\textsc{2sg.imp} cry.\textsc{ipfv}.\textsc{imp}\\
    \glt `Do not cry.' (Several disparate areas, yellow areas on \href{https://ibl.bas.bg/lib/bda4/#page/171/mode/1up}{Map 145})
  \ex Немой плакати.\label{ex:bul:imp-6a}\\ 
  	\gll Nemoj plaka-ti.\\
        do.not.\textsc{2sg.imp} cry.\textsc{ipfv}-\textsc{inf}\\
    \glt `Do not cry.' (Southern, red area on \href{https://ibl.bas.bg/lib/bda4/#page/171/mode/1up}{Map 145})
  \ex Стига плакал.\label{ex:bul:imp-6b}\\
	\gll Stiga plakal.\\
        enough cry.\textsc{ipfv}.\textsc{active.ptcp.m.sg}\\
    \glt `Do not cry.'\footnote{The glossing in the source is \textit{minalo svǝršeno dejatelno pričastie} which should have resulted in a gloss such as \textsc{pfv.active.ptcp}. However, one of the reviewers rightfully points out that the verb is imperfective, hence the change of gloss. It should be noted that it is formed on the aorist stem, which for this particular group of verbs ends in \slash k\slash. The imperfective stem ends in \slash č\slash so the imperfective past participle is \textit{плачел\slash plačel}. As far as I can tell, the latter form is used mostly in evidential constructions.} (Western areas, green areas on \href{https://ibl.bas.bg/lib/bda4/#page/171/mode/1up}{Map 145})
  \ex Недей/Немой плачеш.\label{ex:bul:imp-6c}\\
	\gll Nedej/nemoj plač-eš.\\
 do.not.\textsc{2sg.imp} cry.\textsc{ipfv}-\textsc{2sg.prs}\\
    \glt `Do not cry.' (Western dialects, light blue areas on \href{https://ibl.bas.bg/lib/bda4/#page/171/mode/1up}{Map 145})
  \z 
\z

As indicated on \href{https://ibl.bas.bg/lib/bda4/#page/171/mode/1up}{Map 145} in \citet{AntonovaEtAl2016}, the use of a specific auxiliary does not seem to be geographically conditioned, except for the \textit{Stiga}-construction which appears especially common in some non-standard variants west of Sofia. On the other hand, the use of a specific form of the lexical verb does show a geographically conditioned distribution. Specifically, the Eastern dialects opt for some form of the infinitive while in the west a finite form of the lexical verb is clearly preferred. The use of a regular imperative in the complex clause prohibitive construction is observed in delimited areas all over the Bulgarian speaking territory. It should be noted that the \textit{njama}-constructions (the negated declarative future) which can be used as a prohibitive is not observed in this function in non-standard variants of Bulgarian as cited by \citet{AntonovaEtAl2016}.

Hortatives are formed periphrastically  using the same template as the one we have seen for the complex predicate prohibitive. The particles \textit{hajde} or \textit{neka} are used as auxiliaries followed by a proposition introduced by the subordinating conjunction \textit{da} as in (\ref{ex:bul:hort-1a}). Another possibility is to omit these particles as in (\ref{ex:bul:hort-1b}). The construction in (\ref{ex:bul:hort-1b}) is referred to by \citet[379]{Scatton1980} as the \textit{compound imperative}.

\ea \label{ex:bul:hort-1}
  \ea Хайде/Нека да тичаме!\label{ex:bul:hort-1a}\\ 
  	\gll Hajde/Neka da tič-ame!\\
        Let's/Let's \textsc{conj} run.\textsc{ipfv}-\textsc{1pl.prs}\\ 
    \glt `Let's run!' (constructed example)
  \ex Да тичаме!\label{ex:bul:hort-1b}\\ 
	\gll Da tič-ame!\\
        \textsc{conj} run.\textsc{ipfv}-\textsc{1pl.prs}\\
    \glt `Let's run!' (constructed example)
  \z 
\z

Both hortative constructions in (\ref{ex:bul:hort-1}) can be negated by the particle \textit{ne} as shown in (\ref{ex:bul:hort-2}). 

\ea \label{ex:bul:hort-2}
  \ea Хайде/Нека да не тичаме!\label{ex:bul:hort-2a}\\ 
  	\gll Hajde/Neka da ne tič-ame!\\
        Let's/Let's \textsc{conj} \textsc{neg} run.\textsc{ipfv}-\textsc{1pl.prs}\\ 
    \glt `Let's not run!' (constructed example)
  \ex Да не тичаме!\label{ex:bul:hort-2b}\\ 
	\gll Da ne tič-ame!\\
        \textsc{conj} \textsc{neg} run.\textsc{ipfv}-\textsc{1pl.prs}\\
    \glt `Let's not run!' (constructed example)
  \z
\z

Based on the data in (\ref{ex:bul:hort-1}) and (\ref{ex:bul:hort-2}) negation in hortatives can be classified as symmetric. It should be noted, however, that usage based data, obtained from searches in the Bulgarian National Corpus and on social media highlight the fact that the particles \textit{neka} and \textit{hajde} differ with regard to their uses under negation and that affirmative and negated hortatives have very different discourse functions. These generalizations are further substantiated below.

First, the particles \textit{neka} and \textit{hajde} are not equal with regard to their frequency with negation. In the Bulgarian National Corpus, \textit{neka}  shows 2271 hits in contexts of negation while \textit{hajde} occurs with negation about five times less frequently, in 522 contexts.\footnote{Searches in social media yield even more striking proportions, specifically 20,600,000 for \textit{neka} with negation and  3,900,000 hits for \textit{hajde}. However, since the data on social media are not as controlled as the data in the Bulgarian National Corpus, it is probably safer to rely on the frequencies obtained from the latter.} 
    
Second, completely different verbs are used in positive versus negated hortatives. Specifically, positive hortatives typically encode expressions of  well wishes and good will as illustrated in (\ref{ex:bul:hort-3a}) and (\ref{ex:bul:hort-3b}). It should be noted that the constructions in (\ref{ex:bul:hort-3a}) and (\ref{ex:bul:hort-3b}), which are fairly common hortatives, are rarely, if ever, negated. Negated hortatives, on the other hand, are often admonitions and mitigated commands, which  include verbs such `forget', `limit', `hurry' as in (\ref{ex:bul:hort-3c}) and (\ref{ex:bul:hort-3d}). Thus the functions of the positive and the negated hortative are completely different even if their encoding appears symmetric. 
    
\ea Frequently occurring hortatives\label{ex:bul:hort-3}
  \ea Нека да е лято!\label{ex:bul:hort-3a}\\ 
  	\gll Neka da e ljato!\\
     let \textsc{conj} be.\textsc{3sg.prs} summer.\textsc{n.sg}\\
\glt `Let it be summer!' (popular song by the Tonika Family, a music group)

\ex Хайде да поговорим.\label{ex:bul:hort-3b}\\ 
	\gll Hajde da po-govor-im.\\
     let's \textsc{conj}  \textsc{ipfv}-talk.\textsc{ipfv}-\textsc{1pl.prs}\\
\glt `Let us talk.' (social media)

\ex Хайде да не бързаме.\label{ex:bul:hort-3c}\\ 
	\gll Hajde da ne bərz-ame.\\
     let's \textsc{conj}  \textsc{neg} hurry.\textsc{ipfv}-\textsc{1pl.prs}\\
\glt `Let's not hurry.' (social media)

\ex Нека (да) не забравяме... \label{ex:bul:hort-3d}\\ 
	\gll Neka (da) ne zabrav-jame...\\
     let's (\textsc{conj}) \textsc{neg}  forget-\textsc{1pl.prs}\\
\glt `Let's not forget...' (social media)

\z \z 

A comment on the use of the conjunction \textit{da} together with \textit{neka} is in order. The examples in (\ref{ex:bul:hort-3a}) and in (\ref{ex:bul:hort-3d}) may give the impression that it is obligatory in the affirmative and optional with negation. This is not the case: the conjunction \textit{da} is optional in both contexts but it would be odd to leave it out in (\ref{ex:bul:hort-3a}) which is a lyric from a song, highly entrenched for the generation who know the group. 

\subsubsection{Negative questions} \label{sec:bul:negative_questions}
Polar questions are formed by the question particle \textit{li} which follows the main verb in the clause and is typically clause final. Negative polar questions can be formed by either the standard negator \textit{ne} or by the negative existential \textit{njama}; their use follows the restriction to non-future vs. future tense as described in \sectref{sec:bul:sn}. The negators \textit{ne} and \textit{njama} can be used as short negative answers as well, see examples (\ref{ex:bul:qn-1a}), (\ref{ex:bul:qn-1b}), (\ref{ex:bul:qn-1c}). The repetition of the full verb sentence in all answers is optional and a matter of politeness: the more complete the answer, the higher the degree of politeness.

%\hc{I changed the structure of ex 12-13 (it was all in one before).}\lv{I had a point of having them all in one. I am afraid I do approve of this. Changing the structure of my examples is not ok.}\hc{link to ex 11-12 missing from the text} \lv{Yes, since there were changes in the structure of the examples.}
%\hc{05.12.2024: is it ok to keep the structure as it is now (this is how it is in (most?)/all other chapters for question-answers)? If yes, you can delete all these comments and add a reference to 11-12 in the paragraph preceding. If not, leave me a comment here that I should change it back to how it was (11 all on one line, 12 also all one one line), but we also need to add a reference to 11-12 in the preceding paragraph anyway.}

% \ea Polar questions\label{ex:bul:qn-1}%this is not needed anymore because i copied it below in another format
\ea\label{ex:bul:qn-1a}
    \begin{xlist}[B: ]
    \exi{A: } Пишеш ли?\\
        \gll Piš-eš li? \\ 
             write.\textsc{ipfv}-\textsc{2sg.prs} \textsc{q} \\ 
        \glt `Are you writing?'
    \exi{B: } Не, не пиша.\\
        \gll Ne, ne piš-a. \\ 
             No, \textsc{neg} write.\textsc{ipfv}-\textsc{1sg.prs} \\
        \glt `No, I am not writing.' (constructed example)
    \end{xlist}
\z 

\ea\label{ex:bul:qn-1b}
    \begin{xlist}[B: ]
    \exi{A: } Не пишеш ли? \\
        \gll Ne piš-eš li? \\ 
             \textsc{neg} write.\textsc{ipfv}-\textsc{2sg.prs} \textsc{q} \\ 
        \glt `Are you not writing?'
    \exi{B: } Не, (не пиша).\\
        \gll Ne, (ne piš-a). \\ 
             No, (\textsc{neg} write.\textsc{ipfv}-\textsc{1sg.prs}) \\
        \glt `No, I am not (writing).' (constructed example)
    \end{xlist}
\z 

\ea\label{ex:bul:qn-1c}
    \begin{xlist}[B1: ]
    \exi{A: } Няма ли да дойдеш? \\
        \gll Njama li da dojd-eš? \\ 
             \textsc{negex} \textsc{q} \textsc{conj} come.\textsc{pfv}-\textsc{2sg.prs} \\ 
        \glt `Aren't you going to come?'
    \exi{B1} Не, няма (да дойда).\\
        \gll Ne, njama (da dojd-a). \\ 
             No, \textsc{negex} (\textsc{conj} come.\textsc{pfv}-\textsc{1sg.prs}) \\
        \glt `No, I won't \slash No, I am not going to come.' (constructed example)
 \exi{B2} Няма.\\
        \gll Njama. \\ 
             \textsc{negex} \\
        \glt `No, I won't.' (constructed example)
    \end{xlist}
\z 
Negative polar questions can also be formed by using the particle \textit{nima} as in (\ref{ex:bul:qn-nima}). Such questions express either surprise, (\ref{ex:bul:qn-2a}), or are rhetorical in nature, cf. (\ref{ex:bul:qn-2b}). They belong to formal registers or high literary style as for instance, poetry.\footnote{The example in (\ref{ex:bul:qn-2b}) is from a famous Bulgarian poet, Pencho Slaveykov, who was active at the turn of the 19th and beginning of the 20th century. He remains a significant, debated and controversial figure in Bulgarian literature. The line cited by \citet{PopovEtAl1983Synt} is from the poem \href{https://chitanka.info/text/36978-izgorata-na-shemzedin-hafisa}{\textit{Izgorata na Šemzedin Hafisa}`The love of Šemzedin Hafisa'}.}

\ea \label{ex:bul:qn-nima} 
  \ea Нима не ги беше видял никой?\label{ex:bul:qn-2a}\\ 
  	\gll Nima ne gi beše vid-jal nikoj?\\
     \textsc{neg.q} \textsc{neg} \textsc{3pl.obj} be.\textsc{3sg.impf} see.\textsc{pfv}-\textsc{ptcp.pst.m.sg} nobody\\
\glt `Didn't anyone see them?' \citep[46]{PopovEtAl1983Synt}

\newpage
\ex Нима така поета люби?  \label{ex:bul:qn-2b}\\
	\gll Nima taka poet-a ljub-i?\\
     \textsc{neg.q} so poet.\textsc{m.sg}-\textsc{def.m.sg} love-\textsc{3sg.prs}\\
\glt `A poet does not love this way.' (lit. `Doesn't the poet love this way?'\footnote{A generic use of the definite article is a characteristic feature of several Balkan languages.}) \citep[46]{PopovEtAl1983Synt}
\z \z


\subsection{Negation in stative predications} \label{sect:stative}
Predications of identity, class membership and property assignment are negated by the standard negator \textit{ne}. Even though negation in these predications is structurally identical to negation in verbal predications, it is functionally different in that it is mostly used for contrastive purposes cf. (\ref{ex:bul:stat-1b}), see also \citet{Mettouchi2006} for the contrastive uses of negation in ascriptive predications. Constrative negation is further discussed in \sectref{sec:bul:scope} and \sectref{sec:bul:contrastive_neg}.

\ea  \label{ex:bul:stat-1}
  \ea Стефан ми е брат.\label{ex:bul:stat-1a}\\ 
  	\gll Stefan mi e brat.\\
        Stefan \textsc{1sg.dat/poss} be.\textsc{3sg.prs} brother\\
    \glt `Stefan is my brother' (lit. 'Stefan is a brother to me'.) (constructed example)
  \ex Стефан не ми е брат, ами братовчед.\label{ex:bul:stat-1b}\\ 
  	\gll Stefan ne mi e brat, ami bratovčed.\\
        Stefan \textsc{neg} \textsc{1sg.dat/poss} be.\textsc{3sg.prs} brother, but cousin\\
    \glt `Stefan is not my brother, he is my cousin' (lit. 'Stefan is not a brother to me, but a cousin'.) (constructed example)
  \ex Стефан е висок.\label{ex:bul:stat-1c}\\ 
    \gll Stefan e visok.\\
         Stefan be.\textsc{3sg.prs} tall.\textsc{m.sg}\\
    \glt `Stefan is tall.' (constructed example)
  \ex Стефан не е висок.\label{ex:bul:stat-1d}\\ 
	\gll Stefan ne e visok.\\
        Stefan \textsc{neg} be.\textsc{3sg.prs} tall.\textsc{m.sg}\\
    \glt `Stefan is not tall.' (constructed example)
  \z 
\z

Locative predications can be negated either by the particle \textit{ne} or by the negative existential \textit{njama} as shown in (\ref{ex:bul:loc-2}).

\ea  \label{ex:bul:loc-2}
  \ea Тодор (си) е в къщи.\label{ex:bul:loc-2a}\\ 
  	\gll Todor (si) e v kəšti.\\
     Todor (\textsc{refl}) be.\textsc{3sg.prs} in at.home\\
\glt `Todor is at home.' (constructed example)


  \ex Тодор не (си) е в къщи, (а на кино)\label{ex:bul:loc-2b}\\ 
  	\gll Todor ne (si) e v kəšti (a na kino).\\
     Todor \textsc{neg} (\textsc{refl}) be.\textsc{3sg.prs} in at.home (but at cinema)\\
\glt `Todor is not at home (but at the movies).' (constructed example)

  \ex Тодор го няма в къщи.\label{ex:bul:loc-2c}\\ 
  	\gll Todor go njama v kəšti.\\
     Todor \textsc{3sg.obj} \textsc{negex} in at.home\\
\glt `Todor is not at home.' (constructed example)

\z \z

Some noteworthy facts about negation of locative predications are as follows: (i) The reflexive marker in (\ref{ex:bul:loc-2a}) and (\ref{ex:bul:loc-2b}) is optional which is why it is within brackets; its addition makes the sentence more informal. (ii) Negation by the SN marker \textit{ne} can be used with contrastive focus as indicated in (\ref{ex:bul:loc-2b}). When the negative existential \textit{njama} is used, there is no such possibility. The information that Todor is at some place other than home has to be supplied in a new clause. This falls in line with a general cross-linguistic tendency for negative existentials to be mutually exclusive with focus (cf. \cite{Veselinova2013neg.exist}). (iii) When the negative existential \textit{njama} is used for the negation of locative predications, the subject has to be marked with an object clitic as shown in (\ref{ex:bul:loc-2c}). If the locative adverbial \textit{v kəšti} is omitted, the sentence would still be grammatical. However, depending on the situation, the utterance may mean either that Todor is absent from a specific location, which can be deduced from context, or that Todor is dead.

Predications of existence and possession are negated by the negative existential \textit{njama} which replaces pertinent forms of the habeo verb \textit{ima} `have'.

\ea  \label{ex:bul:exn-1}
  \ea Има питомни лъвове.\label{ex:bul:exn-1a}\\ 
  	\gll Ima pitomn-i ləv-ove.\\
     have.\textsc{3sg.prs} tame-\textsc{pl} lion\textsc{m}-\textsc{m.pl}\\
\glt `There are tame lions.' / `Tame lions exist' (constructed example)

 \ex Няма питомни лъвове.\label{ex:bul:exn-1b}\\ 
  	\gll Njama pitomn-i ləv-ove.\\
    \textsc{negex}.\textsc{3sg.prs} tame-\textsc{pl} lion.\textsc{m}-\textsc{m.pl}\\
\glt `There are no tame lions.' / `Tame lions do not exist' (constructed example)

\ex (Ние) имаме кола. \label{ex:bul:exn-1c}\\
\gll (Nie) ima-me kola.\\
(\textsc{1pl}) have-\textsc{1pl.prs} car.\textsc{f.sg}\\
\glt `We have a car.' (constructed example)

  \ex (Ние) нямаме кола. \label{ex:bul:exn-1d}\\
  	\gll (Nie) njama-me kola.\\
     (\textsc{1pl}) not.have-\textsc{1pl.prs} car.\textsc{f.sg}\\
\glt `We don't have a car.' (constructed example)

\z \z

As indicated by the data (\ref{ex:bul:exn-1a}) and (\ref{ex:bul:exn-1b}), the existential predicate, either affirmative or negative, is clause initial, which is a marked position for a predicate in Bulgarian. Furthermore, it has defective morphology in that it appears in a frozen third person singular form; it does not inflect for number but does inflect for tense. Conversely, when used in possessive predications as in (\ref{ex:bul:exn-1c}) and (\ref{ex:bul:exn-1d}), both the affirmative \textit{ima} and the negative \textit{njama} occur in clause-medial position and agree with their syntactic subjects. It is easy to see that the negative existential \textit{njama} is a univerbation between the negator \textit{ne} and the verb \textit{ima}. However, as discussed in \sectref{sec:bul:diachrony}, \textit{njama} is a single meaningful unit; it is not justified to postulate two morphemes in it.

\subsection{Negation in non-main clauses} 
Negation in non-main clauses is encoded in the same way as negation in main clauses, by \textit{ne} and \textit{njama} as shown in (\ref{ex:bul:nonmain-1}).

\ea  
  Стефан каза че няма да кара защото не е спал.\\ \label{ex:bul:nonmain-1}
  	\gll Stefan kaza če njama da kara zaštoto ne e spal.\\
     Stefan said that \textsc{negex} \textsc{conj} drive.\textsc{3sg.prs} because \textsc{neg} be.\textsc{3sg.prs} sleep.\textsc{m.sg.pst.ptcp}\\
\glt `Stefan said he wouldn't drive because he hadn't slept.' (constructed example)

\z
% djv 119 in Bulgarian syntax treats обособени обстоятелства. It says that that the most "typical" set-aside circumstance is by the active adverbial participle. Is it correct to say that this notion corresponds to dependent non-finite terminology in Enlgish? There are a bunch of other set-aside constituents. Are they dependent non-finite clauses as well? I guess I should read carefully through the section on обосебени части and decide on an appropriate approach.

The closest equivalent to dependent non-finite clauses in Bulgarian are constructions which are headed by a form labeled \textit{active adverbial participle}\slash Bulgarian \textit{деепричастие}. It is formed from the present/imperfective stem of the verb and the suffix \textit{-йки\slash -jki}. The final vowel of the stem can be either /a/ or /e/; its quality depends on the inflectional class of the verb, see  \citet[382]{Pǝrvev1983} and data in \tabref{extab:bul:deep-1-otherpossibility}.

%\eabox{\label{ex:bul:deep-1} Examples of active adverbial participles\\%I also copied this example below with different syntax, see comment right above
 %   \begin{tabular}{ll}
  %     каз-в-ам\slash kaz-v-am `say-\textsc{ipfv-1sg.prs}'  & казва-йки\slash kazva-jki `saying'\\ 
   %     пея\slash peja `sing.\textsc{ipfv.1sg.prs}' & пее-йки\slash pee-jki `singing'\\
    %    вярв-ам\slash vjarv-am `believe.\textsc{ipfv.1sg.prs}' & вярва-йки\slash vjarva-jki `believing'\\
    %\end{tabular}
%} 

\begin{table}
% \eabox{
\caption{Examples of active adverbial participles}
\label{extab:bul:deep-1-otherpossibility}
% \footnotesize  %HC same example as right before, with different syntax. this is the best readable one, but too small text
\fittable{
    \begin{tabular}{llllll}
    \lsptoprule
       каз-в-ам &  kaz-v-am & `say-\textsc{ipfv-1sg.prs}'  & казв-айки & kazv-ajki & `saying'\\ 
        пея & peja & `sing.\textsc{ipfv.1sg.prs}' & пее-йки & pee-jki & `singing'\\
        вярв-ам & vjar-v-am & `believe-\textsc{ipfv-1sg.prs}' & вярв-айки & vjarv-ajki & `believing'\\
    \lspbottomrule
    \end{tabular}
    }
%     }
\end{table}

This active adverbial participle is used to indicate action that is often simultaneous with the action indicated by the main verb; the clause headed by it is described as an adverbial constituent (cf. \citealt[238--239]{PopovEtAl1983Synt}). Dependent non-finite clauses are negated by the SN marker \textit{ne} as shown in (\ref{ex:bul:nonmain-2}).

\ea  Не вярвайки на собствените си очи, аз се приближих до будката.\\ \label{ex:bul:nonmain-2}
  	\gll Ne vjar-v-ajki na sobstven-i-te si oči, az se približ-ix do budka-ta.\\
        \textsc{neg} believe-\textsc{ipfv}-\textsc{adv.ptcp} \textsc{prep} own-\textsc{pl}-\textsc{def.pl} \textsc{refl} eye.\textsc{pl}, I \textsc{refl} approach.\textsc{pfv}-\textsc{pst} \textsc{prep} kiosk.\textbf{f.sg}-\textsc{def.f.sg}\\
    \glt `Not believing my own eyes, I approached the kiosk.' (Bulgarian National Corpus)
\z
It has to be said that non-finite clauses headed by a gerund-like form are negated very seldom. Out of the 205061 hits of such clauses in the Bulgarian National corpus, there are only 1279 where the active adverbial participle (деепричастие) is negated.\footnote{The seed for the corpus search uses the morphological tagging of the Bulgarian National Corpus. Thus I used regular expressions including the tag for  active adverbial participle, *\textsc{pos=v ivf}=z to locate dependent clauses headed by it. That yields 205601 hits. To locate such clauses in the context of negation I used `не *\textsc{pos=v ivf}=z'. This yields 1488 hits; however in 209, we do not observe negation of the active adverbial participle; rather the dependent clause occurs after a standalone negative answer or precedes a negated constituent such as a negated object or a place adverbial.}

\subsection{Negative lexicalizations}\label{sec:bul:l}
Negative lexicalizations include negative existential \textit{njama} and the special prohibitive forms \textit{nedej} and \textit{nemoj}, cf. \sectref{sec:2.2} as well as \sectref{sect:stative} for examples. In addition, there is a special form \textit{невем/nevem} `I don't know' observed in the non-standard dialect of Trojan (Balkan Mountain, Central Bulgaria), cf. \href{https://ibl.bas.bg/lib/ber_4_000-1004/#page/585/mode/1up}{\citet[v.4:585]{GeorgievIlčev1971}}. 

All of these lexicalizations fall in line with the cross-linguistic generalizations outlined for this phenomenon, cf. \citet{Veselinova2013lexicalization}. The origin of these lexicalizations is discussed in \sectref{sec:bul:diachrony}.

%\subsection{Other clausal negation constructions} \label{sec:bul:other-clausal_negators}
%The negators \textit{nito...nito} and \textit{ni...ni} can be used on their own for the negation of verbal predications which is is why they can be described as an alternative clausal negation strategy. However, they are always observed in coordinated structures, hence most examples and discussion of these markers are placed in \sectref{sec:bul:coord}.

\section{Non-clausal negation}\label{sec:bul:non-clausal-negation}
\subsection{Negative replies}\label{sec:bul:negative-replies}
Both the standard negator \textit{ne} as well as the negative existential \textit{njama} can be used as short negative replies. Examples (\ref{ex:bul:qn-1a}--\ref{ex:bul:qn-1b}) above show \textit{ne} in this function while (\ref{ex:bul:qn-2}) illustrates the same for \textit{njama}, see also discussion and data in \ref{sec:bul:negative_questions}.

\ea \label{ex:bul:qn-2}
    \begin{xlist}[B: ]
    \exi{A: }Ще дойдеш ли? \\
        \gll Šte dojd-eš li? \\ 
             \textsc{fut} come.\textsc{pfv}-\textsc{2sg.prs} \textsc{q}\\ 
        \glt `Are you going to come?'
    \exi{B: }Не\slash Няма!  \\ 
        \gll Ne\slash Njama! \\ 
             \textsc{neg}\slash \textsc{negex} \\
        \glt `No!' (constructed example)
    \end{xlist}
\z 

To change the polarity of a negative question, as in (\ref{ex:bul:qn-1b}), the adverb/sentence tag \textit{naprotiv} `on the contrary' must be used. The affirmative particle \textit{da} `yes' cannot be used in answers to negative polarity questions.

\subsection{Negative indefinites and quantifiers}
Negative indefinite pronouns are formed by prefixing the particle or prefix \textit{ni} to content question words such as \textit{koj} `who', \textit{kakvo} `what', \textit{kakəv} `what.\textsc{adj}', \textit{kəde} `where', \textit{кога/koga} `when' etc. Some examples are shown in (\ref{ex:bul:indef-1}).

\ea\label{ex:bul:indef-1}Negative indefinite pronouns\\
    \begin{tabular}{ll}
        никой/nikoj & `no one'\\
        никакъв/nikakəv & `none'\\
        никъде/nikəde & `nowhere'\\
        никога/nikoga & `never'\\
    \end{tabular}
\z


The forms listed in (\ref{ex:bul:indef-1}) inflect according to number and gender of their referent when applicable. In addition  to these forms, there is also the negative quantifier \textit{нищо/ništo} `nothing' which is a lexicalized form resulting from the fusion of \textit{ni} and \textit{što} `what'. The form \textit{što} is used as a question word in fixed phrases and non-standard variants of Bulgarian only; \textit{kakvo} `what' is the standard content question word.

Indefinite pronouns co-occur with predicate negation as demonstrated in (\ref{ex:bul:indef-2}). On this parameter Bulgarian follows a wide-spread cross-linguistic tendency (cf. \citealt{Haspelmath2013-wals115}).

  \ea Никой никъде няма да ходи. \\ \label{ex:bul:indef-2}
  	\gll Nikoj nikəde njama da xodi.\\
     no.one no.where \textsc{negex} \textsc{conj} go.\textsc{ipfv.3sg.prs}\\
\glt `No one will go anywhere.' (constructed example)

\z 
\subsection{Negative derivation and case marking}
Case marking is not relevant for Bulgarian, therefore, only derivation is discussed in this section. Negative derivation is done by using the particle \textit{ne} and by the preposition \textit{bez} `without'. The particle \textit{ne} is typically prefixed to a positive stem, see data in (\ref{ex:bul:der-1}). The suffixes -\textit{ив}\slash -\textit{iv} and \textit{-ен}\slash \textit{-en} are used for the derivation of adjectives regardless of polarity, that is they derive both affirmative and negative adjectives.

\ea\label{ex:bul:der-1} Negative derivations by using the particle \textit{ne} (constructed examples)

    \begin{tabular}{lll} 
        щастие & štastie  & `happiness, bliss' \\
        нещастие & ne-štastie & `misery, a stroke of bad luck' \medskip \\ 
        справедлив & spravedliv  & `just' \\
        несправедлив & ne-spravedliv & `unjust, partial' \medskip \\ 
        доволен & dovolen  & `content, happy, satisfied' \\
        недоволен & ne-dovolen & `unhappy, dissatisfied' \\
    \end{tabular}
\z

The derivational operation shown in (\ref{ex:bul:der-1}) is very productive. As is common with derivation, there are also cases of mismatches. For instance, the derivation of \textit{ne-štasten} `unhappy' presents an example of a stem mismatch since it is derived from a stem different from the one used for its positive correspondent \textit{štastliv} `happy'. A semantic mismatch can be exemplified by the derivation of \textit{neumoren} `energetic, somebody who never gets tired' from \textit{umoren} `tired'. The positive adjective refers to a resulting state, the derived negative adjective refers to a constant property of a person's disposition; thus even if there is a seeming symmetry in form, the senses of the two lexemes do not form an opposition.

Privative or caritive marking is encoded by prepositional phrases where the preposition \textit{bez} `without' is used together with a nominal; the resulting prepositional phrase indicates the absence of that nominal. These prepositional phrases are further univerbated and used as bases for the derivation of adjectives with a similar meaning as shown in data in (\ref{ex:bul:der-2}).

\protectedex{%HC style here, but I don't remember in which chapter there was something similar. But it looks good here, so let's keep it as it is.
\ea Privative expressions with \textit{bez} \label{ex:bul:der-2}
  \ea без дом $\rightarrow$ бездомен\label{ex:bul:der-2a}\\ 
  	\gll bez dom $\rightarrow$ bez-dom-en\\
        without home $\rightarrow$ without-home-\textsc{adj}\\
    \glt `without (a) home' $\rightarrow$ `homeless' \citep[156]{StojanovEtAl1983}
  \ex без смисъл $\rightarrow$ безсмислен\label{ex:bul:der-2b}\\ 
    \gll bez smisəl $\rightarrow$ bez-smisl-en\\
         without sense $\rightarrow$ without-sense-\textsc{adj}\\
    \glt `without (a) sense' $\rightarrow$ `senseless' \citep[following][156]{StojanovEtAl1983}
  \ex без съвест $\rightarrow$ безсъвестен\label{ex:bul:der-2c}\\ 
    \gll bez səvest $\rightarrow$ bez-səvest-en\\
         without conscience $\rightarrow$ without-conscience-\textsc{adj}\\
    \glt `without (a) conscience/scruples' $\rightarrow$ `unscrupulous' \citep[following][156]{StojanovEtAl1983}
  \ex без утеха $\rightarrow$ безутешен\label{ex:bul:der-2d}\\ 
    \gll bez uteha $\rightarrow$ bez-uteš-en\\
         without consolation $\rightarrow$ without-consolation-\textsc{adj}\\
    \glt `without (a) consolation' $\rightarrow$ `inconsolable' \citep[following][156]{StojanovEtAl1983}
  \z
\z }

The preposition \textit{bez} `without' can also be prefixed directly to some adjectives as shown in (\ref{ex:bul:der-3}). 

\TabPositions{.33\textwidth,.66\textwidth}

\ea полезен\footnote{The adjective \textit{polezen} `useful' is a derivation from \textit{polza} `use' by adding an epenthetic vowel to the stem and the adjectivizer \textit{-en}.} $\rightarrow$ безполезен\label{ex:bul:der-3}\\ 
  	\gll polezen   $\rightarrow$ bez-polezen\\
      useful  $\rightarrow$ without-useful\\
\glt `useful'  $\rightarrow$ `useless' (constructed example)
\z%HC format? but i guess it could stay like this too, unless i find the way i did it in another chapter.

\section{Other aspects of negation}

\subsection{The scope of negation} \label{sec:bul:scope}
The scope of negative markers may differ. When used as SN, the particle \textit{ne} typically scopes over the entire proposition as in (\ref{ex:bul:scope-1}). The same applies for the negative lexicalizations listed in \sectref{sec:bul:l}.

\ea Не знам какво да кажа.\label{ex:bul:scope-1}\\ 
  	\gll Ne znam kakvo da kaža.\\
      \textsc{neg} know.\textsc{1sg.prs} what \textsc{conj} say.\textsc{1sg.prs}\\
\glt `I do not know what to say.' (constructed example)
\z

The particle \textit{ne} can also be used with a narrow scope, that is, as a constituent negator. For instance, in (\ref{ex:bul:scope-2}), it scopes only over the adverbial phrase \textit{na šega}, `as a joke' rendering the sense `not as a joke'. Other negative particles with narrow scope include \textit{ni} and \textit{nito}, cf. (\ref{ex:bul:reinf-1b}) in \sectref{sec:bul:reinf} below.

\ea Райко се уплаши не на шега и взе да се взира отново в шумата.\label{ex:bul:scope-2}\\ 
  	\gll Rajko se u-plaši ne na šega i vze da se vzira otnovo v šuma-ta.\\
     Rajko \textsc{refl} \textsc{pfv}-be.scared.\textsc{3sg.pst} \textsc{neg} on joke and take.\textsc{pfv.3sg.pst} \textsc{conj} \textsc{refl} stare.\textsc{pfv.3sg.aor} again in forest.leaves.\textsc{f.sg}-\textsc{def.f.sg}\\
\glt `Rajko was scared, not as a joke, and started staring into the forest leaves again.' \citet[492]{StojanovEtAl1983} 
\z
In many cases of contrastive negation, the negated construction is emphasized by a special intonation as it would be in (\ref{ex:bul:scope-3}) when spoken. However, I cannot think of cases where the contrast is achieved by prosodic means only. \citet{Lashkova1980} offers a detailed discussion on the use of negation and parsing of the predication.

\ea Казах ти да купиш пиле, не петел!\label{ex:bul:scope-3}\\ 
  	\gll Kaza-x ti da kup-iš pile, ne petel!\\
      say-\textsc{aor.1sg} \textsc{2sg.dat} \textsc{sub.conj} buy-2\textsc{sg} chicken, \textsc{neg} rooster!\\
\glt `I told you to buy a chicken, not a rooster!' (constructed example)
\z
\subsection{Negative polarity} \label{sec:bul:negative.polarity}

In the strict sense of a negative polarity item, that is, words which only occur with negation and other irrealis contexts, Bulgarian has only two, \textit{gək} and \textit{hič} `none, not at all'.\footnote{I am very grateful to the reviewer who brought up \textit{hič}, which I had missed, and likewise prompted me to examine the adverbs \textit{izobšto\slash vəobšte}.} In addition, there are two adverbs \textit{izobšto\slash vəobšte} `generally, at all, (ever)' that are used very frequently in irrealis contexts which is why it is fitting to mention them when negative polarity items are discussed.

The form \textit{gək} occurs exclusively in negative contexts and typically together with the emphatic negators \textit{nito} or \textit{ni}.

\ea Стефан не каза нито гък цяла вечер.\label{ex:bul:polarity-1}\\ 
  	\gll Stefan ne kaza nito gək tsjala večer.\\
     Stefan \textsc{neg} say.\textsc{pfv.3sg.aor} \textsc{neg.emph} \textit{gək} entire evening\\ 
\glt `Stefan did not say a single word the entire evening.' (constructed example) 
\z
In the Bulgarian Etymological Dictionary (\href{http://ibl.bas.bg/lib/ber/#page/394/mode/2up}{\cite[v. 1: 300]{GeorgievIlčev1971}}), \textit{gək} is described as an onomatopoetic form. I am not aware of any referential meaning of this word. It occurs mostly with verbs of utterance such as \textit{кажа/kaža} `say' or the more archaic \textit{река/reka} `say'. Occasionally, it is also observed with the verb of hearing \textit{чувам/čuvam} `hear' as shown in (\ref{ex:bul:polarity-2}) where \textit{gək} is used in a ban to utter any kind of sound/say anything.

\ea Да не чувам нито гък!\label{ex:bul:polarity-2}\\ 
  	\gll Da ne ču-v-am nito gək!\\
     \textsc{conj} \textsc{neg} hear-\textsc{ipfv}-\textsc{1sg.prs} \textsc{neg.emph} gək!\\
\glt `May I not hear a single word/sound!' = `Not a single sound!'  (Bulgarian National Corpus)
\z

The form \textit{gək} is described in Bulgarian dictionaries as an interjection. It has to be said that it invariably occurs in an object position, so based on syntactic criteria, it can also be said to have nominal properties. This contrasts with the form \textit{hič} `none, not at all' which is always used in a modifying function and is thus classified as an adverb, see also Dictionary of the Bulgarian Language (\href{http://ibl.bas.bg/rbe/lang/bg/%D1%85%D0%B8%D1%87/}{\citealt{IBE1993}}).
It is a loan from Turkish. Some examples follow in (\ref{ex:bul:polarity-hich00}).

\ea \label{ex:bul:polarity-hich00}
    \ea Хич не ме интересува.\label{ex:bul:polarity-hich01}\\
        \gll Hič ne me interesuva.\\
            not.at.all \textsc{neg} \textsc{1sg.acc} interest.\textsc{ipfv.3sg.prs}\\
        \glt `[It\slash this] does not interest me at all.'\\(Dictionary of the Bulgarian Language \href{http://ibl.bas.bg/rbe/lang/bg/%D1%85%D0%B8%D1%87/}{\cite{IBE1993}}) 
    \ex Хич не му мига окото.\label{ex:bul:polarity-hich02}\\
        \gll Hič ne my miga oko-to.\\
            not.at.all \textsc{neg} \textsc{3sg.dat} blink.\textsc{3sg.prs} eye.\textsc{n.sg}-\textsc{def.n.sg}\\
        \glt lit. `To him the eye never blinks.' = `He never blinks an eye.' = \\`He does not care.' \\(Dictionary of the Bulgarian Language \href{http://ibl.bas.bg/rbe/lang/bg/%D1%85%D0%B8%D1%87/}{\cite{IBE1993}})
    \z 
\z 

The adverbs \textit{izobšto\slash vəobšte} `generally, at all, entirely' are very close in meaning, with \textit{izobšto} showing a broader polysemy than \textit{vəobšte}, see the entries for \textit{izobšto} 
%\citep{bul:dict:izobsto} 
and \textit{vəobšte} %\citep{bul:dict:veobste} 
in the \href{https://ibl.bas.bg/rbe/}{Dictionary of the Bulgarian Language online}. Examples are shown in (\ref{ex:bul:izobshto_00}) and (\ref{ex:bul:vaobshte_00}).
    % \lv{Let's talk about reference and these links during the meeting}
    % \hc{These outcommented links are to put in the bibliography, i haven't done it yet}
    % \hc{20241205: done, I also added a reference to them in the paragraph. But I do not really know what to add in the bibliography, for now there is only these links and "Dictionary of the Bulgarian Language" as booktitle. You can add the urldate around line 10112-10130 in the localbibliography file, and ideally also author,publisher,year. Also, if you don't bring it to work as it should, please let me know how it should look like here in the text as well as in the list of references.}%I got inspired by this page to cite a dictionary entry, but I don't have all necessary information yet. https://tex.stackexchange.com/questions/200643/how-to-cite-a-dictionary-entry

\ea \label{ex:bul:izobshto_00}
    \ea Двигателят изгасна и колата изобщо не даваше признаци за живот.\label{ex:bul:izobshto-negative}\\
        \gll Dvigatel-jat iz-gas-na i kola-ta izobšto ne dava-še priznatsi za život.\\
            engine.\textsc{m.sg}-\textsc{def.m.sg} \textsc{pfv}-extinguish-\textsc{pst} and car.\textsc{f.sg}-\textsc{def.f.sg} not.at.all \textsc{neg} give.\textsc{ipfv}-\textsc{impf} indication\textsc{.m.pl} for life.\textsc{m.sg} \\ 
        \glt `The engine died and the car showed no signs of life at all.' \\(Bulgarian National Corpus)
    \ex Но хепатитът често протича изобщо без пожълтяване. \label{ex:bul:izobshto-positive}\\
        \gll No hepatit-ət često protiča izobšto bez po-žəlt-javane.\\
            But hepatitis\textsc{m.sg}-\textsc{def.m.sg} often occur.\textsc{ipfv.prs.3sg} at.all without \textsc{pfx}-yellow-\textsc{nmlz}\\
        \glt `But hepatitis often occurs without yellowing at all.' \\(Bulgarian National Corpus)
    \z 
\z

\ea \label{ex:bul:vaobshte_00}
    \ea Въобще не се е постарал за това ние да вникнем в дилемата му...\label{ex:bul:vaobshte-negative}\\
        \gll Vəobšte ne se e postar-al za tova nie da vnikn-em v dilema-ta mu...\\
            at.all \textsc{neg} \textsc{refl} be.\textsc{3sg.prs} make.an.effort-\textsc{pst.ptcp.m.sg} for this \textsc{1pl} \textsc{conj} get.an.insight.\textsc{pfv}-\textsc{1pl.prs} in dilemma.\textsc{sg.f}-\textsc{def.sg.f} \textsc{3sg.dat\slash poss}\\
        \glt `[Apparently] [he] made no effort for us to get an insight into his dilemma...' \\(Bulgarian National Corpus)
    \ex Въобще селото изглежда като след война. \label{ex:bul:vaobshte-positive}\\
        \gll Vəobšte selo-to izgležda kato sled vojna.\\
            at.all village.\textsc{sg.n}-\textsc{def.sg.n} look.like.\textsc{ipfv.prs.3sg} like after war\\
        \glt `Generally, the village looks like after a war.' \\(Bulgarian National Corpus)
    \z 
\z

As mentioned above, both \textit{izobšto} and \textit{vəobšte}  are very frequently used in irrealis contexts. Analyses of their contextual preference based on random samples of 200 sentences extracted from the Bulgarian National Corpus are presented in \tabref{tab:bul:izobshto}.

\begin{table}
    \begin{tabular}{lrr}
        \lsptoprule
        {{Context}} & \textit{{izobšto}}& \textit{{vəobšte}}\\ 
        \midrule
        negative\slash irrealis   & 74,50\% & 67\%\\
        positive  & 25,50\% & 33\%\\
        \midrule
         & 100\% & 100\%\\
        \lspbottomrule
    \end{tabular}
    \caption{Contexts of use for \textit{izobšto} and \textit{vəobšte} in a random sample of 200 sentences}
    \label{tab:bul:izobshto}
\end{table}

As indicated in \tabref{tab:bul:izobshto}, in one-fourth of its uses, \textit{izobšto} appears in positive contexts; this is valid for one-third of the uses \textit{vəobšte}. Both adverbs show a clear preference for negative\slash irrealis contexts but they are also used often enough in the affirmative domain. This, in turn, precludes classifying them as full-fledged negative polarity items. In contrast, \textit{gək} and \textit{hič} are entirely restricted to the negative domain.

\subsection{Marking of NPs under the scope of negation} \label{sec:4.3}

Noun phrases under the scope of negation do not receive any special marking as shown by the data in (\ref{ex:bul:NP-1a}) and (\ref{ex:bul:NP-1b}). The only exception is the subject of a negated locative predication when the negative existential is used for its negation. In that case, the subject is marked by an object clitic, cf. example (\ref{ex:bul:loc-2c}) and pertinent discussion. 

Full noun phrases in object position follow the predicate and the standard negator as shown in (\ref{ex:bul:NP-1b}). 
The following can be said about pronominal objects. In affirmative sentences, they occur between the overt subject and the predicate as in (\ref{ex:bul:NP-1е}). If the subject is dropped as in (\ref{ex:bul:NP-1о}), the object pronoun follows the verb. In negated sentences, pronominal objects are placed between the negator and the lexical verb as in (\ref{ex:bul:NP-1c}). There is dialectal variation on this parameter. Within the entire Southeastern Slavic continuum, the pronoun is placed before the entire predicate phrase as in (\ref{ex:bul:NP-1d}) mainly in Central-Western varieties with a single pocket to the East, see blue areas on \href{https://ibl.bas.bg/lib/bda4/#page/169/mode/1up}{Map 143} \citep{AntonovaEtAl2016}. It is noteworthy that the varieties where the pronominal object occurs between the negator and the predicate surround the varieties where it is placed before the negator.

 \protectedex{\ea\label{ex:bul:NP}
    \ea Аз чувам телефона.\label{ex:bul:NP-1a}\\ 
      	\gll Az ču-v-am telefon-a.\\
         1\textsc{sg} hear-\textsc{ipfv}-\textsc{1sg.prs} telefon-\textsc{def.m.sg}\\
    \glt `I hear the phone.' = `I am hearing the phone ring.' (constructed example) 
    
    \ex Не чувам телефона.\label{ex:bul:NP-1b}\\ 
      	\gll Ne ču-v-am telefon-a.\\
         \textsc{neg} hear-\textsc{ipfv-1sg.prs} phone.\textsc{m.sg}-\textsc{def.m.sg}\\
    \glt `I don't hear the phone' (constructed example) 
    
    \ex Аз те чувам.\label{ex:bul:NP-1е}\\ 
      	\gll  Az te ču-v-am.\\
         1\textsc{sg} \textsc{2sg.obj} hear-\textsc{ipfv}-\textsc{1sg.prs}\\
    \glt `I hear you.' (constructed example) 
    
    \ex Чувам те.\label{ex:bul:NP-1о}\\ 
      	\gll ču-v-am te.\\
         hear-\textsc{ipfv}-\textsc{.1sg.prs} \textsc{2sg.obj}\\
    \glt `I hear you.' (constructed example)
    
     \ex Аз не те чувам. \label{ex:bul:NP-1c} \\ 
       	\gll Az ne te ču-v-am.\\
          \textsc{1sg} \textsc{neg} \textsc{2sg.obj} hear-\textsc{ipfv-1sg.prs}\\
     \glt `I don't hear you.'  (Central Eastern dialects within Bulgarian territory, red areas on the \href{https://ibl.bas.bg/lib/bda4/page/169/mode/1up}{Map 143} in \cite{AntonovaEtAl2016})%HC original link was https://ibl.bas.bg/lib/bda4/#page/169/mode/1up with hashtag before page but it creates an error
    
     \ex Аз те не чувам.\label{ex:bul:NP-1d}\\ 
       	\gll Az te ne ču-v-am.\\
          \textsc{1sg} \textsc{2sg.obj} \textsc{neg} hear-\textsc{ipfv-1sg.prs} \\
     \glt `I don't hear you.'  (Western dialects within Bulgarian territory, blue areas on the \href{https://ibl.bas.bg/lib/bda4/page/169/mode/1up}{Map 143} in \cite{AntonovaEtAl2016})%HC original link was https://ibl.bas.bg/lib/bda4/#page/169/mode/1up with hashtag before page but it creates an error
     \z 
 \z
}%


The data presented in \citet{AntonovaEtAl2016} include only sentences with overt subject. In standard Bulgarian, the only possibility is for the object pronoun to occur in between the negator and the predicate even when the subject is not overtly expressed, e.g. \textit{Не те чувам\slash Ne te ču-v-am}, similarly to (\ref{ex:bul:NP-1c}).

\subsection{Reinforcing negation} \label{sec:bul:reinf}

There are various ways to reinforce negation: by the special emphatic negative markers \textit{ni} and \textit{nito} as  in (\ref{ex:bul:reinf-1}), by topicalizing a negated object as shown in (\ref{ex:bul:reinf-1a}), by indefinite pronouns and negative concord as in (\ref{ex:bul:reinf-2}), and by the use of a minimizer together with \textit{nito} or \textit{ni}, see (\ref{ex:bul:polarity-1}) and (\ref{ex:bul:polarity-2}) above and (\ref{ex:bul:reinf-3}) below. The use of the adverb \textit{hič} `not at all' is another strategy to reinforce negation, see (\ref{ex:bul:polarity-hich00}) in \sectref{sec:bul:negative.polarity}.

The special emphatic negative markers \textit{ni} and \textit{nito} are used together with the predicate negator as in (\ref{ex:bul:reinf-1}). They are frequently used in coordinated constructions, cf. \sectref{sec:bul:coord}. The negative structures they produce are, however, always emphatic.

\ea \label{ex:bul:reinf-1}
\ea Ни думица горката му невеста не смея да каже.\\ \label{ex:bul:reinf-1a}
  	\gll Ni dum-itsa gorka-ta mu nevesta ne smeja da kaže.\\
     \textsc{neg.emph} word-\textsc{dim.f} poor.\textsc{f.sg}-\textsc{def.f.sg} \textsc{3sg.m.dat} bride \textsc{neg} dare.\textsc{pfv.3sg.aor} \textsc{conj} say.\textsc{ipfv.3sg.aor}\\
\glt `Not a single little word did his poor bride dare say.' \citep[492]{StojanovEtAl1983}

\ex Не чакай нито минута! \\\label{ex:bul:reinf-1b}
  	\gll Ne čak-aj nito minuta!\\
     \textsc{neg} wait-\textsc{imp.2sg} \textsc{neg.emph} minute\\
\glt `Do not wait a single minute!' (constructed example) 

\z \z

\ea \label{ex:bul:reinf-2}
\ea Никой нищо не знае.\\ \label{ex:bul:reinf-2a}
  	\gll Nikoj ništo ne znae.\\
     nobody nothing \textsc{neg} know.\textsc{3sg.prs}\\
\glt `No one knows anything/a single thing' (constructed example) 


\ex Нищо не чувам \\ \label{ex:bul:reinf-2b}
  	\gll Ništo ne ču-v-am.\\
     nothing \textsc{neg} hear-\textsc{ipfv}-\textsc{1sg.prs} \\
\glt `I don't hear no nothing.' (constructed example) 
\z \z

\newpage
\ea Тя не хапва нито трошичка.\\ \label{ex:bul:reinf-3}
  	\gll Tja ne xap-v-a nito troš-ička.\\
     \textsc{3sg.f} \textsc{neg} bite-\textsc{ipfv}-\textsc{3sg.prs} \textsc{neg.emph} crumb-\textsc{dim.f.sg}\\
\glt `She does not eat anything.'\\
lit. `She does not bite a single little crumb.' (constructed example) 

\z

\subsection{Negation, coordination and complex clauses} \label{sec:bul:coord}
Bulgarian has a special construction for emphatic negative coordination. Following \citet[664, in passim]{Salaberri2022}, it can be classified as having an \textsc{enc} construction encoded by the particles \textit{nito...nito} or their shortened variant \textit{ni...ni}. They can be the sole predicate negator as in (\ref{ex:bul:coord-1a}) or appear together with predicate negation as in (\ref{ex:bul:coord-1b}). In this respect, Bulgarian differs from a closely related language, Croatian (ISO: \href{https://iso639-3.sil.org/code/hrv}{hrv}, Glottocode: \href{https://glottolog.org/resource/languoid/id/croat1245}{croat1245}), where the longer and the shorter variants of the construction have different distributional properties. Specifically, in Croatian, the longer variant, \textit{niti...niti} scopes over the entire predication while the shorter one, \textit{ni...ni} is used for the negation of noun phrases only, see \citet[670]{Salaberri2022} citing \citet{AuweraKrasnoukhova2020}. In Bulgarian the shorter variant \textit{ni...ni} can also be used for predicate negation, see (\ref{ex:bul:coord-1c}) below.\footnote{The differences between Bulgarian and Croatian might be correlated with the fact that Croatian, unlike Bulgarian has a special negation strategy for idenfication and property predications which is also closely connected to special strategies for contrastive negation.}

%In the BulgarianNC I find \textit{ni--ni} more often as phrasal negator but uses with a scope over a whole proposition/predicate/predication are possible too. In the corpus I searched for \textit{ni} as a conjunction.

\ea \label{ex:bul:coord-1}
\ea Нито съм го чула, нито съм го видяла.\\ \label{ex:bul:coord-1a}
  	\gll Nito səm go ču-la, nito səm go vid-jala.\\
     \textsc{enc} be.\textsc{1sg.prs} \textsc{3sg.obj} hear.\textsc{pfv}-\textsc{pst.ptcp.f.sg} \textsc{enc} be.\textsc{1sg.prs} \textsc{3sg.obj} see.\textsc{pfv}-\textsc{pst.ptcp.f.sg}\\
\glt `I have neither heard, nor seen him.' (constructed example) 

\ex Не искам нито едното, нито другото. \\ \label{ex:bul:coord-1b} 
  	\gll Ne iska-m nito edno-to, nito drugo-to.\\
     \textsc{neg} want-\textsc{1sg.prs} \textsc{enc} one.\textsc{n.sg}-\textsc{def.n.sg} \textsc{enc} other.\textsc{n.sg}-\textsc{def.n.sg}\\
\glt `I don't want either one or the other.' (constructed example) 

\ex Стъкленият блясък в очите на Гъси говореше, че ни приема, ни предава.\\ \label{ex:bul:coord-1c}

\gll Stǝkl-en-ijat bljasǝk v oči-te na Gǝsi govore-še, če ni priem-a, ni predav-a\\
glass-\textsc{adj.m.sg}-\textsc{def.m.sg} glint in eye.\textsc{pl}-\textsc{def.pl} \textsc{prep} Gussie speak-\textsc{impf.3sg} that \textsc{enc} accept-\textsc{prs.3sg} \textsc{enc} send.out-\textsc{prs.3sg}\\
\glt `The glassy look in Gussie's eyes suggested he neither took in [anything], nor was he communicating.' (lit. `The glassy glint in Gussie's eyes spoke to [the fact] that he was neither accepting [anything], nor was he sending out [anything].' (Bulgarian National Corpus)
\z \z

In \citegen{Salaberri2022} classification, Bulgarian falls into the most populated α1 type since the \textsc{enc} construction is expressed by correlative markers without the addition of any other element. 

\subsection{Miscellaneous aspects of negation} \label{sec:bul:other_aspects}
In this section I discuss negative transport in \sectref{sec:bul:neg-transport}, contrastive negation in \sectref{sec:bul:contrastive_neg}, and metalinguistic negation in \sectref{sec:bul:metalinguistic_neg}. Non-negative uses of negation markers are presented in \sectref{sec:bul:non-negative}. I also offer a brief discussion on the origin of some of the expressions of negation discussed in this article in \sectref{sec:bul:diachrony}.

\subsubsection{Negative transport} \label{sec:bul:neg-transport}
The phenomenon of negative transport refers to the interpretation of negation in a complex clause structure. Specifically,  the marking of negation in the higher/main clause is interpreted as negation scoping over the predicate of the subordinate clause. For instance, in English, when verbs such as \textit{think}, \textit{believe}, \textit{expect} are negated as predicates in the main clause, the implication is that their complement clause is negated as in \textit{I don't think they will come}=\textit{I think they will not come}.

Negative transport does not occur in Bulgarian. In (\ref{ex:bul:trasport-2a}) below the negator scopes over the act of thinking, not over the action described in the subordinate clause. To negate that action, the negator has to be placed in the subordinate clause as in (\ref{ex:bul:trasport-2b}).

\ea \label{ex:bul:trasport-2}
\ea Не мисля че ще дойдат.\\ \label{ex:bul:trasport-2a}
  	\gll Ne misl-ja če šte dojd-at.\\
     \textsc{neg} think.\textsc{ipfv}-\textsc{1sg.prs} that \textsc{fut} come.\textsc{pfv}-\textsc{3pl.prs}\\
\glt `I don't think they will come.' (constructed example) 

\ex Мисля че няма да дойдат.\\ \label{ex:bul:trasport-2b} 
  	\gll Misl-ja če njama da dojd-at.\\
     think.\textsc{ipfv}-\textsc{1sg.prs} that \textsc{negex} \textsc{conj} come.\textsc{pfv}-\textsc{3pl.prs}\\
\glt `I think they will not come' (constructed example) 
\z \z

\subsubsection{Contrastive negation} \label{sec:bul:contrastive_neg}
As discussed in \sectref{sec:bul:scope}, several of the already mentioned expressions of negation, i.e. the SN marker \textit{ne} as well as \textit{nito/ni}, can be used for the encoding of \textit{contrastive negation}, that is, the opposition between a negated element and an alternative. This is shown in (\ref{ex:bul:contrastive-1}) where the SN marker \textit{ne} is used as a constituent negator; see (\ref{ex:bul:coord-1b}) for an example of \textit{nito...nito} as both a constituent and a contrastive negator.

\ea Следобед пия чай, не кафе.\\ \label{ex:bul:contrastive-1} 
  	\gll Sledobed pi-ja čaj, ne kafe.\\
  	Afternoon drink.\textsc{ipfv}-\textsc{1sg.prs} tea \textsc{neg} coffee\\
\glt `In the afternoon, I drink tea, not coffee' (constructed example) 
\z

As noted in \sectref{sect:stative}, when the SN marker \textit{ne} is used as an ascriptive negator, it is also typically marking contrastive negation as well, see (\ref{ex:bul:stat-1b}) as well as pertinent discussion in \citet{Mettouchi2006}, also \citet{Mettouchi2021}.\footnote{Mettouchi \citeyear{Mettouchi2021} suggests the term \textit{attributive-equative} for this kind of negator.}

\subsubsection{Metalinguistic negation} \label{sec:bul:metalinguistic_neg}
Metalinguistic negation is shown in (\ref{ex:bul:metalinguistic}). By using negation the speaker objects to the choice of \textit{časovnik} `clock', a term for a broader category, and opts for the more specific term \textit{budilnik} `alarm clock' which is a subcategory of the broader one. 

%Horn (1989:363), is “a device for objecting to a previous utterance,” whereby the speaker repeats in the negative a formulation used by the addressee, and proposes a correction. 

\ea Не часовник, ами цял будилник ще ти намеря.\\ \label{ex:bul:metalinguistic} 
  	\gll Ne časovnik, ami tsjal budilnik šte ti namerja.\\
  	\textsc{neg} clock, but whole alarm\_clock  \textsc{fut} \textsc{2sg.obl} find.\textsc{pfv.1sg.prs}\\
\glt lit. `Not a clock, but I will find you a whole alarm clock' =\\ 'Not [just] a clock, but I will find you an alarm clock' (Bulgarian National Corpus) 
\z
%\lv{punctuation and translation of Bulgarian titles}
%\ea \label{ex:bul:meta-1}
 %   \begin{xlist}[B: ]
  %  \exi{A: }\label{ex:bul:meta-1a} Дара не изглади ризите! \\
   %     \gll Dara ne iz-glad-i riz-i-te! \\ 
    %         Dara \textsc{neg} \textsc{pfv}-iron-\textsc{3sg.aor} shirt-\textsc{pl}-\textsc{def.pl} \\ 
     %   \glt Dara did not iron the shirts!'
    %\exi{B: }\label{ex:bul:meta-1b} Не само ги изглади ами ги сгъна и ги прибра в гардероба \\
     %   \gll Ne samo gi iz-glad-i ami gi s-gən-a i gi pribr-a v garderob-a \\ 
      %       \textsc{neg} only \textsc{3pl.obj} \textsc{pfv}-iron-\textsc{3sg.aor} but \textsc{3pl.obj} \textsc{pfv}-fold-\textsc{3sg.aor} and \textsc{3pl.obj} put.\textsc{pfv}-\textsc{3sg.aor} into wardrobe-\textsc{def.m.sg} \\
       % \glt Not only did she iron them but folded them and put them in the wardrobe' (constructed example) 
    %\end{xlist}
%\z 

\subsubsection{Non-negative uses of negatives and expletive negation} \label{sec:bul:non-negative}
% see the section on negation in the Bulgarian grammar, volume 3 syntax. There I can pick up examples of negation where the authors state that a positive content is actually meant. They also have examples of constituent negation where the subject is negated, Page 22 in the djv file.
%If I need more material on negative questions, I can go to Page 23 in the djv file.

Expletive negation refers to the use of a negative marker when no negative meaning is intended. Among the expressions of negation discussed in this article, only \textit{ne} can be used as an expletive negator.  This is typically triggered by the adverbial subjunction \textit{dokato} `until' as shown in (\ref{ex:bul:nonneg-2}).\footnote{See \citet{Žereva2012} for further discussion of non-negative uses of negation markers in Bulgarian.}

\ea Няма да тръгна докато не заспиш.\\ \label{ex:bul:nonneg-2} 
  	\gll Njama da trəgna dokato ne za-spi-š.\\
  	\textsc{negex} \textsc{conj} leave.\textsc{pfv.1sg.prs} until \textsc{neg} \textsc{pfv}-sleep-\textsc{2sg.prs}\\
\glt `I won't leave until you fall asleep' (constructed example) 
\z

Negative polar questions shown in (\ref{ex:bul:nonneg-2a}) and also (\ref{ex:bul:qn-1a}--\ref{ex:bul:qn-nima}) are also often interpreted as rhetorical questions; in these cases the negative markers used in them \textit{ne} and \textit{nima} can be said to have a non-negative function. Or we could say that the entire negative question is used to convey a non-negative meaning (Matti Miestamo, p.c.).

\ea Не улових ли те два пъти?\\ \label{ex:bul:nonneg-2a}
  	\gll Ne u-lov-ix li te dva pət-i?\\
     \textsc{neg} \textsc{pfv}-catch-\textsc{1sg.pst} \textsc{q} \textsc{2sg.obj} two.\textsc{m} way.\textsc{m}-\textsc{pl}\\
\glt `I caught you twice.' (lit. `Did I not catch you twice') \citep[492]{StojanovEtAl1983}
\z

\subsubsection{Diachronic notes and observations} \label{sec:bul:diachrony}
In this section I present diachronic information about some of the negation markers discussed in this chapter and also outline hypotheses about their development. I start with a discussion on the evolution of \textit{stiga} as a prohibitive auxiliariy, proceed with the lexicalizations of negation, and close with a discussion on the non-operation of the Negative Existential Cycle in Bulgarian although examining the modern data only would suggest otherwise.

The following can be said about the form \textit{stiga}. Morphologically, it is the third person singular present of the verb \textit{stigam} which is a polysemous verb. One of its senses is motion `reach, arrive', another is `reach to X, be able to touch'. Yet another sense is `be enough, be satisfied', for this sense the verb is used with a dative subject. The regular imperative \textit{stigaj} occurs only with the sense of motion, e.g. reaching a specific place; it is a form of very low frequency, with only 18 occurrences in the Bulgarian National Corpus. 

The form \textit{stiga} is formally part of the paradigm of the verb \textit{stigam} '1. reach, arrive; 2. be enough'. Yet, it is best described as a special form, split from this verb. In Bulgarian dictionaries it is described as a particle. It is used to indicate that the action performed by the listener should stop immediately, its literal meaning is `Enough!/End!'. It is the most frequent form of this verb. In the Bulgarian National Corpus a search for all forms of the verb \textit{stigam} yields 46 904 hits. Among them, the form \textit{stiga} alone occurs 32 999 times, so 70\% of all occurrences of \textit{stigam}. The uses of the form \textit{stiga} are examined in greater detailed based on a random sample of 200 sentences extracted from all hits with the randomizing tool of SketchEngine. The analysis of the random sample allows for the postulation of four broad categories/groups. The first group covers examples where \textit{stiga} is used as a prohibitive auxiliary as in (\ref{ex:bul:imp-stiga}), see also (\ref{ex:bul:stiga-stop}).  The second group includes examples where it is part of a complex conjunction \textit{stiga da} `as long as, if' shown in (\ref{ex:bul:stiga-da}). The third group covers examples where \textit{stiga} is used as a verb of motion that includes the reaching of a goal `reach, arrive', either literally or metaphorically, see (\ref{ex:bul:stiga-reach}). The fourth group includes examples where \textit{stiga} is used with the sense `enough' in various constructions, one of which is shown in (\ref{ex:bul:stiga-enough}). 

\ea \label{ex:bul:stiga-other}
    \ea Всяка криза предлага възможности, стига да ги изобретиш.\label{ex:bul:stiga-da}\\
        \gll Vsjak-a kriza predlaga vǝzmožnost-i, stiga da gi izobret-iš.\\
            every-\textsc{sg.f} kriza.\textsc{sg.f} offer.\textsc{ipfv.3sg.prs} opportunity-\textsc{pl}, enough \textsc{conj} \textsc{3pl.obj} invent.\textsc{ipfv}-\textsc{2sg.prs}\\
        \glt `Every crisis offers opportunities as long as you invent them.' (Bulgarian National Corpus)
    
    \ex  До плажната ивица се стига по черен път.\label{ex:bul:stiga-reach}\\
        \gll Do plažna-ta ivitsa se stiga po čeren pǝt.\\
            \textsc{prep} beach.\textsc{adj.sg.f}-\textsc{def.sg.f} stripe.\textsc{sg.f} \textsc{refl} reach \textsc{prep} black\textsc{sg.m} road.\textsc{sg.m}\\
        \glt `The beach is reached by a dirt road.' (Bulgarian National Corpus)
        
    \ex За тях най-важното е пенсията да им стига за хляб, кисело мляко, лекарства, [...]. \label{ex:bul:stiga-enough}\\
        \gll Za tjah naj-važno-to e pensija-ta da im stiga za hljab, kiselo mljako, lekarstv-a [...]\\
            for \textsc{3pl.obj} most-important.\textsc{3sg.n}-\textsc{def.3sg.n} be.\textsc{3sg.prs} pension.\textsc{sg.f}-\textsc{def.sg.f} \textsc{conj} \textsc{3pl.dat} be.enough.\textsc{3sg.prs} for bread, sour.\textsc{sg.n} milk.\textsc{sg.n}, medicine-\textsc{pl}\\
        \glt `For them, the most important thing is that their pension is enough for bread, yogurt, medicine,[...]' (Bulgarian National Corpus)
    \z 
\z 

A summary of the frequency of occurrence of these groups in the random sample is presented in \tabref{tab:bul:stiga} below.


\begin{table}
    \begin{tabular}{lrr}
        \lsptoprule
        \textbf{Function} & \textbf{Number of hits}& \textbf{Percent}\\
        \midrule
        \textit{стига \slash stiga} `stop, prohibitive auxiliary'  & 75 & 37.5\%\\
        \textit{стига се (до)\slash stiga se (do)} `arrive at, reach'  & 53 & 26.5\%\\
        \textit{стига да\slash stiga da} `as long as, if' & 48 & 24.0\%\\
        \textit{стига \slash stiga}  `be enough' (various constructions) & 24 & 12.0\%\\
        \midrule
         & 200 & 100\%\\
        \lspbottomrule
    \end{tabular}
    \caption{Functions of the form \textit{стига\slash stiga} in a random sample of 200 sentences}
    \label{tab:bul:stiga} 
\end{table}

As indicated in \tabref{tab:bul:stiga}, the form \textit{stiga} is used with both grammatical and lexical meanings. Its uses as a prohibitive auxiliary are the most interesting ones here given the topic of this article. \textit{Stiga} is used with this function in close to 40\% of the randomized sample. It is observed in several different constructions: in a complex predicate construction as shown in (\ref{ex:bul:imp-stiga}), with nominal constituents as in (\ref{ex:bul:stiga-stop-whining}) or in elliptic sentences on its own as in (\ref{ex:bul:stiga-stop-prosentence}).

\ea\label{ex:bul:stiga-stop}
    \ea Стига хленчене!\\ \label{ex:bul:stiga-stop-whining} 
        \gll Stiga hlenč-ene!\\
            enough whine-\textsc{nmlz}\\
        \glt `Stop whining!' (lit. `enough whining') (Bulgarian National Corpus)
        \ex Айде стига!\\ \label{ex:bul:stiga-stop-prosentence}
        \gll Ajde stiga!\\
            c'mon enough\\
        \glt `Come on, stop!' (Bulgarian National Corpus)
    \z 
\z


Conceptually, we can suggest grammaticalization paths such as those outlined in (\ref{ex:bul:stiga1}) and (\ref{ex:bul:stiga2}).

\protectedex{
\ea reach, arrive > satisfaction/fullness > stop/end \label{ex:bul:stiga1} 
\ex reach, arrive > stop/end \label{ex:bul:stiga2} 
\z }

The fact that the form \textit{stiga} appears in elliptical sentences with the sense `enough\slash stop' and also that imperative morphology is preserved for the motion sense suggest that the construal of stopping an action or banning it builds on the sense of reaching satifaction or fullness rather than on directed motion alone. As pointed out by \citet{Auwera2006}, the semantic connection between the sense `enough' and prohibition is easy to understand when we think of uses of expressions meaning `enough' which in many languages also indicate that the action should stop. In other words, the synchronic data seem to give more support to (\ref{ex:bul:stiga1}). However, more research is necessary where historical data are analyzed as well. Besides, as one of the reviewers suggests, the grammaticalization of \textit{stiga} as a prohibitive auxiliary has to be contrasted with the evolution of the complex subjunction \textit{stiga da} `as long as, if'.

The lexicalizations of negation cited in \sectref{sec:bul:l} together with several of the prohibitive auxiliaries are all fairly transparent univerbations of the negative particle \textit{ne} and a positive stem, see data  in (\ref{ex:bul:diach-1}).

\ea \label{ex:bul:diach-1}
    \ea няма/njama  `not.have, not.exist' < ne + Old Church Slavonic имѣти, имамь, имаши/iměti, imamь, imeši `have, hold'\\ \label{ex:bul:diach-1a} \href{https://ibl.bas.bg/lib/ber_4_000-1004/#page/732/mode/1up}{\citet[v.4:732]{GeorgievIlčev1971}}, see also \citealt{OldChurchSlavonicOnline2020}
      	
    \ex недей/nedej `do not' < ne +Old Church Slavonic дѣꙗти, дѣѭ, дѣѥши/dějati, dějǫ, děješi `do, commit' \\ \label{ex:bul:diach-1b} \href{https://ibl.bas.bg/lib/ber_4_000-1004/#page/597/mode/1up}{\citet[v.4:597]{GeorgievIlčev1971}} see also \citealt{OldChurchSlavonicOnline2020}
     
    \ex немой/nemoj `do not' \\ < ne +Old Church Slavonic  мошти, могѫ, можеши/mošti, mogǫ, možeši `be able, can' \\ \label{ex:bul:diach-1c} \href{https://ibl.bas.bg/lib/ber_4_000-1004/#page/610/mode/1up}{\citet[v.4:610]{GeorgievIlčev1971}} see also \citealt{OldChurchSlavonicOnline2020}
    
    \ex невем/nevem `I don't know' \\ < ne + Old Church Slavonic  вѣдѣти, вѣмь, вѣси/věděti, věmь, věsi `see, know' \\ \label{ex:bul:diach-1d} \href{https://ibl.bas.bg/lib/ber_4_000-1004/#page/585/mode/1up}{\citet[v.4:585]{GeorgievIlčev1971}} see also \citealt{OldChurchSlavonicOnline2020}
    \z 
\z
%\hc{Hei! So here i would link the source from the reference instead of the link to the website directly. Because like it is now, the source is not even in the references. it would be just \citealt{OldChurchSlavonicOnline2020}}\lv{ok}\hc{to do}

It should be noted that although the forms cited in (\ref{ex:bul:diach-1}) clearly originate from the fusion of two morphemes, synchronically they constitute single meaningful units. As demonstrated by the data in (\ref{ex:bul:diach-1}), the erstwhile morpheme boundaries have been blurred to various degrees. For instance, any further morphological analysis and morphemic segmentation would be misguided with \textit{njama} `not.have, not.exist'. Morpheme boundaries are still discernible with the remaining lexicalizations. However, it has to be said that the stems following the prefixed \textit{ne}, while still having an identifiable sense, are effectively restricted to the forms cited in (\ref{ex:bul:diach-1b}) through (\ref{ex:bul:diach-1d}).

Among the lexicalizations listed in (\ref{ex:bul:diach-1}) only \textit{njama} has a positive counterpart \textit{ima} `have.\textsc{3sg.prs}, exist'. The Old Church Slavonic дѣꙗти, дѣѭ, дѣѥши \textit{dějati, dějə, dějəši} `do, commit' is not used in the positive in Modern Bulgarian while \textit{мошти, могѫ, можеши}/\textit{mošti, mogǫ, možeši} `be able can' survives as the modal verb \textit{moga} `be able to, can' but it does not have an imperative. As for \textit{вѣдѣти, вѣмь, вѣси}/\textit{věděti, věmь, věsi} `see, know', its stem appears only in compounds or derived forms. Some of them express knowledge/expertise of X as in (\ref{ex:bul:diach-2a}) and (\ref{ex:bul:diach-2b}) but many of them are negative derivations, which typically express negative content as in (\ref{ex:bul:diach-2c}) and (\ref{ex:bul:diach-2d}).

\ea \label{ex:bul:diach-2}
\ea литератур-о-вед\\ \label{ex:bul:diach-2a}
\gll literatur-o-ved\\
 literature-\textsc{link}-expert/one.who.knows\\
\glt `literature expert, literature /scholar' (constructed example)\\

\ex език-о-вед\\ \label{ex:bul:diach-2b}
\gll ezik-o-ved\\
 language-\textsc{link}-expert/one.who.knows \\
\glt `language expert, language scholar' / `linguist' (constructed example)\\

\ex не-вед/не-веж\\ \label{ex:bul:diach-2c}
\gll ne-ved/ne-vež\\
 \textsc{neg}-expert/one.who.knows \\
\glt `not-knowing, one who does not know, clueless, illiterate'

\ex не-веде-ние\\ \label{ex:bul:diach-2d}
\gll ne-vede-nie\\
 \textsc{neg}-know-\textsc{nmlz} \\
\glt `(state of) lack of knowledge, lack of information'

\z \z

\noindent 
Another issue relevant for the diachrony of negation in Bulgarian is the operation of the Negative Existential Cycle. This diachronic model was first formulated by \citet{Croft1991} to model the evolution of standard negation markers from negative existentials. The synchronic data may suggest that in Bulgarian we observe a partial operation of the cycle because of use of the negative existential \textit{njama} for the negation of verbal predications marked for future tense (cf. (\ref{ex:bul:prs-3b}) above, repeated as (\ref{ex:bul:diach-3}) below for convenience). However, the diachronic data do not support this hypothesis. 

As indicated in (\ref{ex:bul:diach-4}), in Old Church Slavonic there were several ways to encode future tense. One of the affirmative future constructions resulted from the grammaticalization of the verb хотѣти\slash \textit{xotěti} `want', shown in (\ref{ex:bul:diach-4b}), into the current particle \textit{ще/šte}, shown in (\ref{ex:bul:diach-6}). Another Old Church Slavonic future construction involved \textit{iměti} `have' + infinitive of the lexical verb, cf. (\ref{ex:bul:diach-4a}). This construction, shown in (\ref{ex:bul:diach-5}), was frequently used with negation, cf. \citet[201]{Đorđić1975}. Consequently, the construction we see in Modern Bulgarian in (\ref{ex:bul:diach-3}) does not result from an expansion of the uses of the negative existential but represents an inherited, modernized version of the Old Church Slavonic construction, cf. \citet{Veselinova2010,Veselinova2014,Veselinova2016} for detailed discussions.

\ea Different ways to encode future tense periphrastically in Old Church Slavonic\\ \label{ex:bul:diach-4}
  \ea имѣти/iměti `have' + infinitive\\ \label{ex:bul:diach-4a}
  \ex хотѣти/xotěti `want'  + infinitive\\ \label{ex:bul:diach-4b}
\ex начати/načati `begin' + infinitive\\
\ex бѫдѫ/bǫdǫ `be.\textsc{fut}, become' + infinitive \\
(\cite[418]{Duridanov1991}, see also \cite[288-294]{Slavova217})
\z \z

\ea\label{ex:bul:diach-5} Не иматъ остати сьде камень на камени.\\ 
  	\gll Ne imat ostati s'de kamen na kamen-i.\\
  	\textsc{neg} have.\textsc{3sg.prs.ind} leave.\textsc{inf} here stone \textsc{prep} stone-\textsc{m.loc.sg}\\
    \glt `there shall not be left one stone upon another'\\ (\cite[418]{Duridanov1991} citing Mark 13.2)\footnote{The glossing follows the one offered in the Proiel Project, specifically for this verse, \href{http://foni.uio.no/proiel/sentences/37108}{http://foni.uio.no/proiel/sentences/37108}; note that a registration, free of charge, is required in order to access the data in the project. The translation is cited from the King James Bible Online, \href{https://thekingsbible.com/Bible/41/13}{https://thekingsbible.com/Bible/41/13}.}
\z 

\ea С̇тефан ще тича.\label{ex:bul:diach-6} \\ 
  	\gll Stefan šte tič-a. \\
     Stefan \textsc{fut} run.\textsc{ipfv}-\textsc{3sg}.\textsc{prs}\\
\glt `Stefan will run.' (constructed example)
  \z
  
\ea С̇тефан няма да тича.\\ \label{ex:bul:diach-3}
  	\gll Stefan njama da  tič-a.\\
  	Stefan \textsc{negex} \textsc{conj} run.\textsc{ipfv}-\textsc{3sg}.\textsc{prs} \\
    \glt `Stefan will not run.' (constructed example)
\z 

Generally, the Bulgarian facts cited above highlight the need to test diachronic hypotheses by diachronic data whenever possible.

\section{Summary}
Negation in Bulgarian operates on clause, constituent or word level. The language follows several wide-spread cross-linguistic trends: its SN marker is a pre-verbal particle, it has a special negative existential, and several dedicated encodings of the prohibitive, and negative indefinite pronouns are abundantly used with predicate negation. A complex predicate construction, seemingly identical to the Balkan subordinate clause construction, is used for the encoding of predications set in the future as well as for several variants of the prohibitive and also for the hortative. The negative lexicalizations identified in the language also fall into known cross-linguistic tendencies: specifically, we observe expressions for nonexistence, lack of ability and lack of knowledge. Features that are  perhaps typologically less common include the presence of special emphatic negators which also function as coordinative negators as well as the presence of a special negative question particle. However, this conclusion may be subject to revision given that world-wide typologies of these features are still in demand. The main expressions of negation in Bulgarian are summarized in \tabref{tab:bul:negsummary} below.

\begin{table}
 \small
    \begin{tabularx}{\textwidth}{llQ}
        \lsptoprule
        \textbf{Negative expression} & \textbf{Word class} & \textbf{Function}\\
        \midrule
        \textit{не/ne}  & particle & standard negation non-future, negative reply, constituent negator, expletive negator\\
        \textit{няма/njama}  & auxiliary, verb & standard negation future, negative existential, negative reply\\
        \textit{недей/nedej}  & auxiliary & prohibitive auxiliary\\
        \textit{немой/nemoj}  & auxiliary & prohibitive auxiliary (non-standard variants) \\
        \textit{стига/stiga} `enough'  & auxiliary & prohibitive auxiliary\\
        \textit{невем/nevem}  & invariable verb form & `I don't know' (non-standard variant)\\
        \textit{нито/ни / nito/ni}  & particle & emphatic negator, coordinative negator\\
        \textit{нима/nima}  & particle & negative question particle\\
        \lspbottomrule
    \end{tabularx}
    \caption{Expressions of negation in Bulgarian}
    \label{tab:bul:negsummary}
\end{table}


\section*{Acknowledgements}
I would like to thank Matti Miestamo, Björn Wiemer and two anonymous reviewers for their suggestions, questions and constructive criticism, all of which substantially improved this article. Needless to say, all shortcomings remain my own.

Many thanks also to Nina Knobloch and Sebastian Nordhoff for patiently answering my questions about various LaTeX issues. A big thank you too to Héloïse Calame for her meticulous work and careful proofreading.

The first version of this article was written while locked up at home during the COVID19 pandemic. Many thanks to my mom, Zhivka Veselinova, and to my daughter, Vera-Kathleen, for not uttering a single sound (\textit{гък/gək}) while I worked.

% \lv[inline,size=\scriptsize,color=blue!40]{@LSP: This is about the bibliography: long URLs that stretch across two lines do not work since only the first part is picked up when clicking on them. Please advise. In addition, the publisher for the very last reference, Zhereva, Maria. 2012, should be Akademichno Izdatelstvo "Prof. Marin Drinov". However, if we put the double quotes in the bibliography post by \", that combination produces an umlaut mark over P in Prof. This is obviously wrong so I removed the double quotes.}

\section*{Abbreviations}

\begin{tabularx}{.45\textwidth}{lQ}
\textsc{active} & active \\
\textsc{adj} & adjective\\
\textsc{adv} & adverb(ial)\\
\textsc{aor} & aorist\\
\textsc{conj} & conjunction\\
\textsc{dat} & dative\\
\textsc{def} & definite\\
\textsc{dim} & diminutive\\
\textsc{emph} & emphatic\\
\textsc{enc} & emphatic negative coordination\\
\textsc{f} & feminine\\
\textsc{fut} & future\\
\textsc{imp} & imperative\\
\textsc{impf} & imperfect\\
\textsc{inf} & infinitive\\
\textsc{ipfv} & imperfective\\
\textsc{ivf} & infinite verb Form\\
\textsc{link} & linking morpheme\\
\end{tabularx}
\quad
\begin{tabularx}{.45\textwidth}{lQ}

\textsc{m} & masculine\\
\textsc{n} & neuter\\
\textsc{neg} & negation\slash negator\\
\textsc{negex} & negative existential\\
\textsc{nmlz} & nominalization\\
\textsc{obj} & object\\
\textsc{obl} & oblique\\
\textsc{pfv} & perfective\\
\textsc{pl} & plural\\
\textsc{poss} & possessive\\
\textsc{pfx} & prefix\\
\textsc{prep} & preposition\\
\textsc{prs} & present tense\\
\textsc{pst} & past tense\\
\textsc{ptcp} & participle\\
\textsc{q} & question\\
\textsc{refl} & reflexive\\
\textsc{sg} & singular\\
\textsc{sn} & standard negation\\
\textsc{sub} & subordinator\\
\end{tabularx}


\printbibliography[heading=subbibliography,notkeyword=this]

\end{document}
