\documentclass[output=paper,colorlinks=true, citecolor=brown,draftmode]{langscibook}
\ChapterDOI{10.5281/zenodo.18233768}
\author{Rozenn Guérois\affiliation{CNRS-LLACAN, Paris; University of KwaZulu-Natal}}
\title{Negation in Cuwabo}
\abstract{This chapter provides a descriptive analysis of the negative constructions in Cuwa\-bo (Bantu P34) on both clausal and non-clausal levels. Formal properties such as type, position and number of negative markers are dealt with, as well as the structural differences between affirmatives versus negatives. On the clausal level, distinction is made between declaratives, non-declaratives, verbless predication and non-main clauses. Non-clausal negation, usually poorly described in Bantu grammars, involves negative replies, negative indefinites, and negative derivation. This paper finally addresses several semantic-pragmatic aspects linked with negation. Of particular interest is the recurrent use of ironic negative constructions in order to convey emphatic affirmative statements.

\keywords{negation strategies, (non-)clausal negation, paradigmatic (a)symme\-try, ironic negation, Cuwabo (Bantu)}}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 

\section{The language}\label{sec:cuw:1}
%Cuwabo is an Eastern Bantu language, coded as chw (ISO 693-3), chuw1238 (Glottolog) and P34 in \citet{Guthrie1967} referential classification. 
Cuwabo (ISO: \href{https://iso639-3.sil.org/code/chw}{chw}, Glottocode: \href{https://glottolog.org/resource/languoid/id/chuw1238}{chuw1238}), pronounced /ʧuwabo/, is an Eastern Bantu language, coded as P34 in \citegen{Guthrie1967} referential classification.
It is spoken in the Zambézia province of Mozambique by more than one million speakers (\citealt{InstitutoNacionaldeEstatísticas2017}). The number and names of the different Cuwabo varieties differ depending on the sources. The Ethnologue (\citealt{LewisEtAl2014})\footnote{This dialectal census could not be checked in the later editions of the Ethnologue because the access to any informative page have become a paid (and quite expensive!) service.}  distinguishes five main varieties: central Cuwabo, Karungu, Mayindo, Marale, and Nyaringa, whereas \citet{Guérois2015} excludes Marale (seemingly closer to Sena) and adds Manyawa, spoken around Lugela. The data used in this chapter belong to the central variety, simply referred to as Cuwabo, and located around Quelimane, the capital city of the Zambézia province. Quelimane corresponds both traditionally and geographically to the homeland of the Cuwabo people. 

Central Cuwabo has 31 consonants and a five-vowel system, which distinguishes short and long vowels. In this chapter, I mostly follow the standardized orthographic system adopted in \citet{NgungaFaquir2011}, where most graphemes are equivalent to the phonemes, except for the following: tt = /ʈ/, dd = /ɖ/, h = /ʔ/, dh = /ð/, ng = /ŋg/, ñg = /ŋ/, and lr = /ɽ/. As conventionally practiced in Bantu studies, the graphemes j and y stand for the palatal stop /ɟ/ and the palatal approximant /j/, respectively. For each example provided in this chapter, the first tier presents a broad phonetic representation written in phonetic symbols. It includes surface morphophonological processes, including resyllabification processes and tonal processes of doubling and spreading. The second tier, written in graphemes, provides the morphological segmentation of each utterance. On this tier, only the underlying High tones are represented. 

Cuwabo is a tone language with a binary distinction between High-toned moras (indicated with an acute accent on the segment) and toneless moras (zero-marked). The tone-bearing unit is the mora. Tones are both lexically and grammatically significant. There are two tone rules. The most frequent rule is High-Tone doubling, which consists of copying an underlying H only onto the next mora to the right, both word-internally and across word boundaries. Plateauing whereby at least two toneless moras become H between two underlying H tones word-internally, is the second tone rule. 

As most Bantu languages,\footnote{Among sources defining what typical Bantu languages are, the reader is referred to general works such as \citet[11--42]{WilliamsonBlench2000}, \citet{NursePhilippson2003} and \citet{VandeVeldeEtAl2019}.} Cuwabo has a predominantly head-marking morphology characterized by an extensive system of noun class agreement numbered from 1 to 18. Among them, class 15 prefix \textit{o}{}- serves to form infinitives. In addition to noun class prefixes, Cuwabo nouns carry an “augment tone” whose presence or absence is mostly conditioned by pragmatic factors, such as referentiality. This issue is discussed in \sectref{sec:cuw:4.3}.

Cuwabo verbal morphology is highly agglutinative. Its verbal complex consists of a string of grammatical morphemes that appear in a strict order organized around a stem nucleus, which is in turn built upon a verbal root optionally hosting one or several derivational suffixes. The elements that appear in these verbal slots convey a range of grammatical information, as can be seen on examination of \tabref{tab:cuw:1}. With regard to polarity, three slots are of importance in Cuwabo: the pre-initial slot, the post-initial slot and the final slot.

\begin{table}
\begin{tabularx}{\textwidth}{lQl}
\lsptoprule
{slots} & {functions} & {negators} \\
\midrule
{\textbf{pre}{}-\textbf{initial}} & {TAM, \textbf{negation}} & {\textit{ka}{}-} \\
{initial} &{subject} & \\
{\textbf{post}{}-\textbf{initial}} &{TAM, AM, \textbf{negation}} & {\textit{hi}{}- \textit{/ náá}{}-}\\
{pre-radical} &{object, reflexive} &  \\
{radical} &{verb root} &  \\
{pre-final} &{derivation, TA} &  \\
{\bfseries final} &{TAM, \textbf{negation}} & {{}-\textit{i}}\\
{post-final} & {plural imperative suffix, locative, comitative/instrumental, restrictive, intensive, subject pronouns in relatives} &  \\
\lspbottomrule
\end{tabularx}
\caption{The morphological structure of the Cuwabo verb (based on \citeauthor{Güldemann1999}'s \citeyear{Güldemann1999} terminology)}
\label{tab:cuw:1}
\end{table}

Cuwabo negation is mostly preverbal. In this, it conforms to the universal preference to place negation before the verb stem (\citealt{Dahl1979}; \citealt{Dryer1988, Dryer2011-wals143}). Pre-stem negative prefixation is the pattern standardly found across Bantu languages, most likely inherited from Proto-Bantu (\citealt{Westphal1958}; \citealt{KambaMuzenga1981}; \citealt{Güldemann1999}). See \sectref{sec:cuw:4.6.5} for diachronic observations.

Variation in Bantu negation strategies has been the subject of a number of previous studies (see \citealt{Westphal1958}; \citealt{KambaMuzenga1981}; \citealt{Contini-Morava1989}; \citealt{Güldemann1999}; \citealt{DevosAuwera2013}; \citealt{GuéroisEtAlForthcoming}, for comparative studies). This chapter aims to provide an exhaustive and typologically-informed
% \lv{@LSP: Alena Witzlak-Makarevich is asking for the line break to be fixed.}
picture of the different negative constructions attested in Cuwabo, on both clausal and non-clausal levels.

The outline of this chapter follows the questionnaire provided by the editors (\citetv{chapters/questionnaire}):
% \mm{@LSP: cross-ref to the questionnaire})
\sectref{sec:cuw:2} explores negation on the clausal level, covering both verbal and non-verbal predication; \sectref{sec:cuw:3} examines non-clausal negation, i.e. negative items or constructions such as negative replies, negative indefinites, and negative derivation; \sectref{sec:cuw:4} describes other aspects of negation, including the scope of negation, negative polarity, the pragmatics and marking of NPs under the scope of negation, negation reinforcers, negative coordinators, negative transport, metalinguistic negation, ironic negation and diachronic considerations. A summary of the chapter is provided in \sectref{sec:cuw:5}.

Most of the data presented in this paper come from the author’s fieldwork, conducted around Quelimane between 2011 and 2013. For each example, the type of source (story, conversation, image, elicitation) is provided between square brackets. The reader is referred to \citet{Guérois2015,Guérois2019} for a full-fledged description of the language. 

\section{Clausal negation}\label{sec:cuw:2}
This section explores clausal negation constructions in Cuwabo. Following the questionnaire, standard negation, i.e. negation in declarative main clauses, is first described in \sectref{sec:cuw:2.1}, as opposed to negation in non-declarative main clauses, which is covered in \sectref{sec:cuw:2.2}. \sectref{sec:cuw:2.3} and \sectref{sec:cuw:2.4} are concerned with negation in non-verbal clauses and negation in non-main clauses, respectively. No discussion of negative lexicalizations \citep{Veselinova2013lexicalization} is provided as they were not observed on the basis of my corpus.

\subsection{Standard negation} \label{sec:cuw:2.1}
Negation in declarative verbal main clauses is invariably marked by the pre-initial negator \textit{ka}{}-. This is seen with the past imperfective \textit{kaálóga} and the past perfective \textit{kaaziwílé} in \REF{ex:cuw:1} and the hypothetical \textit{kamugaabélél̩le} in \REF{ex:cuw:2}.\footnote{Subject prefixes of 3\textsuperscript{rd} person singular and plural are expressed by means of noun classes, hence they are glossed \textsc{sbj.1} and \textsc{sbj.2}. Speech act participants are, in turn, glossed \textsc{sbj.1sg/sbj.2sg/sbj.1pl/sbj.2pl}.}
% \hc{@LSP: if this first line is not in italics, the author wishes for |xxx| (because it is phonetic notation)}
\ea\label{ex:cuw:1} mwáán’ óóĺ̩le kaálóga. kaaziwíl’ óologā\\
    \gll mú-aná ól̩le \textbf{ka}{}-\textbf{á}{}-\textbf{log}{}-\textbf{a} \textbf{ka}{}-\textbf{a}{}-\textbf{ziw}{}-\textbf{íle} o-log-á\\
          1-child 1.\textsc{dem.dist} \textsc{neg.sbj1-pst.ipfv}{}-speak-\textsc{fv} \textsc{neg.sbj1-pst}{}-know-\textsc{pfv} \textsc{htd.15-}speak-\textsc{fv}\\
    \glt ‘That child did not speak. She was not able to speak.’ [Mute.Girl.Story]
\z 

\ea\label{ex:cuw:2} koʔíkálagá koɖó, kamugaabélél̩̩le \\ 
    \gll ka-o-hí-kal-ag-á koddo \textbf{ka}{}-\textbf{mu}{}-\textbf{gaa}{}-\textbf{abélel}{}-\textbf{ile}\\
          \textsc{cntf-15-neg}{}-be-\textsc{hab-fv} \textsc{htd.}9a.war \textsc{neg-sbj.2pl-hyp}{}-swim-\textsc{pfv}\\
    \glt ‘If it was not on account of war, you would not (be able to) swim.’ [Elicitation]
\z  

All declarative TA categories are listed in \tabref{tab:cuw:2}, both in their affirmative and negative shapes. Whenever possible, a cross-referenced example is indicated in parentheses. 

\begin{table} %HC @LSP I used shortstacks, I'm sorry if it's not the right solution:/
\begin{tabularx}{\textwidth}{XXX}
\lsptoprule
{label}    & {Affirmative}    & {Negative}\\
\midrule 
\multirow{2}{*}{\shortstack[l]{{progressive present} \\ {‘we are Ving’}}}   & \multirow{2}{*}{\textsc{sbj}-\textit{ni-o}-\textsc{vb}-\textit{a} \textsc{[dj]} \REF{ex:cuw:4b}}  & \multirow{2}{*}{\textit{{\textbf{ka}}}{}-\textsc{sbj}-\textit{ni-o}-\textsc{vb}-\textit{a} \REF{ex:cuw:56b}} \\ 
&&\\
\midrule 
\multirow{2}{*}{\shortstack[l]{{(hab.) present}  \\  {‘we V’}}} & \multirow{2}{*}{{\textsc{sbj}-\textit{ni}-\textsc{vb}-\textit{a} \textsc{[cj]} \REF{ex:cuw:3}}} & \multirow{2}{*}{{{\textit{\textbf{ka}}{}-\textsc{sbj}-\textit{ni}-\textsc{vb}-\textit{a}}} \REF{ex:cuw:38a}}\\
&&\\
\midrule
\multirow{2}{*}{\shortstack[l]{{perfective/perfect}  \\ {‘we (have) Ved’}}}  & {\textsc{sbj}-\textit{hi}-\textsc{vb}-\textit{a} \textsc{[dj]} \REF{ex:cuw:4a}}    & \multirow{2}{*}{{{\textit{\textbf{ka}}{}-\textsc{sbj}-\textsc{vb}-\textit{ile}}} \REF{ex:cuw:23b}} \\ 
  & {\textsc{sbj}-\textsc{vb}-\textit{ile} \textsc{[cj]}}    & \\
\midrule
\multirow{2}{*}{\shortstack[l]{{past perfective}  \\ {‘we (had) Ved’}}}  & {\textsc{sbj}-\textit{á}-\textsc{vb}-\textit{a} \textsc{[dj]}}    & \multirow{2}{*}{{\textit{\textbf{ka}}{}-\textsc{sbj}-\textit{a}-\textsc{vb}-\textit{íle} \REF{ex:cuw:1}}} \\ 
   & {\textsc{sbj}-\textit{a}-\textsc{vb}-\textit{ile} \textsc{[cj]}}  &  \\
\midrule
\multirow{2}{*}{\shortstack[l]{{past imperfective}  \\ {‘we Ved/were Ving’}}} & {\textsc{sbj}-\textit{a-ni}-\textsc{vb}-\textit{a} \textsc{[dj]}} & \multirow{2}{*}{\textit{\textbf{ka}}{}-\textsc{sbj}-\textit{á}-\textsc{vb}-\textit{a} \REF{ex:cuw:1}} \\ 
   & {\textsc{sbj}-\textit{á}-\textsc{vb}-\textit{a} \textsc{[cj]}}    & \\
\midrule
\multirow{2}{*}{\shortstack[l]{{future}  \\ {‘we will V’}}} & {\textsc{sbj}-\textit{naa-ilá-o}-\textsc{vb}-\textit{a} \textsc{[dj]}} & \multirow{2}{*}{{\textit{\textbf{ka}}-\textsc{sbj}-\textit{náa}-\textsc{vb}-\textit{e}} \REF{ex:cuw:40}} \\ 
  & {\textsc{sbj}-\textit{náa}-\textsc{vb}-\textit{e} \textsc{[cj]}}  & \\
\midrule
\multirow{2}{*}{\shortstack[l]{{future imperfective}  \\ {‘we will be Ving’}}} & {\textsc{sbj}-\textit{gá-ni}-\textsc{vb}-\textit{a} \textsc{[dj]}}   & \multirow{2}{*}{{\textit{\textbf{ka}}{}-\textsc{sbj}-\textit{gá}-\textsc{vb}-\textit{a}}} \\ 
  & {\textsc{sbj}-\textit{gá}-\textsc{vb}-\textit{a} \textsc{[cj]}}   & \\ 
\midrule
\multirow{2}{*}{\shortstack[l]{{hypothetical}  \\ {‘we would have Ved’}}}    & {\textsc{sbj}-\textit{gaa-hí}-\textsc{vb}-\textit{a} \textsc{[dj]}}    & \multirow{2}{*}{\textit{\textbf{ka}}{}-\textsc{sbj}-\textit{gaa}-\textsc{vb}-\textit{íle} \REF{ex:cuw:2}} \\ 
  & {\textsc{sbj}-\textit{gaa}-\textsc{vb}-\textit{ile} \textsc{[cj]}}& \\
\midrule
\multirow{2}{*}{\shortstack[l]{{past progressive}  \\ {‘we were Ving’}}} & \multirow{2}{*}{{\textsc{sbj}-\textit{á}-\textsc{aux} \textit{o}-\textsc{vb}-\textit{a}}}  & \multirow{2}{*}{{\textit{\textbf{ka}}{}-\textsc{sbj}-\textit{á}-\textsc{aux} \textit{o}-\textsc{vb}-\textit{a} \REF{ex:cuw:5}}}\\
 & & \\
\midrule
\multirow{2}{*}{\shortstack[l]{{counter-expectational}  \\ {‘we did V’}}}    & \multirow{2}{*}{{\textsc{sbj}-\textsc{aux} \textit{o}-\textsc{vb}-\textit{a} \REF{ex:cuw:6}}}    & \multirow{2}{*}{\textit{\textbf{ka}}{}-\textsc{sbj}-\textsc{aux} \textit{o}-\textsc{vb}-\textit{a} \REF{ex:cuw:6}}\\
&&\\
\midrule
{‘not yet’}  &   & {\textit{\textbf{ka}}{}-\textsc{sbj}-\textit{ná}{}-\textsc{vb}-\textit{a} \REF{ex:cuw:7}}\\
\lspbottomrule
\end{tabularx}
\caption{Declarative TA categories}
\label{tab:cuw:2}
\end{table} 

A common typological criterion when dealing with verbal negation is symmetry between affirmative and negative constructions.
According to \citet[556]{Miestamo2007}, “symmetric negative constructions do not differ from non-negatives in any other way than by the presence of the negative marker(s)”. This is partly the case in Cuwabo. Looking more closely at the affirmative column in \tabref{tab:cuw:2}, most declarative TA categories are subdivided into a pair of complementary conjugations, known since \citet{Meeussen1959} as “conjoint” (CJ) and “disjoint” (DJ).\footnote{Such a distinction is observed mostly in Eastern and Southern Bantu (see van der \citealt{WalHyman2017}).} This CJ/DJ alternation is conditioned by information structure. CJ verb forms are used to introduce a focused element, which necessarily occupies the Immediate-After-Verb position and, depending on its grammatical category, loses its (first) underlying High tone, a process glossed as HTD for High Tone Deletion (see \sectref{sec:cuw:4.3} for more details). This is shown in \REF{ex:cuw:3} with the noun \textit{mulobwana} ‘man’ (underlyingly \textit{múlobwana}), which loses its underlying H after the conjoint verb \textit{ddiḿ̩fúná} (see \REF{ex:cuw:57a} for another example). In disjoint verb constructions, no special relationship between the verb and the following constituent applies \REF{ex:cuw:4a}. This means that, unlike conjoint forms, a disjoint verb can stand on its own, as in \REF{ex:cuw:4b}.

\ea\label{ex:cuw:3} supeéyó supeéyo, míyó ɖiḿ̩fúná mulobwana\\
    \gll  supeéyo supeéyo míyo ddi-\textbf{ní}{}-fun-a mu-lobwana\\
           \textsc{htd}.9a.mirror \textsc{htd}.9a.mirror \textsc{pro1sg} \textsc{sbj.1sg}{}-\textbf{\textsc{ipfv}}\textbf{.\textsc{cj}}{}-want-\textsc{fv} \textsc{htd.1}{}-man\\
    \glt  ‘Mirror, mirror, I want a man.’ [Maria.Story]
\z 
           

\ea\label{ex:cuw:4}
    \ea\label{ex:cuw:4a}múlóbwan’ oókosáː masárápiʈw’ aábále  \\ 
        \gll mú-lobwana o-\textbf{hí}{}-kos-á ma-sárapitto ábale\\
               1-man \textsc{sbj1}{}-\textbf{\textsc{pfv}}\textbf{.\textsc{dj}}{}-do-\textsc{fv} 6-magic 6.\textsc{dem.dist}\\
        \glt  ‘The man made that magic.’ [Mbílri.Story]
    \ex\label{ex:cuw:4b}ánófúlá \\
        \gll á-\textbf{ni}{}-ó-ful-á \\
              \textsc{sbj2}{}-\textbf{\textsc{ipfv.dj}}\textsc{{}-15}{}-wash-\textsc{fv}\\
        \glt ‘They are washing.’ [Ddoolrinddo.Story]
    \z
\z 
              

Negative TAs in Cuwabo do not have the CJ/DJ alternation. Instead, only one form, based on the conjoint morphology and marked by the pre-initial negative marker \textit{ka}{}-, is used to obtain negative polarity. This means that both CJ/DJ affirmative forms are negated by means of a single negative form. As a result, constructional negation symmetry only applies with conjoint affirmatives: negative verb forms are formally equivalent to the affirmative conjoint verb forms.\footnote{If the symmetry invariably holds on the segmental level, in certain TAMs, tonal differences between affirmatives and negatives occur on the supra-segmental level. See \citet{KisseberthGuérois2014} for more details on melodic High tones in Cuwabo verbs.} The disjoint morphology is lost when negation takes place. The only exception is the progressive present form \textit{ka}-\textsc{sbj}-\textit{ni-o}-\textsc{vb}-\textit{a} ‘we are not Ving’, which functions as the negative counterpart of the disjoint form \textsc{sbj}-\textit{ni-o}-\textsc{vb}-\textit{a} ‘we are Ving’ (first line in \tabref{tab:cuw:2}). This is most likely due to the aspectual difference between the disjoint present and conjoint present in Cuwabo: whereas the disjoint form is progressive, the conjoint form \textsc{sbj}-\textit{ni}-\textsc{vb}-\textit{a} is used to refer to a general or even habitual present. In this sense, the CJ/DJ alternation for present is further specified with aspectual distinctions as compared with the six other TAs displaying the alternation.

The absence of CJ/DJ alternation in negatives underlies a paradigmatic asymmetry between affirmative and negative declarative TA categories. More precisely, this case of syncretism where two (or more) affirmative paradigms share a common negative counterpart is referred to by \citet{Miestamo2005} as “paradigmatic A/Cat asymmetry”. 
\largerpage
The last three declarative TA categories in \tabref{tab:cuw:2} do not have the CJ/DJ alternation. Both past progressive \REF{ex:cuw:5} and counter-expectational\footnote{Counter-expectational, commonly attested in narrative texts, is used to express unexpected or unplanned events, in both affirmative and negative polarities. It is aspectually close to the perfective, since the event is seen as a whole, but unlike the perfective, the counter-expectational involves a surprising effect.} \REF{ex:cuw:6}, built upon the auxiliary \textit{ila} (grammaticalized from \textit{ila} ‘say/do’), show symmetry with their affirmative equivalent.

\ea\label{ex:cuw:5} íyééné kaál’ uuʈolon’ úkó kéél’ oofulá?  \\
    \gll íyeéne ka-á-li o-ttolo=ni óko \textbf{ka}{}-\textbf{á}{}-\textbf{ila} \textbf{o}{}-\textbf{ful}{}-\textbf{á}\\
          \textsc{pro3sg} \textsc{neg.sbj1}-\textsc{pst.ipfv}{}-be \textsc{htd.17}{}-well=\textsc{loc} 17.\textsc{dem.ref} \textsc{neg.sbj1}-\textsc{pst.ipfv-aux} \textsc{htd.15-}wash-\textsc{fv}\\
    \glt ‘Wasn’t she there at the well, wasn’t she washing?’ [Ddoolrinddo.Story]
\z 

\ea\label{ex:cuw:6}
    \begin{xlist}[A: ]
    \exi{A: }\label{ex:cuw:6A} múkwáníya óyévíʔáwa ðaáwo? \\
        \gll  mú-kwaní=ya ó-yev-íh-a=wa dhaáwo\\ 
              3-greenery=\textsc{def} 15-be.small-\textsc{caus-fv}=\textsc{def15} like.that.\textsc{ii} \\ 
        \glt ‘(Why) so little greenery?’
    \exi{B: }\label{ex:cuw:6B} kaɖiíl’ óogula, ɖil’ óóváʔíwa\\
        \gll \textbf{ka}{}-\textbf{ddi}{}-\textbf{íla} \textbf{o}{}-\textbf{gul}{}-\textbf{á} ddi{}-lá ó{}-vah{}-íw{}-a\\ 
            \textsc{neg-sbj.1sg-cnte} \textsc{htd.15-}buy-\textsc{fv} \textsc{sbj.1sg-cnte} 15-give-\textsc{pass-fv} \\
        \glt ‘I did not buy it, I was given it.’ [Elicitation]
    \end{xlist}
\z 

As for ‘not yet’, which combines \textit{ka}{}- with the counter-expectational prefix \textit{ná}{}- as in \REF{ex:cuw:7}, it has no affirmative equivalent. There is no dedicated morpheme to express the concept ‘already’ in the language, although perfective verb forms may convey such meaning \citep{Guérois2021}. Notably, the counter-expectational prefix \textit{ná}{}- is also attested with the post-initial negator \textit{hi}{}- in non-main clauses to express ‘before Ving’. Such polysemy is common for \textit{nondum} constructions \citep{VeselinovaEtAlFC}. This is further discussed in \sectref{sec:cuw:2.4.2}. 

{\ea\label{ex:cuw:7} om̩mála wóóda. só kanákwa  \\
    \gll o-hi-mál-a ó-odd-á só \textbf{ka}{}-\textbf{ná}{}-\textbf{kw}{}-\textbf{a}\\
          \textsc{sbj1}-\textsc{pfv.dj}{}-finish-\textsc{fv} \textsc{15-}be.thin-\textsc{fv} but \textsc{neg.sbj1}-\textsc{cnte}{}-die-\textsc{fv}\\
    \glt ‘She has turned thin, but she had not died yet.’ [Ddoolrinddo.Story]
\z }

\subsection{Negation in non-declaratives}\label{sec:cuw:2.2} 
This section is concerned with negation in non-declaratives. Negation in interrogative sentences does not differ from negation in other clause types; see, for instance, examples \REF{ex:cuw:5}, \REF{ex:cuw:52}, \REF{ex:cuw:56b}, \REF{ex:cuw:66}. It is worth noting that question tags are expressed by means of the non-verbal negator \textit{kahíye} {\textasciitilde} \textit{kahíyo} discussed at length in \sectref{sec:cuw:3.3} below.
An example is given in \REF{ex:cuw:additional} with the variant \textit{kahíyo}. This pro-sentential use of negative existential is cross-linguistically very common (cf. \citealt{Veselinova2013neg.exist}).
% MM: here's the def of q-tags, just in case: i.e. interrogative fragments which seek to confirm the truth value of the main statement or assertion the speaker declares,
% LV: It is, however, worth noting that question
\ea\label{ex:cuw:additional} Aíssa ooʈúkúla m̩píka, kaáyo? \\ 
    \gll  Aíssa o-hi-ttúkul-a mu-píka \textbf{kahíyo}\\
          Aissa \textsc{sbj1-pfv.dj}{}-take\textsc{{}-fv} 3-bucket \textsc{nvb.neg}\\
    \glt ‘Aissa took the bucket, didn’t she?’ [Elicitation]
\z 


% \setlength{\parindent}{12pt}
The rest of this section is concerned with imperative and other modality-linked notions such as hortative, jussive, optative and desiderative.

% \setlength{\parindent}{15pt}
Commands in the second person singular or plural are expressed by means of the pre-initial imperative prefix \textit{ka}{}-, i.e. a grammaticalized form of the associated motion marker \textit{ka}{}- (\citealt{GuéroisEtAl2021}).\footnote{A second form of imperative, consisting of a bare stem (e.g. \textit{gulá!} ‘buy!’ < \textit{ógulá} ‘to buy’), also exists in Cuwabo, but has largely fallen into disuse.} This is illustrated in \REF{ex:cuw:8}. As in many Bantu languages \citep[192]{Nurse2008}, there is no formal negative counterpart to the imperative in Cuwabo. The only strategy to express negative commands or prohibitives is the use of the negative subjunctive. As in most Bantu languages, the subjunctive in Cuwabo is marked by the final vowel -\textit{e} \citep[261]{Nurse2008}. Remarkably, there are three different ways of negating subjunctives in Cuwabo, two of which are attested in non-declaratives (see \sectref{sec:cuw:2.4.1} for a discussion of the third negator). The first strategy involves the pre-initial negator \textit{ka}{}-, the same as used in standard negation. This is illustrated in \REF{ex:cuw:9}, where \textit{ka}{}- is merged with the second person singular subject prefix \textit{o}{}-, resulting in \textit{ku}{}-. The second strategy consists in prefixing the post-initial negator \textit{náa}{}-, as in \REF{ex:cuw:10} for second person plural or \REF{ex:cuw:12a} for second person singular. According to my main consultant, these two strategies are interchangeable. Research through my corpus seems to confirm their occurring in similar contexts.

\ea\label{ex:cuw:8} kaðówá vatákûlu!  \\
    \gll ka-dhów-a va-tákulu\\
          \textsc{imp}{}-go\textsc{{}-fv} 16-9a.home \\ 
    \glt ‘Go (sg) home!’ [Ddoolrinddo.Story]
\z

{\ea\label{ex:cuw:9} kulime munɖ’ áwó muzógwé aʔirúbwîle  \\
    \gll \textbf{ku}{}-lim-e mú-ndda ó-awo mu-zógwe a-hi-rúbw-ile\\
          \textsc{neg.sbj.2sg-}cultivate-\textsc{sbjv} 3-field \textsc{agr}3-\textsc{poss2sg} 1{}-rain \textsc{sbj1-neg-}rain-\textsc{pfv.seq}  \\
    \glt ‘Do (2\textsc{sg}) not farm your field if it has not rained.’ [Elicitation]
\z }

\ea\label{ex:cuw:10} kááʈiyeń̩na áányā. munééye!\footnote{Note in \REF{ex:cuw:10} that when an object prefix is added to an affirmative imperative verb form, as in \textit{kááʈiyeń̩na}, the final vowel \textit{{}-a} changes to \textit{{}-e}. This is a regular change in Eastern Bantu (see also \citealt{chapters/ruuli}).}\\
% \hc{@LSP: link to other chapter from NegBook (in footnote)} \\
    \gll ká-a-ttiy-e=ní=na á-anyá mu-\textbf{náa}{}-iy-e \\
          \textsc{imp-obj2}{}-leave\textsc{{}-fv}=\textsc{pl.addr=com} 2-owner \textsc{sbj.2pl-neg}{}-steal\textsc{{}-sbjv}\\
    \glt ‘Leave (2pl) it for/with the owners. Do not steal!’ [Elicitation]
\z 

The lack of paradigmatic symmetry between (positive) imperatives and prohibitives is commonly observed cross-linguistically \citep{Auwera2006}. 

Other non-declarative verbal clauses in Cuwabo are all built on the subjunctive morphology, in both affirmative and negative polarities. Apart from its recurrent use in complement clauses (see \sectref{sec:cuw:2.4.1}), the subjunctive found in non-declaratives may be used as a suggestion or mild imperative \REF{ex:cuw:11a}, an optative/desiderative \REF{ex:cuw:11b}, or a hortative/jussive \REF{ex:cuw:11c}. 
The negative equivalents are provided in \REF{ex:cuw:12}.

\ea\label{ex:cuw:11}
    \ea\label{ex:cuw:11a} muróme ókúpúʈ’ óotákúlúwa, mwiinávólowa mucélánímwa \\
        \gll  \textbf{mu}{}-\textbf{róm}{}-\textbf{e} ó-kupútt-a o-tákulúwa mu-hi-ná-volow-a mu-céla=ní=mwa\\
              \textsc{sbj.2}\textsc{pl}{}-start-\textsc{sbjv} 15-scrub-\textsc{fv} 17-9a.courtyard\textsc{} \textsc{sbj.2pl-neg-cnte}{}-enter-\textsc{fv} 18{}-well=\textsc{loc=def18}\\ 
        \glt ‘(You should) Start scrubbing the courtyard before getting into the well.’ [Elicitation] 
    \ex\label{ex:cuw:11b} ɖikáló w’ oomáríʔa  \\
        \gll ddi-kál-e o-a o-márih-a\\
              \textsc{sbj.1sg}{}-be-\textsc{sbjv} \textsc{agr1-conn} 15-finish-\textsc{fv}\\
        \glt ‘May I be the last one.’ [Mbílri.Story]
    \ex\label{ex:cuw:11c} n̩ðów’ óóʈólóni n̩ðówé náfûle  \\
        \gll ni-dhów-e o-ttólo=ni ni-dhów-e ni-á-ful-e\\
               \textsc{sbj.1}\textsc{pl}{}-go-\textsc{sbjv} 17-well=\textsc{loc} \textsc{sbj.1}\textsc{pl}{}-go-\textsc{sbjv} \textsc{sbj.1}\textsc{pl}{}-\textsc{amot}{}-wash-\textsc{sbjv}\\
        \glt  ‘Let’s go to the well, let’s go and wash (do the laundry).’ [Ddoolrinddo.Story]
    \z
\z 


\ea\label{ex:cuw:12}
    \ea\label{ex:cuw:12a} onáávégé wiíko, ookálá anyákôko  \\ 
        \gll  o-\textbf{náa}{}-veg-e o-íko o-hi-kál-a a-nyákoko\\
              \textsc{sbj.2sg-neg}{}-play-\textsc{sbjv} 17-river \textsc{sbj17}{}-\textsc{pfv.dj}{}-be-\textsc{fv} 2-crocodile\\
        \glt ‘You should not play at the river, there are crocodiles.’ [Elicitation]
    \ex\label{ex:cuw:12b} kaɖikale w’ oomáríʔa  \\
        \gll    \textbf{ka}{}-ddi-kal-e o-a o-márih-a \\
              \textsc{neg-sbj.1sg}{}-be-\textsc{sbjv} \textsc{agr}1-\textsc{conn} 15-finish-\textsc{fv}\\
        \glt ‘May I not be the last one.’ [Elicitation]
    \ex\label{ex:cuw:12c} kaniðow’ ooʈólóni  \\
        \gll   \textbf{ka}{}-ni-dhow-e o-ttólo=ni \\
              \textsc{neg-sbj.1}\textsc{pl}{}-go-\textsc{sbjv} 17-well=\textsc{loc} \\
        \glt  ‘Let’s not go to the well.’ [Elicitation]
    \z
\z 

\noindent The pre-initial \textit{ka}- is also used in non-declaratives to express incapacity, as in \REF{ex:cuw:13}.
\newpage

\ea\label{ex:cuw:13} óɖó n̩ládá kagaye!  \\ 
    \gll  óddo ni-láda \textbf{ka}{}-gay-e \\
           1.\textsc{dem.ref} 5-hand \textsc{neg.sbj1-}balance-\textsc{sbjv}\\
    \glt  ‘That Mr. Hand cannot balance itself!’ [Body.Story]
\z 

The morphological formulae of both affirmative and negative non-declarative mood categories are summarized in \tabref{tab:cuw:3}. 

\begin{table}
\begin{tabularx}{.6\textwidth}{XXl}
\lsptoprule
{label} & {affirmative} & {negative}\\
\midrule 
{imperative} & {\textit{ka}-\textsc{vb}{}-\textit{a}} & {✘} \\
{subjunctive} & {\textsc{sbj}{}-\textsc{vb}{}-\textit{e}} & {\textit{\textbf{ka}}{}-\textsc{sbj-vb}{}-\textit{e}} \\
& & {\textsc{sbj}{}-\textit{\textbf{náa-}}\textsc{vb}{}-\textit{e}}\\
\lspbottomrule
\end{tabularx}
\caption{Non-declarative mood categories}
\label{tab:cuw:3}
\end{table}

As explained above, negating the imperative involves subjunctive morphology. The lack of inherent negative construction makes imperative paradigmatically asymmetrical. In contrast, both negative subjunctives are symmetric to their affirmative equivalent on the segmental level: either the pre-initial negator \textit{ka}{}- or the post-initial \textit{náa-} is added. However, tone patterns differ on each of these constructions. Affirmative subjunctives have an underlying H on the first mora of the stem in case of bimoraic stems (as in \REF{ex:cuw:11}), or on the second mora with trimoraic stems. \textit{ka}{}- negative subjunctives are all low, whereas \textit{náa-} negative subjunctives are subjected to tone plateauing (if no further prosodic constraints apply, see \citealt{Guérois2019}: 739) starting from the negator (e.g. \textit{munáávégé} in \REF{ex:cuw:12a}).

In terms of distribution, \textit{ka}{}- is restricted to main clauses only, whereas \textit{náa-} may be used in both main clauses, as shown in this section, and complement clauses (see \sectref{sec:cuw:2.4.1}), but with a very limited distribution, as it is restricted to the subjunctive mood.

\subsection{Negation in stative predications}\label{sec:cuw:2.3} 
Cuwabo has two distinct non-verbal negators. The first, \textit{kahíye} (with \textit{kahíyo} as a free variant), is invariable and translated in English as ‘(there/it) is not’. Apart from the initial negative marker \textit{ka}{}-, the rest of the underlying morphological structure is difficult to identify synchronically. The segment \textit{hi}{}- might represent the perfect(ive) marker, commonly used in Cuwabo to express a result state. The origin of the final segment \textit{yo {\textasciitilde} ye} is unknown. Another puzzle is that no verb root is discernible. Since they are not morphologically transparent, these forms are glossed ‘\textsc{nvb.neg}’. As several examples in this section show, \textit{kahíye}{\textasciitilde}\textit{kahíyo} may also be produced on the surface as \textit{kaáye} and \textit{kaáyo}, respectively, whereby \textit{hí}{}- totally assimilates with the preceding low vowel. 

The second non-verbal negator derives from \textit{kahíye}{\textasciitilde}\textit{kahíyo}, but inflects for subject marking and also contains one of the three locative clitics. The resulting shape is as follows: \textit{ka}{}-\textsc{sbj}{}-\textit{hí}=\textsc{loc}. This negator is limited to the context of locative-existential predication, hence it is glossed ‘\textsc{negex}’. 

An alternative strategy to these existential negators is the copula verb \textit{li} ‘be’ used in its negative form.\footnote{In many Bantu languages, \textit{li} has a suppletive negative form \textit{si} ‘not to be’. This is not the case in Cuwabo.} As shown throughout this section, its use is broad and it often comes as a free alternative to one of the two existential negators. The copula verb \textit{li} ‘be’, reconstructed to Proto-Bantu as *di, is highly defective: it does not take the default final vowel and only occurs in (zero-marked) present and \textit{a}{}- past. 

% \setlength{\parindent}{11pt}
Affirmative non-verbal predication is in turn expressed in three different ways. The first, restricted to class 1/2 third-person subjects, involves the deletion of the augment tone on the predicate. For instance, in \textit{Maríyá muyaná} ‘Marie is a woman’, the citation form \textit{múyaná} (HØH) surfaces as \textit{muyaná} (ØØH). High-Tone Deletion triggers a predicative interpretation. The second strategy consists in using class-inflected copulas, such as class 1 \textit{ddi}, class 2 \textit{ba}, class 3 \textit{bu}, etc., grammaticalized from demonstrative pronouns (\citealt{Guérois2019}: 762; \citealt{GibsonEtAl2019}).\footnote{More specifically, the copula is formally identical to the proximate demonstrative stems (with the exception of the class 1 copula form \textit{ddi}).} Finally, the copula verb \textit{li} ‘be’ appears as the third recurrent strategy. Multiple examples for each strategy are provided along the section. The reader is also referred to \citet{GuéroisFC} for a detailed study of non-verbal predication in Cuwabo.

% \setlength{\parindent}{12pt}
This section explores negation in non-verbal predicates along their different subtypes, i.e. equational (\sectref{sec:cuw:2.3.1}), ascriptive (\sectref{sec:cuw:2.3.2}), locative-existen\-tial (\sectref{sec:cuw:2.3.4}) and possessive (\sectref{sec:cuw:2.3.5}).

\subsubsection{Equational predication}\label{sec:cuw:2.3.1}
Equational predicates establish a relation of identity between two referential NPs. This is commonly expressed in Cuwabo by means of the copula verb \textit{li} ‘be’, as seen in \REF{ex:cuw:14}. In \REF{ex:cuw:14A}, speaker A speaks in the first person singular and asserts she is Maria. Speaker B denies it.
Two negation strategies are observed: the negated copula verb \textit{kulí} (where the negative marker \textit{ka}{}- is merged with the second singular subject prefix \textit{o}{}-) is used in a context of contrastive negation, while the non-verbal negator \textit{kahíye}{\textasciitilde}\textit{kahíyo} (henceforth \textit{kahíye} for simplicity) is used in a non-focus context. It remains to be checked whether the selection of negation strategy is necessarily determined by information structure. Notably, \textit{kahíye} is located between the subject and the predicative complement, leading to constructions that can literally be interpreted as ‘X, it is not Y’.

\ea\label{ex:cuw:14}
     \begin{xlist}[A: ]
    \exi{A: }\label{ex:cuw:14A} míyó ɖili Maríya \\
         \gll  míyo \textbf{ddi}{}-\textbf{li} Maríya \\ 
                \textsc{pro1sg} \textsc{sbj.1sg-}be Maria \\ 
         \glt  ‘I am Maria.’
    \exi{B: }\label{ex:cuw:14B} wéyó kulí Máriya, kúzívale, Maríyá káyí’ wéyo \footnote{The segment \textit{hi} in \textit{kahíye} in example \REF{ex:cuw:14B} is most frequently heard with the palatal approximant/glide (here noted as /y/) rather than with the glottal stop /ʔ/.} \\ 
        \gll  wéyo \textbf{ku}{}-\textbf{lí} M. kú-zival-e M. \textbf{kahíye} wéyo \\
             \textsc{pro2sg} \textsc{neg}.\textsc{sbj.2sg}{}-be M. \textsc{neg}.\textsc{sbj.2sg}{}-lie-\textsc{sbjv} M. \textsc{nvb.neg} \textsc{pro2sg} \\
        \glt  ‘You are not Maria, do not lie, Maria (it) is not you.’ [Maria.Story]
     \end{xlist}
 \z 

% \setlength{\parindent}{10pt}
Affirmative equational predication is also expressed by means of class-inflected copulas, such as class 1 \textit{ddi} in \REF{ex:cuw:15}, or by the loss of the augment tone through High-Tone Deletion, as in \REF{ex:cuw:16} with \textit{rapaási} ‘is a boy’ (versus \textit{rápaási} in citation form). In this example, which constitutes a typical way to open stories in Cuwabo, the predicate \textit{rapaási} expresses a contrastive focus as opposed to the specific but indefinite referent \textit{óddo} ‘that one’ introduced (again) by the existential negator \textit{kaáye} (where \textit{kaáye} is a reduced form of \textit{kahíye}).

\ea\label{ex:cuw:15} oń̩lógá ɖi míyo Féŕnáânda\\ 
    \gll o-ní-log-á \textbf{ddi} míyo Férnaánda\\
         \textsc{agr1-ipfv}{}-speak\textsc{{}-fv.rel} \textsc{cop1} \textsc{pro1sg} Fernanda \\
    \glt ‘The one speaking is me Fernanda.’ [Ddingí.Story]
\z 

\ea\label{ex:cuw:16} weéŕ̩lé kááy’ ɖo, weéŕ̩lé rapaási…  \\
    \gll  o-ér-ilé \textbf{kaáye} óddo o-ér-ilé \textbf{rapaási} \\
          \textsc{agr1-}do/say-\textsc{pfv}.\textsc{rel} \textsc{nvb.neg} 1.\textsc{dem.ref} \textsc{agr1-}do/say-\textsc{pfv}.\textsc{rel} \textsc{htd}.1a.boy \\
    \glt ‘There once was a boy...’ (lit. ‘the one who did/said (it) is not that one, the one who did/said is a boy’) [Ddoolrinddo.Story]
\z 

\subsubsection{Ascriptive predication}\label{sec:cuw:2.3.2}
Predicates that ascribe a property to an individual or an entity are known as ‘ascriptive’ (\citealt{Hengeveld1990,Hengeveld1992}). This property can be a character trait, a physical or intrinsic property of an individual or an entity.\footnote{Result states, which ensue from a prior event giving rise to them, as well as certain physical properties (e.g. \textit{ókóddéla} ‘be beautiful’) are in turn expressed by perfect(ive)-inflected verbs (see \citealt{Guérois2015}: 363–365).} Adjective as a grammatical category is almost non-existent in Cuwabo, as typical for many Bantu languages. Only a handful of adjectival items are attested in the language. Instead, the semantic content of ascriptive predication is conveyed by an infinitive verb or by a noun in adnominal possessor function introduced by the genitive linker \textit{a} (aka “connective” in Bantu studies, cf. \citealt{VandeVelde2013}), itself preceded by a class-inflected copula.\footnote{Notably, in this construction, the connective lacks the expected agreement marker co-referring to the head noun, as in e.g. \REF{ex:cuw:18} with \textit{wa} referring to class 3 \textit{músolro} ‘head’. Such absence, which might be viewed as an instance of haplology, points to a grammaticalization process of this construction. A comparative analysis would be needed to reach some historical explanation.} This is illustrated in the  examples \REF{ex:cuw:17} to \REF{ex:cuw:19}.

\ea\label{ex:cuw:17} míyó ɖa mwaaná  \\ 
    \gll míyo \textbf{ddi}{}-\textbf{a} \textbf{mu}{}-\textbf{aná}  \\
          \textsc{pro1sg} \textsc{cop1-conn} \textsc{htd.1}{}-child \\
    \glt  ‘I am a child.’ [Elicitation]
\z 

\ea\label{ex:cuw:18} músóɽo wa mwááná óɖu b’ oolápa\\ 
    \gll  mú-solro o-a mú-aná óddu \textbf{bu}{}-\textbf{a} \textbf{o}{}-\textbf{láp}{}-\textbf{a}  \\
         3-head \textsc{agr3}{}-\textsc{conn} 1-child 1.\textsc{dem.prox} \textsc{cop3-conn} \textsc{15}{}-be.big-\textsc{fv} \\
    \glt ‘The head of this child is big.’ [Elicitation]
\z 

\ea\label{ex:cuw:19} bírí\'{ŋ}kw’ iísí sa wóóru  \\ 
    \gll bírińku ési \textbf{si}{}-\textbf{a} \textbf{ó}{}-\textbf{oru}  \\
           10a.earring 10.\textsc{dem.prox} \textsc{cop}10-\textsc{conn} 11-gold  \\ 
    \glt  ‘These earrings are (made) of gold.’ [Elicitation]
\z 

As already said above, class 1/2 third-person subjects do not need a copula item between the subject and the predicative complement. However, the latter is necessarily subjected to High-Tone Deletion, as seen in both \REF{ex:cuw:20} with the adjective \textit{munddimúwa}, and \REF{ex:cuw:21} with the noun \textit{namapuja}.
\newpage
\ea\label{ex:cuw:20} Mosambíki munɖimúwa  \\ 
    \gll Mosambíki \textbf{mu}{}-\textbf{nddimúwa}\\
        1a.Mozambique \textsc{htd.agr1}{}-big \\
    \glt ‘Mozambique is big.’ [Elicitation]
\z 


\ea\label{ex:cuw:21} namárógoló namapuja   \\
    \gll namárogolo \textbf{namapuja} \\
         1a.hare \textsc{htd}.1a.joker \\
    \glt ‘The hare is a joker.’ [Ddingí.Story]
\z 

Similarly to equational predicates, ascriptive predicates are negated by using the non-verbal negator \textit{kahíye} located between the subject and the predicative complement. This is shown in \REF{ex:cuw:23a}, \REF{ex:cuw:24}, \REF{ex:cuw:25} and \REF{ex:cuw:26}. These constructions show how the use of non-verbal negators has extended into predicative copular constructions. The presence of the copula-connective \textit{dda} in \REF{ex:cuw:22a}, leading to an existential-copular sequence, is not clear. Such a sequence, however, seems to be restricted to speech act participants. Again, the negated copula verb \textit{li} can be used as an alternative. This is shown in \REF{ex:cuw:22b}, \REF{ex:cuw:25b} and \REF{ex:cuw:26b}. Interestingly, my main consultant interprets \REF{ex:cuw:25a} and \REF{ex:cuw:25b} differently. More specifically, \REF{ex:cuw:25b} may be conceived as the denial of a previous statement asserting the country’s greatness. It could then easily be completed by ‘it is weak’. In contrast, the same consultant does not denote any difference in meaning in the pairs of sentences provided in \REF{ex:cuw:22} and \REF{ex:cuw:26}, and if the use of the copula verb \textit{li} in both \REF{ex:cuw:22b} and \REF{ex:cuw:26b} may originally be associated to contrastive negation, its synchronic use does not seem to be restricted to specific pragmatic contexts. Conversely, it may be that negative-inflected \textit{li} is specializing as a contrastive negator due to the pervasiveness of the non-verbal negator \textit{kahíye} in copular constructions. Another distinction between these paired sentences is the presence versus absence of the augment tone: HTD only occurs on the predicative complement that follows negative-inflected \textit{li}. Finally, when the modifier is a verb, the latter may also receive a result state reading through perfect(ive) inflection \REF{ex:cuw:23b}. Here too, there is no difference in meaning between the two constructions.

\ea\label{ex:cuw:22}
    \ea\label{ex:cuw:22a} míyó kayíyé ɖa mwááná\\
        \gll  míyo \textbf{kahíye} \textbf{ddi}{}-\textbf{a} mú-aná \\
              \textsc{pro1sg} \textsc{nvb.neg} \textsc{cop1-conn} 1{}-child\\
        \glt ‘I am not a child.’
\newpage

    \ex\label{ex:cuw:22b} míyó kaɖilí mwaaná \\
        \gll  míyo \textbf{ka}{}-\textbf{ddi}{}-\textbf{lí} mu-aná \\
              \textsc{pro1sg} \textsc{neg-}\textsc{sbj.1sg-}be \textsc{htd.1}{}-child\\
        \glt ‘I am not a child.’ [Elicitation]
    \z
\z 

\ea\label{ex:cuw:23}
    \ea\label{ex:cuw:23a} músóɽo wa mwááná óɖu kayiy’ oolápa\\ 
        \gll mú-solro o-a mú-aná óddu \textbf{kahiye} o-láp-a \\
              3-head \textsc{agr}3-\textsc{conn} 1-child 1.\textsc{dem.prox} \textsc{nvb.neg} 15{}-be.big-\textsc{fv}\\
        \glt  ‘The head of this child (it) is not big.’ [Elicitation]
    \ex\label{ex:cuw:23b} músóɽo wa mwááná óɖu kulápîle \\
        \gll  mú-solro o-a mú-aná óddu \textbf{ku}-\textbf{láp}-\textbf{ile}\\
              3-head \textsc{agr}3-\textsc{conn} 1-child 1.\textsc{dem.prox} \textsc{neg.sbj3}{}-be.big-\textsc{prf}\\
        \glt  ‘The head of this child is not big.’ [Elicitation]
    \z
\z 

{\ea\label{ex:cuw:24} bírí\'{ŋ}kw’ iísí ðiɖigúlél̩le wóóː, kaa’ ða wóóru? \\
    \gll  bírińku ési dhi-ddi-gúl-el-ile wéyo \textbf{kaaye} dhi-a ó-oru\\
          10a.earring 10.\textsc{dem.prox} \textsc{agr}10-\textsc{obj}1\textsc{sg}{}-buy-\textsc{appl}{}-\textsc{pfv}.\textsc{rel} \textsc{pro2sg} \textsc{nvb.neg} \textsc{agr}10-\textsc{conn} 11-gold \\
    \glt ‘These earrings you bought me, aren’t they made of gold?’ [Mbílri.Story]
\z }

\ea\label{ex:cuw:25}
    \ea\label{ex:cuw:25a} Póŕtúgálí kayíyé múnɖímúwa\\
        \gll  Pórtugáli \textbf{kahíye} mú-nddimúwa\\
              1a.Portugal \textsc{nvb.neg} \textsc{agr1}-big\\
        \glt ‘Portugal (it) is not big.’
    \ex\label{ex:cuw:25b} Póŕtúgálí kalí munɖimúwa \\
        \gll Pórtugáli \textbf{ka}{}-\textbf{lí} mu-nddimúwa \\
             1a.Portugal \textsc{neg.sbj1-}be \textsc{htd.agr1}{}-big \\
        \glt ‘Portugal is not big.’ (Interpreted as ‘Portugal is no longer respected/powerful.’) [Elicitation]
    \z
\z 

\ea\label{ex:cuw:26}
    \ea\label{ex:cuw:26a} namárógoló kayíye namápûja\\ 
        \gll namárogolo \textbf{kahíye} namápuja \\
              1a.hare \textsc{nvb.neg} 1a.joker\textsc{} \\
        \glt ‘The hare (it) is not a joker.’ [Elicitation]
    \ex\label{ex:cuw:26b} namárógoló kalí namapuja \\
        \gll namárogolo \textbf{ka}{}-\textbf{lí} namapuja \\
              1a.hare \textsc{neg.sbj1}{}-be \textsc{htd}.1a.joker\\
        \glt ‘The hare is not a joker.’ [Elicitation]
    \z
\z 

\subsubsection{Locative-existential predication}\label{sec:cuw:2.3.4}\label{sec:cuw:2.3.3}
The negative locative-existential \textit{ka}{}-\textsc{sbj}{}-\textit{hí}=\textsc{loc} is formally very similar to \textit{kahíye}: both constructions have the negative marker \textit{ka}{}- and the (seemingly) perfect(ive) prefix \textit{hi}{}-. However, the former is more transparent than the latter in that it clearly inflects for subject marking and also contains one of the three locative clitics (class 16 =\textit{vo}, class 17 =\textit{wo}, class 18 =\textit{mo}). The full paradigms are provided in \tabref{tab:cuw:4}. 

\begin{table}
\begin{tabularx}{\textwidth}{lXXX}
\lsptoprule
Person/class & \textit{ka}-\textsc{sbj}-\textit{hí}=\textsc{loc16} & \textit{ka}-\textsc{sbj}-\textit{hí}=\textsc{loc17} & \textit{ka}-\textsc{sbj}-\textit{hí}=\textsc{loc18} \\
 & ‘is not here’ & ‘is not there’ & ‘is not inside’ \\
\midrule 
\scshape 1sg & \textit{kaddihívo} & \textit{kaddihíwo} & \textit{kaddihímo}  \\
\scshape 2sg & \textit{kawihívo} & \textit{kawihíwo} & \textit{kawihímo} \\
\scshape 1pl  & \textit{kanihívo} & \textit{kanihíwo} & \textit{kanihímo} \\
\scshape 2pl  & \textit{kamwiívo} & \textit{kamwihíwo} & \textit{kamwihímo} \\
1 & \textit{kaávo} & \textit{kaáwo} & \textit{kaámo} \\
2 & \textit{kaávo} & \textit{kaáwo} & \textit{kaámo} \\
3 & \textit{kawihívo} & \textit{kawihíwo} & \textit{kawihímo} \\
4 & \textit{kadhihívo} & \textit{kadhihíwo} & \textit{kadhihímo} \\
5 & \textit{kanihívo} & \textit{kanihíwo} & \textit{kanihímo} \\
6 & \textit{kaávo} & \textit{kaáwo} & \textit{kaámo} \\
9 & \textit{kayihívo} & \textit{kayihíwo} & \textit{kayihímo} \\
10 & \textit{kadhihívo} & \textit{kadhihíwo} & \textit{kadhihímo} \\
11, 14 & \textit{kawihívo} & \textit{kawihíwo} & \textit{kawihímo} \\ 
16 & \textit{kavihívo} &  &  \\
17 &  & \textit{kawihíwo} &  \\
18 &  &  & \textit{kamwihímo} \\
\lspbottomrule
\end{tabularx}
\caption{Inflected negative locative-existentials}
\label{tab:cuw:4}
\end{table}

This locative-existential negator is used for both locative predication ‘is not here’ and existential predication ‘there is no’. In both interpretations, the spatial reference is contained in the locative enclitic. In pragmatically-neutral contexts, the subject-referent to which the location is assigned is preverbal, as in \REF{ex:cuw:28}. However, in the case of a sentence-focus construction, i.e. when both the predicate and the subject are new information, VS order applies, as in \REF{ex:cuw:29}. In such thetic sentences, the postverbal element is subjected to HTD, be it the subject NP \REF{ex:cuw:29a} or---more unexpectedly---the (headless) subject relative \REF{ex:cuw:29b}. 

{\ea\label{ex:cuw:28}
    \ea\label{ex:cuw:28a} naámbéɖ’ oónójúwáávi, ma namárógolo kaawó!\\
        \gll  naámbedde ó-ni-o-j-uw-á=vi mas namárogolo \textbf{kaá}=\textbf{wo} \\
             1a.maize \textsc{sbj1}-\textsc{ipfv.dj-15}{}-eat-\textsc{pass}{}-\textsc{fv=restr} but 1a.hare \textsc{negex.sbj1=loc17}  \\
        \glt ‘Maize keeps being eaten, but the hare is not here!’ [Ddingí.Story]

    \ex\label{ex:cuw:28b} mwááðíye kaávo, ánááye kaávo, ðoóbó ðaye kaðiyívo\\
        \gll  mú-adhí=ye \textbf{kaá}=\textbf{vo} á-aná=ye \textbf{kaá}=\textbf{vo} dhi-óbo dhi-aye \textbf{kadhihí}=\textbf{vo}  \\
             1-wife=\textsc{poss3sg} \textsc{negex.sbj1}=\textsc{loc16} 2-child=\textsc{poss3sg} \textsc{negex.sbj2}=\textsc{loc16} 10-furniture \textsc{agr}10-\textsc{poss3sg} \textsc{negex.sbj1}0=\textsc{loc16}  \\
        \glt ‘His wife is not here, his children are not here, his furniture is not here.’ [Maria.Story]
    \z
\z }

{\ea\label{ex:cuw:29}
    \ea\label{ex:cuw:29a} [elábw’ ééjí y’ ooténe] kaávó m̩bava w’ oopíʈa míyo. \\ 
        \gll  \textbf{kaá}=\textbf{vo} mu-bava o-a o-pítt-a míyo \\
              \textsc{negex.sbj1}=\textsc{loc16} \textsc{htd.1}{}-thief \textsc{agr}1-\textsc{conn} \textsc{15}{}-surpass\textsc{{}-fv} \textsc{pro1sg} \\
        \glt ‘[In this whole place], there is no thief better than me.’ [Maria]
    \ex\label{ex:cuw:29b} kaáwó oðileewóː \\
        \gll  \textbf{kaá}=\textbf{wo} o-dh-ile=wo\\
                \textsc{negex.sbj1}=\textsc{loc17} \textsc{agr}1-come-\textsc{pfv.rel.htd}=\textsc{loc17}\\
        \glt ‘Nobody came.’ (lit. ‘There is not who came there.’ / ‘He is not here the one who came here.’) [Elicitation] 
    \z
\z }

Again, the copula verb \textit{li} ‘be’ may also serve to express locative-existential predication. In this case, it always appears in thetic VS constructions, but unlike inflected negative locative-existentials as in \REF{ex:cuw:29}, the postverbal subject does not control verb agreement. Instead, class 16 subject prefix \textit{va}{}- appears as default agreement. Three examples are provided in \REF{ex:cuw:30}. Again, the postverbal subject is subjected to HTD. 

{
\ea\label{ex:cuw:30}
    \ea\label{ex:cuw:30a} ńněː, ápa kávalí mwáʔa \\
        \gll  ńne ápa \textbf{ká}{}-\textbf{va}{}-\textbf{lí} mwaha\\
             no 16.\textsc{dem.prox} \textsc{neg-sbj1}6-be \textsc{htd.3}.reason \\
        \glt ‘No, here there is no problem.’ [Maria]
\newpage
    \ex\label{ex:cuw:30b} kavalí vogóː van̩rórómeleya asúpâyi\\
        \gll  \textbf{ka}{}-\textbf{va}{}-\textbf{lí} vogo va-ni-róromel-ey-a a-súpayi\\
            \textsc{neg-sbj1}6-be \textsc{htd}.16.somewhere \textsc{agr}16-\textsc{ipfv}{}-believe-\textsc{n-fv} 2-policeman \\
        \glt ‘Nowhere are the policemen trustworthy.’ [Elicitation]

    \ex\label{ex:cuw:30c} ápató kaval’ íísavi\\
          \gll ápa=to \textbf{ka}{}-\textbf{va}{}-\textbf{lí} e-savi \\
            16.\textsc{dem.prox}=then \textsc{neg-sbj1}6-be \textsc{htd.9}{}-relish  \\
        \glt ‘Here then, there is no relish.’ [Elicitation]
    \z
\z }

Affirmative equivalents to locative-existential predication are based on the copula verb \textit{li} ‘be’, for both locative \REF{ex:cuw:31a} and existential predication \REF{ex:cuw:31b}. Another common strategy is to use the semi-copula verb \textit{kála} ‘be, live’, especially as part of thetic sentences \REF{ex:cuw:32}.

\ea\label{ex:cuw:31}
    \ea\label{ex:cuw:31a} mwámúna Maríya, ol’ úumabásâni \\ 
        \gll mwámúna Maríya \textbf{o}{}-\textbf{lí} o-ma-bása=ni \\
               1.husband Maria \textsc{sbj1-}be 17-6-work=\textsc{loc} \\
        \glt  ‘Maria’s husband is at work.’ [Maria.Story]

    \ex\label{ex:cuw:31b} [weeŕ̩légé keeŕ̩lége ðaweene,] waáli Mádáː \'{E}rúgúlúː na Manyálo. \\ 
        \gll \textbf{o}{}-\textbf{á}{}-\textbf{li} má-da é-rugulu na ma-nyálo\\
              \textsc{sbj1-}\textsc{pst}{}-be 6-hand 9-belly and 6-foot  \\
        \glt  ‘[Once upon a time,] there were Mr. Hands, Mr. Belly and Mr. Feet.’ [Body.Story]
    \z
\z

\ea\label{ex:cuw:32} oʈólón’ uúkúl’ ookálá fúlóór’ iinííbá\\ 
    \gll  {o-ttólo=ni}           {ókule} \textbf{o}{}-\textbf{hi}{}-\textbf{kála}        {fúloóri}   {e-ní-ib-á}\\
          {17-well=\textsc{loc}}  {17.\textsc{dem.dist}}    {\textsc{sbj1}7-\textsc{pfv}.\textsc{dj}{}-be} {9a.flower} {\textsc{agr}9-\textsc{ipfv-}sing-\textsc{fv.rel}}\\
    \glt ‘There at the well there is a flower which sings.’ [Ddoolrinddo.Story]
\z 

\subsubsection{Possessive predication}\label{sec:cuw:2.3.5}
Two ‘have’ verb stems exist in Cuwabo: \textit{káana} and \textit{na}. Etymologically, the first, \textit{káana}, can be analysed as the verb stem \textit{kála} ‘be, live’ + the comitative preposition \textit{na} ‘with’. This combination was reanalyzed as a full-fledged verb, lending itself to normal TAM marking around the root \textit{kaan}. The second, \textit{na} ‘have’, also involves the reanalysis of the comitative preposition \textit{na} but in association with a subject index and a deleted copula. Albeit more restricted than \textit{káana} in terms of inflectional marking, \textit{na} may still host the pre-final habitual suffix \nobreakdash-\textit{ag}, which supports its reanalysis as an irregular verb base.

\ea\label{ex:cuw:33}
    \ea\label{ex:cuw:33a} míyô maásíkiní ɖíkaánê dálá ðaaví, ójâ múkucéː? n̩nééː!\\
        \gll  míyo maásikiní \textbf{ddí}{}-\textbf{kaán}{}-\textbf{e} dála dhaavi ó-j-a mú-kucé n̩ne  \\
            \textsc{pro1sg} even.if \textsc{sbj1}-have-\textsc{sbjv} 9a.hunger how \textsc{15}{}-eat-\textsc{fv} 3-leftover no \\
        \glt ‘I am very hungry, but to eat the leftovers? No way!’ [Elicitation]

    \ex\label{ex:cuw:33b} kaázîwá ɖí ḿ̩báaláwa óĺ̩lé waánága péle\\
        \gll  ka-á-ziw-á ddi mú-baali-áwa ól̩le \textbf{o}{}-\textbf{á}{}-\textbf{naga} péle \\
              \textsc{neg.sbj2}-\textsc{pst}.\textsc{ipfv}{}-know-\textsc{fv} \textsc{cop1} 1-sister-\textsc{poss3pl} 1.\textsc{dem.dist} \textsc{agr}1-\textsc{pst.ipfv}{}-have.\textsc{hab.rel} 1a.scabies \\
        \glt ‘They did not know she was their sister, the one who had scabies.’ [Maria.Story] 
    \z
\z 

‘Not have’ simply requires the addition of a verbal negator, either \textit{ka}{}- or \textit{hi}{}- depending on the type of clause. The verb \textit{na} seems to be used by default \REF{ex:cuw:34b}. The only negative form of \textit{káana} available in my database is provided in \REF{ex:cuw:34a}. 

{\ea\label{ex:cuw:34}
    \ea\label{ex:cuw:34a} muʔinákáan’ éébaribarí munááváʔe koóbíɽiða\\ 
        \gll   \textbf{mu}{}-\textbf{hi}{}-\textbf{ná}{}-\textbf{káaná} e-baribari mu-náa-vah-e koóbilri=dha  \\
              \textsc{sbj.2pl-neg-cnte}{}-have \textsc{htd.9}{}-truth \textsc{sbj.2pl-neg}{}-give-\textsc{sbjv} 10a.money=10.\textsc{def}  \\
        \glt  ‘As long as you are not sure, do not give the money.’ [Elicitation]
    \ex\label{ex:cuw:34b}   íyééne kaáná maare mééna oókósā\\ 
        \gll íyeéne \textbf{ka}{}-\textbf{á}{}-\textbf{na} ma-are má-ina a-a ó-kos-á \\ 
              \textsc{pro3sg} \textsc{neg.sbj1}-\textsc{pst.ipfv}{}-have \textsc{htd.6}{}-idea 6-other \textsc{agr}6-\textsc{conn} 15-do-\textsc{fv}  \\
        \glt  ‘She had nothing else to do.’ [Maria.Story]
    \z
\z}

\subsection{Negation in non-main clauses}\label{sec:cuw:2.4}
In this section, negation is discussed in three types of non-main clauses: clausal complementation (\sectref{sec:cuw:2.4.1}), temporal and condition adverbial clauses (\sectref{sec:cuw:2.4.2}), and relative clauses (\sectref{sec:cuw:2.4.3})

\subsubsection{Clausal complementation}\label{sec:cuw:2.4.1}
Complement clauses are of three types in Cuwabo: they can be subjunctive, indicative or infinitive. Subjunctive and indicative complements are typically introduced by the complementizer \textit{wíilá} ‘that’ (often shortened to \textit{wii} or \textit{wi} in oral speech), which constitutes a grammaticalized form derived from the verb \textit{wíilá} ‘say, do’. Its main function is to introduce a subordinate clause. 

Subjunctive complements are typically used as purpose clauses. When negated, one of the two post-initial negators is required, either \textit{hi}{}- \REF{ex:cuw:35} or \textit{náa}\textbf{\textit{{}-}} \REF{ex:cuw:36}. Recall that \textit{náa}\textbf{\textit{{}-}} is restricted to the subjunctive mood and is also used in non-declaratives (see \sectref{sec:cuw:2.2}). The absence of the complementizer in \REF{ex:cuw:36b} shows that it is not mandatory.

{\ea\label{ex:cuw:35} ooɖíkúkusela owánínyu wíí ɖiʔikwentéː?!\\
    \gll o-hi-ddí-kukus-el-a owáninyu wíi \textbf{ddi}{}-\textbf{hi}{}-\textbf{kwent}{}-\textbf{e} \\
         \textsc{sbj.2sg-pfv-obj.1sg}{}-take\textsc{{}-appl-fv} 17.at.home.\textsc{poss2pl} \textsc{comp} \textsc{sbj.1sg-neg}{}-copulate-\textsc{sbjv} \\
    \glt ‘You took me to your house to not have sex?!’ [Elicitation]
\z }

\ea\label{ex:cuw:36}
    \ea\label{ex:cuw:36a} amosambikánó awúúbúwéla wááʈamagíʔa ázúgu wíílá anáátólócémó ðoorééla ðéʔu\\
        \gll  a-mosambikano a-hí-ubuwél-a ó-a-ttamag-íh-a á-zugu wíilá \textbf{a}{}-\textbf{náa}{}-\textbf{toloc}{}-\textbf{e}=\textbf{mo} dhi-oréela dhi-éhu \\
              2-Mozambican \textsc{sbj2-pfv.dj-}think-\textsc{fv} \textsc{15-sbj2-}run\textsc{{}-caus-fv} 2-white.person \textsc{comp} \textsc{sbj2-neg}{}-destroy-\textsc{sbjv=loc18} \textsc{10-}belonging \textsc{agr10-poss1pl} \\
        \glt ‘The Mozambicans expelled the whites so that they do not destroy our belongings.’ [Image 01\_Gui]

    \ex\label{ex:cuw:36b} Rósa ónóʈápwáʈeða náma, enáálúle\\
        \gll  Rósa ó-ni-o-ttápwattedh-a náma \textbf{e}{}-\textbf{náa}{}-\textbf{lul}{}-\textbf{e} \\
              Rosa \textsc{sbj1-ipfv.dj-15-}cook-\textsc{fv} 9a.meat \textsc{sbj9}{}-\textsc{neg-}spoil\textsc{{}-sbjv} \\
        \glt  ‘Rosa is cooking the meat in salty water, so that it does not spoil.’ (\citealt{FestiValler1994}: 317, my glosses)
    \z
\z 

The subjunctive is also used as the complement of the volitional verb \textit{ófuná} ‘to want’ in order to convey wishes. While this construction occurs frequently in my corpus with affirmative verb forms, the negative examples in \REF{ex:cuw:37} were elicited.

\ea\label{ex:cuw:37}
    \ea\label{ex:cuw:37a} ábále míyó ɖiḿ̩fúná ɖíiyaʈukule \\
        \gll  ábale míyo \textbf{ddi}{}-\textbf{ní}{}-\textbf{fun}{}-\textbf{á} \textbf{ddi}{}-\textbf{hi}{}-\textbf{a}{}-\textbf{ttukul}{}-\textbf{e}  \\
             2.\textsc{dem.dist} \textsc{pro1sg} \textsc{sbj.1sg-ipfv.cj-}want-\textsc{fv} \textsc{sbj.1sg-neg-obj2-}take-\textsc{sbjv} \\
        \glt ‘Those ones, I don’t want to take them.’ [Elicitation]

    \ex\label{ex:cuw:37b} ɖaáfúna ɖiimwe múróbo óbu\\
        \gll \textbf{ddi}{}-\textbf{á}{}-\textbf{fun}{}-\textbf{a} \textbf{ddi}{}-\textbf{hi}{}-\textbf{mw}{}-\textbf{e} mú-robo óbu  \\
             \textsc{sbj.1sg-pst.ipfv.cj-}want-\textsc{fv} \textsc{sbj.1sg-neg-}drink-\textsc{sbjv} 3-medicine 3.\textsc{dem.prox}  \\
        \glt ‘I didn’t want to take this medicine (but I did).’ [Elicitation]
    \z
\z 

In indicative complements, the verb form of the complement clause exhibits the same TA categories as verbs in declarative main or independent clauses. As a result, they use the same negation strategy as that found in standard negation, i.e. through the pre-initial standard negator \textit{ka}\textbf{\textit{{}-}}. This is illustrated in \REF{ex:cuw:38a} with present, and in \REF{ex:cuw:38b} with hypothetical.

{
\ea\label{ex:cuw:38}
    \ea\label{ex:cuw:38a} pa ápa vań̩lógííʔu wíílá mírála ða anámwáʔuno kaðiń̩gónééla\\ 
        \gll  pa ápa va-ní-log-á=ehu wíilá mí-rala dhi-a a-námwahuno \textbf{ka-dhi-ní-goneél-a} \\
             \textsc{cop16} \textsc{16.dem.prox} \textsc{sbj16-ipfv-}say-\textsc{fv.rel=pro1pl} \textsc{comp} 4-home \textsc{agr}4-\textsc{conn} \textsc{2-}young.generation \textsc{neg-sbj4-ipfv-}last-\textsc{fv} \\
        \glt ‘This is how we say that young couples do not last long.’ [Wedding.Story]

    \ex\label{ex:cuw:38b} kuulogúwága ɖaábaála ɖiíziw’ wi káɖigaataábúwile \\
        \gll  ka-o-log-úw-ag-a ddi-a-hí-baál-a ddi-hí-ziw-a wi \textbf{ka-ddi-gaa-taábuw-ile} \\
             \textsc{cntf-15-}tell-\textsc{pass-hab-fv} \textsc{sbj.1sg-pst-pfv.dj-}give.birth-\textsc{fv} \textsc{sbj.1sg-pfv.dj}{}-know-\textsc{fv} \textsc{comp} \textsc{neg-sbj.1sg-hyp}{}-suffer-\textsc{pfv} \\
        \glt ‘If I had children, I know that I would not suffer (i.e. they would sustain me).’ [Elicitation]
    \z
\z }

Finally, infinitive complements are negated by means of the post-initial negator \textit{hi}\textbf{\textit{{}-}}, which follows the class 15 infinitive prefix \textit{o}\textbf{\textit{{}-}}, as shown in \REF{ex:cuw:39}.
\newpage
\ea\label{ex:cuw:39} oówóðá oʔímútela múʈáamb’ úúĺ̩le  \\ 
    \gll  o-hí-wodh-á \textbf{o}{}-\textbf{hí}{}-\textbf{mu}{}-\textbf{tel}{}-\textbf{a} mú-ttaambí ól̩le \\
        \textsc{sbj1-pfv.dj}{}-succeed-\textsc{fv} \textsc{15-neg}{}-\textsc{obj}1-marry-\textsc{fv} 1-liar 1.\textsc{dem.dist}  \\
    \glt ‘She succeeded in not marrying that liar.’ [Elicitation]
\z 

\subsubsection{Temporal and conditional adverbial clauses}\label{sec:cuw:2.4.2}
Five TA categories relate to temporality and/or conditionality: situative, counter-expectational, sequential, narrative and counterfactual. These categories are summarized in \tabref{tab:cuw:5} in both their affirmative and negative shapes. As can be seen, all are invariably negated by the post-initial marker \textit{hi}{}-. The situative further employs the final vowel -\textit{i}, which replaces -\textit{a}. Counter-expectational expressing \textit{before} sentences as in \REF{ex:cuw:42} has no affirmative counterpart.\footnote{\textit{Before} counter-expectational differs from the counter-expectational expressing ‘not yet’ declaratives (discussed in \sectref{sec:cuw:2.1}) which is constructionally symmetric (see \tabref{tab:cuw:2}).} The last three TA categories, i.e. sequential, narrative and counterfactual, display constructional and paradigmatic symmetry with their affirmative equivalents. Each category is discussed in more detail after the table.

\begin{table}
\begin{tabularx}{\textwidth}{QllQ}
\lsptoprule
{Label} & {Affirmative} & {Negative} & {English} \\
\midrule 
situative & \textsc{sbj}-\textit{a}-\textsc{vb}-\textit{a} & \textsc{sbj}-\textit{a-\textbf{hi}}-\textsc{vb}-\textit{\textbf{i}} & ‘if/when we V’ \\
 & \textsc{sbj}-\textit{gaa}-\textsc{vb}-\textit{a} & \textsc{sbj}-\textit{ga(a)-\textbf{hi}}-\textsc{vb}-\textit{\textbf{i}} &  \\
\tablevspace
 ‘before’ counter-expectational  & - & \textsc{sbj}-\textit{\textbf{hi}-na}-\textsc{vb}-\textit{a} & ‘before we V’ \\
\tablevspace
sequential & \textit{ba}-\textsc{sbj}-\textsc{vb}-\textit{a} & \textit{ba}-\textsc{sbj}-\textit{\textbf{hi}}-\textsc{vb}-\textit{a} & ‘and we V/Ved’ \\
 & \textit{ba}-\textsc{sbj}-\textsc{vb}-\textit{ile} & \textit{ba-}\textsc{sbj}-\textit{\textbf{hi}}-\textsc{vb}-\textit{ile} & ‘when/while we are Ved’  \\
 & \textit{ba}-\textsc{sbj}-\textit{gaa}-\textsc{vb}-\textit{a} & \textit{ba}-\textsc{sbj}-\textit{ga(a)}-\textit{\textbf{hi}}-\textsc{vb}-\textit{a} & ‘while Ving’ \\
\tablevspace
narrative & \textit{o}-\textsc{vb}-\textit{a} & \textit{o-\textbf{hi}}-\textsc{vb}-\textit{a} & ‘and we Ved’ \\
\tablevspace
counterfactual  & \textit{ka-o}-\textsc{vb}-\textit{a} & \textit{ka-o-\textbf{hi}}-\textsc{vb}-\textit{a} & \mbox{‘if we (had) Ved’} \\
\lspbottomrule
\end{tabularx}
\caption{Temporal/conditional TA categories}
\label{tab:cuw:5}
\end{table}

The situative is used to express a logical or temporal precondition, best translated in English by if/when-clauses. Two situative markers exist in the language: the inherited prefix \textit{a}\textbf{\textit{{}-}} \REF{ex:cuw:40} and the borrowed prefix \textit{gaa}\textbf{\textit{{}-}} \REF{ex:cuw:41} \citep{Guérois2017}. The negative counterpart to the situative consists of the negative post-initial prefix \textit{hi}\textbf{\textit{{}-}} and the final vowel \textbf{\textit{{}-}}\textit{i}, which substitutes the default final vowel {}-\textit{a}. 


\ea\label{ex:cuw:40} ɖaaʔíríntigí kabálayáː, bwenɖéná kan̩náálíbe\\ 
    \gll  \textbf{ddi}{}-\textbf{a}{}-\textbf{hí}{}-\textbf{rint}{}-\textbf{ig}{}-\textbf{i} kabála=ya bwenddé=na ka-ni-náa-lib-e\\
          \textsc{sbj.1sg-sit-neg}{}-weave-\textsc{hab-neg} 9a.rope=\textsc{def} 5a.mat=\textsc{def5} \textsc{neg-sbj5-fut}{}-be.strong-\textsc{fv}\\
    \glt  ‘If I do not weave this rope, this mat will not be strong.’ [Elicitation]
\z 

\ea\label{ex:cuw:41} mugaʔíðów’ oomusikáː, kamunááje \\ 
    \gll  \textbf{mu}{}-\textbf{gaa}{}-\textbf{hí}{}-\textbf{dhow}{}-\textbf{i} o-mu-siká ka-mu-náa-j-e \\
        \textsc{sbj.2pl}{}-\textsc{sit-neg}{}-go\textsc{{}-neg} 17-\textsc{htd.3}{}-market \textsc{neg-sbj.2pl-fut}{}-eat-\textsc{fv} \\
    \glt ‘If you do not go to the market, you will not eat.’ [Elicitation]
\z 

In these forms, it is difficult to decide whether \textbf{\textit{{}-}}\textit{i} should be analyzed as a final negative marker or rather as a tense marker. Since this high final vowel only appears in the context of negative situatives and not affirmative situatives, I choose here to analyze it as a negative marker. This means that the situative is the only TAM category in Cuwabo whose negation is marked twice.

The counter-expectational prefix \textit{ná}\textbf{\textit{{}-}} (already discussed in ‘not yet’ declaratives of the shape \textit{ka}\textbf{\textit{{}-}}\textit{sbj}{}-\textit{ná}{}-\textit{VB}{}-\textit{a}, see \sectref{sec:cuw:2.1}), when combined with the negator \textit{hi}{}- serves to express ‘before Ving’ or ‘as long as + \textsc{neg}’ temporal clauses. This is very much in line with functions of ‘not yet’ expressions cross-linguistically (\citealt{Baar1997}; \citealt{Auwera1998}; \citealt{Plungian1999}). Two examples can be found in \REF{ex:cuw:11a} and \REF{ex:cuw:34a} above, and a third one is provided in \REF{ex:cuw:42}.

\ea\label{ex:cuw:42} kapíyáni míímbw’ iísó ðiinányâla \\ 
    \gll  ka-píy-a=ni mí-imbu éso \textbf{dhi}{}-\textbf{hi}{}-\textbf{ná}{}-\textbf{nyal}{}-\textbf{a} \\
        \textsc{imp}{}-cook-\textsc{fv}=\textsc{pl.addr} 4-pod 4.\textsc{dem.ref} \textsc{sbj}4-\textsc{neg-cnte}{}-wither-\textsc{fv} \\
    \glt ‘Cook these peas before they fade away (as long as they are not withered).’ [Elicitation]
\z 

The sequential is recurrently heard in narratives to mark subsequent actions. When the sequential marker \textit{ba}{}- is used on its own, it is best translated as ‘and he Ved’, as in \REF{ex:cuw:43}. When combined with the perfect(ive) suffix -\textit{ile}, it refers to a simultaneous state in the past \REF{ex:cuw:44}. Simultaneity is also conveyed with sequential situatives, but with a progressive reading translated as ‘while Ving’ \REF{ex:cuw:45}.\footnote{Sequential is part of the TA categories where class 1 subject prefix \textit{o}{}- surfaces as \textit{a}{}-. The resulting sequence \textit{ba-a}{}- may then merge and surface as short \textit{ba}{}-, but this is not systematic.} 
\newpage
\ea\label{ex:cuw:43} baakumbiɽa baaʔiðawo \\ 
    \gll ba-a-kumbilr-a \textbf{ba}{}-\textbf{a}{}-\textbf{hi}{}-\textbf{dh}{}-\textbf{a=wo} \\
        \textsc{seq-sbj1-}ask-\textsc{fv} \textsc{seq-sbj1-neg}{}-come-\textsc{fv=loc17}  \\
    \glt ‘And he asked not to come here.’ (lit. ‘And he asked and he did not come here.’) [Conversation.05]
\z 

\ea\label{ex:cuw:44} wáám̩pambíy’ óoɖikoséla áje, ókú baɖiimuté\^{l}̩le\\ 
    \gll  ó-a-ni-pamb-íh-a o-ddi-kos-él-a áje óku \textbf{ba}{}-\textbf{ddi}{}-\textbf{hi}{}-\textbf{mu}{}-\textbf{tél}{}-\textbf{ile} \\
        \textsc{sbj.2sg-pst-ipfv.dj}{}-worry-\textsc{caus-fv} \textsc{htd.15-obj.1sg}{}-do-\textsc{appl-fv} 9a.jealousy 1.\textsc{dem.prox} \textsc{seq-sbj.1sg-neg}{}-\textsc{obj}1-marry-\textsc{pfv}  \\
    \glt ‘You were preoccupied in making me jealous (lit. ‘you were preoccupied it is making me jealous’), when/while I was not engaged with you.’ [Elicitation]
\z 

\ea\label{ex:cuw:45} bagávirávóː, bagaʔílôga \\ 
    \gll ba-gá-vir-á=vo \textbf{ba}{}-\textbf{ga}{}-\textbf{hí}{}-\textbf{log}{}-\textbf{a}  \\
        \textsc{seq.sbj1}-\textsc{sit}{}-pass-\textsc{fv}=\textsc{loc16} \textsc{seq.sbj1}-\textsc{sit-neg}{}-talk-\textsc{fv}  \\
    \glt ‘As he was going, he was not talking.’ [Elicited from Mbílri.Story]
\z 

Like the sequential, the narrative is frequently used in a sequence of events, once the time frame was overtly marked on the first verb form. In this sense the narrative is dependent on the temporal context previously established. Morphologically, the narrative is a non-finite verb form, segmentally identical to the infinitive, but supra-segmentally different: in narrative forms, the first underlying High tone (historically the augment tone), does not surface. Thus, \textit{ohimugulihédha} in \REF{ex:cuw:46}, interpreted as ‘I did not sell (it) to him’ differs in tones from the negative infinitive \textit{ohímúgulihédha} ‘to not sell (something) to someone’. While affirmative narrative is frequently heard, negative narrative seems much rarer.\footnote{Cross-linguistically, it is common for non-finite forms to occur in a variety of dependent clauses performing different functions \citep{Cristofaro2003}.}

\ea\label{ex:cuw:46} muɖaákúlelííyé káḿbo, míyó oʔimuguliyéða \\ 
    \gll mu{}-ddi-ákul-el-é=iye kámbo míyo \textbf{o}{}-\textbf{hi}{}-\textbf{mu}{}-\textbf{gul}{}-\textbf{ih}{}-\textbf{édh}{}-\textbf{a}  \\
         \textsc{agr18-obj.1sg}{}-answer-\textsc{appl-pfv.rel=pro3sg} 9a.insult \textsc{pro1sg} \textsc{narr-neg-obj1}{}-buy-\textsc{caus-appl-fv} \\
    \glt ‘When he replied (to me) by an insult, I did not sell (it) to him.’ [Elicitation]
\z 

Finally, the counterfactual is found in speculative conditional sentences known to be false or contrary to the facts. It is formally composed of the counterfactual prefix \textit{ka}{}- followed by the infinitive verb. The negator \textit{hi}{}- is placed between the infinitive prefix and the verbal stem, as shown in \REF{ex:cuw:47}.

\ea\label{ex:cuw:47} kooʔígúl’ ooba kavagaajúwe \\ 
    \gll  \textbf{ka}{}-\textbf{o}{}-\textbf{hí}{}-\textbf{gúl}{}-\textbf{a} o-ba ka-va-gaa-j-úw-(il)e \\
         \textsc{cntf-15-neg}{}-buy-\textsc{fv} \textsc{htd.11}{}-fish \textsc{neg-sbj16-hyp}{}-eat-\textsc{pass-pfv} \\
    \glt ‘If I had not bought fish, there would be no meal at home.’ [Elicitation]
\z 

\subsubsection{Relative clauses}\label{sec:cuw:2.4.3}
Cuwabo exhibits a relativization strategy, also attested in certain North-Western Bantu languages \citep{Nsuka-Nkutsi1982}, where relativized verb forms look like participial modifiers and retain all verbal properties except subject agreement. Thus, the initial verbal slot dedicated to subject marking in independent clauses is occupied by a pronominal prefix in relative clauses, which expresses agreement with the head noun, just like any modifier in the language (see \citealt{GuéroisCreissels2020}). This is illustrated in the non-subject relative in \REF{ex:cuw:48}, where the relativized verb \textit{dhiḿ̩fúná} agrees with class 10 noun \textit{guwo} ‘clothes’.

{\ea\label{ex:cuw:48} oðówé wágúlé guwóðáwo ðiḿ̩fúná wéyo\\
    \gll  o-dhów-e o-á-gul-é \textbf{guwó}=dhi-awo \textbf{dhi}{}-ní-fun-á wéyo  \\
          \textsc{sbj.2sg}{}-go-\textsc{sbjv} \textsc{sbj.2sg-amot-}buy-\textsc{sbjv} 10a.cloth=\textsc{agr}10-\textsc{poss2sg} \textsc{agr10-ipfv-}want-\textsc{fv.rel} \textsc{pro2sg}  \\
    \glt  ‘Go and buy (your) clothes that you want.’ [Women.Duty]
\z }

Not all TAM categories can be relativized in Cuwabo. In fact, only the seven TAMs which display the CJ/DJ alternation (discussed in \sectref{sec:cuw:2.1} above) have relative equivalents. From a morphological perspective, affirmative relative verbs are on the surface identical to conjoint forms, but differ in their distribution: relative verbs may appear sentence-finally, whereas conjoint verbs cannot. Furthermore, in the case of transitive constructions, conjoint forms trigger HTD on the following (nominal) constituent as already seen in \REF{ex:cuw:3} above, whereas the object following a relative verb appears in its citation form, e.g. \textit{balílo} ‘wood poles’ in \REF{ex:cuw:49}.
\newpage
\ea\label{ex:cuw:49} ḿ̩púle m̩muʈáʈi munááremííwe balílo\\
    \gll  ḿ̩pule mu-mu-ttátti mu-náa-rem-e=iwe balílo  \\
           18.\textsc{dem.dist} \textsc{18-3-}mangrove \textsc{agr18-fut}{}-cut-\textsc{fv.rel=pro2sg} 10a.wood.pole \\ 
    \glt ‘There in the mangrove, where you are going to cut wood poles.’ [House.Building]
\z 

Another distinction between conjoint verb forms and affirmative relatives is the way they encode negative polarity. Conjoint verb forms function as independent TA forms and consequently resort to the pre-initial negative marker \textit{ka}{}- (see \tabref{tab:cuw:2}). Their relative counterparts are dependent and encode negation via the negator \textit{hi}{}-. Both affirmative and negative relative forms are provided in \tabref{tab:cuw:6}.

\begin{table}
\begin{tabularx}{\textwidth}{XXl}
\lsptoprule
{Label} & {Affirmative} & {Negative}\\
\midrule 
{\textsc{rel} present} & {\textsc{sbj}{}-\textit{ni}-\textsc{vb}{}-\textit{a}} & {\textsc{sbj}{}-\textit{\textbf{hi}{}-ni}-\textsc{vb}{}-\textit{a}}\\
{\textsc{rel} perfective} & {\textsc{sbj}{}-\textsc{vb}{}-\textit{ile}} & {\textsc{sbj}{}-\textit{\textbf{hi}}{}-\textsc{vb}{}-\textit{ile}}\\
{\textsc{rel} past perfective} & {\textsc{sbj}{}-\textit{a}-\textsc{vb}{}-\textit{ile}} & {\textsc{sbj}{}-\textit{a-\textbf{hi}}{}-\textsc{vb}{}-\textit{ile}}\\
{\textsc{rel} past imperfective} & {\textsc{sbj}{}-\textit{a}-\textsc{vb}{}-\textit{a}} & {\textsc{sbj}{}-\textit{a-\textbf{hi}}{}-\textsc{vb}{}-\textit{a}}\\
{\textsc{rel} future} & {\textsc{sbj}{}-\textit{náa}-\textsc{vb}{}-\textit{e}} & {\textsc{sbj}{}-\textit{\textbf{hi}{}-náa}-\textsc{vb}{}-\textit{e}}\\
{\textsc{rel} future imperfective} & {\textsc{sbj}{}-\textit{gá}-\textsc{vb}{}-\textit{a}} & {\textsc{sbj}{}-\textit{\textbf{hi}{}-gá}-\textsc{vb}{}-\textit{a}}\\
{\textsc{rel} hypothetical} & {\textsc{sbj}{}-\textit{gaa}-\textsc{vb}{}-\textit{ile}} & {\textsc{sbj}{}-\textit{gaa-\textbf{hi}}{}-\textsc{vb}{}-\textit{ile}}\\
\lspbottomrule
\end{tabularx}
\caption{TA categories available for relative clauses}
\label{tab:cuw:6}
\end{table}

Symmetry operates perfectly on both constructional and paradigmatic levels: each affirmative form has a corresponding negative form, and there is a one-to-one formal match between the two paradigms. An example of a negative relative construction is provided in \REF{ex:cuw:50}.

{
\ea\label{ex:cuw:50} kaáʈíya muʈu óðówá otélá n̩lógó niʔiń̩zíwééye \\ 
    \gll  ka-á-ttiy-a mu-ttu ó-dhow-á o-tél-a ni-lógo \textbf{ni}{}-\textbf{hi}{}-\textbf{ní}{}-\textbf{ziw}{}-\textbf{á}=\textbf{eye} \\
          \textsc{neg.sbj2-pst.ipfv-}leave-\textsc{fv} \textsc{htd.1}{}-person \textsc{15}{}-go-\textsc{fv} \textsc{15}{}-marry-\textsc{fv} \textsc{5}{}-family \textsc{agr}5-\textsc{neg-ipfv}{}-know-\textsc{fv}\textsc{.rel=pro3sg} \\
    \glt ‘One would not leave someone marry some unknown family.’ (lit. ‘family that they do not know of’) [Marriage.Tradition]
\z }


An alternative strategy to express negation in relatives is to use the auxiliary verb stem \textit{koódda} ‘refuse’ followed by the main verb in its infinitive form. This is shown in \REF{ex:cuw:51}.


\ea\label{ex:cuw:51}
    \ea\label{ex:cuw:51a} káɖívaʔeðé sábáá eekoóɖíl’ oosígíʔiwá \\ 
        \gll  ká-ddi-vah-edh-é sábaá \textbf{e}{}-\textbf{a}{}-\textbf{koódd}{}-\textbf{ile} o-sígih-iw-á\\
              \textsc{imp-obj.1sg}{}-give-\textsc{appl-fv} 9a.soap \textsc{agr}9-\textsc{pst}{}-refuse-\textsc{pfv.rel} 15-apply-\textsc{pass-fv}\\
        \glt ‘Give me the soap which has not been used.’ [Elicitation]
    \ex\label{ex:cuw:51b} eń̩kóóɖélíim’ úwáákuláː, [...]  \\
        \gll \textbf{e}{}-\textbf{ní}{}-\textbf{koodd}{}-\textbf{él}{}-\textbf{a=imi} o-ú-akul-á \\
            \textsc{agr}9-\textsc{ipfv}{}-refuse-\textsc{appl}{}-\textsc{fv.rel=pro}1\textsc{sg} \textsc{15}{}-\textsc{obj}2\textsc{sg}{}-reply-\textsc{fv} \\
        \glt ‘If I do not reply to you, ...’ [Elicitation]
    \z
\z 

\section{Non-clausal negation}\label{sec:cuw:3}
\subsection{Negative replies}\label{sec:cuw:3.1}
Negative replies to polar questions may be expressed in two different ways. The default strategy consists in repeating the inflected verb in its negative form, that is the ‘echo system’ in \citegen{SadockZwicky1985} typology.\footnote{See \citet{EnfieldEtAl2018} and \citet{Moser2019} for more recent typologies.} This is seen in \REF{ex:cuw:52} where the negative verb form \textit{kanuwíídhi} ‘we don’t know you’ is first used to answer the affirmative question \textit{munóddíziwá}?\footnote{Two lexical verb stems for ‘know’ exist in Cuwabo: the regular root \textit{ziw} and the defective verb \textit{idhi}.} ‘do you know me?’. Still in \REF{ex:cuw:52}, \textit{kanuwíídhi} is repeated again, but this time in echo with the negative question \textit{kamuddíídhí}? ‘don’t you know me?’, so as to confirm the content of the question. Affirmative replies to polar questions require the same strategy, as shown in \REF{ex:cuw:53}, where \textit{ɖiwííwa} ‘I heard’ is used as an affirmative reply to \textit{owííwá?} ‘did you hear?’.

\ea\label{ex:cuw:52} Maríy’ óowáávuúza: “míyó munóɖíziwáː?”  \\ 
    \gll  Maríya o-hí{}-a-vuúz-a {\db}míyo mu-ni-ó{}-ddi-ziw-a\\
          Maria \textsc{sbj1-pfv.dj-obj2}{}-ask-\textsc{fv} {\db}\textsc{pro1sg} \textsc{sbj.2pl-ipfv.dj-15}{}-\textsc{obj.1sg}{}-know-\textsc{fv} \\
    \glt ‘Maria asked them: “Do you know who I am?”
    
\medskip áwééne: “yiíí! kanuwííði” \\
    \gll  áweéne {\db}yiíí \textbf{ka}{}-\textbf{ni}{}-\textbf{ú}{}-\textbf{idhi} \\
          \textsc{pro3pl} {\db}\textsc{inj} {\db}\textsc{neg-sbj.1pl-obj.2sg}{}-know  \\
    \glt    They: “We don’t know you.” 
    
\medskip “míyó kamuɖííðíː?” \\
    \gll  míyo ka-mu-ddí{}-idhi \\
           \textsc{pro1sg} \textsc{neg-sbj.2pl-obj.1sg}{}-know   \\
    \glt  [Maria] “You don’t know me?”
    
\medskip belá: “kánuwííði”\\
    \gll  ba-il-á {\db}\textbf{ka}{}-\textbf{ni}{}-\textbf{ú}{}-\textbf{idhi}  \\
          \textsc{seq.sbj1}{}-do/say-\textsc{fv} {\db}\textsc{neg-sbj.1pl-obj.2sg}{}-know   \\
    \glt   [They] “We don’t know you.”\,’ [Maria.Story]
\z

{\ea\label{ex:cuw:53} “owííwá Maríyá?” Maríya: “ɖiwííwa” \\
    \gll o-hí-iw-á Maríya Maríya \textbf{ddi}{}-\textbf{hí}{}-\textbf{iw}{}-\textbf{a}  \\
          \textsc{sbj.2sg-pfv.dj}{}-hear-\textsc{fv} Maria Maria \textsc{sbj.1sg-pfv.dj}{}-hear-\textsc{fv} \\
    \glt ‘“Did you hear, Maria?” Maria: “I did” (lit. ‘I heard’).’ [Maria.Story]
\z }

Less frequently, one-word replies to polar questions are used, matching \citegen{SadockZwicky1985} ‘yes/no system’. Three negative particles are attested in Cuwabo, \textit{nada}, \textit{nne} and \textit{não}, which all serve as ‘no’. The most common and spontaneously used in discourse is \textit{nada}, borrowed from Portuguese \textit{nada} ‘nothing’. It usually stands on its own, as the two naturalistic examples below show. 

\ea\label{ex:cuw:54} owáávuúz’ áábál’ aálîvo. \\ 
    \gll  o-hí{}-a-vuúz-a ábale a-á{}-li=vo  \\
          \textsc{sbj1-pfv.dj-obj2}{}-ask-\textsc{fv} 2.\textsc{dem.dist} \textsc{sbj2-pst}{}-be=\textsc{loc16}  \\
    \glt ‘He asked those who were there: \\
\medskip  “ósál’ uúb’ úup\H{a}t\H{u}wa?” múkwááye: “náda!”\\
    \gll ó-salu óbu o-hi-pátuw-a mú-kwe=aye \textbf{nada} \\
           11-thread 11.\textsc{dem.prox} \textsc{sbj11-pfv.dj}{}-break-\textsc{fv} 1-friend=\textsc{poss3sg} no \\
    \glt “Did this thread break?” His friend: “No!”\,’ [Mbílri.Story]
\z 

\ea\label{ex:cuw:55}
    \begin{xlist}[A: ]
    \exi{A: }\label{ex:cuw:55A} ókúléwa ókúle munofíyâwo? \\ 
        \gll  ókule=wa ókule mu-ni-o-fíy-a=wo \\ 
             17.\textsc{dem.dist=def} \textsc{17.dem.dist} \textsc{sbj.2pl-ipfv-15}{}-arrive-\textsc{fv=loc17}\\
        \glt  ‘There, are you going to arrive there?’ 

    \exi{B: }\label{ex:cuw:55B} nada, nínga múkwáága kanáða kaɖinónile\\ 
        \gll   \textbf{nada} nínga mú-kwe=aga ka-ná-dh-a ka-ddi-nón-ile \\
               no like 1-friend-\textsc{poss1sg} \textsc{neg.sbj1-cnte-}come-\textsc{fv} \textsc{neg-sbj.1sg}{}-know-\textsc{pfv} \\
        \glt ‘No, as my friend has not arrived yet, I don’t know.’ [Conversation.3]
    \end{xlist}
\z 

The negative particle \textit{nne} is not attested in my naturalistic data. Examples in (\ref{ex:cuw:56a}--\ref{ex:cuw:56b}) were elicited. Comparing \REF{ex:cuw:56a} and \REF{ex:cuw:56b}, we see that \textit{nne} ‘no’ disagrees with the content rather than with the polarity of the question. In both \REF{ex:cuw:56a} and \REF{ex:cuw:56b}, answering \textit{nne} ‘no’ means that the woman is not cultivating. On the other hand, answering \textit{ndde} ‘yes’ functions as a positive answer, i.e. the woman \textit{is} cultivating. In this sense, \textit{nne} and \textit{ndde} have the same connotation as in English: they refer to the confirmation or negation of the content, and thus do not function as polarity reversing particles \citep{Holmberg2015}. 

\ea\label{ex:cuw:56a}
    \begin{xlist}[B2: ]
        \exi{A: }\label{ex:cuw:56aA} múyaná ónólímá?\\ 
            \gll mú-yaná ó-ni-ó-lim-a\\ 
                  1-woman \textsc{sbj1-ipfv.dj-inf}{}-cultivate-\textsc{fv}\\
            \glt  ‘Is the woman cultivating?'
        \exi{B1: }\label{ex:cuw:56aB1} ńnêː! \\ 
            \gll \textbf{nne}\\
                    no \\
            \glt ‘No!’
        \exi{B2: }\label{ex:cuw:56aB2} ńɖêː! \\ 
            \gll  \textbf{ndde}\\
                    yes\\
            \glt ‘Yes!’
    \end{xlist}
\z
\ea\label{ex:cuw:56b}
    \begin{xlist}[B2: ]
        \exi{A: }\label{ex:cuw:56bA} muyáná kánólimá? \\ 
            \gll mú-yaná ká-ni-ó-lim-a\\ 
                   1-woman \textsc{neg.sbj1-ipfv.dj-inf}{}-cultivate-\textsc{fv}\\
            \glt  ‘Isn’t the woman cultivating?'
        \exi{B1: }\label{ex:cuw:56bB1} ńnêː! \\
            \gll   \textbf{nne}\\
                    no \\
            \glt ‘No! (= she is not)’
        \exi{B2: }\label{ex:cuw:56bB2} ńɖêː! \\ 
            \gll \textbf{ndde}\\
                    yes\\
            \glt ‘Yes! (= she is)’
    \end{xlist}
\z 


The use of the negative particle \textit{não} (< Portuguese \textit{não} ‘no’) appears very marginal in Cuwabo. The only naturalistic occurrence in my corpus is an example of ironic negation, provided in \REF{ex:cuw:87} below (see \sectref{sec:cuw:4.6.4}).

\subsection{Negative indefinites and quantifiers}\label{sec:cuw:3.2}
Cuwabo does not have indefinite pronouns such as ‘somebody’, ‘something’ or ‘somewhere’. The same holds for their negative counterparts ‘nobody’, ‘nothing’ and ‘nowhere’, respectively. Instead, generic nouns denoting ‘person’, ‘thing’ and ‘place’ are used together with clausal negation. The absence of negative indefinite pronouns and the construction whereby affirmative (pro)nouns occur with negated verbs seems to be a widespread feature in African languages \citep[2]{Cyffer2009}. Thus, the noun \textit{múttu} ‘person’ is used to express both ‘somebody’ and ‘nobody’. The interpretation ‘somebody’ is obtained with affirmative polarity \REF{ex:cuw:57a}, while ‘nobody’ requires negation marking on the verb and the resulting construction is usually a thetic sentence with VS word order \REF{ex:cuw:57b}. 

\ea\label{ex:cuw:57} 
    \ea\label{ex:cuw:57a} ɖiníḿ̩funá múʈ’ uuń̩zíwá wííyá vaɖíííɖi\\ 
        \gll ddi-ní-mu-fun-á \textbf{mu}{}-\textbf{ttu} o-ní-ziw-á ó-iy-á vaddíddi \\
              \textsc{sbj.1sg}{}-\textsc{ipfv.cj-obj}1-want-\textsc{fv} \textsc{htd.1}{}-person \textsc{sbj1}-\textsc{ipfv}{}-know-\textsc{fv} \textsc{15}{}-steal-\textsc{fv} much  \\
        \glt  ‘I want someone who steals a lot.’ [Maria.Story]
    \ex\label{ex:cuw:57b} kaðílévó muʈu \\
        \gll  \textbf{ka}{}-dh-íle=vo \textbf{mu}{}-\textbf{ttu}\\
               \textsc{neg.sbj1}-come-\textsc{pfv=loc16} \textsc{htd.1}{}-person\\
        \glt  ‘Nobody came.’ (lit. ‘s/he did not come here someone’) [Elicitation]
    \z
\z  

Alternatively, the use of a cleft construction combining a negative existential and a free relative clause also serves to express ‘nobody’. In \REF{ex:cuw:58}, \textit{kaáwó odhileewó}, literally translated as ‘there is not who came’, is semantically equivalent to \textit{kadhílévó muttu} in \REF{ex:cuw:57b}. 
\newpage
\ea\label{ex:cuw:58} kaáwó oðileewóː \\ 
    \gll kaá=wo o-dh-ile=wo \\
           \textsc{negex.sbj1}=\textsc{loc17} \textsc{agr}1-come-\textsc{pfv.rel.htd}=\textsc{loc17}\\
    \glt  ‘Nobody came.’ (lit. ‘there is not who came here’) [Elicitation]
\z 

While the use of a generic word (equivalent to strategy 1 in \citegen{Kahrel1996} typology) is a widespread alternative in substitution for indefinite pronouns crosslinguistically \citep[53]{Haspelmath1997}, the negative existential construction as in \REF{ex:cuw:58} (Kahrel’s strategy 5) seems less reported worldwide and mostly attested in Oceanic languages (see \citealt[52--57]{Haspelmath1997} and \citealt[118--123]{Veselinova2013neg.exist}) for discussion).

The concept \textsc{‘}something’ is expressed through the noun \textit{élôbo} (or its reduced variant \textit{élo}) ‘thing’. Similarly to ‘somebody/nobody’, its negative interpretation ‘nothing’ is obtained by means of negative verb forms, as illustrated in \REF{ex:cuw:59}. 

{
\ea\label{ex:cuw:59}
    \ea\label{ex:cuw:59a} míyó kaɖináálógéː ɖihiváʔíwe elobo\\ 
       \gll míyo ka-ddi-náa-log-e ddi-\textbf{hi}{}-váh-iw-e \textbf{e}{}-\textbf{lobo}\\
          \textsc{pro1sg} \textsc{neg-sbj.1sg-fut}{}-speak-\textsc{fv} \textsc{sbj.1sg-neg-}give-\textsc{pass-sbjv} \textsc{htd.9}{}-thing \\
        \glt  ‘I will not speak if I am not given anything.’ [Elicitation]
    \ex\label{ex:cuw:59b} kaɖiḿ̩fún’ éelo \\
        \gll  \textbf{ka}{}-ddi-ní-fun-á \textbf{e}{}-\textbf{lo}\\
              \textsc{neg-sbj.1sg-ipfv}{}-see-\textsc{fv} \textsc{htd.9}{}-thing  \\
        \glt  ‘I do not want anything.’ [Elicitation]
    \z
\z }

Three locative nouns exist to express the meaning ‘somewhere’: class 16 \textit{vógo} ‘somewhere \textit{at} some place’, class 17 \textit{wógo} ‘somewhere \textit{to} some place’ \REF{ex:cuw:60a}, and class 18 \textit{mógo} ‘somewhere \textit{in} some place’ \REF{ex:cuw:60b}. Again, their negative counterpart, ‘nowhere’, is obtained through negative polarity on verbs integrated into thetic sentences \REF{ex:cuw:60c}.

\ea\label{ex:cuw:60}
    \ea\label{ex:cuw:60a} ɖaáðówá wógóː waáɖámana na bára\\ 
        \gll  ddi-á-dhow-á \textbf{wogo} o-áddam-an{}-a na bára \\
               \textsc{sbj.1sg-pst.cj}{}-go-\textsc{fv} \textsc{htd.17}.somewhere \textsc{15-}be.next.to-\textsc{rec-fv} with 5a.sea\\
        \glt  ‘I went somewhere near the sea.’ [Elicitation]

    \ex\label{ex:cuw:60b} ðoóbóː oʈiyilé mógo gaani? \\
        \gll dhi-óbo o-ttiy-ilé \textbf{mogo} gaani  \\
              10-basket \textsc{sbj1}-leave-\textsc{pfv.cj} \textsc{htd.18}.somewhere which \\
        \glt  ‘In which place did he leave the baskets?’ [Elicitation]

    \ex\label{ex:cuw:60c} kavalí vogóː van̩rórómeleya asúpâyi\\
        \gll  \textbf{ka}{}-va-lí \textbf{vogo} va-ni-róromel-ey-a a-súpayi \\
            \textsc{neg-sbj1}6-be \textsc{htd.16}.somewhere \textsc{agr}16-\textsc{ipfv}{}-trust-\textsc{n-fv.rel} 2-policeman \\
        \glt   ‘Nowhere are the policemen trustworthy.’ (lit. ‘there is no place where policemen are trustable) [Elicitation]
 
    \z
\z 

All the affirmative lexical items presented so far in this section have otherwise retained affirmative polarity uses in the language, which means that when they occur without negation, they never acquire a negative interpretation all by themselves (in the way \textit{personne} ‘person’ in French has for instance, by acquiring the parallel interpretation ‘nobody’). As an expected result, they cannot function as an elliptic answer ‘Nobody!’, ‘Nothing!’ or ‘Nowhere!’ to questions ‘Who came?’, ‘What did you eat?’ or ‘Where did you go?’, respectively.

Finally, the time frequency value ‘never’ is most commonly associated with the counter-expectational TAM category, expressed by means of the prefix \textit{ná}{}-. Counter-expectational conveys both unexpectedness and perfect aspectual meaning. In \REF{ex:cuw:61}, the described event ‘go to the beach’ has so far never occurred, and appears as some surprising and unexpected piece of information.

\ea\label{ex:cuw:61} míyó kaɖináðówag’ oobara\\ 
    \gll míyo ka-ddi-\textbf{ná}{}-dhow-ag-a o-bara\\
          \textsc{pro1sg} \textsc{neg-sbj.1sg-cnte}{}-go-\textsc{hab-fv} \textsc{htd.17}{}-9a.sea \\
    \glt  ‘I have never been at the beach.’ [Elicitation]

\z 

More rarely, a dedicated temporal adverb in the shape of \textit{cípo} is used to express ‘never’. It can be found in addition to counter-expectational TAM-marking \REF{ex:cuw:62a}, but may also easily be associated with other TAM categories, such as present imperfective \REF{ex:cuw:62b} or future \REF{ex:cuw:63}. This temporal adverb is inherently negative, but still requires to be embedded in a negative polarity context, provided by verbal negative markers. Unlike the ‘nobody/nothing/nowhere’ constructions discussed above, where negation is formally and semantically expressed once, the constructions based on \textit{cípo} ‘never’ are marked by a negative concord in the sense of \citet[9]{Zanuttini1997}, i.e. one single negative meaning is expressed twice, through the default negation marker (in the shape of a negative affix in Cuwabo) and through an inherently negative pronoun or adverb. In transitive constructions \REF{ex:cuw:63}, \textit{cípo} may be located clause-initially or clause-finally, but cannot emerge between the verb and the postverbal object. All the examples below were elicited. 

\ea\label{ex:cuw:62}
    \ea\label{ex:cuw:62a} kaɖinápíyága cípo \\ 
        \gll  ka-ddi-\textbf{ná}{}-piy-ag-a \textbf{cípo}  \\
             \textsc{neg-sbj.1sg-cnte}{}-cook-\textsc{hab-fv} never  \\
        \glt  ‘I have never cooked.’ [Elicitation] 

    \ex\label{ex:cuw:62b} kaɖim̩píya cípo\\
        \gll  ka-ddi-ni-píy-a \textbf{cípo} \\
               \textsc{neg-sbj.1sg-ipfv}{}-cook-\textsc{fv} never\\
        \glt  ‘I never cook.’ [Elicitation]
    \z
\z 

\ea\label{ex:cuw:63} (cípó) kaɖin̩náátágíʔe (*cípo) ebáríbariya (cípo) \\ 
    \gll (\textbf{cípo}) ka-ddi-mu-náa-tagih-e {} e-báribari=ya (cípo)  \\
          never \textsc{neg-sbj.1sg-obj1-fut}{}-repeat-\textsc{fv} {} 9-truth=\textsc{def} never  \\
    \glt  ‘I will never tell him the truth.’ [Elicitation]

\z 


\subsection{Negative derivation and case marking}\label{sec:cuw:3.3}
Case marking, including abessive, caritive or privative, does not exist in Cuwabo, and only some remnants of negative derivation are attested synchronically. The equivalent of English ‘un-’ is expressed lexically through dedicated verbs, most of which are historically derived by means of the (no longer productive) suffix {}-\textit{ul}, which is a reflex of the reconstructed Proto-Bantu form \textit{*}{}-\textit{ʊd}, interpreted as a ‘reversive’ (\citealt{Meeussen1967}: 87, 92) or ‘separative’ (\citealt{Schadeberg1982, Schadeberg2003}). In all cases, the original underived verb is no longer attested in the language, hence each pair of verbs in \REF{ex:cuw:64} displays different roots.

\TabPositions{.33\textwidth,.1\textwidth,.4\textwidth}
\ea\label{ex:cuw:64} 
ókosá ‘to do’ \tab  vs. \tab  ógujúla ‘to undo’ \\
ókuneéla ‘to cover’ \tab  vs.\tab  ópumbunúla ‘to uncover’ \\
osásánya ‘to build’ \tab  vs.\tab  ógújúla ‘to unbuild’ \\
okúna/ofúnya ‘to fold’ \tab  vs.\tab  otápúula ‘to unfold’ \\
ómagá ‘to tie’ \tab  vs. \tab otápúula ‘to untie’ \\
\z 

English ‘without’ or ‘-less’ is rendered through the adposition \textit{sé} (< Portuguese \textit{sem} ‘without’), used both with nouns \REF{ex:cuw:65} and infinitive forms of verbs \REF{ex:cuw:66}.


\ea\label{ex:cuw:65} íyééne waálí sé guwo\\
    \gll íyeéne o-á-li \textbf{sé} guwo \\
           \textsc{pro3sg} \textsc{sbj1-pst.ipfv-}be\textsc{} without \textsc{htd.}10a.cloth\\
    \glt ‘He remained without clothes.’ [Elicitation]
\z 

\ea\label{ex:cuw:66}
    \begin{xlist}[A: ]
    \exi{A: }\label{ex:cuw:66A} niigúmána mas sssssh, yaaká ngáasi niigumánîle? \\ 
        \gll ni-hi-gúman-a mas sssssh e-aká ngáasi ni-hi-gumán-ile \\ 
             \textsc{sbj.1pl-pfv.dj}{}-meet-\textsc{fv} but \textsc{inj} 9-year how.many \textsc{sbj.1pl-neg}{}-meet-\textsc{pfv}\\
        \glt ‘We have already met, but ooooh, how many years without meeting each other?’  
    \exi{B: }\label{ex:cuw:66B} s’ óógúmána \\ 
        \gll \textbf{sé} ó-gumán-a\\
          without \textsc{15}{}-meet-\textsc{fv}\\
    \glt ‘Without meeting each other.’ [Conversation.7]
    \end{xlist}
\z 

\section{Other aspects of negation}\label{sec:cuw:4}
This section explores other aspects of negation, which are more directly concerned with semantic and pragmatic issues. The discussion covers the scope of negation (\sectref{sec:cuw:4.1}), negative polarity items (\sectref{sec:cuw:4.2}), the marking of NPs in the scope of negation (\sectref{sec:cuw:4.3}), the reinforcement of negation (\sectref{sec:cuw:4.4}), negation in sentence coordination (\sectref{sec:cuw:4.5}) and other aspects of negation (\sectref{sec:cuw:4.6}).

\subsection{The scope of negation}\label{sec:cuw:4.1}
The scope of negation represents the area semantically affected by a negation cue, i.e. all the negated concepts. On the clausal level, negation cues and scopes are discontinuous: while negation is invariably marked on the verb (highlighted in bold in \REF{ex:cuw:67}), the scope further operates on the different clause constituents, indicated between square brackets in \REF{ex:cuw:67}. Among these constituents, the one immediately following the verb (here the object \textit{attu} ‘people’) is marked by High Tone Deletion (HTD), as further discussed in \sectref{sec:cuw:4.3}. The other constituents under negation scope are not subjected to any particular prosodic marking. 
\newpage
\ea\label{ex:cuw:67} ogaaʔíléjeeleg’ aaʈu m̩múrúɖááni kuń̩léméélîwa\\ 
    \gll  [o]-gaa-\textbf{hí}{}-[lejeel]-ege [a-ttu] [mu-mú-ruddá-ni] ku-ní-lemeél-iw-a \\ 
          \textsc{sbj.2sg-sit-neg-}greet-\textsc{fv.hab} \textsc{htd.2-}person 18-3-village-\textsc{loc} \textsc{neg.sbj.2sg-prs-}adapt-\textsc{pass-fv} \\
    \glt ‘If you do not greet the people of the community, you will not be integrated.’ [Elicitation]
\z 

It is also possible to narrow the scope of subject constituents by inverting the canonical order SV to VS, as in \REF{ex:cuw:68}, repeated from \REF{ex:cuw:57}. In this construction too, the postverbal element, here the subject \textit{muttu} ‘person’, is marked by HTD. 

\ea\label{ex:cuw:68} kaðílévó muʈu  \\ 
    \gll  ka-dh-íle=vo mu-ttu \\
          \textsc{neg.sbj1}-come-\textsc{pfv=loc16} \textsc{htd.1}{}-person\\
    \glt ‘Nobody came.’ (lit. ‘s/he did not come here someone’) [Elicitation]

\z 

\subsection{Negative polarity items}\label{sec:cuw:4.2}
Items restricted to negative polarity are very limited in Cuwabo. These consist of the two non-verbal negators discussed and illustrated in \sectref{sec:cuw:2.3}, i.e. the general non-verbal negator \textit{kahíye} and the negative locative-existential \textit{ka}{}-\textsc{sbj}{}-\textit{hí}=\textsc{loc.} 

On the basis of my corpus, the verb \textit{koodda} ‘refuse’ (in combination with a main infinitive verb) qualifies as another example of items that only occur in negative polarity contexts. This was shown in relative constructions such as \REF{ex:cuw:51}. In the question in \REF{ex:cuw:69}, two negative independent clauses follow each other: while the second exhibits standard negation in \textit{ka}{}-, the first illustrates \textit{koodda} + infinitive negation. 

\ea\label{ex:cuw:69} ogákóoɖ’ oóða kunámál’ oōsíńtínááríː?\\  
    \gll  \textbf{o}{}-\textbf{gá}{}-\textbf{koodd}{}-\textbf{a} ó-dh-a ku-ná-mal-a ó-sintinaári  \\
          \textsc{sbj.2sg}{}-\textsc{fut.ipfv.cj}{}-refuse-\textsc{fv} \textsc{15}{}-come-\textsc{fv} \textsc{neg.sbj.2}\textsc{sg}{}-\textsc{cnte-}finish-\textsc{fv} \textsc{15}{}-defecate  \\
    \glt  ‘Are you not coming, are you not done yet with your necessities?’ [Páaká.Story]

\z 

\subsection{Marking of NPs in the scope of negation}\label{sec:cuw:4.3}
It is cross-linguistically well-established that negation may have an effect on the marking of grammatical categories in NPs (see \citealt{Miestamo2014} for a typological survey). For instance, indefinite NPs under the scope of negation tend to be interpreted as non-referential, and in some languages this new interpretation is grammatically marked, either through explicit non-referential marking or through partitive. In a number of Bantu languages, the semantic-pragmatic category of referentiality is morphologically encoded by what Bantuists call the “augment”. The augment is a grammatical morpheme that most commonly appears as a vowel-shaped pre-prefix attached onto nouns and certain nominal modifiers. Very pervasive in Bantu, the augment is often considered as a determiner, a default marker for referentiality. Overall, augmented nouns assert the existence of the referent: ‘a/the (particular, individual) X’, ‘a (specific) X’. Augmentless nouns do not assert such existence (without, however, denying it) and tend to express ‘any, anything like, anything that can be classified as X’. It is however difficult to obtain a uniform picture on the use(s) of the augment throughout Bantu, as considerable cross-linguistic variation prevails \citep{Velde2019}. A common trend, however, is that augmentless forms tend to appear in syntactic environments where referentiality is not aimed at. The reader is referred to \citet{Blois1970} and \citet{Velde2019} for more comprehensive accounts of the Bantu augment.

In Cuwabo, the augment is not realized segmentally as typically observed in Bantu languages, but merely by means of tones, with two anchoring positions depending on the presence versus absence of a lexical H tone on nominal stems. In the absence of lexical H, the augment tone anchors onto the first mora of the stem, i.e. the mora following the noun class prefix, as in \REF{ex:cuw:70a}. In the presence of lexical H (invariably assigned to the penultimate mora or the ultimate in case of bimoraic stems), the augment tone cannot appear on the stem and instead retracts onto the noun class prefix, as shown in \REF{ex:cuw:70b}, where the lexical H is bolded. 

% \hc{LSP: the bold doesn't print well in \ref{ex:cuw:70b}, do you have a better solution?}
\TabPositions{.23\textwidth,.66\textwidth}
\ea\label{ex:cuw:70}
    \ea\label{ex:cuw:70a} Nouns without lexical H \\
    e-r{ú̲}ru \tab ‘chin’ \\
    ni-g{ó̲}goddo  \tab ‘bone’ \\
    mu-k{ó̲}lrófitto  \tab ‘loincloth’ \\
    \ex\label{ex:cuw:70b} Nouns with lexical H \\
    n{í̲}{}-ful\textbf{ó} \tab  ‘anvil’\\
    m{ú̲}{}-ká\textbf{á}tte \tab  ‘jug’\\
    {é̲}{}-sárá\textbf{á}ro \tab ‘rope’\\
    \z 
\z 

In Cuwabo, four different contexts trigger the deletion of the first H tone (HTD), i.e. the tone historically associated with the augment. Augmentless nouns appear: i) in non-verbal predication (see \sectref{sec:cuw:2.3}); ii) when used as vocatives (see examples \REF{ex:cuw:3} and \REF{ex:cuw:92}); iii) after conjoint verb forms \REF{ex:cuw:3}; and iv) under the scope of negation. For ease of reading, the three aforementioned examples \REF{ex:cuw:50}, \REF{ex:cuw:59} and \REF{ex:cuw:61} illustrating HTD after negation (context iv) are repeated below. In all these examples the postverbal object is non-referential. \textit{muttu} ‘person’ in \REF{ex:cuw:71} may denote any human being; \textit{elobo} ‘thing’ in \REF{ex:cuw:72} denotes any type of thing; and \textit{obara} ‘at the beach’ in \REF{ex:cuw:73} refers to any place that can be classified as a beach. All these items are interpreted as negatively quantified and the resulting constructions can be understood as ‘NEG-V any NP’.

\ea\label{ex:cuw:71} kaáʈíya muʈu óðówá otélá n̩lógó niʔiń̩zíwééye \\ 
    \gll \textbf{ka}{}-á-ttiy-a \textbf{mu}{}-\textbf{ttu} ó-dhow-á o-tél-a ni-lógo ni-hi-ní-ziw-á=eye \\
          \textsc{neg.sbj2-pst.ipfv-}leave{}-\textsc{fv} \textsc{htd.1}{}-person \textsc{15}{}-go-\textsc{fv} \textsc{15}{}-marry-\textsc{fv} \textsc{5}{}-family \textsc{agr}5-\textsc{neg-ipfv}{}-know-\textsc{fv.rel=pro3sg} \\
    \glt ‘One would not let someone marry some unknown family.’ (lit. ‘family that they do not know of’) [Marriage.Tradition]

\z 

{\ea\label{ex:cuw:72} míyó kaɖináálógéː ɖihiváʔíwe elobo \\
    \gll  míyo ka-ddi-náa-log{}-e ddi-\textbf{hi}{}-váh-iw-e \textbf{e}{}-\textbf{lobo}\\
          \textsc{pro1sg} \textsc{neg-sbj.1sg-fut}{}-speak-\textsc{fv} \textsc{sbj.1sg-neg-}give-\textsc{pass-sbjv} \textsc{htd.9}{}-thing \\
    \glt  ‘I will not speak if I am not given anything.’ [Elicitation]
\z }

\ea\label{ex:cuw:73} míyó kaɖináðówag’ oobara \\ 
    \gll míyo \textbf{ka}{}-ddi-ná-dhow-ag-a \textbf{o}{}-\textbf{bara}\\
          \textsc{pro1sg} \textsc{neg-sbj.1sg-cnte}{}-go-\textsc{hab-fv} \textsc{htd.17}{}-9a.sea \\
    \glt  ‘I have never been at the sea (meaning: any sea).’ [Elicitation]
\z 

However, in (most?) cases where the postverbal NP is meant as definite/specific, the augment tone remains present. In \REF{ex:cuw:74}, \textit{mwámúni} ‘husband’ is unambiguously made definite by the context and spontaneously translated with the possessive by my consultant; in \REF{ex:cuw:75}, \textit{koóbilri} ‘money’ is overtly marked for definiteness; and in \REF{ex:cuw:76}, repeated from \REF{ex:cuw:39}, \textit{múttaambí} ‘liar’ is followed by the demonstrative. All these augmented NPs clearly refer to particular individuals or entities.

{\ea\label{ex:cuw:74} mwááðíye kaáḿ̩váʔa mwámúni\\
    \gll mú-adhí=ye \textbf{ka}{}-á-mu-váh-a \textbf{mwámuni}  \\
         1-wife=\textsc{poss3sg} \textsc{neg.sbj1}-\textsc{pst.ipfv-obj}1-give-\textsc{fv} 1.husband  \\
    \glt ‘His wife did not give (it, i.e. food) to her husband.’ [Mbílri.Story]
\z }

\ea\label{ex:cuw:75} muʔinákáan’ éébaribarí munááváʔe koóbíɽiða  \\
    \gll  mu-hi-ná-kaaná e-baribari mu-\textbf{náa}{}-vah-e \textbf{koóbilri=dha}\\
           \textsc{sbj.2pl-neg-cnte}{}-have \textsc{htd}.9-truth \textsc{sbj.2pl-neg}{}-give-\textsc{sbjv} 10a.money=\textsc{def10}\\
    \glt  ‘As long as you are not sure, do not give the money.’ [Elicitation]
\z 

\ea\label{ex:cuw:76} oówóðá oʔímútelá múʈáamb’ úúĺ̩le \\ 
    \gll o-hí-wodh-á o-\textbf{hí}{}-mu-tel-á \textbf{mú}{}-\textbf{ttaambí} \textbf{ól̩le}\\
          \textsc{sbj1-pfv.dj}{}-succeed-\textsc{fv} \textsc{htd.15-neg}{}-\textsc{obj}1-marry-\textsc{fv} 1-liar 1.\textsc{dem.dist} \\
    \glt ‘She succeeded in not marrying that liar.’ [Elicitation]
\z 


Notably, the presence of a modifier is not a sufficient condition to obtain a definite/specific reading. In \REF{ex:cuw:77}, albeit \textit{guwo} ‘cloth’ is understood as belonging to a particular individual through possessive modification, no reference to a particular/specific cloth is made and in this context the NP is better interpreted as ‘any cloth of his’, hence the augment omission. 

\ea\label{ex:cuw:77} agaʔífúlege guwo ðaáyéː agáń̩nuka \\ 
    \gll  a-gaa-\textbf{hí}{}-ful-eg-e \textbf{guwo} \textbf{dhi}{}-\textbf{áye} a-gá-ni-nuk-a \\
          \textsc{sbj1}{}-\textsc{sit}{}-\textsc{neg}{}-wash{}-\textsc{hab}{}-\textsc{fv} \textsc{htd}.10a.cloth \textsc{agr}10-\textsc{poss3sg} \textsc{sbj1}{}-\textsc{fut.ipfv}{}-\textsc{ipfv.dj}{}-smell.bad-\textsc{fv}\\
    \glt ‘If he does not wash (any of) his clothes, he will smell bad.’ [Elicitation]
\z 

The absence of the augment under the scope of negation is attested in several other Bantu languages, such as Zulu (\citealt{Doke1961}; \citealt{Buell2006}), Luganda (\citealt{HymanKatamba1990}), Ruuli (Ruppert, Namyalo \& Witzlack-Makarevich this volume), and seems to always be (originally) associated with non-referentiality.

\subsection{Reinforcing negation}\label{sec:cuw:4.4}
The invariable morpheme \textit{wéene} or its agreeing cliticized form \textit{=ene} are frequent\-ly attested in Cuwabo to express an intensive (‘\textsc{ints}’) meaning. They may appear on different constituents, including verbs. When these are marked for negation, \textit{wéene} {\textasciitilde} \textit{=ene} reinforces negation and assumes an adverbial function best translated in English as ‘really’. Compare \REF{ex:cuw:78b} with \REF{ex:cuw:78a}. As a general intensifier, \textit{wéene} {\textasciitilde} \textit{=ene} also functions in affirmative contexts, e.g. to express ‘the meat is \textit{really} cooked’. Cuwabo has no dedicated adverb to reinforce negation, such as English ‘at all’. Instead the language resorts to the negative particle \textit{nne} ‘no’ (see \sectref{sec:cuw:3.1}), either before or after the verb, as shown in \REF{ex:cuw:78c}. In this context, as well as with one-word negative replies to polar questions, the borrowed variant \textit{nada} ‘no’ (< Portuguese \textit{nada} ‘nothing’) may substitute \textit{nne}. The use of \textit{nne} or \textit{nada} as reinforcers of negation remains infrequent in the language (but see \sectref{sec:cuw:4.6.4} for ironic negation).

\ea\label{ex:cuw:78}
    \ea\label{ex:cuw:78a}námá kipiyéyîle\\ 
        \gll náma \textbf{ki}{}-píy-ey-ile  \\
              9a.meat \textsc{neg.sbj9-}cook-\textsc{n-pfv}\\
        \glt ‘The meat is not cooked.’
    \ex\label{ex:cuw:78b}námá kipiyéyíle wééne\\
        \gll  náma \textbf{ki}{}-píy-ey-ile\textbf{} \textbf{wéene}\\
              9a.meat \textsc{neg.sbj9-}cook-\textsc{n-pfv} \textsc{ints}\\
        \glt ‘The meat is really not cooked.’
    \ex\label{ex:cuw:78c} námá ńnéː kipiyéyîle {\textasciitilde} kipiyéyíle ńnêː\\
        \gll  náma \textbf{nne} \textbf{ki}{}-píy-ey-ile  \\
              9a.meat no \textsc{neg.sbj9-}cook-\textsc{n-pfv} \\
        \glt ‘The meat is not cooked at all.’  [Elicitation] 
    \z
\z 

\subsection{Negation, coordination and complex clauses}\label{sec:cuw:4.5}
Negative coordinators ‘neither… nor…’ are rendered by \textit{nne… nne…} (negative counterpart of \textit{óbe} ‘or’). Most often, \textit{nne} coordinates two nouns as in \REF{ex:cuw:79}. Note that negative polarity is also obligatorily marked on the verb.

\ea\label{ex:cuw:79} kaɖiḿ̩fúná omuʈeba nne \'{E}rúgúlú, nné viínagwíya Máda \\
    \gll \textbf{ka}{}-ddi-ní-fun-á o-mu-tteb-a \textbf{nne} é-rugulu \textbf{nne} viína-gwíya má-da\\
          \textsc{neg-sbj.1}\textsc{sg}{}-\textsc{ipfv}.\textsc{cj}{}-want-\textsc{fv} \textsc{htd.15}{}-\textsc{obj1}{}-carry-\textsc{fv} nor 9-belly nor too-again 6-hand  \\
    \glt  ‘I do not want to carry neither Mr. Belly, nor Mr. Hand.’ [Body.Story]

\z 

The particle \textit{nne} may also function as a clause coordinator \REF{ex:cuw:80a}, albeit such constructions are more commonly expressed by clause juxtaposition, with optional addition of the adverb \textit{viina} ‘too’ \REF{ex:cuw:80b}. 

\ea\label{ex:cuw:80}
    \ea\label{ex:cuw:80a} kaɖiḿ̩fúnáː osuńzáː, nné ógóná\\ 
        \gll \textbf{ka}{}-ddi-ní-fun-á o-suńz-a \textbf{nne} ó-gon-á  \\
               \textsc{neg-sbj.1sg-ipfv}{}-want-\textsc{fv} \textsc{htd.15}{}-study-\textsc{fv} \textsc{neg} \textsc{15}{}-sleep-\textsc{fv}\\
        \glt  ‘I don’t want to study, nor do I want to sleep.’ [Elicitation]

    \ex\label{ex:cuw:80b} kaɖiḿ̩fúnáː osuńzáː, viínáː kaɖiḿ̩fúnáː ogoná \\
        \gll  \textbf{ka}{}-ddi-ní-fun-á o-suńz-a viina \textbf{ka}{}-ddi-ní-fun-á o-gon-á\\
               \textsc{neg-sbj.1sg-ipfv}{}-want-\textsc{fv} \textsc{htd.15}{}-study-\textsc{fv} too \textsc{neg-sbj.1sg-ipfv}{}-want-\textsc{fv} \textsc{htd.15}{}-sleep-\textsc{fv}\\
        \glt  ‘I don’t want to study, nor do I want to sleep.’ [Elicitation]
    \z
\z 

\subsection{Miscellaneous aspects of negation}\label{sec:cuw:4.6}
\subsubsection{Negative transport}\label{sec:cuw:4.6.1}
Negative transport, whereby negation formally marked in a first (higher) clause semantically applies to the second (lower) clause predicate (see, e.g., \citealt{Lakoff1969}; \citealt{Lindholm1969}; \citealt{Horn1989}), does occur in Cuwabo with different kinds of predicates. Starting with verbs of feeling, the raised version ‘I don’t hope he comes’ in \REF{ex:cuw:81b} is semantically tantamount to ‘I hope he does not come’ in \REF{ex:cuw:81a}, where negation is in the lower clause.

\ea\label{ex:cuw:81}
    Context:  ‘Oh no, your husband is annoying!’ %yiii, mámúnááwo onorúcá vaɖíɖi!
    \ea\label{ex:cuw:81a} ɖinojééðéla wi kanááðéwo\\ 
        \gll  ddi-ni-o-jéedh-el-a wi \textbf{ka}{}-náa-dh-e=wo \\
               \textsc{sbj.1sg-ipfv-15}{}-hope-\textsc{appl-fv} \textsc{comp} \textsc{neg.sbj1-fut-}come-\textsc{fv=loc17} \\
        \glt  ‘I hope he doesn’t come.’ [Elicitation]

    \ex\label{ex:cuw:81b} kaɖin̩jééðéla wi onoóðáwo \\
        \gll  \textbf{ka}{}-ddi-ni-jéedh-el-a wi o-ni-ó-dh-a=wo \\
             \textsc{neg-sbj.1sg-ipfv-}hope-\textsc{appl-fv} \textsc{comp} \textsc{sbj1-ipfv-15-}come-\textsc{fv}\textsc{=loc17}  \\
        \glt  ‘I don’t hope he comes.’
    \z
\z 

The same cases of anticipated negation are observed with verbs of volition \REF{ex:cuw:82} and verbs of opinion \REF{ex:cuw:83}.
\newpage
{\ea\label{ex:cuw:82}
    Context:  ‘I appreciate your company.’ %ɖinozívéliwa okálá na wéyo
    \ea\label{ex:cuw:82a} ɖiḿ̩fúná oʔiðowe\\ 
        \gll  ddi-ní-fun-á o-\textbf{hi}{}-dhow-e \\
              \textsc{sbj.1sg-ipfv.cj}{}-want-\textsc{fv} \textsc{sbj.2sg-neg}{}-go-\textsc{sbjv.htd} \\
        \glt  ‘I want you not to go.’ [Elicitation]

    \ex\label{ex:cuw:82b} kaɖiḿ̩fúná wí oðówe\\
        \gll  \textbf{ka}{}-ddi-ní-fun-á wí o-dhów-e\\
               \textsc{neg-sbj.1sg-ipfv}{}-want-\textsc{fv} \textsc{comp} \textsc{sbj.2sg-}go-\textsc{sbjv}\\
        \glt  ‘I don’t want you to go.’ 
    \z
\z }

\ea\label{ex:cuw:83}
    Context: ‘It is late.’ %vaaríba. 
    \ea\label{ex:cuw:83a} n̩nóóná wí kanááðewo\\ 
        \gll ddi-ní-on-á wí \textbf{ka}{}-náa-dh-e=wo \\
               \textsc{sbj.1sg-ipfv.cj}{}-see-\textsc{fv} \textsc{comp} \textsc{neg.sbj1-fut-}come-\textsc{fv=loc17}\\
        \glt  ‘I think they are not coming.’ [Elicitation]

    \ex\label{ex:cuw:83b} kaɖinóónámó wí anoóða \\
        \gll \textbf{ka}{}-ddi-ní-on-á=mo wí a-ni-ó-dh-a \\
               \textsc{neg-sbj.1sg-ipfv}{}-see-\textsc{fv=loc18} \textsc{comp} \textsc{sbj1-ipfv-15-}come-\textsc{fv}\\
        \glt  ‘I don’t think they are coming.’
    \z
\z 

Cuwabo has no verbs of perception such as ‘seem’, ‘appear’, ‘look like’, ‘sound like’, or ‘feel like’. Instead, the comparison adverb \textit{nínga} ‘like, as’ is used, as illustrated in \REF{ex:cuw:84}. 

\ea\label{ex:cuw:84} Context: ‘This land is dry.’\\% Maʈáká ábá aʔúúma.
        nínga karúbíle muzógwe \\ 
    \gll  \textbf{nínga} \textbf{ka}{}-rúb-ile mu-zógwe\\
          like \textsc{neg.sbj1}{}-rain-\textsc{pfv} 1-rain\\
    \glt ‘It seems it did not rain.’ {\textasciitilde} ‘It does not seem it rained.’ [Elicitation]
\z 

\subsubsection{Metalinguistic negation}\label{sec:cuw:4.6.2}
Metalinguistic negation is a marked use of negation conveying reactive and corrective speech act. In \citeauthor{Horn1989}’s terms (1989: 363), it consists in “objecting to a previous utterance on any grounds whatever, including the conventional or conversational implicata it potentially induces, its morphology, its style or register, or its phonetic realization”. Typically, an utterance involving metalinguistic negation is followed by a rectifying assertion that not only contrasts with the corrected segment, but also provides or clarifies the motivation for the objection. In Cuwabo, metalinguistic negation does not differ from ordinary/descriptive negation employed on verbs. In \REF{ex:cuw:85}, the verb stem \textit{ziveliwa} ‘like’ used twice is negated in the first clause, then repeated in the second clause, but affirmatively and along with the emphatic adverb \textit{vaddíddi} ‘much’. This metalinguistic use of negation marks a reassessment supported by an overbid: the speaker does more than like coconut water: s/he loves it, literally translated as ‘like very much’. 

\ea\label{ex:cuw:85} míyó kaɖin̩zívéliwa n̩lanya, ɖinozívéliwa vaɖíííɖi! \\ 
    \gll  míyo \textbf{ka}{}-ddi-ni-zíveliw-a ni-lanya ddi-ni-o-zíveliw-a vaddíddi \\
           \textsc{pro1sg} \textsc{neg-sbj.1sg-ipfv}{}-like-\textsc{fv} \textsc{htd.5}{}-coco.water \textsc{sbj.1sg-ipfv.cj-15-}like-\textsc{fv} much\\
    \glt  ‘I don’t like coconut water, I love it!’ [Elicitation]
\z 

\subsubsection{Phasal polarity}\label{sec:cuw:4.6.3}
Phasal polarity in \citegen{Kramer2021} terms involves two positive and two negative concepts. The negative ones are ‘not yet’ and ‘no longer’. The first combines the counter-expectational prefix \textit{ná}{}- with one of the two main negators, \textit{ka}{}- in declarative verbal main clauses (\sectref{sec:cuw:2.1}) and \textit{hi}{}- in temporal adverbial clauses (\sectref{sec:cuw:2.4.2}). The second concept ‘no longer’ marks the end of a situation which used to hold before the changing point. In Cuwabo, it is expressed by means of the adverb \textit{viina} ‘too, as well’ used together with standard negation, as exemplified in \REF{ex:cuw:86}. See \citet{Guérois2021} for a detailed analysis of phasal polarity in Cuwabo.

\ea\label{ex:cuw:86} n̩níngó neetéêne kaneéddíle víîna \\  
    \gll  ni-níngo ni-eté=ene \textbf{ka}{}-ni-édd-ile \textbf{víina}  \\
          5-body 5-all=\textsc{ints} \textsc{neg-sbj}5-walk-\textsc{pfv} too \\
    \glt ‘The whole body no longer worked (lit. ‘walked’).’ [Body.Story]
\z 

\subsubsection{Non-negative uses of negatives}\label{sec:cuw:4.6.4}
It is common in discourse that negative verbal and non-verbal forms are used to express emphatic affirmative clauses. This deliberate use of language which states the opposite of the truth serves the communicative goal of irony. When applied to negative constructions, this rhetoric device accentuates affirmative/positive connotation. Resulting constructions read as negative statements but in actual fact mean the opposite. In \REF{ex:cuw:87}, ironic negation is used with the negative copula \textit{kayiye}, supplemented with the negative particle \textit{não} (see \sectref{sec:cuw:3.1}), although \textit{nne} ‘no’ is also possible in this context. 

\ea\label{ex:cuw:87}
    \begin{xlist}[A: ]
    \exi{A: }\label{ex:cuw:87A} oótélá. áyáná oná gwéeː! \\ 
         \gll o-hí-tel-á á-yaná o-ná gwée\\
          \textsc{sbj1-pfv.dj}{}-marry-\textsc{fv} 2-woman \textsc{sbj1}{}-have many \\
    \glt ‘He got married. Women, he has lots of them.’
    \exi{B: }\label{ex:cuw:87B}
        \gll yii!\\ 
              \textsc{inj}\\
        \glt (Surprise)
    \exi{A: }\label{ex:cuw:87A2} kayíye áyaná não. niimwálána\\ 
         \gll \textbf{kayíye} á-yaná \textbf{não} ni-hi-mwálan-a\\
          \textsc{nvb.neg} 2-woman \textsc{neg} \textsc{sbj.1pl-pfv.dj}{}-break.up-\textsc{fv}\\
        \glt ‘He has very many women. We broke up already.’ (lit. ‘There are no women no.’) [Conversation.7]
    \end{xlist}
\z 

According to my main consultant, the presence of the negative particle is obligatory. \textit{Kayíye áyaná} would be interpreted literally as ‘There are no women (but only men, for instance)’. Double negation, originally used as a reinforcing device (see \sectref{sec:cuw:4.4}), thus seems to have semantically evolved into an emphatic affirmative marker. However, the two ironic negative constructions in \REF{ex:cuw:88} and \REF{ex:cuw:89} do not exhibit double negation. Albeit my main consultant still prefers the presence of a final negative particle in these contexts, these two examples, produced by two different Cuwabo story tellers, clearly illustrate that double negation is not a mandatory condition to convey ironic negation.

\ea\label{ex:cuw:88} áʈú kayíyo óðów’ óoʈólon’ uúkúle kayíy’ óópítáála\\ 
    \gll á-ttu \textbf{kahíyo} \textbf{ó}{}-\textbf{dhow}{}-\textbf{á} o-ttólo=ni ókule \textbf{kahíyo} \textbf{ó}{}-\textbf{pitaál}{}-\textbf{a}\\
          2-person \textsc{nvb.neg} \textsc{15-}go\textsc{{}-fv} 17-well=\textsc{loc} 17.\textsc{dem.dist} \textsc{nvb.neg} \textsc{15-}hide\textsc{{}-fv} \\
    \glt  ‘People went to that well and hid themselves.’  (lit. ‘People are not to go to that well, are not to hide.’) [Ddoolrinddo.Story]
\z 

\ea\label{ex:cuw:89} rapáríga ɖabunó kayíy’ óokálá ðiiðoː \\ 
    \gll  rapáriga ddabunó \textbf{kahíyo} \textbf{o}{}-\textbf{kál}{}-\textbf{a} dhiidho \\
           1a.girl\textsc{} then \textsc{nvb.neg} \textsc{15-}be-\textsc{fv} naked \\
    \glt  ‘Then the mistress remained naked.’ (lit. ‘The mistress then is not to be naked’) [Mbílri.Story]
\z 

The conditions for the presence of the negative particle \textit{nne} {\textasciitilde} \textit{não} are equally unclear with verbal predicates. While it is considered necessary in \REF{ex:cuw:90}, which is semantically equivalent to the affirmative \textit{ddinozívéliwa vaddíííddi} ‘I like [it] a lot’ in \REF{ex:cuw:85} above, its addition in both \REF{ex:cuw:91} and \REF{ex:cuw:92} is unaccepted, even by my main consultant. 

\ea\label{ex:cuw:90} kaɖínóziveliwa ńněː!\\
    \gll \textbf{ka}{}-ddí-ni-o-ziv-el-iw-a \textbf{nne}  \\
           \textsc{neg-sbj.1sg-ipfv-15}{}-be.tasty-\textsc{appl}{}-\textsc{pass-fv} \textsc{neg} \\
    \glt  ‘[I don’t like coconut water,] I love it!’ (lit. ‘I don't like/love it no’) [Elicitation]
\z 

\ea\label{ex:cuw:91}
    \ea\label{ex:cuw:91a} ɖaáɖúwaála kí mááɲjé kâgubúl̩léː, maŋgánɖwéyá kaɖaareyíle. \\ 
        \gll  ddi-a-hí-dduwaál-a kí má-anjé \textbf{ka}{}-\textbf{gubúl}{}-\textbf{ile} (\textbf{*nne}) ma-ngánddwe=ya ka-ddi-a-rey-íle \\
               \textsc{sbj.1sg-pst-pfv.dj}{}-forget-\textsc{fv} \textsc{emph} 6-water \textsc{neg.sbj6-}rise-\textsc{pfv} \textsc{neg} 6-fish.trap=\textsc{def} \textsc{neg-sbj.1sg-pst}{}-set-\textsc{pfv} \\
        \glt  ‘I had forgotten, the tide rose (lit. ‘the tide didn't rise') and I had not set up the fish traps.’ [Elicitation]

    \ex\label{ex:cuw:91b} Affirmative alternative: \\
            mááɲjé aágúbúla\\
        \gll má-anjé a-hí-gubúl-a \\
             6-water \textsc{sbj6-pfv.dj}{}-rise-\textsc{fv} \\
        \glt ‘the tide rose' 
    \z
\z 

\ea\label{ex:cuw:92}
    \ea\label{ex:cuw:92a} m̩bwaná! múgóɖo, en̩zívéliwééye! woóne: kimáĺ̩le!  \\ 
        \gll   mu-bwaná mú-goddo e-ni-zível-iw-a=íye o-ón-e \textbf{ki-mál-ile} (\textbf{*nne}) \\
              \textsc{htd.1}{}-friend 3-rice.cake \textsc{agr}9-\textsc{ipfv}{}-please-\textsc{pass-fv.rel=pro3sg} \textsc{sbj.2sg}{}-see-\textsc{sbjv} \textsc{neg.sbj9}{}-finish-\textsc{pfv} \textsc{neg} \\
        \glt  ‘Friend! The rice cake, it’s something he likes! Look, it’s all finished!’ (lit. ‘Look, it's not finished') [Conversation.2]

    \ex\label{ex:cuw:92b} Affirmative alternative: \\
            woóne: eemála! \\
        \gll  o-ón-e \textbf{e}{}-\textbf{hi}{}-\textbf{mál}{}-\textbf{a}  \\
              \textsc{sbj.2sg}-see-\textsc{sbjv} \textsc{sbj9-pfv.dj}{}-finish-\textsc{fv} \\
        \glt ‘Look, it's finished!'
    \z
\z 

The examples above show that the presence or absence of the negative particle may in certain cases be indicative of intra-speaker variation (examples \REF{ex:cuw:88} and \REF{ex:cuw:89}). In other cases, it may reveal an ongoing process whereby the negative particle would preferably be added to avoid any pragmatic ambiguity (examples \REF{ex:cuw:87} and \REF{ex:cuw:90}). However, it is difficult at this stage to understand why examples such as \REF{ex:cuw:91} and \REF{ex:cuw:92} preclude the presence of the negative particle. More research is needed.

Ironic negative constructions are not freely expressed with any TAM category. In addition to already-illustrated habitual present \REF{ex:cuw:90} and perfective (\ref{ex:cuw:91}--\ref{ex:cuw:92}), only present progressive \REF{ex:cuw:93a}, past progressive \REF{ex:cuw:94a} and counter-expectational \REF{ex:cuw:95a} seem to authorize ironic negative constructions. On the other hand, negation used in past imperfective \REF{ex:cuw:96a} and future \REF{ex:cuw:97a} may only be interpreted as face value, and the expression of an emphatic positive statement may only obtain through affirmative polarity, as seen in \REF{ex:cuw:96b} and \REF{ex:cuw:97b}. Further research is required to figure out exactly how ironic negation and TAMs do interact. It may be that the future, as an irrealis, blocks the effect of the pragmatic device. The restriction observed with the past imperfective remains, however, more puzzling. All the examples below were elicited.

{\ea\label{ex:cuw:93}
    \ea\label{ex:cuw:93a} João kánózivala ńnêː! \\ 
        \gll  João \textbf{ká}{}-ni-o-zival-a \textbf{nne} \\
               João \textsc{neg.sbj1-ipfv-15}{}-lie\textsc{{}-fv} \textsc{neg} \\
        \glt  ‘João is lying a lot!’ (lit. ‘John is not lying no’)
    \ex\label{ex:cuw:93b} João onozívála vaɖíɖi! \\
        \gll João o-ni-o-zival-a vaddíddi \\
            João \textsc{sbj1-ipfv-15}-lie-\textsc{fv} much \\
        \glt ‘João is lying a lot!’ (Affirmative alternative)
    \z
\z
}

\ea\label{ex:cuw:94}
     \ea\label{ex:cuw:94a} João kéél’ ozivala ńnêː!  \\
        \gll  João \textbf{ka}{}-á-ila o-zival-a \textbf{nne}  \\
               João \textsc{neg.sbj1-pst.ipfv-aux} \textsc{htd.15}{}-lie\textsc{{}-fv} \textsc{neg} \\
        \glt  ‘João was lying a lot!’ (lit. ‘John wasn’t lying no’)
    \ex\label{ex:cuw:94b} João wééloozívâla vaɖíɖi! \\
        \gll João ó-a-ila o-zíval-a vaddíddi \\
            João \textsc{sbj1-pst.ipfv-aux} 15-lie-\textsc{fv} much \\
        \glt ‘João was lying a lot!’ (Affirmative alternative)
    \z
\z
\newpage
{\ea\label{ex:cuw:95}
  \ea\label{ex:cuw:95a}João keél’ ózivala ńnêː. \\
        \gll  João \textbf{ka}{}-íla o-zival-a \textbf{nne} \\
            João \textsc{neg.sbj1-aux} \textsc{htd.15-}lie-\textsc{fv} \textsc{neg} \\
        \glt  `João (unexpectedly) lied.’ (lit. ‘John didn’t (unexpectedly) lie no’)
    \ex\label{ex:cuw:95b} João olóózíválá vaɖíɖi! \\
        \gll João o-íla o-zival-a vaddíddi \\
            João \textsc{sbj1-aux} \textsc{htd}.15-lie-\textsc{fv} much \\
        \glt `João (unexpectedly) lied.’ (Affirmative alternative)
    \z
\z }
% \lv[inline,size=\scriptsize,color=blue!40]{@LSP: please make sure that single quotes are shown correctly as opening and closing. For instance, for the literal translation of example \REF{ex:cuw:95}, they look like they do in the code but not in the compiled version. Rozenn is asking for the opening single quote to be reverted.}
{\ea\label{ex:cuw:96}
    \ea\label{ex:cuw:96a}João kaázívalaga ńnêː! \\ 
        \gll  João \textbf{ka}{}-á-zival-ag-a \textbf{nne}  \\
               João \textsc{neg.sbj1-pst.ipfv}{}-lie-\textsc{hab-fv} \textsc{neg} \\
        \glt  *‘João lied/used to lie a lot!’ → ‘João did not lie.’

    \ex\label{ex:cuw:96b} João wáán̩zivalaga vaɖíɖi! \\
        \gll João ó-a-ni-zival-ag-a vaddíddi \\
              João \textsc{sbj1-pst-ipfv}{}-lie-\textsc{hab-fv.dj} much  \\
        \glt  ‘João lied/used to lie a lot!’
    \z
\z }

\ea\label{ex:cuw:97}
    \ea\label{ex:cuw:97a} João kanáázíválége ńnêː!\\
        \gll  João \textbf{ka}{}-náa-zival-eg-e \textbf{nne} \\
               João \textsc{neg.sbj1-fut}{}-lie-\textsc{hab-fv} \textsc{neg} \\
        \glt  *‘João will lie a lot!’ → ‘João will never lie.’

    \ex\label{ex:cuw:97b} João agáń̩zivala vaɖíɖi.\\
        \gll João a-gá-ni-zival-a vaddíddi \\
               João \textsc{sbj1-fut.dj-ipfv}{}-lie-\textsc{fv} much \\
        \glt  ‘João will lie a lot!’

    \z
\z 

While ironic negative constructions such as ‘he’s not very bright’ used in the sense ‘he’s stupid’ may easily be perceived as a common rhetoric device cross-linguistically (albeit they have not been the subject of much investigation), ironic negative constructions such as those found in Cuwabo, which point to an emphatic affirmative state of affairs reversing a literally negative phrasing (‘he’s not bright’ = ‘he’s very bright’), are certainly more exceptional. Similar constructions exist in French, as illustrated in \REF{ex:cuw:98}, but these represent very colloquial expressions, where: i) the innovated negative particle \textit{pas} is the single negator in the sentence, as commonly heard in verbal speech; ii) informal contractions are overtly pronounced (\textit{chuis} < \textit{je suis}, \textit{t’es} < \textit{tu es}, \textit{z’êtes} < \textit{vous êtes}); iii) very informal words are used (\textit{merde}, \textit{chiant}, \textit{veinard}); iv) a specific prosody applies, which unambiguously underlines irony. 

\ea\label{ex:cuw:98}
    \ea\label{ex:cuw:98a}
        \gll Chuis pas dans la merde! \\
             \textsc{pro1sg.}be.\textsc{prs} \textsc{neg} in \textsc{def} shit  \\
        \glt  ‘I am in big trouble!’ (lit. ‘I’m not in trouble.’) 

    \ex\label{ex:cuw:98b}
        \gll T’es pas chiant! \\
             \textsc{pro2sg.}be.\textsc{prs} \textsc{neg} boring \\
        \glt  ‘You’re boring!’ (lit. ‘You’re not boring.’)

    \ex\label{ex:cuw:98c}
        \gll Z’êtes pas des veinards!\\
              \textsc{pro2pl.}be.\textsc{prs} \textsc{neg} \textsc{indf}.\textsc{pl} lucky.person.\textsc{pl} \\
        \glt  ‘You’re lucky people!’ (lit. ‘You’re not lucky people.’)
    \z
\z 

In contrast, Cuwabo ironic negative constructions do not differ formally, prosodically and stylistically from truth-functional negative constructions. Only situational factors may help in realizing that~the negative sentence is ironic. These are clear cases whereby the meaning of a sentence cannot be exhaustively interpreted on the basis of the literal meaning of its constituents.

\subsubsection{Diachronic observations}\label{sec:cuw:4.6.5}
Diachronically, the two pre-stem negators \textit{ka}{}- and \textit{hi}{}- have been reconstructed to Proto-Bantu as \textit{*ka}{}- and \textit{*ti}{}-, respectively \citep{Meeussen1967}. The final vowel -\textit{i}, not uncommon through (Eastern) Bantu, has also been observed by \citet{Meeussen1967}, but not reconstructed. The post-initial \textit{náa}{}- (restricted to the subjunctive mood in Cuwabo) is most likely related to the form \textit{na}{}-, which is a recurrent negative marker in zone P languages \citep[115]{KambaMuzenga1981}, i.e. Bantu languages spoken in south Tanzania and north Mozambique. 

Both non-verbal negators \textit{kahíye} and \textit{ka}{}-\textsc{sbj}{}-\textit{hí}=\textsc{loc}, discussed in \sectref{sec:cuw:2.3}, are based on the initial negative marker \textit{ka}{}-. The use of standard negation places Cuwabo onto stage A of the model of the Negative Existential Cycle as set out by \citet{Croft1991} and retaken in \citet{Veselinova2013neg.exist,Veselinova2014,Veselinova2016}. This conforms to the tendency observed across Bantu, where the majority of existential negators belong to the earlier stages of the cycle \citep{BernanderEtAl2022}, as well as to the cross-linguistic generalizations provided in Croft’s and Veselinova’s typologies.

\section{Summary}\label{sec:cuw:5}

\tabref{tab:cuw:7} summarizes all the negative constructions of Cuwabo discussed in this chapter, as well as the contexts in which they appear. 

% \hc{@LSP could you find a way to make it clear that some lines belong together (echo system + negative particles both belong to negative replies in the second column, same with the next two lines.}

\begin{table} %new table with separate column for section

\begin{tabularx}{\textwidth}{lQp{13mm}}
\lsptoprule
{Negator\slash Construction} & {Function} & Section\\
\midrule
{\textit{ka}{}-} & {standard negation } & {\sectref{sec:cuw:2.1}}\\
\tablevspace
{\textit{ka-} (with subjunctive)} & {non-declarative main clauses } & {\sectref{sec:cuw:2.2}}\\
\tablevspace
{\textit{na\'{} a}{}- (with subjunctive)} & {non-declarative main clauses } & {\sectref{sec:cuw:2.2}} \\
&{complement clauses } & {\sectref{sec:cuw:2.4.1}}\\
\tablevspace
{\textit{kahíyo} {\textasciitilde} \textit{kahíye}} & {non-verbal predication } & \sectref{sec:cuw:2.3.1} \& \sectref{sec:cuw:2.3.2}\\
\tablevspace
{\textsc{neg} + copula verb \textit{li}}  & {non-verbal predication } & {\sectref{sec:cuw:2.3}}\\
\tablevspace
{\textit{ka}{}-\textsc{sp}{}-\textit{hí}=\textsc{loc}} & {negative locative-existential } & {\sectref{sec:cuw:2.3.3}}\\
\midrule
{\textit{hi}{}-}  & {non-main clauses (default marker) } & {\sectref{sec:cuw:2.4}}\\
\tablevspace
{\textit{hi}{}- + \textsc{fv} negator {}-\textit{i}}  & {situative TA category } & {\sectref{sec:cuw:2.4.2}}\\
\midrule
{echo system} & \multirow{2}{*}{negative replies } & {\sectref{sec:cuw:3.1}}\\
{negative particles \textit{nada}{\textasciitilde}\textit{nne}{\textasciitilde}\textit{não}} &  & \\
\midrule
{generic noun + \textsc{neg-v}} & \multirow{2}{*}{negative indefinites } & {\sectref{sec:cuw:3.2}}\\
{existential construction + free relative} &  & \\
\midrule
{adposition \textit{sé} ‘without’} &  {negative derivation } & {\sectref{sec:cuw:3.3}}\\
\tablevspace
{\textit{koodda} ‘refuse’ + infinitive} & {negative polarity } & {\sectref{sec:cuw:4.2}}\\
{drop of augment} & {marking of NPs after negation } & {\sectref{sec:cuw:4.3}}\\
\tablevspace
{adverb \textit{weéne} or particle \textit{nne}} & {reinforcing negation } & {\sectref{sec:cuw:4.4}}\\
\tablevspace
{\textit{nne… nne…}} & {negative coordinators } & {\sectref{sec:cuw:4.5}}\\
\lspbottomrule
\end{tabularx}
\caption{Summary of Cuwabo negative items/constructions}
\label{tab:cuw:7}
\end{table}

%\begin{table}%old table
%\begin{tabularx}{\textwidth}{lX}
%\lsptoprule
%{Context} & {Negator/Construction}\\
%\midrule 
%{Standard negation} & {\textit{ka}{}-} \\
%{Non-declarative clauses} & {\textit{hi}{}- or \textit{naa}{}-}\\
%{Non-verbal clauses} & {negative existential \textit{kahíyo} {\textasciitilde} \textit{kahíye}}\\
% & {negative locative-existential \textit{ka}{}-\textsc{sbj}{}-\textit{hí}=\textsc{loc}}\\
% & {\textsc{neg} + copula verb \textit{li}} \\
%{Non-main clauses} & {\textit{hi}{}- (by default)}\\
% & {\textit{hi}{}- or \textit{naa}{}- (with subjunctive)}\\
% & {\textit{hi}{}- + \textsc{fv} {}-\textit{i} (with situatives)}\\
%{Negative replies} & {Primary strategy: echo system} \\
% & {Secondary strategy: negative particles \textit{nada} {\textasciitilde} \textit{nne} {\textasciitilde} \textit{não}}\\
%{Negative indefinites} & {Primary strategy: generic noun + \textsc{neg-v}} \\
% & {Secondary strategy: existential construction + free relative}\\
%{Negative derivation} & {only adposition \textit{sé} ‘without’}\\
%{Scope of negation} & {verb-dependent} \\
%{Negative polarity} & {\textit{koodda} ‘refuse’ + infinitive}\\
%{Marking of NPs after negation} & {drop of augment}\\
%{Reinforcing negation} & {adverb \textit{weéne} or particle \textit{nne}} \\
%{Negative coordinators}  & {\textit{nne… nne…} }\\
%\lspbottomrule
%\end{tabularx}
%\caption{Summary of Cuwabo negative items/constructions}
%\label{tab:cuw:7a}
%\end{table}

Two areas of Cuwabo negation in particular require further research. First, it would be interesting to dig deeper into the connection between negation and the marking of referentiality on the postverbal constituent through the presence or absence of the augment tone, and then relate the results with other grammatical contexts involving augmentless forms. The existence of ironic negative constructions in order to postulate an emphatic affirmative statement is a second fascinating aspect of Cuwabo negation. This pragmatic device is recurrently attested in natural speech, and does not appear as colloquial as ironic negation observed in a language like French. Particularly interesting in these constructions is the possible role played by double negation, as well as the interaction with TAM categories.

\section*{Acknowledgements}
I am grateful to the audience of the Workshop on Negation across Languages of the 8\textsuperscript{th} Conference on Syntax of the World’s Language (SWL8). I would also like to extend my gratitude to Matti Miestamo and Ljuba Veselinova as editors as well as the anonymous reviewers who provided helpful feedback and comments on earlier versions of the chapter. 







\section*{Abbreviations}
\begin{tabularx}{.54\textwidth}{lQ}
1, 2, 3, etc & noun class \\
 1a, 5a, 9a, 10a & subclasses of classes 1, 5, 9, 10\\
\scshape 1sg & first person singular \\ 
\scshape 2sg & second person singular \\ 
\scshape 3sg & third person singular \\ 
\scshape 1pl & first person plural \\ 
\scshape 2pl & second person plural\\ 
\scshape 3pl & third person plural \\ 
\scshape addr & addressee \\
\scshape agr & agreement prefix\\
\scshape amot & associated motion \\
\scshape appl & applicative \\
\scshape aux & auxiliary  \\
\scshape caus & causative \\
\scshape cnte & counter-expectational \\
\scshape cntf & counterfactual \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape cj & conjoint \\
\scshape com & comitative \\
\scshape comp & complementizer \\
\scshape conn & connective \\
\scshape cop & copula \\
\scshape def & definite\\
\scshape dem & demonstrative \\ 
\scshape dist & distal  \\ 
\scshape dj & disjoint \\
\scshape emph & emphatic \\
\scshape fut & future \\
\scshape fv & final vowel \\
\scshape h & high tone \\
\scshape htd & (first) high tone deletion \\
\scshape hab & habitual \\
\scshape hyp & hypothetical \\
\scshape imp & imperative \\
\end{tabularx}

\begin{tabularx}{.45\textwidth}{lQ}
\scshape indf & indefinite \\
\scshape inf & infinitive \\
\scshape ints & intensive/intensifier \\
\scshape inj & interjection \\
\scshape ipfv & imperfective \\
\scshape loc & locative \\
\scshape narr & narrative \\
\scshape neg & negative \\
\scshape negex & existential negation \\ 
\scshape n & neuter \\
\scshape nvb & non-verbal \\
\scshape obj & object prefix \\
\scshape pass & passive \\
\scshape pfv & perfective \\
\scshape pl & plural \\
\scshape poss & possessive\\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape prf & perfect \\
\scshape pro & pronoun \\
\scshape prox & proximal \\ 
\scshape prs & present \\
\scshape pst & past \\
\scshape recp & reciprocal \\
\scshape ref & referential \\ 
\scshape rel & relative \\
\scshape restr & restrictive \\
\scshape sbj & subject prefix \\
\scshape sbjv & subjunctive \\
\scshape seq & sequential  \\
\scshape sit & situative \\
\scshape ta(m) & tense-aspect(-mood)\\
\scshape vb & verbal base \\
\\
\end{tabularx}


\sloppy\printbibliography[heading=subbibliography,notkeyword=this]

\end{document} 
