\documentclass[output=paper,colorlinks=true, citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233772}
\author{Paulette Roulon-Doko\affiliation{CNRS-LLACAN}}
\title{Negation in Northwest Gbaya}
\abstract{In Northwest Gbaya there is a single dedicated negative marker \textit{ná} which is reconstructed in Proto-Gbaya. It is placed after the verb, in the very last position of the verb phrase. For standard negation, negative verbal constructions are symmetric in realis mood, as the negator is simply added with no changes, but there is a paradigmatic asymmetry as perfective forms are blocked in the negative. The paradigm of the imperfective forms is narrowed to three forms. In the virtual mood, both perfective and imperfective  forms carry the negator, but these paradigms are narrowed to the two neutral forms. Attributive verbal predication can be negated but this is not frequent. In the command mood, the construction is the same as in Realis mood, but there are two new forms dedicated to negative commands (vetitive and prohibitive). All non-verbal predications (identificational, locative, existential and presentative) add the potential verb \textit{be} ‘can be’ when negated. This potential verb carries the negator which is placed clause-finally. Finally, there is a constituent negator to negate any element except a predicate.

\keywords{negation, verbal predication, non-verbal predication, Niger-Congo, Gbaya  }}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 

\section{The language}\label{sec:gba:1}

Northwest Gbaya (ISO: \href{https://iso639-3.sil.org/code/gya}{gya}, Glottocode: \href{https://glottolog.org/resource/languoid/id/nort2775}{nort2775}) is part of the main linguistic group Gbaya-Mandja-Ngbaka, an Ubangian language family (a branch of the Niger-Congo phylum, Adamawa Ubangi). It is spoken in the northwest of the Central African Republic (CAR) and in the central{}-eastern part of Cameroon. It is subdivided in six dialects: four in CAR, referred to as [ɓòdòè, ɓòkpàn, ɓòɲìnà, ɓùgùì], and usually collectivley named Gbaya-Kara, and two in Cameroon [ɓòyà] and [yàáyùwèè]. In 1996, there were 265,000 speakers: 200,000 in the western part of CAR and 65,000 in the central-eastern part of Cameroon. This paper, which deals with the Gbáyá \MakeUppercase{ɓ}òdòè, a Northwest Gbaya dialect spoken in the Central African Republic, is based on a corpus of spontaneous speech collected in the field between 1970 and 2013. The statistics are drawn from my current 4.5-hour oral corpus.
%\lv{This might not be idiomatic English. Would be possible to have the paper proofread by a native speaker of English prior to the next submission?}\aut{change to: is based on a corpus of spontaneous speech collected in the field between 1970 and 2013. The statistics are drawn from my current 4.5-hour oral corpus.} of~4h30 oral corpus).

Gbaya is the mother-tongue of all the inhabitants of the region under study. It is used in the home and for all interactions of daily life. Most of the men and women can speak Sango, the national and official language of CAR, but it is not used in everyday life. They can also speak some Fulfulde when meeting Mbororo people (nomadic cattle-herders, a subgroup of the Fulani ethnic group). Gbaya is mainly a spoken language, although there is a Bible translation in the Yaayuwee dialect (Glottocode \href{https://glottolog.org/resource/languoid/id/yaay1238}{yaay1238}).

Gbaya verbs are always the expression of a process since there are no verbs of quality. Verbs have no lexical tones, so tones on the verb are grammatical elements. Depending on the aspect-mood (AM) constructions, the verb attests a bare form {\textsc{vb}} and two bound forms \textsc{vb}\textit{i} and \textsc{vb}\textit{a}, which will be simply glossed as \textsc{vb} in the word-for-word examples. The Gbaya verbal paradigm consists of 21 constructions whose semantic (or conceptual) domain is structured on two axes, mood and aspect (AM). Verbs do not indicate temporal reference. There are three moods (realis, virtual and command) and two aspects (perfective and imperfective). The AM constructions are complex forms including a tonal scheme. Most often they appear in pairs combining an imperfective form and a perfective form, depending on the chosen perspective. Furthermore, some AM constructions (in italics in \tabref{tab:gba:1}) are linked forms that identify the clause they predicate as a subordinate clause without any subordinate conjunction. 

In the glosses, the names of conjugated forms (AM constructions) are most often compound nouns, the mention of the aspect being systematically added for all those which have an aspectual opposition (\textsc{prog.ipfv}, \textsc{prog.pfv}, \textsc{vrt.ipfv}, \textsc{vrt.pfv}, etc.). Only the neutral forms of the realis mood are reduced to the mention of the aspect (\textsc{ipfv} and \textsc{pfv}), whereas the mention of the aspect is not indicated for those which have only one aspectual form (\textsc{know}, \textsc{hypo}, \textsc{cons}, \textsc{inch}).
Additionally, note that AM is sometimes marked by tone only, either on the verb, or on both the verb and preceding personal pronoun.
%
% \hc{27.11.2024: We suggest to add a paragraph here with explanations of what the different kinds of parentheses are used for, for clarity to the readers}\hc{author's answer is now in the footnote, to be checked, and some things to be deleted I guess.}\footnote{%\textbf{In the article's text}, parentheses (x) are used to indicate a point providing clarification, such as cf. p. vii § after (7) or p. xxiv after (80), to introduce a cross-reference or citation, such as cf. p. vi before (4), and to enclose example numbers (x).

% \textbf{In the translation line of an example}, parentheses (…) are used
% \begin{itemize}
%     \item to enclose the literal meaning (lit. ‘XXX’) (4) (5)(43) (44) (47) (48) (51)(75)(79)(81)
%     \item to clarify or specify the meaning conveyed (= xxx) en (9) (52) (53)(92)
%     %\item to enclose example numbers (x)
% \end{itemize}

% while brackets […] are used
% \begin{itemize}
%     \item to indicate the continuation or preceding part of the sentence in the original text (3) (8)(9) (12) (13) (65)(80) (91) (95)(98)
%     \item to provide an explanation of the construction (60)(61) or to indicate what a proper noun refers to (79). to enclose the preceding text of the example, identified as [Context: xxx] (3) (9) (91) (92) (93)
% \end{itemize}

% \textbf{In the transcription line of an example}, Brackets are used to indicate a variant. (81)

% A single use of {…} in Table 5, which I suggest replacing with brackets.}
%LV: removing the arrows as per recommendation of the LSP proofreader; ${\downarrow}$ I 
% \newpage
% \lv{@LSP: would it be possible to align the heading Aspects in the table below so that it encompasses both \textsc{ipfv} and \textsc{pfv}?}
\begin{table}
\fittable{
\begin{tabular}{lllll}
\lsptoprule
{Moods } & \multicolumn{2}{c}{AM constructions } & \multicolumn{2}{c}{{Aspects }}\\
\cmidrule(l){2-3}\cmidrule(l){4-5}
& Names & Glosses & \scshape ipfv & \scshape pfv\\
\midrule
Realis & Neutral & ø & \textsuperscript{\textsc{h}}\textsc{vb} & \textsuperscript{\textsc{lh}}\textsc{vb}\textit{a}\textsuperscript{\textsc{h}}\\
& Progressive & \scshape prog & \textit{bítí}-\textsuperscript{\textsc{l}}\textsc{vb} & \textit{ɓɔ̀náá}{}-\{\textsuperscript{\textsc{l}}\textsc{vb}...~{\textasciitilde} \textsuperscript{\textsc{lh}}\textsc{vb}\textit{i}\} \\
& Inchoative & \scshape inch & \textit{bítí}-(\textit{mɛ})-\textsuperscript{\textsc{l}}\textsc{vb} & \\
& Known & \scshape known &  & \textsuperscript{\textsc{l}}\textsc{vb}\\ 
& Permanent & \scshape perm & \textit{bàá}-\textsuperscript{\textsc{l}}\textsc{vb} & \\
& \textit{Temporal condition} & \scshape cond & \textsuperscript{\textsc{h}}\textsc{vb} \textsuperscript{\textsc{l}}\textsc{vb} & \textsuperscript{\textsc{h}}\textsc{vb} \textit{tɛɛ}-\textsuperscript{\textsc{l}}\textsc{vb}\\
& \textit{Hypothetical} & \scshape hypo &  & {\textit{nɛ́}-\textsuperscript{\textsc{lh}}\textsc{vb}\textit{a}\textsuperscript{\textsc{h}}}\\
& \itshape Consecutive & \scshape cons &  & {\textit{nɛ́}-\textsuperscript{\textsc{l}}\textsc{vb}}\\
Virtual & Neutral & \textsc{vrt} & \textit{tɛ́}-\textsuperscript{\textsc{l}}\textsc{vb} &\textit{tɛ̀ɛ́}-\textsuperscript{\textsc{l}}\textsc{vb}\\
& Imminent & \scshape imn & \textit{dɛ́-mɛ́}-\textsuperscript{\textsc{l}}\textsc{vb} & \textit{dɛ̀ɛ̀-mɛ́}-\textsuperscript{\textsc{l}}\textsc{vb}\\
& \itshape Contingent & \scshape ctg & \textit{mɛ́}-\textsuperscript{\textsc{l}}\textsc{vb} & \textit{mɛ́-tɛ̀ɛ́}-\textsuperscript{\textsc{l}}\textsc{vb}\\
Command & Neutral & \textsc{ijt} & {\textsc{pers}\textsuperscript{\textsc{l}}   \textsuperscript{\textsc{h}}\textsc{vb}} & {\textsc{pers}\textsuperscript{\textsc{l}}   \textsuperscript{\textsc{l}}\textsc{vb}} \\ 
& Imperative & \scshape imp & \textsc{[pers = Ø]}   \textsuperscript{\textsc{h}}\textsc{vb} & \textsc{[pers = Ø]}   \textsuperscript{\textsc{l}}\textsc{vb} \\
\lspbottomrule
\end{tabular}
}
\caption{The 21 AM categories of verb paradigms in affirmative clauses}
\label{tab:gba:1}
\end{table}
% \hc{27.11.2024: I have changed [] to \{ \} in the PFV column of the second line (progressive), because these are two different options, as you also use later on in the chapter}


%     \aut[inline,size=\scriptsize,color=yellow!40]{I used colons to indicate the presence of two elements that cannot be separated or only partially. This applies between the TAM marker and the following verbal base, as well as between a term and the determiner that follows it, merging with it.}
%     \hc[inline,size=\scriptsize]{27.11.2024: Based on your answers (question 1, 2, 5 (and 6)), we understand that a and i are part of the verb stem, which is why we delete the colon from the notation (\textsc{vb}\textit{a}). This reflects that this letter is part of the verb stem in a more straightforward way.}
% \lv[inline,size=\scriptsize,color=blue!40]{2) What is final \textit{a}? Is it possibly a verb ending?} 
%     \aut[inline,size=\scriptsize,color=yellow!40]{No, a as i are bound forms of the VB, that is vb:i and vb:a as mentionned previoulsy (Depending on the Aspect-Mood (AM) constructions, the verb attests a bare form vb and two bound forms vb:i and vb:a, which will be simply glossed}
% \lv[inline,size=\scriptsize,color=blue!40]{3) What is the difference between tone in superscript and tone indicated prior to segments?}
%     \aut[inline,size=\scriptsize,color=yellow!40]{All tons are superscript in transcription, however, in the case of verbs without lexical tones, the tone they carry and that is written on them comes from the obligatory TAM form. This is why, in the glosses, this tone appears either on its own (e.g., H:VB for IPFV) or in association with the segmental marker of the TAM (e.g., tɛ́-L:VB for VIRT.IPFV)}
%     \hc[inline,size=\scriptsize]{27.11.2024: We do not want to have tone marked in different ways. In order to make it consistent, regardless of its position in the form, we have therefore chosen to use superscripts everywhere, keeping them before the verb stem whenever it was before the stem.}
    
% \lv[inline,size=\scriptsize,color=blue!40]{4) What does the hash tag mean, that is, the symbol \#? (= dièse)}
%     \aut[inline,size=\scriptsize,color=yellow!40]{This hash means that the TAM marker is distributed with the first part on the subject and the second part on the verb.}
%     \hc[inline,size=\scriptsize]{27.11.2024: We wonder whether marking the hashtag is necessary. Would it be possible to write it without? All the information is already there without it. We could add a sentences before the table to make it clear to readers. e.g. "Additionally, note that TAM is sometimes marked by tone only, either on the verb, or both on the verb and preceding personal pronoun."}
% \lv[inline,size=\scriptsize,color=blue!40]{5) what do i and a mean in \textsc{vb}:i and \textsc{vb}:a?}
%     \aut[inline,size=\scriptsize,color=yellow!40]{according to the AM form, VB is short or long, either VB:a or VB:i see answer to 2)}
% \lv[inline,size=\scriptsize,color=blue!40]{6) We noticed that you only use these underlining tone marks in the tables. Would it be possible to use the surface marking with accents instead? If not, please add an explanation that you use a different notation between tables and examples.}
%     \aut[inline,size=\scriptsize,color=yellow!40]{In the tables, I separate what pertains to the TAM from the verb base, which is lexically toneless. As a result, I introduce the notation for the tone that will attach to the verb base and be marked in superscript. To display this in the examples, an additional line would need to be added, which is generally not planned. Some examples
% Ex 1  bá > H.ba > IPFV:take		VB (short form) = ba
% Ex 3 bàà > BH.ba.H > PFV:take:TM	VB:a (long form) = baa
% Ex 4 gàsá > BH.gasi > PFV:grow  	VB (short form)= gasi,  VB:a (long form)= gasa
% Ex 6 bàá-kɔ̰̀ > bàá B kɔ̰ > PERM:want	VB (short form) = kɔ̰
% Ex 10 tɛ́-fè > tɛ́ B fe > VRT.IPFV:die	VB (short form) = fe}
% \hc[inline,size=\scriptsize]{27.11.2024: ok, we will keep these two ways of marking tone using letters in tables and text (superscript as mentioned above) and the accents in the examples.}


    
In addition to these AM markers placed before and on the verbal base (\textsc{vb}),\footnote{To keep the preverbal position of the AM markers and the associated tonal pattern, instead of a \textsc{vb}{\textbackslash}\textsc{lh}{}-a\textsuperscript{\textsc{h}} notation I chose the \textsuperscript{\textsc{lh}}\textsc{vb}\textit{a}\textsuperscript{\textsc{h}} notation.} the verb may have other optional markers which are placed in strict order after the verb. Among these, the negative morpheme is always in the very last position of the verb phrase in clause final position, eventually being followed only by some sentence particles, adverbs and interrogatives, as mentioned earlier.

Gbaya is an isolating language with very little morphology and no agreement at all. The word order is very strict. It is an SVO language which also makes use of non-verbal predication. It has verbs, non-verbal predicates, nouns and four categories of adjectives. The verbal system is organized in three moods (realis, virtual and command) and two aspects (perfective and imperfective) \citep[see][]{Roulon1975,Roulon-Doko1993}. Verbs always express a process where the obligatory subject is either external to the process (transitive construction = transitive voice) or included in the process (intransitive construction = middle voice). It is a tonal language with two levels and four tones. The high tone can mark grammatical distinctions (functioning as a supra-segmental marker, \textsc{tm}, in glosses). It occurs between various elements. When it occurs between two elements, it modifies the tonal melody of the first element. How the tones are realized depends on the tone of the second element. Gbaya phonology has 3 glottalized consonants, 3 labio-velars and a complete range of nasals and semi-nasals; it has nasal vowels too.

The paper is organized according to the questionnaire provided by the editors (\citetv{chapters/questionnaire}). In \sectref{sec:gba:2}, I present clausal negation first in verbal clauses, then in non-verbal clauses. In \sectref{sec:gba:3}, I focus on non-clausal negation including negative replies and a discontinuous negation. In \sectref{sec:gba:4}, I address pragmatic phenomena and specific devices used in connection with negation. A summary in  \sectref{sec:gba:5} concludes the chapter.

\section{Clausal negation}\label{sec:gba:2}

Gbaya has a single dedicated negative marker, the morpheme \textit{ná} (\textsc{neg}). It was reconstructed in Proto-Gbaya showing its great stability throughout the Gbaya-Manza-Ngbaka group \citep{Monino1995}.
This negative morpheme occurs with verbs only. It is always placed after the verb, in the very last position of the verb phrase, and after all the complements and adverbs following the verb, if there are any. This clause-final position may be shared with some other morphemes that can follow the negative morpheme \textit{ná} in strict order, as (i) two adverbs, a restrictive one and an additive one (cf. \sectref{sec:gba:4.4}), (ii) sentence particles which manifest the attitude of the speaker towards the utterance and the addressee, and in the last position, (iii) some interrogative particles (cf. \sectref{sec:gba:4.5}). This brief characterization is given here just to put into perspective the clause-final position of the negative morpheme. There are different constructions used to express negation according to the type of clause, verbal clause opposed to non-verbal clause. 

\subsection{Standard negation}\label{sec:gba:2.1}
\subsubsection{Negation in the realis mood}\label{sec:gba:2.1.1}



In the realis mood, aspect serves to differentiate between a non-delimited process for the imperfective and a delimited process for the perfective.
The perfective AM categories (\textsc{pfv}) cannot be negated; the verb must be in an imperfective form to be negated. So there is asymmetry in the paradigm as all perfective forms are blocked in the negative. Furthermore, only two imperfective forms can be negated (the neutral and the permanent); the others (the progressive, the inchoative, and the temporal condition) are only used in affirmative clauses. The negative marker is added to a corresponding affirmative construction and placed in clause-final position. So in both cases the constructions are symmetric. 

The neutral realis imperfective is the most frequent form, attested in 48,8\% of all the realis AM forms in my corpus. As the perfective form is blocked, the negative clause \REF{ex:gba:1} corresponds to both the imperfective and perfective forms that the speaker can use in affirmative as \REF{ex:gba:2} and \REF{ex:gba:3} to express his\textbackslash her point of view. Such a choice is not possible in negative clauses; the opposition perfective / imperfective is neutralized. The structure of the negative clause \REF{ex:gba:1} is symmetric to the corresponding imperfective positive construction \REF{ex:gba:2} and the negative marker \textit{ná} is simply added in clause-final position. 

\ea\label{ex:gba:1}
    \gll Nàá-ngɔ̀ɗì bá sóyó kɔ́í ná \\ 
         \textsc{pn} \textsc{ipfv}:take seat of:\textsc{log.sg} \textsc{neg} \\
    \glt ‘Naa-ngɔɗi does not take her seat.’ 
\z 

\protectedex{
\ea\label{ex:gba:2}
    \gll ʔèé Nàá-ngɔ̀ɗí bá sóyó kɔ́í  \\ 
         and:\textsc{tm} \textsc{pn} \textsc{ipfv}:take seat of:\textsc{log.sg}  \\
    \glt ‘Then Naa-ngɔ̀ɗi takes her seat [and sits down on the side of the road].’
\z }


\ea\label{ex:gba:3}[Context: The speaker has convinced the man to enter her house.]\newline
    \gll ká Nàá-ngɔ̀ɗí bàà sóyó kɔ́í \\ 
         then \textsc{pn} \textsc{pfv}:take:\textsc{tm} seat of:\textsc{log.sg}  \\
    \glt ‘Then Naa-ngɔ̀ɗi takes her seat [and they enter the house].’
\z 

For some verbs in an intransitive construction such as \textit{gasi} ‘grow’  the English translation as a verb of quality can be misleading. Such verbs also refer to a process and their perfective forms express the result of the process that can be a quality as in \REF{ex:gba:4}. This is especially the case of verbs from which verbal adjectives are derived and whose resultative process corresponds to a quality. These verbs can be negated, but when they are negated they still refer to processes as in \REF{ex:gba:5}\footnote{A note on the placement of diacritics is in order. In this chapter, nasal diacritics appear under vowels, tone diacritics are placed above the vowels.}. They are not stative verbs. Stative constructions would be expressed with the stative verb \textit{ʔɔ} ‘be’, or with constructions without any dedicated predicate (see \sectref{sec:gba:2.3.2}).

\ea\label{ex:gba:4}
    \gll sààyé hɛ̰̰̀ gàsá\\ 
         village \textsc{prox} \textsc{pfv}:grow \\
    \glt ‘This village is big.’ (lit. ‘has grown’)
\z 
 
\ea\label{ex:gba:5}
    \gll sààyé hɛ̰̰̀ gásí ná  \\ 
        village \textsc{prox} \textsc{ipfv}:grow \textsc{neg} \\
    \glt ‘This village is not big.’ (lit. ‘has not grown’)
\z 

The permanent, which has only an imperfective form, can be negated without any change. The permanence of the process meaning “always having" for the affirmative value \REF{ex:gba:6} becomes “never having" for the negative value \REF{ex:gba:7}. The permanent is used with negation very seldom, most probably because of its very specific meaning; it has 0,1\% of all its occurrences in my corpus, that is, 7 occurrences, only one of which is negated.

    % \mm[inline,size=\tiny,color=green!40]{Glossing [in ex7]: PERM:want with colon but hyphen on the example line, either delete hyphen or replace colon with hyphen.}
    % \aut[inline,size=\tiny,color=yellow!40]{Ex 6 and 7. There is no morpheme-by-morpheme correspondence due to the tonal pattern that every conjugation marker brings to the verbal base. 
    % Placing a hyphen between PERM and the verbal base removes the tonal element (low tone) that the PERM form applies to the verbal base. Indeed, the hyphen in the transcription isolates the morphemic part of PERM bàá, which necessarily carries a low tone attached to the verbal base kɔ̰ ‘to want’ (bàá-tɔ̰̀ = bàá-B tɔ̰ = PERM want), in the same way as for the perfective of the verb gasi (gàsà = Bha gasi = PFV grow). It might be better to use semicolons between the AM form and the verbal base instead of colons, if you prefer, but all AM forms had to be modified : IPFV;VB au lieu de IPVV:VB.}
    %\hc{Based on the description you sent us, we decided to delete the morpheme boundary in the Gbaya line. The colon in the gloss line shows that these two meanings are expressed in the form, and they are not }
    % \hc[inline,size=\tiny]{Il y a deux possibilités ici: nous pouvons soit enlever le tiret dans la ligne en Gbaya, soit changer le double-point et mettre un tiret aussi dans la ligne des glosses. La règle que nous suivons ici est qu'il doit y avoir le même nombre d'éléments sur chaque ligne dans les exemples. Pour l'instant, j'ai changé l'exemple 6 en y mettant le tiret dans les deux lignes, et pour l'exemple 7 je l'ai considéré comme un mot non segmentable, donc sans limite de morphème. Qu'est-ce que vous préférez entre l'option présentée en 6 et celle présentée en 7?}
    % \hc{Par ailleurs, le double-point est utilisé dans les LGR quand il est possible de ségmenter le mot en plusieurs morphèmes, mais l'auteur(e) ne veut pas les séparer, pour une raison ou une autre. Le point-virgule est utilisé si ce n'est pas possible de ségmenter le mot en plusieurs morphèmes. Le simple point peut être utilisé dans tous les cas.}
    % \aut[inline,size=\tiny,color=yellow!40]{I prefer to keep the hyphen in both as in ex 6. \\ The issue with TAM forms lies in the fact that they consist of an isolable segmental element, which also carries the tone of the verb, though the tone itself is not isolable. For simple TAM forms (PFV, IPFV, HYPO, etc.), colons are used to link them to the verbal base they affect (IMPV:VB = English term for the verb). For all complex TAM forms, I follow the proposed use of a period (e.g., VRT.PFV or INF.IPFV, PERM, etc.) and always mark the boundary between the TAM form and the verbal base with a hyphen (e.g., VRT.PFV-VB = English term for the verb), to maintain an artificial correspondence between transcription and glosses (a hyphen in both). However, when the complex TAM form itself is composed of multiple elements (e.g., prohibitive yak-ká-L VB or imminent dɛ́-mɛ́-L VB), there are two hyphens in the transcription: one between the TAM components and another before the verbal base. In these cases, I use periods between the TAM components and a hyphen before the verbal base. No morpheme-to-morpheme correspondence is possible, similar to the case of nominal compounds discussed in example 34.}
    % \hc{I can delete this whole bunch of comments later, we keep them for the time of the meeting in case we need them later on on another page.}

\ea\label{ex:gba:6}
    \gll ʔéí bàá-kɔ̰̀ kéí ʔó bóíbémáà \\ 
         \textsc{log.sg} \textsc{perm}-want as\_for:\textsc{log.sg} \textsc{pl} boy:\textsc{def} \\
    \glt ‘He always wanted boys.’
\z 

\ea\label{ex:gba:7}
    \gll ʔéí bàá-kɔ̰̀ kéí ʔó bóíbémáà ná \\ 
         \textsc{log.sg} \textsc{perm}-want as\_for:\textsc{log.sg} \textsc{pl} boy:\textsc{def} \textsc{neg} \\
    \glt ‘He never wanted boys.’
\z 


Within the linked AM forms of realis mood, only the hypothetical can be negated. Since in the affirmative the hypothetical ‘\textsc{hypo}’ {\textit{nɛ́}-\textsuperscript{\textsc{lh}}\textsc{vb}\textit{a}\textsuperscript{\textsc{h}}}
    % \mm[inline,size=\tiny,color=green!40]{We have slightly modified the notation of tones here, is it fine with you?}
    % \aut[inline,size=\tiny,color=yellow!40]{keep the form as in table 1: {nɛ́-lh.VB:ah }}
    % \hc[inline,size=\tiny]{the old form here was: \textit{nɛ́}-\textsc{lh}a.VB.\textsc{h}. The new we suggested is \textit{nɛ́}{}-a\textsuperscript{\textsc{lh}}.\textsc{vb}\textsuperscript{\textsc{h}}. I have put the old form again for now.}
    %\hc[noline,size=\tiny]{If the author agrees with this change, I'll do it everywhere for all tones in texts: tone in superscript after the syllable instead of smallcaps and dot before} 
    % \aut[inline,size=\tiny,color=yellow!40]{No. The last H is not on the VB, could you use the form given in the table 1 :    nɛ́-lh.VB:ah    It is the same form as PFV of neutral realis lh.VB:ah   postponed to nɛ́-}
    % \aut[inline,size=\tiny,color=yellow!40]{Not to be changed: the tone in smallcaps and dot, correspond to the tonal part of the conjugated form. Tone cannot be following the VB. {nɛ́-\textsc{l}.VB} and not {nɛ́-VBL} as in Table 3.}
    %\hc{I have put the old form again for now (also lower in the paragraph). But the notation that you had needs to be modified in some way.}
    %\aut{ nɛ́-LV.B:a.H as in table 1} 
    % \hc{These comments relate to the discussion we’ve had about the notation. The main questions are elsewhere, so what is to be done here on p. 7 depends on answers to those questions.}
    % \mm[inline,size=\tiny,color=green!40]{And we have an additional question regarding "a" here: what does it stand for? If it is language material, it should be italicized just like nɛ́- before it. And why is it separated from VB with a dot and not with a hyphen? Can you please clarify?}
    % \aut[inline,size=\tiny,color=yellow!40]{The verb attest, as mentionned previously, a bare form VB and two bound forms VB:i and VB:a, It is this form VB:a that is the basis for realis neutral PFV and for Hypothetical form. In addition, these forms support the tonal marker -H}
    % \hc{do this as we decided on page 3-4. So here I will use L.VBa.H, put superscript and delete dots... }
    % \hc{We also took the curly brackets away here and elsewhere because they are not needed in these contexts. I will go through the chapter.}
is always a perfective form (a process that is considered delimited), see \REF{ex:gba:8}, there is no corresponding imperfective form to carry the negative morpheme. However, it is possible to hypothesise a negated process, and for this purpose a special form is created, \textit{nɛ́}-\textsuperscript{\textsc{l}}\textsc{vb}, as in 
    % \aut{change to {\textit{nɛ́}-\textsc{l.vb}} as written in table 3}
    % \hc{the old form was: \textit{nɛ́}-\textsc{l.vb}. The form we suggested was \textit{nɛ́}{}-\textsc{vb}\textsuperscript{\textsc{l}}. I have put the old one for now}
\REF{ex:gba:9}. This form reuses the form of the consecutive, highlighting a usual practice in Gbaya syntax: one form used with different meanings in two mutually exclusive contexts. I call this special form hypothetical negative ‘\textsc{hypo.neg}’. This form could fill the empty place of the imperfective form of the Hypothetical in \tabref{tab:gba:1}. However, as it is only attested in negative clauses, it cannot be opposed to the perfective form in the affirmative and is therefore only mentioned in \tabref{tab:gba:2}. In my corpus, the hypothetical negative has 4 occurrences, i.e. 0.06\%, while in affirmative clauses the hypothetical has 44 occurrences, i.e. 0.6\%. Both therefore represent 0,66\% of all AM forms.

\ea\label{ex:gba:8}
    \gll kà̰à̰  nɛ́-kɔ̰̀ám   \\ 
         when:\textsc{3sg} \textsc{hypo}-want:\textsc{1sg} \\
    \glt ‘If he loves me [let him get a dowry to marry me].’
\z 

\ea\label{ex:gba:9}[Context: The two girls ask him to marry them.] \\
    \gll ʔá̰    kɔ̀mɛ̀ nɛ́-kɔ̰̀ ná \\ 
         and\_now when:2\textsc{sg} \textsc{hypo.neg}-want \textsc{neg} \\
    \glt ‘And if you do not want (= love us) [choose the one you love].’
\z 

To sum up, in the realis mood, the constructions are symmetric as the negator is simply added with no further structural changes. All perfective forms are blocked in the negative and so there is a paradigmatic asymmetry. Moreover, the paradigm of the imperfective forms which can carry the negative morpheme is narrowed to three forms, the neutral form being the most frequent. These imperfective forms represent 91,3\% of all negative predications in my corpus (see \tabref{tab:gba:3}).

\subsubsection{Negation in the virtual mood}\label{sec:gba:2.1.2} 

The virtual mood in Gbaya contrasts with the realis by encoding, for example, plans or possibilities. In this mood, aspect distinguishes between events that may be realized (possible) using imperfective forms \REF{ex:gba:11} or will not be realized (irrealis) using perfective forms \REF{ex:gba:13}. Both can be negated just by adding the negative morpheme at the end of the clause. So in both cases the constructions are symmetric. 

Only the neutral form of the virtual mood is attested in negative clauses. The other forms (the imminent and the contingent as mentioned in \tabref{tab:gba:1}) are not possible because of their semantic nature. The imminent focuses on the short time frame leading up to the action. The contingent is a linked form expressing the process as a final purpose or a possibility to be taken into account. Both the imminent and the contingent are only used in affirmative clauses. In the virtual mood the paradigm of the negative forms is narrowed to the two neutral forms, imperfective \REF{ex:gba:10} and perfective \REF{ex:gba:12}.

\ea\label{ex:gba:10}
    \gll ʔéí tɛ́-fè ná  \\ 
          \textsc{log.sg} \textsc{vrt.ipfv}-die \textsc{neg}  \\
    \glt ‘He will not die.’
\z 

\ea\label{ex:gba:11}
    \gll ʔéí tɛ́-fè kɔ́í ʔín ʔá̰    yè \\ 
         \textsc{log.sg} \textsc{vrt.ipfv}-die for:\textsc{log.sg} \textsc{com:tm} \textsc{3sg} finally \\
    \glt ‘As for him, he finally will    die with her.’
\z 

\ea\label{ex:gba:12}
    \gll Wàntò tɛ̀ɛ́-kpà mà mbûl ná \\ 
         \textsc{pn} \textsc{vrt.pfv}-find \textsc{indf} spider\_web \textsc{neg} \\
    \glt ‘[If it hadn't been like that] Wanto would not find some spider web.’
\z 
 
\ea\label{ex:gba:13}
    \gll ʔám tɛ̀ɛ́-kpà kàà wí ʔèém nɛ́nɛ́ \\ 
         1sg \textsc{vrt.pfv}-find \textsc{rpt.add} person help:\textsc{nmlz}:\textsc{1sg} going \\
    \glt ‘[If possible] I would find someone to go with me.’
\z 

To sum up, in the virtual mood, both imperfective and perfective AM categories can be negated, but the paradigm is limited to neutral perfective and imperfective forms. They represent 4,8\% of all negative predications in my corpus. 

\subsubsection{Negation in serial verb constructions}\label{sec:gba:2.1.3}

In serial verb constructions (SVC) the negative morpheme is always attached to the first conjugated serialized verb. The second serialized verb has an infinitive form. There are three forms of the infinitive: an imperfective ‘\textsc{inf.ipfv}’, a perfective ‘\textsc{inf.pfv}’ and a virtual ‘\textsc{inf.vrt}’.
Most often SVCs are affirmative clauses. In my corpus, SVCs that carry negation can have any verb as their first serialized verb when the second serialized verb has a perfective infinitive form \REF{ex:gba:14}. But when the second serialized verb has an imperfective infinitive form, only the modal verb \textit{kɔ̰́} ‘want’ is attested in the corpus as the first serialized verb, see \REF{ex:gba:15}. Finally, when the second serialized verb has a virtual infinitive form, only the verbs \textit{kɔ̰́} ‘want’ and \textit{ɲɛm} ‘be suitable to’ are attested as the first serialized verbs in the corpus \REF{ex:gba:16}, as summarized in \tabref{tab:gba:2}. While in such affirmative SVCs the first serialized verb can have an imperfective or a perfective realis form, in the negative counterpart only the neutral imperfective is attested for this verb that falls under the scope of the negation.

\begin{table}
\begin{tabularx}{\textwidth}{lXXXl}
\lsptoprule
\textsc{vb}\textsubscript{1} + \textsc{ipfv} & \multicolumn{3}{c}{\textsc{vb}\textsubscript{2} +} & in final position\\
& \scshape inf.pfv & \scshape inf.ipfv & \scshape inf.vrt & \\
\midrule 
{any verb} & x\footnote{ x = attested; --- = unattested } & --- & --- & {+ \textit{ná} \textsc{neg}}\\ 
\textit{kɔ̰} ‘want’ & x & x & x & + \textit{ná} \textsc{neg}\\ 
\textit{ɲɛm} ‘be suitable to’ & x & --- & x & + \textit{ná} \textsc{neg}\\
\lspbottomrule
\end{tabularx}
    \caption{Types of verbs in negated SVCs }
    \label{tab:gba:2}
\end{table}


\ea\label{ex:gba:14}
    \gll sɔ́sɔ́ɔ̀ hɛ̰̀ ʔá̰    nɛ́ zɔ́k pɛ́r ná \\ 
         today here \textsc{3sg} \textsc{ipfv}:go \textsc{inf.pfv}:see trap \textsc{neg} \\
    \glt ‘Today he is not going to visit his traps.’
\z 
 
 \ea\label{ex:gba:15}
    \glll ʔá̰  kɔ̰́ gíí gbɔ̀lɔ̀ wó ná \\ 
         \textsc{3sg} \textsc{ipfv}:want \textsc{inf.ipfv}:cook {Corchorus} really \textsc{neg} \\ 
         {} {V1} {V2}\\
    \glt ‘She really does not want to cook gooey leaves.’
\z 

\ea\label{ex:gba:16}
    \gll ʔám kɔ̰́ mɔ́-tɔ̰̀ tò ná \\ 
         \textsc{1sg} \textsc{ipfv}:want \textsc{inf.vrt}-say tale \textsc{neg} \\
    \glt ‘I do not want to tell a tale.’
\z 

 
\subsection{Negation in non-declaratives}\label{sec:gba:2.2}

In the command mood, imperfective forms are used to encode commands that are either feasible or not feasible to be carried on by the hearer. This is in contrast with the performative command, which is interpreted as already realized and thus belongs to the perfective aspect. As for realis mood, all perfective forms are blocked in the negative.

%and a performative command, considered as already realized, in the perfective forms. 

 

Within imperfective command forms, only the impersonal form i.e. the imperative can be negated by adding the negative morpheme at the end of the clause without any change (\ref{ex:gba:20}--\ref{ex:gba:21}). The construction is symmetric. Such a form is common after a vocative.


\ea\label{ex:gba:20}
    \gll ʔó nɔ́kɔ́ hṵ́\'{ı̰}  yíknɛ́ ná  \\ 
         \textsc{pol} maternal\_uncle \textsc{imp}:open eye:\textsc{2pol} \textsc{neg}  \\
    \glt ‘Uncle, do not open your eyes.’ 
\z 
 
\ea\label{ex:gba:21}
    \gll hṵ́\'{ı̰}  yíknɛ́ \\ 
         \textsc{imp.ipfv}:open eye:\textsc{2pol}  \\
    \glt ‘Open your eyes.’
\z 

But most often, there are two new forms dedicated to negative commands: the vetitive or injunctive negative (\textsc{ijt.neg}), and a stronger form, the prohibitive (\textsc{proh}). 

The vetitive, constructed in the same way as all the personal forms of the command mood, is encoded by both the verbal base and the personal pronoun that precedes it. The verbal base has a low tonal scheme as the injunctive perfective, and the personal pronoun has a low tonal scheme as for both injunctive forms, to which is added a vowel with an identical vocal quality and a high tone: \mbox{\textsc{pers}\textsc{\textsuperscript{l}[\textsuperscript{h}V]}   \textsuperscript{\textsc{h}}\textsc{vb}}
\REF{ex:gba:22}. This construction is the most common one to express a negative command. It corresponds, in the affirmative, to the injunctive imperfective \REF{ex:gba:23}, but the constructions are not symmetric. The vetitive is glossed as injunctive negative ‘\textsc{ijt.neg}’. 

\ea\label{ex:gba:22}
    \gll wòóyèé mɛ̀ɛ́ hèì kɔ̰̀á̰ ná  \\ 
         no \textsc{2sg:ijt.neg} \textsc{ijt.neg:}pour tears \textsc{neg}  \\
    \glt ‘No, do not cry.’
\z 

\ea\label{ex:gba:23}
    \gll mɛ̀ héí kɔ̰̀á̰ \\ 
         \textsc{2sg:ijt} \textsc{ijt.ipfv:}pour tears \\
    \glt ‘Cry.’
\z 

The prohibitive uses the verb \textit{yak} ‘leave’ as an auxiliary verb. It is followed by the temporal coordinative \textit{ká} ‘then’ \REF{ex:gba:24}, or by the noun \textit{tɛ̀} ‘entity, body’ \REF{ex:gba:25} carrying the high tone of the tonal marker \textit{tɛ̀\textsuperscript{\textsc{h}}}, or by the quality noun \textit{tí} ‘identity, front’. Any of these terms (\textit{ká}, \textit{tɛ̀}\textsuperscript{\textsc{h}} or \textit{tí}) precede the verb which has the low tonal scheme [\textsuperscript{\textsc{l}}\textsc{vb}] and is regularly in a non-finite form used for any verb in complex forms and for the nominalization of a verb within a nominal phrase \REF{ex:gba:26}. The prohibitive has thus three variants: (i) \textit{yak-ká}{}-\textsuperscript{\textsc{l}}\textsc{vb}, (ii)~\textit{yak-tɛ̀}\textsuperscript{\textsc{h}}-\textsuperscript{\textsc{l}}\textsc{vb} and (iii)~\textit{yak-tí}{}-\textsuperscript{\textsc{l}}\textsc{vb}. The auxiliary verb is attested with three AM markers: a high tonal scheme for \textsc{imp.ipfv}, \textsc{ijt.ipfv}, and a low tonal scheme for \textsc{ijt.neg}. 

\ea\label{ex:gba:24}
    \gll yák-ká-tɔ̰̀ wèn ná  \\ 
         \textsc{imp.ipfv-proh}-say speech \textsc{neg}  \\
    \glt ‘You must not say anything.’
\z 

\ea\label{ex:gba:25}
    \gll ʔàm yák-tɛ́-kùr ná \\ 
         \textsc{1sg:ijt} \textsc{ijt.ipfv-proh}-get\_up \textsc{neg} \\
    \glt ‘I must not get up.’
\z 

\ea\label{ex:gba:26} [Context: The squirrel pretends to be ill and commands the lion not to disturb him.]\\ 
    \gll ʔɛ̀nɛ̀ɛ́ yàk-ká-sɔ̌m géè ná \\ 
         \textsc{2pol:ijt.neg} \textsc{ijt.neg-proh}-disturb:1\textsc{sg} needlessly \textsc{neg}  \\
    \glt ‘You must not disturb me for any reason.’
\z 
To sum up, in the command mood, as in the realis mood, only the imperfective forms can be negated; all the perfective forms are blocked. For the imperative{}-imperfective, the negative construction is symmetric, but most often two new forms dedicated to negative command, the vetitive and the prohibitive, are used within non-symmetric constructions. All these forms represent 3,9\% of all negative predications in my corpus. 

In the case of negation in an interrogative sentence, where the question marker \textit{ndé} is always placed at the absolute end of the sentence, the negation morpheme \textit{ná} is positioned at the end of the clause, just before the question marker, as illustrated in \sectref{sec:gba:4.6.4} (see \ref{ex:gba:100}--\ref{ex:gba:101}).

To sum up this presentation of AM categories in verbal clauses, \tabref{tab:gba:3} shows the negative counterpart of AM categories for affirmative clauses shown in \tabref{tab:gba:1}. The specialized forms dedicated to carry negation are in bold in \tabref{tab:gba:3}.

\begin{table}
\fittable{
\begin{tabular}{lp{20mm}ll@{}lr}
\lsptoprule
Moods $\downarrow$ & {AM forms $\downarrow$} &  & {\scshape ipfv} & {\scshape pfv} & {frequency}\\
\midrule
Realis & {Neutral} & {ø} & \textsuperscript{\textsc{h}}\textsc{vb} &  & {\textbf{91,3\%}}\\
\tablevspace
& {Permanent} & {\scshape perm} & {\textit{bàá}-\textsuperscript{\textsc{l}}\textsc{vb}} &  & \\
\tablevspace
& \bfseries Hypothetical negative & \bfseries\scshape hypo.neg & \textbf{\textit{nɛ́}-\textsuperscript{\textsc{l}}\textsc{vb}} &  & \\
\midrule
Virtual & Neutral & \textsc{vrt} & \textit{tɛ́}-\textsuperscript{\textsc{l}}\textsc{vb} & \textit{tɛ̀ɛ́}-\textsuperscript{\textsc{l}}\textsc{vb} & \textbf{4,8\%}\\ 
\midrule
\multicolumn{1}{c}{Command} & Imperative & \scshape imp & {\textsc{pers =}{ø}   \textsuperscript{\textsc{h}}\textsc{vb}} &  &  {\textbf{3,9\%}}\\
\tablevspace
& \bfseries Injunctive negative & \bfseries\scshape ijt.neg & {\textbf{\textsc{pers}}\textbf{\textsc{\textsuperscript{l}[\textsuperscript{h}V]}} \textsuperscript{\textbf{\textsc{h}}}\textbf{\textsc{vb}}} &  & \\
\tablevspace
 & \bfseries Prohibitive negative & \bfseries\scshape proh & \textbf{\textit{yak}-\{\textit{ká \textasciitilde\space tɛ̀}\textsuperscript{\textsc{h}} \textasciitilde\space \textit{tí}\}-\textsuperscript{\textsc{l}}\textsc{vb}} &  & \\
\lspbottomrule
\end{tabular}
}
\caption{The 7 AM categories of the verbal paradigm in negative clauses}
\label{tab:gba:3}
\end{table}
% \hc{28.11.2024: last line of the table, IPFV column: are these 3 options in the brackets? I would change the / to {\textasciitilde} as elsewhere when there are different options}
% \aut{%to indicate the three options, you could use  {\textasciitilde} as previoulsy in the paragraph before (58) tá  {\textasciitilde} tɛ́, or the semi-colon as I proposed too. 
% Simple brackets is all right here yák-[ká  {\textasciitilde}  tɛ̀H  {\textasciitilde} tí]-LVB  It may be necessary to put  {\textasciitilde} in the abbreviations for 'alternative form'.}
% \hc{I did not change the kind of parentheses, keeping the weird ones instead of the square ones the author wanted > to modify in the list in the new footnote at the beginning.}
% \aut{hyphen that I use for compound terms do not refer to an option. Here it is an option, may be a semi-colon would be good or a  {\textasciitilde} as I explain in the comment in the table. > HC: this does not seem to be related to what we were asking, keeping this comment for now but will delete later}

\subsection{Negation in stative predications}\label{sec:gba:2.3}

Gbaya stative predications are divided into several types presented below. They include identificational, locative, existential and presentative predications, which have a dedicated term each (see \sectref{sec:gba:2.3.1}). Possessive, similative, qualitative, adverbial and numeral predications are encoded by juxtaposition with no dedicated predicates. Adding the potential verb to carry the negative morpheme is the dedicated negative construction common to all non-verbal clauses. They represent a quarter of all negated clauses in my corpus (see \sectref{sec:gba:2.3.2}). Predication of property is also  encoded by clauses where the copular element is a position verb (see \sectref{sec:gba:2.3.3}). The negation of each of subtype of stative clauses is discussed in the respective sections.


\subsubsection{Clauses with a non-verbal predicate}\label{sec:gba:2.3.1}

There are four types of clauses whose predicate is non-verbal: identificational, locative, existential and presentative.

\begin{table}
\fittable{
\begin{tabular}{llll}
\lsptoprule
Predications & in affirmative clauses & in negative clauses & with a non-verbal predicate\\
\midrule 
Identificational & [NP] \textit{nɛ́} NP  & [NP] \textit{bé nɛ́} NP ná & \textit{nɛ́} \textsc{id}\\ 
Presentatives & NP  \textit{ʔá̰à̰  }& NP \textit{bé   ʔá̰à̰ ná }& \textit{ʔá̰à̰  }\textsc{prst-loc} \\
& NP \textit{ʔá̰-h\`{ı̰}\'{ı̰}  } & \textsc{NP} \textit{bé   ʔá̰-h\`{ı̰}\'{ı̰}  ná} & \textit{ʔá̰-h\`{ı̰}\'{ı̰}  } \textsc{prst-cons}\\
Locative & NP \textit{ʔá̰  } NP & NP \textit{bé}  locative NP \textit{ná} & \textit{ʔá̰  }\textsc{loc}\\
Existential & NP \textit{ʔá̰-sɛ̀nɛ́}  & NP \textit{bé}  \textit{ná} & \textit{ʔá̰-sɛ̀nɛ́ } \textsc{loc}-inside\slash \textsc{ex}\\
\lspbottomrule
\end{tabular}
}
\caption{Negation of clauses with a non-verbal predicate}
\label{tab:gba:4}
\end{table}

In affirmative clauses, identificational predication uses a dedicated predicate: the identificator \textit{nɛ́} ‘\textsc{id}’ that most often links two noun-phrases within a three-constituent clause NP \textit{nɛ́} NP \REF{ex:gba:27}. It can also sometimes be reduced to a two-constituent clause \textit{nɛ́} NP that refers to an impersonal construction, often in response to \REF{ex:gba:29}. For the negation of this construction, the potential verb \textit{be} ‘can be’ is placed just before the identificator and the negation morpheme is placed, like in verbal predications, clause-finally (\ref{ex:gba:28}, \ref{ex:gba:30}). 

\ea\label{ex:gba:27}
    \gll  mɛ́ nɛ́ bêm  \\ 
         2\textsc{sg} \textsc{id} child  \\ 
    \glt ‘You are a child.’
\z 

\ea\label{ex:gba:28}
    \gll mɛ́ bé nɛ́ bêm ná \\ 
         2\textsc{sg} \textsc{ipfv}:can\_be \textsc{id} child \textsc{neg} \\
    \glt ‘You are not a child.’
\z 

\ea\label{ex:gba:29}
    \gll nɛ́ tùà kɔ́m  \\ 
         \textsc{id} house of:1\textsc{sg}  \\
    \glt ‘It's my house.’
\z 

\ea\label{ex:gba:30}
    \gll bé nɛ́ tùà kɔ́m ná \\ 
         \textsc{ipfv}:can\_be \textsc{id} house of:1\textsc{sg} \textsc{neg} \\
    \glt ‘It's not my house.’
\z 

Among the presentatives which are placed after the term they present \REF{ex:gba:31}, only two of them, the locative ‘be here’ \textit{ʔá̰à̰ } \textsc{prst.loc} and the consecutive ‘be like that’ \textit{ʔá̰-h\`{ı̰}\'{ı̰} } \textsc{prst.cons} are attested in a negative clause. The anaphoric locative ‘be there’ \textit{ʔà̰\'{ı̰} } \textsc{prst.anaph} is not attested in a negated sentence. For the negation of these constructions, the potential verb \textit{be} ‘can be’ is placed just before the presentatives and the negation morpheme is placed, like in verbal predications, clause-finally, in the same way as for the identificator (\ref{ex:gba:32}--\ref{ex:gba:34}).

\ea\label{ex:gba:31}
    \gll bêm ʔá̰à̰  \\ 
         child \textsc{prst.loc}  \\
    \glt ‘Here is the child.’ 
\z 
 
\ea\label{ex:gba:32}
    \gll bêm bé ʔá̰à̰  ná  \\ 
         child \textsc{ipfv}:can\_be \textsc{prst.loc} \textsc{neg} \\
    \glt ‘The child is not there.’
\z 

\ea\label{ex:gba:33}
    \gll wí héí mɔ́-ʔɔ̀ ʔîn bé ʔá̰à̰  ná \\ 
          person to:\textsc{log.sg} \textsc{ctg.ipfv}-lay \textsc{com} \textsc{ipfv}:can\_be \textsc{prst.loc} \textsc{neg}\\
    \glt ‘Someone you can sleep with cannot be this one here.’
\z 

\ea\label{ex:gba:34}
    \gll wéì kéí ndèì péí ʔínà bé ʔá̰-h\`{ı̰}\'{ı̰}  ná yè \\ 
         husband of.\textsc{log.sg} \textsc{rel.log.sg} \textsc{ipfv:}revenir \textsc{com.tm.3sg} \textsc{ipfv:}can\_be \textsc{prst-cons} \textsc{neg} finally \\
    \glt ‘My husband, with whom I have returned, cannot be like that after all.’
\z 

In affirmative clauses, locative predication uses a dedicated predicate: the locative predicate \textit{ʔá̰} ‘\textsc{loc}’ to introduce a noun phrase refering to a location \REF{ex:gba:35}. For the negation of this construction, the potential verb \textit{be} ‘can be’ replaces the locative predicate \textit{ʔá̰ } ‘\textsc{loc}’ and the negative morpheme is, like in verbal predications, placed clause-finally \REF{ex:gba:36}.


\ea\label{ex:gba:35}
    \gll béí ʔá̰  kútùà  \\ 
         people \textsc{loc} house \\
    \glt ‘People are in the house.’
\z 

\ea\label{ex:gba:36}
    \gll béí bé kútùà ná  \\ 
         people \textsc{ipfv}:can\_be house \textsc{neg} \\
        \glt ‘People are not in the house.’
\z 

In affirmative clauses, the dedicated predicate to express an existential is a compound term: \textit{ʔá̰-sɛ̀nɛ́} (\textsc{loc}-inside), literally ‘is inside’, which follows the NP \REF{ex:gba:37}. For the negation of this construction, the potential verb \textit{be} 'can be' replaces the existential predicate as it does for the locative predicate \REF{ex:gba:38}. The negative morpheme is placed, like in verbal predications, clause-finally.  

\ea\label{ex:gba:37}
    \gll wèn ʔá̰-sɛ̀nɛ́  \\ 
          word \textsc{loc-}inside \\
    \glt ‘There is a problem.’
\z 
    

\ea\label{ex:gba:38}
    \gll wèn bé ná  \\ 
         word \textsc{ipfv}:can\_be \textsc{neg}  \\
    \glt ‘There is no problem.’
\z 

\subsubsection{Clauses with no dedicated predicate} \label{sec:gba:2.3.2}

In Gbaya the mere juxtaposition of two specific constituents encodes predication. The first constituent is always a NP, and the second one is either a prepositional noun phrase or a simple term. These constructions express various predications such as possessive \REF{ex:gba:39}, similative, qualitative, and numeral predications. For the negation of these constructions, the potential verb \textit{be} ‘can be’ is placed between the two constituents. It hosts the negative morpheme \textit{ná}, which is placed in clause-final position. The addition of the potential verb is comparable to what happens with the locative predicate and the existential predicate which are not used in negative statements. In addition, adverbial predications, which have only one constituent that functions as a predicate, are negated in the same way, the potential verb being placed before the adverb and the negative morpheme after it. The potential verb which is added in all these clauses is negated by the standard negator \textit{ná} (see \tabref{tab:gba:5}). 

\begin{table}
\fittable{
\begin{tabular}{llll}
\lsptoprule
{Predications} & {in affirmative clauses} & {in negative clauses} & {with preposition}\\
\midrule
{Possessive} & {NP \textit{nɛ̀} NP}  & {NP \textit{bé nɛ̀} NP \textit{ná}} & {\textit{nɛ̀} ‘with’}\\
{Attributive} & {NP \textit{nɛ̀} \textsc{adj.} \textsc{mnlz} } & {NP \textit{bé nɛ̀} \textsc{adj.} \textsc{nmlz} \textit{ná}} & {\textit{nɛ̀} ‘as’}\\
{Similative} & {NP \textit{hé màá gá}}  & {NP \textit{bé hé màá gá ná}} & {\textit{hé… gá} ‘like’}\\
{Numeral}  & {NP \textsc{num}} & {NP \textit{bé}   \textsc{num} \textit{ná}} & {---}\\
{Adverbial}  & {\textsc{adv}} & {\textit{bé}   \textsc{adv} \textit{ná}} & {---}\\
\lspbottomrule
\end{tabular}
}
\caption{Negation of clauses with no dedicated predicate}
\label{tab:gba:5}
\end{table}
In the possessive, similative and attributive clauses the second constituent is a prepositional noun phrase as in the examples below. 
Possessive clauses are common (\ref{ex:gba:39}--\ref{ex:gba:41}).

\ea\label{ex:gba:39} [Context: He is trying to find out if anyone has goats in the village. Somebody warns him.] \\
    \gll Sìngà nɛ̀ dùà  \\ 
          \textsc{pn} with goat \\
    \glt ‘Singa has goats.’
\z 
 
\ea\label{ex:gba:40}
    \gll Sìngà bé nɛ̀ dùà ná  \\ 
         \textsc{pn} \textsc{ipfv}:can\_be with goat \textsc{neg} \\
    \glt ‘Singa has no goats.’
\z 

\ea\label{ex:gba:41}
    \gll tɛ́ mà ngìmbì ʔám bé nɛ̀ tùà ná \\ 
         by\textsc{:tm} \textsc{indf} time 1\textsc{sg} \textsc{ipfv}:can\_be with house \textsc{neg} \\
    \glt ‘At that time I didn't have a house.’
\z 

Attribution is a domain most often used affirmatively (\ref{ex:gba:43}, \ref{ex:gba:45}) and relatively seldom negatively (\ref{ex:gba:42}, \ref{ex:gba:44}). Attributive clauses are not frequent. In this construction the adjectives are nominalized and introduced by the preposition \textit{nɛ̀} used here with a comparative sense ‘as’. Here the use of \textit{nɛ̀} as a comparative marker introducing adjectives \REF{ex:gba:43}---\REF{ex:gba:45} is different from the \textit{nɛ̀} ‘with’ which is part of the predicative possessive construction \REF{ex:gba:39}---\REF{ex:gba:41}. I present two examples, one with a verbal adjective \textit{dúrá} ‘long’ and one with an adjective \textit{tɔ́r} ‘raw’.

\ea\label{ex:gba:43}
    \gll wáŋá hɛ̰̀ nɛ̀ dúráà \\ 
         leaf \textsc{prox} as long:\textsc{nmlz} \\
    \glt ‘This leave is long.’
\z 

\ea\label{ex:gba:42} 
    \gll wáŋá hɛ̰̀  bé nɛ̀ dúráà ná \\ 
         leaf \textsc{prox} \textsc{ipfv}:can\_be as long:\textsc{nmlz} \textsc{neg}  \\
    \glt ‘This leaf is not long.’ 
\z 

\ea\label{ex:gba:45}
    \gll sàɗì hɛ̰̀ nɛ̀ tɔ́ráà  \\ 
         animal \textsc{prox} as raw:\textsc{nmlz}  \\
    \glt ‘This meat is raw.’
\z 

\ea\label{ex:gba:44}
    \gll sàɗì hɛ̰̀ bé nɛ̀ tɔ́ráà ná  \\ 
          animal \textsc{prox} \textsc{ipfv}:can\_be as raw:\textsc{nmlz} \textsc{neg}  \\
    \glt ‘This meat is not raw.’
\z 

Similative clauses encode similarity between two constituents, the second one being surrounded by the discontinous similative adposition \textit{hé...gá} ‘\textsc{simv}…\textsc{simv}’ \REF{ex:gba:47}. Only similative clauses where the second constituent is the reciprocal pronoun \textit{màá} ‘\textsc{recp}’ and which express a comparison of equality are negated in the corpus, \REF{ex:gba:46}.

\ea\label{ex:gba:46}
    \gll yíkwà bé hé màá gá ná \\ 
         face:\textsc{tm}:3\textsc{pl} \textsc{ipfv}:can\_be \textsc{simv} \textsc{recp} \textsc{simv} \textsc{neg} \\
    \glt ‘Their faces are not the same.’ (lit. ‘not like one to the other’) 
\z 

\ea\label{ex:gba:47}
    \gll yíkwà hé màá gá fɛ́t \\ 
          face:\textsc{tm}:3\textsc{pl} \textsc{simv} \textsc{recp} \textsc{simv} all \\
    \glt ‘Their faces are exactly the same.’ (lit. ‘like one to the other’)
\z 

In numeral clauses the second constituent is always a numeral (\ref{ex:gba:49}, \ref{ex:gba:51}), and the negative construction is common (\ref{ex:gba:48}, \ref{ex:gba:50}). 

\ea\label{ex:gba:48}
    \gll tòyó bé náár ná  \\ 
         dog \textsc{ipfv}:can\_be four \textsc{neg}  \\
    \glt ‘The dogs are not four.’
\z 

\ea\label{ex:gba:49}
    \gll tòyó náár  \\ 
         dog four \\
    \glt ‘The dogs are four.’
\z 

\ea\label{ex:gba:50}
    \gll  bêm kɔ̰́à̰  bé tààr ná \\ 
         child of:3\textsc{sg} \textsc{ipfv}:can\_be three \textsc{neg} \\
    \glt ‘He doesn't have three children.’ (lit. ‘his children are not three’)
\z 

\ea\label{ex:gba:51}
    \gll bêm kɔ̰́à̰  tààr \\ 
         child of:3\textsc{sg} three \\
    \glt ‘He has three children.’ (lit. ‘his children are three’)
\z 

There are only a few adverbs that are used predicatively, either as the second element in a juxtaposition construction, or alone in a holophrastic use. Most of these adverbs in a predicative function are found in affirmative constructions, and some are found in interrogative constructions. Only the adverb \textit{hɔ̰́gɔ̰́} ‘like this’ is observed with negation in my corpus (\ref{ex:gba:52}--\ref{ex:gba:53}). This adverb is very commonly negated.


\ea\label{ex:gba:52}
    \gll  bé hɔ̰́gɔ̰̰́ ná  \\ 
         \textsc{ipfv}:can\_be like\_that \textsc{neg}  \\
    \glt ‘It is not like that.’
\z 

\ea\label{ex:gba:53}
    \gll hɔ̰́gɔ̰̰́  \\ 
         like\_that  \\
    \glt ‘It is like that.’
\z 

To sum up, under negation the locative predicate \textit{ʔá̰} and the existential predicate  \textit{ʔá̰-sɛ̀nɛ́} are replaced by the potential verb \textit{be} ’can be’ and the standard negator \textit{ná} in clause-final position. In all other stative predications, the potential verb \textit{be} ’can be’ is placed before the non-verbal predicate or between the two constituents in a juxtaposition construction. With adverbial predications, the potential verb \textit{be} and the standard negator \textit{ná} surround the adverb as in \REF{ex:gba:52}.

%To sum up, except for the locative predicate \textit{ʔá̰} and the existential predicate \textit{ʔá̰-sɛ̀nɛ́} which are removed and replaced by the potential verb, in all other cases, the potential verb is added and placed before the non-verbal predicate, or between the two constituents of a juxtapositive construction, or surrounding the adverb \textit{hɔ̰́gɔ̰́}.
In all cases the negative morpheme is placed clause-finally and the potential verb \textit{be} 'can be' is negated by standard negation. Thus adding it to carry the negative morpheme is the dedicated negative construction for non-verbal clauses.


\subsubsection{Attributive verbal predication}\label{sec:gba:2.3.3}

Gbaya uses the verb \textit{ʔɔ} ‘stand, lay’ with adjectives in a predicative function. In this context, \textit{ʔɔ} is the stative verb ‘be’. 
    %\aut{change [two first sentences] to: Qualitative predication in Gbaya is predominantly verbal, alongside a much rarer non-verbal qualitative predication, which will be presented in section 2.3.2. This qualificative verbal predication reuses the verb ʔɔ, whose basic meaning is "to lie down, to extend," to introduce an adjective, becoming in this usage a stative verb glossed as " be."}
This verb can carry the negative morpheme in the same way as all the processive verbs, but such negative clauses are rather rare \REF{ex:gba:17}. 

\ea\label{ex:gba:17}
    \gll Nàná ʔɔ́ dèngbèŋ ná \\ 
           \textsc{pn} \textsc{ipfv:}be slow \textsc{neg} \\
    \glt ‘The river Nana does not flow slowly.’ (lit. ‘is not slow’)
\z 

The negative morpheme \textit{ná} is simply added to the corresponding affirmative clause \REF{ex:gba:18}, as in the example below about the belly of an animal after being cooked \REF{ex:gba:19}\footnote{Words such as  \textit{ɓɔ̀lɓɔ̀l} \REF{ex:gba:18}---\REF{ex:gba:19}, \textit{zɛ́nbgɛ́zɛ́nbgɛ́} 'fragile' \REF{ex:gba:60}, \textit{yɔ̀kàìyɔ̀kàì} 'silly' \REF{ex:gba:103}, \textit{pálpál} 'clean' \REF{ex:gba:104} are historically reduplications. However, in Modern Northwest Gbaya, the elements they consist of have no referential meaning which is why reduplication is not considered operational on a synchronic level.}.

\ea\label{ex:gba:18}
    \gll  zàŋáà ʔɔ́ ɓɔ̀lɓɔ̀l \\ 
           belly:\textsc{tm}:3\textsc{sg} \textsc{ipfv}:be soft  \\
    \glt ‘Its belly is soft.’ (= ‘fully cooked’) 
\z 

\ea\label{ex:gba:19}
    \gll zàŋáà ʔɔ́ ɓɔ̀lɓɔ̀l ná \\ 
         belly:\textsc{tm}:3\textsc{sg} \textsc{ipfv}:be soft \textsc{neg} \\
    \glt ‘Its belly is not soft.’ (= ‘not cooked enough’)
\z 
To sum up, attributive verbal predications constructed with the verb \textit{ʔɔ} ‘be’ introducing an adjective can be negated, but this is not frequent. Conversely, it is common for properties that correspond to a resultative process to be supported by verbs that are negated in the same way as all verbs (cf. \sectref{sec:gba:2.1.1}).


\subsection{Negation in non-main-clauses}\label{sec:gba:2.4}

Subordinate clauses (SC) introduced by a conjunction can be negated by standard negation, but the possibility of negation of the SC depends on the position of the SC in front of or behind the main clause.

When a SC is in front of the main clause, it can be negated by standard negation \REF{ex:gba:54}. The negative morpheme is placed in final position of the SC and the main clause follows it. 

\ea\label{ex:gba:54}
    \gll  kɔ̀mɛ̀ ʔɔ́ nù ná kɔ́mɛ́ Wàntò kàɗà ʔé \\ 
          when:\textsc{2sg} \textsc{ipfv}:lay ground \textsc{neg} then:\textsc{2sg} \textsc{pn} \textsc{pfv}:finish:\textsc{tm} yet \\
    \glt ‘If you don’t lie down then it will be the end of you, Wanto.’
\z 

When a SC follows the main clause, the negative morpheme always scopes over the verb of the main clause; the SC cannot be negated, see \REF{ex:gba:55}.

\ea\label{ex:gba:55}
    \gll ʔéí ɲɛ́m kèì ʔá yù ʔín fɔ̀rɔ̀ ná \\ 
         \textsc{log.sg} \textsc{ipfv:}be\_able so\_that:\textsc{losg} \textsc{ipfv}:put race \textsc{com}:\textsc{tm} elephant \textsc{neg} \\
    \glt ‘He is not able to race with the elephant.’
\z 

\ea\label{ex:gba:56}
    \gll ʔám ʔíŋ kɔ́m mà mɔ̀ há̰  mɔ́-dɛ̀ sɛ̌n ʔòyáà ná \\ 
         \textsc{1sg} \textsc{ipfv}:know as\_for:\textsc{1sg} \textsc{indf} thing so-as-to \textsc{ctg.ipfv}-do in:\textsc{tm} remainder:\textsc{def} \textsc{neg} \\
    \glt ‘Personally, I don’t know what I will do to correct this mistake.’
\z


\ea\label{ex:gba:57}
    \gll ʔám gbɛ́ sàɗì há̰  kóò kɔ́m ɲɔ́ŋ ná \\ 
         \textsc{1sg} \textsc{ipfv}:kill animal so-as-to wife of:1\textsc{sg} \textsc{ipfv}:eat \textsc{neg} \\
    \glt ‘I did not kill any animal to feed my wife.’
\z 

Purpose SC as \REF{ex:gba:56} and \REF{ex:gba:57} above, which is placed behind the main clause, can be negated by changing the conjunction. To introduce a negative purpose clause, there is a new conjunction specific to negation \textit{tá} {\textasciitilde} \textit{tɛ́} ‘so as not to’ or ‘lest’ that replaces the conjunction \textit{há̰} ‘so-as-to’ of the affirmative counterpart. In this use, the negative morpheme \textit{ná} which is placed clause-finally at the very end of the sentence always and refers to the verb of the SC, is optional. Most often the negative morpheme is absent \REF{ex:gba:59}, and its presence is redundant with the negative purpose conjunction \REF{ex:gba:58}.

\ea\label{ex:gba:58}
    \gll  ʔà̰	ɓɔ́n nù ʔínàm hɛ̰̀ tá̰à̰  sí ná \\ 
          \textsc{3sg} \textsc{ipfv}:remain ground \textsc{com:tm:1sg} here so\_as\_not\_to:\textsc{3sg} \textsc{ipfv}:go\_away \textsc{neg} \\ 
    \glt ‘She stays here with me so that she does not leave.’
\z 

\ea\label{ex:gba:59}
    \gll  wà hùɲà náŋwà tá kórò ŋmáwá \\ 
          \textsc{3pl} \textsc{pfv:}detach:\textsc{tm} foot.\textsc{3pl} so\_as\_not\_to rain \textsc{ipfv}.hit.\textsc{3pl} \\
    \glt ‘They hurried so that the rain did not get them wet.’
\z 

The linked AM forms that create dependence without any conjunction do not accept negation, except for the Hypothetical form. The negative hypothetical clause is always placed in front of the main clause as it has been presented (cf. \REF{ex:gba:8} and \REF{ex:gba:9}).

A relative clause (RC), which, in Gbaya, is always introduced by the relative marker \textit{nɛ̀} \textsc{rel,} can be negated. As for any subordinate clause, it depends on the position of the RC within the main clause. When the noun that the RC modifies is the subject of the main verb, the RC is placed in front of the main clause verb and the negative morpheme is placed at the end of the RC. The RC is negated by standard negation \REF{ex:gba:60}. Most often the antecedent of the relative marker is the subject of the main clause verb, but it may also be a topicalized term placed before the verb of the main clause \REF{ex:gba:61}. 

\ea\label{ex:gba:60}
    \gll mà làà nɛ̀ ndótí ná ʔɔ́ zɛ́ngbɛ́zɛ́ngbɛ́ \\ 
         \textsc{indf} garment \textsc{rel} \textsc{ipfv}:thicken \textsc{neg} \textsc{ipfv}:be fragile \\
    \glt ‘A garment which is not thick is fragile.’ [antecedent = subject]
\z 
 %\mm{tilde for reduplication as in ex 53, please check!}
\ea\label{ex:gba:61}
    \gll ndàyá mɔ̀ nɛ̀ dìlà zéí gɛ́r yérè náìí ʔá̰-nàm dɛ́-mɛ́-tɔ̰̀ \\ 
         bottom:\textsc{tm} thing \textsc{rel} lion \textsc{ipfv}:hear voice buffalo \textsc{neg:anaph} \textsc{top-1sg:tm} \textsc{imn-ipfv}-say \\
    \glt ‘The reason why the lion does not hear the voice of the buffalo I am about to say.’ [antecedent = topicalized object]
\z 
%\lv[inline,size=\scriptsize,color=blue!40]{For the form ʔá̰-nàm there is only one complex gloss \textsc{top:1sg:tm}. Please indicate how to fix this.}

But when the noun the RC modifies is placed behind the verb of the main clause, as for any SC in the same position, the negative morpheme placed at the very end of the sentence refers to the verb of the main clause (\ref{ex:gba:62}--\ref{ex:gba:64}). 

\ea\label{ex:gba:62}
    \gll ʔà̰   ʔíŋ fàrà nàá yàkà sí sɛ̀nɛ́ ná ʔòé \\ 
         3\textsc{sg} \textsc{ipfv}:know place \textsc{rel}:3\textsc{sg} \textsc{pfv}:leave:\textsc{tm} \textsc{inf.pfv}:go inside \textsc{neg} too \\
    \glt ‘He does not even know in which place he went.’
\z 

\ea\label{ex:gba:63}
    \gll  ʔám zéí wèn kɔ́mɛ́ nɛ̀mɛ̀ tɔ̰́ìí ná \\ 
         1\textsc{sg} \textsc{ipfv}:hear speech of:2\textsc{sg} \textsc{rel}:2\textsc{sg:tm} \textsc{ipfv}:say:\textsc{anaph} \textsc{neg} \\
    \glt ‘I do not understand what you are saying.’
\z 

\ea\label{ex:gba:64}
    \gll ʔám ʔíŋ tí mɔ̀ nɛ̀ dɛ́m ná \\ 
          1\textsc{sg} \textsc{ipfv}:know identity thing \textsc{rel} \textsc{ipfv}:do:1\textsc{sg} \textsc{neg} \\
    \glt ‘I do not know what is happening to me.’
\z 
 
\subsection{Negative lexicalizations}\label{sec:gba:2.5}

There are two verbs, \textit{mbosi} ‘be absent, be missing’ \REF{ex:gba:74} and \textit{ndaŋ} ‘miss, lack’ \REF{ex:gba:72} (in relation to the numeral \textit{ndáŋ} ‘one’) that express a lack. They can, however, be negated by standard negation (\ref{ex:gba:71}, \ref{ex:gba:73}). 

\ea\label{ex:gba:71}
    \gll sàdì ndáŋ ná  \\ 
          animal \textsc{ipfv}:miss \textsc{neg}  \\
    \glt ‘[After checking] No game is missing.’
\z 

\ea\label{ex:gba:72}
    \gll kàfɛ́ɛ̀ ndàngám mɔ́ tútùyè \\ 
         coffee \textsc{pfv}:miss:\textsc{tm:1sg} in morning \\
    \glt ‘I did not get coffee this morning.’
\z 

\ea\label{ex:gba:73}
    \gll yák-tɛ́-mbòsì mɛ́ mɔ̀ɔ́sá̰  ná  \\ 
         \textsc{imp.pfv-proh}-be\_absent in tomorrow \textsc{neg}  \\
    \glt ‘You must not be absent tomorrow.’
\z 
%\mm{@LSP: for ex 68 we need a way to get rid of the underscore, which is not possible on the example line; a space between the two words would be fine, but inserting a space will cause the LSP latex template to align the glosses wrongly.}
\ea\label{ex:gba:74}
    \gll wànyé mbòsá \\ 
         chief \textsc{pfv}:be\_absent \\
    \glt ‘The village chief is absent.’
\z 

\subsection{Other clausal negation constructions}\label{sec:gba:2.6}

Gbaya uses a complex clause to express impossibility. The main clause with the potential verb \textit{be} ‘can be’ is followed by a subordinate clause introduced by the conjunction \textit{kà} ‘so that’ expressing a consequence \REF{ex:gba:55}. The negative morpheme is clause-final. In such constructions, the two clauses have the same subject that is mentioned before each verb and the potential verb always takes a realis neutral imperfective form: S\textsubscript{1} \textit{bé kà} S\textsubscript{1} \textsc{vb} \textit{ná} (\ref{ex:gba:65}--\ref{ex:gba:67}). 

\ea\label{ex:gba:65}
    \gll ʔám bé kàm nɛ́ ná \\ 
         1\textsc{sg} \textsc{ipfv}:can\_be so\_that:\textsc{1sg} \textsc{ipfv}:go \textsc{neg} \\
    \glt ‘I cannot go.’
\z 

\ea\label{ex:gba:66}
    \gll ʔɛ́ɛ́ bé kɛ̀ɛ̀ bá wéì ná \\ 
         2\textsc{pl} \textsc{ipfv}:can\_be so\_that:\textsc{2pl} \textsc{ipfv}:take husband \textsc{neg} \\
    \glt ‘We cannot get married.’
\z 

\ea\label{ex:gba:67}
    \gll  ʔà̰	bé	kà̰à̰	  yák	ná\\ 
         3\textsc{sg} \textsc{ipfv}:can\_be so\_that:\textsc{3sg} \textsc{ipfv}:leave \textsc{neg} \\
    \glt ‘He can’t leave.’
\z 

Sometimes, the subject of the main clause is not mentioned in the SC, and the construction becomes a set phrase \textit{bé kà} \REF{ex:gba:68}. This construction shows low frequency.


\ea\label{ex:gba:68}
    \gll mɛ́ bé kà bá ná \\ 
         2\textsc{sg} \textsc{ipfv}:can\_be so\_that \textsc{ipfv:}take \textsc{neg} \\
    \glt ‘You cannot take it.’
\z 

It is interesting to point out that unlike Gbaya \MakeUppercase{ɓ}odoe, in Gbaya Yaayuwee, another Northwest Gbaya language, such a set phrase, \textit{bé kà}, is the only construction used. So in Gbaya Yaayuwee this set phrase has become a verbal marker of impossibility \textit{bé{}-kà,} which is not the case in Gbaya \MakeUppercase{ɓ}odoe \citep[see][]{Noss1981}.

\section{Non-clausal negation}\label{sec:gba:3}

\subsection{Negative replies}\label{sec:gba:3.1}

Negative replies to polar questions are not related to the negative morpheme \textit{ná}. There are two negative replies: \textit{wòóyèé} {\textasciitilde} \textit{ʔòóyèé}, \textit{ʔṵ̀ṵ́ʔṵ̀} meaning ‘no’ which are interjections in Gbaya. They correspond to the affirmative replies \textit{ʔı̰́ı̰̀ } {\textasciitilde} \textit{ʔṵ́ṵ̀, h\'{ı̰}\`{ı̰}   }meaning ‘yes’. In response to a question, these negative replies always disagree with the content of the question and never with its polarity.

\subsection{Negative indefinites and quantifiers}\label{sec:gba:3.2}

There are no negative indefinites and quantifiers in Gbaya. To say ‘nobody’ \REF{ex:gba:69} or ‘nothing’ \REF{ex:gba:70} Gbaya uses the nouns \textit{béí} ‘person’ and \textit{mɔ̀} ‘thing’ in a negative clause, most often with the indefinite determiner \textit{mà}.


\ea\label{ex:gba:69}
    \gll béí tɛ́ ná \\ 
         person \textsc{ipfv}:come \textsc{neg} \\
    \glt ‘Nobody comes.’
\z 
 
\ea\label{ex:gba:70}
    \gll mà mɔ̀ bé kútùà hɛ̰̀ ná \\ 
         \textsc{indf} thing \textsc{ipfv}:can\_be house \textsc{prox} \textsc{neg} \\
    \glt ‘There is nothing in this house.’
\z 

\subsection{Negative derivation and case marking}\label{sec:gba:3.3}

No negative derivation is attested in Gbaya, which has very little derivation but very productive compounding. There is no adposition ‘without’ either, but there are a lot of adjectives-adverbs (AA) to express lack. These AA are original terms corresponding to derived terms in other languages: \textit{bɛ́ngɛ́ɛ́} ‘ineffective, without result’, \textit{wálásáyá} ‘without stopping’, \textit{fà̰ṵ̀-fà̰ṵ̀} ‘nonstop’, \textit{kpáká-kpáká} ‘thoughtless, without thinking’, \textit{ɗɔ́k-ɗɔ́k} ‘speechless’, etc. This way of expressing lack takes in account a lot of contexts as for example ‘noiseless’: \textit{ʔíyɔ́ɔ́} (standing or walking), \textit{yɔ̀kɔ̀kɔ̀-yɔ̀kɔ̀kɔ̀} (specific to the way you put your foot on the ground), \textit{géɗééŋ} (gait of an animal), \textit{yɛ́kɛ́t} (quiet), \textit{yɛ́lɛ́lɛ́} (acting slowly). So negative meaning expressing lack is lexicalized into specific AAs. Verbal expressions of absence are discussed in \sectref{sec:gba:2.5}.

There is also a new conjunction specific to negation, \textit{tá} {\textasciitilde} \textit{tɛ́} ‘so as not to’ or ‘lest’, that replaces the conjunction \textit{ha̰} ‘so as to’ of the affirmative counterparts (see \sectref{sec:gba:2.4}).

\section{Other aspects of negation}\label{sec:gba:4}

\subsection{The scope of negation}\label{sec:gba:4.1}

In clausal negation the predicate is always in the scope of negation whether it is the verb of the clause (see \sectref{sec:gba:2.1} and \sectref{sec:gba:2.2}) or the potential verb for non-verbal predications (see \sectref{sec:gba:2.3.1} and \sectref{sec:gba:2.3.2}).

To narrow the scope of negation to a specific constituent, Gbaya creates a discontinuous negation that reuses the potential verb \textit{be} ‘can be’ into a fixed set where this verb has no longer any predicative function. In this construction, the potential verb, which has an imperative imperfective form carrying the polar tone of insistence modality (\textit{béè} [Ø \textsuperscript{\textsc{h}}\textit{be}.\textsc{ins]}), is in clause-initial position and the negative morpheme is in clause-final position. This construction \textit{béè…ná}, which surrounds the clause and which can target any element except the predicate, functions as a constituent negator. This discontinuous negation objects to something else but not to the content of the verb, whatever is the AM form of the verb in the clause. Depending on the context, the constituent in the scope of this negation is most often an adverb (\ref{ex:gba:75}--\ref{ex:gba:76}); an oblique object (\ref{ex:gba:77}--\ref{ex:gba:78}); a way to point out, in case of a verb in realis neutral perfective, that is a delimited process has to be put in perspective \REF{ex:gba:79}, or an element topicalized or focalized (\ref{ex:gba:80}--\ref{ex:gba:81}).

\ea\label{ex:gba:75}
    \gll béè ʔà̰   ʔìŋá mɔ̀ fɛ́t ná \\ 
         \textsc{neg} 3\textsc{sg} \textsc{pfv}:know:\textsc{tm} thing all \textsc{neg} \\
    \glt ‘She does not know everything.’ (lit. ‘She knows things but not all.’)
\z 

\ea\label{ex:gba:76}
    \gll béè ʔám tɛ́-nɛ̀ bí mbàɓì fɛ̀rɛ̀ ná \\ 
         \textsc{neg} 1\textsc{sg} \textsc{vrt.ipfv}-go \textsc{inf.pfv}:play dice near \textsc{neg} \\
    \glt ‘I will go play dice not nearby.’
\z 

\ea\label{ex:gba:77}
    \gll nɛ̀ béè Wàntò gbòó tɛ̀ gíà zǎŋbɛ́ɛ́ ná wó \\ 
         so \textsc{neg} \textsc{pn} \textsc{pfv}:go\_out:\textsc{tm} in\textsc{:tm} hunting bush \textsc{neg} really \\
    \glt ‘So Wanto went in the bush but not really to hunt.’
\z 

\ea\label{ex:gba:78}
    \gll nɛ̀ béè ʔéí tɛ̀ɛ̀ kéí wěn wà ná \\ 
         so \textsc{neg} \textsc{log.sg} \textsc{pfv:}come:\textsc{tm} as\_for:\textsc{log.sg} because\_of:\textsc{tm} 3\textsc{pl} \textsc{neg} \\
    \glt ‘So he came not for them.’
\z 

\ea\label{ex:gba:79}
    \gll ʔá̰  Kàɲátà béè ɲɛ̀ɲá ná \\ 
         and\_now \textsc{pn} \textsc{neg} \textsc{pfv}:be\_distant \textsc{neg} \\
    \glt ‘And [the mount] Kaɲata is not very far away.’ (lit. ‘is far away but not very far away’)
\z 

\ea\label{ex:gba:80}
    \gll béè màrá mɔ̀ mɔ́ ʔɛ̰̀ ʔám tɛ̀ɛ́-zɔ̀k ná \\ 
         \textsc{neg} sort:\textsc{tm} thing \textsc{appos} this\_one \textsc{1sg} \textsc{vrt.pfv}-see \textsc{neg} \\
    \glt ‘[If I was gone], such a thing I will not have seen.’
\z 

When the topicalized term is a modified noun as in \REF{ex:gba:81}, \textit{béè} is repeated surrounding the topic and \textit{ná} placed clause finally while a personal pronoun (\textit{tɛ́w\`{a}}) fills the empty place (note that there is another variant for \textit{mɔ̀ɔ́}: \textit{mɔ̀ ʔá̰}).


\ea\label{ex:gba:81}
    \gll béè dìlà ndɛ̀ɛ̀ tɛ́ ʔîn hɛ̰̀ béè mà {mɔ̀ɔ́} tɛ́wà ná \\ 
         \textsc{neg} lion \textsc{rel:1pl:tm} \textsc{ipfv}:come \textsc{com} here \textsc{neg} \textsc{indf} thing:\textsc{loc} body:\textsc{tm:3pol} \textsc{neg} \\
    \glt ‘The lion I come here with is unable to do something.’ (lit. ‘there is nothing in himself’)
\z 
 
In addition, I want to mention the role that perfective forms play in some cases to relate the negation to another verb. As we recall, perfective forms are blocked in the negative. 
The discussion in \sectref{sec:gba:2.4} explains that when a negative morpheme is placed at the end of a sentence, it is linked to the verb of the main clause if the main clause appears before a subordinate or relative clause. However, this linkage is blocked if the verb of the main clause is in the realis perfective, as this form cannot support negation. In such cases, if a relative clause follows the main clause, the negation at the end of the sentence will instead apply to the verb of the relative clause \REF{ex:gba:82}. Another case of incompatibility between the perfective and negation is observed between coordinated clauses sharing the subject of the first clause; the negation at the end of the second clause cannot apply to its verb if that verb is in the realis perfective.
Therefore, negation scopes over the verb of the first clause, see \REF{ex:gba:83}.


%, as in the case of a postposed RC \REF{ex:gba:82}, or in a coordinated proposition \REF{ex:gba:83}. \mm{If this is mentioned it needs to be explained as well. The reader cannot understand what is intended just by looking at the examples.}
%\aut[inline,size=\tiny]{I propose to end the first sentence after 'another verb. Then replace the following sentence with : Going against the treatment presented in 2.4, where the negative morpheme placed at the end of the sentence is linked to the verb of the main clause when it is placed before a subordinate or relative clause, the presence of the realis perfective on the verb of the main clause, a form that cannot support negation, prevents this linkage. Thus, in the case of a relative clause placed after a main clause whose verb is in the realis perfective, the negation at the end of the clause can only be related to the verb of the RC (82). Similarly, in the case of two coordinated clauses connected by the sharing of the subject of the first clause, the negation placed at the end of the second clause, which should apply to its verb, will not be possible because that verb is in the realis perfective. Therefore, the negation will be related to the verb of the first clause (83)." I propose, if you agree, to change this formulation in another one that add an example with the existential which plays the same role as the perfective forms of the verb to modify the scope of negation (see my mail)}

\ea\label{ex:gba:82}
    \gll ʔɛ́nɛ́ zɔ̀kà béí yàrà yíítòó nɛ̀ kàm ɓír ná. \\ 
         \textsc{2pl} \textsc{pfv}:see:\textsc{tm} person day two \textsc{rel} cassava\_dough \textsc{ipfv}:pass \textsc{neg} \\
    \glt ‘You have seen someone who hasn’t eaten for two days.’
\z 

\ea\label{ex:gba:83}
    \gll mà béí mbíí kùtí mɔ̀ sɛ́ká pı̰̀ı̰̀   gúá nù ná \\ 
         \textsc{indf} person \textsc{ipfv}:sweep first:\textsc{tm} thing and\_then \textsc{pfv}:throw:\textsc{tm} firewood ground \textsc{neg} \\
    \glt ‘Nobody sweeps the place and then throws firewood on the ground.’ 
\z 
 
\subsection{Negative polarity}\label{sec:gba:4.2}

No negative polarity items have been found in Gbaya.

\subsection{Marking of NPs in the scope of negation}\label{sec:gba:4.3}

The marking of NPs is not affected by negation in Gbaya.

\subsection{Reinforcing negation}\label{sec:gba:4.4}

In a negative clause, unlike all the other adverbs, two adverbs \textit{ʔòé} {\textasciitilde} \textit{ʔòí} ‘ also’ and \textit{ʔèá} ‘only’ can be placed right after the negative morpheme. The sequence \textit{ná ʔòé} ‘neither, not also’ (\ref{ex:gba:84}--\ref{ex:gba:85}) is more frequent than the sequence \textit{ná ʔèá} ‘not even’ (\ref{ex:gba:86}--\ref{ex:gba:87}), but both are well-attested in my corpus.

\ea\label{ex:gba:84}
    \gll ʔéí kpá tí sàɗì ná ʔòé \\ 
         \textsc{imps} \textsc{ipfv}:find entity:\textsc{tm} animal \textsc{neg} also \\
    \glt ‘There is no game either.’
\z 

\ea\label{ex:gba:85}
    \gll mà tɔ́rsì bé ná ʔòé wá \\ 
         \textsc{indf} electric\_torch \textsc{ipfv}:can\_be \textsc{neg} also unfortunately \\
    \glt ‘Unfortunately, he does not have a flashlight either.’
\z 

\ea\label{ex:gba:86}
    \gll mɔ̀ ndèì ʔíŋ ndàyá mɔ̀ ná ʔèá \\ 
         thing \textsc{rel:log.sg} \textsc{ipfv}:know bottom:\textsc{tm} thing \textsc{neg} only \\
    \glt ‘Something you do not even know anything about...’
\z 
 
\ea\label{ex:gba:87}
    \gll zɔ̀ɔ̀mí kéí bá kóò héí ná ʔèá \\ 
         elder\_brother of:\textsc{log.sg} \textsc{ipfv}:take wife to:\textsc{log.sg} \textsc{neg} only \\
    \glt ‘His elder brother didn’t even get a wife for him.’
\z 

We can compare the regular order \textit{ʔèá ná} (attested only for \textit{ʔèá}, but impossible for \textit{ʔòé}) to the opposite order \textit{ná ʔèá,} in order to bring out the difference of meanings: \textit{ʔèá ná} (only \textsc{neg}) means ‘not only’ \REF{ex:gba:88}, and \textit{ná ʔèá} (\textsc{neg} only) means ‘not even’ (\ref{ex:gba:86}--\ref{ex:gba:87}).

\ea\label{ex:gba:88}
    \gll ʔàm dɛ́ nɛ̀ tǒm kɔ̀fɛ̀ hɔ̰́gɔ̰́ fà̰à̰   ʔèá ná \\ 
         1\textsc{sg} \textsc{ipfv}:do \textsc{circ} work:\textsc{tm} son-in-law like\_that always only \textsc{neg} \\
    \glt ‘I will not keep on doing the son-in-law’s work like that forever.’
\z 

Placed after the negative morpheme, the scope of these two adverbs seems to refocus on the negative morpheme and no longer on the verb of the clause. 

\subsection{Negation, coordination and complex clauses}\label{sec:gba:4.5}

Only for subordinate purpose clauses, there is a new conjunction specific to negation \textit{tá} {\textasciitilde} \textit{tɛ́} ‘so as not to’ or ‘lest’ that replaces the conjunction \textit{ha̰} ‘so-as-to’ of the affirmative counterparts (see \sectref{sec:gba:2.4}).

\subsection{Miscellaneous aspects of negation}\label{sec:gba:4.6}

\subsubsection{Negative transport}\label{sec:gba:4.6.1}

In Gbaya, there is no negative transport. When a subordinate clause follows the main clause it cannot be negated (see \sectref{sec:gba:2.4}).

\subsubsection{Metalinguistic negation}\label{sec:gba:4.6.2}

No examples of metalinguistic negation have been found in the data.

\subsubsection{Non-negative uses of negatives and expletive negation}\label{sec:gba:4.6.3}

In an isolating language like Gbaya, word order is very important and changing the place of an item can lead to giving it a new function and a new signification, as described in \sectref{sec:gba:2.1.1} for the AM forms of the hypothetical{}-negative and the consecutive. 
When the negative morpheme \textit{ná} is not placed clause-finally, it is used with two functions: (i) a discourse particle or (ii) an element of quantifier adverbs.

{(i)} There are two contexts where \textit{ná} is a discourse particle, depending on its position in the clause: after a noun or after a verb. First, when the negative morpheme \textit{ná} is no longer placed clause-finally but right after a noun, it becomes a discourse marker that identifies this noun as the addressee (\ref{ex:gba:89}--\ref{ex:gba:90}). Doing so, it keeps its negative semantic feature that contributes to the identification of the addressee as not being the person who speaks, but the person to whom something is said. As a discourse marker, \textit{ná} is an affirmative item, ‘\textsc{adre}’ in the glosses \citep[see][]{Roulon-Doko2008}.


\ea\label{ex:gba:89}
    \gll wà ná wà dɛ́ gìmà kpɔ́k kpɔ́k kpɔ́k. \\ 
         \textsc{3pl} \textsc{adre} \textsc{3pl:inj} \textsc{inj.ipfv}:do song one one one \\
    \glt ‘They are told to sing one by one.’
\z 

\ea\label{ex:gba:90}
    \gll kó kàó Ɗà̰à̰   mí ná ʔám dúk.  \\ 
         then:\textsc{pol} maternal\_uncle \textsc{pn} \textsc{1sg} \textsc{adre} \textsc{1sg} \textsc{ipfv}:stay \\
    \glt ‘Then uncle \MakeUppercase{ɗ}à̰à̰   told me to stay.’
\z 

Secondly, \textit{ná} can be placed after the verbs \textit{zei} ‘hear’ and \textit{zɔk} ‘see’ in a specific construction of two clauses, which is different from complement clauses. In Gbaya, complement clauses follow the verb as a direct object without any conjunction. In the first clause these verbs are followed by the neutral direct object \textit{mɔ̀} ‘thing’ while the second one is a coordinate clause introduced by \textit{ká} ‘then’. In both cases \textit{ná} keeps its negative semantics. With the verb \textit{zɔk} ‘see’, \textit{ná} indicates the possibility of seeing X or being aware of the presence\slash coming of X as in \REF{ex:gba:91}; when used with the verb \textit{zei} ‘hear’, \textit{ná} does not indicate hearing but rather understanding as in \REF{ex:gba:92}.

\ea\label{ex:gba:91}[Context: She is with her older sister who takes care of her.]\\
    \gll ʔám zɔ́k mɔ̀ ná kó dàà tɛ̀á \\ 
         \textsc{1sg} \textsc{ipfv}:see thing \textsc{disc} then:\textsc{pol} older\_sister \textsc{pfv}:come \\
    \glt ‘I become aware that the older sister comes [to warn me].’
\z 

In the next example \REF{ex:gba:92}, the first clause is a temporal subordinate and the main clause a coordinate clause. Compare with a clause without \textit{ná} \REF{ex:gba:93}. 

\ea\label{ex:gba:92}[Context: She decides to go to her older sister's house because her husband is never at home.]\newline 
    \gll kàm tɛ́-zɔ̀k mɔ̀ ná wó kà̰à̰   gbòó nɛ̀ tútùyè ɗòŋám \\ 
         when:\textsc{1sg} \textsc{vrt.ipfv}-see thing \textsc{disc} really then:\textsc{3sg} \textsc{pfv}:go\_out in morning after:\textsc{tm:1sg} \\
    \glt ‘When I realize that he has arrived in the morning at home (= behind me).’
\z 

\ea\label{ex:gba:93}[Context: They all think that the older sister is dead. The family is eating.] \newline
    \gll ʔá̰-nɛ̀ bé kàyáìí  kà̰à̰  zɔ́k mɔ̀ wó ká ʔà̰á̰  tɛ́ìí \\ 
         and\_now small eldest:\textsc{anaph}  when:\textsc{3sg} \textsc{ipfv}:see thing really then \textsc{3sg:top} \textsc{ipfv}:come:\textsc{anaph} \\
    \glt ‘Then, as the eldest child looks up, there she is, coming back.’
\z 

I would like to point out that there are some other discourse markers based on the negative morpheme \textit{ná} and the interrogative marker \textit{ndé} that function alike but will not be discussed in this paper.

{(ii)} There are seven quantifier adverbs that are compound words. All of them include the negative morpheme \textit{ná} as the last element of the compound. The first element of these compound adverbs may be an adverb, a nominal phrase or, only in one case (\textit{ɲɛ́m-ná}), a verb which is followed by the negative morpheme, see \tabref{tab:gba:6} for a full listing of these expressions.
%{XXlX}
\begin{table}
\begin{tabularx}{\textwidth}{llQQ}
\lsptoprule
{Adverbs} & {Component stems} & {Literal meaning} & {Meaning}\\
\midrule 
\textit{gèɗɛ̀k-ná\slash ɗɛ́k-ná} & {(a little-\textsc{neg)}} & {‘not a little}’ & {‘a lot’, ‘very’}\\
\textit{sór-mɔ̀-ná} & {(\textsc{dim}-thing-\textsc{neg})} & {‘not a very small thing}’ & {‘very little’}\\
\textit{zɔ̌m-mɔ̀-ná} & {(\textsc{aug.tm}-thing-\textsc{neg})} & {‘not that big}’ & {‘very much’}\\
\textit{ɲɛ́m-ná} & {(\textsc{ipfv.}suit-\textsc{neg})} & {‘not suitable}’ & {‘very much’, ‘too much’}\\
\textit{bé-mɔ̀-ná} & {(small-thing-\textsc{neg})} & {no negative meaning} & {‘a little’}\\
\textit{bé-ɗàŋ-ná\slash bé-ɗàŋ} & {(small-bad-\textsc{neg}\slash small-bad)} & {no negative meaning} & {‘a little’} \\
\textit{bé-nàŋ-ná\slash bé-nàŋ} & {(small-?-\textsc{neg}\slash small-?)} & {no negative meaning} & {‘a little’} \\
\lspbottomrule
\end{tabularx}
\caption{The compound quantifier adverbs}
\label{tab:gba:6}
\end{table}
%\hc{in the second column of table 6, we changed "/" to hyphens since these are glosses of the first column}

% Except \textit{ɲɛ́m-ná} which includes a verb, all the other adverbs are ungrammatical
%     \lv{What does \textit{agrammatical} mean here? Do you mean non-compositional?} 
%     \aut{I mean constructions that do not follow the regular pattern and are therefore considered ungrammatical. As these compound adverbs with the negative morpheme which is not supported by a component which is not a verb as usually.}
%     %\hc{what about 'irregular', or 'not transparent'?}
%     \hc{we could delete the first sentence in this paragraph because its content is already mentioned in the last sentence before the table.}
% constructions. 
Most of the forms cited in \tabref{tab:gba:6} keep a negative semantic feature that contributes to their literal meaning but the reading is a positive one, as mentioned previously. For three of them the negative semantic feature is no longer present (\ref{ex:gba:94}--\ref{ex:gba:95}) and, in addition, two of these have a variant without the negative morpheme with the same meaning.
    % \hc{we deleted the first sentence of this paragraph since all information was already in the last paragraph before the table. We moved the form \textit{ɲɛ́m-ná} there. We made slight changes in the paragraph following the table }


%Most of them keep a negative semantic feature that contributes to their literal meaning but produces a positive meaning, as already mentioned previously. For three of them the negative semantic feature is no longer understood (\ref{ex:gba:94}--\ref{ex:gba:95}) and, in addition, two of these have a variant without the negative morpheme with the same meaning. 

\ea\label{ex:gba:94}
    \gll ʔɛ́ɛ́ dɛ̀ɛ́ tòm ɗɛ́kná \\ 
         \textsc{2pl} \textsc{pfv}:do\textsc{:tm} work a\_lot \\
    \glt ‘We worked hard.’
\z 
%\mm{I have changed the glosses according to what is given in the table, as the previous glosses did not conform to the LGR in terms of the use of hyphens and underscore. I hope this is ok.}
\ea\label{ex:gba:95}
    \gll  há̰  késí bémɔ̀ná \\ 
         so\_that \textsc{ipfv}:dry a\_little \\
    \glt ‘[He puts locusts next to the fire] so that they dry a little.’
\z 

In this section, I presented the non-negative uses of the negative morpheme. They illustrate a common syntactic and pragmatic feature of Gbaya that highlights the polyfunctionality of the negative morpheme.
%Using the negative morpheme out of scope of negation as it has been presented, describes a usual device in the Gbaya syntax that highlights an important feature of their conceptualization.
%\lv[inline,size=\scriptsize,color=blue!40]{It is hard for me to understand this sentence. Here is what I think it might mean:\newline In this section, I presented the non-negative uses of the negative morpheme. They illustrate a common syntactic and pragmatic feature of Gbaya that highlights the polysemy of the negative morpheme.}

\subsubsection{Elements that can follow the negative marker}\label{sec:gba:4.6.4}

There are seven sentence particles which, in affirmative as in negative clauses, are placed clause-finally, just after the negative morpheme and the adverbs \textit{ʔèá} or \textit{ʔòé} when appropriate \REF{ex:gba:88}. These particles show the point of view of the speaker, or of the actor of the event (\ref{ex:gba:96}--\ref{ex:gba:98}).

\ea\label{ex:gba:96}
    \gll ʔá̰  tɔ̰́ ná wá \\ 
         3\textsc{sg} \textsc{ipfv}:say \textsc{neg} unfortunately \\
    \glt ‘Unfortunately, she does not say anything.’
\z 
 
\ea\label{ex:gba:97}
    \gll mà mɔ̀ bé kútùà h\`{ı̰}\'{ı̰}  ná wó \\ 
         \textsc{indf} thing \textsc{ipfv}:can\_be house \textsc{loc.anaph} \textsc{neg} really \\
    \glt ‘There is really nothing in that house.’
\z 

\ea\label{ex:gba:98}
    \gll ʔéí tɛ́-dɛ̀ tùà ná yè \\ 
         \textsc{log.sg} \textsc{vrt.ipfv}-do house \textsc{neg} finally \\
    \glt ‘[Since there is a cave] finally I will not build a house.’
\z 
In negative clauses too, sentence particles can combine like they usually do in affirmative clauses \REF{ex:gba:99}.

\ea\label{ex:gba:99}
    \gll sàɗì mɔ́ ʔɛ̰̀ ʔɛ́nɛ́ ɲɔ́ŋ ná ʔòé wó yɛ̀ \\ 
         animal \textsc{appos} this\_one 2\textsc{pl} \textsc{ipfv}:eat \textsc{neg} also really indeed \\
    \glt ‘Indeed, this animal you do not actually eat either.’
\z 

The interrogative marker \textit{ndé} glossed ‘\textsc{q}’ is always at the last position of a clause; nothing can follow it. In negative clauses too, it is placed after the negative morpheme and put at the last position, clause-finally (\ref{ex:gba:100}--\ref{ex:gba:101}). For negative interrogatives, see \sectref{sec:gba:2.2}. 

\ea\label{ex:gba:100}
    \gll ʔɛ́nɛ́ zɔ́káà ná ndé? \\ 
         2\textsc{pl} \textsc{ipfv}:see:3\textsc{sg} \textsc{neg} \textsc{q} \\
    \glt ‘Have you not seen him?’
\z 

\ea\label{ex:gba:101}
    \gll zúà tɛ́-gbà ná ndé? \\ 
         head:\textsc{tm}:3\textsc{sg} \textsc{vrt.ipfv}-split \textsc{neg} \textsc{q} \\
    \glt ‘Won't he have a headache?’
\z 
 
Two other interrogative adverbs \textit{hégè} ‘how’ \REF{ex:gba:102} and \textit{wěn-gè} ‘why’ have the same clause-final position.

\ea\label{ex:gba:102}
    \gll ʔám tɛ́-tɔ̰̀ bɔ̀ɔ̀ ná hégè? \\ 
         1\textsc{sg} \textsc{vrt.ipfv}-say thus \textsc{neg} how \\
    \glt ‘Why would I not say it?’ 
\z 

There are some cases of right detachment when an adjunct is placed outside the boundaries of the clause, i.e. after the negative morpheme. It may be an adjective \REF{ex:gba:104}, an adverb \REF{ex:gba:103} or a NP that adds \textit{a posteriori} some clarifications to the content of the clause \REF{ex:gba:105}. In Gbaya, this is not a device for topicalization \citep[120]{Creissels2006}.

\ea\label{ex:gba:103}
    \gll mɛ́ ʔíŋ ʔòó mɔ̀ ná yɔ̀kàìyɔ̀kàì géè \\ 
         2\textsc{sg} \textsc{ipfv}:know remainder:\textsc{tm} thing \textsc{neg} silly simply \\
    \glt ‘You do not know anything, you are simply silly.’
\z 
%\mm{Use of tilde for reduplication as in 53, please check!}
\ea\label{ex:gba:104}
    \gll kánáŋáà bá ndîŋ ná pálpál \\ 
         ankle:\textsc{tm}:3\textsc{sg} \textsc{ipfv}:take dirty \textsc{neg} clean \\
    \glt ‘His ankles are not dirty, they are clean.’ 
\z 

\ea\label{ex:gba:105}
    \gll ʔá̰  gbó kɔ̰́à̰  zân ɓɔ̀ ná mɔ́-dɛ̀ tòm kɔ́ wànyé \\ 
         \textsc{sg} \textsc{ipfv}:go\_out as\_for:3\textsc{sg} bush again \textsc{neg} \textsc{inf.vrt}-do work of chief \\
    \glt ‘He no longer goes to the bush to work for the chief.’
\z 

\section{Summary}\label{sec:gba:5}

In Gbaya, only verbs can be negated by the negative morpheme \textit{ná}. Most often the constructions are symmetric, but there is an asymmetry in the paradigm in the realis and command moods, as all perfective forms are blocked in the negative. There are different forms in non-declaratives, i.e. the command mood. In order to negate stative predications, the potential verb \textit{be} `can be' has to be used. There is a constituent negator that negates any element except a predicate. 
The systematic clause-final position of the negative morpheme and its last position in the verbal phrase characterizes the negation system in Gbaya. 
The negation strategies of this language are summarized in \tabref{tab:gba:7}.



\begin{sidewaystable}\small
\begin{tabularx}{\textwidth}{lXllll}
\lsptoprule
\multicolumn{3}{l}{\bfseries Negation strategy} & {\bfseries Function} & {\bfseries Chapter} & {\bfseries Examples}\\
\midrule
\multicolumn{3}{l}{negative morpheme \textit{ná} in clause-final position} & realis mood 2 forms (\textsc{ipfv}, \textsc{perm}) & \sectref{sec:gba:2.1.1} & {\REF{ex:gba:1}, \REF{ex:gba:5}, \REF{ex:gba:7}, \REF{ex:gba:9}} \\
&&& Virtual mood 2 forms (\textsc{vrt}) & \sectref{sec:gba:2.1.2} & {\REF{ex:gba:10}, \REF{ex:gba:12}} \\
&&& \textsc{hypo-neg}  & \sectref{sec:gba:2.1.2} & \REF{ex:gba:9} \\
&&&  Command mood (\textsc{imp-ipfv}) & \sectref{sec:gba:2.2} & \REF{ex:gba:23} \\
&&&  \textit{Vetitive} \textsc{ijt-neg} & \sectref{sec:gba:2.2} & \REF{ex:gba:25} \\
&&&  \textit{Prohibitive} \textsc{proh} & \sectref{sec:gba:2.2}& {\REF{ex:gba:27}, \REF{ex:gba:28}, \REF{ex:gba:29}} \\
&&&  Stative verb ʔɔ (\textsc{ipfv,} \textsc{vrt)}  & \sectref{sec:gba:2.2} & {\REF{ex:gba:20}, \REF{ex:gba:22}} \\
\midrule
\multirow{9}{*}{\shortstack[l]{potential verb \textit{be} \\ + negative \\ morpheme \textit{ná} \\in clause-final \\position}} & \multirow{2}{\linewidth}{verb \textit{be} add to the predicate} & nɛ̀ \textsc{id} & Identificational & \sectref{sec:gba:2.3.1}& \REF{ex:gba:31}, \REF{ex:gba:33}\\
\cmidrule{3-6}
 &  & \shortstack[l]{ ʔà̰á̰ \textsc{prst-loc},\\ ʔá̰-hḭ̀ḭ́  \textsc{prst-cons}} & Presentatives~ & \sectref{sec:gba:2.3.1} & \REF{ex:gba:35}, \REF{ex:gba:36}, \REF{ex:gba:37}\\
\cmidrule{2-6}
& \multirow{2}{\linewidth}{verb \textit{be} replace the predicate} & ʔá̰ \textsc{loc} & Locative & \sectref{sec:gba:2.3.1} & \REF{ex:gba:39}\\
 &  &   ʔá̰-sɛ̀nɛ́ \textsc{ex} & Existential & \sectref{sec:gba:2.3.1} & \REF{ex:gba:41}\\
\cmidrule{2-6}
& \multirow{4}{\linewidth}{verb \textit{be} between the two  juxtaposed elements} & NP nɛ̀ NP & Possessive & \sectref{sec:gba:2.3.2} & \REF{ex:gba:43}, \REF{ex:gba:44}\\
&  & NP nɛ̀ \textsc{adj.nmlz} & Attributive & \sectref{sec:gba:2.3.2} & \REF{ex:gba:45}, \REF{ex:gba:47}\\
&  & NP hé NP gá & Similative & \sectref{sec:gba:2.3.2} & \REF{ex:gba:49}\\
&  & NP \textsc{num} & Numeral & \sectref{sec:gba:2.3.2} & \REF{ex:gba:51}, \REF{ex:gba:53}\\
\cmidrule{2-6}
& holophrastic use & \textsc{adv} (= hɔ̰́gɔ̰́) & Adverbial & \sectref{sec:gba:2.3.2} &  \REF{ex:gba:55}\\
\midrule

\multicolumn{3}{l}{P1 + verb potential \textit{be} \textit{kà} P2 + negative morpheme \textit{ná}} & Impossibility & \sectref{sec:gba:2.6} & (\ref{ex:gba:65}--\ref{ex:gba:68})\\
\midrule
\multicolumn{3}{l}{fixed set: \textit{béè} C \textit{ná}} & \shortstack[l]{targeting any constituent (C) \\except the verb} & \sectref{sec:gba:4.1} & (\ref{ex:gba:75}--\ref{ex:gba:81})\\
\midrule
\multicolumn{6}{l}{{\bfseries Negative lexicalizations} } \\
\multicolumn{3}{l}{\itshape mbosi} & ‘be absent, be missing’ &  \sectref{sec:gba:2.5} & \REF{ex:gba:71} \\
\multicolumn{3}{l}{{\itshape ndaŋ}} & ‘miss, lack’ &  & \REF{ex:gba:73} \\
\multicolumn{3}{l}{conjunction \textit{tá} {\textasciitilde} \textit{tɛ}} & {‘so as not to’ or ‘lest’} & \sectref{sec:gba:2.4} & \REF{ex:gba:58}, \REF{ex:gba:59}\\
\multicolumn{3}{l}{\textit{wòóyèé} {\textasciitilde} \textit{ʔòóyèé}, \textit{ʔṵṵʔṵ}} & {negative replies ‘no’} & \sectref{sec:gba:3.1} & \\
\lspbottomrule
\end{tabularx}
\caption{Summary of negative strategies in Gbaya}
\label{tab:gba:7}
\end{sidewaystable}





\section*{Abbreviations}

\begin{tabularx}{.45\textwidth}{lQ}
\scshape 1 & first person\\ 
\scshape 2 & second person\\ 
\scshape 3 & third person\\ 
\scshape Ø & empty place\\ 
\scshape aa & adjective-adverb\\ 
\scshape add & additive\\ 
\scshape adj & adjective\\ 
\scshape adre & addressee\\ 
\scshape adv & adverb\\ 
AM & aspect-mood\\ 
\scshape anaph & anaphoric\\ 
\scshape appos & appositive preposition\\ 
\scshape aug & augmentative\\ 
\scshape circ & circumstantial\\ 
\scshape com & comitative\\ 
\scshape cond & condition\\
\scshape cons & consecutive\\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape ctg & contingent\\ 
\scshape def & definite\\ 
\scshape dim & diminutive\\ 
\scshape disc & discourse marker\\ 
\scshape ex & existential\\ 
\scshape h & high tone\\ 
\scshape hypo & hypothetical\\ 
\scshape id & identificator \\ 
\scshape ijt & injunctive\\ 
\scshape imn & imminent\\ 
\scshape imp & imperative\\ 
\scshape imps & impersonal pronoun\\ 
\scshape inch & inchoative\\ 
\scshape indf & indefinite\\ 
\scshape inf & infinitive\\ 
\scshape ipfv & imperfective\\ 
\scshape known & known\\ 
\end{tabularx}


\begin{tabularx}{.5\textwidth}{lQ}
\scshape l & low tone\\ 
\scshape lh & low-high tone\\ 
\scshape loc & locative\\ 
\scshape log & logophoric\\ 
\scshape neg & negation, negative\\ 
\scshape nmlz & nominalizer, nominalization\\ 
%NP & noun phrase\\ 
\scshape pn & proper name\\ %hc change to NAME
\scshape num & numeral\\ 
\scshape perm & permanent\\ 
\scshape pers & personal pronoun\\ 
\scshape pfv & perfective\\ 
\scshape pl & plural\\ 
\scshape pol & polite form\\ 
\scshape prog & progressive\\ 
\scshape proh & prohibitive\\ 
\scshape prox & proximal \\ 
\end{tabularx}
\begin{tabularx}{.5\textwidth}{lQ}
\scshape prst & presentative\\ 
\scshape q & interrogative particle\\ 
\scshape rc & relative clause\\ 
\scshape recp & reciprocal\\ 
\scshape rel & relative\\ 
\scshape rpt & repetitive\\ 
\scshape sc & subordinate clause\\ 
\scshape sg & singular \\ 
\scshape simv & similative\\ 
\scshape svc & serial verb construction\\ 
\scshape tm & tonal marker \\ 
\scshape top & topic \\ 
\scshape V & vowel \\ 
\scshape \textsc{vb} & verbal base \\ 
\scshape vrt & virtual \\
\\
\end{tabularx}


\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
\end{document} 
