\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233977}

\author{Sami Honkasalo\affiliation{University of Helsinki}}
\title{Negation in Geshiza}  
\abstract{This chapter offers a typologically and functionally motivated description of negation in Geshiza, a Horpa language of the Trans-Himalayan (Sino-Tibetan) language family. Geshiza has a wide range of negative morphology and negative constructions, including four distinct negator prefixes and two dedicated negative verbs. 
Morphologically, negation in Geshiza belongs to the realm of verbs, contrasting with nouns and adjectives that lack dedicated means for being negated category-internally.
The large inventory of negative morphology and particular constructions, e.g. the linking of negation and controllability, make negation in Geshiza a topic of interest to typologists. At the same time, the conservative negative morphology of Geshiza contributes towards understanding the history of negation in the Trans-Himalayan language family.

\keywords{negation, Geshiza, Horpa, Gyalrongic, languages of China}
}


\IfFileExists{../localcommands.tex}{
 \addbibresource{../localbibliography.bib}
 \input{../localpackages}
 \input{../localcommands}
 \input{../localhyphenation}
 \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle
\section{The language}


Geshiza (Glottocode: \href{https://glottolog.org/resource/languoid/id/gesh1238}{gesh1238}) is a Horpa language of the Gyalrongic branch of the Trans-Himalayan/Sino-Tibetan language family (see \citealt{Sun2000a, Sun2000b}  for establishing Gyalrongic as a sub-branch of the language family). Geshiza remains understudied, the language being spoken in Danba County of Sichuan Province in the People’s Republic of China. In the Horpa language cluster, Geshiza belongs to the Central Horpa branch \citep{Sun2019}, which makes it closely related to Stau, another partially documented Horpa lect. The language has approximately 8000 speakers who are officially classified as Tibetan by the Chinese State and also self-identify with the label. The Geshiza are predominantly argiculturalists who inhabit a mountain valley with prominent altitudinal differences. The geographical features of the homeland are encoded in the Geshiza language, most prominently in the form of ``orientational" prefixes in the verbal system.

Geshiza is one of the two main Horpa varieties spoken in Geshiza valley, the other one being Dandong. The two form a dialect continuum with a high degree of mutual comprehensibility. This chapter uses Geshiza as a shorthand for Eastern Geshiza (Glottocode: \href{https://glottolog.org/resource/languoid/id/east2851}{east2851}) spoken in Balang (Ge. \textit{bəra}) Village at the proximity of Danba County Town. Consequently, it is worth noting that not everything discussed herein necessarily applies to all Geshiza dialects. 


When examined against the \citet{UNESCO2003} criteria, Geshiza appears endangered. This, however, has taken place only recently, at least in its more dramatic forms. The endangerment of Geshiza reflects an ongoing widespread loss of languages in China that is especially happening in the Tibetosphere \citep [24]{Roche2017}. As a result of extensive ongoing language contact with Chinese, Geshiza speakers are increasingly shifting into Mandarin Chinese, a change that is most prominently seen among the young and children. While Sichuanese Mandarin is gaining ground, the use of Standard Mandarin is also expanding due to education and media. In contrast, the Geshiza generally lack knowledge of any Tibetic languages in spoken or written form, even though Tibetic languages, such as a Kham lect known as the ``Twenty-four villages’ patois," are spoken in the immediate neighborhood of the Geshiza homeland.

Grammatically, Geshiza exhibits many features known from earlier work on Gyalrongic and more broadly, Qiangic languages, a proposed branch of Trans-Himalayan (see  \citealt{Chirkova2012} for a critique on Qiangic as a valid genealogical unit). Geshiza has an extensive phoneme inventory comprising 8 full vowel phonemes and 37 full consonant phonemes, but tone plays no distinctive function in the language. While the morphology of nouns and other related word classes relies primarily on enclitics, Geshiza verbs show extensive morphological operations by means of affixation, stem modification, reduplication, and suppletion. 

\citet{Honkasalo2019} is a culturally anchored descriptive grammar aiming at sketching Geshiza's main grammatical features. The present work builds on this grammar, the only work of previous research that extensively discusses negation in a Horpa language. Negation is also briefly mentioned in \citet{Duoerji1998}, a grammatical sketch of Eastern Geshiza written in Chinese.
The description and analysis offered here is based on the author's \textit{in situ} fieldwork among the Geshiza. The source materials include recorded texts from various speech genres, supplemented by monolingual elicitation and overheard examples. The monolingual fieldwork strategies include grammaticality judgements and asking Geshiza speakers to generate examples with selected words and expressions.

From the viewpoint of negation, Geshiza is of interest due to its multitude of negative constructions and rich negative morphology. Also, since the negative morphology is generally conservative, it also contributes towards understanding the history of the Trans-Himalayan language family. The description of negation in Geshiza in this chapter is based on the questionnaire provided by the editors (\citetv{chapters/questionnaire}) and the structure of the chapter follows the model given in the questionnaire.

\section{Clausal negation} 
This section discusses clausal negation in Geshiza. Six primary means for clausal negation exist in the language. The inventory includes four negative suffixes and two dedicated negative verbs: the negative copula \ipa{mɲa} \textasciitilde \ \ipa{mja} and the negative existential verb \ipa{ma}. 

\subsection{Standard negation}
\subsubsection{General principles of standard negation}
Standard negation, namely the general and productive way(s) of negating verbal declarative main clauses \citep{Miestamo2005, Payne1985} is expressed by two verbal prefixes in Geshiza: \ipa{mi-} and \ipa{mɛ-}. Standard negation is paradigmatically symmetric in the language, since all grammatical categories, including evidentiality, also occur in negative contexts. 

When both a main verb and an auxiliary verb are present, the negators adjoin the main verb if the auxiliary is bound, and the auxiliary if the auxiliary is independent. To illustrate, with the bound causative auxiliary \ipa{-pʰə}, the main verb hosts the negator, as in the example pair (\ref{placement3}) and (\ref{placement4}). It should be noted that the examples include modal discourse enclitics, also known as ``(sentence-)final particles" in the literature, conveying the speaker’s attitude, emotional state, and certainty (see \citealt[574--579]{Honkasalo2019} for an overview). Further research is needed to analyze the ways in which the ubiquitous modal discourse enclitics interact with polarity in Geshiza.

\ea 
\label{placement3}
\gll {vʑa=tʰə}\textasciitilde {tʰə} 		{jindɔ} 	{spa-pʰə-ræ=mɔ}.\\  
     warm.season=\textsc{top\textasciitilde red}  strongly be.thirsty-\textsc{caus.aux.npst.}3-\textsc{sens=mod}\\
\glt `One feels very thirsty in the warm season.' (elicited)
\z

\ea
\label{placement4}
\gll {rtso=nɔ}	{mi-spa-pʰə-ræ=je.}\\  
     cold.season=\textsc{top.cntr} \textsc{neg-}be.thirsty-\textsc{caus.aux.npst.}3-\textsc{sens=mod}\\
\glt `One does not feel very thirsty in the cold season.' (elicited)
\z

\noindent
In turn, the experiential perfect is a complex aspect formed with a self-standing auxiliary \ipa{zda} `to have done something' and a main verb in the perfective aspect, as in (\ref{placement1}). The negator attaches to the independent auxiliary when the perfective aspect is negated, as in (\ref{placement2}).

\ea
\label{placement1}
\gll {sopʰo} {æ-li} {dæ-ɕʰoŋ} {zda}. \\  
     \textsc{topn} one-\textsc{clf}.time  \textsc{pfv}-go.\textsc{pst.1} \textsc{aux.exp.perf}\\
\glt `I have been to Suopo once.' (personal history)
\z

\ea
\label{placement2}
\gll {sopʰo} {dæ-ɕʰoŋ} {mi-zda}. \\  
     \textsc{topn}  \textsc{pfv}-go.\textsc{pst.1} \textsc{neg-aux.exp.perf}\\
\glt `I haven't been to Suopo.' (accepted during elicitation; see \ref{placement1})'
\z

\subsubsection{Distribution of the negators in standard negation}
To understand the distributional properties of \ipa{mi-} and \ipa{mɛ-}, it is helpful to briefly investigate negation in Gyalrongic languages and the tense-aspect system of Geshiza. In East Gyalrongic (also known as Core Gyalrongic) languages, the selection of a standard negator is usually aspectually determined. For instance, \citet [529]{Prins2011} interprets the difference of the negators \ipa{ma-} and \ipa{ɟi-} in Jiaomuzu Gyalrong  in terms of aspectual contrast in which the former is used for the imperfective and the latter for the perfective aspect. 
Also, in a recent description of Luhuo Stau, a Horpa language related to Geshiza, \citet[381--385]{Tunzhi2019} argues that the difference between the negators \ipa{ma=} and \ipa{mi=} is generally determined on the basis of tense, \ipa{ma=} appearing in past and \ipa{mi=} in non-past constructions.
Such aspect and tense based systems are typologically common and widely attested.


The distribution of \ipa{mi-} and \ipa{mɛ-} is also aspectually determined, although the distributive conditions differ from other previous descriptions of Gyalrongic languages. The language has three prefixally expressed primary aspects: perfective (\ipa{dæ-}), imperfective (\ipa{gæ-}), and prospective (\ipa{zə-}): \ipa{dæ-ŋgu} `I ate',  \ipa{gæ-ŋgu} `I was eating', \ipa{zə-ŋgu} `I am about to eat'. In the whole aspectual repertoire, the primary aspects appear most frequently. They are morphologically non-compositional and set the frame for the structurally more complex secondary aspects. Geshiza also has a ``binary tense system" \citep [48--49]{Comrie1985} comprising the past and non-past tenses expressed through aspiration alternation in the consonants of the verb stem. It should be noted that since not all consonants participate in aspiration alternation, many verbs with no tense distinctions exists in Geshiza. As shown in \tabref{tab:1:distneg}, aspect and tense are strongly linked in the language, allowing no free variation. For instance,  the perfective aspect always co-occurs with the past tense. The imperfective aspect also always occurs with the past tense, but as some of the following examples show, its use is not strictly limited to a past time frame. Such scenarios are nevertheless formally marked as being in the past tense.

\begin{table}
\caption{Distribution of standard negation prefixes in Geshiza}
\label{tab:1:distneg}
 \begin{tabular}{lll} 
  \lsptoprule
    Negative prefix & Aspect & Tense\\ 
  \midrule
    \ipa{mi-}    &   aspectually neutral    &   non-past\\
    \ipa {mɛ-}   &   prospective      &   non-past\\
    \ipa {mɛ-}   &   imperfective     &   past\\
    \ipa {mɛ-}   &   perfective       &   past\\
  \lspbottomrule
 \end{tabular}
\end{table}

As \tabref{tab:1:distneg} illustrates, the distribution of the two negators is explainable as follows. The negator \ipa{mi-} adjoins predicates that lack primary aspect marking and may thus be considered aspectually neutral. In contrast, the negator \ipa {mɛ-} appears in aspectually marked contexts where one of the primary aspect prefixes is present. Alternative explanations, e.g. in terms of tense, fail to explain the distribution of the negators. Examples (\ref{distri1}--\ref{distri4}) illustrate the distribution of negators in aspectually unmarked and prospective contexts. In terms of frequency, negated prospective predicates occur only rarely in Geshiza, contrasting with the relatively common negated imperfective and perfective predicates. Examples of standard negation with the imperfective aspects are offered in (\ref{distri5}) and (\ref{distri6}). The perfective aspect presenting a structural asymmetry is discussed subsequently.\\


\noindent No aspectual marking:

\ea
\label{distri1}
\gll {tʰævære} {lŋa=ɲə} {braŋgu} {dʑi-ræ}. \\  
     recently    child=\textsc{pl} \textsc{topn} \textsc{exvb.}3-\textsc{sens}\\
\glt `Recently, (our) children are (living) in Danba County Town.' (conversation)
\z

\ea
\label{distri2}
\gll {tʰævære} {lŋa=ɲə} {braŋgu} {mi-dʑi-ræ}. \\  
     recently    child=\textsc{pl} \textsc{topn} \textsc{neg-exvb.}3-\textsc{sens}\\ 
\glt `Recently, (our) children are not (living) in Danba County Town.' (accepted during elicitation; see \ref{distri1})
\z


\noindent Prospective aspect:
\ea
\label{distri3}
\gll {tʰævæ=tʰə} 	{zə-stʰoŋ=bɔ.}   \\  
     now=\textsc{top}	\textsc{prosp}-finish.\textsc{npst.1pl=mod}\\ 
\glt `We are about to get finished now.' (accepted during elicitation; see \ref{distri4})
\z

\ea
\label{distri4}
\gll {tʰævæ=tʰə} 	{zə-mɛ-stʰoŋ=bɔ.}\\  
     now=\textsc{top}	\textsc{prosp-asp.neg}-finish.\textsc{npst.1pl=mod}\\ 
\glt `We are not about to get finished now.' (chronicle)
\z

{
\noindent Imperfective aspect:
\ea
\label{distri5}
\gll {tʰævæ=tʰə\textasciitilde tʰə}  {leska=be}     {skəu}  {gæ-rkʰa}.\\  
     now=\textsc{top}\textasciitilde \textsc{red} work=too more \textsc{ipfv}-be.hard.\textsc{pst}\\ 
\glt `Now, work too is more tiring (than it used to be).' (accepted during elicitation; see \ref{distri6})
\z
}

\ea
\label{distri6}
\gll {tʰævæ=tʰə\textasciitilde tʰə}  {leska=be}     {skəu}  {gæ-mɛ-rkʰa}.\\  
     now=\textsc{top}\textasciitilde \textsc{red} work=too more \textsc{ipfv-asp.neg}-be.hard.\textsc{pst}\\ 
\glt `Now, work too is less tiring (than it used to be).' (lit. `Now, work too is being more non-hard.') (local history) 
\z

\subsubsection{Standard negation and symmetry}
Negation in Geshiza largely maintains structural symmetry.  In other words, the affirmative and the corresponding negative constructions are identical, save the presence of a negator in the latter. This can be seen in examples (\ref{symmetry3}--\ref{symmetry4}) with the imperfective aspect (see also the previous examples above).

\ea
\label{symmetry3}
\gll {bəvi}	{skɛ} {gæ-tɕoŋ-sʰə=mde}.\\  
     this.year      more    \textsc{ipfv}-can.\textsc{pst.1-infr=mod}\\
\glt `This year I have been earning more than usual.' (lit. `This year I have been more able.') (accepted during elicitation; see \ref{symmetry4})
\z

\ea
\label{symmetry4}
\gll {bəvi}	{skɛ} {gæ-mɛ-tɕoŋ-sʰə=mde}.\\  
     this.year      more    \textsc{ipfv-asp.neg}-can.\textsc{pst.1-infr=mod}\\
\glt `This year I have been earning less than usual.' (lit. `This year I have been less able.') (conversation)
\z

The perfective aspect, however, is exceptional, since asymmetry commonly occurs in this context. The perfective aspect always requires prefixal marking. The standard perfective marker \ipa{dæ-} is incompatible with the negator \ipa{mɛ-}. Consequently, it is deleted in this context, which results in structural asymmetry. In (\ref{symmetryX}), the speaker negates \ipa{dæ-v-ɕa} `he took' as  \ipa{mɛ-v-ɕa} `he did not take', the symmetrical form  *\ipa{dæ-mɛ-v-ɕa} being ungrammatical:

{\ea
\label{symmetryX}
\gll    {tɕʰu} 		{æqɛ=ræ} 		{wnæ-sqʰa}, 		{wnæ-sqʰa-wtɕʰəu} 	{ŋuə-ræ}.	<{ləupər}> 		{lməu=tʰə} 		{wnæ-sqʰa-wʑæ=zɔ} 		{dæ-v-ɕa.}    {sʰo}			{kʰrə}			{wne}     {mɛ-v-ɕa.}\\  	
so	all=\textsc{lnk}		two-ten		two-ten-six \textsc{cop}.3-\textsc{sens}		shopkeeper		\textsc{3.erg=top} two-ten-four=only	\textsc{pfv-inv}-take.\textsc{pst}.3
	more			yuan			two			\textsc{asp.neg-inv}-take.\textsc{pst}.3.\\ 
\glt `(The bill) was 26 yuan. The shopkeeper took only 24 yuan. He did not take the two yuan.' (chronicle)
\z
}

As the previous examples have shown, the other two aspects compatible with \ipa{mɛ-}, namely the  imperfective and prospective, always retain overt aspectual marking in negation. Therefore, the sole appearance of \ipa{mɛ-} must always be interpreted as negation of the perfective aspect, not as a neutralization of aspectual distinctions in this context. The reasons behind the phenomenon remain unknown. In the closely related Stau language, perfective negation is achieved symmetrically.

Importantly, in contrast to the default perfective prefix  \ipa{dæ-}, no asymmetries arise in negation when the perfective aspect is expressed by secondary markers of perfectivity. In addition to the directionally neutral perfective prefix \ipa{dæ-}, prefixes with orientational value, such as \ipa{rə-} `upwards, away from the river' have a secondary function of marking perfectivity. However, no asymmetry in negations of such instances takes place, as seen in (\ref{symmetry5}) and (\ref{symmetry6}).

\ea
\label{symmetry5}
\gll    {xaræ}	{grə=nɔ} 			{rə-v-læ-sʰi}		{ŋuə-mə-ræ.}\\
\textsc{conj}   boat=\textsc{loc}	\textsc{pfv.dir-inv}-let.3-\textsc{nmlz}	\textsc{cop}.3-\textsc{ep-sens}\\ 
\glt `But (the boatman) took them in.' (accepted during elicitation; see \ref{symmetry6})
\z

\ea
\label{symmetry6}
\gll    {xaræ}	{grə=nɔ} 			{rə-mɛ-v-læ-sʰi}		{ŋuə-mə-ræ.}\\
\textsc{conj}   boat=\textsc{loc}	\textsc{pfv.dir-asp.neg-inv}-let.3-\textsc{nmlz}	\textsc{cop}.\textsc{3-ep-sens}\\ 
\glt `But (the boatman) did not take them in.' (mythical local history)
\z

\subsubsection{Controllability and deviation from standard negation}
In aspectually neutral negation, a negator \ipa{mə-} surfaces in 18 identified cases where \ipa{mi-} would be expected (see \tabref{tab:ges:add2} for a list).\footnote{In the table, verbs that are included with the negator \textit{mə}- attached can only occur in a negative form while the rest can be used both with or without the negative prefix.}
In the context of \ipa{mɛ-}, the set of these 18 verbs behaves regularly. As shown in the pair (\ref{dev}) and (\ref{dev.wrong}), the aspectually neutral negation of the verb \ipa{v-se} `to know' always takes the form \ipa{mə-}, \ipa{mi-} being ungrammatical.

\begin{table}
    \begin{tabularx}{\textwidth}{lX ll}
    \lsptoprule
        \multicolumn{2}{l}{\textbf{Modal verbs}} & \multicolumn{2}{l}{\textbf{Other verbs of low controllability}} \\
    \midrule
        \ipa{mə-grə} &`to not be able' &  \ipa{dzo} &`to fit in' \\ 
        \ipa{mɲə} &`to be able'& \ipa{dʑo} &`to bear, able to put up with' \\ 
        \ipa{ŋi} &`to be fine, all right'& \ipa{mə-kʰe} &`to envy' \\ 
        \ipa{sko} &`to manage, be able'& \ipa{mnæ} &`to (be able) to reach' \\ 
        \ipa{snə} &`to dare'& \ipa{v-ri} &`to find' \\ 
        \ipa{tɕʰa} &`to be able'& \ipa{v-se} &`to know' \\ 
        \ipa{vɕe} &`to need, have to' & \ipa{stʰæ} &`to finish' \\ 
        \ipa{mə-ske} & `should not'   & \ipa{tɕɔ} &`to feel comfortable' \\
    \tablevspace
        \multicolumn{2}{l}{\textbf{Opportunistic perception verbs}} & \\
    \midrule
        \ipa{ntʰje} &`to hear'& & \\ 
        \ipa{vdo} &`to see' & & \\ 
        \lspbottomrule
    \end{tabularx}
    \caption{List of verbs negated with the negator \textit{mə}-}
    \label{tab:ges:add2}
\end{table}

\ea
\label{dev}
\gll    {dæ-lmu-sʰi}  {tɕʰu}  {æqe}  {mə-səu=mɔ.}\\
\textsc{pfv}-forget.\textsc{1sg-infr}    so	all	\textsc{mod.neg}-know.\textsc{npst.1sg=mod}\\ 
\glt `I have forgotten, so I don't know all of it.' (conversation)
\z

\ea[*]{
\label{dev.wrong}
\gll {dæ-lmu-sʰi}  {tɕʰu}  {æqɛ}  {mi-səu=mɔ.}\\
\textsc{pfv}-forget.\textsc{1sg-infr}    so	all	\textsc{neg}-know.\textsc{npst.1sg=mod}\\ 
\glt Intended meaning: `I have forgotten, so I don't know all of it.' (rejected during elicitation)}
\z

Since the distribution of \ipa{mə-} cannot be explained as an automatic phonological process, the subset of 18 verbs requires explaining. 

The negator occurs together with modal verbs (\textit{snə} ‘to dare’, \textit{tɕʰa} ‘to be able’), \textsc{opportunistic perception verbs} (\textit{ntʰje} ‘to hear’,  \textit{vdo} ‘to see’; see \citealt[647]{Honkasalo2019} for details), and several other verbs (\ipa{v-se} `to know', \ipa{tɕɔ} `to be, feel comfortable'). This distribution is hardly accidental. Rather, all three groups denote low level of control by the subject. Consequently, \ipa{mə-} appears as a negator of modal verbs and several verbs of uncontrollable action. As a shorthand it is glossed as \textsc{mod.neg} `negator of modal verbs' here. Previous studies on Trans-Himalayan languages, such as \citet{Sandman2016} on Wutun, have shown that controllability is sometimes embedded into the grammatical systems of these languages. Geshiza negation reflects this as well. The existing forms may be remnants of a more productive historical pattern that has largely disappeared from the language.

A formally and functionally identical cognate of \ipa{mə-} may also exists in some dialects of the closely related Stau language, although vowel harmony that operates in the language, and thus alters the surface level realizations of vowels, presents some challenges for confirming this (personal fieldwork on the Poxiu dialect, possibly 15 identified cases of verbs negated with \ipa{mə-}). The sets of verbs using \ipa{mə-} in both languages largely overlap. In addition, a cognate is attested in the extinct Tangut language where \citet{Jacques2011} reports \ipa{mjɨ¹} ‘negation of modal verbs’ formally resembling \ipa{mji¹} with the function of ‘basic negation’. East Gyalrongic languages, however, lack a corresponding phenomenon. The modal negator consequently constitutes another feature highlighting the closeness of Tangut to West Gyalrongic languages, an ongoing topic of Gyalrongic research.

\subsection{Negation in non-declaratives}
Negation of non-declaratives in Geshiza is determined by the reality status of the underlying mood. Reality status is coded in the language's grammar where two main categories, namely irrealis and realis moods, are identified. Except for the imperative that uses the same prefixes as the indicative, individual moods are in turn marked by vowel prefixes that fuse with the multifunctional ``directional" prefixes of verbs. Realis moods consist of indicative, non-actual realis, and interrogative. Non-actual realis is a term for a Geshiza realis mood referring to habitual action or planned non-initiated action in the future (see \citealt[563--564]{Honkasalo2019} for more). In turn, imperative and optative are classified under irrealis. The irrealis moods have a dedicated structure different from standard negation.

Like many languages of the region, Geshiza general imperatives use a multifunctional verbal prefix that marks orientation, aspect, and mood, as illustrated in (\ref{imp}). Prohibitives (i.e., negative imperatives) cannot be formed with the standard negation strategy, as illustrated in the pair  (\ref{irr.neg}) and (\ref{irr.neg.wrong}). Instead, the construction requires the irrealis negator \ipa{di-} \textasciitilde \ \ipa{dʑi-} that occurs paradigmatically in the same slot with the other negators in the Geshiza verbal template. Note that as illustrated in the examples, the preferred prefix in general imperatives and prohibitives may also differ, a topic that needs further research.

\ea
\label{imp}
\gll {dʑa}    {wə-tʰi.}\\  
    tea  \textsc{imp}-drink.\textsc{npst.2sg}\\ 
\glt `Have some tea!' (addressing the author)
\z

\ea
\label{irr.neg}
\gll {ɕʰældæn}    {noŋ}    {tɕʰæræ}	{mi-ndʑə.}    {dæ-di-tʰi.}\\  
     bottle     inside  thing   \textsc{neg-exvb}.3 \textsc{imp}-\textsc{irr.neg}-drink.\textsc{npst.2sg}\\ 
\glt `There is nothing inside the bottle. Don't drink from it!' (overheard)
\z

\ea[*]{
\label{irr.neg.wrong}
\gll {ɕʰældæn}    {noŋ}     {tɕʰæræ}	{mi-ndʑə.}    {dæ-mi-tʰi.}\\  
     bottle     inside  thing   \textsc{neg-exvb}.3 \textsc{imp}-\textsc{neg}-drink.\textsc{npst.2sg}\\ 
\glt Intended meaning: `There is nothing inside the bottle. Don't drink from it!' (rejected during elicitation; see \ref{irr.neg})}
\z

From a formal viewpoint, the prohibitive exhibits lack of parallelism vis-à-vis the other negative prefixes. Unlike all  other negative prefixes that share the initial \ipa{m} and vary in their vowels, the irrealis negator has a different initial and maintains the vowel \ipa{i} constantly. Seen from a semantic perspective, the prohibitive is generally restricted to negating actions and processes that can be controlled by a sentient being capable of changing its behavior. Consequently, stative verbs, such as \ipa{nəu} `to be deep' and \ipa{pʰru} `to be white', are usually not negated with the prohibitive.

Finally, examined against the typology of \citet{AuweraEtAl2013}, Geshiza belongs to type 2, namely ``the verbal construction of the second singular imperative and a sentential negative strategy not found in (indicative) declaratives." This cross-linguistically most common type is  typical for languages of East and Southeast Asia. 

Optatives, another irrealis mood, are formed with the prefix \ipa{-a-} fusioning with a verbal multifunctional prefix. Semantically the optative carries cupitive and prescriptive tones, the former illustrated in (\ref{optative}). Optatives in Geshiza are frequently used in what \citet{Aikhenvald2010} calls non-canonical imperatives, namely commands and requests to someone other than the addressee. Like the imperative, optatives must be negated with the irrealis negator (\ref{optative.neg}), since the standard negation is deemed ungrammatical, as in (\ref{optative.neg.wrong}).

\ea
\label{optative}
\gll  {n-a-mdær}\\  
       \textsc{pfx}-\textsc{opt}-bite\\ 
\glt `Let (the hornet) bite you!' (adressing the author, joke)
\z

\ea
\label{optative.neg}
\gll  {n-a-di-mdær}\\  
       \textsc{pfx}-\textsc{opt}-\textsc{irr.neg}-bite\\ 
\glt `Let (the hornet) not bite you!' (accepted during elicitation; see \ref{optative})
\z

\ea[*]{
\label{optative.neg.wrong}
\gll  {n-a-mi-mdær}\\  
       \textsc{pfx}-\textsc{opt}-\textsc{neg}-bite\\ 
\glt Intended meaning: `Let (the hornet) not bite you!' (rejected during elicitation; see \ref{optative.neg})}
\z


Cognates for the Geshiza irrealis negator \ipa{di-} \textasciitilde \ \ipa{dʑi-} are widely attested in Trans-Himalayan languages where the proto-form is usually reconstructed as {*ta} \textasciitilde \ {da} and labeled as prohibitive \citep [see][586]{Matisoff2003}. The prohibitive function has been inherited into many modern Trans-Himalayan languages as well. In Geshiza, analyzing  \ipa{di-} \textasciitilde \ \ipa{dʑi-} as a dedicated prohibitive marker would be somewhat misleading. First, as shown above, the  suffix is required by all irrealis moods, including the optative. Second, it also appears in counterfactual irrealis scenarios where no other formal markers of irrealis status are present, as shown in (\ref{irr.scenario}). For this reason, I analyze it more broadly as an irrealis marker, rather than as a dedicated prohibitive. If more languages having reflexes of {*ta} \textasciitilde \ {da} behaving similar to Geshiza are documented, it will be possible to postulate that even in the proto-language, a primary irrealis function likely existed. At the current stage of research the analysis remains inconclusive.


\ea
\label{irr.scenario}
\gll {æpa=ke}     {bɔ} 	{dæ-dʑi-joŋ=ræ} 	{dæ-dʑi-a̯məmoŋ=ræ} {mə-ŋi-ræ=gæ}. 		{ŋɛ}			{væ} 		{ŋuə} 	{tɕʰu} 	{joŋ-mə.} \\  
     father=\textsc{dat}  thus   \textsc{pfx}-\textsc{irr.neg}-say.1=\textsc{lnk} \textsc{pfx}-\textsc{irr.neg}-discuss.1=\textsc{lnk} \textsc{mod.neg}-be.allright-\textsc{sens}=\textsc{mod}
     \textsc{1sg.gen}   dad \textsc{cop}.3 \textsc{dm}	say.1-\textsc{ep}\\ 
\glt `It is not fine for me not to tell thus (e.g. all these things) to my father and discuss (them with him). He is my dad, I am saying!' (conversation) 
\z

As discussed above, questions in Geshiza are realis, with the result that they are negated with the standard negation prefixes in the limited contexts where this is allowed. Negative questions are never used as a strategy for gaining information, shown in the pair (\ref{question}) and (\ref{question.wrong}). 

\ea
\label{question}
\gll  {g-i-vjin}.\\  
      \textsc{ipfv}-\textsc{q}-be.hungry.2\\ 
\glt `Are you hungry?' (overheard, addressing the author, multiple occurrences)
\z

\ea[*]{
\label{question.wrong}
\gll  {g-i-mɛ-vjin}.\\  
      \textsc{ipfv}-\textsc{q}-\textsc{asp.neg}-be.hungry.2\\ 
\glt Intended meaning: `Are you not hungry?' (rejected during elicitation; see \ref{question})}
\z

Rather, negative questions occur in the pragmatic context of rhetorical questioning and ``auto-interrogation" (`I wonder'). This is illustrated in (\ref{auto}) that would be rendered ungrammatical if the auto-interrogation enclitic \textit{=goŋ} was absent.

\ea
\label{auto}
\gll  <{tsoɲe}> {mɛ-stəu} {tɕʰu} 	<{ləusə}>{=wo} 	{æ-mi-v-dæ=goŋ}.\\  
homework	\textsc{asp.neg}-finish.\textsc{pst.1sg}		so		teacher=\textsc{erg}	\textsc{q-neg-inv}-punish.3=\textsc{q}\\ 
\glt `I haven’t finished the homework, so I wonder if my teacher will (lit. won’t) punish me.' (elicited) 
\z

\subsection{Negation in stative predications}\label{neg.stative}
Geshiza existential and copular predications use both standard negation and dedicated expressions. Copular predications in the language contain a copula together with a copular subject and a copular complement. Their functional range covers identity or (in)equation, group (dis)membership or inclusion/exclusion, and attribution. The copula \ipa{ŋuə} `to be' conjugates like an ordinary verb and has a dedicated negative counterpart \ipa{mɲa \textasciitilde \  mja} `not to be'. As shown in the example set (\ref{cop.aff}--\ref{cop.neg.wrong}), the negative copula cannot be replaced by standard negation in Geshiza.

\ea
\label{cop.aff}
\gll  {bɔ}    {gæ-tɕʰæ} {ŋoŋ-ræ=gæ.}\\ 
      thus  \textsc{adjz}-big   \textsc{cop.1-sens=mod}\\ 
\glt `We are that big.' (conversation)
\z

\ea
\label{cop.neg}
\gll  {bɔ}    {gæ-tɕʰæ} {mɲoŋ-ræ=gæ.}\\
      thus  \textsc{adjz}-big   \textsc{neg.cop.1-sens=mod}\\
\glt `We aren't that big.' (accepted during elicitation; see \ref{cop.aff})
\z

\ea[*]{
\label{cop.neg.wrong}
\gll  {bɔ}    {gæ-tɕʰæ} {mi-ŋoŋ-ræ=gæ.}\\
      thus  \textsc{adjz}-big   \textsc{neg-cop.1-sens=mod}\\
\glt Intended meaning: `We aren't that big.' (rejected during elicitation; see \ref{cop.neg})}
\z


While synchronically non-analytical, the Geshiza negative copula possibly originates from a coalescence of a standard negation marker attaching the affirmative copula: {*mV-ŋV} (here V represents unclear exact vowel values of the reconstructed forms). 
The recently sketched Lamo language related to Geshiza and other Gyalrongic languages shows that in some related languages, standard negation can be applied to the copula. Nevertheless, if the fusional hypothesis has historical validity, its future elaborations should offer an explanation for the somewhat unexpected  phonological outcomes of the fusion.

Existential and locative verbs abound in languages of Southwestern China \citep [170]{Sun2001}. Their inventory in Geshiza includes six verbs, the selection of which depends on several factors, such as the  animacy of the referent (animate, inanimate); locus of existence (containment, attachment, etc.); and alienability (alienable, inalienable). \tabref{exvT} summarizes the attested existential and locative verbs together with their negations and basic properties.

\begin{table}
\caption{Existential and locative verbs and their properties in Geshiza}
\label{exvT}
 \begin{tabular}{lllll} 
  \lsptoprule
    Verb & Negation & Animacy & Locus & Alienability\\ 
  \midrule
    \ipa{də}    &   \ipa{ma}    &  inanimate & general & not relevant\\
    \ipa {wi}   &   \ipa{mi-wi}         &   inanimate & attachment & inalienable\\
    \ipa {ndʑə}   &   \ipa{mi-ndʑə}       &   not relevant & container & not relevant\\
   	 \ipa {dʑi}   &    \ipa{mi-dʑi}         &   animate & general & not relevant\\
\ipa {mdʑi}   &    \ipa{mi-mdʑi}         &   animate & group & not relevant\\
  \lspbottomrule
 \end{tabular}
\end{table}

As can be seen, negative existential and locative predications are largely formed analytically by means of standard negation. To illustrate, standard negation applies to the existential-locative verb \ipa {dʑi} in (\ref{exv.standard.neg}).

{%
\ea
\label{exv.standard.neg}
\gll  {e} {bændi}-{næmkʰa=tʰə} {mi-dʑi-ræ} {tʰə} {pʰe} {æ-ɣi=tʰə} {dʑi}.\\ 
      \textsc{inj} \textsc{pn}-\textsc{pn}=\textsc{top}  \textsc{neg-exvb.3-sens}  \textsc{dem} other one-\textsc{clf}.person=\textsc{top}  \textsc{exvb}.3\\ 
\glt `{Bændi}-{næmkʰa} is not (at home), but another person (of his family) is.' (chronicle)
\z
}

In contrast, the general inanimate existential-locative verb \ipa{də} cannot be negated with standard negation (*\ipa{mi-də}), but has a dedicated non-analytic negative counterpart \ipa{ma}, illustrated in (\ref{exv}) and (\ref{negex}). As discussed in \citet{Veselinova2013neg.exist}, this conforms to a general cross-linguistic tendency of applying a special negation strategy for negative existentials.

\ea
\label{exv}
\gll  {vdzi} {æ-ɣi=je} {rjəu} {æ-vtɕa} {bɔtʰə=tʰə} {ɳʐɖolu} {də-ræ}.\\ 
      man  one-\textsc{clf}.person=\textsc{gen}    wife  one-\textsc{clf}.pair like.that=\textsc{top}  custom \textsc{exvb-sens}\\ 
\glt `There is a custom of a man marrying two women.' (ethnographic explanation)
\z

\ea
\label{negex}
\gll  {vdzi} {æ-ɣi=je} {rjəu} {æ-vtɕa} {bɔtʰə=tʰə} {ɳʐɖolu} {ma-ræ}.\\ 
      man  one-\textsc{clf}.person=\textsc{mod}    wife  one-\textsc{clf}.pair like.that=\textsc{top}  custom \textsc{negex-sens}\\ 
\glt `There is no custom of a man marrying two women.' (accepted during elicitation; see \ref{exv})
\z


\subsection{Negation in non-main clauses}
Non-main clauses in Geshiza use standard negation, and the language shows no dedicated negative constructions in this context. Three main subordinate clause types exist in the language: adverbial, complement, and relative clauses. Syntactically, subordinate clauses precede main clauses. Both finite and non-finite subordinate clauses are attested, depending on the construction. Negation in such complex clauses frequently appears in the main clause, rather than in a subordinate clause, such as `I saw that she's not coming'. In reported speech, however, a subtype of complementation with verbs of speech and cognition as the main clause predicates, the subordinate clause commonly appears as negated, as in (\ref{neg.rep.speech}). 

\ea
\label{neg.rep.speech}
\gll  {braŋgu} {mi-ɕoŋ=bɔ} {dæ-joŋ}.\\ 
      \textsc{topn} \textsc{neg}-go.\textsc{npst.1}=\textsc{mod} \textsc{pfv}-say.1\\ 
\glt `I said that I am not going to Danba County Town.' (chronicle)
\z

Negative relative clauses frequently  appear in a double negative construction discussed in \sectref{double}.

\subsection{Negative lexicalizations}
\label{neglex}
Geshiza has relatively few negative lexicalizations, such as \ipa{mkʰuə} `to lack'. Most prominently, the affirmative copula \ipa{ŋuə} has a negative counterpart \ipa{mɲa} \textasciitilde \ \ipa{mja}. In a similar fashion, the language has a dedicated negative existential verb \ipa{ma} that corresponds to the affirmative inanimate existential verb \ipa{də}. These two negative lexicalizations are discussed in detail in \sectref{neg.stative}. The lexicon of the language also includes three identified verbs that must always be negated and can thus be seen as borderline cases of negative lexicalizations (see \sectref{compneg}).

The modal auxiliary verb \ipa{vɕe} `to want, need' has an irregular stem in aspectually neutral negation where \ipa{mə-ɕe} in lieu of *\ipa{mə-vɕe} appears, as in example (\ref{irreg.neg}). In aspectual negation, the fully regular form \ipa{mɛ-vɕʰe} is used, as shown in (\ref{reg.neg}).

\ea
\label{irreg.neg}
\gll  {stɕæn} 		{mə-ɕe}.\\ 
      be.afraid.\textsc{npst.2} \textsc{mod.neg}-need.\textsc{npst}\\ 
\glt `You need not be afraid.' (folktale)
\z

\ea
\label{reg.neg}
\gll  {lmæ=ɲə=ke} {kʰoŋ} {mɛ-vɕʰe} {tɕʰa} {sʰo} {ru-ræ=gæ.}\\ 
     3=\textsc{pl}=\textsc{dat} give.\textsc{npst.1pl} 
     \textsc{asp.neg}-need.\textsc{pst} \textsc{cond}
     more find.\textsc{1sg-sens}=\textsc{mod}\\ 
\glt `If we didn't have to give them (a part of the earnings), I would earn even more.' (conversation)
\z

The irregular negative form of the modal auxiliary cannot be explained as a result of morphophonemics. Rather, it is better viewed as an irregular quasi-fusional form that has some properties of negative lexicalizations. Typologically, negative lexicalizations frequently develop from fusional negated forms. As a well known example, 
the Latin verb \textit{nolō} ‘to wish not, not want’ originates from the negation of \textit{volō} ‘wish, want’ \citep [829] {WaldeHofmann1938}. Geshiza \ipa{mə(-)ɕe} is similar to this. Finally, \citet{Veselinova2013lexicalization} shows that negative lexicalizations are not random, but surface in limited semantic domains. `To not want' discussed above is one such domain, and Geshiza thus reflects this typological tendency.

\section{Non-clausal negation}
This section addresses non-clausal negation in Geshiza. Most negative morphology in the language pertains to the domain of verbs. Save one exceptional case, no dedicated negative indefinites and quantifiers have been identified. Also, non-verbal word classes cannot be morphologically negated, except when individual lexical items derive from verbs.

\subsection{Negative replies}
Geshiza has a negative interjection  \ipa{Ɂ\~əx\~ə} used for forming negative replies to polar questions. The language nevertheless prefers a verb-echo (repetition-type) strategy (see \citealt{Holmberg2016}), namely echoing the verb of the question, either in affirmative or negative form. With a sample of 14 languages, \citet{EnfieldEtAl2019} show all of the analyzed languages allowed both interjection-type answers and repetition-type answers to polar questions, but showed different frequencies for the two. To illustrate the repetition-strategy more common in the source materials collected from Geshiza, the question in (\ref{neg.reply.question}) that serves as a conventional greeting among the Geshiza in the evening when people return home from work is answered both in the affirmative (\ref{neg.reply.question.answer2}) and in the negative (\ref{neg.reply.question.answer1}) by echoing the stative verb \ipa{rka} `to be tired':

\ea
\label{neg.reply.question}
\gll  {d-i-rkʰa}.\\ 
        \textsc{pfx-q}-be.tired.\textsc{pst}\\
\glt `Are you tired?' (lit. `Did you get tired?') (overheard)
\z

\ea
\label{neg.reply.question.answer2}
\gll  {æ-ntsʰæ} {rka-ræ=bɔ.}\\ 
        one-\textsc{clf}.little.bit be.tired.\textsc{npst-sens}=\textsc{mod}\\
\glt `I am a little bit tired.' (addressing the author, notice also how the answer is in the non-past tense)
\z

\ea
\label{neg.reply.question.answer1}
\gll  {mɛ-rkʰa}.\\ 
        \textsc{asp.neg}-be.tired.\textsc{pst}\\
\glt `I'm not tired.' (lit. `I did not get tired.') (overheard)
\z

As mentioned previously, Geshiza never uses negative questions to elicit information. Consequently, in the language it cannot be determined whether negative replies negate the proposition or disagree with the polarity of the question.

\subsection{Negative indefinites and quantifiers}
\label{neg.indef}
Geshiza has few if any dedicated indefinites and quantifiers. Most basic, i.e. non-derived, interrogative pro-forms of Geshiza also carry an indefinite function, a typologically well-known tendency (see \citealt [401]{Dixon2012}, \citealt [170--179]{Haspelmath1997}, \citealt{Kahrel1996}). They are further compatible with negative predicates, but must be accompanied with the contrastive topic marker (\ref{indef1}--\ref{indef2}). 

\ea
\label{indef1}
\gll  {ɲi}	{sʰədə} 	{rə-tjin}.\\ 
        \textsc{2sg} when    \textsc{dir}-come.\textsc{npst}.2\\
\glt `When are you coming?' (elicited)
\z

\ea
\label{indef2}
\gll  {sʰədə=nɔ} 	{rə-mi-tjan=bɔ}.\\ 
        when=\textsc{top.cntr}    \textsc{dir-neg}-come.\textsc{npst}.1=\textsc{mod}\\
\glt `I will never come!' (elicited)
\z

\noindent
The echo-like partially reduplicated \ipa{sʰədəsʰəɲəu} `never' derives from \ipa{sʰədə} `when'  and requires a negative predicate. Exceptionally, \ipa{sʰədəsʰəɲəu} is the only attested negative indefinite pronoun, other interrogative pronouns lacking corresponding derivations. Example (\ref{indef3}) semantically corresponds to (\ref{indef2}).


\ea
\label{indef3}
\gll  {sʰədəsʰəɲəu} 	{rə-mi-tjan=bɔ}.\\ 
        never    \textsc{dir-neg}-come.\textsc{npst}.1=\textsc{mod}\\
\glt `I will never come!' (elicited)
\z

\subsection{Negative derivation and case marking}
Geshiza uses case enclitics for coding semantic roles. The grammar of the language lacks dedicated cases for expressing the absence of something, such as the Finnish Abessive \textit{-ttA} in \textit{ongelmitta} `without problems'. Equally, a without-postposition does not exist in the repertoire of postpositions in Geshiza.

Nouns and adjectives cannot be negated category-internally in Geshiza. Only adjectives and nouns that are derivations may rarely appear with negation in the language. In other words, negating the source category, namely verbs, offers the only pathway for deriving negative nouns and adjectives. The process is consequently represented as verb > negated verb > nominalization or adjectivization, as in \REF{negandderivation}. The nominalizer is almost always the productive agentive nominalizer \ipa{-me}, the rare agentive nominalizer \ipa{-pa} occurring only in few non-productive instances, such as in \ipa{mi-ntɕʰo-pa}, an alternative to \ipa{mi-ntɕʰo-me} `poor person'. 

\ea
\label{negandderivation}
\gll  \ipa{ntɕʰo} > \ipa{mi-ntɕʰo}  > \ipa{mi-ntɕʰo-me}\\ 
have  > \textsc{neg}-have  > \textsc{neg}-have-\textsc{nmlz}\\
\glt `to have’ (verb) > `not to have' (negated verb) > `poor person' (nominalization of the negated verb)
\z

% \TabPositions{.15\textwidth,.4\textwidth,.2\textwidth}
% \ea \label{negandderivation}
% \ipa{ntɕʰo} \tab > \ipa{mi-ntɕʰo} \tab > \ipa{mi-ntɕʰo-me}\\ 
% have   \tab > \textsc{neg}-have \tab > \textsc{neg}-have-\textsc{nmlz}\\
% `to have’  \tab > `not to have' \tab > `poor person' \\ 
% (verb)  \tab {\db}  (negated verb) \tab {\db} (nominalization of the negated verb)\\ 
% \z 

The derivation of negative adjectives is subject to more restrictions. Like other Horpa languages for which research exists, Geshiza has three types of adjectives: prefixed (\ipa{gæ-tɕʰæ} `big'), reduplicated (\ipa{dəu}\textasciitilde \ipa{dəu} `small'), and non-marked (\ipa{rɲoŋba} `old'). Of these, the prefixed and reduplicated adjectives derive from verbs. Only the prefixed adjectives appear with a negator in negative adjectivizations (\ref{negandderivation2}). Similar formations from reduplicated and non-marked adjectives are deemed ungrammatical, with the result that they can only be negated with the negative copula (see \sectref{neg.stative}), a strategy equally applicable to prefixed adjectives as well.

\ea
\label{negandderivation2}
\gll  \ipa{mdze} > \ipa{mi-mdze} > \ipa{gæ-mi-mdze}\\ 
be.beautitul  > \textsc{neg}-be.beautiful > \textsc{adjz-neg}-be.beautiful\\
\glt `to be beautiful’ (verb) > `not to be beautiful' (negated verb) > `non-beautiful' (adjectivalization of the negated verb)
\z

% \TabPositions{.21\textwidth,.5\textwidth,.2\textwidth}
% \ea
% \label{negandderivation2}
% \ipa{mdze} \tab > \ipa{mi-mdze} \tab > \ipa{gæ-mi-mdze}\\ 
% be.beautitul \tab > \textsc{neg}-be.beautiful \tab > \textsc{adjz-neg}-be.beautiful\\
% `to be beautiful’ \tab > `not to be beautiful'\tab > `non-beautiful' \\
% (verb) \tab {\db} (negated verb) \tab {\db} (adjectivalization of the negated verb)\\
% \z

\section{Other aspects of negation}
This section discusses major phenomena that occur in connection with negation in Geshiza. Several of such phenomena rely on pragmatic factors coded by information structure enclitics.

\subsection{The scope of negation}
The scope of negation in Geshiza is manipulated by the use of the information structure enclitics, which often leads into contrastive interpretations, as in (\ref{scope}) with the topic enclitic \ipa{=tʰə}.

\ea
\label{scope}
\gll  {goŋma=tʰə} {mɲa-ræ}. {pʰe} {a̯pən}, {tʰə} {a̯pən} {ŋuə-ræ.}\\ 
emperor=\textsc{top} \textsc{neg.cop}.3-\textsc{sens} other leader \textsc{dem} leader \textsc{cop}.3-\textsc{sens}\\
\glt `He is not an emperor. (He is) a different leader. He is a leader.' (local history)
\z

\subsection{Negative polarity}
Negative polarity refers to expressions attested only in the negative context. In addition to \ipa{sʰədəsʰəɲəu} `never' discussed in \sectref{neg.indef}, the Geshiza source materials contain only one negative polarity item, namely the  discourse intensifier \textit{=m(d)e} `only, no more/other than, apart from, besides’ that requires a negated predicate (\ref{only1}--\ref{only2}). 

\ea
\label{only1}
\gll  {bjəwa} 		{æ-lə=je} 					{tsʰetsæ=mde} 		{mi-ntɕʰo-ræ=mɔ}.\\ 
    mouse		one-\textsc{clf.indf=gen}		life.span=only  \textsc{neg}-have.\textsc{npst.}3-\textsc{sens=mod}\\
\glt `He only has the life span of a mouse.' (i.e., `He will die very soon.') (folktale) 
\z

\ea
\label{only2}
\gll  
{tʰævæ=tʰə} {<sentsʰæntui>} 	{wsʰu=mde}		{mɛ-stoŋ=wo} 	{sʰo} 			{sʰo}		{<sentsʰæntui>} 	{æ-vtɕa}	{ntɕʰoŋ=mɔ.}\\ 
    			now=\textsc{top} production.unit	three=only	\textsc{asp.neg}-finish.\textsc{pst.1pl}=cause	more    more
    			production.unit	one-\textsc{clf}.pairs		have.\textsc{npst.1pl=mod}\\
\glt `Because we have finished only three production units, we still have two more production units (to finish).' (procedure)
\z

In some languages, such as Japanese with \textit{sika} and Tibetan with \textit{ma gtogs}, the morpheme `only' is compulsorily accompanied by a negator, and Geshiza follows this pattern. An interpretation of \ipa{=mde} as an intensifier `even', as in `He doesn't even have the life span of a mouse' for (\ref{only1}) cannot be justified. In (\ref{only3}), the speaker narrates his daily activities when the author was visiting his village. Since the day had been busy for both, there was time to eat cup noodles only, unlike the more elaborate meals that the Geshiza frequently enjoy. Consequently, the speaker conveys the message that no other kind of food was available. Examples like this strongly support the semantic interpretation of `only' for \textit{=m(d)e}.

\ea
\label{only3}
\gll  
{bəsni=tʰə} {mdzo=ke} {<fænbiænmiæn>=mde} {mɛ-ŋgoŋ-sʰə-mə.}\\
today=\textsc{top} lunch=\textsc{dat} cup.noodles=only \textsc{asp.neg}-eat.\textsc{1pl-infr-ep}\\
\glt `Today we had only cup noodles for lunch.' (chronicle)
\z

\subsection{Marking of NP in the scope of negation}
As mentioned previously, negation does not affect case marking in Geshiza. In terms of information structure enclitics, however, negation commonly cooccurs with noun phrases marked with the contrastive topic marker \ipa{=nɔ} (see \sectref{neg.indef}, \sectref{reforce}).

\subsection{Reinforcing negation}
\label{reforce}
Negation in Geshiza is frequently reinforced with two main strategies: the use of an adverb and a versatile discourse enclitic. Geshiza has an inherently reduplicated adverb \ipa{kəŋkən} that together with a negative predicate conveys the meaning `at all', as in (\ref{reinforcing1}).

\ea
\label{reinforcing1}
\gll  {kəŋkən} {lməu} {rejuɕəmæ} {mi-stɕʰə-mə-ræ-jə.}\\
\textsc{adv} \textsc{3.erg} front.teeth \textsc{neg}-show.\textsc{npst}.3-\textsc{ep-sens-rep}\\
\glt `It is said that she does not show her front teeth at all.' (folktale)
\z

While frequently appearing paired with a negative predicate, \ipa{kəŋkən} is not a negative polarity item. In interrogative contexts, the adverb expresses the notion of `on earth', illustrated in (\ref{reinforcing2}). In positive polarity contexts, it conveys the notion of `fully, completely', as in (\ref{reinforcing7}).

\ea
\label{reinforcing2}
\gll  {ɲi}			{kʰɔ}			{kəŋkən}		{sʰə=ke=kʰa}			{stɕæn} 		{jə}.\\
\textsc{2sg} \textsc{inj} \textsc{adv}     who=\textsc{dat}=about fear.\textsc{npst.2} say.3\\
\glt `\ ``Who on earth are you afraid of?'' he said.' (joke)
\z

\ea
\label{reinforcing7}
\gll  {e} {sme} {kəŋkən=tʰə} {wmərja=z}ɔ {dæ-tʰje-sʰi} {ŋuə-ræ=bɔ}, {æ-ŋuə-ræ}.\\
\textsc{dem} woman \textsc{adv}=\textsc{top}     hair=only \textsc{pfv}-become.\textsc{pst}.3-\textsc{nmlz} \textsc{cop}.3-\textsc{sens}=\textsc{mod} \textsc{q}-\textsc{cop}.3-\textsc{sens}\\
\glt `That lady had become completely covered with hair (because of living in isolation away from other human beings and civilization), right?' (folktale)
\z

In addition, the information structure enclitic \ipa{=nɔ} is frequently used to reinforce negation. The enclitic covers a wide semantic range, but I analyze the marking of contrastive topic as the central function. To illustrate, in (\ref{reinforcing3}) involving a conversation of two participants, \ipa{=nɔ} indicates that while the speaker finds Balang Village close to Danba County Town pleasant to live in, other villages further away from the County Town are less so in his opinion.

\ea
\label{reinforcing3}
\gll  {bəra=nɔ} {tɕɔ-re}. {pʰe} {tɕædʑi} {mə-tɕɔ-ræ}.\\
    \textsc{topn=top.cntr} be.pleasant.\textsc{npst-sens.mod} other distant \textsc{mod.neg}-be.pleasant.\textsc{npst-sens}\\
\glt `Balang is pleasant! Other (villages) far away aren't pleasant.' (conversation)
\z

\noindent
The enclitic is frequently found adjoining the constituent preceding the predicate verb, as in (\ref{reinforcing4}) where it emphasizes \ipa{æ-ntsʰæ} `a little bit':

\ea
\label{reinforcing4}
\gll  {ŋa=tʰə} 		{æ-ntsʰæ=nɔ} 					{mɛ-vdo=je.}\\
	\textsc{1sg=top}	one-\textsc{clf}.little.bit=\textsc{top.cntr}	\textsc{asp.neg}-see.\textsc{1sg=mod}\\
\glt `I haven't seen him around at all!' (conversation)
\z

\noindent
When putting emphasis on the predicate verb itself, the verb must be repeated, as illustrated in (\ref{reinforcing5}).

\ea
\label{reinforcing5}
\gll  {ɣæ-ŋguæ} {lxua=nɔ} {mi-lxua-ræ.}\\
    ten-five   appear.3=\textsc{top.cntr}	\textsc{neg}-appear.3\textsc{-sens}	\\
\glt `There is no way I will get 150,000 yuan!' (conversation; 15 refers to units of 10,000)
\z

\noindent
Finally, a clitic may attach directly to a negator, in which case it is detached from the predicate and also replicated in its canonical position as a verbal prefix (\ref{reinforcing6}).

\ea
\label{reinforcing6}
\gll  {tʰə} 	{ŋæ=ŋu} 	{tɕʰəsʰo} 	{mə=nɔ}		{mə-san.}\\
	\textsc{dem}    \textsc{1=pl.erg}   \textsc{dm}
	\textsc{mod.neg=top.cntr}  \textsc{mod.neg}-know.\textsc{npst.1pl}\\
\glt `We (the younger generations) have no idea at all about that.' (local history)
\z


\subsection{Negation, coordination and complex clauses}
Negation does not affect coordination in Geshiza (\ref{complex1}).

\ea
\label{complex1}
\gll  {dəu\textasciitilde dəu} {tʰu} {æ-li} {d-ə-v-dæ} {æ-li} {d-ə-mi-v-dæ.} \\
	\textsc{red.adjz}\textasciitilde small \textsc{dem.erg}
	one-\textsc{clf}.time \textsc{pfx}-\textsc{nactl-inv}-do.3
	one-\textsc{clf}.time 
	\textsc{pfx}-\textsc{nactl-neg-inv}-do.3
	\\
\glt `The small one (i.e. younger brother) carried out the tasks sometimes and sometimes he did not.' (folktale)
\z

Geshiza lacks any dedicated negative coordinators. The conjunction \ipa{mɛ-mɲa} \textasciitilde \ \ipa{mɛ-mja} `either...or' is etymologically related to the negated negative copula. Like all double negation in Geshiza, it follows the general principle of \textit{duplex negatio affirmat} \REF{complex2}.

\ea
\label{complex2}
\gll  <{xoko}>, 	{mɛmja} 	<{xoko}> 	{ŋgə}			{d-ə-ɕoŋ}. {mɛmja} 	<{ʈʂʰuænʈʂʰuæntɕan}> 	{ŋgə}	{ɕoŋ}.\\
hot.pot		either		hot.pot		eat.\textsc{inf} \textsc{pfx-nactl}-go.\textsc{npst}.1
either		meat.skewers					eat.\textsc{inf} go.\textsc{npst}.1\\
\glt `Let’s go to eat Chinese hotpot or Sichuanese style meat skewers.' (conversation)
\z

\subsection{Miscellaneous aspects of negation}
\label{otheraspects}
\subsubsection{Double negation}
\label{double}
Geshiza frequently applies double negation that follows the principle \textit{duplex negatio affirmat}, viz. double negations always carry positive semantics, rather than its logical opposite \textit{duplex negatio negat}. It commonly surfaces in a strong necessitative construction. Geshiza forms a permissive construction with the non-argument nominaliser \ipa{-rgui} ‘permission’ and the existential verb \ipa{də} (see \sectref{neg.stative}), as in (\ref{permissive}). This in turn is subject to negation, illustrated in (\ref{permissive.neg}). Finally, as shown in (\ref{permissive.neg.neg}), changing the polarity value of the nominalization results in a strong necessitative, literally `
there is no permission not to V'. Necessitative parallels to the Geshiza double negation construction exist in other Asian languages, e.g. the Chinese \textit{bùdé bù} V `to have to V' and the Japanese V\textit{-nakereba naranai} `to have to V'.

\ea
\label{permissive}
\gll  <{ʈʂʰetsə}> {mdʑə-rgui} {də-ræ=gæ}.\\ 
car		change-\textsc{nmlz}:permission \textsc{exvb-sens=mod}\\
\glt `Changing the car is allowed!' (conversation)
\z

\ea
\label{permissive.neg}
\gll  <{ʈʂʰetsə}> {mdʑə-rgui} {ma-ræ=gæ}.\\ 
car		change-\textsc{nmlz}:permission \textsc{negex-sens=mod}\\
\glt `Changing the car isn't allowed!' (accepted; see \ref{permissive})
\z

\ea
\label{permissive.neg.neg}
\gll  <{ʈʂʰetsə}> {mi-mdʑə-rgui} {ma-ræ=gæ}.\\
car		\textsc{neg}-change-\textsc{nmlz}:permission \textsc{negex-sens=mod}\\
\glt `We really need to change the car!' (accepted; see \ref{permissive})
\z

While double negation is commonly attested with such nominalizations, finite verbs also follow the same principle of double negation yielding affirmative semantics (\ref{finite.neg.neg}). This reflects a pattern attested in some Trans-Himalayan languages whereby a negated verb appears together with a negated modal verb with a resulting affirmative meaning. 

\ea
\label{finite.neg.neg}
\gll  {ŋa} {a̯su=læ} {mə-tɕʰan} {tɕʰu} {ŋa=læ} {xo} {ŋɛ} {stʰi} {rə-di-ɕoŋ} {mə-ŋi-ræ} 	[...]\\
\textsc{1sg}    rear.\textsc{inf}=\textsc{foc} \textsc{mod.neg}-can.\textsc{npst}.2 so \textsc{1sg=foc} \textsc{dem.loc} \textsc{1.sg.gen} place \textsc{dir-irr.neg}-go.\textsc{npst}.1 \textsc{mod.neg}-be.all.right-\textsc{sens} [...]\\		
\glt `You cannot take care of me, so I really need to go back to my place.' (lit. `You cannot take care of me, so if I do not go back to my place, things will not be fine.') (folktale)
\z

\subsubsection{Verbs with compulsory negation}
\label{compneg}
The source materials include three Geshiza verb stems that always occur with a negative prefix: \ipa{mə-grə} ‘to be unable’, \ipa{mə-ske} ‘should not’, \ipa{mə-kʰe} ‘to envy’ (meaning of the morpheme \ipa{kʰe} remains unclear). These can be interpreted as borderline cases of negative lexicalizations (see \sectref{neglex}). To illustrate, (\ref{should.not}) and (\ref{should.wrong}) show that \ipa{mə-ske} ‘should not’ cannot appear without the negative prefix.

\ea
\label{should.not}
\gll  {sponqæal} 		{sʰæ}		{mə-ske=mɔ}.\\ 
      frog kill.\textsc{inf}    \textsc{mod.neg}-should=\textsc{mod}\\ 
\glt `One should not kill frogs.' (conversation)
\z

\ea[*]{
\label{should.wrong}
\gll  {sponqæal} 		{sʰæ}		{ske=mɔ}.\\ 
      frog kill.\textsc{inf} should=\textsc{mod}\\ 
\glt Intended meaning: `One should kill frogs.' (rejected during elicitation; see \ref{should.not})}
\z

\subsubsection{Disagreement and self-repair}
    The pragmatic functions of expressing disagreement and repairing one's utterance during conversation are achieved by means of the negative copula \ipa{mɲa \textasciitilde \ mja}. In (\ref{disagreement1}), the husband in an elderly couple is narrating a folktale. His wife interrupts him, since she disagrees with her husband concerning the correct plot of the folktale.

\ea 
    \begin{xlist}
    \exi{A: }\label{disagreement1}
        \gll  {tsʰoŋpən=ke}		{oja}		{ɲi} 		{rji}			{dæ-ŋuə-sʰi}		{tɕʰu}    {xaræ} {kʰuæn=bɔ}.		{dæ-v-tɕʰi=ræ}		{dæ-ɕʰə-sʰə-mə-ræ}.\\
        merchant=\textsc{dat}		\textsc{inj}	\textsc{2sg.gen}	horse		\textsc{pfv-cop}.3-\textsc{infr}		so			\textsc{conj} 	give.\textsc{npst}.2=\textsc{mod}	\textsc{pfv-inv}-ride.\textsc{pst}.3=\textsc{lnk}		\textsc{pfv}-go.\textsc{pst}.3-\textsc{infr-ep-sens}\\
        \glt `He (\textit{A khu ston pa}) said to the merchant: ``Yes, it is your horse, so I will give it (back) to you," and he rode the horse, going away.'    
    \exi{B: }\label{disagreement2}
        \gll  {mɲa}. 		{rə-ʑe-sʰə-mə-ræ}.\\
        \textsc{neg.cop}.3	\textsc{pfv.dir}-come.3-\textsc{ep-ep-sens}\\
        \glt `No. He came back (to the merchant).' (folktale mixed with conversation)
    \end{xlist}
\z

In a closely related fashion, the negative copula \ipa{mɲa} \textasciitilde \  \ipa{mja} carries the pragmatic function of self repair when speakers change their minds, modify, or even negate their previous statement. In (\ref{disagreement3}), the speaker suddenly realizes his mistake in identifying a person as a lama, immediately followed by the negative copula and self-repair.

\ea
\label{disagreement3}
\gll  {tʰo} 	{næmkʰæ} 	{jə-me} 		{æ-lə}						{dæ-dʑi-sʰi}, {læma}.		{mɲa}.	{læma=tʰə}		
{mɲa-ræ=bɔ}.\\
\textsc{dem.loc}    \textsc{pn}		say-\textsc{nmlz}	one-\textsc{clf.indf} 	\textsc{pfv-exvb}.3-\textsc{sens} lama \textsc{neg.cop.}3	lama=\textsc{top}		\textsc{neg.cop.3}=\textsc{mod}\\
\glt `There was a certain person called \ipa{Næmkʰæ}, a lama, there. No. He is not a lama.' (personal history)
\z

\section{Summary}
This chapter introduced the principal morphology and constructions pertaining to negation in Geshiza. The findings concerning negative morphology and negative verbs are summarized in \tabref{summary} below. 

\begin{table}
\caption{Negative morphology in Geshiza}
\label{summary}
\fittable{
 \begin{tabular}{llll} 
  \lsptoprule
    Domain & Type & Form & Description \\ 
  \midrule
    verbal    &   \ipa{mi-}    &  prefix & aspectually neutral standard negator\\
    verbal   &   \ipa{mɛ-}         &   prefix & aspectual standard negator\\
    verbal   &   \ipa{mə-}       &  prefix  & unproductive aspectually neutral negator\\
   verbal  &    \ipa{di-} \textasciitilde \ \ipa{dʑi-}         &  prefix & irrealis negator\\
existential   &    \ipa{ma}         &   verb & negative inanimate existential verb\\
copular & \ipa{mɲa} \textasciitilde \  \ipa{mja} & verb & negative copula\\
nominal & \ipa{mi-}, \ipa{mə-} & prefix & verbal negators present in nominalizations\\
adjectival & \ipa{mi-}, \ipa{mə-} & prefix & verbal negators present in adjectivalizations\\
  \lspbottomrule
 \end{tabular}
 }
\end{table}

Negation in Geshiza  pertains strictly to the domain of verbs, which means that nouns and adjectives are only subject to negation when they derive from verbs and have already been negated in that word class prior to derivation. The language deviates from other described Gyalrongic languages in terms how aspect and negation interact. Geshiza includes an unproductive negator \ipa{mə-} that is used for modal verbs and several other verbs denoting non-controllable action. Additionally, the language includes a dedicated negative prefix \ipa{di- \textasciitilde \ dʑi-} for irrealis scenarios, such as the prohibitive.


\section*{Abbreviations}
The glossing of examples uses abbreviations from the Leipzig Glossing rules. New abbreviations for categories not included in the rules have been adopted. In all, the abbreviations used are listed below. To ease their identification, recent Chinese loanwords are placed inside angle brackets: <loanword>. Omissions are indicated with [...].\medskip\noindent


\begin{tabularx}{.45\textwidth}{lQ} %HC look at how to layout it so that there is no big space before the list. but it will also depend on the rest of the text
 \scshape 1 & first person\\
 \scshape 2 & second person \\
 \scshape 3 & third person\\
 \scshape adjz  & adjectivizer\\
 \scshape adv  & adverb\\
 \scshape asp  & aspectual\\
 \scshape aux  & auxiliary\\
 \scshape cntr  & contrastive\\
 \scshape caus  & causative\\
 \scshape clf  & classifier\\
 \scshape cond  & conditional\\
 \scshape conj  & conjunction\\
 \scshape cop  & copula\\
 \scshape dat  & dative\\
 \scshape dem  & demonstrative\\
 \scshape dir  & direction, orientation \\
 \scshape dm  & discourse marker\\
 \scshape ep  & epistemic suffix\\
 \scshape erg  & ergative\\
 \scshape exp  & experiential\\
 \scshape exvb  & existential verb\\
 \scshape foc  & focus\\
 \scshape gen  & genitive\\
 \scshape infr  & inferential \\
 \scshape imp  & imperative\\
 \scshape indf  & indefinite\\
 \scshape inf  & infinitive\\
 \scshape inj  & interjection\\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
 \scshape inv  & inverse\\
 \scshape ipfv  & imperfective\\
 \scshape irr  & irrealis\\
 \scshape lnk  & linker\\
 \scshape loc  & locative\\
 \scshape mod  & modal\\
 \scshape nactl  & nonactual, realis mood\\
 \scshape neg  & negation\\
 \scshape negex  & negative existential \\
  \scshape nmlz  & nominalizer\\
 \scshape npst  & non-past \\
 \scshape opt  & optative\\
 \scshape perf  & perfect\\
 \scshape pfv  & perfective\\
 \scshape pl  & plural\\
 \scshape pn  & personal name\\
 \scshape pfx  & multifunctional verbal prefix\\
 \scshape prosp  & prospective\\
 \scshape pst  & past \\
 \scshape q  & interrogative\\
 \scshape red  & reduplication\\
 \scshape rep  & reportative \\
 \scshape sens  & sensory \\
 \scshape sg  & singular\\
 \scshape top  & topic\\
 \scshape topn  & toponym\\
 \\
\end{tabularx}

\printbibliography[heading=subbibliography,notkeyword=this]

\end{document}
