\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233958}
\author{Omri Amiraz\affiliation{}}
\title{Negation in Modern Hebrew}
\abstract{This chapter provides an overview of negation in Modern Hebrew, an Afroasiatic language of the Semitic branch. Standard negation is typically expressed by a construction where the negative particle \textit{lo} precedes the finite verb. However, in formal registers, the negative copula \textit{en}, historically derived from a negative existential particle, is employed instead of \textit{lo} in the present tense. This negative copula can appear in two distinct standard negation constructions, with no difference in meaning: (i) in an inflected form (person, number, and gender) positioned before the finite verb; or (ii) in an uninflected form in clause-initial position. Beyond clausal negation, particular attention is paid to negative indefinites and polarity items.
\keywords{Hebrew, standard negation, negative copula, constituent negation, negative indefinites, polarity items}
}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 

\section{The language} \label{sec:heb:1}
\largerpage
Modern Hebrew (ISO: \href{https://iso639-3.sil.org/code/heb}{heb}, Glottocode: \href{https://glottolog.org/resource/languoid/id/hebr1245}{hebr1245}) belongs to the Semitic branch of the Afroasiatic family. It is spoken by over 5 million native speakers, and it is the official language of Israel \citep{EberhardEtAl2023}. Pre-modern Hebrew ceased to be a spoken language at the end of the second century CE but continued to be used in liturgical, scholarly, legal, and literary writings by Jewish communities world-wide. It is commonly classified into several periods based on existing corpora: (i) Biblical Hebrew (the Hebrew Bible); (ii) Mishnaic Hebrew (the Mishna and parts of the Talmud); (iii) Medieval Hebrew (religious, philosophical, poetic, and scientific texts written during the period of Islamic rule in the Middle East, North Africa, and the Iberian Peninsula); (iv) Late Rabbinic Hebrew and Hasidic Hebrew (religious writings of Jewish communities in Europe); and (v) Maskilic Hebrew (secular writings which began to appear in Central and Eastern Europe during the Enlightenment, starting from the middle of the eighteenth century). The language of the latter period, which was usually characterized by a linguistic mixture of different historical styles, had considerable influence on the development of Modern Hebrew. Another major influence was the Hebrew spoken by Jewish communities in Palestine in the late nineteenth century. Their native languages were usually Arabic, Judeo-Spanish, or Yiddish, but they used Hebrew as a lingua franca for inter-community communication. Jewish immigrants who came to Palestine from Europe at the end of the nineteenth century adopted Hebrew as their primary community language and became the first speakers of Modern Hebrew. Hebrew later became one of the official languages of Mandatory Palestine in 1922 and the state of Israel in 1948 \citep[2--4]{Doron2015}.

Due to an extended history of bilingualism, Hebrew had long-standing contact with Yiddish (Indo-European), Slavic languages (Indo-European), and Arabic (Afroasiatic) \citep[7]{Doron2015}. In recent decades, Hebrew’s main contact languages are English (Indo-European) and Palestinian Arabic. Modern Hebrew shows almost no regional variation \citep[313]{Berman1997}. However, it exhibits sociolinguistic variation in phonology, syntax, semantics, and the lexicon \citep{Henshke2013}. 

The basic word order is SVO, though VS is a possible alternative in certain contexts, especially in intransitive clauses. Hebrew is predominantly head-initial and dependent-marking. As in other Semitic languages, all verbs and many nouns and adjectives are formed from a triconsonantal root and, sometimes, additional affixes \citep[317]{Berman1997}. Adjectives and demonstratives agree in number and gender with the noun that they modify, and numerals agree in gender. Past and future tense verbs inflect for person, number, and gender. Present tense verbs inflect for number and gender but not person because this form is historically derived from the present participle and has retained its adjective-like morphology \citep[205]{Tzivoni1993}.

Many aspects of negation in Modern Hebrew have been previously described in the literature. \citet{Tzivoni1993} is a comprehensive description (written in Hebrew) of the system of negation, which is mostly based on written corpora. \citet{Shor2020} studies the expression of negation in spontaneous conversation. In addition, chapters about negation can be found in grammars such as \citet[293--307]{Glinert1989}, \citet[225--230]{Rosén1977}, and \citet[145--146, 220--224]{Berman1978}.

The linguistic examples in this paper represent both formal and informal registers.\footnote{The examples are transcribed using the following phonetic notation: “c” stands for a voiceless alveolar affricate /t͡s/, “š” stands for a voiceless post-alveolar fricative /ʃ/, “y” stands for a palatal approximant /j/, “x” stands for a voiceless uvular fricative /χ/, and “r” stands for a voiced uvular approximant /ʁ̝/. The transcription represents standard General Israeli Hebrew.} They are drawn from several sources, which are indicated in the glosses: (i) previous literature, primarily \citet{Tzivoni1993} and \citet{Glinert1989}; (ii) the Beʔofen Tivʔi Hebrew Textual Corpus, which is based on an internet forum dedicated to homeschooling and other alternative lifestyles; (iii) web examples (URLs are given in footnotes); and (iv) constructed examples. The structure of this chapter follows the structure of the questionnaire provided by the editors (\citetv{chapters/questionnaire}).

\section{Clausal negation} \label{sec:heb:2}
\subsection{Standard negation} \label{sec:heb:2.1}

Standard negation is expressed by a construction in which the negative particle \textit{lo} appears before the finite verb. This construction is symmetric with the corresponding affirmative \REF{ex:heb:1}, according to the criteria of \citet{Miestamo2005}. \textit{lo} is considered a particle based on \citegen{Dryer2013-wals112} criteria since it carries its own stress and does not inflect.

\ea\label{ex:heb:1} 
    \gll Hem (lo) makir-im ot-a. \\ 
          3\textsc{pl.m} (\textsc{neg}) know.\textsc{prs-pl.m} \textsc{acc-3sg.f} \\
    \glt  ‘They (don’t) know her.’ [constructed]
\z

In formal registers, present tense forms are not negated by \textit{lo} but by the negative copula \textit{en} \REF{ex:heb:2}. On the other hand, past and future tense verbs are negated by \textit{lo}, as in the spoken language. The negative copula \textit{en} typically inflects for person, number, and gender \citep[205]{Tzivoni1993}, as in \REF{ex:heb:2a}. However, \textit{en} is not a verb for two reasons. First, it does not inflect for tense. Second, it inflects for person, whereas present tense verbs do not inflect for person (see \sectref{sec:heb:1}). Syntactically, \textit{en} belongs to a small class of particles, which also includes the presentatives \textit{harey} and \textit{hine}, the latter having also grammaticalized into the copula \textit{hin}{}-, which is the affirmative counterpart of \textit{en} in formal registers (see \sectref{sec:heb:2.3}). These particles are unique in that they can either precede a clitic pronoun, as in \REF{ex:heb:2a}, or an independent pronoun, as in \REF{ex:heb:2b}, while other grammatical constructions may only occur with one type of pronoun or the other. The construction with inflecting \textit{en} \REF{ex:heb:2a} may be considered an instance of constructional asymmetric negation because \textit{en} is followed by a clitic pronoun, which is absent in the corresponding affirmative. The construction with clause-initial, non-inflecting \textit{en} \REF{ex:heb:2b}, by contrast, is symmetric.

\ea\label{ex:heb:2} 
    \ea\label{ex:heb:2a}
        \gll Hem (en-am) makir-im ot-a.\\
             3\textsc{pl.m} \textsc{(neg-3pl.m)} know.\textsc{prs-pl.m} \textsc{acc-3sg.f} \\
        \glt ‘They (don’t) know her.' [constructed]
    \ex\label{ex:heb:2b}
        \gll (En) hem makir-im ot-a.\\
             \textsc{(neg)} 3\textsc{pl.m} know.\textsc{prs-pl.m} \textsc{acc-3sg.f} \\
        \glt ‘They (don’t) know her.' [constructed]
    \z 
\z 

\subsection{Negation in non-declaratives} \label{sec:heb:2.2}

Prohibitives are expressed by a construction in which the negative particle \textit{al} precedes the future tense form \REF{ex:heb:3a}. The dedicated imperative form cannot occur in this construction. However, in most cases, prohibitives are symmetric with the corresponding affirmative because the imperative paradigm has been largely replaced by the future tense paradigm in the spoken language \citep[284--285]{Glinert1989}. Note that several verbs have preserved the imperative form, most of these verbs having a “weak” root, i.e., one of the consonants of the root is not pronounced in part of the verb conjugation \citep{TheHebrewAcademy2017}. In this case, then, negation is asymmetric (\ref{ex:heb:3b}--\ref{ex:heb:3c}). In indirect speech, the standard negation particle \textit{lo} is used instead of \textit{al} \REF{ex:heb:3d}.

\ea\label{ex:heb:3}
    \ea\label{ex:heb:3a}
        \gll (Al) ta-kšiv-${\emptyset}$ l-o!\\
             (\textsc{proh}) 2\textsc{.fut}{}-listen-\textsc{sg}.\textsc{m} \textsc{dat-3sg.m}\\
        \glt ‘(Don’t) listen to him!’ [constructed]
    \ex\label{ex:heb:3b}
        \gll Lex-${\emptyset}$! \\
             go.\textsc{imp-sg.m}\\
        \glt ‘Go!’ [constructed]
    \ex\label{ex:heb:3c}
        \gll Al te-lex-${\emptyset}$!\\
             \textsc{proh} \textsc{2.fut}{}-go\textsc{{}-sg.m}\\
        \glt ‘Don’t go!’ [constructed] 
    \ex\label{ex:heb:3d}
        \gll Ta{}-gid{}-i l-a še-lo te-lex{}-${\emptyset}$!\\
             \textsc{2.fut-}tell{}-\textsc{sg.f} \textsc{dat-3sg.f} \textsc{comp-neg} \textsc{3.f.fut}{}-go{}-\textsc{sg}\\
        \glt ‘Tell her not to go!’ [constructed]
    \z
\z 

In formal registers, \textit{al} may also occur in a jussive construction, in which it is either followed by a dative noun phrase or pronoun and a non-finite clause \REF{ex:heb:4a} or a future tense verb \REF{ex:heb:4b}.

\ea\label{ex:heb:4}
    \ea\label{ex:heb:4a}
        \gll Al l-anu lehitʔarev.\\
             \textsc{neg} \textsc{dat-1pl} get.involved.\textsc{inf}\\
        \glt ‘We shouldn't get involved.' [Beʔofen Tivʔi] 
    \ex\label{ex:heb:4b}
        \gll Al ni-štameš be-mil-im loʔazi-ot. \\
             \textsc{neg} \textsc{1pl.fut-}use in-word-\textsc{pl} foreign-\textsc{pl}\\
        \glt ‘Let us not use foreign words.'\footnote{\url{https://www.daat.ac.il/daat/ezrachut/begin/48_5-2.htm} (Accessed on 2022-08-22)}
    \z 
\z

The particle \textit{lo} can be used in another jussive construction with a finite clause introduced by the complementizer \textit{še} \citep{Bar-Adon1966}. This construction usually has a flavor of a warning \REF{ex:heb:5a}. It is symmetric with the affirmative, which is usually used as an optative in the second person \REF{ex:heb:5b}.

\ea\label{ex:heb:5} 
    \ea\label{ex:heb:5a}
        \gll Še-lo ti-škax-${\emptyset}$ lavo maxar!\\
             \textsc{comp{}-neg} \textsc{2.fut}{}-forget-\textsc{sg.m} come.\textsc{inf} tomorrow\\
        \glt ‘Don't forget to come tomorrow!' \citep[410]{Bar-Adon1966}
    \ex\label{ex:heb:5b}
        \gll Še-ta-mut-${\emptyset}$!\\
             \textsc{comp{}-2.fut-}die\textsc{{}-sg.m}\\
        \glt ‘Drop dead!' \citep[411]{Bar-Adon1966}
    \z
\z 

Negative obligation is expressed by a construction in which the non-inflecting modal \textit{asur} ‘forbidden’ precedes a non-finite clause \REF{ex:heb:6a}. This modal is also used to express a general prohibition (e.g., \textit{It is forbidden to smoke here}), and in this context the standard negation particle \textit{lo} and the formal negative particle \textit{en} may be used as well \REF{ex:heb:6b}. Despite the fact that these three items appear to be interchangeable in \REF{ex:heb:6b}, they give rise to different types of speech acts: when \textit{asur} and \textit{en} are used, \REF{ex:heb:6b} is an assertion, but when \textit{lo} is used, \REF{ex:heb:6b} is a directive. Evidence for this comes from the fact that only \textit{lo} is compatible with \textit{na} and \textit{bevakaša} ‘please' \REF{ex:heb:6c}. Note that the construction in which \textit{en} is followed by a non-finite clause may express not only prohibition \REF{ex:heb:6b} but also impossibility \REF{ex:heb:6d}.

\ea\label{ex:heb:6}
    \ea\label{ex:heb:6a}
        \gll Asur l-i ledaber it-xa.\\
             forbidden \textsc{dat-1sg} talk.\textsc{inf} with-\textsc{2sg.m}\\
        \glt ‘I’m not allowed to talk to you.’ [constructed]
    \ex\label{ex:heb:6b}
        \gll {\{Asur / Lo / En\}} lehaʔaxil et ha-xay-ot. \\
             forbidden/\textsc{neg/neg} feed.\textsc{inf} \textsc{acc} \textsc{def}{}-animal-\textsc{pl}\\
        \glt ‘Do not feed the animals!’ [constructed]
    \ex\label{ex:heb:6c}
        \gll Na lo lehaʔaxil et ha-xay-ot!\\
             please \textsc{neg} feed.\textsc{inf} \textsc{acc} \textsc{def}{}-animal-\textsc{pl}\\
        \glt ‘Please do not feed the animals!' [constructed]
    \ex\label{ex:heb:6d}
        \gll En ladaʔat matay.\\
             \textsc{neg} know.\textsc{inf} when\\
        \glt ‘There is no knowing when.' \citep[376]{Tzivoni1993}
    \z
\z 

\subsection{Negation in stative predications} \label{sec:heb:2.3}

In existential and possessive predication, \textit{lo} is in complementary distribution with the negative particle \textit{en}. In the present tense, \textit{en} is used (\ref{ex:heb:7a}--\ref{ex:heb:7b}), and in other tenses \textit{lo} is used, followed by the copular verb \textit{haya} ‘be’ (\ref{ex:heb:7c}--\ref{ex:heb:7d}) \citep[173]{Tzivoni1993}. The construction with \textit{lo} is symmetric with the corresponding affirmative, while \textit{en} contrasts with the affirmative existential particle \textit{yeš}. The use of a special negative existential which replaces the affirmative one reflects a wide-spread cross-linguistic tendency \citep{Veselinova2013neg.exist}.

\ea\label{ex:heb:7}
    \ea\label{ex:heb:7a}
        \gll {\{Yeš / En\}} oxel.\\ 
             \textsc{ex/neg.ex} food\\
        \glt ‘There is (no) food.’ [constructed]
    \ex\label{ex:heb:7b}
        \gll {\{Yeš / En\}} l-i oxel.\\
             \textsc{ex/neg.ex} \textsc{dat-1sg} food\\
        \glt ‘I have (no) food.’ [constructed]
    \ex\label{ex:heb:7c}
        \gll (Lo) hay-a oxel.\\
             \textsc{(neg)} be.\textsc{pst{}-3sg.m} food\\
        \glt ‘There was (no) food.’ [constructed]
    \ex\label{ex:heb:7d}
        \gll (Lo) hay-a l-i oxel.\\
             \textsc{(neg)} be.\textsc{pst{}-3sg.m} \textsc{dat-1sg} food\\
        \glt ‘I had (no) food.’ [constructed]
    \z 
\z 

In the case of other non-verbal predicates such as equation \REF{ex:heb:8a}, proper inclusion \REF{ex:heb:8b}, attribution \REF{ex:heb:8c}, and locative predication \REF{ex:heb:8d}, \textit{lo} is used in all tenses. However, in formal registers, the negative copula \textit{en} is used instead of \textit{lo} in the present tense \citep[201]{Tzivoni1993}. There is a related use of \textit{en} in locative predication where the location is not specified \REF{ex:heb:8e}. In such cases, the argument can be understood to be absent either temporarily or permanently \citep[181]{Tzivoni1993}. Notice that the negative particle precedes the copular verb \textit{haya} ‘be’ \REF{ex:heb:8f} but follows non-verbal copulas such as those in (8a-b), which are derived from third-person independent pronouns. Finally, in formal registers, affirmative copular clauses may be formed with the inflecting copula \textit{hin-} \REF{ex:heb:8g}, which is the affirmative counterpart of \textit{en} (see \sectref{sec:heb:2.1}). This copula cannot be negated.

\ea\label{ex:heb:8}
    \ea\label{ex:heb:8a}
        \gll Gila (hi) {\{lo / en-a\}} ima šel-i.\\
            Gila \textsc{cop.sg.f} \textsc{neg/neg-3sg.f} mother \textsc{poss-1sg}\\
        \glt ‘Gila is not my mother.’ [constructed]
    \ex\label{ex:heb:8b}
        \gll Dani (hu) {\{lo / en-o\}} rofe.\\
             Dani \textsc{cop.sg.m} \textsc{neg/neg-3sg.m} doctor\\
        \glt ‘Dani is not a doctor.’ [constructed]
    \ex\label{ex:heb:8c}
        \gll Ha-bdixa {\{lo / en-a\}} macxik-a.\\
             \textsc{def{}-}joke \textsc{neg/neg-3sg.f} funny-\textsc{sg.f}\\
        \glt ‘The joke isn’t funny.’ [constructed]
    \ex\label{ex:heb:8d}
        \gll Ha-patiš {\{lo / en-o\}} b-a-argaz.\\
             \textsc{def-}hammer \textsc{neg/neg-3sg.m} in\textsc{{}-def}{}-box\\
        \glt ‘The hammer is not in the box.’ [constructed]
    \ex\label{ex:heb:8e}
        \gll Ha-yeled en-enu.\\
             \textsc{def-}child \textsc{neg-3sg.m}\\
        \glt ‘The child is not present.’ or: ‘The child is gone [dead].’ \citep[181]{Tzivoni1993}
    \ex\label{ex:heb:8f}
        \gll Ha-patiš lo hay-a b-a-argaz.\\
             \textsc{def-}hammer \textsc{neg} be.\textsc{pst{}-3sg.m} in\textsc{{}-def}{}-box\\
        \glt ‘The hammer was not in the box.’ [constructed]
    \ex\label{ex:heb:8g}
        \gll Ha-neʔešam-im hin-am hor{}-e{}-ha šel ha-yalda.\\
             \textsc{def-}defendant-\textsc{pl} \textsc{cop{}-3pl.m} parent-\textsc{pl-2sg.f} \textsc{poss} \textsc{def-}girl\\
        \glt ‘The defendants are the girl’s parents.’ [Beʔofen Tivʔi]
    \z 
\z 

Usually, \textit{en} follows the subject and inflects for person, number and gender \REF{ex:heb:9a}, but it can also precede the subject, in which case it occurs in an uninflected form \REF{ex:heb:9b} \citep[211]{Tzivoni1993}.

\ea\label{ex:heb:9}
    \ea\label{ex:heb:9a}
        \gll Ha-hanhala en-a axraʔi-t le-ovdan xafac-im.\\
             \textsc{def}{}-management \textsc{neg-3sg.f} responsible-\textsc{sg}.\textsc{f} for-loss.of item-\textsc{pl}\\
        \glt ‘The management is not responsible for lost items.’\footnote{\url{https://www.kfar-saba.muni.il/?CategoryID=776} (Accessed on 2019-09-26)}
    \ex\label{ex:heb:9b}
        \gll En ha-hanhala axraʔi-t le-ovdan xafac-im.\\
             \textsc{neg} \textsc{def}{}-management responsible-\textsc{sg}.\textsc{f}  for-loss.of item-\textsc{pl}\\
        \glt ‘The management is not responsible for lost items.’\footnote{\url{https://www.rehovot.muni.il/238/} (Accessed on 2019-09-26)}
    \z 
\z

The negative copula \textit{en} cannot be used with predicates that do not have a subject argument \REF{ex:heb:10a}. When the subject position is filled by a non-deictic demonstrative \textit{ze} ‘this', \textit{en} has to precede the subject (\ref{ex:heb:10b}--\ref{ex:heb:10c}) \citep[297]{Glinert1989}.


\ea\label{ex:heb:10}
    \ea\label{ex:heb:10a}
        \gll  {\{Lo / *En\}} kar l-i.\\ 
              \textsc{neg/neg} cold \textsc{dat-1sg}\\
        \glt ‘I’m not cold.’ [constructed]
    \ex\label{ex:heb:10b}
        \gll Ze {\{lo / *en-o\}} yaf-e. \\
              this \textsc{neg/neg-3sg.m} nice-\textsc{sg.m}\\
        \glt  ‘It’s not nice.’ [constructed]
    \ex\label{ex:heb:10c}
        \gll En ze yaf-e.\\
            \textsc{neg} this nice-\textsc{sg.m}\\
        \glt ‘It’s not nice.’ \citep[216]{Tzivoni1993}
    \z
\z 


Some copular constructions exhibit symmetric negation, while others exhibit asymmetric negation. In past and future tense, the negative particle \textit{lo} precedes the finite copular verb \textit{haya} ‘be’ \REF{ex:heb:8f}, which is also used in the affirmative. Hence, these constructions are symmetric. In the present tense, a copula is usually used in the affirmative in the case of equation \REF{ex:heb:8a} and proper inclusion \REF{ex:heb:8b} unless the subject is an independent pronoun. The constructions with \textit{lo} are symmetric with the affirmative because the copula is used in the negative as well, but those with \textit{en} are asymmetric because the copula is often omitted in the negative. Stative predicates expressing attribution \REF{ex:heb:8c} and location \REF{ex:heb:8d} usually do not occur with a copula in the affirmative, and hence the negative only differs from the affirmative by the presence of the negative particle or inflecting copula.

\subsection{Negation in non-main clauses} \label{sec:heb:2.4}

Negation in finite \REF{ex:heb:11a} and non-finite \REF{ex:heb:11b} dependent clauses is normally expressed by the standard negation particle \textit{lo}. The formal negative copula \textit{en} may only be used in the former case \REF{ex:heb:11a}. Privative prepositions such as \textit{bli} ‘without’ may take a non-finite clause as an argument too (see \sectref{sec:heb:3.3.1}).

\ea\label{ex:heb:11}
    \ea\label{ex:heb:11a}
        \gll Ze ma še-ani {\{lo / en-i\}} yodaʔ{}-at.\\
        this what \textsc{comp}{}-1\textsc{sg} \textsc{neg/neg-1sg} know.\textsc{prs-sg.f}\\
        \glt ‘This is what I don’t know.’ [Beʔofen Tivʔi]
    \ex\label{ex:heb:11b}
        \gll Ani maʔadif-a lo ladaʔat.\\
             \textsc{1sg} prefer.\textsc{prs-3sg.f} \textsc{neg} know.\textsc{inf}\\
        \glt ‘I prefer not to know.' [Beʔofen Tivʔi]
    \z 
\z 

However, Hebrew has several negative counterfactual conditional conjunctions, which express negation by themselves: \textit{ilule, lule,} and \textit{ilmale} \REF{ex:heb:12}. The last two can also take a nominal complement, e.g., \textit{ilmale ha-gešem} ‘if it weren’t for the rain’. Other types of negative conditionals (e.g., hypothetical) are expressed by the ordinary conditional conjunction \textit{im} ‘if’ and the negative particle \textit{lo} \citep[378--381]{Tzivoni1993}.

\ea\label{ex:heb:12}
    \gll Ilmale hay-ta l-a-šnaim haškafat olam dom-a, [...]\\ 
        if.not be.\textsc{pst-3sg.f} \textsc{dat}{}-\textsc{def}{}-two view.of world similar-\textsc{sg.f} \\
    \glt ‘If the two of them didn’t have a similar world view, [Barak wouldn’t be able to implement his novel approaches].' \citep[381]{Tzivoni1993}
\z 

\subsection{Negative lexicalizations} \label{sec:heb:2.5}

The phasal polarity adverb \textit{terem} ‘not yet' (also ‘before'), which is restricted to formal registers, is used in a clausal negation construction where it precedes the finite verb \REF{ex:heb:13a}. This adverb is interchangeable with \textit{adain lo} ‘not yet' (lit. ‘still not'), except for the fact that \textit{terem} cannot be used with negative indefinites of the \textit{af/šum}{}-series \REF{ex:heb:13b} (see \sectref{sec:heb:3.2.1}). The adverb \textit{terem} can also be used as a negative reply to a polar question \REF{ex:heb:13c} (see \sectref{sec:heb:3.1} for a discussion of negative replies).

\ea\label{ex:heb:13}
    \ea[]{\label{ex:heb:13a}
        \gll Terem nifgaš{}-nu.\\
             not.yet meet.\textsc{pst-1pl}\\
        \glt ‘We haven't met yet.' [constructed]}
    \ex[*]{\label{ex:heb:13b}
        \gll Af exad terem higia-${\emptyset}$.\\
            even one not.yet arrive.\textsc{pst}{}-3\textsc{sg.m}\\
        \glt (‘Nobody has arrived yet.') [constructed]}
    \z
\z

\ea\label{ex:heb:13c}
    \begin{xlist}[A: ]
        \exi{Q: }
            \gll  Gamar-ta kvar et ha-avoda?\\ 
                 finish.\textsc{pst-2sg.m} alreday \textsc{acc} \textsc{def}{}-job\\ 
            \glt ‘Have you finished the job yet?'
        \exi{A: }
             \gll  Terem.\\ 
                    not.yet\\
             \glt ‘Not yet.' \citep[335]{Tzivoni1993} 
    \end{xlist}
\z 

\section{Non-clausal negation} \label{sec:heb:3}
\subsection{Negative replies} \label{sec:heb:3.1}

Negative replies to polar questions are expressed by the negative particle \textit{lo} \REF{ex:heb:14}, while affirmative replies are expressed by the affirmative particle \textit{ken} ‘yes’. In the spoken language, a negative reply can also be expressed by a dental click, which can occur as a freestanding element or followed by \textit{lo} \citep[614--615]{Shor2020}. This interjection is probably borrowed from Arabic.

\ea\label{ex:heb:14}
    \begin{xlist}[A: ]
    \exi{A: }\label{ex:heb:14a}
        \gll Dibar-ta it-o? \\
            talk.\textsc{pst-2sg.m} with-\textsc{3sg.m} \\
        \glt ‘Did you talk to him?' 
    \exi{B: }\label{ex:heb:14b}
        \gll Lo.\\
            \textsc{neg} \\
        \glt ‘No.’ [constructed]
    \end{xlist}
\z 

\citet{Burstein1999} collected and examined thousands of question-and-answer pairs from interviews that appeared in the Israeli press. She observes that negative replies to negative questions agree with the negative proposition in the question \REF{ex:heb:15}.

\ea\label{ex:heb:15}
    \begin{xlist}[A: ]
    \exi{A: }\label{ex:heb:15A}
        \gll  Moše lo omer-${\emptyset}$ l-ax še-at rak-a miday?\\ 
              Moše \textsc{neg} say.\textsc{prs-sg.m} \textsc{dat-2sg.f} \textsc{comp{}-}you soft-\textsc{sg.f} too\\ 
        \glt ‘Does Moshe not say to you that you are too soft?'
    \exi{B: }\label{ex:heb:15B}
        \gll  Lo.\\ 
               \textsc{neg}\\
        \glt ‘No [he doesn’t].’
    \end{xlist}
\z 

\citet[146]{Burstein1999} reports that she didn't find any example in the corpus of a single word \textit{ken} ‘yes' reply to a negative question. On the other hand, standalone negative replies to negative questions are to be found, e.g., \REF{ex:heb:15}. Positive replies to negative questions are often used when the question has a positive bias \citep[387]{Burstein1999}, as in \REF{ex:heb:16}.

\ea\label{ex:heb:16}
    \begin{xlist}[A: ]
    \exi{Q: }\label{ex:heb:16Q}
        \gll  Lo mafxid-${\emptyset}$? ibad-eta et ha-regel.\\ 
              \textsc{neg} scary-\textsc{sg.m} lose.\textsc{pst-2sg.m} \textsc{acc} \textsc{def}{}-leg\\ 
        \glt ‘Isn't it scary? You lost your leg.'
    \exi{A: }\label{ex:heb:16A}
        \gll  Ken. Aval ata roʔ-e še-ata  xay-${\emptyset}$.\\ 
               yes but you see.\textsc{prs-sg.m} \textsc{comp}{}-you alive-\textsc{sg.m}\\
        \glt ‘Yes [it is]. But you see that you're alive.' \citep[388]{Burstein1999}
    \end{xlist}
\z 

When a question has a negative bias, a positive reply is interpreted as confirming this bias \REF{ex:heb:17} unless the affirmative particle is preceded by an adverb like \textit{davka} ‘actually' \REF{ex:heb:18}, which is similar in meaning to German \textit{doch}.

\ea\label{ex:heb:17}
    \begin{xlist}[B2: ]
    \exi{A: }\label{ex:heb:17A}
        \gll  Lo koʔev-${\emptyset}$, naxon?\\ 
              \textsc{neg} hurt.\textsc{prs-sg.m} right\\ 
        \glt ‘It doesn't hurt, right?'
    \exi{B1: }\label{ex:heb:17B}
        \gll  Ken, bixlal lo. \\ 
               yes at.all \textsc{neg} \\
        \glt ‘Yes, not at all  [constructed]
    \exi{B2: }\label{ex:heb:17C}
        \gll  \#Ken, koev-${\emptyset}$.\\ 
             yes hurt.\textsc{prs-sg.m}\\
        \glt \#Yes, it hurts.' [constructed]
    \end{xlist}
\z 

\ea\label{ex:heb:18}
    \begin{xlist}[A: ]
    \exi{A: }\label{ex:heb:18A}
        \gll  Lo koʔev-${\emptyset}$, naxon?\\ 
              \textsc{neg} hurt.\textsc{prs-sg.m} right\\ 
        \glt ‘It doesn't hurt, right?'
    \exi{B: }\label{ex:heb:18B}
        \gll  Davka ken. Kcat.\\ 
               actually yes a.little\\
        \glt ‘Actually, it does. A little.'\footnote{\url{http://stage.co.il/Stories/445839} (Accessed on 2019-09-19)}
    \end{xlist}
\z 

\subsection{Negative indefinites and quantifiers} \label{sec:heb:3.2}
\subsubsection{Overview}
The most common series of negative indefinites is the \textit{af/šum-}series (see \sectref{sec:heb:3.2.1}). In formal registers, the \textit{iš}{}-series is an alternative, and it has a slightly broader distribution than the \textit{af/šum-}series (see \sectref{sec:heb:3.2.2}). Finally, the {}-\textit{šehu}{}-series, which is primarily used in non-negative contexts, can also co-occur with negation when the indefinite pronoun is modified by a relative clause or other modifiers (see \sectref{sec:heb:3.2.3}). 
Negative indefinite pronouns obligatorily co-occur with clausal negation, regardless of their position with respect to the verb (\ref{ex:heb:19a}--\ref{ex:heb:19b}). However, in rare cases, the negative marker is omitted for emphatic purposes \REF{ex:heb:19c}, possibly under the influence of English.
For negated quantifier phrases, see \REF{sec:heb:4.6.4}.

\ea\label{ex:heb:19}
    \ea\label{ex:heb:19a}
        \gll {\{Klum / Davar\}} lo kar-a.\\
            nothing/thing \textsc{neg} happen.\textsc{pst-3sg.m}\\
        \glt ‘Nothing happened.’ [constructed]
    \ex\label{ex:heb:19b}
        \gll Lo kar-a {\{klum / davar\}}.\\
            \textsc{neg} happen.\textsc{pst-3sg.m} nothing/thing\\
        \glt ‘Nothing happened.’ [constructed]
    \ex\label{ex:heb:19c}
        \gll Hirgaš-ti klum ve-šum davar.\\
            feel.\textsc{pst-1sg} nothing and{}-no thing\\
        \glt ‘[In my last pregnancy I felt that I just loved my embryo, and this time when I first saw the embryo in an ultrasound scan,] I felt absolutely nothing.’ [Beʔofen Tivʔi]
    \z 
\z 

\subsubsection{The \textit{af/šum}{}-series} \label{sec:heb:3.2.1}

The \textit{af/šum-}series is restricted to negative sentences. The forms \textit{af} ‘no' (lit. ‘even') and \textit{šum} ‘no' function as determiners, e.g., \textit{af yeled} ‘no child'. The main difference between \textit{af} and \textit{šum} is that \textit{af} can only combine with singular count nouns, while \textit{šum} is also compatible with mass nouns and plurals, e.g., \textit{šum tlunot} ‘no complaints’ (\citealt{Tzivoni1993}: 112, 118). Both \textit{af} and \textit{šum} can combine with generic nouns to form indefinite pronouns such as \textit{af exad} ‘nobody’ (lit. ‘even one'), \textit{af paʔam} ‘never’ (lit. ‘even time'), \textit{šum makom} ‘nowhere’ (lit. ‘no place'), \textit{šum davar} ‘nothing’ (lit. ‘no thing'). This series also includes \textit{klum} ‘nothing’ and the formal \textit{meʔum(a)} ‘nothing’, which only have pronominal forms. This series is unrelated to the positive {}-\textit{šehu}{}-series, which is derived from interrogative pronouns followed by parametric concessive conditional clause \citep[298]{Haspelmath1997}.

The \textit{af/šum}{}-series can co-occur with clausal negation \REF{ex:heb:19}, negative modals \REF{ex:heb:20a} and the privative preposition \textit{bli} ‘without’ \REF{ex:heb:20b}.

\ea\label{ex:heb:20}
    \ea\label{ex:heb:20a}
        \gll Asur lesaper et ze le-af exad.\\
             forbidden tell.\textsc{inf} \textsc{acc} this \textsc{dat}{}-even one\\
        \glt ‘We mustn’t tell anyone about it.’ \citep[336]{Tzivoni1993} 
    \ex\label{ex:heb:20b}
        \gll Bli šum haxana mukdem-et.\\
            without no preparation prior-\textsc{sg.f}\\
        \glt ‘Without any prior preparation.’ [Beʔofen Tivʔi]
    \z 
\z 

The \textit{af/šum}{}-series cannot co-occur with implicitly negative predicates such as \textit{sirev} ‘refuse’ \REF{ex:heb:21}, \textit{besafek} ‘doubt’ and \textit{hikxiš} ‘deny’ \citep[306]{Glinert1989}. Note that these predicates fall under the category of \textsc{indirect} \textsc{negation} in \citet{Haspelmath1997}, just like \textit{bli} ‘without’. The fact that the \textit{af/šum}{}-series can co-occur with the latter but not the former suggests that the category of \textsc{indirect} \textsc{negation} should be split into at least two subcategories. 

\ea[*]{\label{ex:heb:21}
    \gll Hu sirev-${\emptyset}$ laʔasot klum.\\
        3\textsc{sg.m} refuse\textsc{.pst-3sg.m} do\textsc{.inf} nothing\\
    \glt (‘He refused to do anything.’) [constructed] }
\z 
%\hc{other chapters do not have parentheses around ungrammatical sentences' translation. Some have nothing, others have e.g. "Attempted:" (see murrinhpatha) or "Intented reading:" (see Zapotec). In my opinion that difference isn't important, if you(M/L) agree you can delete this message, if that should also be harmonized throughout all chapters, let me know which version to use} \lv{I think "Intended reading:" is best. However, I suspect, it's a lot of work to go through all chapters and harmonize this. Perhaps put this on a the list of items to be discussed when we meet in August?} \mm{The option with parentheses is what the langsci style sheet has.}
The \textit{af/šum}{}-series can co-occur with superordinate negation when the matrix predicate is neg-raising \REF{ex:heb:22a}, and more rarely when it is non-neg-raising but also non-factive \REF{ex:heb:22b} (see \sectref{sec:heb:4.6.1} for a discussion of neg-raising). However, the {}-\textit{šehu-}series (see \sectref{sec:heb:3.2.3}) is preferred in this context. If the matrix predicate is factive \REF{ex:heb:22c} or a saying predicate \REF{ex:heb:22d}, only the {}-\textit{šehu-}series can be used.

\ea\label{ex:heb:22}
    \ea\label{ex:heb:22a}
        \gll Ani lo xošev-et še-af exad beʔemet hevin-${\emptyset}$ ma kor-e it-i. \\
            \textsc{1sg} \textsc{neg} think.\textsc{prs-sg.f} \textsc{comp}{}-even one really understand.\textsc{pst-3sg.m} what happen.\textsc{prs-sg.m} with-1\textsc{sg}\\
        \glt ‘I don’t think anyone really understood what was going on with me.’ [Beʔofen Tivʔi]
    \ex\label{ex:heb:22b}
        \gll Ani lo batuax{}-${\emptyset}$ še-af exad b-a-xeder kan yodea-${\emptyset}$ mi ze ha-koʔalicya ve-mi ze ha-opozicya. \\
            \textsc{1sg} \textsc{neg} sure-\textsc{sg.m} \textsc{comp}{}-even one  in-\textsc{def}{}-room here know.\textsc{prs-sg.m} who  \textsc{cop} \textsc{def}{}-coalition and-who \textsc{cop} \textsc{def}{}-opposition\\
        \glt ‘I’m not sure anyone in this room knows who the coalition is and who the opposition is.’\footnote{\url{https://www.nevo.co.il/law_html/law103/ksafim2005-02-25.htm} (Accessed on 2019-09-26)}
    \ex\label{ex:heb:22c}
        \gll Lo yada-ti še-mi-šehu moxer-${\emptyset}$ b-a-arec.\\
            \textsc{neg} know.\textsc{pst-1sg} \textsc{comp}{}-who-\textsc{indf} sell.\textsc{prs-sg.m} in-\textsc{def}{}-country \\
        \glt ‘I didn’t know someone sells it in Israel.’ [Beʔofen Tivʔi]
    \ex\label{ex:heb:22d}
        \gll Makabi tel{}-aviv af paʔam lo taʔan-a še-mi-šehu hitʔa-${\emptyset}$ ot-a. \\
            Makabi Tel-Aviv even time \textsc{neg} claim.\textsc{pst-3sg.f} \textsc{comp}{}-who-\textsc{indf} mislead.\textsc{pst{}-3sg.m} \textsc{acc-3sg.f}\\
        \glt ‘Maccabi Tel Aviv (basketball club) has never claimed that someone misled it.’\footnote{\url{http://10tv.nana10.co.il/Article/?ArticleID=577708} (Accessed on 2019-09-26)} 
    \z 
\z 

The \textit{af/šum}{}-series can be used as a negative reply without the negative particle \textit{lo} \REF{ex:heb:23}. Prescriptivists maintain that \textit{lo} should also be used in this context (see \citealt{Tzivoni1993}: 143 for references). For instance, the answer in \REF{ex:heb:23} would be \textit{lo xlum} (where /k/ is pronounced as [x] due to post-vocalic spirantization).

\ea\label{ex:heb:23}
    \begin{xlist}[A: ]
    \exi{A: }\label{ex:heb:23A}
        \gll  Ma amar-ta?\\ 
               what say.\textsc{pst-2sg.m}\\ 
        \glt ‘What did you say?'
    \exi{B: }\label{ex:heb:23B}
        \gll  Klum.\\ 
                nothing\\
        \glt ‘Nothing.’ [constructed]
    \end{xlist}
\z 

In very formal registers, \textit{lo xlum} ‘nothing’ can also be used in full declarative clauses, but clausal negation is still obligatory \REF{ex:heb:24}.

\ea\label{ex:heb:24}
    \gll En-a os-a ve-lo xlum kedey limnoa…\\ 
        \textsc{neg}{}-3\textsc{sg.f} do.\textsc{prs-3sg.f} and-\textsc{neg} nothing  in.order prevent.\textsc{inf}\\
    \glt ‘[The government] doesn’t do anything to prevent…’ \citep[144]{Tzivoni1993} 
\z 

\subsubsection{The \textit{iš}{}-series} \label{sec:heb:3.2.2}


The formal \textit{iš}{}-series is almost entirely restricted to negative sentences, such as \REF{ex:heb:25}, but it also rarely occurs in questions and conditionals. This series consists of \textit{iš} ‘anyone’ (lit. ‘person') and \textit{davar} ‘anything' (lit. ‘thing'). The determiner \textit{kol} ‘any’ can also be considered part of this series, but it can also function as a free-choice item \citep[298]{Haspelmath1997}.\footnote{Note that \textit{meʔolam} in \REF{ex:heb:25} is not part of the \textit{iš}{}-series since it can occur in superlatives (see \sectref{sec:heb:4.2.1}).}

\ea\label{ex:heb:25} 
    \gll David Levi meʔolam lo hivtiax-${\emptyset}$ le-iš davar. \\
         David Levi never \textsc{neg} promise.\textsc{pst-3sg.m} \textsc{dat}{}-person thing \\
    \glt ‘David Levi has never promised anything to anyone.’ \citep[111]{Tzivoni1993}
\z 

The \textit{iš}{}-series has a slightly broader distribution than the \textit{af/šum}{}-series. For instance, the \textit{iš}{}-series can co-occur with implicitly negative predicates \REF{ex:heb:26a} and with the negative phasal polarity adverb \textit{terem} ‘not yet’ \REF{ex:heb:26b} (compare with \REF{ex:heb:21} and \REF{ex:heb:13b}, respectively).

\ea\label{ex:heb:26}
    \ea\label{ex:heb:26a}
        \gll Ax ha-yehudi sirev{}-${\emptyset}$ lekabel davar. \\
             but \textsc{def}{}-Jew refuse.\textsc{pst-3sg.m}  accept.\textsc{inf} thing \\
        \glt ‘But the Jew refused to accept anything [as a reward].’\footnote{\url{http://www.chabad.org.il/Magazines/Article.asp?ArticleID=13712 & CategoryID=2215}\\ (Accessed on 2019-09-29)}
    \ex\label{ex:heb:26b}
        \gll Ax terem neʔes-a davar.\\
            but not.yet do\textsc{.pass.pst-3sg.m} thing\\
        \glt ‘But nothing has been done yet.’\footnote{\url{https://www.israelhayom.co.il/article/678211} (Accessed on 2019-09-26)}
    \z
\z 

The \textit{iš}{}-series occurs very rarely in non-negative contexts such as questions and conditionals. It seems that questions with \textit{iš}{}-series indefinites have a negative bias, as in \REF{ex:heb:27a}, but there aren’t enough examples to confirm this conjecture. It also seems that conditionals with \textit{iš}{}-series indefinites have a flavor of a warning, as in \REF{ex:heb:27b}.

\ea\label{ex:heb:27}
    \ea\label{ex:heb:27a} 
        \gll Axrey kol ha-mil-im […] haʔim notar-${\emptyset}$, davar haʔim notr-a l-a ahava?\\
         after all \textsc{def}{}-word-\textsc{pl} {} \textsc{q} left-\textsc{sg.m} thing \textsc{q} left-\textsc{sg.f} \textsc{dat}{}-3\textsc{sg.f} love?\\
        \glt ‘After all these words […] is there anything left, is there any love left?’\footnote{\url{https://simania.co.il/showReview.php?reviewId=49727} (Accessed on 2019-09-29)}
    \ex\label{ex:heb:27b}
        \gll Ki im te-saper-${\emptyset}$ le-iš al ha-mikre, te-kabel-${\emptyset}$ hetkef lev. \\
            \textsc{comp} if \textsc{2.f.fut}{}-tell-\textsc{sg} \textsc{dat}{}-person about \textsc{def}{}-event \textsc{2.f.fut}{}-get{}-\textsc{sg} attack heart \\
        \glt ‘[The defendant told the plaintiff] that if she tells anyone about the incident, she will get a heart attack.'\footnote{\url{https://www.news1.co.il/Archive/001-D-270469-00.html} (Accessed on 2019-09-29)}
    \z 
\z 

Indefinite pronouns of the \textit{iš}{}-series cannot be used as freestanding negative replies \citep[135, 139]{Tzivoni1993}, as demonstrated by \REF{ex:heb:28}. It is unclear if the reason for this is that speakers do not consider the \textit{iš}{}-series as inherently negative or that this series is rarely used in spoken language, which is where one expects to find negative replies.

\ea\label{ex:heb:28} 
    \begin{xlist}[A: ]
    \exi{A: }[]{\label{ex:heb:28A}
        \gll  Ma asi-ta?\\ 
              what do.\textsc{pst-2sg.m}\\ 
        \glt ‘What did you do?'}
    \exi{B: }[*]{\label{ex:heb:28B}
        \gll  Davar!\\ 
              thing\\
        \glt ‘Nothing!’}
    \end{xlist}
\z 

\subsubsection{The {}-\textit{šehu}{}-series} \label{sec:heb:3.2.3}

According to \citet[135--136]{Haspelmath1997}, the -\textit{šehu}-series is derived from a concessive conditional clause, headed by an interrogative pronoun. For instance, \textit{mišehu} ‘someone' consists of \textit{mi} ‘who', \textit{še} ‘that', and the third person singular masculine pronoun \textit{hu} literally meaning ‘who that it (is)' or ‘whoever it may be'.
The -\textit{šehu}{}-series is primarily used in non-negative contexts \citep[298--299]{Haspelmath1997}. However, a {}-\textit{šehu} indefinite can occur in the same clause with negation when it is modified by a relative clause \REF{ex:heb:29}, an adjective, or a prepositional phrase.

\ea\label{ex:heb:29}
    \gll Ani lo makir{}-${\emptyset}$ mi{}-šehu še-xolek-${\emptyset}$ al ze. \\
         \textsc{1sg} \textsc{neg} know-\textsc{sg.m} who-\textsc{indf.sg.m} \textsc{comp}{}-dispute-\textsc{sg.m} on this \\
    \glt ‘I don’t know anyone who disputes that.’ [Beʔofen Tivʔi]
\z 

However, unlike the \textit{af/šum}{}-series and the \textit{iš}{}-series, indefinite pronouns of the {}-\textit{šehu}{}-series have to follow negation in order to be interpreted in its scope. For instance, in \REF{ex:heb:30} \textit{mišehu} ‘someone’ precedes negation and can only be interpreted outside its scope.

\ea\label{ex:heb:30}
    \gll Mi{}-šehu še-ani makir-a lo sagar-${\emptyset}$ et ha-delet. \\
        who-\textsc{indf} that-\textsc{1sg} know-\textsc{sg.f} \textsc{neg} close.\textsc{pst-3sg.m} \textsc{acc} \textsc{def}{}-door \\
    \glt ‘Someone I know didn’t close the door.’ [constructed]
\z 

\subsection{Negative derivation and case marking} \label{sec:heb:3.3}

\subsubsection{Privative markers} \label{sec:heb:3.3.1}


Privative prepositions include \textit{bli} ‘without’, sometimes preceded by another preposition, and the negative particle \textit{lo}, which is always preceded by another preposition, e.g., \textit{le-lo} and \textit{be-lo} \citep[68]{Tzivoni1993}. Bare \textit{bli} can either take a noun phrase \REF{ex:heb:31a} or a non-finite clause \REF{ex:heb:31b}. When \textit{bli} is preceded by another preposition, it can only take a clausal complement \REF{ex:heb:31c} (see \sectref{sec:heb:2.4} for a discussion of negation in non-main clauses).

\ea\label{ex:heb:31}
    \ea\label{ex:heb:31a}
        \gll bli simxa\\
            without happiness\\
        \glt ‘without happiness’
    \ex\label{ex:heb:31b}
        \gll bli laševet\\
            without sit.\textsc{inf}\\
        \glt ‘without sitting’
    \ex\label{ex:heb:31c}
        \gll mi-bli laševet\\
             \textsc{abl}{}-without sit.\textsc{inf}\\
        \glt ‘without sitting’ \citep[68]{Tzivoni1993}
    \z 
\z 

The negative particle \textit{lo} most often combines with the dative/allative preposition \textit{le-}. It can take either a noun phrase \REF{ex:heb:32a} or a non-finite clause \REF{ex:heb:32b}. When \textit{lo} combines with \textit{be}{}- ‘in’, only a noun phrase complement is possible.

\ea\label{ex:heb:32}
    \ea\label{ex:heb:32a}
        \gll le-lo simxa\\
            \textsc{dat{}-neg} happiness\\
        \glt ‘without happiness’
    \ex\label{ex:heb:32b}
        \gll le-lo nua\\
            \textsc{dat{}-neg} move.\textsc{inf}\\
        \glt ‘without moving’ \citep[68]{Tzivoni1993}
    \z 
\z 

Neither \textit{bli} nor \textit{lelo} can take personal pronouns as complements. Instead, the preposition \textit{bilʔadey} ‘without’ is used \REF{ex:heb:33}. This is the sole function of \textit{bilʔadey} in the spoken language, but in formal registers it can also take a definite noun phrase as a complement \citep[74]{Tzivoni1993}. 

\ea\label{ex:heb:33}
    \gll Bilʔadey{}-ha lo i-hiy-e sipur.\\
         without-3\textsc{sg.f} \textsc{neg} \textsc{3.m.fut-}be-\textsc{sg} story\\
    \glt ‘Without her, there will be no story.’ \citep[74]{Tzivoni1993}
\z 

Hebrew has another set of privative markers, which are derived from participial forms. These markers can only modify noun phrases: \textit{xasar} (e.g., \textit{xasar buša} ‘shameless’), \textit{neʔedar} (e.g., \textit{neʔedar haskala} ‘uneducated’), and \textit{netul} (e.g., \textit{netul} \textit{kafeʔin} ‘caffeine free’). The former two forms have nominal equivalents, e.g., \textit{xoser buša} ‘shamelessness’ and \textit{heʔeder haskala} ‘lack of education’ \citep[55--66, 77--81]{Tzivoni1993}.

\subsubsection{Opposition and antonymy} \label{sec:heb:3.3.2}

Markers of opposition and antonymy include \textit{bilti}, \textit{lo}, \textit{i{}-}, and \textit{al}. \textit{bilti} modifies adjectives, e.g., \textit{bilti} \textit{racuy} ‘undesirable’. \textit{bilti} usually cannot modify adjectives that have lexical antonyms. For instance, *\textit{bilti tov} (‘ungood’) is arguably blocked by \textit{ra} ‘bad’. On the other hand, \textit{lo} is not subject to such a restriction, e.g., \textit{lo tov} ‘not good’. Another restriction is that \textit{bilti} cannot modify adjectives such as color terms, which do not have antonyms, e.g., \textit{*bilti lavan} ‘unwhite’ \citep[31]{Tzivoni1993}.

Since \textit{lo} is also used in stative predication (see \sectref{sec:heb:2.3}), there is often a structural ambiguity in the present tense between clausal negation and negative derivation \REF{ex:heb:34a}. On the other hand, structural ambiguity does not arise in other tenses due to the presence of the verbal copula (\ref{ex:heb:34b}--\ref{ex:heb:34c}).

\ea\label{ex:heb:34a} Clausal negation / negative derivation\\
    \gll Ze lo menumas.\\
        this \textsc{neg} polite\\
    \glt ‘It’s not polite.’ or: ‘It’s impolite.’ [constructed]
\z
    
\ea\label{ex:heb:34}
    \ea\label{ex:heb:34b} Clausal negation\\
        \gll Ze lo hay-a menumas.\\
            this \textsc{neg} be.\textsc{pst-3sg.m} polite\\
        \glt ‘It wasn’t polite.’ [constructed]
    \ex\label{ex:heb:34c} Negative derivation\\
        \gll Ze hay-a lo menumas.\\
             this be.\textsc{pst-3sg.m} \textsc{neg} polite\\
        \glt ‘It was impolite.’ [constructed]
    \z 
\z 

The prefix \textit{i{}-} modifies abstract nouns, e.g., \textit{i{}-racon} ‘unwillingness’, including action nominals, e.g., \textit{i{}-havana} ‘misunderstanding’ \citep[43--44]{Tzivoni1993}.
The prefix \textit{al}{}- modifies adjectives, e.g., \textit{al{}-xuti} ‘wireless’, and nouns, e.g., \textit{al-mavet} ‘immortality’ \citep[25--26]{Tzivoni1993}. This prefix is no longer productive.

\section{Other aspects of negation} \label{sec:heb:4}
\subsection{The scope of negation} \label{sec:heb:4.1}

Contrastive negation is expressed by employing prosodic stress on the focused constituent \REF{ex:heb:35a} or by placing the negative particle \textit{lo} immediately before that constituent \REF{ex:heb:35b} \citep[599]{Shor2020}.


\ea\label{ex:heb:35}
    \ea\label{ex:heb:35a} 
        \gll Ve-gam lo xani-ta \textsc{po}, xani-ta lemala.\\
             and-also \textsc{neg} park.\textsc{pst-2sg.m} here park.\textsc{pst-2sg.m} up\\
        \glt ‘And you didn’t park here either, you parked up [the street].’ \citep[599]{Shor2020}

    \ex\label{ex:heb:35b}
        \gll Yoce sof{}-šavua lo ha-karov, ha-ba.\\
            come.out.\textsc{sg.m} end{}-week \textsc{neg} \textsc{def}{}-upcoming \textsc{def}{}-next\\
        \glt ‘It will be not the upcoming weekend, but the next.’ \citep[599]{Shor2020}
    \z 
\z 

A focused element may also be preposed together with the negative particle \REF{ex:heb:36a}. In fact, preposing is obligatory when the negative particle is not in its normal pre-verbal position unless the clause is followed by a corrective \textit{but} clause. For instance, in \REF{ex:heb:36b} the first clause is odd without the presence of the second clause. Note that the conjunction that is used in corrective clauses is \textit{ela} ‘but’ rather than the ordinary adversative conjunction \textit{aval} ‘but’.

\ea\label{ex:heb:36} 
    \ea\label{ex:heb:36a}
        \gll Lo \textsc{it-i} dibar-ta.\\
            \textsc{neg} with-\textsc{1sg} talk.\textsc{pst-2sg.m}\\
        \glt ‘It wasn’t me you talked to.’ [constructed]
    \ex\label{ex:heb:36b}
        \gll Dibar-ta lo \textsc{it-i}, ela im ax šel-i.\\
            talk.\textsc{pst-2sg.m} \textsc{neg} with-1\textsc{sg} but with brother \textsc{poss{}-1sg}\\
        \glt ‘You didn’t talk to \textsc{me}, you talked to my \textsc{brother}.’ [constructed]
    \z 
\z 

\subsection{Polarity items} \label{sec:heb:4.2}
\subsubsection{Overview}
Negative Polarity Items (NPIs) are items that can co-occur with negation but cannot occur in a positive declarative episodic sentence, i.e., a sentence that relates a specific event. Hebrew NPIs include:

\begin{itemize}
\item 
phasal polarity adverbs, e.g., \textit{yoter} ‘anymore’

\item 
maximizers, e.g., \textit{meʔolam} ‘ever (in the past)’ (lit. ‘from eternity’), \textit{baxaim} ‘ever’ (lit. ‘in life’)

\item 
minimizers, e.g., \textit{nakaf ecba} ‘lift a finger’, \textit{acam ain} ‘sleep a wink’

\item 
vulgar idioms, e.g., \textit{lo meziz li} ‘I don’t give a damn’ (lit. ‘it doesn’t move [the tip of my belly button/finger/etc.]’)

\item 
idiomatic expressions, e.g., \textit{lo bixdi} ‘not for naught’

\end{itemize}

Positive Polarity Items (PPIs) are items that can occur in a positive declarative episodic sentence but cannot co-occur with negation in the same clause and be interpreted in its scope. Hebrew PPIs include: 

\begin{itemize}
\item 
phasal polarity adverbs, e.g., \textit{kvar} ‘already’, \textit{adain} ‘still’

\item 
minimizers, e.g., \textit{kcat} ‘a little’, \textit{tipa} ‘a little bit’ (lit. ‘a drop’)

\item 
approximators, e.g., \textit{dey} ‘rather’, \textit{beʔerex} ‘approximately’

\end{itemize}

The rest of this section focuses on those items that appear to be of typological interest.

\subsubsection{\textit{meʔolam} and \textit{leʔolam} ‘ever’} \label{sec:heb:4.2.1}


Hebrew has two adverbs that are built on the word \textit{olam} ‘eternity’: \textit{meʔolam} ‘ever (in the past)’ and \textit{leʔolam} ‘ever (in the present or future)’. Their literal meaning is ‘from eternity’ and ‘to eternity’, respectively. In the spoken language, both of these items are NPIs.

The adverb \textit{meʔolam} has a broader distribution than \textit{leʔolam}. For example, the former, but not the latter, can occur in superlatives \REF{ex:heb:37}, questions and conditionals.

\ea\label{ex:heb:37}
    \ea[]{\label{ex:heb:37a}
        \gll Ha-maim haxi meraʔanen-im še-taʔam-ti meʔolam.\\
            \textsc{def}{}-water most refreshing-\textsc{pl} that-taste.\textsc{pst-1sg} ever\\
        \glt ‘The most refreshing water I have ever tasted.’ [Beʔofen Tivʔi]}
    \ex[*]{\label{ex:heb:37b}
        \gll Ze ha-seret haxi tov še-ti-rʔ-e leʔolam!\\
            it \textsc{def}{}-movie most good that-2\textsc{.fut}{}-see-\textsc{sg.m} ever\\
        \glt (‘It’s the best movie you’ll ever see!’) [constructed]}
    \z 
\z 

However, in formal registers \textit{leʔolam} can also be used in affirmative sentences, where it means ‘forever’ \REF{ex:heb:38}. \textit{meʔolam} is used in such contexts mostly in the idiom \textit{meʔaz u-meʔolam} ‘since time immemorial’.

\ea\label{ex:heb:38}
    \gll Leʔolam e-zkor l-ahen xesed ziknat-an.\\
         forever \textsc{1sg.fut}{}-remember \textsc{dat-3pl.f} devotion old.age-\textsc{3pl.f}\\
    \glt ‘I will always remember the devotion of their old age.’ [Beʔofen Tivʔi]
\z 

\subsubsection{\textit{tipa} ‘a little bit’} \label{sec:heb:4.2.2}

Minimizers are idiomatic expressions of minimal amount, e.g., \textit{lift a finger, sleep a wink, say a word} \citep[1662]{Giannakidou2011}. Minimizers are usually associated with NPIs, but in Modern Hebrew some of them are PPIs. \textit{tipa} ‘a little bit’ (lit. ‘a drop’) is particularly interesting because in one construction it behaves like a PPI, and in another like an NPI. As an adverb, \textit{tipa} is a PPI (\ref{ex:heb:39a}--\ref{ex:heb:39b}). However, note that \textit{tipa} can co-occur with negation when it is in the focus of the scalar additive particle \textit{afilu} ‘even’ \REF{ex:heb:39c}.

\ea\label{ex:heb:39}
    \ea[]{\label{ex:heb:39a}
        \gll Ani tipa meʔuxzav-${\emptyset}$.\\
             1\textsc{sg} drop disappointed-\textsc{sg.m}\\
        \glt ‘I’m a little bit disappointed.’ [Beʔofen Tivʔi]}
    \ex[*]{\label{ex:heb:39b}
        \gll Ani lo tipa meʔuxzav-${\emptyset}$.\\
             1\textsc{sg} \textsc{neg} drop disappointed-\textsc{sg.m}\\
        \glt (‘I’m not one bit disappointed.’) [constructed]}
    \ex[]{\label{ex:heb:39c}
        \gll Ani afilu lo tipa mešuʔašaʔ{}-at.\\
             1\textsc{sg} even \textsc{neg} drop amused-\textsc{sg.f}\\
        \glt ‘I’m not even one bit amused.’\footnote{\url{http://www.themarker.com/misc/smartphonexmlfeed/1.2655748} (Accessed on 2019-09-29)}}
    \z 
\z 

On the other hand, when \textit{tipa} functions as a noun in a pseudo{}-partitive construction (i.e., ‘a drop of’), it is an NPI \REF{ex:heb:40}.

\ea\label{ex:heb:40}
    \ea[]{\label{ex:heb:40a}
        \gll En l-i tipat pratiyut.\\
            \textsc{neg.ex} \textsc{dat-1sg} drop.of privacy.\\
        \glt ‘I don’t have any privacy whatsoever.’ [Beʔofen Tivʔi]}
    \ex[*]{\label{ex:heb:40b}
        \gll Yeš l-i tipat pratiyut.\\
            \textsc{ex} \textsc{dat-1sg} drop.of privacy.\\
        \glt (‘I have a little privacy.’) [constructed]}
    \z 
\z 

\subsubsection{\textit{ey paʔam} ‘ever’ and \textit{paʔam} ‘some time, once’} \label{sec:heb:4.2.3}

The adverbs \textit{ey paʔam} ‘ever’ (lit. ‘which time’) and \textit{paʔam} ‘some time, once’ (lit. ‘time’) are interchangeable in most contexts, e.g., questions \REF{ex:heb:41a}, but \textit{ey paʔam} ‘ever’ is an NPI, and \textit{paʔam} ‘some time, once’ is a PPI. In positive episodic sentences, only \textit{paʔam} can be used \REF{ex:heb:41b}. In comparatives and superlatives, only \textit{ey paʔam} can be used \REF{ex:heb:41c}. Interestingly, \textit{ey paʔam} cannot occur in the same clause with negation \REF{ex:heb:41d} (\citealt{Tzivoni1993}: 155, \citealt{FrancezGoldring2012}: 366). In this context, \textit{af paʔam} ‘never’ (see \sectref{sec:heb:3.2.1}) or \textit{baxaim} ‘ever’ are used instead. 


\ea\label{ex:heb:41}
    \ea[]{\label{ex:heb:41a}  
        \gll Nahag-ta (ey) paʔam l-a-avoda?\\
             drive.\textsc{pst-2sg.m} which time to-\textsc{def}{}-work\\
        \glt ‘Have you ever driven to work?’ (\citealt{FrancezGoldring2012}: 366)}
    \ex[]{\label{ex:heb:41b}
        \gll Pagaš-ti ot-o paʔam b-a-kanyon.\\
             meet.\textsc{pst-1sg} \textsc{acc-3sg} time in-\textsc{def}{}-mall\\
        \glt ‘I’ve met him once at the mall.’\footnote{\url{https://www.ynet.co.il/Ext/App/TalkBack/CdaViewOpenTalkBack/0,11382,L-4944298-2,00.html} (Accessed on 2019-09-29)}}
    \ex[]{\label{ex:heb:41c}
        \gll Zot ha-memšala haxi grua ey paʔam.\\
             this.\textsc{sg.f} \textsc{def}{}-government most bad.\textsc{sg.f} which time\\
        \glt ‘This is the worst government ever.’ (\citealt{FrancezGoldring2012}: 366)}
    \ex[?]{\label{ex:heb:41d}
        \gll Ani lo nahag-ti (ey) paʔam l-a-avoda.\\
             I \textsc{neg} drive.\textsc{pst-1sg} which time to-\textsc{def}{}-work\\
        \glt (‘I haven't ever driven to work.’) (\citealt{FrancezGoldring2012}: 366)}
    \z 
\z 

\subsection{Marking of NPs in the scope of negation} \label{sec:heb:4.3}

Marking of NPs is not affected by negation.

\subsection{Reinforcing negation} \label{sec:heb:4.4}

Negation can be reinforced by intensifiers such as \textit{bixlal} ‘at all' \REF{ex:heb:42a} and minimizers \REF{ex:heb:42b}.

\ea\label{ex:heb:42}
    \ea\label{ex:heb:42a}
        \gll Ani bixlal lo maskim-a.\\
             \textsc{1sg} at.all \textsc{neg} agree.\textsc{prs-sg.f}\\
        \glt ‘I don't agree at all.' [Beʔofen Tivʔi]
    \ex\label{ex:heb:42b}
        \gll Lo amar-ti mila.\\
             \textsc{neg} say.\textsc{pst-1sg} word\\
        \glt ‘I didn’t say a word.' [Beʔofen Tivʔi]
    \z 
\z

\subsection{Negation, coordination and complex clauses} \label{sec:heb:4.5}

Coordination of a positive and a negative clause may involve verb phrase ellipsis \REF{ex:heb:43}. This type of ellipsis is also available in coordination of two positive clauses.

\ea\label{ex:heb:43}
    \gll Anaxnu yodʔ{}-im ve-atem lo [yodʔ{}-im]. \\
        1\textsc{pl} know.\textsc{prs-pl.m} and-you.\textsc{pl} \textsc{neg} know.\textsc{prs-pl.m} \\
    \glt ‘We know and you don’t [know].’ [Beʔofen Tivʔi]
\z 

Hebrew does not have negative coordinators. However, there is a construction in which the negative particle \textit{lo} precedes each of the conjuncts \REF{ex:heb:44a}. This construction is more emphatic than its unmarked counterpart \REF{ex:heb:44b}.

\ea\label{ex:heb:44}
    \ea\label{ex:heb:44a}
        \gll Lo hay-a l-i ixpat lo me-ha-yehudi-m ve-lo me-ha-aravi-m. \\
            \textsc{neg} be.\textsc{pst-3sg.m} \textsc{dat-1sg} care \textsc{neg} from-\textsc{def}{}-Jew-\textsc{pl} and-\textsc{neg} from-\textsc{def}{}-Arab-\textsc{pl}\\
        \glt ‘I cared neither about the Jews nor about the Arabs.’ \citep[353]{Tzivoni1993}
    \ex\label{ex:heb:44b}
        \gll Lo hay-a l-i ixpat me-ha-yehudi-m o me-ha-aravi-m.\\
            \textsc{neg} be.\textsc{pst-3sg.m} \textsc{dat-1sg} care from-\textsc{def}{}-Jew-\textsc{pl} or from-\textsc{def}{}-Arab-\textsc{pl}\\
        \glt ‘I didn’t care about the Jews or the Arabs.’ [constructed]
    \z 
\z

When the verb precedes the conjuncts, e.g., \REF{ex:heb:44a}, clausal negation is obligatory. However, when the verb follows the conjuncts \REF{ex:heb:45}, clausal negation is optional \citep[354]{Tzivoni1993}.

\ea\label{ex:heb:45}
    \gll Lo ha-doktor ve-lo ha-profesor (lo) dibr-u al šituf našim b-a-šilton.\\
        \textsc{neg} \textsc{def}{}-doctor and-\textsc{neg} \textsc{def}{}-professor \textsc{neg} talk.\textsc{pst-3pl} about participation.of women in-\textsc{def}{}-administration\\
    \glt ‘Neither the doctor nor the professor […] talked about letting women take part in the administration.’ \citep[354]{Tzivoni1993}
\z 

\subsection{Miscellaneous aspects of negation} \label{sec:heb:4.6}

\subsubsection{Negative transport} \label{sec:heb:4.6.1}


Negative transport is a phenomenon where negation of a higher clause is interpreted as negating a subordinate clause. Neg-raising predicates are listed below based on the classification in \citet{Horn1978}.

\begin{itemize}
    \item \textbf{Opinion:} \textit{xošev} ‘think', \textit{maʔamin} ‘believe', \textit{savur} ‘of the opinion that', \textit{mešaʔer} ‘suppose', \textit{meniax} ‘assume', \textit{metaʔer leʔacmi} ‘I imagine', \textit{nirʔe li} ‘it seems to me' \citep[275]{Tzivoni1993}, \textit{nidme li} ‘it seems to me', and \textit{cofe} ‘predict'. 
    \item \textbf{Perception}: \textit{nirʔe} ‘look like', \textit{nišma} ‘sound like', and \textit{margiš} ‘feel like'.
    \item \textbf{Probability}: \textit{savir} ‘likely'.
    \item \textbf{Intention/volition:} \textit{roce} ‘want' \citep[280]{Tzivoni1993}, \textit{xošev} ‘think of' \citep[273]{Tzivoni1993}, \textit{ba li} ‘I want'\textit{, meʔunyan} ‘interested in', \textit{mitkaven} ‘intend', and \textit{metaxnen} ‘plan'. 
    \item \textbf{Judgment/obligation:} \textit{amur} ‘supposed', \textit{carix} ‘need', and \textit{mamlic} ‘recommend'. 
\end{itemize}

Some of the aforementioned categories also include non-neg-raising predicates:

\begin{itemize}
    \item \textbf{Opinion:} \textit{menaxeš} ‘guess', \textit{mešuxna} ‘convinced', and \textit{batuax} ‘sure'.
    \item \textbf{Intention/volition:} \textit{mekave} ‘hope' and \textit{maʔadif} ‘prefer'.
    \item \textbf{Judgment/obligation:} \textit{xayav} ‘must'.
\end{itemize}

The difference between a neg-raising predicate and a non-neg-raiser is exemplified in \REF{ex:heb:46} in the domain of necessity modals. Example \REF{ex:heb:46a} contains the neg-raising \textit{carix} ‘need', and its salient reading is one where the modal takes wide scope over negation: it is necessary that children not be exposed to such horrors. By contrast, \textit{xayav} ‘must' in \REF{ex:heb:46b} is not a neg-raiser, and negation takes wide scope over the modal: it is not necessary that children be exposed to such horrors. Also see \sectref{sec:heb:3.2} for a discussion of the relevance of neg-raising to licensing of negative indefinites.

\ea\label{ex:heb:46}
    \ea\label{ex:heb:46a}
        \gll Yelad-im lo crix-im lehexasef le-zvaʔ-ot kaʔele.\\
            child-\textsc{pl} \textsc{neg} need{}-\textsc{pl.m} be.exposed.\textsc{inf} \textsc{dat-}horror-\textsc{pl.f} such\\
        \glt ‘Children shouldn’t be exposed to such horrors.’\footnote{\url{https://www.haaretz.co.il/gallery/television/tv-review/1.1866196} (Accessed on 2020-07-03)}
    \ex\label{ex:heb:46b}
        \gll Yelad-im lo xayav-im lehexasef le-zvaʔ{}-ot kaʔele.\\
            child-\textsc{pl} \textsc{neg} must{}-\textsc{pl.m} be.exposed.\textsc{inf} \textsc{dat-}horror-\textsc{pl.f} such\\
        \glt ‘Children don’t have to be exposed to such horrors.’ [constructed]
    \z 
\z

\subsubsection{Metalinguistic negation} \label{sec:heb:4.6.2}

Metalinguistic negation is a use of negation where what is negated is not the assertive content of the proposition but the way it is expressed \citep[363]{Horn1989}. Metalinguistic negation looks like ordinary negation in Hebrew.

\ea\label{ex:heb:47}
    \gll Ze lo tov, ze meʔule.\\
        this \textsc{neg} good this great\\
    \glt ‘It’s not good, it’s great.’ [constructed]
\z 

\subsubsection{Non-negative uses of negatives and expletive negation} \label{sec:heb:4.6.3}

Expletive negation—i.e., a non-negative use of a negative marker—is found in several contexts, and it is always optional. First, expletive negation is often used in free relative clauses \REF{ex:heb:48}. In this context, expletive negation gives rise to an interpretation which is similar to English \textit{{}-ever}, i.e., speaker indifference or ignorance \citep[41]{Eilam2007}.

\ea\label{ex:heb:48}
    \gll Ma še-dani lo katav-${\emptyset}$, hitparsem-${\emptyset}$ b-a-iton. \\
        what that-Dani \textsc{neg} write.\textsc{pst}{}-3\textsc{sg.m} publish.\textsc{pass.pst-}3\textsc{sg.m} in-\textsc{def}{}-newspaper\\
    \glt ‘Whatever Danny wrote was published in the newspaper.’ \citep[41]{Eilam2007}
\z 

A related use of expletive negation is in exclamative rhetorical questions \citep[167]{RubinsteinEtAl2015}.

\ea\label{ex:heb:49}
    \gll Be-ma lo heʔešim-u ot-o!\\
        in-what \textsc{neg} blame.\textsc{pst}{}-3\textsc{pl} \textsc{acc-3sg.m}\\
    \glt ‘The things he’s been blamed for!’ (lit. ‘What hasn't he been blamed for?!') \citep[167]{RubinsteinEtAl2015}
\z 

Second, expletive negation is used in clausal complements of \textit{ad} ‘until’ \REF{ex:heb:50}, \textit{bli} ‘without’ and, less often, \textit{lifney} ‘before’ \citep[168]{RubinsteinEtAl2015}.

\ea\label{ex:heb:50}
    \gll Ad še-lo ya-gia-${\emptyset}$ milyarder rusi, nyukasel ti-tkaše-${\emptyset}$. \\
        until that-\textsc{neg} \textsc{3.m.fut-}arrive{}-\textsc{sg} billionaire Russian Newcastle 3\textsc{.f.fut}{}-have.trouble-\textsc{sg}\\
    \glt ‘Newcastle will be in trouble until a Russian billionaire comes along.’ \citep[168]{RubinsteinEtAl2015}
\z 

Expletive negation is not used in complements of emotive predicates such as ‘fear’ and ‘doubt’. However, it is used in complements of negated ‘surprise’ predicates \citep[169]{RubinsteinEtAl2015}.

\ea\label{ex:heb:51}
    \gll Ani lo e-tpale im hu lo yazam-${\emptyset}$ et kol ha-takrit ha-mexoʔer-et ha-zo.\\
        1\textsc{sg} \textsc{neg} 1\textsc{sg.fut}{}-be.surprised if 3\textsc{sg.m} \textsc{neg} initiate.\textsc{pst-3sg.m} \textsc{acc} all \textsc{def}{}-incident \textsc{def}{}-ugly-\textsc{sg.f} \textsc{def}{}-this\\
    \glt ‘I wouldn’t be surprised if he is the one behind this ugly incident.’ \citep[169]{RubinsteinEtAl2015} 
\z 

\subsubsection{Negated quantifier phrase constructions} \label{sec:heb:4.6.4}

Modern Hebrew has two types of negated quantifier phrase constructions: (i) \textit{lo kol} ‘not all (NP)’ and \textit{lo harbe} ‘not many (NP)’; (ii) \textit{lo meʔat} ‘quite a few’. The difference between the two types is that the former may only occur pre-verbally (\ref{ex:heb:52a}--\ref{ex:heb:52b}), whereas the latter may also occur post-verbally (\ref{ex:heb:53a}--\ref{ex:heb:53b}). Another difference is related to their interaction with prepositions. In the case of \textit{lo kol}, the negative particle \textit{lo} has to precede the entire prepositional phrase \REF{ex:heb:52c}. On the other hand, a preposition takes \textit{lo meʔat} in its entirety as its complement \REF{ex:heb:53c}.

\ea\label{ex:heb:52}
    \ea[]{\label{ex:heb:52a}
        \gll Lo kol ha-student-im ba-u.\\
             \textsc{neg} all \textsc{def}{}-student-\textsc{pl} come.\textsc{pst-3pl}\\
        \glt ‘Not all of the students came.’ [constructed]}
    \ex[*]{\label{ex:heb:52b}
        \gll Ba-u lo kol ha-student-im.\\
             come.\textsc{pst-3pl} \textsc{neg} all \textsc{def}{}-student-\textsc{pl} \\
        \glt (‘Not all of the students came.’) [constructed]}
    \ex[]{\label{ex:heb:52c}
        \gll Lo le-kol ha-student-im yeš mexonit.\\
             \textsc{neg} \textsc{dat}{}-all \textsc{def}{}-student-\textsc{pl} \textsc{ex} car\\
        \glt ‘Not all of the students have a car.’ [constructed]}
    \z 
\z

\ea\label{ex:heb:53}
    \ea\label{ex:heb:53a}
        \gll Lo meʔat student-im ba-u.\\
             \textsc{neg} few student-\textsc{pl} come.\textsc{pst-3pl}\\
        \glt ‘Quite a few students came.’ [constructed]
    \ex\label{ex:heb:53b}
        \gll Ba-u lo meʔat student-im.\\
             come.\textsc{pst-3pl} \textsc{neg} few student-\textsc{pl} \\
        \glt ‘Quite a few students came.’ [constructed]
    \ex\label{ex:heb:53c}
        \gll Le{}-lo meʔat student-im yeš mexonit.\\
             \textsc{dat}{}-\textsc{neg} few student-\textsc{pl} \textsc{ex} car\\
        \glt ‘Quite a few students have a car.’ [constructed]
    \z 
\z

\subsubsection{Diachronic notes and observations} \label{sec:heb:4.6.5}

The expansion of the functions of the negative existential \textit{en} into the verbal domain is an instance of the Negative Existential Cycle (\citealt{Croft1991}; \citealt{Veselinova2014,Veselinova2016}). Hebrew exhibits an overlap of two types in the cycle: A{\textasciitilde}B and B{\textasciitilde}C. On the one hand, it belongs to type A{\textasciitilde}B because \textit{en} is used in existential negation only in the present tense, whereas the standard negator \textit{lo} is used in other tenses (see \sectref{sec:heb:2.3}). On the other hand, it belongs to type B{\textasciitilde}C because \textit{en} may be used to negate present tense verbs, which are historically derived from the present participle (see \sectref{sec:heb:2.1}). This overlap dates back to Biblical Hebrew \citep{NaudéEtAl2019,NaudéEtAl2022}.

\section{Summary} \label{sec:heb:5}

In clausal negation, the choice of negative particle depends on three criteria, namely sentence type (declarative or imperative), type of predication, and tense. 
The main negation strategies in Hebrew are summarized in \tabref{tab:heb:1}. 

\begin{table}
    \centering\small
    \begin{tabularx}{\textwidth}{lQl}
         
\lsptoprule
Negative marker & Function & Section\\
\midrule
lo (negative particle) 
    & verbal predication & \ref{sec:heb:2.1} \\
\tablevspace
    & nominal, adjectival, and locative predication  & \ref{sec:heb:2.3}  \\
\tablevspace
    & existential and possessive predication, except in the present tense  & \ref{sec:heb:2.3}  \\
\tablevspace
    & negative reply  & \ref{sec:heb:3.1}   \\
\tablevspace
    & constituent negation  & \ref{sec:heb:4.1}, \ref{sec:heb:4.6.2}  \\
\tablevspace
    & impersonal prohibition  & \ref{sec:heb:2.2}  \\
\midrule
\textit{en} (negative copula) 
    & existential and possessive predication in the present tense  & \ref{sec:heb:2.3}  \\
\tablevspace
    & verbal predication in the present tense, formal register  & \ref{sec:heb:2.1}  \\
\tablevspace
    & nominal, adjectival, and locative predication in the present tense, formal register  & \ref{sec:heb:2.3}  \\
\tablevspace
    & impersonal prohibition, formal register  & \ref{sec:heb:2.2}  \\
\midrule
\textit{al} (negative particle) 
    & prohibitives  & \ref{sec:heb:2.2}  \\
\tablevspace
    & jussive construction  & \ref{sec:heb:2.2}  \\
\midrule
\textit{terem} (adverb) 
	& phasal polarity adverb ‘not yet’ & \ref{sec:heb:2.5}  \\
\lspbottomrule
    \end{tabularx}
    \caption{Summary of negation strategies in Hebrew}
    \label{tab:heb:1}
\end{table}

The constructions with \textit{lo} and \textit{terem} are symmetric with the affirmative. On the other hand, existential and possessive constructions with \textit{en} are asymmetric with the affirmative because \textit{en} is in contrastive distribution with the affirmative existential particle \textit{yeš}. Copular constructions with \textit{en} typically exhibit a constructional asymmetry because \textit{en} is followed by a clitic pronoun, which is absent in the affirmative. However, \textit{en} may also occur clause-initially rather than after the subject, in which case it does not co-occur with a clitic pronoun, and hence there is no asymmetry. Finally, prohibitives exhibit a paradigmatic asymmetry in those cases where a verb has retained the dedicated imperative form, which is restricted to the affirmative (see \sectref{sec:heb:2.2}).

In addition to the above-mentioned negative markers, Modern Hebrew has several negative indefinites (see \sectref{sec:heb:3.2}), a wide variety of negative derivations (see \sectref{sec:heb:3.3}), and several negative counterfactual conditional conjunctions (see \sectref{sec:heb:2.4}).

 

\section*{Abbreviations}
\largerpage[1.5]
\begin{tabularx}{.45\textwidth}{lQ}
{\scshape 1} & first person\\
{\scshape 2} & second person\\
{\scshape 3} & third person\\
{\scshape abl} & ablative\\
{\scshape acc} & accusative\\
{\scshape comp} & complementizer\\
{\scshape cop} & copula\\
{\scshape dat} & dative\\
{\scshape def} & definite\\
{\scshape ex} & existential\\
{\scshape f} & feminine\\
{\scshape fut} & future\\
{\scshape imp} & imperative\\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
{\scshape indf} & indefinite\\
{\scshape inf}  & infinitive\\
{\scshape m} & masculine\\
{\scshape neg} & negation\\
{\scshape pass} & passive\\
{\scshape pl} & plural\\
{\scshape pn} & proper name\\
{\scshape poss} & possessive\\
{\scshape proh} & prohibitive\\
{\scshape prs} & present\\
{\scshape pst} & past\\
{\scshape q} & question particle\\
{\scshape sg} & singular\\
\end{tabularx}

\newpage
\sloppy\printbibliography[heading=subbibliography,notkeyword=this]

\end{document} 
