\documentclass[output=paper,colorlinks,citecolor=brown,draft]{langscibook}
\ChapterDOI{10.5281/zenodo.18234069}
\author{Eline Visser\affiliation{Uppsala University}}
\title{Negation in Kalamang}  
\abstract{The Papuan language Kalamang has standard clausal negation with a predicate clitic \textit{=nin}. Prohibitives are formed with a suffix \textit{-mun} on the pronoun, and a clitic \textit{=in} on the verb. Non-verbal clauses may be negated with propositional negator \textit{ge} `no; not' or with negative existential \textit{saerak} `\textsc{neg.ex}'. Kalamang has dedicated negative verbs with the meaning `to not want', `cannot' and `to not know'. Negative replies are made with \textit{ge} `no; not'. \textit{Ge} or derivatives thereof are used in other negative constructions or expressions such as polar question tag \textit{ye ge} `or not', negative conditional \textit{get me} `if not' and negative answer \textit{ge mera} `oh, nothing'.

\keywords{Papuan languages, negation, verbal morphology, negative lexicalization, prohibitive}
}


\IfFileExists{../localcommands.tex}{
   \input{../localpackages}
   \input{../localcommands}
   \input{../localhyphenation}
  \addbibresource{../localbibliography.bib}
    \togglepaper[23]
}{}

\begin{document}
\maketitle

\section{The language}
Kalamang (also known as Karas, ISO: \href{https://iso639-3.sil.org/code/kgv}{kgv}, Glottocode: \href{https://glottolog.org/resource/languoid/id/kara1499}{kara1499}) is a Papuan language of the Greater West Bomberai family spoken by around 130 people in eastern Indonesia, in West Papua province. Kalamang is spoken on the biggest of the three Karas Islands. This island is home to two villages, Mas and Antalisa. In both villages, Kalamang is the main indigenous language. All Kalamang speakers are bilingual in Papuan Malay (a cover term for local varieties of Malay used in the provinces Papua and West Papua). Papuan Malay is the main language in mixed-language marriages, in village meetings, at school and for interaction with anyone who is not a Kalamang speaker. Kalamang is an endangered language: there are no fluent speakers born after 1990.

Besides Kalamang, the Greater West Bomberai family is composed of Iha and Mbaham, which form their own subgroup, and a subgroup formed by the Timor-Alor-Pantar languages \citep{UsherSchapper2021}, see \figref{fig:kal:1}.

\begin{figure}
    \centering
\begin{forest} 
[Greater West Bomberai
    [Kalamang]
    [Mbaham-Iha]
    [Timor-Alor-Pantar
        [Bunaq]
        [East Timor]
        [Alor-Pantar]
    ]
]
\end{forest}
    \caption{Greater West Bomberai language family}
    \label{fig:kal:1}
\end{figure}


In this chapter, I describe aspects of negation in Kalamang following the questionnaire provided by the editors \parencitetv{chapters/questionnaire}. First, I describe clausal negation in \sectref{sec:ninneg}, including standard negation, the negation of declarative verbal main clauses \citep{Miestamo2005}. Non-clausal negation is described in \sectref{sec:nonclneg}, including negative replies, negative indefinites and negative derivations. I treat other aspects of negation in \sectref{sec:othneg}, including among other topics the scope of negation in multi-verb constructions, the negation of quantifiers and reinforcement of negation.

To understand the examples in this chapter, it is good to know that Kalamang has two verb classes: regular and irregular verbs. Regular verbs may end in any phoneme and have an invariable root. Irregular verbs have a vowel-final root, with variants ending in \textit{-n} and \textit{-t}, depending on the inflection the verb carries. Irregular verbs also have two variants when they are uninflected: one vowel-final and one with \textit{-n}, which seem to be used interchangeably. Within the class of irregular verbs there is some minor variation, but most verbs pattern like \textit{bara} `to descend', illustrated in Table~\ref{tab:kal:verbcomp}, where it is contrasted with the regular verb \textit{muap} `to eat'.

\begin{table}[ht!]
\centering
	\caption{Behaviour of the regular verb \textit{muap} `to eat' and the irregular verb \textit{bara} `to descend'}
		\begin{tabular}{l l l l }
			\lsptoprule
			 && \textit{muap} & \textit{bara}/\textit{bara\textbf{n}}\\
			 && `eat' & `descend'\\ \midrule
			\textit{=i} & \textsc{plnk} & \textit{muap=i} & \textit{bara\textbf{n}=i}\\
			\textit{=et} & \glet & \textit{muap=et} & \textit{bara\textbf{t}=et} \\
			\textit{=kin} & \glkin  & \textit{muap=kin}& \textit{bara\textbf{t}=kin} \\
			\textit{=nin} & \textsc{neg} & \textit{muap=nin} & \textit{bara\textbf{t}=nin}\\
			\textit{=in} & \textsc{proh} & \textit{muap=in} & \textit{bara\textbf{t}=in}\\
			\textit{=te} & \textsc{imp} & \textit{muap=te} & \textit{bara=te}\\ 	\lspbottomrule
		\end{tabular}
	\label{tab:kal:verbcomp}
\end{table}

Another important aspect of Kalamang grammar is that almost anything can be the predicate (including nouns, nouns inflected with the locative postposition, numerals and demonstratives), and that mood, aspect and modality enclitics attach to the predicate. There is no tense marking. Note also that Kalamang has no adjective word class: properties like colours and attributes are expressed with monovalent verbs. Attributive use of verbs is indicated with the attributive clitic \textit{=ten}.

\largerpage
All data in this chapter is drawn from the Kalamang corpus, which is based on eleven months of field work carried out by the author between 2015 and 2019. The corpus contains spontaneous conversations and narratives (abbreviated “conv” or “narr”), stimulus-based conversations such as director-matcher tasks and stimulus-based narratives such as the pear story (both abbreviated “stim”) and elicitation (abbreviated “elic”). Examples from recorded natural speech are labelled with the genre and number of the recording and the time stamp of the utterance, e.g. [conv8\_2:03], while references to elicited examples code the topic of elicitation, e.g. [elic\_dem]. All examples can be found with help of these codes in the corpus.\footnote{The Kalamang corpus can be accessed at \url{ http://hdl.handle.net/10050/00-0000-0000-0003-C3E8-1}. The example tags in this chapter are clickable links that lead to the relevant corpus entry.} The natural spoken part of the corpus contains 15.5 hours of transcribed recordings, with almost 70000 words in total. A reference grammar of Kalamang is published as \citet{Visser2022}.


\section{Clausal negation}
\largerpage
\label{sec:ninneg}
This section starts with a description of standard negation, the negation of declarative verbal main clauses. It then goes on to discuss negation in non-declarative, non-verbal and non-main clauses, in \sectref{sec:negnondecl}\textendash\ref{sec:nonmainneg}. Dedicated negative verbs with the meanings `not know', `not like' and `cannot' are discussed in \sectref{sec:lexicalneg}.

\subsection{Standard negation}
\label{sec:standneg}
Standard negation is achieved by cliticising negator \textit{=nin} to the predicate. (\ref{exe:kal:sem}) contrasts an affirmative and a negated clause, showing that the only difference between the two is the negator clitic.\footnote{I do not use punctuation in the Kalamang line, because it is the underlying (phonemic) form and because it is often taken from the bigger context of a narrative or conversation, where it is more often than not unclear where the sentence boundaries are. The English translation includes interpunction to increase readability. The interpunction, like the translation itself, is a free interpretation of the original Kalamang.} 


\ea \label{exe:kal:sem}
\ea \label{exe:kal:masem}{
	\gll ma sem\\
	\textsc{3sg} be.afraid\\
	\glt `She's afraid.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA2-F}{[conv11\_2:14]}}}

\ex \label{exe:kal:SV}{
	\gll ma sem=\textbf{nin}\\
	\textsc{3sg} be.afraid=\textsc{neg}\\
	\glt	`She's not afraid.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_16:33]}}}
\z 
\z  

Examples (\ref{exe:kal:negv}) and (\ref{exe:kal:negadj}) show a negated transitive verb and a negated intransitive stative verb. In both cases, the constituent order of the non-negated counterpart is the same, as is shown in the a-examples. 

\ea 	\label{exe:kal:negv}	
\ea \gll ma tumun=at kona\\
\textsc{3sg} child=\textsc{obj} see\\
\glt `He sees the child.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_25:19]}}
\ex \gll ma yuon=at konat=\textbf{nin}\\
\textsc{3sg} sun=\textsc{obj} see=\textsc{neg}\\
\glt 	`He didn't see the sun.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_22:55]}}
\z
\z 

\ea \label{exe:kal:negadj}	
\ea \gll wa me kang\\
\textsc{prox} {\glme} sharp\\
\glt `This one is sharp.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB6-D}{[conv18\_5:50]}}
\ex \gll wa me mang=\textbf{nin}\\
\textsc{prox} {\glme} bitter=\textsc{neg}\\
\glt	`This one isn't bitter.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB1-3}{[narr34\_3:10]}}
\z \z 

In general, the negative construction is symmetrical to the affirmative. Negative clauses differ from affirmatives in that there are some co-occurrence restrictions with other markers on the predicate.

The clitic \textit{=kin}, used in affirmative clauses for the expression of volition and about-to-happen situations (glossed \textsc{vol}), is incompatible with predicate negator \textit{=nin}. This is illustrated in example~(\ref{exe:kal:kinneg}). The range of semantic distinctions that \textit{=kin} can express is lost, or underspecified, once a clause is negated. Thus, (\ref{exe:kal:turnin}) is ambiguous between a reading that the child does not want to eat, or any other of the meanings expressed by \textit{=kin} in affirmative clauses, and a plain negation of the verb `to eat', i.e. `the child does not eat'. Note that for explicitly expressing `not want' or `not agree', the Kalamang speaker can use a dedicated expression \textit{suka}-\textsc{poss} \textit{ge}, as discussed in \sectref{sec:lexicalneg}.

\ea 	\label{exe:kal:kinneg}
\ea \gll tumun muap=kin\\
child eat=\textsc{vol}\\
\glt `The child wants to eat.'
\ex [*]	{\label{exe:kal:turkinin}
	\gll tumun muap=\textbf{kin=nin}\\
	child eat={\glkin}\textsc{=neg}\\
	\glt 	`The child doesn't want to eat.' }

\ex 	[]	{\label{exe:kal:turnin}
	\gll tumun muap=\textbf{nin}\\
	child eat=\textsc{neg}\\
	\glt	`The child doesn't (want to) eat.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg]}}	}
\z \z 

Irrealis mood, indicated by the enclitic \textit{=et} on the predicate, is compatible with negator \textit{=nin}, following it. It always gets the reading of a negative conditional, whereas irrealis mood in affirmative clauses is not exclusively used for conditionals. (For a discussion of the form \textit{eranun} `cannot' in \ref{exe:kal:newerninet} see \sectref{sec:lexicalneg}.)

\ea 	\label{exe:kal:neweret}
\ea \label{exe:kal:newerpos}
	\gll ki newer=\textbf{et} eba in kahetmat=et\\
	\textsc{2pl} pay={\glet} then \textsc{1pl.excl} open={\glet}\\
	\glt `If you pay, then we open.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B85-F}{[narr5\_3:46]}}
\ex \label{exe:kal:newerninet}
	\gll  ki tok newer=\textbf{nin=et} ki sabua nerungga rat-un eranun\\
	\textsc{2sg} yet pay=\textsc{neg}={\glet} \textsc{2pl} tent inside.\textsc{lat} move-\textsc{nmlz} cannot\\
	\glt `If you haven't paid yet, you cannot go inside the tent.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B85-F}{[narr5\_2:18]}}
\z \z 

Two aspectual markers are compatible with negated verbs, but change in meaning. The aspectual particle \textit{tok} `yet; still; first' can be combined with a negated verb to form the meaning `not yet', and iamitive \textit{se} can be combined with a negated verb to form the meaning `not any more'.

The word \textit{tok} can mean `still', `first', or `yet'. The former two meanings are illustrated in~(\ref{exe:kal:tokna}) and~(\ref{exe:kal:harustok}) and are used in positive clauses. The use of \textit{tok} with the meaning `yet' is with negated or inherently negative verbs, expressing non-realised expectations, and is called nondum \citep{VeselinovaDevos2021}. Thus, in~(\ref{exe:kal:tokbotnin}), the expectation is that Nyong's father will go, but at the time of speaking he has not left.
 
\ea \label{exe:kal:tok}
\ea \label{exe:kal:tokna}
 	\gll ma \textbf{tok} nawanggar\\
 	\textsc{3sg} still wait\\
 	\glt `He still waits.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD4-C}{[stim29\_1:34]}}
\ex \label{exe:kal:harustok}
 	\gll harus an \textbf{tok} bo=et eciet=et eba\\
	must \textsc{1sg} first go={\glet} return={\glet} then\\
	\glt `I must go first, when [I] return, then...' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA6-6}{[conv13\_11:15]}}
\ex \label{exe:kal:tokbotnin}
	\gll Nyong esun \textbf{tok} bot=\textbf{nin} \\ 
 	Nyong father.\textsc{3poss} yet go=\textsc{neg} \\
 	\glt `Nyong's father doesn't go yet.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA3-3}{[conv10\_7:16]}} 
\z \z

The iamitive particle \textit{se}, often translatable as `already' in positive clauses,\footnote{Iamitives are ``more or less grammaticalised markers that have functions shared by `already' and the perfect'' \citep[][4]{Olsson2013}. Their semantics also include speaker and cultural expectations. For example, in Kalamang (as in Indonesian) one cannot ask the question `Are you married?' or `Do you have children?' without using the iamitive.} can be combined with a negated verb to form the meaning `not any more'. This is illustrated in~(\ref{exe:kal:se}). 
 
\ea \label{exe:kal:se}
\ea \label{exe:kal:inse}
 {\gll in \textbf{se} mia\\
 	\textsc{1pl.excl} {\glse} come\\
 	\glt `We had already come.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BDC-D}{[conv8\_2:34]}}}
 
\ex 	\label{exe:kal:heparuotnin}
 {\gll ma \textbf{se} paruot=\textbf{nin}\\
 	\textsc{3sg} {\glse} do=\textsc{neg}\\
 	\glt `He won't do it any more.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_28:36]}}}
\z \z 
 
Table~\ref{tab:kal:setok} shows the meanings associated with \textit{se} and \textit{tok} in affirmative and negative clauses. As also described in \cite{Baar1997}, `no longer' is rendered by negation of `already' and `not yet' is rendered by negation of `still'. The negative meanings given in Table~\ref{tab:kal:setok} hold for use with any negative verb, whether negated with \textit{=nin}, or whether inherently negative or prohibitive.
 
 \begin{table}
 	\caption{Aspectual particles \textit{se} and \textit{tok} in affirmative and negative clauses.}
 	\label{tab:kal:setok}
 	\begin{tabular}{lll} 
 		\lsptoprule
 		& iamitive \textit{se} & nondum \textit{tok}\\ 
 		\midrule
 		affirmative  &   perfect; already  &  yet; still; first    \\
 		negative  &   no longer &  not yet  \\
 		\lspbottomrule
 	\end{tabular}
 \end{table}

As stated in \sectref{sec:negind}, the meanings `ever' and `never' are underspecified. Sentences with `ever' are translated with iamitive \textit{se}, and those with `never' with nondum \textit{tok} and a verb negated with \textit{=nin} in elicitation. In the natural spoken corpus one may find sentences that express actions or states that have never been realised before, but that nevertheless are expressed with a nondum, making reference to cultural expectations. After marriage, a woman has to be welcomed with a short ritual in every new house she enters. In~(\ref{exe:kal:tokbara}) a woman who has grown up in the Kalamang community and has been married for more than ten years expresses her surprise at the fact that there are still houses she has never officially entered.

\ea \label{exe:kal:tokbara}
\gll ba an me \textbf{tok} wandi mengga barat=\textbf{nin} \textbf{tok} kewe mengga sarat=\textbf{nin}\\
but \textsc{1sg} {\glme} yet like.this \textsc{dist.lat} descend=\textsc{neg} yet house \textsc{dist.lat} ascend=\textsc{neg}\\
\glt `But as for me, I have never gone down there like this, never went up to that house.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_22:14]}}
\z

There is another, minor, strategy to express the meaning `not anymore' than the combination of iamitive \textit{se} and a negated predicate. The use of \textit{koi} `again' with a negated predicate has the same semantics (parallel to Indonesian \textit{tidak lagi} `not again'), as can be seen in \REF{exe:kal:koinin}.

\ea \label{exe:kal:koinin}
\gll pi konenen=nin koi ran mian=nin\\
\textsc{1pl.incl} remember=\textsc{neg} again go come=\textsc{neg}\\
\glt `We don't have to remember to go back and forth anymore.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BCD-C}{[conv3\_3:55]}}
\z 

Give-constructions, with the zero morpheme `give', are negated by attaching \textit{=nin} to the zero morpheme, as~(\ref{exe:kal:kaanin}) illustrates \citep[288]{Visser2022}.

\ea \label{exe:kal:kaanin}
\ea \gll an kon-i ka ∅ some\\
\textsc{1sg} one-\textsc{quant.obj} \textsc{2sg} give \textsc{encrg}\\
\glt `I gave you one, come on!' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_17:12]}}
\ex \gll 	an me ka \textbf{an} ∅=\textbf{nin} o\\
\textsc{1sg} {\glme} \textsc{2sg} \textsc{1sg} give=\textsc{neg} \textsc{emph}\\
\glt `Me, you didn't give me!' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_17:43]}}
\z \z

When the recipient in a give-construction is expressed as a noun instead of as a pronoun, it is followed by benefactive postposition \textit{=ki}.\footnote{The use of causative \textit{di=} is optional both when the recipient is expressed as a noun or as a pronoun.} In a negated clause, the zero morpheme follows the benefactive clitic, and \textit{=nin} is attached to the zero morpheme, as illustrated in~(\ref{exe:kal:kinin}).

\ea	\label{exe:kal:kinin}
\ea \gll guru dodon=at di=tumtum=ki ∅\\
	teacher thing=\textsc{obj} \textsc{caus}=child=\textsc{ben} give\\
	\glt 	`The teacher gives the children the gift.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_giv]}}
\ex \gll guru 		tok 		dodon=at 	di=tumtum=ki ∅=\textbf{nin}\\
	teacher yet thing=\textsc{obj} \textsc{caus}=child=\textsc{ben} give=\textbf{\textsc{neg}}\\
	\glt 	`The teacher hasn't given the children the gift yet.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_giv]}}
\z \z

Standard negation is also used to negate certain non-verbal predicates, as discussed in \sectref{sec:negnonv}.                                            

\subsection{Negation in non-declaratives}
\label{sec:negnondecl}
Polar questions are formed by adding \textit{ye ge}, literally `or not', at the end of an affirmative clause, as in~(\ref{exe:kal:yege1}). Leaving \textit{ye ge} out but keeping question intonation is possible, but this is pragmatically weird, because \textit{ye ge} is neutral in the sense that there is no expectation of either a positive or negative answer. (It is possible to only leave out \textit{ye} `or' and keep \textit{ge} `not', but this results in a tag question, as illustrated in example~\ref{exe:kal:tumun} in \sectref{sec:negrep}).

\ea \label{exe:kal:yege1}
\gll mu kat gonggung ter-nan \textbf{ye} \textbf{ge}\\
\textsc{3pl} \textsc{2sg.obj} call tea-drink or not\\
\glt `Are they calling you for tea?' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_20:16]}}
\z 	

If one wants to express a negative expectation to an answer, one may use predicate negator \textit{=nin} as in \REF{exe:kal:natnin}. Note that there are no occurrences of this type in the natural spoken corpus.

\ea\label{exe:kal:natnin}{
	\gll ma pasa=at nat=\textbf{nin}
	\\
	\textsc{3sg} rice=\textsc{obj} eat=\textsc{neg}\\
	\glt `Doesn't he eat rice?' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg]}}
}
\z 

Prohibitive constructions are marked by the combination of a pronominal suffix \textit{-mun} and a predicate clitic \textit{=in}. The combination of \textit{-mun} and predicate negator \textit{=nin} is ungrammatical. Example (\ref{exe:kal:proh}) contrasts an imperative and a prohibitive clause.

\ea \label{exe:kal:proh}
\ea \label{exe:kal:eware}
{\gll ka ewa=te\\
	\textsc{2sg} speak=\textsc{imp}\\
	\glt `You speak!' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C75-D}{[stim15\_0:41]}}}

\ex \label{exe:kal:ewain}	{\gll ka-\textbf{mun} se reidak-i ewa=\textbf{in}\\
	\textsc{2sg-proh} {\glse} much-\textsc{quant.obj} speak=\textsc{proh}\\
	\glt 	`Don't you speak much anymore!' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_21:00]}}
}
\z \z 

Other persons than second person singular may be the subject of prohibition, such as `smoke' in~(\ref{exe:kal:sarein}). Because \textit{-mun} cannot be attached to nouns, the noun is proceeded by a pronoun.

\ea \label{exe:kal:sarein}
{\gll ma-\textbf{mun} dugar sara=\textbf{in}\\
	\textsc{3sg-proh} smoke go.up=\textsc{proh}\\
	\glt 	`The smoke must not rise!' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_8:01]}}
}
\z 

As illustrated in~(\ref{exe:kal:sarein}), \textit{-mun} cannot be suffixed to nouns. To deal with this, when a speaker wishes to use someone's title or name, the pronoun with \textit{-mun} follows the noun, as exemplified in~(\ref{exe:kal:esaproh}).

\ea 	\label{exe:kal:esaproh}
\ea [*]
{\gll \textbf{esa-mun} sara=in\\
	father-\textsc{proh} go.up=\textsc{proh}\\
	\glt `Father, don't go up!' 
}
\ex []
{\gll \textbf{esa} ka-\textbf{mun} sara=in\\
	father \textsc{2sg-proh} go.up=\textsc{proh}\\
	\glt `Father, don't (you) go up!' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_proh]}}
}
\z \z 


When the pronoun is elided, \textit{-mun} is elided as well, such as in~(\ref{exe:kal:tin}).


\ea \label{exe:kal:tin}
{\gll bo kuet=te tik=\textbf{in}\\
	go bring={\glte} take.long=\textsc{proh}\\
	\glt `Don't [you] bring it for a long time!' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BDD-5}{[narr42\_34:19]}}
}
\z 

All mood, aspect and modality marking is incompatible is incompatible with the prohibitive, except for the iamitive and nondum.
Example~(\ref{exe:kal:ewain}) above illustrates the combination of the iamitive \textit{se} and the prohibitive, which results in the meaning `not anymore'. The nondum \textit{tok} plus a prohibitive results in the meaning `not yet', illustrated in~(\ref{exe:kal:muntok}).

\ea 	\label{exe:kal:muntok}
{\gll ki-\textbf{mun} tok na=\textbf{in}\\
	\textsc{2pl-proh} yet eat=\textsc{proh}\\
	\glt `Don't you eat yet!' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA2-F}{[conv11\_3:41]}}}
\z 


In multi-verb constructions, it is the last verb that carries prohibitive \textit{-in}. It has scope over the whole construction. Example (\ref{exe:kal:ewainmelelu}) is one clause. 

\ea \label{exe:kal:ewainmelelu}
\gll ka-\textbf{mun} melelu ewa=\textbf{in}\\
\textsc{2sg-proh} sit speak=\textsc{proh}\\
\glt `Don't you sit speaking!' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg19]}}
\z 

One may combine a prohibitive and an imperative in one message as follows in~(\ref{exe:kal:melelure}), which is biclausal as indicated by square brackets.

\ea \label{exe:kal:melelure}
\gll [ka-\textbf{mun} ewa=\textbf{in}] [melelu=te]\\
\textsc{2sg-proh} speak=\textsc{proh} sit=\textsc{imp}\\
\glt `Don't you speak, sit!' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg19]}}
\z 

General prohibition, not directed at a specific person, is expressed with \textit{ka-mun} or \textit{ki-mun}, second person singular or plural, respectively. One could use this for example on a prohibition sign. In spoken Kalamang, one may also leave out the verb, and just say \textit{kamun!} `don't!' (for example when warning a child against doing something naughty).

It is unclear what the origin of the pronominal suffix \textit{-mun} is. Perhaps there is a relation with the verb \textit{mamun} `leave' (as in `leave be; leave in place'). \citet{KutevaEtAl2019} give `leave' as a source for imperatives and negation, but not for prohibitives. It does not seem strange, in any case, that a language would develop a prohibitive marker on the pronoun, seen that prohibition is both urgent and person-oriented. By signalling that one wants to convey a prohibition on the first constituent of the clause, a pronoun, it is immediately clear for the addressee that action (or rather: no action) must be undertaken.

The Kalamang corpus does not contain any examples of ``soft prohibitives'' akin to Indonesian \textit{tak usah} `\textsc{neg} need'.



\subsection{Negation in stative predications}
\label{sec:negnonv} 
Kalamang clauses may have a noun phrase, demonstrative or quantifier as predicate, or a noun inflected with the locative or similative postposition. Locative and similative predicates are negated with predicate negator \textit{=nin}, as illustrated in~(\ref{exe:kal:negloc}), (\ref{exe:kal:wisenggap}) and~(\ref{exe:kal:inggonggonin}). This also counts for adverbial demonstratives, such as the lative demonstrative \textit{mengga} `to/from there' in~(\ref{exe:kal:mengganin}) or the manner adverb \textit{mindi} `like that' in~(\ref{exe:kal:mindinin}).

\ea \label{exe:kal:negloc}
\ea \gll ma tok kit=ko\\
\textsc{3sg} still top=\textsc{loc}\\
\glt `He is still on top.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B94-F}{[stim21\_3:08]}}
\ex \gll ma ewun naun-kit=ko=\textbf{nin}\\
\textsc{3sg} root.\textsc{3poss} soil-top=\textsc{loc}=\textsc{neg}\\
\glt	`Its root is not on top of the soil.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB1-3}{[narr34\_4:58]}}
\z \z 

\ea \label{exe:kal:wisenggap}
\ea \gll ka kalamang emumur=kap to\\
		\textsc{2sg} Kalamang woman.\textsc{pl}=\textsc{simv} right\\
		\glt `You are like the Kalamang women, aren't you?' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_6:51]}}
\ex \gll wise=kap=\textbf{nin}\\
long.time.ago=\textsc{simv=neg}\\
\glt `It's not like it used to be.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_32:32]}}
\z \z

\ea \label{exe:kal:inggonggonin}
\ea \gll ma se mahara Mustafa mu=konggo\\
\textsc{3sg} {\glse} ascend Mustafa \textsc{3pl=an.loc}\\
\glt `He went up to Mustafa's.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA3-3}{[conv10\_3:20]}}
\ex \gll pengalaman kon yume sampi in=konggo=\textbf{nin}\\
experience one \textsc{dist} until \textsc{1pl.excl=an.loc=neg}\\
\glt `That experience hasn't reached us yet.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C98-7}{[narr26\_15:54]}}
\z \z

\ea \label{exe:kal:mengganin}
\ea \gll an mengga\\
\textsc{1sg} \textsc{dist.lat}\\
\glt `I went there.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_24:01]}}
\ex \gll tete Manggan mu tok mengga=\textbf{nin} weinun\\
grandfather Manggan \textsc{3pl} yet \textsc{dist.lat=neg} too\\ 
\glt `[I] haven't yet been to grandfather Manggan's either.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_24:27]}}
\z \z

\ea \label{exe:kal:mindinin}
\ea \gll in-ninggan mindi\\
\textsc{1pl.excl}-all like.that\\
\glt `All of us [were invited] like that.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD8-4}{[narr1\_0:47]}}
\ex \gll mindi=\textbf{nin}\\
like.that=\textsc{neg}\\
\glt `It's not like that.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC3-B}{[conv7\_10:11]}}
\z  \z

In equational clauses such as (\ref{exe:kal:emage}) and (\ref{exe:kal:guruge}), where the predicate is a noun, \textit{ge} `not' is used because the whole proposition is negated. \textit{Ge} `not' is clause-final.

\ea 	\label{exe:kal:emage}
\ea \gll ma me ema anggon\\
\textsc{3sg} {\glme} mother \textsc{1sg.poss}\\
\glt `She is my mother.'  {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg]}}
\ex \gll me me ema anggon \textbf{ge}\\
\textsc{dist} {\glme} mother \textsc{1sg.poss} not\\
\glt `That is not my mother.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg]}}
\z \z 

\ea 	\label{exe:kal:guruge}
\ea \gll kon se guru tumtum kon guru\\
	one {\glse} teacher children one teacher\\
\glt `One is already teacher, one child [one of my children] is teacher.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_9:15]}}
\ex \label{exe:kal:gurugeneg} \gll dun-an me me guru \textbf{ge}\\
cross.sex.sibling-\textsc{1sg.poss} \textsc{dist} {\glme} teacher not\\
\glt `My sibling is not a teacher.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg]}}
\z \z

Clauses with proper inclusion can alternatively be negated with predicate negator \textit{=nin} as in~(\ref{exe:kal:gurunin}). It is unclear how the meaning of this construction differs from that in~(\ref{exe:kal:gurugeneg}).

\ea 	\label{exe:kal:gurunin}
\gll dun-an me me guru=\textbf{nin}\\
cross.sex.sibling-\textsc{1sg.poss} \textsc{dist} {\glme} teacher=\textsc{neg}\\
\glt `My sibling is not a teacher.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg]}}
\z 

Quantifiers are not negated. 
Finally, existential and possessive predication clauses are negated with negative existential \textit{saerak}, which also comes clause-finally. Examples~(\ref{exe:kal:sorsaerak}) and~(\ref{exe:kal:saerak}) show existential negation. \textit{Saerak} is the opposite of existential \textit{mambon}, as shown in~(\ref{exe:kal:mamsae}), which together with~(\ref{exe:kal:tangunme}) illustrates the negation of possessive clauses.

\ea	\label{exe:kal:sorsaerak}
\begin{xlist}
\exi{A:} 	{
	\gll ba sor mambon\\
	but fish \textsc{ex}\\
	\glt	`But is there fish?'
}

\exi{B:}{
    \gll sor \textbf{saerak}\\
    fish \textsc{neg.ex}\\
    \glt `There is no fish.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC9-2}{[conv5\_6:49]}}
}
\end{xlist}
\z
\ea \label{exe:kal:saerak}
\gll karena pak \textbf{saerak} karek=ki=a kanie\\
because nail \textsc{neg.ex} string=\textsc{ins=foc} tie\\
\glt	`Because there were no nails [we] tied with string.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB3-0}{[narr7\_1:03]}}
\z 
\ea \label{exe:kal:mamsae}
\gll utkon kadok-un \textbf{mambon} utkon muin \textbf{saerak}\\
some cloth-\textsc{3poss} \textsc{ex} some \textsc{3pl.poss} \textsc{neg.ex}\\
\glt	`Some had their cloths, some didn't have theirs.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_19:28]}}
\z 
\ea \label{exe:kal:tangunme}
\gll tang-un me hal<al>us=ten me \textbf{saerak}\\
seed-\textsc{3poss} {\glme} fine<\textsc{ints}>=\textsc{attr} {\glme} \textsc{neg.ex}\\
\glt	`It doesn't have fine seeds.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA6-6}{[conv13\_5:55]}}
\z 

It is ungrammatical to combine either \textit{mambon} or \textit{saerak} with predicate negator \textit{=nin}. \textit{Saerak} is more common than \textit{mambon}. For negated possessive clauses, \textit{saerak} is obligatory, whereas for affirmative possessive clauses \textit{mambon} only stresses the presence of a possessed item, but is not necessary to express possession. Compare \citet{Veselinova2013neg.exist}, who finds that negative existentials state non-existence or absolute absence rather than negate positive existence. It is unclear what the origin of \textit{saerak} is, and whether it is diachronically composed of several morphemes. 


\subsection{Negation in non-main clauses}
\label{sec:nonmainneg}
Negative conditional clauses are introduced by \textit{get me} `if not', which is likely derived from \textit{ge} `no' and irrealis clitic \textit{=et}. \textit{Get me} `if not' is illustrated in~(\ref{exe:kal:getme}).

\ea	\label{exe:kal:getme}
\gll kacamata-an rat=et eba kawas=at kosiaur=et sin=ka \textbf{get} \textbf{me} eranun\\
glasses-\textsc{1sg.poss} move.path={\glet} then yarn=\textsc{obj}  put.in={\glet} needle=\textsc{lat} not {\glme} cannot\\
\glt `If I put on my glasses, then [I can] put the thread through the needle, if not, [I] cannot [do it].' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA6-6}{[conv13\_4:50]}}
\z

Kalamang has only one other type of non-main clause: complement clauses with direct or reported speech. There is no special strategy for negating non-main clauses. This is illustrated with complement clauses introduced by \textit{toni} `say' and \textit{kona} `think' in~(\ref{exe:kal:matoni}) and~(\ref{exe:kal:ankona}), respectively.

\ea \label{exe:kal:matoni}
\ea \gll ma toni tumun-an boda\\
\textsc{3sg} say child-\textsc{1sg.poss} stupid\\
\glt `She says my child is stupid.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg\_35]}}
\ex \gll ma toni tumun-an boda=\textbf{nin}\\
\textsc{3sg} say child-\textsc{1sg.poss} stupid\\
\glt `She says my child is not stupid.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg\_36]}}
\z \z


\ea \label{exe:kal:ankona}
\ea \gll an kona ma-nan bara ming-jie\\
\textsc{1sg} think \textsc{3sg}-too descend oil-buy\\
\glt `I though he went down to buy oil.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC3-B}{[conv7\_5:23]}}
\ex \gll an kona in barat=\textbf{nin}\\
\textsc{1sg} think \textsc{1pl.excl} go.down=\textsc{neg}\\
\glt `I thought we didn't go down.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA4-1}{[conv16\_28:32]}}
\z \z



\subsection{Negative lexicalizations}\label{sec:lexicalneg}
Apart from negative existential \textit{saerak} (the opposite of existential \textit{mambon}), discussed in \sectref{sec:negnonv}, there are three inherently negative verbs: `not want; not agree', `cannot' and `not know'. They are listed with their closest positive counterparts in (\ref{ex:kal:36}--\ref{ex:kal:38}).

\TabPositions{.33\textwidth,.66\textwidth}
\ea\label{ex:kal:36}
	\ea \textit{lo} \tab `to want; to agree' \\
	\ex \textit{=kin}  \tab volitional verbal clitic \\
	\ex \textit{suka}-\textsc{poss} \textit{ge}  \tab `to not want; to not agree' {\href{https://doi.org/10.5281/zenodo.4297474}{[dictionary]}}
\z \z
\ea 
	\ea \textit{bisa}  \tab `can' \\
	\ex \textit{eranun}  \tab `cannot; to not be possible' {\href{https://doi.org/10.5281/zenodo.4297474}{[dictionary]}}
\z \z 
\ea\label{ex:kal:38}
	\ea \textit{gonggin}  \tab `to know' \\
	\ex \textit{komahal}  \tab `to not know' {\href{https://doi.org/10.5281/zenodo.4297474}{[dictionary]}}
\z \z


\textit{Suka}-\textsc{poss} \textit{ge} is a dedicated negative construction that means `not want; not agree'. The construction is made with the borrowed positive verb \textit{suka},\footnote{Compare Indonesian \textit{suka} `to like', also used as habitual marker in colloquial varieties of Indonesian \citep{ConnersBrugman2014,Jeoung2020}. \textit{Suka} is not used with either of those meanings in Kalamang} nominalised by inflection with a possessive suffix and followed by \textit{ge} `no; not'. The standard possessive suffix is third person \textit{-un}, which can be used for all persons. A speaker can, however, choose to inflect \textit{suka} with another person. An example with a third person subject \textit{lek} `goat' is given in~(\ref{exe:kal:lek}). Inflection for all persons are given in Table~\ref{tab:kal:sukaungge}.

\ea 	\label{exe:kal:lek}
\gll canam opa me mat narorar ba lek \textbf{suka-un} \textbf{ge}\\
man {\glopa} {\glme} \textsc{3sg.obj} lead but goat want-\textsc{3poss} no\\
\glt `That man is leading it, but the goat doesn't want.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BD1-D}{[stim31\_0:51]}}
\z

\begin{table}
	\caption{Inflections of \textit{suka}-\textsc{poss} \textit{ge} `to not want; to not agree'.}
	\label{tab:kal:sukaungge}
	\begin{tabular}{ll} 
		\lsptoprule
			\textsc{1sg} & \textit{an suka-n-an ge}\\
			\textsc{2sg}& \textit{ka suka-ca ge}\\
			\textsc{3sg}& \textit{ma suka-un ge}\\
			\textsc{1pl.incl}& \textit{pi suka-pe ge}\\
			\textsc{1pl.excl}& \textit{in suka-un ge}\\
			\textsc{2pl} &\textit{ki suka-ce ge}\\
			\textsc{3pl}& \textit{mu suka-un ge}\\
		\lspbottomrule
	\end{tabular}
\end{table}

It is perhaps best to paraphrase \textit{suka}-\textsc{poss} \textit{ge} as `is not of \textsc{subject}'s liking'. As for the lexical opposite of `not want; not agree', there is a verb \textit{lo} `want', but it is rather uncommon. An example is given in~(\ref{exe:kal:lo}). \textit{Lo} cannot be negated with \textit{=nin}. Wishes and desires are more commonly expressed with volitional \textit{=kin}, see~(\ref{exe:kal:natkin}). As illustrated in \sectref{sec:standneg}, \textit{=kin} and predicate negator \textit{=nin} are incompatible with each other. \textit{Suka}-\textsc{poss} can also be used in combination with predicate negated with \textit{=nin}, as in~(\ref{exe:kal:parinin}). In that case, propositional negator \textit{ge} is not needed.


\ea 
\ea \label{exe:kal:lo}
\gll an bo mu et-un=at paning paning=i koyet mu se \textbf{lo}\\
\textsc{1sg} go \textsc{3sg} canoe-\textsc{3poss-obj} ask.for ask.for={\gli} finished \textsc{3pl} {\glse} want\\
\glt `I went to ask for their canoe, after asking, they consented.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C99-E}{[narr8\_0:12]}}
\ex \label{exe:kal:natkin}
\gll bal napikir an sor=at nat=\textbf{kin}\\
dog think \textsc{1sg} fish-\textsc{obj} eat={\glkin}\\
\glt `The dog thinks: ``I want to eat the fish.''' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB9-6}{[stim1\_0:34]}}
\ex \label{exe:kal:parinin}
\gll sontum haus ba an \textbf{suka-n-an} parin=\textbf{nin}\\
people thirsty but \textsc{1sg} want-{\glnn}-\textsc{1sg.poss} sell=\textsc{neg}\\
\glt `People are thirsty (keen to buy) but I don't want to sell.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BDD-5}{[narr42\_34:03]}}
\z \z 
	
In the pair \textit{bisa} `can'/\textit{eranun} `cannot' we are again dealing with an Indonesian loan, but this time for the positive form: \textit{bisa} also means `can' in Indonesian. But whereas \textit{bisa} in Indonesian is negated with standard negation marker \textit{tidak} (\textit{tidak bisa} `cannot'), Kalamang has a dedicated form \textit{eranun}. The use of \textit{eranun} triggers nominalisation of the negated proposition with \textit{-un}. \textit{Eranun} is clause-final. They are illustrated in ~(\ref{exe:kal:bisera}).

\ea \label{exe:kal:bisera}
\ea \label{exe:kal:bisa}
\gll ka marua hukat yume \textbf{bisa}\\
\textsc{2sg} towards.sea fishing.net \textsc{dist} can\\
\glt `Can you go seawards with the fishing net?' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BCB-5}{[conv1\_2:56]}}
\ex	\label{exe:kal:ningkabor}
\gll ning-kabor pasiengga bor-un \textbf{eranun}\\
sick-stomach to.beach go-\textsc{3poss} cannot\\
\glt 	`[If you have] stomachache [you] cannot go to the toilet.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BCA-4}{[conv20\_15:58]}}
\z \z

An alternative translation of the construction \textit{V-un eranun} might be `V-ing isn't possible', such that~(\ref{exe:kal:ningkabor}) translates as `toilet-going isn't possible.'

The pair \textit{gonggin} `to know'/\textit{komahal} `to not know' is illustrated in~(\ref{exe:kal:gongkom}). Note also \textit{yeso} `I don't know' in~(\ref{exe:kal:gonggin}). This is a clause-initial interjection uttered at high pitch which expresses indignation, and not a verb like \textit{komahal}. It can only be used with reference to a first person subject.

\ea \label{exe:kal:gongkom}
\ea \label{exe:kal:gonggin}
\gll yeso ka-tain=a \textbf{gonggin}\\
I.don't.know \textsc{2sg}-alone=\textsc{foc} know\\
\glt `I don't know, only you know.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_18:43]}}
\ex \label{exe:kal:komahal}
\gll ema tamatko eh ema in-nan \textbf{komahal}\\
mother where \textsc{inj} mother \textsc{1pl.excl}-too not.know\\
\glt `Where's mother?' `Eh, we don't know either.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_5:19]}}
\z \z

Negation of any of the verbs discussed in this section, whether they are positive or negative forms, with predicate negator \textit{=nin} is ungrammatical. The negative verbs and constructions in this section behave like other negated verbs with aspectual particles iamitive \textit{se} and nondum \textit{tok}, giving them the meanings `not anymore' and `(not) yet', as illustrated in~(\ref{exe:kal:tokkom}).

\ea \label{exe:kal:tokkom}
	\gll an tok komahal\\
	\textsc{1sg} yet not.know\\
	\glt `I don't know yet.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C97-F}{[stim27\_7:32]}}
\z

\section{Non-clausal negation}
\label{sec:nonclneg}
In this section I discuss constructions expressing negation other than clausal negation, starting with negative replies in \sectref{sec:negrep}. Negative indefinites and quantifiers are dealt with in \sectref{sec:negind} and negative derivation is treated in \sectref{sec:negder}.

\subsection{Negative replies}
\label{sec:negrep}
The negative reply is \textit{ge}, which is the same as the propositional negator, and serves to negate the preceding proposition. The affirmative reply \textit{ya} or \textit{yo} is less frequent, perhaps because there are more strategies to agree with someone or with a preposition, such as \textit{haidak} `true' and \textit{yor} `true'. A piece of discourse with a negative reply with (a repeated) \textit{ge} is given in~(\ref{exe:kal:gedia}).

\ea	\label{exe:kal:gedia}
\begin{xlist}
\exi{A:} 	{\label{exe:kal:antonika}
	\gll an toni ka ∅ to {\ob}...{\cb}\\
	\textsc{1sg} think \textsc{2sg} give right {\ob}...{\cb}\\
	\glt	`I thought I gave [it] to you, right {\ob}...{\cb}?'
}

\exi{B:}{\label{exe:kal:gege}
	\gll \textbf{ge} \textbf{ge}\\
	no no\\
	\glt `No, no.'
}
\end{xlist}

\begin{xlist}
\exi{A:}{\label{exe:kal:tamandiage}
	\gll tamandi=a \textbf{ge}\\
	how=\textsc{foc} no\\
	\glt `What do you mean ``no''?'
}

\exi{B:}{\label{exe:kal:kadianin}
	\gll ka di=an ∅=nin\\
	\textsc{2sg} \textsc{caus=1sg} give=\textsc{neg}\\
	\glt `You didn't give [it] to me.'{\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_17:05]}}
}
\end{xlist}
\z   

An affirmative example from the same text is~(\ref{exe:kal:yo}).

\ea	\label{exe:kal:yo}
\begin{xlist}
\exi{A:}{
	\gll manadu naun tem∼temun eh\\
	taro fruit \textsc{ints}∼big \textsc{tag}\\
	\glt	`A big taro, eh?'
}

\exi{B:}{
	\gll \textbf{yo}\\
	yes\\
	\glt `Yes.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_0:23]}}
}
\end{xlist}
\z 

\textit{Ge} seems to disagree with the content of the question, not with its polarity. That is, whether or not there is negation in the question seems to be irrelevant. Consider the four questions in~(\ref{exe:kal:tumun}). For all of them, if they are answered with \textit{ge}, it means that the child is not crying. As there are no examples of negated questions answered with \textit{ge} in the corpus, this had to be elicited.

\ea \label{exe:kal:tumun}
\begin{xlist}
    \ex\label{exe:kal:tumun:a}
        \gll tumun wa me ecuan\\
        child \textsc{prox} {\glme} cry\\
        \glt `Is this child crying?'
    \ex\label{exe:kal:tumun:b}
        \gll tumun wa me ecuat=nin\\
        child \textsc{prox} {\glme} cry=\textsc{neg}\\
        \glt `Is this child not crying? / Isn't this child crying?' 
    \ex\label{exe:kal:tumun:c}
        \gll tumun wa me ecuan, ge\\
        child \textsc{prox} {\glme} cry no\\
        \glt `This child is crying, isn't it?'
    \ex\label{exe:kal:tumun:d}
        \gll tumun wa me ecuan ye ge\\
        child \textsc{prox} {\glme} cry or not\\
        \glt `Is this child crying or not?' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg]}}
\end{xlist}
\z

If these are answered with an affirmative reply, it means the child is crying. For example \REF{exe:kal:tumun:a}, the affirmative answer may be just \textit{ya/yo} `yes', \textit{yor} `true' or \textit{haidak} `true'. For examples (\ref{exe:kal:tumun:b}--\ref{exe:kal:tumun:d}), seen that they already contain a negation, it is more natural for the answer to (also) contain a repetition of the clause, as in~(\ref{exe:kal:yege}).

\ea	\label{exe:kal:yege}
\begin{xlist}
\exi{A:}{
	\gll yua me pulisi=a wa ye ge\\
	\textsc{prox} {\glme} police=\textsc{foc} \textsc{prox} or not\\
	\glt	`This one, is this the police or not?'
}

\exi{B:}{
	\gll pulisi=a wa\\
	police=\textsc{foc} \textsc{prox}\\
	\glt `This is the police.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA9-9}{[stim6\_11:24]}}
}
\end{xlist}
\z 

There are two common answers involving \textit{ge} to the questions \textit{nebara paruo} `what are you doing', \textit{tamanggara bot} and `where are you going', which are used as greetings. They are illustrated in~\REF{exe:kal:geo}.

\ea	\label{exe:kal:geo}
\begin{xlist}
\exi{A:}
{\gll nebara paruo / tamanggara bot\\
	what.\textsc{obj.foc} do / where.\textsc{lat} go\\ 
	\glt 	`What are [you] doing?' / `Where are you going?'
}
\exi{B:}
{\gll \textbf{ge} o / \textbf{ge} mera\\
	no \textsc{emph} / no \textsc{inj}\\
	\glt `Oh, nothing/nowhere.'  [overheard]}
\end{xlist}
\z

Both \textit{ge o} and \textit{ge mera} mean `nothing' or `nowhere' and downplay the importance of the speaker's activity or destination.\footnote{\textit{Mera} is an interjection expressing downplay, obviousness or encouragement. In combination with \textit{ge} `no' it always expresses downplay.} It is wrong to answer just *\textit{ge} `no' in this context. By using \textit{ge o}, the speaker affirms that he or she is doing nothing of importance, or nothing that he or she wants to elaborate on. (\ref{exe:kal:gemera}) is another illustration of \textit{ge mera}, used in a conversation acted out by one speaker. \textit{Ge mera} can also be used as an answer to other questions, as illustrated in~(\ref{exe:kal:ramie}), where it functions as an ironic or denigrating comment on the yield of the speaker's husband. 

\ea 	\label{exe:kal:gemera}
\glll ``Ka tamanggara bot?'' {``Ge mera!''} ``Ka nebara paruo?'' ``{Ge mera}, in pasakajie reba!''\\
ka tamangga=ta bot \textbf{ge.mera} ka neba=at=a paruo \textbf{ge.mera} in pasa-kajie=teba\\
\textsc{2sg} to.where={\glta} go nothing \textsc{2sg} what=\textsc{obj=foc} do nothing \textsc{1pl.excl} rice-pick=\glteba\\
\glt	`{``}Where are you going?'' ``Nowhere!'' ``What are you doing?'' ``Oh, nothing, just picking rice!''' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA6-6}{[conv13\_10:44]}}
\z

\ea	\label{exe:kal:ramie}
\begin{xlist}
\exi{A:} 	{\label{exe:kal:wis}
	\gll wis ma ramie\\
	yesterday \textsc{3sg} pull\\
	\glt	`Did he catch any fish yesterday?'
}

\exi{B:}{\label{exe:kal:tagierun}
	\gll \textbf{ge.mera} tagier-un et-kon\\
	nothing mackerel-\textsc{3poss} \textsc{anml.clf}-one\\
	\glt `Nothing, he had one mackerel.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BA3-3}{[conv10\_8:24]}}
}
\end{xlist}
\z

\textit{Tok} `still; yet; first' (discussed in \sectref{sec:standneg}) can also be an interjection meaning either `yes (still)' or `no (not yet)', depending on whether the question contained a negative or not. This contrast is illustrated in~(\ref{exe:kal:stillschool}) and~(\ref{exe:kal:notschool}).

\ea \label{exe:kal:stillschool}
\begin{xlist}
	\exi{A:}
		{\gll ka tok sekola\\
			\textsc{2sg} still go.to.school\\
			\glt `Do you still go to school?' }
	\exi{B:}
		{\gll tok\\
			still\\
			\glt `Yes [I still go to school].' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_wc19]}}}
\end{xlist}		
\z 
		
\ea	\label{exe:kal:notschool}	
\begin{xlist}
	\exi{A:}
		{\gll ka tok sekola=nin\\
			\textsc{2sg} yet go.to.school=\textsc{neg}\\
			\glt `Don't you go to school yet?' }
	\exi{B:}
		{\gll tok\\
			not.yet\\
		\glt `Not yet.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_wc19]}}}
\end{xlist}
\z

\subsection{Negative indefinites and quantifiers}
\label{sec:negind}
Negative indefinites, meanings related to English nobody, no-one, nowhere, never, none and no \citep{Haspelmath1997}, are expressed with a number of strategies. Most strategies are the same as positive indefinites in combination with a negated verb. Some indefinite strategies seem to be unique to negatives. These are the constructions with \textit{-barak} `any', which must be combined with a negated verb, and constructions with \textit{don konkon V=nin} `not any(thing)'. Table~\ref{tab:kal:negindef} summarises the constructions and their translational equivalents discussed in this section. Sometimes there are two translational equivalents for one construction. For each construction is indicated whether it is found in the natural spoken corpus or not. It is also indicated whether the construction is parallel to the positive indefinite (except for the negator, of course).


\begin{table} 
	\caption{Indefinite constructions. Q = question word, V = verb, N = noun (phrase)}
	\label{tab:kal:negindef}
	\footnotesize 
	\begin{tabular}{p{45mm}lll}
		\lsptoprule 
		construction and gloss & translational equivalent & corpus? & same as\\
		&&& positive?\\
		\midrule
        \textit{biar} Q=\textit{taet-o} V=\textit{nin} & \multirow{2}{*}{not anybody, not anywhere} & \multirow{2}{*}{no} & \multirow{2}{*}{yes} \\
		although Q-again=\textsc{cond} V=\textsc{nin} &&& \\
\tablevspace
		N-\textit{barak} V=\textit{nin} & \multirow{2}{*}{no-one, not any} & no-one: no, & \multirow{2}{*}{no}\\
		N-any V-\textsc{neg}&&not any: yes&\\
\tablevspace
		Q-Q V=\textit{nin} & \multirow{2}{*}{not anywhere} & \multirow{2}{*}{no} & \multirow{2}{*}{yes}\\
		Q∼\textsc{red} V-\textsc{neg}&&&\\
\tablevspace
		\textit{don konkon} V=\textit{nin} & \multirow{2}{*}{not any(thing)} & \multirow{2}{*}{yes} & \multirow{2}{*}{no}\\
		thing one∼\textsc{red} V-\textsc{neg}&&&\\
\tablevspace
		V=\textit{nin=i koyet} & \multirow{2}{*}{not any} & \multirow{2}{*}{no} & \multirow{2}{*}{no}\\
		 V-\textsc{neg}={\gli} finished&&&\\
		\lspbottomrule	
	\end{tabular}
\end{table}

% \begin{table} 
% 	\caption{Indefinite constructions. Q = question word, V = verb, N = noun (phrase)}
% 	\label{tab:kal:negindef}
% 	\footnotesize 
% 	\begin{tabular}{p{45mm}lp{16mm}l}
% 		\lsptoprule 
% 		construction and gloss & translational equivalent & corpus? & same as\\
% 		&&& positive?\\
% 		\midrule
%     \vspace{-5mm}{\begin{xlist} \exi{} \hspace{-5mm}\gll \textit{biar} Q=\textit{taet-o} V=\textit{nin} \\
% 		  although Q-again=\textsc{cond} V=\textsc{nin} \\
%     \end{xlist} }       & {not anybody, not anywhere} & {no} & {yes} \\
		
% 	\vspace{-5mm}{\begin{xlist} \exi{} \hspace{-5mm}\gll N-\textit{barak} V=\textit{nin} \\
% 		  N-any V-\textsc{neg} \\
%     \end{xlist} }\vspace{-3mm}       & {no-one, not any} & no-one: no, not any: yes& {no}\\
		
% 	\vspace{-5mm}{\begin{xlist} \exi{} \hspace{-5mm}\gll Q-Q V=\textit{nin} \\
% 		  Q∼\textsc{red} V-\textsc{neg}\\
%     \end{xlist} }       & {not anywhere} & {no} & {yes}\\
		
% 	\vspace{-5mm}{\begin{xlist} \exi{} \hspace{-5mm}\gll \textit{don konkon} V=\textit{nin} \\
% 		 thing one∼\textsc{red} V-\textsc{neg}\\
%     \end{xlist} }       & {not any(thing)} & {yes} & {no}\\
				
% 	\vspace{-5mm}{\begin{xlist} \exi{} \hspace{-5mm}\gll V=\textit{nin=i koyet} \\
% 		 V-\textsc{neg}={\gli} finished \\
%     \end{xlist} }       & {not any} & {no} & {no}\\
% 		\lspbottomrule	
% 	\end{tabular}
% \end{table}




Constructions with \textit{naman=taet=o} `anybody' or \textit{tamatko=taet=o} `anywhere' do not occur in the corpus, and elicited examples are complicated strategies rather than set constructions, relying partly on Indonesian loans, suggesting that these indefinites are not well-established. Consider~(\ref{exe:kal:anybody}) and~(\ref{exe:kal:anywhere1}), which make use of the Indonesian loan \textit{biar} `although',\footnote{In Indonesian, but not in Kalamang, \textit{biar} can also mean `let'.} \textit{-taet} `more; again' and conditional \textit{-o} on a question word, and relativiser \textit{-ten} and \textit{-saet} `only; exclusively; all' on a verb. In the negated versions of these sentences, two strategies are adopted. In~(\ref{exe:kal:nobody}), a suffix \textit{-barak} `any' is added to the NP, followed by a negative verb. In~(\ref{exe:kal:nowhere1}), the same strategy as in the positive examples is applied, at least with \textit{biar} and \textit{-taet} and \textit{-o} on the question word, followed by a negative verb.

\ea \label{exe:kal:anybnob}
\ea \label{exe:kal:anybody}{
	\gll biar naman=taet=o bisa paruon pi=nininggan bisa-ten=saet\\
	although who-again=\textsc{cond} can do \textsc{1pl.incl}-all can-\textsc{attr}=only\\
	\glt `Anyone can do it, we all can do it.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_ind]}}}

\ex \label{exe:kal:nobody}{
	\gll som kon-\textbf{barak} karajang-un \textbf{eranun}\\
	person one-any work-\textsc{nmlz} cannot\\
	\glt `No-one can do the work.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_ind]}}}
\z \z 

\ea \label{exe:kal:anywhnowh}
\ea \label{exe:kal:anywhere1}{
	\gll sor kerkap-ten wa me tamatko-taet=o mambon-ten=saet\\
	fish red-\textsc{attr} \textsc{prox} {\glme} where-again=\textsc{cond} \textsc{ex}-\textsc{attr}=only\\
	\glt `There's red fish everywhere.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_ind]}}}

\ex \label{exe:kal:nowhere1}{
	\gll lawan wa me \textbf{biar} tamatko-\textbf{taet}=\textbf{o} \textbf{saerak}\\
	grouper \textsc{prox} {\glme} although? where-again=\textsc{cond} \textsc{neg.ex}\\
	\glt `There are no groupers anywhere.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_ind]}}}
\z \z 

The strategy with \textit{-barak} `any' is found several times in the corpus, such as in (\ref{exe:kal:lampurbarak}). \textit{-barak} `any' is a negative polarity item that has to be accompanied by a negated verb, such as \textit{konatnin} `not see' in~(\ref{exe:kal:lampurbarak}). It is also used as `even' in negative clauses. \textit{-barak} `any' was also elicited with the meaning `not any', see~(\ref{exe:kal:konbarak}) (although this is arguably the same meaning as in~(\ref{exe:kal:nobody}), which could be paraphrased as `not any person').

\ea \label{exe:kal:lampurbarak}
\gll lampur-\textbf{barak} ma konat=\textbf{nin}\\
lamp-any \textsc{3sg} see=\textsc{neg}\\
\glt `He didn't see any lamp.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_29:22]}}
\z 

\ea \label{exe:kal:konbarak}
\gll sairarar reidak ba kon-\textbf{barak} an jiet=\textbf{nin}\\
lobster many but one-any \textsc{1sg} get=\textsc{neg}\\
\glt `There were many lobsters but I didn't get any.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg]}}
\z 


Another strategy to create indefinites, both negative and positive, is by full reduplication (a common cross-linguistic strategy, \citealt[][179]{Haspelmath1997}). Example (\ref{exe:kal:anyno}) illustrates the meanings `anywhere' and `not anywhere', which are both created by reduplicating the question word \textit{tamangga} `to/from where'. The negated version is made by negating the verb.

\ea \label{exe:kal:anyno}
\ea \label{exe:kal:anywhere}
{\gll an marmar=i tamangga∼tamangga bisa bot\\
	\textsc{1sg} walk={\gli} to.where∼\textsc{red} can go\\
	\glt `I can walk anywhere.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg]}}
}

\ex \label{exe:kal:nowhere}
{\gll \textbf{tamangga}∼\textbf{tamangga}=ta an bot=\textbf{nin}\\
	to.where∼\textsc{red}={\glta} \textsc{1sg} go=\textsc{neg}\\
	\glt `I don't (want to) go anywhere.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg]}}
}
\z \z

Reduplication is also used to create yet another instance of `not any', as in \REF{exe:kal:donkonkon}.


\ea \label{exe:kal:donkonkon}
\gll ma \textbf{don} \textbf{kon}∼\textbf{kon} paruot=\textbf{nin}\\
\textsc{3sg} thing one∼\textsc{red} do=\textsc{neg}\\
\glt `He didn't do anything.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BAA-C}{[stim7\_25:50]}}
\z 

Note that negator \textit{=nin} can also be added directly to \textit{don konkon} to create the meaning `it doesn't matter', as in \REF{exe:kal:donkonkonin}. There is no positive variant of either \textit{don konkon V=nin} or \textit{don konkonin}.
   
\ea \label{exe:kal:donkonkonin}
\gll ma miat=nin=et eh \textbf{don} \textbf{kon∼kon=nin}\\
\textsc{3sg} come=\textsc{neg}={\glet} \textsc{inj} thing one∼\textsc{red=neg}\\
\glt `She didn't come, it doesn't matter.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC3-B}{[conv7\_11:27]}}
\z 

One strategy to create the meaning `no-one' or `nobody', with \textit{-barak} `any' and a negated verb, was shown in~(\ref{exe:kal:nobody}). Another strategy is to negate the construction \textit{=i koyet}, literally the predicate linker and the verb `to be finished', but here used to mean `all'. This is shown in \REF{exe:kal:allnone}.

\ea \label{exe:kal:allnone}
\ea \label{exe:kal:all}
{\gll tabaruop-bon=i koyet\\
	lead-\textsc{com}={\gli} finished\\
	\glt `[The sinkers are] all with lead.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C76-B}{[stim16\_1:28]}}
}

\ex \label{exe:kal:none}
{\gll an jiet=\textbf{nin=i} \textbf{koyet}\\
	\textsc{1sg} get=\textsc{neg}={\gli} finished\\
	\glt `I didn't get any.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg]}}
}
\z \z

The meanings `ever' and `never' are underspecified. Sentences with `ever' are translated with iamitive \textit{se}, and those with `never' with nondum \textit{tok} and a verb negated with \textit{=nin}. These constructions are also used to create meanings such as `already' and `not yet' (\sectref{sec:standneg}).

The quantifier \textit{reingge} `not much/many' is the negation of quantifiers \textit{reidak} `many, much', \textit{ikon} `some, a few' and \textit{taukon} `some, a few'. Consider~(\ref{exe:kal:reingge}).

\ea 	\label{exe:kal:reingge}
\gll mu buoksarun=at paruo ba bolodak to sontum tok \textbf{reingge}\\
\textsc{3pl} offering=\textsc{obj} make but just.little right people yet not.many\\
\glt	`They made the offerings, but just a little, right, there weren't many people yet.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BB3-0}{[narr7\_0:49]}}
\z 

\textit{Reingge} `not much/many' is a contraction of a morpheme \textit{rei-} with unknown meaning (though compare \textit{reidak} `much/many') and propositional negator and negative answer \textit{ge}. 

\subsection{Negative derivation}
\label{sec:negder}
There are no morphological ways to derive negative forms from nouns. The only kind of negative derivation is with the inherently negative verb \textit{eranun} `cannot', which triggers nominalisation of the verb it follows. Consider~(\ref{exe:kal:kuarun}). Perhaps a better paraphrasis of the example is `cooking is not possible'. \textit{Eranun} `cannot' is also discussed in \sectref{sec:lexicalneg}.

\ea \label{exe:kal:kuarun}
\gll pi pasor kuar-un eranun\\
\textsc{1pl.incl} fry cook-\textsc{nmlz} cannot\\
\glt `We fry, [we] cannot cook.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BCE-D}{[conv4\_3:46]}}
\z 




\section{Other aspects of negation} 
\label{sec:othneg}
Here I will discuss aspects of negation that have not been treated in \sectref{sec:ninneg} or \sectref{sec:nonclneg}. I will describe the marking and scope of negation in multi-verb constructions (\sectref{sec:negSVCscope}), negative polarity (\sectref{sec:negpol}), marking of NPs under negation (\sectref{sec:nps}), reinforcing negation (\sectref{sec:reinfneg}), and, finally, negation in complex clauses (\sectref{sec:complexneg}).
\subsection{The scope of negation}
\label{sec:negSVCscope}
In negated multi-verb constructions,\footnote{Because multi-verb constructions may be linked by the predicate linker \textit{=i} or non-final \textit{=te} on the first verb, these constructions are not defined as serial verb constructions, following the criteria in \citet{Haspelmath2016}.} predicate negator \textit{=nin} comes on the last verb and has scope over the entire construction. When \textit{=nin} comes on the first of a series of two verbs, or when both verbs are negated with \textit{=nin}, the construction is biclausal. This is marked with intonation, with the repetition of the subject, and/or with conjunction \textit{ba} `but', preferably with all three. Example (\ref{exe:kal:marmarsvc}) illustrates all possible placements of predicate negator \textit{=nin} in a multi-verb construction with two verbs.

\ea \label{exe:kal:marmarsvc}
\ea \label{exe:kal:marmarmia}
\gll canam marmar mia\\
man walk come\\
\glt		`The man comes walking.'

\ex
\gll canam marmar miat=\textbf{nin}\\
man walk come=\textsc{neg}\\
\glt 	`The man doesn't come walking.'	

\ex
\gll canam mian ba ma marmar=\textbf{nin}\\
man come but \textsc{3sg} walk=\textsc{neg}\\
\glt	`The man comes but he isn't walking.'	

\ex \label{exe:kal:miatninbamarmar}
\gll canam miat=\textbf{nin} ba ma marmar=teba\\
man come=\textsc{neg} but \textsc{3sg} walk=\glteba\\
\glt	`The man doesn't come, but he's walking.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg]}}
\z \z 

Speakers disprefer clauses where the first in a series of two verbs was negated, or where both verbs were, but where the two verbs were not separated with repetition of the subject or with \textit{ba}. That is, forms like \textit{marmarnin mia}, \textit{miatnin marmar} and \textit{marmarnin miatnin} were usually rejected.

Negated multi-verb constructions in the corpus are rare, but~(\ref{exe:kal:marmarieratnin}) is an example.

\ea 	\label{exe:kal:marmarieratnin}
\gll pi nabestai nain mengga \textbf{marmar=i} \textbf{erat=nin}\\
\textsc{1pl.incl} well like \textsc{dist.lat} walk={\gli} go.uphill=\textsc{neg} \\
\glt	`Usually we don't walk up the hill.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_24:36]}}
\z 


\subsection{Negative polarity}
\label{sec:negpol}
Except for \textit{-barak} `any' and the constructions with \textit{don konkonin} `it doesn't matter' and \textit{don konkon} V=\textit{nin} `not any', discussed in \sectref{sec:negind}, there are no negative polarity items.

\subsection{The marking of NPs in the scope of negation}
\label{sec:nps}
NPs in the scope of negation are not marked any differently that those in affirmative clauses.

\subsection{Reinforcing negation}
\label{sec:reinfneg}
The generic emphatic interjection \textit{o} can be added to basically anything. It is also seen after \textit{ge} `no; not' and predicate negator \textit{=nin}. Example (\ref{exe:kal:nino}) illustrates a reinforced negation with \textit{=nin}, and~(\ref{exe:kal:mmgeo}) illustrates a reinforced negation with \textit{ge}, as part of an exchange between two people doing a picture matching task; (\ref{exe:kal:mmgeo}) also has reinforcer \textit{o}.


\ea 	\label{exe:kal:nino}
\gll an paruot=\textbf{nin} \textbf{o}\\
\textsc{1sg} do=\textsc{neg} \textsc{emph}\\
\glt 		`I won't do it!' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBD-5}{[conv12\_10:09]}}
\z 

\ea 	\label{exe:kal:mmgeo}
\begin{xlist} 
	\exi{A:}
	{\gll coba an tok sanggarat=et anggon=at\\
		try \textsc{1sg} still search={\glet} \textsc{1sg.poss=obj}\\
		\glt `Try, I'm still searching mine.'
	}
	\exi{B:} 
	{\gll tok mambara opa me mm \textbf{ge} \textbf{o}\\
		still standing {\glopa} {\glme} \textsc{inj} no \textsc{emph}\\
		\glt `That one where [he's] still standing, hmm, no.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BE7-5}{[stim42\_13:02]}}
		
	}
\end{xlist}
\z 



\subsection{Negation and complex clauses}
\label{sec:complexneg}
There are no special ways to negate complex clauses. Two coordinated negatives or a coordinated positive and negative do not have any special effects. Thus, each clause in a coordination can be negated with predicate negator \textit{=nin}, as illustrated in~(\ref{exe:kal:negneg}). There are no negative conjunctions. Note that the conjunction \textit{ba} `but' is only used when there is a mismatch between the two actions, i.e. when there is negation in one clause but not in the other, as in~(\ref{exe:kal:b}) and~(\ref{exe:kal:c}).
% \hc{@LSP: make sure that the first line is italicised in the following examples, not the second one}

\ea 	\label{exe:kal:negneg}
\ea  \label{exe:kal:a}
\itshape Tumun ma me ewanin, ma tirinin.\\
\gll tumun ma me ewa=\textbf{nin} ma tiri=\textbf{nin}\\
child \textsc{3sg} {\glme} speak=\textsc{neg} \textsc{3sg} run=\textsc{neg}\\ 
\glt	`That child, he doesn't speak and he doesn't run.'

\ex \label{exe:kal:b}
\itshape Tumun ma me ewa ba, ma tirinin.\\
\gll tumun ma me ewa ba ma tiri=\textbf{nin}\\
child \textsc{3sg} {\glme} speak but \textsc{3sg} run=\textsc{neg}\\ 
\glt	`That child, he speaks but he doesn't run.'

\ex \label{exe:kal:c}
\itshape Tumun ma me ewanin ba, ma tiri.\\
\gll tumun ma me ewa=\textbf{nin} ba ma tiri\\
child \textsc{3sg} {\glme} speak=\textsc{neg} but \textsc{3sg} run\\ 
\glt	`That child, he doesn't speak but he runs.' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1C60-A}{[elic\_neg]}}
\z \z 

\subsection{Miscellaneous aspects of negation} 
The propositional negator and negative answer \textit{ge} `no' is used in two non-negative contexts. First, it is used as a confirmation tag, as in~(\ref{exe:kal:getag}). Second, it is used to form polar questions, as in~(\ref{exe:kal:yege2}). Note that \textit{mu sem ge} in~(\ref{exe:kal:getag}) cannot mean `they are not afraid', as \textit{sem} is a stative verb `to be afraid' which has to carry the verbal negation marker \textit{=nin} in order to be negated.

\ea \label{exe:kal:getag}
\gll mu sem \textbf{ge}\\
\textsc{3pl} afraid no.\textsc{tag}\\
\glt 	`They were afraid, you see?' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BBB-2}{[narr40\_15:32]}}
\z 

\ea 	\label{exe:kal:yege2}
\gll ma sungsung=at napaki \textbf{ye} \textbf{ge}\\
\textsc{3sg} pants=\textsc{obj} use or not\\
\glt	`Is he wearing pants?' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1BC6-C}{[narr17\_0:45]}}
\z 

An apparent derivative from \textit{ge} is \textit{gen} `maybe'. There is no productive morphology that explains the \textit{-n} in \textit{gen}. An example is given in~(\ref{exe:kal:gen}). \textit{Gen} is either clause-initial or follows the subject.

\ea 	\label{exe:kal:gen}
\gll ma \textbf{gen} sara tabai-jie\\
\textsc{3sg} maybe go.up tobacco-buy\\
\glt	`Maybe he came up to buy tobacco?' {\href{http://hdl.handle.net/10050/00-0000-0000-0004-1B9F-F}{[conv9\_22:03]}}
\z


\section{Summary}
Kalamang has three main morphemes that aid in the construction of negative clauses and negation: predicate negator \textit{=nin}, propositional negator and negative answer \textit{ge}, and existential negator \textit{saerak}. The forms and functions are summarised in Table~\ref{tab:kal:sum}.


\begin{table}
	\caption{Summary of negation in Kalamang}
	\label{tab:kal:sum}

	\begin{tabularx}{\textwidth}{l Q}
		\lsptoprule 
		Form& Function \\
		\midrule
		\textsc{pred}=\textit{nin} & standard negator \\
		\textit{ge} & propositional negator, negative answer \\
		\textit{saerak}& existential negator, negator of possessive clause \\
		\textsc{pro}=\textit{mun} \textsc{pred}=\textit{in} & prohibitive \\
		\lspbottomrule	
	\end{tabularx}
\end{table}


Standard negation is done with negator \textit{=nin}. This is a clitic that attaches to the predicate, and which is thus also used for non-verbal predicates like locatives. \textit{Ge} is the propositional negator, used in for example negated equational clauses. It is also the negative answer. \textit{Saerak} is the existential negator, and is also used for negation of possessive clauses. There is little that sets positive and negative clauses apart, with the NP remaining the same and most clausal modifiers available also under negation. The prohibitive is formed with two markers at different places in the clause: a suffix \textit{-mun} on the pronoun and a dedicated morpheme \textit{=in} on the predicate. There do not seem to be well-established negative indefinite constructions, except for the reduplication of \textit{kon} `one'. The only negative derivation happens with use of the dedicated negative verb \textit{eranun} `cannot', which triggers nominalisation of the preceding verb. Iamitive \textit{se} and nondum \textit{tok} interact will all kinds of negation, be it standard negation, inherently negative verbs or otherwise negated clauses to form the meanings `not anymore' and `not yet'. Nondum \textit{tok}, like \textit{ge}, can also be used as a negative answer. \textit{Ge} has several derivatives, such as \textit{get me}, which introduces negative conditionals, and \textit{gen} `maybe'. It is also used non-negatively as a confirmation tag and in polar questions.

\section*{Abbreviations}
\largerpage
\begin{tabularx}{.54\textwidth}{lQ}
1 & first person\\
2 & second person\\
3 & third person\\
\textsc{an} & animate\\
\textsc{attr} & attributive\\	
\textsc{anml} &  animals \\ 
\textsc{ben} & benefactive\\
\textsc{caus} & causative\\
\textsc{clf} & classifier\\
\textsc{cond} & conditional\\
\textsc{com} & comitative \\
\textsc{dem} & demonstrative\\
\textsc{dist} & distal\\
\textsc{emph} & emphasis marker \\
\textsc{encrg} & interjection of (annoyed) encouragement\\
\textsc{excl} & exclusive\\
\textsc{ex} & existential \\
\textsc{foc} & focus\\
\textsc{iam} & iamitive \\
\textsc{imp} & imperative\\
\textsc{incl} & inclusive\\
\textsc{ins} & instrumental\\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}%HC
\textsc{inj} & interjection\\
\textsc{ints} & intensifier\\
\textsc{irr} & irrealis\\
\textsc{lat} & lative\\
\textsc{loc} & locative\\
\textsc{neg} & negator, negative\\
\textsc{nmlz} & nominaliser\\
\textsc{obj} & object marker\\
\textsc{prog} & progressive \\
\textsc{pl} & plural\\
\textsc{plnk} & predicate linker\\
\textsc{poss} & possessive\\
\textsc{pred} & predicate \\
\textsc{pro} & pronoun \\
\textsc{proh} & prohibitive\\
\textsc{prox} & proximal\\
\textsc{quant} & quantifier\\
\textsc{red} & reduplication \\
\textsc{simv} & similative \\
\textsc{sg} & singular\\
\textsc{tag} & sentence tag\\
\textsc{top} & topic\\
\textsc{vol} & volitional\\
\end{tabularx}

\newpage
\largerpage
\printbibliography[heading=subbibliography,notkeyword=this]
\largerpage


\end{document}
