\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233770}
\author{Jakob Lesage\affiliation{Humboldt-Universität zu Berlin; 
    CNRS-LLACAN; 
    INALCO, Paris}}
\title{Negation in Kam}
\abstract{This chapter provides a comprehensive analysis of the negation system of Kam, a Niger-Congo language spoken in central-east Nigeria. The analysis is based on data from various sources, including elicitation, narratives and conversations. It follows the same structure as all other chapters in this volume, treating negation in different types of simple clauses, in complex clauses (including questions of scope and Neg-raising), and negative replies. Some notable features of negation in Kam are the following. Morphological tone plays a major role in expressing negation, as it does in other aspects of Kam. Negation interacts in complex ways with tense and aspect, showing a clear distinction under negation between future and non-future tense which is rare in the area where Kam is spoken. Negation is also expressed differently in clauses with an object and clauses without an object. There is a special intensity construction, which features the clausal negator used expletively, expressing an emphatically positive meaning instead. In other areas, e.g. in non-verbal predication, Kam shows common typological patterns. The chapter also discusses related constructions, including nondum and discontinuative constructions. The conclusion gives a brief overview of most markers that are discussed. Some markers and syntactic devices appear to be optional and require more research.

\keywords{Negation, Kam language, Niger-Congo, Adamawa languages, Grammatical tone}}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 

\section{The language}\label{sec:kam:1}

Kam (ISO: \href{https://iso639-3.sil.org/code/kdx}{kdx}, Glottocode: \href{https://glottolog.org/resource/languoid/id/kamm1249}{kamm1249}) is spoken in central-east Nigeria by the Kam ethnic group. The name of the language in Kam is \textit{àngwɔ̀m} and the speakers call themselves \textit{nyí ngwɔ̀m} ‘Kam people’. \textit{Kam} is probably an exonym, but the Kam have adopted it and their oral tradition states that it was coined by a Kam person. Most speakers live in around thirty villages in Bali Local Government Area in Taraba State, in an area bounded by the Kamajim mountain to the east and the Kam river to the south. The largest of these villages is Mayo Kam. There are also communities of Kam speakers in large Nigerian towns and cities such as Lagos, Abuja and Jalingo, the state capital of Taraba. I estimate that in 2020 there were between 8,000 and 15,000 speakers of Kam \citep[6-7]{Lesage2020}.

Kam has been classified as an Adamawa language, but published evidence for this claim is meager, based on 100 words published in the early 20\textsuperscript{th} century (\citealt[547--550]{Meek1931}, \citealt{Greenberg1963}, \citealt{Boyd1989}). In the absence of more convincing evidence, Kam is better considered an unclassified Niger-Congo language. Its Niger-Congo membership is fairly well established \citep{Lesage2019}. In the past, Kam was in close contact with one or more Central Jukunoid languages (Jukun Takum or Wapan, Benue-Congo, Niger-Congo). It shows minimal lexical links with Mumuye and the Yendangic languages (Adamawa, Niger-Congo) such as Wam and Nyesam \citep{Lesage2019}. Hausa (Chadic, Afro-Asiatic) is the lingua franca of the region and Fulfulde (Atlantic, Niger-Congo) has been a politically important language for some centuries due to past Islamic conquests initiated by Fula leaders and due to British indirect rule. The Kam associate closely with the Jirim, who are considered preferred marriage partners and who participate in some traditional Kam ceremonies. The Jirim speak a language closely related to Chamba Daka (Benue-Congo, Niger-Congo).\footnote{ISO 639-3 codes and Glottocodes for languages mentioned are the following: 
    Jukun Takum: \href{https://iso639-3.sil.org/code/jbu}{jbu}, \href{https://glottolog.org/resource/languoid/id/juku1254}{juku1254}, 
    Wapan : \href{https://iso639-3.sil.org/code/juk}{juk}, \href{https://glottolog.org/resource/languoid/id/wapa1235}{wapa1235},
    Mumuye : \href{https://iso639-3.sil.org/code/mzm}{mzm}, \href{https://glottolog.org/resource/languoid/id/nucl1240}{nucl1240}, 
    Wam or Kugama : \href{https://iso639-3.sil.org/code/kow}{kow}, \href{https://glottolog.org/resource/languoid/id/kuga1239}{kuga1239}, 
    Nyesam : \href{https://iso639-3.sil.org/code/pbn}{pbn}, \href{https://glottolog.org/resource/languoid/id/kpas1242}{kpas1242}, 
    Hausa : \href{https://iso639-3.sil.org/code/hau}{hau}, \href{https://glottolog.org/resource/languoid/id/haus1257}{haus1257}, 
    Adamawa Fulfulde : \href{https://iso639-3.sil.org/code/fub}{fub}, \href{https://glottolog.org/resource/languoid/id/adam1253}{adam1253}, 
    Jirim : \href{https://iso639-3.sil.org/code/dir}{dir}, \href{https://glottolog.org/resource/languoid/id/diri1253}{diri1253}, 
    Nigerian Pidgin : \href{https://iso639-3.sil.org/code/pcm}{pcm}, \href{https://glottolog.org/resource/languoid/id/nige1257}{nige1257}, 
    English : \href{https://iso639-3.sil.org/code/eng}{eng}, \href{https://glottolog.org/resource/languoid/id/stan1293}{stan1293}, 
    Tiv : \href{https://iso639-3.sil.org/code/tiv}{tiv}, \href{https://glottolog.org/resource/languoid/id/tivv1240}{tivv1240}.}

Most Kam speakers are multilingual. Apart from Kam, they speak at least Hausa. Many also speak Jirim, and some speak Nigerian Pidgin and/or English. Others speak Mumuye, Jukun Takum or Wapan, Fulfulde and other languages spoken in the wider area such as Tiv (Benue-Congo, Niger-Congo). 

Most Kam speakers live in small communities. The Kam people have little representation in the state government. More urban areas (such as Mayo Kam) show a tendency for language shift to Hausa. One might therefore consider Kam a vulnerable language. It will likely be affected by demographic changes currently happening in the region. This is particularly relevant considering the rapid rate of population growth and urbanization in Nigeria as well as the presence of Boko Haram in the north-east of the country, which has led to an influx of refugees among the welcoming Kam community.

\citet{Lesage2020} provides the most up-to-date grammatical description of Kam. Its basic constituent order is SVO. Kam has prepositions and modifiers always follow heads in the noun phrase. Kam has cross-linguistically rare features such as labial-velar stops (/kp/ and /gb/), logophoric pronouns, tense-aspect-negation marking on subject pronouns, clause-final negation, and a plural word. Other features of Kam are a prevalence of multi-verb constructions and tonal morphology. Many of these typological features are characteristic of the linguistic area called the Macro-Sudan Belt \citep{Güldemann2008,Güldemann2018} that covers large parts of West and Central Africa, including the area where Kam is spoken. Segmental inflectional morphology is scarce and Kam does not have gender agreement, an otherwise widespread feature of Niger-Congo languages. Repetition and reduplication play a striking role in different aspects of Kam grammar, including future tense constructions (see \sectref{sec:kam:2.1.2}). As this chapter will illustrate, many constructions in Kam involve multiple grammatical markers (and tones) simultaneously marking the same grammatical category in different places in the clause. Tones are essential to negation in Kam and to the interpretation of the examples presented in this chapter. For this reason, I briefly introduce the properties of tone that are most relevant here. Kam has three tone levels: L (\textit{à}), M (\textit{ā}) and H (\textit{á}). M and H tones spread one position to the right, regardless of morpheme boundaries \REF{ex:kam:1}.\footnote{Tone spread is optional: in careful speech, tones do not always spread, and it is more prominent in some speakers’ speech than in others.}\textsuperscript{,}\footnote{When a tone spreads to a prepausal syllable, this results in a falling HL, HM or ML contour tone, as both the spreading tone and the tone originally attached to the targeted syllable are realized on this syllable. HL contour: àlíb \# nyì → àlíb nyî [head thing] ‘something’s head’; HM contour: àlíb \# yō → àlíb yo᷇  [head \textsc{pl}] ‘heads’.}


\ea\label{ex:kam:1}
    M tone: ǹc\={ã}  \# àtàr → ǹc\={ã}  ātàr [star stone] ‘star of stone’\\
    H tone: àlíb \# dìnà → àlíb dínà [head mosquito] ‘head of mosquito’
\z

Occasionally, a floating L tone occurs: a L tone that follows the segmental form of a morpheme and only surfaces on the following syllable. A floating L tone does not surface when the morpheme in question is prepausal, as in \REF{ex:kam:2}, where the floating L tone marking the possessive pronoun \textit{ú\textsuperscript{L}} ‘3\textsc{sg:poss’} remains unassigned. The second line in the example displays the underlying lexical tones. Superscript \textsuperscript{L} indicates a floating L tone.

\ea\label{ex:kam:2} 
    \glll àná ú.\\
          àná ú-\textsuperscript{L}\\
          father 3\textsc{sg-poss}\\
    \glt ‘His/her father.’
\z 

A \textsuperscript{L}{}-tone may affect the realization of the following syllable. When a \textsuperscript{L}{}-tone precedes a H or M tone, it causes downstep, lowering the pitch of the following and all subsequent tones \REF{ex:kam:3}. I indicate downstep with a downstep arrow ꜜ. When a \textsuperscript{L}{}-tone follows a H or M tone, it blocks tone spread.

\ea\label{ex:kam:3} 
    \glll ngmgbàb àná \textbf{ú} \textbf{ꜜkə́} bùn.\\
        ngmgbàb àná ú-\textsuperscript{L} kə́ àbùn\\
        shoot father \textsc{3sg-poss} \textsc{ins} arrow\\
    \glt ‘Shoot his father with an arrow.’\footnote{An initial \textit{à-} in nouns is typically reduced to \textit{ə̀} or can be dropped altogether in utterance-internal positions. When a H tone precedes and would spread an \textit{à-}prefix, and the \textit{à} is dropped because of this rule, this H-spread does not surface anywhere, as in the realization of \textit{àbùn} ‘arrow’ in \REF{ex:kam:3}.}
\z 

When a H or M tone spreads over a L tone, or a H tone spreads over a M tone, this L or M tone is delinked, or ‘pushed’ rightwards, and acts as a floating \textsuperscript{L}-tone, causing downstep \REF{ex:kam:4}.

\ea\label{ex:kam:4} 
    sírá \# àlíb → sírá áꜜlíb [hair head] ‘head hair’
\z 

Morphological tone (or \textit{replacive} tone) occurs widely across Kam negation constructions. When a morpheme carries morphological tone, the glosses indicate this with a backslash ({\textbackslash}). In the second line of examples, a backslash is followed by the tonal morpheme in question (H, M, L). In the third line, a backslash is followed by the abbreviation for the grammatical category expressed. \REF{ex:kam:5} shows the verb \textit{gɔ̀m} ‘pay’ marked by H tone as a negative future verb.

\ea\label{ex:kam:5}
    \glll ... gɔ́m ú ...\\
        {} gɔ̀m{\textbackslash}\textbf{H} ú \\
        {} pay{\textbackslash}\textbf{\textsc{neg}.\textsc{fut}} 3\textsc{sg} \\
    \glt ‘(It will) not pay him ...’
\z 

This chapter follows the structure of the negation questionnaire provided by the editors (\citetv{chapters/questionnaire}).
%\mm{@LSP: cross-ref to the questionnaire}.
I describe clausal negation in \sectref{sec:kam:2}, which covers standard verbal negation (\sectref{sec:kam:2.1}), negative imperatives and questions (\sectref{sec:kam:2.2}), negation in stative predications (\sectref{sec:kam:2.3}), and negation in subordinate clauses (\sectref{sec:kam:2.4}). The section on non-clausal negation (\sectref{sec:kam:3}) contains information on negative replies in \sectref{sec:kam:3.1}. I discuss negative indefinites (\sectref{sec:kam:3.2}) and negative derivation (\sectref{sec:kam:3.3}) very briefly, since these topics are not prominent in Kam. Other aspects of negation (\sectref{sec:kam:4}) I discuss briefly are the scope of negation (\sectref{sec:kam:4.1}), negative polarity items (\sectref{sec:kam:4.2}), reinforcing negation (\sectref{sec:kam:4.4}), negative transport (\sectref{sec:kam:4.5}), metalinguistic negation (\sectref{sec:kam:4.6.1}), and an emphatic affirmative construction that features the clause-final negator \textit{hǹ} without expressing negative meaning (\sectref{sec:kam:4.6.2}). Finally, \sectref{sec:kam:5} provides an overview of negative markers in the clause. Where a cross-linguistically interesting construction is not present in Kam, or where there’s not enough data to make claims on a topic, I mention this in the relevant sections. Research on Kam is ongoing and further research may change some of the findings presented in this paper.

\section{Clausal negation}\label{sec:kam:2}
\subsection{Standard negation}\label{sec:kam:2.1}
\subsubsection{Introduction}\label{sec:kam:2.1.1}

Standard negation interacts in complex ways with tense and aspect in Kam. Most affirmative constructions have no simple negative counterpart. In other words, there is constructional asymmetry between affirmative and negative constructions. Conversely, most negative constructions cannot straightforwardly be transformed into an affirmative construction. Some tense and aspect distinctions in affirmative constructions disappear under negation: there is paradigmatic asymmetry. These asymmetries are summarized in \tabref{tab:kam:1} below (see also \citealt{Miestamo2005}). Future constructions provide a good example of constructional asymmetry. One type of affirmative future construction optionally repeats the main verb after an object, as in \REF{ex:kam:6}.

\ea\label{ex:kam:6} 
    \gll ì \textbf{jùt} ù \textbf{jūtī}.\\
        \textsc{2pl} \textbf{give}:\textsc{fut} 3\textsc{sg} \textbf{give}\\
    \glt ‘We gave it to him.’ (2023-01-23, Babangida Audu, elicitation)
\z 

Negating this statement yields a construction that is asymmetric with the affirmative construction: there is no repetition but there are two markers that have no affirmative equivalents (as in \REF{ex:kam:7}; note that the second marker is optional, cf. \sectref{sec:kam:2.1.2} below).

\ea\label{ex:kam:7} 
    \gll ì \textbf{rə́} ꜜjút ú \textbf{rə́} hǹ.\\
        1\textsc{pl} \textbf{\textsc{neg.fut}} give:\textsc{neg.fut} 3\textsc{sg} \textbf{\textsc{neg.fut}} \textsc{neg}\\
    \glt ‘We will not give it to him.’ (2023-01-23, Babangida Audu, elicitation)
\z 

Negation, tense and aspect values are generally marked by more than one marker in the clause. I give an overview of all negation markers in \sectref{sec:kam:5} below. The only marker that consistently recurs in standard negation constructions is the clause-final marker \textit{hǹ}. Apart from \textit{hǹ}, (a) there is replacive tone on verbs (see \tabref{tab:kam:3} below), (b) there are preverbal markers, (c) there is a marker that follows the object and (d) there are separate sets of negative 1\textsc{sg} and 2\textsc{sg} subject pronouns (see \tabref{tab:kam:2}). The negative pronouns qualify as \textsc{stamp}{}-morphs, which simultaneously mark the subject, tense, aspect, mood and polarity of a clause (cf. \citealt{Anderson2016}). Because Kam only has these for 1\textsc{sg} and 2\textsc{sg} categories and they occur in the subject pronoun slot in the clause, I refer to them as pronouns. 1\textsc{sg} and 2\textsc{sg} (and 1\textsc{pl)} subject pronouns cannot occur independently.

The next sections give an overview of the different affirmative and corresponding negative constructions in declarative clauses. The basic distinctions relevant to declarative verbal negation in Kam are the following:

\begin{itemize}
    \item Tense: future vs. non-future tense
    \item Aspect: perfective vs. imperfective aspect
    \item The presence of an object vs. the absence of an object 
    \item Polarity: positive vs. negative polarity
\end{itemize}


The third point, presence or absence of an object, does not only involve the difference between transitive and intransitive clauses. It also distinguishes transitive clauses where the object is expressed from those where the object is not expressed. There are no third person inanimate pronouns in Kam. As a result, inanimate objects are often left unexpressed (see examples \REF{ex:kam:23}, \REF{ex:kam:25} and \REF{ex:kam:68}).

Nondum and discontinuative constructions are variations on standard negation constructions but have some important differences. For this reason, I discuss them as other clausal negation constructions in \sectref{sec:kam:2.6}.

\subsubsection{Future tense constructions}\label{sec:kam:2.1.2}

There are two affirmative future constructions in Kam. Which one is used depends on whether there is an object in the clause or not. There is only one negative future construction, which does not distinguish object-less from object-having clauses. The negative future construction is very different from the positive future constructions.

The positive future with an object is expressed by the structure in \REF{ex:kam:8}, where \textit{\'{N}} and \textit{á} are the first and second person singular subject pronouns.

\ea\label{ex:kam:8}
    Future tense construction with an object\\
    \'{N}, á or NP\textsuperscript{Subject} \hspace{3mm} + V\textsuperscript{L} \hspace{3mm}+ NP\textsuperscript{Object} \hspace{3mm}(+ V-i)
\z 

In other words, this construction involves a H tone \textsc{1sg} or \textsc{2sg} subject pronoun (or any other noun phrase in subject function with its lexical tone), a floating \textsuperscript{L} tone after the main verb, and optional repetition of the verb after the object. The repeated verb carries the final vowel \textit{{}-i} if its stem ends in a consonant, as in \REF{ex:kam:9}.\footnote{This final vowel obligatorily occurs in a wide range of constructions, including on deverbal attributive adjectives, on speech and thought reporting verbs in reported speech constructions, in the intensity construction with partial reduplication (see \sectref{sec:kam:4.6.2}). Sometimes it occurs without a clear function \citep[246--251]{Lesage2020}.}

\ea\label{ex:kam:9} 
    \glll \textbf{ḿ} \textbf{búg} à \textbf{būg-ī}.\\
            \textbf{\'{N}} būg-\textbf{\textsuperscript{L}} á \textbf{būg-i}\\
            \textsc{1sg.sbj.\textbf{fut}} pinch-\textbf{\textsc{fut}} \textsc{2sg.obj}\textbf{\textsuperscript{}} \textbf{pinch-\textsc{fv}}\\
    \glt ‘I will pinch you.’ (session\_345, Danjuma Bello, elicitation)
\z 
Object-less future tense clauses \REF{ex:kam:10} have the same H tone \textsc{1sg} and \textsc{2sg} subject pronouns. Instead of full repetition of the verb, however, the verb is partially reduplicated. This reduplication involves an initial copy of the first syllable of the verb with a prespecified high vowel (original \textit{a}, \textit{e} and \textit{ɛ} become \textit{i}; original \textit{ɔ} and \textit{o} become \textit{u}). There is no replacive tone. The final vowel -\textit{i} is added to the reduplicated verb if its stem ends in a consonant (e.g. in \REF{ex:kam:11}).

\ea\label{ex:kam:10} 
    Future tense construction without an object\\
    \'{N}, á or NP\textsuperscript{Subject} + V{\textasciitilde}V(-i)
\z 

\ea\label{ex:kam:11} 
    \glll \textbf{ń} \textbf{síssàrí}.\\
        \textbf{\'{N}} \textbf{sàrí{\textasciitilde}}sàrí\\
        \textsc{1sg.sbj}\textbf{.\textsc{fut}} \textbf{\textsc{fut}}\textbf{{\textasciitilde}}peel \\
    \glt ‘I will peel (it).’ (session\_345, Danjuma Bello, elicitation)
\z 
The negative future construction (\ref{ex:kam:12}--\ref{ex:kam:14}) is formed by the same H tone pronouns found in affirmative future constructions. However, it involves a H melody on verbs instead of the lexical tone or a floating \textsuperscript{L} tone, a preverbal marker \textit{rə́}\textsuperscript{L} and an optional clause-final marker \textit{rə̀}. The negator \textit{hǹ} is placed at the end of the clause. There is no repetition or reduplication and there is no final vowel \textit{{}-i} on the verb.

\ea\label{ex:kam:12} 
    Negative future tense construction\\
    \'{N}, á or NP\textsuperscript{Subject} + rə́\textsuperscript{L} + V{\textbackslash}H + NP\textsuperscript{Object} (+ rə̀) + hǹ
\z 

\ea\label{ex:kam:13} 
    \glll \textbf{ń} \textbf{rə́} \textbf{ꜜjút} ú \textbf{rə́} \textbf{hǹ}.\\
        \textbf{ń} \textbf{rə́\textsuperscript{L}} jùtí\textbf{{\textbackslash}H} ú \textbf{rə̀} \textbf{hǹ}\\
        \textsc{1sg}.\textbf{\textsc{neg}}\textbf{.\textsc{fut}} \textbf{\textsc{neg}}\textbf{.\textsc{fut}} give{\textbackslash}\textbf{\textsc{neg}}\textbf{.\textsc{fut}} \textsc{3sg} \textbf{\textsc{neg}}\textbf{.\textsc{fut}} \textbf{\textsc{neg}}\\
    \glt ‘I will not give it to him/her.’ (2023-01-23, Babangida Audu, elicitation)
\z 
\ea\label{ex:kam:14} 
    \glll \textbf{rə́} \textbf{ꜜkábrí} \textbf{rə́} \textbf{hǹ}.\\
        \textbf{rə́\textsuperscript{L}} kàbrí{\textbackslash}\textbf{H} \textbf{rə̀} \textbf{hǹ}\\
        \textbf{\textsc{neg}}\textbf{.\textsc{fut}} break{\textbackslash}\textbf{\textsc{neg}}\textbf{.\textsc{fut}} \textbf{\textsc{neg}}\textbf{.\textsc{fut}} \textbf{\textsc{neg}}\\
    \glt ‘(Whatever you put in this bag,) (it) will not break.’ (session\_290, David Mark, elicitation)
\z 
\subsubsection{Non-future tense constructions}\label{sec:kam:2.1.3}

Affirmative declarative clauses in Kam specify future tense, perfective aspect or imperfective aspect. Under negation, however, aspectual distinctions are neutralized: only the future-non-future distinction remains. There are two negative non-future constructions, differentiated by the tone on the verb. One construction marks negative non-future clauses with an expressed object, the other construction marks negative non-future clauses without an expressed object. This distinction is not apparent in affirmative non-future constructions. I present the two affirmative non-future constructions first, and then describe the two negative non-future constructions. 

Perfective aspect is encoded by the schema in \REF{ex:kam:15}, where \textit{\'{N}} and \textit{á} are the 1\textsc{sg} and 2\textsc{sg} pronouns. \textit{à} is an optional post-verbal perfective marker \REF{ex:kam:16}. It is not always present, especially when there is adverbial modification \REF{ex:kam:17} and sometimes in multi-verb constructions (where its tone may still be present on or after the first verb, blocking H tone spread \REF{ex:kam:18}).\footnote{An alternative explanation for the floating \textsuperscript{L} tone after \textit{pə́} in \REF{ex:kam:18} is that \textit{pə̀} does not carry aspect marking but simply carries a H tone that spreads from the preceding syllable (\textit{wú}), which delinks the lexical L tone of \textit{pə} which then acts as a floating \textsuperscript{L} tone. \citet[309]{Lesage2020} follows this analysis. This issue requires further work.} It does not occur in imperfective aspect constructions and in future tense constructions.

\ea\label{ex:kam:15} 
    Perfective aspect construction\\
    \'{N}, á or NP\textsuperscript{Subject} + V{\textbackslash}H + NP\textsuperscript{Object} (+ à)
\z 

\ea\label{ex:kam:16} 
    \glll \textbf{ḿ} \textbf{fá} kúmé \textbf{â}.\\
        \textbf{\'{N}} fà{\textbackslash}\textbf{H} kúmé \textbf{à}\\
        \textsc{1sg}.\textsc{sbj}.\textbf{\textsc{pfv}} raise{\textbackslash}\textbf{\textsc{pfv}} chicken \textbf{\textsc{pfv}}\\
    \glt ‘I raised chickens.’ (session\_365, Babangida Audu, elicitation)
\z 

\ea\label{ex:kam:17} 
    \glll ú \textbf{ngmgbáb} ú kə́ bùn. \\
        ú ngmgbàb{\textbackslash}\textbf{H} ú kə́ àbùn\\
        \textsc{3sg} shoot{\textbackslash}\textbf{\textsc{pfv}} \textsc{3sg} \textsc{ins} arrow\\
    \glt ‘He shot him with an arrow.’ (session\_048, Isa Sarkin Dawa, narrative)
\z 

\ea\label{ex:kam:18} 
    \glll àwú \textbf{pə́} yī lū bì.\\
        àwú pə̀{\textbackslash}\textbf{H} \textbf{à?} yī lù àbì\\
        \textsc{3sg} come{\textbackslash}\textbf{\textsc{pfv}} \textbf{\textsc{pfv}}? arrive lie down\\
    \glt ‘He came and lay down.’ (session\_181, Babangida Audu, narrative)
\z 

Imperfective aspect is expressed by L tone \textsc{1sg} and \textsc{2sg} subject pronouns and the pre-verbal marker \textit{rə̀}. There is no replacive tone on verbs, see (\ref{ex:kam:19}--\ref{ex:kam:21}).

\ea\label{ex:kam:19} 
     {Imperfective}  {aspect}  {construction}\\
    \textbf{\`{N}}, \textbf{à} or NP\textsuperscript{Subject} + rə̀ + V + NP\textsuperscript{Object}
\z 

\ea\label{ex:kam:20} 
    \gll \textbf{à} \textbf{rə̀} bɛ́n ántɛ̄rī.\\
        \textsc{2sg}.\textsc{sbj}\textbf{.\textsc{ipfv}} \textbf{\textsc{ipfv}} say truth\\
    \glt ‘You are telling the truth.’ (session\_159, Solomon Ahmadu, elicitation)
\z 

\ea\label{ex:kam:21} 
    \gll \textbf{ǹ} \textbf{rə̀} būnī nyɛ́b.\\
        \textsc{1sg.sbj}\textbf{\textsc{.ipfv}} \textbf{\textsc{ipfv}} cross go\\
    \glt ‘I am crossing (the road).’ (session\_311, Danjuma Bello, elicitation)
\z 

The distinction between perfective and imperfective aspect is neutralized under negation. Independent from this, there are two negative non-future constructions. One construction negates clauses with an expressed object (\ref{ex:kam:26}--\ref{ex:kam:28}), the other negates clauses without an expressed object (\ref{ex:kam:22}--\ref{ex:kam:25}). Both constructions have L tone \textsc{1sg} and \textsc{2sg} subject pronouns (as opposed to H tone in affirmative perfective constructions) and the clause-final negator \textit{hǹ}. There is also an optional pre-verbal negative marker \textit{màg} (see below). Only the tone on the verb is different: in the object-less construction, there is no grammatical tone and the verb is left with its lexical tone \REF{ex:kam:22}. In the construction with an object, the verb is marked with a replacive H tone \REF{ex:kam:26}. Neither construction can carry the preverbal affirmative imperfective marker \textit{rə̀} or the postverbal affirmative perfective marker \textit{à}.

\ea\label{ex:kam:22} 
    Negative non-future construction without object: no replacive H tone on verbs\\
    \`{N}, à or NP\textsuperscript{Subject} (+ màg) + V + hǹ
\z 

\ea\label{ex:kam:23} 
    \gll \textbf{ǹg} gɔ̀m \textbf{hǹ.}\\
        \textsc{1sg}.\textbf{\textsc{sbj}}\textbf{.\textsc{neg}}\textbf{.\textsc{nfut}} pay \textbf{\textsc{neg}}\\
    \glt ‘I did not pay.’ (session\_364, David Mamuda, elicitation)
\z 

\ea\label{ex:kam:24} 
    \gll \textbf{à} pə̀ \textbf{hǹ}.\\
        \textsc{2sg}.\textbf{sbj.\textsc{neg}}\textbf{.\textsc{nfut}} come \textbf{\textsc{neg}}\\
    \glt ‘You did not come.’ (session\_364, David Mamuda, elicitation)
\z 

\ea\label{ex:kam:25} 
    \gll ì dɔ́g \textbf{hǹ}.\\
        \textsc{1pl} see \textbf{\textsc{neg}}\\
    \glt ‘We did not see (it).’ (session\_364, David Mamuda, elicitation)
\z 

\ea\label{ex:kam:26} 
    Negative non-future construction with object: replacive H tone on verbs\\
    \`{N}, à or NP\textsuperscript{Subject} (+ màg) + V{\textbackslash}H + NP\textsuperscript{Object} + hǹ
\z 

\ea\label{ex:kam:27} 
    \glll \textbf{ǹg} \textbf{gɔ́m} ú h\^{n}.\\
        \textbf{\`{N}} gɔ̀m{\textbackslash}\textbf{H} ú hǹ\\
        \textsc{1sg}.\textsc{sbj}\textbf{.\textsc{neg}}\textbf{.\textsc{nfut}} pay{\textbackslash}\textbf{\textsc{neg}}\textbf{.\textsc{nfut}} \textsc{3sg} \textsc{neg}\\
    \glt ‘I did not pay him.’ (session\_364, David Mamuda, elicitation)
\z 

\ea\label{ex:kam:28} 
    \glll nyò \textbf{gbɛ́b} ú \textbf{hǹ}.\\
        nyò gbɛ̀b\textbf{{\textbackslash}H} ú \textbf{hǹ}\\
        \textsc{2pl} beat{\textbackslash}\textbf{\textsc{neg}}\textbf{.\textsc{nfut}} \textsc{3sg} \textbf{\textsc{neg}}\\
    \glt ‘You people did not beat him.’ (session\_364, David Mamuda, elicitation)
\z 

Negative non-future constructions may feature an optional pre-verbal negator \textit{màg}, as in (\ref{ex:kam:29}--\ref{ex:kam:32}).\footnote{An alternative pronunciation, \textit{mà}, elides the final velar. This pronunciation is homophonous with the Hausa word \textit{mà} ‘also, too’, which is also often found in Kam discourse (e.g. \REF{ex:kam:71}). When my consultants were confronted with this pronunciation in a clause in isolation, they often could not tell for certain which \textit{mà} was intended when not presented with the English translation.}  It does not change the semantics of a proposition when inserted or left out. One could hypothesize that it serves an important function in those clauses where the only obligatory negation marking is the clause-final marker \textit{hǹ}. Some (but not all) of my consultants said that \textit{màg} is never obligatory but that it is always preferable, yet in elicited clauses it only occasionally occurs. Its frequency in natural discourse remains to be assessed. I observed \textit{màg} mostly in perfective contexts: there is only one example in my data where it occurs with an imperfective reading \REF{ex:kam:32}. There is no evidence that \textit{màg} sets apart perfective from imperfective constructions.

\ea\label{ex:kam:29} 
    \glll \textbf{\`{m}} \textbf{(màg)} \textbf{ním} àfàm nág \textbf{h\^{n}}.\\
        \textbf{\`{N}} \textbf{(màg)} nìm{\textbackslash}\textbf{H} àfàm nág \textbf{hǹ}\\
        \textsc{1sg}.\textsc{sbj}.\textbf{\textsc{neg}}\textbf{.\textsc{nfut}} \textbf{(\textsc{neg}}\textbf{.\textsc{nfut}}) eat{\textbackslash}\textbf{\textsc{neg}}\textbf{.\textsc{nfut}} meat cow \textbf{\textsc{neg}}\\
    \glt ‘I did not eat cow meat.’ (session\_285, Babangida Audu, elicitation)
\z 

\ea\label{ex:kam:30} 
    \glll Gàmbó \textbf{(mâg)} \textbf{kím} ḿfàm ú \textbf{hǹ}.\\
        Gàmbó \textbf{(màg)} kìm{\textbackslash}\textbf{H} \`{m}fàm ú-\textsuperscript{L} \textbf{hǹ}\\
        Gambo \textbf{(\textsc{neg}}\textbf{.\textsc{nfut}}) chew{\textbackslash}\textbf{\textsc{neg}}\textbf{.\textsc{nfut}} meat \textsc{3sg-poss} \textbf{\textsc{neg}}\\
    \glt ‘Gambo did not eat his meat.’ (session\_285, Babangida Audu, elicitation)
\z 

\ea\label{ex:kam:31} 
    \glll àwú \textbf{(mág)} (ꜜ)\textbf{pə́} līlē \textbf{hǹ}.\\
        àwú \textbf{(màg)} pə̀{\textbackslash}\textbf{H} àlīlē \textbf{hǹ}\\
        \textsc{3sg} (\textbf{\textsc{neg}}\textbf{.\textsc{nfut}}) come{\textbackslash}\textbf{\textsc{neg}}\textbf{.\textsc{nfut}} yesterday \textbf{\textsc{neg}}\\
    \glt ‘He did not come yesterday.’ (session\_285, Babangida Audu, elicitation)
\z 

\ea\label{ex:kam:32}
    \glll \textbf{(màg)} sín á \textbf{hǹ} yí?\\
        \textbf{(màg)} sìn{\textbackslash}\textbf{H} á \textbf{hǹ} yí?\\
        \textbf{\textsc{neg}}\textbf{.\textsc{nfut}} do{\textbackslash}\textbf{\textsc{neg}}\textbf{.\textsc{nfut}} 2\textsc{sg.obj} \textbf{\textsc{neg}} \textsc{q}\\
    \glt ‘Isn’t it doing (it for) you?’ (i.e. ‘Aren’t you satisfied?’) (session\_271, Anonymous, conversation) 
\z 
\subsubsection{Summary}\label{sec:kam:2.1.4}

\tabref{tab:kam:1} summarizes the tense, aspect, and negation values that are obligatorily expressed in Kam clauses, referring to relevant examples in the preceding text.

\begin{table}\footnotesize
    \centering
    \begin{tabularx}{\textwidth}{XXlll}
    \lsptoprule
    Polarity value & With or without object & Future & \multicolumn{2}{c}{Non-future}\\
    &  &  & Perfective & Imperfective\\
    \midrule 
    Positive polarity & With object & \REF{ex:kam:9} & \REF{ex:kam:16}, \REF{ex:kam:18} & \REF{ex:kam:20}, \REF{ex:kam:21}\\
    & Without object & \REF{ex:kam:11} &  & \\
    Negative polarity & With object & \multirow{2}{*}{\REF{ex:kam:13}, \REF{ex:kam:14}} & \multicolumn{2}{c}{\REF{ex:kam:29}}\\
    & Without object &  & \multicolumn{2}{c}{\REF{ex:kam:25}}\\
    \lspbottomrule
    \end{tabularx}
    \caption{Tense, aspect, and negation values in declarative clauses}
    \label{tab:kam:1}
\end{table}

\tabref{tab:kam:1} also highlights some paradigmatic asymmetries (cf. \citealt{Miestamo2005}): aspectual distinctions (perfective vs. imperfective aspect) are lost under negation, as well as the distinction between an object-having and an object-less future. One grammatical distinction is relevant in negative constructions but is neutralized in affirmative constructions: the distinction between object-having vs. object-less non-future clauses.


\subsection{Negation in non-declaratives}\label{sec:kam:2.2}

Kam imperatives are expressed by the bare stem of the verb without added markers or morphology: \textit{lù} ‘sleep!’; \textit{jɛ̄rī} ‘look!’. Negative imperatives are formed by a discontinuous negation construction involving a pre-verbal marker \textit{kə́} and a clause-final marker \textit{gà}. This construction is used for 2\textsc{sg} addressees and 2\textsc{pl} addressees alike. When the 2\textsc{pl} pronoun \textit{nyò} occurs, it precedes the preverbal marker \textit{kə́}, as in examples (\ref{ex:kam:33}--\ref{ex:kam:35}).

\ea\label{ex:kam:33} 
    \gll \textbf{kə́} lú \textbf{gà}!\\
        \textbf{\textsc{neg}}\textbf{.\textsc{imp}} sleep \textbf{\textsc{neg}}\textbf{.\textsc{imp}}\\
    \glt ‘Don’t sleep!’ (session\_137, Babangida Audu, elicitation)
\z

\ea\label{ex:kam:34} 
    \gll \textbf{kə́} jɛ́rī núng \textbf{gâ}!\\
        \textbf{\textsc{neg}}\textbf{.\textsc{imp}} look bird \textbf{\textsc{neg}}\textbf{.\textsc{imp}}\\
    \glt ‘Don’t look at the bird!’ (session\_137, Babangida Audu, elicitation)
\z 

\ea\label{ex:kam:35} 
    \gll nyò \textbf{kə́} nyɛ́b ꜜnggwɔ́g wàn nɛ̀ \textbf{gà}!\\
        \textsc{2pl} \textbf{\textsc{neg}}\textbf{.\textsc{imp}} go house chief \textsc{emph} \textbf{\textsc{neg}}\textbf{.\textsc{imp}}\\
    \glt ‘Don’t go to the house of the chief!’ (to more than one person) (Tope\_Bitrus\_06, Garba Abubakar Bako, elicitation) 
\z 

Negative polar questions can be used to ask rhetorical questions or to ask for confirmation of a suspicion. There are two kinds of polar questions in Kam: those with the clause-final interrogative marker \textit{yí} \REF{ex:kam:36} and those that are marked by a clause-final H tone \REF{ex:kam:37}. The question marker \textit{yí} is placed after the negative clause that is questioned.

\ea\label{ex:kam:36} 
    \gll lɛ́m nyó \textbf{hǹ} \textbf{yí}?\\
            be.good:\textbf{\textsc{neg}}\textbf{.\textsc{nfut}} \textsc{2pl} \textbf{\textsc{neg}} \textbf{\textsc{q}}\\
    \glt ‘(They are helping, and you are still complaining.) Isn’t it good for you people?’ (session\_271, Anonymous, conversation)
\z 

A question marked by a clause-final H tone causes the L tone of the clause-final negator \textit{hǹ} to be realized as H, as in \REF{ex:kam:37}.

\ea\label{ex:kam:37} 
    \glll àkp\'{õ}  ꜜngwún, àkp\'{õ}  ꜜngwún á \textbf{hń}?\\
        àkp\'{õ}  ǹgwún àkp\'{õ}  ǹgwún á \textbf{hǹ{\textbackslash}H}\\
        twenty five twenty five \textsc{prst} \textbf{\textsc{neg}}\textbf{{\textbackslash}}\textbf{\textsc{q}}\\
    \glt ‘One hundred, isn't it one hundred?’ (session\_034, Isa Sarkin Dawa, conversation)
\z 

Negative content questions of the type ‘who/what/... did not do X’ are expressed by relative clauses of the type ‘who/what/where/how is it that did not do X?’ (see \sectref{sec:kam:2.4}). Grammatically, there is no difference between a negative content question, where the negation targets the propositional content of the clause as a whole, and a content question containing a negated relative clause, where the negation targets only the relative clause (\REF{ex:kam:38}, cf. \sectref{sec:kam:4.6.2}).

\ea\label{ex:kam:38} 
    \gll ànyí mín\textbf{ā} màg kím búwú \textbf{hǹ} \textbf{à}?\\
        person which\textbf{:\textsc{rel}} \textsc{neg.nfut} chew:\textsc{neg.nfut} one:\textsc{3sg:poss} \textbf{\textsc{neg}} \textbf{\textsc{rel}}\\
    \glt ‘Who did not eat their piece?’ or ‘Who is the person that did not eat their piece?’ session\_285, Babangida Audu, elicitation)
\z 

\subsection{Negation in stative predications}\label{sec:kam:2.3}

\subsubsection{Negation of equation and proper inclusion}\label{sec:kam:2.3.1}

Affirmative nominal predication is expressed by simple juxtaposition of the subject noun phrase and the predicate noun phrase. The predicate noun phrase is marked by a clause-final presentative marker \textit{á}. There are no tense and aspect distinctions in equation and proper inclusion constructions.

Nominal predication is negated by the regular clause-final verbal negator \textit{hǹ} (\ref{ex:kam:39}--\ref{ex:kam:41}). Other than this, there is no difference between affirmative nominal predication constructions and negative nominal predication constructions. The other negative markers available in verbal clauses cannot occur in the nominal predication construction: there is no verb, there are no grammatical tense or aspect distinctions, and the 1\textsc{sg} and 2\textsc{sg} pronouns in nominal predication constructions are independent pronouns (\textit{nyìm} ‘1\textsc{sg.idp}’, \textit{nyà} ‘2\textsc{sg.idp}’), which do not mark different tense, aspect and negation values.

\ea\label{ex:kam:39} 
    Negative nominal predication construction\\
    NP\textsuperscript{\textsc{verbless.subject}} + NP\textsuperscript{\textsc{complement}} + á + hǹ
\z

\ea\label{ex:kam:40} 
    \gll Jɔ́n, àwú málám á \textbf{h\^{n}}, àwú nyí nyí ꜜíbārī á.\\
        John \textsc{3sg} teacher \textsc{prst} \textbf{\textsc{neg}} \textsc{3sg} person thing farm:\textsc{adjz} \textsc{prst}\\
    \glt ‘John, he is not a teacher, he is a farmer.’ (session\_285, Babangida Audu, elicitation)\\
\z 

\ea\label{ex:kam:41} 
    \gll ànyàgbàn à mē ḿ ꜜá \textbf{h\^{n}}. àngwán ḿ ꜜá.\\
        woman \textsc{prox} mother \textsc{1sg:poss} \textsc{prst} \textsc{neg} in.law \textsc{1sg:poss} \textsc{prst}\\
    \glt ‘This woman is not my mother. She’s my (mother-)in-law.’ (session\_285, Babangida Audu, elicitation)
\z 
\subsubsection{Existential, locative and possessive negation}\label{sec:kam:2.3.2}

The copula \textit{də́} ‘to be, to exist’ expresses existence and location. In existential clauses (but not usually in locational clauses), \textit{də́} is usually followed by the emphatic marker \textit{bhə́}, which elsewhere also functions as a presentative marker. There are no tense and aspect distinctions in copula constructions (see \ref{ex:kam:42}--\ref{ex:kam:43}).

\ea\label{ex:kam:42} 
    \gll nyìsīrī \textbf{də́} \textbf{bhə́}\\ 
        money \textbf{be} \textsc{emph}\\
    \glt ‘There is money.’
\z 

\ea\label{ex:kam:43} 
    \gll ànggwɔ́g á \textbf{ꜜdə́} lɛ́b ə̀sh\'{ẽ} \\
        house \textsc{prox} \textbf{be} inside bush\\
    \glt ‘The house is in the bush.’ (Tope\_Bitrus\_06, Garba Abubakar Bako, elicitation)
\z 

Speakers use the negative existential \textit{sìrí} combined with the clause-final mar\-ker \textit{ríg} to say that something does \textit{not} exist or that something is \textit{not} in a specific place. \textit{ríg} has an emphatic function in affirmative clauses (e.g. \REF{ex:kam:68}). I did not find the presentative marker \textit{bɛ̄} in negative existential clauses (see \ref{ex:kam:44}--\ref{ex:kam:45}).

\ea\label{ex:kam:44} 
    \gll ànggwɔ́g  \textbf{sìrí} lɛ́b ə́ꜜsh\'{ẽ}  \textbf{ríg}.\\ 
        house  \textbf{be.\textsc{neg}} inside bush \textbf{\textsc{neg}}\\
    \glt ‘The house is not in the bush.’ (2024-11-03-Rahab, Rahab Garba Precious, elicitation)
\z 


\ea\label{ex:kam:45} 
    \gll nyì \textbf{sìrí} \textbf{ríg}.\\
        thing \textbf{be.\textsc{neg}} \textbf{\textsc{neg}}\\
    \glt ‘There is nothing.’ (session\_253, Anonymous, narrative)
\z 
There are two predicative possession constructions in Kam. The first construction is similar to locational predication constructions in Kam, but instead of a locational prepositional phrase (\textit{lɛ́b ə}́ꜜ\textit{sh\'{ẽ}} ‘in the bush’ in \REF{ex:kam:44}), the possessive construction has a comitative prepositional phrase (\textit{rə̀ wó} ‘with a dog’ in \REF{ex:kam:46}).


\ea\label{ex:kam:46} 
    \gll ń \textbf{də́} \textbf{rə́} \textbf{ꜜ}wó\\
        1\textsc{sg.sbj} \textbf{be} \textbf{\textsc{com}} dog\\
    \glt ‘I have a dog.’ (session\_071, Isa Sarkin Dawa, elicitation)
\z 

This construction is negated by the same negative copula \textit{sìrí} and the clause final negator \textit{ríg}, as in locative predications. Usually, there is no comitative marker in the negative construction \REF{ex:kam:47}. Occasionally, it still occurs, as in \REF{ex:kam:48}.

\ea\label{ex:kam:47} 
    \gll àwú \textbf{sírí} ꜜnggwɔ́g \textbf{ríg}.\\ 
        \textsc{3sg} \textbf{be.\textsc{neg}} house \textbf{\textsc{neg}}\\
    \glt ‘He does not have a house.’ (session\_208, Babangida Audu, elicitation)
\z 

\ea\label{ex:kam:48} 
    \gll Àúdú \textbf{sìrí} \textbf{rí} cì càr \textbf{ríg}.\\
        Audu \textbf{be.\textsc{neg}} \textbf{\textsc{com}} child three \textbf{\textsc{neg}} \\
    \glt ‘Audu doesn’t have three children.’ (session\_287, Babangida Audu, elicitation)
\z 

Predicative possession can also be expressed by the regular verb \textit{kàr} ‘to keep, to have’. This construction is negated by standard negation, including negative subject pronouns, pre-verbal negators, replacive tone, post-verbal negators and the clause-final negator \textit{hǹ} (see \sectref{sec:kam:2.1} and \REF{ex:kam:72}).

The expression of predicative possession and its negation in Kam completely aligns with cross-linguistic patterns that are already very well studied in the typological literature, e.g. \citet{Stassen2009}; \citet{Heine1997}; \citet{KutevaEtAl2019}, and \citet{Veselinova2013neg.exist}.

\subsubsection{Negation of adjectival predication (attribution)}\label{sec:kam:2.3.3}

Affirmative predicative adjectives in Kam are derived from verbs by means of partial reduplication (\REF{ex:kam:49}, see also \citealt[264--265]{Lesage2020}).

\TabPositions{.27\textwidth,.66\textwidth} 
\ea\label{ex:kam:49}
kág ‘to crunch’ \tab → kíkkág ‘to be crunched’\\
būm ‘to spoil’ \tab → būmbūm ‘to be spoiled’\\
sɔ̀ngí ‘to make tasty’ \tab → sùnsɔ́ꜜngí ‘to be tasty (of food)’ \\ 
\z 

To mark tense or aspect distinctions or to negate adjectival predication constructions, speakers use the simplex verb instead of the reduplicated form (\ref{ex:kam:50}--\ref{ex:kam:51}). Standard negation constructions are used (see \sectref{sec:kam:2.1}). One could say that the stative/dynamic distinction is lost under negation.

\ea\label{ex:kam:50} 
    \gll mìngkángá \textbf{dág} \textbf{h\^{n}.}\\
        water \textbf{be.cold} \textbf{\textsc{neg}}\\
    \glt ‘The water is not cold.’ (session\_163, Solomon Ahmadu, elicitation)
\z 

\ea\label{ex:kam:51} 
    \gll àtábé bá \textbf{lɛ̀m} k-ánīmī \textbf{h\^{n}}\\
        fish \textsc{prox} \textbf{be.good} \textsc{advz}{}-eat:\textsc{adjz} \textbf{\textsc{neg}}\\
    \glt ‘This fish is not good to eat.’ (lit: ‘This fish is not good as food.’) (session\_169, Babangida Audu, elicitation)
\z 

If a reduplicated form of the adjective is used in combination with \textit{hǹ}, the result is a completely different construction with an emphatically positive meaning rather than a negative meaning. I discuss this construction in \sectref{sec:kam:4.6.2} below.

\subsection{Negation in non-main clauses}\label{sec:kam:2.4}

Relative clauses in Kam follow the noun and involve two relativizers: the marker \textit{ā}, which follows the noun and precedes the relative clause, and the marker \textit{à} which follows the relative clause. Relative clauses are negated like main clauses, by the clause-final negator \textit{hǹ} and any other negators enclosed by the two relativizers. The relative clause in \REF{ex:kam:52} involves standard negation in a non-future tense construction (cf. \sectref{sec:kam:2.1.3}). The relative clause in \REF{ex:kam:53} involves negation of property assignment (cf. \sectref{sec:kam:2.3.3}).

\ea\label{ex:kam:52} 
    \gll ànyí \textbf{mínā} \textbf{màg} \textbf{kím} \textbf{búwú} \textbf{hǹ} \textbf{à}?\\
        person \textbf{which:\textsc{rel}} \textbf{\textsc{neg}}\textbf{.\textsc{nfut}} \textbf{chew:\textsc{neg}}\textbf{.\textsc{nfut}} \textbf{one:3\textsc{sg}}\textbf{:\textsc{poss}} \textbf{\textsc{neg}} \textbf{\textsc{rel}}\\
    \glt ‘(Who is) the person that did not eat their piece?’ (session\_285, Babangida Audu, elicitation)\\
\z 

\ea\label{ex:kam:53} 
    \gll àcì gbɛ́b wó yág ꜜkə́ \textbf{kwí} \textbf{ā} \textbf{nád} \textbf{hń} \textbf{à}.\\
        child beat:\textsc{pfv} dog carry \textsc{ins} \textbf{stick} \textbf{\textsc{rel}} \textbf{be.thick} \textbf{\textsc{neg}} \textbf{\textsc{rel}}\\
    \glt ‘The child beat the dog with a thin stick (a stick that is not thick).’ (session\_342, Isa Sarkin Dawa, elicitation)
\z 
Subordinate clauses, including relative clauses, adverbial clauses, and complement clauses, are negated like main clauses. If a subordinate clause involves the standard clause-final negator \textit{hǹ}, it occurs at the end of the complete sentence. The scope of negation is apparent from non-final negators and other markers when these are present (viz. subject pronouns, pre-verbal markers, post-object markers and replacive tone on the verb). \sectref{sec:kam:4.1} and \sectref{sec:kam:4.5} contain more discussion and examples of negation in complex sentences involving subordinate clauses.

Conditional clauses in Kam are expressed by a clause-final conditional marker \textit{kā}. When a negated clause is expressed as a condition (\ref{ex:kam:54}--\ref{ex:kam:55}), \textit{kā} simply follows the whole clause, including the negator \textit{hǹ} in verbal clauses, and the emphatic marker \textit{rīg} in existential or locational clauses \REF{ex:kam:56}.

\ea\label{ex:kam:54} 
    \gll nyí nánī \textbf{h\={n}} \textbf{kā}...\\
        people be.many \textbf{\textsc{neg}} \textbf{\textsc{cond}} \\
    \glt ‘If people aren’t many, (then what do they say)?’ (session\_034, Samuel U. Adams, conversation)
\z 

\ea\label{ex:kam:55} 
    \gll nyò dɔ́g nyí ꜜínīmī \textbf{hǹ} \textbf{kā}...\\
        \textsc{2pl} see:\textsc{neg}.\textsc{nfut} thing eat:\textsc{adjz} \textbf{\textsc{neg}} \textbf{\textsc{cond}} \\
    \glt ‘If you people do not see food ... (you people will not go to school).’ (session\_271, Anonymous, conversation)
\z 

\ea\label{ex:kam:56} 
    \gll àkpò \textbf{sìrí} \textbf{ríg} \textbf{ká}...\\
        food \textbf{be.\textsc{neg}} \textbf{\textsc{neg}} \textbf{\textsc{cond}} \\
    \glt ‘If there is no food ... (you people will not go).’ (session\_292, David Mark, elicitation)
\z 

The negative conditional coordinator \textit{sīkā} has an apprehensive reading similar to English ‘lest’ in some of its occurrences. It contrasts with the coordinator \textit{ànyáꜜlíb} that simply translates as polarity-neutral ‘because’. In \REF{ex:kam:57}, \textit{sīkā} may simply be translated as ‘if not’: the king would be caught if he were not hidden. In \REF{ex:kam:58} only an apprehensive ‘lest’ reading is possible. 

\ea\label{ex:kam:57} 
    \gll ì wɔ́ɡ ə̀wàn à, \textbf{sīkā} í pə́ ꜜcúɡ ə́wàn...\\
        \textsc{1pl} hide{\textbackslash}\textsc{fut} king \textsc{rel} \textbf{lest} \textsc{3pl} come{\textbackslash}\textbf{\textsc{fut}} catch king \\
    \glt ‘(that) we will hide the king, lest they will come and catch (our) king.’ (session\_271, Isa Sarkin Dawa, narrative)
\z 

\ea\label{ex:kam:58} 
    \gll í tín gɔ́psí, \textbf{sīkā} àyí sámrì kɔ̀m sámrí.\\
        \textsc{log} move separate, \textbf{lest} \textsc{3pl} spoil:\textsc{fut} place spoil\\
    \glt ‘(The Kam say that) they moved away, lest they (the Jukun) will ruin the place.’ (session\_271, Isa Sarkin Dawa, narrative)
\z 

In \REF{ex:kam:58}, there is no logical connection between the Kam leaving and the Jukun ruining the place. It is assumed that the Jukun will ruin the place in any case, and the Kam want to leave to avoid this ruin.
 
\subsection{Negative lexicalizations}\label{sec:kam:2.5}

The only inherently negative lexical verb in Kam is the negative existential copula \textit{sìrí} ‘not to be’, which is not marked for tense or aspect. It is always accompanied by the negative existential marker \textit{ríg} (see \sectref{sec:kam:2.3.2}). The marker \textit{ríg} has a negative existential meaning in existential and locational predication, but it has an emphatically affirmative meaning in other contexts (e.g. \REF{ex:kam:66}). This polysemy requires more discussion in further research.

\subsection{Other clausal negation constructions}\label{sec:kam:2.6}

It is not within the scope of this chapter to provide a complete treatment of phasal polarity constructions in Kam.\footnote{The continuative (‘still’) is expressed by different adverbs: \textit{táncɔ̀ng} ‘still, always’, \textit{kpɔ̀ng} ‘always, usually’, or \textit{kpàn} ‘always, usually’. Inchoative constructions (‘already’) may be left unmarked or expressed by a coordinated clause with \textit{càŋrí} ‘to finish’ or \textit{lɔ̀grí} ‘to end’. See \citet[334--337]{Lesage2020}.} Two values, however, are relevant to negation in Kam and are expressed differently to other negation constructions. These values are nondum (‘not yet’, cf. \citealt{VeselinovaDevos2021}) and the discontinuative (‘no longer’) \citep{Auwera1998}, which are viewed as negative tenses in some languages \citep{Veselinova2017}. \Citet{Auwera1998} calls nondum constructions \textit{negative continuative}, but because this term is very similar to \textit{discontinuative} (‘no longer’) and is also used with other functions elsewhere, I follow \citeapo{VeselinovaDevos2021} terminology. 


\subsubsection{Clause-final nondum marker: \textit{jɛ̀n} ‘not yet’}\label{sec:kam:2.6.1}

\textit{jɛ̀n}, the clause-final nondum marker, is semantically related to such adverbs as \textit{táncɔ̀ng} ‘still’ and \textit{kpàn} ‘always, usually’. Such adverbs can also occur in other positions and can occur as minimal clauses on their own. \textit{jɛ̀n} only occurs clause-finally, however, and can be considered a grammatical marker rather than an adverb.

Unlike standard negation clauses, nondum clauses can combine with the different affirmative aspect and future constructions without constructional asymmetries. \REF{ex:kam:59} is an imperfective nondum clause. Standard negation constructions do not distinguish perfective and imperfective aspect (\sectref{sec:kam:2.1}).

\ea\label{ex:kam:59} 
    \glll Ráháb \textbf{rə́} ꜜgbání Gárínhāmzà \textbf{jɛ̀n}.\\
        Ráháb \textbf{rə̀} gbán-i Gárínhāmzà \textbf{jɛ̀n}\\
        Rahab \textsc{ipfv} arrive-\textsc{fv} Garin.Hamza \textbf{\textsc{ndum}}\\
    \glt ‘Rahab is not yet arriving in Garin Hamza.’ (session\_288, Solomon Ahmadu, elicitation)
\z 

\REF{ex:kam:60} is a future nondum construction. Standard negative future constructions would have a preverbal negative future marker \textit{rə́}ꜜ, which is absent in this case. Standard negative future verbs have a replacive H tone instead of the floating \textsuperscript{L} tone in \REF{ex:kam:60} that is characteristic of affirmative future tense constructions.

\ea\label{ex:kam:60} 
    \glll Ráháb gbán-ì Gárínhāmzà pə́ꜜkáng \textbf{jɛ́n}.\\
        Ráháb gbán\textbf\textsuperscript{L}\textbf{{}-}i Gárínhāmzà pə́\textsuperscript{L}káng \textbf{jɛ̀n}\\
        Rahab arrive{\textbackslash}\textbf{\textsc{fut}}\textbf{{}-}\textsc{fv} Garin.Hamza tomorrow \textbf{\textsc{ndum}}\\
    \glt ‘Rahab will not yet arrive in Garin Hamza tomorrow.’ (session\_288, Solomon Ahmadu, elicitation)
\z 

While imperfective and future nondum clauses do not follow the standard negation pattern, perfective nondum clauses do. Perfective nondum clauses contain negative subject pronouns (\textit{\`{m}} ‘1\textsc{sg}.\textsc{sbj.neg}.\textsc{nfut}’ in \REF{ex:kam:61}) and the optional preverbal negator \textit{màg} (in both \REF{ex:kam:61} and \REF{ex:kam:62}). In these cases, \textit{jɛ̀n} may be considered as a sub-type of clause-final negator, since it occupies the slot usually taken by the negator \textit{hǹ}.

\ea\label{ex:kam:61} 
    \glll \textbf{\`{m}} \textbf{màg} dɔ́g ú dúdꜜdɔ́gí \textbf{jɛ́n}.\\
            \textbf{\`{m}} \textbf{màg} dɔ́g\textbf{{\textbackslash}H} ú dɔ́g{\textasciitilde}dɔ́g-i \textbf{jɛ̀n}\\
            \textbf{\textsc{1sg.sbj.neg}}\textbf{.\textsc{nfut}} \textbf{\textsc{neg}}\textbf{:\textsc{nfut}} see\textbf{{\textbackslash}}\textbf{\textsc{neg}}\textbf{.\textsc{nfut}} \textsc{3sg} \textsc{exp}{\textasciitilde}see\textbf{{}-}\textsc{exp} \textbf{\textsc{ndum}}\\
    \glt ‘I have never seen him until now.’ (Literally: ‘I have never seen him yet.’) (session\_271, Danjuma Bello, conversation)\\
\z 

\ea\label{ex:kam:62} 
    \glll àyí \textbf{mág} ꜜdɔ́g ńtèkù \textbf{jɛ̀n}.\\
        àyí \textbf{màg} dɔ́g{\textbackslash}\textbf{H} ǹtèkù \textbf{jɛ̀n}\\
        \textsc{3pl} \textbf{\textsc{neg}}\textbf{.\textsc{nfut}} see{\textbackslash}\textbf{\textsc{neg}}\textbf{.\textsc{nfut}} toad \textbf{\textsc{ndum}}\\
    \glt ‘They haven’t seen the toad yet.’ (session\_181, Babangida Audu, narrative)
\z 

\subsubsection{Discontinuative: \textit{nàng hǹ/nàng ríg} ‘no longer’}\label{sec:kam:2.6.2}

A discontinued action is expressed by regular negation constructions with a marker \textit{nàng} preceding the clause-final negator \textit{hǹ} (\ref{ex:kam:63}--\ref{ex:kam:64}). \textit{nàng} is only found elsewhere as part of the adverb \textit{nàngkò} ‘first’, and nowhere else. It is not clear what the function of -\textit{kò} is in this adverb and the exact function and full distributional properties of \textit{nàng} are matter for further research. It is not clear whether \textit{nàng} should be considered a fully grammaticalized discontinuative marker or a lexical item.

\ea\label{ex:kam:63} 
    \gll Mɛ́rí ním nyí \textbf{nàng} \textbf{hǹ}.\\
        Mary eat:\textsc{neg}.\textsc{nfut} thing \textbf{\textsc{discnt}} \textbf{\textsc{neg}}\\
    \glt ‘Mary is no longer eating.’ (session\_288, Solomon Ahmadu, elicitation)
\z 

\ea\label{ex:kam:64} 
    \gll Dɛ́yvíd cúg ə́ꜜnyág \textbf{nàng} \textbf{hǹ}.\\
        David do:\textsc{neg}.\textsc{nfut} work \textbf{\textsc{discnt}} \textsc{\textbf{neg}}\\
    \glt ‘David is no longer working.’ (session\_288, Solomon Ahmadu, elicitation)
\z 

\textit{nàng} also occurs in negative existential and locational negation constructions, preceding the clause-final negator \textit{ríg} (\REF{ex:kam:65}, cf. \sectref{sec:kam:2.3.2}).

\ea\label{ex:kam:65} 
    \gll Bàbánggídá síꜜrí Jàlìngò \textbf{nàng} \textbf{ríg}.\\
        Babangida be.\textsc{neg} Jalingo \textbf{\textsc{discnt}} \textsc{\textbf{neg}}\\
    \glt \textsc{‘}Babangida is no longer in Jalingo.’ (session\_288, Solomon Ahmadu, elicitation)
\z 

\section{Non-clausal negation}\label{sec:kam:3}
\subsection{Negative replies}\label{sec:kam:3.1}

Kam has two synonymous positive response words that can be used to respond positively to a question: \textit{m} and \textit{ooy}.\footnote{For this part of the chapter, I collected data with the questionnaire from \citet{Moser2018}.} Both are toneless but usually occur with a relatively high pitch. Both words are considered ‘proper Kam’ and not borrowed by speakers though \textit{m} also occurs in other languages in the area. \textit{ooy} seems to be more exclusive to Kam. Added to a proper name, \textit{ooy} also functions as a vocative marker. Elsewhere it can be used as a call to attract attention or as a greeting from a distance. To respond negatively to a question, a reduplicated form of either positive response words is used, with a glottal stop separating the copy from the reduplicated segment: \textit{mʔm} or \textit{oʔo}. Note that \textit{oʔo} does not have the final glide of its affirmative counterpart. The negative response words are also toneless and are found with all possible melodies.

Kam has a truth system for negative and positive replies: a positive reply signals agreement with the polarity of the question; a negative reply signals disagreement with the polarity of the question (cf. \citealt{Holmberg2015}; \citealt{Moser2018}). Response words do not indicate agreement or disagreement with the proposition in the question.
In \REF{ex:kam:66}, a positive reply to the question “They cannot speak English?” acts as confirmation that “they” indeed do not speak English (A1). A negative reply expresses that, contrary to the negative polarity value of the question, “they” \textit{do} speak English (A2). 
    
\ea\label{ex:kam:66}
    \begin{xlist}[A2: ]
    \exi{Q: }\label{ex:kam:66Q}
        \gll àyí bɛ́n túráncì màsí hń ꜜyí?\\
            \textsc{3sg} speak:\textsc{neg}.\textsc{nfut} English know \textsc{neg} \textsc{q}?\\
        \glt ‘They cannot speak English?’
    \exi{A1: }\label{ex:kam:67}
        \gll \textbf{ḿ,} àyí bɛ́n màsí h\^{n}.\\
            \textbf{yes} \textsc{3sg} speak know \textsc{neg}\\
        \glt ‘Indeed, they cannot speak English.’
    \exi{A2: }\label{ex:kam:68}
        \gll \textbf{\`{m}ʔḿ,} àyí bɛ́n màsí ríg.\\
            \textbf{no} \textsc{3sg} speak:\textsc{pfv} know \textsc{emph}\\
        \glt  ‘No, they \textit{can} speak (it).’ (session\_304, Rahab Garba Precious, elicitation)
    \end{xlist}
\z 

A negative response to a negative question often includes the emphatic marker \textit{rīg} to emphasize the reversed positive polarity value of the reply, as in (\ref{ex:kam:66}, A2).

Another way of responding positively or negatively to a question is to simply repeat the proposition with the correct polarity value, as in \REF{ex:kam:69}. Reply words are not obligatory.
 
\ea\label{ex:kam:69}
    \begin{xlist}[A2: ]
    \exi{Q: }\label{ex:kam:69Q}
        \gll kɔ̀m sísɔ́b nà yí?\\
              place \textsc{pred}:be.hot today \textsc{q}\\
        \glt ‘Is it hot today?’
    \exi{A1: }\label{ex:kam:70}
        \gll \textbf{(mʔm,)} kɔ̀m sɔ́b h\^{n}.\\
            \textbf{(no,)} place be.hot \textsc{neg}\\
        \glt ‘(No,) it is not hot today.’ (session\_304, Rahab Garba Precious, elicitation)
    \end{xlist}
\z 

The expression \textit{màsīrī} is an emphatic negative reply roughly equivalent to English \textit{never}. It indicates that a time someone just asked about does not exist (\ref{ex:kam:71}--\ref{ex:kam:72}). While it is not synchronically possible to segment the expression into smaller parts, it may derive from a combination of the optional negative marker \textit{màg} used in non-future constructions and the adverb \textit{sīrī} ‘completely’, or the existential negator \textit{sìrə̄} (\sectref{sec:kam:2.3.2}).

\ea\label{ex:kam:71}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:kam:71a}
        \gll á zúzꜜzúbí dúdꜜdɔ́gí kɔ̀m gɔ́mnátí yí? \\ 
             \textsc{2sg.sbj.fut} \textsc{fut}:steal \textsc{exp} place government \textsc{q} \\
        \glt ‘Would you ever steal from the government?’
    \exi{A: }\label{ex:kam:71b}
        \gll \textbf{màsīrī}, mà àyí rə́ ꜜcúg ḿ rə́ hǹ kā. \\
             \textbf{never}, also 3\textsc{sg} \textsc{neg}.\textsc{fut} catch:\textsc{neg}.\textsc{fut} 1\textsc{sg}.\textsc{obj} \textsc{neg}.\textsc{fut} \textsc{neg} if \\
        \glt ‘Never, (not) even if they would not catch me.’ (session\_286, Babangida Audu, elicitation)\\
    \end{xlist}
\z 

\ea\label{ex:kam:72}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:kam:72a}
        \gll kə́ bə́rí mīnā, nyí Sɛ́rkíndáwá rə̀ nìmsí nág á ꜜyí? \\ 
             \textsc{temp} time which:\textsc{rel} people Sarkin.Dawa \textsc{ipfv} feed cow \textsc{rel} \textsc{q} \\
        \glt ‘At what time do the people of Sarkin Dawa feed their cows?’ (session\_286, Babangida Audu, elicitation)\\
    \exi{A: }\label{ex:kam:72b}
        \gll \textbf{màsīrī}, nyí Sɛ́rkíndáwá màg kár nág h\^{n}. \\
            \textbf{never} people Sarkin.Dawa \textsc{neg}.\textsc{nfut} have:\textsc{neg}.\textsc{nfut} cow \textsc{neg}\\
        \glt ‘Never, the people of Sarkin Dawa don’t have cows.’ (session\_286, Babangida Audu, elicitation)
    \end{xlist}
\z 

Kam speakers use other response words , but these are not considered proper Kam by speakers. Speakers identify the following responses as Hausa, although they are probably much more widespread than Hausa:

\begin{itemize}
    \item ɛɛ (no associated tone); positive response word
    \item aʔa (no associated tone); negative response word
    \item ɛ̥↓ (voiceless ingressive ɛ; can also be other sounds, sometimes \textit{m̥↓}, a voiceless ingressive bilabial nasal; both eyebrows are raised while inhaling, and the head is tilted lightly backwards for a moment); positive response word
    \item ʞ (back-released uvular click with closed lips); positive response word \citep[cf.][]{Lionnet2020}
    \item ʞʞ (two back-released uvular clicks with closed lips); negative response word
\end{itemize}

\subsection{Negative indefinites}\label{sec:kam:3.2}

Negative indefinites and negative quantifiers do not exist as separate categories in the language. 
To express the corresponding meanings, variations of clausal negation constructions are used. 

Negative location is encoded by clausal negation and an indefinite noun phrase in the negated clause \REF{ex:kam:73}.

\ea\label{ex:kam:73} 
    \gll \`{m} màg nyɛ́b \textbf{kɔ́m} \textbf{mí} Yùrɔ̀b \textbf{hǹ}.\\
        \textsc{1sg.sbj.neg.nfut} \textsc{neg}.\textsc{nfut} go:\textsc{neg}.\textsc{nfut} \textbf{place} \textbf{some} Europe \textbf{\textsc{neg}}\\
    \glt ‘I have not been to a{\textbackslash}some place in Europe.’, ‘I have been nowhere in Europe.’ or ‘I have not been anywhere in Europe.’ (session\_286, Babangida Audu, elicitation)
\z 
Negative time, is expressed by standard negation constructions \REF{ex:kam:74}. It can also be expressed by the negative response word \textit{màsīrī} ‘never’ but then it forms an utterance of its own and does not act as an adverbial modifier (\sectref{sec:kam:3.1}).

\ea\label{ex:kam:74} 
    \gll kùrùkí ꜜvɛ́g hǹ.\\
        God die \textsc{neg}\\
    \glt ‘God does not die.’ or ‘God never dies.’ (whatsapp\_01, Babangida Audu, elicitation)
\z 
\subsection{Negative derivation}\label{sec:kam:3.3}

I did not find any instances of negative derivation. Equivalent meanings are rendered by negators on the clause level. As exemplified in \REF{ex:kam:53}, the equivalent of English negative adjectives (e.g. ‘un-happy’ or ‘un-true’) is expressed by a relative clause using the regular negator \textit{hǹ} (‘a story that is not true’).

Privative constructions, are expressed by clausal negation constructions, ne\-gating a clause with a comitative prepositional phrase \REF{ex:kam:75}.

\ea\label{ex:kam:75} 
    \gll à cúg ə́ꜜnyág rə́ nyìm hǹ.\\
        2\textsc{sg.sbj}.\textsc{neg}.\textsc{nfut} do:\textsc{neg}.\textsc{nfut} work \textsc{com} 1\textsc{sg.idp} \textsc{neg}\\
    \glt ‘You did not work with me.’ or ‘You worked without me.’ (session\_354, Rahab Garba Precious, elicitation)
\z 

\section{Other aspects of negation}\label{sec:kam:4}
\subsection{The scope of negation}\label{sec:kam:4.1}

The scope of negation over noun phrases can be indicated by cleft constructions. Cleft constructions indicate focus in Kam, including contrastive focus \citep[375--378]{Lesage2020}. In \REF{ex:kam:76}, the patient (‘fish’) is in focus, which in a pragmatically unmarked clause would follow the verb. The cleft construction moves the patient into a focus position at the beginning of the utterance. In \REF{ex:kam:77}, the agent role (‘the man’) is brought into focus with a cleft construction. 

\ea\label{ex:kam:76} 
    \gll àtábí á \textbf{ꜜ}nyí ꜜván á cùg hǹ, núng á.\\
        fish \textsc{rel} person male \textsc{prox} catch \textsc{neg} bird \textsc{prst}\\
    \glt ‘It is not a fish that the man caught, it is a bird.’ (lit. ‘It is a fish that the man did not catch, it is a bird.’ (2023-01-31, Babangida Audu, elicitation)
\z 

\ea\label{ex:kam:77} 
    \gll ànyí ꜜván á \textbf{ꜜ}míná cúg ə̀tábí h\^{n}, ànyà gbàn á, ú cúg ə́\textbf{ꜜ}tábí.\\
        person male \textsc{rel} like.this catch fish \textsc{neg} person female \textsc{prst} \textsc{3sg} catch fish\\
    \glt ‘It is not the man who caught the fish, it is the woman, she caught the fish.’ (2023-01-31, Babangida Audu, elicitation)
\z 

In my data and elicitation attempts, I found no way to restrict the scope of negation to a verb or to tense or aspect markers in a simple clause. I also did not detect a differences in prosodic prominence or intonation that could mark the scope of negation.

In complex sentences involving subordinate clauses, the main clause and dependent clause can be negated separately, or both clauses can be negated simultaneously. In all cases, the standard negator \textit{hǹ} is placed just once at the end of the complete sentence. As such, it does not disambiguate which of the clauses is negated. Non-clause-final markers and constructional asymmetries for different tense-aspect combinations may, then, indicate the scope of negation (cf. \sectref{sec:kam:2.1}). In \REF{ex:kam:78}, for example, the matrix clause (‘I don’t know.’) is negative and the subordinated clause (‘He will come.’) is affirmative. That the matrix clause is negated is apparent from the negative subject pronoun (\textit{\`{m}} ‘1\textsc{sg}.\textsc{sbj}.\textsc{neg}.\textsc{nfut}’). That the dependent clause is affirmative is apparent from the future construction with the reduplicated verb \textit{pə́ꜜpə́rí}, which only marks future tense in affirmative clauses.

\ea\label{ex:kam:78} 
    \gll \textbf{\`{m}} màsí kô ú \textbf{pə́ꜜpə́rí} h\^{n}.\\
        \textbf{\textsc{1sg}}\textbf{.\textsc{sbj}}\textbf{.\textsc{neg}}\textbf{.\textsc{nfut}} know \textsc{comp} \textsc{3sg} \textbf{\textsc{fut}}\textbf{{\textasciitilde}come} \textsc{neg}.\\
    \glt ‘I don’t know if he will come.’  (session\_286, Babangida Audu, elicitation)
\z 

\sectref{sec:kam:2.1.3} describes the optional preverbal non-future negator \textit{màg}. In complex constructions, \textit{màg} may acquire a disambiguating function. Since both affirmative perfective and negative non-future verb forms are marked by a replacive H tone, and the clause-final negator \textit{hǹ} does not indicate whether just one or both clauses are negative, \REF{ex:kam:79} is in principle ambiguous between having an affirmative matrix clause (‘He said that he will not come.’) or a negative matrix clause (‘He did not say that he will not come.’).

%\ea\label{ex:kam:79} 
 %   \gllll {àwú \textbf{bɛ́ní} íyí ꜜrə́ ꜜpə́ rə́ \textbf{hǹ}.\\
  %      {àwú \textbf{bɛ́n{\textbackslash}H}{}-i íyí\textsuperscript{L}{}-í\textsuperscript{L} rə́\textsuperscript{L} pə̀{\textbackslash}H rə̀ \textbf{hǹ}} \\
   %    {\textsc{3sg} {\textbf{say{\textbackslash}}\textbf{\textsc{neg}}\textbf{.\textsc{nfut}}{}-\textsc{quot}} \textsc{quot}{}-\textsc{log} \textsc{neg}.\textsc{fut} come{\textbackslash}\textsc{neg}.\textsc{nfut} \textsc{neg}.\textsc{fut} \textsc{neg}} \\ 
    %    {{} {\textbf{say{\textbackslash}}\textbf{\textsc{pfv}}{}-\textsc{quot}} {} {} {} {} {}}\\ 

    %\glt ‘He said that he will not come.’ or ‘He did not say that he will not come.’ (session\_286, Babangida Audu, elicitation)
%\z 

\ea\label{ex:kam:79} 
   \gllll \small{àwú} \small{\textbf{bɛ́ní}} \small{íyí} \small{ꜜrə́} \small{ꜜpə́} \small{rə́} \small{\textbf{hǹ}}. \\
        \small{àwú} \small{\textbf{bɛ́n{\textbackslash}H}{}-i} \small{íyí\textsuperscript{L}{}-í\textsuperscript{L}} \small{rə́\textsuperscript{L}} \small{pə̀{\textbackslash}H} \small{rə̀} \small{\textbf{hǹ}} \\
       \small{\textsc{3sg}} \small{{\textbf{say{\textbackslash}}\textbf{\textsc{neg}}\textbf{.\textsc{nfut}}{}-\textsc{quot}}} \small{\textsc{quot}{}-\textsc{log}} \small{\textsc{neg}.\textsc{fut}} \small{come{\textbackslash}\textsc{neg}.\textsc{nfut}} \small{\textsc{neg}.\textsc{fut}} \small{\textsc{neg}} \\ 
        {} \small{{\textbf{say{\textbackslash}}\textbf{\textsc{pfv}}{}-\textsc{quot}}} {} {} {} {} {}\\ 

    \glt \small{‘He did not say that he will not come.’ or ‘He said that he will not come.’ (session\_286, Babangida Audu, elicitation)}
\z 


To resolve this ambiguity, the marker \textit{màg} can be used to mark that the matrix clause is negative. \REF{ex:kam:80} shows \textit{màg} in a disambiguating position. In \REF{ex:kam:80}, where \textit{màg} is used, the main clause is negated. In \REF{ex:kam:81}, where \textit{màg} is absent, the main clause is not negated. Note also that the clause-final perfective marker \textit{à} in \REF{ex:kam:81} indicates that the subordinated clause is affirmative, since \textit{à} does not occur under negation (cf. \sectref{sec:kam:2.1.3}). The absence of \textit{à} in the subordinate clause in \REF{ex:kam:80} signals that it is negative.

\ea\label{ex:kam:80} 
    \gll àwú \textbf{mág} lɛ̀g íyí jɛ́nɛ́rɛ́ytɔ̀r sámr \textbf{á} yī hn᷆. \\
        \textsc{3sg} \textbf{\textsc{neg}}\textbf{.\textsc{nfut}} hear \textsc{quot} generator spoil:\textsc{pfv} \textbf{\textsc{pfv}} stand \textsc{neg}\\
    \glt ‘He did not hear that the generator had exploded.’ (session\_286, Babangida Audu, elicitation)
\z 

\ea\label{ex:kam:81} 
    \gll ń dɔ́g jɛ́nɛ́rɛ́ytɔ̀r \textbf{sámrí} yí hǹ.\\
        \textsc{1sg.sbj.pfv} see:\textsc{pfv} generator \textbf{spoil:\textsc{neg}}\textbf{.\textsc{nfut}} stand \textsc{neg}\\
    \glt ‘I saw that the generator had not exploded.’ (session\_286, Babangida Audu, elicitation)
\z 

When the scope of negation is restricted to a relative clause, the negator is enclosed in the relative clause and is follwed by the clause-final relativizer \textit{à} (see \sectref{sec:kam:2.4}).

\subsection{Negative polarity}\label{sec:kam:4.2}

The discontinuative marker or adverb \textit{nàng} (see \sectref{sec:kam:2.6.2}) has an unclear meaning elsewhere in the language and can be considered a negative polarity item, since it is not a negator by itself but only occurs in negative contexts. Elsewhere, \textit{nàng} only occurs as part of the adverb \textit{nàngkò} ‘first’. More research is needed to understand whether it occurs anywhere else.

The conditional conjunction \textit{sīkā} ‘if not/lest’ (\sectref{sec:kam:2.4}) carries inherently negative meaning. The clauses it connects are not negated with standard negation, however. As such, it is not a typical negative polarity item, but could be considered related to negative polarity.

\subsection{Marking of NPs in the scope of negation}\label{sec:kam:4.3}

Kam noun phrases do not behave differently under negation. Information structure has not yet been studied in detail, so it is possible that special constructions or restrictions will be discovered in the future.

\subsection{Reinforcing negation}\label{sec:kam:4.4}

Not all negative markers are fully obligatory in Kam. In negative future tense constructions, for example, there is an optional negative marker \textit{rə̀} that comes right before the clause final negator \textit{hǹ} (see \sectref{sec:kam:2.1.2}). In negative non-future tense constructions, there is an optional pre-verbal negator \textit{màg} (see \sectref{sec:kam:2.1.3}), which can also be used to indicate the scope of negation in a complex clause (see \sectref{sec:kam:4.1}). It is not clear whether these markers can also be used to reinforce negation or whether they have other information structural functions. Information structure and emphasis are not yet well understood in Kam (see \citealt{Lesage2020}: 375--386). More research is needed to understand how negation may be reinforced.

\sectref{sec:kam:3.1} mentions the emphatic negative response word \textit{màsīrī} ‘never’, which reinforces negative answers.

\subsection{Negation, coordination and complex clauses}\label{sec:kam:4.5}

Negation in complex clauses is similar to negation in simple clauses. The clause-final negator \textit{hǹ} comes at the end of the complex sentence and the other markers of negation (negative pronouns, pre-verbal negators, replacive tone, post-verbal markers) mark negation in the constituent clauses. Examples \REF{ex:kam:78} to \REF{ex:kam:81} above illustrate how negation works in complex sentences.

The constituent clauses in a complex sentence may be ambiguous between an affirmative and a negative reading. This happens in perfective object-having clauses without 1\textsc{sg} or 2\textsc{sg} subject pronouns that do not have the optional preverbal negator \textit{màg}. The replacive tone on the verb for both an affirmative perfective clause and a negative nonfuture clause is H, so it does not allow disambiguation. \REF{ex:kam:82} is an elicited example and does not contain enough context to disambiguate between different readings for the main clause. 

\ea\label{ex:kam:82} 
    \gllll àwú bɛ́ní íyí ꜜrə́ ꜜpə́ rə́ hǹ.\\
        àwú bɛ́n{\textbackslash}H-i íyí\textsuperscript{L}{}-í\textsuperscript{L} rə́\textsuperscript{L} pə̀{\textbackslash}H rə̀ hǹ\\
        3\textsc{sg} {say{\textbackslash}\textsc{neg.nfut-quot}} \textsc{quot-log} \textsc{neg.fut} come{\textbackslash}\textsc{neg.fut} \textsc{neg.fut} \textsc{neg}\\
        {} {say{\textbackslash}\textsc{pfv}{}-\textsc{quot}} {} {} {} {} {}\\

    \glt ‘He did not say that he will not come.’ or ‘He said that he will not come.’ (session\_286, Babangida Audu, elicitation)
\z

%\ea\label{ex:kam:82} 
 %   \gllll àwú bɛ́ní íyí ꜜrə́ ꜜpə́ rə́ hǹ.\\
  %      àwú bɛ́n{\textbackslash}H-i íyí\textsuperscript{L}{}-í\textsuperscript{L} rə́\textsuperscript{L} pə̀{\textbackslash}H rə̀ hǹ\\
   %     3\textsc{sg} {say{\textbackslash}\textsc{pfv}{}-\textsc{quot}} \textsc{quot-log} \textsc{neg.fut} come{\textbackslash}\textsc{neg.fut} \textsc{neg.fut} \textsc{neg}\\
    %    3\textsc{sg} {say{\textbackslash}\textsc{neg.nfut-quot}} \textsc{quot-log} \textsc{neg.fut} come{\textbackslash}\textsc{neg.fut} \textsc{neg.fut} \textsc{neg}\\

    %\glt ‘He said that he will not come.’ or ‘He did not say that he will not come.’ (session\_286, Babangida Audu, elicitation)
%\z 


The current data does not contain examples of coordinated clauses in which only the first clause is negative or in which both clauses are negative. Coordination typically involves parataxis and in some cases the coordinators \textit{àmā} ‘and/but’ and \textit{kō} ‘or’, both borrowed from Hausa. In all cases, the constituent clauses behave identical to other clauses in Kam. There do not appear to be any special negative coordinators or coordination-specific negation constructions.


Sometimes, the syntactic scope of negation is narrower than its semantic reach. What happens in Kam is related to the phenomenon that has been called Negative Transport(ation) or Neg-raising: negation on one level of a complex clause is ‘transported’ or ‘raised’ to another level of the clause \citep[312--313]{Horn2001}. In \REF{ex:kam:83}, the negation within the relative clause semantically targets the whole sentence rather than just the relative clause. The intended meaning is ‘We do not want bad things to happen.’, where the negation semantically covers the verb ‘want’. The literal interpretation of the sentence, however, is that ‘We want what is not bad.’ 
    

\ea\label{ex:kam:83} 
    \gll ì wúr\textbf{á} ꜜsámrí hǹ \textbf{à}.\\
        \textsc{1pl} want:\textsc{pfv:rel} bad \textsc{neg} \textsc{rel}\\
    \glt ‘We don’t want something bad (‘that which is bad’).’ (lit. ‘We want what is not bad.’) (session\_051, Isa Sarkin Dawa, narrative)
\z 

Examples \REF{ex:kam:84a}, \REF{ex:kam:84b}, \REF{ex:kam:84c} like \REF{ex:kam:82} above, illustrate how negators in Kam are often not syntactically bound to one level of a complex clause. In isolation, \REF{ex:kam:84a}, \REF{ex:kam:84b}, \REF{ex:kam:84c} would be ambiguous. It is not clear which of the clauses is negated or whether they are both negated. However, the context provides enough information to decide that the reading is not ‘we \textbf{don’t} want that blood does \textbf{not} drip on the ground’. In principle, the clause remains ambiguous between ‘we \textbf{don’t} want that blood drips on the ground’ and ‘we want that blood does \textbf{not} drip on the ground’.

\ea\label{ex:kam:84a} 
    \glll ì wúrí  kə́ bá wò nyár  ə́bí à rə̀ hǹ.\\ 
        ì wūrī{\textbackslash}H kə́ ba᷇  wò nyàr{\textbackslash}H àbìbí à rə̀ hǹ \\ 
        \textsc{1pl} {want{\textbackslash}\textsc{neg.nfut}} \textsc{temp} \textsc{rel} blood {drip{\textbackslash}\textsc{pfv}} ground \textsc{rel} \textsc{emph} \textsc{neg}\\

    \glt lit. ‘We don’t want that blood drips down.’
\z 

\ea\label{ex:kam:84b} 
    \glll ì wúrí  kə́ bá wò nyár  ə́bí à rə̀ hǹ.\\ 
        ì wūrī{\textbackslash}H kə́ ba᷇  wò nyàr{\textbackslash}H àbìbí à rə̀ hǹ \\ 
        \textsc{1pl} {want{\textbackslash}\textsc{pfv}} \textsc{temp} \textsc{rel} blood {drip{\textbackslash}\textsc{neg.nfut}} ground \textsc{rel} \textsc{emph} \textsc{neg}\\

    \glt lit. ‘We want that blood does not drip down.’
\z 

\ea\label{ex:kam:84c} 
    \glll ì wúrí  kə́ bá wò nyár  ə́bí à rə̀ hǹ.\\ 
        ì wūrī{\textbackslash}H kə́ ba᷇  wò nyàr{\textbackslash}H àbìbí à rə̀ hǹ \\ 
        \textsc{1pl} {want{\textbackslash}\textsc{neg.nfut}} \textsc{temp} \textsc{rel} blood {drip{\textbackslash}\textsc{neg.nfut}} ground \textsc{rel} \textsc{emph} \textsc{neg}\\

    \glt lit. ‘We want that blood does not drip down.’ \\
    (free translation for \REF{ex:kam:84a}, \REF{ex:kam:84b}, \REF{ex:kam:84c}: ‘We don’t want there to be bloodshed.’; in isolation this could also mean ‘We don’t want there not to be bloodshed.’) (session\_191, Isa Sarkin Dawa, narrative) 
\z 

\subsection{Miscellaneous aspects of negation}\label{sec:kam:4.6}

\subsubsection{Metalinguistic negation}\label{sec:kam:4.6.1}

Metalinguistic negation follows the normal rules of negation in the language. No different behavior has been observed. \REF{ex:kam:85} features the regular negative possession construction, for example (\sectref{sec:kam:2.3.2}) 

\ea\label{ex:kam:85} 
    \gll Àúdú sìrí rə́ cì càr ríg. àwú də́ rə́ cì ǹnár.\\
        Audu be.\textsc{neg} \textsc{com} child three \textsc{neg} \textsc{3sg} be \textsc{com} child four\\
    \glt ‘Audu doesn’t have three children. He has four children.’ (session\_287, Babangida Audu, elicitation)
\z 

\subsubsection{Non-negative uses of negatives and expletive negation}\label{sec:kam:4.6.2}

There is an \textit{intensity} \textit{construction} in Kam that involves the standard negator \textit{hǹ} but is emphatically affirmative instead of negative. It features a partially reduplicated form of the verb that is followed by the final vowel \textit{{}-i}.\footnote{Such reduplicated forms of verbs also occur in affirmative (but not negative) future constructions without an object (see \sectref{sec:kam:2.1.2} above) and in the perfective forms of some verbs expressing bodily functions (e.g. \textit{gúgꜜgúrí} ‘vomit’, \textit{kpiꜜkpírí} ‘cough’). See \citet[266--268]{Lesage2020}.} The verb receives a replacive HM tone. I only found this construction in elicited data so far. Its functions are not well understood and require more consideration in a larger corpus of natural language use. The construction can be schematized as in \REF{ex:kam:86}. See examples (\ref{ex:kam:87}--\ref{ex:kam:89}).

\ea\label{ex:kam:86} 
    Intensity constructions\\
    NP + {\textasciitilde}V{\textbackslash}HM-i + hǹ
\z

\ea\label{ex:kam:87} 
    \glll  ànggwɔ́g á \textbf{ꜜtíntām-ī} \textbf{hn}\raisebox{.7pt}{᷆}.\\
        ànggwɔ́g à \textbf{tám{\textasciitilde}tám{\textbackslash}HM-i} \textbf{hǹ}\\
        house \textsc{prox} \textsc{ints}{\textasciitilde}grow{\textbackslash}\textsc{ints}{}-\textsc{fv} \textsc{neg}\\
    \glt ‘The house is very big.’ (not ‘The house is not big’) (session\_071, Isa Sarkin Dawa, elicitation)
\z 

\ea\label{ex:kam:88} 
    \glll ànyìván á dùb h\^{n}, àmā, \textbf{àwú} \textbf{dúddūb-ī} \textbf{hn}\raisebox{.7pt}{᷆}.\\
        ànyìvān ā dùb hǹ àmā \textbf{àwú} \textbf{dùb{\textasciitilde}dùb{\textbackslash}HM-i} \textbf{hǹ}\\
        man \textsc{prox} be.tall \textsc{neg} but \textbf{\textsc{3sg}} \textbf{\textsc{ints}}\textbf{{\textasciitilde}be.tall{\textbackslash}}\textbf{\textsc{ints}}\textbf{{}-}\textbf{\textsc{fv}} \textbf{\textsc{neg}}\\
    \glt ‘This man is not (just) tall, but he is very tall.’ (session\_287, Babangida Audu, elicitation)
\z 

\ea\label{ex:kam:89} 
    \glll ànggwɔ́g mō ā \textbf{díddāg-ī} \textbf{hn}\raisebox{.7pt}{᷆}.\\
        ànggwɔ́g á\textsuperscript{L}mō à \textbf{dág{\textasciitilde}dág{\textbackslash}HM-i} \textbf{hǹ}.\\
        house good \textsc{prox} \textbf{\textsc{ints}}\textbf{{\textasciitilde}cold{\textbackslash}}\textbf{\textsc{ints}}\textbf{{}-}\textbf{\textsc{fv}} \textbf{\textsc{neg}}\\
    \glt ‘The good house is very cold.’ (not ‘The good house is not cold.’) (session\_342, Isa Sarkin Dawa, elicitation)
\z 
At least two possible diachronic scenarios may explain the structure of this construction. 

The negator \textit{hǹ} may have originally been an intensity marker that was used with other negators in the clause (for example, the optional pre-verbal negator \textit{màg} discussed in \sectref{sec:kam:2.1.3}). The marker \textit{hǹ} could then, over time, have acquired a negative meaning. The intensity construction may be a rare instance preserving an earlier non-negative meaning which may have been something like ‘really, truly, absolutely’. This scenario would be an example of the Jespersen Cycle, observed in many other languages, where intensifying markers evolve into negation markers over time \citep[e.g.][]{Jespersen1917,Dahl1979,Auwera2009}.

This construction could also derive from an ironic negative construction where a negative clause was used to express an emphatically affirmative proposition. This ironic negative construction could then have grammaticalized into an emphatic affirmative construction and lost its negative meaning altogether. See \citet{Shmelev2016} on the conventionalization of irony and other pragmatic effects.\footnote{See also the Lingtyp discussion on Ironic negative constructions in January 2020 at \url{http://listserv.linguistlist.org/pipermail/lingtyp/2020-January/007568.html}.}


I did not encounter other instances of expletive negation. Complex sentences have sometimes been difficult to elicit, however. And relevant constructions may be fairly rare in the Kam corpus. Future research may uncover more data relevant to this topic.

\section{Summary}\label{sec:kam:5}
 \largerpage[-1]
Tone plays a role in negative marking in declarative verbal negation constructions. Different tones on 1\textsc{sg} and 2\textsc{sg} subject pronouns (\tabref{tab:kam:2}) as well as on lexical verbs (\tabref{tab:kam:3}) signal different tense, aspect, and negation values.

\begin{table}
\small
    \begin{tabularx}{\textwidth}{llXX}
    \lspbottomrule
        \multicolumn{2}{l}{{Tense/aspect value}} & {Positive polarity} & {Negative polarity}\\
        \midrule
        \multirow{2}{*}{Non-future} & Perfective & H (\textsc{1sg}: \'{N}, \textsc{2sg}: \textit{á}) & \multirow{2}{*}{L (\textsc{1sg}: \`{N}, \textsc{2sg}: \textit{à})}\\
        & {Imperfective} & L (\textsc{1sg}: \`{N}, \textsc{2sg}: \textit{à}) & \\
        \midrule
        \multicolumn{2}{l}{Future} & \multicolumn{2}{c}{H (\textsc{1sg}: \'{N}, \textsc{2sg}: \textit{á})}\\
        \lspbottomrule
    \end{tabularx}
    \caption{Negative and affirmative pronouns}
    \label{tab:kam:2}
\end{table}

\begin{table}\small 
    \centering
        \begin{tabularx}{\textwidth}{lllXX}
        \lsptoprule
        \multicolumn{2}{l}{Tense/aspect value}          & Positive polarity                     & \multicolumn{2}{c}{Negative polarity} \\
                                        &               &                                       &  with object  & without object \\
        \midrule 
        \multirow{2}{*}{{Non-future}}   & {Perfective}  & H                                     &  H            &   lexical tone (${\emptyset}$) \\
                                        & Imperfective  & Lexical tone (${\emptyset}$)          &  H            &   lexical tone (${\emptyset}$)\\
        \midrule
        Future                          &               & Lexical tone +\textsuperscript{L}     &  H            & H \\
        \lspbottomrule
        \end{tabularx}
    \caption{Replacive tone on verbs}
    \label{tab:kam:3}
\end{table}

The eight segmental markers and one copula involved in negation in simple clauses in Kam are summarized in \tabref{tab:kam:4} according to the position they occupy in a clause. Optional markers are written between brackets.

\begin{table}
    \footnotesize
    \begin{tabularx}{\textwidth}{Qp{10mm}p{7mm}QQp{7mm}l@{}}
         \lsptoprule
            Slot & Clause-initial & Pre-verbal & Verb or copula & After \textsc{obj} and multi-verb \mbox{constructions} & Clause-final& After  negator\\
%             & initial & verbal & & multi-verb & final  & negator\\
%             & & & & constructions & & \\
         \cmidrule{1-6}
            non-future & Subject                  & (\textit{màg}) &  with \textsc{obj}: \textsc{vb}\textbackslash H  & - & \textit{hǹ } & \multirow{11}{*}{\shortstack[l]{\textit{kā} (conditional) \\ \textit{yī} (question)}} \\
                       & \textsc{1sg}: \textit{Ń}; &                &  without \textsc{obj}: \textsc{vb}                &   &                               & \\
                       & \textsc{2sg}: \textit{à}  &                &                              &   &                               & \\
           \cmidrule{1-6}
            future & Subject                   & \textit{rə́}\textsuperscript{L} & \textsc{vb}\textbackslash H & (\textit{rə̀}) & \textit{hǹ } & \\
                   & \textsc{1sg}: \textit{Ń}; &                                &                    &               &  & \\
                   & \textsc{2sg}: \textit{á}  &                                &                    &               &  & \\
           \cmidrule{1-6}
            imperative  & \mbox{\textsc{2pl}: \textit{nyo}̀}  & \textit{kə́} & \textsc{vb} & - & \textit{gà} & \\
                        & \mbox{or nothing}                  &             &   &   &             & \\
           \cmidrule{1-6}
            \multirow{2}{*}{nondum} & \multicolumn{4}{c}{Either: (1) a negative non-future clause,} & \multirow{2}{*}{\textit{jɛ̀n}} & \\
                                    & \multicolumn{4}{c}{or (2) an affirmative declarative clause}  &                                   & \\
           \cmidrule{1-6}
            existential & Subject & - & \textit{sìrí} & - & \textit{ríg} & \\
        \lspbottomrule
    \end{tabularx}
    \caption{Negative markers in Kam}
    \label{tab:kam:4}
\end{table}

Other relevant markers, not included in \tabref{tab:kam:4}, are the negative conditional and apprehensive conjunction \textit{sīkā} ‘lest’ (\sectref{sec:kam:2.4}) and \textit{mʔm} and \textit{oʔo}, the negative response words (\sectref{sec:kam:3.1}). 

\largerpage
\section*{Acknowledgments}
I gratefully acknowledge the financial support of the AdaGram project financed by the “Emergence(s)" programme of the city of Paris. I am also thankful to the workshop organizers and the participants of the workshop that lead to this volume for their interest in my work and for their valuable feedback. The editors of the volume and two anonymous reviewers provided feedback that made this chapter stronger. I wish to thank them for their support. I also want to thank Héloïse Calame for the helpful final checks and for her patience.


\section*{Abbreviations}
\largerpage[3]
\begin{tabularx}{.45\textwidth}{lQ}
1 & first person \\
2 & second person \\
3 & third person \\
\scshape adjz & adjectivizer \\
\scshape advz & adverbializer \\
\scshape com & comitative \\
\scshape comp & complementizer \\
\scshape cond & conditional \\
\scshape discnt & discontinuative \\
\scshape emph & emphasis \\
\scshape exp & experiential \\
\scshape fut & future \\
\scshape m & mid tone \\
\scshape fv & final vowel \\
\scshape h & high tone \\
\scshape idp & independent pronoun \\
\scshape imp & imperative \\
\scshape ins & instrumental \\
\scshape ints & intense \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape ipfv & imperfective \\
\scshape \textsuperscript{l} & floating low tone \\
\scshape l & low tone \\
\scshape log & logophoric pronoun \\
\scshape m & mid tone \\
\scshape n & homorganic nasal \\
\scshape neg & negator \\
\scshape nfut & non-future \\
\scshape ndum & nondum \\
NP & noun phrase \\
\scshape obj & object \\
\scshape pfv & perfective \\
\scshape pl & plural \\
\scshape poss & possessive \\
\scshape m & mid tone \\
\scshape pred & predicative \\
\scshape prox & proximal \\
\scshape prst & presentative \\
\scshape q & question marker \\
\end{tabularx}

\begin{tabularx}{.45\textwidth}{lQ}
\scshape quot & quotative \\
\scshape rel & relativizer \\
\scshape sbj & subject \\
\scshape sg & singular \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape stamp & subject, tense, aspect, mood, polarity \\
\scshape temp & temporal \\
\scshape v & verb \\
\end{tabularx}

\vspace{5mm}

\section*{Special characters}
\begin{tabularx}{\textwidth}{lQ}
ꜜ & downstep \\
- & affix boundary \\
$\#$ & potential pause \\
( ) & optional \\
. & gloss boundary but no morpheme boundary \\
... & omitted text from example \\
: & composed of multiple morphemes but not segmented in this example \\
\textbackslash  & replacive tone in gloss   \\
\textasciitilde & reduplication in gloss \\
ø & not marked \\
\end{tabularx}


\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
\end{document} 
