\documentclass[output=paper,colorlinks,citecolor=brown,chinesefont]{langscibook}
\ChapterDOI{10.5281/zenodo.18233970}
\author{Benjamin Brosig\affiliation{Academia Sinica [2018--2019]; University of Bern [2020--2023]; Nazarbayev University [2023--]}}
\title{Negation in Khalkha Mongolian}
\abstract{This chapter offers an introduction to the negation system of Khalkha Mongolian structured according to typological and functional categories. The negation of declarative predicates, nominalized elements and adverbial clauses is all done by a former existential negator (=\textit{güi}). For locative negation, the same negator generally combines with the copula. Alternatively, a new locational negator (\textit{alaɢ} ‘absent’) can be used for locative, dative-possessive and marginally privative-possessive negation. Possession is negated either by a dative-based or a privative-based pattern. Nouns are negated by the former adjective \textit{biš} ‘other’, which usually also negates adjectival predicates and is used with participles for metalinguistic negation. Second person imperatives are almost exclusively negated by a preverbal negator (\textit{bitgii}) which likely goes back to a verb form meaning ‘desist!’. Negative one-word answers can be formed by both \textit{(ü)güi} and \textit{biš}, while lone \textit{bitgii} can be used to turn down somebody else's proposed action. Postverbal negators (\textit{=güi}, \textit{biš}) take scope over preceding elements within the predicate, while there tend to be strong preferences for the position of the negator in sentences with a matrix and subordinate predicate. 
The negation of indefinites is expressed by a focus clitic in combination with clausal negation. 
This paper also addresses negative lexicalizations and other aspects of negation.

\keywords{Standard negation, prohibitive, ascriptive negation, negative derivation, negative modal adjectives}
}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 


\section{The language}\label{sec:kha:1}
This paper is intended as a comprehensive introduction to the negation system of Khalkha Mongolian (ISO: \href{https://iso639-3.sil.org/code/khk}{khk}, Glottocode: \href{https://glottolog.org/resource/languoid/id/halh1238}{halh1238}) as spoken in the Mongolian state. The paper starts out from the author’s general understanding of Mongolian negation, based on existing literature and experience as a second language user. To extend this limited basis, all negative forms within a 134,191 word (18 hours) Spoken Corpus in a partially phonetic transcription (“SC”, \citealt{ZolzargalBrosig2012} [=SC1], \citeyear{ZolzargalBrosig2012-2017} [=SC2]) were annotated and classified. Since this corpus underrepresents several forms that a literate user of Khalkha Mongolian would consider part of her negation system, several individual patterns were additionally evaluated on the basis of a 34,642,687 word Internet Corpus (“IC”, \citealt{OstlingBrosig2011}).
Regardless of whether data is taken from spoken, written or secondary sources, this paper employs a unified transcription trimmed towards phonemic accuracy.\footnote{The correspondences between phonemes and their transcriptions are as follows: ɑ-a, ɔ-o, ʊ-u, e-e, ɵ-ö, ʉ-ü, i-i, j-y, w-w, ʲ-ʲ, p-b, pʰ-p, t-d, tʰ-t, k-g, ɢ-ɢ, kʰ-k, ts-z, tsʰ-c, tʃ{}-ž, tʃʰ{}-č, f-f, s-s, ʃ{}-š, {х-x, m-m, n-n, ŋ-ŋ,} ɮ{}-l, ɬ{}-lh, r-r.} In the bibliography and for citing text other than examples, a strict transliteration of Mongolian Cyrillic is used.\footnote{The Cyrillic letters were transcribed as follows: {а}{}-a, {б}{}-b, {в}{}-v, {г}{}-g, {д}{}-d, {е}{}-e, {ё}{}-jo, {ж}{}-ž, {з}{}-z, {и}{}-i, {й}{}-j, {к}{}-k, {л}{}-l, {м}{}-m, {н}{}-n, o-{о}, {ө}{}-ö, {п}{}-p, {р}{}-r, {с}{}-s, {т}{}-t, {у}{}-u, {ү}{}-ü, {ф}{}-f, {х}{}-x, {ц}{}-c. {ч}{}-č, {ш}{}-š, {ъ}{}-’’, {ы}{}-y, {ь}{}-’, {э}{}-\.e, {ю}{}-ju, {я}{}-ja.} In examples from the non-televised part of SC2, person names are anonymized. Capitalization is retained in IC, but not imposed on SC. The description of negation in Khalkha in this chapter is based on the questionnaire provided by the editors (\citetv{chapters/questionnaire}) and the structure of the chapter follows the model given in the questionnaire.%HC This creates what looks like a mess in capitalization throughout the chapter, but the author doesn't want this to change. Since it is "explained" here, I guess it's OK (eventhough it is a bit against LSP's guidelines). I think we can leave it like this and if LSP isn't OK, they'll let us know.

\largerpage
Khalkha Mongolian is part of the Mongolic language family. Prehistorically, “Common Serbi-Mongolic” split into Proto-Serbi-Avar (including Xianbei, Taghbach, T’u-yü-hun and Kitan, first mentioned for the 1\textsuperscript{st} century CE, linguistically attested from 6\textsuperscript{th} century CE onwards) and Pre-Proto-Mongolic (\citealt[25, 84--85, 116--118, 135--136]{Shimunek2015}). By the early 13\textsuperscript{th} century, it had evolved into Middle Mongol (in the eastern parts of the contemporary Mongolian state), the ancestor of all contemporary Mongolic languages. In the 13\textsuperscript{th}{}\textemdash14\textsuperscript{th} century, Middle Mongol as the language of the Chingisid conquerors was used as an official language in large parts of Eurasia. From there, the Central Mongolic branch evolved that consisting of the demographically major branches Buryat (Southern Siberia), Oirat (Dzungaria, Qinghai, western Mongolia, Kalmykia), Khorchin-Kharchin (western Manchuria) and Khalkha-Chakhar, along with a few smaller varieties (see \citealt{Luvsanvandan1959,Janhunen2006}). Chakhar is spoken in central Inner Mongolia in north China, while Khalkha has historically been spoken in most of the current Mongolian state, excluding its west (Oirat, Kazakh, Tuvinian) and north-east (Buryat). A fraction of Khalkha (“Chakhar”) speakers are found on the Chinese side of the border (cf. \citealt[143]{SvantessonEtAl2005}). 

In the present-day Mongolian state (except for Bajan-Ölgij Province with its Kazakh majority), Khalkha Mongolian is de facto the only official language, restricting Buryat, Oirat, Tuvinian and Mongolian Sign Language to private settings. It is used at all levels of administration, education and media. Foreign movies are overdubbed in Mongolian. The assimilation of other dialects has enriched the vocabulary of literary Khalkha, but this is much less notable for the spoken language. Ulaanbaatar Khalkha has developed an urban slang (\citealt{SambuudorzErdenesuvd2002}). The first foreign language learned in school is now English, and previously was Russian. These are also the most important donor languages for recent loanwords, while previously important donor languages include Turkic (neighboring tribes), Tibetan (religion) and Chinese (trade). Emigration is common, with larger communities in e.g. South Korea, the Czech Republic and Sweden (as manual labor force) and also in USA, Germany and China (as students and subsequent employees).

Khalkha Mongolian has the default word order subject-object-predicate. The relative order of subjects and direct objects is relatively loose, and repair strategies allow for arguments (esp. subjects) that follow the predicate. Subordinate clauses precede matrix clauses. Adverbials precede verbs, attributes (except some quantifiers) precede modified nouns, auxiliaries follow main predicates. All non-reduplicative affixing processes are suffixal and obey vowel harmony. Obligatory verbal suffixes yield three different verb form types: \textsc{Converbs} form adverbial clauses or connect clauses or verbs; finite uses in ellipsis or afterthought are possible to a limited extent. \textsc{Participles} are used as attributes and sentence-finally. Only they and other nominals co-occur with modal particles and postposed negators. \textsc{Finite} \textsc{verbs} can only be used sentence-finally or before the quotative verb \textit{ge}{}-. They are subdivided into declaratives (coding tense and evidentiality) and mood forms (imperatives and hortatives). 

\section{Clausal negation}\label{sec:kha:2}
\subsection{Standard negation}\label{sec:kha:2.1}

Standard negation is marked by the postclitic =\textit{güi} (orthographically <güj>, realized as =[ɡwʉi], =[ɡwiː] or -[ɡʉː]/[ɡʊː]), which attaches to participles, but not to finite verbs (an opposition that will be briefly discussed at the end of this section). It also functions as an existential negator (see \sectref{sec:kha:2.3.2}) and privative case marker (see \sectref{sec:kha:3.3.2}), functions that are older than its predicate-negating use. As a standard negator, it usually attaches to a verbal predicate, as in \REF{ex:kha:1} and \REF{ex:kha:2}. This includes auxiliaries other than the copular auxiliary verb \textit{bai}{}- which are always connected to the main verb by converbal suffixes, e.g. \textit{üz-ež amž-ax=güi} ‘don’t manage (in time) to see’. The negator =\textit{güi} is classified here as a clitic since it still resists vowel harmony quite often (and consistently so in orthography).\footnote{In SC, a majority of critical examples are transcribed as conforming to vowel harmony. However, to ensure that these transcriptions are accurate, one would have to conduct systematic acoustic or perceptual phonetic measurements.}The second reason is that the additive focus clitic =\textit{č} ‘even, also’ can be used in-between participle and negator as in \REF{ex:kha:3}.

\ea\label{ex:kha:1}
    \gll aaa teg-ež ix id-ex=\textbf{küü}. \\ 
         \textsc{inj} do.like.that-\textsc{cvb} much eat-\textsc{npst.ptcp=neg} \\
    \glt ‘Well, I won’t / don’t eat that much.’ (SC2: Z01-1)\footnote{The cited source only contains the Mongolian example, while its glossing and translation were created specifically for this paper.}
\z 

\ea\label{ex:kha:2}
    \gll öglöön-ii xool id-ee=\textbf{güü}. \\ 
         morning-\textsc{gen} meal  eat-\textsc{res.ptcp=neg} \\
    \glt ‘[I] haven’t eaten / didn’t eat breakfast.’ (SC2: Z08-7)
\z 

\ea\label{ex:kha:3}
    \gll bii bol üz-ež amž-aa=č=\textbf{güi}. \\ 
         \textsc{1sg} \textsc{top} see-\textsc{cvb} succeed(in.time)-\textsc{res.ptcp=add.foc=neg} \\
    \glt ‘I have not even managed to see [that movie].’ (SC2: Z08-5)
\z 

The only morphological forms that can be negated by =\textit{güi} as an original existential negator (cf. \citealt{Veselinova2014}) are nominals. The only class of verbal flexional suffixes that allow for this negator are participial markers. There are four participial forms to which negative markers can attach. The negated non-past participle \textit{-x} attached to activity verbs refers to events that at reference time are not prone to happen, including actions that the speaker doesn’t intend to commit as in \REF{ex:kha:1}. With some stative verbs, it refers to the absence of possibility, knowledge etc. as in \REF{ex:kha:4}. Verbs in \textit{-x} cannot be used in absolute sentence-final position without any subsequent morphemes. Accordingly, \textit{-x=güi} in negative sentences roughly corresponds to the suffix \textit{-n} that expresses potential, sometimes habitually instantiated developments in positive sentences (unless it attaches to the copular auxiliary \textit{bai}{}-).
The negated habitual participle \textit{-dag} in \REF{ex:kha:5} indicates that a certain action does not take place habitually, repeatedly or as a disposition, thus merely negating its positive values. While \textit{-x=güi} and \textit{-dag=güi} may often replace one another (like \textit{-n} and \textit{-dag} in positives, see \citealt{Brosig2015Aspect}), they differ in their aspectual perspective, which is a non-occurring habitual development vs. a non-occurring potential development in the present. The negated “resultative participle” \textit{-aa=güi} in \REF{ex:kha:2} indicates that an event has not or not yet happened, while the negated “perfect participle” \textit{-saŋ=güi} indicates that the event didn’t happen at a certain point in the past. However, \textit{-aa=güi} has almost turned into a neutral way of negating past events, while \textit{-saŋ=güi} is now largely restricted to contexts in which the event can no longer take place in a way that would still be meaningful to the interlocutors.\footnote{I partially follow \citet{Hashimoto2007}. However, he analyzes \textit{-aa=güi} and \textit{-saŋ=güi} as prototypes of equal status, thus failing to recognize their disparity in favor of \textit{-aa=güi}. Several native speakers      rejected his (38b) and a few other examples.} %FYI 38b here is not related to anything within the chapter but in Benjamin's source. Nothing to be changed here
In \REF{ex:kha:6}, the elderly speaker with four daughters will not try to produce additional children. In \REF{ex:kha:7}, the TV anchor is introducing a fully prepared report during broadcasting, so he could no longer meaningfully change his mind about how to prepare it. Only \textit{-aa=güi}, but not \textit{-saŋ=güi}, can be used in the scope of \textit{bai}{}- (e.g. in the immediately perceived negated past \textit{-aa=güi bai-n} in \REF{ex:kha:12} below). In addition, =\textit{güi} also negates the non-aspecto-temporal quasi-participles in \textit{-maar} (intention)/\textit{-xaar} (inclination), as in \REF{ex:kha:8}. 

\ea\label{ex:kha:4}
    \gll deegii med-ex=\textbf{küü}. \\ 
         \textsc{name} realize+know-\textsc{npst.ptcp=neg} \\
    \glt ‘Deegii doesn’t know.’ (SC2: Z01-1)
\z 
\ea\label{ex:kha:5}
    \gll bii taxʲaan{}-ii max id{}-deg=\textbf{güü}. \\ 
         \textsc{1sg} chicken-\textsc{gen} meat eat-\textsc{hab.ptcp=neg} \\
    \glt ‘I don’t eat chicken.’ (SC2: Z05-1)
\z 
\ea\label{ex:kha:6}
    \gll uulan=č xüü ɢara-ɢ{}-č-ii ge-ž bod-žii-soŋ. ɢar-saŋ=\textbf{güi}.\\ 
         actually=\textsc{add.foc} son come.out-\textsc{caus-compl-hort} \textsc{quot-cvb} think-\textsc{prog-est.pst} come.out-\textsc{prf.ptcp=neg} \\
    \glt ‘Actually, I was also thinking, “Let’s get a son.” [But] I didn’t get one.’ (SC1: TH, Parik Jagaa)
\z 
\ea\label{ex:kha:7}
    \gll aa bid büxeŋ bol (…) iŋg-ež beld-ii ge-ž bod-soŋ=\textbf{guu}.\\ 
         \textsc{inj} \textsc{1pl} all \textsc{top} (…) do.like.this-\textsc{cvb} prepare-\textsc{hort} \textsc{quot-cvb} think-\textsc{prf.ptcp=neg} \\
    \glt ‘[They usually speak in such a manner on TV programs, the TV programs usually prepare (this kind of interviews.)] Ah, as for our team, we didn’t think, “Let’s (…) prepare it like this.”’ (SC1: Derik, Yanhan1)
\z 
\ea\label{ex:kha:8}
    \gll bii daxʲ-až yum uu-maar=\textbf{guu} bai-n. \\ 
         \textsc{1sg} do.again-\textsc{cvb} thing drink-wish.\textsc{ptcp=neg} \textsc{cop-imm.prs} \\
    \glt ‘I don’t want to drink more.’ (SC2: Z05-5)
\z 

Participles are required for forming negated declarative, attributive and nominalized forms, while positive finite main clauses employ both finite and participial suffixes. For instance, in the past tense there is a distinction (in order of decreasing frequency) between an established past in \textit{-saŋ}, immediate sensory and inferential pasts in \textit{-laa} and \textit{-ž(ee)}, an interrogative-leaning peripheral past in \textit{-w} and a particle-assisted peripheral dubitative past in \textit{-aa} (for which see \citealt{Brosig2015Aspect, Brosig2018, BrosigSkribnik2018}). These distinctions, which among other things relate to evidentiality and time of acquisition, are lost under simple negation and replaced by the two-way distinction between \textit{-aa} and \textit{-saŋ} plus =\textit{güi} that relates to completion. Overall, asymmetric negation in Mongolian entails the common loss of finiteness and TAME distinctions under negation (cf. \citealt{Miestamo2005}).

Putative correspondences between positive and negative forms get even more complicated for aspectually complex constructions. In these, the copular auxiliary \textit{bai}{}- follows a converbal or participial form of the main verb. The positive progressive takes the form \textit{-ž bai-n} > \textit{-žai-n} (for an immediately perceived event) or \textit{-ž bai-ɢaa} > \textit{-žai-ɢaa} (for an event already known to be taking place) as in \REF{ex:kha:9}. But if =\textit{güi} is added directly to this latter form as in \REF{ex:kha:10}, the resulting sentence expresses that its subject has never experienced the event in a period that ranges from the past until the time of speech, i.e. a negated experiential perfect. A better candidate for a negated progressive is shown in \REF{ex:kha:11}, which expresses either that speaker B has observed the subject for a period of time and seen that s/he doesn’t eat or that speaker B is drawing an inference to this effect from currently observable evidence. The marker \textit{-n}, which indicates immediate perception when attached to the copula \textit{bai}{}-, thus has the negation in its scope. On the other hand, the simple state of not having occurred yet as immediately observed by the speaker would be expressed by the pattern \textit{-aa=güi bai-n} as in \REF{ex:kha:12}.\footnote{In contrast to \textit{-aa=güi bai}{}- (n\textsubscript{IC}=14390, including false positives and repetitions), forms like \textit{\nobreakdash-saŋ=güi bai}-(n\textsubscript{IC}=29) and \textit{-saŋ bai}{}-\textsc{ptcp}=\textit{güi} (n\textsubscript{IC}=3 for \textit{bai-ɢaa=güi} and n\textsubscript{IC}=16 for \textit{bai-dag=güi}) don’t play a role in the Khalkha negation system.}

\ea\label{ex:kha:9}
    \gll öröön-d=öö yum id-žai-ɢaa. \\ 
         room-\textsc{dat=rposs} thing eat-\textsc{prog-est.prs} \\
    \glt ‘He is eating something in his room.’ (SC2: Z04-3)
\z 
\ea\label{ex:kha:10}
    \gll xezee=č ard\_tümeŋ (…) cad-tal=aa id-ež bai-ɢaa=\textbf{güi}. \\ %HC I still need to find a solution for these (...). Ehkä [...] ?
         when=\textsc{add.foc} people {} become.satiated-\textsc{cvb(}until)=\textsc{rposs} eat-\textsc{cvb} \textsc{cop-res.ptcp=neg} \\
    \glt ‘Never have the people (…) been eating their fill.’ (IC)
\z 
\ea\label{ex:kha:11}
    \begin{xlist}
    \exi{A: }\label{ex:kha:11A}
        \gll ügü=ee, naa-d=čin   xool=oo \\ 
             no=\textsc{emph} \textsc{2prox.dem-attr=2poss} meal=\textsc{rposs} \\ %HC second word changed by BB, original was 3sg(adr.prox)
        \glt 
    \exi{B: }\label{ex:kha:11B}
        \gll ugaasaa id-ex=\textbf{küü} bai-n. \\ 
              fundamentally eat-\textsc{npst.ptcp=neg} \textsc{cop-imm.prs} \\
        \glt ‘No, s/he doesn’t eat her meal at all (these days / as I currently realize).’ [A starts the sentence, B completes it.] (SC2: Z08-7)
    \end{xlist}
\z 
\ea\label{ex:kha:12}
    \gll Bi zöw ge-ž xel-ee=\textbf{güi} bai-n=l=daa. \\ 
         \textsc{1sg} correct \textsc{quot-cvb}   say-\textsc{res.ptcp=neg} \textsc{aux-imm.prs=add.foc=dp} \\
    \glt ‘I haven’t said (as you should realize) that it’s right/correct.’ (IC)
\z 

The way in which positive and negative TAME forms might, partially, correspond to each other cannot be discussed here in any detail. \textit{Biš} as a negator of main clause predicates is discussed in \sectref{sec:kha:4.1}, \sectref{sec:kha:4.6.2}.


\subsection{Negation in non-declaratives} \label{sec:kha:2.2}

Among Khalkha non-declaratives, imperatives use distinct negators. Khalkha uses three imperative suffixes: plain \textit{-Ø}, imploring \textit{-aač}, and politely requesting \textit{-aarai}. These are negated by the preverbal negation particles \textit{bitgii} or rarely \textit{büü}. The patterns in \REF{ex:kha:13} and \REF{ex:kha:14} illustrate simple and aspectually complex negated imperatives, while \REF{ex:kha:15} illustrates their positive counterparts (though the first sentence in \REF{ex:kha:15} actually fulfills a prohibitive function).

\ea\label{ex:kha:13}
    \gll yanz yanz-iiŋ nöx-düüd-iig \textbf{bitgii} aw-aač=ee. \\ 
         sort sort-\textsc{gen} comrade-\textsc{pl-acc} \textsc{neg} take-\textsc{imp.impl=dp} \\
    \glt ‘Don’t bring along all kinds of guys!’ (SC1: Derik, Yanhan3)
\z 
\ea\label{ex:kha:14}
    \gll \textbf{bitgii} ineedtei yum yarʲ-aad bai-Ø! \\ 
         \textsc{neg} ridiculous thing speak-\textsc{pfv.cvb} \textsc{cop-imp} \\
    \glt ‘Don’t keep on talking nonsense!’ (SC1: Derik, Yanhan3)
\z 
\ea\label{ex:kha:15}
    \gll aa tenegt-eed bai-Ø! tegeed ɢudamž-ruu=ɢaa or-Ø! \\ 
         ah behave.stupidly-\textsc{pfv.cvb} \textsc{cop-imp} then street-\textsc{all=rposs}  enter-\textsc{imp} \\
    \glt ‘Ah, [just] keep on talking nonsense (I dare you)! Then get out onto your street!’ (SC2: Z08-1)
\z 

Positionally, \textit{bitgii} is placed between the subject, if overt, and the predicate (cf. \citealt[82--84]{Yu1991}). In \REF{ex:kha:14}, it precedes an indefinite direct object, while in \REF{ex:kha:13} it follows a specific direct object. \citet[87--88]{Yu1991} claims that prohibitives can be embedded if \textit{bitgii} modifies a case-marked participle, but no such examples were found in IC. \textit{Bitgii} and rarely \textit{büü} are also used in connection with the desiderative suffix -\textit{aasai}, e.g. \textit{bitgii exl-eesei} 'may it not begin' (IC). First and third person hortatives (\textit{-ii}/\textit{-y=aa}, \textit{-g}) don’t seem to combine with \textit{bitgii}, but \textit{büü} V-\textit{ii} occurs in a few rare idioms. The intention of a first person not to commit an action can be expressed by a negated nonpast participle in the scope of \textit{bai-y} as in \REF{ex:kha:16}. 

\ea\label{ex:kha:16}
    \gll Getel ter=maan genet xoy-uul=aa {ger bül} bol-tl=oo seks xii-x=\textbf{güi} bai-y ge-seŋ. \\ 
          but that=\textsc{stc} suddenly two-\textsc{coll=rposs} family become-\textsc{npst.ptcp=neg} sex do=\textsc{npst.ptcp} \textsc{aux-hort} \textsc{quot-est.pst} \\
    \glt ‘But suddenly s/he said: “Let’s not have sex until we marry.”’ (IC)
\z 

The old imperative negator \textit{büü} is rarely used, occasionally showing up in solemn exhortations like \textit{büü mart-aarai} ‘Don’t forget!’ or \textit{büü ai-Ø} ‘Don’t fear!’. It is the only option in the idiomatic “pseudo-prohibitive” pattern \textit{büü med-Ø} ‘no clue’ (cf. \citealt[73--74]{Yu1991}).

Negated interrogatives, by contrast, are participles modified by negators and question particles, i.e. they use the same type of negator as declaratives. Functionally, negated questions in \textit{=güi=yüü} cover a range from exhortations as in \REF{ex:kha:17} to guesses and rhetorical questions. Questions formed with \textit{biš=üü} ‘isn’t it so that’ (based on the ascriptive negator \textit{biš}) cover a range from guesses as in \REF{ex:kha:18} to rhetorical questions as in \REF{ex:kha:19} (cf. \citealt[14, 18]{Bjambasan2001}).

\ea\label{ex:kha:17}
    \gll nem-eed piiw uu-č-x=\textbf{güi}=yüü? aay \\ 
         \textsc{add-pfv.cvb} beer drink-\textsc{compl-npst.ptcp=neg=q.plr} \textsc{inj} \\
    \glt ‘Drink some more beer, won’t you?’ (SC2: Khalkha0018)
\z

\ea\label{ex:kha:18}
    \gll či xarʲaŋ tur-c=iin \textbf{biš=üü}? \\ 
         \textsc{2sg} but become.thin-\textsc{compl.est.pst=asrt} \textsc{neg=q.plr} \\
    \glt ‘But haven’t you become thinner?’ (SC2: Z08-7)
\z 
\ea\label{ex:kha:19}
    \gll en xarʲaŋ   ɢoy yum \textbf{biš=üü}? \\ 
          this but nice thing/\textsc{asrt} \textsc{neg=q.plr} \\
    \glt ‘But ain’t this (a) nice (thing)!’ (SC2: Z08-9)
\z 

One common way of forming alternative questions is to combine a regular polar question with \textit{(ü)güi}, here still used as a clause-forming negative particle, plus the interrogative clitic =\textit{uu} as in \REF{ex:kha:20}. Instead of \textit{(ü)güi}, full verbs negated by =\textit{güi} can sometimes be used. \textit{Güi=yüü} is also used when following up a declarative with a tag question as in \REF{ex:kha:21}.

\ea\label{ex:kha:20}
    \gll budaa xii-xžii-n=uu \textbf{güi=yüü}? \\ 
         rice make-\textsc{prosp-imm.prs=q.plr} \textsc{neg=q.plr} \\
    \glt ‘Are you about to prepare rice or not?’ (SC1: TH, Ganbaatar)
\z 
\ea\label{ex:kha:21}
    \gll amʲdral saixaŋ bai-n \textbf{güi} \textbf{yuu}? \\ 
         life beautiful \textsc{cop-imm.prs} \textsc{neg} \textsc{q.plr} \\
    \glt ‘Life is beautiful, isn’t it?’ (SC1: TH, Ganbaatar)
\z 

 
\subsection{Negation in stative predications}\label{sec:kha:2.3}
\subsubsection{Ascriptive predications} \label{sec:kha:2.3.1}

In stative non-temporal non-verbal ascriptive predications (cf. \citealt[110--111]{Veselinova2013neg.exist}, following \citealt{Hengeveld1992}), the only negator is \textit{biš} (< ‘other’). It denies nominal predications such as identity in \REF{ex:kha:22} or proper inclusion. In positive predications, \textit{biš} corresponds to zero, as in the second sentence of \REF{ex:kha:22}. Identity predications occasionally employ the existential-identificational particle \textit{mön} ‘same’ to confirm identity, but this in turn can be denied by \textit{mön} \textit{biš} ‘not the same’.

\ea\label{ex:kha:22}
    \gll aan, eŋxbayar \textbf{biš}. en pürewžaw. \\ 
         ah \textsc{name} \textsc{neg} this \textsc{name} \\
    \glt ‘Ah, [it] is not Enkhbayar. It is Purevjav.’ (SC2: Khalkha0001)
\z 

For negating generic or permanent adjectival property predications, \textit{biš} is the most common negator. It denies that a proposition holds, for whatever reason, as in \REF{ex:kha:23}. Relatively rarely, =\textit{güi} is used instead of \textit{biš}. It signals that a property is absent or present to a lesser extent (\citealt[43]{Sarangguua2007}) as in \REF{ex:kha:24}. This pattern is skewed heavily towards certain stems, e.g. it is much more frequent with \textit{saiŋ} ‘good’ (n\textsubscript{IC}=256) than with \textit{saixaŋ} ‘beautiful’ (n\textsubscript{IC}=8) or \textit{ɢoy} ‘nice, pretty’ (n\textsubscript{IC}=0).

\protectedex{\ea\label{ex:kha:23} 
    \begin{xlist}
        \exi{BB:}\label{ex:kha:23a}
            \glt ‘Gambling is bad.’ \\
        \exi{A:}\label{ex:kha:23b}
            \gll ügü=ee, muu \textbf{biš=ee}. \\ 
                 \textsc{neg=emph} bad   \textsc{neg=emph} \\
            \glt ‘No, it is not bad (at all) [but more exciting].’ (SC2: Khalkha0017)
    \end{xlist}
\z }
\ea\label{ex:kha:24}
    \gll man-ai xos bid xoyor bas muu=\textbf{güi} šüü.  \\ 
         \textsc{1pl-gen} pair \textsc{1pl}   two also bad=\textsc{neg} \textsc{dp} \\
    \glt ‘[You think you can take the price?] The two of us are not bad either.’ (IC)
\z 


In predications that overtly specify aspect, tense, time of acquisition or evidentiality, non-verbal predicates are followed by the copular auxiliary verb \textit{bai}{}-. In this setting, \textit{biš} or (only with adjectives) \textit{=güi} may attach to the nominal, so that a negative state is located in time (or qualified epistemologically) as in \REF{ex:kha:25a}. Alternatively, the copula verb can be negated. This is normally done by the standard negator \textit{=güi}  as in \REF{ex:kha:25b}, claiming that a positive situation did not occur at the specified point in time. Alternatively, negating the copula with \textit{biš} would deny a potential conceptualization of the situation as accurate and point to an alternative as in \REF{ex:kha:25c} (cf. \sectref{sec:kha:4.1}, \sectref{sec:kha:4.6.2}).

\ea\label{ex:kha:25}
   \ea\label{ex:kha:25a}
        \gll Xarʲaŋ bi beleŋ \textbf{biš} bai-saŋ yum. \\ 
             but \textsc{1sg} ready \textsc{neg} \textsc{cop-est.pst} \textsc{asrt} \\
        \glt ‘But I was not ready.’ (IC)
    \ex\label{ex:kha:25b}
        \gll Tüüx-eŋ en öörčlölt-öd bidn-ii xeŋ=č beleŋ bai-ɢaa=\textbf{güi}.\\ 
             history-\textsc{attr} this   change-\textsc{dat} \textsc{1pl-gen} who=\textsc{add.foc} ready \textsc{cop-res.ptcp=neg} \\
        \glt ‘For this historical change, none of us were ready (yet).’ (IC)
    \ex\label{ex:kha:25c}
        \gll Odoo-giiŋ-x   šig büx yum beleŋ bai-saŋ \textbf{biš}. \\ 
             now-\textsc{gen-nmlz} like all thing ready \textsc{cop-prf.ptcp} \textsc{neg} \\
        \glt ‘[We would write our play. For every role, we would make a picture of the intended clothes and sew it.] It is not that all things were readily available, like it is today.’\footnote{Ž. Tujaac\.ec\.eg: Zarimdaa jamar ix t\.evč\.e\.ert\.ej jum b\.e g\.ež boddog. Interview with the wife of the high-ranking politician M. \.Enxbold) \url{http://www.olloo.mn/n/43845.html}, online 2017-6-20, retrieved 2019-3-16}
    \z 
\z 


\subsubsection{Locative, possessive and existential predications} \label{sec:kha:2.3.2}

In Khalkha, the locative pattern is also used as the main possessive pattern which is preferably used with topicalized possessa. In \REF{ex:kha:26}, both the “existential” {\textasciitilde} inverse locational (cf. \citealt[1]{Creissels2014}) interpretation in \REF{ex:kha:26a} and the possessive interpretation in \REF{ex:kha:26b} are possible since the dative-marked noun can be interpreted as either a location (a shop as a place) or a possessor (a shop as a corporation). In \REF{ex:kha:27}, the spatial demonstrative can only bear locational meaning, while the dative-marked person pronoun in \REF{ex:kha:28} is normally conceptualized as a possessor. In all of these examples, a participial form of the copula \textit{bai}{}- is negated, the possessed or located entity is in the unmarked nominative, and the possessor/location receives dative-locative case. 

\ea\label{ex:kha:26}
        \gll Delgüür-t dawsn-aas öör yum {\textbf{bai-x=güi}...} \\ 
             shop-\textsc{dat}   salt-\textsc{abl} other thing \textsc{cop-hab.ptcp=neg} \\
    \ea\label{ex:kha:26a}
        \glt ‘In the shop, there is nothing but salt.’ (IC, decontextualized) 
    \ex\label{ex:kha:26b}
        \glt ‘The shop doesn’t have ({\textasciitilde}sell) anything but salt.’
    \z 
\z 
\ea\label{ex:kha:27}
    \gll Aztaŋ bol-soŋ Xišigsaixaŋ=maan odoo-goor end \textbf{bai-x=güi}[.]  \\ 
         lucky.winner become-\textsc{prf.ptcp} \textsc{name=1poss} now-\textsc{ins} here \textsc{cop-npst.ptcp=neg} \\
    \glt [A photo shows a woman around 55 standing in front of a rack with new shoes and holding a paper bag with something inside. The text below reads:] ‘The winner, Khishigsaikhan, is currently not here. [Her mother came and took her present. Congratulations again!]’\footnote{\href{https://www.globuya.com/MN/Dalan-Zadagad/Өмнөговь-"emerald"-Гутлын-Дэлгүүр}{https://www.globuya.com/MN/Dalan-Zadagad/Өмнөговь-"emerald"-Гутлын-Дэлгүүр}, retrieved 2020-2-13}
\z 

\ea\label{ex:kha:28}
    \begin{xlist}
    \exi{A: }\label{ex:kha:28A}
        \gll čam-d tamaɢ bai-x=\textbf{kuu} biz=dee? \\ 
             \textsc{2sg-dat} ace \textsc{cop-npst.ptcp=neg} \textsc{dp=dp} \\
        \glt  ‘You don’t have an ace, do you?’   
    \exi{B: }\label{ex:kha:28B}
        \gll bai-x=\textbf{kuu}. \\
            \textsc{cop-npst.ptcp=neg} \\
        \glt ‘No, I don’t.’ (SC2: Khalkha0018)
    \end{xlist} 
\z

Instead of the nonpast participle \textit{baixgüi}, it is also possible to use the young locative-possessive negator \textit{alaɢ} ‘absent/missing’ as in \REF{ex:kha:29}. As a predicate, it indicates that an entity is absent despite the speaker’s expectation or wish, that the entity is temporarily absent or that an intention is absent (see the excellent functional account of \citealt{Umetani2004}, summarized in English in \citealt[97--98]{Brosig2015Negation}; for its morphology, see \citealt{Monx-Amgalan1999}). The old existential negator \textit{ügüi} is sometimes still used if preceded by the additive focus particle as in \REF{ex:kha:30}, but without a clitic, as in \REF{ex:kha:31}, this pattern is exceedingly rare. To convey non-existence without regard for location, speakers often add a converbal form of the quotative verb \textit{ge}{}- in its naming function as in \REF{ex:kha:32}. However, this is not required if the non-existing entity is designated by a complex attribute as in \REF{ex:kha:33}. The positive equivalents of \REF{ex:kha:26} and \REF{ex:kha:32} would use habitual \textit{bai-dag}, and (\ref{ex:kha:27}--\ref{ex:kha:29}) would use established present \textit{bai{}-ɢaa}, without \textit{yamar=č} in \REF{ex:kha:29}.

\ea\label{ex:kha:29}
    \gll nadad yamar=č zaw \textbf{alaɢ}. \\ 
         \textsc{1sg.dat} what.kind=\textsc{add.foc} free.time absent \\
    \glt ‘I don’t have any time at all.’ (SC1: Derik, Yanhan2)
\z 
\ea\label{ex:kha:30}
    \gll … üül=\textbf{č} \textbf{ügüi}, boroo=\textbf{č} \textbf{ügüi} … \\ %HC ...
         {} cloud=\textsc{add.foc} \textsc{neg} rain=\textsc{add.foc} \textsc{neg} \\
    \glt ‘… there were no clouds and there was no rain …’ (\citealt{Bjambasan2001}: 9, quoting the writer D. Nacagdorž)
\z 
\ea\label{ex:kha:31}
    \gll Xorwoo delxii-d tüün-ees=min saiŋ, saixaŋ xüŋ \textbf{ügüi} bai-saŋ.\\ 
         (fleeting)world globe-\textsc{dat} that-\textsc{abl-1poss} good   beautiful person \textsc{neg} \textsc{cop-est.pst}\\
    \glt ‘In the [entire] world, there was no better, more wonderful person than him/her.’ (IC)
\z 
\ea\label{ex:kha:32}
    \gll burxaŋ ge-ž bai-x=\textbf{güi}. \\ 
         god \textsc{quot-cvb} \textsc{cop-npst.ptcp=neg} \\
    \glt ‘There is no god.’ (lit. ‘What is called “God” is not.’) (IC)
\z 
\ea\label{ex:kha:33}
    \gll man-ai dorž-iig güic-ex xüŋ \textbf{bai-x=gü=ee}. \\ 
         \textsc{1pl-gen} \textsc{name}-\textsc{acc} match-\textsc{npst.ptcp} person \textsc{cop-npst.ptcp=neg=emph} \\
    \glt ‘There’s nobody who matches our Dorj [in card games].’ (SC2: Z08-7)
\z 

Possession can also be expressed by a construction that uses nominative possessors. Possessees are marked by either comitative-possessive or privative case as in \REF{ex:kha:34}. The use of \textit{alaɢ} in this pattern, which for most speakers requires dative-marked possessors, is apparently non-standard, but not uncommon if a dummy noun \textit{yum} with a preceding participial construction is the possessed, as in \REF{ex:kha:35} (see also \citealt[101]{Brosig2015Negation}).


\ea\label{ex:kha:34}
    \gll öör=öö möŋg=\textbf{güü}, ör-\textbf{tei} yuu ta? \\ 
         self\textsc{=rposs} money-\textsc{priv} debt-\textsc{com} \textsc{q.plr} \textsc{2sg.hon} \\
    \glt ‘Are you without money and with debts?’ (SC1: TH, Parik Jagaa)
\z 
\ea\label{ex:kha:35} 
    \gll tan-ai cai wol ünen-d=ee (.) yarʲ-ax yum alaɢ \\ 
         \textsc{2sg-gen} tea \textsc{top} truth-\textsc{dat=rposs} {} speak-\textsc{npst.ptcp} thing absent \\
    \glt ‘There is truly nothing [bad] to say about your tea.’ (SC1: TH, Ganbaatar)
\z 

 
\subsection{Negation in non-main clauses}\label{sec:kha:2.4}
\subsubsection{Attributive and argument clauses}\label{sec:kha:2.4.1}

In attributive and nominalized clauses, the distinction between the negated incomplete past in \textit{-aa=güi} and the negated completed past in \textit{-saŋ=güi} is lost. Only forms in \textit{-aa=güi} are used as attributes as in \REF{ex:kha:36} or as nominalized argument clauses as in \REF{ex:kha:37}. Habitual \textit{-dag=güi}, nonpast \textit{-x=güi}, \textit{-xaar=güi} ‘not likely to’ and nominal attributes in \textit{=güi} are common, too. 
Contrastive attributes with adjectival, participial etc. predicates negated by \textit{biš} as in \REF{ex:kha:38} are rare, and attributes in \textit{biš} without a contrasting positive form are even less common. For adjectives and nouns, this latter function is sometimes fulfilled by \textit{bus} (see \sectref{sec:kha:3.3.1}). Direct nominalizations of \textit{biš} as in \REF{ex:kha:39} are very rare.

\ea\label{ex:kha:36}
    \gll tan-ii xairl-aa=\textbf{kuu} xüüxed tan-iig ilüü xairal-žii-n=aa. \\ 
         \textsc{2sg.hon-gen} love-\textsc{res.ptcp=neg} child \textsc{2sg.hon-acc} more love-\textsc{prog-imm.prs=hon} \\
    \glt ‘The child that you didn’t love is loving you more [than your other son].’ (SC2: Z04-2)
\z 
\ea\label{ex:kha:37}
    \gll Xʲatad-ad bai-dag naiz-uud   ir-ee=\textbf{güi}-g med-ež   bai-ɢaa. \\ 
         China-\textsc{dat} \textsc{cop-hab.ptcp} friend\textsc{{}-pl} come-\textsc{res.ptcp=neg-acc} know-\textsc{cvb} \textsc{cop-est.prs}\\
    \glt ‘[I] know that the friends who are in China have/did not come.’ (IC)
\z 
\ea\label{ex:kha:38}
    \gll Muu \textbf{biš}, saiŋ tal-iiŋ šüümžlel   erg-eed ür\_düŋg=ee ög-dög yum bai-n. \\ 
         bad \textsc{neg} good side-\textsc{gen} critique turn-\textsc{pfv.cvb} result=\textsc{rposs} give-\textsc{hab.ptcp} \textsc{asrt} \textsc{cop-imm.prs} \\
    \glt ‘Not bad, but good critique gives results in return.’ (IC)
\z 
\ea\label{ex:kha:39}
    \gll Bi tiim=č xel am-tai, eldew aaš-tai büsgüi \textbf{biš} šüü. Xöörxöŋ \textbf{biš}-iig ud-ax=güi xar-aarai za.\\ 
         \textsc{1sg} such=\textsc{add.foc} tongue mouth-\textsc{com} various character-\textsc{com} woman \textsc{neg} \textsc{dp} cute \textsc{neg-acc} take.long-\textsc{npst.ptcp=neg} see-\textsc{imp(hon)} \textsc{tag} \\
    \glt ‘I’m really not such a gossipy, ill-tempered woman. [But] take a look at [the one who] is not pretty [=me] soon, okay?’ (IC)
\z 
 

\subsubsection{Negation in adverbial clauses}\label{sec:kha:2.4.2}
\largerpage[2]

Khalkha Mongolian has inherited a large number of converbs. Due to the demise of preverbal negation, these cannot be negated directly, but a number of strategies have evolved that circumvent this issue. Firstly, the converbal suffix may attach to the copula \textit{bai}{}-, while the negator attaches to the main verb as in \REF{ex:kha:40a}, as compared to positive \REF{ex:kha:40b}. This entails specific aspectual meaning, namely reference to an ongoing time period. Secondly, aspectually neutral subordinating conjunctions as in \REF{ex:kha:41} may be used. Conjunctions are preceded by the same set of negated attributive forms that modify nouns (see \sectref{sec:kha:2.4.1}).\footnote{Historically, conjunctions evolved from verbs (\textit{bol} ‘if’ < \textit{bol-bol} 'become-\textsc{cond.cvb}', via haplology), nouns (\textit{učr-aas} ‘because’ < 'reason-\textsc{abl}') or are intransparent (\textit{tul} < \textit{tula} ‘because’).}  In rare cases that involve conditionals, writers may also choose to retain the old preverbal negator (see discussion on \textit{es} in \sectref{sec:kha:3.3.1}). Finally, some forms cannot be negated. This includes converbs in \textit{-ŋɢuut} as in \REF{ex:kha:42} which express that the matrix clause action takes place as soon as the subordinate clause action occurs. The progressive-stative interpretation of using a negated participle with \textit{bai}{}- is incompatible with its punctual meaning (hence *V-\textsc{ptcp}=\textit{güi bai-ŋɢuut} is unattested). No conjunctions that involve this medium-frequency suffix seem to have evolved. 

\ea\label{ex:kha:40}
    \ea\label{ex:kha:40a}
        \gll usl-ax-\textbf{guu} {bai-wal} (…) ösölt=ön zogs-n=oo=doo. \\ %HC ...
             water-\textsc{npst.ptcp-neg} \textsc{cop-cond.cvb} {} growth=\textsc{3poss} stop-\textsc{pot=hon=dp} \\
        \glt ‘If you are not watering it (over a period of time), its growth will stop.’ (SC1: BTV, 16 nemex)
    \ex\label{ex:kha:40b}
        \gll usal-ž bai-wal  \\ 
             water-\textsc{cvb} \textsc{cop-cond.cvb} \\
        \glt ‘if you are watering it’ (constructed)
    \z 
\z 
\ea\label{ex:kha:41}
    \gll uulz-ax-\textbf{guu} bol al-an xʲad-an. \\ 
         meet-\textsc{npst.ptcp-neg} if kill-\textsc{pot} massacre-\textsc{pot} \\
    \glt ‘If [you] don’t meet up [with us], [we] will kill [you].’ (SC1: Derik, Gazar) \\ cf. \textit{uulz-wal} … ‘if [you] meet up [with us]’ (constructed) 
\z 
\ea\label{ex:kha:42}
    \gll utɢ-ii=n ol-oŋɢuut=aa nad-ad šuud xel-žai-ɢaa baixkuuyuu.\\ 
         meaning-\textsc{acc=3poss} find-\textsc{cvb}(as.soon.as)=\textsc{rposs} \textsc{1sg-dat} direct {say-}\textsc{prog-est.prs} \textsc{dp} \\
    \glt ‘Once [he] understands its meaning, [he] directly tells me, you see.’ (SC2: Z08-7)
\z 

There is also a special negative converbal suffix \textit{-lgüi}. Diachronically, it consists of the deverbal noun-forming suffix \textit{-l} (e.g. \textit{medr}{}- ‘feel’ > \textit{medr-el} ‘perception’) and the standard negator in its (older) use as a privative case marker =\textit{güi}, but it can be formed from verbs to which simple \textit{-l} does not attach and it cannot be used attributively \citep{Bjambasan1985}. It expresses that one action took place without another action being undertaken as in \REF{ex:kha:43}. Its use is partially similar to the negated nonpast participle \textit{-x=güi} in its adverbial use, shown in \REF{ex:kha:44}, which specifies that the subordinate action was not started throughout the main action. The adverbial status of a predication is sometimes clarified by attaching the instrumental case to participles as in \REF{ex:kha:45}, nouns, or even unambiguous converbs in \textit{-lgüi}. The precise circumstances under which participles negated by =\textit{güi} can be used as adverbials are not fully clear.

\ea\label{ex:kha:43}
    \gll ertn-ii en urlag (…)   žiŋxeŋ  utɢ=aa ald-\textbf{algüi} bidn-ii önöö-giiŋ cag-t ulamžl-agd-aŋ ir-seŋ. \\ 
         early-\textsc{gen} this art {} real meaning=\textsc{rposs} lose-\textsc{neg.cvb} \textsc{1pl-gen} now-\textsc{gen} time-\textsc{dat} inherit-\textsc{pass-mnr.cvb} come-\textsc{est.pst} \\
    \glt ‘This old art form has (…) been passed down to us without losing its real meaning.’ (SC1: BTV, 16 nemex)
\z 
\ea\label{ex:kha:44}
    \gll en (…) šön   unt-uul-\textbf{kuu} uil-aad bai-x=iin. \\ 
         this {} night sleep-\textsc{caus-npst.ptcp.neg} cry-\textsc{pfv.cvb} \textsc{cop-npst.ptcp=asrt} \\
    \glt ‘S/he keeps on crying all night without [ever] allowing [you] to sleep.’ (SC2: Z07-6)
\z 
\ea\label{ex:kha:45}
    \gll Texeer en bügd-iig xel-\textbf{küü-geer} unt-až bol  čad-kuu=n bai-n=aa.\\ 
         thus this all-\textsc{acc} say-\textsc{npst.ptcp.neg-ins} sleep-\textsc{cvb} \textsc{top} can\textsc{-npst.ptcp.neg} \textsc{cop-imm.prs=hon}\\
    \glt ‘Accordingly, I feel that I cannot sleep without telling [you] all this.’ (SC1: TH, Badaruugan)
\z 

In a closely related pattern, negated participles collocate with several auxiliary-like matrix verbs, depicting the state of the subject during the passage of time. For resultative participles, the most common of these quasi-auxiliaries is \textit{ud}{}- ‘take time, endure’ as in \REF{ex:kha:46}, while for nonpast participles it is \textit{yaw}{}- ‘go (forth, on etc.)’.  

\ea\label{ex:kha:46}
    \gll bii yöröösöö uu-ɢaa-\textbf{guu} ud-žii-n=aa. \\ 
         \textsc{1sg} at.all drink-\textsc{res.ptcp-neg} take.time-\textsc{prog-imm.prs=hon} \\
    \glt ‘I’ve not been drinking at all for a long time.’ (lit. ‘[It] is taking a long time with I not drinking at all.’) (SC2: Z05-3)
\z 
 
Negated (non-clausal) adjectival adverbials seem to be dispreferred, but if they are formed, the negator is mostly =\textit{güi} as in \REF{ex:kha:47} rather than \textit{biš}. 

\ea\label{ex:kha:47}
    \gll Xüite-wter us (...) caŋɢaa-g muu=\textbf{güi} tail-dag. \\ 
         cold-\textsc{dim} water {} thirst-\textsc{acc} bad=\textsc{neg} relieve.of-\textsc{hab.ptcp} \\
    \glt ‘Coldish water quenches thirst quite well.’ (IC)
\z 
 
\subsection{Negative lexicalizations}\label{sec:kha:2.5}

Among apparent lexicalizations of negative concepts, the only word that does not contain a known negator (besides \textit{alaɢ} ‘absent’, see \sectref{sec:kha:2.3.2}) is the verb \textit{yad}- ‘be unable to, to bungle something up’ (< ‘be in dire straits’, cf. \textit{yaduu} ‘poor’). However, \textit{yad}{}- is restricted to a fixed set of collocations, of which only \textit{üz-eŋ yad}{}- ‘cannot see > hate’ is common. Other collocations such as \textit{tes-č yad-} ‘cannot endure’ or \textit{učr=aa} \textit{ol-ž yad-} ‘cannot find one’s cause’ are confined to written style. 

\section{Non-clausal negation}\label{sec:kha:3}
\subsection{Negative replies}\label{sec:kha:3.1}

When providing an informative answer to regular open [and, in particular, alternative] questions, it is possible to answer with a verb echoing the question, either in positive or negated form (cf. \citealt[21--22]{Guntsetseg2016}), and with or without additional arguments. These answers still function as assertions of their own right, since the addressee is free to switch aspectual forms, in this case answering not that he doesn’t drink habitually, as asked, but that he doesn’t drink at all. Alternatively, positive items such as adjectival \textit{tiim(ee)} and its interjectional forms \textit{te} {\textasciitilde} \textit{tx} ‘such’ as well as \textit{mxm} (predominantly used for signaling attentiveness) or negative items like the existential adjective into particle \textit{(ü)güi} can be used in interjection-type answers, particularly to confirm \citep[94--95]{Binnick2012} or contradict presumptive questions, as in \REF{ex:kha:49}, or affirm or reject assertions, as in \REF{ex:kha:50}. While a proper quantitative comparison of answer strategies in SC cannot be undertaken in this paper, some preliminary checks suggest that interjection-type patterns such as \textit{güü} in \REF{ex:kha:49b} and \textit{ügüü} in \REF{ex:kha:50b} are overall more frequent as compared to the verb in \REF{ex:kha:48b} are overall more frequent. A similar observation has been offered by \citealt{EnfieldEtAl2019} for most languages of their convenience sample.


\ea\label{ex:kha:48}
    \ea\label{ex:kha:48a}
        \gll arʲax uu-dag=uu? \\ 
             liquor drink-\textsc{hab.ptcp=q.plr} \\
        \glt ‘Do you drink liquor?’
    
    \ex\label{ex:kha:48b}
        \gll uu-x=\textbf{güi} \\
             drink-\textsc{npst.ptcp=neg} \\
        \glt ‘No.’ (cf. \textit{uu-dag}: ‘Yes.’) (SC1: TH, Ganbaatar)
    \z
\z 
\ea\label{ex:kha:49}
    \ea\label{ex:kha:49a}
        \gll ter (.) baicaa=n xʲamd-xaŋ bai-dag biz=dee? \\ 
             that {} cabbage=\textsc{3poss} cheap-\textsc{dim} \textsc{aux-hab.ptcp} \textsc{mp=dp} \\
        \glt ‘That cabbage is usually cheap, right? [I claim, you will agree?]’
    
    \ex\label{ex:kha:49b}
        \gll \textbf{güü}, xarʲaŋ ün-tei wai-n. xil=en dörwöm mʲaŋɢ. \\
             no but price-\textsc{com} \textsc{aux-imm.prs} kg=\textsc{3poss} four thousand \\
        \glt ‘No, but [just the opposite], it’s expensive. One kilo of it is 4000.’ (SC2: Z01-1)
    \z
\z 
\ea\label{ex:kha:50}
    \ea\label{ex:kha:50a}
        \gll bii   kolon bol-on ge-ž yöröösöö yarʲ-aa-\textbf{guu}=šdee. \\ 
             \textsc{1sg} colony become-\textsc{pot} \textsc{quot-cvb} on.the.whole say-\textsc{res.ptcp-neg=dp} \\
        \glt ‘I didn’t say at all that [Mongolia] would become a colony [then].’
    
    \ex\label{ex:kha:50b} 
        \gll \textbf{ügüü}=šdee. say či tex-s=ii=šd. \\
             no=\textsc{dp} recently \textsc{2sg} do.like.that-\textsc{est.pst=asrt=dp} \\ 
        \glt ‘No, you just said so.’ (SC2: Z08-7)
    \z
\z 

Orders, requests or suggestions to act can be accepted by \textit{za} ‘well’ (see \citealt{Purev-Ocir2009}) or declined by \textit{(ü)güi} as in \REF{ex:kha:51}, and sole \textit{bitgii} as in \REF{ex:kha:52} can be used to prevent the interlocutor from acting. If the goal is to correct the addressee and give an unmediated alternative explanation, \textit{biš} (< ‘other’) can be used instead, as in \REF{ex:kha:53}. The interjection used for negative replies and a subsequent negative explication don’t have to use the same form: In \REF{ex:kha:54}, the speaker objects to the addressee’s statement using \textit{biš}, but follows it up with an assertion in standard negation.

\ea\label{ex:kha:51}
    \ea\label{ex:kha:51A}
        \gll bii xamt xii-č-ii! \\ 
             \textsc{1sg} together do-\textsc{compl}{}-\textsc{hort} \\
        \glt ‘I shall prepare it together [with you]!’
    
    \ex\label{ex:kha:51B}
        \gll \textbf{güi} \textbf{bitgii} xamt xii-Ø! \\
             no \textsc{neg} together do-\textsc{imp} \\
        \glt ‘No, don’t prepare it together [with me]!’ (SC1: TH, Parik Jagaa)
    \z
\z 

\ea\label{ex:kha:52}
    \gll \textbf{Ügüi}, yaa-ž teg-ež bol-dog yum!? \textbf{Bitgii}! Či=minʲ taiwšir-Ø! \\
         no do.what-\textsc{cvb} do.like.that-\textsc{cvb}  may-\textsc{hab.ptcp} \textsc{asrt} \textsc{neg} \textsc{2sg=1poss} become.calm-\textsc{imp}\\ 
    \glt ‘No, how could one do such a thing!? Don’t! Calm down, my dear!’\footnote{Xolboony surguulijn ulaan d\.e\.elt xuv’ zajaagaa urlagtaj xolboson tüüx (interview with actress Lxamžavyn Dolgor shortly after her 90th birthday by L.D\.emidbaatar) \url{http://entertainment.mongolnews.mn/Архив/tdq}, published 2012-12-19, retrieved 2019-03-09}
\z 

\ea\label{ex:kha:53}
    \ea\label{ex:kha:53Aa}
        \gll nöxör-tei yum=uu   či? \\ 
             husband-\textsc{com} \textsc{asrt=q.plr} \textsc{2sg} \\
        \glt ‘Do you [thus] have a husband?’
    
    \ex\label{ex:kha:53B}
        \gll \textbf{biš=ee}, man-ai naiz. \\
             no=\textsc{emph} \textsc{1pl-gen} friend \\
        \glt ‘No, [this is just] my friend.’ (SC1: Derik, Yanhan2)
    \z
\z 

\ea\label{ex:kha:54}
    \begin{xlist}
    \exi{BB: }\label{ex:kha:54a}
        \glt  ‘Now I can play a five!’
    \exi{A: }\label{ex:kha:54b}
        \gll \textbf{biš=ee}, iŋg-ež ɢar-\textbf{kuu}. \\
         no=\textsc{emph} do.like.this-\textsc{cvb} come.out-\textsc{npst.ptcp.neg} \\
        \glt ‘No, you don’t play [your cards] like this.’ (SC2: Khalkha0017)
    \end{xlist}
\z  

When answering negative questions, the responder seems to use different strategies in different communicative situations. In several conversational contexts, the responders match the polarity of the question in their answers. For instance, to a biased question intended to elicit a positive answer, the responder can retain her position via denial using \textit{(ü)güi} as in \REF{ex:kha:55} or concede with an answer like \textit{xarʲaŋ teg-seŋ} ‘indeed I did so’. This encompasses negative questions with \textit{biš=üü} ‘isn’t it so that’ as in \REF{ex:kha:56}. If the question is meant to suggest a course of action e.g. for a third person, the responder can express disagreement via \textit{(ü)güi} or agree by forms like \textit{tii=š=d} ‘precisely’. However, the situation is different if the asker is perplexed between two conflicting pieces of information and uses a more or less open negative question to seek confirmation for the newly acquired position that the event did actually not happen. In this case, the responder may agree with the asker’s underlying negative presupposition by using plain \textit{tiiŋ} ‘yes, in this way’ as in \REF{ex:kha:57}, while using \textit{(ü)güi} to confirm the negation would be precariously ambiguous, though this could be resolved by adding a negated verb form like \textit{med-küü}. If the responder instead wants to contest the negative proposition of the asker, she can immediately answer with assertive forms like reduplicated \textit{med-n=ee med-en} ‘[they] know it alright’ or \textit{xarʲaŋ med-en} ‘yeah, [they] indeed know it’. If the responder considers the asker’s doubt as unjustified, a sharper, more authoritative answer would be \textit{med-ii=š=d}.

\ea\label{ex:kha:55}
    \begin{xlist}
    \exi{A:} ‘How come you brand people as thieves?’
    \exi{B:}\label{ex:kha:55a} ‘Brand people as thieves?’ \\
        \gll ta say end-ees xulɢai xii-gee=\textbf{güi} \textbf{yuu}? \\ 
             \textsc{2sg.hon} recently here-\textsc{abl} theft do-\textsc{res.ptcp} \textsc{q.plr} \\
        \glt ‘Have you not just stolen [something] from here?’
    \exi{A:}\label{ex:kha:55b}
        \gll \textbf{gü=ee}, zügeer yum sonʲorx-oo=l yaw-žii-[n]=šd. \\ 
             no=\textsc{emph} just thing be.interested.in-\textsc{pfv.cvb=lim.foc} go-\textsc{prog-imm.prs=dp} \\
        \glt ‘No, I was just walking around looking at (lit. being interested in) things.’ (SC1: Derik, Narantuul)
    \end{xlist}
\z 

\ea\label{ex:kha:56} A: I was walking alone. \\ 
    \begin{xlist}
    \exi{B: }\label{ex:kha:56a}
        \gll xoy-uul=aa yaw-žii-s=iin \textbf{biš=üü}? \\
             two=\textsc{coll=rposs} go-\textsc{prog-est.pst=asrt} \textsc{neg=q.plr} \\
        \glt ‘Wasn’t it rather the case that you were walking as a pair?’
    
    \exi{A: }\label{ex:kha:56b} 
        \gll \textbf{gü=ee} \textbf{güü}. \\
             no=\textsc{emph} no \\
        \glt ‘No, no.’ (SC1: Derik, Narantuul)
    \end{xlist}
\z 

\ea\label{ex:kha:57}
    \begin{xlist}
    \exi{A: }\label{ex:kha:57a}
        \gll xüŋ med-\textbf{küü} yuu? \\ 
             person know-\textsc{npst.ptcp.neg} \textsc{q.plr} \\
        \glt ‘[So] people don’t know [about this]?’ 
    
    \exi{B: }\label{ex:kha:57b}
        \gll tiiŋ. \\
             such \\
        \glt ‘Yes [they don’t know].’ (SC1: Derik, Yanhan1) 
    \end{xlist}
\z  

Rather than using interjections or negated verbs, a construction that is occasionally used in negative answers is by using the non-reduced form of the old existential negator \textit{ügüi} plus copula as in \REF{ex:kha:58} and \REF{ex:kha:59}. These clauses are used to state, from immediately accessed evidence, that something is or was not the case, i.e. that a negative state obtains. These small pro-clauses may start with sentence adverbials and end in modal particles, but contain no arguments.

\ea\label{ex:kha:58}
    \begin{xlist}
    \exi{A: }\label{ex:kha:58a}
        \gll ɢurl=an tomd-ood bai-n=uu güi=yüü? \\ 
             flour=3\textsc{poss} get.too.big-\textsc{pfv.cvb} \textsc{cop-imm.prs=q.plr} \textsc{neg=q.plr} \\
        \glt ‘Is the flour getting too big or not?’
    
    \exi{B: }\label{ex:kha:58b}
        \gll  \textbf{ügüi} \textbf{bai}-n. \\
             \textsc{neg} \textsc{cop-imm.prs} \\
        \glt ‘It is not.’ (SC2: Z07-5)
    \end{xlist}
\z

\ea\label{ex:kha:59}
    \gll Iiŋxüü   asuudal šiid-egd-seŋ=üü ge-wel  \textbf{ügüi} \textbf{bai}-n. \\
         such problem decide-\textsc{pass-est.pst=q.plr} \textsc{quot-cond.cvb} \textsc{neg} \textsc{cop-imm.prs}\\
    \glt ‘If one were to state whether this question has been decided, it is not the case.’ (IC)
\z 
  \label{bkm:Ref12913}\textit{} 


\subsection{Negative indefinites and quantifiers}\label{sec:kha:3.2}

Khalkha lacks specific negative indefinites. Positive indefinites are recruited from a few regular nouns (\textit{yum} ‘(some)thing’, cf. \REF{ex:kha:9}) or formed from interrogative pronouns by adding \textit{neg-eŋ} ‘one-\textsc{attr}’ (e.g. \textit{xeŋ neg-eŋ} ‘somebody, anybody’) or the additive focus clitic =\textit{č} ‘even, also’ (\textit{xeŋ=č} ‘anybody’) (cf. \citealt[132--133]{Janhunen2012}, cf. also 67--68; for =\textit{č} under negation also see \citealt{Hashimoto2009}), %Why not Janunen 2012: 67--68, 132--133? BB wishes to keep it like that
as in \REF{ex:kha:60}. These forms are then used together with a regular verbal negator, as in \REF{ex:kha:61}. =\textit{č}{}-based indefinites most often occur under negation, while \textit{negeŋ}{}-based indefinites are more common in positive sentences, but show up in a limited set of negative contexts, particularly when the set of possible referents is restricted and topical as in \REF{ex:kha:62}. With quantifiers under negation as in \REF{ex:kha:63}, the use of =\textit{č} is optional. 

\ea\label{ex:kha:60}
    \gll Xii-seŋ xerg=ee \textbf{xeŋ=č} üür-ex yos-toi biz=dee. \\ 
         do-\textsc{prf.ptcp} deed=\textsc{rposs} who=\textsc{add.foc} load-\textsc{npst.ptcp} custom-\textsc{com} \textsc{mp=dp} \\
    \glt ‘Anyone should take responsibility for what s/he did, right?’ (IC) 
\z 

\ea\label{ex:kha:61}
    \gll bidn-ii \textbf{xeŋ=č} orolc-oo=\textbf{güi} \\ 
         \textsc{1pl-gen} who=\textsc{add.foc} participate-\textsc{res.ptcp=neg} \\
    \glt  ‘... none of us has taken part...’ (IC)
\z 

\ea\label{ex:kha:62}
    \gll Gexdee   az bol-ž \textbf{xeŋ} \textbf{neg-ŋ=en} bertel gemtel aw-aa=\textbf{güi} yum bai-n. \\ 
         but luck become-\textsc{cvb} who one-\textsc{attr=3poss} injury damage take-\textsc{res.ptcp=neg} \textsc{asrt} \textsc{aux-imm.prs} \\
    \glt ‘But fortunately, apparently \textbf{none} \textbf{of} \textbf{them} [i.e., the aforementioned reporters that were staying at the hotel] got injured [in the course of the armed robbery].’ (IC)
\z 

\ea\label{ex:kha:63} 
    \gll \textbf{ɢanc=ač} xanz med-ex={güü} bai-ž … \\ 
         single=\textsc{add.foc} hanzi know-\textsc{npst.ptcp=neg} \textsc{cop-cvb} \\
    \glt ‘{Not} knowing \textbf{even a single} Chinese character, …’ (SC2: Khalkha0001) 
\z 

\subsection{Negative derivation and case marking}\label{sec:kha:3.3}
\subsubsection{Negative derivation}\label{sec:kha:3.3.1}

There are several lexicalized expressions that involve old or current negators. Most are analytic and contain the old preverbal standard negators \textit{es} and \textit{ül} or the old ascriptive negator \textit{bus}. Some are morphologically derived from or via \textit{ügüi} or from \textit{biš}. 

The old imperfective/non-past negator \textit{ül} (n\textsubscript{IC}${\leq}$9121) is still somewhat common in negative expressions related to willful disregard (e.g. \textit{ül too-} ‘disregard’, \textit{ül xarɢalzaŋ} ‘without regard for’, \textit{ül xündetg-} / \textit{ül xündl-} ‘disrespect’, cf. also \textit{ül xamaar-} ‘be independent of’) or (often negligent) failure to realize (\textit{ül med-} / \textit{ül medegd-} ‘not realize / not be perceived’, \textit{ül oilɢ{}-} / \textit{ül oišoo-} ‘not understand’, \textit{ül tanʲ-} ‘not recognize’, \textit{ül itg-} ‘disbelieve’, \textit{ül aŋxaar-} ‘not notice’). In addition, it also forms part of the non-inflecting lexicalizations \textit{ül xödlöx (ed xöröŋg)} ‘immovable (property)’ and \textit{ül baram} ‘not likely to stop with > not only but also’. 

The old perfective/past negator \textit{es} is 3.5 times less frequent (n\textsubscript{IC}${\leq}$2631), and its distribution appears to be less systematic. Its use in a few verbal expressions like \textit{es zöwšöör-} ‘disagree/reject’, phrasal idioms like \textit{tus-iig es ol}{}- ‘not find help’ or an adverbial like \textit{xüs-seŋ=č es xüs-sen=č}  wish-\textsc{prf.ptcp+add.foc} \textsc{neg} wish-\textsc{prf.ptcp=add.foc} ‘nolens volens’ is relatively uncommon, and it is only used in a few fixed legal expressions such as \textit{es üilde-xüi} \textsc{neg} perform-\textsc{dvn} ‘inaction’ or \textit{šašaŋ šütex, es šütex erx} ‘right to worship and not to worship (i.e. freedom of religion)’. However, it is relatively common with petrified conditional converbs in \textit{-wal} (< \textit{-ba=la} –\textsc{pst=lim.foc}) which have turned into discourse markers such as the clause-connective \textit{-iig} \textit{es toocwol} ‘if one discounts that …’ or the sentence adverbials \textit{es tegwees} / \textit{es tegwel} ‘if one does not do like that’. Conditional negation is the only area in which the closely related Oirat variety Kalmyk retained its cognate of \textit{es} in active usage \citep[27--31]{Baranova2019}. This also includes the frequent conjunction \textit{eswel} {\textasciitilde} \textit{esxül} ‘or’ (< ‘if it is not’). This and the existence of the historical unclear form \textit{esex} ‘or not’ in subordinate clauses (see \sectref{sec:kha:4.5}) suggest a historical verb stem \textit{es}{}-, but this is probably a fallacity (see \citealt[89--90]{Brosig2015Negation}).

The old existential negator \textit{ügüi} is part of several derived or lexicalized forms, e.g. as stem in \textit{ügüi-l}{}- ‘miss (somebody)’ and \textit{ügüi-sg}{}- ‘deny’, as independent word in \textit{ügüi bol}{}- ‘become absent / die’ (cf. \textit{alaɢ bol}{}- ‘disappear’), or as privative in complex forms such as \textit{ažal-güi-d-el} ‘unemployment’ (cf. the rare \textit{ažal-güi-d-} ‘be unemployed’ etc.), \textit{noir-güi-d-el} ‘sleeplessness’, \textit{ür-güi-d-el} ‘childlessness’ or \textit{büt-el-güi-t}{}- succeed-\textsc{v>n-neg-n>v.intr}{}- ‘fail’. Forms derived from \textit{biš} such as \textit{biš-d}{}- ‘go wrong’, by contrast, are relatively rare. The combined \textit{biš-güi} ‘plentiful’ in \REF{ex:kha:64} might originally have been conceptualized from ‘without other’, i.e. without anything comparable [in quantity] (see \citealt[9--10, 14--15]{Bjambasan2001} for additional forms derived from \textit{ügüi} and \textit{biš}).

\ea\label{ex:kha:64}
    \gll Ger\_bül-d iim marɢaaŋ \textbf{bišgüi} ɢar-dag. \\ 
         family-\textsc{dat} such quarrel plentiful emerge-\textsc{hab.ptcp} \\
    \glt ‘In a family, this kind of quarrel comes up abundantly.’ (IC)
\z 


\textit{Bus}, which is cognate with \textit{biš} (< \textit{busi}), is exceedingly common as a negator in recent official and especially administrative terms in attributive use. In these collocations, the first part is an adjective or noun, either bare or with a case suffix (genitive \textit{-iiŋ}, attributive \textit{-aŋ}, short comitative{}-possessive \textit{-t}). The modified noun may be fixed as in \textit{tör{}-iiŋ bus baiɢuulaɢ} state.power-\textsc{gen} \textsc{neg} organization ‘non-governmental organization’ or semantically delimited as in \textit{eelž-et bus xural} / \textit{xuraldaaŋ} / \textit{čuulɢaŋ} / \textit{soŋɢuulʲ} ‘irregular meeting / gathering / assembly / election’. Some expressions allow for a wider range of modified nouns, e.g. \textit{ašg-iiŋ (tölöö) bus} profit-\textsc{gen} (for) \textsc{neg} ‘non-profit’, \textit{xaraa-t bus} ‘independent’ or \textit{xüč-eŋ tögöldör} \textit{bus} ‘invalid’. There are also some official-sounding, but non-administrative collocations like \textit{šudraɢ bus} ‘dishonest/unfair’, \textit{erüül bus} ‘unhealthy’ and the highly frequent \textit{todorxoi} \textit{bus} ‘unclear’. Finally, there are the two colloquial collocations \textit{neg bus udaa} ‘more than once’ and \textit{bas bus} also \textsc{neg} ‘this and that’. For terms ending in \textit{bus}, predicative uses are highly dispreferred. 


\subsubsection{Negative case and negative adjectives}\label{sec:kha:3.3.2}

In one of its oldest attested functions, =\textit{güi} may function as a privative that marks the absence of an entity. It is used adverbially \REF{ex:kha:65}, attributively \REF{ex:kha:66} or sentence-finally, namely in the possessive construction discussed in \REF{ex:kha:34}. The comitative-possessive case \textit{-tai} is used to create roughly equivalent positive patterns. Adjectives are uncommon with the privative (see the discussion about \REF{ex:kha:24}).

\ea\label{ex:kha:65}
    \gll ta nar soloŋɢos-ruu wiz=\textbf{güi} yaw-dag yum=uu? \\ 
         \textsc{2} \textsc{pl} Korea-\textsc{all} visa=\textsc{priv} go-\textsc{hab.ptcp} \textsc{asrt=q.plr} \\
    \glt ‘Can you thus enter Korea without visa?’ (SC2: Khalkha0001)
\z 

\ea\label{ex:kha:66}
    \gll man-ai-x=čan wiz=\textbf{güi} orn-uud-tai, zöndöö=šdee. \\ 
         \textsc{1pl-gen-nmlz=stc} visa=\textsc{priv} country-\textsc{pl-com}   lots.of=\textsc{dp} \\  
    \glt ‘Ours [=Our people] have visa-free countries, lots of them!’ (SC2: Khalkha0001)
\z 

Several nouns that denote concepts like social obligation and inclination can be derived into pairs of “modal adjectives” via comitative and privative suffixes, as shown in \REF{ex:kha:67}.

\TabPositions{.25\textwidth,.45\textwidth,.1\textwidth,.5\textwidth}

\ea\label{ex:kha:67}
    \textit{xereg-tei/güi}       \tab  {‘necessity’}      \tab  >  \tab  {‘(not) need to’} \\
    \textit{yos-toi/güi}         \tab   {‘custom}’        \tab  >  \tab  {‘ought to/mustn’t’} \\
    \textit{šaardlaɢ{}-tai/güi}  \tab  {‘requirement’}    \tab  >  \tab  {‘(not) have to’} \\
    \textit{bolomž-toi/güi}      \tab    {‘possibility’}  \tab  >  \tab  {‘(im)possible’} \\
    \textit{dur-tai/güi}         \tab  {‘desire’}         \tab  >  \tab  {‘(dis)like to’} \\
    \textit{xüsel-tei/güi}       \tab  {‘wish’}           \tab  >  \tab  {‘(not) wish to’} \\
    \textit{sanaa-tai/güi}       \tab  {‘intention’}      \tab  >  \tab  {‘(not) intend to’} \\
\z 

% raw data of the table above
% \textit{xereg-tei/güi}       {‘matter’}         >  {‘(not) need to’}
% \textit{yos-toi/güi}          {‘custom}’        >  {‘ought to/mustn’t’} 
% \textit{šaardlaɢ{}-tai/güi}  {‘requirement’}    >  {‘(not) have to’}
% \textit{bolomž-toi/güi}        {‘possibility’}  >  {‘(im)possible’} 
% \textit{dur-tai/güi}         {‘desire’}         >  {‘(dis)like to’} 
% \textit{xüsel-tei/güi}       {‘wish’}           >  {‘(not) wish to’} 
% \textit{sanaa-tai/güi}       {‘intention’}      >  {‘(not) intend to’} 

Modal adjectives are used as regular adjectives \REF{ex:kha:68} or convey modality for a preceding verb in the nonpast participle \REF{ex:kha:69}. They can be followed by a copula for additional TAME marking. A few adjective pairs only have adjectival uses, e.g. \textit{ašag-tai/güi} ‘profit’ > ‘(dis)advantageous’ and \textit{xamaa-tai/güi} ‘relevance [only found in existential clauses]’ > ‘(ir)relevant’. The modal adjective \textit{maɢadgüi} ‘maybe’ (< Classical Mongolian \textit{maɢad} ‘certain’) doesn’t contrast with a positive form and (in Khalkha dialect) only follows linking converbs in \textit{-ž} \REF{ex:kha:70}.

\ea\label{ex:kha:68}
    \gll žims bol \textbf{xereg=güi}. \\ 
         fruit \textsc{top} necessity=\textsc{neg} \\
    \glt ‘As for fruits, we don’t need any.’ (SC2: Khalkha0001)
\z 
\ea\label{ex:kha:69}
    \gll ɢurʲal aw-ax \textbf{xereg=güi}. \\ 
         flour   take-\textsc{npst.ptcp} necessity=\textsc{neg} \\
    \glt ‘We don’t need to import [wheat] flour.’ (SC2: Khalkha0001)
\z 
\ea\label{ex:kha:70}
    \gll tuŋɢaa tat-saŋ bai-ž \textbf{maɢadɢuu}. \\ 
         \textsc{name} draw-\textsc{prf.ptcp} \textsc{cop-cvb} maybe \\
    \glt ‘Tungaa might have smoked [this fag].’ (SC2: Khalkha0001)
\z 

The morpheme combination \textit{-š=güi} forms deverbal adjectives which express that the occurrence of the event is unlikely or impossible, e.g. \textit{mart-agd-aš=güi tör-söŋ ödor} ‘unforgettable birthday’ or \textit{sal-š=güi naiz} ‘inseparable friends’. These adjectives are mostly used as attributes and only rarely as predicates. An irregular case is \textit{bolz-oš=güi} ‘unarrangable’ in \REF{ex:kha:71} which indicates that some adverse action marked by a converb in \textit{-ž} might conceivably happen. It is also common in predicates. A related positive pattern in \textit{-uuš-tai} forms adjectives that denote that the action is worthwhile doing, e.g. \textit{anxaar-uuš-tai züil} ‘an issue worth taking notice of’. While most of these words are lexicalized and do not come in polarity pairs, the patterns retain some limited productivity. The morph \textit{-(uu)š-} always co-occurs with comitative/privative case.

\ea\label{ex:kha:71}
    \gll yaltaŋ amʲ=aa xorl-ož bolz-oš=\textbf{güi} bol … \\ %HC ...
         prisoner life={\textsc{rposs}} harm-{\textsc{cvb}} arrange-{\textsc{likely=neg}} if  \\
    \glt ‘If a prisoner is likely to commit suicide …’ (IC)
\z 


\section{Other aspects of negation}\label{sec:kha:4}
\subsection{The scope of negation}\label{sec:kha:4.1}
 

The position of a declarative negator within a predicate correlates with its scope. It is subsumed under the scope of all subsequent TAME markers and takes all preceding (necessarily non-finite) TAME markers into its scope. For instance, the negator in \textit{id-ež bai-ɢaa=güi} ‘have not yet been eating’ in \REF{ex:kha:10} takes scope over the progressive, while the copula in \textit{xel-ee=güi bai-n} ‘have not yet said’ in \REF{ex:kha:12} defines a time span during which a certain action is not being performed. If a negator attaches to adjectives as in \REF{ex:kha:24} or \REF{ex:kha:25a}, it only affects the adjectival phrase itself. In imperatives (\sectref{sec:kha:2.2}), the negator precedes and takes scope over the full verbal predicate, sometimes along with some arguments other than the subject.

In Mongolic languages, it is often not possible or feasible to separately negate adjectives used as adverbials to verbal predications \citep[156]{Yu1991}, since =\textit{güi} is uncommon with adjectives (\sectref{sec:kha:2.3.1}) and \textit{biš} rarely forms adverbials. Instead, standard negation negates these adverbials and their predicates as a whole, as in \REF{ex:kha:72} which only denies that the speaker has \textsc{good} knowledge. Since =\textit{güi} entails reduction, \textit{saiŋ} ‘good’ could not readily be replaced with bare \textit{muu} ‘bad’. With the additive focus clitic \textit{=č} (see \sectref{sec:kha:3.2}), \textit{muu=č} could imply that the speaker does not even know the matter badly, but just not at all.

\ea\label{ex:kha:72}
    \gll en-iig bol bii neg saiŋ med-ex=\textbf{küü}. \\ 
         this-\textsc{acc} \textsc{top} \textsc{1sg} one good   know-\textsc{npst.ptcp=neg} \\
    \glt ‘This I don’t know particularly well.’ (SC1: Derik, Benzin)
\z 

To negate an adjunct through its predicate without necessarily reducing or denying the main action, the negator or “{\cn 焦点調整形式}” ‘focus adjustment form’ \citep[75--76]{Mukai2001} \textit{biš} can be used \citep[159--160]{Yu1991}. \textit{Biš} indicates that something is not correct about the proposition, but does not specify what. For instance, \REF{ex:kha:73} is ambiguous with regard to the negated constituent, while in \REF{ex:kha:74} the context specifies that only the adverbial \textit{xüüxed-tei ge-ed} ‘because I have a child’ is negated by \textit{biš}.

\ea\label{ex:kha:73}
    \gll Oyuuŋ bayaŋ ge-ed Dorž tüün-tei gerel-seŋ \textbf{biš}. \\ 
         \textsc{name} rich \textsc{quot-pfv.cvb} \textsc{name} that-\textsc{com} marry-\textsc{prf.ptcp} \textsc{neg} \\
    \glt ‘Dorj didn’t marry Oyuun saying [i.e. because] \textit{Oyuun is rich} (but he did something else/but he married her for another reason).’ \citep[159--160]{Yu1991}
\z 
\ea\label{ex:kha:74}
    \gll Xüüxed-tei ge-ed yaw-ž čad-ax=güi bai-ɢaa=m \textbf{biš} šüü.\\ 
         child-\textsc{com} \textsc{quot-pfv.cvb} go-\textsc{cvb} can-\textsc{npst.ptcp=neg} \textsc{cop-prs.ptcp=asrt} \textsc{neg} \textsc{dp} \\
    \glt  ‘It is really not because I have a child that I cannot leave [my husband]. [I have (…) all things necessary to take my child with me.]’ (IC)
\z 

\textit{Biš} most often attaches to predicates, but it is also used for the contrastive negation of constituents such as arguments \REF{ex:kha:75}, goal complements \REF{ex:kha:76}, and attributes. The denied and affirmed constituent must both be overt \citep[123--124]{Yu1991}, except in some fossilized expressions such as \textit{neg biš udaa / neg bus udaa} ‘on not just one occasion’ (cf. \sectref{sec:kha:3.3.1}).

\ea\label{ex:kha:75}
    \gll Moŋɢol-d \textbf{biš} xarʲ ɢarʲag-t tör-söŋ bol saixaŋ baixgüiyuu. \\ 
         Mongolia-\textsc{dat} \textsc{neg} foreign world-\textsc{dat} be.born-\textsc{prf.ptcp} if beautiful \textsc{dp} \\
    \glt ‘It is great if you were not born in Mongolia, but in the great wide world, you see.’ (IC)
\z 
\ea\label{ex:kha:76}
    \gll Xudaldaa-g xož-ogd-ox   ge-ž \textbf{biš} xož-ox ge-ž xii-deg. \\ 
         trade-\textsc{acc} defeat-\textsc{pass-npst.ptcp} \textsc{quot-cvb} \textsc{neg} defeat-\textsc{npst.ptcp} \textsc{quot-cvb} do-\textsc{hab.ptcp} \\
    \glt ‘One does trade not in order to be defeated, but to defeat.’ (IC)
\z 

\subsection{Negative polarity}\label{sec:kha:4.2}

A few semantically general adverbs are confined to negated contexts, including the intensifier \textit{ogt} ‘(not) at all’ in \REF{ex:kha:77}, \textit{tödiilöŋ} ‘(not) particularly’ [lit. making it that much], and the concessives \textit{yaawč} / \textit{xerxewč} ‘(not) in any regard’ [lit. in spite of doing what]. The comparative-concessive adverbial \textit{tegtlee} [lit. until doing like that] is restricted to negated and interrogative sentences, and \textit{tödii} ‘(only) that much’ prefers, but does not strictly require negated contexts. Indefinite proforms (see \sectref{sec:kha:3.2}) are mostly found in negative, interrogative and permissive contexts.

Some adverbs are more common in negative contexts due to their specific meanings. For instance, \textit{xaraaxaŋ} ‘still’ only collocates with predicates that express that something has not yet become ready or not come to a certain point as in \REF{ex:kha:77}, which also includes positive sentences containing the word \textit{ert} ‘early’ or its derivations. Similarly, \textit{or tas} ‘completely’ collocates with predicates that denote becoming absent, whether this is achieved lexically or via negators.\footnote{All forms discussed in this section were identified by \citet[109]{Kim2016}.}

\ea\label{ex:kha:77}
    \gll  Gexdee   en=en   xüŋ \textbf{ogt} ič-ex yos=güi ge-seŋ üg \textbf{xaraaxaŋ} \textbf{biš}. \\ 
         but this=\textsc{3poss} person entirely feel\textunderscore ashamed-\textsc{npst.ptcp} custom=\textsc{neg} \textsc{quot-prf.ptcp} word still \textsc{neg} \\
    \glt ‘But that does not quite mean that man shouldn’t feel ashamed at all.’ (IC)
\z


\subsection{Marking of noun phrases in the scope of negation}\label{sec:kha:4.3}

Under negation, case marking is left intact, only that comitative case marking in \textit{-tai} is replaced with privative marking in =\textit{güi} if the constituent itself is negated (see \sectref{sec:kha:3.3}). The use and choice of focus markers are affected by polarity, i.e. additive focus marking is used for ‘[not] even [one/once]’ (cf. \REF{ex:kha:63}), but limitative focus marking in =\textit{l} is used in the corresponding positive statement ‘only [one/once]’. 

\subsection{Reinforcing negation}\label{sec:kha:4.4}
 
The negative polarity adverb \textit{ogt} ‘(not) at all’ as in \REF{ex:kha:77} reinforces negation. Apart from this, several sentence-final particles are general means for reinforcing positive or negative statements.

\subsection{Negation, coordination and complex clauses}\label{sec:kha:4.5}

Mongolian tends to embed subordinate into main clauses and thus makes relatively limited use of coordination. Constructions based on \textit{eswel} ‘or’ historically speaking use negation in the scope of conditionality to coordinate constituents (see \citealt[89--90]{Brosig2015Negation}). Coordination also plays a role in alternative questions with \textit{(ü)güi=yüü} (\sectref{sec:kha:2.2}). In a similar vein, it is possible to coordinate two finitely used predicates at the end of a matrix clause. Most often both of these have the same polarity as in \REF{ex:kha:78}, but different forms as in \REF{ex:kha:79} are possible, too. The latter might be seen as a predecessor to negated adverbial clauses (\sectref{sec:kha:2.4.2}). Contrastive constituents can be coordinated by \textit{biš} (\sectref{sec:kha:4.1}), antonym compounds use =\textit{güi} (\sectref{sec:kha:4.6.3}). 

\ea\label{ex:kha:78}
    \gll Getel tern-ii aaw=an ir-ž (…),   tüün-iig bol ömöör-öö=\textbf{güi} bas eež-iig=en bas  zagn-aa=\textbf{güi}. \\ %HC (...)
         but that-\textsc{gen} father=\textsc{3poss} come=\textsc{cvb} {} that-\textsc{acc} \textsc{top} console-\textsc{res.ptcp=neg} also mother-\textsc{acc=3poss} also scold-\textsc{res.ptcp=neg}\\
    \glt ‘But her father came (…) but he did not console her, nor did he scold the mother.’ (lit. and also didn’t scold …) (SC2: Retelling\_EO)
\z 

\ea\label{ex:kha:79}
    \gll neg xözör ɢar-ɢ{}-ax-\textbf{guu}, öŋg-öör=ön ɢar-aɢ{}-čax-Ø!\\
         one trump exit-\textsc{caus-npst.ptcp-neg} color-\textsc{ins=3poss} exit-\textsc{caus-compl-imp}\\
    \glt ‘[One] doesn’t play a trump [here; rather] follow suit [lit. ‘by its color’]!’ (SC2: Khalkha0017) 
\z 

A specific case between coordination and subordination is the negative nominalizer \textit{esex}. It can follow any nominal predication, but most commonly combines with nonpast participles in \textit{-x}, as in \REF{ex:kha:80}. Structurally, this pattern is reminiscent of the formation of alternative questions in that the negator forms something reminiscent of a negated pro-clause, and the tendency to “coordinate” identical TAME forms (if \textit{esex} is taken to historically end in a nonpast participle) supports this explanation. \textit{Esex} as a nominal element can form argument clauses, which would be marked by case suffixes as in \REF{ex:kha:81} or, for subjects, by personal-possessive suffixes (for subjects) as in \REF{ex:kha:80}. Alternatively, \textit{esex} can also be used attributively as in \REF{ex:kha:82}. It is prevalent in formal written texts and entirely absent from SC. 

\ea\label{ex:kha:80}
    \gll Bag=n delxii-ŋ   awarɢ{}-ad orolc-ox \textbf{esex}=n ergelzee-tei. \\ 
         team=\textsc{3poss} world-\textsc{gen} champion-\textsc{dat}   participate-\textsc{npst.ptcp} or.not=\textsc{3poss} doubt-\textsc{com} \\
    \glt ‘It is doubtful whether the team will participate in the world champion[ship].’ (IC)
\z 
\ea\label{ex:kha:81}
    \gll en doloo\_xonog-t baɢt-aa-ŋ D.Zorʲogt said-iig \textit{ogcr-uul-ax} \textbf{esex}-iig šiidverl-ex yos-toi.\\ 
          this week-\textsc{dat} be.contained-\textsc{caus-cvb} \textsc{name} minister-\textsc{acc} resign-\textsc{caus-npst.ptcp} or.not-\textsc{acc} decide-\textsc{npst.ptcp} custom-\textsc{com} \\
    \glt  ‘[We] should decide within this week whether or not [we] relieve minister D. Zorigt of his duty.’ (IC)
\z 
\ea\label{ex:kha:82}
    \gll Erdes\_bayalag erčem\_xüčn-ii said D.Zorʲogt-iig ogcr-uul-ax \textbf{esex} asuudl-iig UIX-iiŋ negd-seŋ čuulɢan-aar xel-elc-lee.\\ 
          minerals energy-\textsc{gen} minister \textsc{name-acc}  resign-\textsc{caus-npst.ptcp} or.not problem-\textsc{acc} Great.State.Assembly\textsc{{}-gen} unite-\textsc{prf.ptcp} assembly-\textsc{ins} discuss\textsc{-recp-imm.fh.pst} \\
    \glt ‘[They] discussed the question of whether or not to relieve the minister for resources and energy, D. Zorigt, of his duty during the plenary session of the Great State Assembly.’ (IC)
\z 

The pattern \textit{(ter=č) büü xel-Ø} ‘not to mention’ with the old imperative negator \textit{büü} also fulfills a connective function, as in \REF{ex:kha:83}, though in structural terms it seems to be adverbial rather than coordinating.

\ea\label{ex:kha:83}
    \gll Tatwar ge-xeer önöödör biznes erxlegč-d-ees exl-eed žiriiŋ neg-eŋ irgeŋ Dorž ɢuai xürtel töwögš-öö-ž, \textbf{ter=č} \textbf{büü} \textbf{xel-Ø} zarʲam=an daramt-and or-ž, stress-t barʲ-agd-saŋ bai-dag. \\ 
         tax {\textsc{quot-cvb}} today business owner-\textsc{pl-abl} begin-\textsc{cvb} normal one=\textsc{attr} citizen \textsc{name} \textsc{hon} until be\textunderscore uncomfortable-\textsc{caus-cvb} that=\textsc{add.foc} \textsc{neg} say-\textsc{imp} some=\textsc{3poss} pressure-\textsc{dat} enter-\textsc{cvb} stress-\textsc{dat} catch-\textsc{pass-prf.ptcp} \textsc{cop-hab.ptcp} \\
    \glt ‘If you speak about taxes, today [people] starting from the business owner up to normal citizen Dorj feel uncomfortable. \textbf{Not} \textbf{to} \textbf{mention} \textbf{that,} some \textbf{even} get into pressure and are caught by stress.’ (IC)
\z

\subsection{Miscellaneous aspects of negation}\label{sec:kha:4.6}
\subsubsection{Negative transport}\label{sec:kha:4.6.1} 
\largerpage

In many languages, subordinate clause predicates with certain semantic characteristics can be negated via a negation that occurs in the matrix clause (e.g. \textit{I don’t think they sing} for \textit{I think they don't sing}). This phenomenon is called negative transport (see \citealt{Horn1989}: 315). In Mongolian, predicates that entail such implicatures include \textit{bod}{}- ‘think’ \REF{ex:kha:84} and the somewhat formal \textit{xüs}{}- ‘wish’ \REF{ex:kha:85}. Regardless of whether the matrix predicate or the embedded predicate is negated, the most accessible interpretation is that the speaker seeks to express that the predication in the embedded clause does not obtain.

\ea\label{ex:kha:84}
    \ea\label{ex:kha:84a}
        \gll Bi margaaš toɢol-ž čad-an ge-ž bod-ox=\textbf{güi} bai-n.\\ 
             \textsc{1sg} tomorrow play\textsc{{}-cvb}   can-\textsc{imm.prs} \textsc{quot-cvb} think-\textsc{npst.ptcp=neg} \textsc{cop-imm.prs} \\
        \glt ‘[I] don’t think I can play tomorrow.’ (IC)
    
    \ex\label{ex:kha:84b}
        \gll inflʲacʲ-ad   tiim=č ix nölööl-öx=\textbf{güi} ge-ž bod-ož bai-n. \\
             inflation-\textsc{dat}   so=\textsc{add.foc}  much influence-\textsc{npst.ptcp=neg} \textsc{quot-cvb} think-\textsc{cvb} \textsc{cop-imm.prs}\\
        \glt ‘[I] think that [this] (…) would not influence inflation that much.’ (IC)
    \z
\z 


\ea\label{ex:kha:85}
    \ea\label{ex:kha:85a}
        \gll Tiim boloxoor bi itgel ald-ax-iig xüs-ex=\textbf{güi} bai-ɢaa. \\ 
             like.that because \textsc{1sg} belief lose-\textsc{npst.ptcp-acc} wish-\textsc{npst.ptcp-neg} \textsc{cop-est.prs} \\
        \glt ‘Because it is like that, I don’t want to lose trust.’ (IC)
    
    \ex\label{ex:kha:85b}
        \gll Wirus eswel teknik-iiŋ gemtl-ees bol-ž čuxal fail-uud=aa ald-ax=\textbf{güi} bai-x-iig xüs-deg.\\
             virus or technic-\textsc{gen} damage-\textsc{abl} become-\textsc{cvb} important file-\textsc{pl=rposs} lose-\textsc{npst.ptcp-neg} \textsc{cop-est.prs-acc} wish-\textsc{hab.ptcp}\\
        \glt ‘We want not to be losing important files because of viruses or technical defects.’ (IC)
    \z
\z 

Structurally, the complement clause in \REF{ex:kha:84} is in the scope of the quotative verb, while in \REF{ex:kha:85} it is nominalized via a case-marked nonpast participle. It is exceedingly rare for negated participles to receive accusative case marking (n\textsubscript{IC:}\textit{\textsubscript{xüs-}}=8, i.e. the verb \textit{xüs}{}- is immediately preceded by –\textsc{ptcp}=\textit{güi=g} on 8 occasions in the Internet Corpus). So if a verb of cognition requires accusative marking, the case suffix rather attaches to a copula that follows the negated participle (n\textsubscript{ IC:}\textit{\textsubscript{xüs-}}=147). However, a copula automatically conveys aspect (e.g. progressive in \REF{ex:kha:85b}). If this is to be avoided, the negation has to be placed in the matrix clause (n\textsubscript{ IC:}\textit{\textsubscript{xüs-}}=1917), as in  \REF{ex:kha:85a}. In contrast to \textit{bod}{}- (\textsc{opinion}) and \textit{xüs}{}- (\textsc{intention}), negated forms of the verb \textit{san-agd}{}- miss/think-\textsc{pass}{}- ‘seem (to the mind) as if’ (\textsc{perception}) only rarely take (positive) participial complements (n\textsubscript{IC}=15) or quoted complement clauses (n\textsubscript{IC}=12). Instead, negation usually takes place within the complement of the positive matrix verb \textit{san-agd-}, e.g. by introducing negated participles by means of the dummy noun \textit{yum} ‘thing’ followed by the similative linker \textit{šig} in a sequence \textsc{-ptcp}\textit{=güi yum šig san-agd} (n\textsubscript{IC}=201). The option to negate either the main or subordinate clause doesn’t seem to be common for predicates of other semantic classes either. \textsc{Probability} is commonly expressed by complex predicates or adverbials rather than by main clause predicates. For \textsc{intention} and \textsc{judgment/weak} \textsc{obligation}, speakers heavily rely on \textsc{modal} \textsc{adjectives}, see (\ref{ex:kha:67}--\ref{ex:kha:70}) above. These are inherently marked for polarity by privative or comitative (possessive) case forms, while subordinate forms take bare nonpast participles. Consequently, they are negated in the matrix clause as in \REF{ex:kha:86} (n\textsubscript{IC}=575), while a structure similar to \REF{ex:kha:85b} is strongly dispreferred  (n\textsubscript{IC}=1). 

\ea\label{ex:kha:86}
    \gll Bi öwöl ger-ees=ee ɢar-ax \textbf{dur=güi}=ee. \\
         \textsc{1sg} winter home-\textsc{abl=rposs} go.out-\textsc{npst.ptcp} desire-\textsc{neg=emph}\\
    \glt ‘I won’t like leaving my place during winter.’ (IC)
\z 

If this preliminary overall picture is correct, “negative transport” in Khalkha Mongolian appears to be a somewhat restricted phenomenon.

\subsubsection{Metalinguistic negation}\label{sec:kha:4.6.2}

While regular negation denotes that a proposition does not hold true, metalinguistic negation is used to object to an utterance on any grounds conceivable, including its apparent implicatures, its register, morphology or phonetics \citep[363]{Horn1989}. This usually requires the speaker to explicitly invoke an alternative contrastive interpretation, something that the ascriptive negator \textit{biš} (< ‘other’) does by necessity (see \sectref{sec:kha:2.3.1}). Consequently, it is \textit{biš} that is used for the metalinguistic readjustment of perspective as in \REF{ex:kha:87}.


\ea\label{ex:kha:87}
    \gll 2009 oŋ-ii soŋɢuulʲ-d C.Elbegdorž ge-deg ulstörč yal-saŋ \textbf{biš}, N.Eŋxbayar ge-deg ulstörč yal-agd-saŋ yum. \\ 
         2009 year-\textsc{gen} election-\textsc{dat} \textsc{name} \textsc{quot-hab.ptcp} politician win-\textsc{prf.ptcp} \textsc{neg} \textsc{name} \textsc{quot-hab.ptcp} politician win-\textsc{pass-prf.ptcp} \textsc{asrt} \\
    \glt ‘In the election of 2009, it was not the politician Ts. Elbegdorj who won, [but rather] the politician N. Enkhbayar who was defeated.’ (IC)
\z 
 

\subsubsection{Non-negative uses of negatives}\label{sec:kha:4.6.3}

A negated nonpast participle plus optional conditional particle may be followed by a negated nonpast participle of \textit{bol}{}- ‘may’ as in \REF{ex:kha:88} to indicate necessity. Both negated nonpast participle in \textit{-x=güi} and negative converbs in \textit{-lgüi} can be followed by \textit{yaa}{}- ‘do what’ as in \REF{ex:kha:89} to indicate a reasonable course of action. Straightforward double negation, yielding a positive statement, is accomplished by =\textit{güi biš}, as in \REF{ex:kha:90}. Several rare forms of double negation that are even absent from IC are discussed by \citet{Pjurbeev2015}.


\ea\label{ex:kha:88}
    \gll bitmiŋ uu-x-\textbf\textit{kuu} bol bol-\textbf{kuu}. \\ 
         vitamin drink-\textsc{npst.ptcp} if may-\textsc{npst.ptcp.neg} \\
    \glt ‘I (really) need to take vitamins.’ (SC2: Z05-1) (lit. ‘It is not possible if [SUBJECT] doesn’t drink vitamins.’) 
\z 
 
\ea\label{ex:kha:89}
    \gll odoo xon-uul-x=\textbf{güü} yaa-x=aw=dee.\\
         now stay.overnight-\textsc{caus-npst.ptcp=neg} do.what-\textsc{npst.ptcp=neg=q.ifr}\\
    \glt ‘Now what else but to let him stay overnight [now that he’s arrived].’ (SC1: TH, Ganbaatar)
\z 
 

\ea\label{ex:kha:90} 
    \begin{xlist}
        \exi{A:} ‘It seems that this light hasn’t been put in yet.’
        \exi{B:}
            \gll güi=ee, čiideŋ-gii=n xii-gee=\textbf{güi} \textbf{biš}=ee, en-čen šat-c=iin. \\ 
                 no=\textsc{emph} light.bulb-\textsc{acc=3poss} do-\textsc{res.ptcp=neg} \textsc{neg=emph} this-\textsc{stc} burn-\textsc{est.pst=asrt} \\
            \glt ‘No, it’s not that the light bulb hasn’t been put in, it has burnt out.’ (IC)
    \end{xlist}
\z 

\newpage
If attached to a pair of antonyms, =\textit{güi} cancels out the relevance of this distinction for the situation at hand, and both sets are conjoined, as in \REF{ex:kha:91}. Most attested antonym collocations are highly conventionalized. 

\ea\label{ex:kha:91}
    \gll ödör šön=\textbf{güü} gedes güi-lg-eed bai-n. \\ 
         day night=\textsc{neg} stomach run-\textsc{caus-pfv.cvb} \textsc{cop-imm.prs} \\
    \glt ‘I’m having diarrhea (without regard for) day and night.’ (SC2: Z05-2)
\z 

As discussed in \sectref{sec:kha:2.2}, negated questions commonly combine a request for information with the speaker’s guess at a positive state. The form \textit{bai-x=güi=yüü}, consisting of the existential/copula verb \textit{bai}{}- as a negated nonpast participle with question clitic, conveys such a meaning in a locative-possessive construction, as in \REF{ex:kha:92}. However, if it follows predicative nominals as in \REF{ex:kha:75} and \REF{ex:kha:93}, it functions as a specialized discourse particle. It signals to the addressee that the speaker is quite convinced of the proposition and, given the explanation just presented, the addressee should also accept it. Alternatively, the speaker might use it to remind the addressee of an important factor which the addressee might not have considered earlier. This expression also implies that this factor would enable the addressee to automatically understand the overall situation.
    
In a sincere or even rhetorical negated verbal question, \textit{=güi=yüü} would attach directly to the participle. The semantically similar discourse particle \textit{šüü}, conveying the speaker’s superior authority, might have a similar origin, going back to \textit{biš=üü}.


\ea\label{ex:kha:92}
    \gll čam-d maas \textbf{bai-x-kuu=yuu}?  \\ 
         2\textsc{sg.dat} pair.of.cards be-\textsc{npst.ptcp=neg=q.plr} \\
    \glt ‘Don’t you have any pair of cards?’ (SC2: Khalkha0017)
\z 

\ea\label{ex:kha:93}
    \gll ter yag iŋg-ež čimx-deg \textbf{baixkuuyuu}. \\ 
         that precisely do.like.this-\textsc{cvb} pinch-\textsc{hab.ptcp} \textsc{mp} \\
    \glt ‘She pinches the dumplings shut in precisely this [efficient and helpful] way, you see.’ (SC2: Z03-1)
\z 
 
The suffix \textit{-lgüi} (see \sectref{sec:kha:2.4.2}) forms yet another complex discourse particle, \textit{bai-lgüi=dee}, as in \REF{ex:kha:94}. It conveys that the occurrence of the event (at a time specified by a preceding participial suffix, mostly in \textit{-x}) is almost automatic and nothing that the addressee should worry about. Since specialized converbs in \textit{-lgüi} can normally not occur in sentence-final position, the construction might have arisen from ellipsis, but native speakers are unable to supply any elided component.


\ea\label{ex:kha:94}
    \gll neg caɢaan sar-aas   ömön uulz-acgaa-x bai-\textbf{lguu} dee. \\ 
         one white month\textsc{-abl} before meet\textsc{-pl-npst.ptcp} \textsc{aux-neg.cvb} \textsc{dp} \\
    \glt ‘Before New Year, we shall certainly meet.’ (SC2: Z08-1)
\z 
 
As mentioned for \REF{ex:kha:53} in \sectref{sec:kha:2.3.1}, the function of \textit{biš} can sometimes be understood in terms of its original meaning ‘other’, and such an interpretation is forced for \REF{ex:kha:95}. It thus does not truly instantiate a non-negative usage of a negative, but rather the not-yet negative use of a young negator that has not acquired a negation function in all contexts. 


\ea\label{ex:kha:95}
    \gll öör   xüüxd-üüd-ees=l asuu-x-aas \textbf{biš} yaa-x=aw? \\ 
         other child-\textsc{pl-abl=lim.foc} ask-\textsc{npst.ptcp-abl} other do.what-\textsc{npst.ptcp-q.ifr} \\
    \glt ‘What should you do, except for asking other children?’ (IC)
\z 

Finally, the construct in \REF{ex:kha:96} appears to be biclausal and ultimately positive. The nominal clause preceding \textit{biš} seems to express an action that the speaker regrets. The second clause conveys the unexpected and undesired consequence of that action. The habitual \textit{-dag} conveys a surprised pseudo-generalization \citep[esp. 93--96]{Brosig2015Aspect}, while \textit{bai-n} signals immediate perception. Alternatively, its predicate could contain the completive \textit{-čix-} indicating lack of speaker control. \REF{ex:kha:96a} is a regular translation and \REF{ex:kha:96b} a hypothetic attempt to translate \textit{biš} as a negator, with elliptic elements that currently cannot be motivated independently.

\protectedex{
\ea\label{ex:kha:96}
        \gll Yar’-saŋ \textbf{biš} ɢai bol-dog bai-n. \\
        speak-\textsc{prf.ptcp} \textsc{neg} misfortune become\textsc{{}-hab.ptcp} \textsc{cop-imm.prs} \\
    \begin{xlist}
        \ex\label{ex:kha:96a}
            \glt  ‘I spoke [about it], and now misfortune has come out of it.’
        \ex\label{ex:kha:96b}
            \glt ‘[If] it had not been for [me] speaking, [everything would have been good. But now,] I see, misfortune evolves.’ (D. Guntsetseg, p.c., 2019-04-24)
    \end{xlist}
\z }

While \textit{biš} (< ‘other’) might still express its original meaning in connective contexts like (\ref{ex:kha:95}--\ref{ex:kha:96}), neither \textit{biš} nor \textit{bus} can express this meaning in regular attributive uses. In this regard, they contrast with the fossilized plural adjective form \textit{bus.ad} ‘the other(s)’ which can be used attributively or as a nominalized form but cannot function as a negator. 

\subsubsection{Diachronic notes and observations}\label{sec:kha:4.6.4}
\largerpage

The Middle Mongol (13\textsuperscript{th}{}-14\textsuperscript{th} century) system was based on several preverbal negators: \textit{ese} for past forms, \textit{ülü} for non-pasts, and \textit{büü} for imperatives. There is defeasible evidence that these were also used for ascriptive negation. \textit{Busi} ‘other’ started to take over ascriptive negation in Late Middle Mongol, while \textit{bü-tügei} ‘desist!’ started its development into an imperative negator even later. The existential negator was \textit{ügei} (back then an adjective-like word with varying forms), which also formed privative adverbials. Non-possession was based on a privative rather than a locative scheme (see \citealt{Brosig2015Negation} and references therein). %\lv{Your response that this section might have been optional does not motivate the omission of the negative existential cycle. Its operation is so pervasive in Khalkha that it has to be mentioned here together with Croft's and my work.}

In the “Negative Existential Cycle” \citep{Croft1991}, syncronically speaking, Middle Mongolian occupies stage B{\textasciitilde}C (an existential negator making first inroads into verbal negation, still closer to stage B). Its daughter Khalkha Mongolian is harder to locate. On the one hand, it occupies a stage in which the cliticized (or suffixed) form =\textit{güi} of a former negative existential has taken over standard negation including the marking of the positive existential, displacing the particle \textit{ügüi} from nearly all of its former functions (B{\textasciitilde}C, almost completely attaining C). On the other hand, the pragmatically specific form \textit{alaɢ} ‘absent’ is making first inroads into the domain of possessive \citep[128-129]{Brosig2015Negation} and, secondarily and somewhat arguably, existential negation \citep[137]{Veselinova2013neg.exist}, suggesting stage C{\textasciitilde}A at a point still very close to C. Between the still poetically or dialectally usable constructions involving existential \textit{ügüi}, the complete reign of =\textit{güi} and the advent of \textit{alaɢ}, one might situate Khalkha at a stage B{\textasciitilde}\textbf{C}{\textasciitilde}A, confirming the high stability of transitionary stages (\citealt{VeselinovaHamari2022intro}: 13).


\section{Summary}\label{sec:kha:5} 

We observe several distinct negation strategies in Khalkha Mongolian, summarized in \tabref{tab:kha:summary}. The most frequent and multi-functional negator is the clitic\textit{ =güi}. Diachronically, it started out as an adjective indicating absence (without reference to a specific location) or possessive negation in predicates, attributes and adverbials. In the course of time, it has taken over standard negation and is currently only marginally used for negation of possession. As a standard negator, \textit{=güi} attaches to participles, thus neutralizing the evidential opposition between different past tense forms.
In the new locative-possessive-existential negation pattern, \textit{=güi} normally attaches to a participial form of the existential verb \textit{bai}-. Its longer and older form \textit{ügüi}, which would be used on its own, not following the positive existential verb, is only marginally attested. Both \textit{=güi} and \textit{ügüi} are used in negative answers and the second clauses of alternative questions. In the second clauses of embedded alternative propositions, the fossilized negator \textit{esex} can be used. The diachronically new negator \textit{alaɢ} ‘absent’ is used as a pragmatically conditioned alternative to \textit{bai-}\textsc{ptcp}\textit{=güi} and is mostly found in locative-possessive predications. The converbal suffix \textit{-lgüi} (based on deverbal noun-forming \textit{-l}) marks negative adverbial clauses.

\begin{table}
    \begin{tabularx}{\textwidth}{lQp{20mm}}
\lsptoprule
Negation marker & Functions & Section\\
\midrule 

V-\textsc{ptcp}=\textit{güi}, N=\textit{güi}
& negates finite predications &\ref{sec:kha:2.1}--\ref{sec:kha:2.2}, \ref{sec:kha:4.1}\\
&negates attributive predications & \ref{sec:kha:2.4.1} \\
&negates adverbial predications & \ref{sec:kha:2.4.2}, \ref{sec:kha:3.3.2}\\
&negates property predications & \ref{sec:kha:2.3.1}  \\
&negates alternative questions & \ref{sec:kha:2.2}   \\
& negative replies & \ref{sec:kha:3.1} \\
&privative case & \ref{sec:kha:3.3.2} \\
&derivational negation & \ref{sec:kha:3.3.1}--\ref{sec:kha:3.3.2}  \\

\midrule
(N) \textit{bai}-\textsc{ptcp}=\textit{güi},\newline
    (N) \textit{ügüi}
& \mbox{negates locative-possessive}-existential predications& \ref{sec:kha:2.3.2}\\
& negates comitative{\hyp}possessive predications & \ref{sec:kha:2.3.2}\\
& alternative questions & \ref{sec:kha:2.2}\\
& negative replies &  \ref{sec:kha:3.1}\\

(N) \textit{ügüi} 
& basis for negative derivation & \ref{sec:kha:3.3.1}  \\

\midrule
-\textit{lgüi} 
&  negates adverbial clauses & \ref{sec:kha:2.4.2}  \\

\midrule
(N) \textit{alaɢ}  %‘absent’
& \mbox{negates locative-possessive}-existential predications & \ref{sec:kha:2.3.2}  \\
& negates comitative-possessive predications & \ref{sec:kha:2.3.2}\\

\midrule
\textit{esex} 
& alternative propositions & \ref{sec:kha:4.5}  \\

\midrule
N \textit{biš} 
& negates identity / property in finite predicates &\ref{sec:kha:2.2}--\ref{sec:kha:2.3.1}, \ref{sec:kha:4.1}, \ref{sec:kha:4.6.2}\\
& metalinguistic negation & \ref{sec:kha:4.6.2}\\
& negative replies & \ref{sec:kha:3.1}\\
&basis for negative derivation  &\ref{sec:kha:3.3.1}\\
&connective ‘other’ & \ref{sec:kha:4.6.3}  \\

N \textit{bus} 
& negates identity / property in attributives & \ref{sec:kha:3.3.1}  \\

\midrule
\textit{bitgii} V-\textsc{2imp}, büü V-\textsc{2imp}
& express prohibition & \ref{sec:kha:2.2}   \\

\textit{bitgii}
&  expresses negative replies & \ref{sec:kha:3.1}  \\

\midrule
\textit{ül} V-, \textit{es} V- 
& derivational negation & \ref{sec:kha:3.3.1}  \\

\textit{es} V-\textit{wal} 
& conditional negation & \ref{sec:kha:3.3.1} \\
\lspbottomrule
    \end{tabularx}
    \caption{Khalkha Mongolian negators }
    \label{tab:kha:summary}
\end{table}

The ascriptive negator \textit{biš} has become the only negator for identity and the most common negator for property predications. It originates from the Middle Mongol word \textit{busi} ‘other’ and, probably due to that, it is the only device for meta\-linguistic negation. It is also used for the negation of participles and as a short standalone negative answer. \textit{Biš} is mostly used in predicates, while attributive identity negation is primarily encoded by the negator \textit{bus} which also derives from \textit{busi}. In addition to their uses as clausal negators, both \textit{(ü)güi} and \textit{biš} may function as nominal bases for derivation.

For imperatives, the preverbal usage of negators has been retained from \mbox{Middle} Mongol, but the innovative form \textit{bitgii} has reduced the original imperative negator \textit{büü} to a rare stylistic variant. If responding to a proposal that the interlocutor would like to act in a particular way, \textit{bitgii} on its own can be used for prohibition. The old pre-verbal standard negators \textit{ül} (imperfective) and \textit{es} (perfective) have been confined to lexical negation and here combine with particular verb stem classes regardless of tense-aspect marking.



\section*{Acknowledgements}
This paper was written during the project “Evidentiality, authority and information status - Investigating constructional networks and grammaticalization paths of quotative indices in Central Mongolic” funded by Academia Sinica (PI: Suying Hsiao). It was revised while working on the project ‘Evidentiality in time and space’ (PI: Fernando Zúñiga, Co-I: Marius Zemp) at the University of Bern and while working as assistant professor at Nazarbayev University. I am grateful to Dolgor Guntsetseg, Vlada Baranova, Hsiao Suying and two anonymous reviewers for their comments on earlier versions of this paper and to T. Erdene-Ochir for our discussion about \sectref{sec:kha:2.3.2} and \sectref{sec:kha:3.1}.

\section*{Abbreviations}
\largerpage[2]
\begin{tabularx}{.5\textwidth}{lQ}%larger than usual
(.) & short pause \\
\scshape 1 & first person \\
\scshape 2 & second person\\
\scshape 3 & third person\\
\scshape abl & ablative\\
\scshape acc & accusative\\
\scshape add & additive (focus)\\ 
\scshape all & allative\\
\scshape asrt & assertive (particle) \\
\scshape attr & attributive\\ 
\scshape aux & auxiliary\\
\scshape caus & causative\\
\scshape coll & collective\\
\scshape com & comitative\\
\scshape compl & completive\\
\scshape cond & conditional\\
\scshape cop & copula\\
\scshape cvb & converb\\
\scshape dat & dative\\
\scshape dim & diminutive\\
\scshape dp & discourse particle\\
\scshape dvn & de-verbal noun\\
\scshape emph & emphatic\\
\scshape est & established (present, past) \\
\scshape ex & existential \\
\scshape f & feminine \\
\scshape fh & firsthand (evidentiality) \\
\scshape foc & focus\\
\scshape fut & future \\
\scshape gen & genitive\\
\scshape hab & habitual\\
\scshape hon & honorific \\
\scshape hort & hortative\\
\scshape IC & Internet Corpus\\
\scshape ifr & information (question) \\
\scshape imfr & inferential \\
\scshape imm & immediate (present, past) \\
\scshape imp & imperative\\
\scshape impl & implore\\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape ins & instrumental\\
\scshape ints & intensifier\\
\scshape intr & intransitive\\
\scshape inj & interjection\\
\scshape ipfv & imperfective \\
\scshape lim & limitative (focus) \\
\scshape loc & locative \\
\scshape mnr & manner (converb) \\
\scshape mp & modal particle\\
\scshape name & proper name \\
\scshape neg & negation\\
\scshape nmlz & nominalizer\\
\scshape npst & nonpast\\
\scshape pass & passive\\
\scshape pfv & perfective\\
\scshape pl & plural\\
\scshape plr & polar (question) \\
\scshape poss & possessive\\
\scshape pot & potential\\
\scshape prf & perfect\\
\scshape prep & preparatory\\
\scshape priv & privative\\
\scshape prog & progressive\\
\scshape prox & proximal\\
\scshape prs & present\\
\scshape prosp & prospective\\
\scshape pst & past\\
\scshape ptcp & participle\\
\scshape q & question marker\\
\scshape quot & quotative verb\\
\scshape recp & reciprocal\\
\scshape res & resultative \\
\scshape rposs & reflexive possessive\\ 
\scshape SC & Spoken Corpus\\
\scshape sg & singular\\
\scshape stc & stance marker\\
\scshape tag & tag (question) \\
\scshape top & topic\\
\\
\end{tabularx}


\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
\end{document} 
