\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18234083}
\author{Jack B. Martin\affiliation{William \& Mary}}
\title{Negation in Koasati}
\abstract{Koasati has a complex negative pattern with three markers on the verb whose placement is determined by verb class. The complexity of the morphological system has hindered understanding of negation as a whole in the language. This paper thus explores the larger topic of Koasati negation in typological perspective, with special attention to tone, aspect, and the origin of the system.
\keywords{negation, Koasati, Muskogean, infixation}
}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 

\section{The language}\label{sec:koa:1}

Koasati (ISO: \href{https://iso639-3.sil.org/code/cku}{cku}, Glottocode: \href{https://glottolog.org/resource/languoid/id/koas1236}{koas1236}) is a Muskogean language spoken by 200 members of the Coushatta Tribe of Louisiana and by a portion of the Alabama-Coushatta Tribe of Texas. It is closely related to Alabama, and more distantly related to Choctaw, Chickasaw, Apalachee, Mikasuki, and Muskogee (Creek). The data in this paper are from elicited data collected while developing pedagogical materials with the Coushatta Heritage Department. The structure of the paper follows the negation questionnaire provided by the editors (\citetv{chapters/questionnaire}).

The vowel phonemes of Koasati are short /a i o/, long /a: i: o:/, and nasal\linebreak /aⁿ iⁿ oⁿ/.
The consonant phonemes are /b č f h k l ɬ m n p s t w y/. /č/ and /y/ are phonetically [ʧ] and [j] in IPA. Within a word, acute, grave, caron, and circumflex accents are used for tone (generally indicating verbal aspect). Phrase-final intonation patterns are indicated with ʹ (high) and \^{} (falling) after a word.

Koasati generally has SOV word order \REF{ex:koa:1}.

\ea\label{ex:koa:1}
    \gll a:t-osi-k ɬaɬo-ⁿ ì:p. \\
         person-\textsc{dim}-\textsc{sbj} fish-\textsc{nsbj} eat.\textsc{lgr}\\
    \glt ‘The child is eating the fish.’
\z

Enclitics appear after noun phrases to indicate case or discourse status: \textit{-k} for subject (with a focused form \textit{-ok}), -\textit{ⁿ} for nonsubject (focused form -\textit{on}), and \textit{-p} for topic, among others. In conversation, subjects, objects, and other phrases may be placed after the verb as afterthoughts.

Verbs are highly synthetic and are marked for the person and number of their arguments. Koasati distinguishes the three series of person markers in (\ref{ex:koa:2}--\ref{ex:koa:4}): agent (\textsc{a}), patient (\textsc{p}), and dative (\textsc{dat}).\footnote{For simplicity, verbs are cited here in future forms ending in -\textit{lahoⁿ}.}

\TabPositions{.33\textwidth,.66\textwidth}
\ea\label{ex:koa:2} Agent series (stem \textit{hiča}- ‘see’): \\ 
        \textit{hiča-\textbf{l-}lahoⁿ} \tab ‘I will see it’\footnote{In \REF{ex:koa:2}, \textit{hiča-l-lahoⁿ} is glossed see-1\textsc{sg}.\textsc{a}{}-\textsc{fut}.\textsc{sf}, and \textit{is-hiča-lahoⁿ} is glossed 2\textsc{sg}.\textsc{a}-see-\textsc{fut}.\textsc{sf}.}  \\
        \textit{\textbf{is-}hiča-lahoⁿ} \tab ‘you will see it’ \\
        \textit{hiča-lahoⁿ} \tab ‘he/she will see it’ \\
        \textit{\textbf{il-}hiča-lahoⁿ} \tab ‘we will see it’ \\
        \textit{\textbf{has-}hiča-lahoⁿ} \tab ‘you all will see it’ \\
\z  

\ea\label{ex:koa:3} Patient series (stem \textit{banna}- ‘want’): \\
    \textit{\textbf{ča-}banna-lahoⁿ} \tab ‘I will want it’\footnote{In \REF{ex:koa:3}, \textit{ča-banna}\textbf{{}-}\textit{lahoⁿ} is glossed 1\textsc{sg.p}{}-want-\textsc{fut}.\textsc{sf}.} \\
    \textit{\textbf{či-}banna-lahoⁿ} \tab ‘you will want it’ \\
    \textit{banna-lahoⁿ} \tab ‘he/she will want it’ \\
    \textit{\textbf{ko-}banna-lahoⁿ} \tab ‘we will want it’ \\
    \textit{\textbf{hači-}banna-lahoⁿ} \tab ‘you all will want it’ \\
\z 

\ea\label{ex:koa:4} Dative series (stem \textit{im-aka:no}- ‘be hungry’):\\ 
    \textit{\textbf{am-}aka:no-lahoⁿ} \tab ‘I will be hungry’\footnote{In \REF{ex:koa:4} \textit{am-aka:no-lahoⁿ} is glossed 1\textsc{sg}.\textsc{dat}{}-be.hungry-\textsc{fut}.\textsc{sf}.} \\
    \textit{\textbf{čim-}aka:no-lahoⁿ} \tab ‘you will be hungry’ \\
    \textit{\textbf{im-}aka:no-lahoⁿ} \tab ‘he/she will be hungry’ \\
    \textit{\textbf{kom-}aka:no-lahoⁿ} \tab ‘we will be hungry’ \\
    \textit{\textbf{hačim-}aka:no-lahoⁿ} \tab ‘you all will be hungry’ \\
\z 

\newpage
Person markers are prefixed in the patient and dative series. In the agent series, the first person singular -\textit{li} (or \textit{-l}) is always a suffix. The other agent person markers may be prefixed, infixed, or suffixed depending on verb class.\footnote{Verb classes are established based on the placement and form of the agent person markers (if the verb is agentive) or, as we will see, on the placement and form of negative \textit{ik-}. Verb class also affects the form of verbal nouns (\citealt[273ff]{Kimball1991}). \citet{Haas1946} described six verb classes in Koasati (I, IIA, IIB, IIC, IIIA, IIIB). \citet{Kimball1991,Kimball1994} recognized ten classes (1A, 1B, etc.), though some of his classes have only a few members. For reasons of space, I focus here on four classes. See also \citet{MontlerHardy1991} for a description of negation in Alabama.}

The verb \textit{hiča-} ‘see’ in \REF{ex:koa:2}, for example, belongs to a small class of verbs (“SEE verbs” or Kimball’s class 1A) that prefix agent person markers other than the first person singular.

Verbs belonging to the SING class (class 2C) differ in infixing these markers. This is seen with the stem \textit{talwa}- ‘sing’ in \REF{ex:koa:5}.

\ea\label{ex:koa:5} 
    \textit{talwa-\textbf{l}-lahoⁿ} \tab ‘I will sing’ \\
    \textit{ta<\textbf{či}>lwa-lahoⁿ} \tab ‘you will sing’ \\
    \textit{talwa-lahoⁿ} \tab ‘he/she will sing’ \\
    \textit{ta<\textbf{li}>lwa-lahoⁿ} \tab ‘we will sing’ \\
    \textit{ta<\textbf{hači}>lwa-lahoⁿ} \tab ‘you all will sing’ \\
\z

Most other verbs are suffixing, but belong to several classes based on the specific form of the markers. The stem \textit{bitli}- ‘dance’ in \REF{ex:koa:6} (class 2Aii) is typical of the large number of verb roots ending in -\textit{li}.\footnote{The same pattern is used for a number of verbs like \textit{isi-} ‘take’ that don’t end in \textit{{}-li} (Kimball’s class 2Ai).}

\ea\label{ex:koa:6} 
    \textit{bitli-\textbf{l}-lahoⁿ} \tab ‘I will dance’ \\
    \textit{bit-\textbf{či}-lahoⁿ} \tab ‘you will dance’ \\
    \textit{bitli-lahoⁿ} \tab ‘he/she will dance’ \\
    \textit{bit-\textbf{hil}-lahoⁿ} \tab ‘we will dance’ \\
    \textit{bit-\textbf{hači}-lahoⁿ} \tab ‘you all will dance’ \\
\z

The stem \textit{na:ɬi:ka}- ‘speak’ in \REF{ex:koa:7} is another suffixing verb, this time typical of verbs ending in -\textit{ka} (class 3A).

\ea\label{ex:koa:7}
    \textit{na:ɬi:ka-\textbf{l}-lahoⁿ} \tab ‘I will speak’ \\
    \textit{na:ɬi:-\textbf{hiska}-lahoⁿ} \tab ‘you will speak’ \\
    \textit{na:ɬi:ka-lahoⁿ} \tab ‘he/she will speak’ \\
    \textit{na:ɬi:-\textbf{hilka}-lahoⁿ} \tab ‘we will speak’ \\
    \textit{na:ɬi:-\textbf{haska}-lahoⁿ} \tab ‘you all will speak’ \\
\z 

Koasati has a system of \textsc{grades} or internal changes typically applying to the penultimate syllable of a verb stem and affecting verbal aspect. As \REF{ex:koa:8} shows, a verb stem like \textit{hača:li-} ‘stand’ can thus appear in the lengthened grade (L-grade, \textsc{lgr}) to indicate an event in progress, or in the geminating-rising tone grade (GR-grade, \textsc{ggr}) to indicate a state resulting from an event. 

\protectedex{\ea\label{ex:koa:8}
    \ea\label{ex:koa:8a}
        \gll hačà:lí-l. \\ 
             stand.\textsc{sg}.\textsc{lgr}-1\textsc{sg.a} \\
        \glt ‘I am standing up.’ (event; L-grade)
    \ex\label{ex:koa:8b}
        \gll haččǎ:li-l. \\
             stand.\textsc{sg}.\textsc{ggr}-1\textsc{sg.a} \\
        \glt ‘I am standing.’ (resulting state; GR-grade)
    \z
\z 
}
Other grades include the rising tone grade (R-grade), the aspirating grade (H-grade), and the nasalizing grade (N-grade).

In questions and emphatic environments, the final vowels of verbs may be nasalized \REF{ex:koa:9a}. Final oral vowels are deleted in statements and commands \REF{ex:koa:9b}.

\ea\label{ex:koa:9}
    \ea\label{ex:koa:9a}
        \gll hačà:lí-ⁿ.\\ 
             stand.\textsc{sg}.\textsc{lgr}-\textsc{sf} \\
        \glt ‘Is he/she standing up?’
    
    \ex\label{ex:koa:9b}
        \gll hačà:l. \\
             stand.\textsc{sg}.\textsc{lgr}\\
        \glt ‘He/she is standing up.’
    \z
\z 

Koasati has a measured tense system with several degrees of remoteness in past time. \tabref{tab:koa:1} shows four past tenses with forms used in statements (with loss of final vowel), questions, and emphatic statements.\footnote{Past 1 is glossed in examples as \textsc{rec.pst} and past 2 is glossed as \textsc{med.pst}.}


\begin{table}
\caption{Past tense suffixes}
\label{tab:koa:1}
\begin{tabular}{lllll}
\lsptoprule
& & Statement & Question & Emphasis \\ 
\midrule 
\multicolumn{2}{l}{Past 1 (today/last night)} &&& \\
& Perfective                      & \textit{-t}    & no mark   & \textit{-tiⁿ\^{}} \\
& Imperfective                    & \textit{-:s}   &  \textit{-:saⁿʹ}   & \\
\multicolumn{2}{l}{Past 2 (yesterday/a year ago)}   &  \textit{-t}   &  -\textit{toⁿʹ}    &\\
\multicolumn{2}{l}{Past 3 (long ago)}              & \textit{-k}    & -\textit{kiⁿʹ}     & -\textit{kiⁿ\^{}} \\
\multicolumn{2}{l}{Past 4 (very long ago)}          & \textit{-toⁿ} & -\textit{to:liⁿʹ} & \\
\lspbottomrule
\end{tabular}
\end{table}


\section{Clausal negation}\label{sec:koa:2}

Many aspects of Koasati negation have been described in \citet{Kimball1991}.\footnote{\citet{Kimball1991} is particularly thorough in investigating verb classes associated with negation. \citet{Kimball1994} is a dictionary indicating class membership.} This description reflects further research on Koasati tone and grades and describes negation in a wider typological context (\citealt{Dahl1979,Miestamo2005,chapters/questionnaire}).

Negation in Koasati is restricted to verbs. It is unusual in being signaled by three separate markers on the verb: an affix \textit{ik-}, a suffix \textit{-o}, and use of rising tone. Of these, the affix \textit{ik-} is the most reliable and salient mark of negation.\footnote{I say this because \textit{{}-o} sometimes deletes and because rising tone can be difficult to hear.}

\subsection{Standard negation}\label{sec:koa:2.1}

Standard negation (negation of indicative clauses) is a morphological operation in Koasati applying to verbs. In \REF{ex:koa:10}, the verb stem is \textit{ipa}- ‘eat’.\footnote{When rising tone falls on a syllable closed by an obstruent (as in \textit{ǐk-p-o-ⁿ} in \REF{ex:koa:10b}), the rise is often carried over to the following syllable: \textit{ìk-p-ó-ⁿ}.}


\ea\label{ex:koa:10}
    \ea\label{ex:koa:10a}
        \gll a:t-osi-k  ɬaɬo-ⁿ  ì:p. \\ 
             person-\textsc{dim}-\textsc{sbj}  fish-\textsc{nsbj}  eat.\textsc{lgr} \\
        \glt ‘The child is eating the fish.’
    \ex\label{ex:koa:10b}
        \gll a:t-osi-k  ɬaɬo-ⁿ \textbf{ǐk}-p-\textbf{o}-ⁿ.\\
             person-\textsc{dim}-\textsc{sbj}  fish-\textsc{nsbj}  \textsc{neg}-eat.\textsc{rgr}-\textsc{neg}-\textsc{sf}\\
        \glt ‘The child is not eating the fish.’
    \z
\z

Three morphological operations apply to make a verb negative:

\begin{enumerate}
    \item a form of the negative affix \textit{ik-} is prefixed, infixed, or suffixed (depending on verb class);
    \item a negative suffix \textit{-o} is added;
    \item the verb stem is put in the rising tone grade. Rising tone is assigned to the syllable before \textit{-o}. If the vowel in that syllable is short and open, it is lengthened.
\end{enumerate}

Negative \textit{ik-} and \textit{-o} are unique to negation and can be considered a type of two-part negation. The final -\textit{o} may delete before another vowel, but a form of \textit{ik}- is always present. The rising tone grade is a grammatical aspect limited to negatives, some nominalizations, and as part of the geminating-rising tone grade (which indicates a resulting state). These three uses suggest the rising tone grade was once a type of imperfective, but the three uses have little to do with each other now, and the rising tone grade helps distinguish a negative verb.

Whether negative \textit{ik-} is prefixed, infixed, or suffixed depends on verb class. The placement of \textit{ik-} is thus parallel to the placement of the agent person markers. In fact, the agent person markers and \textit{ik}- are fused as a separate series of negative agent person markers. Verbs like \textit{hiča}- ‘see’ that prefix the positive agent person markers also prefix the negative agent person markers \REF{ex:koa:11}.\footnote{As \citet[129]{Miestamo2005} has noted, this is an asymmetric system, because agentive person marking is expressed differently in positive and negative clauses.} 

\ea
\label{ex:koa:11}
    \textit{hiča-\textbf{l-}lahoⁿ} \tab ‘I will see it’ \\
    \textit{\textbf{is}-hiča-lahoⁿ} \tab ‘you will see it’ \\
    \textit{hiča-lahoⁿ} \tab ‘he/she will see it’ \\
    \textit{\textbf{il}-hiča-lahoⁿ} \tab ‘we will see it’ \\
    \textit{\textbf{has}-hiča-lahoⁿ} \tab ‘you all will see it’ \\
\vspace{2mm}
    \textit{\textbf{ak}-hǐ:č-o-lahoⁿ} \tab ‘I won’t see it’ \\
    \textit{\textbf{čik}-hǐ:č-o-lahoⁿ} \tab ‘you won’t see it’ \\
    \textit{\textbf{ik}-hǐ:č-o-lahoⁿ} \tab ‘he/she won’t see it’ \\
    \textit{\textbf{kil}-hǐ:č-o-lahoⁿ} \tab ‘we won’t see it’ \\
    \textit{\textbf{hačik}-hǐ:č-o-lahoⁿ} \tab ‘you all won’t see it’ \\
\z 


Verbs like \textit{talwa}- ‘sing’ that infix the agent person markers also infix the negative agent person markers \REF{ex:koa:12}.

\ea
\label{ex:koa:12}
    \textit{talwa-\textbf{l}-lahoⁿ} \tab ‘I will sing’ \\
    \textit{ta<\textbf{či}>lwa-lahoⁿ} \tab ‘you will sing’ \\
    \textit{talwa-lahoⁿ} \tab ‘he/she will sing’ \\
    \textit{ta<\textbf{li}>lwa-lahoⁿ} \tab ‘we will sing’ \\
    \textit{ta<\textbf{hači}>lwa-lahoⁿ} \tab ‘you all will sing’ \\
\vspace{2mm}
    \textit{ta<\textbf{kǎ}>lw-o-lahoⁿ} \tab ‘I won’t sing’ \\
    \textit{ta<\textbf{čikǐ}>lw-o-lahoⁿ} \tab ‘you won’t sing’ \\
    \textit{ta<\textbf{kǐ}>lw-o-lahoⁿ} \tab ‘he/she won’t sing’ \\
    \textit{ta<\textbf{likǐ}>lw-o-lahoⁿ} \tab ‘we won’t sing’ \\
    \textit{ta<\textbf{hačikǐ}>lw-o-lahoⁿ} \tab ‘you all won’t sing’ \\
\z

Verbs like \textit{bitli}- ‘dance’ in \REF{ex:koa:13} and \textit{na:ɬi:ka-} ‘speak’ in \REF{ex:koa:14} that suffix the agent person markers also suffix the negative agent person markers. 

\protectedex{\ea
\label{ex:koa:13}
    \textit{bitli-\textbf{l}-lahoⁿ} \tab ‘I will dance’ \\
    \textit{bit-\textbf{či}-lahoⁿ} \tab ‘you will dance’ \\
    \textit{bitli-lahoⁿ} \tab ‘he/she will dance’ \\
    \textit{bit-\textbf{hil}-lahoⁿ} \tab ‘we will dance’ \\
    \textit{bit-\textbf{hači}-lahoⁿ} \tab ‘you all will dance’ \\
\vspace{2mm}
    \textit{bit-\textbf{tǎkko}-lahoⁿ} \tab ‘I won’t dance’ \\
    \textit{bit-\textbf{čǐkko}-lahoⁿ} \tab ‘you won’t dance’ \\
    \textit{bǐt-\textbf{ko}-lahoⁿ} \tab ‘he/she won’t dance’ \\
    \textit{bit-\textbf{kǐlko}-lahoⁿ} \tab ‘we won’t dance’ \\
    \textit{bit-\textbf{hačǐkko}-lahoⁿ} \tab ‘you all won’t dance’ \\
\z
}

\ea\label{ex:koa:14}
    \textit{na:ɬi:ka-\textbf{l}-lahoⁿ} \tab ‘I will speak’ \\
    \textit{na:ɬi:-\textbf{hiska}-lahoⁿ} \tab ‘you will speak’ \\
    \textit{na:ɬi:ka-lahoⁿ} \tab ‘he/she will speak’ \\
    \textit{na:ɬi:-\textbf{hilka}-lahoⁿ} \tab ‘we will speak’ \\
    \textit{na:ɬi:-\textbf{haska}-lahoⁿ} \tab ‘you all will speak’ \\
\vspace{2mm}
    \textit{na:ɬi:-\textbf{hǎkko}-lahoⁿ} \tab ‘I won’t speak’ \\
    \textit{na:ɬi:-\textbf{čǐkko}-lahoⁿ} \tab ‘you won’t speak’ \\
    \textit{na:ɬi:-\textbf{hǐkko}-lahoⁿ} \tab ‘he/she won’t speak’ \\
    \textit{na:ɬi:-\textbf{kǐlko}-lahoⁿ} \tab ‘we won’t speak’ \\
    \textit{na:ɬi:-\textbf{hačǐkko}-lahoⁿ} \tab ‘you all won’t speak’ \\
\z

Many verbs use the agent series for their subjects, but some use the patient or dative series instead. For these, the person markers are not fused with negative \textit{ik-}. Instead, negation has the same form as the third person negative. The verb \textit{banna}- ‘want’, for example, uses the patient series for its subject, but as \REF{ex:koa:15} shows, negation uses the same pattern as \textit{bitli}- ‘dance’ in the third person.

\ea\label{ex:koa:15}
    \textit{\textbf{ča-}banna-lahoⁿ} \tab ‘I will want it’  \\
    \textit{\textbf{či-}banna-lahoⁿ} \tab ‘you will want it’ \\
    \textit{banna-lahoⁿ} \tab ‘he/she will want it’ \\
    \textit{\textbf{ko-}banna-lahoⁿ} \tab ‘we will want it’ \\
    \textit{\textbf{hači-}banna-lahoⁿ} \tab  ‘you all will want it’ \\
\vspace{2mm}
    \textit{\textbf{ča}-bǎn-\textbf{ko}-lahoⁿ} \tab  ‘I won’t want it’ \\
    \textit{\textbf{či}-bǎn-\textbf{ko}-lahoⁿ} \tab  ‘you won’t want it’ \\
    \textit{bǎn-\textbf{ko}-lahoⁿ} \tab ‘he/she won’t want it’ \\
    \textit{\textbf{ko}-bǎn-\textbf{ko}-lahoⁿ} \tab ‘we won’t want it’ \\
    \textit{\textbf{hači}-bǎn-\textbf{ko}-lahoⁿ} \tab ‘you all won’t want it’ \\
\z

The verb \textit{im-aka:no}- ‘hungry’ uses the dative series for its subject. Again, as \REF{ex:koa:16} shows, negation for this verb is as for \textit{bitli}- ‘dance’.

\ea\label{ex:koa:16}
    \textit{\textbf{am-}aka:no-lahoⁿ} \tab ‘I will be hungry’ \\
    \textit{\textbf{čim-}aka:no-lahoⁿ} \tab ‘you will be hungry’ \\
    \textit{\textbf{im-}aka:no-lahoⁿ} \tab ‘he/she will be hungry’ \\
    \textit{\textbf{kom-}aka:no-lahoⁿ} \tab ‘we will be hungry’ \\
    \textit{\textbf{hačim-}aka:no-lahoⁿ} \tab ‘you all will be hungry’ \\
\vspace{2mm}
    \textit{\textbf{am}-akǎn-\textbf{ko}-lahoⁿ} \tab ‘I won’t be hungry’\footnote{\citet[73]{Kimball1991} records infixation with this verb: \textit{am-aka<}\textbf{\textit{k\v{i}:}}\textit{>n-}\textbf{\textit{o}}\textit{ⁿ} ‘I am not hungry’. Speakers sometimes use different conjugation patterns with the same verb and sometimes also switch patterns between persons. Note that the root has a short vowel in negative forms like \textit{am-akǎn-ko-lahoⁿ} ‘I won’t be hungry’. Koasati does not normally permit long vowels in syllables closed by sonorants.} \\
    \textit{\textbf{č}\textbf{im}-akǎn-\textbf{ko}-lahoⁿ} \tab ‘you won’t be hungry’ \\
    \textit{\textbf{im}-akǎn-\textbf{ko}-lahoⁿ} \tab ‘he/she won’t be hungry’ \\
    \textit{\textbf{kom}-akǎn-\textbf{ko}-lahoⁿ} \tab ‘we won’t be hungry’ \\
    \textit{\textbf{ha}\textbf{č}\textbf{im}-akǎn-\textbf{ko}-lahoⁿ} \tab ‘you all won’t be hungry’ \\
\z 

A third feature of negative clauses is the use of the rising tone grade. As the examples in (\ref{ex:koa:11}--\ref{ex:koa:16}) show, rising tone is assigned to the syllable preceding \textit{-o}, either on the verb root \REF{ex:koa:17} or on an affix \REF{ex:koa:18}.

\ea\label{ex:koa:17}
\textit{banna-lahoⁿ} \tab ‘he/she will want it’ \\
\textit{bǎn-\textbf{ko}-lahoⁿ} \tab ‘he/she won’t want it’ \\
\z

\ea\label{ex:koa:18}
\textit{na:ɬi:-\textbf{hilka}-laho}ⁿ \tab ‘we will speak’ \\
\textit{na:ɬi:-\textbf{kǐlko}-lahoⁿ} \tab ‘we won’t speak’ \\
\z
Rising tone leads to lengthening of short vowels in open syllables \REF{ex:koa:19}.

\ea\label{ex:koa:19}
\textit{hiča-lahoⁿ} \tab ‘he/she will see it’ \\
\textit{\textbf{ik}-hǐ:č-\textbf{o}-lahoⁿ} \tab ‘he/she won’t see it’ \\
\z
The rising tone grade is a restricted form of imperfective (limited to negatives, some nominalizations, and as a component of the geminating-rising tone grade). In the present tense, a positive verb may be in the lengthened grade for an event or in the geminating-rising tone grade for a resulting state. In the negative, however, only the rising tone grade is possible, so this  distinction is lost (\tabref{tab:koa:2}).

\begin{table}
    \centering
    \begin{tabular}{lll} 
\lsptoprule
        Event (L-grade) & \textit{čokò:lí-l} & ‘I’m getting seated’ \\
        Resulting state (GR-grade) & \textit{čokkǒ:li-l} & ‘I’m sitting’ \\
        Negative (R-grade) & \textit{čoko:-tǎkko-ⁿ} & ‘I’m not sitting’ \\ 
\lspbottomrule
    \end{tabular}
    \caption{Present (stem \textit{čoko:li}- ‘get seated’)}
    \label{tab:koa:2}
\end{table}


The past 1 time frame (for earlier today up to last night) differs. If an event was perfective (successfully completed once), a suffix -\textit{ti} is used with the lengthened grade. If an event was ongoing, a resulting state, or a negative, then a suffix \textit{-:sa} is used with different grades (\tabref{tab:koa:3}).\footnote{For reasons of space, I describe only the present and past 1 time frames here.}

\begin{table}
    \centering
    \begin{tabular}{lll} 
\lsptoprule
       Perfective (L-grade + -\textit{ti}) & \textit{čokò:lí-li-t} & ‘I sat down’ \\
        Event (L-grade + -\textit{:sa}) & \textit{čokò:lí-li-:s} & ‘I was getting seated’ \\
        Resulting state (GR-grade + -\textit{:sa}) & \textit{čokkǒ:li-li-:s} & ‘I was seated’ \\
        Negative (R-grade + -\textit{:sa}) & \textit{čoko:-tǎkko-:s} & ‘I didn’t sit’ \\
\lspbottomrule
    \end{tabular}
    \caption{Past 1 (stem \textit{čoko:li}- ‘get seated’)}
    \label{tab:koa:3}
\end{table}


In examining the present and past 1 forms in \tabref{tab:koa:2} and \tabref{tab:koa:3}, it is clear that negative verbs have more limited aspectual distinctions. This may be due to the fact that negative actions are inherently incomplete and thus inherently imperfective. Negative verbs also occur in a specific grade, however, which means that other grade distinctions are not possible.


\subsection{Negation in non-declaratives}\label{sec:koa:2.2}
The same form of negation is used in statements and in questions (though questions have higher intonation at the end) \REF{ex:koa:20}.


\ea\label{ex:koa:20}
    \ea\label{ex:koa:20a}
        \gll ik-hǐ:č-o-ⁿ.\\ 
             \textsc{neg}-see.\textsc{rgr}-\textsc{neg}-\textsc{sf} \\
        \glt ‘He/she doesn’t see it.’
    
    \ex\label{ex:koa:20b}
        \gll ik-hǐ:č-o-ⁿʹ? \\
             \textsc{neg}-see.\textsc{rgr}-\textsc{neg}-\textsc{sf} \\
        \glt ‘Doesn’t he/she see it?’
    \z
\z 

Koasati has a dedicated construction for negative commands \citep[263]{Kimball1991}. In this pattern, a prohibitive suffix -\textit{nna} is added to the agentive second person singular or second person plural. The forms in (\ref{ex:koa:21}--\ref{ex:koa:23}) are of the stem \textit{talwa}- ‘sing’.

\ea\label{ex:koa:21}
    \gll talw\^{}! \\
         sing \\
    \glt ‘Sing!’
\z

\ea\label{ex:koa:22}
    \gll ta<či>lwa-\textbf{nna}-ⁿ\^{}! \\
         sing<2\textsc{sg.a}>-\textsc{proh}-\textsc{sf} \\
    \glt ‘Don’t you sing!’
\z

\ea\label{ex:koa:23}
    \gll ta<hači>lwa-\textbf{nna}-ⁿ\^{}! \\
         sing<2\textsc{pl.a}>-\textsc{proh}-\textsc{sf} \\
    \glt ‘Don’t you all sing!’
\z

\subsection{Negation in stative predications}\label{sec:koa:2.3}
Stative predications in Koasati can be divided into copular predications, resultative stative predications, and existential predications.

Copular predications use the verb \textit{ommi}- ‘be’. The regular negative form is \textit{ǒn-ko-},  but it is often shortened to \textit{ko-}. Copular predications are used for equative sentences \REF{ex:koa:24}:


\ea\label{ex:koa:24}
    \ea\label{ex:koa:24a}
        \gll y-ok  ič-ok  om. \\ 
             this-\textsc{foc}.\textsc{sbj}  deer-\textsc{foc}.\textsc{sbj}  be \\
        \glt ‘This is a deer.’
    
    \ex\label{ex:koa:24b}
        \gll y-ap  ič-ok  ko-ⁿ. \\
             this-\textsc{top}  deer-\textsc{foc}.\textsc{sbj}  be.\textsc{neg}-\textsc{sf} \\
        \glt ‘This is not a deer.’
    \z
\z 

Attributes like \textit{homma} ‘red’ are verbs in Koasati. These words may be used to modify nouns: \textit{čokbani h\v{o}mma} ‘red ant’. When used as stative predicates, however, they are generally turned into participles (deverbal nouns) by adding -\textit{hči}. The verb \textit{ommi}- ‘be’ is then used as with equative sentences \REF{ex:koa:25}:


\ea\label{ex:koa:25}
    \ea\label{ex:koa:25a}
        \gll pokko-k  hommá-hč-ok  om. \\ 
             ball-\textsc{sbj}  red-\textsc{ptcp}-\textsc{foc}.\textsc{sbj}  be \\
        \glt ‘The ball is red.’
    
    \ex\label{ex:koa:25b}
        \gll pokko-k  hommá-hč-ok  ko-ⁿ \\
             ball-\textsc{sbj}  red-\textsc{ptcp}-\textsc{foc}.\textsc{sbj}  be.\textsc{neg}-\textsc{sf} \\
        \glt ‘The ball is not red.’
    \z
\z 

Some verbs use the lengthened grade for an event and the geminating-rising tone grade for a resulting state. The negative is the same for both forms, so this aspectual distinction is lost \REF{ex:koa:26}.



\ea\label{ex:koa:26}
    \ea\label{ex:koa:26a}
        \gll čokò:lí-l. \\ 
             sit.\textsc{sg.lgr}-1\textsc{sg.a} \\
        \glt ‘I am getting seated.’ (event)
    
    \ex\label{ex:koa:26b}
        \gll čokk\v{o}:li-l. \\
             sit.\textsc{sg.ggr}-1\textsc{sg.a}\\
        \glt ‘I am seated.’ (resulting state)
    \ex\label{ex:koa:26c}
        \gll čoko:-t\v{a}kko-ⁿ. \\
             sit.\textsc{sg}-1\textsc{sg.a.neg.rgr-sf} \\
        \glt ‘I am not seated/getting seated.’
    \z
\z  


Existential predications express quantity (‘we are three’), existence, and ‘have’ sentences. To express existence, the verb \textit{nǎ:ho-} ‘exist’ is used \REF{ex:koa:27}, with an unexpected negative form \textit{ǐkso-}. Lexicalized negative existentials of this type are common in other languages, as \citet{Veselinova2013neg.exist} has discussed.


\ea\label{ex:koa:27}
    \begin{xlist}[A: ]
    \exi{Q: }\label{ex:koa:27a}
        \gll hapi-k  nǎ:ho-ⁿʹ? \\ 
             salt-\textsc{sbj}  exist.\textsc{ggr}-\textsc{sf} \\
        \glt ‘Is there salt?’
    \exi{A: }\label{ex:koa:27b}
        \gll hapi-k  ǐkso-ⁿ .\\
             salt-\textsc{sbj}  \textsc{negex}.\textsc{rgr}-\textsc{sf} \\
        \glt ‘There is no salt.’
    \end{xlist}
\z

The same verbs are used to express ‘have’ and ‘not have’ \REF{ex:koa:28}. In this use, a dative prefix \textit{im}- (or \textit{in}-) is added to the verb.

\ea\label{ex:koa:28}
    \ea\label{ex:koa:28a}
        \gll a:t-osi-k  ɬaɬo-k  in-nǎ:ho-:s. \\ 
             person-\textsc{dim}-\textsc{sbj}  fish-\textsc{sbj}  \textsc{dat}-exist.\textsc{ggr}-\textsc{rec.pst}.\textsc{ipfv} \\
        \glt ‘The child had a fish.’ (lit. ‘Child, fish to-her-existed.’)
    
    \ex\label{ex:koa:28b}
        \gll a:t-osi-k  ɬaɬo-k   im-ǐkso-:s. \\
             person-\textsc{dim}-\textsc{sbj}  fish-\textsc{sbj}  \textsc{dat}-\textsc{neg}.exist.\textsc{rgr}-\textsc{rec.pst}.\textsc{ipfv} \\
        \glt ‘The child didn’t have a fish.’
    \z
\z 

Location is generally expressed with posture verbs \REF{ex:koasati:new29}. These posture verbs use standard negation.

\ea\label{ex:koasati:new29}
    \gll pokko-k  mobǐ:la  obǎ:li-ⁿ           čokkǒ:l.\\
         ball-\textsc{sbj}  car          behind-\textsc{nsbj}  sit.\textsc{sg.ggr}\\
    \glt ‘The ball is (sitting) behind the car.’\\
\z 


\subsection{Negation in non-main clauses}\label{sec:koa:2.4}

Any type of clause may be negated in Koasati, and negation is done the same way in main clauses and non-main clauses. Negation is limited to verbs, however. (\ref{ex:koa:29}--\ref{ex:koa:30}) show negation on verbs in adverbial clauses.

\ea\label{ex:koa:29}
    \gll {falaⁿ-}\textbf{hǐkko}-ska-n,  ho<kǐ>kf-o-:s. \\ 
         wake-\textsc{neg}-because-\textsc{ds}  wake.\textsc{neg}.\textsc{rgr}-\textsc{neg}-\textsc{rec.pst}.\textsc{ipfv} \\
    \glt  ‘Because he didn’t wake up, he didn’t get dressed.’
\z 

\ea\label{ex:koa:30}
    \gll o<\textbf{kǐ:}>b-o-:p. \\ 
         rain<\textsc{neg}>.\textsc{rgr}-\textsc{neg}-\textsc{top} \\
    \glt ‘If it doesn’t rain …’
\z 

As \citet[239]{Kimball1991} notes, the suffix -\textit{ha:lo}- ‘while’ usually appears with negatives \REF{ex:koa:31}.


\ea\label{ex:koa:31}
    \gll hopon-tǎkk-o-ha:l-ok… \\ 
         cook-1\textsc{sg.a}.\textsc{neg}-\textsc{neg}-while-\textsc{foc}.\textsc{ss} \\
    \glt ‘Before I cook…’ (lit. ‘While I do not cook ...’)
\z 

Koasati makes little use of non-finite clauses. The verb \textit{banna}- ‘want, try’ is one verb that can appear with both finite and non-finite complements for different meanings (\ref{ex:koa:32}--\ref{ex:koa:33}).


\ea\label{ex:koa:32}
    \gll aɬɬa-l-a:hi:-k  ča-bànnǎ-:s. \\ 
         go.\textsc{sg}-1\textsc{sg.a}-going.to-\textsc{ss}  1\textsc{sg.p}-want.\textsc{lgr}-\textsc{rec.pst}.\textsc{ipfv} \\
    \glt ‘I wanted to go.’ (lit. ‘I wanted that I am going to go.’)
\z 

\ea\label{ex:koa:33}
    \gll aɬɬa-h  ča-bànnǎ-:s. \\ 
         go.\textsc{sg}-\textsc{inf}  1\textsc{sg.p}-want.\textsc{lgr}-\textsc{rec.pst}.\textsc{ipfv} \\
    \glt ‘I tried to go.’
\z 
The complement clause in \REF{ex:koa:32} is marked for first person singular and can be considered finite. The complement clause in \REF{ex:koa:33} has no marking of the subject and no obvious tense or aspect. I therefore consider the complement clause in \REF{ex:koa:33} to be non-finite.

As \REF{ex:koa:34} shows, speakers accept negative versions of non-finite clauses.

\ea\label{ex:koa:34}
    \ea\label{ex:koa:34a}
        \gll Skye-ka-k  tabak-čampoli-ⁿ ipa-h  bànná-t.\\
             Skye-\textsc{ln}-\textsc{sbj}  bread-sweet-\textsc{nsbj} eat-\textsc{inf}  try.\textsc{lgr}-\textsc{rec.pst} \\
        \glt ‘Skye tried to eat the cake.’
    
    \ex\label{ex:koa:34b}
        \gll Skye-ka-k  tabak-čampoli-ⁿ ǐk-p-o-h  bànná-t \\ 
             Skye-\textsc{ln}-\textsc{sbj}  bread-sweet-\textsc{nsbj} \textsc{neg}-eat.\textsc{rgr-neg}-\textsc{inf}  try.\textsc{lgr}-\textsc{rec.pst} \\
        \glt ‘Skye tried not to eat the cake.’
    \z
\z 
In \REF{ex:koa:34b}, \textit{\v{i}k-po-h} ‘not to eat’ has no mark for the person of the subject.

\subsection{Negative lexicalizations}\label{sec:koa:2.5}

As described in \sectref{sec:koa:2.3}, the verb \textit{ǐkso-} ‘not exist, not have’ is a lexical negative (serving as the negative of \textit{n\v{a}:ho}- ‘exist, have’). The negative copula \textit{\v{o}n-ko}- ‘not be (a noun), not be (a property)’ is regular, but is generally shortened to \textit{ko-}. A few words describing properties have had their negative forms lexicalized as antonyms (\sectref{sec:koa:3.3}).


\section{Non-clausal negation}\label{sec:koa:3}
\subsection{Negative replies}\label{sec:koa:3.1}

Replies to yes/no questions \REF{ex:koa:35} can be made with \textit{ay} ‘yes’ or \textit{\v{i}nkoⁿ} ‘no’ \REF{ex:koa:36}. 

\ea\label{ex:koa:35}
    \gll ifa-k  wo:wòhká-ⁿ?\\ 
         dog-\textsc{sbj}  bark.\textsc{lgr-sf} \\
    \glt ‘Is the dog barking?’
\z 

\ea\label{ex:koa:36}
    \ea\label{ex:koa:36a}
        \gll ay,  wo:wòhk. \\ 
             yes  bark.\textsc{lgr} \\
        \glt  ‘Yes, it’s barking.’
    
    \ex\label{ex:koa:36b}
        \gll \v{i}nkoⁿ,    wo:woh-\v{i}kko-ⁿ.\\
             be.not.so  bark-\textsc{neg.rgr-sf} \\
        \glt ‘No, it’s not barking.’
    \z
\z 

A negative yes/no question can be asked with -\textit{t\v{i}kko-ⁿ} ‘it isn’t the case?’ \REF{ex:koa:37}. In that context, \textit{ay} ‘yes’ can be used to mean ‘that’s right’ \REF{ex:koa:38a}, although \textit{\v{i}nkoⁿ} ‘no’ can also be used \REF{ex:koa:38b}.

\ea\label{ex:koa:37} 
    \gll ifa-k  wo:wohka-t\v{i}kko-ⁿ?\\ 
         dog-\textsc{s}  bark-\textsc{neg.rgr-sf} \\
    \glt ‘Isn’t the dog barking?’
\z 
\ea\label{ex:koa:38}
    \ea\label{ex:koa:38a}
        \gll ay,  \v{o}nko-ⁿ.\\ 
             yes  do.\textsc{rgr-neg-sf} \\
        \glt ‘Yeah, he’s not.’  
    \ex\label{ex:koa:38b}
        \gll \v{i}nkoⁿ,    wo:woh-\v{i}kko-ⁿ.\\
             be.not.so  bark-\textsc{neg.rgr-sf} \\
        \glt ‘No, it’s not barking.’
    \z
\z 

% \ea\label{ex:koa:37}%HC that is the Q-A1-A2 version of the two previous exmaples, but the author didn't want me to do the change.
%     \begin{xlist}[A2: ]
%     \exi{Q: }\label{ex:koa:37}
%     \gll ifa-k  wo:wohka-t\v{i}kko-ⁿ \\ 
%          dog-\textsc{s}  bark-\textsc{neg.rgr-sf} \\
%     \glt ‘Isn’t the dog barking?’
%     \exi{A1: }\label{ex:koa:38a}
%         \gll ay,  \v{o}nko-ⁿ \\ 
%              yes  do.\textsc{rgr-neg-sf} \\
%         \glt ‘Yeah, he’s not.’  
%     \exi{A2: }\label{ex:koa:38b}
%         \gll \v{i}nkoⁿ,    wo:woh-\v{i}kko-ⁿ \\
%              be.not.so  bark-\textsc{neg.rgr-sf} \\
%         \glt ‘No, it’s not barking.’    
%     \end{xlist}
% \z 

\subsection{Negative indefinites and quantifiers}\label{sec:koa:3.2}
\citet[423ff]{Kimball1991} discusses indefinite pronouns in Koasati. As he notes, the same word is often used for \textit{some-} or \textit{any-} words in English (‘something’, ‘(not) anything’, etc.), as well as interrogatives (‘what’, etc.) (\ref{ex:koa:39}--\ref{ex:koa:41}).

\ea\label{ex:koa:39}
    \gll na:s-on  hì:čá-l. \\ 
         what/thing-\textsc{foc}.\textsc{nsbj}  see.\textsc{lgr}-1\textsc{sg.a} \\
    \glt ‘I see something.’
\z 

\ea\label{ex:koa:40}
    \gll na:s-on  ak-hǐ:č-o-ⁿ.\\ 
         what/thing-\textsc{foc}.\textsc{nsbj}  1\textsc{sg.a}.\textsc{neg}-see.\textsc{rgr}-\textsc{neg}-\textsc{sf} \\
    \glt ‘I don’t see anything.’
\z 

\ea\label{ex:koa:41}
    \gll na:s-on  is-hì:čá-ⁿ?\\ 
         what/thing-\textsc{foc}.\textsc{nsbj}  2\textsc{sg.a}-see.\textsc{lgr}-\textsc{sf} \\
    \glt ‘What do you see?’
\z 

The following indefinite pronouns are like \textit{na:si} in having ‘{some}-', ‘{any}-', and ‘{wh}-' uses: \textit{nakso-fa} ‘somewhere, (not) anywhere, where'; \textit{nakso-fo:ka} ‘sometime, (not) anytime, when'. The human indefinite pronoun \textit{a:ti} can be used for ‘someone' and for ‘(not) anyone', but never for ‘who'. To express ‘who', the word \textit{nakso} ‘which one' is used (and so is not specifically for humans).\footnote{My description of this section is informed by the work of \citet{Haspelmath1997}.}


Modifiers can be added to these pronouns to express other senses \REF{ex:koa:42}.

\TabPositions{.3\textwidth,.69\textwidth} 
\ea\label{ex:koa:42}
free-choice: \tab \textit{nak(so)-}\textbf{\textit{santik}} ‘which\textbf{ever}’ \\
universal: \tab  \textit{a:t-}\textbf{\textit{ohya}} ‘\textbf{every}one’, \textit{nakso-f-}\textbf{\textit{ohya}} ‘\textbf{every}where’ \\
else: \tab  \textit{a:t-}\textbf{\textit{mi:ta}} ‘someone \textbf{else}’ \\
\z 
\TabPositions{.33\textwidth,.66\textwidth}

\subsection{Negative derivation and case marking}\label{sec:koa:3.3} 
Only verbs may be negated in Koasati. To express ‘without’, the verb \textit{ǐkso}- ‘not exist’ is used \REF{ex:koa:43}.

\ea\label{ex:koa:43}
\textit{ilhič-ǐkso} \tab ‘invisible’ (lit. “appearance-without”) \\ 
\textit{a:ti im-ililla ǐkso} \tab ‘God’ (lit. “person their-death-without”) \\
\z 

Some instances of clausal negation in Koasati have been lexicalized as antonyms \REF{ex:koa:44}.

\ea\label{ex:koa:44}
\textit{kano-} ‘good’ \tab \textit{kǎn-\textbf{k-o}-ⁿ} ‘bad, not good’ \\
\textit{čoba-} ‘big’ \tab \textit{čo<\textbf{kǐ:}>b-\textbf{o}-ⁿ} ‘small, not big’ \\
\textit{čayha-} ‘tall’ \tab \textit{ča<\textbf{kǐ:}>h-\textbf{o}-ⁿ} ‘short, not tall’ \\
\z 

One reason for distinguishing these negatives from clausal negatives is that they may themselves be negated \REF{ex:koa:45}.\footnote{One speaker describes these double negatives as “kind of awkward, but okay in an argument”. The same speaker rejects a double negative in an ordinary (non-lexicalized) negative: *\textit{ho<}\textbf{\textit{kǐ}}\textit{>kf-}\textbf{\textit{o}}\textit{{}-}\textbf{\textit{tǐkk-o}}\textit{{}-ⁿ} ‘it’s not that he isn’t getting dressed’.}

\ea\label{ex:koa:45}
\textit{kǎn-\textbf{k-o-t\v{i}kk-o}-ⁿ} \tab ‘it’s not that he/she/it is bad’ \\
\textit{ča<\textbf{kǐ:}>h-\textbf{o-tǐkk-o}-ⁿ} \tab ‘it’s not that he/she/it is short’ \\
\textit{čo<\textbf{kǐ:}>b-\textbf{o-tǐkk-o}-ⁿ} \tab ‘it’s not that he/she/it is small’ \\
\z 

\section{Other aspects of negation}\label{sec:koa:4}
\subsection{The scope of negation}\label{sec:koa:4.1}

Verbs in Koasati are highly synthetic, with nine prefix slots, fifteen suffix slots, and different grades and internal changes to stems \citep{Kimball1991}. In some cases, negation can be marked in different places within the verb to indicate differences in scope. \citet[196--197]{Kimball1991}, for example, notes that the sentence in \REF{ex:koa:46} has the two corresponding negatives in \REF{ex:koa:47}.\footnote{I have changed the transcription and glossing to match the rest of the paper.}


\ea\label{ex:koa:46}
    \gll ča-tamm-ǎ:pi-t. \\ 
         1\textsc{sg.p}-fall-almost-\textsc{rec.pst}.\textsc{pfv} \\
    \glt ‘I almost fell.’
\z 
\ea\label{ex:koa:47}
    \ea\label{ex:koa:47a}
        \gll ča-tǎn-\textbf{\textit{k}}\textit{-a:pi-t}. \\
             1\textsc{sg.p}-fall.\textsc{rgr-neg}-almost-\textsc{rec.pst}.\textsc{pfv} \\
        \glt ‘I almost didn’t fall.’
    
    \ex\label{ex:koa:47b}
        \gll ča-tamm-a:pi-\textbf{\textit{tǐkko}}\textit{-t}. \\
             1\textsc{sg.p}-fall-almost-\textsc{neg.rgr}-\textsc{rec.pst}.\textsc{pfv} \\
        \glt ‘I didn’t almost fall.’
    \z
\z 

\citet[197]{Kimball1991} observes that if the verb root itself is negated, the regular negative of the root is used (\textit{tǎn-k-o}- ‘not fall’). If a modifier of the verb is negated, then a form of -\textit{tǐkko} is used. I record similar forms, where the negative of \textit{lokba}- ‘be hot’ is \textit{lo<\textbf{kǐ}>kb-\textbf{o}-ⁿ} ‘it is not hot’, but the negative of \textit{lokb-ǎ:ⁿhos} ‘it is very hot’ is \textit{lokb-ǎ:ⁿhos-\textbf{tǐkko}-ⁿ} ‘it is not very hot’. In general, the negative in -\textit{tǐkko}- appears to be used for newer patterns or expanded patterns (what we might call a periphrastic negative).

Negation can have scope over several different phrases within a clause. Examples (\ref{ex:koa:48}--\ref{ex:koa:50}) are contexts in which negation has scope over a subject, complement, and time adverb, respectively.

\ea\label{ex:koa:48}
    \gll Linda-k  aɬkǐ:y-o-t, Houston-ka-fa-p,  nihtaka-p. Eli-ka-k  ommi-t \\
         Linda-\textsc{sbj}  go.\textsc{sg}<\textsc{neg}>.\textsc{rgr}-\textsc{neg}-\textsc{med.pst} Houston-\textsc{ln}-\textsc{loc}-\textsc{top}  yesterday-\textsc{top} Eli-\textsc{ln}-\textsc{sbj}  do.\textsc{lgr}-\textsc{med.pst}\\
    \glt ‘\uline{Linda} didn’t go to Houston yesterday. \uline{Eli} did.’
\z 

\ea\label{ex:koa:49}
    \gll Linda-k  aɬkǐ:y-o-t, Houston-ka-fa-p,  nihtaka-p. Lake Charles-ka-f-on  aɬì:yá-t. \\ 
         Linda-\textsc{sbj}  go.\textsc{sg}<\textsc{neg}>.\textsc{rgr}-\textsc{neg}-\textsc{med.pst} Houston-\textsc{ln}-\textsc{loc}-\textsc{top}  yesterday-\textsc{top} Lake Charles-\textsc{ln}-\textsc{loc}-\textsc{foc}.\textsc{nsbj}  go.\textsc{sg}.\textsc{lgr}-\textsc{med.pst}\\
    \glt  ‘Linda didn’t go to \uline{Houston} yesterday. She went to \uline{Lake Charles}.’
\z 

\ea\label{ex:koa:50}
    \gll Linda-k  aɬkǐ:y-o-t, Houston-ka-fa-p,  nihtaka-p. Tahollo-fo:k-on  aɬì:yá-t. \\ 
         Linda-\textsc{sbj}  go.\textsc{sg}<\textsc{neg}>.\textsc{rgr}-\textsc{neg}-\textsc{med.pst} Houston-\textsc{ln}-\textsc{loc}-\textsc{top}  yesterday-\textsc{top} Sunday-when-\textsc{foc}.\textsc{ds}  go.\textsc{sg}.\textsc{lgr}-\textsc{med.pst}\\
    \glt ‘Linda didn’t go to Houston \uline{yesterday}. She went on \uline{Sunday}.’
\z 

The scope of negation seems to be limited to the clause containing the negation, however. There is no evidence of negative transport, for example (\sectref{sec:koa:4.6.1}). Similarly, in several types of complex clauses, negation is again limited to the negative clause.

Instead of using coordination to join clauses, Koasati makes extensive use of clause chaining.\footnote{See \citet{Martin2024} for a discussion of clause chaining and case in Muskogean languages.}
The same-subject marker -\textit{k} (focused form -\textit{ok}) or different-subject marker \textit{-n} (or -\textit{ⁿ}, focused form \textit{-on}) are used at the ends of dependent non-final chained clauses. The final verb form is an independent form and is marked for tense and mood \REF{ex:koa:51}.

\ea\label{ex:koa:51}
    Chained clauses:
    
    [\textsubscript{S1} [\textsubscript{S2} …VERB-\textit{k/n}] [\textsubscript{S3} …VERB-\textit{k/n}] VERB-TENSE-MOOD]
\z 

The same switch-reference markers appear at the ends of complement clauses \REF{ex:koa:52} and adverbial clauses \REF{ex:koa:53}.

\ea\label{ex:koa:52}
    Complement clauses:
    
    [\textsubscript{S1} [NP] [\textsubscript{VP} [\textsubscript{S2} …VERB-\textit{k/n}] VERB-TENSE-MOOD]
\z 

\ea\label{ex:koa:53}
    Adverbial clauses:
    
    [\textsubscript{S1} [\textsubscript{S2} …VERB-\textit{k/n}] … VERB-TENSE-MOOD] 
\z 

In each of these structures, negation is restricted to the clause containing the negative element. We see this in chained clauses in \REF{ex:koa:54}.

\ea\label{ex:koa:54}
    \ea\label{ex:koa:54a}
        \gll Linda-k  čokòhl-ók  ì:pá-t. \\ 
             Linda-\textsc{sbj}  sit.\textsc{sg}.\textsc{hgr-foc}.\textsc{ss}  eat.\textsc{lgr}-\textsc{rec.pst.pfv} \\
        \glt ‘Linda sat down and ate.’
    
    \ex\label{ex:koa:54b}
        \gll Linda-k  čokòhl-ók ǐk-p-o-:s.\\ 
             Linda-\textsc{sbj}  sit.\textsc{sg}.\textsc{hgr-foc}.\textsc{ss} \textsc{neg}-eat.\textsc{rgr}-\textsc{neg}-\textsc{rec.pst}.\textsc{ipfv}\\
        \glt ‘Linda sat down and didn’t eat.’ (not: ‘Linda didn’t sit down     and eat.’)
    \z
\z 

The same is true for chained clauses in imperatives \REF{ex:koa:55}:

\ea\label{ex:koa:55}
    \ea\label{ex:koa:55a}
        \gll čokòhl-ók  ohomp\^{}! \\ 
             sit.\textsc{sg}.\textsc{hgr}-\textsc{foc}.\textsc{ss}  eat.\textsc{imp} \\
        \glt ‘Sit down and eat!’
    
    \ex\label{ex:koa:55b}
        \gll čokòhl-ók  ohompa-n\^{} \\
             sit.\textsc{sg}.\textsc{hgr}-\textsc{foc}.\textsc{ss}  eat-\textsc{proh} \\
        \glt ‘Sit down and don’t eat!’ (not: ‘Don’t sit down and eat!’)
    \z
\z 

Interestingly, Koasati has another clause type that is used in expressing manner adverbs. These coverbs or coverb phrases are reduced clauses ending in a suffix \textit{-t} that modify a verb \REF{ex:koa:56}.\footnote{\citet[227]{Kimball1991} refers to \textit{{}-t} as a coordinating conjunctive suffix. See \citet{Broadwell2006} for a similar suffix in Choctaw.}


\ea\label{ex:koa:56}
    \textit{-t} phrases:
    
    [\textsubscript{S1} [NP] [\textsubscript{VP} [\textsubscript{S2} …VERB-\textit{t}] VERB-TENSE-MOOD]
\z 

These \textit{-t} phrases are within the scope of a negative verb. We can see this in the indicative examples in \REF{ex:koa:57}.


\ea\label{ex:koa:57}
    \ea\label{ex:koa:57a}
        \gll ɬoyka-t  ì:lá-t. \\ 
             return-\textsc{cvb}  get.here.\textsc{sg}.\textsc{lgr}-\textsc{rec.pst.pfv} \\
        \glt ‘He/she came back.’ (lit. ‘He/she got here returning.’)
    
    \ex\label{ex:koa:57b}
        \gll ɬoyka-t  ǐk-l-o-:s. \\
             return-\textsc{cvb}  \textsc{neg}-get.here.\textsc{sg}.\textsc{rgr}-\textsc{neg}-\textsc{rec.pst}.\textsc{ipfv} \\
        \glt ‘He/she didn’t come back.’ (lit. ‘He/she didn’t get here returning.’)
    \z
\z 
 
Coverbs are also in the scope of negative commands, which use the prohibitive construction (\ref{ex:koa:58}--\ref{ex:koa:59}):



\ea\label{ex:koa:58}
    \gll haččǎ:li-t  is-pa-n\^{}! \\ 
         stand.\textsc{ggr}-\textsc{cvb}  2\textsc{sg.a}-eat-\textsc{proh} \\
    \glt ‘Don’t eat while standing!’ (lit. ‘Don’t eat standing!’)
\z 

\ea\label{ex:koa:59}
    \gll čoko:-či-n\^{},  mat-ča-hì:čá-t! \\ 
         sit.\textsc{sg}-2\textsc{sg.a}-\textsc{proh}  \textsc{dir}-1\textsc{sg.p}-look.\textsc{lgr}-\textsc{cvb} \\
    \glt ‘Don’t sit there looking at me!’
\z 

With the exception of coverbs, then, the scope of negation in Koasati is limited to the clause containing the negative element.

\subsection{Negative polarity}\label{sec:koa:4.2}

I have so far not found any elements that only occur in negative environments.

\subsection{Marking of NPs in the scope of negation}\label{sec:koa:4.3}

In Koasati, the subject and non-subject case markers are often replaced by a “topic” marker \textit{-p} (or \textit{-ap} on pronouns) when a sentence is negative \REF{ex:koa:60}.\footnote{\citet[411]{Kimball1991} describes \textit{{}-p} as a marker of “new topic” indicating “that the topic into which a discussion is about to enter, or which is just finished, is new, or contains new information.” In my experience, \textit{{}-p} often clarifies or resumes an established topic. For now, I will simply refer to it as “topic”.}


\ea\label{ex:koa:60}
    \ea\label{ex:koa:60a}
        \gll akk-ok  if-ok  om. \\ 
              that-\textsc{foc}.\textsc{sbj}  dog-\textsc{foc}.\textsc{sbj}  be \\
        \glt ‘That is a dog.’
    
    \ex\label{ex:koa:60b}
        \gll akk-ap  if-ok  ko-ⁿ.\\
             that-\textsc{top}  dog-\textsc{foc}.\textsc{sbj}  be.\textsc{neg}-\textsc{sf} \\
        \glt ‘That is not a dog.’
    \z
\z 

This use is reminiscent of Japanese \textit{wa}, which is also called a topic marker and which may indicate the target of negation (see, for example, \citealt[149ff]{Hinds1986}; \citealt[45--46]{Nyberg2012}). When I have specifically tested this idea, however, I have found a poor correlation between \textit{-p} and target of negation. This can be seen in (\ref{ex:koa:48}--\ref{ex:koa:50}) above, where \textit{-p} is found several times in negative sentences, but doesn’t correspond with the target of negation. What is most striking about the examples in (\ref{ex:koa:48}--\ref{ex:koa:50}) is that \textit{-p} appears on phrases that have been added after the verb (normally a position for afterthoughts).

My feeling is that \textit{-p} serves to set a phrase apart from the rest of the sentence, whether to establish a frame of reference (‘as for X’, ‘if it is X’) or to fill in incomplete information (‘he’s going to town, John is’). In \REF{ex:koa:61}, for example, we see \textit{-p} used in a constructed phone conversation between two people discussing plans for Thanksgiving.

\ea\label{ex:koa:61}
    \gll kosn-\textbf{ap} ama:-hilka-laho-k om, Texas-ka-f-on, ča-fono:si-fa-ⁿ.\\ 
         we-\textsc{top}  go.\textsc{tpl}-1\textsc{pl.a}-\textsc{fut}-\textsc{ss} be Texas-\textsc{ln}-\textsc{loc}-\textsc{foc}.\textsc{nsbj}  1\textsc{sg.p}-sister-\textsc{loc}-\textsc{nsbj}\\
    \glt ‘Us, we’re going to Texas, to my sister’s.’
\z 
 
This usage seems common in exchanges of this type. One person makes a statement (e.g., ‘my name is Mary’) and then a second person might use \textit{-p} to distinguish himself or herself from the previous speaker (‘me, my name is Betty’).

Why, then, would case markers commonly shift to \textit{-p} in negative sentences? Negative sentences do commonly have a phrase in focus. That means that the other phrases will be out of focus (backgrounded). It may be that \textit{-p} is used in negative sentences to mark the backgrounded phrases, which are often added as afterthoughts, as in \REF{ex:koa:62}.

\ea\label{ex:koa:62}
    \gll ak-hǐ:č-o-:s,  ya-:fa-p. \\ 
         1\textsc{sg.a}.\textsc{neg}-see.\textsc{rgr}-\textsc{neg}-\textsc{rec.pst}.\textsc{ipfv}  here-\textsc{loc}-\textsc{top} \\
    \glt ‘I didn’t see him over here.’
\z 
This would explain the frequency of \textit{-p} in negative sentences, the fact that it may appear on multiple phrases, and the fact that these phrases are often (though not always) postverbal.

\subsection{Reinforcing negation}\label{sec:koa:4.4}

I haven’t yet found any examples where a distinct pattern is used to reinforce or emphasize negation in Koasati.

\subsection{Negation, coordination and complex clauses}\label{sec:koa:4.5}
As noted above, Koasati lacks true coordination, though extensive clause chaining and subordination are used. The scope of negation over various types of subordinate clauses is described in \sectref{sec:koa:4.1}.

\subsection{Miscellaneous aspects of negation}\label{sec:koa:4.6} 
\subsubsection{Negative transport}\label{sec:koa:4.6.1}
As \citet{Fillmore1963} observed, when a verb like \textit{think} is negated in English, the implication is that it is the complement that is negated (thus \textit{I didn’t think that Linda would come} implies \textit{I thought that Linda wouldn’t come}).

Koasati does not have negative transport. In Koasati, the mark of negation goes on the verb of the semantically-negated clause (\ref{ex:koa:63}--\ref{ex:koa:64}).


\ea\label{ex:koa:63}
    \gll Linda-k  o<kǐ>nt-o-laho-ⁿ ǎ:lo-l.\\ 
         Linda-\textsc{sbj}  come<\textsc{neg}>.\textsc{rgr}-\textsc{neg}-\textsc{fut}-\textsc{ds} think.\textsc{ggr}-1\textsc{sg.a} \\
    \glt ‘I don’t think Linda will come.’ (lit. ‘I think Linda will not   come.’)
\z 

\ea\label{ex:koa:64}
    \gll o<kǐ>:b-o-laho-ⁿ  ǎ:lo-l. \\ 
         rain<\textsc{neg}>.\textsc{rgr}-\textsc{neg}-\textsc{fut}-\textsc{ds}  think.\textsc{ggr}-1\textsc{sg.a} \\
    \glt ‘I don’t think it will rain.’ (lit. ‘I think it will not rain.’) 
\z 

\subsubsection{Metalinguistic negation}\label{sec:koa:4.6.2}

Metalinguistic negation is “a device for objecting to a previous utterance on any grounds whatever” \citep[363]{Horn1989}. In Koasati, the verb \textit{tǐkko}- ‘be not the case that’ can be used to focus negation on a particular element (\ref{ex:koa:65}--\ref{ex:koa:66}).

\ea\label{ex:koa:65}
    \gll John-ka-k  in-n\v{a}:ho-ⁿ,  točč\v{i}:na-t  im-atlawista-k. \\ 
         John-\textsc{ln-sbj}  \textsc{dat}-have.\textsc{rgr-sf}  three.\textsc{rgr-ss}  \textsc{dat}-children-\textsc{sbj} \\
    \glt ‘John has three children.’
\z 

\ea\label{ex:koa:66}
    \gll \v{i}nko-ⁿ  točč\v{i}:na-t  t\v{i}kko-hči. ost\v{a}:ka-t  in-n\v{a}:ho-ⁿ.\\ 
         be.not.so.\textsc{rgr-sf}  three.\textsc{rgr-ss}  be.not-\textsc{ptcp} four.\textsc{rgr-ss}  \textsc{dat}-exist.\textsc{rgr-sf} \\
    \glt ‘No, he doesn’t have three. He has four.’
\z 

As shown in (\ref{ex:koa:65}--\ref{ex:koa:66}), the verb \textit{tǐkko}- ‘be not the case that' appears to be used for metalinguistic negation. We have seen the same element used in negative questions (\ref{ex:koa:37}), negation of lexical negatives (\ref{ex:koa:45}), and alternative negative patterns like (\ref{ex:koa:47b}). The common thread seems to be that \textit{tǐkko}- has broader scope than the morphological negative.

\subsubsection{Diachronic notes and observations}\label{sec:koa:4.6.3}

From a cross-linguistic perspective, two aspects of Koasati negation seem especially unusual: a. the use of three separate marks of negation (a form of \textit{ik-}, -\textit{o}, and use of the rising tone grade); and, b. the placement of the primary negative marker \textit{ik-} as a prefix, infix, or suffix. 

The prefixing pattern appears to have been limited to a set of ‘short’ verbs of the shape \#(C)V(:)CV\#. These were negated directly in an earlier form of the language, without an auxiliary, as in \REF{ex:koa:67a}.\footnote{See \citet{Booker1980} and \citet{MartinMunro2005} for more discussion of negation in the Muskogean languages. The development of infixation in Muskogean is discussed in \citet{Yu2007}.} This order of affixes is maintained in the modern language without change, as in \REF{ex:koa:67b}. 

\ea\label{ex:koa:67} %67a and 67b are exactly the same, this IS correct.
    \ea Earlier {\label{ex:koa:67a}\\
    \gll ak-hǐ:č-o-ⁿ.\\ 
         1\textsc{sg.a}.\textsc{neg}-see.\textsc{rgr}-\textsc{neg}-\textsc{sf} \\
    \glt ‘I do not see.’} 
    \ex Modern {\label{ex:koa:67b}\\
    \gll ak-hǐ:č-o-ⁿ.\\
    1\textsc{sg.a}.\textsc{neg}-see.\textsc{rgr}-\textsc{neg}-\textsc{sf} \\
    \glt ‘I do not see.’}
    \z 
\z 

\noindent Short verbs that were inflected directly included \textit{hiča}- ‘see’, \textit{ha:lo}- ‘hear’, \textit{ipa}- ‘eat’, \textit{ila}- ‘get here’, \textit{kaha}- ‘say’, and \textit{noči}- ‘sleep’ (Class 1 verbs).

A larger set of verbs (Class 3) used periphrastic negation. In this pattern, an auxiliary *\textit{k}- was added after the verb, and the auxiliary was inflected for negation \REF{ex:koa:68}.

\ea\label{ex:koa:68}
    \ea Earlier{\label{ex:koa:68a}\\
        \gll hopon-t  ǎk-k-o-ⁿ.\\ 
             cook-\textsc{ss}  1\textsc{sg.a}.\textsc{neg}-\textsc{neg}.\textsc{aux.rgr}-\textsc{neg}-\textsc{sf} \\
        \glt ‘I am not cooking.’} 
    \ex Modern {\label{ex:koa:68b}\\
        \gll hopon-tǎkko-ⁿ.\\
             cook-1\textsc{sg.a}.\textsc{neg}-\textsc{sf} \\
        \glt ‘I am not cooking.’}
    \z
\z 
The modern language preserves this order, but now treats same subject -\textit{t} as part of the negative marker.

The infixing verbs (Class 2) are more difficult to understand. Infixing verbs generally have the shape …C\textsubscript{1}C\textsubscript{2}V\#, where two different consonants appear before the final vowel. I would suggest that infixing arose when speakers began inflecting the final C\textsubscript{1}C\textsubscript{2}V\# as though it were an auxiliary \REF{ex:koa:69}.


\ea\label{ex:koa:69}
    \ea Earlier{\label{ex:koa:69a}\\
        \gll talwa-t ǎk-k-o-ⁿ. \\ 
             sing-\textsc{ss} 1\textsc{sg.a}.\textsc{neg}-\textsc{neg}.\textsc{aux}-\textsc{neg}-\textsc{sf} \\
        \glt ‘I don’t sing.’ } 
    
    \ex Modern {\label{ex:koa:69b}\\
        \gll ta<kǎ>lw-o-ⁿ.\\
             sing<1\textsc{sg.a}.\textsc{neg}>.\textsc{rgr}-\textsc{neg}-\textsc{sf} \\
        \glt ‘I don’t sing.’}
    \z
\z 
Verbs where the final C\textsubscript{1}C\textsubscript{2}V was inflected include \textit{hokfi}- ‘put in’, \textit{hofna}- ‘smell’, \textit{hokfa}- ‘get inside’, \textit{iltohno}- ‘work’, \textit{ohompa}- ‘eat (several things)’, \textit{onti}- ‘come’, and \textit{talwa}- ‘sing’. 


\section{Summary}\label{sec:koa:5}
The various strategies for expressing negation are matched to their functions in \tabref{tab:koa:4}.
Koasati negation is marked on verbs with an affix \textit{ik}- and a suffix -\textit{o}. A negative verb must be in a specific aspect, called the rising tone grade. Verb stems fall into different classes based on the placement of the \textit{ik}- affix, which has different forms depending on verb class and fuses with agent person markers.
Except for one specific coverb construction, the scope of negation is limited to the immediate clause containing the negated verb.

\begin{table}
    \begin{tabularx}{\textwidth}{QQl}
\lsptoprule
Negation strategy & Function & Section\\
\midrule
\textit{ik}- + \textit{-o} + rising tone grade on verb 
& Standard negation & \ref{sec:koa:2.1}, \\
& Negation in non-declaratives & \ref{sec:koa:2.2}, \\
& Negative locative & \ref{sec:koa:2.3},\\
& Negative indefinites and quantifiers & \ref{sec:koa:3.2} \\
\tablevspace
\textit{-nna}, \textit{-n} 
& Negative command & \ref{sec:koa:2.2},   \ref{sec:koa:4.1} \\
\tablevspace
\textit{ko-ⁿ} (lexical negative of \textit{ommi-} ‘be’) 
& Negating nouns in equative clauses & \ref{sec:koa:2.3},\\
& Negative attributive clauses & \ref{sec:koa:2.3}  \\
\tablevspace
\textit{ǐkso-ⁿ} (lexical negative of \textit{nǎ:ho-} ‘exist’) 
& Negative existential, negative ‘have’ & \ref{sec:koa:2.3}  \\
\tablevspace
\textit{ǐnko-ⁿ} ‘no’ 
& Negative reply & \ref{sec:koa:3.1}  \\
\tablevspace
\textit{-tǐkko-ⁿ} ‘be not the case that’ 
& Negative yes/no question & \ref{sec:koa:3.1}, \\
& Negative of lexicalized negative & \ref{sec:koa:3.3}, \\
& Periphrastic negative & \ref{sec:koa:4.1}, \\
& Metalinguistic negation & \ref{sec:koa:4.6.2}  \\
\lspbottomrule
    \end{tabularx}
    \caption{Koasati negation strategies and their functions}
    \label{tab:koa:4}
\end{table}


Koasati has rich verbal morphology. Negation can appear in different places within a verb to indicate that different elements are negated.

Stative predicates are generally negated in the same way as other predicates. The verb \textit{\v{i}kso}- ‘not exist, not have’ is the only lexical negative form: other negatives are formed morphologically, though the verb \textit{ommi}- ‘be’ is developing an irregular negative \textit{ko}-.

Koasati negation, while complex, may have developed from a system like that in Finnish in which a negative auxiliary is inflected for person marking. In the case of Koasati, the system is more complicated in that a few short verbs can be inflected like auxiliaries, and some verbs allow a final prosodic unit (…C\textsubscript{1}C\textsubscript{2}V) to be inflected.


\section*{Acknowledgements}

I am deeply indebted to the Coushatta Language Committee for sharing their knowledge with me. I would especially like to thank Bertney Langley, Lorenda Poncho, and Kathy Poncho for providing most of the Koasati forms in this paper. Linda Langley and Bertney Langley helped make arrangements for me to research this topic. I am also grateful to Ljuba Veselinova, Matti Miestamo, and two reviewers for comments.


\section*{Abbreviations}


\begin{tabularx}{.45\textwidth}{lQ}
\scshape 1 & first person \\ 
\scshape 2 & second person \\ 
\scshape a & agent person marking \\
\scshape aux & auxiliary \\
\scshape cvb & coverb  \\
\scshape dat & dative  \\
\scshape dim & diminutive \\ 
\scshape dir & directional prefix  \\
\scshape ds & different subject \\ 
\scshape foc & focus  \\
\scshape fut & future  \\
\scshape ggr & geminating rising tone grade  \\
\scshape hgr & aspirating grade \\ 
\scshape imp & imperative  \\
\scshape inf & infinitive  \\
\scshape ipfv & imperfective  \\
\scshape lgr & lengthened grade  \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape ln & loan word marker  \\
\scshape loc & locative  \\
\scshape med &  medial \\
\scshape neg & negative  \\
\scshape nsbj & nonsubject  \\
\scshape p & patient person marking  \\
\scshape pfv & perfective  \\
\scshape pl & plural \\  
\scshape ptcp & participle  \\
\scshape proh & prohibitive  \\
\scshape pst & past  \\
\scshape rec & recent  \\
\scshape rgr & rising tone grade  \\
\scshape sbj & subject  \\
\scshape sf & sentence final marker  \\
\scshape sg & singular  \\
\scshape ss & same subject  \\
\scshape top & topic  \\
\scshape tpl & triplural (three or more) \\
\end{tabularx}


\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
\end{document} 
