\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18234097}

\author{Dominique Knuchel}
\title{Negation in Kogi}
\abstract{This paper discusses negation in Kogi, a Chibchan language spoken in Colombia. For standard negation (i.e. negation of main declarative clauses), three major markers can be identified whose use is dependent on the aspectual interpretation of a clause. These markers can further be used in stative predications, as well as most dependent and non-declarative clauses. In addition, there are dedicated negation markers in clauses marked for modality, optative and imperative mood, and two types of dependent clauses (i.e. circumstantial and purpose clauses). A further aspect of negation in Kogi concerns the expression of phasal polarity. Moreover, a negative existential verb can be used in certain stative predications and abilitive modality. Finally, in non-clausal negation, Kogi makes use of negative indefinite pronouns which consist of an interrogative pronoun in combination with an additive focus particle.

\keywords{Negation markers, Kogi language, phasal polarity, standard negation.}
}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 
%\shorttitlerunninghead{}%%use this for an abridged title in the page headers
\section{The language}\label{sec:kog:1}

This paper details the expression of clausal as well as non-clausal negation in Kogi (ISO: \href{https://iso639-3.sil.org/code/kog}{kog}, Glottocode: \href{https://glottolog.org/resource/languoid/id/cogu1240}{cogu1240}), an indigenous language of Colombia. Kogi belongs to the Chibchan language family, where it forms the Arwako subgroup together with its neighboring languages Ika (ISO: \href{https://iso639-3.sil.org/code/arh}{arh}, Glottocode: \href{https://glottolog.org/resource/languoid/id/arhu1242}{arhu1242}) and Damana (ISO: \href{https://iso639-3.sil.org/code/mbp}{mbp}, Glottocode: \href{https://glottolog.org/resource/languoid/id/mala1522}{mala1522}). The Kogi live in the region of the Sierra Nevada de Santa Marta, a mountain range in the north of Colombia. According to the most recent national census of Colombia which was conducted in 2018 \citep{DANE2018}, the group comprises 15,820 individuals. The language name \textit{Kogi} is commonly used in the anglophone linguistic and anthropological fields, while the language is known as \textit{Cogui} in the Spanish-speaking area. The speakers themselves refer to their mother tongue as \textit{Koguian}. The Kogi word \textit{kaggaba} denotes the indigenous group of the Kogi, but also means ‘indigenous person’ in general.

The present study of Kogi negation is based on first-hand language data that was collected during three fieldwork periods in Colombia between September 2016 and December 2018. The questionnaire provided by the editors \parencitetv{chapters/questionnaire} served as a guideline for the investigation of different aspects of negation as well as for the structure of the presentation.

Section \sectref{sec:kog:2} discusses the topic of clausal negation and takes into account declarative main clauses (i.e. standard negation), non-declarative clauses, stative predications as well as dependent clauses. \sectref{sec:kog:3} discusses negation in the non-clausal domain, more specifically negative replies and negative indefinite pronouns. \sectref{sec:kog:4} briefly discusses some other aspects of negation and a summary is provided in \sectref{sec:kog:5}.

In addition to a general description, negative constructions are analyzed in terms of constructional and paradigmatic (a)symmetry, as proposed by \citet{Miestamo2005}. The first type of asymmetry refers to possible structural differences that negation marking causes in a clause. A construction is symmetrical when the negated clause does not differ from the affirmative in any other respect than the presence of a negation marker. As for the second type, a paradigm is symmetrical when the same grammatical categories are available in the affirmative and the negative (i.e. there is no neutralization of distinctions in the negative).

The remainder of this introduction presents some basic points and mentionable traits of Kogi grammar.

The morpho-phonological processes attested in Kogi are numerous and relatively complex, including, for instance, vowel raising, devoicing of intervocalic plosives and fricatives, or the denasalization of morpheme-initial nasal consonants. Some of these morpho-phonological rules are restricted to occur at certain morphological boundaries (e.g. prefix-stem, or stem-anti-clitic).

The basic constituent order is SOV and the language exhibits split ergativity marking on noun phrases conditioned by the referential nature of the noun phrase: personal pronouns follow the nominative-accusative pattern, while non-pronominal noun phrases show ergative-absolutive case marking. 

Nominal morphology is fairly simple, involving only few inflectional and derivational processes. A distinction can be made between alienable and inalienable nouns, inalienable possession being marked by nominal prefixes. Case, number and discourse functions are marked by phrasal suffixes at the end of noun phrases.

Verbal morphology, by contrast, shows more complexity. Verbs can have up to three different stem allomorphs which are morphologically conditioned, i.e. their use depends on whether and what kind of inflectional suffixes are attached. Formally, the stem allomorphs differ in the presence of a stem formative \textit{\nobreakdash-a} ({\textasciitilde} \textit{{}-ya {\textasciitilde} {}-wa}; glossed as \textsc{st)}, or the deletion of final \textit{k}. For example, the verb ‘ascend’ can take the form \textit{nitshik} (e.g. with habitual aspect), \textit{n\'{i}tshika} (e.g. with future tense) or \textit{nitshi} (as an independent verb form).

There are three distinct sets of markers that index core arguments of a clause on the verb. Each of the paradigms is typically associated with a specific argument role, i.e. Set I is preferable with S/A arguments, Set II with P arguments, and Set III with G arguments. 

Certain TAM (tense, aspect, mood/modality) categories are expressed exclusively by synthetic verb forms (e.g. potential modality or habitual aspect), while others are denoted by periphrastic constructions involving different auxiliaries (e.g. obligative modality). A number of grammatical categories have both options available which – in semantic terms – can be used interchangeably (e.g. future tense).

A distinctive trait of Kogi grammar is the expression of engagement, i.e. the encoding of (a)symmetries in the distribution and access to knowledge \citep{EvansEtAl2018}. Engagement is encoded in four verbal prefixes that reflect two different parameters: epistemic authority (i.e. whose knowledge is targeted) and (non)-shared access (i.e. whether it is exclusive to one of the speech act participants or shared by both) \citep{Bergqvist2016}. These markers are not obligatory but are used by a speaker when they wish to express their epistemic stance related to a proposition. The pragmatic functions of engagement markers are complex and their exact interpretation depends on the discourse context. The use of such a marker, however, can influence the choice of other grammatical categories, such as tense. That is, there is a separate tense paradigm involving a three-way distinction in remoteness. 

Besides prefixes, suffixes and clitics, Kogi exhibits a further bound morpheme type which can be labeled anti-clitics. An anti-clitic constitutes the counterpart of the more familiar category of clitics, which are commonly defined as units that constitute independent words on the grammatical level, but are phonologically dependent on another word.\footnote{For the concept of wordhood and the distinction between words on the grammatical and phonological level (g-words and p-words) see for example \citet{DixonAikhenvald2002}.} Anti-clitics, by contrast, constitute independent words on the phonological level, while they cannot be considered independent words on the grammatical level. An anti-clitic is thus similar to an affix in that it forms part of a bigger grammatical unit. In contrast to affixes, however, an anti-clitic shows less (or no) phonological cohesion. In Kogi, this phonological independence is reflected in the fact that not all morphophonological processes that occur at affix boundaries also occur at anti-clitic boundaries. As an example, let us look at the morphophonological rule \textit{n → l / a\_\_} that applies at the affix-stem boundary. This is illustrated in \REF{ex:kog:1} with the inceptive suffix \textit{{}-na}, where the initial \textit{n} of the suffix is denasalized after \textit{a}. 

\ea\label{ex:kog:1} {galá}.\\    
    \gll ga-ná\\
        eat-\textsc{incep}\\
    \glt‘S/he began to eat.'
\z 

Example \REF{ex:kog:2} shows the past marker \textit{\#ne}, an anti-clitic. Note that \# indicates the border between a stem and an anti-clitic. It is evident that the denasalization does not take place at the stem-anti-clitic boundary. This suggests that the verb form in \REF{ex:kog:2} consists of two phonological units, whereas it constitutes a single unit on the grammatical level.

\ea\label{ex:kog:2}{gane}.\\
    \gll ga\#ne\\
         eat\#\textsc{pst}\\
    \glt  ‘S/he ate.’
\z 

A further rule concerns the denasalization of \textit{m}: \textit{m → b / a\_\_.} This rule, too, is observed at stem affix boundaries, but not at boundaries with anti-clitics. This is illustrated in \REF{ex:kog:3}. Note that the anti-clitic here is complex, i.e it consists of \textit{\#ne} ‘\textsc{pst’} with the second person singular prefix \textit{ma-}. It is apparent that the initial \textit{m} of the complex anti-clitic (\#\textit{male)} does not denasalize.\footnote{Note also that, while denasalization of \textit{n} does not take place at stem-anti-clitic boundaries, it does so at the affix-anti-clitic boundary within a complex anti-clitic (\textit{\#ma-ne → male}).}

\ea\label{ex:kog:3}{gamale}.\\
    \gll {ga\#ma-ne}\\
         eat\#2\textsc{sg.I-pst}\\
    \glt  ‘You ate.’
\z 

The distinction between affixes and anti-clitics is crucial for differentiating between the two markers \textit{\#ne} ‘\textsc{pst’} and \textit{{}-ne} ‘\textsc{neg.hab}’, the negative habitual suffix which is detailed in 2.1.3 below. While both markers seem identical at first sight, they differ in their morphophonological behavior. Consider the verb form in \REF{ex:kog:4}, which differs from the one in \REF{ex:kog:3} in only one respect: the \textit{m} of the person marker is denasalized to \textit{b}. Thus, the two markers clearly are distinct, one being an anti-clitic denoting past tense, the other one being a suffix expressing negation.

 \ea\label{ex:kog:4}gabale.\\
    \gll ga-ma-ne\\
         eat-2\textsc{sg.I-neg.hab}\\
    \glt  ‘You never eat.’
\z 

For the purposes of the present discussion, this short illustration, in particular of the distinction between \textit{\#ne} ‘\textsc{pst’} and \textit{{}-ne} ‘\textsc{neg.hab}’, should suffice. The issue of wordhood in Kogi and especially the analysis of anti-clitics is a complex one and the interested reader will find more details in \citet{Knuchel2022}.

\section{Clausal negation}\label{sec:kog:2}

\subsection{Standard negation}\label{sec:kog:2.1}

Standard negation, i.e. the basic way of negating declarative main clauses, involves a number of different bound morphemes that reflect the aspectual value of a clause. A set of three negation markers, subsumed under the term ‘major negation markers’ here, is found in all declarative main clauses, namely general negative \textit{{}-zhá} (\sectref{sec:kog:2.1.1}), \textit{{}-kí} ‘\textsc{neg.ipfv}’
%     \mm{@LSP: how was it again with glosses in text, parentheses rather than quotes?}
(\sectref{sec:kog:2.1.2}) and \textit{{}-ne} ‘\textsc{neg.hab}’ (\sectref{sec:kog:2.1.3}). In addition, a dedicated negation marker \textit{{}-sïng(ze)} can be used with specific modality values and its source, i.e. the negative existential verb \textit{sïng(ze),} features in the negation of abilitive modality (\sectref{sec:kog:2.1.4}). Two further markers, namely {}-\textit{gánga} ‘not yet’ and -\textit{ksá} ‘no longer’, express values of phasal polarity (\sectref{sec:kog:2.1.5}).

\subsubsection{Negation in perfective clauses: -\textit{zhá}}\label{sec:kog:2.1.1}

The negation marker \textit{{}-zhá} occurs in perfective past and future clauses. Note that perfective verbs are not overtly marked for aspect, and contrast with overtly marked imperfective predicates. The marker \textit{{}-zhá} does not express perfective aspect, but occurs on verbs with a perfective meaning. 

As a general rule, \textit{{}-zhá} only attaches to lexical verbs, and never to auxiliaries. A first example featuring the past marker \#\textit{ne} is given in \REF{ex:kog:5}.\footnote{Note that the past anti-clitic \textit{ne} is distinct from the habitual past suffix \textit{{}-ne} for reasons that are apparent in their morphophonological behavior (see \sectref{sec:kog:2.1.3}).}

\ea\label{ex:kog:5}
    \ea\label{ex:kog:5a}Santa Martali nẽnkale.\\
        \gll  {Santa Marta-ni nẽ\#nka-ne}\\ 
               Santa Marta-\textsc{all} go\textsc{\#1pl.I{}-pst}\\
        \glt ‘We went to Santa Marta.’
    \ex\label{ex:kog:5b} Santa Martali nẽyazhánkale.\\
        \gll  Santa Marta-ni nẽ{}-a-\textbf{zhá\#nka-ne}\\
              Santa Marta-\textsc{all} go{}-\textsc{st{}-neg\#1pl.I{}-pst}\\
        \glt  ‘We did not go to Santa Marta.’
    \z
\z 

Along similar lines, the negation marker occurs on the lexical verb in periphrastic past predicates marked for engagement, such as the one in \REF{ex:kog:6}.\footnote{As noted in the introduction, the functions of engagement markers are complex, and their exact interpretation depends on the pragmatic context. In grammatical elicitation focusing on tense and negation, the speaker symmetric form \textit{ni-} was chosen as the least marked variant. As a pragmatic context that would allow for an interpretation of the engagement marker was absent, I do not include it in the free translation.}
 
\ea\label{ex:kog:6}
    \ea\label{ex:kog:6a}  mebák nak nigungú.\\
        \gll  mebák nak ni-gu-ngú \\
        yesterday arrive \textsc{spkr.sym-}do{}-\textsc{rec.pst}\\
        \glt ‘S/he arrived yesterday.’
    \ex\label{ex:kog:6b} mebak nagazhá nigungú.\\
        \gll  mebák nak-a-\textbf{zhá} ni-gu-ngú\\
              yesterday arrive-\textsc{st-neg} \textsc{spkr.sym-aux.pfv}-\textsc{rec.pst}\\
        \glt ‘S/he did not arrive yesterday.’
    \z
\z 

In comparing the lexical verbs in the affirmative clauses with those in the negative ones, it is apparent that the forms differ. In \REF{ex:kog:5}, for example, the stem denoting ‘go’ has the form \textit{nẽ} in the affirmative and \textit{nẽya} (featuring the stem vowel {}-\textit{a}) in the negated clause. They constitute different inflectional stems, which are conditioned morphologically. That is, \textit{nẽ} is used when an anti{}-clitic follows, i.e. \textit{\#nka-ne} ‘\textsc{1pl.I{}-pst’,} while \textit{nẽya} (\textit{nẽ-a}) is required when inflectional suffixes such as {}-\textit{zhá} ‘\textsc{neg’} follow. The same alternation can be observed with \textit{nak} ‘arrive’ in example \REF{ex:kog:6}. It occurs as \textit{nak} without inflectional suffixes, and as \textit{naka} (\textit{nak-a}) before \textit{{}-zhá} ‘\textsc{neg’.}

It is important to note that this is not a consequence of negation marking specifically but rather a general morphological requirement. Therefore, the periphrastic past constructions can be considered symmetric: while there is a structural difference in the form of the verb stem, it is not induced by negation marking.\footnote{A comparable apparent asymmetry is found in Canela-Krahô (Macro-Ge) as discussed by \citet[69]{Miestamo2005}.}

The periphrastic construction in \REF{ex:kog:6b} has a synthetic counterpart, presented in \REF{ex:kog:7a}. However, such constructions cannot take the negation marker, as shown in \REF{ex:kog:7b}.

\ea\label{ex:kog:7}
    \ea[]{\label{ex:kog:7a} mebák ninaggú.\\
        \gll  mebák ni-nak-ngú\\ 
              yesterday \textsc{spkr.sym-}arrive\textsc{{}-rec.pst}\\
        \glt ‘S/he arrived yesterday.’}
    \ex[*]{\label{ex:kog:7b} 
        \gll ni{}-nak{}-a{}-\textbf{zhá}-ngú\\
            \textsc{spkr.sym-}arrive{}-\textsc{st-neg-rec.pst} \\ } 
    \z
\z 

States of affairs with future time reference, similar to the recent past construction, can be expressed by either a synthetic verb or by a periphrastic construction, as illustrated in \REF{ex:kog:8}. In contrast to the past construction, however, the negation marker \textit{\nobreakdash-zhá} can be used in both variants, i.e. either on the synthetic form, as in \REF{ex:kog:9a}, or the lexical verb in a periphrastic construction, as shown in \REF{ex:kog:9b}.\footnote{The alternation (\textit{k {\textasciitilde} g}) between \textit{naka} and \textit{naga} in examples \REF{ex:kog:5} and \REF{ex:kog:6} is morphologically conditioned and should not concern the reader further.}

\ea\label{ex:kog:8}
    \ea\label{ex:kog:8a} nakalika.\\
        \gll  nak{}-a{}-líka\\ 
              arrive{}-\textsc{st{}-fut}\\
        \glt ‘S/he will arrive.’
    \ex\label{ex:kog:8b} nak gualika.\\
        \gll  nak gu-a-líka\\
              arrive \textsc{aux.pfv-st{}-fut}\\
        \glt ‘S/he will arrive.’
    \z
\z 

\ea\label{ex:kog:9}
    \ea\label{ex:kog:9a}nagazhálika.\\
        \gll  nak-a{}-\textbf{{zhá}}-líka \\
               arrive\textsc{{}-st{}-neg-fut}\\
        \glt ‘S/he won't arrive.’ 
    \ex\label{ex:kog:9b}nagazhá gualika.\\
        \gll  nak-a{}-\textbf{zhá} gu-a{}-líka\\
               arrive\textsc{-st{}-neg} \textsc{aux.pfv-st-fut}\\
        \glt ‘S/he won't arrive.’ 
    \z
\z 

In comparing the affirmative future forms in \tabref{tab:kog:1} with their negative counterparts in \tabref{tab:kog:2}, we can observe some idiosyncrasies which point to asymmetries between the two constructions, namely in the forms of second person and first person plural.\footnote{Note that the 1\textsc{pl.I} allomorph -\textit{kua} used in the affirmative exclusively occurs in future tense.} In the second person forms, the person index changes from a pre-stem position to post-stem position in the negated form. First person plural has two possibilities in negation: one in which the person index follows the stem, as in the affirmative, and one in which it precedes the stem. Both forms can be used interchangeably. The verb pairs of first person singular, and third person singular and plural, by contrast, are symmetrical in that the addition of the negative marker does not cause any structural changes. Note that the stem alternation \textit{tuka {\textasciitilde} tuga} is conditioned morphologically and depends on the nature of the following suffix. 

Symmetry can also be observed in paradigmatic terms, as all distinctions made in the affirmative are possible in the negative.

\begin{table}
\begin{tabularx}{.6\textwidth}{lll}
    \lsptoprule
        & {\scshape sg}                  & {\scshape pl}\\
    \midrule
    {1} & {\itshape tukalikue}           & {\itshape tugakualika} \\
        & {\itshape tuk-a-líka-uge}      & {\textit{tuk-a-kua{}-líka}} \\
        & {drink-\textsc{st-fut-1sg.I}}  & {drink-\textsc{st-1sg.I-fut}}\\
    \midrule
    {2} & {\itshape matukalika}                  & {\itshape mintukalika} \\
        & {\textit{ma{}-tuk-a-líka}}             & {\textit{min{}-tuk-a-líka}} \\
        & {\textsc{2sg.I-}drink-\textsc{st-fut}} & {\textsc{2pl.I{}-}drink-\textsc{st-fut}}\\
    \midrule
    {3} & {\itshape tukalika}         & {\itshape atukalika} \\
        & {\itshape tuk-a-líka}       & {\itshape a-tuk-a-líka} \\
        & {drink-\textsc{st-fut}}     & {\textsc{3pl.I{}-}drink- \textsc{st-fut}} \\
    \lspbottomrule
\end{tabularx}
\caption{Affirmative synthetic future tense of \textit{tuk} ‘drink’}
\label{tab:kog:1}
\end{table}

\begin{table}\small 
\begin{tabularx}{\textwidth}{lXXX} 
    \lsptoprule
    & {\scshape sg} & {\scshape pl} & \\
    \midrule 
    {1} & \textit{tuga\textbf{zhá}likue}    & \textit{tukanka\textbf{lá}lika}   & \textit{katuga\textbf{zhá}lika}\\
        & \textit{tuk-a-zhá-lika-uge}       & \textit{tuk-a-nka-zhá-l\'{i}ka}   & \textit{nka-tuk-a-zhá-l\'{i}ka}\\
        & drink-\textsc{st-neg-1sg.I-fut}   & {drink-\textsc{st-1pl.I-neg-fut}} & \textsc{1pl.I-}drink-\textsc{st-neg-fut}\\
    \midrule
    {2} & \textit{tukaba\textbf{lá}lika}    & \textit{tukabi\textbf{ná}lika}    & \\
        & \textit{tuk-a-ma-zhá-l\'{i}ka}    & \textit{tuk-a-min-zhá-l\'{i}ka}   & \\
        & drink-\textsc{st-2sg.I-neg-fut}   & drink-\textsc{st-2pl.I-neg-fut}   & \\
    \midrule
    {3} & \textit{tuga\textbf{zhá}lika}     & \textit{atuga\textbf{zhá}lika}    & \\
        & \textit{tuk-a-zhá-líka}           & \textit{a-tuk-a-zhá-líka}         & \\
        & drink-\textsc{st-neg-fut}         & \textsc{3pl.I-}drink-\textsc{st-neg-fut} & \\
    \lspbottomrule
\end{tabularx}
\caption{Negative synthetic future tense of \textit{tuk} ‘drink’}
\label{tab:kog:2}
\end{table}

In negated future clauses, the negation marker \textit{{}-zhá}~has different allomorphs, as can be seen in \tabref{tab:kog:2}. The allomorph -\textit{zhá} always occurs directly after the stem (which is also the case in past clauses, where the marker exclusively is found in this environment). In the forms in which a person index precedes \textit{{}-zhá}, the negation marker has the allomorphs \textit{{}-ná} after \textit{n} (2\textsc{pl)} and \textit{{}-lá} after \textit{a.} This variation can be explained by the frequent morphophonological process of denasalization \textit{n → l / \_\_a.}\footnote{E.g. \textit{na}{}- ‘1\textsc{sg.inal’} \textsc{+} \textit{nu} ‘elder sister’ → \textit{nalú} ‘my elder sister’.} This points to a historically underlying form \textit{{}-ná} of the negation marker. The process \textit{n} → \textit{zh / i\_}\footnote{E.g. \textit{mi}{}- ‘2\textsc{sg.inal’} \textsc{+} \textit{nu} ‘elder sister’ → \textit{mizhú} ‘your elder sister’} might account for the form \textit{{}-zhá,} yet there is no element \textit{i} present. For this reason, the allomorphy can be considered morphologically conditioned (at least from a synchronic perspective).

Summarizing the observations for the constructions and paradigms that involve the negation marker -\textit{zhá}, asymmetry is present in the synthetic recent past, as this construction cannot be negated, see example \REF{ex:kog:7} above.\footnote{The same holds for hodiernal and remote past morphemes, which form a single paradigm with the recent past marker.} In all other respects, negation with \textit{{}-zhá} is symmetrical.

\subsubsection{Negative imperfective \textit{-kí}} \label{sec:kog:2.1.2}

Imperfective aspect in affirmative clauses is expressed by a number of markers including \textit{{}-tok{\textasciitilde}nok} ‘\textsc{prog’}\footnote{With 1\textsc{pl} and \textsc{2sg/pl} subjects, the marker has the allomorph -\textit{nok}, which is segmentally identical with its source, namely the lexical verb \textit{nok} ‘be located’. The marker retains some verbal properties in that it is extended by the stem formative -\textit{a} in certain instances, such as negation with \textit{{}-kí} (see below in this section).} which expresses progressive aspect, and the two imperfective markers \textit{{}-t\'{e}} ‘\textsc{ipfv}\textsc{\textsubscript{1}}’ and \textit{{}-ĩ} ‘\textsc{ipfv}\textsc{\textsubscript{2}}’. While the functional and distributional properties between the latter two are not yet fully understood, it appears that \textit{{}-té} often conveys inchoative semantics, focussing on the beginning of a state of affairs, while \textit{{}-ĩ} is used with stative verbs (see \citealt{Knuchel2022} for details). Note that this descriptive shortcoming does constitute a substantial problem for the present discussion, as clauses with either of these markers are negated by one and the same negation marker, namely -\textit{kí} ‘\textsc{neg.ipfv’}.

Imperfectives formed with \textit{{}-t\'{e}} ‘\textsc{ipfv}\textsc{\textsubscript{1}}’ and \textit{{}-ĩ} ‘\textsc{ipfv}\textsc{\textsubscript{2}}’can be followed by the imperfective auxiliary \textit{nok}, which is optional in the present and required in other tenses. The negation marker \textit{{}-kí} attaches to the lexical verb, where it replaces the affirmative imperfective marker, as is illustrated in examples (\ref{ex:kog:10}--\ref{ex:kog:11}).

\ea\label{ex:kog:10}
    \ea\label{ex:kog:10a} inzhi munaté (nzhok).\\
        \gll   inzhi mun-a-té ni-nok\\ 
              yucca grow{}-\textsc{st{}-ipfv}\textsc{\textsubscript{1}} \textsc{spkr.sym-aux.ipfv}\\
        \glt ‘The yucca plants are growing.’
    \ex\label{ex:kog:10b}inzhi munakí (nzhok).\\
        \gll  inzhi mun{}-a-\textbf{kí} ni-nok\\
              yucca grow{}-\textsc{st{}-neg.ipfv} \textsc{spkr.sym-aux.ipfv}\\
        \glt  ‘The yucca plants are not growing.’
    \z
\z 

\ea\label{ex:kog:11}
    \ea\label{ex:kog:11a}sui nakzeshĩ.\\
        \gll  sui nak-zek-ĩ\\ 
              cold \textsc{1sg.III{}-}feel-\textsc{ipfv}\textsc{\textsubscript{2}}\\
        \glt ‘I'm cold.’
    \ex\label{ex:kog:11b}sui nakzegakí.\\
        \gll  sui nak-zek{}-a-\textbf{
        kí}\\
               cold \textsc{1sg.III{}-}feel{}-\textsc{st}{}-\textsc{neg.ipfv}\\
        \glt  ‘I'm not cold.’
    \z
\z 

Clauses with the progressive marker \textit{{}-tok{\textasciitilde}nok} have two possible negative counterparts, which appear to be equal in meaning. In \REF{ex:kog:12b}, the negative imperfective marker -\textit{kí} attaches after -\textit{tok}, to a form that is extended with the stem vowel {}-\textit{a}, apparently a retained trait from the marker's source, the full verb \textit{nok} ‘be located’. While affirmative progressive verbs accommodate person and tense markers, this is not the case in negated forms. Instead, these categories can be marked on the imperfective auxiliary \textit{nok} (optional in present tense).

\ea\label{ex:kog:12}
    \ea\label{ex:kog:12a}arróz nugatukkuge.\\ 
        \gll  arróz nuk-a-tok-ka-uge\\ 
              rice cook\textsc{{}-st}{}-\textsc{prog{}-prs-1sg.I}\\
        \glt ‘I am cooking rice.’
    \ex\label{ex:kog:12b}arr\'{o} z nugatogak\'{i} (nzhokú).\\
        \gll  arróz nuk{}-a-tok-a-\textbf{kí} ni-nok-kú\\
              rice cook{}-\textsc{st}{}-\textsc{prog{}-st{}-neg.ipfv} \textsc{spkr.sym-aux.ipfv}{}-\textsc{1sg.I}\\
        \glt ‘I am not cooking rice.’
    \z
\z  

The sentence in \REF{ex:kog:12a} can equally by negated with the clause in \REF{ex:kog:13}, which is the same negative construction as the one with \textit{{}-\~{i} } and -\textit{té} imperfectives, examples \REF{ex:kog:10b} and \REF{ex:kog:11b} above.

\ea\label{ex:kog:13} arroz nugakí (nzhokú).\\
    \gll  arroz nuk{}-a-\textbf{\textit{kí}} \textit{ni-nok-kú}\\ 
          rice cook{}-\textsc{st-neg.ipfv} \textsc{spkr.sym-}be-\textsc{1sg.I.prs}\\
    \glt ‘I am not cooking rice.’
\z 

Negation marking in the imperfective involves constructional asymmetry. First, in the forms with \textit{{}-\~{i} } and \textit{{}-té}, the aspect markers are dropped in negation. Secondly, negation of imperfectives with \textit{{}-tok} causes the loss of person and tense marking on the lexical verb. Furthermore, we encounter paradigmatic asymmetry, as the distinction among the different imperfective forms is largely neutralized in the negative domain and all constructions are negated with the same marker \textit{\nobreakdash-kí.}

\subsubsection{Negative habitual \textit{-ne}}\label{sec:kog:2.1.3}

Clauses with habitual meaning, marked with \textit{{}-ka,} are negated by the suffix {}-\textit{ne} ‘\textsc{neg.hab’}, as illustrated in example \REF{ex:kog:14}. 

\ea\label{ex:kog:14}
    \ea\label{ex:kog:14a}awalinti matukka.\\
        \gll  awalinti ma-tuk-ka\\ 
              aguardiente \textsc{2sg.I{}-}drink{}-\textsc{hab}\\
        \glt  ‘You drink aguardiente (a type of liquor).’
    \ex\label{ex:kog:14b} awalinti tugabale.\\
        \gll  awalinti tuk{}-a{}-ma-\textbf{ne}\\
              aguardiente drink{}-\textsc{st{}-2sg.I{}-neg.hab}\\
        \glt ‘You never drink aguardiente.’
    \z
\z 

The full paradigms of the verb \textit{tuk} ‘drink’ in its affirmative and negative habitual forms are given in \tabref{tab:kog:3} and \tabref{tab:kog:4}.

\begin{table}[t]
\begin{tabularx}{.6\textwidth}{lXX} 
    \lsptoprule
        & {\scshape sg}             & {\scshape pl}\\
    \midrule  
    {1} & {\itshape tukkuge}        & {\itshape katukka} \\
        & {\itshape tuk-ka-uge}     & {\itshape ka-tuk-ka} \\
        & drink-\textsc{hab-1sg.I}  & {\textsc{1pl.I{}-}drink-\textsc{hab}}\\
    \midrule
    {2} & {\itshape ma-tuk-ka}      & {\itshape mintukka} \\
        & {\itshape ma-tuk-ka}      & {\itshape min-tuk-ka} \\
        & {2\textsc{sg.I-}drink-\textsc{hab}} & \textsc{2pl.I-}drink-\textsc{hab}\\
    \midrule
    {3} & {\itshape tukka}          & {\itshape atukka} \\
        & {\itshape tuk-ka}         & {\itshape a-tuk-ka} \\
        & {drink-\textsc{hab}}      & {\textsc{3pl.I{}-}drink-\textsc{hab}} \\
\lspbottomrule
\end{tabularx}
\caption{Habitual paradigm of \textit{tuk} ‘drink’}
\label{tab:kog:3}
\end{table}

\begin{table}[t]
\begin{tabularx}{.8\textwidth}{lXX}
    \lsptoprule
    & {\scshape sg} & {\scshape pl}\\
    \midrule
    {1} & {\itshape tugazhuge}              & {\itshape tukankale} \\
        & \textit{tuk-a-\textbf{ne}-uge}    & \textit{tuka-nka-\textbf{ne}} \\
        & drink-\textsc{st-neg.hab-1sg.I}   & {drink-\textsc{st-1pl.I-neg.hab}} \\
    \midrule
    {2} & {\itshape tugabale}               & {\itshape tugabine} \\
        & \textit{tuk-a-ma-\textbf{ne}}     & \textit{tuk-a-min-\textbf{ne}} \\
        & drink-\textsc{st-2sg.I-neg.hab}   & drink\textsc{-st-2pl.I-neg.hab}\\
    \midrule
    {3} & \textit{tugagale}                 & \textit{atugagale} \\
        & \textit{tuk-a-\textbf{gale}}      & \textit{a-tuk-a-\textbf{gale}} \\
        & drink-\textsc{st-neg.hab}         & \textsc{3pl.I}-drink-\textsc{st-neg.hab}\\
    \lspbottomrule
\end{tabularx}
\caption{Negative habitual paradigm of \textit{tuk} ‘drink’}
\label{tab:kog:4}
\end{table} 

While in the affirmative habitual, subject person is marked by verbal prefixes, the negated forms feature subject suffixes. The habitual negation marker has an allomorph \textit{{}-zhe} in the first person singular. This alternation is reminiscent of the one in negated future verbs (see \tabref{tab:kog:2} above). Similarly to the allomorphy in the negative future, no phonological environment that would account for this change can be determined, as both -\textit{zh(e)} and \textit{{}-ne} occur after \textit{a.} Thus, it can be maintained that the allomorphs are conditioned by morphology. Moreover, third person verbs feature an allomorph \-\textit{gale,} which is attested exclusively in these forms.

The constructions are analyzed as asymmetric, given that the habitual marker \textit{\nobreakdash-ka} is dropped in the negated forms. Habitual aspect is expressed alongside negation by \textit{{}-ne.} Furthermore, a striking structural difference is the position of subject indexes in the forms of first person plural and second person singular and plural. While these markers are prefixed in the affirmative, they occur after the verb stem as suffixes in the negative habitual.

While the marker \textit{{}-ka} can have both a habitual and a present imperfective interpretation, -\textit{ne} only negates habitual states of affairs. In other words, there is no one-to-one correspondence between the affirmative and the negative habitual on formal grounds. Rather, the choice of the negation marker depends on the aspectual/temporal interpretation of a clause. For example, the clause in \REF{ex:kog:15} is marked with the first person present/habitual marker, has a present imperfective interpretation and is negated with \textit{\nobreakdash-kí}.


\ea\label{ex:kog:15}
    \ea\label{ex:kog:15a}nahuli nẽkuge.\\
        \gll  na-hu-ni nẽ-ka-uge\\ 
              \textsc{1sg.inal{}-}house-\textsc{all} go-\textsc{prs-1sg.I}\\
        \glt ‘I'm going home.’
    \ex\label{ex:kog:15b}nahuli nẽyakí.\\
        \gll  na-hu-ni nẽ-a-\textbf{kí}\\
              \textsc{1sg.inal}\textsc{-}house-\textsc{all} go-\textsc{st-neg.ipfv}\\
        \glt ‘I'm not going home.’
    \z
\z 

The negative habitual verb forms can be combined with auxiliaries that denote tense to refer to, for instance, past situations as in \REF{ex:kog:16}. 

\ea\label{ex:kog:16}
    \ea\label{ex:kog:16a}sukkua nalgukuéng ũwã gákuge nane.\\
        \gll  sukkua nal{}-gu{}-ku-éng ũwã  ga{}-ka-uge nal\#ne\\ 
              child be-\textsc{rem.pst-1sg.I{}-sim} meat eat-\textsc{hab-1sg.I} \textsc{aux}\#\textsc{pst}\\
        \glt ‘When I was I child, I had the habit of eating meat.’
    \ex\label{ex:kog:16b}ũwã gazhuge nane.\\
        \gll  ũwã ga-\textbf{ne}-uge nal\#ne\\
              meat eat-\textsc{neg.hab-1sg.I} \textsc{aux}\#\textsc{pst}\\
        \glt ‘I did not have the habit of eating meat.’
    \z
\z 

\subsubsection{Negation in modal constructions}\label{sec:kog:2.1.4}

Kogi marks four types of event modality: obligative, volitive, abilitive (for acquired ability) and potential. While the first three are expressed by a combination of a lexical verb and a modal verb, potential modality is expressed through a synthetic lexical verb carrying a modal marker.

Negated clauses marked for modality involve the standard negation markers described above. However, there are further strategies that can be observed in the negation of modality, for example a negative construction which involves the modal suffix -\textit{sïng(ze)} on lexical verbs. Another construction, namely the one of negative abilitive modality, involves the negative verb \textit{sïng(ze)}.

Negation is marked on the modal verb in the constructions that feature one, i.e. obligative, volitive and abilitive. This is illustrated with the obligative modality clauses in \REF{ex:kog:17}, which involve a lexical verb marked with \textit{{}-kue} and the modal verb \textit{agatsa}.\footnote{The obligative modal verb can be analyzed as consisting of an indirect object marker (typical for experiencer predicates), a possessive marker and the stem \textit{zal} ‘be’: \textit{agatsa} – \textit{ak-a-zal} (\textsc{3sg.III{}-poss}{}-be). This is reminiscent of the verb form in possessive predication which is based on a different lexeme with the meaning ‘be’, namely \textit{nal}. For details about possessive predicates see \sectref{sec:kog:2.3}.}

\ea\label{ex:kog:17}
    \ea\label{ex:kog:17a}colegioli nẽyakue nagatsane.\\
        \gll  colegio-ni nẽ-a-kue nak-a-zal\#ne\\ 
              school-\textsc{all} go\textsc{-st-oblig} \textsc{1sg.III{}-poss-}be\#\textsc{pst}\\
        \glt  ‘I had to go to school.’
    \ex\label{ex:kog:17b}colegioli nẽyakue nagatsalazhane.\\
        \gll  colegio-ni nẽ{}-a-kue nak-a-zal{}-a-\textbf{zhá}\#ne\\
              school-\textsc{all} go\textsc{-st-oblig} \textsc{1sg.III{}-poss-}be{}-\textsc{st{}-neg}\#\textsc{pst}\\
        \glt ‘I didn't have to go to school.’
    \z
\z 

In obligative, volitive and abilitive clauses, markers of standard negation are attached to the modal verb. In this respect, these negative constructions contrast with those unmarked for modality, where the locus of negation marking typically is the lexical verb. Any of the major negation markers can be used, depending on the aspectual value of the clause.

The potential marker \textit{{}-akka} can have a permissive or abilitive interpretation. Potential verbs consist of a verb stem, followed by a person index and the potential marker, as is shown in \REF{ex:kog:18}. Such predicates exclusively accommodate the negative marker \textit{{}-zhá.}

\ea\label{ex:kog:18}
    \ea\label{ex:kog:18a}zú kazïnakka.\\
        \gll  zï{}-nuk{}-a-zïna-akka\\ 
              \textsc{antip{}-}cook\textsc{{}-st-1sg.II{}-pot}\\
        \glt ‘We may cook.’
    \ex\label{ex:kog:18b}zugazházïnakka.\\
        \gll  zï{}-nuk{}-a-\textbf{zhá}-zïna-akka\\
              \textsc{antip{}-}cook\textsc{{}-st-neg-1sg.II{}-pot}\\
        \glt  ‘We may not cook.’
    \z
\z 

Negation of abilitive clauses that refer to an acquired ability involves the negative existential verb \textit{sïngze}. Its affirmative counterpart \textit{kual {\textasciitilde} kue} ‘exist’ is used with the sense of knowing or being capable of something when indirect object markers index the subject. With this meaning, the verb is used as a modal verb in abilitive clauses, as exemplified in \REF{ex:kog:19a}, and is negated by replacing the stem with the negative verb \textit{sïng(ze)} \REF{ex:kog:19b}. Note that this polysemy of existence in modal constructions has been observed as well in other languages, where the same expression is used in abilitive and existential constructions. This is noted, for example, by \citet{Veselinova2013neg.exist} for Hawai’ian. Alternatively, abilitive clauses can be negated with the habitual negative marker, as in \REF{ex:kog:19c}.

% \lv{Please provide non-glossed lines for consistency with all other examples}
\ea\label{ex:kog:19}
    \ea huẽna nakkué. \label{ex:kog:19a}\\
        huẽn-a nak-kué\\
        \gl  swim\textsc{{}-st 1sg.III{}-}exist\\ 
        \glt ‘I can / know how to swim.’
    \ex huẽna naksïngze.\label{ex:kog:19b}\\
        huẽn-a nak-\textbf{sïngze} \\
        \gl  swim\textsc{{}-st 1sg.III{}-neg.ex}\\ 
        \glt ‘I can't / don't know how to swim.’ (lit. ‘Swimming exists / does not exist to me.’)
    \ex huẽna nakkualagále. \label{ex:kog:19c}\\
        huẽn-a nak-kual-a-\textbf{gale}\\
        \gl  swim\textsc{{}-st 1sg.III{}-}exist\textsc{{}-st-neg.hab}\\
        \glt  ‘I can't / don't know how to swim.’
    \z
\z 

Note that the verb \textit{sïngze} is the negative existential, discussed in \sectref{sec:kog:2.3} below.

In addition to negation with the major markers discussed above, obligative, volitive and potential modality verbs may take the suffix \textit{{}-sïng(ze)}, which is a grammaticalized form of the negative existential verb.\footnote{Cross-linguistically, negative existentials are a common source of markers of verbal negation, cf. the “Negative Existential Cycle” discussed by \citet{Croft1991} or \citet{Veselinova2014}, as is the case in Kogi. It must be noted, however, that such developments are often only partial takeovers of the domain of verbal negation, as discussed by \citet{Veselinova2016}.} The forms -\textit{sïng} and -\textit{sïngze} appear to be interchangeable; the source or original function of \textit{ze} is unknown.

The suffix attaches to the lexical verb (rather than to the modal) and is neutral in terms of modal meaning, i.e. it is used for any of the three modalities. The negation marker replaces the one of modality, as illustrated in (\ref{ex:kog:20}--\ref{ex:kog:21}). The type of modality instead is indicated by the modal verb \REF{ex:kog:19b}, or can be inferred from context, as in \REF{ex:kog:21b}.\footnote{In negative modal clauses with -\textit{sïng(ze),} the modal verb is often omitted.}

\ea\label{ex:kog:20}
    \ea\label{ex:kog:20a}matṹ nana nakzeshĩ.\\
        \gll   matũ-nana nak-zek-ĩ\\ 
              \textsc{2sg.II{}-}see-\textsc{vol} \textsc{1sg.III{}-}feel-\textsc{ipfv}\textsc{\textsubscript{2}}\\
        \glt  ‘I want to see you.’ 
    \ex\label{ex:kog:20b}alé ngki tũasḯng akzeshĩ.\\
        \gll  aléng-ki tũa{}-\textbf{sïng} ak-zek-ĩ\\
              3\textsc{sg-top} see\textsc{{}-neg} \textsc{3sg.III{}-}feel-\textsc{ipfv}\textsc{\textsubscript{2}}\\
        \glt ‘She doesn't want to see [him].’ 
    \z
\z 

\ea\label{ex:kog:21}
    \ea\label{ex:kog:21a}tũanakka.\\
        \gll  tũ{}-a-na-akka\\ 
               see\textsc{{}-st-1sg.II{}-pot}\\
        \glt  ‘I can see [it].’
    \ex\label{ex:kog:21b}muldetwa tũasĩng.\\
        \gll muldetwa tũa-\textbf{sïng} \\
              well see\textsc{{}-neg}\\
        \glt ‘I can't see it well.’ 
    \z
\z 

Negation with -\textit{sïng(ze)} can be characterized as asymmetric in both constructional and paradigmatic terms. The negative modality marker replaces affirmative modality markers, which constitutes a structural difference. Moreover, in the case of potential modality (\ref{ex:kog:20}), person marking on the lexical verb is omitted. As for the paradigmatic structure, the distinction between obligative, volitive and potential is neutralized in the negative paradigm.

The different strategies of negation and their compatibility with different modalities are summarized in \tabref{tab:kog:5}. At present, it is not clear what determines the choice of the standard negation markers or the negative existential.

\begin{table}\small 
\begin{tabularx}{\textwidth}{llXll} 
    \lsptoprule
                    & {Structure}               & {Standard}            & {Suffix}                      & {Verb}\\
                    & {(3\textsc{sg,} affirm.)} & {negators}            & {\textit{-sïng(ze)}}          & {\textit{sïng(ze)}}\\
    \midrule 
    {Obligative}    & {V-\textit{kue agatsé}}   & {any, on modal V}     & {V-\textit{sïng(ze) }}        & {–}\\
                    & {‘he has to V’}           &                       & \textit{agatsé}               & \\
    %\midrule
    {Volitive}      & {V-\textit{nana akzesh\~{i}}} & {any, on modal V} & {V\textit{{}-sïng(ze)}}       & {–}\\
                    & {‘he feels like V-ing’}   &                       & \textit{akzeshi}              &\\
    %\midrule
    {Abilitive}     & {V \textit{akkué}}        & {any, on modal V}     & {–}                           & {V \textit{aksïng(ze)}}\\
                    & {‘he knows how to V’}     &                       &                               &\\
    %\midrule
    {Potential}     & {V-\textit{akka}}         & {exclusively \textit{-zhá}} & {V\textit{{}-sïng(ze)}} & {–}\\
                    & {‘s/he can/may V’}        & on lexical verb &&\\
    \lspbottomrule
\end{tabularx}
\caption{Negation in modal clauses}
\label{tab:kog:5}
\end{table}

\subsubsection{Phasal polarity}\label{sec:kog:2.1.5}

Kogi has dedicated markers of phasal polarity, a category that combines polarity distinctions with temporal/aspectual meanings (\citealt{Auwera1998}, \citealt{Kramer2017}). This is illustrated in the examples in \REF{ex:kog:22}. 

\ea\label{ex:kog:22}
    \ea\label{ex:kog:22a} José ya ahuk izhukka.\\
        \gll  José \textbf{ya} a-hu-k i-nok-ka\\ 
              José already \textsc{3sg.inal{}-}house-\textsc{loc} \textsc{loc.appl{}-}be.located-\textsc{prs}\\
        \glt ‘José is already at home.’
    \ex\label{ex:kog:22b}José ahuk izhukkaga.\\
        \gll  José a-hu-k i-nok-ka{}-\textbf{ga}\\
              José \textsc{3sg.inal{}-}house-\textsc{loc} \textsc{loc.appl{}-}be.located-\textsc{prs{}-}still\\
        \glt  ‘José is still at home.’
     \ex\label{ex:kog:22c}José ahuk izhogaksá.\\
        \gll  José a-hu-k i-nok{}-a-\textbf{ksá}\\
              José \textsc{3sg.inal{}-}house-\textsc{loc} \textsc{loc.appl{}-}be.located\textsc{{}-st{}-}no.longer\\
        \glt ‘José is no longer at home.’
    \ex\label{ex:kog:22d}José ahuk izhogag\'{a}nga.\\
        \gll  José a-hu{}-k i-nok{}-a{}-\textbf{g\'{a}nga}\\
              José \textsc{3sg.inal{}-}house-\textsc{loc} \textsc{loc.appl{}-}be.located\textsc{{}-st{}-}not.yet\\
        \glt ‘José is not yet at home.’
    \z
\z 

Expressions of phasal polarity have been characterized as involving “reference points at two related phases implying situations which are contrasted as opposites with different polarity values” \citep[1]{Kramer2017}. More precisely, the expressions \textit{already} and \textit{still} indicate that the proposition \textit{John is at home} holds at a certain reference point, while at the same time referring to a point in time, either preceding (i.e. \textit{already}) or subsequent (i.e. \textit{still}), at which it does not hold. The expressions \textit{no longer} and \textit{not yet,} by contrast, express that the proposition is not the case at a certain reference point, while referring to a reference point, either preceding (i.e. \textit{no longer}) or subsequent (i.e. \textit{not yet}), at which the proposition holds.

In Kogi, the phasal polarity value of ‘already’ is expressed by the Spanish loan \textit{ya} ‘already’ \REF{ex:kog:22a}, and the value ‘still’ by the suffix -\textit{ga} \REF{ex:kog:22b}.\footnote{The marker \textit{{}-ga} is also found with parts of speech other than verbs, e.g. nouns or postpositions, and depending on the context has a number of different functions: in addition to ‘still’, it may be used for emphasis and to denote ‘too, also’ or ‘the same’. This is a common polysemy of markers of ‘still’ (cf. van \citealt{Baar1997}).} Both negative phasal values are expressed by affixes, namely {}-\textit{ksá} ‘no longer’ \REF{ex:kog:22c} and \nobreakdash-\textit{gánga} ‘not yet’ \REF{ex:kog:22d}.

While forms with -\textit{gánga} do not take person marking, this is the case in constructions with -\textit{ksá}, as exemplified in \REF{ex:kog:23}. Instead of subject indexes, the subject is referenced with direct object markers in this construction.

\ea\label{ex:kog:23} huk izhogalaksá.\\
    \gll  hu-k i-nok{}-a-na-\textbf{ksá}\\ 
          house-\textsc{loc} \textsc{loc.appl}{}-be.located-\textsc{st{}-1sg.II{}-}no.longer\\
    \glt  ‘I am no longer at home.’
\z 

With regard to tense, no restrictions in the use of -\textit{ksá} and -\textit{gánga} are attested. In other words, negative phasal polarity affixes can be used in clauses with past and present as well as future time reference. The markers only apply in periphrastic constructions where they attach to the lexical verb while tense is marked on the auxiliary. This is illustrated in \REF{ex:kog:24} and \REF{ex:kog:25}.

\ea\label{ex:kog:24} nuzhí nki mulbatá akzegaksá gualika.\\
    \gll  nuzhín-ki mulbatá ak{}-zek{}-a{}-\textbf{ksá} gu-a{}-líka\\ 
          tomorrow-\textsc{top} sickness \textsc{3sg.III{}-}feel{}-\textsc{st}{}-no.longer \textsc{aux.pfv{}-st}{}-\textsc{fut}\\
    \glt ‘Tomorrow, I will no longer be sick.’
\z 

\protectedex{\ea\label{ex:kog:25} mebakke tuang zeshĩ nki nahate nagagánga nigungú. \\
    \gll  mebák{}-ke tuang zek-ĩ-ki na-hate nak-a-\textbf{gánga} ni-gu-ngú\\ 
          yesterday-\textsc{sub} dark happen-\textsc{st-ipfv}\textsc{\textsubscript{2}}\textsc{-top} \textsc{1sg.inal{}-}father arrive\textsc{{}-st{}-}not.yet \textsc{spkr.sym-aux.pfv}{}-\textsc{rec.pst}\\
    \glt  ‘Last night, my father hadn't arrived yet.’
\z  }

 
Verbs marked with \textit{-ksá} ‘no longer’ necessarily denote a durative state of affairs, which no longer holds at the time of speaking. Thus, predicates marked with -\textit{ksá} always have a perfective reading. States of affairs marked with \textit{{}-gánga} ‘not yet’, by contrast, are neutral in terms of aspectual interpretation. That is, depending on the employed auxiliary, it has a perfective \REF{ex:kog:26a} or imperfective \REF{ex:kog:26b} reading.

\ea\label{ex:kog:26}
    \ea\label{ex:kog:26a}María zugagánga niguá.\\
        \gll  María zï-nuk{}-a-\textbf{gánga} ni-gu-á\\ 
              María \textsc{antip-}cook\textsc{{}-st}{}-not.yet \textsc{spkr.sym-aux.pfv}{}-\textsc{hod.pst}\\
        \glt ‘María did not cook yet.’
    \ex\label{ex:kog:26b}María zugagánga nzhok.\\
        \gll  María zï-nuk-a{}-\textbf{gánga} ni-nok\\
              María \textsc{antip-}cook\textsc{{}-st}{}-not.yet \textsc{spkr.sym-aux.ipfv}\\
        \glt ‘María is not cooking yet.’
    \z
\z 

While English has distinct expressions for each of the values, there are languages in which there is a relationship in terms of negation between two members, i.e. they constitute a pair of the same expression in an affirmative and a negated form. The Spanish expression \textit{todavía no} ‘not yet’, for example, is a negated form of \textit{todavía} ‘still’. \citet{Löbner1989} characterizes the relationships that hold among phasal polarity expressions in terms of internal versus external negation. In internal negation, an element has the negation in its scope, i.e. (x \textsc{not),} whereas in external negation, the element is within the scope of negation, i.e. \textsc{not(}x). Thus, the relationship between Spanish \textit{ya} and \textit{ya no} is an internal one and can be conceptualized as (still \textsc{not)} \textsc{=} \textsc{‘}not yet’. 

Along similar lines, a relationship of internal negation may be postulated to hold between \textit{{}-ga} ‘still’ and -\textit{g\'{a}nga} ‘not yet’, as it seems that the former is included in the latter. That is, -\textit{g\'{a}nga} can be considered a negated form of {}-\textit{ga}, given that one of the syllables is considered a negation marker. While the morpheme \textit{{}-ga} is not attested as a negation marker in any other environment, it is associated with negation in that it is part of the prohibitive marker -\textit{gabba} (see \sectref{sec:kog:2.2.1}), and occurs in the third person allomorph of the negative habitual, i.e. -\textit{gale}. At present, I have no conclusive answer to the historical source of -\textit{gánga.}

(A)symmetries in the structure of paradigms, as introduced by \citet{Miestamo2005}, can also be determined in phasal polarity systems. \citet[16--19]{Kramer2017} makes a distinction between internal and external (a)symmetry. The internal symmetry relates to expressibility (i.e. a symmetric system has expressions for each of the four values) and morphosyntactic status (i.e. in terms of wordhood / morphosyntactic position) of phasal polarity expressions. Kramer’s concept of external symmetry is similar to paradigmatic symmetry as introduced by \citet{Miestamo2005}. That is, the system is symmetric if the same distinctions (e.g. in terms of TAM) in the affirmative can combine with all phasal polarity expressions.

From the internal (a)symmetry perspective, the Kogi system can be considered asymmetric in terms of morphosyntactic status among the expressions. Both negative phasal polarity markers attach to the dependent verb stem. The marker denoting ‘still’ by contrast, occurs in the left most slot of an inflected verb form. Finally, the particle \textit{ya} ‘already’ borrowed from Spanish stands out in that it is an unbound morpheme that is not restricted to only one syntactic slot in the clause. In this respect Kogi is similar to many European languages where expressions for ‘already’ are markedly different from the other phasal polarity expressions (cf. \citealt{Auwera1998}).

In terms of tense, the negative phasal polarity markers show paradigmatic symmetry. That is, all tense distinctions can occur in the corresponding negative polarity constructions. The interactions of negative phasal polarity and aspect are more complex. While \textit{{}-gánga} ‘not yet’ is compatible with both perfective and imperfective aspect, \textit{{}-ksá} ‘no longer’ always has a perfective interpretation, in that it denotes events that are completed.

\subsection{Negation in non-declaratives}\label{sec:kog:2.2}

Negative non-declarative clauses generally feature the same negation markers as declarative clauses. Exceptions are imperative and optative clauses with second person subjects, which are negated with the suffix \textit{{}-gabba.}

\subsubsection{Negation of optatives and imperatives}\label{sec:kog:2.2.1}
 
Optative mood, i.e. the expression of hopes or wishes, is marked by the suffix \textit{{}-lí.}\footnote{The marker -\textit{lí} is generally associated with future reference. It is used as future marker in subordinate verbs and is a component of the future marker -\textit{líka.} When occurring in main clauses, -\textit{lí} denotes an uncertain future.} As exemplified in \REF{ex:kog:27}, the negation marker \textit{{}-zhá} is used in the negative clause \REF{ex:kog:27b}.

\ea\label{ex:kog:27}
    \ea\label{ex:kog:27a}hanshibé  nakalí.\\
        \gll  hanshibé nak{}-a-lí\\ 
              good arrive\textsc{{}-st-opt}\\
        \glt ‘Hopefully, he will arrive safely.’
\newpage
    \ex\label{ex:kog:27b}itshuanazháli.\\
        \gll  itshuan{}-a-\textbf{zhá}-lí\\
              get.lost\textsc{-st-neg-opt}\\
        \glt  ‘Hopefully, he doesn't get lost.’
    \z
\z 

Optative clauses with second person subjects differ in terms of negation as they employ the dedicated negation marker \textit{{}-gabba}, as illustrated in \REF{ex:kog:28}.

\ea\label{ex:kog:28}
    \ea\label{ex:kog:28a}hanshibé malẽyalí.\\
        \gll  hanshibé ma-nẽ-a-lí\\ 
              good \textsc{2sg.I{}-}go\textsc{{}-st-opt}\\
        \glt  ‘Travel safely!’ (lit. ‘May you go well!')
    \ex\label{ex:kog:28b}tashimali mitshuanagabba.\\
        \gll  tashima-ni ma-itshuan{}-a-\textbf{gabba}\\
              woods-\textsc{loc} \textsc{2sg.II{}-}get.lost\textsc{{}-st-neg.opt}\\
        \glt ‘May you not get lost in the woods.’
    \z
\z 

The marker \textit{{}-lí} is also used in the expression of invitations, which can be conceived of as a mitigated form of commands. For the more prototypical expression of commands, namely imperatives, Kogi has two dedicated markers for second singular and second plural clauses, which is cross-linguistically very common (see \citealt{AuweraEtAl2013}). Imperatives featuring {}-\textit{gua} ‘\textsc{imp.sg’} and \textit{{}-guĩ} ‘\textsc{imp.pl’} are illustrated in \REF{ex:kog:29}.

\ea\label{ex:kog:29}
    \ea\label{ex:kog:29a}
        \gll  tũ-gua!\\ 
             look-\textsc{imp.sg}\\
        \glt ‘Look! (\textsc{sg})’
    \ex\label{ex:kog:29b}
        \gll  tũ-guĩ!\\
              look-\textsc{imp.pl}\\
        \glt ‘Look! (\textsc{pl})’
    \z
\z 

Negative imperatives, also called prohibitives (\citealt{AuweraEtAl2013-wals70}), also feature the negation marker \textit{{}-gabba,} as shown in \REF{ex:kog:30}. The Kogi prohibitive belongs to the fourth type defined by \citet{AuweraEtAl2013-wals70}, i.e. it differs structurally from the affirmative imperative and features a negation marker that does not occur in indicative declaratives.

\ea\label{ex:kog:30}
    \gll tũ{}-a-\textbf{gabba}!\\
         look\textsc{{}-st}{}-\textsc{proh}\\
    \glt‘Don't look! (\textsc{sg/pl})’
\z    

Second person optative and imperative clauses are structurally asymmetrical given that the mood marker is dropped when the clause is negated. Moreover, optative and imperative paradigms can be characterized as asymmetrical, as optative clauses have two types of negative constructions depending on the subject. Moreover, the number distinction found in imperative clauses is neutralized in prohibitives.

It is evident that the negation marker -\textit{gabba} in non-declarative clauses is associated with second person. The same suffix is used in negative circumstantial clauses such as in \textit{walk without looking,} which are discussed in \sectref{sec:kog:2.4} below.

 
\subsubsection{Negation in questions}\label{sec:kog:2.2.2}

As for interrogatives, there are no dedicated negation markers and the same constructions as in standard negation are used. It can be maintained that questions tend to be marked on auxiliaries, whereas negation markers more commonly attach to lexical verbs. A question formed by the interrogative marker \textit{{}-é} is illustrated in \REF{ex:kog:31}.\footnote{There are two further interrogative strategies, namely marking only by falling intonation and by engagement markers which target the addressee’s knowledge. All interrogative constructions behave alike in terms of negation.}

\ea\label{ex:kog:31}
    \ea\label{ex:kog:31a}arróz habbi makungué?\\
        \gll  arróz habbi ma-gu-ngú-é\\ 
               rice buy \textsc{2sg.I}{}-\textsc{aux.pfv{}-rec.pst-q}\\
        \glt ‘Did you buy rice?’
    \ex\label{ex:kog:31b}arróz habbiazhá makungué?\\
        \gll  arróz habbi{}-a-\textbf{zhá} ma-gu-ngú-é?\\
              rice buy{}-\textsc{st{}-neg} \textsc{2sg.I}{}-\textsc{aux.pfv{}-rec.pst-q}\\
        \glt ‘Didn't you buy rice?’
    \z
\z 

When expressing a request for information with a negative interrogative construction, the speaker typically has reason to believe that the questioned proposition is not the case. This is illustrated in the question-answer pair in \REF{ex:kog:32}, in which speaker A, who in fact knows about the condition of speaker B’s father, asks for confirmation of his assumption that the father did not come down from the village in the mountains. Speaker B answers the question with the negative reply \textit{nada} and confirms speaker A's assumption.\footnote{\textit{nada} is borrowed from Spanish \textit{nada} ‘nothing’ and is used as a sort of emphatic negative reply (see \sectref{sec:kog:3.1}).}

\ea\label{ex:kog:32}
    \begin{xlist}
    \exi{Q: }\label{ex:kog:32q}mihate zabazhá?\\
        \gll  mi{}-hate zabi{}-a{}-\textbf{zhá}\\ 
              \textsc{2sg.inal{}-}father descend\textsc{{}-st-neg}\\ 
        \glt  ‘Your father did not come down?’
    \exi{A: }\label{ex:kog:32a}nada ẽki kala akzeshĩ.\\
        \gll  nada ẽ-ki kala ak-zek-ĩ\\ 
               no \textsc{dem-top} leg \textsc{3sg.III{}-}hurt-\textsc{ipfv}\textsc{\textsubscript{2}}\\
        \glt ‘No, his legs are hurting.’
    \end{xlist}
\z 

Negative interrogatives are often used as tag questions to seek the interlocutor’s confirmation, as in the following examples (\ref{ex:kog:33}--\ref{ex:kog:34}) from an interactional, picture-based story building task.


\ea\label{ex:kog:33}hika nẽshatukka hana agatsé, \textsc{nalagale shina}?\\
    \gll hika nẽ-sha-tuk-ka hana agatsé nal{}-a-\textbf{gale} shi-nal\\
        something go-\textsc{caus-prog{}-prs} like look be-\textsc{st-neg.hab} \textsc{addr.sym}{}-be\\
    \glt ‘It looks like he is vending something, \textsc{doesn't it}?’ 
\z 

\ea\label{ex:kog:34}hũa, somá nóngutse, \textsc{tũakí shimala}?\\
    \gll hũa somá nóngutse tũ{}-a-\textbf{kí} shi-ma-nal\\
          look baby little see-\textsc{st-neg.ipfv} \textsc{addr.sym-2sg.I}{}-be\\
    \glt‘Look, the baby is little, \textsc{don't you see?}’ 
\z 

\subsection{Negation in stative predications}\label{sec:kog:2.3}

Stative predicates do not exhibit any specialized markers and the major negation markers are used in these constructions. The choice of the marker depends on the aspectual interpretation of a clause, which is often determined by the type of state of affairs in terms of (im)permanence. 

In the affirmative domain, equation and proper inclusion in the present tense are typically expressed by juxtaposition of the subject and the predicate. The copula \textit{nal} ‘be’ is optional in present tense (as in \REF{ex:kog:35a}, whereas it is required when markers of tense or modality are employed. This is illustrated in \REF{ex:kog:35} with an example of an equation clause, where \textit{\-gale} ‘\textsc{neg.hab’} attaches to the copula.

\ea\label{ex:kog:35}
    \ea\label{ex:kog:35a}Maríaki nahaba (na).\\
        \gll  María-ki na{}-haba nal\\ 
              María-\textsc{top} \textsc{1sg.inal{}-}mother \textsc{cop} \\
        \glt  ‘María is my mother.’
    \ex\label{ex:kog:35b}Maríaki nahaba nalagale.\\
        \gll  María-ki na-haba nal{}-a-\textbf{gale}\\
               María-\textsc{top} \textsc{1sg.inal{}-}mother \textsc{cop{}-st}{}-\textsc{neg.hab}\\
        \glt ‘María is not my mother.’
    \z
\z 

With regard to structural differences, the clauses in \REF{ex:kog:35} differ in the obligatoriness of the copula verb. Note however, that this is not due to negation marking specifically but rather a general morphosyntactic requirement for expressing certain inflectional categories such as negation or tense (e.g. past tense in \REF{ex:kog:36} below).

Adjectives in Kogi can be divided broadly into two classes: those that are used independently, and those that always occur with an auxiliary verb in both attributive and predicative function. The independent class consists of a relatively small number of adjectives which denote properties such as age, color or size. Similar to nominal predicates, adjectival predicates in present tense are juxtaposed with the subject, the copula being optional. The use of inflectional markers requires the copula, as in the past tense forms in \REF{ex:kog:36}. The copula is also the base for negation marking, as shown in \REF{ex:kog:36b}.

\ea\label{ex:kog:36}
    \ea\label{ex:kog:36a}zhiklukká nánuge.\\
        \gll   zhiklukká nal\#ne-uge\\ 
              tall \textsc{cop}\#\textsc{pst-1sg.I}\\
        \glt  ‘I was tall.’
    \ex\label{ex:kog:36b} zhiklukká nalazhánuge.\\
        \gll  zhiklukká nal{}-a{}-\textbf{zhá}\#ne-uge\\
              tall \textsc{cop}{}-\textsc{st-neg\#pst-1sg.I}\\
        \glt ‘I was not tall.’
    \z
\z 

Adjectives belonging to the second class cannot occur on their own but must be combined with an auxiliary in both predicative or attributive use, namely \textit{ats\'{a}l} for properties that are visually perceivable and \textit{zal} for properties that are perceived through other senses, e.g. taste. These dependent adjectives express properties that can be either temporary or permanent, a distinction which is reflected in the choice of negation marker. More permanent states are negated with the negative habitual \REF{ex:kog:37b}, while those that hold temporarily take the negative imperfective \REF{ex:kog:38}.

\newpage
\ea\label{ex:kog:37}
    \ea\label{ex:kog:37a}hanshi nagatsé.\\
        \gll  hanshi nak{}-atsal\\ 
               beautiful \textsc{1sg.III{}-aux}\\
        \glt ‘I'm beautiful.’
    \ex\label{ex:kog:37b}hanshi nagatsalagale.\\
        \gll  hanshi nak{}-atsal{}-a-\textbf{gale}\\
              beautiful \textsc{1sg.III{}-aux{}-st}{}-\textsc{neg.hab}\\
        \glt ‘I'm not beautiful.’
    \z
\z 

\ea\label{ex:kog:38}
    \ea\label{ex:kog:38a}zhakua hanshi agatsé.\\
        \gll  zhakua hanshi ak-atsal\\ 
              dress clean \textsc{3sg.III{}-aux}\\
        \glt ‘The dress is clean.’
    \ex\label{ex:kog:38b}zhakua hanshi agatsaldakí.\\
        \gll  zhakua hanshi ak-atsal-a-\textbf{kí}\\
              dress clean \textsc{3sg.III{}-aux-st-neg.ipfv}\\
        \glt  ‘The dress is not clean.’
    \z
\z 

Existential and locative predications are structurally similar in Kogi. That is, both kinds of constructions can feature either a locative verb, the existential verb \textit{kual {\textasciitilde} kue,} or a combination of both. While there are certain characteristics that set the two construction types apart, they are identical in terms of negation, i.e. in both, negation markers attach to the existential verb.

A prototypical existential involves a compound verb consisting of a locative root and the existential verb, as in \REF{ex:kog:39}.

\ea\label{ex:kog:39}baka abaksï zhekue.\\
    \gll baka abaksï zhe-kue\\
         cow black be.located.\textsc{pl-}exist\\
    \glt‘There are black cows.’
\z 

In some instances, the existential verb can be used on its own, e.g. when referencing shapeless objects (e.g. smoke) or objects whose position is not known. An existential construction may also involve solely a locative verb, in which case they are mainly distinguished from locative predicates on the basis of the order of Figure and Ground noun phrases.

When the Ground NP precedes the Figure NP (\ref{ex:kog:40}), only an existential reading is possible. Clauses with the order Figure – Ground are prototypical locative predications; however, they can also be interpreted as existential predications (\ref{ex:kog:41}). 

\ea\label{ex:kog:40}mesak kalta ipá.\\
  \ea \gll mesa-k kalta i{}-pá \\
         table-\textsc{loc} book \textsc{loc.appl-}lie \\
  \ex \gll mesa-k kalta i{}-pa-kue\\
         table-\textsc{loc} book \textsc{loc.appl-}lie\textsc{{}-}exist \\
    \glt ‘There is a book on the table.’ (*‘The book is on the table.’)
  \z 
\z 

\ea\label{ex:kog:41}kalta mesak ipá.\\
    \gll kalta mesa-k i{}-pá\\
         book table-\textsc{loc} \textsc{loc.appl-}lie\\
    \glt‘The book is on the table.’ or ‘There is a book on the table.’
\z 

Locative predicates prototypically have the structure [Figure – Ground – locative verb], as illustrated in \REF{ex:kog:41}, and \REF{ex:kog:42} below. While the existential verb can be used in locative predication instead of a locative verb in some instances \REF{ex:kog:42}, the compound verb, such as the one in \REF{ex:kog:39}, does not occur in these constructions. 

\ea\label{ex:kog:42}haggi canastali akkué.\\
    \gll haggi canasta{}-li ak{}-kué\\
         stone basket-\textsc{loc} \textsc{ine{}-}exist\\
    \glt‘The stone is in the basket.’
\z 

Turning to negation, both existentials and locatives have the same structure, namely imperfective or habitual negation is marked on the existential verb and cannot attach directly to a locative verb. In the case of locative predication, the presence of \textit{kual {\textasciitilde} kue} is a general morphosyntactic requirement for the use of inflectional markers such as tense and also negation, as shown in (\ref{ex:kog:43}--\ref{ex:kog:45}).

\ea\label{ex:kog:43}baka atsïtshi zhekualagale.\\
    \gll baka atsïtshi zhe{}-kual{}-a{}-\textbf{gale}\\
          cow red be.located.\textsc{pl-}exist-\textsc{st-neg.hab}\\
    \glt ‘There are no red cows.’
\z 

\ea\label{ex:kog:44} mesak kalta ipakualakí.\\
    \gll mesa{}-k kalta i{}-pa{}-kual{}-a{}-\textbf{kí}\\
         table-\textsc{loc} book \textsc{loc.appl-}lie\textsc{{}-}exist-\textsc{st-neg.ipfv}\\
    \glt‘There is no book on the table.’
\z 

\newpage
\ea\label{ex:kog:45}kalta mesak ipakualakí.\\
    \ea[]{\gll kalta mesa{}-k i{}-pa{}-kual-a{}-\textbf{kí} \\
         book table-\textsc{loc} \textsc{loc.appl-}lie\textsc{{}-}exist-\textsc{st-neg.ipfv}\\}
    \ex[*]{\gll kalta mesa{}-k i-pa{}-kí\\
         book table-\textsc{loc} \textsc{loc.appl-}lie\textsc{-neg.ipfv}\\
        \glt‘The book is not on the table.’}
    \z 
\z 

In possessive predications, exemplified in (\ref{ex:kog:46}--\ref{ex:kog:47}), either negation marker is attested, depending on the permanence of the situation.

\ea\label{ex:kog:46}
    \ea\label{ex:kog:46a}puelta nakalé.\\
        \gll  puelta nak-a-nal\\ 
              money \textsc{1sg.III{}-poss-}be\\
        \glt ‘I have some money.’
    \ex\label{ex:kog:46b}puelta nakalalakí.\\
        \gll  puelta nak-a-nal{}-a-\textbf{kí}\\
               money \textsc{1sg.III{}-poss-}be{}-\textsc{st}{}-\textsc{neg.ipfv}\\
        \glt ‘I don't have any money (on me right now).’
    \z
\z 

\ea\label{ex:kog:47}
    \ea\label{ex:kog:47a}nalú nakalé.\\
        \gll  na-nu nak-a-nal\\ 
              \textsc{1sg.inal{}-}elder.sister \textsc{1sg.III{}-poss-}be\\
        \glt ‘I have an elder sister.’
    \ex\label{ex:kog:47b}nalú nakalalagale.\\
        \gll  na-nu nak-a-nal{}-a-\textbf{gale}\\
              \textsc{1sg.inal{}-}elder.sister \textsc{1sg.III{}-poss-}be\textsc{{}-st-neg.hab}\\
        \glt ‘I don't have an elder sister.’
    \z
\z 

An alternative strategy for negation features the negative existential verb \textit{sïng(ze).} These constructions are equal in meaning to the negative locative/existential and possessive clauses discussed above. Both forms \textit{sïng} and \textit{sïngze} can be used interchangeably. As illustrated with the existential constructions in \REF{ex:kog:48}, the negative verb replaces the existential verb stem. In \citegen{Miestamo2005} definition, this would thus be an asymmetric construction.

\ea\label{ex:kog:48}
    \ea\label{ex:kog:48a}puttili ni akkué.\\
        \gll  putti-li ni ak-kué\\
              pot-\textsc{ine} water \textsc{ine-}exist \\
        \glt  ‘There is water in the pot.’
    \ex\label{ex:kog:48b}puttili ni aksïngze.\\
        \gll  putti-li ni ak-\textbf{sïngze}\\
               pot-\textsc{ine} water \textsc{ine-}\textsc{neg.ex}\\
        \glt ‘There is no water in the pot.’
    \z
\z 

In locative constructions, \textit{sïng(ze)} is only found in the negation of clauses with the existential verb \textit{kual {\textasciitilde} kue}, see example \REF{ex:kog:42}, or locative verbs containing the root \textit{te,} as in example \REF{ex:kog:49}. The locative verb \textit{te} is typically used for three-dimensional objects or animates in a sitting position.\footnote{In addition to the notion of sitting, \textit{te} also occurs in \textit{akté} ‘stand (upright)’ (negated: \textit{aksïng}) and \textit{agaté} ‘lean on’ (negated: \textit{agasïng}). With persons, \textit{te} can also denote ‘be located’ without information about posture (e.g. in ‘He is in the village.’).}

\ea\label{ex:kog:49}
    \ea\label{ex:kog:49a}ẽ kalokák ité.\\
        \gll  ẽ kaloká-k i-té\\ 
              \textsc{dem} chair-\textsc{loc} \textsc{loc.appl-}sit \\
        \glt ‘He is sitting on the chair.’
    \ex\label{ex:kog:49b}ẽ kalokák isḯngze.\\
        \gll  ẽ kaloká-k i-\textbf{sḯngze}\\
              \textsc{dem} chair-\textsc{loc} \textsc{loc.appl-}\textsc{neg.ex} \\
        \glt ‘He is not sitting on the chair.’
    \z
\z 

Clauses that involve other locative verbs (e.g. \textit{pa} ‘lie (flat objects)’ or \textit{sha} ‘lie (long objects)’) are negated by adding the existential verb and a standard negator, as shown in \REF{ex:kog:50b}.

\ea\label{ex:kog:50}
    \ea\label{ex:kog:50a} plato mesak ipa.\\
        \gll  plato mesa-k i-pa\\ 
              plate table-\textsc{loc} \textsc{loc.appl-}lie\\
        \glt ‘The plate is on the table.’
    \ex\label{ex:kog:50b}plato mesak ipakualakí.\\
        \gll  plato mesa-k i-pa{}-kual-a-kí\\
              plate table-\textsc{loc} \textsc{loc.appl-}lie\textit{{}-}exist-\textsc{st-neg.ipfv}\\
        \glt ‘The plate is not on the table.’
    \z
\z 

Finally, the negative verb also occurs in predicative possession, where it is equal in meaning to verbs negated with major negation markers. That is, the affirmative clause in \REF{ex:kog:46a} may also be negated as shown in \REF{ex:kog:51}. In these contexts, \textit{sïng(ze)} is neutral in terms of aspectual meaning, i.e. the situation presented in \REF{ex:kog:51} may be interpreted both as temporary (present/imperfective) or as more permanent (habitual). As noted already, the negative existential \textit{sïng(ze)} is part of the SN construction as well. Generally, its uses and polysemy patterns follow well-known cross-linguistic tendencies to encode absence and unavailability, see \citet{Veselinova2013neg.exist}.

\ea\label{ex:kog:51} puelta nisḯng.\\
    \gll  puelta na-i-\textbf{sïng}\\ 
          money \textsc{1sg.II{}-loc.appl-neg.ex}\\
    \glt ‘I don't have any money (on me right now).’
\z 

 
\subsection{Negation in dependent clauses}\label{sec:kog:2.4}

In general, dependent clauses (i.e. relative, complement and most adverbial clauses) feature the same negation markers as found in standard negation. Exceptions are negative purpose clauses and negative circumstantial clauses, as exemplified in this section.

The use of major negation markers in dependent clauses is illustrated here by adverbial clauses, which are introduced by a subordinating verbal affix expressing the temporal or causal relation between events. In the temporal adverbial clause (in brackets) in \REF{ex:kog:52}, the negation marker \textit{{}-kí} attaches to the lexical verb \textit{tũ} ‘watch’ in a periphrastic construction. The affirmative counterpart of the adverbial clause, i.e. \REF{ex:kog:52b}, can be synthetic or periphrastic.

\ea\label{ex:kog:52}
    \ea\label{ex:kog:52a}[tũakí nokenki] ezua sukkua bicicletak nak.\\
        \gll  tũ{}-a-\textbf{kí} nok-éng-ki ezua sukkua bicicleta-k nak\\ 
              watch\textsc{{}-st-neg.ipfv} \textsc{aux.ipfv}{}-\textsc{sim{}-top} \textsc{indf} boy bicycle-\textsc{loc} arrive\\
        \glt  ‘While he isn't watching, a boy on a bicycle arrives.’
    \ex\label{ex:kog:52b}tũĩ  nokengki / tũatokéngki sukkua nak.\\
        \gll   tũ{}-ĩ  nok{}-eng-ki / tũ{}-a{}-tok{}-éng-ki sukkua nak\\
              watch{}-\textsc{ipfv}\textsc{\textsubscript{2}} be-\textsc{sim-top} / watch\textsc{{}-st-prog{}-sim{}-top} boy arrive\\
        \glt ‘While he is watching, the boy arrives.’ 
    \z
\z 

While there are dedicated subordinating morphemes that denote simultaneity and succession, the notion of anteriority (‘before’) is expressed with a simultaneous adverbial clause involving the phasal polarity marker \textit{{}-g\'{a}nga} ‘not yet’, as in \REF{ex:kog:53}:
 
 \newpage
\ea\label{ex:kog:53} [huk nagaganga noguéng] níkala zekalá.\\
    \gll   hu-k nak-a-gánga nok-ngu-éng níkala zek-a-ná\\ 
          house-\textsc{loc} arrive\textsc{{}-st-}not.yet \textsc{aux.ipfv}{}-\textsc{pst{}-sim} rain happen-\textsc{st-incep}\\
    \glt  ‘It started raining before s/he arrived at home.’ (lit. ‘S/he still hadn't arrived at home…')
\z 

Note that constructional (a)symmetries in dependent clauses that take major negation markers are determined by the negative construction, as introduced in Sections \sectref{sec:kog:2.1.1} to \sectref{sec:kog:2.1.3} above (e.g. imperfective constructions with \textit{{}-\~{i} } as in \REF{ex:kog:52} are asymmetrical). 

There are two types of purpose clauses. Purpose of motion (e.g. ‘I'm going to the river in order to do laundry.’) is expressed by a non-finite verb marked with -\textit{l.} These non-finite predicates cannot be negated. Other purpose clauses involve an unmarked lexical verb in combination with an auxiliary marked with -\textit{amak}, such as the one in \REF{ex:kog:54}.

\ea\label{ex:kog:54} nahí sukkuakũẽ colegioli nẽ akualiamak hiba atshikuge.\\
    \gll nahí sukkua-kũẽ colegio-ni nẽ a-gu-a-lí-amak hiba atshi-ka-uge \\
          \textsc{1sg.poss} child{}-\textsc{pl} school-\textsc{all} go \textsc{3pl.I{}-aux.pfv-st{}-fut-purp}  work do-\textsc{hab-1sg.I}\\
    \glt ‘I work so that my children can go to school.’
\z 

In the negative counterpart of finite purpose clauses, the uninflectable particle \textit{guakua} is used in combination with an imperfective form of the lexical verb. Even though \textit{guakua} possibly includes the verb \textit{gu} ‘do’, it is considered a particle as it does not allow for any person or tense/aspect marking. Given the different forms of the lexical verb and the use of a particle instead of an auxiliary, the construction is clearly asymmetrical, as exemplified in \REF{ex:kog:55}.

\ea\label{ex:kog:55}
    \ea\label{ex:kog:55a}hugaké akiónuge shentá  hulunĩ  guakua.\\
        \gll  hugaké akió\#ne-uge shentá  huldun-ĩ \textbf{guakua}\\ 
              door close\#\textsc{pst-1sg.I} chicken enter-\textsc{ipfv}\textsc{\textsubscript{2}} \textsc{neg.purp}\\
        \glt  ‘I closed the door so that the chickens would not enter.’
    \ex\label{ex:kog:55b}séngaba nẽgua, mapenĩ guakua.\\
        \gll  séngaba nẽ{}-gua ma{}-pen-ĩ \textbf{guakua}\\
              slowly walk-\textsc{imp.sg} \textsc{2sg.II{}-}fall-\textsc{ipfv}\textsc{\textsubscript{2}} \textsc{neg.purp}\\
        \glt  ‘Walk slowly, so you don't fall.’
    \z
\z 

The negation marker \textit{{}-gabba} that is found in prohibitive and negative optative mood is also used in negative circumstantial clauses, see (\ref{ex:kog:56}--\ref{ex:kog:57}). These clauses express that an event takes place without (or instead of) the occurrence of another event.

\protectedex{\ea\label{ex:kog:56} stugagabba zuketa.\\
    \gll  zï{}-tuk{}-a-\textbf{gabba} zuketa\\ 
           \textsc{antip{}-}drink\textsc{{}-st-neg} pass.by\\
    \glt ‘He goes by without drinking.’
\z }

\protectedex{\ea\label{ex:kog:57}eka zïligagabbaki hiúngulak me nẽ kakungueniki hoklék. \\
    \gll   eka zï-nik{}-a-\textbf{gabba}-ki hiungula-k me nẽ ka-gu-ngu-eni-ki hokldék\\
          \textsc{dem} \textsc{antip{}-}sow\textsc{{}-st-neg-top} road-\textsc{loc} only go \textsc{1pl.I{}-aux.pfv}{}-\textsc{rem.pst-seq-top} play  \\
    \glt ‘Instead of sowing there [on the field], we just went up to the road and played.’ 
\z }

Affirmative equivalents of examples \REF{ex:kog:56} and \REF{ex:kog:57} would imply the realization of two successive events which are expressed by temporal adverbial clauses similar to the one in \REF{ex:kog:45}. As such clauses differ from the ones in \REF{ex:kog:56} and \REF{ex:kog:57} in several respects (e.g. in person indexing, which is not possible on the forms with \textit{{}-gabba}), the constructions can be regarded as asymmetrical. 

\section{Non-clausal negation}\label{sec:kog:3}

In contrast to clausal negation, negation in the non-clausal domain is considerably less complex in Kogi. In what follows, the properties of negative replies (\sectref{sec:kog:3.1}.) and negative indefinite pronouns (\sectref{sec:kog:3.2}.) are discussed. Finally, \sectref{sec:kog:3.3} an abessive construction with the meaning ‘without’.

\subsection{Negative replies}\label{sec:kog:3.1}

Polar questions, i.e. questions that seek either affirmation or denial, are answered by particles that convey the polarity of the intended answer. In this regard, the system can be classified as a yes/no system, which is the most widely attested type cross-linguistically (see \citealt{Moser2018}; \citealt{SadockZwicky1985}).

A polar question is answered affirmatively by \textit{akze} ‘yes’ or the interjection \textit{aha} [ʔaˈha], and negatively by \textit{no} ‘no’ (most likely a loan from Spanish) or interjections with falling intonation \textit{mh mh} [ˈʔ\'{m}ʔ\`{m} ]\textit{, ah ah} [ˈʔáʔà]. In addition, the Spanish loan \textit{nada} ‘nothing’ is employed as an emphatic negative particle with the meaning ‘no way/anything but’, see \REF{ex:kog:32} above. In both negative and affirmative answers, the particle or interjection can be used on its own, but more commonly it is complemented by a repetition of the questioned proposition or some other kind of statement, as in \REF{ex:kog:58}.

\protectedex{\ea\label{ex:kog:58}
    \begin{xlist}
    \exi{Q: }\label{ex:kog:58q}ekí sigí na mawatũne nagabẽne nala shina?\\
        \gll  ekí sigí na ma-a-tũ\#ne nak{}-a-bẽ\#ne nal{}-a shi-nal\\ 
               \textsc{dem} man with \textsc{2sg.II{}-3pl.I}{}-see\#\textsc{pst} \textsc{1sg.III{}-3pl.I}{}-tell\#\textsc{pst} be{}-\textsc{st} \textsc{addr.sym}{}-be\\ 
        \glt ‘They told me they saw you together with a man. Is that so?’
    \exi{A: }\label{ex:kog:58a} no shäbbial nuka nẽnuge nzha.\\
        \gll  \textbf{no} zï{}-habbi{}-a-l nuka nẽ\#ne-uge ni-nal\\ 
               no \textsc{antip}{}-buy-\textsc{st-purp} only go\#\textsc{pst{}-1sg.I} \textsc{spkr.sym}{}-be \\
        \glt ‘No, I just went to buy [something].’
    \end{xlist}
\z }

A negative reply negates the propositional content of the question, rather than its polarity. Thus, with the negative answer to the question in \REF{ex:kog:59}, speaker B negates the proposition (that he went) and thereby affirms speaker A’s assumption (that he didn't go).
% \lv{@LSP: example \REF{ex:kog:59} is split across two pages.}
\ea\label{ex:kog:59}
    \begin{xlist}
    \exi{Q: } José-María Santa Martali nẽyazháne?\label{ex:kog:59q}\\
        \gll  José-María Santa Marta-ldi nẽya-zhá\#ne\\ 
               José-María Santa Marta-\textsc{all} go-\textsc{neg\#pst}\\ 
        \glt  ‘Didn’t José-María go to Santa Marta?’
    \exi{A: } ahah, nẽyzháne naklá.\label{ex:kog:59a}\\
        \gll  \textbf{ahah}, nẽy{}-a{}-zhá\#ne nak{}-ná\\ 
               no go{}-\textsc{st{}-neg\#pst} \textsc{spkr.asym-}be \\
        \glt  ‘No, he didn't go.’
    \end{xlist}
\z 

There are no polarity-reversing particles and the polarity of a negative question as in \REF{ex:kog:53} is simply negated with an affirmative particle.

\subsection{Negative indefinites and quantifiers}\label{sec:kog:3.2}
\largerpage
Indefinite pronouns of the three ontological categories ‘person’, ‘thing’ and ‘time’, which occur in affirmative clauses, are presented in \tabref{tab:kog:6}. To express indefinite reference to persons, Kogi makes use of the numeral \textit{ezua} ‘one’. For things, either the noun \textit{hika} ‘thing’ or the contrastive interrogative pronoun \textit{maile} ‘which one’ is used (typically in contexts of free choice). With the exception of \textit{maile} ‘which one’, all of these forms can be used pronominally, as a phrasal head, and also as a nominal modifier. \textit{Maile}, in contrast, can only function as a modifier.

\begin{table}
\begin{tabularx}{\textwidth}{lQQQ}
    \lsptoprule
    category & indefinite & negative indefinite & interrogative \\
    \midrule
    person & \textit{ezua}\newline ‘someone, some’ & \textit{ni me / me hiega}\newline ‘no one’ & \textit{me}\newline ‘who’ \\
    \tablevspace
    thing & \textit{hika, maile}\newline ‘something, some’ & \textit{ni hi / hi hiega}\newline ‘nothing’ & {\textit{hi}\newline ‘what’} \\
    \tablevspace
    time & \textit{atshuák}\newline ‘sometime’ & \textit{ni mitsák / mitsák hiega}  ‘never’ & \textit{mitsák}\newline ‘when’ \\
    \lspbottomrule
\end{tabularx}
\caption{Indefinite and interrogative pronouns}
\label{tab:kog:6}
\end{table}


\largerpage
The bases of negative indefinite pronouns are interrogative pronouns, which are combined with either of two additive focus particles, namely \textit{hiega} ‘too, also’ or \textit{ni} ‘even’. The former occurs in affirmative contexts, while the latter is exclusive to negative contexts where it co-occurs with verbal negation.

While both variants of a negative indefinite pronoun, e.g. \textit{ni me} and \textit{me hiega} ‘no one’, are used interchangeably, there can be speaker-dependent preferences for either of the forms. Negative indefinite pronouns consist of two phonological words\textemdash forms with \textit{ni} do not undergo the word-internal, morphophonological process of \textit{m} → \textit{b} (e.g. *\textit{nibe}). As for forms with \textit{hiega}, each component retains its own stress.

As shown in (\ref{ex:kog:60}--\ref{ex:kog:61}), negative indefinite pronouns are always used together with verbal negation marking. In this respect, Kogi belongs to the cross-linguistically most widespread type where negative indefinites and clausal negation obligatorily co-occur \citep[202]{Haspelmath1997}. The non-negative set cannot occur in negative clauses.

\ea\label{ex:kog:60}
    \ea\label{ex:kog:60a}ezua nakle.\\
        \gll  ezua nak\#le\\ 
               \textsc{indf} arrive\#\textsc{.pst}\\
        \glt  ‘Someone arrived.’
\newpage
    \ex\label{ex:kog:60b}ni me nakazháne.\\
        \gll  ni me nak-a{}-\textbf{zhá}\#ne\\
              \textsc{add} who arrive\textsc{{}-st{}-neg\#pst}\\
        \glt ‘No one arrived.’
    \z
\z 

\ea\label{ex:kog:61} nahí gamá mani hiega nakzalazhá.\\
    \gll  nahí gamá mani hiega nak{}-zal{}-a{}-\textbf{zhá}\\ 
          1\textsc{sg.poss} bag where \textsc{add} \textsc{1sg.III{}-}be{}-\textsc{st{}-neg}\\
    \glt  ‘I didn't find my bag anywhere.’
\z 

As for quantifiers, the set of forms such as \textit{nuk} ‘all’, \textit{azbuáng} ‘some’ and \textit{nõwa} ‘few’ does not include dedicated negative counterparts. That is, the same forms occur in affirmative and negated clauses.

\subsection{Negative derivation and case marking}\label{sec:kog:3.3}

As for negative derivation or case marking, no strategies are attested. The notion of abessive is conveyed by verbal constructions involving major negation markers, as in (\ref{ex:kog:62}--\ref{ex:kog:63}).
\ea\label{ex:kog:62} sapatu nalakí  nẽkuge.\\
    \gll   sapatu nal{}-a-\textbf{kí} nẽ-ka-uge\\
           shoe be\textsc{{}-st{}-neg.ipfv} walk\textsc{{}-prs-1sg.I}\\
    \glt  ‘I'm walking without shoes.’ (lit. ‘There are no shoes, I am walking.')
\z 

\ea\label{ex:kog:63} Maríaki pulazhá zúkane.\\ 
    \gll María-ki pul{}-a-\textbf{zhá} zï{}-nuk{}-a\#ne\\
          María-\textsc{top} burn{}-\textsc{st{}-neg} \textsc{antip{}-}cook{}-\textsc{st}\#\textsc{pst}\\
    \glt ‘María cooked without getting burnt.’ (lit. ‘She cooked and didn't get burnt.')
\z 

\section{Other aspects of negation}\label{sec:kog:4}

This section  discusses some morphosyntactic, semantic and pragmatic phenomena that are connected to negation but do not constitute negative expressions per se. Note, however, that these areas need further research for a detailed description, as my corpus does not include data that account for every topic of the questionnaire by \textcitetv{chapters/questionnaire}.

\subsection{Scope of negation}\label{sec:kog:4.1}

In order to narrow the scope of negation on a specific constituent, the focus particle \textit{na} can be used in the same way as in affirmative clauses. This is exemplified in \REF{ex:kog:64}. The focused element is fronted and typically made prosodically more prominent by intensity or pitch.

 \ea\label{ex:kog:64}
    \ea\label{ex:kog:64a}Pedro na mebák nakne.\\
        \gll  Pedro na mebák nak\#ne\\ 
              Pedro \textsc{foc} yesterday arrive\#\textsc{pst}\\
        \glt ‘Pedro arrived yesterday.’
    \ex\label{ex:kog:64b} Pedro na mebák nakazháne.\\
        \gll  Pedro na mebák nak-a{}-zhá\#ne\\
              Pedro \textsc{foc} yesterday arrive{}-\textsc{st-neg}\#\textsc{pst}\\
        \glt‘Pedro didn’t arrive yesterday.’
    \ex\label{ex:kog:64c}mebák na Pedro nakazháne.\\
        \gll  mebák na Pedro nak-a{}-zhá\#ne\\
              yesterday \textsc{foc} Pedro arrive-\textsc{st-neg}\#\textsc{pst}\\
        \glt  ‘Pedro didn’t arrive yesterday.’
    \z
\z 

\subsection{Negative polarity}\label{sec:kog:4.2}

As for negative polarity items, with the exception of the additive focus particle \textit{ni} (see \sectref{sec:kog:3.2}), I have not encountered any expressions
that exclusively occur in negative contexts.

\subsection{Marking of noun phrases in the scope of negation}\label{sec:kog:4.3}

Case marking of noun phrases in negative clauses is the same as in affirmative clauses. With regard to determiners, the indefinite article \textit{ezua} (cf. \textit{ezua} ‘one’) is not used in negative clauses, as shown in \REF{ex:kog:65}. As noted in \sectref{sec:kog:3.2}., there are no special determiners that occur in negative clauses, rather the same ones as in affirmatives are used.

\ea\label{ex:kog:65}
    \ea\label{ex:kog:65a}Martahã ezua zhentá habbine.\\
        \gll  Marta{}-hã ezua zhentá habbi\#ne\\ 
             Marta-\textsc{erg} \textsc{indf} chicken buy\#\textsc{pst} \\
        \glt ‘Marta bought a chicken.’
    \ex\label{ex:kog:65b}Martahã zhentá habbiazháne.\\
        \gll  Marta-hã zhentá habbi-a-zhá\#ne\\
               Marta-\textsc{erg} chicken buy{}-\textsc{st-neg}\#\textsc{pst}\\
        \glt  ‘Marta didn’t buy any chicken.’
    \z
\z 

\subsection{Reinforcing negation}\label{sec:kog:4.4}

The verbal marker \textit{{}-shã} can be used to put emphasis on a statement. It occurs in affirmative as well as negated clauses, as illustrated in \REF{ex:kog:66}. Note that \textit{{}-shã} does not attach to the form inflected for tense and negation, but to the auxiliary \textit{nal} (‘be’), which is required for the use of the emphasis marker.

\ea\label{ex:kog:66}
    \ea\label{ex:kog:66a}keigaki hiba atshazhálikue.\\
        \gll  keiga{}-ki hiba atsh{}-a{}-zhá{}-likue\\ 
              today{}-\textsc{top} work do-\textsc{st-neg-fut:1sg.I}\\
        \glt ‘I won’t work today.’
    \ex\label{ex:kog:66b}keigaki hiba atshazhálikue nalashã.\\
        \gll  keiga{}-ki hiba atsh{}-a{}-zhá{}-likue nal-a-shã\\
              today{}-\textsc{top} work do-\textsc{st-neg-fut:1sg.I} \textsc{aux-st-emph}\\
        \glt ‘I won’t work today!’
    \z
\z 

\subsection{Negation, coordination and complex clauses}\label{sec:kog:4.5}

There are no dedicated negative coordinators. Rather, negative clauses each have negation marking on the respective predicate, as in the example in \REF{ex:kog:67}.

\ea\label{ex:kog:67} ahaba akkuezhá ahate akkuezhá.\\
    \gll  a{}-haba akkue{}-zhá a{}-hate akkue{}-zhá\\ 
           \textsc{3sg.inal-}mother remember-\textsc{neg} \textsc{3sg.inal-}father remember-\textsc{neg}\\
    \glt ‘He didn’t remember his mother nor his father.’
\z 

Note also that negation markers may not scope over more than one predicate. As shown in \REF{ex:kog:68}, each of the coordinated predicates is marked for negation.

\ea\label{ex:kog:68}  colegioli nẽyazhá hiba atshazhá gualíka.\\
    \gll  colegio{}-li nẽ{}-ya{}-zhá hiba atsh{}-a{}-zhá gu{}-a{}-líka\\ 
           school-\textsc{all} go-\textsc{st-neg} work do-\textsc{st-neg} \textsc{aux.pfv-st-fut}\\
    \glt  ‘S/he won’t go to school nor work.’
\z 

\section{Summary}\label{sec:kog:5}
Negation in Kogi shows particular complexity in the clausal domain. There is a set of major negation markers that are used in all declarative main clauses depending on the aspectual value. These also occur in stative predications, and most dependent and non-declarative clauses.
\tabref{tab:kog:7} summarizes the various attested negation markers as well as their contexts of use.

\begin{table}
\begin{tabularx}{\textwidth}{llQ}
    \lsptoprule
    \multicolumn{2}{l}{Marker} & {Environment}\\
    \midrule
    \textit{zhá} & \textsc{neg} & {Perfective aspect; past and future tense, non-modal and modal}\\
    \tablevspace
    \textit{kí}  & \textsc{neg.ipfv} & {Imperfective aspect; past, present and future tense; non-modal and modal}\\
    \tablevspace
    \textit{ne}  & \textsc{neg.hab} & {Habitual aspect; past, present and future tense; non-modal and modal}\\
    \tablevspace
    \textit{sïng(ze)}  & \textsc{neg.ex} & {Obligative, potential, volitive modality}\\
    \tablevspace
    \textit{sïng(ze)}  & \textsc{neg.ex} & {Abilitive modality, locative clauses, predicative possession}\\
    \tablevspace
    \textit{sïng(ze)}  & \textsc{neg.ex} & {Existential clauses}\\
    \tablevspace
    \textit{gabba} & \textsc{neg} & {Prohibitive, optative, circumstantial clauses}\\
    \tablevspace
    \textit{gánga}  & {‘not yet’} & {Perfective or imperfective aspect; past, present and future tense}\\
    \tablevspace
    \textit{ksá}  & {‘no longer’} & {Imperfective aspect; past, present and future tense}\\
    \lspbottomrule
\end{tabularx}
\caption{Summary of negation markers}
\label{tab:kog:7}
\end{table}

\begin{table}[p]
\footnotesize
\caption{Summary of constructional and paradigmatic (a)symmetries}
\label{tab:kog:8}

% \subtable[First part]{\label{tab:kog:8a}
\begin{tabularx}{\textwidth}{lQQ}
    \lsptoprule
    \textbf{Types of clauses} & \textbf{{Construction}} & \textbf{Paradigm}\\
    \midrule
    \textbf{Past \textit{ne} (\textsc{pfv)}} & {sym} & {sym} \\
    \textbf{Past (\textsc{pfv)}} &  & \\
    \tabitem{synthetic} & {sym} & {No negated form}\\
    \tabitem{periphrastic} & {sym} & {sym} \\

    \textbf{Future (\textsc{pfv)}} & {sym} &  {sym} \\

    \textbf{Imperfective \textit{{}-i, -té}} & {Loss of \textsc{ipfv} marker} & {Loss of \textsc{ipfv} distinctions}\\

    \textbf{Imperfective \textit{{}-tok}} & {Loss of person/tense marker} & {Loss of \textsc{ipfv} distinctions}\\

    \textbf{Present/habitual \textit{{}-ka}}\\
    \tabitem{habitual context} & {Loss of \textsc{prs/hab} marker, change of person indexes} & {sym}\\
    \tabitem{present context} & {Loss of \textsc{prs/hab} marker} & {sym}\\

    \textbf{Obligative} & & \\
    \tabitem{major} \textsc{neg} & {sym} & {sym}\\
    \tabitem{\itshape {}-sïng(ze)} & {Loss of \textsc{oblig} marker} & {Loss of \textsc{mod} distinctions}\\

    \textbf{Volitive} & & \\
    \tabitem{major \textsc{neg}} & {sym} & {sym}\\
    \tabitem{\textit{-sïng(ze)}} & {Loss of \textsc{vol} marker} & {Loss of \textsc{mod} distinctions}\\

    \textbf{Potential} & & \\
    \tabitem{major \textsc{neg}}  & {sym} & {sym} \\
    \tabitem{Loss of \textsc{pot} and person marker} & {Loss of \textsc{mod} distinctions}\\
%     \lspbottomrule
% \end{tabularx}
% }
% \end{table}

% \pagebreak
% \begin{table}\footnotesize
% \ContinuedFloat
% \subtable[Second part]{\label{tab:kog:8b}
% \begin{tabularx}{\textwidth}{llX}
%     \lsptoprule
%     \textbf{{Types of clauses}} & \textbf{{Construction}} & \textbf{{Paradigm}}\\
%
    \textbf{Abilitive} &&\\
    \tabitem{major \textsc{neg}} & {sym} & {sym}\\
    \tabitem{\textit{sïng(ze)}}  & {Replacement of stem} & {sym}\\

    \textbf{Optative} &&\\
    \tabitem{non-second person}  & {sym} & {Additional person distinction (2 / non-2)}\\
    \tabitem{second person}  & {Loss of \textsc{opt} marker} & Additional person distinction (2 / non-2) \\

    \textbf{Imperative} & {Loss of \textsc{imp} marker} & {Loss of \textsc{num} distinction}\\

    \textbf{Interrogative} & {sym} & {sym}\\

    \textbf{Proper inclusion, equation} & {sym} & {sym}\\

    \textbf{Indep. Adjectives} & {sym} & {sym}\\

    \textbf{Dep. Adjectives} & {sym} & {sym}\\

    \textbf{Existential} &&\\
    \tabitem{{major \textsc{neg}}}  & {sym} & {sym}\\
    \tabitem{\textit{sïng(ze)}}  & {Replacement of stem} & {sym}\\

    \textbf{Locative} &&\\
    \tabitem{{major \textsc{neg}}} & {sym} & {sym}\\
    \tabitem{\textit{sïng(ze)}} & {Replacement of stem} & {Restricted to \textit{te} ‘sit’}\\

    \textbf{Possessive} &&\\
    \tabitem{major \textsc{neg}} & {sym} & {sym}\\
    \tabitem{\textit{sïng(ze)}} & {Replacement of stem} & {sym}\\
    \lspbottomrule
\end{tabularx}
% }
\end{table}



\largerpage[-4]
In \tabref{tab:kog:8}, the constructions discussed in the course of the paper are summarized and their categorization in terms of constructional and paradigmatic symmetry is given. For each asymmetrical construction or paradigm, the differences between the affirmative and negative domains are indicated.\footnote{Note
    that the table does not include non-main clauses. This is due to two reasons. Firstly, in most dependent clauses (e.g. temporal adverbial), the relation in terms of symmetry depends on the type of negation marker in each case. That is, clauses with perfective aspect take \textit{{}-zhá} which is a symmetric construction, while imperfective adverbial clauses are negated with \textit{{}-kí}, which typically implies asymmetry. Secondly, negative purpose and circumstantial clauses cannot explicitly be categorized due to the lack of relevant examples in the data corpus. While it was noted that the constructions can be considered asymmetrical, further research is required for a categorization in terms of paradigmatic symmetry.
} The   most common  asymmetry in constructional terms concerns the loss of one marker when a negative marker is used. Negation may further induce a change in person marking, either an alternation of the position of indexes or the loss thereof.

Paradigmatic asymmetry often involves a reduction of distinctions, more precisely with regard to aspect (i.e. imperfective), number or modality (in negation with \textit{{}-sïng(ze)}).


\section*{Abbreviations}
\begin{tabularx}{.5\textwidth}{lQ}
\# & anticlitic boundary \\
1 & first person \\
2 & second person \\
3 & third person \\
I & set I argument marker (A) \\
II & set II argument marker (P) \\
III & set III argument marker (G) \\
\scshape add & additive focus \\
\scshape addr & addressee epistemic authority \\
\scshape all & allative \\
\scshape antip & antipassive \\
\scshape appl & applicative \\
\scshape aux & auxiliary \\
\scshape asym & asymmetrical epistemic access \\
\scshape caus & causative \\
\scshape cop & copula \\
\scshape dem & demonstrative \\
\scshape emph & emphatic \\
\scshape erg & ergative \\
\scshape ex & existential \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape foc & focus \\
\scshape fut & future \\
\scshape hab & habitual \\
\scshape hod & hodiernal \\
\scshape imp & imperative \\
\scshape inal & inalienable possession \\
\scshape incep & inceptive \\
\scshape indf & indefinite article \\
\scshape ine & inessive \\
\scshape ipfv & imperfective \\
\scshape loc & locative \\
\scshape neg & negative \\
\scshape oblig & obligative modality \\
\scshape opt & optative \\
\scshape pfv & perfective \\
\scshape pl & plural \\
\scshape poss & possessive \\
\scshape pot & potential \\
\scshape prog & progressive \\
\scshape proh & prohibitive \\
\scshape prs & present \\
\scshape pst & past \\
\end{tabularx}

\begin{tabularx}{.45\textwidth}{lQ}
\scshape purp & purposive \\
\scshape q & question \\
\scshape rec & recent \\
\scshape rem & remote \\
\scshape seq & temporal sequence \\
\scshape sg & singular \\
\scshape sim & simultaneous \\
\\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape spkr & speaker epistemic authority \\
\scshape st & stem-forming vowel \\
\scshape sub & subordinate clause \\
\scshape sym & symmetrical epistemic access \\
\scshape top & topic \\
\scshape vol & volitional \\
\end{tabularx}\vspace*{-5mm}



% \hc{@LSP please take a look at this table, it should be only one table split on two pages because there is too much}





\printbibliography[heading=subbibliography,notkeyword=this]
\end{document}
