\documentclass[output=paper,colorlinks,citecolor=brown,draftmode]{langscibook}
\ChapterDOI{10.5281/zenodo.18233758}
\author{Jonathan Moodie\affiliation{University of Melbourne}}
\title{Negation in Lopit}
\abstract{Negation in Lopit is mostly constructed with the negative verb /ɲa/ ‘not be’ in a complex clausal construction. This verb is often used in the form /ɪɲa/, with the same form for first, second and third person. However, when used with modal verbs and in subordinated clauses, it has the pronominal prefixes normally used in the language. The form /ɪɲa/ appears to be a fused or lexicalised form of the verb /ɲa/, as is the form /xɔɲa/, which is used in coordinated constructions.
Most negative constructions comprise the negative verb /ɲa/ and the lexical verb, which is marked with the subordinating prefix and maintains its person marking. The negative verb is clause-initial, and the lexical verb is placed after the subject to give a \textsc{neg} SVO word order. It is likely that the negative construction originated as a bi-clausal construction.
Nearly all negation involves complex clausal constructions. Even when verbless predications are negated, the negative verb is mostly used with a copula verb. The main exception is with non-locative existential predications, which are negated using the negative verb /ɪɲa/ only, with no copula verb.
Negative imperative constructions use the verb /idek/, meaning ‘avoid’ or ‘leave’, in combination with a special prefix /xai-/ on the lexical verb. Negative constructions involving an ongoing or persistive activity use the verb, /ɲei/, which is an alternate form of /ɲa/.

\keywords{verbal negation, Nilotic, stative predications, imperatives, interrogatives}}

\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 
\examplesitalics
\section{The language}

Lopit (ISO: \href{https://iso639-3.sil.org/code/lpx}{lpx}, Glottocode: \href{https://glottolog.org/resource/languoid/id/lopi1242}{lopi1242}) is an Eastern Nilotic language of South Sudan, spoken by around 50,000 people living in the Lopit Hills, north-east of Torit in Eastern Equatoria province. It is part of the Lotuxo sub-group of the Lotuxo-Maa languages. Until recently, the Lopit language has received little descriptive attention. Some observations on Lopit were made by authors working on the related Otuho (Lotuko) language \citep{Muratori1938} and in comparative wordlist data collected by \citet{Driberg1932} and \citet{Vossen1982}. Lopit has only been the focus of linguistic description and documentation in more recent years. The language has six different dialects (Ngaboli, Dorik, Ngutira, Lomiaha, Lohutok and Lolongo), and data collected with speakers of a number of these has led to observations on aspects of Lopit phonetics and phonology  \citep{Stirtz2014, Turner2001} and morphology and syntax (e.g. \citealt{LaduEtAl2014}). A major study of the phonetics and phonology of the Dorik dialect of Lopit has been carried out by \citet{Billington2017}.  A grammar of the Dorik dialect of Lopit has been compiled by \citet{MoodieBillington2020}. 


The data in this paper come from extensive fieldwork with Lopit speakers now living in Melbourne, Australia, and, to a lesser extent, living in the Kakuma Refugee Camp in Northern Kenya. The examples in this paper are taken from the transcriptions of the recorded sessions and are examples of the Dorik dialect.

In common with most other Eastern Nilotic languages, the basic constituent order is VSO.  Lopit has the feature of advanced tongue root (\textsc{atr}), which is common amongst Nilotic languages. It also has both lexical and grammatical tone. Tone is used, among other things, to mark the nominative case and to mark aspect. Lopit has two classes of verbs, one in which the root is consonant-initial (Class \textsc{i}) and one which has a root-initial close front vowel (Class \textsc{ii}). The verbs have a pronominal prefix which marks the number and person of the subject. This pattern is shown in \tabref{tab:lop:1}. The first, second and third person singular markers (/a-/, /ɪ-/, /ɛ-/) are widespread in Nilo-Saharan languages \citep[86]{Greenberg1966Africa}. Note that these are [–\textsc{atr}] in their underlying form but can be realised differently according to \textsc{atr} vowel harmony and mid-vowel assimilation processes. 

\begin{table}
    \begin{tabularx}{0.5\textwidth}{lXX}
    \lsptoprule
    { person} & { singular} & { plural} \\
    \midrule
    {1\textsuperscript{st}} & \textit{a-} & \textit{ɛɪ-, ei-} \\
    {2\textsuperscript{nd}} & \textit{ɪ-, i-} & \textit{ɪ-, i-} \\
    {3\textsuperscript{rd}} & \textit{ɛ-, e-, ɔ-, o-} & \textit{ɛ-, e-, ɔ-, o-} \\ 
    \lspbottomrule 
    \end{tabularx}
    \caption{Number/person marking for subjects}
    \label{tab:lop:1}
\end{table}

There is no grammatical tense in Lopit. Temporal reference is made with a small range of adverbs or is determined from the discourse context. There are many inflectional and derivational possibilities for the verb in Lopit. Verbs can be marked for person, number, aspect, modality, voice, directionality and the presence of causative and applicative derivations. In addition, infinitives, interrogatives and imperatives are marked on the verb. A possible template for Lopit verbal morphology is shown in \figref{fig:lop:2}.

%
% \begin{table} \footnotesize
% \begin{tabularx}{\textwidth}{ll|c|ll|ll|ll|l|ll}
% \lsptoprule
% \multicolumn{2}{c}{initial} & \multicolumn{1}{c}{pronominal} & \multicolumn{2}{c}{modal} & \multicolumn{2}{c}{aspectual} & \multicolumn{2}{c}{causative} & \multicolumn{1}{c}{root} & \multicolumn{2}{c}{suffixes}\\
% \multicolumn{2}{c}{prefixes} & \multicolumn{1}{c}{marker} & \multicolumn{2}{c}{prefixes} & \multicolumn{2}{c}{prefixes} & \multicolumn{2}{c}{prefix} & \multicolumn{1}{c}{} & \multicolumn{2}{c}{}\\
%
% \midrule
% \textit{x-} & \textsc{q} & \multirow{8}{*}{\parbox{11mm}{\centering person and \\ number of \\ subject (and object)}} & \textit{ŋai-} &  \textsc{irr} & \textit{(x)ɪ-} & \textsc{pfv} & \textit{ɪtɪ-} & \textsc{caus} && \textit{-a, -o} & \textsc{ipfv} \\
% \textit{x-} & \textsc{seq} &  &  \textit{ma-} & \textsc{pot} & \textit{ŋa-} & \textsc{pfv} &  & & & \textit{-ítà} & \textsc{cont} \\
% \textit{l-} & \textsc{sub} &  & \textit{mai-} & \textsc{cond} & \textit{lV-} & \textsc{persis} & & &  & \textit{-(V)k} & \textsc{dat/appl} \\
% \multicolumn{2}{c}{} &  & \textit{tV-} & \textsc{oblig} &  &  & &  &  & \textit{-ri, -ije} & \textsc{instr}\\
% \multicolumn{2}{c}{} &  &  &  &  &  &  &  &  & \textit{u, -ni} & \textsc{ven}\\
% \multicolumn{2}{c}{} &  &  &  &  &  &  &  &  & \textit{-ei, -ri} & \textsc{itv}\\
% \multicolumn{2}{c}{} &  &  &  &  &  &  &  &  & \textit{-a, -o} & \textsc{mid} \\
% \multicolumn{2}{c}{} &  &  &  &  &  &  &  &  & \textit{-a, -o} & \textsc{dtr}\\
% \lspbottomrule
% \end{tabularx}
%     \caption{A verb morphology template for Lopit}
%     \label{fig:lop:2}
% \end{table}


\begin{figure} \footnotesize
\begin{tabular}{l@{  }l | c | l@{  }l | l@{ }l | l@{ }l | c@{} | l @{ }l}
\multicolumn{2}{c|}{initial} & \multicolumn{1}{c|}{pronominal} & \multicolumn{2}{c|}{modal} & \multicolumn{2}{c|}{aspectual} & \multicolumn{2}{c|}{causative} & \multicolumn{1}{c|}{root} & \multicolumn{2}{c}{suffixes}\\
\multicolumn{2}{c|}{prefixes} & \multicolumn{1}{c|}{marker} & \multicolumn{2}{c|}{prefixes} & \multicolumn{2}{c|}{prefixes} & \multicolumn{2}{c|}{prefix} & & \\
\hline
\textit{x-} & \textsc{q} & {person and} & \textit{ŋai-} &  \textsc{irr} & \textit{(x)ɪ-} & \textsc{pfv} & \textit{ɪtɪ-} & \textsc{caus} && \textit{-a, -o} & \textsc{ipfv} \\
\textit{x-} & \textsc{seq} & number of &  \textit{ma-} & \textsc{pot} & \textit{ŋa-} & \textsc{pfv} &  & & & \textit{-ítà} & \textsc{cont} \\
\textit{l-} & \textsc{sub} & subject & \textit{mai-} & \textsc{cond} & \textit{lV-} & \textsc{persis} & & &  & \textit{-(V)k} & \textsc{dat/appl} \\
& & (and object) & \textit{tV-} & \textsc{oblig} &  &  & &  &  & \textit{-ri, -ije} & \textsc{instr}\\
\multicolumn{2}{c}{} &  &  &  &  &  &  &  &  & \textit{u, -ni} & \textsc{ven}\\
\multicolumn{2}{c}{} &  &  &  &  &  &  &  &  & \textit{-ei, -ri} & \textsc{itv}\\
\multicolumn{2}{c}{} &  &  &  &  &  &  &  &  & \textit{-a, -o} & \textsc{mid} \\
\multicolumn{2}{c}{} &  &  &  &  &  &  &  &  & \textit{-a, -o} & \textsc{dtr}\\
\end{tabular}
\caption{A verb morphology template for Lopit}
\label{fig:lop:2}
\end{figure}

In this paper, I use broad phonetic transcription in glossed examples and tables. Abbreviations used in glosses are given in the abbreviation list at the end of the chapter. The transcription of vowels in any example reflects the results of any harmony and assimilation processes present. The in-text examples are also given in broad phonetic transcription. Square brackets are used to indicate what was transcribed. Where words or stems are discussed in general terms rather than in an extract from a transcription, they are transcribed without tones, as the tones in any particular utterance depend on a number of factors such as case, aspect, and number. The description of negation in Lopit in this chapter is based on the questionnaire provided by the editors (\citetv{chapters/questionnaire}) and the structure of the chapter follows the model given in the questionnaire.

\section{Clausal negation} \label{sec:lop:2}
\subsection{Standard negation} \label{sec:lop:2.1}
\largerpage

A declarative verbal main clause is given in \REF{ex:lop:1}. Standard negation of this clause is shown in \REF{ex:lop:2}. The negative is expressed with the verb /ɲa/. The lexical verb is placed after the subject and is prefixed with the subordinate marker /l-/, ‘\textsc{sub}’.\footnote{The prefix /l-/ is used on verbs in a number of environments such as the conditional clause, the relative clause, negative constructions, contrastive focus constructions, and interrogatives. In some situations where the verb has the subordinate marker, it will be in a subordinate clause. This is why the term ‘subordinate marker' and the gloss ‘\textsc{sub}’ are used. In other situations, the subordinate marker is used in a clause which may not be subordinate.  However, in these cases, the lexical verb with the subordinate marker is not in the unmarked clause-initial position (see \citealt{Moodie2019}: 271, for further discussion).} The lexical verb [láwú] maintains its first person pronominal marking and the subject maintains its nominative case marking, as shown in \REF{ex:lop:1} and \REF{ex:lop:2}.

\ea\label{ex:lop:1}
    \gll á-wú náŋ à=tòrít \\ 
         1\textsc{sg}{}-go 1\textsc{sg.nom} to=Torit \\
    \glt ‘I’m going to Torit.’ (\textsc{mb1-341} 00:01:06)
\z 

\ea\label{ex:lop:2}
    \gll ɪɲà náŋ l-á-wú à=tòrít  \\
         not.be 1\textsc{sg.nom} \textsc{sub}{}-1\textsc{sg}{}-go to=Torit  \\
    \glt `I’m not going to Torit.' (lit. ‘I am not (that) I’m going to Torit.') (\textsc{mb1}-341 00:01:18)
\z

In the simplest form of standard negation, the negative form /ɪɲa/ does not follow the normal pronominal marking paradigm which is shown in \tabref{tab:lop:1}. Constructions such as example \REF{ex:lop:3}, with the normal first-person pronominal marking, are not possible. The word /ɪɲa/ is used for first, second and third person and I gloss it as ‘not.be’. It might be a lexicalised form, possibly coming from the second person form /ɪ-ɲa/. 
 
\ea[*]{\label{ex:lop:3}
    \gll a-ɲà náŋ l-á-wú à=tòrít \\ 
         1\textsc{sg}-not.be 1\textsc{sg.nom} \textsc{sub}{}-1\textsc{sg}{}-go to=Torit \\
    \glt Attempted: ‘I’m not going to Torit.’    (\textsc{mb1-015} 02:03:25)}
\z 

In some situations, the pronominal marking for /ɲa/ does follow the normal pronominal marking paradigm. This can happen when the negative verb is used in subordinate clauses, such as conditional or modal constructions (\sectref{sec:lop:2.3}) or in focussed constructions (\sectref{sec:lop:4.1}). Sometimes /ɔɲa/ is used in negative constructions instead of /ɪɲa/. This happens in interrogatives (\sectref{sec:lop:2.2.2}) and coordination constructions (\sectref{sec:lop:4.5}). The pronominal marker /ɔ/ is a third person marker.

The negative /ɪɲa/ can be regarded as a verb. The root is /ɲa/ and this verb behaves like other verbs in that it is clause initial; it is marked for person (sometimes); it takes the subordinate and interrogative markers; and it can be marked for (some types) of mode and aspect. 

The use of negative verbs is reasonably common in the world’s languages. In WALS Online, Dryer provides examples of languages with negative verbs (\citeyear{Dryer2013}). He describes them as “negative auxiliary verbs”, which he defines as negative words which occur with the “inflectional morphology associated with verbs in the language” \citep{Dryer2013}.

In \citegen{Miestamo2005} classification, standard negation in Lopit can be described as asymmetric with a finite negative verb. Miestamo states that the “negative verb can be an auxiliary or a higher verb taking a clausal complement”  and the “finiteness of the lexical verb is lost or at least reduced” (\citeyear[81]{Miestamo2005}). In most of the examples of auxiliary negative verbs given by \citet[81--87]{Miestamo2005}, the lexical verb does not have pronominal marking.  In Lopit, however, both the negative verb and the lexical verb are finite in that they both show pronominal marking. The only example given by Miestamo where both verbs are finite is an example from the Salishan language Halkomelem \citep[176, 185]{Galloway1993}, shown in \REF{ex:lop:added}. The main difference for this example is that the lexical verb is marked for irrealis (subjunctive).   

\ea\label{ex:lop:added} Halkomelem (ISO: \href{https://iso639-3.sil.org/code/hur}{hur}, Glottocode: \href{https://glottolog.org/resource/languoid/id/halk1245}{halk1245}, \citealt[176, 185]{Galloway1993} quoted by \citealt[83]{Miestamo2005}) \\
  \ea   \gll lɛ́m-cəl \\
            go-\textsc{1sg} \\
        \glt ‘I go.’ 
  \ex   \gll ʔə́wə-cəl lɛ́m-ɛ̀l \\
            \textsc{neg-1sg} go-\textsc{1sg.sbjn} \\
        \glt ‘I do/will not go.’ 
\z\z

Negative verbs have also been examined by Payne, who groups them into auxiliary negative verbs and higher negative verbs. In the first group, the negative verb “acts as a finite auxiliary to the lexical verb which in turn typically occurs in some non-finite form” \citep[207]{Payne1985}. The Lopit negative verb cannot be regarded as an auxiliary verb since the lexical verb maintains its pronominal marking, i.e. is finite. In Payne’s other group, the negative verbs take a full sentential complement. This is like Miestamo’s second classification of negative verbs, i.e.  “a higher verb taking a clausal complement” as presented above. The Lopit negative verb might belong to this group. For example, in \REF{ex:lop:2}, ‘I’m not going to Torit’ could be interpreted as something like ‘I am not that I am going to Torit’, where the expression [láwú à=tòrít] could be a complement clause. Note that the subject maintains its nominative case marking when the negative is formed. That means that one could not interpret the construction in \REF{ex:lop:2} as ‘It is not me that am going to Torit’. Nor could it be interpreted as ‘It’s not that I’m going to Lopit’.\footnote{If the two clauses were the equivalent of ‘It’s not’ and ‘I’m going to Lopit’, the second clause would be [nàŋ láwú à=tòrít] ‘1\textsc{sg.abs} \textsc{sub-1-}go to=Torit’. That is, the first person pronoun would be in the absolutive and not the nominative case, since there is “no case before the verb” in marked nominative languages \citep{König2006}.} In addition, the use of the first-person pronominal marking in the verb /lawu/ means that \REF{ex:lop:2} could not be interpreted as ‘I am not the one (who is) going to Torit’. 


This kind of construction also occurs with other words that are derived from auxiliary-like verbs. The word /lɔɟɔ/ (or /laɟɔ/) is often used to introduce an adverbial clause as in \REF{ex:lop:4}, where it can be translated as ‘when’. It is probably derived from the verb /ɟɔ/, ‘say’ and could be glossed as /l-a-ɟɔ/, (\textsc{sub-1sg}{}-say), in \REF{ex:lop:4}. It is verb-like in that it marks the subject of the subordinate clause. In addition, the word order differs from the normal \textsc{vso} order and could be described as \textsc{aux} \textsc{vso} word order. The lexical verb in the subordinate clause is prefixed with the subordinate marker and is placed after the subject. The corresponding independent clause would be [áwú náŋ à=tòrìt], ‘I go to Torit’. Thus, the word /lɔɟɔ/ (or /laɟɔ/) shows similar verb-like behaviour to the negative verb in terms of person marking, influence on word order and influence on the use of the subordinate marker on the lexical verb.

\ea\label{ex:lop:4}
    \gll {\ob}làjɔ́ náŋ l-á-wú à=tòrìt] á-bót náŋ à=bòlìs \\ 
         {\db}when.1\textsc{sg} 1\textsc{sg.nom} \textsc{sub-}1\textsc{sg}{}-go to=Torit 1\textsc{sg}{}-go.direct 1\textsc{sg.nom} to=police \\
    \glt ‘When I go to Torit, I will go straight to the police.’ (\textsc{mb1-160} \textsc{00}:14:20) 
\z 

In Lopit, both perfective and imperfective aspect are compatible with negation. Imperfective and perfective constructions with the verb /iwus/, ‘drink (hot li\-quids)’, are shown in \REF{ex:lop:5} and \REF{ex:lop:6} respectively. This is common amongst the world’s languages (\citealt{MiestamoAuwera2011}). 

\ea\label{ex:lop:5}
        \gll ɪ́ɲà náŋ l-a-íwús-ò ʃaɪ̂    \\ 
             not.be 1\textsc{sg.nom} \textsc{sub-1sg}{}-drink-\textsc{ipfv} tea.\textsc{abs}  \\
        \glt ‘I’m not drinking tea.’  (\textsc{mb1-269} \textsc{00}:32:18)
\z

\ea\label{ex:lop:6}
        \gll ɪ́ɲà náŋ l-a-îwús ʃaɪ̂     \\ 
             not.be 1\textsc{sg.nom} \textsc{sub-1sg}{}-\textsc{pfv.}drink tea.\textsc{abs} \\
        \glt ‘I didn't drink tea.’   (\textsc{mb1-269} \textsc{00}:32:50)
\z


In Lopit, there is a particular word, /ɔlɔxɔɲa/ or /ɛlɛxɪɲa/, which includes the negative verb /ɲa/. It can be translated as ‘not yet’. It is used in constructions which are similar to standard negation and is normally used in perfective constructions. The word could be glossed as /ɔ́-lɔ́-x-ɔ̀ɲà/ ‘3-\textsc{persis}{}-\textsc{seq}{}-3-not.be’, where the prefix /lɛ-/ marks the persistive aspect (see \sectref{sec:lop:2.5}). This would support its interpretation as ‘not yet’ or ‘still not’. Thus, the meaning is still somewhat compositional (\textsc{still} \textsc{not}), although it can be regarded as having a fused (or formally altered) construction. 

This word is illustrated in the following examples. Example \REF{ex:lop:7} is a question and \REF{ex:lop:8} is a possible answer. Note that the clause [ɔ̀lɔ̀xɔ́ɲà náŋ lát:ɔ́xɔ̀] has the same structure as standard negation (i.e. \textsc{neg} S V and the lexical verb is prefixed with the subordinate marker /l-/. 

\ea\label{ex:lop:7}
        \gll x-ɪ́-tːɔ̀xɔ́  íjé ɪ̀ɡɛ̀m \\
             \textsc{q}{}-2-finish.\textsc{pfv} 2\textsc{sg.nom} work.\textsc{abs} \\
        \glt ‘Have you finished?’    (\textsc{mb1-330} \textsc{00:}39:02)
\z

\ea\label{ex:lop:8}
        \gll ɪ́ɲà ɔ̀lɔ̀xɔ́ɲà náŋ l-á-tːɔ̀xɔ́ \\ 
             not.be {not.yet} \textsc{1sg.nom} \textsc{sub}{}-1\textsc{sg}{}-finish.\textsc{pfv} \\
        \glt ‘No, I haven’t finished yet.’ (\textsc{mb1-330} \textsc{00:}39:02)
\z

Example \REF{ex:lop:9} is a question which can be answered in two different ways. The answer in \REF{ex:lop:10} uses the /ɔlɔxɔɲa/, ‘not yet’, construction. The more general way of answering is given in \REF{ex:lop:11}.  

\ea\label{ex:lop:9}
        \gll x-ɪ̂-cá íjé dɛ̀=tòrít \\ 
             \textsc{q-2.pfv}{}-dance 2\textsc{sg.nom} at=Torit.\textsc{abs} \\
        \glt ‘Have you danced in Torit?’ or ‘Did you dance in Torit?’       (\textsc{mb1-142} 00:32:12)
\z

\ea\label{ex:lop:10}
        \gll ɔ́lɔ́xɔ̀ɲà náŋ l-â-cá dɛ̀=tòrít   \\ 
             not yet 1\textsc{sg.nom} \textsc{sub-1sg.pfv}{}-dance at=Torit.\textsc{abs} \\
        \glt ‘I have not yet danced in Torit.’       (\textsc{mb1-142} \textsc{00:}34:37)
\z

\ea\label{ex:lop:11}
        \gll ɪ́ɲà náŋ l-â-cá dɛ̀=tòrít   \\ 
             not.be 1\textsc{sg.nom} \textsc{sub-1sg.pfv}{}-dance at=Torit.\textsc{abs} \\
        \glt ‘I have not danced in Torit.’       (\textsc{mb1-142} \textsc{00:}32:36)
\z

This word is sometimes expressed as [ɛ̀lɛ́xɪ̀ɲà], as illustrated in \REF{ex:lop:12}. The word could be glossed as /ɛ̀-lɛ́-x-ɪ̀ɲà/ ‘3-\textsc{persis}{}-\textsc{seq}{}-not.be’. Both /ɔlɔxɔɲa/ and /ɛlɛxɪɲa/ are often shortened to /ɔlɔɲa/ and /ɛlɛɪɲa/ respectively. The elision of the /x/ prefix in Lopit is common.

\ea\label{ex:lop:12}
        \gll a-í-jːé-jːénà náŋ xɪ̀ɾò nà lòpít   ɛ̀lɛ́xɪ̀ɲà xàtí náŋ l-a-î-ijːén bìnó\\ 
             \textsc{1sg-ii-red}{}-know \textsc{1sg.nom} \textsc{vn.}speak of.\textsc{f} Lopit  not.yet however \textsc{1sg.nom} \textsc{sub-1sg-pfv-}know very \\
        \glt ‘I have been learning to speak Lopit, but I have not yet learned much.’ (\textsc{mb1-293} 00:13:08)
\z

\subsection{Negation in non-declaratives}
\label{sec:lop:2.2}
\subsubsection{Negative imperatives}
\label{sec:lop:2.2.1}
For Class I verbs, standard imperatives use the prefix /tɛ-/ or /tɔ-/ for singular imperatives and the form /ɪtɪ-/ for plural imperatives. The singular imperative of the Class I verb /bɔt/, ‘go direct’, is shown in \REF{ex:lop:13}.

\ea\label{ex:lop:13}
        \gll tɔ́-bɔ̀t a=xàŋ náítì \\ 
             \textsc{imp.sg}{}-go.direct to=home.\textsc{abs} my.\textsc{f.abs} \\
        \glt ‘Go straight to my place!’      (\textsc{mb1-098} \textsc{00}:41:41)
\z

For Class II verbs, the normal imperative is the same as the verb stem, except that the singular and plural forms have different tonal patterns. An example of the singular imperative with the verb /ɪɲak/, ‘repeat’, is shown in \REF{ex:lop:14}. 

\ea\label{ex:lop:14}
        \gll ɪ́ɲáx-à  róɾí nànà l-ɪ́-ɟɔ́ íjé   \\ 
             \textsc{imp.sg}.repeat-\textsc{ipfv} words.\textsc{abs} those.\textsc{f.pl} \textsc{sub}{}-2-say 2\textsc{sg.nom} \\
        \glt ‘Repeat those words you saidǃ’     (\textsc{mb1-321} \textsc{00:}17:32)
\z

There is a special construction for negative imperatives.  It involves the word /idek/ and a verb prefix /xaɪ-/. An example is given with the Class II verb /ɪɾɔ/, ‘speak’, in \REF{ex:lop:15}. The prefix /xaɪ-/ appears to be a kind of imperative prefix. Unlike standard imperatives, the negative imperative has no distinction between singular and plural. In addition, as shown in \REF{ex:lop:16}, the infinitive construction cannot be used.

\ea\label{ex:lop:15}
        \gll ídèk xaɪ-ɪ́ɾɔ̀  \\ 
             \textsc{neg.imp} \textsc{neg.imp}{}-speak \\
        \glt ‘Don’t talkǃ’       (\textsc{mb1-014} \textsc{00}:36:40)
\z

\ea[*]{\label{ex:lop:16}
        \gll ídèk  x-ɪ́ɾɔ̀ \\ 
             \textsc{neg.imp} \textsc{inf}{}-speak \\
        \glt Attempted: ‘Don’t talk!’    (\textsc{mb1-014} \textsc{00}:36:40)}
\z

The negative imperative with a Class I verb is illustrated with the verb /ca/ ‘dance' in \REF{ex:lop:17}.

\ea\label{ex:lop:17}
        \gll ídèk xaɪ̀-cá díxà   \\ 
             \textsc{neg.imp} \textsc{neg.imp-}dance like.that \\
        \glt `Don't dance like thatǃ'    (\textsc{mb1-319} \textsc{00}:45:36)
\z

A more complex example is shown in \REF{ex:lop:18} using the verb /isjere/, ‘give'. Note that the use of the second person pronoun (in this case /ɪtɛɪ/, \textsc{{}'2pl'}) is possible, but it is not normally used.

\ea\label{ex:lop:18}
        \gll ídèk (ɪ́tɛɪ́) xaɪ-ísò nàŋ xɪ́ɾɪ́ŋɔ̀ \\ 
             \textsc{neg.imp} (2\textsc{pl.nom}) \textsc{neg.imp}{}-give 1\textsc{sg.abs} meat.\textsc{abs} \\
        \glt `Don't give me meatǃ' (lit ‘Avoid give me meat.')  (\textsc{mb1-324} \textsc{00:}09:38)
\z

The word /idek/ can be glossed as the imperative form of the verb /idek/, ‘leave’ (i.e. as \textsc{imp}.leave).  It can be translated as ‘avoid', ‘leave off', ‘give up', ‘refrain from', as illustrated in \REF{ex:lop:19}.

\ea\label{ex:lop:19}
        \gll a-ídék náŋ rìɲà tìfì xɪ̀ɟɔ́ è-díxà nàŋ xɔ́ɲɛ́xítɛ̀ \\ 
             \textsc{1sg}{}-leave 1\textsc{sg.nom} \textsc{vn}.watch TV \textsc{comp} 3-ache 1\textsc{sg.abs} eyes \\
        \glt ‘I gave up watching TV because it hurts (my) eyes.’       (\textsc{mb1-020} 01:26:14)
\z

The verb /idek/ can also be used in the sense of ‘move away from’ as in \REF{ex:lop:20}.

\ea\label{ex:lop:20}
        \gll a-ídék náŋ xàŋ àrá móítéʔ \\ 
             1\textsc{sg}{}-leave 1\textsc{sg.nom} home.\textsc{abs} \textsc{imm.pst} morning \\
        \glt ‘I left home this morning.’       (\textsc{mb1-020} 01:28:48)
\z

This verb could be regarded as an auxiliary verb when it is used in negative constructions such as \REF{ex:lop:15}, \REF{ex:lop:17}, and \REF{ex:lop:18}. In these constructions, the lexical verb has a finite form (i.e. no pronominal marking).

\subsubsection{Negative interrogatives}
\label{sec:lop:2.2.2}
Negative polar interrogatives can be formed using the negative verb /ɲa/ together with the question marker /x-/. A pair of affirmative and negative polar questions is given in \REF{ex:lop:21} and \REF{ex:lop:22}. The word order with the negative question verb is similar to the word order in negative declarative clauses.  In \REF{ex:lop:22}, the subject [íjé], ‘you’, is placed in front of the lexical verb, which maintains its pronominal prefix and is also prefixed with the subordinate marker /l-/.  The word [xɔ́ɲà] is glossed as ‘\textsc{q}{}-not.be’. This is another generalized pronoun marker where /ɔ-/ is used instead of /ɪ-/, for the first, second or third person.

\ea\label{ex:lop:21}
        \gll x-ɪ́-t:ɔ́xɔ̀ íjé \\ 
             \textsc{q}{}-2-finish.\textsc{n} 2\textsc{sg.nom} \\
        \glt ‘Have you finished?’    (\textsc{mb1-330} \textsc{00:}39:02)
\z

\ea\label{ex:lop:22}
        \gll x-ɔ́ɲà  xàtí íjé l-ɪ́-t:ɔ́xɔ̀ \\ 
             \textsc{q}{}-not.be indeed 2\textsc{sg.nom} \textsc{sub}{}-2-finish \\
        \glt ‘Haven’t you finished (yet)?’   (\textsc{mb1-330} \textsc{00:}39:02)
\z

The word /xatɪ/, ‘but’, ‘however’, ‘indeed’, appears to be always used in negative interrogative constructions and is placed after /xɔɲa/. The word /xatɪ/ is also used in adversative coordination (\citealt{MoodieBillington2020}: 367), although it is not a conjunction. It could be described as an emphatic adverb.

Another example is given with the verb /wu/, ‘go’, in \REF{ex:lop:23}.

\ea\label{ex:lop:23}
        \gll x-ɔ́ɲà  xàtí l-í-wú à=xàŋ \\ 
             \textsc{q}{}-not.be indeed \textsc{sub}{}-2-go to=home.\textsc{abs} \\
        \glt ‘Aren’t you going home?’   (\textsc{mb1-330} \textsc{00:}41:00)
\z

Sometimes /ɪɲa/ is used instead of /xɔɲa/, as shown in \REF{ex:lop:24}. There are insufficient examples to know how common this is.

\ea\label{ex:lop:24}
        \gll ɪ́ɲà xàtí l-í-wú à=xàŋ \\ 
             not.be indeed \textsc{sub}{}-2-go to=home.\textsc{abs} \\
        \glt ‘Aren’t you going home?’ (\textsc{mb1-341} 00:01:53)
\z

The word /xɔɲa/ is also used in tag questions, together with the particle /ifa/, which is normally used to indicate that the event occurred in the past. The interrogative tag, [xɔ́ɲà ífá] is shown in \REF{ex:lop:25}. However, unlike English, the tag does not change for polarity \REF{ex:lop:26} or temporal situation \REF{ex:lop:27}. Given that the expression [xɔ́ɲà ífá] is used in relation to a future event (as in \REF{ex:lop:27}), it appears to be a lexicalised construction.

\ea\label{ex:lop:25}
        \gll ò-wú ɪ́ɲɛ́ à=sùk x-ɔ́ɲà  ífá \\ 
             3-go 3\textsc{sg.nom} to=market.\textsc{abs} \textsc{q}{}-not.be \textsc{pst} \\
        \glt ‘She went to the market, didn’t she?’      (\textsc{mb1-022} 01:04:12)
\z

\ea\label{ex:lop:26}
        \gll ɪ́ɲà ɪ́ɲɛ́ l-ò-wú à=sùk x-ɔ́ɲà  ífá \\ 
             not.be 3\textsc{sg.nom} \textsc{sub}{}-3-go to=market.\textsc{abs} \textsc{q}{}-not.be \textsc{pst} \\
        \glt ‘She didn’t go to the market, did she?' (\textsc{mb1-022} 01:04:26)
\z

\ea\label{ex:lop:27}
        \gll ò-wú ìsó ɪ́ɲɛ́ à=sùk x-ɔ́ɲà  ífá \\ 
             3-go \textsc{fut} 3\textsc{sg.nom} to=market.\textsc{abs} \textsc{q}{}-not.be \textsc{pst} \\
        \glt ‘She will go to the market, won’t she?' (\textsc{mb1-022} 01:04:26)
\z

\subsection{Negation in stative predications}
\label{sec:lop:2.3}
The morpheme /ɪɲa/ is used in all negation in stative predications. Most predications utilize the copula verbs /ɾa/, ‘be’, and /wːɔn/, ‘exist’. As will be shown in this section, negatives in stative predications are mostly of the same form as standard negation. The main exception is with non-locative existential predications where the only verb is the negative verb.

Example \REF{ex:lop:28} can be described as an equative predicate nominal (T. E. \citealt[114]{Payne1997}) or as a referential nominal predicate \citep[233]{Dryer2007}.  The copula verb /ɾa/ has a fused form in most situations.  The form /aɾa/ is used for first, second and third person in these situations. In this paper it is glossed as ‘be’. The other copula verb /wːɔn/ has normal pronominal marking, as shown in \REF{ex:lop:46}.

\ea\label{ex:lop:28}
        \gll áɾá xàb{\U} lɔ̀ɾɛ̀wâ lìtí \\
             be chief\textsc{.nom} husband\textsc{.abs} my.\textsc{m.abs} \\ 
        \glt ‘The chief is my husband.’       (\textsc{mb1-162} \textsc{00:}15:50)
\z

In some situations, the verb /ɾa/ has normal pronominal marking. This is illustrated in \REF{ex:lop:29}, where it is used with the conditional prefix /maɪ/. Thus, the verbs /ɾa/ and /ɪɲa/ are similar in that they both have fused forms in simple declarative clauses but can have normal pronominal marking in subordinate clauses.

\ea\label{ex:lop:29}
        \gll ɪ̀-maɪ̀-ɾá íjé xábʊ́ ɪ́-maɪ́-lwáxá-k íjé ìjòxoí \\ 
             2\textsc{sg}{}-\textsc{cond}{}-be \textsc{2sg.nom} chief\textsc{.abs} 2\textsc{sg}{}-\textsc{cond}{}-help-\textsc{dat} \textsc{2sg.nom} \textsc{1pl.abs} \\
        \glt ‘If you could be the chief, you would help us.’       (\textsc{mb1-010} \textsc{00:}24:00)
\z

The negation of \REF{ex:lop:28} is given in \REF{ex:lop:30}. The construction is similar to standard negation. The negative verb is clause-initial, and the subject is placed after the negative verb and before the copula. 

\ea\label{ex:lop:30}
        \gll ɪ́ɲà xàb{\U} l-áɾá lɔ̀ɾɛ̀wâ lìtí \\
             not.be chief\textsc{.nom} \textsc{sub-}be husband\textsc{.abs} my.\textsc{m.abs} \\
        \glt ‘The chief is not my husband.’  (\textsc{mb1-341} 00:27:52)
\z

Sometimes, this kind of stative predication is expressed without copula verbs. Demonstratives can be used, as shown in \REF{ex:lop:31}. When this verbless copula construction is negated, a copula verb, /ɾa/, ‘be’, is introduced with the negative verb, as shown in \REF{ex:lop:32}. Once again, this is similar to the standard negation construction.

\ea\label{ex:lop:31}
        \gll xábʊ́ ɪ̀lːɛ́ŋ \\ 
             chief\textsc{.abs} this.\textsc{m.nom} \\
        \glt ‘This is the chief.’       (\textsc{mb1-341} 00:26:38)
\z

\ea\label{ex:lop:32}
        \gll ɪ́ɲà ɪ̀lːɛ́ŋ l-áɾá xábʊ́  \\ 
             not.be this.\textsc{m.nom} \textsc{sub-}be chief\textsc{.abs} \\
        \glt ‘This is not the chief.’   (\textsc{mb1-341} 00:26:59)
\z

The following two utterances are examples of proper inclusion (\citealt[114]{Payne1997}) or non-referential nominal predicates, \citep[233]{Dryer2007}.  Example \REF{ex:lop:34} is the negation of \REF{ex:lop:33}. This construction is also similar to standard negation.

\ea\label{ex:lop:33}
        \gll áɾá ɪ́ɲɛ́ xaìtíjːénàní \\ 
             be 3\textsc{sg.nom} teacher.\textsc{abs} \\
        \glt ‘He/she is a teacher.’    (\textsc{mb1-329} \textsc{00}:17:06)
\z

\ea\label{ex:lop:34}
        \gll ɪ́ɲà ɪ́ɲɛ́ l-áɾá xaìtíjːénàní \\ 
             not.be 3\textsc{sg.nom} \textsc{sub}{}-be teacher.\textsc{abs} \\
        \glt ‘He/she is not a teacher.’    (\textsc{mb1-329} \textsc{00}:17:12)
\z

Attributive construction or property assignment predicates are usually expressed with a stative verb construction (or intransitive verbal predicate), as shown in \REF{ex:lop:35}. The same constructions are used for both permanent and temporary property assignment.

\ea\label{ex:lop:35}
        \gll ɛ̀-xálàn ɪ́ɲɛ́ bɪ́nɔ̀  \\ 
             3-be.lazy 3\textsc{sg.nom} very \\
        \glt ‘He is very lazy.’       (\textsc{mb1-128} \textsc{00:}37:11)
\z

This kind of expression is negated using standard negation, as shown in \REF{ex:lop:36}. 

\ea\label{ex:lop:36}
        \gll ɪ́ɲà náŋ l-á-xálàn  \\ 
             not.be 1\textsc{sg.nom} \textsc{sub-1sg}{}-be.lazy \\
        \glt ‘I am not lazy.’       (\textsc{mb1-128} \textsc{00:}36:05)
\z

Standard negation can also be used in locative predicate constructions with the copula verb, /wːɔn/, ‘exist’, as shown in \REF{ex:lop:37} and \REF{ex:lop:38}.

\ea\label{ex:lop:37}
        \gll ɔ̀-wːɔ́n mɔ́lɔ́ŋ dɛ̀=jànì   \\ 
             3-exist baboon\textsc{.nom} in=tree.\textsc{abs} \\
        \glt ‘The baboon is in the tree.’   (\textsc{mb1-162} \textsc{00:38}:14)
\z

\ea\label{ex:lop:38}
        \gll ɪ́ɲà mɔ́lɔ́ŋ l-ɔ̀-wːɔ́n dɛ̀=jànì  \\ 
             not.be baboon\textsc{.nom} \textsc{sub}{}-3-exist in=tree.\textsc{abs} \\
        \glt ‘The baboon is not in the tree.’   (\textsc{mb1-171} \textsc{00}:01:45)
\z

Existential predications can overlap with locative predications in some languages \citep[240]{Dryer2007}. However, in Lopit, there are some existential predicates with no overt attribution of a location.  An example of an existential predication is given in \REF{ex:lop:39}, where the copula verb /wːɔn/, ‘exist’, ‘be available’, is used with a dummy subject. The negative existential predication is given in \REF{ex:lop:40} using the negative verb /ɪɲa/ with no lexical or copula verb. Here, [ɪ́ɲà] can be translated as ‘there is not' or ‘there is no'. Negation of this kind of existential (perhaps described as ‘non-locative existential’) is different from standard negation. 

\ea\label{ex:lop:39}
        \gll ɔ̀-wːɔ́n xɪ̀sʊ́ŋ xʊ́ná l-ɔ́-bwɔ̀r  \\ 
             3-exist cows.\textsc{abs} of.\textsc{f.pl} \textsc{sub-3-}be.white \\
        \glt ‘There are white cows.’ (\textsc{mb1-341} 00:17:51)
\z

\ea\label{ex:lop:40}
        \gll ɪ́ɲà xɪ̀sʊ́ŋ xʊ́ná l-ɔ́-ɲóɾì  \\ 
             not.be cows.\textsc{abs} of.\textsc{f.pl} \textsc{sub-3-}be.green \\
        \glt ‘There are no green cows.’ (\textsc{mb1-341} 00:18:08)
\z

Another example of a non-locative existential is given in \REF{ex:lop:41}.  Once again, there is no copula verb in the negative construction. However, it is worth noting that, since the negative clause is subordinate, the negative verb has the third person pronominal prefix (referencing a dummy subject). This is discussed in \sectref{sec:lop:2.3}.

\ea\label{ex:lop:41}
        \gll l-ɛ́-ɲà dáxá é-jeí xìjò  \\ 
             \textsc{sub-3-}not.be eat.\textsc{vn} 3-die people.\textsc{nom} \\
        \glt ‘If there is no food, people die.’ (\textsc{mb1-341} 00:23:35)
\z

A pair of utterances demonstrating a positive and a negative existential with the same forms as \REF{ex:lop:39} and \REF{ex:lop:40} is shown in \REF{ex:lop:42} and \REF{ex:lop:43}.

\ea\label{ex:lop:42}
        \gll ɔ̀-wːɔ́n xìfjôŋ \\ 
             3-exist water.\textsc{abs} \\
        \glt ‘There is water.’ (\textsc{mb1-341} 00:21:57)
\z

\ea\label{ex:lop:43}
        \gll ɪ́ɲà xìfjôŋ \\ 
             not.be water.\textsc{abs} \\
        \glt ‘There is no water.’    (\textsc{mb1-256} \textsc{00}:07:49)
\z

A consultant was asked if one could say the utterance shown in \REF{ex:lop:44}. He said that it “will sound incomplete” (\textsc{mb1-256} \textsc{00}:22:49) and that an utterance like that in \REF{ex:lop:45} would be more acceptable. Example \REF{ex:lop:45} could be regarded as a locative existential construction. This confirms that standard negation is used in locative predicate constructions but not in non-locative existential constructions. 

\ea\label{ex:lop:44}
        \gll ɪ́ɲà xìfjôŋ l-ɔ̀-wːɔ́n   \\ 
             not.be water.\textsc{abs} \textsc{sub-3-}exist \\
        \glt ?‘There is no water.’ (\textsc{mb1-341} 00:22:29)
\z

\ea\label{ex:lop:45}
        \gll ɪ́ɲà xìfjôŋ l-ɔ̀-wːɔ́n ìní  \\ 
             not.be water.\textsc{abs} \textsc{sub-3-}exist here \\
        \glt ‘There is no water here.’ (\textsc{mb1-341} 00:23:03)
\z

Possessive predicate constructions can also be negated with the negative verb.  In Lopit, the comitative construction /wːɔn xɔ/, ‘exist with’, is often used to express possession, as shown in \REF{ex:lop:46}. The negative form is shown in \REF{ex:lop:47}.

\ea\label{ex:lop:46}
        \gll ɛɪ́-wːɔ́n íjòxoì xò=dòŋìʔ \\ 
             1\textsc{pl}{}-exist 1\textsc{pl.nom} with=drums.\textsc{abs} \\
        \glt ‘We have drums.' (lit. ‘We are with drums.’)    (\textsc{mb1-329} \textsc{00}:06:09)
\z

\ea\label{ex:lop:47}
        \gll ɪ́ɲà íjòxoì l-ɛɪ́-wːɔ́n xò=dòŋìʔ  \\ 
             not.be 1\textsc{pl.nom} \textsc{sub-1pl}{}-exist with=drums.\textsc{abs} \\
        \glt ‘We have no drums.’ (lit. ‘We are not with drums.’)   (\textsc{mb1-329} \textsc{00}:06:09)
\z


\subsection{Negation in non-main clauses}
\label{sec:lop:2.4}
Negation in non-main clauses shows some differences from the standard negation construction. Firstly, the range of pronominal inflections seen on most verbs is usually seen on the negative verb in a subordinate construction. In many subordinate clauses in Lopit, the initial verb is marked with the subordinate marker /l-/, glossed as ‘\textsc{sub’}. If this initial verb is the negative verb, then it will have person marking, unlike the negative verb in a simple declarative clause. This is illustrated in the contrast between [ɪ́ɲà] \REF{ex:lop:48} and [láɲà] in \REF{ex:lop:49}.

\ea\label{ex:lop:48}
        \gll ɪ́ɲà náŋ l-á-ɟíŋ-ú à=xàɟì      \\ 
             not.be 1\textsc{sg.nom} \textsc{sub-1sg}{}-go-\textsc{ven} to=house.\textsc{abs} \\
        \glt ‘I didn’t come into the house.’   (\textsc{mb1-253} \textsc{00:}36:11)
\z

\ea\label{ex:lop:49}
        \gll  á-ɡílò-k náŋ xɪɟɔ í-dáxá  ɪ́tɛɪ́ bì     l-á-ɲà náŋ l-á-ɟíŋ-ú à=xàɟì  \\
             1\textsc{sg}{}-think-\textsc{dat} 1\textsc{sg.nom} \textsc{comp} 2-eat.\textsc{n} 2\textsc{pl.nom} indeed \textsc{sub-1sg}{}-not.be 1\textsc{sg.nom} \textsc{sub-1sg}{}-go-\textsc{ven} to=house.\textsc{abs} \\
        \glt  ‘I thought you were eating, so I didn’t come into the house.’   (\textsc{mb1-253} \textsc{00}:34:07)
\z

In addition, the negative verb can also be marked for modality. An example is given in the conditional clause in \REF{ex:lop:50} (i.e. [áŋaîɲà náŋ lámweí] ‘if I were not sick’). The negative verb [áŋaîɲà] has pronominal marking and the main verb (/mwei/, ‘be sick’,) retains its pronominal marking.

\ea\label{ex:lop:50}
        \gll [á-ŋaî-ɲà náŋ l-á-mweí]\textsubscript{\textsc{p}}     [á-ŋaî-ibóŋ náŋ xò=xìjò x{\U}ná l-ɔ̀-lʊ́ŋà]\textsubscript{\textsc{q}} \\
             1\textsc{sg-irr}{}-not.be 1\textsc{sg.nom} \textsc{sub-1sg}{}-be.sick 1\textsc{sg-irr}{}-meet \textsc{1sg.nom} with=people.\textsc{abs} \textsc{rel.pl} \textsc{sub-3pl}{}-be.many\\
        \glt ‘Were I not sick (if I weren’t sick), I would have met many people.’    (\textsc{mb1-014} \textsc{0}1:25:40)
\z

An example with first-person plural marking is given in \REF{ex:lop:51}. It should be noted that the 1\textsc{pl} pronominal marker /ɛɪ-/ is realised as [ɛ] before the prefixes /ŋa-/ and /ŋai-/ as shown in the negative modal construction [ɛ̀ŋaîɲà] in \REF{ex:lop:51}.

\ea\label{ex:lop:51}
        \gll ɛ̀-ŋaî-ɲà íjòxoì l-eí-mweí       è-ŋaî-ibóŋ íjòxoì xò=xìjò x{\U}ná l-ɔ̀-lʊ́ŋà \\
             1\textsc{pl-irr}{}-not.be 1\textsc{pl.nom} \textsc{sub}{}-1\textsc{pl}{}-be.sick 1\textsc{pl-irr}{}-meet 1\textsc{pl.nom} with=people.\textsc{abs} \textsc{rel.pl} \textsc{sub-3}{}-be.many \\
        \glt ‘Were we not sick, we would have met many people.’    (\textsc{mb1-014} \textsc{0}1:28ː21)
\z

It is also possible for both the protasis (\textsc{p}) and the apodosis (\textsc{q}) to be negative clauses. This is illustrated in \REF{ex:lop:52}. Here, the negative condition is that the hearer does not go to Torit. 

\ea\label{ex:lop:52}
        \gll [ɪ́ɲà náŋ l-á-wú à=tòrìt]\textsubscript{\textsc{q}} [l-ɪ́-ɲá íjé l-í-wú]\textsubscript{\textsc{p}} \\ 
             not.be \textsc{1sg.nom} \textsc{sub-1sg}{}-go to=Torit \textsc{sub}{}-2-not.be 2\textsc{sg.nom} \textsc{sub}{}-2-go \\
        \glt ‘I'm not going to Torit if you don't go.’         (\textsc{mb1-163} \textsc{00}:03:55)
\z

It should be noted that some other negative constructions involving modality do not have normal pronominal marking. Constructions involving the obligatory prefix /tɛ-/ use the pronominal marker /ɛ-/ for first, second and third person, as illustrated in \REF{ex:lop:53}. This is also the case for most non-negative constructions involving the prefix /tɛ-/.

\ea\label{ex:lop:53}
        \gll ɛ́-tɛ̀-ɲà íjé l-í-mórò xìjò  \\ 
             3-\textsc{oblig}{}-not.be 2\textsc{sg.nom} \textsc{sub-2sg}{}-abuse people.\textsc{abs} \\
        \glt ‘You shouldn't abuse people.’  (\textsc{mb1-267} \textsc{00:}55:09)
\z

In complement clauses, the subordinator is not used on the lexical verbs, as is shown in the second clause in \REF{ex:lop:54} with the verb [òlòtú]. Similarly, the negative verb does not have the subordinating prefix /l-/ as shown in \REF{ex:lop:55}.  Here the negative verb [ɪ́ɲà] is placed after the complementizer and has the same, lexicalised form as the negative verb in independent clauses, as shown in \REF{ex:lop:2}. 

\ea\label{ex:lop:54}
        \gll a-íjːén náŋ xɪ̀ɟɔ̀ ò-lòt-ú ɪ́ɲɛ́ \\ 
             1\textsc{sg}{}-know 1\textsc{sg.nom} \textsc{comp} 3-go-\textsc{ven} 3\textsc{sg.nom} \\
        \glt ‘I know that he is coming.’    (\textsc{mb1-329} \textsc{00}:19:00)
\z

\ea\label{ex:lop:55}
        \gll a-íjːén náŋ xɪ̀ɟɔ̀ ɪ́ɲà ɪ́ɲɛ́ l-ò-lòt-ú \\ 
             1\textsc{sg}{}-know \textsc{1sg.nom} \textsc{comp} not.be 3\textsc{sg.nom} \textsc{sub}{}-3-go-\textsc{ven} \\
        \glt ‘I know that he is not coming.’    (\textsc{mb1-329} 00:19:26)
\z

Negation in relative clauses appears to be straightforward. When the action in the relative clause in \REF{ex:lop:56} is negated, the result is \REF{ex:lop:57}. The relative clause in \REF{ex:lop:57} can be compared with the corresponding declarative clause [ìɲá xìwàrú lòtːòxoí ìkúdò] ‘the leopard didn’t kill the squirrel’ in \REF{ex:lop:58}. The changes to form the negative relative clause are similar to those for other relative clauses.  That is, the relative marker /na/ is inserted, the common argument [xìwàrú] is omitted and the verb (in this case [lɔ̀ɲá]) is prefixed with the subordinate marker.

\ea\label{ex:lop:56}
        \gll ó-wòló ɪ́ɲɛ́ xìwàɾ{\U} [nà l-ò-tːòxo-í ìkúdò]  \\
             3-see 3\textsc{sg.nom} leopard.\textsc{abs} \textsc{rel.f} \textsc{sub}{}-3-kill-\textsc{itv} squirrel.\textsc{abs} \\
        \glt ‘He saw the leopard that killed the squirrel.’     (\textsc{mb1-255} \textsc{00:}28:20)
\z

\ea\label{ex:lop:57}
        \gll ó-wòló ɪ́ɲɛ́ xìwàɾ{\U} [nà l-ɔ̀-ɲá l-ò-tːòxo-í ìkúdò] \\
             3-see 3\textsc{sg.nom} leopard.\textsc{abs} \textsc{rel.f} \textsc{sub-3-}not.be \textsc{sub}{}-3-kill-\textsc{itv} squirrel.\textsc{abs} \\
        \glt ‘He saw the leopard that didn’t kill the squirrel.’       (\textsc{mb1-255} \textsc{00}:26:58)
\z

\ea\label{ex:lop:58}
        \gll ó-wòló ɪ́ɲɛ́ xìwàɾ{\U} ɪ́ɲà xɪ́wàɾʊ́ l-ò-tːòxo-í ìkúdò \\
             3-see 3\textsc{sg.nom} leopard.\textsc{abs} not.be leopard.\textsc{nom} \textsc{sub}{}-3-kill-\textsc{itv} squirrel.\textsc{abs} \\
        \glt ‘He saw the leopard.  The leopard didn’t kill the squirrel.’ (\textsc{mb1-342} \textsc{00}:39:19)
\z

An example of a sentence where NP\textsubscript{rel} takes on the object function is given in \REF{ex:lop:59}.  In this example, the head noun /xiwaru/ ‘leopard’ is relativized as the object of the verb /tːoxo/ ‘kill’.  The subject of this verb, [íkùdò], is placed in front of the lexical verb [lòtːòxoí], i.e. in the normal position of a negative clause.

\ea\label{ex:lop:59}
        \gll ó-wòló ɪ́ɲɛ́ xìwàɾ{\U} [nà l-ɔ̀-ɲá íkùdò l-ò-tːòxo-í] \\
             3-see 3\textsc{sg.nom} leopard.\textsc{abs} \textsc{rel.f} \textsc{sub-3-}not.be squirrel.\textsc{nom} \textsc{sub}{}-3-kill-\textsc{itv} \\
        \glt ‘He saw the leopard that the squirrel didn’t kill.’   (\textsc{mb1-255} \textsc{00}:26:58)
\z


\subsection{Negative lexicalizations}
\label{sec:lop:2.5}
Lopit has several verbs which can express negative concepts without the use of the negative verb. These include the verbs /riŋ/ ‘not know’; /tumo/ ‘not know’, ‘be ignorant of’; /miŋa/ ‘not see’; and /diak/ ‘not complete’ and /ɪwɔɪta/ ‘abstain’, ‘not do’. An example with /tumo/ is given in  \REF{ex:lop:60}.  

\ea\label{ex:lop:60}
        \gll ò-túmó ɪ́ɲɛ́ ròɾí x{\U}ná jànì ɪ̀nːá  \\
             3-not.know 3\textsc{sg.nom} things.\textsc{abs} of.\textsc{f.pl} tree.\textsc{abs} this.\textsc{f} \\
        \glt ‘He doesn’t know anything about this tree.’   (\textsc{mb1-341} \textsc{00:29}:23)
\z

A positive counterpart to \REF{ex:lop:60} can be expressed using the verb /ɪjːɛn/, ‘know’, as shown in \REF{ex:lop:61}.

\ea\label{ex:lop:61}
        \gll a-íjːén náŋ ròɾí x{\U}ná jànì ɪ̀nːá \\
             1\textsc{sg}{}-know 1\textsc{sg.nom} things.\textsc{abs} of.\textsc{f.pl} tree.\textsc{abs} this.\textsc{f} \\
        \glt ‘I know something about this tree.’   (\textsc{mb1-341} \textsc{00:30}:12)
\z

It should be noted that verbs like /tumo/ can also be negated, as shown in \REF{ex:lop:62}.

\ea\label{ex:lop:62}
        \gll ɪ́ɲà náŋ l-á-túmó ròɾí x{\U}ná jànì ɪ̀nːá \\
             not.be 1\textsc{sg.nom} \textsc{sub-1}{}-not.know things.\textsc{abs} of.\textsc{f.pl} tree.\textsc{abs} this.\textsc{f} \\
        \glt ‘I know something about this tree.' (lit. `I’m not ignorant about this tree.’)  (\textsc{mb1-341} \textsc{00:29}:57)
\z

An example with the verb /miŋa/, ‘not see’, is given in \REF{ex:lop:63}. 

\ea\label{ex:lop:63}
        \gll é-xì-rwátà  xɪ̀ɟɔ́  íjé x-í-míŋà-rì \\ 
             3-\textsc{pfv}{}-escape and 2\textsc{sg.nom}  \textsc{seq}{}-2-not.see-\textsc{itv} \\
        \glt ‘He escaped and you missed seeing it.’   (\textsc{mb1-307} \textsc{00:}01:59)
\z

The sense of this utterance can also be expressed using the verb /wolo/ ‘see’, in a negative construction. There is, however, a difference in meaning. In \REF{ex:lop:63}, the speaker is saying that the hearer was oblivious or was not paying attention. This meaning is also covered in \REF{ex:lop:64} with the extra possibility that the hearer was, in fact, looking out, but did not see it.

\ea\label{ex:lop:64}
        \gll é-xì-rwátà  x-ɔ̀-ɲá  íjé l-í-wóló \\ 
             3-\textsc{pfv}{}-escape \textsc{seq}{}-3-not.be 2\textsc{sg.nom}  \textsc{sub}{}-2-not.see \\
        \glt ‘He escaped and you didn’t see it.’ (\textsc{mb1-314} \textsc{00}:16:51)
\z

The verb /rita/, ‘be missing’, is used to convey that someone is not present or able to be found, as shown in \REF{ex:lop:65}. 

\ea\label{ex:lop:65}
        \gll ó-ɾúmá ɪ́ɲɛ́ xótóɲí l-é-ɾítà tɛ̀=xàɟì  \\ 
             3-find.\textsc{n} 3\textsc{sg.nom} mother.\textsc{abs} \textsc{sub}{}-3-be.missing from=house.\textsc{abs} \\
        \glt ‘He found his mother missing from the house.’   (\textsc{mb1-284} \textsc{00}:22:13)
\z

An example using the verb /ɪwɔɪta/, ‘abstain’, ‘don’t do’, is given in \REF{ex:lop:66}. The utterance could be made by someone who has just been offered a drink of beer. 

\ea\label{ex:lop:66}
        \gll á-la-ɪ́wɔ̀ɪ́tà náŋ tɛ̀=màtà    \\ 
              1\textsc{sg-persis-}abstain 1\textsc{sg.nom} from=\textsc{vn}.drink \\
        \glt ‘I’m still not drinking.’ (lit. ‘I’m still having a break from drinking.’) (\textsc{mb1-037} \textsc{00:51:21})
\z

\subsection{Other clausal negation constructions}
\label{sec:lop:2.6}
There is another form of negative verb which appears to be a variation of /ɲa/ ‘not.be'. I am calling it the persistive negative verb, /ɲei/, in contrast to the “unmarked” negative verb /ɲa/. An example of the persistive is given in \REF{ex:lop:67} with the verb /fanu/, ‘come.\textsc{pl’}. The prefix /le-/ is used to denote that the action is continuing. The negated form of \REF{ex:lop:67} is given in \REF{ex:lop:68}, where /ɲei/ is used as a higher verb. This construction is similar to standard negation, except that the lexical verb [èléfánù] does not have the subordinate marker. In addition, in contrast to the verb /ɲa/, /ɲei/ has normal third person marking (i.e. /e-/ and not /i-/). It should be noted that, although I have called /ɲei/ the persistive negative verb, it does not appear to be inflected for the persistive aspect. It could perhaps be called the ‘negative verb associated with the persistive’. It might possibly have been formed from the fusion of /ɲa/ with the itive suffix /-ei/, although it is hard to see any connection between the itive and the persistive.

\ea\label{ex:lop:67}
        \gll è-lé-fánù xáɟàŋáʔ \\ 
             3-\textsc{persis}{}-come.\textsc{pl} flies.\textsc{nom} \\
        \glt ‘The flies are still coming.' (or ‘the flies keep coming')    (\textsc{mb1-252} \textsc{00}:42:37)
\z

\ea\label{ex:lop:68}
        \gll è-ɲeî xáɟàŋáʔ  è-lé-fánù   \\ 
             3-not.be flies.\textsc{nom} 3-\textsc{persis}{}-come.\textsc{pl} \\
        \glt ‘The flies have stopped coming.' (lit. ‘The flies are not still coming.')    (\textsc{mb1-252} \textsc{00:}43:29)
\z

The construction in \REF{ex:lop:68} can be contrasted with that in \REF{ex:lop:69}, where the persistive form of the negative is not used. The [ɪɲà] verb is used, and the lexical verb has the subordinate marker, which is the standard negation construction.

\ea\label{ex:lop:69}
        \gll ɪ́ɲà xáɟàŋáʔ  l-è-fánù   \\ 
             not.be flies.\textsc{nom} \textsc{sub}{}-3-come.\textsc{pl} \\
        \glt ‘The flies did not come.’ (or ‘The flies are not coming.’) (\textsc{mb1-252} \textsc{00}:48:42)
\z

Another set of examples is shown in \REF{ex:lop:70} and \REF{ex:lop:71}. They also illustrate the contrast between the negative verb /ɲa/ and the persistive negative verb /ɲei/. In \REF{ex:lop:71}, there is the normal pronominal marker /a-/ ‘1\textsc{sg}', on the verb /ɲei/ and no subordinate marker on the lexical verb [álámátà] in the main clause.

\ea\label{ex:lop:70}
        \gll ɪ́ɲà náŋ l-á-mát-à kòfí l-ò-lót-ù patrik \\ 
             not.be 1\textsc{sg.nom} \textsc{sub-1sg}{}-drink-\textsc{ipfv} coffee.\textsc{abs} \textsc{sub}{}-3-go-\textsc{ven} Patrick \\
        \glt ‘I wasn't drinking coffee when Patrick came.’     (\textsc{mb1-319} \textsc{00:}54:14)
\z

\ea\label{ex:lop:71}
        \gll á-ɲeî náŋ á-lá-mát-à kòfí l-ò-lót-ù patrik \\ 
             1-not.be 1\textsc{sg.nom} 1\textsc{sg-persis}{}-drink-\textsc{ipfv} coffee.\textsc{abs} \textsc{sub}{}-3-go-\textsc{ven} Patrick \\
        \glt ‘I wasn't drinking coffee anymore when Patrick came.’ (lit. ‘I wasn’t still drinking.’)    (\textsc{mb1-319} \textsc{00:}54:29)
\z

It can also be used as a verbal noun, /ɲei/.  As shown in \REF{ex:lop:72}, /ɲei/ can be translated as ‘lacking', or ‘not having'. The noun phrase [ɲeî nà mòtːèɟà] ‘lacking of friends', is the subject of the verb /mir/ ‘defeat'. Note that the NP, which has absolutive case marking, precedes the verb and the verb is marked with the subordinate marker /l-/.  This is a fronted construction. This supports the view that /ɲei/ is a noun in this construction.

\ea\label{ex:lop:72}
        \gll è-sám ɪ́ɲɛ́ àɾá xátì ɲeî nà mòt:èɟà l-è-mír ɪ̀ɲɛ̀ \\ 
             3-be.rich 3\textsc{sg.nom} however \textsc{vn}.not.be.\textsc{abs} of.\textsc{f} friends.\textsc{abs} \textsc{sub}{}-3-defeat 3\textsc{sg.abs} \\
        \glt ‘He is rich but suffers from having no friends.’ (lit. ‘lacking of friends defeats him’)  (\textsc{mb1-273} \textsc{00}:49:38)
\z

\section{Non-clausal negation} \label{sec:lop:3}
\subsection{Negative replies}
\label{sec:lop:3.1}
The word [ɪ́ɲà] is used when answering in the negative, as shown in \REF{ex:lop:74}, which is an answer to \REF{ex:lop:73}.

\ea\label{ex:lop:73}
        \gll x-ɪ̂-cá  ije dè=ɟùbà \\ 
             \textsc{q}{}-2-dance 2\textsc{sg.nom} {in=Juba} \\
        \glt ‘Have you danced in Juba?’      (\textsc{mb1-142} \textsc{00}:28:20)
\z

\ea\label{ex:lop:74}
        \gll ɪ́ŋà â-cá  náŋ dè=tòɾìt ɪ́ŋá xàtì náŋ l-á-cá dè=ɟùbà \\ 
             not.be 1.\textsc{pfv}{}-dance 1\textsc{sg.nom} {in=Torit} not.be however 1\textsc{sg.nom} \textsc{sub}{}-1\textsc{sg}{}-dance {in=Juba} \\
        \glt ‘No, I’ve danced in Torit, but I haven’t danced in Juba.’ (\textsc{mb1-142} \textsc{00}:28:47)
\z

The negative answer [ɪ́ɲà] is also used to negate negative questions, indicating that it is used to negate the content of the question. This is illustrated in examples \REF{ex:lop:75} and \REF{ex:lop:76}. 

\ea\label{ex:lop:75}
        \gll x-ɔ́ɲà  xàtí l-í-wú à=xàŋ \\ 
             \textsc{q}{}-not.be indeed \textsc{sub}{}-2-go to=home.\textsc{abs} \\
        \glt ‘Aren’t you going home?’   (\textsc{mb1-330} \textsc{00:}41:00)
\z

\ea\label{ex:lop:76}
        \gll ɪ́ɲà á-la-ɪ̀tɪ́lá náŋ mɔ̀tːɛ̀ leìtí  \\ 
             not.be \textsc{1sg-persis-}wait 1\textsc{sg.nom} friend.\textsc{abs} my.\textsc{m} \\
        \glt ‘No, I’m still waiting for my friend.’ (\textsc{mb1-341} \textsc{00:}12:32)
\z

Another form of negative reply can be made using the word /ɔlɔxɔɲa/ ‘not yet’, as shown in \REF{ex:lop:77}.  This word is discussed in \sectref{sec:lop:2.1}.

% \ea\label{ex:lop:77}
%         \gll x-ɪ́-tːɔ̀xɔ́  íjé ɪ̀ɡɛ̀m  … ɔ̀lɔ̀xɔ́ɲà \\ 
%              \textsc{q}{}-2-finish.\textsc{pfv} 2\textsc{sg.nom} work.\textsc{abs}  not yet \\
%         \glt ‘Did you finish work?’     ‘Not yet.’      (\textsc{mb1-010} \textsc{0}1:16:24)%HC to be deleted because I have changed it to Q&A
% \z

\ea\label{ex:lop:77}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:lop:77Q}
        \gll x-ɪ́-tːɔ̀xɔ́  íjé ɪ̀ɡɛ̀m\\ 
             \textsc{q}{}-2-finish.\textsc{pfv} 2\textsc{sg.nom} work.\textsc{abs} \\
        \glt ‘Did you finish work?’   
    \exi{A: }\label{ex:lop:77A}
        \gll ɔ̀lɔ̀xɔ́ɲà \\ 
              not.yet \\
        \glt   ‘Not yet.’      (\textsc{mb1-010} \textsc{0}1:16:24)
    \end{xlist}
\z 

Lopit distinguishes ‘not’ from ‘never’ using the word /lɪŋ/.  Example \REF{ex:lop:79} is an answer to the question in \REF{ex:lop:78}. If /lɪŋ/ were omitted in \REF{ex:lop:79} the utterance would be translated as ‘I’m not going to Kapoeta’. Note that /lɪŋ/ can also be placed at the end of the utterance.

\ea\label{ex:lop:78}
        \gll x-ɪ́-wú ìsó  íjé a=kápwoítà   \\ 
             \textsc{q}{}-2-go \textsc{fut} 2\textsc{sg.nom} to=Kapoeta \\
        \glt ‘Will you go to Kapoeta?’ (\textsc{mb1-341} 00:53:16)
\z

\ea\label{ex:lop:79}
        \gll ɪ́ɲà náŋ  lɪ́ŋ l-á-wú a=kápwoítà  \\ 
             not.be \textsc{1sg.nom} \textsc{ideo} \textsc{sub-1sg-}go to=Kapoeta \\
        \glt ‘I’m never going to Kapoeta.’  (\textsc{mb1-341} 00:54:17)
\z

The word /lɪŋ/ can be used in a strongly negative reply as shown in \REF{ex:lop:80}. It is probably an ideophone, because its lexical meaning is not clear and its position in the word order in the clause is variable (see \citealt[347]{MoodieBillington2020}). (See also \sectref{sec:lop:4.4} for more discussion of /lɪŋ/.)

\ea\label{ex:lop:80}
        \gll lɪ́ŋ lɪ́ŋ lɪ́ŋ \\ 
             \textsc{ideo} \textsc{ideo} \textsc{ideo}  \\
        \glt ‘Never, everǃ’ (\textsc{mb1-341} \textsc{00}:57:08)
\z

\subsection{Negative quantifiers and indefinites}
\label{sec:lop:3.2}
Lopit does not have words which correspond to words like ‘nobody’, ‘nothing’, or ‘nowhere’. The equivalent meanings of these words are usually expressed using the negative verb /ŋa/ in combination with words like /nabɔ/ ‘one.\textsc{f}’; /xuɾe/ ‘some.\textsc{f}’, /ŋai/ ‘who’ and /ɲo/ ‘what’. This could be because negation in Lopit appears to be almost exclusively verbal in character and is not observed in affixes or particles.

The Lopit translations of ‘no-one’ or ‘nobody’ can be expressed using the interrogative pronoun /ŋai/ ‘who’ together with the negative verb /ɪɲa/.  This can be used when the gender is unknown or if the referents include both males and females. Thus, in \REF{ex:lop:81}, /ɲo/ could be interpreted as the indefinite pronoun ‘something’ and, in \REF{ex:lop:82}, /ŋai/ could be interpreted as ‘somebody’ or ‘anybody’. The use of interrogatives as indefinite pronouns is cross-linguistically common (\citealt[170--176]{Haspelmath1997}). 

\ea\label{ex:lop:81}
        \gll ɪ́ɲà ɲo l-ɔ̀-wːɔ́n dɛ̀=xàɟì  \\ 
             not.be what \textsc{sub-3-}exist in=house.\textsc{abs} \\
        \glt ‘There is nothing in the house.’ (\textsc{mb1-128} \textsc{00}:27:19)
\z

\ea\label{ex:lop:82}
        \gll á-ɡí-ɡíl-ò náŋ ɪ́ɲà ŋaî l-ò-lòtú \\ 
             1\textsc{sg-red-}think\textsc{{}-n} 1\textsc{sg.nom} not.be who \textsc{sub-}3-come \\
        \glt ‘I think that nobody is coming.’ (\textsc{mb1-123} 00:22:11)
\z

The following example, \REF{ex:lop:83}, shows the use of /xʊlak/, ‘some.\textsc{m}’ together with /ɪɲa/ to express the equivalent of the English ‘not any’ or ‘no’.

\ea\label{ex:lop:83}
        \gll ɪ́ɲà xìfjôŋ x{\U}lák \\
             not.be water.\textsc{abs} some.\textsc{m} \\
        \glt `There wasn't any rain.' (\textsc{mb1-123} \textsc{00:}12:29)
\z

If the gender of the referents is known, then the words /nabɔ/, /lɔbɔ/ can be used with the noun /toxoni/ ‘person’. In the following example, the quantifier is placed in front of the noun to give the phrase [lɔ́bɔ́ tòxónì].  This is a pragmatically marked position since quantifiers normally follow nouns. This gives special emphasis to the word /lɔbɔ/ so that the literal translation of \REF{ex:lop:84} is ‘I think (that) not (even) one man is coming.’

\ea\label{ex:lop:84}
        \gll á-ɡíl náŋ ɪ́ɲà lɔ́bɔ́ tòxónì l-ò-lòt-ù  \\ 
             1\textsc{sg}{}-believe 1\textsc{sg.nom} not.be one.\textsc{m} person.\textsc{nom} \textsc{sub}{}-3-go-\textsc{ven} \\
        \glt ‘I think that nobody (\textsc{m}) is coming.’   (\textsc{mb1-335} 01:17:54)
\z


\subsection{Negative derivation and case marking}
\label{sec:lop:3.3}
Most property concepts in Lopit are expressed using property stative verbs.  This means that the property concept can be negated by negating the stative verb. This is illustrated in positive example in \REF{ex:lop:85} and the negative one in \REF{ex:lop:86}. 

\ea\label{ex:lop:85}
        \gll ó-múnó xítò leítì \\ 
             3-be.happy child.\textsc{nom} my.\textsc{m} \\
        \glt ‘My son is happy.’   (\textsc{mb1-267} 00:51:29)
\z

\ea\label{ex:lop:86}
        \gll l-ɛ́-ɲá xítò leítì  l-ò-múnò ɪ́ɲà náŋ l-á-múnò \\ 
             \textsc{sub-3-}not.be child.\textsc{nom} my.\textsc{m} \textsc{sub-3-}be.happy not.be 1\textsc{sg.nom} \textsc{sub-1sg-}be.happy \\
        \glt ‘If my son is not happy, I'm not happy.’   (\textsc{mb1-267} 00:53:49)
\z

Adverbial notions are also often expressed using stative verbs. There is a verb /laita/ which is used to convey that an event occurred recently. It is difficult to translate directly but conveys something like ‘occurred recently’ or ‘has only recently been in existence’. It is used with a subject, e.g., the nominative form of the pronoun [ɪ́ɲɛ́] in \REF{ex:lop:87}, and always requires a verbal noun or infinitive.  The negation of \REF{ex:lop:87} is shown in \REF{ex:lop:88} and uses the standard negation construction.

\ea\label{ex:lop:87}
        \gll e-laita-k ɪ́ɲɛ́ ŋàíjòŋ     \\ 
             3-be.recent-\textsc{dat} 3\textsc{sg.nom} come\textsc{.vn} \\
        \glt ‘He came recently.’ (\textsc{mb1-341} 00:46:47)
\z

\ea\label{ex:lop:88}
        \gll ɪ́ɲà ɪ́ɲɛ́ l-e-laita-k ŋàíjòŋ \\ 
             not.be 3\textsc{sg.nom} \textsc{sub-}3-be.recent-\textsc{dat} come\textsc{.vn} \\
        \glt ‘He didn't come recently.’ (\textsc{mb1-341} 00:45:14)
\z

In Lopit, meanings of absence, opposition and antonymy are expressed with verbal negation rather than with affixes on adjectives, nouns and adverbs.  An example of how the meaning of absence or lack is expressed is illustrated in example \REF{ex:lop:89}. A coordination construction is used (see \sectref{sec:lop:4.5}) and the sense of ‘without’ is conveyed with the word /xɔɲa/ (which can be translated as ‘and not’) together with the preposition /xɔ/, ‘with’.

\ea\label{ex:lop:89}
        \gll è-ríbó ɪ́ɲɛ́ x-ɔ̀-ɲà xɔ̀=báŋɪ́ \\ 
             3-fight 3\textsc{sg.nom} \textsc{seq}{}-3-not.be with=fear \\
        \glt ‘He fights without fear.’ (lit. ‘He fights and not with fear.’) (\textsc{mb1-342} 00:41:51)
\z


\section{Other aspects of negation} \label{sec:lop:4}
\subsection{The scope of negation}
\label{sec:lop:4.1}
Focussed constructions can be expressed in Lopit using cleft constructions. The unmarked \textsc{vso} word order is shown in example \REF{ex:lop:90}.  When the speaker wants to say that “it was Victor (and no-one else) that I saw”, then an \textsc{ovs} word order is used, as shown in \REF{ex:lop:91}.

\ea\label{ex:lop:90}
        \gll á-wóló náŋ vìktòr \\ 
             1\textsc{sg-}see \textsc{1sg.nom} Victor.\textsc{abs} \\
        \glt ‘I see Victor.’ (\textsc{mb1-341} 00:33:41)
\z

\ea\label{ex:lop:91}
        \gll vìktòr l-á-wóló náŋ \\ 
             Victor.\textsc{abs} \textsc{sub-}1\textsc{sg-}see \textsc{1sg.nom} \\
        \glt ‘(It is) Victor I see.’ (\textsc{mb1-341} 00:36:13)
\z

A focussed object (e.g. Victor) can be the scope of a negative construction, as shown in \REF{ex:lop:92}. Here, the standard negation construction is used, except that the subordinate marker is used on the negative verb [láɲá], since it has moved from the clause-initial position (as was the case with /lawolo/ on example \REF{ex:lop:91}). Note that the negative verb has the first-person marker /a-/.

\ea\label{ex:lop:92}
        \gll vìktòr l-á-ɲá náŋ l-á-wóló \\ 
             Victor.\textsc{abs} \textsc{sub-1sg-}not.be \textsc{1sg.nom} \textsc{sub-}1\textsc{sg-}see \\
        \glt ‘(It is) Victor I do not see.’ (\textsc{mb1-341} 00:36:39)
\z

In \REF{ex:lop:93}, the scope of negation is different to that in \REF{ex:lop:92}. In \REF{ex:lop:93}, the scope of negation is the whole of the utterance in \REF{ex:lop:91}, i.e. [vìktòr láwóló náŋ] ‘Victor I see’. Once again, standard negation is used.

\ea\label{ex:lop:93}
        \gll ɪ́ɲà l-áɾá vìktòr l-á-wóló náŋ dànjèl l-á-wóló náŋ \\ 
             not.be \textsc{sub}{}-3be Victor.\textsc{abs} \textsc{sub-}1\textsc{sg-}see \textsc{1sg.nom} Daniel  \textsc{sub-}1\textsc{sg-}see \textsc{1sg.nom} \\
        \glt ‘It’s not Victor I see, (it’s) Daniel I see.’ (\textsc{mb1-341} 00:40:21)
\z


\subsection{Negative polarity}
\label{sec:lop:4.2}
As mentioned in \sectref{sec:lop:3.2}, Lopit does not have words equivalent to ‘nobody’, ‘none’, ‘anybody’ etc. The words /xuɾeʔ/ ‘some.\textsc{f}’ and /xʊlak/ ‘some.\textsc{m}’, are used in combination with negation to convey the sense of ‘not any’, as shown in \REF{ex:lop:94}.

\ea\label{ex:lop:94}
        \gll ɪ́ɲà íjòxoí l-eí-wːɔ̀n xɔ̀ dòŋìʔ xùɾêʔ \\ 
             not.be 1\textsc{pl.nom} \textsc{sub-1pl-}exist with drums.\textsc{abs} some.\textsc{f} \\
        \glt ‘We don’t have any drums.’ (lit. ‘We do not exist with some drums.’) (\textsc{mb1-}329 00:07:22)
\z

This contrasts with \REF{ex:lop:95}, which is a simpler negative expression.

\ea\label{ex:lop:95}
        \gll ɪ́ɲà íjòxoí l-eí-wːɔ̀n xɔ̀ dòŋìʔ   \\ 
             not.be 1\textsc{pl.nom} \textsc{sub-1pl-}exist with drums.\textsc{abs} \\
        \glt ‘We have no drums.’ (lit. ‘We do not exist with drums.’) (\textsc{mb1-}329 00:06:38)
\z

\subsection{Marking of NPs in the scope of negation}
\label{sec:lop:4.3}

There appears to be no effect of case marking under negation, as illustrated with the object [ʃaɪ̂] ‘tea.\textsc{abs}’ in the affirmative and the negative utterances in \REF{ex:lop:96} and \REF{ex:lop:97} respectively. This is not surprising since there are only two cases in Lopit, nominative and absolutive.

\ea\label{ex:lop:96}
        \gll a-íwús-ò náŋ ʃaɪ̂ \\ 
             \textsc{1sg}{}-drink-\textsc{ipfv} 1\textsc{sg.nom} tea.\textsc{abs} \\
        \glt ‘I’m drinking tea.’  (\textsc{mb1-269} \textsc{00}:29:04)
\z

\ea\label{ex:lop:97}
        \gll ɪ́ɲà náŋ l-a-íwús-ò ʃaɪ̂ \\ 
             not.be 1\textsc{sg.nom} \textsc{sub-1sg}{}-drink-\textsc{ipfv} tea.\textsc{abs} \\
        \glt ‘I’m not drinking tea.’  (\textsc{mb1-269} \textsc{00}:32:18)
\z

\subsection{Reinforcing negation}
\label{sec:lop:4.4}
Several means of reinforcing negation have been observed. The first is with ideophonic or emphatic adverbs (see \sectref{sec:lop:3.1}). The word /lɪŋ/, when used in an utterance such as \REF{ex:lop:98}, can be translated as ‘really’ or ‘definitely’, although it can also be interpreted as ‘never’ as discussed in \sectref{sec:lop:3.1}. To add further emphasis, /lɪŋ/ can be combined with /lɛfɛ/, as shown in \REF{ex:lop:99}.  

\ea\label{ex:lop:98}
        \gll ɪ́ɲà náŋ lɪ́ŋ l-á-wák ìjè \\ 
             not.be 1\textsc{sg.nom} \textsc{ideo} \textsc{sub-1sg-}like 2\textsc{sg.abs} \\
        \glt ‘I really don’t like you.’ (\textsc{mb1-341} 00:54:45)
\z

\ea\label{ex:lop:99}
        \gll ɪ́ɲà náŋ l-á-wák ìjè lɛ́fɛ̀ lɪ́ŋ  \\ 
             not.be 1\textsc{sg.nom} \textsc{sub-1sg-}like 2\textsc{sg.abs} \textsc{ideo}  \textsc{ideo} \\
        \glt ‘I most certainly don’t like you.’ (\textsc{mb1-341} 00:55:23)
\z

Another means is with marked word order. A quantifier will normally follow the noun that it modifies, as shown in \REF{ex:lop:100}. However, if the quantifier precedes the noun, as in \REF{ex:lop:101}, it will add emphasis. The use of marked word order for pragmatic effect is common in Lopit (see \citealt[152]{MoodieBillington2020}). 

\ea\label{ex:lop:100}
        \gll ɪ́ɲà xìfjôŋ x{\U}lák l-ɔ̀-wːɔ́n  \\
             not.be water.\textsc{abs} some.\textsc{m} \textsc{sub-3-}exist \\
        \glt ‘There isn’t any water.’ (\textsc{mb1-342} 00:09:40)
\z

\ea\label{ex:lop:101}
        \gll ɪ́ɲà x{\U}lák xìfjôŋ l-ɔ̀-wːɔ́n  \\
             not.be some.\textsc{m} water.\textsc{abs} \textsc{sub-3-}exist \\
        \glt ‘There isn’t \textbf{any} water.’ (\textsc{mb1-342} 00:10:06)
\z

\subsection{Negation, coordination and complex clauses}
\label{sec:lop:4.5}
The negative verb /ɲa/ can also be used in clause coordination with the sequential marker /x-/. The sequential marker is prefixed to the verb in the second clause of a pair of coordinated clauses, as shown in \REF{ex:lop:102} with the verb /wolo/, ‘see’. In \REF{ex:lop:103}, the sequential marker is prefixed to the negative verb to negate the second clause and provide an adversative coordination construction.  In this situation, the pronominal marker /ɔ-/ is used.  As discussed in \sectref{sec:lop:4.6.3}, the form /xɔɲa/ is used in Dorik Lopit and not /xɪɲa/. The lexical verb in the negated clause, [xɔ̀ɲà lòwòló kɛ̀ɾà], is now marked with the subordinating prefix /l-/, as is the case in standard negation. As shown in \REF{ex:lop:104}, it is also possible to express the same meaning as \REF{ex:lop:103} using the standard negation form /ɪɲa/. In this case, coordination is achieved by clause juxtaposition.

\ea\label{ex:lop:102}
        \gll ò-wòló náŋóɾúò  xɪ́sʊ́ŋ x-ò-wòló  kɛ̀ɾà  \\ 
             3-see.\textsc{pfv} wife.\textsc{nom} cattle.\textsc{abs} \textsc{seq}{}-3-see.\textsc{pfv}  sheep.\textsc{pl.abs} \\
        \glt ‘The wife saw cattle and she saw sheep.’    (\textsc{mb1-064} 01:04:20)
\z

\ea\label{ex:lop:103}
        \gll ò-wòló náŋóɾúò  xɪ́sʊ́ŋ x-ɔ̀-ɲà l-ò-wòló  kɛ̀ɾà \\ 
             3-see.\textsc{pfv} wife.\textsc{nom} cattle.\textsc{abs} \textsc{seq}{}-3-not.be \textsc{sub}{}-3-see.\textsc{pfv}  sheep.\textsc{pl.abs} \\
        \glt ‘The wife saw cattle but didn't see sheep.’      (\textsc{mb1-064} 01:04:22)
\z

\ea\label{ex:lop:104}
        \gll ò-wòló náŋóɾúò  xɪ́sʊ́ŋ ɪ́ɲà xàtí l-ò-wòló  kɛ̀ɾà \\ 
             3-see.\textsc{pfv} wife.\textsc{nom} cattle.\textsc{abs} not.be indeed \textsc{sub}{}-3-see.\textsc{pfv}  sheep.\textsc{pl.abs} \\
        \glt ‘The wife saw cattle; she didn't see sheep.’      (\textsc{mb1-064} 01:04:10)
\z

Lopit has a type of negation called contrastive negation, which refers to “expressions which are combinations of affirmation and negation in which the focus of negation is replaced in the affirmative part of the expression” \citep[10]{Silvennoinen2019}.  These can be negative-first or negative-second constructions. The Lopit examples given here are all negative-second constructions, i.e. [Y not X], where the focus of the affirmative is replaced in the negative part of the expression. Thus, in \REF{ex:lop:105}, the focus of the affirmation is [dɛ̀=nàŋ] ‘with me’ and this is changed to [dè=ìjè] in the negation expression.  In \REF{ex:lop:106}, the focus of the affirmation is [dê=xìtó nà lòmweí], ‘to the girl who is sick’. 

\ea\label{ex:lop:105}
        \gll á-gí-gílò náŋ e-ìɟóríà ɪ́ɲɛ́ dɛ̀=nàŋ    ìɲá l-áɾá dè=ìjè   \\
             1\textsc{sg-red-}think 1\textsc{sg.nom} 3-be.angry \textsc{3sg.nom} {with=1\textsc{sg.abs}} not.be \textsc{sub}{}-be  {to=2\textsc{sg.abs}} \\
        \glt ‘I think he is angry with you, not me.’ (\textsc{mb1-342} 00:23:36)
\z

\ea\label{ex:lop:106}
        \gll ɔ́-wák ɪ́ɲɛ́ x-ɪ̀ɾɔ̀ dê=xìtó nà l-ò-mweí ìɲá l-áɾá dê=xìtó nà l-ò-múnò bí \\ 
             3-want \textsc{3sg.nom} \textsc{inf}{}-speak to=child \textsc{rel.f} \textsc{sub}{}-3-be.sick not.be \textsc{sub}{}-be to=child.\textsc{abs} \textsc{rel.f} \textsc{sub}{}-3-be.happy \textsc{att} \\
        \glt ‘He wants to speak to the sick girl not to the happy one.’        (\textsc{mb1-257} \textsc{00:}09:35)
\z

One interesting feature of these constructions is that the negation involves the expression /ɪɲa lara/. The verb /laɾa/ cannot be omitted, although it is usually shortened to /la/ in normal discourse. The literal translation of \REF{ex:lop:106} could be something like ‘He wants to speak to the sick child, it is not with the happy one’ or ‘He wants to speak to the sick child, it is not that it is with the happy one.’ It appears to be a rather complex construction, perhaps originating from clausal juxtaposition. More research is required to understand it.

Constructions of the type ‘neither nor’ in Lopit have not been observed.

\subsection{Miscellaneous aspects of negation}
\label{sec:lop:4.6}
\subsubsection{Negative transport}
\label{sec:lop:4.6.1}

Negative transport (neg-raising) can occur in Lopit. Examples are given using the verb /ɡilo/ ‘believe’. The first example, \REF{ex:lop:107}, has a higher-clause negative, which can be interpreted as negating the lower-clause predicate (i.e. negating [òlòtú ɪ́ɲɛ́]). That is, \REF{ex:lop:107} can be interpreted as \REF{ex:lop:108}.

\ea\label{ex:lop:107}
        \gll ɪ́ɲà náŋ l-á-ɡí-ɡílò xɪ̀ɟɔ̀ ò-lòt-ú ɪ́ɲɛ́ \\ 
             not.be \textsc{1sg.nom} \textsc{sub-}1\textsc{sg}{}-\textsc{red-}believe \textsc{comp} 3-go-\textsc{ven} 3\textsc{sg.nom} \\
        \glt ‘I don't think that he is coming.’ (\textsc{mb1-329} 00:22:47)
\z

\ea\label{ex:lop:108}
        \gll á-ɡí-ɡílò náŋ xɪ̀ɟɔ̀ ɪ́ɲà ɪ́ɲɛ́ l-ò-lòt-ú \\ 
             1\textsc{sg}{}-\textsc{red-}believe \textsc{1sg.nom} \textsc{comp} not.be 3\textsc{sg.nom} \textsc{sub}{}-3-go-\textsc{ven} \\
        \glt ‘I think that he is not coming.’    (\textsc{mb1-329} 00:22:26)
\z

However, it appears that few verbs allow this construction. Examples \REF{ex:lop:107} and \REF{ex:lop:108} are repeated in \REF{ex:lop:109} and \REF{ex:lop:110} with the verb /ijːen/ in place of the verb /ɡilo/. Both examples are acceptable, but they have different meanings. The higher-clause negative in \REF{ex:lop:109} does not negate the lower-clause predicate [òlòtú ɪ́ɲɛ́].

\ea\label{ex:lop:109}
        \gll ɪ́ɲà náŋ l-á-ijːén xɪ̀ɟɔ̀ ò-lòt-ú ɪ́ɲɛ́ \\ 
             not.be \textsc{1sg.nom} \textsc{sub-}1\textsc{sg}{}-know \textsc{comp} 3-go-\textsc{ven} 3\textsc{sg.nom} \\
        \glt ‘I don't know that he is coming.’ (\textsc{mb1-329} 00:19:47)
\z

\ea\label{ex:lop:110}
        \gll á-ijːén náŋ xɪ̀ɟɔ̀ ɪ́ɲà ɪ́ɲɛ́ l-ò-lòt-ú \\ 
             1\textsc{sg}{}-know \textsc{1sg.nom} \textsc{comp} not.be 3\textsc{sg.nom} \textsc{sub}{}-3-go-\textsc{ven} \\
        \glt ‘I know that he is not coming.’    (\textsc{mb1-329} 00:19:26)
\z


\subsubsection{Metalinguistic negation}
\label{sec:lop:4.6.2}

Two examples of metalinguistic negation have been recorded and are shown in \REF{ex:lop:111} and \REF{ex:lop:112}.  More research is needed to determine if these types of expressions are common.

\ea\label{ex:lop:111}
        \gll ɪ́ɲá  íjòxoì l-a-wak bálú e-rok íjòxoì bálú  \\ 
             not.be 1\textsc{pl.nom} \textsc{sub-1-}want beer\textsc{.abs} \textsc{3-}love 1\textsc{pl.nom} beer\textsc{.abs} \\
        \glt ‘We don't like beer, we love itǃ’ (\textsc{mb1-342} 00:18:18)
\z

\ea\label{ex:lop:112}
        \gll ɪ́ɲá  ɪ́ɲɛ́  l-ɛ̀-jɛìtà ŋòɾúò wùník ɛ̀-jɛìtà ŋòɾúò mjèt \\ 
             not.be 3\textsc{sg.nom} \textsc{sub-3-}have women.\textsc{abs} three.\textsc{abs} 3-have woman.\textsc{abs} five.\textsc{abs} \\
        \glt ‘He doesn’t have three wives, he has fiveǃ’ (\textsc{mb1-342} 00:19:14)
\z


\subsubsection{Diachronic notes and observations}
\label{sec:lop:4.6.3}
There is not much data available on how negation in Lopit might have developed. It is possible that the form /ɪɲa/ might have been lexicalised from the verb /ɲa/ with the second person prefix /ɪ-/.  Lexicalisation commonly occurs with the third-person prefix.  This is observed with the form /xɔɲa/ used in coordination constructions (\sectref{sec:lop:4.5}). The alternation of /ɔlɔxɔɲa/ and /ɛlɛxɪɲa/ and the use of normal pronominal marking in subordinate clauses and negative modal verbs suggests that this lexicalization has only partially taken place or there is some kind of formal division based on different functions or clause types.

Eastern Nilotic languages use several methods for negation, including a separate verb, a negative prefix and a particle.  These are shown for a range of languages in \tabref{tab:lop:3}.\footnote{References: Otuho \citep{Muratori1938}; Lokoya \citep{Westermann1944}; Maa \citep{TuckerEtAl1955}; Turkana \citep{Dimmendaal1983}; Ateso \citep{Barasa2017}; Bari \citep{Spagnolo1933}} There does not appear to be much in common amongst the forms of negation.

% \begin{table} 
%     \begin{tabularx}{\textwidth}{XXXXXp{2cm}p{1.7cm}l}
%     & {\bfseries Lopit} & {\bfseries Otuho} & {\bfseries Lokoya} & {\bfseries Maa} & {\bfseries Turkana} & {\bfseries Ateso} & {\bfseries Bari}\\
%     \lsptoprule
%     {verb} & \textit{ɲa} & \textit{beng} &  &  & {\textit{mamaka} \newline (lack)} & {\textit{mameun} \newline (be absent)} & \\
%     {prefix} &  &  &  & \textit{m-} & \textit{na-; pe-; nyi-} &  & \\
%     {particle} &  &  & \textit{aŋ} & \textit{ḿmɛ} &  & \textit{mam} & \textit{ti; tine}\\
%     \lspbottomrule
%     \end{tabularx}
% \caption{Negation in Eastern Nilotic languages}
% \label{tab:lop:3}
% \end{table}

\begin{table}
    \begin{tabularx}{.85\textwidth}{llll}
\lsptoprule
                        & {particle}& {prefix} & {verb} \\
    \midrule
    { Lopit}   &               &              & \textit{ɲa}     \\
    { Otuho}   &               &              & \textit{beng}  \\
    { Lokoya}  & \textit{aŋ}   &              & \\
    { Maa}     & \textit{ḿmɛ}  & \textit{m-}  & \\
    { Turkana} & {\textit{mamaka} (lack)} & \textit{na-; pe-; nyi-}&\\
    { Ateso}   & \textit{mam}  &              & {\textit{mameun} (be absent)}\\
    { Bari}    & \textit{ti; tine} &          &\\
    \lspbottomrule
    \end{tabularx}
\caption{Negation in Eastern Nilotic languages} 
\label{tab:lop:3}
\end{table}


Lopit is not the only Eastern Nilotic language which has a verbal negator. Otuho, a neighbouring language, has the negative verb \textit{beng}, although the two negative verbs appear to be unrelated. The Otuho negative verb has pronominal marking when used in the simple form. Thus, \textit{abeng} in \REF{ex:lop:113} contrasts with [ɪ́ɲà] in \REF{ex:lop:2}. Although the word order is the same as in Lopit, the Otuho construction differs in the sense that the lexical verb, \textit{alo}, ‘\textsc{1sg-}go’, is not inflected with dependent morphology.

\ea\label{ex:lop:113} Otuho (ISO: \href{https://iso639-3.sil.org/code/lot}{lot}, Glottocode: \href{https://glottolog.org/resource/languoid/id/otuh1238}{otuh1238})  \\ 
        \gll a-beng dwo ni a-lo \\ 
             1\textsc{sg}{}-not.be \textsc{imm.pst} 1\textsc{sg} 1\textsc{sg}{}-go \\
        \glt ‘I didn’t go there.’ \citep[339]{Muratori1938}
\z

\section{Summary}

Negative constructions and elements are listed in \tabref{tab:lop:4}, together with their respective functions. Negation in Lopit almost always involves a verbal construction in a complex clause. In declarative, interrogative and subordinate constructions, the verb /ɲa/ is used to negate the lexical verb. The construction is bi-clausal in that there are two finite verbs, a ‘higher negative’ verb and a finite lexical verb. The negative verb is often expressed as /ɪŋa/, which is probably a fused form. In some cases (e.g. subordinate constructions) it has normal pronominal marking and it can also be marked for modality. The exact nature of the bi-clausal construction is not clear, but it is similar to other auxiliary-like constructions in the language.

\begin{table}[t]
\small
\begin{tabularx}{\textwidth}{lXl}
\lsptoprule
{\bfseries construction} & {\bfseries function} & {\bfseries section}\\
\midrule
\multirow{3}{*}{\textit{ɪɲa}+ \textsc{sub-}LV} & {negation of declarative clauses} & {\sectref{sec:lop:2.1}} \\
& {negation of complement clauses} & {\sectref{sec:lop:2.3}}\\
& {negation of stative property verbs} & {\sectref{sec:lop:2.3}, \ref{sec:lop:3.3}}\\
\cmidrule{1-3}
{\textsc{sub-}\textsc{pro}-\textit{ɲa} +   \textsc{sub-}LV} & {negation in adverbial, relative and conditional clauses}  & {\sectref{sec:lop:2.3}}\\
\midrule
\multirow{2}{*}{\textit{xɔɲa+} \textsc{sub-}LV} & {negation in coordinated clauses} & {\sectref{sec:lop:4.5}}\\
& {negation of interrogatives} & {\sectref{sec:lop:2.2.2}}\\
\midrule
{\textsc{pro}-modal.prefix-\textit{ɲa}  +  \textsc{sub-}LV} & {negation involving modality} & {\sectref{sec:lop:2.3}}\\
{\textsc{pro}-\textit{ɲei} +   \textsc{pro}-\textsc{persis-}LV.root} & {negation of the persistive aspect} & {\sectref{sec:lop:2.5}}\\
{\textit{idek} + \textit{xaɪ}-verb.root} & {negative imperatives} & {\sectref{sec:lop:2.2.1}}\\
{\textit{ɪɲa+} \textsc{sub-}copula.verb} & {negation of most stative predications} & {\sectref{sec:lop:2.3}}\\
{\textit{ɪɲa}} & {negation of non-locative existentials} & {\sectref{sec:lop:2.3}}\\
{\textit{ɪɲa+} \textsc{q.pro} +   \textsc{sub-}LV} & {negative quantifier (‘nobody’, ‘nothing’)} & {\sectref{sec:lop:3.2}}\\
{\textit{ɪɲa+} quantifier + \textsc{sub-}LV} & {negative quantifier (‘none’, ‘nobody’)} & {\sectref{sec:lop:3.2}}\\
\textit{ɪɲa} & {negative answer (‘no’)} & {\sectref{sec:lop:3.1}}\\
\textit{lɪŋ} & {negative intensifier, never} & {\sectref{sec:lop:3.1}, \ref{sec:lop:4.4}}\\
\textit{ɔlɔxɔɲa, ɛlɛxɪɲa} & {‘not yet’} & {\sectref{sec:lop:2.1}, \ref{sec:lop:3.1}} \\
\lspbottomrule
\end{tabularx}

\caption{Negative constructions and their functions}
\label{tab:lop:4}
\end{table}

In imperatives, a different construction is used. This involves the verb /idek/, ‘stop’, leave’, together with the prefix /xaɪ-/ on the lexical verb. 

In stative predications, the copula verbs /ɾa/ ‘be’ and /wːɔn/ ‘exist’ are often used and can be negated with /ɲa/. There are also some verbless stative predicates.  These are normally negated with the standard negation construction with the introduction of a copula verb. There is one significant exception to this. Non-locative existential predications are negated using the negative verb /ɪɲa/ only, with no lexical or copula verb. Locative existential predicates are negated with the negative verb and the copula verb /wːɔn/.  

In Lopit, there appear to be no negative quantifiers or indefinites. The equivalent meanings of these kinds of words are usually expressed using negation in combination with positive quantifiers and indefinites. Some indefinites in negative constructions are expressed using the negation and an interrogative pronoun.

Most property concepts in Lopit are expressed using property stative verbs.  The property concept can be negated by negating the stative verb. Verbal negation is used rather than affixes on noun, adjectives and adverbs.

% \begin{table}
% \small
% \begin{tabularx}{\textwidth}{lXl}
% \lsptoprule
% {\bfseries construction} & {\bfseries function} & {\bfseries section}\\
% \midrule
% \multirow{3}{*}{\textit{ɪɲa}+ \textsc{sub-}LV} & {negation of declarative clauses} & {\sectref{sec:lop:2.1}} \\
% & {negation of complement clauses} & {\sectref{sec:lop:2.3}}\\
% & {negation of stative property verbs} & {\sectref{sec:lop:2.3}, \ref{sec:lop:3.3}}\\
% \cmidrule{1-3}
% {\textsc{sub-}\textsc{pro}-\textit{ɲa} +   \textsc{sub-}LV} & {negation in adverbial, relative and conditional clauses}  & {\sectref{sec:lop:2.3}}\\
% \midrule
% \multirow{2}{*}{\textit{xɔɲa+} \textsc{sub-}LV} & {negation in coordinated clauses} & {\sectref{sec:lop:4.5}}\\
% & {negation of interrogatives} & {\sectref{sec:lop:2.2.2}}\\
% \midrule
% {\textsc{pro}-modal.prefix-\textit{ɲa}  +  \textsc{sub-}LV} & {negation involving modality} & {\sectref{sec:lop:2.3}}\\
% {\textsc{pro}-\textit{ɲei} +   \textsc{pro}-\textsc{persis-}LV.root} & {negation of the persistive aspect} & {\sectref{sec:lop:2.5}}\\
% {\textit{idek} + \textit{xaɪ}-verb.root} & {negative imperatives} & {\sectref{sec:lop:2.2.1}}\\
% {\textit{ɪɲa+} \textsc{sub-}copula.verb} & {negation of most stative predications} & {\sectref{sec:lop:2.3}}\\
% {\textit{ɪɲa}} & {negation of non-locative existentials} & {\sectref{sec:lop:2.3}}\\
% {\textit{ɪɲa+} \textsc{q.pro} +   \textsc{sub-}LV} & {negative quantifier (‘nobody’, ‘nothing’)} & {\sectref{sec:lop:3.2}}\\
% {\textit{ɪɲa+} quantifier + \textsc{sub-}LV} & {negative quantifier (‘none’, ‘nobody’)} & {\sectref{sec:lop:3.2}}\\
% \textit{ɪɲa} & {negative answer (‘no’)} & {\sectref{sec:lop:3.1}}\\
% \textit{lɪŋ} & {negative intensifier, never} & {\sectref{sec:lop:3.1}, \ref{sec:lop:4.4}}\\
% \textit{ɔlɔxɔɲa, ɛlɛxɪɲa} & {‘not yet’} & {\sectref{sec:lop:2.1}, \ref{sec:lop:3.1}} \\
% \lspbottomrule
% \end{tabularx}
%
% \caption{Negative constructions and their functions}
% \label{tab:lop:4}
% \end{table}

% \clearpage
\section*{Abbreviations}


\begin{tabularx}{.45\textwidth}{lQ}
1 & first person\\
2 & second person\\
3 & third person\\
\scshape abs	 & absolutive\\
\scshape atr	 & advanced tongue root\\
\scshape aux	 & auxiliary\\
\scshape caus  & 	causative\\
\scshape comp & 	complementizer\\
\scshape cond  & 	conditional\\
\scshape cont  & 	continuative\\
\scshape dat	 & dative (applicative)\\
\scshape dtr & 	detransitivizer\\
\scshape f & 	feminine \\
\scshape fut & 	future\\
\scshape ideo & 	ideophone\\
\scshape II	 & class II\\
\scshape imm	 & immediate\\
\scshape imp & 	imperative\\
\scshape inf & 	infinitive\\
\scshape instr  & 	instrumental\\
\scshape ipfv & 	imperfective\\
\scshape irr  & 	irrealis\\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape itv & 	itive\\
LV & lexical verb \\
\scshape m & 	masculine\\
\scshape mid  & 	middle voice\\
\scshape n & 	neutral\\
\scshape neg & 	negative\\
\scshape nom	 & nominative\\
\scshape oblig & 	obligative\\
\scshape persis & 	persistive\\
\scshape pfv & 	perfective\\
\scshape pl & 	plural\\
\scshape pot  & 	potential\\
\scshape pro & pronoun \\
\scshape pst	 & past\\
\scshape q & 	question marker\\
\scshape red & 	reduplication\\
\scshape rel	 & relative marker\\
\scshape sub & 	subordinator\\
\scshape seq	 & sequential (or narrative)\\
\scshape sg & 	singular\\
\scshape ven	 & venitive\\
\scshape vn & 	verbal noun\\
\end{tabularx}


\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
\end{document} 
