\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233960}
\author{Jussi Ylikoski\affiliation{University of Turku, Sámi University of Applied Sciences} and Olle Kejonen\affiliation{Uppsala University}}
\title{Negation in Lule Saami}
\abstract{This chapter surveys negation in Lule Saami, a Uralic language of northern Norway and Sweden. In standard negation, a negative auxiliary verb is inflected for person, number and tense, and the lexical verb is in a non-finite connegative form. In the past tense, several competing constructions are used, encoding tense on the auxiliary or the lexical verb, or sometimes both. The negative auxiliary also has dedicated imperative forms inflected for person and number. The negation of the copula, used in stative predications, is irregular. Remnants of the former abessive case are seen in a number of derivations expressing absence, as well as in a postposition with the meaning ‘without’. The chapter further addresses other issues related to the morphology and syntax of negation, including negative indefinites and a negative converb.
\keywords{Negation, Lule Saami, negative auxiliary verb, negative derivation}
}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 

\section{The language} \label{sec:lul:1}
Lule Saami (\textit{julevsámegiella} or \textit{julevsábme}, also \textit{sámegiella} or \textit{sábme}; ISO: \href{https://iso639-3.sil.org/code/smj}{smj}, Glottocode: \href{https://glottolog.org/resource/languoid/id/lule1254}{lule1254}) is one of the westernmost languages of the westernmost branch of the Uralic language family, spoken by approximately 700 (mostly L1) speakers in a relatively large area in the northern regions of Norway and Sweden. Within the Saami languages, Lule Saami is geographically and genealogically located between North Saami in the northeast and Pite Saami in the southwest. As the language has been spoken in the vicinity of Scandinavian languages for centuries, most speakers have been bilingual in Norwegian or Swedish for generations. In the northern and eastern parts of the area, the language has also been in contact with Meänkieli, a Uralic language previously treated as a variety of Finnish, but since the year 2000 recognized as a national minority language of Sweden.

Despite being the second most spoken Saami language, Lule Saami is relatively little known outside Saami linguistics. One of the reasons is the closeness between Lule Saami and North Saami, the best known representative of the Saami branch. The language is however even more closely related to the critically endangered Pite Saami, to the degree that many of the most prominent experts on Lule Saami have regarded the significantly smaller Pite Saami as a dialect of “Lule Saami in the wide sense” (cf. \citealt[41–43ff.]{Rydving2013} for a detailed discussion). Today, however, both Lule Saami and Pite Saami have officially recognized orthographies and are acknowledged as Saami languages of Norway and Sweden along with North Saami, Ume Saami and South Saami; in Norway, Skolt Saami was also officially recognized by the Saami Parliament in 2022. Despite the low number of speakers, some children learn Lule Saami in the home, and the language is used in modern media as well as in various other domains of society, including higher education in which master’s theses on Saami issues are written in the language (e.g., \citealt{Mikkelsen2011}).

The first descriptions of Lule Saami grammar were published in the late 1800s and early 1900s (see, in particular, \citealt{Wiklund1891} and \citeyear{Wiklund1915}, first edition 1901), but the only modern general descriptions of the language are the school grammars by Spiik in Swedish (\citeyear{Spiik1989}, first edition 1977) and Nystø and Johnsen in Norwegian (\citeyear{NystøJohnsen2011}); for a grammar sketch in English, see \citet{Ylikoski2022}. Negation in Lule Saami has not been a subject of special research earlier, but the observations presented here share many features with negation in South Saami (\citealt{BloklandInaba2015}) and Skolt Saami (\citealt{MiestamoKoponen2015}). In the big picture, Lule Saami is structurally – as well as taxonomically and geographically – intermediate between these two languages on the Saami dialect continuum (see \citealt[6--38]{Sammallahti1998} on the classification of the Saami languages).

Main typological traits of the language include a medium-sized phoneme inventory, nominative-accusative alignment and a relatively free word order featuring both (S)VO and (S)OV. Inflectional morphology is characterized by extensive use of stem-internal morphophonological processes (consonant gradation and vowel alternations) in addition to suffixes. A cross-linguistically rare feature of the language, shared with several other Saami varieties, is a ternary length distinction for consonants: \textit{biese} /piesie/ ‘nests’ vs. \textit{biesse} /piessie/ ‘nest’ vs. \textit{biesse} /piesːsie/ ‘birch bark’; note however that the standard orthography does not distinguish between long /CC/ and overlong /CːC/ consonants.

This survey of negation in Lule Saami concentrates on the contemporary literary standard language, adopted in 1983 and based on the so\nobreakdash-called central dialects spoken in the municipalities of Jåhkåmåhkke/Jokkmokk (Sweden), Hábmer/Hamarøy and Áhkánjárgga/Narvik (Norway); for a brief overview of Lule Saami dialects, see \citet[21--22]{Sammallahti1998} and \citet[130--132]{Ylikoski2022}. Unless otherwise stated, examples come from the multigenre texts of Lule Saami made available by the SIKOR corpus at UiT – The Arctic University of Norway (\citealt{SIKOR2021}, about 1,810,000 words). The corpus contains texts originally written in Lule Saami, as well as translations from Norwegian, Swedish and North Saami, including a full translation of the New Testament; examples from the latter are further indicated by book, chapter and verse number. Additional examples (\ref{ex:lul:iii}--\ref{ex:lul:iv}, \ref{ex:lul:19}--\ref{ex:lul:21}, \ref{ex:lul:30}) have been constructed for the present discussion and confirmed by native speakers, while examples (\ref{ex:lul:5}, \ref{ex:lul:24}--\ref{ex:lul:25}) stem from a spoken corpus of the Lule Saami Forest dialect \citep{KejonenFC}.

The overview of derivation in  \sectref{sec:lul:3.3} draws on \citet{Kintel1991}, as well as on the dictionaries \citet{Spiik1994}, \citet{Korhonen2005} and \citet{Kintel2012}, in addition to the sources mentioned above. The graphemes ‹æ› and ‹ä›, used for /æː/ in Norway and Sweden respectively, have been retained in example clauses; in paradigms and the main text, the former is used.
Due to the largely fusional nature of Lule Saami inflection, morphemes have not been segmented in tables and examples. For reasons of space, the indicative mood and present tense of verbs are not explicitly glossed in example clauses. Similarly, nominative case and singular number are not explicitly glossed for nominals (with the exception of personal pronouns where number is always indicated). The description of negation in Lule Saami in this chapter is based on the questionnaire provided by the editors (\citetv{chapters/questionnaire}) and the structure of the chapter follows the model given in the questionnaire.

\section{Clausal negation} \label{sec:lul:2}
\subsection{Standard negation} \label{sec:lul:2.1}

In Lule Saami, standard negation is expressed by constructions consisting of the negative auxiliary and the lexical verb in a non-finite form known as the connegative. The negative auxiliary has only finite indicative and imperative forms. Disregarding the imperative mood(s) used in prohibitive sentences (\sectref{sec:lul:2.2}), the negative auxiliary in declarative sentences does not express mood, but the three other categories usually coded in the Lule Saami finite verb occur on the auxiliary: person, number (singular, dual and plural) and tense (past and present). Standard negation is thus asymmetric in the sense of \citet{Miestamo2005}, with negation affecting the finiteness of verbal elements, and the negative marker being an auxiliary verb.

The finite verbs can be divided into inflectional classes based on the number of syllables and the vowel quality of the final syllable. A sample of two verbs in the affirmative indicative is presented in \tabref{tab:lul:1} 
in which \textit{oahppat} ‘learn’ represents the most common type, the verbs undergoing consonant gradation (e.g., \textit{{}-hpp{}-} : \textit{{}-hp{}-}) and often also vowel alternation (e.g., \textit{oa{}-} : \textit{å{}-}). The so-called trisyllabic verbs such as \textit{åhpadit} ‘teach’ (causative derivation from \textit{oahppat} ‘learn’) have little stem-internal alternation, but somewhat different inflectional suffixes. Throughout this chapter, these two representative verbs are used in all tables.

\begin{table}
\begin{tabularx}{.45\textwidth}{lll}
\lsptoprule
 & \textit{oahppat} ‘learn’ & \\
 & {Present} & {Past}\\
 \midrule
{\textsc{1sg}} & {\itshape åhpav} & {\itshape oahppiv}\\
{\textsc{2sg}} & {\itshape åhpa} & {\itshape oahppi}\\
{\textsc{3sg}} & {\itshape oahppá} & {\itshape åhpaj}\\
{\textsc{1du}} & {\itshape oahppin} & {\itshape åhpajma}\\
{\textsc{2du}} & {\itshape oahppabihtte} & {\itshape åhpajda}\\
{\textsc{3du}} & {\itshape oahppaba} & {\itshape åhpajga}\\
{\textsc{1pl}} & {\itshape oahppap} & {\itshape åhpajma}\\
{\textsc{2pl}} & {\itshape oahppabihtit} & {\itshape åhpajda}\\
{\textsc{3pl}} & {\itshape oahppi} & {\itshape oahppin}\\
\lspbottomrule
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lll}
\lsptoprule
 & \textit{åhpadit} ‘teach’ & \\
 & {Present} & {Past}\\
\midrule 
{\textsc{1sg}} & {\itshape åhpadav} & {\itshape åhpadiv}\\
{\textsc{2sg}} & {\itshape åhpada} & {\itshape åhpadi}\\
{\textsc{3sg}} & {\itshape åhpat} & {\itshape åhpadij}\\
{\textsc{1du}} & {\itshape åhpadin} & {\itshape åhpadijma}\\
{\textsc{2du}} & {\itshape åhpadihppe} & {\itshape åhpadijda}\\
{\textsc{3du}} & {\itshape åhpadibá} & {\itshape åhpadijga}\\
{\textsc{1pl}} & {\itshape åhpadip} & {\itshape åhpadijma}\\
{\textsc{2pl}} & {\itshape åhpadihpit} & {\itshape åhpadijda}\\
{\textsc{3pl}} & {\itshape åhpadi} & {\itshape åhpadin}\\
\lspbottomrule
\end{tabularx}
 \caption{Indicative paradigms of the verbs \textit{oahppat} ‘learn’ and \textit{åhpadit} ‘teach’}
    \label{tab:lul:1}
\end{table}

%MM: This is the original table without the addition of the verbs on top:
%\begin{table}
%\begin{tabularx}{.45\textwidth}{lll} & {Present} & {Past}\\
%\lsptoprule
%{\textsc{1sg}} & {\itshape åhpav} & {\itshape oahppiv}\\
%{\textsc{2sg}} & {\itshape åhpa} & {\itshape oahppi}\\
%{\textsc{3sg}} & {\itshape oahppá} & {\itshape åhpaj}\\
%{\textsc{1du}} & {\itshape oahppin} & {\itshape åhpajma}\\
%{\textsc{2du}} & {\itshape oahppabihtte} & {\itshape åhpajda}\\
%{\textsc{3du}} & {\itshape oahppaba} & {\itshape åhpajga}\\
%{\textsc{1pl}} & {\itshape oahppap} & {\itshape åhpajma}\\
%{\textsc{2pl}} & {\itshape oahppabihtit} & {\itshape åhpajda}\\
%{\textsc{3pl}} & {\itshape oahppi} & {\itshape oahppin}\\
%\lspbottomrule
%\end{tabularx}
%\begin{tabularx}{.45\textwidth}{lll} & {Present} & {Past}\\
%\lsptoprule
%{\textsc{1sg}} & {\itshape åhpadav} & {\itshape åhpadiv}\\
%{\textsc{2sg}} & {\itshape åhpada} & {\itshape åhpadi}\\
%{\textsc{3sg}} & {\itshape åhpat} & {\itshape åhpadij}\\
%{\textsc{1du}} & {\itshape åhpadin} & {\itshape åhpadijma}\\
%{\textsc{2du}} & {\itshape åhpadihppe} & {\itshape åhpadijda}\\
%{\textsc{3du}} & {\itshape åhpadibá} & {\itshape åhpadijga}\\
%{\textsc{1pl}} & {\itshape åhpadip} & {\itshape åhpadijma}\\
%{\textsc{2pl}} & {\itshape åhpadihpit} & {\itshape åhpadijda}\\
%{\textsc{3pl}} & {\itshape åhpadi} & {\itshape åhpadin}\\
%\lspbottomrule
%\end{tabularx}
% \caption{Indicative paradigms of the verbs \textit{oahppat} ‘learn’ and \textit{åhpadit} ‘teach’}
%    \label{tab:lul:1}
%\end{table}

The negative counterparts of the affirmative verb forms in \tabref{tab:lul:1} are formed by using the negative auxiliary followed by the connegative forms \textit{åhpa} [learn\textsc{.cng}] and \textit{åhpada} [teach\textsc{.cng}], e.g. \textit{iv åhpa} ‘I’m not learning’ and \textit{ejga åhpada} ‘They (two) didn’t teach’ (\tabref{tab:lul:2}).\footnote{As seen in this and the following section, connegative forms like \textit{åhpa} could, strictly speaking, be characterized and glossed as indicative-cum-imperative connegatives or alternatively as non-potential connegatives, cf. \textit{åhpatja} [learn.\textsc{pot.cng}].} Further examples are given in (\ref{ex:lul:1}--\ref{ex:lul:4}) below.

\begin{table}
\begin{tabularx}{.7\textwidth}{llllX}
\lsptoprule
& {Present} & {Past} & {+} & {learn\textsc{.cng} \textsc{/} teach\textsc{.cng}}\\
\midrule
{\scshape 1sg} & {\itshape iv} & {\itshape ittjiv} &  & {\textit{åhpa} / \textit{åhpada}}\\
{\textsc{2sg}} & {\textit{i}} & {\textit{ittji}} &  & {\textit{åhpa} / \textit{åhpada}}\\
{\textsc{3sg}} & {\textit{ij}} & {\textit{ittjij}} &  & {\textit{åhpa} / \textit{åhpada}}\\
{\textsc{1du}} & {\textit{en}} & {\textit{ejma}} &  & {\textit{åhpa} / \textit{åhpada}}\\
{\textsc{2du}} & {\textit{æhppe}} & {\textit{ejda}} &  & {\textit{åhpa} / \textit{åhpada}}\\
{\textsc{3du}} & {\textit{æbá}} & {\textit{ejga}} &  & {\textit{åhpa} / \textit{åhpada}}\\
{\textsc{1pl}} & {\textit{ep}} & {\textit{ejma}} &  & {\textit{åhpa} / \textit{åhpada}}\\
{\textsc{2pl}} & {\textit{ehpit}} & {\textit{ejda}} &  & {\textit{åhpa} / \textit{åhpada}}\\
{\textsc{3pl}} & {\textit{e}} & {\textit{ettjin}} &  & {\textit{åhpa} / \textit{åhpada}}\\
\lspbottomrule
\end{tabularx}
   \caption{Indicative paradigm of the negative auxiliary, combined with connegative forms of \textit{oahppat} ‘learn’ and \textit{åhpadit} ‘teach’}
    \label{tab:lul:2}
\end{table}

For \textit{oahppat} ‘learn’ and \textit{åhpadit} ‘teach’, the connegatives \textit{åhpa} and \textit{åhpada} are identical to the \textsc{2sg} present forms (\tabref{tab:lul:1}), but for still another class of verbs such as \textit{mujttet} ‘remember’ the two forms are not identical: the connegative is \textit{mujte} whereas the \textsc{2sg} present is \textit{mujtá}. However, for all verbs except the copula the connegative is identical to the \textsc{imp.2sg} forms (see \sectref{sec:lul:2.2}).

Although the language has a relatively free word order \citep{Mikkelsen2011}, the negative auxiliary always precedes the verb in the connegative. The two verb forms often occur side by side (\ref{ex:lul:1}--\ref{ex:lul:2}), but for purposes of information structuring they can be disconnected by virtually all kinds of words and phrases (\ref{ex:lul:3}--\ref{ex:lul:4}). Coordinated verbs usually share the negative auxiliary \REF{ex:lul:4}.

\ea\label{ex:lul:1} 
    \gll Mån \textbf{iv} \textbf{mujte} suv. \\ 
         \textsc{1sg} \textsc{neg.1sg} remember.\textsc{cng} \textsc{3sg.acc} \\
    \glt  ‘I don’t remember him/her.’ \citep{SIKOR2021}
\z 

\ea\label{ex:lul:2} 
    \gll Madin oajvvefágajn barggiv, de \textbf{ittjiv} \textbf{åhpada} soames jagev. \\ 
          while master’s.degree.\textsc{com} work.\textsc{pst}.\textsc{1sg} then \textsc{neg.pst.1sg} teach.\textsc{cng} some year.\textsc{acc}\\
    \glt ‘When I was doing my master’s degree, I didn’t teach for a year or so.’ \citep{SIKOR2021}
\z 

\ea\label{ex:lul:3} 
    \gll Æjgáda \textbf{e} gielajt \textbf{åhpada} sæmmi láhkáj degu åhpadiddje skåvlån dahki. \\ 
         parent.\textsc{pl} \textsc{neg.3pl} language.\textsc{pl.acc} teach.\textsc{cng} same in.manner as teacher.\textsc{pl} school.\textsc{ine} do.\textsc{3pl}\\
    \glt  ‘Parents don’t teach languages the way the teachers at school do.’ \citep{SIKOR2021}
\z 
 
\ea\label{ex:lul:4} 
    \gll \textbf{Ij} aktak \textbf{tjále} jalik \textbf{sáhkada} åvtåláhkáj iellemájgenis. \\ 
         \textsc{neg.3sg} one.\textsc{npm} write.\textsc{cng} nor speak.\textsc{cng} similarly lifetime.\textsc{ine.3sg} \\
    \glt ‘No one writes or speaks the same way all their life.’ \citep{SIKOR2021}
\z 

Reality is, however, more complex than the picture suggested by the data in (\ref{ex:lul:1}–\ref{ex:lul:4}) and Tables \ref{tab:lul:1}–\ref{tab:lul:2}, which follow the existing modern descriptions of the language (\citealt{Spiik1989,NystøJohnsen2011}).
The formation of past negatives has been described exclusively as combinations of the negative auxiliary in the past tense and the tense-neutral connegative form (\tabref{tab:lul:2}). While Lule Saami shares this feature with all Saami languages to its south, its sister languages in the north lack all traces of a tense-inflected auxiliary. For example, North Saami as the immediate neighbor of Lule Saami resorts to the use of the past participle as a kind of past-tense connegative instead.\footnote{Within Saami linguistics and Uralic linguistics in general, the existence or lack of past-tense forms of the negative auxiliary is considered one of the most significant morphosyntactic differences that divide the Saami branch in two. The tense distinction can be regarded as inheritance from Proto-Saami and even Proto-Uralic or at least a common ancestor of Saami and Finnic, and the loss of it in the varieties northeast of Lule Saami can be considered an innovation, possibly influenced by similar development in the neighboring Finnic languages \citep[76–77, 80, 224–225]{Sammallahti1998}.}

However, past tense forms of the negative auxiliary appear to have fallen out of use also in the northern variety of Lule Saami spoken in the Unna Tjerusj reindeer herding district (Helena Omma, p.c.) and in the Forest dialect of the Jiellevárre district \citep{KejonenFC}, both in Sweden (for a map of the Lule Saami area, see \citealt[131]{Ylikoski2022}). Instead, these dialects favor the same strategy as in North Saami, i.e. marking tense on the lexical verb rather than on the negative auxiliary. An example from the Forest dialect is given in \REF{ex:lul:5}.

\ea\label{ex:lul:5} 
    \gll Mån \textbf{iv} \textbf{máhttám} mavge dárogielav. \\ 
         \textsc{1sg} \textsc{neg.1sg} know.\textsc{pst.ptcp} any\textsc{.acc.npm} Swedish.\textsc{acc} \\
    \glt ‘I didn’t know any Swedish (when starting school).’ (\citealt{KejonenFC}, transcription adapted) 
\z 

In this clause, the present-tense form \textit{iv} occurs with the past participle (or past connegative) \textit{máhttám} for a past tense reading. In the standard language, past tense would rather be encoded on the negative auxiliary, which would then occur with the lexical verb in the connegative, as in \REF{ex:lul:6}.

\ea\label{ex:lul:6} 
    \gll Vájku \textbf{ittjiv} \textbf{máhte} muohtá ja mága lunna årrot, de sunnuv álu guossidav. \\ 
         although \textsc{neg.pst.1sg} can.\textsc{cng} aunt.\textsc{gen} and uncle.\textsc{gen} at live.\textsc{inf} then \textsc{3du.acc} often visit.\textsc{1sg} \\
    \glt  ‘Although I couldn’t live with my aunt and her husband, I often visit them.’ \citep{SIKOR2021}
\z 

Constructions of the kind seen in \REF{ex:lul:5} are however by no means limited to the Unna tjerusj and Forest dialects, but occur also in the more central dialects of the language, where they coexist with tense-inflected forms of the negative auxiliary. Consider Example \REF{ex:lul:7} from the first major novel published in Lule Saami by Lars Matto \citet{Tuolja1987}, a speaker of the central dialect of Lule Saami spoken in Jåhkåmåhkke/Jokkmokk. This example contains two verb phrases with past tense interpretation, using both of the strategies described above.

\ea\label{ex:lul:7} 
    \gll Ieddnerávkke giehtoj Árjakluoven lij boadnjásj mij  \textbf{ittjij} \textbf{sidá} hiejttet noajdástallamis ja \textbf{ij} \textbf{miededam} sån lij mejdek boasstot dahkam [...] \\ 
         deceased.mother tell.\textsc{pst.3sg} Árjepluovve.\textsc{ine} be.\textsc{pst.3sg} old.man \textsc{rel} \textsc{neg.pst.3sg} want.\textsc{cng} quit.\textsc{inf} practice.sorcery.\textsc{inf} and \textsc{neg}.\textsc{3sg} admit.\textsc{pst.ptcp}  \textsc{3sg} be.\textsc{pst.3sg} any\textsc{.acc.npm} wrong do.\textsc{pst.ptcp}\\
    \glt ‘My late mother told that in Árjepluovve/Arjeplog there was an old man who didn’t want to quit sorcery and didn’t admit to having done anything wrong.’ \citep{SIKOR2021}
\z 

What is more, expressions of past negatives are not confined to the two alternatives seen above, but there are also constructions in which the “southern” strategy (\textsc{neg.pst} \textsc{+} \textsc{cng}) and the “northern” strategy (\textsc{neg} \textsc{+} \textsc{pst.ptcp}) are mixed into a hybrid where the past tense is coded both in the auxiliary and the lexical main verb (\textsc{neg.pst} \textsc{+} \textsc{pst.ptcp}), see \REF{ex:lul:8}.

\ea\label{ex:lul:8} 
    \gll Gå \textbf{ejga} \textbf{diehtám} mij jali ma båhtin, de balos tjiehkádijga suojnij sisi.\\ 
         as \textsc{neg.pst.3du} know.\textsc{pst.ptcp} \textsc{rel} or \textsc{rel.pl} come.\textsc{pst.3pl} then fear.\textsc{ela} hide.\textsc{pst.3du} hay.\textsc{pl}.\textsc{gen} into\\
    \glt ‘As they didn’t know what was or were coming, they hid themselves in the hay in fear.’ \citep{SIKOR2021}
\z 

While hybrid constructions such as the one in \REF{ex:lul:8} are not very common in the corpus, they are evidently used by some speakers, and are even sporadically attested in data on the language from more than a century ago (\citealt{KejonenFC}; cf. \citealt[273]{Wiklund1891}).

Lule Saami also makes use of analytic perfect and pluperfect tenses consisting of an auxiliary copula (inflected for present and past, respectively) and the lexical main verb in the past participle. When negated, it is the copula that appears in the connegative form while the lexical verb remains in the past participle. For example, the negative counterpart of \textit{lav oahppam} [be.\textsc{1sg} learn.\textsc{pst.ptcp}] ‘I have learned’ is \textit{iv la oahppam} [\textsc{neg}.\textsc{1sg} be.\textsc{cng} learn.\textsc{pst.ptcp}] ‘I have not learned’. However, formation of the pluperfect as well as \textsc{2sg} and \textsc{3pl} perfect forms is different from that of the general picture presented above. As the reason for this lies in the exceptional inflection of the copula verb, these forms will be described in \sectref{sec:lul:2.3} on stative predications.

After the above forms it is worthwhile to pay attention to another morphological category. As argued by \citet{Ylikoski2016b}, the so-called potential mood in Lule Saami differs from its cognates and namesakes in other Saami languages in having many semantic features in common with a future tense, rather than with an epistemic mood, as it is traditionally described. The potential forms in \textit{{}-}(\textit{t})\textit{tj{}-} : \textit{{}-sj} seldom depict merely “potential” events, but a more decisive feature of these forms is that they refer to future states of affairs, even in contexts where there is no question about the certainty of these events \REF{ex:lul:9}.

\ea\label{ex:lul:9} 
    \gll De Jubmel javlaj: “Guodo, guodo, \textbf{i} dån huoma \textbf{gallánittja}."\\ 
         then God say.\textsc{pst.3sg} graze.\textsc{imp.2sg} graze.\textsc{imp.2sg} \textsc{neg.2sg} \textsc{2sg} nevertheless get.full.\textsc{pot.cng}\\
    \glt ‘Then God said: “Graze, graze, but you nevertheless won’t get full.”’ (\citealt{SIKOR2021}, cf. \citealt[141]{Grundstrom1926})
\z 

As seen in the negated potential clause \REF{ex:lul:9}, the negative auxiliary is in the present tense indicative, and the potential is encoded only on the lexical verb, which is in a dedicated potential connegative form. \tabref{tab:lul:3} provides affirmative and negative potential forms of the verbs \textit{oahppat} ‘learn’ and \textit{åhpadit} ‘teach’.

\begin{table}

\begin{tabularx}{\textwidth}{llX ll@{}}
\lsptoprule
& \multicolumn{2}{l}{Affirmative} & \multicolumn{2}{l}{Negative} \\
\midrule
& {‘learn’}  & {‘teach’} & {\textsc{neg}} & {learn.\textsc{pot.cng} / teach\textsc{.pot.cng}}\\
{\textsc{1sg}} & {\textit{åhpatjav}} & {\textit{åhpadittjav}} & {\textit{iv}} & {\textit{åhpatja} / \textit{åhpadittja}}\\
{\textsc{2sg}} & {\itshape åhpatja} & {\textit{åhpadittja}} & {\textit{i}} & {\textit{åhpatja} / \textit{åhpadittja}}\\
{\textsc{3sg}} & {\itshape åhpasj} & {\textit{åhpadisj}} & {\textit{ij}} & {\textit{åhpatja} / \textit{åhpadittja}}\\
{\textsc{1du}} & {\itshape åhpatjin} & {\textit{åhpadittjin}} & {\textit{en}} & {\textit{åhpatja} / \textit{åhpadittja}}\\
{\textsc{2du}} & {\itshape åhpatjihppe} & {\textit{åhpadittjabihtte}} & {\textit{æhppe}} & {\textit{åhpatja} / \textit{åhpadittja}}\\
{\textsc{3du}} & {\itshape åhpatjibá} & {\textit{åhpadittjaba}} & {\textit{æbá}} & {\textit{åhpatja} / \textit{åhpadittja}}\\
{\textsc{1pl}} & {\itshape åhpatjip} & {\textit{åhpadittjap}} & {\textit{ep}} & {\textit{åhpatja} / \textit{åhpadittja}}\\
{\textsc{2pl}} & {\itshape åhpatjihpit} & {\textit{åhpadittjabihtit}} & {\textit{ehpit}} & {\textit{åhpatja} / \textit{åhpadittja}}\\
{\textsc{3pl}} & {\itshape åhpatji} & {\textit{åhpadittji}} & {\textit{e}} & {\textit{åhpatja} / \textit{åhpadittja}}\\
\lspbottomrule
\end{tabularx}
    \caption{Affirmative and negative potential forms of \textit{oahppat} ‘learn’ and \textit{åhpadit} ‘teach’}
    \label{tab:lul:3}
\end{table}


Additional examples of negated potential forms are given in (\ref{ex:lul:i}--\ref{ex:lul:ii}), using the verbs \textit{boahtet} ‘come’ and \textit{dahkat} ‘do’.

\ea\label{ex:lul:i}
\gll  \textbf{Ij} gåjt boargál uddni \textbf{båditja}.\\
    \textsc{neg.3sg} anyway shopkeeper today come.\textsc{pot.cng} \\
\glt ‘Surely the shopkeeper won’t come today, anyway.’  \citep{SIKOR2021}\\
\z

\ea\label{ex:lul:ii}
\gll \textbf{Ij} dal biernna midjij majdik \textbf{dagátja}, javlaj Máddjá.\\
    \textsc{neg.3sg} now bear \textsc{1pl.ill} any.\textsc{acc.npm} do.\textsc{pot.cng} say.\textsc{pst.3sg} Máddjá \\
    \glt ‘Surely the bear won’t do anything to us, Máddjá said.’  \citep{SIKOR2021}\\
\z 


Regardless of whether the potential is labeled as a mood without tense distinctions or a tense in itself, this category is functionally in complementary distribution with the present and past indicative, as well as with the imperative.

\subsection{Negation in non-declaratives} \label{sec:lul:2.2}

One of the most interesting – yet most poorly described – areas of Lule Saami grammar is the use of imperative forms. Traditional descriptions of the language posit two distinct imperative moods, labeled as the first and second imperative. 
The former of these inflectional categories is typically described as conveying commands or requests, whereas the second imperative (characterized by the stem-final vowel \textit{{}-u-}) is referred to as more polite or modest than the imperative proper (\citealt[279]{Wiklund1891}; \citeyear[114]{Wiklund1915}; \citealt[1851]{Grundström1946}; \citealt[103]{Spiik1989}). The actual use and functions of these categories, however, has not been subject to detailed study, and today it may well be that the second imperative is not a productive mood for some speakers of the language \citep[139]{Ylikoski2022}. For the purposes of this paper, the second imperative is glossed with the admittedly simplified label \textit{polite imperative} (\textsc{imp.pol}).

The negative imperatives consist of specialized prohibitive forms of the negative auxiliary (corresponding to a suppletive stem *\textit{alle-}) followed by the common connegative form or, in the second imperative, a dedicated connegative in \textit{{}-u}. As seen in Tables \ref{tab:lul:4}--\ref{tab:lul:5}, both connegatives are identical to their corresponding \textsc{2sg} form, e.g. \textit{åhpada} [teach.\textsc{imp.2sg} = teach.\textsc{imp.cng}].
In the typology of \citet{AuweraEtAl2013}, Lule Saami second person singular prohibitives belong to Type 2, which uses a form of the negated verb which is identical to the affirmative imperative and a sentential negative strategy that is not found in declaratives.

\begin{table}

\begin{tabularx}{\textwidth}{llX ll}
\lsptoprule
& \multicolumn{2}{l}{Affirmative} & \multicolumn{2}{l}{Negative} \\
& {‘learn’} & {‘teach’} & {\textsc{neg.imp}} & {learn.\textsc{cng} / teach.\textsc{cng}}\\
\midrule
{\textsc{1sg}} & {–} & {–} & {–} & {–} \\
{\textsc{2sg}} & {\itshape åhpa} & {\textit{åhpada}} & {\textit{ale}} & {\textit{åhpa} / \textit{åhpada}}\\
{\textsc{3sg}} & {\itshape oahppis} & {\textit{åhpadis}} & {\textit{allis}} & {\textit{åhpa} / \textit{åhpada}}\\
{\textsc{1du}} & {\itshape oahppon} & {\textit{åhpadinno}} & {\textit{allon}} & {\textit{åhpa} / \textit{åhpada}}\\
{\textsc{2du}} & {\itshape oahppe} & {\textit{åhpadihtte}} & {\textit{alle}} & {\textit{åhpa} / \textit{åhpada}}\\
{\textsc{3du}} & {\itshape oahppiska} & {\textit{åhpadiska}} & {\textit{alliska}} & {\textit{åhpa} / \textit{åhpada}}\\
{\textsc{1pl}} & {\itshape oahppop} & {\textit{åhpadip}} & {\textit{allop}} & {\textit{åhpa} / \textit{åhpada}}\\
{\textsc{2pl}} & {\itshape oahppit} & {\itshape åhpadit} & {\textit{allit}} & {\textit{åhpa} / \textit{åhpada}}\\
{\textsc{3pl}} & {\itshape oahppisa} & {\textit{åhpadisá}} & {\textit{allisa}} & {\textit{åhpa} / \textit{åhpada}}\\
\lspbottomrule
\end{tabularx}
    \caption{Affirmative and negative imperative I forms of \textit{oahppat} ‘learn’ and \textit{åhpadit} ‘teach’}
    \label{tab:lul:4}
\end{table}

\begin{table}
\begin{tabularx}{\textwidth}{llX lQ}
\lsptoprule
& \multicolumn{2}{l}{Affirmative} & \multicolumn{2}{l}{Negative} \\
& {‘learn’} & {‘teach’} & {\textsc{neg.imp.pol}} & {learn.\textsc{imp.pol.cng}/\newline  teach.\textsc{imp.pol.cng}}\\
\midrule
{\textsc{1sg}} & {–} & {–} & {–} & {–} \\
{\textsc{2sg}} & {\itshape oahppu} & {\textit{åhpadu}} & {\textit{allu}} & {\textit{oahppu} / \textit{åhpadu}}\\
{\textsc{3sg}} & {\itshape oahppus} & {\textit{åhpadus}} & {\textit{allus}} & {\textit{oahppu} / \textit{åhpadu}}\\
{\textsc{1du}} & {\itshape oahppun} & {\textit{åhpadunno}} & {\textit{allun}} & {\textit{oahppu} / \textit{åhpadu}}\\
{\textsc{2du}} & {\itshape oahppuda} & {\textit{åhpaduhtu}} & {\textit{alluda}} & {\textit{oahppu} / \textit{åhpadu}}\\
{\textsc{3du}} & {\itshape oahppuska} & {\textit{åhpaduska}} & {\textit{alluska}} & {\textit{oahppu} / \textit{åhpadu}}\\
{\textsc{1pl}} & {\itshape oahppup} & {\textit{åhpadup}} & {\textit{allup}} & {\textit{oahppu} / \textit{åhpadu}}\\
{\textsc{2pl}} & {\itshape oahpput} & {\textit{åhpaduhtut}} & {\textit{allut}} & {\textit{oahppu} / \textit{åhpadu}}\\
{\textsc{3pl}} & {\itshape oahppusa} & {\textit{åhpadusá}} & {\textit{allusa}} & {\textit{oahppu} / \textit{åhpadu}}\\
\lspbottomrule
\end{tabularx}
  \caption{Affirmative and negative imperative II forms of \textit{oahppat} ‘learn’ and \textit{åhpadit} ‘teach’}
    \label{tab:lul:5}
\end{table}

The imperative is mostly used for second person, as in the command \textit{ale duoh\-tada} ‘don’t touch’ \REF{ex:lul:10}. The second imperative forms in \textit{allu-} may be used for warnings, requests or wishes, as in \REF{ex:lul:11}.

\ea\label{ex:lul:10} 
    \gll \textbf{Ale} \textbf{duohtada} tjalmijt, njunjev jali njálmev. \\ 
         \textsc{neg}.\textsc{imp.2sg} touch.\textsc{cng} eye.\textsc{pl.acc} nose.\textsc{acc} or mouth.\textsc{acc} \\
    \glt ‘Don’t touch your eyes, nose or mouth.’  \citep{SIKOR2021}
\z 

\ea\label{ex:lul:11} 
    \gll \textbf{Allu} niejddam oadde mánáv \textbf{duohtadu}! \\ 
         \textsc{neg.imp.pol.2sg} girl\textsc{.1sg} sleep\textsc{.prs.ptcp} child.\textsc{acc} touch.\textsc{imp.pol.cng} \\
    \glt ‘Please don’t touch the sleeping child, my girl!’ (\citealt{SIKOR2021}, originally from \citealt[530]{Qvigstad1929})
\z 

In addition to imperatives in the second person, Lule Saami also has first- and third-person imperative forms (except for the first person singular).\footnote{Older descriptions of the language feature a few tokens of first person singular forms in the second imperative, including a prohibitive \textit{alluv} [\textsc{neg.imp.pol.1sg}] \citep[79]{Wiklund1915}; such forms are lacking in modern data.} First person imperatives typically have a hortative function, whereas third person imperatives may have a concessive reading (cf. \citealt[114]{Wiklund1915}). In the present corpus, however, first- and third-person prohibitives are only attested in the markedly elevated style of Bible translations (\ref{ex:lul:12}--\ref{ex:lul:13}).

\ea\label{ex:lul:12} 
    \gll \textbf{Allup} \textbf{vájbbu} buorev dahkamis. \\ 
         \textsc{neg}.\textsc{imp.pol.1pl} get.tired\textsc{.imp.pol.cng} good\textsc{.acc} do\textsc{.inf} \\
    \glt ‘Let us not grow weary in doing good.’ (\citealt{SIKOR2021}, Gal. 6:9)
\z 

\ea\label{ex:lul:13} 
    \gll Guhti ittjij birratjuohpaduvá, \textbf{allus} \textbf{birratjuohpaduvvu}. \\ 
         who \textsc{neg.pst.3sg} be.circumcised.\textsc{cng} \textsc{neg.imp.pol.3sg} be.circumcised.\textsc{imp.pol.cng} \\
    \glt ‘Whoever was not circumcised, let him not be circumcised.’ (\citealt{SIKOR2021}, 1 Cor. 7:18)
\z 

Unlike lexical verbs such as \textit{oahppat} ‘learn’ and \textit{åhpadit} ‘teach’, the copula has distinct connegative forms both in the first imperative (\textit{liege {\textasciitilde} åro}) and in the second imperative (\textit{lehku {\textasciitilde} årru}), cf. the negation of the copula in the indicative and potential (\sectref{sec:lul:2.3}). The forms \textit{liege} and \textit{lehku} are more common in Norway, whereas the suppletive forms \textit{åro} and \textit{årru} (from \textit{årrot} ‘live, reside’) are preferred in Sweden (cf. \citealt[69]{Spiik1989}). Examples of the use of the connegative forms \textit{liege} and \textit{lehku} are given in \REF{ex:lul:14} and \REF{ex:lul:15}.

\ea\label{ex:lul:14} 
    \gll \textbf{Ale} \textbf{liege} håjen. \\ 
         \textsc{neg.imp.2sg} be.\textsc{imp.cng} sad \\
    \glt ‘Don’t be sad.’ \citep{SIKOR2021}
\z 

\ea\label{ex:lul:15} 
    \gll Lábttje l mieskas; \textbf{allu} gievrra \textbf{lehku}! \\ 
         rein be.\textsc{3sg} rotten \textsc{neg.imp.pol.2sg} strong be.\textsc{imp.pol.cng} \\
    \glt  ‘The rein is rotten; please don’t be strong!’ (\citealt{SIKOR2021}, originally from \citealt[544]{Qvigstad1929})
\z 

Polar questions are usually marked by the question particle (or clitic) \textit{gus} ({\textasciitilde} \textit{gu} {\textasciitilde} \textit{=k}) following the initial word of a polar interrogative clause \REF{ex:lul:16}. However, at times the mere inverted word order is enough to make the verb-initial clause appear as a question \REF{ex:lul:17}. On a more general level, both features are identical to the way how affirmative polar questions are formed in Lule Saami (cf. \citealt[95]{Spiik1989}; \citealt[145]{Ylikoski2022}), and therefore negative polar question do not substantially differ from affirmative ones (cf. \citealt[373]{MiestamoKoponen2015} for the analogous clause types in Skolt Saami).

\ea\label{ex:lul:16} 
    \gll \textbf{I} \textbf{gus} dasti \textbf{mujte}? \\ 
          \textsc{neg.2sg} \textsc{q} anymore remember.\textsc{cng} \\
    \glt ‘Don’t you remember anymore?’ \citep{SIKOR2021}
\z 

\ea\label{ex:lul:17} 
    \gll \textbf{Ähppe} \textbf{la} dåj vuoddjám? \\ 
         \textsc{neg.2du} be.\textsc{cng} \textsc{2du} drive.off.\textsc{pst.ptcp} \\
    \glt  ‘Haven’t you driven off yet?’ \citep{SIKOR2021}
\z 

\subsection{Negation in stative predications} \label{sec:lul:2.3}

In Lule Saami, stative predications show a syntactic structure similar to the verbal clauses described in the preceding sections. The all-purpose verb for stative clauses is \textit{liehket} ‘be’ (also with a suppletive infinitive \textit{årrot}), the most frequent but also the most irregular verb of the language. The most remarkable exceptions to regular conjugation are seen in the negative forms this verb. There is no common connegative for present and past, but instead, the present-tense forms of the negative auxiliary are followed by \textit{la}, which must be labeled as a specifically present-tense connegative of the copula, whereas the past-tense forms of the negative auxiliary are combined with \textit{lim}, a specific past-tense connegative. As a result, the present vs. past distinction is marked on both parts of the predicate, but only in the dual (all persons) and plural (first and second person); for \textsc{1sg} and \textsc{3sg}, the present-tense forms \textit{iv} and \textit{ij} are used instead of \textit{ittjiv} and \textit{ittjij} (see \tabref{tab:lul:2} above).

What is more, in the second person singular and third person plural, the present forms of the negative auxiliary have fused with the connegative into amalgamated forms, such as \textit{illa} ‘(you) aren’t’, from \textit{i la} ‘(you) are not’ (bolded in \tabref{tab:lul:6}).\footnote{The \textsc{3sg} form \textit{l} ‘is’ in \tabref{tab:lul:6} is a postvocalic counterpart of \textit{la}. The copula also has parallel forms in \textit{le-} : \textit{læ-}, such as \textit{lev} ‘I am’ \citep[70]{Spiik1989}. Note further that the prescribed \textsc{3pl} potential negative \textit{ællittja} is also written \textit{ællitja} in the corpus (cf. \citealt[218]{Ylikoski2016b}).} Potentially, such forms could be treated as general stative negators (cf. \citealt[362–363]{MiestamoKoponen2015} on similar forms in Skolt Saami), but seeing that they are transparently composed of the negative auxiliary and the connegative, and that they occur in the same paradigm as non-fused forms, they are treated here as functionally identical to their non-amalgamated counterparts.

Examples of positive and negative predications of class inclusion and identity are given in (\ref{ex:lul:iii}--\ref{ex:lul:iv}).

\ea\label{ex:lul:iii}
    \ea
    \gll Tjáhppa l bena.\\
        Tjáhppa be.\textsc{3sg} dog\\
    \glt ‘Tjáhppa is a dog.’\\
    \ex 
    \gll Tjáhppa ij la bena. \\
        Tjáhppa \textsc{neg.3sg} be.\textsc{cng} dog\\
    \glt ‘Tjáhppa isn’t a dog.’ (constructed examples)
\z \z 

\ea\label{ex:lul:iv}
    \ea
    \gll Bena l Tjáhppa.\\
        dog be.\textsc{3sg} Tjáhppa\\
    \glt ‘The dog is Tjáhppa.’\\
    \ex 
    \gll Bena ij la Tjáhppa.\\
        dog \textsc{neg.3sg} be.\textsc{cng} Tjáhppa \\
    \glt ‘The dog isn’t Tjáhppa.’ (constructed examples)
\z \z 

\begin{table}
\small 
\begin{tabularx}{\textwidth}{lXlXlXl@{~}}
\lsptoprule
& \multicolumn{2}{c}{{Present indicative}} & \multicolumn{2}{c}{{Past indicative}} & \multicolumn{2}{c}{{Potential}}\\
%\midrule 
& {Affirmative } & {Negative} & {Affirmative } & {Negative} & {Affirmative } & {Negative}\\
\midrule 
{\scshape 1sg} & {\itshape lav} & {\textit{iv la}} & {\itshape lidjiv}  & {\textit{iv lim}} & {\itshape littjav} & {\textit{iv littja}}\\
{\textsc{2sg}} & {\itshape la} & {\bfseries\itshape illa} & {\itshape lidji}  & {\bfseries\itshape illim} & {\itshape littja} & {\bfseries\itshape illittja}\\
{\textsc{3sg}} & {\textit{la}, \textit{l}} & {\itshape ij la} & {\itshape lij} & {\itshape ij lim} & {\itshape lisj} & {\itshape ij littja}\\
{\textsc{1du}} & {\itshape lin}  & {\itshape en la} & {\itshape lijma}  & {\itshape ejma lim} & {\itshape littjin} & {\itshape en littja}\\
{\textsc{2du}} & {\itshape lihppe}  & {\itshape æhppe la} & {\itshape lijda}  & {\itshape ejda lim} & {\itshape littjabihtte} & {\itshape æhppe littja}\\
{\textsc{3du}} & {\itshape libá}  & {\itshape æbá la} & {\itshape lijga}  & {\itshape ejga lim} & {\itshape littjaba} & {\itshape æbá littja}\\
{\textsc{1pl}} & {\itshape lip}  & {\itshape ep la} & {\itshape lijma}  & {\itshape ejma lim} & {\itshape littjap} & {\itshape ep littja}\\
{\textsc{2pl}} & {\itshape lihpit}  & {\itshape ehpit la} & {\itshape lijda}  & {\itshape ejda lim} & {\itshape littjabihtit} & {\itshape ehpit littja}\\
{\textsc{3pl}} & {\itshape li} & {\bfseries\itshape ælla} & {\itshape lidjin} & {\bfseries\itshape ællim} & {\itshape littji} & {\bfseries\itshape ællittja}\\
\lspbottomrule
\end{tabularx}
  \caption{Inflectional paradigm of the copula in the present, past and potential}
    \label{tab:lul:6}
\end{table}

As regards the past-tense connegative \textit{lim}, it is worth noting that although Examples \REF{ex:lul:5}, \REF{ex:lul:7} and \REF{ex:lul:8} in \sectref{sec:lul:2.1} demonstrated the use of the past participle in non-standard past-tense predicates, \textit{lim} is a specialized past-tense connegative that is different from the past participle \textit{læhkám} ({\textasciitilde} \textit{årrum}) – the only of its kind in the language. In \REF{ex:lul:18}, the negated pluperfect contains both forms at the same time.

\ea\label{ex:lul:18} 
    \gll Jåvnni \textbf{ij} \textbf{lim} goassak nav ávon \textbf{læhkám}. \\ 
         Jåvnni \textsc{neg.3sg} be.\textsc{pst.cng} when.\textsc{npm} so joy.\textsc{ine} be.\textsc{pst.ptcp} \\
    \glt ‘Jåvnni had never been so happy.’ \citep{SIKOR2021}
\z 

The verb \textit{liehket} ‘be’ is used in all major types of stative clauses, such as those of locational \REF{ex:lul:19}, existential \REF{ex:lul:20} and possessive predication \REF{ex:lul:21}. 
These types of clauses are identical in terms of argument marking, although they may differ from each other as regards word order. As seen in these examples, locations and possessors are in the inessive case, while themes and possessees that function as subject arguments are in the nominative.


\ea\label{ex:lul:19} 
    \ea\label{ex:lul:19a} 
        \gll Bena l sjaljon.\\ 
          dog be.\textsc{3sg} yard.\textsc{ine}  \\
    \glt ‘The dog is in the yard.’ 
    \ex\label{ex:lul:19b} 
        \gll Bena ij la sjaljon. \\ 
          dog \textsc{neg.3sg} be.\textsc{cng} yard.\textsc{ine} \\
    \glt ‘The dog isn’t in the yard.’ (constructed examples)
    \z
\z 

\ea\label{ex:lul:20}
    \ea\label{ex:lul:20a}
        \gll Sjaljon la bena. \\ 
             yard.\textsc{ine} be.\textsc{3sg} dog  \\
        \glt  ‘There’s a dog in the yard.’ 
    \ex\label{ex:lul:20b}
        \gll Sjaljon ij la bena.  \\ 
             yard.\textsc{ine} \textsc{neg.3sg} be.\textsc{cng} dog \\
        \glt ‘There isn’t a dog in the yard.’  (constructed examples)
    \z
\z 
 
\ea\label{ex:lul:21}
    \ea\label{ex:lul:21a}
        \gll Sikkán la bena.  \\ 
             Siggá\textsc{.ine} be.\textsc{3sg} dog  \\
        \glt ‘Siggá has a dog.’
    
    \ex\label{ex:lul:21b}
        \gll Sikkán ij la bena. \\ 
             Siggá\textsc{.ine} \textsc{neg.3sg} be.\textsc{cng} dog \\
        \glt ‘Siggá doesn’t have a dog.’  (constructed examples)
    \z
\z 

As was mentioned in \sectref{sec:lul:2.1}, the exceptional inflection of the copula verb also affects the formation of negated perfect and pluperfect forms of other predicates: instead of the expected negative auxiliary such as \textit{i} [\textsc{neg.2sg}] and the following copula in the connegative such as \textit{la} [be.\textsc{cng}] (cf. \textit{i åhpa} ‘you don’t learn’ in \tabref{tab:lul:2}), the lexical main verb is preceded by the amalgamated form \textit{illa} \REF{ex:lul:22}.

\ea\label{ex:lul:22} 
    \gll “Jus dån \textbf{illa} \textbf{mannam}, {iv} {la} mån {mannam} danna,” javllá riebij.\\ 
         if \textsc{2sg} be.\textsc{neg.2sg} go.\textsc{pst.ptcp} \textsc{neg.1sg} be.\textsc{cng} \textsc{1sg} go.\textsc{pst.ptcp} there say.\textsc{3sg} fox\\
    \glt ‘“If you haven’t been there, I haven’t been there,” the fox says.’ (\citealt{SIKOR2021}, originally from \citealt[472]{Qvigstad1929})
\z 

The fusion between the negative auxiliary and the copula is also reflected in the fact that the copula seems to be the only verb that may appear between the negative auxiliary and the question particle \textit{gus} \REF{ex:lul:23}; for other verbs, the question particle occurs in second position, immediately after the negative auxiliary (cf. \REF{ex:lul:16} in \sectref{sec:lul:2.2}).

\ea\label{ex:lul:23} 
    \gll \textbf{Ij} \textbf{la} \textbf{gus} geldulasj? \\ 
         \textsc{neg.3sg} be.\textsc{cng} \textsc{q} exciting \\
    \glt  ‘Isn’t it exciting?’ \citep{SIKOR2021}
\z 

Finally, it can be mentioned that in the critically endangered Lule Saami Forest dialect, amalgamated forms of the negative auxiliary and the copula can in the plural take regular person suffixes, effectively resulting in what resembles a negative copula. Examples of this phenomenon are given in \REF{ex:lul:24}, where the suffix \textit{{}-bit} encodes \textsc{2pl} present, and in \REF{ex:lul:25}, where \textit{{}-n} encodes \textsc{3pl} past.

\ea\label{ex:lul:24} 
    \gll Dij \textbf{ällebit} skihpasa. \\ 
         \textsc{2pl} be.\textsc{neg.2pl} sick.\textsc{pl} \\
    \glt ‘You aren’t sick.’ (\citealt{KejonenFC}, transcription adapted)
\z 

\ea\label{ex:lul:25} 
    \gll Da \textbf{ellin} boatjojhoajdon. \\ 
         that\textsc{.pl} be.\textsc{neg.pst.3pl} reindeer.herding.\textsc{ine} \\
    \glt  ‘They weren’t in reindeer herding.’ (\citealt{KejonenFC}, transcription adapted)
\z 

Forms as the ones given in \REF{ex:lul:24} and \REF{ex:lul:25} are however not attested in other varieties of Lule Saami, and from the point of view of the standard language, they remain an interesting but local exception in the negation of the copula.

\subsection{Negation in non-main clauses} \label{sec:lul:2.4}

When combining clauses, Lule Saami makes use of both finite and non-finite subordinate clauses. Finite subordinate clauses are more frequent and more variegated than non-finite ones, and the functions of finite dependent clauses are expressed by a number of subordinating conjunctions and complementizers as well as relative pronouns introducing relative clauses. The internal structure of such finite clauses does not differ from that of main clauses, and negation works the same way in finite subordinate clauses as it does in finite main clauses.
Examples are given in (\ref{ex:lul:v}--\ref{ex:lul:vi}), where subordinate clauses are introduced by the relative pronoun \textit{gut} ‘who’ and the conjunction \textit{vaj} ‘so that’, respectively.

\ea\label{ex:lul:v}
\gll Mån lav dåssju sahte ulmusj \textbf{gut} \textbf{iv} majdik 
\textbf{máhte} divvot.\\
    \textsc{1sg} be.\textsc{1sg} only ordinary person who \textsc{neg.1sg} any.\textsc{acc.npm} can.\textsc{cng} fix.\textsc{inf} \\
\glt ‘I am but a simple man who cannot improve anything.’  \citep{SIKOR2021} \\
\z

\ea\label{ex:lul:vi}
    \gll Na manen de æddnáj tsabmi, \textbf{vaj} måj Jåvvåjn \textbf{ejma} \textbf{galga} gullat. \\
        well why then mother.\textsc{ill} whisper.\textsc{pst.2sg} so.that \textsc{1du} Jåvvå.\textsc{com} \textsc{neg.pst.1du} will.\textsc{cng} hear.\textsc{inf}\\
    \glt ‘Well why then did you whisper to mom, so that Jåvvå and I would not hear?’  \citep{SIKOR2021}\\
\z 

However, the language also has about ten non-finite verb forms that can be characterized as infinitives, participles and converbs. One of them is the non-finite in \textit{{}-k} ({\textasciitilde} \textit{{}-dak}), traditionally called the “verb abessive” in Saami linguistics, but better labeled as a negative converb. This form is most often used as an adverbial non-finite with the vague meaning that can be characterized as negative concomitance, often preceded by the preposition-like particle \textit{váni} ‘without’ \REF{ex:lul:26}; this particle is however not obligatory \REF{ex:lul:27}. As many distinctions between affirmative categories are often neutralized in negation, the Lule Saami negative converb is an example of this cross-linguistically common phenomenon; the converb in question can be used to negate the meanings expressed with the instrumental-concomitant and temporal converbs in the language (cf. \citealt[38ff.]{Ylikoski2009} for North Saami).

\ea\label{ex:lul:26} 
    \gll Danna sån lij juo váni \textbf{oadek} tjåhkkåham gålmmå jánndura, \textbf{båråk} ja \textbf{jugák}.\\ 
         there \textsc{3sg} be.\textsc{pst}.\textsc{3sg} already without sleep.\textsc{neg.cvb} sit.\textsc{pst.ptcp} three.\textsc{acc} day.and.night.\textsc{gen} eat.\textsc{neg.cvb} and drink.\textsc{neg.cvb} \\
    \glt  ‘There he had sat, without sleeping, three days and nights already, without eating and drinking.’ \citep{SIKOR2021}
\z 

\ea\label{ex:lul:27} 
    \gll Vehtsit nubbe nuppev \textbf{njimmurik}. \\ 
         be.hospitable.\textsc{imp.2pl} \textsc{recp} \textsc{recp.acc} complain.\textsc{neg.cvb} \\
    \glt ‘Show hospitality to one another without complaining.’ (\citealt{SIKOR2021}, 1 Peter 4:9)
\z 

As seen in \REF{ex:lul:28}, the negative converb may take an object, such as the accusative \textit{mavga} ‘anything’. Notably, the agent of the converb may also be encoded by means of the genitive case, such as \textit{Áhtjáda} ‘your Father’s’ in \REF{ex:lul:29}; see also \citet[110]{Spiik1989}.

\ea\label{ex:lul:28} 
    \gll Gålmmå tijma maŋŋela bådij áhkkás sisi váni \textbf{mavga} \textbf{diedek}. \\ 
         three.\textsc{acc} hour.\textsc{gen} later come.\textsc{pst} wife.\textsc{3sg} in without any\textsc{.acc.npm} know\textsc{.neg.cvb}\\
    \glt  ‘Three hours later his wife came in without knowing anything.’ (\citealt{SIKOR2021}, Acts 5:7)
\z 

\ea\label{ex:lul:29} 
    \gll Valla dajs aktak ij ednamij gahtja váni \textbf{Áhtjáda} \textbf{diedek}.\\ 
         but that\textsc{.pl.ela} one.\textsc{npm} \textsc{neg.3sg} earth.\textsc{ill} fall.\textsc{cng} without father.\textsc{gen.2pl} know.\textsc{neg.cvb}  \\
    \glt  ‘But none of them will fall to the ground without your Father’s knowing it.’ (\citealt{SIKOR2021}, Matt. 10:29)
\z 

The same form is also used as a general negative participle comprising both
present and past as well as active and passive functions. In phrases like \textit{vállduk ålmåj} ‘unmarried man’ (\ref{ex:lul:vii}) and \textit{basák giedaj} ‘with unwashed hands’ (\ref{ex:lul:viii}), the non-finites can be analyzed as non-finite relative clauses (participials), and they can always be replaced by finite relative clauses.

\ea\label{ex:lul:vii}
\gll Sån lij \textbf{vállduk} \textbf{ålmåj}, ja skåvlån åroj.\\
    \textsc{3sg} be.\textsc{pst.3sg} marry.\textsc{neg.cvb} man and school.\textsc{ine} live.\textsc{pst.3sg}\\
\glt ‘He was an unmarried man, and he lived at the school.’  \citep{SIKOR2021}\\
\z 

\ea\label{ex:lul:viii}
\gll Bårråmus \textbf{basák} \textbf{giedaj} ij almatjav báde.\\
    meal wash.\textsc{neg.cvb} hand.\textsc{pl.com} \textsc{neg.3sg} man.\textsc{acc} soil.\textsc{cng}\\
\glt ‘Eating with unwashed hands does not defile a man.’ (\citealt{SIKOR2021}, Matt. 15:20)\\
\z 

On the use of the so-called second infinitive in -\textit{mis} with \textit{váni} ‘without’, see \REF{ex:lul:45} in \sectref{sec:lul:3.3}. 


\section{Non-clausal negation} \label{sec:lul:3}
\subsection{Negative replies} \label{sec:lul:3.1}

When a polar question is followed by a negative reply, the reply may in principle consist of a full negated sentence \REF{ex:lul:30a}, but a more unmarked way is to reply with the negated verb only \REF{ex:lul:30b}, and the least marked reply consists of the mere negative auxiliary, which is inflected in person, number and tense \REF{ex:lul:30c}.

\ea\label{ex:lul:30} 
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:lul:30q}
        \gll Vuojnij gus Siggá muv? \\ 
             see.\textsc{pst.3sg} \textsc{q} Siggá \textsc{1sg.acc} \\
        \glt  ‘Did Siggá see me?’
    \end{xlist}
    \ea\label{ex:lul:30a}
        \gll Siggá ittjij duv vuojne. \\ 
             Siggá \textsc{neg}.\textsc{pst.3sg} \textsc{2sg.acc} see\textsc{.cng} \\
        \glt ‘Siggá didn’t see you.’
    \ex\label{ex:lul:30b}
        \gll Ittjij vuojne. \\
             \textsc{neg}.\textsc{pst.3sg} see\textsc{.cng} \\
        \glt  ‘No, she didn’t.’
    \ex\label{ex:lul:30c}
        \gll Ittjij. \\
             \textsc{neg}.\textsc{pst.3sg} \\
        \glt ‘No.’ (constructed examples)
    \z
\z 

In colloquial discourse, however, the simplest type of reply is often the borrowed negative particle \textit{næ} ‘no’ (cf. Norwegian \textit{næ}, Swedish \textit{nä} ‘no’). The meaning of a negative reply is independent from the polarity of the question, and therefore the replies in \REF{ex:lul:30} would have the same interpretation even if the question was formulated with negation (e.g., \textit{Ittjij gus Siggá vuojne muv?} ‘Didn’t Siggá see me?’).

Older descriptions of the language (\citealt[185]{Wiklund1891}; \citealt[19]{Collinder1938}) further mention an emphatic form \textit{ihij} of the \textsc{3sg} present negative auxiliary \textit{ij}, which can be used in negative replies (cf. Finnish \textit{ehei}, the emphatic analogue of \textit{ei} \textsc{neg.3sg}, \citealt[658–659]{Penttilä1963}, as well as Swedish \textit{nähä}, from the above-mentioned \textit{nä} ‘no’). While \textit{ihij} and similar emphatic disyllabic forms are used also in contemporary spoken Lule Saami (Torgny Hedström, p.c.), such forms are not attested in the corpus.

\subsection{Negative indefinites and quantifiers} \label{sec:lul:3.2}

Lule Saami has no negative indefinites or quantifiers with negative force in themselves, but instead negative indefinites are composed of interrogative{}-relative and indefinite pronouns and adverbs followed by the clitic =\textit{k}, which functions as a negative polarity marker (see \sectref{sec:lul:4.2}); these forms then occur together with the negative auxiliary. For example, words like \textit{akta} ‘one’, \textit{goabbá} ‘which (of two)’ and \textit{goassa} ‘when’ yield the third person singular forms \textit{ij} \textit{aktak} ‘no one’, \textit{ij goabbák} ‘neither (of two)’ and \textit{ij} \textit{goassak} ‘never’. For monosyllabic words such as the accusative \textit{mav} ‘what\textsc{’}, the negative polarity marker has a form \textit{=ga} {\textasciitilde} \textit{=ge}, yielding e.g. \textit{ij mavga} ‘nothing’.

Third person singular forms as the ones above are however not the only possibility, but in clauses with an indefinite the negative auxiliary agrees with the subject in person and number, as seen in the first person dual \textit{en} \REF{ex:lul:31}. The auxiliary is also inflected for mood and tense; an imperative \textit{ale} is given in \REF{ex:lul:32}, and a past tense form \textit{ittjij} is provided in \REF{ex:lul:33}. As seen in \REF{ex:lul:34}, the scope of negation expressed in a main clause is able to extend to a subordinate clause without overt marking of verbal negation.

\ea\label{ex:lul:31} 
    \gll Måj \textbf{en} suv \textbf{goassak} gávna. \\ 
         \textsc{1du} \textsc{neg.1du} \textsc{3sg.acc} when.\textsc{npm} find.\textsc{cng} \\
    \glt ‘We’ll never find her/him.’ \citep{SIKOR2021}
\z 
 
\ea\label{ex:lul:32} 
    \gll Härrá, dunji dårvustaláv, \textbf{ale} muv \textbf{goassak} guode! \\ 
         master \textsc{2sg.ill} rely.\textsc{1sg} \textsc{neg.imp.2sg} \textsc{1sg.acc} when.\textsc{npm} leave.\textsc{cng} \\
    \glt ‘Lord, I put my faith in you, don’t ever leave me!’ \citep{SIKOR2021}
\z 

\ea\label{ex:lul:33} 
    \gll Kátján \textbf{ij} lim \textbf{aktak} ráddna ja \textbf{ittjij} ga hálijda \textbf{avtajnik} ståhkat.\\ 
         Kátjá.\textsc{ine} \textsc{neg.3sg} be.\textsc{pst.cng} one.\textsc{npm} friend and \textsc{neg.pst.3sg} \textsc{npm} want.\textsc{cng} one.\textsc{com.npm} play.\textsc{inf}\\
    \glt ‘Kátjá didn’t have any friends and didn’t want to play with anyone either.’ \citep{SIKOR2021}
\z 

\ea\label{ex:lul:34} 
    \gll \textbf{Ij} le galla \textbf{goassak} åvddåla gullum \textbf{aktak} tjalmijt rahpá \textbf{gässtak} guhti le tjalmedibmen riegádam.\\ 
         \textsc{neg.3sg} be.\textsc{cng} indeed when.\textsc{npm} before be.heard.\textsc{pst.ptcp} one.\textsc{npm} eye.\textsc{pl.acc} open.\textsc{3sg} who.\textsc{ela.npm} who be.\textsc{3sg} blind.\textsc{ess} be.born.\textsc{pst.ptcp}\\
    \glt  ‘Never earlier has it been heard that anyone opened the eyes of someone who was born blind.’ (\citealt{SIKOR2021}, John 10:32)
\z 

Generally, the negative auxiliary precedes an indefinite, as in (\ref{ex:lul:31}--\ref{ex:lul:34}). The opposite order of constituents is however also possible, although this appears to be most common in religious texts: consider the phrase \textit{aktak ij} ‘no one’ in \REF{ex:lul:35} and the synonymous \textit{ij aktak} in \REF{ex:lul:4} in \sectref{sec:lul:2.1}; an additional example is found in \REF{ex:lul:29} in \sectref{sec:lul:2.4}.

\ea\label{ex:lul:35} 
    \gll \textbf{Aktak} \textbf{ij} buvte guokta härrá dievnnut. \\ 
         one.\textsc{npm} \textsc{neg.3sg} can.\textsc{cng} two.\textsc{acc} master.\textsc{gen} serve.\textsc{inf} \\
    \glt ‘No one can serve two masters.’ (\citealt{SIKOR2021}, Matt. 6:24)
\z 

The first school grammar of Lule Saami mentions the possibility of a negative auxiliary in the present tense to occur twice in clauses with a negative indefinite, i.e. a case of negative concord (\citealt[116]{Wiklund1915}; cf. \citealt[52--53]{Blokland2022}). While this phenomenon is common in the neighboring North Saami language (cf. \citealt[207]{Svonni2018}), the present corpus contains no clear instances of such constructions in Lule Saami.

\subsection{Negative derivation} \label{sec:lul:3.3}
%mmbkm3
Lule Saami has a relatively rich morphology, and negation is also coded in derivation in many ways. In fact, a lexicalized instance of such morphology was seen in \REF{ex:lul:34} above where the adjective \textit{tjalmedibme} ‘blind’ is actually a morphologically transparent derivative in which the noun \textit{tjalmme} ‘eye’ is followed by the derivative suffix \textit{{}-dibme} ‘{}-less’. Just like the derivative for ‘eyeless’ is used for ‘blind’, \textit{biellje} ‘ear’ has yielded \textit{bieljedibme} ‘deaf’ (“earless”), and \textit{giella} ‘language’ in turn \textit{gieladibme} ‘mute’ (“languageless”). Semantically less opaque examples include \textit{namádibme} ‘nameless; anonymous’ (\leftarrow\space \textit{namma} ‘name’) and \textit{biebmodibme} ‘foodless’ (\leftarrow\space \textit{biebbmo} ‘food’).

The derivational suffix \textit{{}-dibme} – and its attributive, adnominal counterpart \textit{\nobreakdash-dis} in phrases like \textit{namádis tjálle} ‘anonymous writer’ – is attached to disyllabic nouns, the most common type of nouns in the language. So-called trisyllabic stems, such as \textit{gahper} ‘hat’, may take an entirely different suffix, \textit{{}-ahtes}, with the identical function of creating denominal negative adjectives like \textit{gahperahtes} ‘hatless’, \textit{jubmelahtes} ‘godless’ and \textit{miejnigahtes} ‘meaningless’ (\leftarrow\space \textit{jubmel} ‘god’, \textit{miejnik} ‘meaning’).

The suffix \textit{{}-ahtes} is also used for other types of negative derivatives. The adjectives \textit{ájlis} ‘holy’ and \textit{rievtuk} ‘fair; honest’ yield the antonyms \textit{ájlisahtes} ‘unholy’ and \textit{rievtugahtes} ‘unfair; dishonest’.
Verbs derived with this suffix include \textit{ánssidahtes} ‘undeserved, unearned’ (\leftarrow\space \textit{ánssidit} ‘deserve, earn’) and \textit{dádjadahtes} ‘uncomprehending’ (\leftarrow\space \textit{dádjadit} ‘understand’).
There are also adjectives like \textit{gållnåmahtes} ‘unwithering’ and \textit{nåhkåmahtes} ‘never-ending, unending’, corresponding to the verbs \textit{gållnåt} ‘wither’ and \textit{nåhkåt} ‘end’ (or, rather, to the nominalizations \textit{gållnåm} ‘withering’ and \textit{nåhkåm} ‘ending’). When based on transitive verbs, such forms usually have a passive reading ‘un-V-able’, for example \textit{tjielggimahtes} ‘unexplainable’ and \textit{vuojttemahtes} ‘undefeatable’ (cf. \textit{tjielggit} ‘explain’, \textit{vuojttet} ‘win’).
The suffix -\textit{dihke} derives adjectives from verbs, e.g. \textit{sárnodihke} ‘unspeakable’ (\leftarrow\space  \textit{sárnnot} ‘say, tell’) and \textit{sáŋardihke} ‘regretless’ (\leftarrow\space  \textit{sáŋardit} ‘regret’).

Lule Saami also has a possibility to turn nouns to caritive verbs with the meaning ‘lose N’. The suffix \textit{{}-duvva{}-} can be used to create verbs like \textit{fámoduvvat} ‘lose one’s strength’ (\leftarrow\space \textit{fábmo} ‘strength’) in \REF{ex:lul:36} and \textit{mieladuvvat} ‘lose one’s mind’ (\leftarrow\space \textit{miella} ‘mind’) in \REF{ex:lul:37}. The suffix \textit{{}-duhtte{}-} yields analogous causative verbs such as \textit{ráfeduhttet} ‘make lose one’s tranquility; unsettle’ (\leftarrow\space  \textit{ráfe} ‘peace’) in \REF{ex:lul:38}.

\ea\label{ex:lul:36} 
    \gll Dat rájes \textbf{fámoduváj} ja gahtjaj. \\ 
         that\textsc{.ela} since lose.strength.\textsc{pst.3sg} and fall.\textsc{pst.3sg} \\
    \glt ‘From that on, he lost his strength and fell down.’  \citep{SIKOR2021}
\z 

\ea\label{ex:lul:37} 
    \gll Peter iesj nav balláj váj \textbf{mieladuváj}. \\ 
         Peter \textsc{refl} so get.scared.\textsc{pst.3sg} so.that lose.mind.\textsc{pst.3sg} \\
    \glt ‘Peter himself got so scared that he lost his mind.’  \citep{SIKOR2021}
\z 

\ea\label{ex:lul:38} 
    \gll Vuojnnin juoga lij dav \textbf{ráfeduhttám}. \\ 
         see.\textsc{pst.3pl} something be.\textsc{pst.3sg} that\textsc{.acc} unsettle.\textsc{pst.ptcp} \\
    \glt  ‘They saw that something had unsettled it.’  \citep{SIKOR2021}
\z 

There are further a number of denominal adverbs with privative meaning. Their morphology is quite varied, which can be explained as a result of the relatively recent loss and restructuring of the abessive case, an ancient Uralic morphological case with the meaning ‘without’. Suffixes of abessive-like adverbs include \textit{{}-ahtá}, \textit{{}-ahti}, \textit{{}-dak}, \textit{{}-dagá} and \textit{{}-dagi}, although the latter two forms are usually better analyzed as postpositions. On the other hand, \textit{{}-ahtá} can sometimes be interpreted as an adverbial counterpart of the suffix \textit{{}-ahtes} seen above: words like \textit{gahperahtes} ‘hatless’ are often used in adnominal positions such as \textit{gahperahtes máná} ‘hatless children’, whereas forms like \textit{gahperahtá} ‘without hat’ \REF{ex:lul:39} are used as adverbials.

\ea\label{ex:lul:39} 
    \gll Ferasuollujn máná ettjin oattjo årrot \textbf{gahperahtá}, ietján de guovsagisá boalldin sijás vuobdajt.\\ 
         Faroe.Islands.\textsc{pl.ine} child.\textsc{pl} \textsc{neg.pst.3pl} be.allowed.\textsc{cng} be.\textsc{inf} without.hat otherwise then aurora.\textsc{pl} burn.\textsc{pst.3pl} \textsc{3pl.ela} hair.\textsc{pl.acc}\\
    \glt  ‘In the Faroe Islands children were not allowed to be without hats, lest the northern lights burn their hair.’  \citep{SIKOR2021}
\z 

Adjectives like \textit{namádibme} ‘nameless’, based on disyllabic nouns like \textit{namma} ‘name’ may have adverbial counterparts like \textit{namádak} \REF{ex:lul:40}, but these are more like exceptions from the normal situation in which sequences like \textit{namá dagi} \REF{ex:lul:41} are interpreted and written as postpositional phrases in which \textit{dagi} ‘without’ governs the genitive (see below).

\ea\label{ex:lul:40} 
    \gll Le gus \textbf{namádak}? \\ 
         be.\textsc{3sg} \textsc{q} without.name \\
    \glt ‘Doesn’t she have a name?’   \citep{SIKOR2021}
\z 

\largerpage
\ea\label{ex:lul:41} 
    \gll Andreassen almodij avijsav \textbf{namá} \textbf{dagi}, ja iesj vargga gájkka artihkkalijt tjálij. \\ 
         Andreassen publish.\textsc{pst.3sg} newspaper.\textsc{acc} name.\textsc{gen} without and \textsc{refl} almost all article.\textsc{pl.acc} write.\textsc{pst.3sg} \\
    \glt ‘Andreassen published the newspaper anonymously, and wrote almost all articles by himself.’  \citep{SIKOR2021}
\z 

In addition to the relatively productive suffixal derivations described above, Lule Saami also has a somewhat productive derivational prefix, the borrowed \textit{iehpe{}-} ‘un{}-, dis-’ (cf. Finnish/Meänkieli \textit{epä{}-} id.). This morpheme can be added to adjectives and nouns, e.g. \textit{iehpeluondulasj} ‘unnatural’ (\leftarrow\space \textit{luondulasj} ‘natural’), \textit{iehpeformálalasj} ‘informal’ (\leftarrow\space \textit{formálalasj} ‘formal’) and \textit{iehpejáhkko} ‘disbelief’ (\leftarrow\space \textit{jáhkko} ‘belief’). It can also be added to at least one non-finite verb form, the passive past participle in \textit{{}-}(\textit{d})\textit{um} (on this form, see \citealt{KejonenFC}), as in \textit{iehpeorganiseridum} ‘unorganized’ (\leftarrow\space \textit{organiseridum} ‘organized’).

There is also a more marginal borrowed prefix \textit{u-} ‘un-, dis-’ (cf. Swedish \textit{o{}-}, Norwegian \textit{u-} id.). This element occurs in a small number of borrowed adjectives, such as \textit{urievtesferduk} ‘unrighteous’ (cf. Swedish \textit{orättfärdig}, Norwegian \textit{urettferdig} id.), but it is also attested with native stems, such as \textit{ugullogis} ‘disobedient’ (\leftarrow\space \textit{gullogis} ‘obedient’); the latter word coexists with a synonym \textit{gullogahtes}, derived with the above-mentioned suffix \textit{{}-ahtes}. The prefix \textit{u{}-} is also found on a few nouns, such as \textit{urievtesvuohta} ‘misdeed’ and the plural form \textit{umáná} ‘naughty children’ (lit. ‘unchildren’, \citealt{Grundström1946} see entry \textit{umānāh}).

The borrowed prefix \textit{vádna-} ‘dis-’ (cf. Swedish/Norwegian \textit{van-} id.) can be added to nouns and adjectives to form privatives such as \textit{vádnajáhkko} ‘disbelief’ (cf. the synonym \textit{iehpejáhkko} above) and \textit{vádnagieres} ‘loveless’ (\leftarrow\space \textit{gieres} ‘beloved’), although such forms are not very common.

In recent neologisms, the least marked negative auxiliary form \textit{ij} [\textsc{neg.3sg}] is used as a prefix for ‘non-, in-’ added to adjectives or nouns, e.g. \textit{ij-formálalasj} ‘informal’ (\leftarrow\space \textit{formálalasj} ‘formal’; cf. the synonym \textit{iehpeformálalasj} above), \textit{ij-gierddisvuohta} ‘intolerance’ (\leftarrow\space \textit{gierddisvuohta} ‘tolerance’). This prefix is also attested with the passive past participle, e.g. \textit{ij-hiebadum} ‘non-adapted’ (\leftarrow\space \textit{hiebadum} ‘adapted’).

A summary of the derivations outlined above is given in Tables \ref{tab:lul:i}--\ref{tab:lul:ii}. As seen in these tables, derivations using suffixes often produce forms belonging to other parts of speech than the base from which they are derived. Derivations using prefixes, on the other hand, retain the word class of the base.

\begin{table}
    \centering\small 

\begin{tabularx}{\textwidth}{p{18mm}lXl}
\lsptoprule
Suffix  & Derives  & Example & Base\\
        & X from Y&&\\
\midrule
{}-\textit{ahtá}, -\textit{ahti} & Adv \leftarrow\space N & \textit{gahperahtá} ‘without hat’ & \leftarrow\space \textit{gahper} ‘hat’\\
{}-\textit{ahtes} & Adj \leftarrow\space N & \textit{gahperahtes} ‘hatless’ & \leftarrow\space \textit{gahper} ‘hat’\\
& Adj \leftarrow\space Adj & \textit{ájlisahtes} ‘unholy’ & \leftarrow\space \textit{ájlis} ‘holy’\\
& Adj \leftarrow\space V & \textit{dádjadahtes} ‘uncomprehending’ & \leftarrow\space \textit{dádjadit} ‘understand’\\
{}-\textit{dak}, -\textit{dagá}, \newline -\textit{dagi} & Adv \leftarrow\space N & \textit{namádak} ‘without name’ & \leftarrow\space \textit{namma} ‘name’\\
{}-\textit{dibme} & Adj \leftarrow\space N & \textit{namádibme} ‘nameless’ & \leftarrow\space \textit{namma} ‘name’\\
\textit{{}-dihke} & Adj \leftarrow\space V & \textit{sárnodihke} ‘unspeakable’ & \leftarrow\space \textit{sárnnot} ‘say’\\
{}-\textit{duhtte}{}- & V \leftarrow\space N & \textit{ráfeduhttet} ‘unsettle’ & \leftarrow\space \textit{ráfe} ‘peace’\\
{}-\textit{duvva}{}- & V \leftarrow\space N & \textit{mieladuvvat} ‘lose one’s mind’ & \leftarrow\space \textit{miella} ‘mind’\\
\lspbottomrule
\end{tabularx}
\caption{Negative derivations using suffixes}
    \label{tab:lul:i}
\end{table}

\begin{table}
    \centering\small 
\begin{tabularx}{\textwidth}{llXl}
\lsptoprule
Prefix & Part of  & Example & Base\\
        & speech && \\
\midrule
\textit{iehpe}{}- & Adj & \textit{iehpeluondulasj} ‘unnatural’ & \leftarrow\space \textit{luondulasj} ‘natural’\\
& N & \textit{iehpejáhkko} ‘disbelief’ & \leftarrow\space \textit{jáhkko} ‘belief’\\
& V & \textit{iehpeorganiseridum} ‘unorganized’ & \leftarrow\space \textit{organiseridum} ‘organized’\\
\textit{ij}{}- & Adj & \textit{ij-formálalasj} ‘informal’ & \leftarrow\space \textit{formálalasj} ‘formal’\\
& N & \textit{ij-gierddisvuohta} ‘intolerance’ & \leftarrow\space \textit{gierddisvuohta} ‘tolerance’\\
& V & \textit{ij-hiebadum} ‘non-adapted’ & \leftarrow\space \textit{hiebadum} ‘adapted’\\
\textit{u}{}- & Adj & \textit{ugullogis} ‘disobedient’ & \leftarrow\space \textit{gullogis} ‘obedient’\\
& N & \textit{urievtesvuohta} ‘misdeed’ & \leftarrow\space \textit{rievtesvuohta} ‘right’\\
\textit{vádna}{}- & Adj & \textit{vádnagieres} ‘loveless’ & \leftarrow\space \textit{gieres} ‘beloved’\\
& N & \textit{vádnajáhkko} ‘disbelief’ & \leftarrow\space \textit{jáhkko} ‘belief’\\
\lspbottomrule
\end{tabularx}
\caption{Negative derivations using prefixes}
    \label{tab:lul:ii}
\end{table}





As mentioned above, \textit{namá dagi} ‘without a name’ \REF{ex:lul:41} is a postpositional phrase, and indeed, \textit{dagi} ({\textasciitilde} \textit{dagá}) is the default Lule Saami adposition for ‘without’. From a diachronic perspective, the postposition is a rare instance of a former inflectional suffix –~the Proto-Saami abessive case ending – which has been degrammaticalized into a postposition \citep{Ylikoski2016a}. In other words, the morpheme is no longer suffixed to a preceding noun, but rather it occurs as a free word governing the genitive, the case typical of other complements of adpositions in the language. The non-suffixal nature of this former affix is perhaps most clearly seen in cases of conjunction reduction, as in \REF{ex:lul:42}.

\ea\label{ex:lul:42} 
    \gll Dievnastiddje galggá gåtsedit, gájkka dáhpáduvvá  \textbf{ságastallama} \textbf{ja} \textbf{stuojme} \textbf{dagi} báhtjaj gaskan.\\ 
         servant shall.\textsc{3sg} supervise.\textsc{inf} everything happen.\textsc{3sg} discussion.\textsc{gen} and fuss.\textsc{gen} without boy.\textsc{pl.gen} between \\
    \glt ‘The servant shall take care that everything is done without discussion or fuss between the boys.’ \citep{SIKOR2021}
\z 

Postpositional phrases with \textit{dagi} ‘without’ may often be preceded by the reinforcing preposition-like particle \textit{váni} ‘without’ \REF{ex:lul:43}, see, for instance, the similar use of \textit{váni} in non-finite negative clauses with the converb in \textit{-k} (\sectref{sec:lul:2.4}).
Occasionally, however, \textit{váni} also occurs as an undisputed preposition without \textit{dagi} \REF{ex:lul:44}. Moreover, the so-called second infinitive in \textit{{}-mis} (also known as the ``action elative'') may also occur as a complement of \textit{váni}, as seen in \REF{ex:lul:45}; unlike the negative converb, however, this form does not have an inherently negative reading, but can be used also in other functions.

\ea\label{ex:lul:43} 
    \gll Ienemusát buorrán váhkon \textbf{váni} \textbf{dálkudime} \textbf{dagi}. \\ 
         mostly get.well\textsc{.3sg} week.\textsc{ine} without medication.\textsc{gen} without \\
    \glt ‘Usually it gets better in a week without medication.’ \citep{SIKOR2021}
\z 
 
\ea\label{ex:lul:44} 
    \gll Sån la mánájgárden barggam, ja åhpadiddjen \textbf{váni} \textbf{åhpadiddjeoahpo}.\\ 
         \textsc{3sg} be.\textsc{3sg} kindergarten.\textsc{ine} work.\textsc{pst.ptcp} and teacher.\textsc{ess} without teacher.education.\textsc{gen}\\
    \glt ‘She has worked in a kindergarten, and as a teacher without a degree.’ \citep{SIKOR2021}
\z 

\ea\label{ex:lul:45} 
    \gll Gatjádav \textbf{váni} \textbf{vásstádusáv} \textbf{vaddemis}. \\ 
         ask.\textsc{1sg} without answer.\textsc{acc} give.\textsc{inf} \\
    \glt ‘I’m asking without giving an answer.’ \citep{SIKOR2021}
\z 

In the Forest dialect, a borrowed preposition \textit{ilman} {\textasciitilde} \textit{ilmán} ‘without’ is also attested, cf. Finnish/Meänkieli \textit{ilman}  ‘without’ (\citealt[121, 123]{Kejonen2022}; cf. \citealt{Grundström1946} see entry \textit{ilmā}).

\newpage
\section{Other aspects of negation} \label{sec:lul:4}
\subsection{The scope of negation} \label{sec:lul:4.1}

Lule Saami has a relatively free word order. From a historical-geographical perspective, the language has an intermediate position between North Saami (SVO) and South Saami (SOV): both (S)OV and (S)VO are used to the extent that neither clearly dominates \citep{Mikkelsen2011}. Even more variation is possible for focusing purposes, and this is one of the main means to narrow or emphasize the scope of negation. In spoken language, elements that are focused by giving them sentence-initial position also receive more prominent stress, as if to underline the focused constituent. Fronting like this simultaneously emphasizes the status of the constituent within the scope of negation \REF{ex:lul:46}.

\ea\label{ex:lul:46} 
    \gll \textbf{Goabddáj} \textbf{árvov} mij \textbf{ep} uddni ållåsit dádjada. \\ 
         ceremonial.drum.\textsc{pl.gen} value.\textsc{acc} \textsc{1pl} \textsc{neg.1pl} today entirely understand.\textsc{cng} \\
    \glt ‘As regards ceremonial drums, we don’t entirely understand their value today.’ \citep{SIKOR2021}
\z 

The scope of negation can also be emphasized and narrowed to specific constituents by elements such as \textit{valla} ‘but’ in contrastive constructions of the type “not A but B” \REF{ex:lul:47}.

\ea\label{ex:lul:47} 
    \gll SOV sjaddá bissot, \textbf{ij} \textbf{vuodobáhkogárggán} \textbf{valla} \textbf{báhkogárggán} gånnå le sierra dæddo værbbaj.\\ 
         SOV will.\textsc{3sg} remain.\textsc{inf} \textsc{neg.3sg} basic.word.order.\textsc{ess} but word.order.\textsc{ess}  where be.\textsc{3sg} special stress verb.\textsc{ill}  \\
    \glt ‘SOV will remain, not as the basic word order, but as a word order where the verb is emphasized.’ \citep{SIKOR2021}
\z

Unfortunately, systematic descriptions of information structuring as well as discoursal and focusing functions of prosody in Lule Saami are non-existent and must be left for future studies.

\subsection{Negative polarity} \label{sec:lul:4.2}
As seen in \sectref{sec:lul:3.2}, negative indefinites usually consist of interrogative and indefinite elements followed by the element =\textit{k} ({\textasciitilde} \textit{=ga} {\textasciitilde} =\textit{ge}), which functions as a negative polarity marker. This element also occurs as a separately written particle \textit{ga} ({\textasciitilde} \textit{ge}) in negation clauses in the meaning ‘(n)either’ (\ref{ex:lul:48}--\ref{ex:lul:49}), yet without an explicit negative power in itself; see also \sectref{sec:lul:4.4} on the use of \textit{ga} for reinforcing negation.

\ea\label{ex:lul:48} 
    \gll \textbf{Ittjiv} \textbf{ga} \textbf{diede} jus iehtja dasi rat jáhkij. \\ 
         \textsc{neg.pst.1sg} \textsc{npm} know.\textsc{cng} if \textsc{refl} that\textsc{.ill} really believe.\textsc{pst.3sg} \\
    \glt ‘Neither did I know whether he really believed in it himself.’ \citep{SIKOR2021}
\z 

\ea\label{ex:lul:49} 
    \gll Mån iv máhte tjállet ja \textbf{iv} \textbf{ga} låhkåt \textbf{máhte} sámegiellaj [...]\\ 
         \textsc{1sg} \textsc{neg}.\textsc{1sg} can.\textsc{cng} write.\textsc{inf} and \textsc{neg}.\textsc{1sg} \textsc{npm} read.\textsc{inf} can.\textsc{cng} Saami.language.\textsc{ill}\\
    \glt ‘I can’t write and I cannot read in Saami either.’ \citep{SIKOR2021}
\z 

Words like \textit{ájn} ‘still’ and \textit{vil {\textasciitilde} vilá} ‘still’ remain unchanged when occurring in negation clauses, where their meaning is ‘(not) yet’ – compare the affirmative \REF{ex:lul:50} and the negative \REF{ex:lul:51}.

\ea\label{ex:lul:50} 
    \gll Ällosujtto le \textbf{vilá} ájnas sáme kultuvran. \\ 
         reindeer.herding be.\textsc{3sg} still important Saami culture.\textsc{ine} \\
    \glt ‘Reindeer herding is still important in Saami culture.’ \citep{SIKOR2021}
\z 

\ea\label{ex:lul:51} 
    \gll Dárojduhttem ij la \textbf{vilá} jábmám. \\ 
         assimilation \textsc{neg.3sg} be.\textsc{cng} yet die.\textsc{pst.ptcp} \\
    \glt ‘The assimilation (Norwegianization/Swedification of the Saami people) has not yet died.’ \citep{SIKOR2021}
\z 

Lule Saami also has a number of words that appear to be restricted to negative clauses only. One of these words is \textit{rat} ‘(not) really; (not) especially’ in \REF{ex:lul:48} above; others include \textit{dasti} ({\textasciitilde} \textit{desti}) ‘anymore; (no) longer (of time)’ (see  \REF{ex:lul:16} in \sectref{sec:lul:2.3}), \textit{heva} ‘(not) especially good, well’ and \textit{vuojga} ‘(not) especially; (not) that much’.

\subsection{Marking of NPs in the scope of negation} \label{sec:lul:4.3}

Negation does not affect marking of NPs in any way, see (\ref{ex:lul:19}--\ref{ex:lul:21}) in \sectref{sec:lul:2.3}.

\subsection{Reinforcing negation} \label{sec:lul:4.4}

Many negative polarity items mentioned in \sectref{sec:lul:3.2} and \sectref{sec:lul:4.2} may have a function of reinforcing negation. An example is given in \REF{ex:lul:52}, in which \textit{ga} gives special emphasis to the immediately preceding constituent. The reinforcing function of \textit{ga} is even more apparent in \REF{ex:lul:53}, where the same element appears to be repeated superfluously.

\ea\label{ex:lul:52} 
    \gll Ja ij la dal diedon, buohtastahttema diehti, nav hálbbe \textbf{Vuonan} \textbf{ga} viessot. \\ 
         and \textsc{neg.3sg} be.\textsc{cng} now of.course comparison.\textsc{gen} for so cheap Norway.\textsc{ine} \textsc{npm} live.\textsc{inf} \\
    \glt ‘And for comparison, neither is it that inexpensive to live in Norway, of course.’ \citep{SIKOR2021}
\z 

\ea\label{ex:lul:53} 
    \gll Ij lim rávgga, \textbf{ij} \textbf{lim} \textbf{ga} \textbf{hávgga} \textbf{ga}. \\ 
         \textsc{neg.3sg} be.\textsc{pst.cng} ghost \textsc{neg.3sg} be.\textsc{pst.cng} \textsc{npm} pike \textsc{npm} \\
    \glt ‘It wasn’t a ghost, nor was it a pike, either.’ \citep{SIKOR2021}
\z 

Other negation-reinforcing elements not yet mentioned include the particles \textit{åbbå} and \textit{åvvå} with the meaning ‘whole; wholly; entire(ly)’ in affirmative contexts, but ‘(not) even; (not) at all’ in negative clauses. The related forms \textit{åbbånis} and \textit{åvvånis} \REF{ex:lul:54} as well as the particle \textit{luoddnok} ‘under no circumstances’ \REF{ex:lul:55} are used exclusively in negative clauses.

\ea\label{ex:lul:54} 
    \gll \textbf{Ælla} vuojn dåppe \textbf{åvvånis} sierga jalik skierre. \\ 
         be.\textsc{neg.3pl} you.see there at.all willow.\textsc{pl} nor dwarf.birch.\textsc{pl} \\
    \glt ‘There are no willows or dwarf birches there at all, you see.’ \citep{SIKOR2021}
\z 

\ea\label{ex:lul:55} 
    \gll \textbf{Iv} \textbf{luoddnok} dasi guorrasa. \\ 
          \textsc{neg.1sg} under.no.circumstances that\textsc{.ill} agree.\textsc{cng} \\
    \glt  ‘By no means do I agree.’ \citep{SIKOR2021}
\z 

On the emphatic \textsc{3sg} present form \textit{ihij} of the negative auxiliary, see \sectref{sec:lul:3.1}.

\subsection{Negation, coordination and complex clauses} \label{sec:lul:4.5}

In clause combining, negation does not differ from that in simple clauses, except for the negative converb described in \sectref{sec:lul:2.4}. However, although the particle \textit{ga} ‘(n)either’ discussed above (\sectref{sec:lul:4.2} and \sectref{sec:lul:4.4}) may be interpreted as somewhat redundant if accompanied by the general coordinator \textit{ja} ‘and’ (see \REF{ex:lul:49} above), alternatively it may also occur without \textit{ja} and has in fact a function of a coordinator \REF{ex:lul:56}.

\ea\label{ex:lul:56} 
    \gll Indiánaj giela ælla almulasj giela \textbf{e} \textbf{ga} oajválattja dajt goassak ane.\\ 
          Indian.\textsc{pl.gen} language.\textsc{pl} be.\textsc{neg.3pl} official language.\textsc{pl} \textsc{neg.3pl} \textsc{npm} authority.\textsc{pl} that\textsc{.pl.acc} when.\textsc{npm} use.\textsc{cng}\\
    \glt ‘The languages of the Indians are not official languages, and authorities never use them either.’ \citep{SIKOR2021}
\z 

Lule Saami also has a special corrective coordinator \textit{ájnat} ‘but instead’ (\ref{ex:lul:57}--\ref{ex:lul:58}), which appears to be replaceable by the general adversative coordinator \textit{valla} ‘but’ (see \REF{ex:lul:47} in \sectref{sec:lul:4.1}), but the former is used only in connection with contrastive negation, following a negative proposition expressed by the preceding clause (cf., e.g., German \textit{sondern}, Swedish \textit{utan} and Finnish \textit{vaan}).

\ea\label{ex:lul:57} 
    \gll Ale guoktálatta, \textbf{ájnat} jáhke! \\ 
         \textsc{neg.imp.2sg} doubt.\textsc{cng} but.instead believe.\textsc{imp.2sg} \\
    \glt ‘Don’t doubt, but believe instead!’ (\citealt{SIKOR2021}, John 20:27)
\z 

\ea\label{ex:lul:58} 
    \gll Mij ep sidá kultuvrramujtton årrot, \textbf{ájnat} oassen viesso kultuvras. \\ 
         \textsc{1pl} \textsc{neg.1pl} want.\textsc{cng} cultural.heritage.\textsc{ess} be.\textsc{inf}  but.instead part.\textsc{ess} live.\textsc{prs.ptcp} culture.\textsc{ela} \\
    \glt ‘We don’t want to be a cultural heritage, but a part of a living culture.’  \citep{SIKOR2021}
\z 

Moreover, the language also has a negative disjunctive conjunction, \textit{jalik} ‘nor’, already seen in \REF{ex:lul:4} and \REF{ex:lul:54} above. This particle is transparently composed of \textit{jali} ‘or’ and the element =\textit{k} seen in negative indefinites and quantifiers (\sectref{sec:lul:3.2}).

\subsection{Miscellaneous aspects of negation}\label{sec:lul:4.6}

Negative transport is possible with cognitive verbs such as \textit{jáhkket} ‘believe’, but the semantic and pragmatic limits of the phenomenon have not been properly studied. In \REF{ex:lul:59}, the negation in the main clause is interpreted as negating the subordinate predicate \textit{gåjkki} ‘dry’.

\ea\label{ex:lul:59} 
    \gll Iv jáhke de uddni gåjkki. \\ 
         \textsc{neg.1sg} believe.\textsc{cng} then today dry.\textsc{3pl} \\
    \glt ‘I don’t believe that they will dry today.’ (= ‘I believe that they won’t dry today.’) \citep{SIKOR2021}
\z 

\section{Summary} \label{sec:lul:5}

The above sections have provided a general description of negation in Lule Saami. \tabref{tab:lul:8} presents a synopsis of the main features of the phenomena discussed in this chapter.

\begin{table}
    \centering\footnotesize
\begin{tabularx}{\textwidth}{lp{32mm}X}
\lsptoprule
{\bfseries Form} & \textbf{Morphosyntactic} \textbf{status} & {\bfseries Function}\\
\midrule
\textit{ij} etc. & negative auxiliary & Used with a connegative verb in standard negation (\sectref{sec:lul:2.1}), polar questions (\sectref{sec:lul:2.2}), stative predications (\sectref{sec:lul:2.3}), finite subordinate clauses (\sectref{sec:lul:2.4}) and used with or without a connegative verb in negative replies (\sectref{sec:lul:3.1})\\
\textit{ij}{}- & derivational prefix & Derives negative adjectives, nouns and verbs (\sectref{sec:lul:3.3})\\
\textit{ale} etc. & negative auxiliary & Used with a connegative verb in imperative (prohibitive) clauses (\sectref{sec:lul:2.2})\\
{}-\textit{k}, -\textit{dak} & negative converb & Used in non-finite subordinate clauses and as a negative participle (\sectref{sec:lul:2.4})\\
{\itshape næ} & negative particle ‘no’ & Used in negative replies (\sectref{sec:lul:3.1})\\
\textit{ga}, \textit{ge} & negative polarity marker & Used in the meaning ‘(n)either’ (\sectref{sec:lul:4.2}), as a negative reinforcer (\sectref{sec:lul:4.4}) and as a negative coordinator (\sectref{sec:lul:4.5})\\
=\textit{k}, =\textit{ga}, =\textit{ge} & negative polarity marker & Forms negative indefinites and quantifiers when used with the negative auxiliary (\sectref{sec:lul:3.2})\\
{}-\textit{ahtá}, -\textit{ahti} & derivational suffix & Derives negative adverbs from nouns (\sectref{sec:lul:3.3})\\
{}-\textit{ahtes} & derivational suffix & Derives negative adjectives from nouns, adjectives and verbs (\sectref{sec:lul:3.3})\\
\textit{dagi}, \textit{dagá} & postposition ‘without’ & Used with the genitive of nominals (\sectref{sec:lul:3.3})\\
{}-\textit{dak}, \nobreakdash-\textit{dagá}, \nobreakdash-\textit{dagi} & derivational suffix & Derives negative adverbs from nouns (\sectref{sec:lul:3.3})\\
{}-\textit{dibme} & derivational suffix & Derives negative adjectives from nouns (\sectref{sec:lul:3.3})\\
{\itshape {}-dihke} & derivational suffix & Derives negative adjectives from verbs (\sectref{sec:lul:3.3})\\
{}-\textit{duhtte}{}- & derivational suffix & Derives negative verbs from nouns (\sectref{sec:lul:3.3})\\
{}-\textit{duvva}{}- & derivational suffix & Derives negative verbs from nouns (\sectref{sec:lul:3.3})\\
\textit{iehpe}{}- & derivational prefix & Derives negative adjectives, nouns and verbs (\sectref{sec:lul:3.3})\\
\textit{u}{}- & derivational prefix & Derives negative adjectives and nouns (\sectref{sec:lul:3.3})\\
\textit{vádna}{}- & derivational prefix & Derives negative adjectives and nouns (\sectref{sec:lul:3.3})\\
{\itshape váni} & preposition, particle ‘without’ & Used with the genitive of nominals and second infinitive of verbs (\sectref{sec:lul:3.3})\\
{\itshape ájnat} & coordinator ‘but instead’ & Used in adversative coordination (\sectref{sec:lul:4.5})\\
{\itshape jalik} & coordinator ‘nor’ & Used in disjunctive coordination (\sectref{sec:lul:4.5})\\
\lspbottomrule
\end{tabularx}
\caption{Summary of Lule Saami negation}
    \label{tab:lul:8}
\end{table}


% \begin{table}[]%old table
% \small 
% \begin{tabularx}{\textwidth}{lX}
% \lsptoprule
% {\bfseries Clausal negation} & \\
% {Standard negation} & {Negative auxiliary (\textit{ij} \textsc{neg.3sg} etc.) + connegative verb}\\
% {Prohibitives} & {Negative (prohibitive) auxiliary (\textit{ale} \textsc{neg.imp.2sg} etc.) + connegative verb}\\
% {Negation in stative predications} & {Negative auxiliary + connegative verb (copula \textit{la}/\textit{lim})}\\
% {Negation in non-main clauses} & {Negative auxiliary + connegative verb}\\
% & {Negative converb (\textit{{}-k})}\\
% \midrule 
% {\bfseries Non-clausal negation} & \\
% {Negative replies} & {Elliptical negative clauses; \textit{næ} ‘no’}\\
% {Negative indefinites and quantifiers} & {Indefinites and quantifiers + clitic =\textit{k} (and negative auxiliary)}\\
% {Negative derivation and case marking} & {Suffixes and prefixes for various parts of speech}\\
% {Other negative expressions} & {Adpositions (\textit{dagi} ‘without’, \textit{váni} ‘without’)}\\
% \midrule 
% {\bfseries Other aspects of negation} & \\
% {The scope of negation} & {Emphasis by word order}\\
% {Negative polarity} & {Particle \textit{ga}}\\
% {Reinforcing negation} & {Particle \textit{ga}}\\
% {Negation and coordination} & {Coordinators, including \textit{jalik} ‘nor’ and the corrective coordinator \textit{ájnat} ‘(not A) but instead (B)’}\\
% \lspbottomrule
% \end{tabularx}
%   \caption{Summary of Lule Saami negation}
%     \label{tab:lul:8}
% \end{table}



\section*{Acknowledgments}

We wish to thank Torgny Hedström, Inga Lill Sigga Mikkelsen, Helena Omma and Siri Nystø Ráhka for enlightening discussions regarding Lule Saami in general and negation in particular.
This research was partially funded by the Research Council of Finland (Ylikoski, Project 349818).


\section*{Abbreviations}

\begin{tabularx}{.45\textwidth}{lQ}
\textsc{1} & first person \\
\textsc{2} & second person \\
\textsc{3} & third person \\
\textsc{acc} & accusative \\
\textsc{cng} & connegative \\
\textsc{com} & comitative \\
\textsc{cvb} & converb \\
\textsc{du} & dual \\
\textsc{ela} & elative \\
\textsc{ess} & essive \\
\textsc{gen} & genitive \\
\textsc{ill} & illative \\
\textsc{imp} & imperative \\
\textsc{ine} & inessive \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\textsc{inf} & infinitive \\
\textsc{neg} & negative (auxiliary) \\
\textsc{npm} & negative polarity marker \\
\textsc{pl} & plural \\
\textsc{pol} & polite \\
\textsc{pot} & potential \\
\textsc{prs} & present \\
\textsc{pst} & past \\
\textsc{ptcp} & participle \\
\textsc{q} & question particle \\
\textsc{recp} & reciprocal \\
\textsc{refl} & reflexive \\
\textsc{rel} & relative \\
\textsc{sg} & singular \\
\end{tabularx}

\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
\end{document} 
