\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233774}
\author{Aurore Montébran\affiliation{CNRS-LLACAN}}
\title{Negation in Mankanya}
\abstract{This chapter investigates the field of negation in Mankanya, a Bak language (Atlantic group) spoken in Guinea-Bissau, Senegal and Gambia. The analysis follows the framework developed by Miestamo and Veselinova, which allows to study negation from a typological point of view. Like the other Atlantic languages, Manka\-nya has a very rich verbal morphology, and negation tends to be integrated into the TAM-system. Negation in Mankanya shows two major types of asymmetry: the first one concerns the marking of TAM, and the second concerns the marking of subject. Negation also involves a specific intonation pattern. This chapter provides a systematic study of negation, analysing the correspondences between affirmative and negative clauses, as well as the interaction between TAM and negation.

\keywords{Negation, verbal morphology, Atlantic, Bak}
}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 
% \mm{@LV: Don't send to author yet! This is a paper that got quite a lot of critique from the reviewers, so we should also check whether the reviewers' comments have been taken into account properly and still discuss the status of this paper. A couple of things that were criticized by the reviewers and that I noted might still be problematic: We still learn very little about the negative intonation, is it enough? And there is still confusion about clausal vs. verb-internal structure regarding the schematic presentations.}
%\lv{@MM: we should ask Aurore to send a report on which comments she addressed and which she didn't. She should do the same with the comments we left to her after the reviews came in.}
\section{The language}\label{sec:mank:1}

Mankanya (ISO: \href{https://iso639-3.sil.org/code/knf}{knf}-{SEN}, Glottocode: \href{https://glottolog.org/resource/languoid/id/mank1251}{mank1251}),  is a Bak language from the Atlantic family (Niger-Congo phylum). Originally spoken in Guinea-Bissau, it is nowadays spoken from the north of Guinea-Bissau to the south of Senegal (in Casamance); there are also some large communities around the Senegalese cities of Dakar and Thiès, and in Gambia. In 2016, there were about 50,000 speakers in Guinea-Bissau, and around 85,200 speakers in totality \citep{EberhardEtAl2019}.

The first overview of the Mankanyan grammar has been provided by \citet{Trifcovič1969}. Then, \citet{Maliman2005} and \citet{Bayepar2013} have led phonological and morphological studies. Recently, \citet{Gaved2007,Gaved2015,Gaved2020} produced the most complete analysis on Mankanya, and \citet{Montébran2021} included Mankanya in a comparative study of negation in the Atlantic languages with regard to tense aspect and mood (TAM). 

There are two main varieties of Mankanya; this chapter focuses on the Hula variety, named after the city Bula (Guinea-Bissau) where it was originally spoken. The varieties of Mankanya (Hula and Co) are mutually intelligible; the differences are principally phonological and related to a small amount of lexemes. It is noteworthy that Mankanya has been strongly influenced by the local Portuguese Creole (see \citealt{Bayepar2013} for more details), but also by Mandinka (Mande) and Wolof (Atlantic), depending on the area.

%The Mankanya people are mostly farmers, organised in patriarchal families. The huge majority is Christian with some remnant of animism beliefs. The Manka\-nya people are still strongly attached to their traditions and are governed by a king living in Bula.\lv{to Yvonne: you suggest to delete this paragraph. I think it's nice to have; Mankanya is not known to the general public. On paperhive Aurore said yes to deletion. I can still delete it.}

Although a written official standard has been set up for Mankanya in Senegal since 2005, the language remains mostly spoken and the majority of speakers do not manage the writing system. Moreover, Mankanya is only used in restricted domains (within the family, but not in professional life). French is prominent at school, and Wolof is used as a vehicular language in Senegal; English in Gambia and in Guinea-Bissau the vehicular language is the Portuguese Creole. Thus, young people leaving villages for the city to get a job are slowly losing the use of the language. This phenomenon is also accelerated with exogamous marriages. 

Mankanya is an agglutinative language. The word order (SVO) is very strict and the modifiers follow the head word. It is a non-tonal language, even though the prosody has an important role including in the TAM system. The verb obligatorily agrees with its subject (the agreement marker is prefixed); if the subject is not an NP, the agreement marker can be the sole subject. Similarly to other Atlantic languages, Mankanya has very rich and complex verbal morphology. The verbal system includes three moods (indicative, injunctive, and irrealis) and an infinitive form of the verb (marked by a prefix \textit{p}{}-). Within the indicative mood, the main opposition concerns the perfective and imperfective aspects. These two aspects are mostly distinguished by the intonation pattern. Concerning the tenses: present is unmarked whereas past and future are marked by auxiliaries. The structure of the TAM system massively employs auxiliary verbs, which appear in serial verb clauses. In these structures, where several verbs are sharing the same subject (typically the auxiliary verb and the main verb), the first verb takes the standard subject agreement marker, and the other verbs have the serial agreement prefix \textit{a}{}-. These features are listed in \tabref{tab:mank:1}.

To sum-up, the finite verbal stem (\textsc{vb}) always has a subject agreement marker (\textsc{sbj}) prefixed and may have TAM markers suffixed. A pronominal object (\textsc{obj}) will be suffixed after the TAM marker and in a few cases, may amalgamate with the TAM marker. In serial verb constructions, the auxiliary has the standard subject agreement prefix and the main verb has the serial subject agreement prefix. The main verb may also have the imperfective prefix. The linear structures of the main clauses is \textsc{sbj-vb-tam-obj}. 
In subordinate clauses, a marker -\textit{uŋ} is suffixed to the verb of the non-main clause. 

Negation is marked by a prefix occurring between the subject agreement prefix and the stem of the finite verb in the perfective; the linear structure of the clause being \textsc{sbj-neg-vb}. In imperfective negative clauses, the aspect is marked by a prefix which appears immediately before the finite verb stem, after the subject agreement marker and the negation marker: \textsc{sbj-neg-ipfv{}-vb}.

\begin{table}
\begin{tabularx}{\textwidth}{llX}
\lsptoprule
Mood & Tense-aspect & Marked by \\
\midrule
Indicative & Perfective & Morphologically unmarked, prosody: slightly rising\\
& Imperfective & Morphologically unmarked, prosody: neutral \newline 
    In serial verb constructions: prefix \textit{k}{}- \newline
    In negative clauses: prefix \textit{k-}\\
&  Past & Auxiliary verb \textit{p-bi} + perfective\\
&  Future & One of the three auxiliary verbs (\textit{p-bi, p-ia} or \textit{p-luŋ}) + imperfective\\
& Durative & Reduplication of the verbal stem\\
& Habitual & Auxiliary verb \textit{p-ji}\\
& Ingressive & Auxiliary verb \textit{p-do} + past\\  
\midrule
Injunctive & Imperative & Suffix -\textit{an} {\textasciitilde} -\textit{ni} {\textasciitilde} -\textit{n}\\
& Obligative & The copula \textit{p-wo} ‘be’ followed by a serial verb construction\\
\midrule
Irrealis & Conditional & Conjunction \textit{wol} + auxiliary verb \textit{p-şaŋ} or the morpheme \textit{le}\\
\lspbottomrule
\end{tabularx}
\caption{TAM markers and constructions}
\label{tab:mank:1}
\end{table}

The data\footnote{In the near future, the collected material will be archived on the platform CorpOrAn (\url{https://corporan.huma-num.fr/})}  used for this chapter were mainly collected during field trips in Goudomp (Senegal), which took place in 2015, 2016 and 2017, and were especially dedicated to the study of negation.\footnote{The examples collected in the field will be mentioned as (Montébran, fieldnotes).} The consultants were interviewed, following the questionnaire provided by the editors (\citetv{chapters/questionnaire}),
% \lv{@LSP: please cross-ref to the questionnaire as a chapter in the book}
and Dahl’s TAM questionnaire \citep{Dahl2000}.
Some texts were also collected, but do not reflect the entire negative system. In addition to the elicited data, some of the examples also come from observed spontaneous speech. Lastly, some examples are from \citet{Gaved2004,Gaved2007,Gaved2020}.

\section{Clausal negation}\label{sec:mank:2}
\subsection{Standard Negation}\label{sec:mank:2.1}

In declarative main clauses, negation is encoded by a verbal prefix {}-\textit{VN}{}-, in which V is a vowel determined by the vowel of the preceding subject agreement marker and N a homorganic nasal consonant.\footnote{The homorganic nasal consonant is articulated based on the consonant that it is followed by: m before m, p, b; n before n, t, d; ŋ before k, g etc.} This prefix appears between the subject agreement marker and the verbal root. 

Negation in declarative main clauses shows three types of asymmetries: one concerning the marking of imperfective aspect, a second concerning subject agreement marking for \textsc{1sg}\footnote{Asymmetries concerning the TAM system are entering into the category A/Cat/TAM and asymmetries concerning person marking into the category A/Cat/PNG in \citegen{Miestamo2005} terminology.}, and a third concerning the intonation pattern.

In negative clauses, the subject agreement marker for \textsc{1sg} is the prefix \textit{m}{}- (glossed \textsc{1sg.neg}), whereas in affirmative clauses the prefix \textit{d}{}- occurs. The vowel of the negation marker is determined by the vowel of the subject agreement marker, which may on the surface appear as a long vowel: \textsc{2sg} has an agreement marker \textit{i}{}-, so the negation marker is \textit{iN}{}-; \textsc{3sg} and the serial marker (occurring in serial verb constructions keeping the same subject for all verbs of the series) have an agreement marker \textit{a}{}-, so the negation marker is \textit{aN}{}-. \textsc{2pl} and \textsc{3pl} have an agreement marker ending with the vowel \textit{a}, so the negation marker is also \textit{aN}{}-. If the subject agreement marker does not end with a vowel (\textsc{1sg} and \textsc{1pl}), the negation marker is \textit{əN}{}-.

Concerning the subject agreement marker, it is noteworthy to add that \textsc{1sg}, \textsc{2sg}, \textsc{3sg}, \textsc{1pl}, \textsc{2pl}, \textsc{3pl}, all refer to human subjects. For non-human \textsc{3sg} and \textsc{3pl}, the agreement marker is based on the noun class prefix, which may already end with a vowel (\textit{a} for the classes 1, 1a, 3, 4, 5 or \textit{u} for class 2) but if it ends with a consonant (classes 6, 7, 8, 9 and 10), a vowel \textit{a} will be added. Therefore, the negation marker is \textit{aN}{}- or \textit{uN}{}-.  These features are summarized in \tabref{tab:mank:2}.

\begin{table}
\begin{tabularx}{\textwidth}{lQQ}
\lsptoprule
& affirmative subject agreement marker & Negative subject agreement marker followed by the negation marker\\
\midrule
\textsc{1sg} & d- & m-əN-\\
\textsc{2sg} & i- & i-iN-\\
\textsc{3sg} & a- & a-aN-\\
\textsc{3sg} non-human & \textsc{cl}-(a) & \textsc{cl}-(a){}-a/uN-\\
\textsc{1pl} & ŋ- & ŋ-əN-\\
\textsc{2pl} & na- & na-aN-\\
\textsc{3pl} & ba- & ba-aN-\\
\textsc{3pl} non-human & \textsc{cl}-(a) & \textsc{cl}-(a){}-a/uN-\\
Serial & a- & a-aN-\\
\lspbottomrule
\end{tabularx}
\caption{Subject agreement marking and negation}
\label{tab:mank:2}
\end{table}

The asymmetry concerning the marking of imperfective aspect, can be explained as follows: in affirmative clauses, perfective and imperfective are morphologically unmarked and are prosodically distinguished. But in negative clauses, both perfective and imperfective have the same intonation pattern (a specific negative one) and the distinction between the aspects is morphological: a prefix \textit{k}{}- occurs in the imperfective, whereas the perfective remains morphologically unmarked. 

The linear structure of the minimal perfective clause is \textsc{sbj-vb},\footnote{It is noteworthy to add concerning the affirmative perfective main clauses that if an intransitive verb occurs without any complement (including time or locative adverbs for example) in pre-pausal positions (typically before the end of the clause or sentence), a completive suffix -\textit{i} occurs systematically on the verbal stem \citep[for more details, see][80]{Gaved2020}.} with a slightly rising intonation pattern. In negative perfective clauses, the subject agreement marker is followed by the negation prefix \textit{VN}{}-. The subject agreement marker for \textsc{1sg} changes to \textit{m}{}- under negation. The linear structure of the verb is \textsc{sbj(.neg){}-neg-vb}.
Negative perfective clauses are also marked by a specific intonation pattern (rising and falling, stressing the negation morpheme). Thus, negation of the perfective presents A/Cat/PNG asymmetry restricted to \textsc{1sg} as well as an asymmetry in the intonation pattern. Examples (\ref{ex:mank:1}--\ref{ex:mank:4}) show the affirmative-negative opposition for perfective.


\ea\label{ex:mank:1}
    \gll d-de taakal.\\ 
         \textsc{1sg}-eat yesterday\\
    \glt ‘I ate yesterday.’ (Montébran, fieldnotes) \\
\z

\ea\label{ex:mank:2}
    \gll d-de-i.\\
         \textsc{1sg}-eat-\textsc{compl}\\
    \glt ‘I ate.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:3}
    \gll m-ən-de taakal.\\
         \textsc{1sg.neg-neg-}eat yesterday\\
    \glt ‘I did not eat yesterday.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:4}
    \gll m-ən-de-i. \\
        \textsc{1sg.neg-neg-}eat\textsc{{}-compl} \\
    \glt ‘I did not eat.’ (Montébran, fieldnotes)\\
\z

In simple declarative main clauses, the imperfective aspect is morphologically unmarked and has a neutral intonation pattern. The structure of the imperfective verb is \textsc{sbj-vb}. The negation morpheme is the prefix {}-\textit{VN}{}- and the specific negative intonation pattern (rising and falling, stressing the negation morpheme) also arises in negative imperfective clauses; thus intonation does not distinguish imperfective and perfective in negative clauses. The \textsc{1sg} subject agreement marker also changes from \textit{d}{}- (in affirmative sentences) to \textit{m}{}- (in negative sentences). Negative imperfective clauses are systematically morphologically marked for aspect by the morpheme \textit{k}{}- occurring immediately before the verb root. The linear structure of the verb is \textsc{sbj(.neg){}-neg-ipfv-vb}. Since the negation morpheme always precedes the prefix \textit{k}{}- and contains a homorganic nasal \textit{N}, it systematically appears as a \textit{Vŋ}{}-. Thus, negation of the imperfective presents A/Cat/TAM, A/Cat/PNG (restricted to \textsc{1sg}) and intonational asymmetries. Examples \REF{ex:mank:5} and \REF{ex:mank:6} show the affirmative-negative opposition for the imperfective. 

\ea\label{ex:mank:5}
    \gll d-de. \\
         \textsc{1sg}-eat \\
    \glt ‘I eat.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:6}
    \gll  m-əŋ-k-de.\\
         \textsc{1sg.neg-neg-ipfv}-eat\\
    \glt ‘I do not eat.’ (Montébran, fieldnotes)
\z

To sum up, the negation of both the perfective and the imperfective uses the same negation morpheme, the prefix \textit{VN}{}-. The negation of the perfective shows two types of constructional asymmetry: A/Cat/PNG for \textsc{1sg} and the intonation pattern; and the negation of the imperfective exhibits three types of constructional asymmetry: A/Cat/PNG for \textsc{1sg}, A/Cat/TAM and the intonation pattern. Both of these aspects show patterns which occur in the majority of the other indicative paradigms.

Past tense is marked by a periphrastic construction employing the auxiliary verb \textit{p-bi} ‘come’ in the perfective aspect, such as in example \REF{ex:mank:7}. The structure of the clause is \textsc{sbj-pst vb} . In these clauses, the main verb does not agree with the subject (only the auxiliary does). In negative past clauses, the standard negation morpheme is prefixed to the auxiliary. The structure of the negative clause is \textsc{sbj(.neg){}-neg-pst vb}, with the same asymmetries as in the negative perfective, such as in example \REF{ex:mank:8}.

\ea\label{ex:mank:7}
    \gll a-bi lɛmp.\\
         \textsc{3sg-pst} work\\
    \glt ‘He worked.’ (Montébran, fieldnotes)\\
\z

\ea\label{ex:mank:8}
    \gll  a-am-bi lɛmp.\\
         \textsc{3sg{}-neg-pst} work\\
    \glt ‘He did not work.’ (Montébran, fieldnotes)
\z

Mankanya has three future tenses, all periphrastic constructions, using one of the following auxiliary verbs \textit{p-ia} ‘go’ (immediate future), \textit{p-bi} ‘come’ (probability or possibility in the future) or \textit{p-luŋ} ‘swing’ (assertive future). These periphrastic constructions are serial verb constructions and the main verb is morphologically marked for imperfective by the prefix \textit{k}{}- followed by a serial agreement marker \textit{a}{}-. The structure of the clause is \textsc{sbj-fut} \textsc{ipfv-ser-vb}. Future tenses are negated by the standard negation morpheme, associated with the negative intonation pattern. Since the finite verb is the auxiliary, it takes the negation marker and the imperfective prefix. The structure of a negative future clause is \textsc{sbj(.neg){}-neg-ipfv-fut ipfv-ser-vb}. Thus, in the negation of the future, the same asymmetries show up as in the imperfective negative.  Examples (\ref{ex:mank:9}--\ref{ex:mank:14}) show the affirmative-negative opposition for the three future paradigms.

\ea\label{ex:mank:9} 
    \gll d-ia k-a-lɛmp.\\
         \textsc{1sg-fut} \textsc{ipfv-ser}-work\\
    \glt ‘I will work (this afternoon).’ (Montébran, fieldnotes).
\z

\ea\label{ex:mank:10}
    \gll m-əŋ-k-ia k-a-lɛmp. \\
        \textsc{1sg.neg-neg-ipfv-fut} \textsc{ipfv-ser}-work \\
    \glt ‘I will not work (this afternoon).’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:11}
    \gll d-bi k-a-lɛmp.\\
         \textsc{1sg-fut} \textsc{ipfv-ser}-work\\
    \glt ‘I will work (probably).’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:12}
    \gll m-əŋ-k-bi k-a-lɛmp. \\
         \textsc{1sg.neg-neg-ipfv-fut} \textsc{ipfv-ser}-work \\
    \glt ‘I will not work (probably).’ (Montébran, fieldnotes)
\z 

\ea\label{ex:mank:13}
    \gll d-luŋ k-a-lɛmp. \\
         \textsc{1sg-fut} \textsc{ipfv-ser}-work\\
    \glt ‘I will work (certainly).’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:14}
    \gll m-əŋ-k-luŋ k-a-lɛmp.\\
         \textsc{1sg.neg-neg-ipfv-fut} \textsc{ipfv-ser}-work\\ 
    \glt ‘I will not work (certainly).’ (Montébran, fieldnotes)
\z

The durative is characterized by the reduplication of the verb, following the structure \textsc{sbj-vb-vb}, such as in example \REF{ex:mank:15}. The negative counterpart has the following structure \textsc{sbj(.neg){}-neg-ipfv-vb-vb}, and shows the same asymmetries as in the negative imperfective, such as in example \REF{ex:mank:16}.

\ea\label{ex:mank:15}
    \gll a-lɛmp-lɛmp. \\
         \textsc{3sg}-work-work\\
    \glt ‘He is working.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:16}
    \gll m-əŋ-k-lɛmp-lɛmp.\\
         \textsc{1sg.neg-neg-ipfv}-work-work \\
    \glt ‘I am not working.’ (Montébran, fieldnotes)
    \z

The habitual is a periphrastic construction, marked by the auxiliary verb \textit{p-ji} ‘say’, but the grammaticalisation process is not complete: not only the auxiliary but also the main verb has a subject agreement prefix. However, this agreement marker has specific features depending on the person \citep[for details see][178--179]{Gaved2020}. \tabref{tab:mank:3} sums up the subject agreement marker in habitual.

\begin{table}
\begin{tabularx}{\textwidth}{p{50mm}XX}
\lsptoprule
 Subject type & \multicolumn{2}{l} {Subject agreement marker}\\
   & on the auxiliary & on the main verb \\
\midrule 
Human \textsc{1sg} & standard & prefix \textit{ka}{}- \\
Human \textsc{2sg} & standard & prefix \textit{k}{}- \\
Human \textsc{3sg} & standard & zero prefix\\
Human plural (\textsc{1pl}, \textsc{2pl} or \textsc{3pl}) & standard & standard\\
Non-human (\textsc{sg} and \textsc{pl}) & standard & standard\\
\lspbottomrule
\end{tabularx}
\caption{Subject agreement marker for habitual}
\label{tab:mank:3}
\end{table}

The habitual follows the structure \textsc{sbj{}-hab sbj.hab{}-vb}. The negative counterpart has the following structure \textsc{sbj(.neg){}-neg-hab sbj.hab{}-vb}; the same asymmetries show up as in the negative perfective. Examples \REF{ex:mank:17} and \REF{ex:mank:18} show the affirmative-negative opposition for the habitual.

\ea\label{ex:mank:17}
    \gll d-ji ka-lɛmp. \\
         \textsc{1sg-hab} \textsc{1sg.hab}-work\\
    \glt ‘I usually work.’ (Montébran, fieldnotes)
\z
   
\ea\label{ex:mank:18}
    \gll m-ən-ji ka-lɛmp.\\
         \textsc{1sg.neg-neg-hab} \textsc{1sg.hab}-work \\
    \glt ‘I do not work usually.’ (Montébran, fieldnotes)
\z

The meaning of `not yet' is the negative counterpart of the ingressive construction (marked by the auxiliary \textit{p-do} ‘do’ and the past tense marker \textit{bi}-. Again, only the auxiliary verb agrees with the subject and uses the standard negation morpheme. The structure of the clause is \textsc{sbj(.neg){}-neg-ingr pst vb}. In this negation, the same asymmetries appear as in the negative perfective, see examples \REF{ex:mank:19} and \REF{ex:mank:20}.

\ea\label{ex:mank:19}
    \gll ba-do bi lɛmp.\\
         \textsc{3pl-ingr} \textsc{pst} work \\ 
    \glt ‘They already have worked.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:20}
    \gll ba-an-do bi lɛmp. \\
        \textsc{3pl{}-neg-ingr} \textsc{pst} work \\
    \glt ‘They have not worked yet.’ (Montébran, fieldnotes)
\z

\subsection{Negation in non-declarative clauses}\label{sec:mank:2.2}
\subsubsection{Prohibitive}\label{sec:mank:2.2.1}

For the second and third persons (singular and plural) the verb is marked by an imperative suffix -\textit{an} (-\textit{n} or -\textit{ni}, depending on the structure of the verbal stem). The second person singular has a zero subject agreement marker, whereas the other persons present the regular subject agreement marker. The linear order of the verb is \textsc{sbj-vb-imp.} For the first person plural, the verb does not take the imperative suffix, the construction is unmarked \textsc{sbj-vb}.

There are two types of prohibitive clauses in Mankanya. The first one is periphrastic, employing the imperative form of the negative auxiliary \textit{p-wut} ‘leave’ followed by the main verb marked for the imperfective and prefixed by the serial subject marker.
This construction is used by older speakers, or in formal speech. Examples (\ref{ex:mank:21}--\ref{ex:mank:25}) show the affirmative-negative opposition for imperative and formal prohibitive.


\ea\label{ex:mank:21}
    \gll ø-tan-an!\\
         \textsc{2sg}-tie-\textsc{imp} \\
    \glt ‘Tie!’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:22}
    \gll ŋ-tan!\\
         \textsc{1pl}-tie\\
    \glt ‘Let us tie!’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:23}
    \gll na-tan-an!\\
         \textsc{2pl}-tie-\textsc{imp}\\
    \glt ‘You tie!’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:24}
    \gll ø-wut-an k-a-de wa!\\
         \textsc{2sg}-leave-\textsc{imp} \textsc{ipfv-ser-}eat \textsc{cl.3.sg.obj} \\ 
    \glt ‘Do not eat it!’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:25}
    \gll na-wut-an k-a-de wa!\\
         \textsc{2pl}-leave-\textsc{imp} \textsc{ipfv-ser}-eat \textsc{cl.3.sg.obj}\\
    \glt ‘Do not eat it!’ (Montébran, fieldnotes)
\z

The second prohibitive construction, used in common speech, has a structure \textsc{sbj-ipfv-vb}, see examples (\ref{ex:mank:26}--\ref{ex:mank:28}). This construction appears (also from the speakers’ point of view) as a shortened form of the periphrastic prohibitive. There is no overt negative marker in this clause.

\ea\label{ex:mank:26}
    \gll ø-k-tan!\\
         \textsc{2sg-ipfv}-tie \\
    \glt ‘Do not tie!’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:27}
    \gll na-k-tan!\\
         \textsc{2pl{}-ipfv}-tie \\
    \glt ‘Do not tie!’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:28}
    \gll ba-k-tan!\\
         \textsc{3pl-ipfv}-tie \\
    \glt ‘Do not let them tie!’ (Montébran, fieldnotes)
\z

To sum up the prohibitive: a specific negation morpheme occurs (the prohibitive auxiliary \textit{p-wut} ‘leave’),  and since the same person-number distinctions are made in the prohibitive and the affirmative imperative, there is no paradigmatic asymmetry.
The main verb loses its finitness with the introduction of a new finite element.

There is a specific impersonal prohibitive construction employing the infinitive of the verb \textit{p-wut} ‘leave’ followed by the infinitive of the main verb, as in examples \REF{ex:mank:29} and \REF{ex:mank:30}. There is no affirmative equivalent of this structure, which is a paradigmatic asymmetry.

\ea\label{ex:mank:29}
    \gll p-wut p-fumaar ţi.\\
         \textsc{inf}-leave \textsc{inf}-smoke here\\
    \glt ‘Smoking is not allowed here.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:30}
    \gll p-wut p-haabus u-janel.\\
         \textsc{inf}-leave \textsc{inf}-open \textsc{cl.2.sg}-window\\
    \glt ‘Do not open the window.’ (Montébran, fieldnotes)
\z

\subsubsection{Negative obligative}\label{sec:mank:2.2.2}

Obligatives are built with the copula \textit{p-wo} followed by a verb marked for imperfective and with a serial subject agreement marker. The structure of the clause is \textsc{sbj-cop} \textsc{ipfv-ser-vb}, as in example \REF{ex:mank:31}. The negative counterpart of obligatives applies the standard negation morpheme on the copula; the structure of the clause is \textsc{sbj(.neg){}-neg-cop ipfv-ser-vb}, as in example \REF{ex:mank:32}. In the negative obligative, the same asymmetries appear as in the negative perfective.


\ea\label{ex:mank:31}
    \gll i-wo k-a-lɛmp. \\
         \textsc{2sg-cop} \textsc{ipfv-ser}-work \\
    \glt ‘You must work.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:32}
    \gll i-in-wo k-a-win-a. \\
         \textsc{2sg{}-neg-cop} \textsc{ipfv-ser}-see-\textsc{3sg.obj} \\
    \glt ‘You must not see him.’ (Montébran, fieldnotes)
\z

\subsubsection{Hypothetical constructions}\label{sec:mank:2.2.3}

The marker for hypothetical is \textit{p-şaŋ}. Semantically, this auxiliary means that the process should have been done, but was prevented. The clause follows the structure \textsc{sbj-hyp} \textsc{ser-vb}, as in example \REF{ex:mank:33}. There is no dedicated negative marker for the hypothetical: the standard negation morpheme occurs with the same asymmetries as in the negative perfective. The structure of the negative clause is \textsc{sbj(.neg){}-neg-hyp ser-vb}, as in example \REF{ex:mank:34}.

\ea\label{ex:mank:33}
    \gll i-şaŋ a-lɛmp. \\
         \textsc{2sg-hyp} \textsc{ser}-work \\
    \glt ‘You should have worked.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:34}
    \gll  i-in-şaŋ a-lɛmp.\\
         \textsc{2sg{}-neg-hyp} \textsc{ser}-work\\
    \glt ‘You should not have worked.’ (Montébran, fieldnotes)
\z

\subsubsection{Negative questions}\label{sec:mank:2.2.4}

Overt questions are marked by the particle \textit{i} occurring at the end of the clause. Interrogative clauses do not have a specific verbal construction, therefore, the negation depends on the paradigm used. There is no dedicated negation morpheme, the standard negation morpheme also occurs in negative questions, see examples \REF{ex:mank:35} and \REF{ex:mank:36}. The negative subject agreement marker for \textsc{1sg} and the negative intonation pattern always occur (rising then falling, stressing the negation marker, and rising at the end of the clause for the question), in some cases (imperfective or future clauses) also TAM marking asymmetry.

\ea\label{ex:mank:35}
    \gll na-ntohi a-an-daan biere i?\\
         \textsc{cl.1.sg}-old\_man \textsc{3sg-neg}-drink beer \textsc{q}\\
    \glt ‘Hasn't the old man drunk beer?’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:36}
    \gll i-iŋ-k-ia p-win şaaş i?\\
         \textsc{2sg-neg-ipfv-fut} \textsc{inf}-see father \textsc{q}\\
    \glt ‘Will you not see your father?’ (Montébran, fieldnotes)
\z

\subsection{Negation in stative predication}\label{sec:mank:2.3}
\subsubsection{Possession}\label{sec:mank:2.3.1}

Predicative possession is expressed by a verbal predication using the verb \textit{p-ka} ‘have’, which is normally inflected in the affirmative and the negative, as shown in examples \REF{ex:mank:37} and \REF{ex:mank:38}. The same asymmetries arise as in declarative clauses.

\ea\label{ex:mank:37}
    \gll Dama a-ka na-poţ. \\
         \textsc{pn} \textsc{3sg}-have \textsc{cl.1.sg}-child\\
    \glt ‘Dama has a child.’ \citep[125]{Gaved2020}
\z

\ea\label{ex:mank:38}
    \gll Dama a-aŋ-ka na-poţ. \\
         \textsc{pn} \textsc{3sg{}-neg}-have \textsc{cl.1.sg}-child \\
    \glt ‘Dama does not have children.’ \citep[125]{Gaved2020}
\z

\subsubsection{Attribution clauses}\label{sec:mank:2.3.2}

Attributive clauses are standard verbal constructions using the copula \textit{p-wo} ‘be’, which is normally inflected in the affirmative and in the negative, as in examples \REF{ex:mank:39} and \REF{ex:mank:40}. The standard negation morpheme occurs on the copula.

\ea\label{ex:mank:39}
    \gll m-ən-wo na-hula.\\
         \textsc{1sg.neg-neg-cop} \textsc{cl.1.sg}-hula \\
    \glt ‘I am not hula (Mankanya).’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:40}
    \gll i-in-wo na-jukan. \\
         \textsc{2sg{}-neg{}-cop} \textsc{cl.1.sg}-teacher \\
    \glt ‘You are not teacher.’ (Montébran, fieldnotes)
\z

\subsubsection{Existential and locative predication}\label{sec:mank:2.3.3}

Existential and locative clauses use the copula \textit{p-wo} ‘be’ which is normally inflected in the affirmative and in the negative, as shown in examples (\ref{ex:mank:41}--\ref{ex:mank:44}). 
    
\ea\label{ex:mank:41}
    \gll ţi da-wo na ba-poţ. \\
         here \textsc{cl.9.sg}\textsc{-cop} \textsc{prep} \textsc{cl.1.pl}-children \\
    \glt ‘There are children here.’ (lit. ‘Here is with children.’) (Montébran, fieldnotes)
\z

\ea\label{ex:mank:42}
    \gll ţi da-an-wo na ba-poţ. \\
         here \textsc{cl.9.sg{}-neg-cop} \textsc{prep} \textsc{cl.1.pl}-children \\
    \glt ‘There are no children here.’ (lit. ‘Here is not with children.’) (Montébran, fieldnotes)
\z

\ea\label{ex:mank:43}
    \gll ba-poţ ba-wo ţi. \\
         \textsc{cl.1.pl}-children \textsc{cl.1.pl{}-cop} here \\
    \glt ‘Children are here.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:44}
    \gll ba-poţ ba-an-wo ţi.\\
         \textsc{cl.1.pl}-children \textsc{cl.1.pl{}-neg-cop} here\\
    \glt ‘Children are not here.’ (Montébran, fieldnotes)
\z

The second existential construction is based on the verb \textit{p-ka} ‘have’ in an impersonal use \textit{uka} (\textit{u-ka} \textsc{cl.2.sg}{}-have; the subject agreement marker is \textit{u}{}- which can be bound to \textit{u{}-ko} ‘something’.\footnote{For more details see \citet[208]{Gaved2020}}) This impersonal construction is exemplified in \REF{ex:mank:45}. This has a negative counterpart \textit{uŋka}~(\textit{u-ŋ-ka} \textsc{cl.2.sg-neg}{}-have) which may occur in a nominal clause or negative verbal clause, see examples \REF{ex:mank:46} and \REF{ex:mank:47}.

\ea\label{ex:mank:45}
    \gll u-ka du u-kalabus... \\
         \textsc{cl.2.sg}{}-have \textsc{loc} \textsc{cl.2.sg}{}-prison\\
    \glt ‘There was in the prison...’ \citep[209]{Gaved2020}
\z

\ea\label{ex:mank:46}
    \gll Uŋka na-duttur ţi u-ntabaŋka. \\
         \textsc{neg.ex} \textsc{cl.1.sg}-doctor \textsc{loc} \textsc{cl.2.sg}-village\\
    \glt ‘There is no doctor in the village.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:47}
    \gll Uŋka ñaaŋ a-an-hil-uŋ p-do u-ko wi. \\
         \textsc{neg.ex} someone \textsc{3sg{}-neg}-can-\textsc{sub} \textsc{inf}-do \textsc{cl.2.sg}-thing \textsc{dem}\\
    \glt ‘Nobody can do that.’ (lit. ‘There isn't a person able to do that.’) (Montébran, fieldnotes)
\z

Presentative constructions are marked by the reduplication of demonstrative articles, as shown in example \REF{ex:mank:48}. The negation morpheme occurring in these clauses is \textit{mënţ}, which is added to the affirmative clauses without other modifications, such as in example \REF{ex:mank:49}.

\ea\label{ex:mank:48}
    \gll ba-buk naan biki biki.\\
         \textsc{cl.1a.sg}-children \textsc{poss.1sg} \textsc{dem} \textsc{dem}\\
    \glt ‘Here are my children.’ \citep[124]{Gaved2020}
\z

\ea\label{ex:mank:49}
    \gll mënţ ba{}-buk naan biki biki. \\
         \textsc{neg.nml} \textsc{cl.1a.sg}-children \textsc{poss.1sg} \textsc{dem} \textsc{dem} \\
    \glt ‘These are not my children.’ \citep[126]{Gaved2020}
\z

Locative constructions use the copula \textit{p-wo} ‘be’ which is normally inflected in the affirmative and in the negative. There is no specific negative morpheme dedicated to  locative predications, the standard negation morpheme occurs on the copula, as shown in example \REF{ex:mank:50}.
States or permanent properties are encoded lexically by a verb, and are thereby negated by the standard negation strategy. The verbal encoding of property predications is cross-linguistically common \REF{ex:mank:add1}.

\ea\label{ex:mank:50}
    \gll b-maŋa ba{}-an-wo ţi b-lay. \\
         \textsc{cl.7.sg}-mango\_tree \textsc{cl.7.sg{}-neg-cop} \textsc{loc} \textsc{cl.5.sg}-yard\\
    \glt ‘The mango tree is not in the yard.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:add1}
\gll ŋ-ya u-ţeeh woli b-nuur ba-an-yiik-i \\
    1\textsc{pl}-go \textsc{cl.2.sg}-field if \textsc{cl.7.sg}-sunlight \textsc{cl.5.sg-neg}-be\_hot-\textsc{compl}\\ 
\glt 'We're going to the field if it's not hot' \citep[121]{Gaved2020} \\
\z

\subsection{Negation in non-main clauses}\label{sec:mank:2.4}

There is no specific negation dedicated to non-main clauses; the standard negation marker (with the negative subject agreement marker for \textsc{1sg} and the negative intonation pattern) also occurs in relative and complement clauses. In the negative, imperfective and future clauses are marked by the prefix \textit{k}{}-. 

Negation in relative clauses involves the copula \textit{p-wo}, which takes the negation marker. It is noteworthy that the suffix -\textit{uŋ} indicating non-main clauses occurs on the finite verb of the dependent clause in the affirmative and on the copula in negative clauses. The verb loses its finiteness, as for example between the sentences \REF{ex:mank:51} and \REF{ex:mank:52}. 

\ea\label{ex:mank:51}
    \gll u-buş, u-ţoo-uŋ, u-joţ-i.\\
         \textsc{cl.2.sg}-dog \textsc{cl.2.sg}-seat{}-\textsc{sub} \textsc{cl.2.sg}-be\_nasty-\textsc{compl}\\ 
    \glt ‘The seated dog is nasty.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:52}
    \gll u-buş, u-un-wo-uŋ u-uŋ-ţoo, u-joţ-i. \\
         \textsc{cl.2.sg}-dog \textsc{cl.2.sg{}-neg-cop-sub} \textsc{cl.2.sg{}-neg}-seat \textsc{cl.2.sg}-be\_nasty-\textsc{compl} \\
    \glt ‘The dog, which is not seated, is nasty.’ (Montébran, fieldnotes)
\z

Negation of complement clauses depends on the paradigm employed (perfective or imperfective), as shown in examples \REF{ex:mank:53} and \REF{ex:mank:54}. 

\ea\label{ex:mank:53}
    \gll a-ji kë i-im-bi taakal.\\
         \textsc{3sg}-say \textsc{coord} \textsc{2sg{}-neg}-come yesterday\\
    \glt ‘He says that you did not come yesterday.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:54}
    \gll d{}-ji kë i-iŋ-k-bi faan.\\
         \textsc{1sg}-say \textsc{coord} \textsc{2sg-neg-ipfv}-come tomorrow\\
    \glt ‘I say that you will not come tomorrow.’ (Montébran, fieldnotes)
\z

Conditional clauses are initiated by the conjunction \textit{wol}. There are perfective and imperfective conditional clauses. 

In perfective conditional clauses, negation proceeds the same way as in the negation of perfective indicative. A/Cat/PNG asymmetry for \textsc{1sg} and the specific negative intonation pattern arise.

Imperfective conditional clauses are introduced by the conjunction \textit{wol} and closed with the morpheme \textit{le}. Negation proceeds the same way as the negation of imperfective indicative. Asymmetry appears in the TAM marking, specifically A/Cat/TAM, with subject agreement A/Cat/PNG, and in the intonation pattern. 

 Examples \REF{ex:mank:55} and \REF{ex:mank:56} show the negative construction for the conditional, perfective and imperfective.

\ea\label{ex:mank:55}
    \gll wol m-ən-lɛmp-i...\\
         \textsc{cond} \textsc{1sg.neg-neg-}work\textsc{{}-compl}\\
    \glt ‘If I have not worked...’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:56}
    \gll wol m-əŋ-k-lɛmp le [...]\\
         \textsc{cond} \textsc{1sg.neg-neg-ipfv-}work \textsc{cond}\\
    \glt ‘If I do not work [...]’ (Montébran, fieldnotes)
\z

\section{Non-clausal negation}\label{sec:mank:3}
\subsection{Negative replies}\label{sec:mank:3.1}

The interjection meaning ‘no’ is \textit{ʔaʔa} (\textit{aa} means ‘yes’). Strictly speaking, \textit{ʔaʔa} means disagreement with the polarity of the question and \textit{aa} means agreement with the polarity of the question. Both can be followed by a negative clause, such as in examples \REF{ex:mank:57} and \REF{ex:mank:58}.

\ea\label{ex:mank:57}
    \glt ‘Is she back home?’ \medskip\newline 
    \gll ʔaʔa a-an-do bi tiiş.\\
         no \textsc{3sg-neg-ingr} \textsc{pst} come\_back\\
    \glt ‘No, she did not come back yet.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:58}
    \glt ‘She is not back home yet?’ \medskip \newline 
    \gll aa a-an-do bi tiiş.\\
        yes \textsc{3sg-neg-ingr} \textsc{pst} come\_back \\
    \glt ‘Yes, she did not come back yet.’ (Montébran, fieldnotes)
\z

\subsection{Negative indefinites}\label{sec:mank:3.2}

Negative indefinites are based on the negation morpheme \textit{niin}: \textit{niin ñaaŋ} ‘no-one, nobody’, \textit{niin u-ko} ‘nothing’, \textit{niin u-ko u-loŋ} ‘nothing, anything’. These negative indefinites occur in affirmative sentences, as in example \REF{ex:mank:59}.

\ea\label{ex:mank:59}
    \gll p-wala ka-ŧemp pa-an-wo niin u-ko u-loŋ. \\
         \textsc{inf}{}-come\_down \textsc{cl.3.sg}{}-circumcision \textsc{cl.3.sg{}-neg}{}-be \textsc{neg} \textsc{cl.2.sg}{}-thing \textsc{cl.2.sg}{}-\textsc{indf}\\
    \glt ‘To be circumcised is nothing.’ \citep[142]{Gaved2020}
\z


\subsection{Negative derivation}\label{sec:mank:3.3}

Negative derivation has not been found in Mankanya.

\citet[12]{Gaved2007} also notes that in some cases, a perfective verb may appear as qualifier in an attributive clause to express some properties of the subject. This verb may be in the negative perfective (with all characteristics of the negation of the perfective), as in example \REF{ex:mank:60}. 

\ea\label{ex:mank:60}
    \gll ba-wo ba-am-buk-i.\\
         \textsc{3pl-cop} \textsc{3pl{}-neg}-bear-\textsc{compl}\\
    \glt ‘They are without children.’ \citep[12]{Gaved2007}
\z

\section{Miscellaneous aspects of negation}\label{sec:mank:4}
\subsection{The scope of negation}\label{sec:mank:4.1}

In Mankanya, the negative marker always occurs immediately before the negated constituent.

The negation of the focused subject or object is realised with the morpheme \textit{mënţ} appearing clause initially and followed by the subject or the object. This morpheme \textit{mënţ} is typically used as an anaphoric demonstrative \REF{ex:mank:add2}, but in this specific construction it occurs as a negative marker, as shown in the opposition between \REF{ex:mank:61} and \REF{ex:mank:62}.

\ea\label{ex:mank:add2}
\gll u-ko mënţ u-de Şompi maakan. \\
    \textsc{cl.2.sg}-think that \textsc{cl.2.sg}-eat \textsc{pn} very \\
\glt ‘This thing worried Shompi a lot.’ \citep[125]{Gaved2020}\\
\z 

\ea\label{ex:mank:61}
    \gll nji d-wo-uŋ ţi.\\
         \textsc{1sg.indep} \textsc{1sg-cop-sub} here\\
    \glt ‘It is me living here.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:62}
    \gll mënţ nji d-wo-uŋ ţi.\\
         \textsc{neg.nml} \textsc{1sg.indep} \textsc{1sg-cop-sub} here\\
    \glt ‘It is not me living here.’ (Montébran, fieldnotes)
\z


The focused object (negated or not) appears at the beginning of the clause, it is followed by a relative clause (note that in these specific relative clauses the first person subject appears as a homorganic nasal \textit{N}{}-), such as in examples \REF{ex:mank:63} and \REF{ex:mank:64}\footnote{Editorial note: Sadly, the author passed away during the final revision process of this chapter. While all efforts have been made to ensure its quality, the glossing of \REF{ex:mank:63} and \REF{ex:mank:64} has not been verified by the author.}.


\ea\label{ex:mank:63}
    \gll ka-hot ki ŋ-kit-iŋ.\\
         \textsc{cl.3.sg}-leg \textsc{rel} \textsc{1sg.rel}-break-\textsc{sub}\\
    \glt ‘It is my leg that is broken.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:64}
    \gll mënţ ka-hot ki ŋ-kit-iŋ.\\
         \textsc{neg.nml} \textsc{cl.3.sg}-leg \textsc{rel} \textsc{1sg.rel}-break-\textsc{sub}\\
    \glt ‘It is not my leg that is broken.’ (Montébran, fieldnotes)
\z

\subsection{Negative polarity}\label{sec:mank:4.2}

Negative polarity items have not been found in the data so far. 


\subsection{Marking of NPs in the scope of negation.}\label{sec:mank:4.3}

No effects on the marking of NPs have been found in the data.


\subsection{Reinforcing negation}\label{sec:mank:4.4}

Mankanya has a particle \textit{i}, used to reinforce the obligative construction (see example \ref{ex:mank:65}), which may also occur in negation, as in example \REF{ex:mank:66}.


\ea\label{ex:mank:65}
    \gll i-wo i k-a-win-a.\\
         \textsc{2sg-cop} \textsc{renf} \textsc{ipfv-ser}-see-\textsc{3sg.obj}\\
    \glt ‘You absolutely must see him.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:66}
    \gll i-in-wo i k-a-win-a. \\
         \textsc{2sg-neg-cop} \textsc{renf} \textsc{ipfv-ser}-see-\textsc{3sg.obj}\\
    \glt ‘You absolutely must not see him.’ (Montébran, fieldnotes)
\z

\subsection{Negation, coordination and complex clauses}\label{sec:mank:4.5}

The negative morphemes \textit{mënţ} and \textit{niin} occur in the negation of nominal constituents of the clause, namely to negate a focused subject or object; or with the meaning ‘{neither}… {nor}…’, in this case, they occur in negative clauses. If it concerns the subject, the structure is similar to the negation of emphatic clauses, and the negative morpheme \textit{mënţ} occurs, as in example \REF{ex:mank:67}.

\ea\label{ex:mank:67}
    \gll mënţ papa a, mënţ nni a, bukal bŧi ba-tëb ba-am-bi-i.\\
         \textsc{neg.nml} father \textsc{3sg.indep.obj} \textsc{neg.nml} mother \textsc{3sg.indep.obj} \textsc{3pl.indep} all \textsc{cl.1.pl}-two \textsc{cl.1.pl-neg}-come-\textsc{compl}\\ 
         
    \glt ‘Neither my father nor my mother came.’ (lit. ‘Neither my father, nor my mother, all the two of them did not come.’) (Montébran, fieldnotes)
\z

If the negation concerns the object, without any emphasis, the negative morpheme will be \textit{niin}, as in example \REF{ex:mank:68}.

\ea\label{ex:mank:68}
    \gll m-ən-du niin papa niin nni.\\
         \textsc{1sg.neg-neg}-call \textsc{neg} father \textsc{neg} mother\\
    \glt ‘I did not call neither my father nor my mother.’ (Montébran, fieldnotes)
\z

The verbal emphasis in affirmative sentences is realised by stressing the verb. The negative counterpart uses the standard negation construction, and is stressed by negative intonation. This construction is mostly used in contrastive sentences (as in examples \REF{ex:mank:69} and \REF{ex:mank:70}), but may also be used as answers.

\ea\label{ex:mank:69}
    \gll m-ən-leir-i, d-piiŧ-piiŧ. \\
         \textsc{1sg.neg-neg}-read-\textsc{compl} \textsc{1sg}-write-write\\
    \glt ‘I did not read, I was writing.’ (Montébran, fieldnotes)
\z

\ea\label{ex:mank:70}
    \gll m-əŋ-k-ia k-a-de, d-daan-daan.\\
         \textsc{1sg.neg-neg-ipfv-fut} \textsc{ipfv-ser}-eat \textsc{1sg}-drink-drink\\
    \glt ‘I will not eat, I am drinking.’ (Montébran, fieldnotes)
\z

\subsection{Other aspects of negation}\label{sec:mank:4.6}

One occurrence of expletive negation has been found in the data. It is a nominal clause, with the negation morpheme \textit{mënţ}, used as an answer to the question ‘who did that?’. The answer expresses the certainty of the speaker, despite the use of negation, as in example \REF{ex:mank:71}. Even more, the use of negation reinforces the certainty of the speaker, even if it may be perceived as ironic.

\ea\label{ex:mank:71}
    \gll mënţ ul a!\\
         \textsc{neg.nml} \textsc{3sg.indep} \textsc{3sg}\\
    \glt ‘That’s him!’ (lit. ‘That is not him!’) (Montébran, fieldnotes)
\z

\section{Summary}\label{sec:mank:5}
\largerpage
To conclude, we observed that there are five negation morphemes in Mankanya, summarized in \tabref{tab:mank:4}. Two of them occur in verbal clauses, the other three in nominal clauses.
The standard negation morpheme is a verbal prefix \textit{VN}{}-, which negates all verbal structures, including existential, stative and possessive ones, except prohibitive constructions. These are negated by the negative auxiliary \textit{p-wut}. 

Verbal clauses always present a subject agreement marker. In negative clauses (except in the prohibitive, which is defective), the subject agreement marking differs for \textsc{1sg}. 

In Mankanya, the aspectual marking differs between affirmative and negative clauses. Intonation patterns distinguish perfective and imperfective constructions in affirmative clauses, but are always identical in negative clauses. Negative clauses with the standard negation morpheme are characterized by a specific intonation pattern. In negative clauses, perfective and imperfective are morphologically distinct (imperfective is marked by a prefix \textit{k}{}-, perfective is unmarked), whereas both are morphologically unmarked in affirmative clauses. 

 Prohibitive is a specific negative counterpart to imperatives and commands. There are two possible forms for the prohibitive: a longer one used in formal speech, and a shortened one, more commonly used. The negation morpheme used to negate commands and imperatives is a negative auxiliary. This paradigm is defective and has no \textsc{1sg} form.

Negation of nominal elements has three specific markers \textit{mënţ}, \textit{niin}, and \textit{uŋka}. The use of each of them is clearly defined. \textit{Mënţ} occurs in focused contexts, and not \textit{niin}. Negation of stative prediction mostly uses standard negation, in the case of existential negation another morpheme (\textit{uŋka}) may arise. \tabref{tab:mank:4} sums-up on the characteristics of each of the paradigms.

\begin{table}
\begin{tabularx}{\textwidth}{p{13mm}lQ}
\lsptoprule
Negative marker & Negative construction  & Used for\\
\midrule
\scshape \textit{VN}- &\textsc{sbj(.neg)-neg-vb} & Perfective, past, habitual, obligative\\
\tablevspace
\scshape \textit{VN}- &\scshape sbj(.neg)-neg-ipfv-vb & Imperfective, future, durative\\
\tablevspace
 \textit{p-wut} &\textsc{sbj-}leave\textsc{{}-imp ipfv-ser-vb} & Prohibitive\\
\tablevspace
 ∅ & \scshape sbj-ipfv-vb & Prohibitive (shortened)\\
\tablevspace
 \textit{mënţ} &\scshape neg.nml sbj & Presentative, focused subject or object\\
\tablevspace
 \textit{uŋka} &\scshape neg.ex sbj & Negative existential\\
\tablevspace
 \textit{niin} &immediately before the negated constituent & neither… nor \\
\tablevspace
 \textit{ʔaʔa} &  \textit{ʔaʔa}& negative reply\\
\lspbottomrule
\end{tabularx}
\caption{The negative constructions of Mankanya}
\label{tab:mank:4}
\end{table}

\newpage

\section*{Abbreviations}
\begin{tabularx}{.5\textwidth}{lQ}%HC extended to .5
\scshape 1--10, 1a &number of noun class \\ 
\scshape 1sg & first person singular \\ 
\scshape 2sg & second person singular \\ 
\scshape 3sg & third person singular \\ 
\scshape 1pl & first person plural\\ 
\scshape 2pl & second person plural\\ 
\scshape 3pl & third person plural\\ 
\scshape cl & noun class\\ 
\scshape cond &conditional \\ 
\scshape coord &coordination \\
\scshape cop &copula \\ 
\scshape compl &completive \\ 
\scshape dem &demonstrative \\ 
\end{tabularx}
\begin{tabularx}{.5\textwidth}{lQ}%HC extended to .5
\scshape ex & existential \\ 
\scshape fut &future \\ 
\scshape hab &habitual \\ 
\scshape hyp &hypothesis \\ 
\scshape imp &imperative \\ 
\scshape indep &independent subject marker \\
\scshape indf &indefinite \\ 
\scshape inf &infinitive \\ 
\scshape ingr &ingressive \\  
\scshape ipfv &imperfective \\ 
\scshape loc &locative \\ 
\scshape nml &nominal  \\ 
\scshape neg &negation \\ 
\end{tabularx}

\begin{tabularx}{.45\textwidth}{lQ}
\scshape np &noun phrase \\ 
\scshape obj &object \\ 
\scshape pst &past \\ 
\scshape pl &plural \\ 
\scshape pn &proper noun \\ 
\scshape poss &possession \\ 
\scshape q &question marker \\ 
\scshape rel &relative \\ 
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape renf &reinforcement particle \\ 
\scshape sbj & subject \\
\scshape ser &serial subject agreement marker \\ 
\scshape sg &singular \\ 
\scshape sub &subordinate \\ 
\scshape tam & tense, aspect, mood \\
\scshape vb & verb \\
\end{tabularx}

\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
\end{document} 
