\documentclass[output=paper,colorlinks,citecolor=brown,draft]{langscibook}
\ChapterDOI{10.5281/zenodo.18234105}
\author{Fernando Zúñiga\affiliation{University of Bern}}
\title{Negation in Mapudungun}
\abstract{This paper outlines the basics of negation in Mapudungun (South America). Three negative suffixes occur in verbal clauses (-\textit{la}, -\textit{ki}, and \nobreakdash-\textit{nu}), and one negative particle occurs in nonverbal clauses (\textit{nu}). The distribution of the negative suffixes is governed by finiteness and mood; tense, aspect, and other grammatical factors are irrelevant for the above allomorphy. Most stative predications are expressed by verbal clauses. However, equational predications can also be expressed by nonverbal clauses. With stative predications, therefore, the occurrence of the negators follows the verbal-nonverbal clause divide. Other than the negator -\textit{nu} also occurring in negative indefinites, the language does not show any negative derivation. Lastly, there is an interesting distribution of the verbs \textit{ngen} and \textit{mülen} in existential predications; the former is also an individual-level copular predicate and the latter also a stage-level one, but with existentials this opposition appears to be neutralized and \textit{ngen} seems to be the preferred option in the negative, while \textit{mülen} is employed in the affirmative.

\keywords{finiteness, existential predication, Mapudungun, nonverbal predication, stative predication}
}

\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}
 

\begin{document}
\maketitle 

\section{The language}

The Mapuche language or Mapudungun (ISO: \href{https://iso639-3.sil.org/code/arn}{arn}, Glottocode: \href{https://glottolog.org/resource/languoid/id/mapu1245}{mapu1245}) is a linguistic isolate and is currently spoken mostly in south-central and central Chile, but also in central Argentina (\citealt{SimonsFennig2018}). The total number of speakers is somewhat uncertain, ranging from 100,000 when considering only high-fluency speakers to some 250,000 when including semi-speakers; passive speakers seem to be numerous (\citealt{ZúñigaOlate2017}). The language had some contact with (southern) Quechuan around the time of the Spanish invasion; since then, and especially since the late 19th century, contact with Chilean and Argentinian Spanish has been intense. Its name is spelled \textit{Mapuzugun} or \textit{Mapuzungun} when following other orthographic conventions found in the contemporary literature; I use the Chilean version of the convention called \textit{Alfabeto Mapuche Unificado} for the data presented here. There are several regional varieties of the language, but differences among them are rather modest, despite occasional claims to the contrary – which is one of several reasons for not following the \textit{Ethnologue} or \textit{Glottolog} in postulating an “Araucanian” family. I address the dialect called Central Mapudungun in the present paper.

Mapudungun morphology ranges from synthetic to polysynthetic. Nominal morphology is relatively simple, with virtually no inflection and only limited derivational affixation, but with productive compounding capabilities. Verbal morphology, by contrast, is highly complex: despite lexical affixation being rather limited, there are numerous templatic slots, productive NP incorporation, and multi-radical verb stems \citep{Zúñiga2017}. The morphology is predominantly suffixing and “agglutinative” (i.e., there is no flexivity, and most morphemes show simple exponence and combine concatenatively to form words). With respect to syntax, there are both nominal and verbal clauses. The latter distinguish two kinds of transitive clauses, viz. direct and inverse (\citealt{Zúñiga2006}: Ch. VII; \citealt{HaudeZúñiga2016}). Constituent order patterns seem to have been predominantly VSO in the past; possibly under the influence of Spanish, SVO patterns have become more frequent of late (\citealt{Zúñiga2006}: Ch. VII).

Based upon original data and the extant descriptions of Mapudungun, this paper presents an outline of negation in the language according to the guiding principles detailed in \citet{Veselinova2014} and \citet{Miestamo2017}, as well as the questionnaire prepared by the editors of this volume  \parencitetv{chapters/questionnaire}. Essentially the same analysis —but a slightly different framing— is presented in a companion article published by the author and colleagues \citep{OlateEtAl2020}. (Unless otherwise specified, examples come from my own field notes.) \sectref{sec:mapu:2} and \sectref{sec:mapu:3} present clausal and non-clausal negation, respectively. \sectref{sec:mapu:4} closes the paper with suggestions for further research.

\section{Clausal negation}
\label{sec:mapu:2}

Mapudungun has four clausal negative morphemes, viz. three verbal suffixes (-\textit{la}, \nobreakdash-\textit{ki}, and -\textit{nu}) and one marker usually thought of as negative particle or negative adverb (\textit{nu}).\footnote{The old negator \textit{müg} does not seem to be in use anymore; see \sectref{sec:mapu:3.1} for more details.} While -\textit{la} and \nobreakdash-\textit{ki} are only found with indicative and imperative verb forms, respectively (\sectref{sec:mapu:2.1} and \sectref{sec:mapu:2.2}), \nobreakdash-\textit{nu} is found with all other verbs (e.g., with subjunctive and nonfinite forms; \sectref{sec:mapu:2.4}), and \textit{nu} appears with nonverbal elements (e.g., in nominal clauses; \sectref{sec:mapu:2.3.6}).\footnote{\citet[23]{Valdivia1606} mentions that the variety spoken in Santiago by the turn of the 16th century showed \nobreakdash-\textit{nu} with some persons in indicative forms as well. I have not found this phenomenon in present-day varieties.} The markers’ distribution is essentially based on mood and finiteness, as summarized in \figref{figtab:mapu:1} below.

% \hc{@Sebastian/LSP: could you either make the table look clearer, or if you take the tree, make the subj + nonfinite come together to a single \textit{-nu}. See outcommented explanation right below. Thanks!}
% \begin{table} %HC show the author how we changed it. Does he want 1 or 2 "-nu" for subjunctive and nonfinite.
%FZ answer: Ideally, subjunctive and nonfinite would lead to a single nu. If this is not feasible, what is the problem with leaving the original table? What does “does not work so good in LaTeX” mean, exactly? %Sebastian could you do the single -nu for subjunctive and nonfinite? (overall FZ would prefer tables than forests if it's possible. he had clear lines delimiting each cell in the word document, which helped to understand the table). I added the \cmidrule's to try make it look better. What do you think Sebastian?

% \begin{tabular}{ccccc}
% \lsptoprule
% \multicolumn{4}{l}{Verbal} & {Nonverbal} \\
% \cmidrule(lr){1-4}\cmidrule(lr){5-5}
% \multicolumn{3}{l}{Finite} & {Nonfinite} & \\
% \cmidrule(lr){1-3}\cmidrule(lr){4-4}
% \multicolumn{1}{l}{Indicative} & \multicolumn{1}{l}{Imperative} & \multicolumn{1}{l}{Subjunctive} &  & \\
% \cmidrule(lr){1-1}\cmidrule(lr){2-2}\cmidrule(lr){3-3}
% \multicolumn{1}{l}{\textit{-la}} & \multicolumn{1}{l}{\textit{-ki(-nu)}} & \multicolumn{2}{c}{\textit{-nu}} & \textit{nu}\\
% \lspbottomrule
% \end{tabular}

% this is the same table with vertical lines too, 
% \begin{tabular}{|ccccc|}
% \lsptoprule
% \multicolumn{4}{c|}{{Verbal}} & {Nonverbal} \\
% \cmidrule(lr){1-4}\cmidrule(lr){5-5} 
% \multicolumn{3}{c|}{{Finite}} & \multicolumn{1}{c|}{Nonfinite} & \\
% \cmidrule(lr){1-3}\cmidrule(lr){4-4} 
% \multicolumn{1}{c|}{Indicative} & \multicolumn{1}{c|}{{Imperative}} & \multicolumn{1}{c|}{{Subjunctive}} & \multicolumn{1}{c|}{} & \\
% \cmidrule(lr){1-1}\cmidrule(lr){2-2}\cmidrule(lr){3-3}
% \multicolumn{1}{c|}{\textit{-la}} & \multicolumn{1}{c|}{\textit{-ki(-nu)}} & \multicolumn{2}{c|}{\textit{-nu}} & \textit{nu}\\
% \lspbottomrule
% \end{tabular}

\begin{figure}
\begin{forest} for tree={grow'=east,delay={where content={}{shape=coordinate}{}}},   forked edges,  
[negator
    [verbal
        [finite
            [indicative
                [\textit{-la}, tier=word, no edge]
            ]
            [imperative
                [\textit{-ki(-nu)}, tier=word, no edge]
            ]
            [subjunctive,tier=cat
                [{$\left.\begin{array}{c}
                        ~\\
                        ~\\
                        ~\end{array}\right\}$\textit{-nu}}, tier=word, no edge, yshift=-5mm]
            ]
        ]
        [nonfinite,tier=cat
        ]
    ]
    [nonverbal,tier=cat
        [\textit{nu}, tier=word, no edge]
    ]
]
\end{forest}
\caption{Overview of clausal negators}
\label{figtab:mapu:1}
\end{figure}

\subsection{Standard negation}
\label{sec:mapu:2.1}
Declarative main clauses are headed by a verb in the \textit{i}{}-marked indicative mood and are negated by suffixing -\textit{la} to the verbal predicate, as in \REF{ex:mapu:1}.

\ea\label{ex:mapu:1}
    \ea\label{ex:mapu:1a}
        \gll La-le-i tüfa-chi üñüm.\\
             die-\textsc{res-ind[3sg]} this-\textsc{attr} bird \\
        \glt ‘This bird has died.’
    \ex\label{ex:mapu:1b}
        \gll La-le-la-i tüfa-chi üñüm. \\
             die-\textsc{res-neg-ind[3sg]} this-\textsc{attr} bird \\
        \glt ‘This bird has not died.’
    \z
\z 

The standard negator \nobreakdash-\textit{la} shows neither phonologically nor lexically conditioned allomorphy (unlike other derivational and inflectional suffixes on the verb, which do have phonologically conditioned variants). Moreover, aspectual and/or temporal distinctions do not appear to interact with the standard negator in any way. The only notable exception is the bimorphemic expression of the cessative (‘no longer, not anymore’), which consists of -\textit{we} ‘already’ and negative -\textit{la} (cf. analogous Spanish \textit{ya no}) \REF{ex:mapu:2a}. There is no dedicated \textit{nondum} ‘not yet’ element \citep{Veselinova2017}; this meaning is expressed by combining \textit{petu} ‘still’ with negative \nobreakdash-\textit{la} (cf. analogous Spanish \textit{aún no}) \REF{ex:mapu:2b}.

\ea\label{ex:mapu:2}
    \ea\label{ex:mapu:2a}
        \gll Küpa-tu-mu-we-la-a-n yod. \\
             come-\textsc{appl-2>1.els-}already-\textsc{neg-fut-1sg.ind} more\\
        \glt ‘You (\textsc{nsg}) do not keep coming for me.’\footnotemark
        \footnotetext{This future indicative form can be used either to describe a state of affairs (and be translated in the indicative) or as a polite way to avoid the more direct and potentially rude imperative (and be translated as ‘do not keep coming for me!’).} \citep[215]{Salas2006}
    \ex\label{ex:mapu:2b}
        \gll Petu la-le-la-i. \\
             still die-\textsc{res}{}-\textsc{neg-ind[3sg]} \\
        \glt ‘S/he has not yet died.’
    \z
\z 

Lastly, standard negation is symmetrical, both constructionally (i.e., the morphological make-up of indicative verb forms does not change in any other way when negated) and paradigmatically (i.e., the same grammatical categories are available in the affirmative and the negative).

\subsection{Negation in non-declarative main clauses}
\label{sec:mapu:2.2}

In Mapudungun, structurally special non-declarative main clause types exist for commands, but not for questions. Commands are typically expressed with the zero-marked imperative mood \REF{ex:mapu:3a} and were presumably negated with -\textit{ki} in an erstwhile transparent form that is no longer in use \REF{ex:mapu:3b}.\footnote{The only prohibitive forms given in \citet[23a--b]{Valdivia1606} have negated subjunctive forms, cf. \textit{elu-l-m-i} ‘if  you (\textsc{sg}) give’ vs. \textit{elu-ki-l-m-i} ‘don’t (\textsc{sg}) give’. It may well be the case, of course, that this “hybrid” morphological make-up was the original. See also Footnote 6.}

\ea\label{ex:mapu:3}
    \ea[]{\label{ex:mapu:3a}
        \gll Langüm-fi-nge tüfa-chi üñüm! \\
             kill-\textsc{3obj-2sg.imp} this-\textsc{attr} bird \\
        \glt ‘Kill this bird!’}
    \ex[*]{\label{ex:mapu:3b}
        \gll Langüm-ki-fi-nge tüfa-chi üñüm!\\
             kill-\textsc{neg-3obj-2sg.imp} this-\textsc{attr} bird\\
        \glt (Reconstructed, intended meaning: ‘Don’t kill this bird!’)}
    \z
\z 

Actual present-day negated forms follow two basic patterns, namely one (“sub-junctive-like”) for all person-number values in which the form takes subjunctive \nobreakdash-\textit{l} and subjunctive person-number markers \REF{ex:mapu:4a}, and one (“hybrid”) for \textsc{2sg} and 3rd person forms in which subjunctive \nobreakdash-\textit{l} co-occurs with (affirmative) imperative person-number markers \REF{ex:mapu:4b}.\footnote{\citet[146]{Salas2006} distinguishes \textit{nu-l}{}-forms from \textit{ki(-nu)-l}{}-forms using the terms “(negative) hypothetical subjunctive” and “(negative) volitive subjunctive,” respectively. I am grateful to Aldo Olate for reminding me of this alternative terminology.} Both patterns show variation regarding whether \nobreakdash-\textit{ki} or \nobreakdash-\textit{ki-nu} is used as negator; the double marker \nobreakdash-\textit{ki-nu} is often found with subjunctive-like forms and occasionally found with hybrid forms.\footnote{To judge from the presentation and indirect comments found in older sources (\citealt[23--24]{Valdivia1606} and \citealt[166--167]{Augusta1903}), the subjunctive-like pattern might be older than the hybrid one, and the single negator \nobreakdash-\textit{ki} definitely appears to be older than the double negator \nobreakdash-\textit{ki-nu}. \citet[124--126]{Moesbach1962} treats the marker \textit{\nobreakdash-ki\nobreakdash-nu} as a third monomorphemic allomorph in its own right (with a different first vowel, viz. “\textit{kenú} or \textit{kenó}”).} 

\ea\label{ex:mapu:4}    
    \ea\label{ex:mapu:4a}
        \gll Langüm-ki-(nu-)fi-l-m-i tüfa-chi üñüm!\\
             kill-\textsc{neg-neg-3obj-sbjv-2-sg} this-\textsc{attr} bird\\
        \glt
    \ex\label{ex:mapu:4b}
        \gll Langüm-ki-(nu-)fi-l-nge tüfa-chi üñüm!\\
             kill-\textsc{neg-neg-3obj-sbjv-2sg.imp} this-\textsc{attr} bird \\
        \glt Both: ‘Don’t kill this bird!’
    \z
\z 

The morphological make-up of monopersonal forms (i.e., those marking only the person of their subject) is schematically summarized in Tables \ref{tab:mapu:2a}-\ref{tab:mapu:2b} below. Note in \tabref{tab:mapu:2a} the essentially agglutinative pattern for most markers built upon a mood marker (\nobreakdash-\textit{l} ‘\textsc{sbjv}’, zero imperative), a person marker (-\textit{i} ‘1’, -\textit{m} ‘2’ and zero third), and a number marking (-\textit{i} ‘\textsc{sg’}, -\textit{u} ‘\textsc{du}’ and \nobreakdash-\textit{n} ‘\textsc{pl’}). The portmanteaus that deviate from this are most of the \textsc{1sg/2sg} markers: \nobreakdash-\textit{li} ‘\textsc{1sg.sbjv’}, \nobreakdash-\textit{chi} ‘\textsc{1sg.imp}’, and \nobreakdash-\textit{nge} ‘\textsc{2sg.imp’.} 3rd-person markers show some portmanteaus as well (\nobreakdash-\textit{le} ‘\textsc{3sg.sbjv}’ and \nobreakdash-\textit{pe} ‘3.\textsc{imp’}), and \textsc{3nsg} forms include a \textit{ng} element to which the dual and plural endings are affixed, which is bound to the verb in the subjunctive but still somewhat detached in the imperative (see \citealt{Zúñiga2014} for details). The 3rd-person hybrid forms in \tabref{tab:mapu:2b} show some variation between affixing the bare mood marker \nobreakdash-\textit{l} and the 3-person-imperative portmanteau \nobreakdash-\textit{le}.


\begin{table}
\caption{Selected monopersonal endings}
    \label{tab:mapu:2} %also needed because once there is a link to table 2 (i.e. probably table 2a + 2b)
    \subtable[Selected monopersonal endings I]{\label{tab:mapu:2a}
    \begin{tabularx}{.8\textwidth}{llXll}
    
    \lsptoprule
    & \multicolumn{2}{l}{{subjunctive}} & \multicolumn{2}{l}{{imperative (\textsc{sbjv}{}-like)}}\\
    \cmidrule(lr){2-3}\cmidrule(lr){4-5}
    & {\textsc{aff}} & {\textsc{neg}} & {\textsc{aff}\footnote{Rather than being imperatives in the strict sense of the term, first-person forms and third-person forms are hortatives and jussives, respectively. The \textsc{1sg} form usually has an optative or a “declarative” value, e.g. \textit{tripa-chi!} ‘I’d like to go out; OK, I’m going to get out now!’ \citep[96]{Salas2006}.}} & {\textsc{neg}}\\
    \midrule 
    \textsc{1sg} & \textit{-li} & \textit{-nu-li} & \textit{{}-chi} & {(unattested)}\\
    \textsc{1du} & \textit{-l-i-u} & \textit{-nu-l-i-u} & \textit{-i-u} & \textit{-ki(-nu)-l-i-u}\\
    \textsc{1pl} & \textit{-l-i-n} & \textit{-nu-l-i-n} & \textit{-i-n} & \textit{-ki(-nu)-l-i-n}\\
    \textsc{2sg} & \textit{-l-m-i} & \textit{-nu-l-m-i} & \textit{-nge} & \textit{-ki(-nu)-l-m-i}\\
    \textsc{2du} & \textit{-l-m-u} & \textit{-nu-l-m-u} & \textit{-m-u} & \textit{-ki(-nu)-l-m-u}\\
    \textsc{2pl} & \textit{-l-m-n} & \textit{-nu-l-m-n} & \textit{-m-n} & \textit{-ki(-nu)-l-m-n}\\
    \textsc{3sg} & \textit{-le} & \textit{-nu-le} & \textit{-pe} & \textit{-ki(-nu)-le}\\
    \textsc{3du} & \textit{-l-ng-u} & \textit{-nu-l-ng-u} & \textit{-pe eng-u} & \textit{-ki(-nu)-le eng-u}\\
    \textsc{3pl} & \textit{-l-ng-n} & \textit{-nu-l-ng-n} & \textit{-pe eng-n} & \textit{-ki(-nu)-le eng-n}\\
    \lspbottomrule
    \end{tabularx}    
    }%end subtable
    
    
    \subtable[Selected monopersonal endings II]{\label{tab:mapu:2b}  
        \begin{tabularx}{.85\textwidth}{llXll}
    \lsptoprule
    & \multicolumn{2}{l}{{subjunctive}} & \multicolumn{2}{l}{{imperative (hybrid)}}\\
    \cmidrule(lr){2-3}\cmidrule(lr){4-5}
    & {\scshape aff} & {\scshape neg} & {\scshape aff} & {\scshape neg}\\
    \midrule 
    {\scshape 2sg} & \textit{{}-l-m-i} & \textit{{}-nu-l-m-i} & \textit{{}-nge} & \textit{{}-ki(-nu)-l-nge}\\
    {\scshape 3sg} & \textit{-le} & \textit{-nu-le} & \textit{-pe} & \textit{-ki(-nu)-l(e)-pe}\\
    {\scshape 3du} & \textit{-l-ng-u} & \textit{-nu-l-ng-u} & \textit{-pe eng-u} & \textit{-ki(-nu)-l(e)-pe eng-u}\\
    {\scshape 3pl} & \textit{-l-ng-n} & \textit{-nu-l-ng-n} & \textit{-pe eng-n} & \textit{{}-ki(-nu)-l(e)-pe eng-n}\\
    \lspbottomrule
    \end{tabularx}
    }%end subtable
\end{table}

Mapudungun negation can be classified as belonging to Type IV in \citegen{AuweraEtAl2013} model: \textsc{2sg} prohibitives use a construction other than the \textsc{2sg} affirmative imperative and a sentential negator not found in the indicative. Only the reconstructed forms \REF{ex:mapu:3b} are constructionally symmetrical; both subjunctive-like and hybrid prohibitives show mood-person-number marking that differs from their affirmative counterpart’s. As to paradigmatic asymmetry, it is also important to note that some dedicated polypersonal imperative forms are no longer in use (either subjunctive-like or hybrid, either affirmative or negative); for instance, \textsc{2sg>1sg} interactions are routinely expressed by the future indicative (see the footnote to \REF{ex:mapu:2a} above and \citealt{Smeets2008}: 185f for more details).\footnote{An analysis of the complex make-up of Mapudungun polypersonal forms lies beyond the scope of the present paper. The interested reader is referred to \citet[Ch. VII]{Zúñiga2006} and the appropriate sections in \citet{Augusta1903}, \citet{Salas2006} and \citet{Smeets2008}.}


\subsection{Negation in stative predications}
\label{sec:mapu:2.3}
\subsubsection{Introduction}\label{sec:mapu:2.3intro}
First note that there are two stative copular/existential verbs in Mapudungun, which play an important role in what follows. One is \textit{ngen} ‘be/exist\textsubscript{1}’; it is an individual-level (i.e., permanent-state) predicate and roughly corresponds to Ibe\-ro-Romance \textit{ser}. The other is \textit{mülen} ‘be/exist\textsubscript{2}’; it is a stage-level (i.e., temporary-state) predicate and roughly corresponds to Ibero-Romance \textit{estar}.\footnote{There are two homophonous \textit{n}{}-suffixes in Mapudungun, one of them finite (‘\textsc{1sg.ind’}) and the other non-finite (‘\textsc{inf’}). \textit{Ngen, mülen}, and other verbs in the running text are given in the usual infinitive citation form.} (The opposition between \textit{ngen} and \textit{mülen} is unlikely to be contact-induced, due to its age -- both verbs appear with the present-day values in the oldest description of the language (1606) -- and apparent consistency -- I am not aware of any significant variation regarding the semantics of these verbs, other than some uses of \textit{mülen} that are not parallel to Spanish \textit{estar}.)

Equation is expressed by nominal clauses in Mapudungun and therefore always uses \textit{nu}; all other subtypes of stative constructions (i.e., proper inclusion, property attribution, existence, location, and possession) are expressed by verbal clauses and therefore show the mood- and finiteness-related allomorphy shown in \sectref{sec:mapu:2.1}--\ref{sec:mapu:2.3}. 
\subsubsection{Proper inclusion}
\label{sec:mapu:2.3.1}

Proper inclusion is typically expressed via verbalization with \nobreakdash-\textit{nge} (< \textit{nge-} ‘be\textsubscript{1}’ in \REF{ex:mapu:5} and \REF{ex:mapu:6}; but see also \sectref{sec:mapu:5} below).

\ea\label{ex:mapu:5}
    \ea\label{ex:mapu:5a}
        \gll Witran-nge{}-i.\\
             {visitor-be\textsubscript{1}{}-\textsc{ind[3sg]}} \\ 
        \glt ‘S/he is a visitor.'
    
    \ex\label{ex:mapu:5b}
        \gll Witran-nge-la-i. \\
             {visitor-be\textsubscript{1}{}-\textsc{neg-ind[3sg]}} \\
        \glt ‘S/he is not a visitor.’

    \ex\label{ex:mapu:5c}
        \gll Witran-nge-le{}-pa-i.\\
             visitor-be\textsubscript{1}{}-\textsc{prog{}-cis-ind[3sg]}\\
        \glt ‘S/he is (behaving formally, as) a visitor here.’ (\citealt{Smeets2008}: 171, 283)

    \ex\label{ex:mapu:5d}
        \gll Witran-nge-le-la-pa-i.\\
             visitor-be\textsubscript{1}{}-\textsc{prog-neg-cis-ind[3sg]}\\
        \glt ‘S/he is not (behaving formally, as) a visitor here.’
    \z
\z 


\ea\label{ex:mapu:6}
    
    \ea\label{ex:mapu:6a}
        \gll {Müna-tremo-pichi-we-che-nge-i-m-i.} \\ 
             very-beautiful-little-young-person-be\textsubscript{1}{}-\textsc{ind-2-sg}\\
        \glt ‘You (\textsc{sg}) are a very beautiful boy.’
    \ex\label{ex:mapu:6b}
        \gll {Müna-tremo-pichi-we-che-nge-la-i-m-i.} \\
             very-beautiful-little-young-person-be\textsubscript{1}{}-\textsc{neg-ind-2-sg}\\
        \glt ‘You (\textsc{sg}) are not a very beautiful boy.’ \citep[54]{Moesbach1962}
    \z
\z 
 
\subsubsection{Property attribution}

Adjectival stems (which are invariably also inchoative verbal stems) are used with the verbalizer \nobreakdash-\textit{nge} (< \textit{nge}{}- ‘be\textsubscript{1}’, e.g. in
(\ref{ex:mapu:7a}--\ref{ex:mapu:7a'}{\small $^{'}$}), cf. \sectref{sec:mapu:2.3.1}) and aspectual \nobreakdash-\textit{le} (< \textit{müle}{}- ‘be\textsubscript{2}’, e.g. in
(\ref{ex:mapu:7b}--\ref{ex:mapu:7b'}{\small $^{'}$}));
although individual-level forms originally took \nobreakdash-\textit{nge}, they appear increasingly unmarked nowadays
(e.g. in (\ref{ex:mapu:7c}--\ref{ex:mapu:7c'}{\small $^{'}$})).


\ea\label{ex:mapu:7}
    \ea\label{ex:mapu:7a}
        \gll  Kurü-nge-i ti ruka.\\ 
              blacken-be\textsubscript{1}\textsc{{}-ind[3sg] art} house\\
        \glt ‘The house is [always] black.’ (Sp. \textit{ser})
    
    % \exp{ex:mapu:7a}\label{ex:mapu:7a'}
    %\exi{a$'$.}\label{ex:mapu:7a'}
    \exi{a\parbox{0mm}{\small$^{'}$}.}\label{ex:mapu:7a'}
        \gll Kurü-nge-la-i ti ruka. \\
             blacken-be\textsubscript{1}{}-\textsc{neg-ind[3sg]} \textsc{art} house \\
        \glt ‘The house is not [always] black.’ (Sp. \textit{ser})
    
    \ex\label{ex:mapu:7b}
        \gll Kurü-le-i ti ruka. \\ 
             blacken-\textsc{res-ind[3sg]} \textsc{art} house \\
        \glt ‘The house is [now] black.’ (Sp. \textit{estar})
    
%    \exp{ex:mapu:7b}\label{ex:mapu:7b'}
    \exi{b\parbox{0mm}{\small$^{'}$}.}\label{ex:mapu:7b'}
        \gll Kurü-le-la-i ti ruka. \\
             blacken-\textsc{res-neg-ind[3sg]} \textsc{art} house \\
        \glt ‘The house is not [now] black.’ (Sp. \textit{estar})
        
    \ex\label{ex:mapu:7c}
        \gll Kurü-i ti ruka. \\ 
             black(en)\textsc{{}-ind[3sg]} art house \\
        \glt ‘The house blackened / is black.’ (Sp. \textit{ser})
    
%    \exp{ex:mapu:7c}\label{ex:mapu:7c'}
    \exi{c\parbox{0mm}{\small$^{'}$}.}\label{ex:mapu:7c'}
        \gll Kurü-la-i ti ruka. \\
             black(en)-\textsc{neg-ind[3sg]} \textsc{art} house \\
        \glt ‘The house did not blacken / is not black.’ (Sp. \textit{ser}) 
    \z
\z 


\subsubsection{Location}

In addition to \textit{müle}{}- ‘be\textsubscript{2}’, three productive suffixes express the same notion and are commonly found together with other verbs, namely \nobreakdash-\textit{le} (< \textit{müle}{}- ‘be\textsubscript{2}’), \nobreakdash-\textit{nie} (< \textit{nie}{}- ‘have’, cf. \sectref{sec:mapu:2.3.5}), and \nobreakdash-\textit{künu} (< \textit{künu}{}- ‘leave’). Example \REF{ex:mapu:8} shows only an affirmative verb form, but the negative would use the standard negator -\textit{la} (\textit{payla-künu-tu-rke-la-fi}).

\ea\label{ex:mapu:8}   
    \ea\label{ex:mapu:8a}
        \gll Müle-ka-i ruka mew. \\ 
             be\textsubscript{2}{}-\textsc{cont-ind[3sg]} house \textsc{posp} \\
        \glt ‘S/he is still in the house.’ (Sp. \textit{estar})
    \ex\label{ex:mapu:8b}
        \gll Müle-ka-la-i ruka mew. \\
             be\textsubscript{2}{}-\textsc{cont-neg-ind[3sg]} house \textsc{posp}\\
        \glt ‘S/he is still not in the house.’ (Sp. \textit{estar})
    \z
\z 

\ea\label{ex:mapu:9}
    \gll Kisu chi wentru payla-künu-tu-rke-fi ta chi domo. \\
         \textsc{dp} \textsc{art} man get.on.one’s.back-leave-\textsc{rest}{}-\textsc{rep-3obj[}3\textsc{sg]} \textsc{ptcl} \textsc{art} woman\\
    \glt ‘The man left the woman on her back again, they say.’ \citep[267]{Salas2006}
\z

\subsubsection{Existence} 
\label{sec:mapu:2.3.4}
We already saw the two existential verbs \textit{ngen} and \textit{mülen} at the beginning of this section. Now also note that they stand in a polarity-based opposition that is still not well understood. Compare the question and the answer in \REF{ex:mapu:10}, which tend to use different verbs; the affirmative question uses \textit{mülen}, whereas the negative answer apparently favors \textit{ngen} (although \textit{mülen} was also possible, and perhaps still is). 

\ea\label{ex:mapu:10}
    \begin{xlist}
    \exi{Q: }\label{ex:mapu:10a}
        \gll Müle-i  chadi? \\ 
             exist\textsubscript{2}{}-\textsc{ind[3sg]} salt \\
        \glt ‘Is there any salt?’ \citep[55]{Augusta1916}
    \exi{A: }\label{ex:mapu:10b}
        \gll Nge-la-i. / Müle-la-i. \\
             exist\textsubscript{1}{}-\textsc{neg-ind[3sg]} {} exist\textsubscript{2}{}-\textsc{neg{}-ind[3sg]}\\
        \glt ‘No, there isn’t any.’ (\citealt[55, 179]{Augusta1916})
    \end{xlist}
\z 

See also both assertions in \REF{ex:mapu:11}, which use more closely related elements (“the same verb” \textit{ngen}) but different constructions (\REF{ex:mapu:11a} being the default clause with a verbal predicate and a nominal argument, and \REF{ex:mapu:11b} being an example of verbalization via \nobreakdash-\textit{nge}; cf. \sectref{sec:mapu:2.3.5}).

\ea\label{ex:mapu:11}
    \ea\label{ex:mapu:11a}
        \gll Nge-la-i chadi. \\ 
             exist\textsubscript{1}{}-\textsc{neg-ind[3sg]} salt \\
        \glt ‘There is no salt.’ \citep[125]{Smeets2008}
    \ex\label{ex:mapu:11b}
        \gll Chadi-nge-la-i. \\
             salt-have\textsubscript{2}{}-\textsc{neg-ind[3sg]} \\
        \glt ‘It is not salty / it does not have salt.’ \citep[126]{Smeets2008}
    \z
\z 

That the opposition is not triggered by the question-answer frame can be seen from example \REF{ex:mapu:12}, which shows negative \textit{ngelan} and affirmative \textit{mülen} occurring in the same (non-interrogative) sentence.

\ea \label{ex:mapu:12}
    \gll Nge-la-i i-al, welu petu müle-we-i putu-al. \\
        exist\textsubscript{1}{}-\textsc{neg-ind[3sg]} eat-\textsc{fut.nfin} but still exist\textsubscript{2}{}-already-\textsc{ind[3sg]} drink-\textsc{fut.nfin}\\
    \glt ‘There is no food, but there still is [something] left to drink.’ \citep[200]{Smeets2008}
\z

Descriptive studies on Mapudungun have not addressed this (partial) polarity-based opposition so far, and my informants produce somewhat inconsistent data and have only rather vague intuitions as to what is at work here, so more research is needed in order to ascertain how advanced \textit{nge-la}- might be on its way to becoming a negative existential. Assuming Mapudungun does not (yet) distinguish negators for existential and other verbal expressions, it belongs to Type A in \citegen{Croft1991} model (see also \citealt{Veselinova2014}). 


\subsubsection{Possession}
\label{sec:mapu:2.3.5}

Functionally mirroring the two copular/existential predicates, but showing an unexpected formal distribution, there are two ways to express predicative possession in Mapudungun.\footnote{See \citet{OlateEtAl2018} for a comprehensive treatment of possession in Mapudungun.} One is with the verb \textit{nien} ‘have\textsubscript{1}’ \REF{ex:mapu:13a}; it is an individual-level default predicate used for permanent possession. The other is \textit{ngen} ‘have\textsubscript{2}’ \REF{ex:mapu:13b}; it is a stage-level suffixal predicate used for temporary possession. 
\textit{ngen} shows a tendency to be used for temporary possession but this is not so clearcut as it is also used with kinship terms (see below).
Both can be covered by Spanish \textit{tener} (\textit{ngen} ‘have\textsubscript{2}’ can also be rendered colloquially as \textit{estar con}, lit. ‘be with’) and English \textit{have}.

\ea\label{ex:mapu:13}
    \ea\label{ex:mapu:13a}
        \gll Nie-i epu kawell. \\ 
             have\textsubscript{1}{}-\textsc{ind[3sg]} two horse \\
        \glt ‘S/he has two horses.’ (individual level: unmarked possession)
    \ex\label{ex:mapu:13b}
        \gll Kawellu-nge-i. \\
             horse-have\textsubscript{2}{}-\textsc{ind[3sg]} \\
        \glt ‘S/he has a horse (right now).’ (stage level; marked possession, Sp. lit. \textit{está con caballo})
    \z
\z 

Notably, temporary-possession -\textit{nge} is also found with some kinship nouns and even with body-part nouns, e.g. \textit{kure-nge-} (wife-have\textsubscript{2}{}-) ‘be married (to a woman)’, \textit{epu-namun{}-nge-} (two-foot-have\textsubscript{2}{}-) ‘be two-footed/legged’.

With nouns where possession expresses ownership (e.g., \textit{ruka} ‘house’ in \REF{ex:mapu:14a}), permanent-possession \textit{nien} ‘have\textsubscript{1}’ is the default option. Temporary-possession -\textit{nge} is occasionally used with unmovable objects like houses as well (e.g. in \REF{ex:mapu:14b}).

\ea\label{ex:mapu:14}
    \ea\label{ex:mapu:14a}
        \gll Nie-la-i kiñe ruka. \\ 
             have\textsubscript{1}{}-\textsc{neg}{}-\textsc{ind[3sg]} one house \\
        \glt ‘S/he does not have a house.’
    \ex\label{ex:mapu:14b}
        \gll Kiñe-ruka-nge-la-i. \\
             one-house-have\textsubscript{2}{}-\textsc{neg-ind[3sg]} \\
        \glt ‘She does not have a house (and her husband has more than one wife).’ (based on \citealt[9]{Augusta1903})
    \z
\z 

\subsubsection{Equation / identity predication}
\label{sec:mapu:2.3.6}

Nominal clauses are invariably used for equational constructions (i.e., ‘NP\textsubscript{X} ${\equiv}$ NP\textsubscript{Y}’), and occasionally for proper-inclusion constructions (i.e., ‘NP\textsubscript{X} ${\in}$ \{NP\textsubscript{Y}\}’). To judge from the extant lexicographic evidence, the latter usage seems to have appeared or spread in the 20th century; it may have originated in elliptical constructions (e.g., NP\textsubscript{X} NP\textsubscript{Y}{}-\textit{nge-i} > NP\textsubscript{X} NP\textsubscript{Y}). In general, nominal clauses often include a particle/proclitic \textit{ta}, especially when the second NP is definite. The exact formal (wordhood-related) properties of \textit{nu} in such constructions are still to be investigated.


The examples in \REF{ex:mapu:15} illustrate equational constructions; \REF{ex:mapu:15a} is the default construction and \REF{ex:mapu:15b} is a fronted variant that conveys some sort of emphasis (presumably: contrastive focus).

\ea\label{ex:mapu:15}
    \ea\label{ex:mapu:15a}
        \gll [Fey-chi domo]\textsubscript{X} [(ta) ñi ina-n lamngen]\textsubscript{Y}.\\
             \textsc{dem-attr} woman \textsc{ptcl} \textsc{1sg.poss} follow-\textsc{nfin} sister \\
        \glt ‘That woman is my youngest sister.’ \citep[143]{Smeets2008} 
    \ex\label{ex:mapu:15b}
        \gll [Iñche]\textsubscript{Y} \textit{nu}, [ta ti]\textsubscript{X}. \\
            \textsc{1sg} \textsc{neg} \textsc{ptcl} \textsc{art} \\
        \glt ‘That is not me/mine.’ \citep[144]{Smeets2008}
    \z
\z 

Examples (\ref{ex:mapu:16a}--\ref{ex:mapu:16b}) illustrate negated proper-inclusion constructions (see \sectref{sec:mapu:2.3.1} above). \REF{ex:mapu:16c} illustrates a slightly different kind of negated equational construction, where the first member of the equation is covert.

\ea\label{ex:mapu:16}    
    \ea\label{ex:mapu:16a}
        \gll {Fey}\textsubscript{X} [wentru]\textsubscript{Y} nu. \\ 
             \textsc{dem} man \textsc{neg} \\
        \glt ‘He is not a man.’ \citep[144]{Smeets2008} (Also: \textit{fey wentru-nge-la-i,} cf. \sectref{sec:mapu:2.3.1})
    
    \ex\label{ex:mapu:16b}
        \gll [Tüfa-chi pu che kom]\textsubscript{X} [mapuche]\textsubscript{Y} nu? \\
             this-\textsc{attr} \textsc{pl} person   all M.   \textsc{neg}\\
        \glt ‘Are these people not all Mapuche?’ \citep[144]{Smeets2008}
    
    \ex\label{ex:mapu:16c}
        \gll Tunte kulli-lle-li rume, Ø\textsubscript{X} [mütewe]\textsubscript{Y} nu. \\
        how.much pay-\textsc{aff-1sg.sbjv} ever {} much \textsc{neg} \\
        \glt ‘No matter how much I paid, it was not much.’ \citep[144]{Smeets2008}
    \z
\z 

\subsubsection{Summary}

\tabref{tab:mapu:3} below schematically summarizes the stative constructions presented above. The symbol Σ represents verbal stems (which may be complex in that they may consist of more than just roots, and/or they may include several roots/stems), and \{\textsc{neg}\} represents the set of negator allomorphs \nobreakdash-\textit{ki}, -\textit{la}, and \nobreakdash-\textit{nu}. In addition to the semantic kinds location, attribution, existence, possession, and proper inclusion, individual-level and stage-level predication have to be distinguished in order to properly account for the distribution of the forms. The functional status of \textit{(-)nge(-)} ‘be’ is quite complex, appearing as it does in both individual-level and stage-level predications of different kinds.

% \hc{@Sebastian /LSP: could you make this table 3 look better? At least now the multirows are correct}
\begin{table} %HC @Sebastian: the second part of the table is a bit difficult. now it does not look good, but at least the multirows are correct and it looks more or less like it should
    \small 
    \begin{tabularx}{\textwidth}{Xllll}
    \lsptoprule
   & \multicolumn{2}{l}{{Individual-level (= unmarked)}} & \multicolumn{2}{l}{{Stage-level (= temporary)}}\\
   \cmidrule(lr){2-3}\cmidrule(lr){4-5}
    & {Affirmative} & {Negative} & \multicolumn{1}{l}{{Affirmative}} & {Negative}\\
\midrule 
    {Location} & --- & --- & \multicolumn{1}{l}{{Σ-}} & {Σ-\{\textsc{neg\}-}}\\
    \midrule
    {Attribution} & {Σ(-\textit{nge})-} & {Σ(-\textit{nge})-\{\textsc{neg\}-}} & \multicolumn{1}{l}{{Σ-\textit{le}{}-}} & {Σ-\textit{le}{}-\{\textsc{neg\}-}}\\
    \midrule
    {Existence} & --- & --- & \multicolumn{1}{l}{{\textit{müle}{}-\textit{/nge-}}} & {\textit{nge}{}-\{\textsc{neg\}-}}\\
\midrule
    {Possession} & {\textit{nie}{}-} & {\textit{nie}{}-\{\textsc{neg\}-}} &  {NP-\textit{nge}{}-} &  {NP-\textit{nge}{}-\{\textsc{neg\}-}}\\
\cmidrule(lr){1-3   }
    \multirow{2}{*}{Proper inclusion} & {NP-\textit{nge}{}-} & {NP-\textit{nge}{}-\{\textsc{neg}\}-} & {NP-\textit{nge}{}-}  & {NP-\textit{nge}{}-\{\textsc{neg\}-}}\\[.5em]
    &  {NP  NP} &  {NP  NP  \textit{nu}} &  {---} &  {---} \\
\cmidrule{1-1}\cmidrule{4-5}
    {Equation} & NP  NP & NP  NP  \textit{nu} & {---} &{---} \\
    \lspbottomrule
    \end{tabularx}
%    \renewcommand\thetable{3} %needed here because of table 2a and 2b before
    \caption{Summary of Stative Predication Constructions (Verbal/Nonverbal)}
    \label{tab:mapu:3}
\end{table}


\subsection{Negation in dependent clauses}
\label{sec:mapu:2.4}

Dependent clauses are headed either by verbs in the \textit{l}{}-marked subjunctive mood \REF{ex:mapu:17} (see also \tabref{tab:mapu:2} above) or by nonfinite verbs that are unmarked for mood and take a specialized ending (e.g., \textit{\nobreakdash-el, \nobreakdash-lu, -n, -am}; \REF{ex:mapu:18} illustrates a \textit{lu}{}-form). Both kinds are negated by suffixing -\textit{nu} to the predicate.


\ea\label{ex:mapu:17}
    
    \ea\label{ex:mapu:17a}
        \gll Feyengün  aku-fu-le! \\ 
             \textsc{3pl} arrive-\textsc{ri-3.sbjv} \\
        \glt ‘If only they (\textsc{pl}) would arrive!’
    
    \ex\label{ex:mapu:17b}
        \gll Feyengün  aku-nu-fu-le! \\
             3\textsc{pl} arrive-\textsc{neg-ri-3.sbjv} \\
        \glt ‘If only they (\textsc{pl}) would not arrive!’ \citep[184]{Smeets2008}
    \z
\z 
 

\ea\label{ex:mapu:18}
    
    \ea\label{ex:mapu:18a}
        \gll Ngilla-la-a-i kofke  nie-lu plata. \\ 
             buy-\textsc{neg-fut-ind[3sg]} bread have-\textsc{nfin} money \\
        \glt ‘He who has money will not buy bread.’
    
    \ex\label{ex:mapu:18b}
        \gll Ngilla-la-a-i kofke  nie-nu-lu plata.\\
             buy-\textsc{neg-fut-ind[3sg]} bread  have-\textsc{neg-nfin} money\\
        \glt  ‘He who has no money will not buy bread.’ \citep[189]{Smeets2008}
    \z
\z 


Like standard negation, dependent-clause negation is constructionally and paradigmatically symmetrical. Nevertheless, the negator does show some (free) variation: it can occur either as \nobreakdash-\textit{nu} or as \nobreakdash-\textit{no}.


\section{Non-clausal negation}
\label{sec:mapu:3}

\subsection{Negative replies}
\label{sec:mapu:3.1}

The old nonverbal negator \textit{müg}, which also occurred as \textit{mü}, \textit{mu} or \textit{me}, was used – possibly until the mid-20th century – as negative answer to questions, i.e., as a counterpart to affirmative \textit{may}.\footnote{I am grateful to Aldo Olate for reminding me of the existence of the old negator \textit{müg}.} Nowadays the default negator \textit{nu {\textasciitilde} no}, possibly under the influence of the Spanish negator \textit{no}, has replaced \textit{müg} in this function. (See \citealt[166--167]{Moesbach1962}: for more details.) The usual answer to a question also includes the verbal predicate, as seen in example \REF{ex:mapu:19}.

\ea\label{ex:mapu:19}
    \ea\label{ex:mapu:19a}
        \begin{xlist}
        \exi{A: }\label{ex:mapu:19a1}
            \gll Allkü-i-m-i? \\ 
                 hear-\textsc{ind-2-sg} \\
            \glt ‘Did you (\textsc{sg}) hear?’ 
        \exi{B: }\label{ex:mapu:19a2}
            \gll (Allkü-n) may. \\ 
                 hear-\textsc{1sg.ind} \textsc{aff} \\
            \glt ‘Yes, I did.’ \citep[166]{Moesbach1962}
        \end{xlist}
    \ex\label{ex:mapu:19b}
        \begin{xlist}
        \exi{A: }\label{ex:mapu:19b1}
            \gll Allkü-la-i-m-i nga? \\
                 hear-\textsc{neg-ind-2-sg} \textsc{dp} \\
            \glt ‘Didn’t you (\textsc{sg}) hear?’
        \exi{B: }\label{ex:mapu:19b2}
            \gll Mü  (allkü-la-n). \\
                 \textsc{neg} hear-\textsc{neg-1sg.ind} \\
            \glt ‘No, I didn’t.’ \citep[166]{Moesbach1962} 
        \end{xlist}
    \z 
\z

Straightforward negative questions like \REF{ex:mapu:19b} above appear to be somewhat disfavored in normal discourse; they are extremely rare in narratives, and speakers seem to prefer to embed negative expressions in affirmative questions and/or reported-speech structures, especially if the clause consists of more than a more predicate, but more research is needed here.

Both affirmative and negative replies above refer to the propositional content, rather than to the polarity of the question. However, see the following example of the usage of \textit{may} in \REF{ex:mapu:20b}, in which it agrees with the polarity of the negative assertion in \REF{ex:mapu:20a} made in a previous section of a reported dialog taken from \citet[226--227]{Salas2006}.


\ea\label{ex:mapu:20}    
    \ea\label{ex:mapu:20a}
        \gll Eymu  ta  ka  fele-la-fu-i-m-u    ka. \\ 
             \textsc{2du} \textsc{ptcl} \textsc{ptcl} be.so-\textsc{neg-ri-ind-2-du} \textsc{dp} \\
        \glt ‘You two would not be like that, either.’
   
    \ex\label{ex:mapu:20b}
        \gll Fele-i may. Iñchiw  may  fele-la-fu-i-i-u. \\
             be.so-\textsc{ind[3sg]} \textsc{aff} \textsc{1du} \textsc{aff} be.so\nobreakdash-\textsc{neg-ri-ind-1-du}\\
        \glt ‘It is true. We both were certainly not like that.’
    \z
\z 

It is the postverbal element \textit{ka} (cf. \ref{ex:mapu:20a}) that is used to refer to the polarity of the question, similarly to the German polarity-reversing particle \textit{doch} (see \citealt{Moser2019});\footnote{Like \textit{rume} (see \sectref{sec:mapu:3.2} and \sectref{sec:mapu:4.5}), \textit{ka} is notoriously polyfunctional: as a verbal suffix, it expresses continuity; as an adjectival root, it means ‘(an)other’; as a conjunction, it means ‘and’; preposed to a verb, it usually means ‘again’ (but apparently not in 20a).} see example \REF{ex:mapu:21} \citep[based on][71]{Augusta1916}. 

\ea\label{ex:mapu:21} 
    \begin{xlist}
    \exi{Q.}\label{ex:mapu:21a}
        \gll Amu-la-i-m-i    nga? \\ 
             go-\textsc{neg-ind-2-sg}  \textsc{dp} \\
        \glt ‘Didn’t you (\textsc{sg}) go?’
    
    \exi{A.}\label{ex:mapu:21b}
        \gll Amu-n      ka. \\
             go-\textsc{1sg.ind}     \textsc{dp} \\
        \glt ‘I did go.’
    \end{xlist}
\z 

\subsection{Negative indefinites}
\label{sec:mapu:3.2}

As in other South American languages (e.g., in Quechuan), the negator \textit{nu} co-occurs with \textit{rume} ‘-ever, even, at least’ in negative indefinites, as shown in \REF{ex:mapu:22}.\footnote{See also \sectref{sec:mapu:4.5} for other uses of the negators and \textit{rume}. The latter element has a wide variety of (arguably related) functions: as a verbal suffix, it means ‘suddenly’; as an adverb preceding adjectives, it means ‘very, much, exceedingly’; as a bimorphemic verb stem, it means ‘pass there’.} 

\ea\label{ex:mapu:22}
    \ea\label{ex:mapu:22a}
        \gll \textit{iney-rume}   ‘who(m)ever’\\ 
             \textit{iney-nu-rume} ‘nobody’  (cf. \textit{iney} ‘who’)\\

    \ex\label{ex:mapu:22b}
        \gll \textit{chem-rume} ‘whatever’ \\
             \textit{chem-nu-rume} ‘nothing’  (cf. \textit{chem} ‘what’) \\

    \ex\label{ex:mapu:22c}
        \gll \textit{chew-rume} ‘wherever’ \\
             \textit{chew-nu-rume} ‘nowhere’  (cf. \textit{chew} ‘where’) \\
    \z
\z 
 
Note the opposition in \REF{ex:mapu:23} between the negator appearing only on the verb \REF{ex:mapu:23a}, the negator appearing only in the indefinite expression \REF{ex:mapu:23b}, and the co-occurrence of two negators in the same clause \REF{ex:mapu:23c}. (The latter is possibly found with an emphatic function, but more research is needed on this.)

\ea\label{ex:mapu:23}
    \ea\label{ex:mapu:23a}
        \gll Doy  chem-rume  nie-ke-la-fu-i-i-n. \\ 
             more  what-ever  have-\textsc{hab-neg-ri-ind-1-pl} \\
        \glt ‘We (\textsc{pl}) did not have anything more.’\footnote{The marker -\textit{fu} ‘\textsc{ruptured} \textsc{implicature}’ is a “metapragmatic operator that acts to warn against, block, or rupture conventional implicature in speech” \citep[240]{Golluscio2000}. It can have a temporal interpretation (‘it is no longer the case that [\textsc{clause}]’), but it is often orthogonal to temporal considerations (‘whatever we expected to happen because of [\textsc{clause}] did not come to pass’).} \citep[244]{Smeets2008}
    
    \ex\label{ex:mapu:23b}
        \gll Doy  chem-nu-rume  nie-ke-fu-i-i-n. \\
             more  what-\textsc{neg-}ever  have-\textsc{hab-ri-ind-1-pl}\\
        \glt ‘We (\textsc{pl}) had nothing more.’ \citep[244]{Smeets2008}
        \ex\label{ex:mapu:23c}
        \gll Chumkaw-nu-rume    feypi-la-a-n. \\
             how-\textsc{neg}{}-ever    say-\textsc{neg-fut-1sg.ind} \\
        \glt ‘I won’t say it under any circumstances.’ (\citealt{Catrileo2010}: 129; orthography adapted)
    \z
\z 
 
 \subsection{Negative derivation and case marking}

There do not appear to exist any lexically idiosyncratic negatives in the language. By a similar token, none of the negators are used as the sole means to derive lexemes. In particular, \textit{nu} is not used as English \nobreakdash-\textit{less} with nominal or adjectival stems, or as Latin/Spanish \textit{dis}{}- {\textasciitilde} \textit{des}{}- {\textasciitilde} \textit{de-} (e.g., \textit{disculpar} ‘excuse’) \textit{an}{}- or \textit{iN}{}- (e.g., \textit{analfabeto} ‘illiterate’), or \textit{no} (e.g., \textit{tratado de no proliferación (de armas nucleares)} ‘non-proliferation treaty (for nuclear weapons)’), with nominal or verbal stems. Moreover, Mapudungun speakers do not normally borrow Spanish negative-prefix items but prefer to resort to the vernacular means of expression, i.e. verbal negation. Consider, for instance, that the language expresses ‘be difficult’ and ‘not be difficult / be easy’, respectively, via verbalization: \textit{küdaw-nge}{}- (work-be\textsubscript{1}{}-) and \textit{küdaw-nge-la}{}- (work-be\textsubscript{1}{}-\textsc{neg}{}-). By a similar token, a negated nonfinite \textit{el}{}-verb form is turned into an attributive adjective in \REF{ex:mapu:24}.

\ea\label{ex:mapu:24}
    \gll rakiduam-nie-no-el-chi    antü  mew \\ 
         think-have-\textsc{neg-nfin-attr} day  \textsc{posp} \\
    \glt ‘someday soon’ (lit. ‘on the day not thought of’, possibly from Spanish \textit{el día menos pensado}) \citep[119]{Augusta1916}
\z 

Perhaps unsurprisingly, the only element comparable to the English or Spanish negative prefixes is originally verbal, namely \textit{nge-no-} ‘not being’ (cf. \textit{ngeno longko} ‘headless’, lit. ‘without a head’). In most nonverbal instances, \textit{ngeno} seems to be an adposition (‘without’) rather than a prefix (‘\nobreakdash-less’), cf. \textit{ngeno düngu} ‘without cause, for no reason’ vs. \textit{ngeno fücha düngu} ‘for no good reason’ \citep[67]{Augusta1916}. \citet{Augusta1916} gives a handful of adverbial nonfinite forms that might be good candidates for lexicalization (e.g. \textit{ngeno-rakiduam-n} ‘inconsiderately’, lit. ‘not thinking’; \textit{ngeno-yewe-n} ‘shamelessly’, lit. ‘not being ashamed’), but this area needs much more research. My impression is that present-day speakers regularly use nominalizations that possibly show univerbation, like \textit{ñi newen-nge-no-n} (\textsc{1sg.poss} strength-be\textsc{\textsubscript{1}}\textsc{{}-neg-nfin}) ‘my weakness’, lit. ‘my not having strength’ \citep[102]{Augusta1916}. 

\section{Other aspects of negation}
\label{sec:mapu:4}

Sections \sectref{sec:mapu:2} and \sectref{sec:mapu:3} above outline our current knowledge about negative structures in Mapudungun. Nevertheless, several additional features of Mapudungun grammatical and lexical structure need to be explored in greater detail, both for the language in general and in the context of negation, in order to arrive at a really comprehensive picture. For the most part, I limit myself in the following subsections to making explicit suggestions for further research.

\subsection{The scope of negation}

Any elements other than verbs that need to be negated take the constituent negator \textit{nu {\textasciitilde} no}. Example \REF{ex:mapu:25} illustrates a negated adverbial phrase.

\ea\label{ex:mapu:25}
    \gll Dewma  mari  tripantu  nie-el    tripa-n ñi    küdaw-a-el,    welu  müte kamapu  nu.\\ 
         already  ten  year    have-\textsc{nfin} exit-\textsc{1sg.ind} \textsc{1sg.poss} work-\textsc{fut-nfin} but  very far.away  \textsc{neg}\\
    \glt ‘When I was already ten years old, I went away to work, but not very far away.’ \citep[245]{Smeets2008}
\z 
 
Not much is known about scope relations in general in Mapudungun. The scope of negation needs to be addressed by further research.

\subsection{Negative polarity}
Despite some pioneering work on particles in \citet[Ch.32]{Smeets2008}, Mapudungun particles are still only poorly understood, so at least some negative or affirmative polarity items might be found there once those items are analyzed thoroughly and exhaustively.  

Mapudungun equivalents of English pairs like \textit{no longer / any longer} make transparent use of negators, but equivalents of \textit{somewhat, at all}, etc. do not seem to exist in the language. The main element to take into account here is \textit{(-)rume} ‘\nobreakdash-ever, even, at least’. In addition to what is outlined in Sections 3.2 and 4.5, see for instance the elements \textit{chumül/chumel} ‘when’ and \textit{chumngechi} ‘how’, which can be used interrogatively and in assertions, either with or without negation, to signify ‘always’/’never’ and ‘however’/‘in no way whatsoever’, respectively. It is worth noting that these items differ as to their status as negative polarity items.

Two items related to \textit{chumül/chumel} ‘when’ merit special attention in the context of polarity items, namely \textit{rumel} and \textit{turpu}. They both seemed to have been used interchangeably a century ago, with the meaning ‘always’ with an unmarked verb and ‘never’ with a negated verb (as unmistakably exemplified in \citealt{Augusta1903,Augusta1916}), and without the elements \nobreakdash-\textit{no-rume} (found on \textit{chumülnorume} ‘never’, for instance). Nevertheless, present-day speakers appear to prefer the distribution described in \citet{Smeets2008}, where \textit{rumel} is invariably used with unmarked verbs \REF{ex:mapu:26a} and \textit{turpu} only occurs with negated verbs \REF{ex:mapu:26b} (and is customarily translated as ‘never’).\footnote{The specialization process seems to have begun a long time ago. In \citet{Coña1995}, which collects narratives by a Mapuche chieftain of the early 20th century, \textit{turpu} appears only with negated verbs, and \textit{rumel} appears five times with an unmarked verb and only once with a negated verb.}

\ea\label{ex:mapu:26}
    
    \ea\label{ex:mapu:26a}
        \gll Rumel    küpal-a-i      yewün. \\ 
             always\textsubscript{1}  bring-\textsc{fut-ind[3pl]} present(s) \\
        \glt ‘They always bring presents.’ \citep[555]{Smeets2008}
    
    \ex\label{ex:mapu:26b}
        \gll Eymi  turpu    tripa-ke-la-i-m-i. \\
             \textsc{2sg} always\textsubscript{2}  go.out-\textsc{hab-neg-ind-2-sg} \\
        \glt  ‘You (\textsc{sg}) never go out.’ \citep[559]{Smeets2008}
    
    \z
\z 

\subsection{Marking of NPs in the scope of negation}

Mapudungun does not have case morphology or adpositions that mark either core or peripheral grammatical relations. The only candidate for case status is the enclitic postposition \textit{mew} {\textasciitilde} \textit{mu} {\textasciitilde} \textit{mo}, but it does not interact with negation in any special way,\footnote{See \citet[61--67]{Smeets2008} who treats the semantically void postposition \textit{mew} as “instrumental case suffix”.}
so no case alternations of the Russian, Finnish, or Basque type are found in the language. To the best of my knowledge, neither spatial, temporal nor causal adpositions interact with negation in any way.

Neither the conditions governing the use of determiners nor the regularities shown by information-structure categories like topics and focuses are well understood for Mapudungun, so this is an area where much more research is needed.

\subsection{Reinforcing negation}

Very little is known about Mapudungun prosody, so future research will hopefully show how negation can be prosodically reinforced, if at all. The only pattern I am aware of that may turn out to be related to reinforcement is the co-occurrence of two negators in the same clause (example \REF{ex:mapu:23c} above). See also example \REF{ex:mapu:29} below. 

\subsection{Negation, coordination and complex clauses}
\label{sec:mapu:4.5}

With affirmative clauses, \textit{ka} ‘and’ can, but need not, be used as coordinator. The Mapudungun equivalent to expressions with negative coordinators (like English \textit{neither … nor…}, or Spanish \textit{ni… ni…}) uses the element \textit{rume} ‘even, at least’ (see \sectref{sec:mapu:3.2}) in the units following the first one (if they are only two, in the second unit). See in example \REF{ex:mapu:27} from \citet[253]{Augusta1916} how all units may structurally parallel, each with the same negator \REF{ex:mapu:27a}, or only the first is a complete clause while the others are elliptical \REF{ex:mapu:27b}.

\ea\label{ex:mapu:27}
    
    \ea\label{ex:mapu:27a}
        \gll I-ke-la-n,      umag-ke-la-n rume. \\ 
             eat-\textsc{hab-neg-1sg.ind} sleep\textsc{{}-hab-neg-1sg.ind} even \\
        \glt ‘I neither eat nor sleep.’
    
    \ex\label{ex:mapu:27b}
        \gll Juan  “razón  nie-i”  pi-la-a-i-m-i-mew, Pedro    no  rume.\\
             Juan  reason    have-\textsc{ind[3sg]} say-\textsc{neg-fut-ind-2-sg-3A} P. \textsc{neg} even\\
        \glt ‘Neither Juan nor Pedro will agree with you (\textsc{sg}).’ (lit. ‘Juan won’t tell you “It’s correct,” [and] neither will Pedro.’)
    
    \z
\z 
 
The language has an adversative conjunction \textit{welu} ‘but’, equivalent to English \textit{but} (and Spanish \textit{pero}) in \textit{this is not a cat, but it is a dog}, but there is no equivalent to English \textit{but} (and Spanish \textit{sino}) in \textit{this is not a cat but a dog}. The usual way to express [\textit{not} \textsc{proposition}\textsubscript{1} \textit{but} \textsc{proposition}\textsubscript{2}] is simply to express the first proposition in the negative and the second in the affirmative, and juxtapose both, as in example \REF{ex:mapu:28}.
 
\ea\label{ex:mapu:28}
    \gll Juan  fem-la-i,      Pedro    fem-i. \\ 
         Juan  do.so-\textsc{neg-ind[3sg]} \textsc{P.} do.so-\textsc{ind[3sg]} \\
    \glt ‘Juan did not do it, but Pedro did.’ \citep[356]{Augusta1916}

\z 
 
I am not aware of any element equivalent to the English negative conjunction \textit{lest} in Mapudungun.

\subsection{Miscellaneous aspects of negation}

Unfortunately, not much is known about clause linkage idiosyncrasies, and metalinguistic targeting (e.g., “metalinguistic negation”). Raising/control phenomena have attracted some attention in the literature (see some notes on argument marking with affirmative raising and control predicates in \citealt{Zúñiga2019}, and references therein), but the question of negative transport (“negation raising”) has not been explored yet.

Co-occurrence of different negators in different loci of the same clause is found in corpora and can be easily elicited, but the factors conditioning this phenomenon, as well as its possible interaction with parallel co-occurrences in non-standard Southern Cone Spanish, have not been studied yet (as mentioned in \sectref{sec:mapu:3.2}). It seems that reinforcement/emphasis of some sort is involved; the idea conveyed by \textit{chumkawnorume} in \REF{ex:mapu:29} is ‘no matter how hard they tried’.


\ea\label{ex:mapu:29}
    \gll Kiñe  kentu    tu-i-ngün    toki. Feymew  ulel-fi-i    tachi  pu  orkon katr-a-fu-el    engün,  welu chumkaw-no-rume    katrü-la-fi-i-ngün.\\ 
         one  group    take-\textsc{ind-3pl} axe with.them  hit-\textsc{3obj-ind[3]}  \textsc{art}  \textsc{pl} post cut-\textsc{fut-ri-nfin} \textsc{3pl} but how-\textsc{neg-}ever    cut-\textsc{neg-3obj-ind-3pl}\\
    \glt ‘A group (of people) took some axes. They used them to hit the posts in order to cut them, but they did not succeed.’ (lit. ‘they did not cut them.’) (\citealt{Salas2006}: 248; orthography adapted)
\z 

 It is worth noting that the verbal negators co-occur with some other markers in combinations that have drawn some attention in the previous literature and will hopefully be addressed systematically soon \citep{ChandiaEtAl-inprep}. The markers -\textit{we} ‘already’ and -\textit{pe} ‘\textsc{dubitative}’ are particularly salient in this regard. I have glossed the combined sequence \textit{{}-we-la} as cessative elsewhere \citep{Zúñiga2006}, because it expresses the meaning ‘no more, not anymore’ \REF{ex:mapu:30a}; it is compositional, however, and partly unexpected only because other languages belong to the ‘no more’ type instead of the ‘already not’ type \citep{Auwera1998}. The sequence \textit{-pe-la}, by contrast, is not compositional: it expresses confidence/likelyhood, i.e., an epistemic modal meaning not simply derivable from the negated dubitative \REF{ex:mapu:30b}.
% 
\ea\label{ex:mapu:30}
    \ea\label{ex:mapu:30a}
        \gll Doy    wangkü-l-nge-we-la-i. \\ 
             more    bark-\textsc{appl-pass-}already-\textsc{neg-ind[3sg]} \\
        \glt ‘They did not bark at him anymore.’ \citep[262]{Salas2006}
    
    \ex\label{ex:mapu:30b}
        \gll Aña-pela-a-i. \\
             hurt-\textsc{mod-fut-ind[3sg]} \\
        \glt ‘It is probably going to hurt.’ \citep[250]{Smeets2008}\\
     (cf. \textit{aña-pe-a-i} ‘it may hurt’)
    \z
\z 
 
Finally, as to the diachrony of negation, nothing is known about the negators’ origins – which is unlikely to change in the future, unless some documented languages can be shown to be genealogically related to Mapudugun and some linguistic reconstruction can be conducted. The oldest grammatical description \citep{Valdivia1606} already records \nobreakdash-\textit{nu} in the same form as it is found today, making it unlikely for the marker itself to be a Spanish loan. This may also hold for the particle \textit{nu}.

The apparent expansion of -\textit{nu} in the imperative paradigms is possibly due to, or may at least have been reinforced by, contact with Southern Cone Spanish. In the latter language, the only dedicated imperative form is \textsc{2sg} affirmative \textit{canta} ‘sing!’;\footnote{This also holds for Rioplatense Spanish and other similar varieties, which have, e.g., a \textit{voseante} \textsc{2sg} imperative \textit{cantá} alongside or instead of the etymological, non-\textit{voseante}, \textit{canta}. See \citet{BaqueroWestphal2014} for a recent analysis and references.} all others are (originally) subjunctive forms (see \tabref{tab:mapu:4}). (Note that imperative and subjunctive forms still differ in the affirmative with respect to the admissible location of the clitic group (\textsc{clt}).)


\begin{table}
    \small
    \begin{tabularx}{\textwidth}{Xcccc}    
    \lsptoprule
    & \multicolumn{2}{c}{{Affirmative}} & \multicolumn{2}{c}{{Negative}} \\    
    \cmidrule(lr){2-3}\cmidrule(lr){4-5}
    & \multicolumn{1}{c}{{Imperative(=\textsc{clt})}} & {(\textsc{clt}=)Subjunctive} & \multicolumn{1}{c}{{(\textsc{clt}=)Imperative}} & {(\textsc{clt}=)Subjunctive}\\
    \midrule 
    {\scshape 2sg} & \textit{cant-a} & \textit{cant-e-s} & \multicolumn{2}{c}{\textit{no cant-e-s}} \\
    {\scshape 3sg} & \multicolumn{2}{c}{\textit{cant-e}} & \multicolumn{2}{c}{\textit{no cant-e}} \\
    {\scshape 1pl} & \multicolumn{2}{c}{\textit{cant-e-mos}} & \multicolumn{2}{c}{\textit{no cant-e-mos}} \\
    {\scshape 2pl/3pl} & \multicolumn{2}{c}{\textit{cant-e-n}} & \multicolumn{2}{c}{\textit{no cant-e-n}} \\
    \lspbottomrule
    \end{tabularx}
%    \renewcommand\thetable{4} %needed here because of table 2a and 2b before
    \caption{Southern Cone Spanish imperative and present subjunctive (\textit{a}{}-class verbs)}
    \label{tab:mapu:4}
\end{table}

\section{Summary}
\label{sec:mapu:5}

The negators presented and discussed in this paper are summarized in \figref{figtab:mapu:5} below. The standard negator \nobreakdash-\textit{la} and the imperative negative \nobreakdash-\textit{ki} have a limited occurrence vis-à-vis the default negators \nobreakdash-\textit{nu} (bound) and \textit{nu} (possibly more autonomous), which occur in a variety of contexts. Rather than transitivity, tense, or aspect, the factors conditioning the observed allomorphy in Mapudungun are finiteness and modality (with clausality and verbality playing a background role, so to speak).

% \hc{@Sebastian /LSP: Can you make that table look good? That would be the author's first choice. If that is not possible, then the tree would be possible too}
\begin{figure} %HC table looks bad --> let's do a forest here too. FZ is ok (but overall he would prefer tables than forests if it is possible). I modified the table a bit, so maybe the table is good after all? @Sebastian what do you think?
% \begin{tabularx}{\textwidth}{XXXXX}
%     \lsptoprule
%     \multirow{5}{*}{{Clausal}} & \multirow{4}{*}{{Verbal}} & \multirow{3}{*}{{Finite}} & {Indicative} & \textit{-la}\\
% \cmidrule{4-5}
%     &  &  & {Imperative} & \textit{-ki(-nu)}\\
% \cmidrule{4-5}
%     &  &  & {Subjunctive} & \textit{-nu}\\
% \cmidrule{3-5}
%     &  & {Nonfinite} &  & \textit{-nu}\\
% \cmidrule{2-5}
%     & \multicolumn{3}{l}{{Nonverbal (equation)}} & \textit{nu}\\
% \cmidrule{1-5}
%     \multirow{2}{*}{{Nonclausal}} & {Indefinite} &  &  & \textit{-nu}\\
% \cmidrule{2-5}
%     & {Else} &  &  & \textit{nu}\\
%     \lspbottomrule
% \end{tabularx}

\fittable{
\begin{forest} for tree={grow'=east,delay={where content={}{shape=coordinate}{}}},   forked edges,
[negators
    [clausal
        [verbal
            [finite
                [indicative 
                    [\textit{-la}, tier=word]
                ]
                [imperative 
                    [\textit{-ki(-nu)}, tier=word]
                ]
                [subjunctive 
                    [\textit{-nu}, tier=word]
                ]
            ]
            [nonfinite 
                [\textit{-nu}, tier=word]
            ]
        ]
        [nonverbal (equation)
            [\textit{nu}, tier=word]
        ]
    ]
    [nonclausal
        [indefinite
            [\textit{-nu}, tier=word]
        ]
        [else
            [\textit{nu}, tier=word]
        ]
    ]
]
\end{forest}
}

%    \renewcommand\thetable{5} %needed here because of table 2a and 2b before
    \caption{Overview of Mapudungun negators}
    \label{figtab:mapu:5}
\end{figure}

\section*{Abbreviations}
\begin{tabularx}{.45\textwidth}{lQ}
1 & first person \\
2 & second person \\
2>1 & second person acting on first person\\ 
3 & third person \\
A & agentive participant of bivalent predicate\\  
\textsc{aff} & affirmative\\ 
\textsc{appl} & applicative\\ \textsc{art} & article \\ 
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\textsc{attr} & attributive \\
\textsc{clt} & clitic group\\
\textsc{cis} & cislocative\\
\textsc{cont} & continuative\\
\textsc{dem} & demonstrative\\
\textsc{dp} & discourse particle\\
\textsc{du} & dual\\
\textsc{els} & extended local scenario\\
\textsc{fut} & future\\
\textsc{hab} & habitual\\
\end{tabularx}

\begin{tabularx}{.45\textwidth}{lQ}
\textsc{imp} & imperative\\ 
\textsc{ind} & indicative \\
\textsc{inf} & infinitive\\ 
\textsc{mod} & modality\\
%N & noun\\ 
\textsc{neg} & negation \\ 
\textsc{nfin} & nonfinite\\ 
NP & noun phrase\\ 
\textsc{nsg} & nonsingular\\ 
\textsc{obj} & object\\ 
\textsc{ptcl} & particle\\
\textsc{pass} & passive\\ 
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\textsc{pl} & plural\\ 
\textsc{posp} & postposition\\ 
\textsc{prog} & progressive\\ 
\textsc{psr} & possessor\\ 
\textsc{rep} & reportative\\
\textsc{res} & resultative \\ 
\textsc{rest} & restitutive\\ 
\textsc{ri} & ruptured implicature\\
\textsc{sbjv} & subjunctive \\ 
\textsc{sg} & singular\\ 
\\
\end{tabularx}

\sloppy\printbibliography[heading=subbibliography,notkeyword=this]

\end{document} 
