\documentclass[output=paper,colorlinks,citecolor=brown,draft]{langscibook}
\ChapterDOI{10.5281/zenodo.18234095}
\author{Carmen Dagostino\affiliation{California State University, San Bernardino}}
\title{Negation in Chuxnabán Mixe}
\abstract{Chuxnabán Mixe, a Mixe-Zoquean language spoken in Oaxaca, Mexico, has a complex grammar of negation with different constructions for different types of negative expressions and negative constructions interacting with the grammatical structure of the clause. This chapter zeroes in on the functions and uses of the negative particles \textit{ka’ap} and \textit{këtii}, the negative prefix \textit{ka-}, and the proclitic \textit{nii=}. Further, the interaction with other structures and multiple expressions of negation are examined, in addition to special forms in prohibitives, negative existentials, and multiple negation strategies in non-verbal clauses. 
In general, the negative particle \textit{ka’ap} can be used in all types of negation. It always precedes the predicate, and its presence triggers a special verb form, including a separate set of person markers. The verb may also include the negative prefix \textit{ka-} instead of or in addition to \textit{ka’ap}. The negative particle \textit{ka’ap} can also directly precede a constituent to be negated. Constituent negation may also be expressed with the negative proclitic \textit{nii=}.  A special negative particle \textit{këtii} is used in prohibitives, in clauses with ‘neither’, in negative polar questions (also non-verbal and existential polar questions), and in tag questions. There is no formal distinction between prohibitives and negative polar questions. Interesting is the fact that non-verbal clauses exhibit multiple strategies for both affirmatives and negatives with different verbalizers and the irrealis enclitic \textit{=ëch} in some negative forms
\keywords{Mixe-Zoquean languages, negative particles, prohibitives, non-verbal negation, negative prefix}
}
\IfFileExists{../localcommands.tex}{
 \addbibresource{../localbibliography.bib}
 \input{../localpackages}
 \input{../localcommands} 
 \input{../localhyphenation} 
 \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 
\section{The language}\label{sec:mix:1}

Chuxnabán Mixe (ISO: \href{https://iso639-3.sil.org/code/pxm}{pxm}, Glottocode: \href{https://glottolog.org/resource/languoid/id/chux1234}{chux1234}) is a Mixe-Zoquean language spoken by about nine hundred people in one village, San Juan Bosco Chuxnabán, in Mexico (latitude/longitude: 17°01’09.5”N 95°49’48.2”W). Linguistic vitality is \textit{unsafe}, \textit{vulnerable} or \textit{at risk} following the UNESCO framework \citep{BrenzingerEtAl2003}, given that education and literacy development occur almost exclusively in the dominant language Spanish. Nevertheless, all community members still speak the language, and there is intergenerational language transmission.

The Mixean territory includes two hundred and ninety communities divided into nineteen municipalities \citep{TorresCisneros1997}. The dialects of the villages within a single municipality are generally mutually intelligible, although speakers identify dialectal differences. Geographically, the region differentiates highlands, midlands, and lowlands. The geographic groupings are somewhat reflected in the genetic division of the languages. The Mixe-Zoquean language family includes seventeen languages, according to Ethnologue, subdivided into a Mixean and a Zoquean branch \citep{EberhardEtAl2019}. Chuxnabán Mixe, a Oaxaca Mixean language, belongs to Quetzaltepec Mixe, which is also the municipality of the village. Looking at diachronic developments within the language family, \citet[10]{Wichmann1995} identifies Lowland, Midland, South Highland, and North Highland Mixe. Chuxnabán Mixe belongs to Midland Mixe. Overall, the Mixean dialects spoken in the two hundred ninety communities have been classified for the most part as eight distinct languages with some differences between the sources. 

The data for this paper stems from personal fieldwork in the village in 2008 and 2011 where I collected 60 oral narratives (1--15 minutes in length) and 2 dinner-table conversations (30 and 50 minutes in length). I transcribed both conversations and 11 narratives that yielded over 250 examples of negation from the naturally occurring data. This data was complemented with elicitation sessions in 2018 to obtain positive clauses and to fill gaps. Data from grammars of related languages, i.e. Ayutla Mixe \citep{Romero-Méndez2009} and Sierra Popoluca \citep{Boudreault2018}, was also studied. Unfortunately, negation has received little attention in the grammars, and no detailed study of negation in any Mixean language is available. This paper constitutes the first detailed treatment of the topic within the language family.

The description of negation in Chuxnabán Mixe in this chapter is based on the questionnaire provided by the editors \parencitetv{chapters/questionnaire} and mirrors its structure.

\subsection{Typological profile}\label{sec:mix:1.2}

Chuxnabán Mixe can be characterized as polysynthetic and head-marking with a hierarchical and inverse alignment system based on person, animacy, and topicality. There are independent and dependent verb paradigms for person and aspect and modal markers, although the dependent inflection is a highly language specific category and not related to clausal dependency. These verb inflections also involve verb stem alterations. There is no nominal case marking, except for the locative; and there is optional number marking for animates. Word order is relatively flexible, but generally verb-final. In what follows, I briefly present the person marking system, the dependent and independent verb inflections, and the TAM system. 

The most prominent participant in an event is marked or cross-referenced on the predicate, following a hierarchical and inverse grammatical system. The hierarchy is determined by person: first > second > third, animacy: animate > inanimate, and topicality: human/topical > human/non-topical. The position of a participant on the hierarchy is further signaled by direct or inverse alignment. When participants in a clause are such that the actor outranks the undergoer, there is direct alignment. If the opposite occurs, there is inverse alignment, indicated with a special inverse suffix \textit{-ë} (only marked if a third person acts on a second or if a less topical third person acts on a more topical third person). The grammatical indexing system is summarized in \tabref{tab:mix:1} and the indexes in \tabref{tab:mix:2}. 

\begin{table}
\begin{tabularx}{0.55\textwidth}{lll}
\lsptoprule
{Relation} & {Marking} & {Inverse marker}\\
\midrule
{1 > 2, 1 > 3} & {1.A} & {{}-}\\
{2 > 1, 3 > 1} & {1.P} & {no}\\
{2 > 3} & {2} & {{}-}\\
{3 > 2} & {2} & {yes}\\
{3 > 3’} & {3.A} & {{}-}\\
{3’ > 3} & {3.P} & {yes}\\
\lspbottomrule
\end{tabularx}
\caption{Grammatical indexing system}
\label{tab:mix:1}
\end{table}

\begin{table}[b]
\begin{tabularx}{\textwidth}{lllX llll}
\lsptoprule
\multicolumn{4}{c}{{Independent paradigm}} & \multicolumn{4}{c}{{Dependent paradigm}}\\
\cmidrule(lr){1-4}\cmidrule(lr){5-8}
{Person} & {S} & {A} & {P} & {Person} & {S} & {A} & {P}\\
\midrule
\textsc{1sg} & {ø-} & \textit{n-} & \textit{x-} & \textsc{1sg} & \textit{n-} & \textit{n-} & \textit{x-}\\
\textsc{2sg} & \textit{m-} & \textit{m-} & \textit{m-} & \textsc{2sg} & \textit{m-} & \textit{x-} & \textit{m-}\\
\textsc{3sg} & {ø-} & \textit{y-} & {ø-} & \textsc{3sg} & \textit{y-} & \textit{t-} & \textit{y-}\\
\textsc{1pl.incl} & {ø- {}\textit{-ëm}} & \textit{n- {}-ëm} & \textit{x- {}-ëm} & \textsc{1pl.incl} & \textit{n- {}-ëm} & \textit{n- {}-ëm} & \textit{x- {}-ëm}\\
\textsc{1pl.excl} & {ø- {}\textit{-të}} & \textit{n- {}-të} & \textit{x- {}-të} & \textsc{1pl.excl} & \textit{n- {}-të} & \textit{n- {}-të} & \textit{x- {}-të}\\
\textsc{2pl} & \textit{m- {}-të} & \textit{m- {}-të} & \textit{m- {}-të} & \textsc{2pl} & \textit{m- {}-të} & \textit{x- {}-të} & \textit{m- {}-të}\\
\textsc{3pl} & {ø- {}\textit{-të}} & \textit{y- {}-të} & {ø- {}\textit{-të}} & \textsc{3pl} & \textit{y- {}-të} & \textit{t- {}-të} & \textit{y- {}-të}\\
\lspbottomrule
\end{tabularx}
\caption{Grammatical indexes}
\label{tab:mix:2}
\end{table}
% \begin{table}%HC this is two tables instead of the big one above.
%     \begin{tabularx}{0.5\textwidth}{lXXX}
%     \lsptoprule
%     {Person} & {S} & {A} & {O} \\
%     \midrule
%     \textsc{1sg} & {ø-} & \textit{n-} & \textit{x-} \\
%     \textsc{2sg} & \textit{m-} & \textit{m-} & \textit{m-} \\
%     \textsc{3sg} & {ø-} & \textit{y-} & {ø-} \\
%     \textsc{1pl.incl} & {ø- {}\textit{-ëm}} & \textit{n- {}-ëm} & \textit{x- {}-ëm} \\
%     \textsc{1pl.excl} & {ø- {}\textit{-të}} & \textit{n- {}-të} & \textit{x- {}-të} \\
%     \textsc{2pl} & \textit{m- {}-të} & \textit{m- {}-të} & \textit{m- {}-të} \\
%     \textsc{3pl} & {ø- {}\textit{-të}} & \textit{y- {}-të} & {ø- {}\textit{-të}} \\
%     \lspbottomrule
%     \end{tabularx}
%     \caption{Grammatical indexes: Independent paradigm}
%     \label{tab:mix:2.1}
% \end{table}
% \begin{table}
%     \begin{tabularx}{0.5\textwidth}{lXXX}
%     \lsptoprule
%     {Person} & {S} & {A} & {O}\\
%     \midrule
%     \textsc{1sg} & \textit{n-} & \textit{n-} & \textit{x-}\\
%     \textsc{2sg} & \textit{m-} & \textit{x-} & \textit{m-}\\
%     \textsc{3sg} & \textit{y-} & \textit{t-} & \textit{y-}\\
%     \textsc{1pl.incl} & \textit{n- {}-ëm} & \textit{n- {}-ëm} & \textit{x- {}-ëm}\\
%     \textsc{1pl.excl} & \textit{n- {}-të} & \textit{n- {}-të} & \textit{x- {}-të}\\
%     \textsc{2pl} & \textit{m- {}-të} & \textit{x- {}-të} & \textit{m- {}-të}\\
%     \textsc{3pl} & \textit{y- {}-të} & \textit{t- {}-të} & \textit{y- {}-të}\\
%     \lspbottomrule
%     \end{tabularx}
%     \caption{Grammatical indexes: Dependent paradigm}
%     \label{tab:mix:2.2}
% \end{table}


In addition to person, singular and plural forms are distinguished, as shown in \REF{ex:mix:1}. 

{\ea\label{ex:mix:1} Person hierarchy 1 > 2 > 3
    \ea\label{ex:mix:1a}
        \glll xyuujxp. \\ 
             \textbf{x-}yuujx-p \\
             1\textsc{p}{}-wake.up-\textsc{asp.ind} \\
        \glt ‘You/s/he wake(s) me up.’
    
    \ex\label{ex:mix:1b}
        \glll myuujxëp. \\
              \textbf{m}{}-yuujx-\textbf{ë}{}-p \\
              2\textsc{o}{}-wake.up-\textbf{\textsc{inv}}\textsc{{}-asp.ind}\\
        \glt  ‘S/he wakes you up.’
    \ex\label{ex:mix:1c}
        \glll myuujxtëp.\\
             \textbf{m}{}-yuujx-\textbf{të-}p \\
             2\textsc{a}{}-wake.up-\textbf{\textsc{pl}}\textsc{{}-asp.ind}\\
        \glt ‘You wake them up.’
    \z
\z}

\largerpage[2]
The person prefixes show two inflections (see \tabref{tab:mix:2}), one for the independent and one for the dependent construction. As with other Mixean languages, all predicates are treated as either independent or dependent, each with its own set of inflectional person and aspect and mood markers, in addition to verb stem changes. In \REF{ex:mix:2a} and \REF{ex:mix:2b}, the second person actor is indicated with a different prefix in each construction, \textit{m-} and \textit{x-} respectively. Equally, the example illustrates the two different verb stem shapes and aspect markers.

{\ea\label{ex:mix:2}
    \ea\label{ex:mix:2a}Independent:\newline
        \glll maatsyüüjchpy. \\ 
              \textbf{m}{}-yaa-\uline{tsyüüjch}-\textbf{p} \\
              \textbf{\textsc{2a.ind}}\textsc{{}-caus-}hurt-\textbf{\textsc{asp.ind}} \\
        \glt ‘You hurt him.’ 
    

    \ex\label{ex:mix:2b}Dependent:\newline 
        \glll ka’ap xyaatsyüch. \\
              ka’ap \textbf{x}{}-yaa-\uline{tsyüts}\textbf{-y} \\
             \textsc{neg} \textbf{\textsc{2a.dep}}\textsc{-caus}-hurt-\textbf{\textsc{asp.dep}} \\
        \glt ‘You don’t hurt him.’
    \z
\z }

Dependent inflection is triggered when a non\nobreakdash-core constituent is placed before the predicate. This includes locative or temporal adverbs (e.g., \textit{axë’ëy} ‘yesterday’, \textit{xëmme} ‘always’ or \textit{yaa} ‘here’), temporal particles (e.g., \textit{tëë} ‘before now’, \textit{oojts} ‘past’), the negative particle \textit{ka’ap} and certain negative words formed with the negator \textit{nii=}, the dubitative \textit{wä’än}, and in questions where the interrogative word is not a core argument (e.g., \textit{ma} ‘where’). The labeling of this inflectional distinction is somewhat unfortunate, given that it is not related to dependency, but it conforms to the terminology used in previous work \citep{Romero-Méndez2009,Wichmann1995}. I have not examined the tense-aspect-mood system in detail, and the analysis presented here is based on the overall data collected (i.e., not including targeted elicitation of tense-aspect-mood), \citegen{Wichmann1995} reconstruction of Mixe-Zoquean, and \citeapo{Romero-Méndez2009} analysis of Ayutla Mixe. The different forms as they appear in Ayutla Mixe are presented in \tabref{tab:mix:3}. Note that “Stem A, B, C” are arbitrary labels used by the author to describe the stem alternations in Ayutla Mixe.

\begin{table}
\begin{tabularx}{\textwidth}{lQl@{}}
\lsptoprule
Type of clause & Independent inflection & Dependent inflection \\
\midrule 
\multirow{4}{3.2cm}{{Neutral aspect\newline \mbox{(incompletive  aspect} in  other Mixean  languages)}}
& {Stem B + {}\textit{-p} (intransitive)} & {Stem A + {}\textit{-y} (singular)} \\
& {Stem B + {}\textit{-yp} (transitive)} & {Stem B + \textit{-t} (plural)} \\
& {Stem B + {}\textit{-të} + {}\textit{-p} (plural)} &\\
\\
\midrule
 {Completive aspect} & {Stem C} & {Stem C + {}\textit{-y}}\\
& {Stem C + {}\textit{-t} (plural)} &\\
\midrule
 {Irrealis mood} & {Stem A$^{'}$ + {}\textit{-p}} & {Stem A$^{'}$ + {}\textit{-t}} \\
& {Stem A$^{'}$ + {}\textit{-të} + {}\textit{-p}} & {Stem A$^{'}$ + + {}\textit{-të {}-t}}\\
\midrule
{Perfect (e.g., neutral)} & {Stem B + {}\textit{-në} + {}\textit{-të} + {}\textit{-p} (plural)} & {Stem B + {}\textit{-në} + \textit{-t} (plural)}\\
\lspbottomrule
\end{tabularx}
\caption{Verb stems in Ayutla Mixe (from \citealt{Romero-Méndez2009})}
\label{tab:mix:3}
\end{table}
    
It is worth mentioning that in addition to the verbal suffixes there are particles encoding tense, aspect, and mood. These are independent words that occur in pre-verbal position, but otherwise show flexible word order. They include the perfective \textit{tëë} ‘before now’, the past tense marker \textit{oojts}, the imperfective \textit{iich} or \textit{tii iich}, the dubitative \textit{wä’än}, and the hypothetical \textit{keexy}.

\section{Clausal negation}\label{sec:mix:2}

In this section, I examine negation strategies based on the type of the negative marker(s) and the structure of the negative clause as a whole focusing on symmetric versus asymmetric negation in the sense of \citet{Miestamo2007,Miestamo2009neg.pragmatics,Miestamo2013-wals114,Miestamo2013-wals113,Miestamo2017} for both constructional and paradigmatic asymmetries. As has been shown \citep{Horn2010,KahrelBerg1994,Payne1997}, negation strategies may interact with marking on other constituents and may result in multiple expression of negation within a clause. The latter may result in negative concord, i.e. multiple negators in a single negation \citep{Horn2010}; as is the case with Chuxnabán Mixe. I distinguish standard negation, i.e. the negation of declarative verbal main clauses \citep{Miestamo2005}, from non-standard negation, which includes negation strategies in prohibitives, existentials, non\nobreakdash-verbal clauses, and subordinate clauses.

\subsection{Standard negation}\label{sec:mix:2.1}

It is generally acknowledged that all languages have grammatical means to negate clauses, so every language has one or more constructions that can be identified as standard negation (\citealt{Miestamo2005,Miestamo2007}). In Chuxnabán Mixe, in general and most frequently, the negative particle \textit{ka’ap} is used in standard negation. The negative particle \textit{ka’ap} always triggers dependent inflection (see also \sectref{sec:mix:2.1.1}), as shown in \REF{ex:mix:3} and \REF{ex:mix:4}.

\ea\label{ex:mix:3}
    \ea\label{ex:mix:3a}
        \glll kääpy. \\
             y-kay-p \\
             \textsc{3a.ind}{}-eat-\textsc{asp.ind} \\
        \glt ‘He eats/ate (something).’
    \ex\label{ex:mix:3b}
        \glll  ka’ap tkaanyë.\\ 
              \textbf{ka’ap} t-kay-në\\
              \textsc{neg} \textsc{3a.dep}{}-eat-\textsc{asp.dep}\\
        \glt ‘He doesn’t/didn’t eat (something).’
    \z
\z 

\ea\label{ex:mix:4}
    \ea\label{ex:mix:4a}
        \glll ja ja’ay tëëjkëtëp. \\
             ja ja’ay ø-tëëjkë-të-p \\
             \textsc{dem} people 3\textsc{s.ind}{}-enter-\textsc{pl-asp.ind}  \\
        \glt ‘The people entered.’    
    \ex\label{ex:mix:4b}
        \glll ka’ap ja ja’ay chimtë’këtë.\\ 
             \textbf{ka’ap} ja ja’ay y-tsimtë’kë{}-të \\
             \textsc{neg} \textsc{dem} people 3\textsc{s.dep}{}-enter-\textsc{pl} \\
        \glt ‘The people didn’t enter.’
    \z
\z 

The negative particle occurs most often clause-initially. It can be preceded by arguments and adverbials, as in \REF{ex:mix:5} and \REF{ex:mix:6}. 

\ea\label{ex:mix:5}
    \glll  tääpë Tinnë ka’apës oojts xnyiikaapxtu’utë.\\
         tääpë Tinnë \textbf{ka’ap}=ës oojts x-niikaapxtu’ut-të\\
         this Tino \textsc{neg-1.emph} \textsc{pst} \textsc{1p.dep}-defend-\textsc{pl}\\
    \glt ‘This Tino didn’t defend us.’
\z

\ea\label{ex:mix:6}
    \glll tää jyä’chë yaa ka’ap myo’k t’xuujtë.\\
         tää jyä’ch yaa \textbf{ka’ap} y-mo’k t-’xuuj-të\\
         before.now many here \textsc{neg} \textsc{3poss}.corn \textsc{3a.dep}{}-put.chemicals-\textsc{pl} \\
    \glt ‘Many (people) here don’t put chemicals in their corn.’
\z

In \REF{ex:mix:5} and \REF{ex:mix:6}, \textit{ka’ap} can be moved before or after any noun phrase or particle, as long as it remains in pre-verbal position. Post-verbal position is ungrammatical. Emphatic or evidential clitics can be attached to it, as in \REF{ex:mix:5} and \REF{ex:mix:7b}. In addition, a demonstrative can cliticize to the negative particle; this often seems to occur when evidential clitics are present, and the demonstrative follows the evidential, as in \REF{ex:mix:7b}. The negator \textit{ka’ap} also occurs with the particle \textit{jëk} ‘already’ to express a meaning of ‘not anymore’, as in \REF{ex:mix:8b} and \REF{ex:mix:9b}.

\ea\label{ex:mix:7}
    \ea\label{ex:mix:7a}
        \glll yë’ë pëtsë’ëmp. \\
             yë’ë ø-pëtsë’ëmp-p \\
             \textsc{dem} \textsc{3s.ind-}go-\textsc{asp.ind} \\
        \glt ‘She is going out.’
    \ex\label{ex:mix:7b}
        \glll ka’apëkjä’ä ma pyëtsiimp.\\ 
             \textbf{ka’ap}=ëk=jä’ä ma y-pëtsiimp \\
             \textsc{neg=evid=dem} there \textsc{3s.dep}{}-go.\textsc{dep} \\
        \glt ‘She is not going out.’
    \z
\z 

\ea\label{ex:mix:8}
    \ea\label{ex:mix:8a}
        \glll jak kääpy.\\
             jak ø-kääy-p \\
             still \textsc{3s.ind-}eat-\textsc{asp.ind} \\
        \glt ‘He is still eating.’
    \ex\label{ex:mix:8b}
        \glll ka’apëk tkaanyë.\\ 
             \textbf{ka’ap=jëk} t{}-kaay-në \\
             \textsc{neg=}already \textsc{3a.dep-}eat\textsc{{}-asp.dep} \\
        \glt ‘He is not eating anymore.’
    \z
\z 
 
\ea\label{ex:mix:9}
    \ea\label{ex:mix:9a}
        \glll miimpëk jä’ä mo’kx. \\
             ø-miim-p=jëk jä’ä mo’kx \\
             \textsc{3s.ind}{}-come-\textsc{asp.ind=}already \textsc{dem} opossum \\
        \glt ‘The opossum was already coming.’
    \ex\label{ex:mix:9b}
        \glll ka’apëk jä’ä mo’kx jëkmyinnë.\\ 
             \textbf{ka’ap=jëk} jä’ä mo’kx jëk=y-min-në \\
             \textsc{neg}{}-already \textsc{dem} opossum already=\textsc{3s.dep-}come-\textsc{asp.dep} \\
        \glt ‘The opossum wasn’t coming anymore.’
    \z
\z

Chuxnabán Mixe further has a negative verbal prefix \textit{ka’-} that occurs occasionally in standard negation, as in \REF{ex:mix:10} and \REF{ex:mix:11}. It is significantly less frequent than the negative particle \textit{ka’ap} in the data and mostly occurs in non-standard negation in combination with other negative markers (see \sectref{sec:mix:3.2}). Note that the verbal prefix \textit{ka’-} does not provoke the dependent paradigm. In \REF{ex:mix:10a} and \REF{ex:mix:10b}, the dependent marking is triggered by the presence of the subordinator \textit{ko} ‘since’;\footnote{\textit{Ko} is a multifunctional subordinator stemming from Spanish \textit{como} 'since' and Spanish \textit{que} 'that'. It is used frequently in narratives and conversations as follows: as a relative pronoun 'that' (the same as Spanish \textit{que}), as a subordinating complementizer 'that' (the same as Spanish \textit{que}), as a subordinating conjunction for adverbial clauses (temporal 'when'; reason “because, since”; condition 'if'; purpose 'so that'), and as an interrogative adverb 'how, why'.} in \REF{ex:mix:11a}, dependent paradigm appears due to the presence of the preverbal particle \textit{nii}. 


 
\ea\label{ex:mix:10}
    \ea\label{ex:mix:10a}
        \glll ko jëtu’un ja ja’ay ntaa’oyëtë.\\
             ko jëtu’un ja ja’ay t-naa-’oyë-të \\
             since so \textsc{dem} people 3\textsc{a.dep-caus}{}-fix-\textsc{pl} \\
        \glt ‘Since the people fixed it.’ 
    \ex\label{ex:mix:10b}
        \glll ko jëtu’un ja ja’ay tka’ja’oyëtë. \\ 
             ko jëtu’un ja ja’ay t-\textbf{ka’}{}-ja-’ooyë-të \\
             since so \textsc{dem} people 3\textsc{a.dep-neg}{}-\textsc{caus-}fix-\textsc{pl} \\
        \glt ‘Since the people didn’t fix it.’
    \z
\z 
 
\ea\label{ex:mix:11}
    \ea\label{ex:mix:11a}
        \glll yaakëëjxpy iich ja kääky. \\
             y-yaa-këëjxpy iich ja kääky \\
             \textsc{3a.ind-caus-}end \textsc{pst} \textsc{dem} tortilla \\
        \glt ‘They finished the tortilla.’
    \ex\label{ex:mix:11b}
        \glll nii yë’ë tkajakëxy ja kääky.\\ 
              nii yë’ë t-\textbf{ka’}{}-ja-këxy ja kääky \\
              \textsc{pst} \textsc{dem} \textsc{3a.dep}{}-\textsc{neg-caus}{}-end \textsc{dem} tortilla\\
        \glt ‘They didn’t finish the tortilla.’ 
    \z
\z 

The negative particle and the negative prefix likely derive from the same proto-form and possibly represent different stages in the grammaticalization cycle. \citet[573]{Wichmann1995} reconstructs a proto-Mixean particle \textit{*ka:h} for 'no' and a proto-Oaxaca-Mixean particle \textit{*kah} with loss of vowel length. It could be speculated that an emphatic form of the negative marker (i.e. \textit{ka’ap}) was used in addition to the original negator and then started to take over (see \citealt{Gelderen2016}). It is also possible that \textit{ka’ap} is the result of univerbation or clitizing of the original negator \textit{ka’-} with an adverbial that triggers dependent marking, since \textit{ka’ap} prompts the dependent paradigm while \textit{ka’-} does not. Now the new negator (\textit{ka’ap)} is taking over, and the old negator (\textit{ka’-)} is used mostly in sentences with constituent negation (see \sectref{sec:mix:3.2}). 

The prefix \textit{ka’-} can also be identified in contructions with \textit{nëm} 'still' yielding a 'not yet' translation, as in \REF{ex:mix:12b}, most likely as a lexicalized form.

\ea\label{ex:mix:12}
    \ea\label{ex:mix:12a}
        \glll tëës nkay.\\
            tëe=ës n-kay-y \\
             before.now=1.\textsc{emph} \textsc{1s.dep-}eat\textsc{-asp.dep} \\
        \glt ‘I have already eaten.’
    \ex\label{ex:mix:12b}
        \glll ka’anëmës nkay.\\ 
             \textbf{ka’a=nëm}=ës n-kay-y \\
             \textsc{neg-}still\textsc{-1.emph} \textsc{1s.dep-}eat\textsc{-asp.dep} \\
        \glt ‘I haven’t eaten yet.’

    \z
\z

Similar forms are common in Mixean languages. In fact, \citet[342]{Wichmann1995} proposes a proto-Oaxaca Mixean form *\textit{ka’h-näm} for “not yet”. 

In fast-paced speech, the negative particle \textit{ka’ap} can surface as \textit{ka’a}, occasionally turning into a proclitic \textit{ka’=}, as in \REF{ex:mix:13}. 

\ea\label{ex:mix:13}
    \ea\label{ex:mix:13a}
        \glll miijts ka’mnëëkx. \\ 
             miijts \textbf{ka’=}m-nëëkx \\
             \textsc{2sg} \textsc{neg=2s}{}-go.\textsc{dep} \\
        \glt ‘You didn’t go.’
    \ex\label{ex:mix:13b}
        \glll miijts ka’mkay. \\
              miijts \textbf{ka’=}m-kay-y  \\
              \textsc{2sg} \textsc{neg=2s-}eat-\textsc{asp.dep}\\
        \glt  ‘You don’t eat.’
    \z
\z 

The ordering of morphemes in the predicate is different in \REF{ex:mix:13} from the previous examples. While the person marker occupies the first slot in the verb template, as in \REF{ex:mix:10} and \REF{ex:mix:11}, in \REF{ex:mix:13a} and \REF{ex:mix:13b}, the negative marker \textit{ka’=} takes that position, thus clearly distinguishing itself from the prefix. The proclitic seems to be an innovation.

A native speaker confirmed in elicitation sessions that the verb may carry the negative prefix \textit{ka’-} instead of or in addition to the negative particle \textit{ka’ap}, as in \REF{ex:mix:14}. Having both markers present as negative concord \citep{Hoeksema2009} or double negation \citep{Dryer2005-wals112} is therefore grammatical, although it does not occur very often in the data. However, negative concord is the norm with constituent negation (see \sectref{sec:mix:3.2}).

\ea\label{ex:mix:14}
    \glll niimääjtsk tää ka’ap yë’ë tka’atsu’tsnëtë. \\
          nii-määjtsk tää \textbf{ka’ap} yë’ë t-\textbf{ka’-}atsu’tsn-ë-të  \\
          \textsc{an-}two then \textsc{neg} \textsc{dem} 3\textsc{a.dep-neg}{}-eat-\textsc{inv-pl} \\
    \glt ‘[...] then he didn’t eat the two [...]’
\z

While there do not seem to be any restrictions in the use of the negative particle \textit{ka’ap}, \citet[453]{Romero-Méndez2009} discusses constraints to the use of the negative prefix \textit{ka’-} in Ayutla Mixe. He shows that its use as a sole negator is restricted to clauses where only the subject is overtly expressed, as opposed to the subject and an object. The same holds true for Chuxnabán Mixe, as shown in \REF{ex:mix:15}. The negative prefx \textit{ka’-} can be an additional negator to the negative particle \textit{ka’ap} in clauses where subject and object are overtly expressed \REF{ex:mix:15a}, but it is ungrammatical as the sole negator in such clauses \REF{ex:mix:15c}. The negative particle \textit{ka’ap}, however, can still be used as the sole negator \REF{ex:mix:15b}.

\ea\label{ex:mix:15}
    \ea[]{\label{ex:mix:15a}
        \glll Juank ka’ap ja kääky tka’kay. \\ 
              Juank \textbf{ka’ap} ja kääky t-\textbf{ka’-}kay \\
              Juan \textsc{neg} \textsc{dem} tortilla \textsc{3a.dep}{}-\textsc{neg-}eat.\textsc{dep} \\
        \glt ‘Juan doesn’t eat tortilla.’}
    \ex[]{\label{ex:mix:15b}
        \glll Juank ka’ap ja kääky tkay. \\
             Juank \textbf{ka’ap} ja kääky t-kay \\
             Juan \textsc{neg} \textsc{dem} tortilla \textsc{3a.dep}{}-eat.\textsc{dep} \\
        \glt ‘Juan doesn’t eat tortilla.’}
    \ex[*]{\label{ex:mix:15c}
        \glll Juank ja kääky tka’kay. \\
              Juank ja kääky t-\textbf{ka’-}kay  \\
              Juan \textsc{dem} tortilla \textsc{3a.dep}{}-\textsc{neg-}eat.\textsc{dep} \\
        \glt ‘Juan doesn’t eat tortilla.’}
    \z
\z 

\citet[453]{Romero-Méndez2009} states that the same restriction applies to locative phrases in clauses with motion verbs, thus extending this limitation to non-core arguments and adjuncts. A similar restriction does not apply in Chuxnabán Mixe. In \REF{ex:mix:16} the single negator can be either the negative particle \textit{ka’ap} or the verbal prefix \textit{ka’-}.

\ea\label{ex:mix:16}
    \ea\label{ex:mix:16a}
        \glll yë’ë \textbf{ka}’nëëkxp kyamoch. \\ 
              yë’ë ø-\textbf{ka}’-nëëkx-p y-kam-och \\
              \textsc{dem} \textsc{3s.ind}{}-\textsc{neg-}go-\textsc{asp.ind} \textsc{3poss}{}-field\textsc{{}-loc}\\
        \glt ‘He is not going to his field.’
    \ex\label{ex:mix:16b}
        \glll yë’ë \textbf{ka’ap} nyiikx kyamoch. \\
              yë’ë \textbf{ka’ap} y-niikx-y y-kam-och \\
             \textsc{dem} \textsc{neg} \textsc{3s.dep}{}-go-\textsc{asp.dep} \textsc{3poss}{}-field\textsc{{}-loc} \\
        \glt ‘He is not going to his field.’
    \z
\z 

Other occurrence or co-occurrence restrictions have not been identified for standard negation. The presence of the negative particle, but not the negative prefix, triggers dependent inflection on the predicate. As a result, the structure of the negative clause differs from the affirmative in these instances (cf. asymmetric negation following \citealt{Miestamo2005,Miestamo2007}). However, if the negative prefix \textit{ka’-} acts as the sole negator, the structure stays the same (cf. symmetric negation following \citealt{Miestamo2005,Miestamo2007}).

\subsubsection{Constructional and paradigmatic asymmetries}\label{sec:mix:2.1.1}

Following \citet{Miestamo2005,Miestamo2007,Miestamo2009neg.pragmatics,Miestamo2014,Miestamo2017}, negative constructions can be categorized into symmetric and asymmetric negation by examining the structural differences between negative and corresponding affirmative clauses. Chuxnabán Mixe generally belongs to the category of languages with partial asymmetric negation, given that the most frequent negative marker \textit{ka’ap} is not simply added to a corresponding positive clause. There can be several accompanying constructional differences, i.e. different sets of verbal affixes related to the dependent paradigms. As with other Mixean languages (see \citealt{Romero-Méndez2009}), in Chuxnabán Mixe predicates in a clause are treated as either independent or dependent, each with its own set of inflectional person indexes, verb stems, and aspectual/mood suffixes. However, dependent paradigms do not specifically relate to negation. Dependency is triggered if a non\nobreakdash-core constituent, such as an adverb, a temporal or locative particle, or the negative particle \textit{ka’ap}, precedes the predicate. The distinction between dependent and independent constructions with the negator \textit{ka’ap} is shown in \REF{ex:mix:17}.

\ea\label{ex:mix:17}
    \ea\label{ex:mix:17a}Independent \newline
        \glll maatsyüüjchpy.\\ 
              \textbf{m-}yaa-tsyüüjch-\textbf{py} \\
              \textbf{\textsc{2a.ind}}\textsc{{}-caus}{}-hurt-\textbf{\textsc{asp.ind}}\\
        \glt ‘You hurt him.’ 
    \ex\label{ex:mix:17b}Dependent\newline
        \glll ka’ap xyaatsyüch. \\
             \textbf{ka’ap} \textbf{x-}yaa-tsyüts-\textbf{y} \\
             \textbf{\textsc{neg}} \textsc{\textbf{2a.dep}-caus}-hurt-\textbf{\textsc{asp.dep}} \\
        \glt ‘You don’t hurt him.’
    \z
\z 

But a temporal particle preceding the predicate, for instance, also triggers dependent marking in an affirmative clause. Therefore, if we negate such an affirmative clause, we find symmetry between the affirmative and the negative, as in \REF{ex:mix:18}.
 
\ea\label{ex:mix:18}
    \ea\label{ex:mix:18a}Dependent\newline
        \glll oojts tiij tuny. \\ 
              oojts tiij t-tun-y \\
              \textsc{pst} something \textsc{3a.dep}{}-do-\textsc{asp.dep}\\
        \glt ‘He did something.’
    \ex\label{ex:mix:18b}Dependent\newline 
        \glll ka’ap oojts tiij tuny.\\
              \textbf{ka’ap} oojts tiij t-tun-y\\
              \textbf{\textsc{neg}} \textsc{pst} something \textsc{3a.dep}{}-do-\textsc{asp.dep} \\
        \glt ‘He didn’t do anything.’
    \z
\z 

In clauses with the negative prefix \textit{ka’-} as the sole negator, dependent marking does not occur (unless it is prompted by some other element), given that the prefix is part of the predicate and thus not a separate constituent preceding the predicate, as in \REF{ex:mix:19} and \REF{ex:mix:20}. 

\ea\label{ex:mix:19}
    \glll pën kyamëëmëto’opy ja myaestrë? \\
          pën \textbf{y-ka’}{}-yaa-mëëmëto’o-\textbf{py} ja y-maestrë \\
          who \textbf{\textsc{3a.ind}}{}-\textsc{neg-caus}{}-obey-\textbf{\textsc{asp.ind}} \textsc{dem} 3\textsc{poss}{}-teacher \\
    \glt ‘Who doesn’t obey his/her teacher?’
\z

\ea\label{ex:mix:20}
    \glll yë’ë ka’nëëkxp kyamoch. \\
          yë’ë \textbf{ø-ka}’-nëëkx-\textbf{p} y-kam-och \\
          \textsc{dem} \textbf{\textsc{3s.ind}}\textbf{{}-}\textsc{neg-}go-\textbf{\textsc{asp.ind}} \textsc{3poss}{}-field\textsc{{}-loc} \\
    \glt ‘He is not going to his field.’
\z

In addition to a special set of person prefixes and aspect/mood suffixes, predicates show stem variations in dependent and independent inflections, as in \REF{ex:mix:16}. Since the predicates in negative clauses are most often inflected as dependent, negative clauses exhibit the following derived asymmetries to corresponding affirmative clauses that are inflected as independent \REF{ex:mix:extra}.

\ea\label{ex:mix:extra}
Properties of negative clauses in dependent inflection
\begin{itemize}[noitemsep]%HC added the noitemsep to make it look better
    \item Separate set of person prefixes (dependent person prefixes)
    \item Separate set of aspect/mood suffixes (for neutral, completive, and irrealis)
    \item Different verb stem 
\end{itemize}
\z

Given that these properties are associated with dependent marking and not specifically with negation, they are derived asymmetries and not considered asymmetries of the subtype A/Cat \citep{Miestamo2013-wals114} that involve changes in the marking of tense, aspect, mood, person, or any (other) verbal categories etc. Rather, there is partial asymmetry in standard negation, depending on all constituents in a clause.

In addition to constructional asymmetries, a language may also display paradigmatic asymmetries \citep{Miestamo2005,Miestamo2007,Miestamo2009neg.pragmatics,Miestamo2014,Miestamo2017} whereby some grammatical categories are not available in negatives. No such restrictions are found for standard negation in Chuxnabán Mixe. Negatives clauses can occur with any aspect or tense markers. 

\subsection{Negation in non-declaratives}\label{sec:mix:2.2}

Negation in non-declarative clauses, such as imperatives and interrogatives, often exhibits special forms not found in standard negation. This section examines prohibitives and negation in interrogatives.

\subsubsection{Negative imperatives}\label{sec:mix:2.2.1}

In Chuxnabán Mixe, there are multiple ways of expressing prohibitives. Generally, they are formed with a special negative particle \textit{këtii} (or its shortened form \textit{kii}) preceding the negated verb, and the latter may optionally be marked with the irrealis enclitic \textit{=ëch}, as in \REF{ex:mix:23} and \REF{ex:mix:24}. This strategy differs from standard negation and thus represents the first type of asymmetry identified. It also differs from affirmative imperatives which are formed by using the bare stem, except in clauses with a first person undergoer. The first person undergoer is marked in both affirmative and negative imperatives as an emphatic enclitic \textit{=ës}, shown in \REF{ex:mix:23} and \REF{ex:mix:24}. In prohibitives, the predicate also takes person prefixes, while there are no person markers in positive imperatives. Plural marking occurs in both, but in prohibitives it is omitted if the irrealis is present, as in \REF{ex:mix:24}. The positive counterparts in the examples below were obtained through elicitation (\ref{ex:mix:21a}, \ref{ex:mix:22a}, \ref{ex:mix:23a}, \ref{ex:mix:24a}), while the prohibitive examples stem from narratives or conversations (\ref{ex:mix:21b}, \ref{ex:mix:22b}, \ref{ex:mix:23b}, \ref{ex:mix:24b}). 

\ea\label{ex:mix:21}
    \ea\label{ex:mix:21a}
        \glll tsë’ëkë! \\
              tsë’ëkë \\
              be.afraid\\
        \glt ‘Be afraid!’ (\textsc{sg}) 
    \ex\label{ex:mix:21b}
        \glll kii mtsë’ëkë!\\ 
              \textbf{kii} m-tsë’ëkë \\
              \textsc{neg.imp} \textsc{2s.dep}{}-be.afraid \\
        \glt ‘Don’t be afraid!’ (\textsc{sg})
    \z
\z 

\ea\label{ex:mix:22}
    \ea\label{ex:mix:22a}
        \glll tsë’ëkëtë! \\
              tsë’ëkë-të \\
              be.afraid-\textsc{pl} \\
        \glt ‘Be afraid!’ (\textsc{pl})
    \ex\label{ex:mix:22b}
        \glll këtii mëtsë‘ëkëtë!\\ 
              \textbf{këtii} m-tsë‘ëkë-të\\
              \textsc{neg.imp} \textsc{2s.dep}{}-be.afraid-\textsc{pl} \\
        \glt ‘Don’t be afraid!’ (\textsc{pl})
    \z
\z 
 
\ea\label{ex:mix:23}
    \ea\label{ex:mix:23a}
        \glll  tukpaat mtëts!\\
              tukpaat m-tëts \\
              touch \textsc{2poss}-tooth \\
        \glt ‘Touch me with your tooth!’ (\textsc{sg})
    \ex\label{ex:mix:23b}
        \glll kiis mtëts xukpaatëch! \\ 
              \textbf{kii=s} m-tëts x-tukpaat=\textbf{ëch} \\
              \textsc{neg.imp}=1.\textsc{emph} \textsc{2poss}-tooth \textsc{1p.dep}-touch=\textsc{irr}\\
        \glt ‘Don’t touch me with your tooth!’ (\textsc{sg})
    \z
\z 

\ea\label{ex:mix:24}
    \ea\label{ex:mix:24a}
        \glll tukpa’tëtës mtëts!\\
              tukpa’të-të=ës m-tëts\\
              touch-\textsc{pl=}1.\textsc{emph} \textsc{2poss}{}-tooth\\
        \glt ‘Touch me with your tooth!’ (\textsc{pl})
    \ex\label{ex:mix:24b}
        \glll kiis xukpaatëch mtëts! \\ 
              \textbf{kii=s} x-tukpaat=\textbf{ëch} m-tëts \\
              \textsc{neg.imp}=1.\textsc{emph} \textsc{1p.dep-}touch=\textsc{irr} \textsc{2poss}{}-tooth\\
        \glt ‘Don’t touch me with your tooth!’ (\textsc{pl}) 
    \z
\z 

The special prohibitive negator always occurs clause-initially in the data; however, it is possible to have adverbs such as \textit{cham} ‘now’ or \textit{yaa} ‘here’ preceding it. There is a difference between second person singular prohibitive and second person plural prohibitive; in the latter, a plural suffix -\textit{të} is added, as in \REF{ex:mix:22}. In general, the verb takes the same person prefixes as in standard negation and in affirmative declaratives. The pattern follows the independent/dependent distinction outlined in \sectref{sec:mix:1.2}. Given that in prohibitives there is a negative particle preceding the predicate at all times, the person prefixes on the verb, as well as as the aspect/mood suffixes, always follow dependent inflection. The ambiguity from having the same form for the person indexes for 2\textsc{a} and 1\textsc{p} in the dependent set (i.e. \textit{x-}) is settled by having an obligatory first person undergoer enclitic, if applicable, as in \REF{ex:mix:25}. While the first-person enclitic is used optionally for emphasis in other types of clauses, it is obligatory in prohibitives.

\ea\label{ex:mix:25}
    \ea\label{ex:mix:25a}
        \glll këtiis xkyöx! \\ 
             këtii\textbf{=s} \textbf{x}{}-kyöx \\
             \textsc{neg.imp}\textbf{=1} \textsc{\textbf{1p}}-hit \\
        \glt ‘Don’t hit me!’
    \ex\label{ex:mix:25b}
        \glll këtii xkyöx! \\
              këtii \textbf{x}{}-kyöx \\
              \textsc{neg.imp} \textbf{\textsc{2a}}-hit \\
        \glt ‘Don’t hit him!’
    \z
\z 

As mentioned above, the irrealis enclitic \textit{=ëch} is optional in these prohibitive constructions, as in \REF{ex:mix:26a} and \REF{ex:mix:27a}. The presence of the irrealis is accompanied by a change in the verb stem. 

\ea\label{ex:mix:26}
    \ea\label{ex:mix:26a}
        \glll këtiis xkyoxëch!\\ 
              këtii=s x-\textbf{kyox=ëch} \\
              \textsc{neg.imp}=1 \textsc{1p.dep-}hit=\textsc{irr} \\
        \glt ‘Don’t hit me!’
    \ex\label{ex:mix:26b}
        \glll këtiis xkyöx! \\
              këtii=s x-\textbf{kyöx} \\
              \textsc{neg.imp}=1 \textsc{1p.dep-}hit \\
        \glt ‘Don’t hit me!’
    \z
\z 
 
\ea\label{ex:mix:27}
    \ea\label{ex:mix:27a}
        \glll këtii mkë’ë xpyujëch! \\ 
              këtii m-kë’ë x-\textbf{pyuj=ëch} \\
              \textsc{neg.imp} \textsc{2poss}{}-hand \textsc{2a.dep-}wash=\textsc{irr}\\
        \glt ‘Don’t wash your hands!’ 
    \ex\label{ex:mix:27b}
        \glll këtii mkë’ë xpyüüj!\\
              këtii m-kë’ë x-\textbf{pyüüj}\\
              \textsc{neg.imp} \textsc{2poss}{}-hand \textsc{2a.dep-}wash \\
        \glt ‘Don’t wash your hands!’
    \z
\z 

In some clauses with other negative particles, such as \textit{niika’ana} ‘never’ or \textit{niimaa} ‘nowhere’, the prohibitive particle \textit{këtii} cannot be used. This occurs because no other negative particle can precede the prohibitive \textit{këtii}, but it can follow it, as in \REF{ex:mix:29a} and \REF{ex:mix:29b}. The prohibitive particle can also be omitted, but in these cases the negative prefix \textit{ka’-} is obligatorily added to the predicate, as well as the irrealis \textit{=ëch}, as in \REF{ex:mix:28a}. The irrealis then is the only marker signaling a prohibitive meaning, as in \REF{ex:mix:30a}. The irrealis is never used in positive imperatives.

\ea\label{ex:mix:28}
    \ea[]{\label{ex:mix:28a}
        \glll niika’na’as xkya’mëkaajpxëch! \\ 
              \textbf{niika’na’a=s} x-\textbf{ka’-}mëkaajpx=\textbf{ëch} \\
             \textbf{never}=1 \textsc{1p.dep-}\textbf{\textsc{neg}}{}-talk=\textbf{\textsc{irr}} \\
        \glt ‘Never talk to me!’}
    \ex[*]{\label{ex:mix:28b}
        \glll niika’na’as kii/këtii xkya’mëkaajpxëch! \\
              niika’na’a=s kii/këtii x-ka’-mëkaajpx=ëch \\
              never=1 \textsc{neg.imp} \textsc{1p.dep-neg}{}-talk=\textsc{irr} \\
        \glt ‘Never talk to me!’}
    \z
\z 
 
\ea\label{ex:mix:29}
    \ea[]{\label{ex:mix:29a}
        \glll kiis xkyoxëch niika’na’a! \\ 
              \textbf{kii}=s x-kyox=ëch \textbf{niika’na’a}  \\
              \textbf{\textsc{neg.imp}}\textsc{=1} \textsc{1p.dep-}hit=\textsc{irr} \textbf{never} \\
        \glt ‘Never hit me!’}
    \ex[]{\label{ex:mix:29b}
        \glll kiis \textbf{niika’na’a} xkyoxëch!\\ 
              \textbf{kii}=s \textbf{niika’na’a} x-kyox=ëch \\
            \textbf{\textsc{neg.imp}}\textsc{=1} \textbf{never} \textsc{1p.dep-}hit=\textsc{irr}\\
        \glt ‘Never hit me!’}
    \ex[*]{\label{ex:mix:29c}
        \glll niika’na’a kiis xkyoxëch!  \\ 
              niika’na’a kii=s x-kyox=ëch \\
               \textbf{never} \textsc{\textbf{neg.imp}=1} \textsc{1p.dep-}hit=\textsc{irr} \\
        \glt ‘Never hit me!’}
    \z
\z 

\ea\label{ex:mix:30}
    \ea\label{ex:mix:30a}
        \glll niimaa mka’në’ëkxëch! \\ 
              \textbf{niimaa} m-\textbf{ka’-}në’ëkx=\textbf{ëch}  \\
              \textbf{nowhere} \textsc{2s-}\textbf{\textsc{neg}}{}-go=\textbf{\textsc{irr}}  \\
        \glt ‘Don’t go anywhere!’ 
    \ex\label{ex:mix:30b}
        \glll  \textbf{niimaa} m\textbf{ka’}nëëkx. \\
              niimaa m-\textbf{ka’}{}-nëëkx \\
              \textbf{nowhere} \textsc{2s-}\textbf{\textsc{neg}}{}-go \\
        \glt ‘You are not going anywhere.’
    \z
\z 

A single negative particle cannot be used for multiple prohibitives; rather, each verb must be negated separately, as in \REF{ex:mix:31}.

\ea\label{ex:mix:31}
    \ea\label{ex:mix:31a}
        \glll kii mtsë’ëkëtë! kii maychë mtaajtë!\\ 
              \textbf{kii} m-tsë’ëkë-të \textbf{kii} maychë m-taaj-të \\
              \textsc{neg.imp} 2\textsc{s.dep}{}-be.afraid-\textsc{pl} \textsc{neg.imp} much 2\textsc{s.dep}{}-think-\textsc{pl} \\
        \glt ‘Don’t be afraid! Don’t worry (too) much!’ (\textsc{pl})
    \ex\label{ex:mix:31b}
        \glll kiis mtëts xchukpaatëch! kiis xpyü’ookëch!  \\
              \textbf{kii=s} m-tëts x-chukpaat=ëch \textbf{kii=s} x-pyü’ook=ëch \\
              \textsc{neg.imp=1} 2\textsc{poss}{}-tooth \textsc{1p.dep}{}-touch=\textsc{irr} \textsc{neg.imp=1} 2\textsc{a.dep}{}-chew=\textsc{irr} \\
        \glt ‘[Swallow me entirely!] Don’t touch me with your tooth! Don’t chew me!’
    \z
\z 

Following \citegen{AuweraEtAl2013} distinction of four types of languages with regard to the marking of prohibitives, Chuxnabán Mixe falls under Types III and IV, which include prohibitives with a verbal construction different from the second person singular imperative and a negation strategy identical (Type III) or different (Type IV) from standard negation. Based on \citet{MiestamoAuwera2007}, various asymmetries can be identified in Chuxnabán Mixe prohibitive constructions. Asymmetries referring to differences in negation strategies between declaratives and imperatives include the special prohibitive \textit{kii/këtii} and the irrealis \textit{=ëch}, the latter being obligatory only in clauses without the prohibitive particle \textit{kii/këtii}. The second type of asymmetry comprises differences in the verbal constructions used in positive and negative imperatives. Positive imperatives are formed by using the bare verb stem. The only marking the predicate can take in these constructions is the plural suffix and the first-person enclitic, while prohibitives take person and aspect markers following the same patterns as declaratives, in addition to the obligatory first person undergoer enclitic. To complicate matters further, Chuxnabán Mixe allows another strategy to form prohibitives, like the negation strategy used in declaratives. In addition to constructions with \textit{kii/këtii}, prohibitives can be formed using the negator \textit{ka’ap} and the irrealis \textit{=ëch}, as in \REF{ex:mix:32a} and \REF{ex:mix:33a}. In these instances, the irrealis is obligatory.

\ea\label{ex:mix:32}
    \ea\label{ex:mix:32a}
        \glll ka’ap mintëch!\\ 
              \textbf{ka’ap} m-min-të=\textbf{ëch}\\
              \textbf{\textsc{neg}} \textsc{2s.dep}{}-come-\textsc{pl}=\textbf{\textsc{irr}}\\
        \glt ‘Don’t come!’ 
    \ex\label{ex:mix:32b}
        \glll këtii mintëch!\\
              \textbf{këtii} m-min-të(=\textbf{ëch}) \\
              \textbf{\textsc{neg.imp}} \textsc{2s.dep}{}-come-\textsc{pl}(=\textbf{\textsc{irr)}}\\
        \glt ‘Don’t come!’
    \ex\label{ex:mix:32c}
        \glll ka’ap mintë.\\
             ka’ap m-min-të \\
             \textbf{\textsc{neg}} \textsc{2s.dep}{}-come-\textsc{pl} \\
        \glt ‘They are not coming.’
    \z
\z 
 
\ea\label{ex:mix:33}
    \ea\label{ex:mix:33a}
        \glll ka’apës xchë’ëkëch! \\ 
              \textbf{ka’ap}=ës x-chë’ëk=\textbf{ëch}\\
              \textbf{\textsc{neg}}\textsc{=1.emph} \textsc{1p.dep}{}-afraid=\textbf{\textsc{irr}} \\
        \glt ‘Don’t be afraid of me!’
    \ex\label{ex:mix:33b}
        \glll këtiis xchë’ëkëch! \\
             këtii=s x-chë’ëk(=ëch)\\
             \textbf{\textsc{neg.imp}}\textsc{=1} \textsc{1p.dep}{}-afraid(=\textsc{irr)} \\
        \glt ‘Don’t be afraid of me!’
    \ex\label{ex:mix:33c}
        \glll ka’apës xchë’ëkë.\\
             ka’ap=ës x-chë’ëkë\\
             \textbf{\textsc{neg}}\textsc{=1.emph} \textsc{1p.dep}{}-afraid\\
        \glt ‘You are not afraid of me.’
    \z
\z 

\REF{ex:mix:32} and \REF{ex:mix:33} illustrate the two strategies used in prohibitives, as well as the structural difference between prohibitives with \textit{këtii} in \REF{ex:mix:32b} and \REF{ex:mix:33b}, prohibitives with \textit{ka’ap} in \REF{ex:mix:32a} and \REF{ex:mix:33a}, and standard negation with \textit{ka’ap} in \REF{ex:mix:32c} and \REF{ex:mix:33c}. \tabref{tab:mix:4} summarizes the strategies used for imperatives and prohibitives in Chuxnabán Mixe.


\begin{table}\small
\begin{tabularx}{\textwidth}{lQQlllQllQ}
\lsptoprule
% & {Bare stem} & {Person prefix} & {first person undergoer} & \textit{këtii} & \textit{ka’ap} & {Other negator (e.g. ‘never’)} & \textit{ka’-} & {\scshape irr} & {\scshape sg/pl} ø/-\textit{të}\\
& {Bare stem} & {Person prefix} & 1U & \textit{këtii} & \textit{ka’ap} & {Other \mbox{negator} \mbox{(e.g. ‘never’)}} & \textit{ka’-} & {\scshape irr} & {\scshape sg/pl} ø/-\textit{të}\\
\midrule 
{Imperative} & {x} & & {x} & & & & & & ~{x}\\
{Prohibitive} & & {x} & {x} & {x} & & {(x)} & & {(x)} & {(x)}\\
{Prohibitive} & & {x} & {x} & & & ~{x} & {x} & ~{x} & \\
{Prohibitive} & & {x} & {x} & {x} & & ~ {x} & & {(x)} & {(x)}\\
{Prohibitive} & & {x} & {x} & & {x} & & & ~{x} & \\
\lspbottomrule
\end{tabularx}
\caption{Summary of imperative and prohibitive strategies. 1U=first person undergoer.}
\label{tab:mix:4}
\end{table}


While the strategy used to form positive imperatives is very similar to the one found in the closely related Ayutla Mixe, the strategies used to form prohibitives are very distinct. Ayutla Mixe exhibits no special negation strategy for prohibitives \citep[307]{Romero-Méndez2009}. In fact, the same negative particle as in standard negation is used with an irrealis marker, like the strategy with \textit{ka’ap} in Chuxnabán Mixe. In the related Sierra Popoluca, there is a special negator for prohibitives and optatives \citep{Boudreault2018}. The prohibitive particle in Chuxnabán Mixe appears also in other constructions, such as negative polar questions and tag questions, discussed in \sectref{sec:mix:2.2.2}, and in clauses with ‘neither’, covered in \sectref{sec:mix:4.5}.

\subsubsection{Questions and answers}\label{sec:mix:2.2.2}

This section covers only polar questions. Negative interrogatives are formed using one of two different strategies. In naturally occurring data, the negative particle \textit{ka’ap} is used most often, the same as in standard negation and the negative existential, as in \REF{ex:mix:34a}. It is also possible to have the negative prefix \textit{ka’}{}-, as in standard negation \REF{ex:mix:35}.

\ea\label{ex:mix:34}
    \ea\label{ex:mix:34a}
        \glll ka’apap pënë? \\ 
              \textbf{ka’ap} jap pën-ë\\
              \textsc{neg} there someone-\textsc{q}\\
        \glt ‘Isn’t s/he here?'
    \ex\label{ex:mix:34b}
        \gll ka’ap jap pën. \\
             \textsc{neg} there someone \\
        \glt ‘Nobody is here.’ = ‘There isn’t anyone.’
    \ex\label{ex:mix:34c}
        \gll täp jä’ä.\\
              \textsc{ex} \textsc{dem}\\
        \glt  ‘Yes, s/he is here.’
    \z
\z 

\ea\label{ex:mix:35}
    \glll jamp tii xkya’oyjaawë? \\
         jamp tii x-\textbf{ka’}{}-yaa’-oyjaaw-ë \\
         there what \textsc{2a.dep}{}-\textsc{neg-caus}{}-feel.good-\textsc{q}\\
    \glt ‘Is there anything that doesn’t make you feel good?’
\z

Another strategy in negative polar questions is to use \textit{kii/këtii}, the same particle found in prohibitives. As with prohibitives, \textit{kii} is the shortened form of \textit{këtii}, and the two can be used interchangeably. The only difference between a standard negation and the negative interrogative is the choice of the respective particle, as can be seen comparing \REF{ex:mix:36c} and \REF{ex:mix:36d}. There is no formal distinction between the prohibitive and the negative interrogative, as in \REF{ex:mix:36d}.
 
\ea\label{ex:mix:36}
    \ea\label{ex:mix:36a}
        \glll këtii iich miny?\\ 
              \textbf{këtii} iich m-miny-y\\
              \textsc{neg} \textsc{pst} \textsc{2s.dep-}come-\textsc{asp.dep} \\
        \glt  ‘You didn’t come?’
    \ex\label{ex:mix:36b}
        \glll këtii miny?\\
             \textbf{këtii} m-miny-y\\
             \textsc{neg} \textsc{2s.dep-}come-\textsc{asp.dep}\\
        \glt ‘You are not coming?’
    \ex\label{ex:mix:36c}
        \glll ka’ap miny.\\ 
              \textbf{ka’ap} m-miny-y\\
              \textsc{neg} \textsc{2s.dep-}come-\textsc{asp.dep}\\
        \glt ‘You are not coming.’    
    \ex\label{ex:mix:36d}
        \glll kii miny! \\
             \textbf{kii} m-miny-y \\
             \textsc{neg} \textsc{2s.dep-}come-\textsc{asp.dep}\\
        \glt ‘Don’t come!’ = ‘You are not coming?’
    \z
\z 

The same construction occurs in non-verbal negative interrogatives and in negative existential questions, as in (\ref{ex:mix:37}--\ref{ex:mix:39}). 

\ea\label{ex:mix:37}
    \ea\label{ex:mix:37a}
        \glll kii iich ja tëëjk chapsëch?\\ 
              \textbf{kii} iich ja tëëjk y-tsaps=ëch\\
              \textsc{neg} \textsc{pst} \textsc{dem} house \textsc{3s.dep-}red=\textsc{irr}\\
        \glt ‘Wasn’t the house red?’
    \ex\label{ex:mix:37b}
        \glll tsaps iich ja tëëjkë? \\
             ø-tsaps iich ja tëëjk-ë \\ 
             \textsc{3s.ind-}red \textsc{pst} \textsc{dem} house-\textsc{q}\\
        \glt ‘Was the house red?’
    \z
\z 

\ea\label{ex:mix:38}
    \glll  kii tii?\\
         \textbf{kii} tii\\
         \textsc{neg} what\\
    \glt ‘Isn’t there anything?’
\z

\ea\label{ex:mix:39}
    \glll kii jap pën?\\
         \textbf{kii} jap pën\\
        \textsc{neg} there someone \\
    \glt ‘Isn’t s/he (here)?’
\z

In addition, the same negative particle is used in tag questions. The tag question is added to a statement to request a confirmation or disconfirmation of the statement made. In many languages tag questions can consist of a positive or a negative. In Chuxnabán Mixe, it is the negative (\ref{ex:mix:40}, \ref{ex:mix:41}).

\ea\label{ex:mix:40}
    \glll xiipë camepsinë i'ta’any, këtiijë pa’ch?\\
        xiipë camepsinë i'ta’any \textbf{këtii=jë’ë} pa’ch\\
        this.one farmer will.be \textsc{neg=dem} buddy\\
    \glt ‘This one (Salvador) will be a farmer, right buddy?’
\z

\ea\label{ex:mix:41}
    \glll jä’ä ko tëë niiwimpi’të, këtii?\\
        jä’ä ko tëë y-niiwimpi’të \textbf{këtii}\\
        \textsc{dem} because before.now \textsc{3s.dep}{}-repeat \textsc{neg} \\
    \glt ‘Because he repeated (the school year), right?’ 
\z

In tag questions, the negative particle always appears clause-finally, unlike in negative interrogatives and prohibitives. As a result, it can be easily distinguished from its other uses.

\subsection{Negation in stative predications}\label{sec:mix:2.3} 

This section examines negation strategies in non-verbal clauses, including the following clause types: equation, proper inclusion, attribution, locative predication, possessive predication, and existential predication, following \citet[111--128]{Payne1997} and \citet{Veselinova2014}. There are several different strategies in Chux\-nabán  Mixe to form these types of clauses in the affirmative and the negative, including juxtaposition, a verbalizer copula, and an existential copula. For some of the clause types, there are multiple negation strategies. Negation strategies in equation and attribution overlap, as well as in locative and existential predication.

\subsubsection{Equative predication (predicate nominals)}\label{sec:mix:2.3.1}

Equative constructions refer to sentences where two entities are equated with each other. In Chuxnabán Mixe, equative predication the subject and the non-verbal predicate are juxtaposed with no copula present. In the negative, the negative particle \textit{ka’ap} is added before the non\nobreakdash-verbal predicate, and the non-verbal constituent takes a person marker and the irrealis enclitic \textit{=ëch}, as in (\ref{ex:mix:42}--\ref{ex:mix:44}). This is the equivalent of the A/NonReal asymmetry identified by \citet[296--297]{Miestamo2013-wals113} for standard negation, but in non-standard negation here. 

\ea\label{ex:mix:42}
    \ea\label{ex:mix:42a}
        \glll yë’ë to’oxychëëjk yë’ë maestrë.\\ 
              yë’ë to’oxychëëjk yë’ë maestrë\\
              \textsc{dem} woman \textsc{dem} teacher\\
        \glt ‘This woman is the teacher.’
    \ex\label{ex:mix:42b}
        \glll  yë’ë to’oxychëëjk ka’ap yë’ëch maestrë. \\
             yë’ë to’oxychëëjk \textbf{ka’ap} \textbf{y-}yë’ë=\textbf{ëch} maestrë\\
             \textsc{dem} woman \textbf{\textsc{neg}} \textsc{\textbf{3s.dep-}dem=\textbf{irr}} teacher\\
        \glt ‘This woman is not the teacher.’
    \z
\z 
 
\ea\label{ex:mix:43}
    \ea\label{ex:mix:43a}
        \glll ëëjts mëëtëkëëk.\\ 
              ëëjts mëëtëkëëk\\
              \textsc{1sg} third\\
        \glt ‘I am the third one.’
    \ex\label{ex:mix:43b}
        \glll ka’apës ëëjts nmëëtëkëëkëch.\\
             \textbf{ka’ap=}ës ëëjts \textbf{n-}mëëtëkëëk=\textbf{ëch}\\
             \textbf{\textsc{neg=1.emph}} \textsc{1sg} \textbf{\textsc{1s.dep}}\textbf{{}-}third=\textbf{\textsc{irr}}\\
        \glt ‘I am not the third one.’
    \z
\z 
 
\ea\label{ex:mix:44}
    \ea\label{ex:mix:44a}
        \glll  chaapë may jyä’äxë’ë.\\ 
              chaapë may jyä’äxë’ë\\
              this much wood\\
        \glt ‘This is a lot of wood.’
    \ex\label{ex:mix:44b}
        \glll chaapë ka’ap myäy jyä’äxëch. \\
             chaapë \textbf{ka’ap} \textbf{y-}may jyä’äx=\textbf{ëch}\\
             this \textbf{\textsc{neg}} \textbf{\textsc{3s.dep}}\textbf{{}-}much wood=\textbf{\textsc{irr}} \\
        \glt ‘This is not a lot of wood.’
    \z
\z 

The person marker is always prefixed to the first constituent of the noun phrase or adjectival phrase that functions as the predicate, while the irrealis enclitic can be attached to either constituent. Although the word order in these clauses cannot be changed, the two constituents in equation can generally be switched. This negation strategy is like the one found in Ayutla Mixe in clauses with adjectival or nominal predicates \citep[408]{Romero-Méndez2009}. \citet[418]{Romero-Méndez2009} states, however, that negative equative sentences in Ayutla Mixe differ from other types of non-verbal predication in that there is no person marker and the predicate takes the suffix \textit{{}-ëp}. He clarifies that only sentences in which there are two referential full noun phrases behave like that. This special negation strategy has not been identified in Chuxnabán Mixe.

However, there is a second type of construction in equative clauses, seen in (\ref{ex:mix:45}--\ref{ex:mix:47}). It involves a verbalizer attached to the predicate in both the negative and the corresponding affirmative clause. The verbalizer has different shapes in the affirmative and the negative, and it takes aspect suffixes and the plural suffix. The verbalized predicate is inflected for person. 

\ea\label{ex:mix:45}
    \ea\label{ex:mix:45a}
        \glll  yë’ës ntaajk’aajchpy. \\ 
              yë’ë=s n-taajk-’aajch-p\\
             \textsc{dem=1.emph} \textsc{1poss-}mother-\textsc{vbz-asp.ind} \\
        \glt ‘She is my mother.’
    \ex\label{ex:mix:45b}
        \glll ka’apës yë’ë ntaajk’achy.\\
             \textbf{ka’ap}=ës yë’ë \textbf{n}{}-n-taajk-\textbf{’ach}{}-y\\
             \textbf{\textsc{neg}}\textsc{=1.emph} \textsc{dem} \textbf{\textsc{1a.dep}}\textsc{{}-1poss-}mother-\textbf{\textsc{vbz}}\textsc{{}-asp.dep}\\
        \glt ‘She is not my mother.’
    \z
\z 
 
\ea\label{ex:mix:46}
    \ea\label{ex:mix:46a}
        \glll chääpë yë’ë ntëëjk’aa’täämpy. \\ 
              chääpë yë’ë n-tëëjk-’aa’tääm-p\\
              this \textsc{dem} \textsc{1poss}{}-house-\textsc{vbz-asp.ind}\\
        \glt ‘This is going to be my house.’
    \ex\label{ex:mix:46b}
        \glll chääpë ka’ap yë’ë xchëëjk’aa’ta’antë. \\
             chääpë \textbf{ka’ap} yë’ë \textbf{x}{}-y-tëëjk-’aa’ta’antë\\
             this \textbf{\textsc{neg}} \textsc{dem} \textbf{\textsc{1p.dep}}{}-1\textsc{poss}{}-house-\textbf{\textsc{vbz}}\\
        \glt ‘This is not going to be my house.’
    \z
\z 
 
\ea\label{ex:mix:47}
    \ea\label{ex:mix:47a}
        \glll chääpëch yë’ë chuump’aajtëp.\\ 
              chääpëch yë’ë y-tuump-’aajt-të-p\\
              these \textsc{dem} \textsc{3poss}{}-worker-\textsc{vbz-pl-asp.ind}\\
        \glt ‘These are his workers.’
    \ex\label{ex:mix:47b}
        \glll chääpëch ka’ap yë’ë tuump’a’të.\\
             chääpëch \textbf{ka’ap} yë’ë \textbf{t-}tuump-‘\textbf{a’t}{}-të\\
             these \textbf{\textsc{neg}} \textsc{dem} \textbf{\textsc{3a.dep}}{}-worker-\textbf{\textsc{vbz}}\textsc{{}-pl}\\
        \glt ‘These are not his workers.’
    \z
\z 

It must be noted that all the equated clauses in (\ref{ex:mix:45}--\ref{ex:mix:47}) contain a demonstrative and a possessed element. The verbalizer takes different shapes in the present, past, and future tense, as well as in the affirmative and the negative, possibly related to verbal stem apophony, as with lexical verbs. The forms are shown in \tabref{tab:mix:5}.

\begin{table}
\begin{tabularx}{\textwidth}{lllll} 
\lsptoprule
& {Affirmative \textsc{sg}} & {Affirmative \textsc{pl}} & {Negative \textsc{sg}} & {Negative \textsc{pl}}\\
\midrule 
{Present/Past} & \textit{{}-’aajch} & \textit{{}-’aajt} & \textit{{}-’ach} & \textit{{}-’a’t}\\
{Future} & {\textit{-’aa’täämp}} &{\textit{-’aa’täämp}} &  {\textit{-’aa’ta’antë}}&  {\textit{-’aa’ta’antë}}\\
\lspbottomrule
\end{tabularx}
\caption{Summary of verbalizer forms in equative clauses}
\label{tab:mix:5}
\end{table}


The verbalized forms can take plural and aspect marking in both the affirmative and the negative. Negative constructions include the negative particle \textit{ka’ap} and the person prefixes from the dependent set on the verbalized element. Both negation strategies used in equative clauses are also found in attributive clauses with adjectival predicates.

\subsubsection{Attribution (adjectival predicates)}\label{sec:mix:2.3.2}

Attribution refers to clauses with an adjectival predicate where the predicate represents a stable quality or a temporary state \citep{Hengeveld1992}. Following \citet[120]{Payne1997}, attributive clauses usually exhibit identical or similar structures to that of predicate nominal clauses. This is also true for Chuxnabán Mixe. The same two negation strategies presented in the previous section occur in these types of clauses (\ref{ex:mix:48}--\ref{ex:mix:51}). 
 
\ea\label{ex:mix:48}
    \ea\label{ex:mix:48a}
        \glll yë’ë ‘uk awa’an.\\ 
              yë’ë ‘uk awa’an\\
              \textsc{dem} dog wild\\
        \glt ‘The dog is wild.’
    \ex\label{ex:mix:48b}
        \glll  ‘uk ka’ap y’awa’anëch. \\
             ‘uk \textbf{ka’ap} \textbf{y-}’awa’an=\textbf{ëch}\\
             dog \textbf{\textsc{neg}} \textbf{\textsc{3s.dep}}{}-wild=\textbf{\textsc{irr}}\\
        \glt ‘The dog is not wild.’
    \z
\z 
 
\ea\label{ex:mix:49}
    \ea\label{ex:mix:49a}
        \glll Juank yë’ë iich wij.\\ 
              Juank yë’ë iich wij\\
              Juank \textsc{dem} \textsc{pst} intelligent\\
        \glt ‘Juan was intelligent.’     
    \ex\label{ex:mix:49b}
        \glll Juank ka’ap iich wyijëch.\\
              Juank \textbf{ka’ap} iich \textbf{y}{}-wij=\textbf{ëch}\\
              Juank \textbf{\textsc{neg}} \textsc{pst} \textbf{\textsc{3s.dep}}\textbf{{}-}intelligent=\textbf{\textsc{irr}}\\
        \glt ‘Juan was not intelligent.’
    \z
\z 

\ea\label{ex:mix:50}
    \ea\label{ex:mix:50a}
        \glll pejyës ëëjts.\\ 
              pejy=ës ëëjts\\
              skinny=1.\textsc{emph} \textsc{1sg} \\
        \glt ‘I am skinny.’    
    \ex\label{ex:mix:50b}
        \glll  ka’apës (ëëjts) npejëch.\\
             \textbf{ka’ap}=ës (ëëjts) \textbf{n}{}-pej=\textbf{ëch}\\
             \textbf{\textsc{neg}}\textsc{=1.emph} \textsc{(1sg)} \textbf{\textsc{1s.dep}}{}-skinny=\textbf{\textsc{irr}}\\
        \glt ‘I am not skinny.’
    \z
\z 

\ea\label{ex:mix:51}
    \ea\label{ex:mix:51a}
        \glll Juank wij i'tä’äny.\\ 
              Juank wij i'tä’äny\\
              Juank intelligent \textsc{fut}\\
        \glt ‘Juan is going to be intelligent.’ 
    
    \ex\label{ex:mix:51b}
        \glll Juank ka’ap wij i’tä’äny.\\
              Juank \textbf{ka’ap} wij \textbf{i'tä’äny}\\
              Juank \textbf{\textsc{neg}} intelligent \textbf{\textsc{fut}}\\
        \glt ‘Juan is not going to be intelligent.’
    \z
\z 
 
Like equative clauses, attributive clauses can be formed with a verbalizer. The future tense, however, exhibits a slightly different structure: clauses include a future marker \textit{i’tä’äny} in both the affirmative and the negative. The only difference in the negative is the presence of the negative particle \textit{ka’ap}. The latter strategy parallels the one in existentials.

\subsubsection{Proper inclusion (predicate nominals)}\label{sec:mix:2.3.3}

Proper inclusion refers to clauses with a nominal predicate where the predicate is included in a certain class \citep{Veselinova2013neg.exist}. More specifically, these are clauses in which "a specific entity is asserted to be among the class of items specified in the nominal predicate" \citep[114]{Payne1997}. The negation strategy used in these types of clauses includes the negative particle preceding the predicate, in addition to the predicate taking person prefixes from the dependent set. The verbalized predicate further takes tense and aspect suffixes in the affirmative and the negative, following the sets for the independent and dependent inflection. This negation strategy is identical to the one in equative and attributive clauses except for the absence of the irrealis \textit{=ëch} (\ref{ex:mix:52}--\ref{ex:mix:54}).
 
\ea\label{ex:mix:52}
    \ea\label{ex:mix:52a}
        \glll Juank yë’ë (tu’uk) mëëtuumpë.\\ 
              Juank yë’ë (tu’uk) mëëtuum-pë\\
              Juank \textsc{dem} (one) worker-\textsc{vbz}\\
        \glt ‘Juan is a worker.’    
    \ex\label{ex:mix:52b}
        \glll Juank ka’ap yë’ë myëëtuny.\\
              Juank \textbf{ka’ap} yë’ë \textbf{y-}mëëtun-y\\
              Juank \textbf{\textsc{neg}} \textsc{dem} \textbf{\textsc{3s.dep}}{}-worker-\textsc{asp.dep}\\
        \glt ‘Juan is not a worker.’
    \z
\z 
 
\ea\label{ex:mix:53}
    \ea\label{ex:mix:53a}
        \glll Juank mëëtuump iich.\\ 
              Juank mëëtuum-pë iich\\
              Juank worker-\textsc{vbz} \textsc{pst}\\
        \glt ‘Juan was a worker.’
    \ex\label{ex:mix:53b}
        \glll Juank ka’ap iich myëëtuny. \\
              Juank \textbf{ka’ap} iich \textbf{y-}mëëtun-y\\
              Juank \textbf{\textsc{neg}} \textsc{pst} \textbf{\textsc{3s.dep}}\textbf{{}-}worker-\textsc{asp.dep}\\
        \glt ‘Juan was not a worker.’
    \z
\z 

\ea\label{ex:mix:54}
    \ea\label{ex:mix:54a}
        \glll Juank yë’ë mëëtun’aamp.\\ 
              Juank yë’ë mëëtun-’aamp\\
              Juank \textsc{dem} worker-\textsc{fut.ind}\\
        \glt ‘Juan is going to be a worker.’
    \ex\label{ex:mix:54b}
        \glll Juank ka’ap myëëtun’a’any.\\
              Juank \textbf{ka’ap} \textbf{y-}mëëtun-’a’any\\
              Juank \textbf{\textsc{neg}} \textbf{\textsc{3s.dep}}{}-worker-\textsc{fut.dep}\\
        \glt ‘Juan is not going to be a worker.’
    \z
\z 

So far, I have described negation in ascriptive clauses. \citet{Veselinova2013neg.exist} states that the negation strategies used in ascriptive and existential predication may overlap. The following section examines negation in existential predication.

\subsubsection{Existential predication}\label{sec:mix:2.3.4}

Existential clauses state the plain existence of an object, and negative existentials often display nonstandard strategies for negation. In Chuxnabán Mixe, negative existentials are formed using the negative existential \textit{ka’ap}, which is the negative particle used in standard negation. In the affirmative, existential clauses are formed with the existential \textit{täp} in the present and past tense, as in (\ref{ex:mix:55}--\ref{ex:mix:57}). Adding a location to the clause does not change this structure, as in \REF{ex:mix:56}. In the negative, the negative existential \textit{ka’ap} replaces the positive existential (see \citealt{Veselinova2013neg.exist,Veselinova2016}) and a pronoun \textit{tii} ‘what, something’ is added, except in the future. The pronoun cannot be omitted in the present and past tense. The future exhibits a different existential: \textit{i’taamp/i’tantëp} for singular and plural, respectively, and the negative existential \textit{ka’ap} is also formed differently in that it does not simply substitute the positive existential but rather appears in addition to a form of the future existential, as in \REF{ex:mix:58} and \REF{ex:mix:59}.


 
\ea\label{ex:mix:55}
    \ea\label{ex:mix:55a}
        \glll täp kwee.\\ 
              täp kwee\\
              \textsc{ex} coffee \\
        \glt ‘There’s coffee.’    
    \ex\label{ex:mix:55b}
        \glll  ka’ap tii kwee.\\
             \textbf{ka’ap} \textbf{tii} kwee\\
             \textbf{\textsc{neg} } \textbf{what} coffee\\
        \glt ‘There isn’t any coffee.’
    \z
\z 
 
\ea\label{ex:mix:56}
    \ea\label{ex:mix:56a}
        \glll täp kwee kamoch. \\ 
              täp kwee kam-och\\
              \textsc{ex} coffee field-\textsc{loc}\\
        \glt ‘There’s coffee in the field.’    
    \ex\label{ex:mix:56b}
        \glll ka’ap tii kwee kamoch.\\
             \textbf{ka’ap} \textbf{tii} kwee kam-och\\
             \textbf{\textsc{neg}} \textbf{what} coffee field-\textsc{loc}\\
        \glt ‘There isn’t any coffee in the field.’
    \z
\z 
 
\ea\label{ex:mix:57}
    \ea\label{ex:mix:57a}
        \glll täp iich kwee.\\ 
              täp iich kwee\\
              \textsc{ex} \textsc{pst} coffee\\
        \glt ‘There was coffee.’    
    \ex\label{ex:mix:57b}
        \glll ka’ap iich tii kwee. \\
             \textbf{ka’ap} \textbf{iich} \textbf{tii} kwee\\
             \textbf{\textsc{neg}} \textsc{\textbf{pst}} \textbf{what} coffee\\
        \glt ‘There wasn’t any coffee.’
    \z
\z 
 
\ea\label{ex:mix:58}
    \ea\label{ex:mix:58a}
        \glll i'taamp kwee.\\ 
              i'taamp kwee\\
              \textsc{ex.fut} coffee\\
        \glt ‘There will be coffee.’    
    \ex\label{ex:mix:58b}
        \glll ka’ap i'ta’any kwee.\\
             \textbf{ka’ap} \textbf{i'ta’any} kwee\\
            \textbf{\textsc{neg}} \textsc{\textbf{ex.fut}} coffee \\
        \glt ‘There won’t be any coffee.’
    \z
\z 
 
\ea\label{ex:mix:59}
    \ea\label{ex:mix:59a}
        \glll i’tantëp may tuumpëch.\\ 
              i’tantëp may tuump-ëch\\
              \textsc{ex.fut.pl} lots worker\textsc{{}-pl}\\
        \glt ‘There will be lots of workers.’
    \ex\label{ex:mix:59b}
        \glll  ka’ap i'ta’antë may tuumpëch.\\
             \textbf{ka’ap} \textbf{i'ta’antë} may tuump-ëch\\
             \textbf{\textsc{neg}} \textsc{\textbf{ex.fut.pl}} lots worker\textsc{{}-pl}\\
        \glt ‘There won’t be lots of workers.’
    \z
\z 

Following \citet{Croft1991} who identifies three types of languages based on the relationship between the verbal negator and the negative existential, Chuxnabán Mixe fits best into type C, as the negative existential has taken over verbal negation, but the negative constructions used for verbal and for existential predications are completely different. \tabref{tab:mix:6} summarizes the existential constructions.

\begin{table}
\begin{tabularx}{\textwidth}{lXX}
\lsptoprule
& {Affirmative} & {Negative}\\
\midrule 
{Present} & {\textit{täp} (+ location)} & {\textit{ka’ap tii} (+ location)}\\
{Past} & {\textit{täp iich} (+ location)} & {\textit{ka’ap iich tii} (+ location)}\\
\multirow{2}{*}{Future} & {\textit{i'taamp} (+ location) \textsc{(sg)}} &{\textit{ka’ap i'ta’any} (+ location) \textsc{(sg)}}\\
& {\textit{i'tantëp} (+ location) \textsc{(pl)}} & {\textit{ka’ap i'ta’antë} (+ location) \textsc{(pl)}}\\
\lspbottomrule
\end{tabularx}
\caption{Affirmative and negative existentials}
\label{tab:mix:6}
\end{table}

Existential and negative existential clauses can also be formed in the same way as ascriptive clauses with a verbalizer, as in \REF{ex:mix:61}. Moreover, there is a third strategy involving an invariant verbalizing suffix \textit{{}-pë} in both the affirmative and the negative, as in \REF{ex:mix:60}. 

\ea\label{ex:mix:60}
    \ea\label{ex:mix:60a}
        \glll  tää nentu’un to’oxychëëjkëch më’këchpyë. \\ 
              tää nentu’un to’oxychëëjk-ëch më’k-ëch-\textbf{pë}\\
              \textsc{ex} like.that women-\textsc{pl} strong-\textsc{pl-vbz} \\
        \glt ‘Like that, strong women exist.’    
    \ex\label{ex:mix:60b}
        \glll  ka’ap nentu’un to’oxychëëjkëch më’këchpyë. \\
             \textbf{ka’ap} nentu’un to’oxychëëjk-ëch më’k-ëch-\textbf{pë}\\
             \textbf{\textsc{neg}} like.that women-\textsc{pl} strong-\textsc{pl-vbz}\\
        \glt ‘Like that, strong women don’t exist.’
    \z
\z 
 
\ea\label{ex:mix:61}
    \ea\label{ex:mix:61a}
        \glll tää nentu’un to’oxychëëjkëch nti’ipë mëk’aajtëpën.\\ 
              tää nentu’un to’oxychëëjk-ëch nti’ipë mëk-\textbf{’aajt-të-pë-}n\\
              \textsc{ex} like.that women-\textsc{pl} which strong-\textsc{vbz-pl-vbz-asp}\\
        \glt ‘Like that, strong women exist.’     
    \ex\label{ex:mix:61b}
        \glll  ka’ap nentu’un to’oxychëëjkëch myëk’a’të.\\
             \textbf{ka’ap} nentu’un to’oxychëëjkëch \textbf{y-}mëk-\textbf{’a’t}{}-të\\
             \textbf{\textsc{neg}} like.that women-\textsc{pl} \textbf{\textsc{3s.dep}}\textbf{{}-}strong-\textbf{\textsc{vbz}}\textsc{{}-pl}\\
        \glt ‘Like that, strong women don’t exist.’
    \z
\z 
 
\ea\label{ex:mix:62}
    \ea[]{\label{ex:mix:62a}
        \glll  ka’ap (tii) kwee pa’akpë. \\ 
              \textbf{ka’ap} \textbf{(tii)} kwee pa’ak\textbf{{}-pë}\\
              \textbf{\textsc{neg}} \textbf{what} coffee sweet-\textsc{vbz} \\
        \glt ‘Sweet coffee doesn’t exist.’}
    \ex[≠]{\label{ex:mix:62b}
        \glll ka’ap kwee pya’akëch.\\
             \textbf{ka’ap} kwee \textbf{y-}pa’ak-\textbf{ëch}\\
             \textbf{\textsc{neg}} coffee \textbf{\textsc{3s.dep}}-sweet-\textsc{pl} \textbf{\textsc{irr}} \\
        \glt ‘The coffee isn’t sweet.’}
    \z
\z 
 
\ea\label{ex:mix:63}
    \ea\label{ex:mix:63a}
        \glll tää iich kwee pa’akpë.\\ 
              tää iich kwee pa’ak-\textbf{pë}\\
              \textsc{ex} \textsc{pst} coffee sweet-\textsc{vbz}\\
        \glt ‘Sweet coffee existed.’
    \ex\label{ex:mix:63b}
        \glll  ka’ap iich (tii) kwee pa’akpë. \\
             \textbf{ka’ap} \textbf{iich} \textbf{(tii)} kwee pa’ak-\textbf{pë}\\
             \textbf{\textsc{neg}} \textsc{pst} \textbf{what} coffee sweet-\textsc{vbz}\\
        \glt ‘Sweet coffee didn’t exist.’
    \z
\z 
 
\ea\label{ex:mix:64}
    \ea\label{ex:mix:64a}
        \glll i'täämp kwee pa’akpë.\\ 
              i'täämp kwee pa’ak-\textbf{pë}\\
              \textsc{ex.fut} coffee sweet-\textsc{vbz}\\
        \glt ‘Sweet coffee will exist.’
    \ex\label{ex:mix:64b}
        \glll  ka’ap i’ta’any kwee pa’akpë.\\
             \textbf{ka’ap} \textbf{i’ta’any} kwee pa’ak-\textbf{pë}\\
             \textsc{neg} \textsc{ex.fut} coffee sweet-\textsc{vbz}\\
        \glt ‘Sweet coffee will not exist.’
    \z
\z 

As can be seen in \REF{ex:mix:60} and \REF{ex:mix:63}, in clauses with the verbalizer \textit{{}-pë} the negative existential is formed with the negative existential \textit{ka’ap} replacing the positive existential \textit{tää} in the present and past. In the future, the existential changes from \textit{i’täämp} to \textit{i’ta’any} in the negative, as in \REF{ex:mix:64}. 

A yet different construction is consistently translated as a negative existential, although its literal meaning is ‘little, hardly any’. While this would not be considered a true existential construction, it is worth noting.



\ea\label{ex:mix:65}
    \glll amaay meeny amaay tiija’ch. \\
        \textbf{amaay} meeny \textbf{amaay} \textbf{tii’-a’ch} \\
        \textbf{little} money \textbf{little} something-\textsc{vbz} \\
    \glt ‘There was no money, there was nothing.’
\z
 
\ea\label{ex:mix:66}
    \ea\label{ex:mix:66a}
        \glll may tiija’ch.\\ 
              may tii-’a’ch\\
              lot something-\textsc{vbz} \\
        \glt ‘There was something.’ 
    \ex\label{ex:mix:66b}
        \glll amaay tiija’ch.\\
              amaay tii-’a’ch\\
             \textbf{little} \textbf{something-\textsc{vbz}} \\
        \glt ‘There was nothing.’
    \z
\z 

In \REF{ex:mix:65} and \REF{ex:mix:66}, the adjective \textit{amaay} ‘little’ is translated as a negative existential. It can occur with the pronoun \textit{tii} ‘what, something’, like the negative existential \textit{ka’ap}. What is interesting is that the verbalizer \textit{{}-‘a’ch} in constructions with \textit{amaay} has the same shape in both the affirmative and the negative, as in \REF{ex:mix:66}, unlike the verbalizer in ascriptive negatives. Further worth mentioning is the pair \textit{may} ‘a lot’ and its antonym \textit{amaay} ‘a little’. It is possible that the initial vowel in \textit{amaay} represents negative derivation. The vowel length of the second vowel in \textit{amaay} could be a parallel development to the stem changes found in predicates. 

\subsubsection{Locative predication}\label{sec:mix:2.3.5}

Locative clauses include structures in which the predicate is a location \citep[112]{Payne1997}. Following \citet[110]{Veselinova2013neg.exist}, locative negation "refers to the negation strategy used in sentences with a locative predicate and a definite subject". These constructions often share features with existentials, but in addition to existence, they specify the location, as is the case in Chuxnabán Mixe. The affirmative includes the existential \textit{täp/tämp} and a tense marker for the future. In addition, the locative can take optional or obligatory demonstratives or pronouns: \textit{tii} ‘what’, \textit{pën} ‘someone’, or \textit{ma} ‘where’. While the first two are generally optional, the latter is obligatory, as in \REF{ex:mix:69}. Moreover, the distal deictic \textit{jap} ‘there’ can be clitized to the negative particle, as in \REF{ex:mix:70}. The distal deictic \textit{jap} ‘there’ and the pronouns \textit{pën} ‘someone’ and \textit{tii} ‘what’ can be omitted (\ref{ex:mix:67}, \ref{ex:mix:68}, \ref{ex:mix:71}). 

\ea\label{ex:mix:67}
    \ea\label{ex:mix:67a}
        \glll yë’ë tuumpë tämp i'ta’any kamoch.\\ 
              yë’ë tuumpë tämp i'ta’any kam-och\\
              \textsc{dem} worker \textsc{ex} \textsc{fut} field-\textsc{loc}\\
        \glt ‘The worker will be in the field.’
    \ex\label{ex:mix:67b}
        \glll  yë’ë tuumpë ka’ap i'ta’any kamoch.\\
             yë’ë tuumpë \textbf{ka’ap} \textbf{i'ta’any} kam-och\\
             \textsc{dem} worker \textbf{\textsc{neg}} \textsc{\textbf{fut}} field-\textsc{loc}\\
        \glt ‘The worker won’t be in the field.’
    \z
\z 
 
\ea\label{ex:mix:68}
    \ea\label{ex:mix:68a}
        \glll tämp tuumpëch i’ta’antë kamoch.\\ 
              tämp tuump-ëch i’ta’antë kam-och\\
              \textsc{ex} worker-\textsc{pl} \textsc{fut} field-\textsc{loc}\\
        \glt ‘The workers will be in the field.’
    \ex\label{ex:mix:68b}
        \glll  tuumpëch ka’ap i'ta’antë kamoch.\\
             tuumpëch \textbf{ka’ap} \textbf{i'ta’antë} kam-och\\
             worker-\textsc{pl} \textbf{\textsc{neg}} \textsc{\textbf{fut}} field-\textsc{loc} \\
        \glt ‘The workers won’t be in the field.’
    \z
\z 
 
\ea\label{ex:mix:69}
    \ea\label{ex:mix:69a}
        \glll jä’ä mok täp i'ta’any ma mesën.\\ 
              jä’ä mok täp i'ta’any ma mes-ën\\
              \textsc{dem} corn \textsc{ex} \textsc{fut} where table-\textsc{loc}\\
        \glt ‘The corn will be on the table.’
    \ex\label{ex:mix:69b}
        \glll jä’ä mok ka’ap i'ta’any ma mesën. \\
             jä’ä mok \textbf{ka’ap} \textbf{i'ta’any} ma mes-ën\\
             \textsc{dem} corn \textbf{\textsc{neg}} \textsc{\textbf{fut}} where table-\textsc{loc}\\
        \glt ‘The corn won’t be on the table.’
    \z
\z 
 
\ea\label{ex:mix:70}
    \ea\label{ex:mix:70a}
        \glll yë’ë tuumpe täpjä’ä kamoch. \\ 
              yë’ë tuumpe täp=jä’ä kam-och\\
              \textsc{dem} worker \textsc{ex}=\textsc{dem} field-\textsc{loc} \\
        \glt ‘The worker is in the field.’
    \ex\label{ex:mix:70b}
        \glll yë’ë tuumpë ka’apap pën kamoch. \\
             yë’ë tuumpë \textbf{ka’ap=jap} \textbf{pën} kam-och \\
             \textsc{dem} worker \textbf{\textsc{neg=}}\textbf{there} \textbf{someone} field-\textsc{loc}\\
        \glt ‘The worker isn’t in the field.’
    \z
\z 
 
\ea\label{ex:mix:71}
    \ea\label{ex:mix:71a}
        \glll jä’ä mok täp ma mesën.\\ 
              jä’ä mok täp ma mes-ën\\
              \textsc{dem} corn \textsc{ex} where table-\textsc{loc}\\
        \glt ‘The corn is on the table.’    
    \ex\label{ex:mix:71b}
        \glll jä’ä mok ka’apap tii ma mesën.\\
             jä’ä mok \textbf{ka’ap=jap} \textbf{tii} \textbf{ma} mes-ën\\
             \textsc{dem} corn \textbf{\textsc{neg=}}\textbf{there} \textbf{what} \textbf{where} table-\textsc{loc} \\
        \glt ‘The corn isn’t on the table.’
    \z
\z  

Overall, negation in locative predications is formed with the existential negator \textit{ka’ap} in addition to optional and obligatory demonstratives, pronouns, and deictics, as summarized in \tabref{tab:mix:7}. %In the future, there seems to be partial overlap with equative clauses. Negation strategies used for locatives-existentials set in the future show a partial overlap with equative clauses.

\begin{table}
\small 
\begin{tabularx}{\textwidth}{Xllll}
\lsptoprule
& \multicolumn{2}{c}{{Affirmative}} & \multicolumn{2}{c}{{Negative}}\\
\midrule
%\uline{Location} &&&&\\
{Present} & \multicolumn{2}{l}{{\textit{täp ({tii})} + location}} & \multicolumn{2}{l}{{\textit{ka’ap jap (tii) (pën)} + location}}\\
\tablevspace
{Past} & \multicolumn{2}{l}{{\textit{täp iich (tii)} + location}} & \multicolumn{2}{l}{{\textit{ka’ap jap iich (tii) (pën)} + location}}\\
\tablevspace
 {Future} & \multicolumn{2}{l}{\textit{tämp i'ta’any} + location \textsc{(sg)}}  & \multicolumn{2}{l}{\textit{ka’ap i'ta’any} + location \textsc{(sg)}} \\
& \multicolumn{2}{l}{\textit{tämp i'ta’antë} + location \textsc{(pl)}}& \multicolumn{2}{l}{{\textit{ka’ap i'ta’antë} + location \textsc{(pl)}}}\\
%\midrule
%\uline{Equation} &&&&\\
% \multicolumn{5}{l}{{Comparison to verbalizer shapes in future tense equative clauses}}\\
% \midrule
%{Future} & \multicolumn{2}{l}{\textit{=’aa’täämp}} & \multicolumn{2}{l}{\textit{=’aa’ta’antë}}\\
\lspbottomrule
\end{tabularx}
\caption{Location in affirmative and negatives}
\label{tab:mix:7}
\end{table}

The future tense shows similarities in shape to negative equative clauses (see \tabref{tab:mix:5} in \sectref{sec:mix:2.3.1}). In equative clauses, the same marker appears in singular and plural, unlike in locative constructions. Negative locative constructions can also be formed in the same way as negative existentials involving the verbalizer \textit{{}-pë}. The constructions are identical to the respective existential clauses, but in addition to predicating existence there is a location, as in \REF{ex:mix:72}.

\ea\label{ex:mix:72}
    \ea\label{ex:mix:72a}
        \glll tää nentu’un to’oxychëëjkëch më’këchpyë ma kaajpnën.\\ 
              tää nentu’un to’oxychëëjk-ëch më’k-ëch-pë \textbf{ma} \textbf{kaajpn-ën}\\
             \textsc{ex} like.that women-\textsc{pl} strong-\textsc{pl-vbz} \textbf{where} \textbf{village-\textsc{loc}}\\
        \glt ‘Like that, strong women exist in the village.’
    \ex\label{ex:mix:72b}
        \glll ka’ap nentu’un to’oxychëëjkëch më’këchpyë ma kaajpnën.\\
             \textbf{ka’ap} nentu’un to’oxychëëjk-ëch më’k-ëch-pë \textbf{ma} \textbf{kaajpn-ën}\\
             \textbf{\textsc{neg}} like.that women-\textsc{pl} strong-\textsc{pl-vbz} \textbf{where} \textbf{village-\textsc{loc}}\\
        \glt ‘Like that, strong women don’t exist in the village.’
    \z
\z 

\citet[123--124]{Veselinova2013neg.exist} notes that there may be a difference between predicating absolute absence, i.e. non-existence anywhere, and predicating absence in a specific location. In some languages, a negative existential used in locatives negates the existence of an entity overall. This can differ from a construction with a different negator in locatives that negates existence in one location but where existence in another location is possible. No such difference has been noted in Chuxnabán Mixe, as illustrated in (\ref{ex:mix:73}--\ref{ex:mix:76}).

\ea\label{ex:mix:73}
    \glll ka’ap u’k awa’an-pë yaa jäämyë waakwimp.\\
        ka’ap u’k awa’an-pë yaa jäämyë waakwimp\\
        \textsc{neg} dog mean-\textsc{vbz} here only Oaxaca\\
    \glt ‘Mean dogs don’t exist here, only in Oaxaca.’
\z

\ea\label{ex:mix:74}
    \glll tää u’k awa’anpë jäämyë waakwimp.\\
        tää u’k awa’an-pë jäämyë waakwimp\\
        \textsc{ex} dog mean-\textsc{vbz} only Oaxaca\\
    \glt ‘Mean dogs exist, but only in Oaxaca.’
\z
 
\ea\label{ex:mix:75}
    \ea\label{ex:mix:75a}
        \glll tää iich kwee pa’akpë.\\ 
              tää iich kwee pa’ak-pë\\
             \textsc{ex} \textsc{pst} coffee sweet-\textsc{vbz} \\
        \glt ‘Sweet coffee existed.’
    \ex\label{ex:mix:75b}
        \glll  ka’ap iich tii kwee pa’akpë.\\
             ka’ap iich tii kwee pa’ak-pë\\
             \textsc{neg} \textsc{pst} what coffee sweet-\textsc{vbz}\\
        \glt ‘Sweet coffee didn’t exist.’
    \ex\label{ex:mix:75c}
        \glll ka’ap iich tii kwee pa’akpë cham tää yë’ë.\\
             ka’ap iich tii kwee pa’ak-pë cham tää yë’ë\\
             \textsc{neg} \textsc{pst} what coffee sweet-\textsc{vbz} now \textsc{ex} \textsc{dem}\\
        \glt ‘Sweet coffee didn’t exist, but now it does (exist).’
    \z
\z 

\ea\label{ex:mix:76}
    \glll yë’ë u’k ka’apap ma tëëjkën täpjä’ä kamoch.\\
        yë’ë u’k ka’ap-jap ma tëëjk-ën täp-jä’ä kam-och\\
        \textsc{dem} dog \textsc{neg}{}-there where house-\textsc{loc} \textsc{ex-dem} field-\textsc{loc}\\
    \glt ‘The dog is not in the house, (the dog) is in the field.’
\z

Ascriptive, existential, and locative predications are sometimes grouped together as stative predication \citep[111]{Veselinova2013neg.exist}. This also includes possessive predications. If the same negator is used in all these predications, it is identified as a ‘stative negator’. 

\subsubsection{Possessive predication}\label{sec:mix:2.3.6}

Languages usually employ existential and/or locational structures to express the notion of possession. This also occurs in Chuxnabán Mixe. Possessives are formed either using the existential construction with \textit{tää/tamp} and a person prefix and verbalizer with the preposition \textit{mëët} ‘with’ (\ref{ex:mix:77}, \ref{ex:mix:79a}), or just the existential, or just the person prefix and verbalizer with the preposition \textit{mëët} ‘with’ (\ref{ex:mix:78}, \ref{ex:mix:80}, \ref{ex:mix:81}). If there is an existential present, then the person prefix in the affirmative is from the dependent set, if not, then it belongs to the independent set, unless dependency is triggered otherwise. In the negative, the existential \textit{tää/tämp} is replaced with the negator \textit{ka’ap} and the pronoun \textit{tii} ‘what’ (\ref{ex:mix:79a}--\ref{ex:mix:79b}); and the person prefix is always from the dependent set. As with existential constructions, there is no existential in the future tense, but the verbalizer has a special form, which is different in the affirmative and the negative, ‘\textit{aa’täämpy} and \textit{‘aa’ta’any} respectively. There is yet another strategy to form possessive constructions. In the affirmative, the possessed predicate is marked with a possessive prefix (\ref{ex:mix:79c}), and in the negative the negator \textit{ka’ap} and the pronoun \textit{tii} ‘what’ are added \REF{ex:mix:79d}.


\ea\label{ex:mix:77}
    \ea\label{ex:mix:77a}
        \glll Xwank tämp tmëët’ach tsuujknë.\\ 
              Xwank tämp t-mëët-’ach tsuujknë\\
              Juan \textsc{ex} \textsc{3a.dep}{}-with-\textsc{vbz} knife\\
        \glt ‘Juan has a knife.’
    \ex\label{ex:mix:77b}
        \glll Xwank ka’ap tii tmëët’ächy tsuujknë.\\
             Xwank \textbf{ka’ap} \textbf{tii} t-mëët-’ächy tsuujknë\\
             Juan \textsc{neg} what \textsc{3a.dep}{}-with-\textsc{vbz} knife\\
        \glt ‘Juan doesn’t have a knife.’
    \z
\z 

\ea\label{ex:mix:78}
    \ea\label{ex:mix:78a}
        \glll  Xwank myëët’aa’täämpy tu’uk tsuujknë.\\ 
              Xwank y-mëët-’aa’täämpy tu’uk tsuujknë\\
             Juan \textsc{3a.ind}{}-with-\textsc{vbz} one knife \\
        \glt ‘Juan will have a knife.’
    \ex\label{ex:mix:78b}
        \glll Xwank ka’ap tmëët’aa’ta’any tsuujknë.\\
             Xwank \textbf{ka’ap} \textbf{t-mëët-’aa’ta’any} tsuujknë\\
             Juan \textsc{neg} \textsc{3a.dep}{}-with-\textsc{vbz} knife\\
        \glt ‘Juan won’t have a knife.’
    \z
\z 
 
\ea\label{ex:mix:79}
    \ea\label{ex:mix:79a}
        \glll yë’ë u’k tämp tmëët’ach toojkx.\\ 
              yë’ë u’k tämp t-mëët-’ach toojkx\\
              \textsc{dem} dog \textsc{ex} \textsc{3a.dep}{}-with-\textsc{vbz} food\\
        \glt ‘The dog has food.’
    \ex\label{ex:mix:79b}
        \glll yë’ë u’k ka’ap tii tmëët’ach toojkx. \\
             yë’ë u’k \textbf{ka’ap} \textbf{tii} \textbf{t-mëët-’ach} toojkx\\
             \textsc{dem} dog \textsc{neg} what \textsc{3a.dep}{}-with-\textsc{vbz} food \\
        \glt ‘The dog doesn’t have food.’
    \ex\label{ex:mix:79c}
        \glll yë’ë u’k choojkx.\\ 
              yë’ë u’k y-toojkx\\
              \textsc{dem} dog \textsc{3poss-}food\\
        \glt ‘The dog has food.’
    \ex\label{ex:mix:79d}
        \glll  yë’ë u’k ka’ap tii choojkx. \\
             yë’ë u’k \textbf{ka’ap} \textbf{tii} y-toojkx\\
             \textsc{dem} dog \textsc{neg} what \textsc{3poss-}food\\
        \glt ‘The dog doesn’t have food.’
    \z
\z 
 
\ea\label{ex:mix:80}
    \ea\label{ex:mix:80a}
        \glll yë’ë u’k myëët’aa’täämpy toojkx. \\ 
              yë’ë u’k y-mëët-’aa’täämpy toojkx\\
              \textsc{dem} dog \textsc{3a.dep}{}-with-\textsc{vbz} food\\
        \glt ‘The dog will have food.’
    \ex\label{ex:mix:80b}
        \glll  yë’ë u’k ka’ap tmëët’aa’ta’any toojkx. \\
             yë’ë u’k \textbf{ka’ap} \textbf{t-mëët’aa’ta’any} toojkx\\
             \textsc{dem} dog \textsc{neg} \textsc{3a.dep}{}-with-\textsc{vbz} food\\
        \glt ‘The dog won’t have food.’ 
    \z
\z 
 
\ea\label{ex:mix:81}
    \ea\label{ex:mix:81a}
        \glll miijts mëët’aachpy meeny.\\ 
              miijts m-mëët-’aachpy meeny\\
              \textsc{2sg} \textsc{2a.ind}{}-with-\textsc{vbz} money\\
        \glt ‘You have money.’
    \ex\label{ex:mix:81b}
        \glll  miijts ka’ap tii xmyëët’ach meeny. \\
             miijts \textbf{ka’ap} \textbf{tii} \textbf{x-mëët-’ach} meeny\\
             \textsc{2sg} \textsc{neg} what \textsc{2a.dep}{}-with-\textsc{vbz} money\\
        \glt ‘You don’t have money.’
    \z
\z 


\largerpage
There are three strategies for possessives in the affirmative and two in the negative, as shown in \tabref{tab:mix:8}.

\begin{table}
\begin{tabularx}{\textwidth}{lQ@{~~}Q@{}}
\lsptoprule
{Tense} & {Affirmative} & {Negative}\\
\midrule 
{Present Strategy \#1} & {\textit{tää/tamp} + \textsc{dep} \textsc{pers} + \textsc{vbz} {}\textit{-‘ach} w/ \textit{mëët} ‘with’} & {\textit{ka’ap tii} + \textsc{dep} \textsc{pers} + \textsc{vbz} {}-\textit{‘ach} w/ \textit{mëët} ‘with’} \\ 
\tablevspace
{Present Strategy \#2} & {\textsc{ind} \textsc{pers} \textsc{+} \textsc{vbz} {}-\textit{‘ach on mëët} ‘with’} & {Same as above} \\
\tablevspace
{Present Strategy \#3} & {\textsc{poss} on possessed} & {\textit{ka’ap tii} + \textsc{poss} on possessed}\\
\tablevspace
{Past} & {Same as above + \textit{iich}} & {Same as above + \textit{iich}}\\
\tablevspace
{Future} & {\textsc{ind} \textsc{pers} \textsc{+} \textsc{vbz} \textit{‘aa’täämpy on mëët} ‘with’} & {\textit{ka’ap (tii)} + \textsc{dep} \textsc{pers} + \textsc{vbz} \textsc{-}\textit{‘aa’ta’any on mëët} ‘with’} \\
\lspbottomrule
\end{tabularx}
\caption{Affirmative and negative possessive clauses}
\label{tab:mix:8}
\end{table}



\begin{sidewaystable}
\small
\rowcolors{2}{white}{black!10!white}
\begin{tabularx}{\textwidth}{>{\raggedright}p{2.9cm}QQp{3cm}}
\lsptoprule
& \textbf{Affirmative} & \textbf{Negative strategy \#1} & \mbox{\textbf{Negative strategy \#2}}\\

{Equation}

{(nominal predicate)} & {juxtaposition} & {\textit{ka’ap} + person + =\textit{ëch}} & {\textit{ka’ap} + person + \textsc{vbz}}\\
{Attribution}

{(adjectival predicate)} & {juxtaposition} 

{(+\textit{iich} in \textsc{pst})}

{(+\textit{i’tä’äny} in \textsc{fut)}} & {\textit{ka’ap} + person + =\textit{ëch}}

{(+\textit{iich} in \textsc{pst})}

{(+\textit{i’tä’äny} in \textsc{fut,} but no =\textit{ëch})} & {\textit{ka’ap} + person + \textsc{vbz}}

{\textit{ka’ap (+i’tä’äny} in \textsc{fut)}}\\

{Proper\newline inclusion}

{(nominal predicate)} & {juxtaposition + {}-\textit{pë}}

{(+\textit{iich} in \textsc{pst})}

{(+{}-\textit{‘aamp} in \textsc{fut,} but no {}-\textit{pë})} & {\textit{ka’ap} + person} 

{(+\textit{iich} in \textsc{pst})}

{(+{}-\textit{‘a’any} in \textsc{fut})} & \\
{Existential \#1} & {\textit{täp} (+ \textit{iich} in \textsc{pst})}

{\textit{i'taamp/i'tantëp} \textsc{(sg,} \textsc{pl)} in \textsc{fut} (but no \textit{täp})} & {\textit{ka’ap tii} (+ \textit{iich} in \textsc{pst})}

{(+\textit{i'ta’any/i'ta’antë} \textsc{(sg,pl)} in \textsc{fut}} & \\
{Existential \#2} & {(\textsc{dem}) \textsc{nml} + {}-\textit{pë}}

{\textit{tää} + {}-\textit{pë}}

{\textit{i’täämp} + {}-\textit{pë} (\textsc{fut})} & {\textit{ka’ap (tii)} + {}-\textit{pë}}

{\textit{ka’ap i’ta’any} + {}-\textit{pë} (\textsc{fut} \textsc{sg})}

{\textit{ka’ap i’ta’antë} + {}-\textit{pë} (\textsc{fut} \textsc{pl})} & {\textit{ka’ap} + person + \textsc{vbz}} 

{–\textit{‘a’t/’a’të} (\textsc{sg/pl})}\\
{Locative \#1 = Existential \#1} & {\textit{täp} + \textit{(tii)} + location/}

{(+ \textit{iich} in \textsc{pst})}

{\textit{tämp i’ta’any/i’ta’antë} + location (\textsc{fut} \textsc{sg/pl})} & {\textit{ka’ap jap} + \textit{(tii/pën)} + location (+ \textit{iich} in \textsc{pst})}

{\textit{ka’ap i’ta’any/i’ta’antë} + location (in \textsc{fut} \textsc{sg/pl)}} & \\
{Locative \#2 = Existential \#2} & {(\textsc{dem}) \textsc{nml} + {}-\textit{pë}}

{\textit{tää} + {}-\textit{pë}}

{\textit{i’täämp} + {}-\textit{pë} (\textsc{fut})} & {\textit{ka’ap (tii)} + {}-\textit{pë}}

{\textit{ka’ap i’ta’any} + {}-\textit{pë} (\textsc{fut} \textsc{sg})}

{\textit{ka’ap i’ta’antë} + {}-\textit{pë} (\textsc{fut} \textsc{pl})} & \\
{Possessive \#1} & \textit{tää/tamp} 

{+ \textsc{dep} \textsc{pers} + \textsc{vbz}} 

{{}-\textit{‘ach on mëët} ‘with’} 

{\textsc{ind} \textsc{pers} \textsc{+} \textsc{vbz} \nobreakdash-\textit{‘ach on mëët}}

{(+ \textit{iich} in \textsc{pst})}

{\textsc{(vbz} \textit{‘aa’täämpy} in \textsc{fut)}} & {\textit{ka’ap tii} + \textsc{dep} \textsc{pers} + \textsc{vbz} {}-\textit{‘ach on mëët} ‘with’} 

{(+ \textit{iich} in \textsc{pst})}

{(\textsc{vbz} \textsc{-}\textit{‘aa’ta’any} in \textsc{fut)}} & \\
{Possessive \#2} & {\textsc{poss} on possessed}

{(+ \textit{iich} in \textsc{pst})} & \textit{{ka’ap tii}} 

{+ \textsc{poss} on possessed}

{(+ \textit{iich} in \textsc{pst})} & \\
\lspbottomrule
\end{tabularx}
\caption{Summary of negation strategies in non-verbal clauses}
\label{tab:mix:9}
\end{sidewaystable}


\tabref{tab:mix:9} provides a summary of negation strategies in non-verbal predication. Overall, equative and attributive clauses exhibit the same structure except for the future forms, and both are formed in the same way as clauses of proper inclusion except for the lack of the irrealis enclitic \textit{=ëch}. While existential and locative constructions are identical, possessive clauses seem to use strategies from both, ascriptive and existential clauses. The existential negator \textit{ka’ap} is the only negator used in all non-verbal clauses.


\subsection{Negation in non-main clauses}\label{sec:mix:2.4}
For the most part, negation in complex and dependent clauses does not exhibit any special negation strategies. The negator \textit{ka’ap} occurs in main and subordinate clauses, which are negated separately. In relative clauses, each clause is negated separately, and both main and subordinate clause can be negated at the same time, as in \REF{ex:mix:82} and \REF{ex:mix:83}.

\ea\label{ex:mix:82}
    \glll ka’ap toxychëëjk piky {\ob}mëti’ipë ka’ap tukjachën tuunk{\cb}. \\ 
        \textbf{ka’ap} toxychëëjk piky {\db}mëti’ipë \textbf{ka’ap} t-tukjats-y-ën tuunk  \\
        \textsc{neg} woman marry the.one \textsc{neg} 3\textsc{a.dep}{}-know-\textsc{asp.dep-dep} work \\
    \glt ‘The woman who didn’t know how to work could not get married.’
\z

\ea\label{ex:mix:83}
    \ea\label{ex:mix:83a}
        \glll ëëjtsës {\ob}mëti’ipë oojts axë’ëy mëkaajpxpo’kxpyën{\cb}.\\ 
              ëëjts=ës {\db}mëti’ipë oojts axë’ëy m-kaajpxpo’kx-p-ën\\
              \textsc{1sg}=1.\textsc{emph} the.one \textsc{pst} yesterday \textsc{2a.ind}{}-greet-\textsc{asp.ind}{}-\textsc{dep}\\
        \glt ‘I am the one you greeted yesterday.’ (lit. ‘I am the one from yesterday you greeted.’)
    \ex\label{ex:mix:83b}
        \glll ka’apës yëëjtsëch {\ob}mëti’ipë oojts xkyaajpxpo’kxën axë’ëy{\cb}. \\
             \textbf{ka’ap=ës} \textbf{y-}ëëjts=\textbf{ëch} {\db}mëti’ipë oojts x-kyaajpxpo’kx-y-ën axë’ëy\\
             \textsc{neg}=1.\textsc{emph} \textsc{3a.dep-1sg}=\textsc{irr} the.one \textsc{pst} \textsc{2a.dep}{}-greet-\textsc{asp.dep-dep} yesterday\\
        \glt ‘I am not the one you greeted yesterday.’
    \ex\label{ex:mix:83c}
        \glll ëëjtsës {\ob}mëti’ipë ka’ap oojts xkyaajpxpo’kxën axë’ëy{\cb}.\\ 
              ëëjts=ës {\db}mëti’ipë \textbf{ka’ap} oojts x-kyaajpxpo’kx-y-ën axë’ëy\\
              \textsc{1}=1.\textsc{emph} the.one \textsc{neg} \textsc{pst} \textsc{2a.dep}{}-greet-\textsc{asp.dep-dep} yesterday \\
        \glt ‘I am the one you didn’t greet yesterday.’
    \ex\label{ex:key83d}
        \glll ka’apës y’ëëjtsëch {\ob}mëti’ipë ka’ap oojts xkyaajpxpo’kxën axë’ëy{\cb}.\\
             \textbf{ka’apës} \textbf{y-}ëëjts=\textbf{ëch} {\db}mëti’ipë \textbf{ka’ap} oojts x-kyaajpxpo’kx-y-ën axë’ëy\\
             \textsc{neg}=1.\textsc{emph} \textsc{3a.dep-1sg}=\textsc{irr} the.one \textsc{neg} \textsc{pst} \textsc{2a.dep}{}-greet-\textsc{asp.dep-dep} yesterday  \\
        \glt ‘I am not the one you didn’t greet yesterday.’
    \z
\z  

The same occurs with complement clauses, as in (\ref{ex:mix:84}--\ref{ex:mix:87}).

\ea\label{ex:mix:84}
    \glll ntë’këpës {\ob}ko yë’ë ka’ap myiita’akëch{\cb}.\\
          n-të’këp=ës {\db}ko yë’ë \textbf{ka’ap} y-miita’akëts-y \\
          1\textsc{a}.\textsc{ind}{}-fear=1.\textsc{emph} \textsc{comp} \textsc{dem} \textsc{neg} 3\textsc{s.dep}{}-win-\textsc{asp.dep}\\
    \glt ‘I’m afraid that he is not going to win.’
\z

\ea\label{ex:mix:85}
    \glll mnä’kxk {\ob}ko yë’ë ka’ap myina’any{\cb}.\\
          n-mnä’kx=k {\db}ko yë’ë \textbf{ka’ap} y-mina’an-y\\
          1\textsc{a}.\textsc{ind}{}-regret=\textsc{evid} \textsc{comp} \textsc{dem} \textsc{neg} \textsc{3s.dep}{}-come-\textsc{asp.dep}\\
    \glt ‘I regret that he can’t come.’
\z

\ea\label{ex:mix:86}
    \glll yë’ë ka’ap oojts ntaatëy {\ob}pen nmiimpyës ëëjts{\cb}. \\
          yë’ë \textbf{ka’ap} oojts t-naatë-y {\db}pen n-miimp-y=ës ëëjts \\
          \textsc{dem} \textsc{neg} \textsc{pst} \textsc{3a.dep}{}-ask-\textsc{asp.dep} \textsc{comp} \textsc{1s.dep-}come-\textsc{asp.dep=1.emph} \textsc{1sg}\\
    \glt ‘He didn’t ask if I’m coming.’
\z

\ea\label{ex:mix:87}
    \glll ka’apës niijaawë {\ob}pen Carlos tëë kääky tkay{\cb}.\\
        \textbf{ka’ap=ës} n-niijaaw-ë {\db}pen Carlos tëë kääky t-kay-y\\
        \textsc{neg=1.emph} \textsc{1a.dep}{}-know-\textsc{asp.dep} \textsc{comp} Carlos before.now tortilla \textsc{3a.dep-}eat-\textsc{asp.dep} \\
    \glt ‘I don’t know whether Carlos ate tortilla.’
\z

In clauses with an epistemic fixed expression, such as ‘I think’ \REF{ex:mix:88a}, however, the negative particle can take scope over either predicate \REF{ex:mix:88b}. To negate both, a different construction must be used \REF{ex:mix:88c}.

\ea\label{ex:mix:88}
    \ea\label{ex:mix:88a}
        \glll miinantëp yo’owë. \\ 
              ø-miinan-të-p yo’owë\\
              \textsc{3s.ind}{}-come-\textsc{pl-asp.ind} I.think\\
        \glt ‘I think they are coming.’
    \ex\label{ex:mix:88b}
        \glll ka’ap yo’owë myiin’a’antë.\\
             \textbf{ka’ap} yo’owë y-miin’a’an-të\\
             \textsc{neg} I.think \textsc{3s.dep}{}-come-\textsc{pl}\\
        \glt ‘I don’t think they are coming.’ = ‘I think they are not coming.’
    \ex\label{ex:mix:88c}
        \glll ka’apës nmëëpëky ko ka’ap myiin’a’antë.\\
             \textbf{ka’ap=ës} n-mëëpëk-y ko \textbf{ka’ap} y-miin’a’an-të\\
             \textsc{neg=1.emph} \textsc{1a.dep-}think\textsc{-asp.dep} \textsc{comp} \textsc{neg} \textsc{3s.dep}-come-\textsc{pl}\\
        \glt ‘I don’t think that they are not coming.’
    \z
\z 

Lexically idionsycratic negatives do not occur. However, negative potential and negative desiderative clauses exhibit a different structure. While the negator \textit{ka’ap} is used, there is also a second negator \textit{nii=}. The negator \textit{nii=} also occurs in constitutent negation (see \sectref{sec:mix:3.2}). In the potential, \textit{nii=} surfaces as part of an invariant form \textit{ntiikeepy} ‘can’t’ (\ref{ex:mix:89a}--\ref{ex:mix:89f}), which in the future changes to \textit{ntiikäp’a’any} (\ref{ex:mix:89g}--\ref{ex:mix:89h}). Used by itself with the negator \textit{ka’ap}, it translates as ‘it can’t be done’, as in \REF{ex:mix:90}. The affirmative \textit{nkeejpy} ‘can’ behaves in the same way. 

\ea\label{ex:mix:89}
    \ea\label{ex:mix:89a}
        \glll  nkeejpyës nëëkxëch.\\
             nkeejpy=ës n-nëëkx=ëch\\
             can=1.\textsc{emph} \textsc{1s.dep}{}-go=\textsc{irr}\\
        \glt ‘I can go.’
    \ex\label{ex:mix:89b}
        \glll ka’apës ntiikeepy nëëkxëch.\\ 
              \textbf{ka’ap=ës} \textbf{ntiikeepy} n-nëëkx=ëch\\
              \textsc{neg=}1.\textsc{emph} \textsc{neg}.can \textsc{1s.dep}{}-go=\textsc{irr}\\
        \glt ‘I can’t go.’
    \ex\label{ex:mix:89c}
        \glll nkeejpy mnëëkxëch.\\
             nkeejpy m-nëëkx=ëch\\
             can \textsc{2s.dep}{}-go=\textsc{irr}\\
        \glt ‘You can go.’
    \ex\label{ex:mix:89d}
        \glll ka’ap ntiikeepy mnëëkxëch. \\ 
              \textbf{ka’ap} \textbf{ntiikeepy} m-nëëkx=ëch\\
              \textsc{neg} \textsc{neg}.can \textsc{2s.dep}{}-go=\textsc{irr}\\
        \glt ‘You can’t go.’
    \ex\label{ex:mix:89e}
        \glll oojts nkeejpyës nëëkxëch. \\
             oojts nkeejpy=ës n-nëëkx=ëch \\
             \textsc{pst} can=1.\textsc{emph} \textsc{1s.dep}{}-go=\textsc{irr}\\
        \glt ‘I could go.’
    \ex\label{ex:mix:89f}
        \glll ka’ap oojts ntiikeepy nëëkxëch. \\ 
             \textbf{ka’ap} oojts \textbf{ntiikeepy} n-nëëkx=ëch \\
             \textsc{neg} \textsc{pst} \textsc{neg}.can \textsc{1s.dep}{}-go=\textsc{irr}\\
        \glt ‘I couldn’t go.’
    \ex\label{ex:mix:89g}
        \glll nkeejpyës nëëkxëch jëpomp. \\
             nkeejpy=ës n-nëëkx=ëch jëpomp\\
             can=1.\textsc{emph} \textsc{1s.dep}{}-go=\textsc{irr} tomorrow\\
        \glt  ‘I can go tomorrow.’
    \ex\label{ex:mix:89h}
        \glll  ka’ap ntiikäp’a’any jëpomp nëëkxëch. \\ 
             \textbf{ka’ap} \textbf{ntiikäp’a’any} jëpomp nëëkx=ëch \\
             \textsc{neg} \textsc{neg}.can.\textsc{fut} tomorrow \textsc{2s.dep}{}-go=\textsc{irr} \\
        \glt ‘I can’t go tomorrow.’
    \z
\z 

\ea\label{ex:mix:90}
    \glll ka’ap (cham) ntiikeepy.\\
        \textbf{ka’ap} (cham) \textbf{ntiikeepy} \\
        \textsc{neg} (now) \textsc{neg}.can \\
    \glt ‘It can’t be done (now).’
\z

\ea\label{ex:mix:91}
    \glll ka’apëkä’ä tniikääpx.\\
          \textbf{ka’ap}=ëk=jä’ä t-\textbf{nii-}kääpx-y \\
          \textsc{neg=evid=dem} 3\textsc{a.dep-neg}{}-speak-\textsc{asp.dep} \\
    \glt ‘(The gopher) could not speak (well).’
\z

\ea\label{ex:mix:92}
    \glll ets niika’apëka pi’kana’k tntëkë’ënë tääkë. \\
          ets \textbf{nii=ka’ap}=ëk=ja’a pi’kana’k t{}-ntëkë’ë-në tääkë \\
          and \textsc{neg=neg=evid=dem} child \textsc{3a.dep}{}-devour-\textsc{asp.dep} like.that  \\
    \glt ‘And he could not devour a child like that.’
\z

The second negator \textit{nii=} can also attach to other predicates or the negator \textit{ka’ap} itself, and the clause assumes a potential or optative meaning, as in \REF{ex:mix:91} and \REF{ex:mix:92}. 

The optative is formed with a change in the verb stem and the same verbal suffixes found with the copula \textit{i’t} in non-verbal negation. These suffixes are glossed as desideratives here. The negator \textit{ka’ap} is used to negate optatives, as in \REF{ex:mix:93b}, \REF{ex:mix:94b} and \REF{ex:mix:95b}. If there is a change in subject between the main clause and the dependent clause, then an irrealis marker is used in both the affirmative and the negative, and the dependent clause can be introduced by an optional complementizer \textit{ko} \REF{ex:mix:97}. If the subject stays the same, the negative desiderative is used in the dependent clause \REF{ex:mix:96}.

\ea\label{ex:mix:93}
    \ea\label{ex:mix:93a}
        \glll nëëkx’aampës.\\
             nëëkx-’aam-p=ës\\
             go-\textsc{des-asp.ind=1.emph}\\
        \glt ‘I want to go.’
    \ex\label{ex:mix:93b}
        \glll ka’apës nëëkx’a’any.\\ 
              \textbf{ka’ap=ës} n-nëëkx-’a’an-y\\
              \textsc{neg-1.emph} \textsc{1s.dep}{}-go-\textsc{des.neg-asp.dep}\\
        \glt ‘I don’t want to go.’
    \ex\label{ex:mix:93c}
        \glll nëëkx’aampës ënech.\\
             nëëkx-’aam-p=ës ënech\\
             go-\textsc{des-asp.ind=1.emph} \textsc{pst}\\
        \glt ‘I wanted to go.’
    \ex\label{ex:mix:93d}
        \glll ka’apës ënech nëëkx’a’any.\\ 
              \textbf{ka’ap=ës} ënech n-nëëkx-’a’an-y\\
              \textsc{neg-1.emph} \textsc{pst} \textsc{1s.dep}{}-go-\textsc{des.neg-asp.dep}\\
        \glt ‘I didn’t want to go.’
    \z
\z 

\ea\label{ex:mix:94}
    \ea\label{ex:mix:94a}
        \glll  ääjts’aampës. \\
             ääjts-’aam-p=ës \\
             dance-\textsc{des-asp.ind=1.emph}\\
        \glt ‘I want to dance.’
    \ex\label{ex:mix:94b}
        \glll ka’apës n’ääjts’a’any.\\ 
              \textbf{ka’ap=ës} n-’ääjts-’a’an-y\\
              \textsc{neg-1.emph} \textsc{1s.dep-}dance-\textsc{des.neg-asp.dep}\\
        \glt ‘I don’t want to dance.’
    \ex\label{ex:mix:94c}
        \glll ääjts’aanëpës. \\
             ääjts-’aanë-p=ës \\
             dance-\textsc{des-asp.ind=1.emph}\\
        \glt ‘I will want to dance (tomorrow).’
    \ex\label{ex:mix:94d}
        \glll ka’apës n’ääjts’a’any. \\ 
              \textbf{ka’ap=ës} n-’ääjts-’a’an-y \\
              \textsc{neg-1.emph} \textsc{1s.dep}{}-dance-\textsc{des.neg-asp.dep}\\
        \glt ‘I won’t want to dance (tomorrow).’
    \z
\z 
 
\ea\label{ex:mix:95}
    \ea\label{ex:mix:95a}
        \glll ntsoojkyëpyës miijts mnëëkxëch.\\
             n-tsoojkyë-py=ës miijts m-nëëkx=ëch\\
             \textsc{1a.ind-}want\textsc{-asp.ind=1} \textsc{2sg} \textsc{2s.dep-}go=\textsc{irr} \\
        \glt ‘I want you to go.’
    \ex\label{ex:mix:95b}
        \glll ka’apës ntsoky miijts mnëëkxëch.\\ 
              \textbf{ka’ap=ës} n-tsok-y miijts m-nëëkx=ëch\\
              \textsc{neg-1.emph} \textsc{1a.dep-}want\textsc{-asp.dep} \textsc{2sg} \textsc{2s.dep-}dance=\textsc{irr}\\
        \glt ‘I don’t want you to go.’ = ‘I want you not to go.’ 
    \z
\z 

\ea\label{ex:mix:96}
    \glll ka’apës ntsoky ka’apës nkay’ä’äny. \\
         \textbf{ka’ap=ës} n-tsok-y \textbf{ka’ap=ës} n-kay-’ä’än-y\\
         \textsc{neg-1.emph} \textsc{1a.dep-}want\textsc{-asp.dep} \textsc{neg-1.emph} \textsc{1a.dep-}eat\textsc{-des.neg-asp.dep} \\
    \glt ‘I don’t want not to eat.’
\z
 
\ea\label{ex:mix:97}
    \ea\label{ex:mix:97a}
        \glll ntsoojkyëpyës (ko) (miijts) ka’ap mkayëch.\\ 
              n-tsoojkyë-py=ës (ko) (miijts) \textbf{ka’ap} m-kay-ëch\\
              \textsc{1a.dep-}want\textsc{-asp.ind=1.emph} \textsc{comp} \textsc{2sg} \textsc{neg} \textsc{2sg-}eat=\textsc{irr}\\
        \glt ‘I want you not to eat.’
    \ex\label{ex:mix:97b}
        \glll ka’apës ntsoky (ko) miijts mkayëch.\\
             \textbf{ka’ap=ës} n-tsok-y (ko) miijts m-kay-ëch \\
             \textsc{neg-1.emph} \textsc{1a.dep-}want\textsc{-asp.dep} \textsc{comp} \textsc{2sg} \textsc{2sg-}eat=\textsc{irr}\\
        \glt ‘I don’t want you to eat.’
    \ex\label{ex:mix:97c}
        \glll ka’apës ntsoky (ko) miijts ka’ap mkayëch.\\
             \textbf{ka’ap=ës} n-tsok-y (ko) miijts \textbf{ka’ap} m-kay-ëch\\
             \textsc{neg-1.emph} \textsc{1a.dep-}want\textsc{-asp.dep} \textsc{comp} \textsc{2sg} \textsc{neg} \textsc{2sg-}eat=\textsc{irr} \\
        \glt ‘I don’t want you not to eat.’
    \z
\z 

In some dependent clauses, the negative prefix \textit{ka’-} occurs instead of the negator \textit{ka’ap}, as in \REF{ex:mix:98}. This is rare, however. 

\ea\label{ex:mix:98}
    \glll ko jä’ä kääky kya’ku’uxë.\\
          ko jä’ä kääky y-\textbf{ka’-}ku’uxë\\
          \textsc{comp} \textsc{dem} tortilla 3\textsc{s.dep-neg}{}-fill.up.\textsc{dep}\\
    \glt ‘[...] (and) when he didn’t fill up with tortillas [...]’
\z

The same negator \textit{ka-} occurs in lest-clauses, as in \REF{ex:mix:99} and \REF{ex:mix:100}.

\ea\label{ex:mix:99}
    \glll je’eyës naaxweech nkoo kos nka’ech.\\
          je’eyë=s n-naaxweech nkoo ko-s n-\textbf{ka’}{}-ech \\
          only=1 \textsc{1s.dep}-sit lest \textsc{comp}-1 \textsc{1s.dep-neg-}dance.\textsc{dep}\\
    \glt ‘I only sat down so that I wouldn’t (have to) dance.’
\z

\ea\label{ex:mix:100}
    \glll je’eyës oo nëëkx nkoo kos nka’mëkääpx.\\
          je’eyë=s oo n-nëëkx nkoo ko-s n-\textbf{ka’-}mëkääpx\\
          only=1 \textsc{pst} \textsc{1s.dep-}go lest \textsc{comp}-1 \textsc{1a.dep-neg-}talk.to.\textsc{dep}\\
    \glt ‘I left so that I wouldn’t have to talk to him/her.’
\z

\subsection{Negative lexicalizations}
No negative lexicalizations have been found in the data.

 \section{Non-clausal negation}\label{sec:mix:3}

This section treats negative replies to polar questions and negative indefinites, as well as negative concord and asymmetries.

\subsection{Negative replies}\label{sec:mix:3.1}
\largerpage
A negative polar question can be answered either with an affirmative or with a negative clause. The negator \textit{ka’ap} can be used as a single word answer to polar questions or as an answer to a polar question followed by a more elaborate answer that contradicts a previous statement, as in \REF{ex:mix:101}. Crosslinguistically, it is common to use negative existentials, such as \textit{ka’ap}, for short, standalone answers 'no' \citep{Veselinova2013neg.exist}, as in \REF{ex:mix:102}. 

\ea\label{ex:mix:101}
    \begin{xlist}
    \exi{A: }\label{ex:mix:101a}
        \glll wi’ix ka’ap oo pën mpuuto’oxy ‘i’të?\\ 
              wi’ix \textbf{ka’ap} oojts pën mpuuto’oxy ‘i’të\\
              what \textsc{neg} \textsc{pst} someone couples there.were\\ 
        \glt ‘What, there weren’t any couples (present)?’
    \exi{B: }\label{ex:mix:101b}
        \glll tämmës to’oxychëëjk. \\ 
              tämmës to’oxychëëjk  \\
              there.were women\\
        \glt ‘Yes, there were women (present).’
    \end{xlist}
\z 

\ea\label{ex:mix:102}
\begin{xlist}[B2: ]
    \exi{A: }\label{ex:mix:102a}
        \glll miimp?\\ 
              m-miim-p\\
              \textsc{2s.ind}{}-come-\textsc{asp.ind}\\ 
        \glt ‘Are you coming?’
    \exi{A: }\label{ex:mix:102b}
        \glll kii miny?\\ 
              kii m-min-y\\
              \textsc{neg} \textsc{2s.dep}{}-come-\textsc{asp.dep}\\ 
        \glt ‘Aren’t you coming?’
     \exi{B1: }\label{ex:mix:102c}
        \gl ka’ap.  \\
        \glt ‘No [I am not coming].’
    \exi{B2: }\label{ex:mix:102d}
        \gl oy. \\
        \glt ‘Yes [I am coming].’
    \end{xlist}
\z %\z 

\ea\label{ex:mix:103}
    \begin{xlist}
    \exi{A: }\label{ex:mix:103a}
        \glll yë’ë kii wi’ix tëë m’anë’ëmexëch? \\ 
              yë’ë kii wi’ix tëë m-’anë’ëmex=ëch\\
              \textsc{dem} \textsc{neg} what \textsc{temp} \textsc{2a.dep}{}-tell=\textsc{irr} \\ 
        \glt ‘He didn’t tell you anything?’
    \exi{B: }\label{ex:mix:103b}
        \glll ka’ap, ka’ap wi’ix tëë inä’äny.\\ 
              ka’ap ka’ap wi’ix tëë inä’äny \\
              \textsc{neg} \textsc{neg} what \textsc{temp} tell\\
        \glt ‘No, he didn’t tell me anything.’
    \exi{B: }\label{ex:mix:103c}
        \glll tëës xukowaanë.\\ 
              tëë=s x-tukowaanë \\
              \textsc{temp}=\textsc{1} \textsc{1p.dep}{}-tell\\
        \glt ‘Yes, he told me.’
    \end{xlist}
\z 

The same positive or negative one-word reply can be used to answer a positive or a negative question retaining the same meaning. In \REF{ex:mix:103}, a short pause after the negative particle \textit{ka’ap} indicates a separate intonation unit and thus a one-word answer elaborated in a longer reply. 

\subsection{Negative indefinites}\label{sec:mix:3.2}

Negative indefinites are formed using a proclitic \textit{nii=}, the same with constituent negation (see \sectref{sec:mix:4.1}), different from the negator \textit{ka’ap}. For instance, in negative indefinites such as ‘nobody’ \textit{niipën} or ‘nowhere’ \textit{niimaa}, the proclitic \textit{nii=} is added to the interrogatives \textit{pën} ‘who’ and \textit{maa} ‘where’, respectively. The presence of a negative indefinite with the proclitic \textit{nii=} generally triggers the negative prefix \textit{ka’-} on the predicate resulting in negative concord, as in (\ref{ex:mix:104}--\ref{ex:mix:106}). It is worth noting that the negative proclitic \textit{nii=} is not to be confused with the homophonous prefix \textit{nii-} signaling animacy.

\ea\label{ex:mix:104}
    \glll niipënëka’a tääpë to’oxychëëjk kya’iixa’chë. \\
          \textbf{nii=pën}=ëk=ja’a tääpë to’oxychëëjk y-\textbf{ka’}{}-’iixa’ch-ë \\
          \textsc{neg=}someone=\textsc{evid=dem} this woman \textsc{3p.dep-neg-}see.meet-\textsc{inv} \\
    \glt ‘Nobody had met (seen) this woman.’
\z

\ea\label{ex:mix:105}
    \glll niimaa kyä’nyëëkxy niiyaa kya’näxy. \\
          \textbf{nii=maa} y-\textbf{ka’}{}-nëëkx-y \textbf{nii=yaa} y-\textbf{ka’}{}-näx-y \\
          \textsc{neg=}where 3\textsc{s.dep-neg}{}-go-\textsc{asp.dep} \textsc{neg=}here 3\textsc{s.dep-neg}{}-walk\textsc{{}-asp.dep } \\ %HC I added the space after dep before the end of textsc just so that the example would separate on two lines. because it had an empty space before the translation...
    \glt ‘She is going nowhere, not even here she is coming through.’
\z

\ea\label{ex:mix:106}
    \glll niipën jyä’ä tka’mënëëkxy. \\
          \textbf{nii=}pën jyä’ä t-\textbf{ka’-}mënëëkxy \\
          \textsc{neg}=someone there \textsc{3a.dep-neg}{}-took-\textsc{asp.dep} \\
    \glt ‘Nobody took it there.’ 
\z

While the three different negators (\textit{nii=, ka’ap, ka’-}) are not attested in a single sentence in naturally\nobreakdash-occurring discourse, \REF{ex:mix:107} is deemed grammatical when elicited from a native speaker. 

\ea\label{ex:mix:107}
    \ea\label{ex:mix:107a}
        \glll niika’na’a ka’ap kyäminy. \\ 
              \textbf{niika’na’a} \textbf{ka’ap} y-\textbf{ka’-}min-y\\
              never \textsc{neg} \textsc{3s.dep}{}-\textsc{neg-}come-\textsc{asp.dep} \\
        \glt ‘S/he is never coming.’
    \ex\label{ex:mix:107b}
        \glll niika’na’a ma niipën kyänëëkxy.\\
             \textbf{niika’na’a} ma \textbf{nii=}pën y-\textbf{ka’}{}-nëëkx-y\\
             never where \textsc{neg}=someone \textsc{3s.dep-neg}{}-go{}-\textsc{asp.dep}\\
        \glt ‘Nobody ever goes anywhere.’
    \z
\z 

The negator \textit{niika’na’a} ‘never’ is a lexicalized form composed of two negators, the negative proclitic \textit{nii=} and the clausal negator \textit{ka‘-}, in addition to the temporal pronoun \textit{na’a} ‘when‘. It can be moved freely in the clause. The negator \textit{nii} occurs also as a separate word \REF{ex:mix:108}, but the same as with \textit{ka’ap}, it must always precede the predicate.

\ea\label{ex:mix:108}
    \glll tää nii yë’ë ja kääky tka’kaychäänyë.\\
          tää \textbf{nii} yë’ë ja’a kääky t-\textbf{ka’-}kaychääny-ë\\
          then \textsc{neg} \textsc{dem} \textsc{dem} tortilla 3\textsc{a.dep-neg}{}-finish.eat-\textsc{inv}\\
    \glt ‘Then they never finished eating the tortilla.’
\z

It is sometimes unclear if the the negator \textit{nii=} represents a Spanish loan (Spanish \textit{ni} ‘not even’ or \textit{ni..ni} ‘neither…nor’). Generally, if it is translated as ‘not even’ or if it is repeated as a separate word, it represents the Spanish loan, as in \REF{ex:mix:109}. The Spanish loan \textit{ni} ‘not even, neither’ can co\nobreakdash-occur with the negator \textit{ka’ap} or with the negative prefix \textit{ka’-}.

\ea\label{ex:mix:109}
    \ea\label{ex:mix:109a}
        \glll ni champaat Godë tka’niijaawë.\\ 
              \textbf{ni} champaat Godë t-\textbf{ka’-}niijaaw-ë\\
              not.even until.now Godo \textsc{3a.dep}-\textsc{neg}{}-know-\textsc{asp.dep}\\
        \glt ‘[...] not even (until) now Godo knows about it.’
    \ex\label{ex:mix:109b}
        \glll ka’ap tmaakääpxy ni comité ni pën.\\
             \textbf{ka’ap} t-maakääpx-y \textbf{ni} comité \textbf{ni} pën\\
             \textsc{neg} \textsc{3a.dep}{}-call-\textsc{asp.dep} neither committee nor someone\\
        \glt ‘He didn’t call anyone, neither the committee nor anyone.’ 
    \z
\z 

\citet[572]{Wichmann1995} identifies \textit{ni} as a negator in proto-Mixe-Zoquean, as in \textit{*ni to} ‘nothing’. The same negator is reconstructed as \textit{*ka ti} for proto-Mixean. Thus, the proclitic \textit{nii=} seems to be of Mixean origin, and the formal resemblance to Spanish \textit{ni} is due to chance. What is interesting is that \textit{nii=/ni} can also attach to the negative particle \textit{ka’ap} in negative concord, as in \REF{ex:mix:110}.

\ea\label{ex:mix:110}
    \glll ets niika’apëka pi’kana’k ntëkë’ënë tääkë. \\
          ets \textbf{nii=ka’ap}=ëk=ja’a pi’kana’k t-nëkë’ën-ë tääkë \\
          and \textsc{neg=neg=evid=dem} child \textsc{3a.dep}{}-devour-\textsc{asp.dep} like.that\\
    \glt ‘And he could not devour a child like that.’
\z

In \REF{ex:mix:110} it is unclear if the negator \textit{nii} represents the Spanish loan \textit{ni} and the sentence thus translates as ‘and he could not even devour a child like that’. 

Negative indefinites can also be expressed using the negator \textit{ka’ap} instead of the proclitic \textit{nii=}, as in (\ref{ex:mix:111}--\ref{ex:mix:112}). In fact, any constituent can be negated using the negator \textit{ka’ap} (see \sectref{sec:mix:4.1}).

\ea\label{ex:mix:111}
    \glll ka’apës pën npaatë tu’unëka inä’äny. \\
          \textbf{ka’ap}=ës pën n-paat-të tu’unëka inä’äny  \\
          \textsc{neg=1} someone 1\textsc{a.excl.dep-}find-\textsc{pl} also they.say \\
    \glt ‘We didn’t find anyone, they also said.’
\z

\ea\label{ex:mix:112}
    \glll ka’apëka ma nyiikxy. \\
          \textbf{ka’ap}=ëk=ja’a ma y-niikx-y\\
          \textsc{neg=evid=dem} where 3\textsc{s.dep}{}-go-\textsc{asp.dep}\\
    \glt ‘She is going nowhere.’
\z

Several negative expressions with a temporal reference (e.g. ‘never, not ever, not anymore, not yet’) are lexicalized. The negator \textit{niika’na’a} or \textit{niina’a} ‘never’ is a lexicalized form composed of either one or two negators, the negative proclitic \textit{nii=} and the clausal negator \textit{ka‘-}, in addition to the temporal pronoun \textit{na’a} ‘when’. The emphatic particle \textit{ntim}/\textit{wiintim/niintim} translated as ‘not ever/never ever’ occurs with the negator \textit{ka’ap} or with other negative words, such as ‘never’ (\ref{ex:mix:113}--\ref{ex:mix:114}). It is unclear if this particle includes the negator \textit{nii=}.  

\ea\label{ex:mix:113}
    \glll wiintim mëku’uk xkyä’kooxtëch niika’ana’a!\\
          \textbf{wiintim} m-mëku’uk x-kä’-koox-të=ëch \textbf{niika’ana’a}\\
          never.ever \textsc{2poss-}friends \textsc{2a.dep-neg}{}-hit-\textsc{pl=irr} never\\
    \glt ‘Never ever hit your friends!’
\z

\ea\label{ex:mix:114}
    \glll ka’ap wiintim jyotkuuk’achy.\\
          \textbf{ka’ap} \textbf{wiintim} y-jotkuuk-’ach-y\\
          \textsc{neg} never.ever \textsc{3s.dep-}happy-\textsc{vbz-asp.dep}\\
    \glt ‘She’s never ever happy.’
\z

Clauses with negative indefinites can exhibit asymmetries with respect to affirmative clauses. As discussed for standard negation, the negative markers generally trigger the dependent verb paradigm, i.e. a separate set of person and aspect markers along with a change in the verb stem. The negator \textit{ka’ap} always triggers the depedent set, while the negative prefix \textit{ka’-} does not.\footnote{This can be explained by the negative existential cycle \citep{Veselinova2016} Negative existentials, such as \textit{ka’ap}, generally form constructions on their own. When they are being used with a verbal predicate, they often require a special form, here the dependent inflection.}. When a negative word with the proclitic \textit{nii=} co-occurs with the negative prefix \textit{ka’-} the predicate can occur either in the dependent or the independent, based on the constituents preceding it. If a non\nobreakdash-core constituent precedes the predicate, dependent inflection occurs, as in \REF{ex:mix:115}; otherwise, independent inflection may be present, as in \REF{ex:mix:116}.

\ea\label{ex:mix:115}
    \glll ka’ap pën chë’ëkë. \\
          \textbf{ka’ap} pën y-tsë’ëk-ë \\
          \textsc{neg} someone \textsc{3s.dep}{}-be.afraid-\textsc{dep.asp} \\
    \glt ‘Nobody was afraid.’
\z

\ea\label{ex:mix:116}
    \glll niipën kyaamëmëtoowëtë. \\
          \textbf{nii=}pën ø-\textbf{ka’-}yaa-mëmëtoow-ë-të \\
          \textsc{neg=}someone \textsc{3p.ind}{}-\textsc{neg-caus}{}-obey-\textsc{inv-pl}\\
    \glt ‘Nobody obeyed him.’
\z

In both examples the constituent ‘nobody’ represents a core argument, but in \REF{ex:mix:115} the negator \textit{ka’ap} triggers the dependent paradigm, while in \REF{ex:mix:116} the negative proclitic \textit{nii=} does not. 

\subsection{Negative derivation and case marking}\label{sec:mix:3.3}

Chuxnabán Mixe does not have an elaborate case system, and no negative derivation has been noted in the data.

\section{Other aspects of negation}\label{sec:mix:4}

\subsection{The scope of negation}\label{sec:mix:4.1}

In terms of scope of negation, it can be said that any of the three negators must preceed the negated constituent or predicate/verb root, and that each constituent or predicate needs to be negated separately (see \sectref{sec:mix:2.1}, \ref{sec:mix:2.2.1}, and \ref{sec:mix:2.4}). 

Constituent negation parallels the construction found in negative indefinites (see \sectref{sec:mix:3.2}). A proclitic \textit{nii=} is added to the negated constituent which triggers the negative prefix \textit{ka’-} on the predicate resulting in negative concord. The proclitic can also occur with numerals, as in \textit{niimääjtsk} ‘not the two’. Any constituent can be negated using the negator \textit{ka’ap}, as in \REF{ex:mix:117} and \REF{ex:mix:118}.

\ea\label{ex:mix:117}
    \glll ka’ap tuky ka’ap wi’ix ka’ap iixtëm ëëjts neychukatuujkëmë. \\
          \textbf{ka’ap} tuky \textbf{ka’ap} wi’ix \textbf{ka’ap} iixtëm ëëjts n-eychukatuujk-ëmë \\
          \textsc{neg} pieces \textsc{neg} like \textsc{neg} as.seen 1\textsc{sg} 1\textsc{a.dep}{}-cut-1\textsc{pl.incl} \\
    \glt ‘[...] not pieces, not like that, not the same way like we cut (the tortillas).’
\z

\ea\label{ex:mix:118}
    \glll miijtsëch ka’ap mtsë’ëkëtëch ka’ap wi’ix mjata’antë. \\
          miijts-ëch ka’ap m-tsë’ëkë-të=ëch \textbf{ka’ap} wi’ix m-jata’an-të  \\
          2-\textsc{pl} \textsc{neg} 2\textsc{s.dep-}be.afraid-\textsc{pl=irr} \textsc{neg} like.that \textsc{2p.dep}-happen-\textsc{pl} \\
    \glt ‘Don’t be afraid, nothing like that will happen to you (pl).’
\z

Effects of intonation and stress have yet to be studied for Chuxnabán Mixe.

\subsection{Negative polarity}\label{sec:mix:4.2}

No negative polarity items have been found in the language.


\subsection{Marking of NPs in the scope of negation}\label{sec:mix:4.3}

Chuxnabán Mixe does not exhibit a robust case marking system, so no effects are noted on the marking of NPs with respect to negation.

\subsection{Reinforcing negation}\label{sec:mix:4.4}

The fact that multiple negators are possible (see \sectref{sec:mix:2.1}) and required in some instances (see \sectref{sec:mix:3.2}) could be considered instances of reinforcing negation, in particular in the rare cases where all three negators occur in the same sentence. It can also be speculated that \textit{ka’ap} can be a univerbation of the negative prefix \textit{ka’-} and a positive element used for emphasis. It can further be speculated that \textit{ka’ap} represents an emphatic form of the negative marker that was used in addition to the original negative prefix \textit{ka-} and then started to take over (see \citealt{Gelderen2016}). Now the old negator \textit{ka’-} is used mostly in sentences with constituent negation.

\subsection{Negation and complex clauses}\label{sec:mix:4.5}

In Chuxnabán Mixe, each clause is negated separately (see \sectref{sec:mix:2.4}). For coordinate clauses, it can be noted that the negator \textit{ka’ap} gets sometimes translated as ‘neither’ or ‘not yet’ when it follows the particle \textit{nen}, whereby \textit{nen} translates as ‘also’, as in \REF{ex:mix:119} and \REF{ex:mix:120}. The prohibitive particle \textit{këtii} is also used to form the negative coordinator \textit{neekëtii} ‘neither’, as in \REF{ex:mix:121}.

\ea\label{ex:mix:119}
    \glll perë nen ka’ap eech agentë jap. \\
          perë \textbf{nen} \textbf{ka’ap} eech agentë jap \\
          but also \textsc{neg} was town.major there \\
    \glt ‘[...] but neither was the town major present.’
\z

\ea\label{ex:mix:120}
    \glll nen kaˈap eech ja jëën noˈoktë.\\
          \textbf{nen} \textbf{kaˈap} eech ja’a jëën n-noˈok-të\\
          also \textsc{neg} was \textsc{dem} fire \textsc{1s.dep}{}-burn-\textsc{pl}\\
    \glt ‘[...] and we didn’t burn fire yet.’
\z

\ea\label{ex:mix:121}
    \glll neekëtii miny. \\
          nee\textbf{këtii} m-miny\\
          neither \textsc{2s.dep-}come\\
    \glt ‘Neither are you coming.’
\z

This topic needs further study. It must be noted that Chuxnabán Mixe often uses Spanish clause conectors, such as \textit{ni…ni} ‘neither…nor’.

\section{Summary}\label{sec:mix:5}

Chuxnabán Mixe, the same as its close relatives, has three main negators: the negative particle \textit{ka’ap}, the negative verbal prefix \textit{ka’-}, and the proclitic \textit{nii=} used in constituent negation. A special negator \textit{këtii/kii} is found in prohibitives. Overall, the negator \textit{ka’ap} can be used for virtually any type of negation, including prohibitives (see \sectref{sec:mix:2.2.1}) and constituent negation (see \sectref{sec:mix:3.2}). In prohibitives, however, the irrealis enclitic \textit{=ëch} is obligatory with \textit{ka’ap}, to distinguish the clause from standard negation; it is optional with the special prohibitive negator (see \sectref{sec:mix:2.2.1}). The formal resemblance between the negative particle and the negative prefix potentially points to a common source, whereby the negative prefix may represent the only negator and the negative particle the new form. 

As expected, negation strategies in non-verbal clauses show many similarities. Overall, the negative existential \textit{ka’ap} is used as the main negator. Negative constructions in equation and attribution overlap, as well as in locative and existential predication (\sectref{sec:mix:2.3}). Interesting is the fact that non-verbal clauses exhibit multiple strategies for both affirmatives and negatives with different verbalizers and the irrealis enclitic \textit{=ëch} in some negative forms.
\tabref{tab:mix:11new} summarizes the negators used for each type of negation.


\begin{table}[b]
\begin{tabularx}{\textwidth}{lXXl}
\lsptoprule
& \textit{ka’ap} & \textit{{ka’}-} & {\textit{nii}=}\\
\midrule
{Dependent} & {x} & & \\
{Dependent} & {x} & {x} & \\
{Dependent} & {x} & & {x}\\
{Dependent} & {x} & {x} & {x}\\
{Dependent/Independent} & & {x} & {x}\\
{Dependent/Independent} & & {x} & \\
{Dependent/Independent} & & & {x}\\
\lspbottomrule
\end{tabularx}
\caption{Summary of dependent/independent marking with different negators}
\label{tab:mix:11}
\end{table}






\begin{table}
\small
\begin{tabularx}{\textwidth}{lQll}
\lsptoprule
\multicolumn{2}{l}{\textbf{Negation in declarative verbal clauses}} & {Chapter} & {Examples}\\
\midrule
{\textit{ka’ap}}    & {preverbal particle triggering special inflection used in independent clauses} & {\ref{sec:mix:2.1}} & {(\ref{ex:mix:3}--\ref{ex:mix:7})}\\
                    & {and in dependent clauses} & {\ref{sec:mix:2.4}} & {(\ref{ex:mix:82}--\ref{ex:mix:87})}\\
{\textit{ka’-}}     & {verbal prefix used in independent clauses} & {\ref{sec:mix:2.1}} & {(\ref{ex:mix:10}--\ref{ex:mix:11})} \\
                    & {and in some dependent clauses} & {\ref{sec:mix:2.4}} & {(\ref{ex:mix:98}--\ref{ex:mix:100})}\\
                    
{\textit{ka’ap + jëk}} & {“not anymore”} & {\ref{sec:mix:2.1}} & {(\ref{ex:mix:8}--\ref{ex:mix:9})}\\
{\itshape nen + ka’ap} & {“neither”, “not yet”} & {\ref{sec:mix:4.5}} & {(\ref{ex:mix:119}--\ref{ex:mix:120})}\\
{\textit{ka’- + nëm}} & {“not yet”} & {\ref{sec:mix:2.1}} & {\REF{ex:mix:12}}\\
{\itshape ka’ap + nii=} & {negative potential} & {\ref{sec:mix:2.4}} & {(\ref{ex:mix:89}--\ref{ex:mix:92})}\\
{\textit{ka’ap} (or \textit{ka’-})} & {negative optative} & {\ref{sec:mix:2.4}} & {(\ref{ex:mix:93}--\ref{ex:mix:97})}\\

\tablevspace
\multicolumn{4}{l}{\textbf{Negation in non-declarative verbal clauses}}   \\
\midrule
{\textit{këtii or kii}} & {preverbal particle (sometimes with optional irrealis enclitic \textit{=ëch}) used in} &&\\
    & {a) prohibitives} &  {\ref{sec:mix:2.2.1}} & {(\ref{ex:mix:21}--\ref{ex:mix:27})} \\
    & {b) negative polar questions (also in non-verbal clauses)} & {\ref{sec:mix:2.2.2}} & {(\ref{ex:mix:36}--\ref{ex:mix:39})}\\
    & {c) clauses with “neither” (\textit{neekëtii})} & {\ref{sec:mix:4.5}} & {\REF{ex:mix:121}}\\

{\itshape ka’ap + =ëch} & {preverbal particle + irrealis enclitic \textit{=ëch} used in prohibitives} & {\ref{sec:mix:2.2.1}} & {(\ref{ex:mix:32}--\ref{ex:mix:33})}\\
{\itshape ka’ap} & {Preverbal particle in negative polar questions (also in non-verbal clauses)} & {\ref{sec:mix:2.2.2}} & {(\ref{ex:mix:34})}\\
{\itshape ka’-} & {Verbal prefix in negative polar questions (also in non-verbal clauses)} & {\ref{sec:mix:2.2.2}} & {\REF{ex:mix:35}}\\

\tablevspace
\multicolumn{4}{l}{{\textbf{Negation in non-verbal clauses (equative, existential, locative, possessive)}}}  \\
\midrule
{\textit{ka’ap} (\textit{+ =ëch})} & {preverbal particle (for details, see \tabref{tab:mix:9})} & {\ref{sec:mix:2.3}} & {(\ref{ex:mix:42}--\ref{ex:mix:81})}\\

\tablevspace
\multicolumn{4}{l}{{\textbf{Non-clausal negation}}}   \\
\midrule 
{\textit{nii= + ka’-}} & {proclitic and verbal prefix used in constituent negation and negative indefinites} & {\ref{sec:mix:3.2}} & {(\ref{ex:mix:104}--\ref{ex:mix:106})}\\
{\itshape ka’ap} & {particle used in constituent negation} & {\ref{sec:mix:4.1}} & {(\ref{ex:mix:117}--\ref{ex:mix:118})}\\
    & {and in negative indefinites} &{\ref{sec:mix:3.2}}& {(\ref{ex:mix:111}--\ref{ex:mix:112})}\\
    & {and in negative replies} & {\ref{sec:mix:3.1}} & {(\ref{ex:mix:102}--\ref{ex:mix:103})}\\
\lspbottomrule
\end{tabularx}
\caption{Summary of negation in Chuxnabán Mixe}
\label{tab:mix:11new}
\end{table}


Negative concord occurs for the most part in clauses where the negator \textit{ka’ap} is not present. In fact, only \textit{ka’ap} shows unrestricted use as the sole negator in a clause. Although the negative prefix \textit{ka’-} may function as the only negator in a clause, this happens rarely, and there are restrictions associated with its use (see \sectref{sec:mix:2.1.1}).

While paradigmatic asymmetries have not been noted for Chuxnabán Mixe, constructional asymmetries are sometimes present. There are two main verb paradigms, the dependent and the independent, each associated with a separate set of person prefixes, aspect suffixes, and verb stem shapes. \tabref{tab:mix:11} summarizes the presence of dependent marking in clauses with different types of negators. The horizontal axis indicates the presence of one or more negative markers in a clause; the vertical axis shows the inflection type on the predicate.


All in all, negation in Chuxnabán Mixe comprises different constructions for different types of negative expressions often with different competing constructions and negative markers interacting with one another and with the grammatical structure of the clause. The main goal of this first comprehensive study of negation in a Mixe-Zoquean language has been to yield a better understanding of negation in Mixe-Zoquean languages to pave the road for future studies, and to contribute to a systematic cross\nobreakdash-linguistic comparison of negation. 

\section*{Abbreviations}
\begin{tabularx}{.45\textwidth}{lQ}
1 & first person \\
2 & second person \\
3 & third person \\
3’ & non-salient third person\\
\scshape a & agent(-like) argument of a transitive verb \\
\scshape an & animate \\
\scshape asp & aspect \\
\scshape caus & causative \\
\scshape comp & complementizer \\
\scshape dem & demonstrative \\
\scshape dep & dependent* \\
\scshape des & desiderative \\
\scshape emph & emphatic \\
\scshape evid & evidential \\
\scshape excl & exclusive \\
\scshape ex & existential \\
\scshape fut & future \\
\scshape imp & imperative \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape incl & inclusive \\
\scshape ind & independent* \\
\scshape inv & inverse \\
\scshape irr & irrealis \\
\scshape loc & locative \\
\scshape neg & negative \\
\scshape nml & nominal \\
\scshape p & patient(like) argument of a transitive verb \\
\scshape pers & person \\
\scshape pl & plural \\
\scshape poss & possessive \\
\scshape pst & past tense \\
\scshape q & question \\
\scshape s & single argument of an intransitive verb \\
\scshape sg & singular \\
\scshape temp & temporal particle \\
\scshape vbz & verbalizer \\
\end{tabularx}
\medskip
\noindent

*Please note that these are language-family specific terms and do not refer to dependency in the sense of a complex sentences (where you have a main clause and a subordinate/dependent clause)



\sloppy\printbibliography[heading=subbibliography,notkeyword=this]

\end{document} 
