\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18234071}
\author{Rachel Nordlinger\affiliation{University of Melbourne}}
\title{Negation in Murrinhpatha}
\abstract{This chapter discusses negation strategies in Murrinhpatha, a non-Pama-Nyungan polysynthetic language from the Daly region of northern Australia. Murrinhpatha is spoken by 2500–3000 people, mainly located in the remote township of Wadeye (Port Keats) in the Northern Territory, where it is the language of daily communication and one of a small number of Australian languages still learned as a first language by all children in the community. Standard negation structures in Murrinhpatha are typical for languages of northern Australia, with an initial negative particle combining with a verb in irrealis mood; therefore of the mixed Type SymAsy category of \citet{Miestamo2013-wals114}. The expression of negative existence and possession requires a distinct construction type involving specialized negative predicates formed from the nominal classifiers. Such a system appears new to the typological literature on negation.

\keywords{Australian languages, nominal classifiers, irrealis, negation, Murrinhpatha}
}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 
%\shorttitlerunninghead{}%%use this for an abridged title in the page headers

\section{The language} \label{sec:mur:1}

This chapter discusses negation strategies in Murrinhpatha (ISO: \href{https://iso639-3.sil.org/code/mwf}{mwf}, Glottocode: \href{https://glottolog.org/resource/languoid/id/murr1258}{murr1258}), a non-Pama-Nyungan polysynthetic language from the Daly region of northern Australia. Murrinhpatha is spoken by 2500–3000 people, mainly located in the remote township of Wadeye (Port Keats) in the Northern Territory. Murrinhpatha is the primary language of daily communication in Wadeye and surrounding communities, and one of a small number of Australian languages still learned as a first language by all children in the community. It is therefore one of the stronger traditional languages in the Australian context. 

In early research Murrinhpatha was assumed to be a language isolate \citep{Tryon1974}, but more recently it has been shown to belong to the Southern Daly family with Ngan’gityemerri \citep{Green2003}. Broader relationships are as yet unestablished, but there are clear areal (or possibly genetic) similarities with the Western Daly family, as discussed in \citet{Nordlinger2017}. Aspects of the grammatical structure of Murrinhpatha have been described by a number of researchers, most notably \citeauthor{Walsh1976} (e.g. \citeyear{Walsh1976}; \citeyear{Walsh1996}; \citeyear{Walsh1997}), \citeauthor{Street1987} (e.g. \citeyear{Street1987}; \citeyear{Street1996}), and more recently \citeauthor{Blythe2009} (e.g. \citeyear{Blythe2009}; \citeyear{Blythe2013}), \citeauthor{Mansfield2014} (e.g. \citeyear{Mansfield2014}; \citeyear{Mansfield2016}; \citeyear{Mansfield2019}) and \citeauthor{Nordlinger2010} (e.g. \citeyear{Nordlinger2010}; \citeyear{Nordlinger2011}; \citeyear{Nordlinger2015}; \citeyear{Nordlinger2017}), but none of these publications have focused in detail on negation. 

Murrinhpatha shares many typological characteristics common to polysynthetic languages, including syntactically free clausal word order, optionality of argument NPs, and a complex verbal template. This means that clauses often consist of just a verb, and that external NPs for arguments are often not present. Verbs in Murrinhpatha are (mostly) complex predicates, made up of two discontinuous stem elements which together define the predicational semantics and argument structure. These are referred to as the ‘classifier stem’ and the ‘lexical stem’. Classifier stems belong to 38 distinct paradigms encoding generic event-classifying semantics and also encode tense/mood and subject person and number. 11 of the 38 classifier stems can occur alone as the sole verbal predicate. Lexical stems provide more specific lexical semantics and can never occur alone without combining with a classifier stem.\footnote{For a more detailed discussion of the system see \citet{Nordlinger2015}, \citet{Mansfield2019}, and the references cited therein.}

The basic templatic structure of the Murrinhpatha verb is shown in \tabref{tab:mur:1};\footnote{\citet{Mansfield2019} notes some variation in morph order on the right edge of this template, especially in the speech of younger speakers, but this variation doesn’t affect the data being discussed in this paper.}
the two stem elements are given here in bold. The only truly obligatory slot in the verb is Slot 1, since some verbs can consist of just a classifier stem, but most verbs are complex predicates, and so have slots 1 and 5 minimally filled.

% \begin{table}
% \begin{tabularx}{\textwidth}{@{}Q|Q|Q|Q|Q|Q|Q|Q|Q@{}}
% % \lsptoprule
% {Slot 1} & {Slot 2} & {Slot 3} & {Slot 4} & {Slot 5} & {Slot 6} & {Slot 7} & {Slot 8} & {Slot 9}\\
% \hline
% \textbf{Classifier stem} (portmanteau with \textsc{subj} person/number and TAM) & {\textsc{subj} Number / \textsc{obj} marker / \textsc{obl} marker} & {Reflexive-reciprocal} & {Incorporated body part / Applicative} & \textbf{Lexical Stem} & {Tense/ aspect/ mood (TAM)} & {Incorporated adverbial} & {Number (\textsc{subj} or \textsc{obj})} & {Incorporated adverbial} \\
% % \lspbottomrule
% \end{tabularx}
% \caption{Murrinhpatha basic verbal template}
% \label{tab:mur:1}
% \end{table}

\begin{sidewaystable}
\begin{tabularx}{\textwidth}{@{}Q|Q|Q|p{20mm}|p{12mm}|p{12mm}|p{20mm}|Q|p{20mm}@{}}
% \lsptoprule
{Slot 1} & {Slot 2} & {Slot 3} & {Slot 4} & {Slot 5} & {Slot 6} & {Slot 7} & {Slot 8} & {Slot 9}\\
\hline
\textbf{Classifier stem} (portmanteau with \textsc{subj} person/number and TAM) & {\textsc{subj} Number / \textsc{obj} marker / \textsc{obl} marker} & {Reflexive-reciprocal} & {Incorporated body part/\newline  Applicative} & \textbf{Lexical Stem} & {Tense/ aspect/ mood (TAM)} & {Incorporated adverbial} & {Number (\textsc{subj} or \textsc{obj})} & {Incorporated adverbial} \\
% \lspbottomrule
\end{tabularx}
\caption{Murrinhpatha basic verbal template}
\label{tab:mur:1}
\end{sidewaystable}

% \begin{table}\footnotesize
% \begin{tabularx}{\textwidth}{@{}Q|Q|Q|Q|Q|Q|Q|Q|Q@{}}
% % \lsptoprule
% {Slot 1} & {Slot 2} & {Slot 3} & {Slot 4} & {Slot 5} & {Slot 6} & {Slot 7} & {Slot 8} & {Slot 9}\\
% \hline
%                                                                                 & {\textsc{subj} Number / \textsc{obj} marker / \textsc{obl} marker} &                        & {Incorporated body part / Applicative} &                       & {Tense/ aspect/ mood (TAM)} &                          & {Number (\textsc{subj} or \textsc{obj})} &                          \\
% \textbf{Classifier stem}   &                                                                    & {Reflexive-reciprocal} &                                        & \textbf{Lexical Stem} &                             & {Incorporated adverbial} &                                          & {Incorporated adverbial} \\
% (portmanteau with \textsc{subj} person/number and TAM) \\
% % \lspbottomrule
% \end{tabularx}
% \caption{Murrinhpatha basic verbal template}
% \label{tab:mur:1}
% \end{table}

%
% \renewcommand{\rotatehead}[2][1cm]{
% %width is the width of the parbox
% %height is the buffer space used to vertically stretch the headere
% \setlength{\rotwidth}{#1}
% \setlength{\rotheight}{.85\rotwidth}
%   \begin{rotate}{60}~ %nbsp shifts the content away from the line underneath
%    \parbox{\rotwidth}{\raggedright #2}
%   \end{rotate}%
%   \rule{0pt}{\rotheight} %add zero width rule to get the right table height
% }
%
%
% \begin{table}\footnotesize
% \begin{tabularx}{\textwidth}{@{}Q|Q|Q|Q|Q|Q|Q|Q|Q@{}}
% % \lsptoprule
% {Slot 1} & {Slot 2} & {Slot 3} & {Slot 4} & {Slot 5} & {Slot 6} & {Slot 7} & {Slot 8} & {Slot 9}\\
% \hline
% \rotatehead[2.5cm]{\textbf{Classifier stem}}\newline  (portmanteau with \textsc{subj} person/number and TAM) &
% \rotatehead[2.5cm]{\textsc{subj} Number/\newline \textsc{obj} marker/\newline \textsc{obl} marker} &
% \rotatehead[2.5cm]{Reflexive-\newline reciprocal} &
% \rotatehead[2.5cm]{Incorporated\newline body part/\newline Applicative} &
% \rotatehead[2.5cm]{\textbf{Lexical stem}} &
% \rotatehead[2.5cm]{Tense/aspect/\newline mood (TAM)} &
% \rotatehead[2.5cm]{Incorporated\newline adverbial} &
% \rotatehead[2.5cm]{Number\newline (\textsc{subj} or \textsc{obj})} &
% \rotatehead[2.5cm]{Incorporated\newline \mbox{adverbial}} \\
% % \lspbottomrule
% \end{tabularx}
% \caption{Murrinhpatha basic verbal template}
% \label{tab:mur:1}
% \end{table}

This chapter aims to provide a detailed overview of negation and negation constructions in Murrinhpatha following the questionnaire  provided by the editors \parencitetv{chapters/questionnaire}. Some features of negation constructions in Murrinhpatha that have not been discussed in the literature before include the polarity reversal that arises from the use of irrealis mood in both negative declarative and negative deontic constructions, which results in a single clause being ambiguous between meanings with opposite truth conditions (e.g. ‘didn’t do’ vs ‘shouldn’t do (but did)’) (\sectref{sec:mur:2.1}); and the existence of negative classifier predicates to express negative existence and negative possession (\sectref{sec:mur:2.3.3}).

The data for this chapter is taken primarily from my fieldwork on the language since 2005; a collaborative Murrinhpatha corpus \citep{MansfieldEtAl2019} with contributions from a number of researchers; a corpus shared with me by Joe Blythe, and published sources on Murrinhpatha. The source of each example is provided – those from my own fieldwork recordings are identified with RN in the source code; those from Joe Blythe’s corpus are marked with JB. All examples provided here are from spontaneous natural speech unless marked as elicited. 

%\hc{for M\&L: They use many smallcaps as glosses, their explanation (at least for LEST, HITHER, MAYBE) is "these ARE glosses - they should not be interpreted in the same way as the regular English word. In many cases the meaning or function or behaviour is different" Also, there is a short explanation at the very end, just after the list of glosses. Here's a list of all of them (thanks to Karoliina): clf:place, clf:human, clf:thing, clf:water, clf:time, yet 7b, lest 21, hither 32, thing 54, maybe 67, poke(19) 1, slash(23) 2, hands(8) 3, go(6) 4, be(4) 5, stand(3) 5, shove(29) 8b, sit(1) 10, snatch(9) 15, travel(7) 22, see(13) 56, have(22) 65), listen(16) 70, sit(1) 75, other 80, but 80, say/do(34) 93, bash(14) 94, what.from 95, hey 95, who 97, where 98,  what 100, again 100, do(34, 105, still 109, foot(7) 110, walk 112} 
%\lv{As far as I understand, we a dealing with a fine-grained classification system. Classifiers are indeed not regular words, so we should leave her glossing as is.}
\section{Clausal negation} \label{sec:mur:2}
\subsection{Standard negation} \label{sec:mur:2.1}

Standard negation in Murrinhpatha is expressed with a negative particle before the verb. The verb must also be inflected with the irrealis mood (see below for further discussion). There are a few different negative particles in the corpus, which appear to vary in use amongst speakers, as described below. The most common negator is \textit{mere}, as in \REF{ex:mur:1b} and \REF{ex:mur:2b}.\footnote{Although analysed as a particle, in fast speech \textit{mere} often is reduced to a proclitic \textit{me=}.}  When speakers are asked to provide the negative equivalent of a declarative sentence (as in \REF{ex:mur:2a}), this is the negator that is typically used (as in \REF{ex:mur:2b}).

\ea\label{ex:mur:1}
    \ea\label{ex:mur:1a}
        \gll  berengunh ngam-rilil\\ 
              already \textsc{1sg.sbj.poke(19).nfut-}write \\
        \glt ‘I wrote it down already.’ (MP-20100917-RN01)
    \ex\label{ex:mur:1b}
        \gll  \textbf{mere} nge-rilil-dha\\
              \textsc{neg} \textsc{1sg.sbj.poke(19).pst:irr}{}-write-\textsc{pst}\\
        \glt ‘I didn't write it down.’ (MP-20100917-RN01)
    \z
\z 

\ea\label{ex:mur:2}
    \ea\label{ex:mur:2a}
        \gll  pan-ngi-bat\\ 
              \textsc{3sg.sbj.slash(23).nfut-1sgO}{}-hit\\
        \glt ‘He hit me.’ (elicited, RN-20090930-002:079)
    \ex\label{ex:mur:2b}
        \gll  \textbf{mere} puy-ngi-bat-dha\\
               \textsc{neg} \textsc{3sg.sbj.slash(23).pst:irr-1sg.obj}{}-hit-\textsc{pst}\\
        \glt ‘He didn't hit me.’ (elicited, MP-20100917-RN01)
    \z
\z 

Other negative particles are \textit{marda} \REF{ex:mur:3} and \textit{manangka} \REF{ex:mur:4}. The choice of negative particle appears to be motivated by sociolinguistic variation, rather than by semantic function: \textit{marda} is reported by \citet[215]{Walsh1976} to be the main clausal negator at the time of his fieldwork, but it is now used by only the oldest speakers; \textit{manangka} is more usually a nominal negator (see \sectref{sec:mur:2.3.1}) but is being extended to clausal negation as well, in free variation with \textit{mere}. Example \REF{ex:mur:4} is from an elicitation session where the speaker was asked how to say ‘he didn’t come’. They replied initially with \REF{ex:mur:4a} and then followed with \REF{ex:mur:4b}, describing the two options as being equivalent. An example of this use in spontaneous speech is given in \REF{ex:mur:5}.

\ea\label{ex:mur:3}
    \gll \textbf{marda} the ma-bath\\
         \textsc{neg} ear \textsc{1sg.sbj.hands(8).fut:irr}{}-have\\
    \glt ‘I don’t know.’ (\citealt{StreetStreet1989} \#marda)
\z 

\ea\label{ex:mur:4} 
    \ea\label{ex:mur:4a}
        \gll  \textbf{mere} wurri-dha\\ 
              \textsc{neg} \textsc{3sg.sbj.go(6).pst:irr-pst}\\
        \glt ‘He didn’t come.’ (elicited, MP-20180718-RN01)
    \ex\label{ex:mur:4b}
        \gll  \textbf{manangka} wurri-dha \\
              \textsc{neg} \textsc{3sg.sbj.go(6).pst:irr-pst}\\
        \glt ‘He didn’t come.’ (elicited, MP-20180718-RN01)
    \z
\z 

\ea\label{ex:mur:5} 
    \gll ba \textbf{manangka} parne-kut-dha-neme, pibim-wirra-puth-neme\\
         oh \textsc{neg} \textsc{3du.sbj.be(4).pst:irr}{}-collect-\textsc{pst-pc.m} \textsc{3du.sbj.stand(3).nfut-3pl.obl}{}-throw\_away-\textsc{pc.m}\\
    \glt ‘Oh they didn’t keep (the gun), they threw it away from them.’ (\citealt{MansfieldEtAl2019}, 2012-06-02\_P-D-W)
\z 

The negative particle almost always immediately precedes the verb, as in the above examples. However, temporal affixes \REF{ex:mur:6} and adverbials \REF{ex:mur:7} come between \textit{mere} and the verb. There are also a small number of examples in which noun class markers \REF{ex:mur:8} and pronouns \REF{ex:mur:9} come between the negative particle and the verb, but as shown by the (b) examples, this is not obligatory and these elements are regularly found to the left of the negative particle in other examples. There are no examples in the corpus of a complex NP or anything more than a single word coming between the two. 

\ea\label{ex:mur:6}
    \gll mere-warda ngena-nham-dha ngay-yu\\
         \textsc{neg-temp} \textsc{1sg.sbj.poke:rr(21).pst:irr}{}-fear-\textsc{pst} \textsc{1sg-dm}\\
    \glt ‘I wasn't frightened anymore.’ (MP-20100916-RN01)
\z

\ea\label{ex:mur:7}
    \ea\label{ex:mur:7a}
        \gll  murrinh mere thathpirr nge-riwatha-dha\\ 
              \textsc{clf:lang} \textsc{neg} correctly \textsc{1sg.sbj.poke(19).pst:irr}{}-do.properly-\textsc{pst}\\
        \glt ‘I didn't really say it the right way.’ (JB20050703JB02.txt)
    \ex\label{ex:mur:7b}
        \gll  mere dangatha ngurri-dha Perrederr-yu\\
              \textsc{neg} \textsc{yet} \textsc{1sg.sbj.go(6).pst:irr-pst} \textsc{P}errederr\textsc{{}-dm} \\
        \glt ‘I’ve not yet been to Perrederr.’ (MP-20180718-RN01)
    \z
\z 

\ea\label{ex:mur:8}
    \ea\label{ex:mur:8a}
        \gll  ngay-ka mere murrinh ma-mpa-bath\\ 
              \textsc{1sg-top} \textsc{neg} \textsc{clf:lang} \textsc{1sg.sbj.hands(8).fut:irr-2sg.obl}{}-have\\
        \glt ‘I don’t believe you.’ (JB20070728JBvid01c.txt)
    \ex\label{ex:mur:8b}
        \gll  murrinh mere puddi-ngarra-riyith-dha \\
              \textsc{clf:lang} \textsc{neg} \textsc{3pl.sbj.shove(29).pst:irr-1pl.excl.obl}{}-explain-\textsc{pst}\\
        \glt ‘They didn't tell us the story.’ (MP-20120724-RN01)
\z
\z 

\ea\label{ex:mur:9}
    \ea\label{ex:mur:9a}
        \gll  awu, mere ngay nge-ngurru-dha\\ 
              no \textsc{neg} \textsc{1sg} \textsc{1sg.sbj.poke(19).pst:irr}{}-provoke-\textsc{pst}\\
        \glt ‘No, I did not threaten him.’ (\citealt{FordMcCormack2011}: 8)
    \ex\label{ex:mur:9b}
        \gll  awu, ngay-ka mere nge-rdarrup-dha\\
              no \textsc{1sg-top} \textsc{neg} \textsc{1sg.sbj.poke(19).pst:irr-}cover-\textsc{pst}\\
        \glt ‘No, I didn't cover it up.’ (\citealt{FordMcCormack2011}: 13)
\z
\z  

In \citegen{Miestamo2005} terms, these negative clauses in Murrinhpatha show constructional symmetry since the only structural distinction between the negative clause and the corresponding affirmative construction is the presence of the negative marker. However, the verb in negative clauses must always be in irrealis mood – realis tense categories are excluded from being in the scope of the negative particle – and thus we have paradigmatic asymmetry: realis affirmative clauses are negated with the irrealis mood, as can be seen in \REF{ex:mur:1} above. In \REF{ex:mur:1a} the classifier stem is in the (realis) non-future form, whereas in the negative equivalent in \REF{ex:mur:1b} the classifier stem is in the past irrealis form (see \citet{NordlingerCaudal2012} for detailed discussion of the tense/aspect/mood system in Murrinhpatha). Thus, Murrinhpatha belongs to the mixed TypeSymAsy category of \citet{Miestamo2013-wals114} and the paradigmatic asymmetry is of the A/NonReal subtype (\citealt{Miestamo2005}; \citealt{Miestamo2013-wals114}), similar to many other northern Australian languages \citep{Miestamo2007}.

In order to mark imperfective aspect, one of a small set of classifier stems is encliticised to the end of the main verb (\citealt{NordlingerCaudal2012}), as shown in \REF{ex:mur:10}. The encliticised classifier must also appear in the irrealis form in negative constructions, showing it to be within the scope of the negative operator. This is shown in the contrast between \REF{ex:mur:10a} and its negative counterpart \REF{ex:mur:10b}: in \REF{ex:mur:10b} both the classifier in the main verb (\textit{ka-}) and the encliticised classifier \textit{=ki} must appear in the future irrealis form.

\ea\label{ex:mur:10}
    \ea\label{ex:mur:10a}
        \gll  dam-rilil=dim\\ 
              \textsc{3sg.sbj.poke(19).nfut-}write\textsc{=3sg.sbj.sit(1).nfut}\\
        \glt ‘He’s writing it down.’ (elicited, MP-20100917-RN01)
    \ex\label{ex:mur:10b}
        \gll  mere ka-rilil=ki\\
              \textsc{neg} \textsc{3sg.sbj.poke(19).fut:irr-}write\textsc{=3sg.sbj.sit(1).fut:irr}\\
        \glt ‘He’s not writing it down.’ (elicited, MP-20100917-RN01)
    \z
\z 

The tense/aspect/mood system of Murrinhpatha makes a different set of tense distinctions in the realis mood than in the irrealis mood (\citealt{NordlingerCaudal2012}). Since negative clauses are restricted to irrealis mood inflections only, negative clauses make different tense distinctions to their affirmative counterparts. This results in a derived asymmetry \citep{Miestamo2013-wals114} as the asymmetry arises from the properties of the tense/aspect/mood system so cannot be attributed to the negative construction directly, but it is nonetheless a characteristic of negative clauses. The traditional pattern is shown in \tabref{tab:mur:2}, and exemplified below.

% \hc{@LSP please check this table}
\begin{table}
\small
\fittable{
    \begin{tabular}{|p{17mm}p{13mm}|p{14mm}p{13mm}cccc|}
        \hline
        & & \multicolumn{6}{c|}{TAM category expressed\strut} \\
        \cline{3-8}
        & & \multicolumn{4}{c|}{realis\strut}  & \multicolumn{2}{c|}{irrealis}   \\
        \cline{3-8}
        & & \multicolumn{1}{l|}{Past\strut } & \multicolumn{1}{l|}{Past } & \multicolumn{1}{c|}{Present} & \multicolumn{1}{c|}{Future} & \multicolumn{1}{c|}{Future} & Past \\
        & & \multicolumn{1}{l|}{imperfective} & \multicolumn{1}{l|}{perfective} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} & \multicolumn{1}{l|}{} &  \\
        \hline
        \multirow{5}{17mm}{{TAM  inflection {on classifier} stem}\newline~} & Positive polarity & \multicolumn{1}{c}{\textsc{pst:ipfv}} & \multicolumn{2}{|c|}{\textsc{nfut}} & \multicolumn{1}{c|}{\textsc{fut}} & \multicolumn{1}{c|}{\textsc{fut:irr}} & \textsc{pst:irr} \\
        \cline{2-8}
        & Negative polarity & \multicolumn{2}{c|}{\textsc{pst:irr}} & \multicolumn{3}{c|}{\textsc{fut:irr}} & \textsc{pst:irr} \\
        \hline
    \end{tabular}
    }
    \caption{Murrinhpatha TAM inflection in positive and negative clauses \citep{NordlingerCaudal2012}}
    \label{tab:mur:2}
\end{table}

For example, in affirmative clauses with telic predicates, the non-future tense is used for present and past events (\ref{ex:mur:11a}, \ref{ex:mur:11b}), and the future tense for future events \REF{ex:mur:11c}:

\ea\label{ex:mur:11}
    \ea\label{ex:mur:11a}
        \gll  dam-rilil=dim\\ 
              \textsc{3sg.sbj.poke(19).nfut-}write\textsc{=3sg.sbj.sit(1).nfut}\\
        \glt ‘He’s writing it down.’ (constructed, for illustration)
    \ex\label{ex:mur:11b}
        \gll  dam-rilil\\
              \textsc{3sg.sbj.poke(19).nfut-}write\\
        \glt ‘He wrote it down.’ (constructed, for illustration)
    \ex\label{ex:mur:11c}
        \gll  pa{}-rilil-nu\\
                 \textsc{3sg.sbj.poke(19).fut}{}-write-\textsc{fut}\\
        \glt ‘He will write it down.’ (constructed, for illustration)
    \z
\z 
 
In the irrealis paradigm, however, present and future tense are expressed with the future irrealis while past tense is expressed with the past irrealis. Thus, while \REF{ex:mur:11a} and \REF{ex:mur:11b} involve the same realis tense form of the classifier stem, their negative counterparts use distinct irrealis forms (\ref{ex:mur:12a}, \ref{ex:mur:12b}). Example \REF{ex:mur:11c}, on the other hand, which is marked with a distinct form in the affirmative, uses the same irrealis classifier stem form as the present tense \REF{ex:mur:12a} in the negative \REF{ex:mur:12c}.

\ea\label{ex:mur:12}
    \ea\label{ex:mur:12a}
        \gll  mere ka{}-rilil=ki\\ 
              \textsc{neg} \textsc{3sg.sbj.poke(19).fut:irr-}write\textsc{=3sg.sbj.sit(1).fut:irr}\\
        \glt ‘He’s not writing it down.’ (constructed, for illustration)
    \ex\label{ex:mur:12b}
        \gll  mere de-rilil-dha\\
              \textsc{neg} \textsc{3sg.sbj.poke(19).pst:irr}{}-write-\textsc{pst} \\
        \glt ‘He didn’t write it down.’ (constructed, for illustration)
    \ex\label{ex:mur:12c}
        \gll  mere ka-rilil-nu\\
              \textsc{neg} \textsc{3sg.sbj.poke(19).fut:irr}{}-write-\textsc{fut}\\
        \glt  ‘He won’t write it down.’ (constructed, for illustration)
\z
\z 

\tabref{tab:mur:2} above presents the traditional system, but in present day Murrinhpatha the distinction between the future irrealis and the past irrealis is being reanalysed, with the future irrealis forms coming to be used as general irrealis forms in all tenses \citep{Mansfield2014}. This means that there is quite a bit of variation in terms of how the past irrealis is expressed both across the community and in the speech of individuals. For example, one speaker in an elicitation session offered the two options in \REF{ex:mur:13} as alternatives meaning ‘he didn’t come’:

\ea\label{ex:mur:13} 
    \ea\label{ex:mur:13a}
        \gll mere kurru-dha\\
             \textsc{neg} \textsc{3sg.sbj.go(6).fut:irr-pst} \\
        \glt ‘He didn’t come.’
    \ex\label{ex:mur:13b}
        \gll mere wurri-dha\\
              \textsc{neg} \textsc{3sg.sbj.go(6).pst:irr-pst} \\
        \glt ‘He didn’t come.’ (elicited, MP-20180718-RN01)
    \z
\z 

The same speaker spontaneously provided the two sequential lines in \REF{ex:mur:14} in a narrative – in the first she uses the future irrealis form in a past context, and in the second she switches to the past irrealis:

\ea\label{ex:mur:14} 
    \ea\label{ex:mur:14a}
        \gll murrinh mere \textbf{kuddu}{}-ngarru-riyith-dha-ngime  \\ 
             \textsc{clf:lang} \textsc{neg} \textsc{3pl.sbj.shove(29).fut:irr}-\textsc{1du.excl.obl-}explain\textsc{{}-pst-pc.f}  \\
        \glt ‘They didn't explain that story to us.’
    \ex\label{ex:mur:14b}
        \gll murrinh mere \textbf{puddi}{}-ngarra-riyith-dha  murrinh nhini-yu \\
              \textsc{clf:lang} \textsc{neg} \textsc{3pl.sbj.shove(29).pst:irr}-\textsc{1pl.excl.obl-}explain\textsc{{}-pst} \textsc{clf:lang} this-\textsc{dm} \\
        \glt ‘They didn't explain that story to us.’ (MP-20120724-RN01)
    \z
\z 

Nonetheless, despite the variation in the irrealis paradigm, the overall generalization that negative polarity clauses require irrealis verb forms holds across all Murrinhpatha speakers. In the examples in the rest of this paper where the speaker is clearly using the future irrealis form in past irrealis contexts, I gloss it simply as \textsc{irr.}

In affirmative constructions, the irrealis categories are used to encode counterfactuals, conditional constructions and deontic constructions as in the following:

\ea\label{ex:mur:15}
    \gll ku beg me-art-dha-ka\\
         \textsc{clf:anim} bag \textsc{1sg.sbj.snatch(9).pst:irr}{}-get-\textsc{pst-top}\\
    \glt ‘I should have brought my bag.’ (JB2004-09-12JB04.txt)
\z 

\ea\label{ex:mur:16}
    \gll kuma-nhi-manpi mardinhpuy\\
         \textsc{3pl.sbj.hands(8).fut:irr-2sg.obj}{}-help young.girl\\
    \glt ‘The young girls should help you.’ (RN20090930-002:079)
\z 

An interesting consequence of the fact that the realis/irrealis distinction is collapsed in negative contexts is that we find ambiguity between standard negative constructions and negative deontic constructions (the negative counterparts of constructions such as \REF{ex:mur:15} and \REF{ex:mur:16}) (\citealt{NordlingerCaudal2012}), since all are marked with the irrealis categories. This is shown in \REF{ex:mur:17} and \REF{ex:mur:18}.

\newpage
\ea\label{ex:mur:17} 
    \gll  marda the-na-mut-dha palngun\\
         \textsc{neg} \textsc{2sg.sbj.poke(19).pst:irr-3sg.m.obl-}give\textsc{{}-pst} woman\\
    \ea\label{ex:mur:17a}
        \glt ‘You didn’t give him that woman.’
    \ex\label{ex:mur:17b}
        \glt ‘You shouldn't have given him that woman.’ (JB2007-08-27JB04c.txt)
    \z
\z 

\ea\label{ex:mur:18}
    \gll nukunu-ka mere kani-ngerren murrinh wiye\\
         \textsc{3sg.m-top} \textsc{neg} \textsc{3sg.sbj.be(4).fut:irr-}speak \textsc{clf:lang} bad\\   
    \ea\label{ex:mur:18a}
        \glt ‘He doesn’t say rude words.’
    \ex\label{ex:mur:18b}
        \glt ‘He shouldn’t say rude words.’ (JB2004\_08\_08JB03b2.txt)
    \z
\z 

This ambiguity results in a type of polarity reversal whereby on one meaning, e.g. \REF{ex:mur:18a}, the interpretation is that the event did not take place (i.e. rude words were not spoken); yet on the other meaning, \REF{ex:mur:18b}, the interpretation is that the event \textit{did} take place (i.e. rude words \textit{were} spoken, but the speaker feels they shouldn’t have been).\footnote{As far as I am aware, this has not been discussed in the literature for other languages, although the same ambiguity is found in other northern Australian languages, such as Bininj Kunwok, and can be particularly difficult when interpreting in legal contexts (Murray Garde, pers. comm.).} 

\subsection{Negation in non-declaratives} \label{sec:mur:2.2}

Imperatives are negated in the same way as second person present/future tense declarative clauses, and therefore belong to Type 1 in the typology of \citet{AuweraEtAl2013}. Examples are provided in \REF{ex:mur:19} and \REF{ex:mur:20}.\footnote{Note that there is no distinction in the verbal paradigms between future realis and future irrealis for second person. However, given that realis tenses are not permitted in negative declarative clauses, I assume the same is true in negative imperatives, and so treat the verb in these examples as being inflected with the future irrealis.}

\ea\label{ex:mur:19}
    \gll  mere thu-ngi-bat!\\
         \textsc{neg} \textsc{2sg.sbj.slash(23).fut:irr-1sg.obj}{}-hit\\
    \glt ‘Don't hit me!’ (MP-20100917-RN01)
\z 

\ea\label{ex:mur:20}
    \gll manangka na{}-rithuk mi pana-yu!\\
         \textsc{neg} \textsc{2sg.sbj.hands(8).fut:irr}{}-pester \textsc{clf:veg} \textsc{rcgn-dm}\\
    \glt ‘Don’t pester him for that food!’ (RN20090606-002:077)
\z 
         
A stronger deontic meaning, including prohibition, is expressed by adding the \textit{{}-nukun} ‘\textsc{lest}’ suffix (see also \sectref{sec:mur:2.6.2}). The \textit{{}-nukun} suffix can be attached either to the verb \REF{ex:mur:21} or to the negative particle \textit{mere} \REF{ex:mur:22} and adds the implication that something bad or undesirable might happen if the order is not followed.


\ea\label{ex:mur:21}
    \gll mere thu{}-ngi-rerda-nukun!\\
          \textsc{neg} \textsc{2sg.sbj.slash(23).fut:irr-1sg.obj}{}-blame-\textsc{lest}\\
    \glt ‘Don’t blame me!’ (\citealt{FordMcCormack2011}: 9)
\z 
 
\ea\label{ex:mur:22}
    \gll da pana-wangu-ka mere{}-nukun nunna{}-ngime \\
         \textsc{clf:place} \textsc{rcgn}-away-\textsc{top} \textsc{neg}-\textsc{lest} \textsc{2du.sbj.travel(7)}\textsc{.fut:irr-pc.f} \\
    \glt ‘Don’t go near that place.’ (MP-20120724-RN01)
\z 

Example \REF{ex:mur:22} was then repeated by the speaker as \REF{ex:mur:23}, showing the two potential positions for \textit{{}-nukun}:

\ea\label{ex:mur:23}
    \gll mere nunna-nukun-ngime \\
         \textsc{neg} \textsc{2du.sbj.travel(7).fut:irr-lest-pc.f}\\
    \glt ‘Don’t go (there).’ (MP-20120724-RN01)
\z 

A special interjection \textit{pirda} ‘don’t, stop it, leave it’ is frequently used in place of, or in conjunction with, negative imperative constructions, especially when admonishing children (\ref{ex:mur:24}, \ref{ex:mur:25}). 

\ea\label{ex:mur:24}
    \gll pirda! \\
         stop\\
    \glt ‘Don’t (touch it), leave it alone!’ (\citealt{MansfieldEtAl2019}, 2015-07-338.1)
\z 

\ea\label{ex:mur:25} 
    \gll pirda, mere numa-rdurt-nukun dedi-yu\\
          stop \textsc{neg} \textsc{2du.sbj.hands(8).fut:irr}{}-wake\_up-\textsc{lest} father-\textsc{dm} \\
    \glt ‘Stop it! Don’t wake up your father.’ (\citealt{MansfieldEtAl2019}, 2015-07-69.1)
\z 
         
There are no dedicated hortative constructions in the language distinct from the imperative constructions illustrated above. The corpus also contains no examples of negative interrogative clauses. 

\subsection{Negation in stative predications} \label{sec:mur:2.3}

There are two different constructions involved in negating stative (non-verbal) predications. Assuming the categories of \citet[111ff]{Payne1997}, equational, proper inclusion, and attributive constructions are negated similarly to verbal clauses, with the addition of the negative particle \textit{mere} or \textit{manangka}. Existential, locative and possessive non-verbal clauses use a unique construction involving negated noun classifier predicates (see \sectref{sec:mur:2.3.3}).

\subsubsection{Equative and proper inclusion clauses} \label{sec:mur:2.3.1}

These clause types are expressed with non-verbal predicates and are negated by placing a negative particle immediately before the predicate. \textit{Mere} is used before pronouns (\ref{ex:mur:26}, \ref{ex:mur:27}), and \textit{manangka} is usually used with nouns (\ref{ex:mur:28}, \ref{ex:mur:29}) (although for some speakers, \textit{mere} is possible with nouns too as in \REF{ex:mur:30}).

\ea\label{ex:mur:26}
    \ea\label{ex:mur:26a}
        \gll  mi ngay-ka!\\ 
              \textsc{clf:veg} \textsc{1sg-top}\\
        \glt ‘That’s my food!’ (elicited, MP-20100917-RN01)
    \ex\label{ex:mur:26b}
        \gll mi \textbf{mere} ngay pana-yu! \\
              \textsc{clf:veg} \textsc{neg} \textsc{1sg} \textsc{rcgn-dm}\\
        \glt ‘That’s not my food!’ (elicited, MP-20100917-RN01)
\z
\z 

\ea\label{ex:mur:27}
    \gll \textbf{mere} ngay\\
         \textsc{neg} \textsc{1sg}\\
    \glt ‘It wasn’t me.’ (\citealt{FordMcCormack2011}: 3)
\z 

\ea\label{ex:mur:28}
    \gll \textbf{manangka} ku were ngay \\
         \textsc{neg} \textsc{clf:anim} dog \textsc{1sg}\\
    \glt ‘That’s not my dog.’ (MP-20180718-RN01)
\z

\ea\label{ex:mur:29} 
    \gll kardu \textbf{manangka} {garbage} {bag} ngay-yu!\\ 
         \textsc{clf:human} \textsc{neg} garbage bag \textsc{1sg-dm}\\
    \glt ‘I’m not a garbage bag!’ (RN-20180725-004:039)
\z 

Examples involving proper inclusion include (\ref{ex:mur:30}--\ref{ex:mur:32}): 

\ea\label{ex:mur:30}
    \gll kardu \textbf{mere} ku teacher pangu-yu\\
          \textsc{clf:human} \textsc{neg} \textsc{clf:anim} teacher there-\textsc{dm}\\
    \glt ‘That man’s not a teacher.’ (elicited, MP-20180718-RN01) 
\z 

\ea\label{ex:mur:31}
    \gll  \textbf{manangka} murrinh\\
         \textsc{neg} \textsc{clf:lang}\\
    \glt ‘That’s not a word.’ (RN-20180724-004:038)
\z 

\ea\label{ex:mur:32} 
    \gll telephone-gathu, nanthi \textbf{manangka} camera pana-yu!\\
         telephone-\textsc{hither} \textsc{clf:thing} \textsc{neg} camera \textsc{rcgn-dm}\\
    \glt ‘Give me the telephone – that’s not a camera!’ (RN20120726-003:037)
\z

\subsubsection{Attributive clauses} \label{sec:mur:2.3.2}

Like equative clauses, these use the standard clausal negators before the attributive predicate, as in (\ref{ex:mur:33}--\ref{ex:mur:35}):

\ea\label{ex:mur:33}
    \gll ku were \textbf{mere} ngala pangu-yu\\
         \textsc{clf:anim} dog \textsc{neg} big there-\textsc{dm}\\
    \glt ‘That dog is not big.’ (elicited, MP-20180718-RN01)
\z 

\ea\label{ex:mur:34}
    \gll \textbf{manangka} dhepdhep-wun-dhay \\
         \textsc{neg} food-\textsc{3pl.obj}{}-mouth\\
    \glt  ‘They are never eating.’ (MP-20180718-RN01)
\z 

\ea\label{ex:mur:35}
    \gll kuguk dangatha \textbf{mere} patha-ngi\\
         wait \textsc{yet} \textsc{neg} good-\textsc{1sg.obj}\\
    \glt ‘Wait a minute, I don’t feel well.’ (JB20040808JB03b1.txt)
\z

\subsubsection{Negative classifier predicates} \label{sec:mur:2.3.3}

The negation of all other stative predications, namely existential, locative and possessive clauses, is expressed using a distinct construction type which involves negative classifier predicates. Negative classifier predicates have not been previously reported in the typological literature on either noun classifiers or negation (e.g. \citealt{Aikhenvald2003}; \citealt{Miestamo2007}; \citealt{Eriksen2011}; \citealt{Veselinova2013neg.exist}).\footnote{Sentani (Trans New Guinea) has four classificatory negators used in existential constructions that appear similar to those described here in some respects: \textit{olo} ‘no person’; \textit{an} ‘no bird, animal’; \textit{ban} ‘no animate’; \textit{u} ‘no inanimate’ (\citealt{Hartzler1994}, and thanks to Ljuba Veselinova for alerting me to this data). However, Sentani does not appear to have a nominal classifier system from which these negators derive and it’s not clear from the description whether they are predicates as we find in Murrinhpatha; they may be more like the negative indefinite pronouns ‘nobody, nothing’ in English, although it is not possible to determine on the data available to me.} While they have been mentioned in previous descriptive work on Murrinhpatha (e.g. \citealt{Walsh1976}; \citealt{Blythe2009}), and briefly in grammatical descriptions of other Daly languages such as Marrithiyel \citep{Green1989} and Ngan’gityemerri \citep{Reid2011}, they have not been previously discussed in detail. 

Murrinhpatha has a set of 10 noun classifiers \citep{Walsh1993}, which co-occur with nouns in an NP as in \REF{ex:mur:36} and \REF{ex:mur:37}, or can occur alone as general nouns \REF{ex:mur:38}. These noun classifiers occur only once in an NP, and do not participate in agreement. 

\ea\label{ex:mur:36} 
    \gll \textbf{kardu} \textbf{wakal} ngurdu-thuk-nu ngarra da-warda\\
         \textsc{clf:human} child \textsc{1sg.sbj.shove}(29).\textsc{fut}{}-send-\textsc{fut} \textsc{loc} \textsc{clf:place-temp} \\
    \glt  ‘I’ll send the little boy home now.’ (RN-20070608-002:037)
\z 

\ea\label{ex:mur:37}
    \gll \textbf{ku} \textbf{tumtum} mam-lerrkperrk\\
         \textsc{clf:anim} egg \textsc{1sg.sbj.hands(8).nfut}{}-destroy\\
    \glt ‘I crushed the egg in my hands.’ (RN-20070608-002:039)
\z 

\ea\label{ex:mur:38}
    \gll \textbf{kardu} mere kurru-lili\\
         \textsc{clf:human} \textsc{neg} \textsc{3sg.sbj.go(6).fut:irr} {}-walk\\
    \glt ‘That person can’t walk.’ (RN-20180724-004:038)
\z 

As is common in such systems, nouns may combine with different classifiers to reflect different semantic construals. The noun \textit{menek} for example combines with the general inanimate (\textsc{thing)} classifier \textit{nanthi} to describe the ironwood tree (\textit{nanthi menek)}, and with the \textsc{spear} classifier \textit{thamul} to refer to a particular type of spear made from the ironwood tree (\textit{thamul menek}). The full set of nominal classifiers is in \REF{ex:mur:39}:

\TabPositions{.15\textwidth,.15\textwidth,.4\textwidth} 
\ea\label{ex:mur:list1}
kardu  \tab ‘{human}’ \tab e.g. kardu pule ‘old man’ \\
ku  \tab ‘{animate}’ \tab e.g. ku lawarnka ‘wallaby’ \\
kura  \tab ‘{water}’ \tab e.g. kura yelyel ‘rain’ \\
thamul  \tab ‘{spear}’ \tab e.g. thamul nguni ‘short light spear’ \\
thu  \tab ‘{weapon}’ \tab e.g. thu paku ‘large club’ \\
thungku  \tab ‘{fire}’ \tab e.g. thungku thay ‘firewood’ \\
mi  \tab ‘{vegetable} {food}’ \tab e.g. mi lawam ‘flour \\
da  \tab ‘{place or time}’ \tab e.g. da pemanhay ‘sandhill’ \\
murrinh  \tab ‘{language}’ \tab e.g. murrinh English ‘English’ \\
nanthi  \tab ‘residue, thing’ \tab e.g. nanthi thelput ‘house’ \\
\z 

Each of these noun classifiers has a special negative form used in the expression of existential, locative and possessive predications. For most of the classes this is formed by prefixing \textit{ma-} to the noun classifier, but for two classes (\textit{da} and \textit{murrinh}) this involves combining the noun classifier with the regular negative particle \textit{manangka}, see \REF{ex:mur:list2}:

\TabPositions{.33\textwidth,.66\textwidth}
\ea\label{ex:mur:list2} 
makardu \tab ‘\textsc{neg:human’} \\
maku \tab ‘\textsc{neg:anim’} \\
makura \tab ‘\textsc{neg:water’} \\
mathamul \tab ‘\textsc{neg:spear’} \\
mathu \tab ‘\textsc{neg:weapon’} \\
mathungku \tab ‘\textsc{neg:fire’} \\
mami \tab ‘\textsc{neg:veg’} \\
manangka da \tab ‘\textsc{neg:place/time’} \\
manangka murrinh \tab ‘\textsc{neg:lang’} \\
mananthi \tab ‘\textsc{neg:thing’} \\
\z

Some examples of these negative classifier predicates are given in (\ref{ex:mur:39}--\ref{ex:mur:42}). Although they are formally based on the nominal classifiers, these negative forms are in fact predicative, not referential, and cannot function as nominal classifiers themselves. In contexts where the presence of a particular noun is being negated, the regular noun classifier must also be used with the noun, as shown in \REF{ex:mur:40} and \REF{ex:mur:42}.

\ea\label{ex:mur:39}
    \gll marra-ka \textbf{makura}{}-warda \\
         now-\textsc{top} \textsc{neg:water}{}-\textsc{temp}\\
    \glt ‘Now there’s no rain.’ \citep[46]{Street1987}
\z 
         
\ea\label{ex:mur:40}
    \gll ngay-ka \textbf{mananthi} \textbf{nanthi} marluk-yu \\
         \textsc{1sg-top} \textsc{neg:thing} \textsc{clf:thing} didgeridoo-\textsc{dm}\\
    \glt ‘I don’t have a didgeridoo.’ \citep[45]{Street1987}
\z 

\ea\label{ex:mur:41}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:mur:41Q}
        \gll  ngarra John-yu?\\ 
               where John-\textsc{dm}\\ 
        \glt ‘Where’s John?’
    \exi{A: }\label{ex:mur:41A}
        \gll  \textbf{makardu}\\ 
                \textsc{neg:human}\\
        \glt ‘(He’s) not here.’ (lit. ‘There is no person.’) (RN, overheard) 
    \end{xlist}
\z 

\ea\label{ex:mur:42}
    \gll  \textbf{ku} were-ka \textbf{maku} ngarra da-yu\\
         \textsc{clf:anim} dog-\textsc{top} \textsc{neg:anim} \textsc{loc} \textsc{clf:place-dm}\\
    \glt ‘The dog’s not at home.’ (elicited, RN-20180724-004:038)
\z 

Example \REF{ex:mur:43} provides further evidence for the predicative status of these negative classifiers since it is not possible for them to appear in regular argument positions with a verbal predicate \REF{ex:mur:43b}, unlike noun classifiers \REF{ex:mur:43c}. 

\ea\label{ex:mur:43}
    \ea[]{\label{ex:mur:43a}
        \gll  makardu{}-warda\\ 
              \textsc{neg:human-temp}\\
        \glt ‘No-one is there.’ (MP-20180718-RN01)}
    \ex[*]{\label{ex:mur:43b}
        \gll  makardu kani/kem\\
              \textsc{neg:human} \textsc{3sg.sbj.be(4).fut:irr/}\textsc{3sg.sbj.be(4).ex}\\
        \glt (both options offered to speaker, but rejected and replaced with \REF{ex:mur:43a}, MP-20180718-RN01)}
    \ex[]{\label{ex:mur:43c}
        \gll kardu kem\\
            \textsc{clf:human} \textsc{3sg.sbj.be(4).ex}\\
        \glt ‘Someone is there.’ (constructed, for illustration)}
    \z
\z 

Negative classifier predicates form a distinct negation construction that is used to express negative locative, existential and possessive predications. In this respect Murrinhpatha is consistent with the claims of \citet[112]{Veselinova2013neg.exist} who found that 66\% of languages in her sample had a distinct construction used in negative existentials with no overlap with the expression of standard negation, and that the most common functions of this special negative construction are negation of existence, possession and location (ibid.: 118).

Examples of negated locative clauses are shown in (\ref{ex:mur:44}--\ref{ex:mur:47}). As is clear from these examples, the negative classifier predicates can be used for both specific and general reference.

\ea\label{ex:mur:44} 
    \gll  Barbara da \textbf{makardu}{}-warda\\
         Barbara \textsc{clf:place} \textsc{neg:human-temp}\\
    \glt ‘Barbara’s not here anymore.’ (When arriving at the dongas and realising Barbara had gone back to Melbourne) (RN20090606-002:075)  
\z 

\ea\label{ex:mur:45}
    \gll kardu wakal \textbf{makardu}{}-ya ngarra truck-yu\\
         \textsc{clf:human} small \textsc{neg:human-dm} \textsc{loc} truck-\textsc{dm}\\
    \glt  ‘The child is not in the truck.’ (MP-20180718-RN01)
\z 

\ea\label{ex:mur:46}
    \gll ngathan ngay-ka Daly River wardarra, kardu \textbf{makardu}{}-warda Darwin-kathu-yu\\
         brother \textsc{1sg-top} Daly River already \textsc{clf:human} \textsc{neg:human-temp} Darwin-\textsc{hither-dm}\\
    \glt ‘My brother is in Daly River already, he’s not in Darwin anymore (heading this way).’ (MP-20180718-RN01) 
\z  

\ea\label{ex:mur:47}
    \gll ku were-ka \textbf{maku} {ngarra da}-yu\\
          \textsc{clf:anim} dog-\textsc{top} \textsc{neg:anim} home\textsc{{}-dm}\\
    \glt ‘The dog’s not at home.’ (elicited, MP-20180718-RN01)
\z

The \textsc{place/time} classifier \textit{da} (also the \textsc{language} classifier \textit{murrinh}) doesn’t have a synthetic negative form, but is just negated with the regular NP negator \textit{manangka}, as shown in \REF{ex:mur:48}:

\ea\label{ex:mur:48}
    \gll manangka da dangatha\\
         \textsc{neg} \textsc{clf:place} \textsc{yet}\\
    \glt ‘It wasn't there at the time.’ (JB20050705JB01.txt:107)
\z 

It can therefore be hard to determine whether these constructions are negative predicate constructions, such as those above, or simple NP negation as illustrated in \sectref{sec:mur:2.3.1} above. However, examples such as \REF{ex:mur:49} suggest that \textit{manangka da} functions predicatively like the other synthetic negative classifier predicates, since it co-occurs with the regular noun classifier when a particular place is being discussed:

\ea\label{ex:mur:49}
    \gll  … ngarra da purtek ngala kanhi-ka \textbf{da} \textbf{manangka} \textbf{da} dangatha \\
         {} \textsc{rel} \textsc{clf:place} ground big this-\textsc{top} \textsc{clf:place} \textsc{neg} \textsc{clf:place} yet\\
    \glt ‘Before the world began...’ (lit. ‘When this entire place was not a place yet’) (\citealt{MansfieldEtAl2019}, 01-Christ-Comes-to-the-World)
\z  

Negative classifier predicates are also the standard way to encode negative existence. Examples include (\ref{ex:mur:50}--\ref{ex:mur:53}):

\ea\label{ex:mur:50}
    \gll bere ku ngalmungku-ka \textbf{maku}\\
         \textsc{so} \textsc{clf:anim} red.kangaroo-\textsc{top} \textsc{neg:anim}\\
    \glt ‘Then the kangaroo was gone.’ (RN Kanamkek-Stanner.wav)
\z 

\ea\label{ex:mur:51} 
    \gll perrkenku \textbf{makardu}{}-warda\\
         two \textsc{neg:human-temp}\\
    \glt ‘(and) the two who have passed away now’ (MP-20120724-RN01)
\z 

\ea\label{ex:mur:52}
    \gll \textbf{makardu}{}-warda \\
          \textsc{neg:human-temp}\\
    \glt ‘No-one is around.’ (MP-20180718-RN01)
\z 

\ea\label{ex:mur:53} 
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:mur:53Q}
        \gll  mi tham-rimut?\\ 
              \textsc{clf:veg} \textsc{2sg.sbj.poke(19).nfut}{}-leave.over\\
        \glt ‘Did you leave some (food)?’
    \exi{A: }\label{ex:mur:53A} %KM comma or period?
        \gll  awu mi ngam-rigerdek \textbf{mami}-warda\\
               no \textsc{clf:veg} \textsc{1sg.sbj.poke(19).nfut}{}-finish \textsc{neg:veg-temp}\\
        \glt  ‘No, I finished it all. There’s nothing left.’ (MP-20110810-RN01)
    \end{xlist}
\z 

In \REF{ex:mur:54}, the existence of the feet is not actually being negated, but rather the awareness of them by the protagonists:

\ea\label{ex:mur:54}
    \gll nanthi me \textbf{mananthi} mere pube-me-ngkardu-dha\\
         \textsc{clf:thing} foot \textsc{neg:thing} \textsc{neg} \textsc{3du.sbj.see(13).pst:irr}{}-foot-see-\textsc{pst} \\
    \glt ‘They didn't see (Thiniminh's) feet (sticking out from under the paperbark).’ (RN Kanamkek-Stanner.wav)
\z

Negative actions expressed by non-verbal clauses are also a type of negated existential predicate (\ref{ex:mur:55}--\ref{ex:mur:56}):

\ea\label{ex:mur:55}
    \gll \textbf{makura}{}-wun-thay\\
         \textsc{neg:water-3pl.obj}-mouth\\
    \glt ‘No drinking, they don’t drink.’ (lit. ‘Their mouths have no water.’) (RN, written on sign in the Wadeye community) 
\z

\ea\label{ex:mur:56} 
    \gll da four month kaka-yu \textbf{kura} \textbf{makura}\\
          \textsc{clf:time} four month uncle\textsc{{}-dm} \textsc{clf:water} \textsc{neg:water}\\
    \glt ‘It's nearly four months that your uncle hasn't been drinking.’ (lit. ‘It’s four months uncle has no water.’) (JB20090707JBvid04.wav)
\z 

However, when the nominal classifier is itself used as predicate \REF{ex:mur:57}, the regular nominal negator must be used instead, as it is with other nominal predicates (see \REF{ex:mur:31} and \REF{ex:mur:32} above, for example):

\ea\label{ex:mur:57} 
    \gll ngankungintha kardu \textbf{manangka} \textbf{kura} \textbf{(*makura)}\\
          \textsc{1du.excl.nsib} \textsc{clf:human} \textsc{neg} drunk\\
    \glt ‘The two of us weren’t drunk.’ (\citealt{MansfieldEtAl2019}, 2015-01-29\_AMN\_NAATI)
\z 

Possessive clauses are also negated with negative classifier predicate constructions as illustrated in (\ref{ex:mur:58}--\ref{ex:mur:63}). The regular NP negator is not possible in these constructions as shown by \REF{ex:mur:58b}. 

\ea\label{ex:mur:58}
    \ea[]{\label{ex:mur:58a}
        \gll  ngay-ka ku were \textbf{maku}\\ 
              \textsc{1sg-top} \textsc{clf:anim} dog \textsc{neg:anim}\\
        \glt ‘I don’t have a dog.’ (MP-20180718-RN01)}
    \ex[*]{\label{ex:mur:58b}
        \gll  ngay-ka manangka ku were\\
              \textsc{1sg-top} \textsc{neg} \textsc{clf:anim} dog\\
        \glt Attempted: ‘I don’t have a dog.’ (MP-20180718-RN01)}
    \z
\z 

\ea\label{ex:mur:59}
    \gll ngay-ka \textbf{maku} ku tumtum \\
          \textsc{1sg-top} \textsc{neg:anim} \textsc{clf:anim} egg\\
    \glt ‘I don’t have any eggs.’ (MP-20180718-RN01)
\z 

\ea\label{ex:mur:60}
    \gll ngay-ka kardu wakal \textbf{makardu}\\
          \textsc{1sg-top} \textsc{clf:human} child \textsc{neg:human}\\
    \glt ‘I don’t have any children.’ (MP-20180718-RN01)
\z 

\ea\label{ex:mur:61}
    \gll kardu wakal dimu \textbf{mananthi}\\
         \textsc{clf:human} child tooth \textsc{neg:thing}\\
    \glt ‘The baby’s got no teeth.’ (MP-20180718-RN01)
\z 

\ea\label{ex:mur:62}
    \gll ngay-ka mi peka \textbf{mami} \\
         \textsc{1sg-top} \textsc{clf:veg} tobacco \textsc{neg:veg}\\
    \glt ‘I don’t have any tobacco.’ (RN20130702-003:054)
\z
         
\ea\label{ex:mur:63}
    \gll nukunu-ka the kanthin-nan ngarra mi \textbf{mami} narnam i nanthi \textbf{mananthi} narnam \\
         \textsc{3sg.m-top} ear \textsc{3sg.sbj.have(22).nfut-2pl.obj} \textsc{rel} \textsc{clf:veg} \textsc{neg:veg} \textsc{2pl.sbj.be(4).nfut} and \textsc{clf:thing} \textsc{neg:thing} \textsc{2pl.sbj.be(4).nfut}\\
    \glt ‘and God knows that you need things (that you don’t have).’ (lit. ‘He knows about you that you are ‘food no food’ and you are ‘thing no-thing’’) (RN, Mt6:24-34-009) 
\z 

Affirmative possessive constructions of this type are very rare in the corpus, but when elicited are generally expressed with a verbal construction as in \REF{ex:mur:64} and \REF{ex:mur:65}. Affirmative existential constructions also use a verbal construction, as in \REF{ex:mur:66} repeated from above. We therefore have another type of structural asymmetry here where the affirmative construction involves a construction with a verbal predicate while the negative equivalent is expressed using a negative classifier predicate.

\ea\label{ex:mur:64}
    \gll nigunu-ka dimu kanthin\\
         3\textsc{sg.m-top} tooth 3\textsc{sg.sbj.have(22).nfut}\\
    \glt ‘He (the baby) has teeth.’ (elicited, Bill Forshaw, pers. comm.)
\z

\ea\label{ex:mur:65}
    \gll nigunu-ka ngathan numi kanthin\\
         3\textsc{sg.m-top} brother one 3\textsc{sg.sbj.have(22).nfut}\\
    \glt  ‘He has one brother.’ (elicited, Bill Forshaw, pers. comm.)
\z 


\ea\label{ex:mur:66} 
    \gll  kardu kem\\ 
          \textsc{clf:human} \textsc{3sg.sbj.be(4).ex}\\
    \glt ‘Someone is there.’ (constructed, for illustration)
\z 


The existence of negative classifier predicates in Murrinhpatha is consistent with \citegen{Veselinova2013neg.exist} claim that negative existence is a separate functional domain, distinct from both standard negation and affirmative existence. However, negative classifier predicates in Murrinhpatha further expand the typology of negative existentials by showing that negative existential strategies can develop from nominal classifiers, something which has not been reported before in typological literature (\citealt{Eriksen2011}; \citealt{Veselinova2013neg.exist}). 

\subsection{Negation in non-main clauses} \label{sec:mur:2.4}

Murrinhpatha doesn’t have non-finite verbal clauses; any non-main verbal clauses have the same structure as main clauses, with the same negation strategies. The negative particle can have scope over just the higher \REF{ex:mur:67} or lower \REF{ex:mur:68} clause; the scope of the negative particle is clearly indicated by irrealis marking on the verb (or lack thereof).

\ea\label{ex:mur:67}
    \gll \textbf{mere} \textbf{the} \textbf{ma-bath} ngarra-kama parnam\\
        \textsc{neg} ear \textsc{1sg.sbj.hands(8).fut:irr}{}-have where-\textsc{maybe} \textsc{3pl.sbj.be(4).nfut} \\
    \glt ‘I don’t know where they are.’ (RN20091006-002:121)
\z 

\ea\label{ex:mur:68}
    \gll murrinh nukunu wurran-ngerren \textbf{mere} \textbf{bi-ma-yepup}\\
          \textsc{clf:lang} \textsc{3sg.m} \textsc{3sg.sbj.go(6).nfut}{}-speak \textsc{neg} \textsc{1sg.sbj.listen(16).fut:irr-appl}{}-hear\\
    \glt ‘He’s talking a language I can't understand.’ (JB20040624JB01.txt)
\z

Embedded non-verbal clauses are negated with negative classifier predicates (\sectref{sec:mur:2.3.3}), as shown in \REF{ex:mur:69}:

\ea\label{ex:mur:69} 
    \gll bam-ngkardu \textbf{nanthi} \textbf{me} \textbf{boot} \textbf{mananthi}, nungam-winharart-yu\\
         \textsc{1sg.sbj.see(13).nfut}{}-see \textsc{clf:thing} foot boot \textsc{neg:thing} \textsc{3sg.sbj.foot(7).nfut}{}-run-\textsc{dm}\\
    \glt ‘I saw him running without boots.’ (lit. ‘I saw him without boots, running.’) (MP-20180718-RN01)
\z

\subsection{Negative lexicalizations} \label{sec:mur:2.5}

Murrinhpatha has a lexicalized interjection \textit{pirda} ‘don’t, stop it, leave it’ (exemplified in \sectref{sec:mur:2.2} and \REF{ex:mur:70})
, a number of lexical stems that encode a negated event as part of their inherent semantics (\ref{ex:mur:71}--\ref{ex:mur:72}) and an idiomatic construction meaning ‘don’t want, don’t like’ (\ref{ex:mur:74}--\ref{ex:mur:76}). The meanings of many of these negative lexicalizations are consistent with the cross-linguistic tendencies identified by \citet{Veselinova2013lexicalization}, especially the lexicalized construction meaning ‘don’t want, don’t like’, which her study found to be very commonly lexicalized across languages in her sample. 

\ea\label{ex:mur:70}
    \gll pirda! \\
         stop\\
    \glt ‘Don’t (touch it), leave it alone!’ (\citealt{MansfieldEtAl2019}, 2015-07-338.1)
\z 

Other negative lexicalizations arise from the meanings of specific lexical stems, which encode a negated state or event as part of their semantics, such as \textit{{}-ngkapi} ‘not recognize’ \REF{ex:mur:71}, and \textit{{}-biye} ‘not realise’ \REF{ex:mur:72}. While these are semantically negative, they are not grammatically negative constructions as shown by the fact that they occur inflected with realis tense categories (\ref{ex:mur:71a}, \ref{ex:mur:72}), and can themselves be negated (in which case they are inflected with irrealis mood) as in \REF{ex:mur:71b}.

\ea\label{ex:mur:71}
    \ea\label{ex:mur:71a}
        \gll  wurdam-\textbf{ngkapi}\\ 
              \textsc{3sg.sbj.shove:rr(30).nfut}{}-not.recognize\\
        \glt ‘He didn’t recognize them.’ (\citealt{MansfieldEtAl2019}, 2015-07-21 210.1)
    \ex\label{ex:mur:71b}
        \gll  mere-warda kurdi-nhi-\textbf{ngkapi}\\
              \textsc{neg-temp} \textsc{3sg.sbj.shove:rr(30).fut:irr-2sg.obj}{}-not.recognize\\
        \glt ‘So he won’t treat you like a stranger.’ (lit. ‘So he won’t not recognize you.’) (\citealt{MansfieldEtAl2019}, 2015-07-21 168.1)
    \z
\z 

\ea\label{ex:mur:72}
    \gll kanam-part=wurran nukunu-ka puy me-nu-the-\textbf{biye}=wurrini yileyile-yu\\
         \textsc{3sg.sbj.be(4).nfut-}leave\textsc{=3sg.sbj.go(6).nfut} \textsc{3sg.m-top} keep.going \textsc{3sg.sbj.hands(8).pst:ipfv-rr}{}-ear-not.realise=\textsc{3sg.sbj.go(6).pst:ipfv} father-\textsc{dm}\\
    \glt ‘She left him and kept going, but her father didn’t realise.’ (\citealt{MansfieldEtAl2019}, CS1-017-A\_07)
\z 

There is also an idiomatic construction for ‘don’t want’ that is different to the construction used to express the positive ‘want’. While the expression for ‘want’ is an idiom which combines the body part noun \textit{marda} ‘belly’ with the verb meaning ‘to get/grab’ \REF{ex:mur:73}, the meaning ‘don’t want’ is formed with the negated clause ‘not be happy’ (\ref{ex:mur:74}--\ref{ex:mur:75}). This negated clause makes use of a regular negated construction and so this is not a negative lexicalization in the strict sense, however it is included here as it is lexicalized as the idiomatic way to encode ‘not want’.

\ea\label{ex:mur:73}
    \gll mi-nu \textbf{marda} \textbf{mangan-art}=dim\\
          \textsc{clf:veg-dat} belly \textsc{3sg.sbj.snatch(9).nfut}{}-get=\textsc{3sg.sbj.sit(1).nfut}\\
    \glt ‘She wants some food.’ (elicited, RN-20180724-004:038)
\z 

\ea\label{ex:mur:74}
    \gll \textbf{manangka} \textbf{le} \textbf{kani} mi dhepdhep\\
         \textsc{neg} happy \textsc{3sg.sbj.be(4).fut:irr} \textsc{clf:veg} food\\
    \glt ‘She doesn’t want any food.’ (lit. ‘She isn’t happy food-wise.’ (elicited, RN-20180724-004:038)
\z 

\ea\label{ex:mur:75}
    \gll \textbf{mere} \textbf{le} \textbf{ngurru}, durdu ma-marda-nukun=ngurru \\
         \textsc{neg} happy \textsc{1sg.sbj.go(6).fut:irr} breathless \textsc{1sg.sbj.hands(8).fut:irr}{}-belly-\textsc{lest=1sg.sbj.go(6).fut}\\
    \glt ‘I don’t want to go, I might get breathless.’ (lit. ‘I won’t go happy’) (JB20040912JB04.txt)
\z 

This construction is also used for ‘don’t like’ \REF{ex:mur:76}:

\ea\label{ex:mur:76}
    \gll \textbf{mere} \textbf{le} \textbf{ngani-na}\\
         \textsc{neg} happy \textsc{1sg.sbj.be(4).fut:irr-3sg.m.obl}\\
    \glt ‘I don’t like him.’ (lit. ‘I am not happy about him.’) (JB20110901\_B\_video\_GYHM100\_01)
\z 

\subsection{Other clausal negation constructions} \label{sec:mur:2.6}
\subsubsection{\textit{wurda} ‘no’ as a negative predicate} \label{sec:mur:2.6.1}

In some examples the negative word \textit{wurda} ‘no, nothing’ (see \sectref{sec:mur:3.1}) is used as a negative predicate in a verbless clause to indicate that something won’t, can’t or didn’t happen, see (\ref{ex:mur:77}--\ref{ex:mur:78}): 

\ea\label{ex:mur:77}
    \gll  mi peka-ka \textbf{wurda-wa} ngarra prison-yu\\
         \textsc{clf:veg} tobacco-\textsc{top} no-\textsc{emph} \textsc{loc} prison-\textsc{dm}\\
    \glt ‘Tobacco is banned in prison.’ (\citealt{MansfieldEtAl2019}, 2015-01-29\_AMN\_NAATI)
\z 

\ea\label{ex:mur:78}
    \gll kardu kigay ngamere-nimin pumpanka-pek-neme nanthi truck-yu, mu nhinhi-ka \textbf{wurda-wa}\\
         \textsc{clf:human} youth few-\textsc{other} \textsc{3du.sbj.go(6).nfut-}destroy\textsc{{}-pc.m} \textsc{clf:thing} truck-\textsc{dm} \textsc{but} \textsc{2sg-top} no\textsc{{}-emph}\\
    \glt ‘The other boys destroyed the truck, but you didn’t.’ (\citealt{FordMcCormack2011}: 1)
\z 
          

\section{Non-clausal negation} \label{sec:mur:3}
\subsection{Negative replies} \label{sec:mur:3.1}

In appropriate contexts, negative replies to polar questions can be expressed with the negative classifier predicates discussed in \sectref{sec:mur:2.3.3}. For example, a negative response to a question like ‘Did you catch any fish?’ might be \textit{maku} ‘\textsc{neg:anim}’; or in response to the question ‘Is anyone inside the house’ a negative reply might be \textit{makardu} ‘\textsc{neg:human}’. It is also possible to use one of two negative words: \textit{wurda} and \textit{awu.} 

\textit{Wurda} is the more general negative word, translating as ‘no, nothing’. It can be used as an interjection to either reinforce a negative statement \REF{ex:mur:86}, answer a question (\ref{ex:mur:87}, \ref{ex:mur:88}), or as a negative tag question \REF{ex:mur:89} (see also \sectref{sec:mur:2.6.1}): 

\ea\label{ex:mur:86}
    \gll manangka lurruth kardu, \textbf{wurda}\\
         \textsc{neg} strong \textsc{clf:human} no\\
    \glt ‘He’s not a strong man, no.’ (\citealt{MansfieldEtAl2019}, 1981\_CS1-03A\_GM)
\z 

\newpage
\ea\label{ex:mur:87} 
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:mur:87A}
        \glt ‘Do you light fires there?’ 
    \exi{A: }\label{ex:mur:87B}
        \gll \textbf{wurda}, mere nga-rdudrurt=ngarni da pangu-yu\\
              no \textsc{neg} \textsc{1pl.sbj.poke(19).fut:irr-}light.fires=\textsc{1pl.sbj.be(4).fut:irr} \textsc{clf:place} there-\textsc{dm}\\
        \glt ‘No, we don’t light fires there.’ (MP-20091008-RN01)
    \end{xlist}
\z 

\ea\label{ex:mur:88}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:mur:88A}
        \glt ‘Will you be taking me home?’ 
    \exi{A: }\label{ex:mur:88B}
        \gll ngay-ka \textbf{wurda-wa}\\
             \textsc{1sg-top} no-\textsc{emph}\\
        \glt ‘Not me (someone else will).’ (MP-20100920-RN01)
    \end{xlist}
\z 

\ea\label{ex:mur:89}
    \gll kura kanhi nukunu wurran=dim mu \textbf{wurda}?\\
         \textsc{clf:water} here \textsc{3sg.m} \textsc{3sg.sbj.go(6).nfut=3sg.sbj.sit(1).nfut} \textsc{or} no\\
    \glt ‘Does he go through this water here or not?’ (\citealt{MansfieldEtAl2019}, 2012-06-20\_landing\_29)
\z 

\textit{Awu} is an interjection and is always initial in the clause. While it is a negative interjection, it has a different pragmatic force to \textit{wurda} and is often used in situations where the utterance is considered to be counter to expectations in some way. 

It can be used to signal disagreement with a prior suggestion or statement, as in \REF{ex:mur:90}:

\ea\label{ex:mur:90}
    \begin{xlist}
    \exi{A: }\label{ex:mur:90Q}
        \gll  kura ti kanhi-ka pa-winhipak-warda\\ 
              \textsc{clf:water} tea here-\textsc{top} \textsc{1incl.sbj.poke(19).fut:irr}{}-spill-\textsc{temp}\\ 
        \glt ‘Let’s tip out this tea.’
    \exi{B: }\label{ex:mur:90A}
        \gll  \textbf{awu}, kura pi-gurdugurduk-ngime\\ 
               no \textsc{clf:water} \textsc{1incl.sbj.sit(1).fut:irr-}drink\textsc{{}-pc.f}\\
        \glt ‘No, let’s drink it\textsc{.’} (\citealt{MansfieldEtAl2019}, 20110828\_JB\_video)
    \end{xlist}
\z 

In some examples it is used to mark contrast with a preceding statement, rather than negation. In the example in \REF{ex:mur:91}, two characters in a story are talking about where they want to go; \textit{awu} is used to contrast with the preceding utterance, rather than to negate or disagree with its content. 

\ea\label{ex:mur:91}
        \gll  nukunu-ka mem-nu Kanamkek-yu \\ 
              \textsc{3sg.m-top} 3\textsc{sg.sbj.hands:rr(10).nfut-rr} Kanamkek-\textsc{dm} \\ 
        \glt ‘Kanamkek said to himself, 
        \medskip \\ 
        \gll   “ngay-ka nga-wintigat-nu kanhi-ngu”\\ 
             {\db}\textsc{1sg-top} \textsc{1sg.sbj.poke(19).fut}{}-descend-\textsc{fut} here-\textsc{way}\\ 
        \glt “I’ll go down (to the water) this way.”
        \medskip \\ 
        \gll  Thinaminh-ka mam \\ 
              Thinaminh-\textsc{top} \textsc{3sg.sbj.say/do(34).nfut} \\
        \glt Thinaminh said,  \\
        \medskip
        \gll  “\textbf{awu} ngay-ka ngi-mardawith-nu kanhi-ngu, palyirr”\\ 
               {\db}no \textsc{1sg-top} \textsc{1sg.sbj.sit(1).fut}{}-ascend-\textsc{fut} here-\textsc{way} hill\\
        \glt “No, I’m going up this way, to the hills.”’ (\citealt{MansfieldEtAl2019}, 2000-11-10\_LKLM) 
\z 

\textit{Awu} is also used in corrections, including self-corrections \REF{ex:mur:92}:

\ea\label{ex:mur:92} 
    \gll piguna-ka pumpanka […] \textbf{awu}! mi bamngutut kuba-birr=kuru\\
         \textsc{3du.sib-top} \textsc{3du.sbj.go(6).nfut} {} no \textsc{clf:veg} boab \textsc{3pl.sbj.bash(14).fut:irr-}plant\textsc{=3pl.sbj.go(6).fut:irr}\\
    \glt ‘The sisters were going […] no! they were all going to plant boab fruit.’ (\citealt{MansfieldEtAl2019}, 2000-11-319.1)
\z 

In some examples \textit{awu} is used in response to a content question. In these examples it may be used to deny the implicature generated by the question. In \REF{ex:mur:93}, for example, B’s use of \textit{awu} denies the implicature that there is anything particularly wrong. 

\ea\label{ex:mur:93}
    \begin{xlist}
    \exi{A: }\label{ex:mur:93A}
        \gll  thangkudha yawu?\\ 
              what.from \textsc{hey}\\ 
        \glt ‘What’s wrong?’
    \exi{B: }\label{ex:mur:93B}
        \gll  \textbf{awu}, mi peka lurruth ngala-wa\\ 
               no \textsc{clf:veg} tobacco strong big-\textsc{emph}\\
        \glt ‘\textit{Awu}, the tobacco is really strong.’ (\citealt{MansfieldEtAl2019}, 20171403\_AN)
    \end{xlist}
\z 

The two negative words can combine to mean ‘definitely not’. In this case, they are always in the order \textit{awu wurda} \REF{ex:mur:94}: 

\protectedex{\ea\label{ex:mur:94} %KM changed 3sS to 3sg.sbj, is that correct?
    \gll  “nhinhi-ka kale ngay?” mam \\ 
          {\db}\textsc{2sg-top} mother \textsc{1sg} \textsc{3sg.sbj.say/do(34).nfut} \\
    \glt ‘“Are you my mother?” he said. \\
    \medskip 
    \gll “\textbf{awu} \textbf{wurda}!” mam ku ngurlmirl-yu  \\
        {\db}no no \textsc{3sg.sbj.say/do(34).nfut} \textsc{clf:anim} fish-\textsc{dm}  \\
    \glt “No, definitely not!” said the fish. \\
    \medskip
    \gll “ngay-ka ku ngurlmirl-wa!”  \\
        {\db}\textsc{1sg-top} \textsc{clf:anim} fish-\textsc{emph} \\
    \glt “I’m a fish!”’ (\textit{Are you my mother?} children’s book)\\ 
\z} 
    
\subsection{Negative indefinites and quantifiers} \label{sec:mur:3.2}

There is a set of indefinite/interrogative pronouns including \textit{nangkarl} ‘someone, who’ \REF{ex:mur:95}, \textit{ngarra} ‘sometime/somewhere, when/where’ \REF{ex:mur:96} and \textit{thangku} ‘something, what’ \REF{ex:mur:97}. When functioning as indefinites these are often found co-occurring with the epistemic suffix -\textit{kama} ‘\textsc{maybe}’ as in \REF{ex:mur:95}.

\ea\label{ex:mur:95}
    \gll \textbf{nangkarl}{}-kama pan-nga-bat\\
         who{}-\textsc{maybe} \textsc{3sg.sbj.slash(23).nfut-1sg.obl-}hit\\
    \glt ‘Someone called me (on the phone).’ (MP-20091008-RN01)
\z

\ea\label{ex:mur:96}
    \gll da pana \textbf{ngarra} purru-lili\\
         \textsc{clf:place} \textsc{rcgn} where \textsc{1incl.sbj.go(6).nfut}{}-walk\\
    \glt ‘We just go walk somewhere.’ (MP-20091008-RN01)
\z 

\ea\label{ex:mur:97}
    \gll ku balli-ya ngarde-wat-ngime-ya ku \textbf{thangku}{}-wu pana ngurlmirl\\
         \textsc{clf:anim} mud.crab-\textsc{dm} \textsc{1du.excl.sbj.be(4).pst:ipfv}{}-go-\textsc{pc.f-dm} \textsc{clf:anim} what{}-\textsc{emph} \textsc{rcgn} fish \\
    \glt ‘We would go for crabs, and, something else, fish.’ (MP-20120724-RN01)
\z 

These indefinite pronouns can be negated with the clausal negator \textit{mere}, as with all other pronouns \REF{ex:mur:98}:

\ea\label{ex:mur:98}
    \gll \textbf{mere} \textbf{thangku} deyida\\
          \textsc{neg} what again\\
    \glt ‘nothing else’ (20110730\_JB\_video\_GYHM100\_04)
\z 

Indefinite pronouns combine with standard clausal negation in a verbal clause to form their negated versions, thus Murrinhpatha belongs to \citegen{Kahrel1996} Type 1 in this respect. Note that while standard clausal negation is used, in these constructions the indefinite pronoun always immediately follows the negative particle, which is not the usual position for nominal arguments in standard negation. Nonetheless, the fact that the verb must appear in the irrealis mood, as with regular clausal negation, shows that the negative marker \textit{mere} has scope over the whole clause, not just the indefinite pronoun (\ref{ex:mur:99}--\ref{ex:mur:101}).

\ea\label{ex:mur:99}
    \gll  \textbf{mere} \textbf{thangku} ku-ngi-mathak=kani\\
         \textsc{neg} what \textsc{3sg.sbj.slash(23).fut:irr-1sg.obj}{}-have.things=\textsc{3sg.sbj.be(4).fut:irr}\\
    \glt  ‘I’ve got nothing (in my pockets).’ (MP-20091001-RN01)
\z

\ea\label{ex:mur:100}
    \gll \textbf{mere} \textbf{nangkarl}-warda kani\\
         \textsc{neg} who{}-\textsc{temp} \textsc{3sg.sbj.be(4).fut:irr}\\
    \glt  ‘No-one is living there.’ (MP-20180718-RN01)
\z 

\ea\label{ex:mur:101}
    \gll \textbf{mere} \textbf{nangkarl} be-rdurt-dha\\
         \textsc{neg} who \textsc{1sg.sbj.bash(14).pst:irr}{}-find-\textsc{pst}\\
    \glt ‘I didn’t see (find) anyone (there).’ (MP-20180718-RN01)
\z 

\textit{Mere ngarra} appears to have been extended into other negative functions as well, including meanings such as ‘not ever, never, can’t’ (\ref{ex:mur:102}--\ref{ex:mur:104}):

\ea\label{ex:mur:102}
    \gll da kura \textbf{mere} \textbf{ngarra} karni-yel da mirrangan-wa kanhi-yu\\
         \textsc{clf:time} \textsc{clf:water} \textsc{neg} what \textsc{3sg.sbj.be(4).fut:irr}{}-rain \textsc{clf:time} dry.season-\textsc{emph} here-\textsc{dm} \\
    \glt ‘It never rains in the dry season (which is now).’ (MP-20180718-RN01)
\z 

\ea\label{ex:mur:103} %KM capital letter/period?
    \gll mu \textbf{manangka} \textbf{ngarra} fix kama-na\\
         \textsc{but} \textsc{neg} what fix \textsc{3sg.sbj.do(34).fut:irr-3sg.m.obl}\\
    \glt ‘but (the doctor) can’t fix him’ (RN, overheard)
\z 

\ea\label{ex:mur:104}
    \gll \textbf{mere} \textbf{ngarra} pani-la\\
         \textsc{neg} what \textsc{1incl.sbj.be(4).fut:irr}{}-climb\\
    \glt ‘We can’t climb (that hill).’ (20110825\_JB\_video\_GYHM100\_01)
\z 


In some examples, \textit{mere ngarra} seems to express lack of volition, rather than meaning ‘never, can’t’. In \REF{ex:mur:105} and \REF{ex:mur:106} the use of \textit{mere ngarra} indicates a choice on the part of the speaker not to undertake the actions described.


\ea\label{ex:mur:105} 
    \gll \textbf{mere} \textbf{ngarra} me-watha-dha, {le wiye}\\
         \textsc{neg} what \textsc{1sg.hands(8).pst:irr-}make\textsc{{}-pst} tired\\
    \glt ‘I didn’t make (any tea), (I’m) tired.’ (MP-20100920-RN01)
\z 

\ea\label{ex:mur:106}
    \gll \textbf{mere} \textbf{ngarra} ngurru\\
          \textsc{neg} what \textsc{1sg.sbj.go(6).fut:irr}\\
    \glt ‘I’m not going.’ (RN-20180725-004:038)
\z 
         
Example \REF{ex:mur:107} shows nicely the interaction of the different negative construction types. Speaker A uses an indefinite pronoun to ask a question. In their response, speaker B answers first with the negative word \textit{wurda} (\sectref{sec:mur:3.1}), then a negative existential construction using a negative classifier predicate (\sectref{sec:mur:2.3.3}), and then reiterates with the use of a negative indefinite construction in which the indefinite pronoun \textit{nangkarl} is used as an argument NP in the negated verbal clause.

\ea\label{ex:mur:107}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:mur:107A}
        \gll  kardu nangkarl-ngadha bam-nhi-ngkardu?\\ 
              \textsc{clf:human} who{}-\textsc{still} \textsc{3sg.sbj.see(13).nfut-2sg.obj-}see\\ 
        \glt ‘Did someone come to see you?’
    \exi{A: }\label{ex:mur:107B}
        \gll  \textbf{wurda}, \textbf{makardu}. \textbf{mere} \textbf{nangkarl} be-ngi-ngkardu-dha\\ 
               no \textsc{neg:human} \textsc{neg} who \textsc{3sg.sbj.see(13).pst:irr-1sg.obj-}see-\textsc{pst}\\
        \glt ‘No, no-one. No-one came to see me.’ (MP-20180718-RN01)
    \end{xlist}
\z 

\subsection{Negative derivation and case marking} \label{sec:mur:3.3}

In general, the meaning of absence at the lexical level – the equivalent to English ‘without’ – is expressed as a non-verbal subordinate clause with a negative classifier predicate as discussed in \sectref{sec:mur:2.3.3}, \sectref{sec:mur:2.4} and illustrated in (\ref{ex:mur:108}--\ref{ex:mur:109}). The use of negative classifier predicates in this way is consistent with \citegen{Veselinova2013neg.exist} finding that negative existential constructions commonly also have privative functions cross-linguistically.

\ea\label{ex:mur:108}
    \gll kardu nungam-pinharart nanthi me boot \textbf{mananthi}\\
         \textsc{clf:human} \textsc{3sg.sbj.foot(7).nfut}{}-run \textsc{clf:thing} foot boot \textsc{neg:thing}\\
    \glt ‘Someone’s running without shoes.’ (lit. ‘A person is running, (s/he is) without shoes’) (elicited, MP-20180718-RN01)
\z 

\ea\label{ex:mur:109} 
    \gll bam-ngkardu nanthi me boot \textbf{mananthi}, nungam-winharart-yu\\
         \textsc{1sg.sbj.see(13).nfut}{}-see \textsc{clf:thing} foot boot \textsc{neg:thing} \textsc{3sg.sbj.foot(7).nfut}{}-run-\textsc{dm}\\
    \glt ‘I saw him running without boots.’ (lit. ‘I saw him, (he was) without boots, he was running.’) (elicited, MP-20180718-RN01)
\z 

The NP negator \textit{manangka} is also possible in this function, for some speakers (\ref{ex:mur:110}--\ref{ex:mur:111}). Note that in these constructions it never combines with a nominal classifier relating to the negated element: 

\ea\label{ex:mur:110}
    \gll \textbf{manangka} me boot parde-wilili-dha-neme\\
         \textsc{neg} foot boot \textsc{3du.sbj.be(4).pst:ipfv-walk-pst-pc.m}\\
    \glt  ‘They walked without any shoes.’ (\citealt{MansfieldEtAl2019}, 20152-07-13\_CW\_verb-suffixes)
\z 

\ea\label{ex:mur:111} 
    \gll kardu \textbf{manangka} pemarr-we\\
         \textsc{clf:human} \textsc{neg} hair-head\\
    \glt ‘person without head hair’ \citep[182]{Walsh1976}
\z 

There is one idiomatic construction in which an associative suffix \textit{{}-ma} is suffixed to the negative classifier predicate \textit{makardu} to indicate absence of a wife or girlfriend (\ref{ex:mur:112}--\ref{ex:mur:113}). This construction is not possible without the \textit{{}-ma} as shown in \REF{ex:mur:114}.

\ea\label{ex:mur:112}
    \gll kardu palngun \textbf{makardu-ma}\\
         \textsc{clf:human} woman \textsc{neg:human-assoc}\\
    \glt ‘single man’ (lit. ‘person without a woman’) (MP-20180718-RN01)
\z 

\ea\label{ex:mur:113} 
    \gll  palngun \textbf{makardu-ngi-ma}\\
         woman \textsc{neg:human-1sg.obj-assoc}\\
    \glt ‘I’ve got no girlfriend/wife.’ (MP-20180718-RN01)
\z

\ea[*]{\label{ex:mur:114}
    \gll palngun makardu\\
         woman \textsc{neg:human}\\
    \glt ‘He’s got no girlfriend/wife.’ (MP-20180718-RN01)}
\z 

In other contexts the nominal suffix \textit{{}-ma} functions as a comitative \citep[183]{Walsh1976} or associative, as in \REF{ex:mur:115} and \REF{ex:mur:116}:

\ea\label{ex:mur:115} 
    \gll  kardu thamul-ma\\
         \textsc{clf:human} spear-\textsc{assoc}\\
    \glt ‘a person habitually associated with spears’ \citep[187]{Walsh1976} 
\z 

\ea\label{ex:mur:116}
    \gll batbat-ngi-ma\\
         right.hand-\textsc{1sg.obj-assoc}\\
    \glt ‘I’m right-handed.’ \citep[166]{Mansfield2019}
\z 

The fact that the construction in \REF{ex:mur:112} is idiomatic is shown by the fact that the comitative/associative \textit{{}-ma} is not possible in other contexts, such as when talking about a woman without a husband. In this case the nominal classifier predicate is not possible (cf. \REF{ex:mur:118} with \REF{ex:mur:112} above), and the regular NP negator must be used instead, as in \REF{ex:mur:117}.

\ea\label{ex:mur:117} 
    \gll kardu manangka nugalin\\
         \textsc{clf:human} \textsc{neg} husband\\
    \glt ‘single woman’ (lit. ‘person without a husband’) (elicited, MP-20180718-RN01) 
\z 

\ea[*]{\label{ex:mur:118}
    \gll kardu nugalin makardu-ma\\
         \textsc{clf:human} husband \textsc{neg:human-assoc}\\
    \glt Attempted: ‘single woman’ (checked in elicitation, rejected and replaced with \REF{ex:mur:117}, MP-20180718-RN01)}
\z 

\section{Other aspects of negation} \label{sec:mur:4}
\subsection{The scope of negation} \label{sec:mur:4.1}

The negative particle generally has scope over the entire clause, reflected in the presence of irrealis mood marking on the verb (\sectref{sec:mur:2.1}). The serialised classifier stem which marks imperfective aspect (\citealt{NordlingerCaudal2012}) must also appear in the irrealis form in negative constructions, showing it to be within the scope of the negative operator \REF{ex:mur:119}.

\ea\label{ex:mur:119}
    \ea\label{ex:mur:119a}
        \gll  \textbf{dam}{}-rilil=\textbf{dim}\\ 
              \textsc{3sg.sbj.poke(19).nfut-}write\textsc{=3sg.sbj.sit(1).nfut}\\
        \glt ‘He’s writing it down.’ (MP-20100917-RN01)
    \ex\label{ex:mur:119b}
        \gll  mere \textbf{ka}{}-rilil=\textbf{ki}\\
              \textsc{neg} \textsc{3sg.sbj.poke(19).fut:irr-}write\textsc{=3sg.sbj.sit(1).fut:irr}\\
        \glt ‘He’s not writing it down.’ (MP-20100917-RN01)
    \z
\z 

As discussed in \sectref{sec:mur:2.4}, in complex sentences the negative particle can have scope over just the higher or lower clause, with the relevant verb inflecting for irrealis mood, as shown in (\ref{ex:mur:120}--\ref{ex:mur:121}) (repeated from above).

\ea\label{ex:mur:120}
    \gll \textbf{mere} \textbf{the} \textbf{ma-bath} ngarra-kama parnam\\
          \textsc{neg} ear \textsc{1sg.sbj.hands(8).fut:irr}{}-have where-\textsc{maybe} \textsc{3pl.sbj.be(3).nfut}\\
    \glt ‘I don’t know where they are.’ (RN20091006-002:121)
\z 

\ea\label{ex:mur:121}
    \gll murrinh nukunu wurran-ngerren \textbf{mere} \textbf{bi-ma-yepup}\\
        \textsc{clf:lang} \textsc{3sg.m} \textsc{3sg.sbj.go(6).nfut}{}-speak \textsc{neg} \textsc{1sg.sbj.listen(16).fut:irr-appl}{}-hear \\
    \glt ‘He’s talking a language I can't understand.’ (JB20040624JB01.txt)
\z 

\subsection{Negative polarity} \label{sec:mur:4.2}

As far as I have been able to determine, there are no negative polarity items in Murrinhpatha. See \sectref{sec:mur:3.2} for a discussion of indefinite pronouns and their interaction with negative particles.

\subsection{Marking of NPs in the scope of negation} \label{sec:mur:4.3}

Case marking on NPs in Murrinhpatha is very limited and mostly not obligatory (see \citealt{Nordlinger2015}; \citealt{Mansfield2019}). There is no special marking of NPs within the scope of negation, although the language does have special negative classifier predicates which are used in existential, locative and possessive constructions as discussed in \sectref{sec:mur:2.3.3}.

\subsection{Reinforcing negation} \label{sec:mur:4.4}

The negative particle \textit{mere} can be inflected with the ‘\textsc{lest}’ suffix \textit{{}-nukun} to emphasise the undesirability of the negated action. This construction is common in imperatives (see \sectref{sec:mur:2.2}), but can be found in declarative clauses too, as in (\ref{ex:mur:122}--\ref{ex:mur:123}).

\ea\label{ex:mur:122}
    \gll nekineme-ka \textbf{mere-nukun} pa-yetmut-neme=purru\\
         \textsc{1pc.m.incl-top} \textsc{neg-lest} 1\textsc{incl.sbj.poke(19).fut:irr}{}-refuse-\textsc{pc.m=1incl.sbj.go(6).fut:irr}\\
    \glt ‘We won’t refuse him.’ (JB20100827\_JB\_video\_02\_modified.wav)
\z 

\ea\label{ex:mur:123}
    \gll da pana-wangu-ka \textbf{mere}{}-\textbf{nukun} nunna-ngime\\
         \textsc{clf:place} \textsc{rcgn-away-top} \textsc{neg-lest} \textsc{2du.sbj.travel(7).fut:irr-pc.f}\\
    \glt ‘Don’t you ever go near that place (warning children).’ (MP-20120724-RN01)
\z 

Reinforced negation is also achieved through combining the two negative words \textit{awu wurda} to mean ‘definitely not’, as discussed in \sectref{sec:mur:3.1}, and shown in \REF{ex:mur:124} (repeated from above):
 
\protectedex{\ea\label{ex:mur:124}
    \gll  “nhinhi-ka kale ngay?” mam  \\ 
          {\db}\textsc{2sg-top} mother \textsc{1sg} \textsc{3sg.sbj.say/do(34).nfut} \\
    \glt ‘“Are you my mother?” he said. \\
    \medskip 
    \gll “\textbf{awu} \textbf{wurda}!” mam ku ngurlmirl-yu  \\
        {\db}no no \textsc{3sg.sbj.say/do(34).nfut} \textsc{clf:anim} fish-\textsc{dm}  \\
    \glt “No, definitely not!” said the fish. \\
    \medskip
    \gll “ngay-ka ku ngurlmirl-wa!”  \\
        {\db}\textsc{1sg-top} \textsc{clf:anim} fish-\textsc{emph} \\
    \glt “I’m a fish!”’ (\textit{Are you my mother?} children’s book)\\ 
\z} 
 
\subsection{Negation, coordination and complex clauses} \label{sec:mur:4.5}

There are no special negative coordinators in the current Murrinhpatha corpus, and no examples where a negator has scope over a coordinated construction. \sectref{sec:mur:2.4} provides examples showing the independent negation of the higher and lower clauses in complex sentences. In the following examples we see that \textit{wurda} ‘no, nothing’ can be used to negate an ellipsed negative declarative in contrastive ‘X but didn’t’ clauses \REF{ex:mur:125}, or those with contrastive subjects \REF{ex:mur:126}.

\ea\label{ex:mur:125}
    \gll me-watha-dha-warda mu \textbf{wurda}\\
         \textsc{1sg.sbj.hands(8).pst:irr-}make\textsc{{}-pst-temp} \textsc{but} no\\
    \glt ‘I tried to make (some tea) but didn’t. (MP-20100920-RN01)
\z

\ea\label{ex:mur:126}
    \gll kardu kigay ngamere-nimin pumpanka-pek-neme nanthi truck-yu, mu nhinhi-ka \textbf{wurda-wa}\\
         \textsc{clf:human} youth few-\textsc{other} \textsc{3du.sbj.go(6).nfut-}destroy\textsc{{}-pc.m} \textsc{clf:thing} truck-\textsc{dm} \textsc{but} \textsc{2sg-top} no\textsc{{}-emph}\\
    \glt ‘The other boys destroyed the truck, but you didn’t.’ (\citealt{FordMcCormack2011}: 1)
\z 


\subsection{Miscellaneous aspects of negation} \label{sec:mur:2.6.2}

Murrinhpatha also has an apprehensive, or ‘lest’ construction, which is encoded with the suffix \textit{-nukun} ‘lest’ added to a verb inflected with the future irrealis. These constructions are not negative constructions, \textit{per se}, but are related to the domain of negation in that they express undesirable consequences that the speaker believes should be avoided. These constructions are different to those discussed in \sectref{sec:mur:2.1}–\sectref{sec:mur:2.3} since they don’t include a negative particle of any kind. Some examples are given in (\ref{ex:mur:79}--\ref{ex:mur:81}): 

\ea\label{ex:mur:79}
    \gll \textbf{pubu}{}-wath-\textbf{nukun}\\
         \textsc{1incl.sbj.17.fut:irr}{}-fall-\textsc{lest}\\
    \glt ‘We might fall down.’ (20110825\_JB\_video\_GHYM100\_01)
\z 

\ea\label{ex:mur:80}
    \gll \textbf{thirra}{}-pe-warda \textbf{ma}{}-nhi-ngkangath-\textbf{nukun} \\
         \textsc{2sg.sbj.stand(3).fut:irr}{}-care.for-\textsc{temp} \textsc{3sg.sbj.hands(8).fut:irr-2sg.obj-}hide\textsc{{}-lest} \\
    \glt ‘Look after her, she might get lost.’ (MP-20100916-RN01)
\z 

\ea\label{ex:mur:81}
    \gll \textbf{ke}{}-nhi-bath-\textbf{nukun}!\\
         \textsc{3sg.sbj.poke:rr(21).fut:irr}\textsc{-2sg.obj-}cook\textsc{{}-lest}\\
    \glt ‘It might burn you!’ (RN20070608-002:035)
\z 

It is possible to add -\textit{nukun} to an NP to create a non-verbal clause expressing an undesirable consequence. The NP+\textit{nukun} encodes the fact that something bad might happen related to the NP. The hearer draws on pragmatics and context to interpret the action that is being negated or cautioned against. These constructions may be imperative \REF{ex:mur:82} or declarative (\ref{ex:mur:83}--\ref{ex:mur:85}):

\ea\label{ex:mur:82}
    \gll mange-nukun!\\
         hand-\textsc{lest}\\
    \glt ‘Look out for your hand (in the car door)! (e.g. it might get jammed)’ (RN, overheard)
\z 

\ea\label{ex:mur:83}
    \gll jection-nukun!\\
         injection-\textsc{lest}\\
    \glt ‘[I’m not coming because] (she) might (give me) an injection!’ (RN, overheard)
\z 

\ea\label{ex:mur:84}
    \gll ku tharnkin-nukun\\
          \textsc{clf:anim} king\_brown\_snake-\textsc{lest} \\
    \glt ‘A King Brown might (come out after us … and we’ll have to get away from it).’ (JB20040808JB03b1.txt)
\z 

\ea\label{ex:mur:85}
    \gll ngay-ka wakay mayern-nukun-warda Nungalinya\\
         1\textsc{sg-top} finish track-\textsc{lest-temp} place\_name\\
    \glt ‘My (money) might all get used up on the trip to Nungalinya.’ (JB20090707JBvid04.wav)
\z 

\section{Summary} \label{sec:mur:5}


\begin{table}
\begin{tabularx}{\textwidth}{p{3cm}Q}
\lsptoprule
{Negation strategy} & {Function} \\
\midrule 
 {\textit{mere}:\newline  negativeparticle} & Standard negation, verbs marked with irrealis mood (\sectref{sec:mur:2.1})\\
\tablevspace
& Negative imperatives (\sectref{sec:mur:2.2})\\
\tablevspace
& Negating pronouns in equative and proper inclusion verbless clauses (\sectref{sec:mur:2.3.1})\\
\tablevspace
& Negative attributive clauses (\sectref{sec:mur:2.3.2})\\
\midrule 
 {\textit{manangka}:\newline negative particle} & Standard negation, verbs marked with irrealis mood (\sectref{sec:mur:2.1})\\
\tablevspace
& Negative imperatives (\sectref{sec:mur:2.2})\\
\tablevspace
& Negating nouns in equative and proper inclusion verbless clauses (\sectref{sec:mur:2.3.1})\\
\tablevspace
& Negative attributive clauses (\sectref{sec:mur:2.3.2})\\
\midrule 
 \mbox{{\textit{mere} ‘\textsc{neg}’ + \textit{{}-nukun}}} ‘\textsc{lest}’ suffix & Emphatic negation (\sectref{sec:mur:4.4})\\
& prohibition (\sectref{sec:mur:2.2})\\
\midrule 
{\textit{pirda}: interjection} & {‘Don’t!, Stop it!’ (\sectref{sec:mur:2.2}})\\
\tablevspace
{negative classifier predicates} & Negating existential, locative, possessive clauses (\sectref{sec:mur:2.3.3})\\
\tablevspace
{\textit{mere/manangka} \mbox{‘\textsc{neg}’ + \textit{ngarra} ‘what’}} & {‘never, can’t’ (\sectref{sec:mur:3.2}})\\
\tablevspace
{\textit{{}-nukun} ‘\textsc{lest}’} & Encoding warnings, undesirable events, admonitives (\sectref{sec:mur:2.6.2})\\
\midrule 
 {\textit{wurda}:\newline  negative word, ‘no’} & Negative replies (\sectref{sec:mur:3.1})\\
& Negative predicate in verbless clauses (\sectref{sec:mur:2.6.1})\\
\midrule 
{\textit{awu}:\newline  negative word, ‘no’} & {Negative replies, negation of expectation (\sectref{sec:mur:3.1}})\\
\lspbottomrule
\end{tabularx}
    \caption{Murrinhpatha negative constructions and their functions}
    \label{tab:mur:3}
\end{table}

The discussion above covers the key aspects of negation and negative construction types in Murrinhpatha, based on our current knowledge about the language. The set of negative constructions and their functions are summarized in \tabref{tab:mur:3}.
Murrinhpatha contributes to our typologies of negation in a number of ways, most notably in the existence of negative classifier predicates, which have not previously been described for any language in the typological literature (\sectref{sec:mur:2.3.3}).

\section*{Acknowledgements}

My fieldwork on Murrinhpatha has been supported by a number of Australian Research Council grants (DP110100961, DP0984419) and by the ARC Centre of Excellence for the Dynamics of Language (CE140100041). I am extraordinarily grateful to all members of the Wadeye community who have welcomed and worked with me over many years, but especially Carmelita Perdjert without whom my work with Murrinhpatha would have been extremely limited and nowhere near as enjoyable. I would also like to thank John Mansfield for being my sounding board on all things Murrinhpatha, Bill Forshaw for help with data questions, two anonymous reviewers, and Matti Miestamo and Ljuba Veselinova for inviting me to contribute to the Negation workshop at the SWL conference in 2018 from which this paper developed, and for many helpful suggestions on earlier drafts. 

\section*{Abbreviations}
\begin{tabularx}{.45\textwidth}{lQ}
1 & first person \\
2 & second person \\
3 & third person \\
\scshape anim & animate \\ 
\scshape appl & applicative \\
\scshape assoc & associative\\
\scshape clf & classifier\\
\scshape dat & dative \\
\scshape dm & discourse \\
\scshape du & dual\\
\scshape emph & emphatic \\
\scshape excl & exclusive\\
\scshape ex& existential \\
\scshape f & feminine \\
\scshape fut & future\\
\scshape incl & inclusive\\
\scshape ipfv & imperfective\\
\scshape irr & irrealis\\
\scshape lang & language \\
\scshape loc & locative\\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape m & masculine \\
\scshape neg & negative\\
\scshape nfut & non-future\\
\scshape nsib & non-siblings\\
\scshape o & object\\
\scshape obl & oblique\\
\scshape pc & paucal \\
\scshape pl & plural\\
\scshape pst & past\\
\scshape rcgn & recognitional demonstrative\\
\scshape rel & relativizer\\
\scshape rr & reflexive/reciprocal \\ 
\scshape s & subject\\
\scshape sg & singular\\
\scshape sib & siblings\\
\scshape temp & temporal \\
\scshape top & topic \\
\scshape veg & vegetable \\
\\
\end{tabularx}
\smallskip

\noindent
Classifier stems are glossed with an indicative semantic gloss where possible (provided in small caps) and a number which corresponds to the traditional description of these elements (e.g. \citealt{BlytheEtAl2007}). Examples are provided in morphemic transcription.
    

\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
\end{document} 
