\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18234075}
\author{Vincent Collette\affiliation{Université du Québec à Chicoutimi}}
\title{Negation in Nakoda}
\abstract{This chapter is a description and analysis of clausal and non-clausal negation in Nakoda, a moribund Siouan language also known as Assiniboine. In Nakoda the enclitics \textit{=šį} and \textit{=ge(n)} negate verbs, adverbs, modality particles, and predicative pronouns, while the suffix \textit{-ni} negates indefinite pronouns and other quantifiers. Conjunctions, postpositions, and nouns (when they are not derived into a stative verb through zero conversion) cannot be negated. The chapter also deals with different topics including the narrow scope of negation, how to reinforce negation, negation in coordination and complex clauses, as well as negative transport. Lastly, synchronic data on negation in Nakoda also help to shed light on the historical development of Proto-Siouan negator \textit{*aši} and adversative \textit{*-šV} in daughter languages.
\keywords{negation, Nakoda, typology, Siouan, diachrony}
}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 

\section{The language}\label{sec:nak:1}

The Nakoda or Nakota language (ISO: \href{https://iso639-3.sil.org/code/asb}{asb}, Glottocode: \href{https://glottolog.org/resource/languoid/id/assi1247}{assi1247}) -- also known as Assiniboine -- is a Siouan language of the Mississippi Valley Siouan branch spoken in southern Saskatchewan (Canada) and northern Montana (USA). Closely related languages of the Dakotan branch includes Dakota, Lakota and Stoney. Although Nakoda is relatively homogeneous, there are two subdialects which reflect minor lexical and phonological differences between the western communities (i.e. Carry The Kettle, Mosquito / Grizzly Bear Head / Leanman, Canada; Fort Belknap, USA) and the eastern communities (i.e. Pheasant Rump, White Bear, Ocean Man, Canada; Fort Peck, USA) (\citealt{ParksDeMallie1992} and \citealt{Collette2022}). 

Nakoda is now seriously endangered and has less than 100 speakers in Montana and Saskatchewan combined. While the language is more viable in Montana where there are still some fluent speakers and a good deal of advanced learners, in Canada the number of fluent speakers does not amount to more than ten, but efforts are being made to teach the language at the local and university levels. The situation can be summarized as such: 1) All competent speakers of Nakoda (seventy years old or more) are bilingual in English and some have rudiments of Dakota, and Plains Cree;\footnote{In the late 19\textsuperscript{th} and early 20\textsuperscript{th} century, Nakoda was still the dominant language in many Canadian reservations but began to lose ground in favor of Plains Cree or English. \citet[447]{McLeod2000} reports, for example, that in Mosquito/Grizzly Bear’s Head/Leanman Reserve (located in the Battleford area in Saskatchewan), the migration of three Plains Cree women from the nearby Red Pheasant Reserve at the turn of the 20\textsuperscript{th} century weakened Nakoda language transmission since the children of these mixed couples learned Plains Cree from their mother despite the fact that their fathers spoke Nakoda.} 2) While Nakoda has diminished communicative viability (e.g. greetings, simple commands), it is nevertheless widely used by some elderly speakers in ceremonies, greetings, songs and prayers, and constitutes the backbone of Nakoda culture and ethnic identity. 

Some typological features of Nakoda are provided here (based on \citealt[viii–ix]{Cumberland2005}; \citealt[40]{Ullrich2018}).
Nakoda is a left-branching, head-marking, and strictly verb final language (SOV), with the exception of post-verbal enclitics. It has no morphological marking of tense, no definite article and no nominal case markers. Core arguments are indexed by means of affixes on the verb, and can optionally be expressed by NPs. Nakoda is a polysynthetic language with an active/stative indexing system (or split intransitivity) that surfaces in the 1\textsuperscript{st} and 2\textsuperscript{nd} persons only; active verbs take volitional agent marking (e.g. \textit{-wa-} ‘I’; \textit{-ma}- ‘me’), while stative verbs take non-volitional patient marking (\textit{-ma}- ‘I’). 3\textsuperscript{rd} person singular is not overtly marked, while 3\textsuperscript{rd} person plural is marked for animate agent (\textit{=bi}) or animate patient (\textit{-wįca-}). The inflectional morphology for pronominal indexes is prefixal or infixal, while negation, aspects and modalities, and sentence types (declarative, interrogative, imperative) are encoded by verbal enclitics and post-verbal particles. 

In this paper I use data from my fieldwork on the Western dialect spoken in Carry The Kettle, textual examples drawn from a series of texts collected in Fort Belknap (\citealt{ParksDeMallie2012a, ParksDeMallie2012b}) as well as \citegen{Cumberland2005} reference grammar. Unreferenced examples stem from my fieldwork. The partly phonemic spelling system used in this chapter is known as the “Fort Belknap orthography”\footnote{The Fort Belknap spelling system uses the American Phonetic Alphabet like the subscript hook \textit{˛} for nasal vowels, \textit{c} for [ʧ\textsuperscript{h}], \textit{š} for [ʃ], \textit{ž} for [ʒ], \textit{ȟ} for [x], \textit{ǧ} for [ɣ] and the apostrophe \textit{’} for the glottal stop [ʔ] and the ejective consonants.} and is the one actually in use in many Nakoda communities. Data collection followed the questionnaire provided  by the editors \parencitetv{chapters/questionnaire}.

\section{Clausal negation}\label{sec:nak:2}

In this section I analyse five types of clausal negations: standard negation (\sectref{sec:nak:2.1}), negation in non-declaratives clauses (\sectref{sec:nak:2.2}), negation in stative clauses (\sectref{sec:nak:2.3}), negation in non-main clauses (\sectref{sec:nak:2.4}), and negative lexicalization (\sectref{sec:nak:2.5}).


\subsection{Standard negation}\label{sec:nak:2.1}

\subsubsection{Negative enclitic \textit{=šį}}\label{sec:nak:2.1.1}

In Nakoda a declarative verbal clause is negated with one of two post-verbal enclitics \textit{=šį} or \textit{=gen}. The choice between the enclitics is motivated by pragmatic factors as shown below. There are no structural differences (or constructional asymmetry) in simple declarative clauses between a negative and the corresponding affirmative, as shown in \REF{ex:nak:1}.

\ea\label{ex:nak:1}
    \ea\label{ex:nak:1a}
        \gll Maǧážu no. \\ 
             rain \textsc{male.decl} \\
        \glt ‘It is raining.’
    
    \ex\label{ex:nak:1b}
        \gll Maǧážu=šį no. \\
             rain=\textsc{neg} \textsc{male.decl}\\
        \glt ‘It is not raining.’
    \z
\z 

\tabref{tab:nak:1} displays the post-verbal position template for enclitics (1--8) and particles (9--11) (adapted from \citealt[310]{Cumberland2005}).\footnote{My placement of enclitics and particles differs from that of \citet[310]{Cumberland2005}. First, the enclitics \textit{=ȟ}, \textit{=ȟtįyą}, \textit{=š}, and \textit{=s} have been included since they occur after the negator \textit{=šį} so an extra slot has been inserted. Second, \textit{hųštá} ‘hearsay’ (originally of position 11) can be assigned to position 10 since both the gender-neutral declarative \textit{c} and the female declarative \textit{’} can occur after \textit{hųštá}. There is no evidence to slot \textit{c} ‘declarative’ in its own terminal position and on semantic grounds it should occur in position 11 with the rest of the sentence type particles.} An example of a verb that shows three enclitics and a male declarative particle is: \textit{ec\'{ų}=bį\textsubscript{3}=kte\textsubscript{6}=šį\textsubscript{7} no\textsubscript{11}} ‘they will not do it’.

\begin{sidewaystable}
\footnotesize
\begin{tabularx}{\textwidth}{@{}l|Q|Q|Q|Q|Q|Q|Q|Q|Q|Q|Q}
%     \lsptoprule
        {STEM} & {1} & {2} & {3} & {4} & {5} & {6} & {7} &  {8} & {9} & {10} & {11} \\
        \hline
        & {\itshape =hą} \newline ‘conti\-nuative aspect’
        & {\itshape =ːga}  \newline ‘durative aspect’ 
        & {\itshape =bi} \newline ‘\textsc{3pl} animate’
        & {\itshape =na} \newline ‘dimin\-utive’
        & {\itshape =s’a} \newline ‘habitual aspect’ 
        & {\itshape =kta} \newline ‘potential mood’ (\textsc{pot)} 
        & {\itshape =šį \newline =gen} ‘negation’
%     \lspbottomrule
% \end{tabularx}
% \begin{tabularx}{\textwidth}{QlQQ}
%     \lsptoprule
        & \textit{=ȟ}\newline ‘inten\-sifier, focus’\medskip

            \textit{=ȟtįyą}\newline ‘augmen\-tative, emphasizer’ focus’\medskip

            \textit{=š}\newline ‘adver\-sative’ focus’\medskip

            \textit{=s}\newline ‘hortative’
        & \textit{jé’e}\newline ‘always’
        & \textit{cá, céyaga}
            \textit{otí’įga}\newline ‘deontic, epistemic modals’  focus’\medskip

            \textit{hųštá}\newline ‘hearsay’
        & \textit{he}\newline ‘inter\-rogative’ focus’\medskip

            \textit{no}\newline ‘declarative (male)’  focus’\medskip

            \textit{c}\newline ‘declarative’ focus’\medskip

            \textit{’}\newline ‘declarative (female)’ focus’\medskip

            \textit{wo}\newline ‘imperative \textsc{sg} (male)’\\
%     \lspbottomrule
\end{tabularx}

\caption{Template positions for post-verbal enclitics and particles}
\label{tab:nak:1}
\end{sidewaystable}


\newpage%added to keep the order text - table - text - examples
The ordering of position 2, 3, and 5 enclitics compared to that of the negative enclitic \textit{=šį} (position 7) is shown in the following examples \REF{ex:nak:2} and \REF{ex:nak:3}.

\ea\label{ex:nak:2} 
    \gll Ná=bį=kte=šį. \\
        2\textsc{sbj}.go.there=\textsc{pl=pot=neg} \\ 
    \glt ‘You all will not go there.’\footnote{In this article I indicate regressive phonetic nasalization, so an enclitic like \textit{=bi} ‘animate plural’ can appear as \textit{bi} or as \textit{bį} if followed by an adjacent or non-adjacent nasal vowel.}
\z 

\ea\label{ex:nak:3}
    \gll (…) ž-e-Ø-Ø-yá=ːga=bi=šį \\
        {} that-\textsc{dst-3p-3a}-say=\textsc{dur=pl=neg} \\
    \glt ‘(…) they haven’t been saying that’\footnote{Nakoda verb stem can be split into two parts by person indexes, and the initial part is usually shorter than the final one. \textsc{dst} stands for the shortest part of a discontinuous stem. It is not a root since it does not have any meaning in itself.} (\citealt[33]{ParksDeMallie2012b})
\z 

Textual examples as well as elicited data contain no paradigmatic asymmetry since the negative \textit{=šį} occurs with TAM (tense-aspect-mood) enclitics of all positions, as well as with person/number, and the diminutive. However, the data reveal some templatic re-ordering when the negative \textit{=šį}\textsubscript{7} co-occurs with the aspectual enclitics \textit{=hą}\textsubscript{1} ‘continuative aspect’ \REF{ex:nak:4}, and \textit{=s’a}\textsubscript{5} ‘habitual aspect’ (\ref{ex:nak:5a} vs \ref{ex:nak:5b}). The enclitic \textit{=bi}\textsubscript{3} ‘\textsc{3}\textsuperscript{rd} animate plural’ also triggers re-ordering\footnote{The following sentence provides evidence of the templatic position of \textit{=hą}\textsubscript{1} (when occurring with \textit{=ːga}\textsubscript{2}), as well as the shift of position for \textit{=bi}\textsubscript{3} ‘3\textsuperscript{rd} animate plural’:

\ea\label{ex:nak:fn}
    \gll \'{E}cen   Ø-yá=bi=hąː=ga. \\ 
         so \textsc{3sbj}-go=\textsc{pl=cont=dur} \\
    \glt ‘So they kept going.’ \citep[315]{Cumberland2005}.
\z 
}

\ea\label{ex:nak:4}
    \gll Duwé į-Ø-Ø-knįge=šį\textsubscript{7}=hą\textsubscript{1}. \\
         people \textsc{dst-3p-3a}-mind=\textsc{neg=cont} \\
    \glt ‘People didn’t mind about her.’
\z 
 
\ea\label{ex:nak:5}
    \ea\label{ex:nak:5a}
        \gll Dóhą-ni Ø-basí=s’e\textsubscript{5}=šį\textsubscript{7}. \\
             when-\textsc{neg} \textsc{3sbj}-drive=\textsc{hab=neg} \\
        \glt ‘She never drives.’ \citep[320]{Cumberland2005}
    
    \ex\label{ex:nak:5b}
        \gll Wįc\'{ą}=bi wa-’á’i’-Ø-a=bi\textsubscript{3}=šį\textsubscript{7}=s’a\textsubscript{5}. \\
             man=\textsc{pl} \textsc{indf}-gossip-\textsc{3sbj-dst}=\textsc{pl=neg=hab} \\
        \glt ‘Men don’t usually gossip.’
    \z
\z 

Some speakers have the expected order \REF{ex:nak:5a}, while others do not \REF{ex:nak:5b}. One of my consultants rejected the re-ordering in \REF{ex:nak:5b} but \citet[171--174]{McBride2016} reports similar re-ordering between the negative enclitic and some aspectual markers in Kansa (a Siouan language of the Dhegiha branch). I do not know if violation of the basic order is motivated by dialectal, idiolectal, or chronolectal differences, elicitation context, or pragmatic or semantic factors like intensification of a preceeding enclitic (see \sectref{sec:nak:4.4}). Nevertheless, the fact that this phenomenon is widely attested in other Siouan languages may indicate that strict templatic ordering might not be the best way to analyse Siouan grammatical morphology \citep[see][186]{RankinEtAl2003}.

Standard negation is also used with three types of monoclausal complex verbsː auxiliary, modal, and adverbial constructions. Monoclausal complex verbs form a single semantic unit that contains a main verb or primary predicate (either an auxiliary or a modal verb) preceded by a complement verb or secondary predicate. These are labelled “compound verbs” by \citet{BoasDeloria1941} and \citet[391--398]{Cumberland2005}, and “secondary predication” by \citet[122]{Ullrich2018}. 

As indicated by \citet[399]{Cumberland2005}, in auxiliary verb constructions, the auxiliary is the head of the phrase and it is inflected for person, number, aspect, modality, and negation. The complement verb, on the other hand, is non-finite and functions as the semantic core of the construction which is modified by the auxiliary verb. The most common auxiliaries areː \textit{\'{ų}} ‘be continuously doing; wandering around’ \REF{ex:nak:6a}; \textit{áya} 1) ‘become (stative); 2) ‘continue doing (active)’ \REF{ex:nak:6b}; \textit{iyáya} ‘become gradually’; and \textit{yągá {\textasciitilde} yįgá} ‘(sit) continuously; keep on’ \REF{ex:nak:6c}. In auxiliary verb constructions, only the auxiliary verb can be negated.

\ea\label{ex:nak:6}
    \ea\label{ex:nak:6a}
        \gll \k{I}ȟá Ø-\'{ų}=bi=šį. \\
             laugh \textsc{3sbj}-be.continuously=\textsc{pl=neg} \\
        \glt ‘They were not continuously laughing.’
    
    \ex\label{ex:nak:6b}
        \gll Nod\'{\k{ı}}n t’á a-Ø-yé=šį. \\
             hunger die \textsc{dst-3sbj}-become=\textsc{neg} \\
        \glt ‘He is not that hungry.’ (lit. ‘He is not starting to die from hunger.’) 
        
    \ex\label{ex:nak:6c}
        \gll \k{I}dé-ša-ya i-Ø-yáye=šį. \\
             face-red-\textsc{adv} \textsc{dst-3sbj}-gradually=\textsc{neg} \\
        \glt ‘He didn’t blush.’
    \z
\z 
 
Modal verbs can also function as primary predicate. As shown by \citet[392]{Cumberland2005} modal verbs have the following featuresː 1) they express degrees of certainty, ability, permission, and obligation; 2) they require a complement marker either \textit{=bi} (similar in function to English infinitival \textit{to}) or \textit{=kta} ‘\textsc{pot} (potential modality)’ or none in some cases (e.g. \textit{ogíhi} ‘be able’); and 3) the subject of the complement verb may be identical to that of the main verb, but often times the implied subject of the complement verb is omitted. As with the auxiliaries only modality verbs like \textit{cįgá} ‘want’ \REF{ex:nak:7} and \textit{snokyá} ‘know’ \REF{ex:nak:8}, \textit{ogíhi} ‘be able’ \REF{ex:nak:9}, and \textit{šk\'{ą}} ‘try’ \REF{ex:nak:10} can be negated (e.g. *\textit{ec\'{ų}=kte=šį šk\'{ą}} ‘try not to do that’).

\ea\label{ex:nak:7}
    \ea\label{ex:nak:7a}
        \gll Eyá=bi Ø-wa-c\'{\k{ı}}ge=šį. \\
             say=\textsc{comp} \textsc{3p-1a}-want=\textsc{neg} \\
        \glt ‘I don’t want/like to say it.’
    
    \ex\label{ex:nak:7b}
        \gll Téhą yįgá=bi Ø-wa-c\'{\k{ı}}ge=šį. \\
             long.time sit=\textsc{comp} \textsc{3p-1a}-want=\textsc{neg} \\
        \glt ‘I don’t want/like to wait for too long.’
    \z
\z 

\ea\label{ex:nak:8}
    \gll Dóken ec\'{ų}=bi snok-Ø-wá-ye=šį. \\
         how do=\textsc{comp} know-\textsc{3p-1a-dst=neg} \\
    \glt ‘I don’t know how to do it.’
\z 

\ea\label{ex:nak:9}
    \gll Maní o-Ø-Ø-gíhi=bi=šį. \\
         walk \textsc{dst}-\textsc{3p-3a}-able=\textsc{pl=neg} \\
    \glt ‘They cannot walk.’
\z 

\ea\label{ex:nak:10}
    \gll Ec\'{ų}=kta šk\'{ą}=šį. \\
         do=\textsc{pot} try\textsc{.2.imp=neg} \\
    \glt ‘Don’t try to do that.’
\z 

The last type of complex monoclausal verb construction is those involving a main verb and what \citet[406]{Cumberland2005} labels “adverbial verb complements” (see also \citealt[122--125]{Ullrich2018}). The morphosyntactic properties of this complex construction are as followsː 1) the verbal complement is non-finite (although it can be negated); 2) it often lacks the final stem-forming \textit{-a} (resulting in the devoicing of voiced consonants); 3) both the verbs describe separate actions which combine to form a single event. Active verbs occur in secondary predication as in \REF{ex:nak:11} which is a simultaneous predicate construction \citep[273--274]{Ullrich2018}. The verbs are adjacent and not truncated, and they share minimally one argument. When the negator \textit{=šį} is added to the verbal complement the construction is translated as ‘do X without Y’. The complement verb is non-finite and no 1\textsuperscript{st} or 2\textsuperscript{nd} person markers can be added.

\ea\label{ex:nak:11}
    \ea\label{ex:nak:11a}
        \gll Wa’óyabi k-nuhé=šį Ø-hí. \\
             book \textsc{poss}-have=\textsc{neg} \textsc{3sbj}-arrive \\
        \glt ‘She arrived without her book.’
    
    \ex\label{ex:nak:11b}
        \gll Yuȟ’ú=šį Ø-Ø-yúda. \\
             peel=\textsc{neg} \textsc{3p-3a}-eat.it. \\
        \glt ‘He is eating it without peeling it.’
    \z
\z 

In \REF{ex:nak:12} the negated verb \textit{dágeyešį} ‘s/he does not say a thing’ can be made into a secondary predicate by adding the adverbial morpheme \textit{-ȟ}. My corpus also contains examples as in \REF{ex:nak:13d} where the negator \textit{=šį} is positioned before the derivational morpheme \textit{-yą} indicating that it has scope only over the derived modifier \textit{teȟíyą} ‘with difficulty’ and not over the primary predicate \textit{mawáni} \REF{ex:nak:13e}.

\ea\label{ex:nak:12}
    \gll Gá wįc\'{\k{ı}}jana žé Ø-giktá, h\'{\k{ı}}kna pam\'{ą}knena-ȟ, dág-eye=šį=ȟ Ø-yągá.\\
         then girl that \textsc{3sbj}-get.up and head.bowed-\textsc{adv} thing-say=\textsc{neg=adv} \textsc{3sbj}-sit \\
    \glt ‘Then this girl got up and with her head bowed, she sat silently.’ \citep[52]{ParksDeMallie2012a}
\z 

\ea\label{ex:nak:13} 
    \ea\label{ex:nak:13a} \textit{teȟí} ‘it is difficult’ (verb) 
    \ex\label{ex:nak:13b} \textit{teȟíyą} ‘with difficulty’ (adverb)
    \ex\label{ex:nak:13c} \textit{teȟ\'{\k{ı}}šį} ‘it is easy; cheap’ (verb) (with regressive nasalization)
    \ex\label{ex:nak:13d} \textit{teȟ\'{\k{ı}}šįyą} ‘easily; without difficulty’
    \ex\label{ex:nak:13e}
        \gll \textit{Teȟ\'{\k{ı}}šį-yą} \textit{ma-wá-ni}.  \\
             difficult.\textsc{neg-adv} \textsc{dst-1sbj}-walk \\
        \glt ‘I walk easily.’
    \z
\z 

Finally, although \textit{=šį} occurs mainly on verbs and deverbal nouns (e.g. \textit{wąyága} ‘s/he sees it’ (\textsc{vb.tr})
    %\hc{vb.tr2 here}
> \textit{wąyágešį} ‘blindness’ (\textsc{n})) it can also be used on some predicative pronouns (e.g. \textit{miyé} ‘it is me, I’ > \textit{miyéšį} ‘it is not me’), adverbs (e.g. \textit{aké} ‘again’ > \textit{akéšį} ‘not again’; \textit{nahąȟ} ‘still, yet’ *\textit{nah\'{ą}ȟci} > \textit{nah\'{ą}ȟtįšį} ‘it is not yet’ (\textsc{vb.intr})), and some modality particles (e.g. \textit{céyaga} ‘should, must’ > \textit{céyagešį} ‘should not, must not’).



\subsection{Negation in non-declaratives}\label{sec:nak:2.2}

Negation in non-declarative clauses like prohibitives (i.e. negated imperative) and negated hortatives are formed by simply negating imperatives. The imperatives are formed by taking the bare stem (which corresponds on the surface with a verb inflected with \textit{-Ø}- a 3\textsuperscript{rd} person agent or subject) as such for a single addresse or adding the plural enclitic \textit{=bi} to indicate a plurality of addresses. Gendered particles are used depending on the speaker only: \textit{m}\textsubscript{11} (neutral for the speaker); \textit{wo}\textsubscript{11} (male speaker talking to a single addressee of both genders) and \textit{bo}\textsubscript{11} (male speaker talking to two addressees or more, of both genders). There are no specific female particles.

Prohibitives (or negated imperatives) are marked simply by adding the negator \textit{=šį} on the positive imperative form. Examples (\ref{ex:nak:18}) show uses of a prohibitive by female speakers and examples (\ref{ex:nak:19}) by speakers of both genders, while examples (\ref{ex:nak:20}) display the equivalent male forms. \citet[332]{Cumberland2005} notes that male imperative particles are used in relationships of respect.

\protectedex{
\ea\label{ex:nak:18}{Female speaker}
    \ea\label{ex:nak:18a}
        \gll Céye=šį! \\
             cry.2.\textsc{imp=neg} \\
        \glt ‘Don’t cry!’
    \ex\label{ex:nak:18b}
        \gll \k{I}kn\'{\k{ı}}ge=šį! \\
             mind.2.\textsc{imp=neg} \\
        \glt ‘Don’t mind it!’
    \ex\label{ex:nak:18c}
        \gll Iyúhana ec\'{ų}=bi=šį! \\
             all do.2.\textsc{imp=pl=neg} \\
        \glt ‘Don’t you all do it!’
    \z
\z 
}

\ea\label{ex:nak:19}{Gender neutral}
    \ea\label{ex:nak:19a}
        \gll “A-n\'{ą}-mi-ji-ȟma=bi=šį m,” e-Ø-yá […] \\
             {\db}\textsc{loc}-\textsc{dst-1p-ben}-hide=\textsc{pl=neg} \textsc{imp} \textsc{dst-3sbj}-say \\
        \glt ‘“Don’t [you all] hide it from me,” she said […]’ (female speaker) \citep[53]{ParksDeMallie2012b}
    \ex\label{ex:nak:19b}
        \gll Mi-c\'{\k{ı}}ja-bi, wóda m,” e-Ø-yá. \\
             1-child-\textsc{pl} eat.2\textsc{.imp} \textsc{imp} \textsc{dst}-\textsc{3sbj.}say \\
        \glt ‘“My children, eat,” he said.’ (male speaker) \citep[87]{ParksDeMallie2012b}
    \z
\z

\ea\label{ex:nak:20}{Male speaker}
    \ea\label{ex:nak:20a}
        \gll Dóhą-ni škáde=šį wo. \\
             when-\textsc{neg} play.2.\textsc{imp}=\textsc{neg} \textsc{male.imp.sg} \\
        \glt ‘Never play with it.’ 
    \ex\label{ex:nak:20b}
        \gll […] eyé=šį ca wo. \\
             {} say.2.\textsc{imp=neg} must \textsc{male.imp.sg} \\ 
        \glt ‘[...] you must not say that.’ \citep[31]{ParksDeMallie2012b}
    \ex\label{ex:nak:20c}
        \gll “Kúwa=bi=šį bo,” e-Ø-yá. \\
             bother.2.\textsc{imp=pl=neg} \textsc{male.imp.pl} \textsc{dst}-\textsc{3sbj}.say \\
        \glt ‘“Don’t bother her,” he said.’ \citep[79]{ParksDeMallie2012b}
    \ex\label{ex:nak:20d}
        \gll A-k-núštą=bi=šį, a-w\'{ą}-k-naga bo. \\
             \textsc{dst-poss}-lose=\textsc{pl=neg} over\textsc{-dst-poss}-see \textsc{male.imp.pl}\\
        \glt ‘Do not lose your own [language], protect it you all.’
    \z
\z 

The hortative construction (\textit{let’s}) is obtained by adding the prefix/infix \textit{ų(g)}- ‘dual’ and the enclitic \textit{=s} ‘hortative’ on the positive imperative form (e.g. \textit{yá} ‘s/he goes there’ > \textit{ų-yá=s} ‘let’s go there’). The negative hortatives (\textit{let’s not}) adds \textit{=šį}\textsubscript{7} before \textit{=s}\textsubscript{8}: (e.g. \textit{yé=šį} ‘s/he does not go there’ > \textit{ų-yé=šį=s} ‘let’s not go there’).\footnote{I wish to thank Tom Shawl (pers. comm. 2020) who has provided these examples. I have not been able to find textual examples of negative hortative verbs.}

\subsection{Negation in stative predications}\label{sec:nak:2.3}

According to \citet[114]{Payne1997} proper inclusion is used when an entity is identified as belonging to a particular group or class of individuals, while equative predication is when an entity (usually the subject) is coreferential to the entity expressed by the predicate nominal, such as in \textit{She} (\textsc{sbj}) \textit{is my mother}. In Nakoda proper inclusion is expressed with a denominal stative verb \REF{ex:nak:21} or with the stative verb \textit{žéca} ‘s/he/it is of that kind’ (\textsc{vb.stat}) that follows a noun which indicates a given trade or ethnic group (\ref{ex:nak:22}). Stative predications are negated by the enclitic \textit{=šį.}

\ea\label{ex:nak:21} {Proper inclusion} \\
    \gll Na-Ø-kóda=šį duká Šahí-Ø-ya. \\
         \textsc{dst-3sbj}-Nakoda=\textsc{neg} but Cree-\textsc{3sbj-dst} \\
    \glt ‘He is not Nakoda but Cree.’
\z 

\ea\label{ex:nak:22}
    \ea\label{ex:nak:22a}
        \gll Sakná žé-Ø-ca. \\ 
             Metis that\textsc{-3sbj}-be \\
        \glt ‘She is Metis.’
    
    \ex\label{ex:nak:22b}
        \gll Sakná žé-Ø-ce=šį. \\
             Metis that\textsc{-3sbj}-be=\textsc{neg}\\
        \glt ‘She is not Metis.’
    \z
\z 

Equative predication is when the subject of a verb is coreferential with a complement \textsc{np}. In Nakoda this is expressed with the copula verb \textit{é} ‘s/he/it is’ (\textsc{vb.stat}), which often occurs attached on a demonstrative as shown in \REF{ex:nak:23a} and \REF{ex:nak:24b}. 

However, with kinship nouns negation of an equative prediction is done using a negated transitive verb instead of \textit{žé’ešį} ‘that one is not’. In \REF{ex:nak:23b} the transitive kinship verb \textit{h\'{ų}guya} ‘she is his/her mother’ is built on the nominal stem \textit{h\'{ų}gu} ‘his/her mother’ on which a relational or causative suffix \textit{-ya} is added. Other transitive verbs of non-kin relation are built like this as shown in \REF{ex:nak:24b}. (Note the suppletive kinship terms for ‘mother’ in \REF{ex:nak:23}.)


\ea\label{ex:nak:23}{Equation}\\
    \ea\label{ex:nak:23a}
        \gll Žé-Ø-’e iná.  \\ 
             that-\textsc{3sbj}-be 1\textsc{sbj}.mother \\
        \glt ‘That’s my mother!’ / ‘She is my mother!’
    \ex\label{ex:nak:23b}
        \gll H\'{ų}gu-Ø-wa-ye=šį. \\
             mother\textsc{-3p-1a-rln=neg} \\
        \glt ‘I don’t have her as a mother.’ (intended: ‘She is not my mother.’)
    \z
\z 

\ea\label{ex:nak:24}
    \ea\label{ex:nak:24a}
        \gll Žé-Ø-’e koná. \\ 
             that-\textsc{3sbj}-be friend \\
        \glt ‘That’s my friend.’
    
    \ex\label{ex:nak:24b}
        \gll Koná-Ø-wa-ye=šį. \\
             friend-\textsc{3p}-\textsc{1a-rln=neg} \\
        \glt ‘I am not friends with him/her.’ (intended: ‘She is not my friend.’)
    \z
\z 

In Nakoda, attributive, locative, existential and possessive predications are all expressed verbally and negated accordingly with the standard negative enclitic \textit{=šį} as seen in (\ref{ex:nak:25b}, \ref{ex:nak:26b}). There is no distinction between permanent properties and temporary ones and no special ways to negate them.


\ea\label{ex:nak:25}{Attributive predication}\\
    \ea\label{ex:nak:25a}
        \gll Wįcá žé wa-snók-Ø-ya. \\
             man that \textsc{indf}-know-\textsc{3sbj-dst} \\
        \glt ‘That man is knowledgeable.’
    
    \ex\label{ex:nak:25b}
        \gll Wįcá žé wa-snók-Ø-ye=šį. \\
             man that \textsc{indf}-know-\textsc{3sbj-dst=neg} \\
        \glt ‘That man is not knowledgeable.’
    \z
\z 



\ea\label{ex:nak:26}{Locative predication} \\
    \ea\label{ex:nak:26a}
        \gll Mína žé cą’ówą én Ø-yįgá. \\
             knife that floor on \textsc{3sbj}-sit \\
        \glt ‘The knife is on the floor.’
    
    \ex\label{ex:nak:26b}
        \gll Mína žé cą’ówą én Ø-yįgé=šį. \\
             knife that floor on \textsc{3sbj}-sit=\textsc{neg} \\
        \glt ‘The knife is not on the floor.’ (lit. ‘The knife is not sitting on the floor.')
    \z
\z 

Predicative possession can be expressed with the intransitive verb \textit{táwa} ‘s/he owns it; his/her thing’ \REF{ex:nak:27} or with the transitive verb \textit{yuhá} ‘s/he has it; carries it’ \REF{ex:nak:28}. Both verbs are negated using the standard negative enclitic \textit{=šį}.

\ea\label{ex:nak:27}{Possessive predication}\\
    \ea\label{ex:nak:27a}
        \gll Tí-bi žé Ø-táwa. \\ 
             live-\textsc{nmlz} that \textsc{3sbj}-own \\
        \glt ‘He owns that house.’
    
    \ex\label{ex:nak:27b}
        \gll Tí-bi žé Ø-táwa=šį. \\
             live-\textsc{nmlz} that \textsc{3sbj}-own=\textsc{neg} \\
        \glt ‘He doesn’t own that house.’
    \z
\z 

\ea\label{ex:nak:28}
    \ea\label{ex:nak:28a}
        \gll Am\'{ų}giya Ø-mn-uhá. \\
             car \textsc{3p-1a}-have \\
        \glt ‘I have a car.’
    
    \ex\label{ex:nak:28b}
        \gll Am\'{ų}giya Ø-mn-uhé=šį. \\
             car \textsc{3p-1a}-have=\textsc{neg} \\
        \glt ‘I don’t have a car.’
    \z
\z 

Traditionally, the label ‘existential predication’ has been used to analyse expressions that predicate the existence of an entity in a given location or in general (i.e. non-locational). This is misleading as shown by \citet{Creissels2014} who highlights the fact that such constructions have discourse turning functions in that they introduce new referents (\textit{Look! There’s Peter.}). \citet[256]{Cumberland2005} showed that the stative verb \textit{yuk\'{ą}} ‘it exists’ (\textsc{vb.stat}) is a general existential verb that predicates the existence of an entity but with no reference to location or position. This verb inflects only for third person singular and is negated with the enclitic \textit{=šį} \REF{ex:nak:29}. Other ontological stative verbs like \textit{dágu} ‘it is a thing’ or \textit{duwé} ‘it is someone’ can also predicate existence and are negated with \textit{=šį} or \textit{=gen} \REF{ex:nak:30}. Negative locative-presentative predication (e.g. ‘there are no opossums here’) is expressed with the negative lexical verb \textit{wan\'{\k{ı}}ja} ‘there is none; s/he lacks something’ (see \sectref{sec:nak:2.5}), which is the most common way to encode negative existence.

\ea\label{ex:nak:29}{Existential predication}\\
    \ea\label{ex:nak:29a}
        \gll Mázaska Ø-yuk\'{ą}. \\ 
             money \textsc{3sbj}-be \\
        \glt ‘There is money.’
    \ex\label{ex:nak:29b}
        \gll Mázaska Ø-yuké=šį. \\
             money \textsc{3sbj}-be=\textsc{neg} \\
        \glt ‘There is no money.’
    \z
\z

\ea\label{ex:nak:30}
    \gll Ø-Dágu-gen. \\ 
         \textsc{3sbj}-thing-\textsc{neg} \\
    \glt ‘It is nothing.’
\z 

\subsection{Negation in non-main clauses}\label{sec:nak:2.4}

In Nakoda complement and subordinate clauses are nominalized clauses that function as a core argument of the verb \citep[415]{Cumberland2005}; they are introduced by a demonstrative such as \textit{žé} ‘that one’ or \textit{né} ‘this one’. As seen in the examples below the negative enclitic \textit{=šį} can occur on the complement verb \REF{ex:nak:31}, on the main verb \REF{ex:nak:32b}, or on both \REF{ex:nak:32c}.

\ea\label{ex:nak:31}
    \gll Wa-n-\'{\k{ı}}c’i-jaǧa=bi=šį žé Ø-wašté. \\
         \textsc{indf}-2\textsc{sbj-refl}-make=\textsc{pl-neg} \textsc{comp} \textsc{3sbj}-good \\
    \glt ‘It’s good you didn’t make it happen; it’s good it wasn’t your fault.’ \citep[415]{Cumberland2005}
\z 

\ea\label{ex:nak:32}
    \ea\label{ex:nak:32a}
        \gll Ø-\'{U}=kte žé snok-Ø-wá-ya. \\ 
             \textsc{3sbj}-come=\textsc{pot} \textsc{comp} know-\textsc{3p-1a-dst} \\
        \glt ‘I know that he is coming.’
    \ex\label{ex:nak:32b}
        \gll Ø-\'{U}=kte žé snok-Ø-wá-ye=šį. \\
             \textsc{3sbj}-come=\textsc{pot} \textsc{comp} know-\textsc{3p-1a-dst=neg} \\
        \glt ‘I didn’t know that he was coming.’
    \ex\label{ex:nak:32c}
        \gll Ø-\'{U}=kte=šį žé snok-Ø-wá-ye=šį. \\
             \textsc{3sbj}-come=\textsc{pot=neg} \textsc{comp} know-\textsc{3p-1a-dst=neg} \\
        \glt ‘I didn’t know that he was not coming.’
    \z
\z 

\subsection{Negative lexicalizations}\label{sec:nak:2.5}

Negative lexicalization deals with lexically idiosyncratic negatives that may fuse a negator and a verb with a positive meaning (although it is not always the case). Some senses are more prone than others to be encoded as separate lexical items. Examples of negative lexicalization in the languages of the world include ‘not yet’, ‘not know’ (e.g. English \textit{dunno}), or ‘not talk’, etc. \citep[see][137]{Veselinova2013neg.exist}. In Nakoda negative existential predication can be expressed with verbs built on the stative root \textit{-nįja}- ‘to not exist; to lack’ (\textit{n\'{\k{ı}}ja} ‘s/he/it lacks’; \textit{man\'{\k{ı}}ja} ‘I lack it’, \textit{nin\'{\k{ı}}ja} ‘you lack it’) which often occurs with locational adverbs. When the indefinite object prefix \textit{wa}- is added on \textit{-nįja}- the resulting verb negates the existence of an entity or indicates the lack of an entity (animate or inanimate) and the resulting meaning of \textit{wan\'{\k{ı}}ja} is ‘s/he/it is nothing, dead; there is none; s/he/it lacks things’, as shown in example \REF{ex:nak:33}.

\ea\label{ex:nak:33}
    \ea\label{ex:nak:33a}
        \gll Wó’ošija Ø-wa-n\'{\k{ı}}ja=bi nén. \\ 
             rattlesnake \textsc{3sbj-indf}-be.none=\textsc{pl} here \\
        \glt ‘There are no rattlesnakes here.’
    \ex\label{ex:nak:33b}
        \gll Pahá Ø-wa-n\'{\k{ı}}ja=bi. \\
             hair \textsc{3sbj-indf}-be.none=\textsc{pl} \\
        \glt ‘They are bald.’
    \ex\label{ex:nak:33c}
        \gll Wóyude Ø-wa-n\'{\k{ı}}ja. \\
             food \textsc{3sbj-indf}-be.none \\
        \glt ‘There is no food.’
    \ex\label{ex:nak:33d}
        \gll Wan\'{ą}gaš Ø-wa-n\'{\k{ı}}ja. \\
             long.ago \textsc{3sbj-indf}-be.none\\
        \glt ‘He passed away a long time ago.’
    \z
\z 

\citet[137]{Veselinova2013neg.exist} has shown that cross-linguistically negative existentials often originate from verbs meaning ‘to lack’, ‘to be absent’, ‘there is not’, ‘to be empty’, or ‘to be dead’. Theses verbs state the non-existence of an entity, rather then negating a positive predicate. The Nakoda example described above conforms to this typological generalization since \textit{wan\'{\k{ı}}ja} has no positive version and cannot be negated; instead the predicate ‘s/he/it did not die’ is expressed by negating the verb \textit{t’á} ‘to die’, \textit{t’éšį} ‘s/he/it did not die’.

\newpage
\subsection{Other clausal negation constructions}\label{sec:nak:2.1.2}
    
As shown in \tabref{tab:nak:1}, position 7 hosts another negative enclitic \textit{=gen} ({\textasciitilde}\textit{=ge}). It is used by fluent speakers although rarely in elicitation context. The enclitic \textit{=gen} also has a wide distribution as it can occur on verbs, quantifiers and indefinite pronouns (see \sectref{sec:nak:3.2}.). \citet[321--322]{Cumberland2005} discusses the negative enclitic \textit{=gen} but does not provide its pragmatic meaning, saying only that it is synonymous with \textit{=šį}. However, in elicitation contexts speakers provide slightly different translations for minimal pairs as seen in \REF{ex:nak:14}.

\ea\label{ex:nak:14}
    \ea\label{ex:nak:14a}
        \gll Ec-Ø-Ø-\'{ų}=kte=šį. \\
             do-\textsc{3p-3a-dst=pot=neg} \\
        \glt ‘He won’t do it.’
    
    \ex\label{ex:nak:14b}
        \gll Ec-Ø-Ø-\'{ų}=kte=gen. \\
             do-\textsc{3p-3a-dst=pot=neg} \\
        \glt ‘He won’t do it anyways.’
    \z
\z 


Nakoda \textit{=ge(n)} is cognate with the Lakota enclitic \textit{=ka} which, according to \citet[69]{Ullrich2011}, expresses “shock in response to something or an attitude of strong contradiction or discredit or negation, often with a touch of irony”. My consultant provided similar meanings for \textit{=gen} and explained the distinction between \textit{=šį} and \textit{=gen} in the following wayː “If someone asks you about something and you have no clue, you say \textit{snokwáyešį} ‘I don’t know’ but if s/he keeps on asking the same question then you say \textit{snokwáyegen} ‘I don’t know [and stop bothering me with it]’”. As seen in \REF{ex:nak:14b} and below the stronger emotional nuances expressed by \textit{=gen} are often translated as ‘not … anyways’ or ‘not … for what I care’ in (\ref{ex:nak:15}--\ref{ex:nak:16}).

\ea\label{ex:nak:15}
    \gll Waȟté-Ø-wa-na=gen!\\
         hate-\textsc{3p-1a}-evaluate=\textsc{neg}\\
    \glt ‘I don’t like it (and I don’t care).’ (the speaker shrugged her shoulders while uttering this sentence)
\z

\ea\label{ex:nak:16}
    \gll Táȟca wąží Ø-Ø-kté=gen.\\
         deer \textsc{indf} \textsc{3p-3a}-kill=\textsc{neg}\\
    \glt ‘He didn’t kill a deer anyways.’
\z 

Like the Lakota enclitic \textit{=ka}, Nakoda \textit{=gen} expresses a stronger negative statement than \textit{=šį} and may express sarcasm, carelessness, or surprise when faced with the unexpected. Textual examples illustrate these pragmatic effects. Example \REF{ex:nak:17} is especially revealing since the sentence starts with \textit{waȟtéšįȟtįyena}, an expression of discouragement or extreme disgust.\footnote{The
    verb \textit{waȟténašį} ‘s/he hates, dislikes sb, smth’ (\textsc{vb.tr}) which occurs in examples \REF{ex:nak:15} and \REF{ex:nak:17} is derived from the stative verb \textit{wašté} ‘s/he/it is good’ through fricative symbolism of the type \textit{š > ȟ} which is common in Siouan languages \citep[181, fn 4]{RankinEtAl2003}. That \textit{waȟté}- does not mean ‘dislike’ by itself is shown by the fact that it always requires the negative enclitic \textit{=šį} and rarely \textit{=ge(n)} as in \textit{Dágu waȟtége!} ‘It’s good for nothing!’
    }
    %\hc{vb.tr1 in footnote} 
%     \hc{@LSP see vertical space before translation because of slightly too large example}

\ea\label{ex:nak:17}
    \gll Waȟté=šį-ȟtįye-na! Mi-tágoš, w\'{\k{ı}}bazoką wé-[ci]-ji-kną=gen.\\
         bad=\textsc{neg-ints}-evaluate \textsc{1}-grandson juneberries \textsc{1a-[2p.1a]-ben}-save=\textsc{neg}\\
    \glt ‘For Pete’s sake. Grandson, I didn’t save those juneberries for you.’ (\citealt[127]{ParksDeMallie2012b})
\z

As suggested by an anonymous reviewer the negators \textit{=ka} in Lakota and \textit{=gen} in Nakoda may stem from an ancestral combination of the distal demonstrative *\textit{ka} ‘that one’ + *\textit{-rį} ‘negation’ (one of the three negatives found in Proto-Siouan). Lakota kept only \textit{=ka} while Nakoda has reflexes of both *\textit{ka} (> \textit{=ge}) and *\textit{karį} (*\textit{*kani > **keni > =gen}). The vowel discrepancy between Nakoda \textit{=g\textbf{e}(n)} and Lakota \textit{=k\textbf{a}} is due to vowel interference from the two other demonstratives \textit{né} ‘this one (near speaker)’ and \textit{žé} ‘that one (near the listener)’.

Although \textit{=gen} is unfrequently used, both orally and in narratives, in the few occurrences I could gather it seems to display paradigmatic asymmetry occurring mainly with enclitics \textit{=bi} ‘3\textsuperscript{rd} plural animate’ and \textit{=kta} ‘potential’. More textual data is needed in order to fully circumscribe the paradigmatic (a)symmetry of \textit{=gen}. 


\section{Non-clausal negation}\label{sec:nak:3}
\subsection{Negative replies}\label{sec:nak:3.1}

In Nakoda a negative reply to a polar question is expressed with the interjections \textit{hiyá} ‘no’. For positive replies the gendered interjections \textit{h\'{ą}} ‘yes’ (male and female), ‘hello’ (female) or \textit{háu} ‘yes, hello’ (male) are used. Although (\ref{ex:nak:34}--\ref{ex:nak:35}) do have the verb repeated in the answer, this is not obligatory, as shown in (\ref{ex:nak:37a}--\ref{ex:nak:37b}).

\ea\label{ex:nak:34}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:nak:34Q}
        \gll Wąží-ȟ é-Ø-n-agu?  \\ 
             one-\textsc{spec} \textsc{dst-3p-2a}-take \\
        \glt ‘Did you take one specifically?’ 
    \exi{A: }\label{ex:nak:34A}
        \gll Hiyá, dágu é-Ø-mn-agu=šį. \\
              no any \textsc{dst-3p-1a}-take=\textsc{neg} \\
        \glt ‘No, I didn’t take any.’
    \end{xlist}
\z
% \ea\label{ex:nak:34}
%     \gll Wąží-ȟ é-Ø-n-agu? Hiyá, dágu é-Ø-mn-agu=šį. \\
%          one-\textsc{spec} \textsc{dst-3p-2a}-take no any \textsc{dst-3p-1a}-take=\textsc{neg} \\
%     \glt ‘Did you take one specifically? -- No, I didn’t take any.’
% \z 

\ea\label{ex:nak:35}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:nak:35a}
        \gll Ø-Iyé=bi he Ø-ú=bi žé?  \\ 
             \textsc{3sbj}-be=\textsc{pl} \textsc{q} \textsc{3sbj}-come.here=\textsc{pl} that  \\
        \glt ‘Is it them coming?’
    \exi{A: }\label{ex:nak:35b}
        \gll H\'{ą}, Ø-iyé=bi. \\ 
             yes \textsc{3sbj}-be=\textsc{pl} \\
        \glt ‘Yes, it’s them.’
    \end{xlist}
\z 
% \ea\label{ex:nak:35}
%     \gll Ø-Iyé=bi he Ø-ú=bi žé? H\'{ą}, Ø-iyé=bi! \\ 
%          \textsc{3sbj}-be=\textsc{pl} \textsc{q} \textsc{3sbj}-come.here=\textsc{pl} that yes \textsc{3sbj}-be=\textsc{pl} \\
%     \glt ‘Is it them coming? -- Yes, it’s them.’
% \z 

In Canada, Nakoda is an endangered language that is not transmitted to children and it is spoken fluently by a handful of elderly people. Nakoda language is only taught a few hours per cycles in school and occasions to hear fluent speakers discuss in their language for more than a few seconds are rare. The fact that traditional texts rarely contain casual conversations combined with the impossibility to document natural conversations -- which are fundamental to elucidate whether Nakoda has content-based or truth-based replies to negative interrogations -- forces me to leave the question open. Data obtained from translations (\ref{ex:nak:36a}--\ref{ex:nak:36b}) seems to parallel English too closely to provide a clear picture of the situation.

\ea\label{ex:nak:36a}
    \begin{xlist}[Q: ]
    \exi{Q: }
        \gll Ø-Dąy\'{ą}=šį he?  \\
             \textsc{3sbj}-good=\textsc{neg} \textsc{q}  \\
        \glt ‘Isn’t it good?’
    \exi{A: }
        \gll Hiyá, Ø-dąy\'{ą}=šį. \\
              no \textsc{3sbj}-good=\textsc{neg} \\
        \glt ‘No, it is not good.’
    \end{xlist} 
\z 
\ea\label{ex:nak:36b}
    \begin{xlist}[Q: ]
    \exi{Q: }
        \gll Ø-Dąy\'{ą}=šį he? \\
             \textsc{3sbj}-good=\textsc{neg} \textsc{q} \\
        \glt ‘Isn’t it good?’
    \exi{A: }
        \gll H\'{ą}, Ø-dąy\'{ą}. \\
              yes \textsc{3sbj}-good\\
        \glt ‘Yes, it is good.’
    \end{xlist}
\z 

However, on comparative grounds one could expect Nakoda to align on Lakota where both \textit{háŋ} ‘yes’ and \textit{hiyá} ‘no’ have polarity reversing properties, see (\ref{ex:nak:37a}--\ref{ex:nak:37b}).

\ea\label{ex:nak:37a}
    \begin{xlist}[Q: ]
    \exi{Q: }
        \gll Ø-L-uhá šni he?   \\
            \textsc{3p-2a}-have \textsc{neg} \textsc{q}  \\
        \glt ‘Do you not have it?’ 
    \exi{A: }
        \gll Háŋ Ø-bl-uhá šni.  \\
             yes \textsc{3p-1a}-have \textsc{neg} \\
        \glt ‘No, I don’t have it. (lit. yes I don’t have it)’ 
    \end{xlist}
\z
\ea\label{ex:nak:37b}
    \begin{xlist}[Q: ]
    \exi{Q: }
        \gll Ø-Hí šni he?   \\
            \textsc{3sbj}-arrive \textsc{neg} \textsc{q} \\
        \glt ‘Did he not come?’ 
    \exi{A: }
        \gll Hiyá, Ø-hí.  \\
            no \textsc{3sbj}-arrive\\
        \glt ‘Yes, he came (lit. no, he came)’ 
    \end{xlist}
\z 



\subsection{Negative indefinite pronouns, and quantifiers}\label{sec:nak:3.2}

In this section I describe the morphology, syntax and semantics of Nakoda negative indefinite pronouns and negative quantifiers. As shown by \citet[21--22]{Haspelmath1997}, in many languages indefinite pronouns come in series and are formed with a root expressing one of the major ontological categories (\textsc{person, thing, place}, etc.) to which is added a marker of indefiniteness, as in English \textit{some-}, \textit{any-}, and \textit{no-}. Nakoda, like other Siouan languages, functions differently since indefinite pronouns do not have a dedicated indefinite morpheme like in English.  Nakoda has a weak noun/verb distinction and nouns for generic categories (e.g. \textit{duwé} ‘person’) can  be converted into stative verbs (\textit{duwé} ‘s/he is someone’; \textit{Madúwe he?} ‘Who am I?’), interrogative pronouns (\textit{duwé} ‘who’), or even indefinite pronouns (\textit{duwé} ‘someone’) with no change in the surface form. Negative indefinite pronouns are formed by adding the negative suffix \textit{-ni} {\textasciitilde}\textit{-na} for some speakers (from Proto-Siouan *\textit{rį}) on the interrogative pronouns as indicated in \tabref{tab:nak:2}.

\begin{table}
\begin{tabularx}{\textwidth}{p{30mm}QQQ@{}}
\lsptoprule
{ontological\newpage  categories} & {thing} & {person} & {place}\\
\midrule 
{noun} & \textit{dágu}\newline ‘thing, any’ & \textit{duwé}\newline ‘person, none’ & {--{}}-\\
\tablevspace
{stative verb} & \textit{dágu}\newline ‘it is something’ & \textit{duwé}\newline \mbox{‘s/he is someone’} & {--{}}-\\
\tablevspace
{interrogative\newpage  pronoun} & \textit{dágu}\newline ‘what’ & \textit{duwé}\newline ‘who’ & \textit{duktén}\newline ‘where’ (static)\\
\tablevspace
{indefinite pronoun} & \textit{dágu}\newline ‘something’ & \textit{duwé}\newline ‘someone’ & \textit{duktéȟ}\newline ‘somewhere’ \newline \textit{dokíya}\newline ‘somewhere’ \\
\tablevspace
{negative indefinite pronoun} & \textit{dágu\textbf{ni}}\newline ‘nothing, none’ (animate) & \textit{duwé\textbf{ni}}\newline ‘nobody’ & \textit{dukté\textbf{ni}}\newline ‘nowhere’ \\
\lspbottomrule
\end{tabularx}
\caption{Nakoda generic nouns and derived indefinite pronouns}
\label{tab:nak:2}
\end{table}

In this section I will treat the forms of \tabref{tab:nak:2} that can be negated, namely the stative verbs and the indefinite pronouns. As stated before \textit{dágu} ‘thing’ and \textit{duwé} ‘person’ can be used as nouns and stative verbs, but it should be added that the negative indefinite pronouns in \textit{-ni} can also be turned into stative verbs: \textit{duwé} ‘person’ (\textsc{n, vb.stat}) > \textit{duwéni} ‘noboby, no one’ (\textsc{pro}) > \textit{duwénišį} ‘there is nobody, no one’ (\textsc{vb.stat}).
Examples in (\ref{ex:nak:38}) illustrates the negative derivations with the noun/pronoun \textit{dágu} ‘thing, something’.
These negative stative verbs can also express existential predication and occur in sentence final position unlike pronouns as in (\ref{ex:nak:39}--\ref{ex:nak:40}). 


\ea\label{ex:nak:38}
    \ea\label{ex:nak:38a} dágu ‘thing, something’ (\textsc{n, pro})
    \ex\label{ex:nak:38b} dágu-ni ‘nothing, none’ (\textsc{pro})
    \ex\label{ex:nak:38e}
        \gll Dágu-ni-ȟ Ø-wa-c\'{\k{ı}}ge=šį. \\
             thing-\textsc{neg-emph} \textsc{3p-1a}-want=\textsc{neg} \\
        \glt ‘I want nothing at all.’
    \ex\label{ex:nak:38c} Ø-dágu-ni=šį ‘it is nothing’ (\textsc{vb.stat}); ‘zero’ (\textsc{n})
    \ex\label{ex:nak:38d}
        \gll Ø-Dágu-ni=šį=ȟtįyą. \\
             \textsc{3sbj}-thing-\textsc{neg=neg=emph} \\
        \glt ‘It is nothing at all.’ (\textsc{vb.stat}) 
    
    
    \z
\z

\ea\label{ex:nak:39}
    \gll \'{O}kšą a-Ø-Ø-kída=bi ųká Ø-duwé-ni=šį. \\
         around \textsc{dst-3p-3a}-look=\textsc{pl} but \textsc{3sbj}-person-\textsc{neg=neg} \\
    \glt ‘They looked around for him but there was nobody.’
\z 

\ea\label{ex:nak:40}
    \ea\label{ex:nak:40a}
        \gll Mázaska Ø-dágu-ni=šį. \\
             money \textsc{3sbj}-thing-\textsc{neg-neg} \\
        \glt ‘There is no money.’
    
    \ex\label{ex:nak:40b}
        \gll Hąwí Ø-dágu-ni=šį. \\
             moon \textsc{3sbj}-thing-\textsc{neg=neg} \\
        \glt ‘There is no moon.’
    \z
\z 

The indefinite pronouns function as an argument of the verb and can be replaced by an NP. They are negated with the suffix \textit{-ni} (or a variant \textit{-na}). The major morphosyntactic feature of negative indefinite pronouns is that the verb must obligatorily be in the negative form as in (\ref{ex:nak:41}--\ref{ex:nak:42}).


\ea\label{ex:nak:41}
    \gll Dágu-ni Ø-wa-c\'{\k{ı}}ge=šį. \\
         thing-\textsc{neg} \textsc{3p-1a}-want=\textsc{neg} \\
    \glt ‘I want nothing.’ (lit. ‘I don’t want nothing.’) 
\z 

\ea\label{ex:nak:42}
    \gll Duwé-ni Ø-hí=šį. \\
         person-\textsc{neg} \textsc{3sbj}-arrive.here=\textsc{neg} \\
    \glt ‘Nobody came.’ (lit. ‘No one did not come.’) 
\z 

Nakoda also has a construction involving the generic nouns \textit{dágu} ‘thing’ and \textit{duwé} ‘person’ and a negated clause. The meaning of these negated clauses is similar to that of \textit{dáguni} ‘nothing’ or \textit{duwéni} ‘no one, nobody’ and a negated verb and these may be in free variation (e.g. \textit{Duwéni híšį.} vs. \textit{Duwé híšį} ‘Nobody came.’) as noted by \citet[361]{Cumberland2005}. One of my consultants proposed that the difference between the two sets was due to formality, where speakers often truncated the redundant \textit{-ni} in fast speech. However, here the single English translation is misleading. Based on comparison with Lakota equivalent constructions I suggest that the indefinite negative pronouns in \textit{-ni} function as pronoun, while the generic nouns are more accurately analysed as nouns functioning as direct object of the verb as seen in (\ref{ex:nak:43}--\ref{ex:nak:45}) -- compare Lakota \textit{\textbf{tákuni} eye šni} ‘She said nothing; She didn’t say nothing’ with \textit{\textbf{táku} eye šni} ‘She didn’t say a thing’ (\citealt[284]{UllrichBlackbear2018}). This is why I translate them with a generic noun (thing, person, place).

\ea\label{ex:nak:43}
    \ea\label{ex:nak:43a}
        \gll Dágu Ø-wa-c\'{\k{ı}}ge=šį. \\
             thing \textsc{3p-1a}-want-\textsc{neg} \\
        \glt ‘I want none of it.’ (lit. ‘I do not want a thing.’)
    
    \ex\label{ex:nak:43b}
        \gll Dágu o-yápte=šį!\\
             thing \textsc{loc}-leave.over.2.\textsc{imp=neg} \\
        \glt ‘Don’t you leave food over!’ (lit. ‘Do not leave a thing over.’)
    \z
\z 

\ea\label{ex:nak:44}
    \ea\label{ex:nak:44a}
        \gll Duwé Ø-hí=šį. \\
             person \textsc{3sbj}-arrive=\textsc{neg} \\
        \glt ‘Nobody came.’ (lit. ‘People did not arrive.’)
    
    \ex\label{ex:nak:44b}
        \gll Duwé wa-’\'{\k{ı}}-Ø-yųǧe=šį. \\
             person \textsc{indf-dst-3sbj}-ask=\textsc{neg} \\
        \glt ‘No one was asking questions.’ (lit. ‘People did not ask questions.’)
    \z
\z 

\ea\label{ex:nak:45}
    \gll Dóki y\'{\k{ı}}=kte=šį. \\
         place \textsc{3sbj.}go=\textsc{pot=neg} \\
    \glt ‘He is going nowhere.’ (lit. ‘He is not going to a place.’)
\z 

When \textit{dágu} and \textit{duwé} (or in some cases the indefinites pronouns \textit{dáguni} ‘nothing’ and \textit{duwéni} ‘no one, nobody’) occur after a noun phrase in a negated clause as in \REF{ex:nak:46}, the reading of the generic noun \textit{dágu} or the indefinite negative pronoun \textit{dáguni} is quantificational ‘no, not … any’ according to \citet[361]{Cumberland2005} and \citet[112]{UllrichBlackbear2018} for the Lakota equivalents.\footnote{The Lakota indefinite pronouns \textit{tákuni} or \textit{tuwéni} occurring after a noun in a negated clause (e.g. \textit{Hokšíla tákuni waŋwičhablake šni} ‘I did not see any boys.’) are labelled “indefinite negative articles” by \citet[812]{Ullrich2011}.} Here it is important to note that \textit{dágu} and \textit{duwé} do not have this quantificational meaning in positive clauses so the construction \textit{generic noun + negated verb} could be considered as a collocation. 

\ea\label{ex:nak:46}
    \ea\label{ex:nak:46a}
        \gll Dáguškina dágu wicá-mn-uhe=šį. \\
             child any \textsc{3p.pl-1a}-have=\textsc{neg} \\
        \glt ‘I don’t have any children.’ \citep[361]{Cumberland2005}
    \ex\label{ex:nak:46b}
        \gll Mázaska dágu ų-k’ú=bi=šį. \\
             money any \textsc{1pl.sbj}-give=\textsc{pass-neg} \\
        \glt ‘We were not given any money.’
    \ex\label{ex:nak:46c}
        \gll  Žéhąc’ehą dágu cótąga Ø-yuké=šį. \\
             back.then any gun \textsc{3sbj}-be=\textsc{neg} \\
        \glt ‘Back then there weren’t any guns.’ (\citealt[78]{ParksDeMallie2012b})

    \z
\z 

A demonstrative (\textit{žé} ‘that; the’ or \textit{né} ‘this’) often occurs with count nouns (\ref{ex:nak:47}--\ref{ex:nak:48}) but this is not obligatory, as seen in \REF{ex:nak:46}. 

\ea\label{ex:nak:47}
    \gll Tasp\'{ą} žé dágu-na tem-Ø-Ø-yá=bi=šį. \\
         apple the thing-\textsc{neg} devour-\textsc{3p-3a-dst=pl=neg} \\
    \glt ‘They didn’t eat any of the apples; they ate none of the apples.’ \citep[360]{Cumberland2005}
\z 

\ea\label{ex:nak:48}
    \gll Mi-s\'{ų}ga-bi né duwé-ni-ȟ Ø-h\'{ą}ske=šį. \\
         1\textsc{sbj}-younger.sibling-\textsc{pl} this person-\textsc{neg-emph} \textsc{3sbj}-tall=\textsc{neg} \\
    \glt ‘None of my younger brothers is tall.’
\z 

Other pronouns and quantifiers can be negated with \textit{-ni} or even \textit{-ge(n)}. This is the case for \textit{umá} ‘one of two; the other(s)’ > \textit{umáni} ‘neither of them’ \REF{ex:nak:49} and the universal quantifier \textit{iyúha} ‘all; every’ > \textit{iyúhagen} ‘any of it’ \REF{ex:nak:50}.

\ea\label{ex:nak:49}
    \gll Umá-ni-ȟ dágu-ni ųg-ógini=bi=šį. \\
         both-\textsc{neg-?} thing-\textsc{neg} \textsc{1pl.sbj}-get=\textsc{pass=neg} \\
    \glt ‘Neither of us got anything.' \citep[360]{Cumberland2005}
\z 

\ea\label{ex:nak:50}
    \ea\label{ex:nak:50a}
        \gll Iyúha-gen Ø-we-ksúye=šį.  \\
             all-\textsc{neg} \textsc{3p-1a}-recall=\textsc{neg} \\
        \glt ‘I don’t remember any of it.’ (\citealt[36]{ParksDeMallie2012a})
    
    \ex\label{ex:nak:50b}
        \gll Iyúha į-Ø-yútą=bi=šį. \\
             all \textsc{ins-3sbj}-try=\textsc{pl=neg}\\
        \glt ‘None of them tried it.’
    \z
\z 

\citet[367]{Cumberland2005} lists the following positive partitives: \textit{ab\'{a}} ‘some, some of’ [count or mass]; \textit{cogán} ‘half’ [count or mass]; \textit{edáhą} ‘some, some of’ [count or mass]; \textit{dóna} ‘some, several, how many’; and \textit{wąžíȟ} ‘any one’. The last affirmative partitive \textit{wąžíȟ} ‘any one’ has a negative counterpart with \textit{-ni} built from the numeral \textit{wąží} ‘one; a’: \textit{wąží}> * \textit{wąžíni > wąžíniȟ} ‘no; not one; not a single one; neither’ \citep[362--363]{Cumberland2005}. The final suffix \textit{-ȟ} ‘augmentative’ \citep[according to][363]{Cumberland2005} is obligatory with \textit{wąžíniȟ}, but its exact semantic input is still obscure as seen in (\ref{ex:nak:51}--\ref{ex:nak:52}) and \REF{ex:nak:49} above.

\ea\label{ex:nak:51}
    \gll Wįcá-bi né wąží-ni-ȟ snok-Ø-Ø-yá=bi=šį. \\
         man-\textsc{pl} \textsc{dem} one-\textsc{neg-?} know-\textsc{3p-3a-dst=pl=neg} \\
    \glt ‘Neither of the men knew it.’
\z 

\ea\label{ex:nak:52}
    \gll Dáguški-bi-na wąží-ni-ȟ Ø=hí=šį. \\
         child-\textsc{pl-dim} one-\textsc{neg-?} \textsc{3sbj}-arrive.here=\textsc{neg} \\
    \glt ‘None of the children came.’
\z 

\subsection{Negative derivation and case marking}\label{sec:nak:3.3}

Nakoda nouns have no case markers for subject, object or any other functions. Plurality is marked with \textit{=bi} but only for animates, while inanimate nouns have no specific or obligatory marking for plurality (besides plural demonstratives or reduplication on the verb). Absentative, privative or antonymic meanings are expressed with postpositions (\ref{ex:nak:53a}--\ref{ex:nak:53d}) or with negated verbs (\ref{ex:nak:54}--\ref{ex:nak:57}).
The postposition \textit{cóna} ‘lacking, deprived of’ is the functional correspondent of the English suffix \textit{-less} and expresses a destitutive or privative meaning which derives an adverb, a derived noun or a stative verb from a simple noun. The postposition \textit{cóna} attaches directly on monosyllabic nouns (\ref{ex:nak:53a}--\ref{ex:nak:53b}) or on contracted nouns such as \textit{hąm}- (<\textit{h\'{ą}ba} ‘shoe’) \REF{ex:nak:53c}. With non-contractible nouns \textit{cóna} provokes a change of stress \REF{ex:nak:53d}. Finally, with nouns of three syllables of more \textit{cóna} carries its own accent and is written separately \REF{ex:nak:53e}.


\ea\label{ex:nak:53}
    \ea\label{ex:nak:53a} \textit{pá} ‘head’ (\textsc{n}) > \textit{pacóna} ‘s/he/it is headless’ (\textsc{vb.stat}) 
    \ex\label{ex:nak:53b} \textit{s\'{i}} ‘foot’ (\textsc{n}) > \textit{sicóna} ‘barefoot’ (\textsc{adv})
    \ex\label{ex:nak:53c} \textit{h\'{ą}ba} ‘shoe’ (\textsc{n}) > \textit{hąmcóna} ‘s/he is without moccasins’ (\textsc{vb.stat}); ‘shoeless’ (\textsc{adv})
    \ex\label{ex:nak:53d} \textit{ókne} ‘sleeve’ (\textsc{n}) > \textit{oknécona} ‘sleeveless vest’ (\textsc{n})
    \ex\label{ex:nak:53e}
        \gll Huȟnáȟyabi, as\'{ą}bi cóna wašté-Ø-wa-na. \\
             coffee milk without good-\textsc{3p-1a}-evaluate \\
        \glt ‘I like coffee without milk.’
    \z
\z

Antonymy, as expressed in English with the lexical prefix \textit{un}- (e.g. \textit{spoken} vs \textit{unspoken}), can be indicated by adding the negative enclitic \textit{=šį} on the positive version of a stative (\ref{ex:nak:54}--\ref{ex:nak:55}) or active verb \REF{ex:nak:56b}.

\ea\label{ex:nak:54} 
    \textit{ecédu} ‘in a common correct way’ (\textsc{vb.stat}) \newline > \textit{ecédušį} ‘in an uncommon way, not the way it is supposed to be’ (\textsc{adv})
\z 

\ea\label{ex:nak:55}
    \gll Huȟnáȟyabi skúye=šį žé Ø-mn-áktą. \\
         coffee sweet=\textsc{neg} \textsc{dem} \textsc{3p-1a}-drink-\textsc{pot} \\
    \glt ‘I’ll drink coffee unsweetened.’
\z 

\ea\label{ex:nak:56}
    \ea\label{ex:nak:56a}
        \gll Wa’óyabi k-nuhá Ø-hí. \\ 
             book \textsc{poss}-have \textsc{3sbj}-arrive \\
        \glt ‘She arrived with her book.’
    
    \ex\label{ex:nak:56b}
        \gll Wa’óyabi k-nuhé=šį Ø-hí. \\
             book \textsc{poss}-have=\textsc{neg} \textsc{3sbj}-arrive \\
        \glt ‘She arrived without her book.’
    \z
\z 

Other antonyms are formed with a negative verb (\ref{ex:nak:57b}), negative verbal compounds (\ref{ex:nak:57a}), zero-derived negative verb (\ref{ex:nak:57c}). Verbs lexicalized with =\textit{šį} often have a meaning that is lexicalized and thus more compositional (\ref{ex:nak:57a}, \ref{ex:nak:58c}). The asterisk (*) as in \REF{ex:nak:57d} indicates that the intermeditate stage in the derivation is not attested synchronically or diachronically as in (\ref{ex:nak:57d}).

% \hc{here too, punctuation and capitaliazion isn't the same in all subexamples, also for 60}
\ea\label{ex:nak:57} {Derived verbs}\\
    \ea\label{ex:nak:57b} \textit{w\'{\k{ı}}yukcą} ‘s/he thinks about things.’ (\textsc{vb.intr.indf}) \newline > \textit{w\'{\k{ı}}yukcąšį} ‘s/he is bold, brave’ (\textsc{vb.intr}) (lit. ‘s/he does not think about things’);
    \ex\label{ex:nak:57a} \textit{tawácį ecédu} ‘his/her mind is right’ (\textsc{vb.stat}) \newline > \textit{tawácį ecédušį} ‘s/he is stupid’ (\textsc{vb.stat}) (lit. ‘his/her mind is not right’);

    \ex\label{ex:nak:57c} \textit{miní} ‘water’ (\textsc{n}) \newline > \textit{miníšį} ‘it is dry, arid, waterless’ (\textsc{vb.stat}) (lit. ‘it is not wet’);
    \ex\label{ex:nak:57d} *\textit{įt\'{ų}} ‘s/he tells the truth’ \newline > \textit{įt\'{ų}šį} ‘s/he is lying’ (\textsc{vb.intr}) (lit. ‘s/he is not telling the truth’).\footnote{The verb \textit{įt\'{ų}} ‘S/he tells the truth’ has never been reported for Nakoda, but must have existed in an anterior stage of the language since it occurs in \citeauthor{Riggs1992}' dictionary of Dakota  (\citeyear{Riggs1992}) where he notes that it is now obsolete.}
    \z
\z 
    %\hc{vb.intr2 in 59a (2x), vb.intr1 in 59d}

Example \REF{ex:nak:58} shows derived nouns that are more or less transparent.

\protectedex{
\ea\label{ex:nak:58} {Derived nouns}\\
    \ea\label{ex:nak:58a} \textit{dąy\'{ą}šį yįgá} ‘s/he is not well seated’ (\textsc{vb.intr}) \newline > \textit{dąy\'{ą}šį yįgá} ‘death bed’ (\textsc{n});
    \ex\label{ex:nak:58b} \textit{wąyága} ‘s/he sees sb, smth’ (\textsc{vb.tr}); ‘s/he sees’ (\textsc{vb.tr}) \newline > \textit{wąyágešį} ‘blindness’ (\textsc{n});
    \ex\label{ex:nak:58c} \textit{ecá} ‘s/he/it is of that kind’ (\textsc{vb.stat}) \newline > *\textit{ecábi} (\textsc{ni}) > \textit{ecábišį} ‘invisible spirit, ghost’ (\textsc{n});
    \ex\label{ex:nak:58d} \textit{i’á} ‘s/he speaks, talks to sb’ (\textsc{vb.tr1}) \newline > \textit{i’éšį} ‘mute’ (\textsc{n});
    \ex\label{ex:nak:58e} \textit{gicízabi} ‘they fight one another’ (\textsc{vb.intr.recp}) \newline > \textit{gicízabišį} ‘peace’ (\textsc{n});
    \ex\label{ex:nak:58f} \textit{šp\'{ą}} ‘it is cooked’ (\textsc{vb.stat}) \newline > \textit{šp\'{ą}šį} ‘it is raw’ (lit. ‘it is not cooked’) (\textsc{vb.stat}) \newline > \textit{šp\'{ą}šį yúda} ‘s/he eats it raw’ (lit. ‘S/he eats it uncooked.’) \newline > \textit{šp\'{ą}šį yúdabi} ‘watermelon’ (\textsc{n}) (lit. that which is eaten uncooked’).
    \z
\z } 
    %\hc{vb.tr2 in 60b, vb.intr3 in 60a and vb.intr1 in 60e}

\section{Other aspects of negation}\label{sec:nak:4}
\subsection{The scope of negation}\label{sec:nak:4.1}

In a negated clause the denial can be neutral (e.g. \textit{he did not go}) or it can have a narrow scope (i.e. \textit{\textbf{he} did not go} [underlying that someone else did]). There are three means of narrowing the scope of negation to a core argument or a constituent in Nakoda: with adversative pronouns, emphatic pronouns, and changes in word order. Adversative pronouns are formed by adding the suffix \textit{-š} ‘adversative’ on the pronominal stem \textit{-iye}- ‘be, self; one’s own’; \textit{m-iyé} ‘I, myself, my own’ > \textit{m-iyé-š} ‘as for me, myself, my own, I on the other hand’. \citet[173--174]{Cumberland2005} describes the suffix \textit{-š} as expressing “[…] a negative contrast between what is expected and what actually occurs.” As seen in (\ref{ex:nak:59}--\ref{ex:nak:60}) the adversative pronoun occurs before the negated verb and narrows the scope of the negation to the subject of the verb.

\ea\label{ex:nak:59}
    \gll Ø-Boh\'{ą} cén Ø-\textbf{iyé-š} \textbf{Ø}-yé=šį iyáme-\textbf{Ø}-ye=šį. Žécen cįh\'{ą}ga-bi iy\'{a}me-Ø-ya=bi. \\
         \textsc{3sbj}-swell thus \textsc{3}-self-\textsc{advs} \textsc{3sbj}-go=\textsc{neg} hunt-\textsc{3sbj}-go=\textsc{neg} then.later older.brother-\textsc{pl} hunt-\textsc{3sbj}-go=\textsc{pl} \\
    \glt ‘It was swelling so \textbf{he} did not go, \textbf{he} did not go hunting. Then later his older brothers went hunting.’ \citep[50]{Wing2002} 
\z 

\ea\label{ex:nak:60}
    \gll \textbf{M-iyé-š} dágu Ø-ká=bi žé o-Ø-wá-pe=šį. \\
         1\textsc{sbj}-self-\textsc{advs} thing \textsc{3sbj}-discuss=\textsc{pl} \textsc{dem} \textsc{dst-3p-1a}-take.part=\textsc{neg} \\
    \glt ‘As for me \textbf{I} didn’t take part in the discussion.’
\z 

Another way to narrow the scope of a negated verb is to use the emphatic pronoun \textit{\'{\k{ı}}š} ‘s/he, it, they too, as well’ or a compounded version \textit{né’įš} ‘this one too’ (with demonstrative \textit{né}) after the word under emphasis. In a clause with a negated verb these pronouns have an adversative meaning similar to the adversative pronoun \REF{ex:nak:61}.

\ea\label{ex:nak:61}
    \ea\label{ex:nak:61a}
        \gll John šųgatąga Ø-Ø-kté=šį. \\
             John horse \textsc{3p-3a}-kill=\textsc{neg} \\
        \glt ‘John killed the horse.’ (neutral scope)
    
    \ex\label{ex:nak:61b}
        \gll John \textbf{né’įš} šųgatąga Ø-Ø-kté=šį. \\
             John this-\textsc{emph} horse \textsc{3p-3a}-kill=\textsc{neg} \\
        \glt ‘\textbf{John} didn’t kill the horse.’ (narrow scope on subject)
    \z
\z

Lastly, constituent negation under a narrow scope can be conveyed by word order modification from the neutral order Adverb + Verb order \REF{ex:nak:62a} to Verb + Adverb final \REF{ex:nak:62b}.\footnote{I thank Tom Shawl (p.c. 2022) who provided examples for the last two ways of expressing narrow scope on negation.}


\ea\label{ex:nak:62}
    \ea\label{ex:nak:62a}
        \gll Hąhébi dóki Ø-y\'{\k{ı}}=kte=šį. \\
             tonight somewhere \textsc{3sbj}-go=\textsc{pot=neg} \\
        \glt ‘He is going nowhere tonight.’ (neutral scope)
    
    \ex\label{ex:nak:62b}
        \gll Mi-h\'{\k{ı}}kna ti’óda ektá Ø-yé=šį \textbf{ȟtánihą}. \\
             1-husband town to \textsc{3sbj}-go=\textsc{neg} yesterday. \\
        \glt ‘My husband didn’t go to town \textbf{yesterday}.’ (narrow scope)
    \z
\z 

\subsection{Negative polarity}

I have not been able to find cases of negative polarity in Nakoda texts, or field work. This is a topic that will require more investigation.


\subsection{Marking of NPs in the scope of negation}\label{sec:nak:4.2}

As mentioned in \sectref{sec:nak:1}, Nakoda does not have nominal case markers and all grammatical relations are indicated on the verb via pronominal affixes. Bare nouns in Nakoda can have definite or indefinite readings in English as with \REF{ex:nak:63} which has a definite meaning. However, definiteness is often expressed with a demonstrative \textit{né} this one’ or \textit{žé} ‘that one; the’ \REF{ex:nak:64}, or a numeral \textit{wąží} ‘one’ that can function as an indefinite article ‘a’ \REF{ex:nak:65}.

\ea\label{ex:nak:63}
    \gll Coníca bus-Ø-Ø-yá jé. \\ 
         dried.meat dry-\textsc{3p-3a-caus} \textsc{hab} \\
    \glt ‘He always dries the meat (under the sun).’
\z 

\ea\label{ex:nak:64}
    \gll Š\'{ų}ga žé wí a-Ø-Ø-néža. \\
         dog \textsc{dem} tent \textsc{loc-3p-3a}-urinate \\
    \glt ‘That dog urinated on the tent.’ 
\z 

\ea\label{ex:nak:65}
    \gll Búza wąží Ø-Ø-k’ú=bi. \\ 
         cat a \textsc{3p-3a}-give=\textsc{pl} \\
    \glt ‘They gave her a cat.’
\z 

In negative constructions there is a tendency to use the bare noun as in \REF{ex:nak:61b} independently with an definite \REF{ex:nak:66} or an indefinite reading \REF{ex:nak:67}.

\ea\label{ex:nak:66}
    \gll Š\'{ų}ga caš-Ø-Ø-t\'{ų}=šį. \\
         dog name-\textsc{3p-3a}-bear=\textsc{neg} \\
    \glt ‘He didn’t name the dog.’
\z 

\ea\label{ex:nak:67}
    \gll Iyécįgayena opé-Ø-wa-tų=šį. \\
         car buy-\textsc{3p-1a-dst=neg} \\
    \glt ‘I didn’t buy a car.’
\z 

With mass nouns the partitive \textit{edáhą} is optional in negative constructions as seen in (\ref{ex:nak:68b}--\ref{ex:nak:68c}) (see \citealt{Miestamo2014} for more details).


\ea\label{ex:nak:68}
    \ea\label{ex:nak:68a}
        \gll Waȟpé edáhą Ø-wá-cįga. \\
             tea some \textsc{3p-1a}-want \\
        \glt ‘I want some tea.’
    \ex\label{ex:nak:68b}
        \gll Waȟpé Ø-wá-cįge=šį. \\
             tea \textsc{3p-1a}-want=\textsc{neg} \\
        \glt ‘I don’t want tea.’
    \ex\label{ex:nak:68c}
        \gll Waȟpé edáhą Ø-wa-c\k{\'{ı}}ge=šį. \\
             tea some \textsc{3p-1a}-want=\textsc{neg} \\
        \glt ‘I don’t want tea.’
    \z
\z 


\subsection{Reinforcing negation}\label{sec:nak:4.3}

Derivational suffixes, phonetic lengthening and less frequently double negators are used to reinforce negation in Nakoda. For the sake of brevity I will concentrate on the intensifying or augmentative functions of \textit{-ȟtįyą} as shown in negative clauses. The formative \textit{-ȟtįyą} occurs as a suffix with stative and active verbs, but also on a handful of lexicalized combinations with nouns, pronouns and adverbs. When the suffix \textit{-ȟtįyą} occurs before the grammatical enclitics it functions as a derivational intensifier of gradable expressions like stative verbs. In other words, it scales upward or downward the degree of a gradable expression and enables the speaker to project his/her subjectivity on the outside world. As can be seen in \REF{ex:nak:69b} the negative \textit{=šį} has scope only over the intensifier \textit{-ȟtįyą} and not over the predicate \textit{dąy\'{ą}} ‘s/he/it is good’ (\textsc{vb.stat}).


\ea\label{ex:nak:69}
    \ea\label{ex:nak:69a}
        \gll Onówą né Ø-dąy\'{ą}=šį. \\
             song this \textsc{3sbj}-good=\textsc{neg} \\
        \glt ‘This song is not good.’
    
    \ex\label{ex:nak:69b}
        \gll Onówą né Ø-dąy\'{ą}-ȟtįye=šį. \\
             song this \textsc{3sbj}-good-\textsc{ints=neg} \\
        \glt ‘This song is really not that good.’
    \z
\z 

However, the formative \textit{=ȟtįyą} can also cliticize on active verbs. In this case it is slotted in position 8 (as seen in \tabref{tab:nak:1}) and has scope over the immediately preceding grammatical enclitic. In \REF{ex:nak:70b} for instance \textit{=ȟtįyą} intensifies, or reinforces \textit{=šį} ‘\textsc{neg}’ and is translated with adverbs such as ‘absolutely’ or ‘completely’:\footnote{Although more work needs to be done in this area of Nakoda enclitics I hypothesize that with active verbs the enclitic \textit{=ȟtįyą} does not signal intensification in the strictest sense of the term (since an active verb is not gradable), but sentence emphasis (i.e. truth-value focus). The speaker seems to imply that there are no chances, even remote ones, for the sentence not to be true. To this effect \citet[157]{deReuse1982} notes that in Lakota when the intensification enclitic \textit{=ȟča} occurs after the negator \textit{=šni}, it is dominated by the sentence and not by the verb.}

\ea\label{ex:nak:70}
    \ea\label{ex:nak:70a}
        \gll O-Ø-wá-gihį=šį. \\
             \textsc{dst-3p-1a}-able=\textsc{neg} \\
        \glt ‘I cannot do it.’
    
    \ex\label{ex:nak:70b}
        \gll O-Ø-wá-gihį=šį=ȟtįyą! \\
             \textsc{dst-3p-1a}-able=\textsc{neg=emph} \\
        \glt ‘I absolutely cannot do it!’
    \z
\z 

Example \REF{ex:nak:71} displays a set of derivatives based on the generic noun \textit{dágu} that highlights the meaning of \textit{-ȟ} and \textit{=ȟtįyą} in negative clauses. In (\ref{ex:nak:71c}, \ref{ex:nak:71e}, and \ref{ex:nak:71f}) \textit{=ȟtįyą} and \textit{-ȟ} accentuate the negative scope of the sentence but without modifiying its content. In \REF{ex:nak:71f}, \textit{=ȟtįyą} emphatically denies that the younger brothers are tall.

\protectedex{\ea\label{ex:nak:71}
    \ea\label{ex:nak:71a} \textit{duwé} ‘person, someone’ (\textsc{n, pro})
    \ex\label{ex:nak:71b} \textit{duwé-ni} ‘nobody, none’ (\textsc{pro})
    \ex\label{ex:nak:71c} \textit{duwé-ni-ȟ} ‘nobody at all; none at all’ (\textsc{pro})
    \ex\label{ex:nak:71d} \textit{Ø-duwé-ni=šį} ‘it is nobody’ (\textsc{vb.stat})
    \ex\label{ex:nak:71e} 
        \gll Ø-Duwé-ni=šį=ȟtįyą. \\
             \textsc{3sbj}-person-\textsc{neg=neg=emph} \\
        \glt ‘It is nobody at all.’ (\textsc{vb.stat})
    \ex\label{ex:nak:71f}
        \gll Mi-s\'{ų}ga-bi né duwé-ni-ȟ Ø-h\'{ą}ske=šį. \\
             1-younger.brother-\textsc{pl} this someone-\textsc{neg-emph} \textsc{3sbj}-tall=\textsc{neg} \\
        \glt ‘None of my younger brothers is tall.’
    \z
\z }

A second mechanism used to reinforce negation is the phonetic lengthening of vowels (shown below with a triangular colon (ː) after the vowel). With negative interrogative pronouns or negative adverbs built with the non-verbal negative suffix \textit{-ni}, emphatic vowel lengthening affects the vowel of the stressed syllable as seen in (\ref{ex:nak:72a}, \ref{ex:nak:72b}).

\ea\label{ex:nak:72}
    \ea\label{ex:nak:72a}
        \gll (…) duwéː-ni-ȟ dág-e-Ø-ye=šį hųštá. \\
             {} person-\textsc{neg-emph} thing-\textsc{dst}-\textsc{3sbj}-say=\textsc{neg} it.is.said \\
        \glt ‘(…) no one at all didn’t say anything, it is said.’ (\citealt[74]{ParksDeMallie2012a})
    
    \ex\label{ex:nak:72b}
        \gll Dóːhą-ni Ø-gišní=bį=kte=šį. \\
             when-\textsc{neg} \textsc{3sbj}-be.well=\textsc{pl=pot=neg} \\
        \glt ‘They will never be well.’ (\citealt[52]{ParksDeMallie2012a})
    \z
\z 

Reduplicated negation (see also \citealt[105]{BoasDeloria1941}) which consists in doubling the standard negator \textit{=šį} on verbs (\textit{=šįšį}) is also used to reinforce negation as in \REF{ex:nak:73} but it is seldom heard, although it appears on some derived adverbs: \textit{ektá} (postposition) ‘at, in, to a location or destination’ > \textit{ektášį} ‘wrongly’ > \textit{ektášįšį} ‘mistakenly’.

\ea\label{ex:nak:73}
    \gll Žé dágu Ø-wašté=šį=šį žé Ø-Ø-ké no. \\
         that thing \textsc{3sbj}-good=\textsc{neg=neg} that \textsc{3p-3a}-indicate \textsc{male.decl} \\
    \glt ‘What he talked about (lit. indicated) is really not good.’ (\citealt[115--116]{ParksDeMallie2012b}; my translation)
\z 

Lastly, as seen in \REF{ex:nak:74} (replicated from \REF{ex:nak:16}), the negative enclitic \textit{=gen} can also pragmatically reinforce a negative statement. Another example includes the interjection \textit{Waȟtéwana\textbf{gen}}! ‘I don’t like it (and I don’t care)’.

\ea\label{ex:nak:74}
    \gll Táȟca wąží Ø-Ø-kté=gen. \\ 
         deer a \textsc{3p-3a}-kill=\textsc{neg} \\ 
    \glt ‘He didn’t kill a deer anyways.’
\z 

\subsection{Negation, coordination and complex clauses}\label{sec:nak:4.4}

Negative coordination is expressed by conjoining the coordinator \textit{eštá} ‘or’ (‘nor’ in negative constuction) \REF{ex:nak:75a} and the post-verbal particle \textit{kó} ‘also, too’ which means ‘either’ when followed by a negated verb \REF{ex:nak:75b}.

\ea\label{ex:nak:75}
    \ea\label{ex:nak:75a}
        \gll W\'{\k{ı}}-Ø-yą eštá wį-Ø-cá. \\
             woman-\textsc{3sbj-dst} or man-\textsc{3sbj-dst} \\
        \glt ‘Was it a woman or a man?’
    \ex\label{ex:nak:75b}
        \gll Wašín-’i’a-bi wąží-na-ȟ eštá, wa-Ø-mn-ápi=šį na-Ø-wá-ȟ’ų=šį kó. \\
             English-speak-\textsc{nmlz} one-\textsc{neg-?} or \textsc{dst-3p-1a}-speak \textsc{dst-3p-1a}-understand=\textsc{neg} either\\
        \glt ‘I didn’t speak a word of English, I couldn’t understand or speak, either.’ (\citealt[127]{ParksDeMallie2012b})
    \z
\z 

I did not find any special constructions that express contrastive negation, except by joining two clauses (with an initial negated verb) with the coordinator \textit{duká} ‘but’, as in \REF{ex:nak:76}. This falls under standard negation.

\ea\label{ex:nak:76}
    \gll Na-Ø-kóda=šį duká Wa-Ø-šíju. \\
         \textsc{dst-3sbj}-Nakoda but \textsc{dst-3sbj}-White\\
    \glt ‘He is not Nakoda but White.’
\z 

\subsection{Miscellaneous aspects of negation}\label{sec:nak:4.5}
\subsubsection{Negative transport}\label{sec:nak:4.5.1}

Negative transport (or negative raising) is the phenomenon whereby the negator of a main clause predicate negates a subordinated clause. \citet{Horn1978} observed that predicates such has \textit{think}, \textit{believe}, \textit{should}, and \textit{want} allow negative transport, while \textit{be able}, \textit{be possible}, \textit{may}, \textit{can}, \textit{know}, \textit{cause}, and \textit{order} do not. For instance, in \textit{I think [he is not sleeping]} the negator \textit{not} negates the lower clause \textit{be sleeping} (and not \textit{I think}) while in \textit{I do not think [he is sleeping]} the negator \textit{not} negates the sleeping as well and not the thinking, although negation is expressed in the main clause. 

Examples (\ref{ex:nak:77}--\ref{ex:nak:78}) show negative transport with the verb \textit{cįgá} ‘s/he wants/likes somebody, something’ (\textsc{vb.tr})
    %\hc{vb.tr1 here}
which is used here as a modal in a compound verb (see \sectref{sec:nak:2.2} above). 

\ea\label{ex:nak:77}
    \gll Kná=bi Ø-wa-c\'{\k{ı}}ge=šį. \\
         go.back=\textsc{comp} \textsc{3p-1a}-want=\textsc{neg} \\
    \glt ‘I don’t want to go back.’
\z

\ea\label{ex:nak:78}
    \gll John Ø-yá=kte žé Ø-wa-c\'{\k{ı}}ge=šį. \\
         John \textsc{3sbj}-go.there=\textsc{pot} \textsc{comp} \textsc{3p-1a}-want=\textsc{neg} \\
    \glt ‘I don’t want John to leave.’
\z

\subsubsection{Metalinguistic negation}\label{sec:nak:4.5.2}

Metalinguistic negation refers to clauses where the speaker objects how negation is expressed, but agrees with its content (e.g. \textit{It’s not hard to do, it’s infeasible.}). Example \REF{ex:nak:79} illustrates metalinguistic negation where the negative clause has scope not over the predicate but on the numeral adverb \textit{n\'{ų}ba} ‘two’.

\ea\label{ex:nak:79}
    \gll N\'{ų}ba wįcá-Ø-yuhe=šį, dóba wįcá-Ø-yuha. \\
         two \textsc{3p.pl-3a}-have=\textsc{neg} four \textsc{3p.pl-3a}-have \\
    \glt ‘She doesn’t have two (children), she has four.’ 
\z 

As said before Nakoda is not used on a daily basis anymore in Canada (except when older speakers visit one another, and some speak only English even in that context). Because of these constraints I have not been able to collect natural utterances illustrating metalinguistic negation and the associated intricacies expressed by differences in prosody or sign language. 


\subsubsection{Diachrony of Siouan negation}\label{sec:nak:4.5.3}
Three negative morphemes have been reconstructed for Proto-Siouan \citep{RankinEtAl2015}: *\textit{-ku}, *\textit{-rį} (> Dakota, Lakota and Nakoda pronominal negator \textit{-ni}) and \mbox{\textit{*-aši}}
(> Osage \textit{aži}; Lakota \textit{=šni}; Nakoda and Stoney \textit{=šį}). \citet{RankinEtAl2015} suggest that the unexpected nasalization of \textit{i} into \textit{į} in both Nakoda and Stoney is due to the long association of \textit{=šį} with reflexes of *\textit{-rį}. While \citet[109]{BoasDeloria1941} put forth the idea that the Dakota/Lakota negative enclitic \textit{-šni} is the adversative suffix \textit{-š} on which is added the negative \textit{-ni}, a closer examination of Siouan comparative data strongly suggests that the segment \textit{š} occurring in Lakota/Dakota \textit{-šni} and the adversative suffix \textit{-š} are in fact two distinct elements. This is shown by their distinct ablauting behavior. On the basis of Dhegiha Siouan languages the initial \textit{a} of *\textit{-aši} can be analysed as part of the suffix (e.g. Osage \textit{aži}) although linguists working on Dakotan dialects segment it as a stem-final vowel. \citet[6]{Rankin2006} suggests that \textit{a} is an ablaut vowel that contracts with the stem-final vowel of verbs and ablauts to \textit{e} in Dakotan dialects, while the adversative enclitic \textit{-š} (< *\textit{-šV}) does not trigger ablaut. Thus, Dakota/Lakota \textit{=šni} combines two negators *\textit{aši} and *\textit{rį} (> *\textit{širį > šni}) just like Mandan \textit{-nix} is a fusion of two negative markers *\textit{-rį} and *\textit{-axi} (plus consonant symbolism \textit{š > x}). \citet[303]{Kasak2019} suggests the following historical development: *\textit{-rį-axi} > *\textit{*-rį-xi} > \textit{=nix}.
In sum, Nakoda data on the morphophonological behavior of these elements reinforces the conclusion that Proto-Siouan *\textit{-aši} ‘negation’ and *\textit{-šV} ‘adversative’ have different etymologies since the negative \textit{=šį} ablauts systematically (\textit{a>e}) while
the adversative \textit{=š} does not: \textit{maǧážukt\textbf{aš} otí’įga} ‘I think it might rain.’ \citep[239]{Cumberland2005} vs. \textit{maǧážukt\textbf{ešį}} ‘it will not rain’.

\section{Summary}\label{sec:nak:5}


To summarize the present chapter, in Nakoda the enclitics  \textit{=šį} and \textit{=ge(n)} negate verbs, adverbs, modality particles, and predicative pronouns, while the suffix \textit{-ni} negates indefinite pronouns and other quantifiers. Conjunctions, postpositions, and nouns (when they are not derived into a stative verb through zero conversion) cannot be negated. \tabref{tab:nak:3} gives an overview of the forms and their functions.


\begin{table}
\begin{tabularx}{\textwidth}{lQll}

\lsptoprule

\multicolumn{2}{l}{} & {  Section} & {  Example}\\
\midrule
{\itshape =šį} & { negative postclitic used with} &&\\

& {   a) clauses} && \\
& {     - simple clauses}& {\ref{sec:nak:2.1.1}} &(\ref{ex:nak:1}--\ref{ex:nak:5})\\
& {     - monoclauses (verb + auxiliary)} & {\ref{sec:nak:2.1.1}} & {\REF{ex:nak:6}} \\
& {     - monoclauses (verb + modal auxiliary)} & {\ref{sec:nak:2.1.1}} & {(\ref{ex:nak:7}--\ref{ex:nak:10})} \\
& {     - monoclauses (adverb + verb)} & {\ref{sec:nak:2.1}} & { (\ref{ex:nak:11}--\ref{ex:nak:13})} \\
& {   b) non-declarative clauses} & {\ref{sec:nak:2.2}} & {(\ref{ex:nak:18}--\ref{ex:nak:20})} \\
& {   c) stative predications} &&\\
& {      - proper inclusion}  & {\ref{sec:nak:2.3}} &  {(\ref{ex:nak:21}--\ref{ex:nak:22})}\\
& {      - equation}  & {\ref{sec:nak:2.3}} & {(\ref{ex:nak:23}--\ref{ex:nak:24})} \\
& {      - attributive predication}  & {\ref{sec:nak:2.3}} & {\REF{ex:nak:25}} \\
& {      - locative}  & {\ref{sec:nak:2.3}} & {\REF{ex:nak:26}} \\
& {      - possessive predication} & {\ref{sec:nak:2.3}} & { (\ref{ex:nak:27}--\ref{ex:nak:28})} \\
& {      - existential predication} & {\ref{sec:nak:2.3}} & {(\ref{ex:nak:29}--\ref{ex:nak:30})} \\
& {   d) non-main clauses}  & {\ref{sec:nak:2.4}} & {(\ref{ex:nak:31}--\ref{ex:nak:32})} \\
& {   e) non-clausal structures}&& \\
& {     - negative replies}  & {\ref{sec:nak:3.1}} & {(\ref{ex:nak:34}--\ref{ex:nak:36b})} \\
& {     - negative derivation} & {\ref{sec:nak:3.3}} & {(\ref{ex:nak:54}--\ref{ex:nak:58})} \\
& {   f) reinforcing negation (with the sentence emphatic \textit{=ȟtįyą})}  & {\ref{sec:nak:4.3}} & {(\ref{ex:nak:69}--\ref{ex:nak:71})} \\
& {   g) coordination and complex clauses} & {\ref{sec:nak:4.4}} & {(\ref{ex:nak:75}--\ref{ex:nak:76})} \\
& {   h) negative transport} & {\ref{sec:nak:4.5.1}} & {(\ref{ex:nak:77}--\ref{ex:nak:78})} \\
& {   i) metalinguistic negation}   & {\ref{sec:nak:4.5.2}} & {\REF{ex:nak:79}}\\

{\itshape =gen} & { negative enclitic used to pragmatically reinforce negation with monoclauses, quantifiers, and indefinite pronouns} & {\ref{sec:nak:2.1.2}} & (\ref{ex:nak:14}--\ref{ex:nak:17})\\
{\itshape -cona} & { suffix meaning ‘without; lacking’ used in negative derivation}  & {\ref{sec:nak:3.3}} & {\REF{ex:nak:53}}\\
{\itshape wan\'{į}ja} & { negative lexicalization ‘it is nothing; there is none’} & {\ref{sec:nak:2.5}} & { \REF{ex:nak:33}}\\
{\itshape -ni}  & { suffix used to negate indefinite pronouns, quantifier and partitives}  & { \ref{sec:nak:3.2}} & { (\ref{ex:nak:38}--\ref{ex:nak:52})}\\
{ \textit{-š}, \textit{\'{į}š}} & { adversative suffix and the emphatic pronoun \textit{\'{į}š} ‘too’ indicate scope relationships}  & { \ref{sec:nak:4.1}} & (\ref{ex:nak:59}--\ref{ex:nak:62})\\
\lspbottomrule
\end{tabularx}
\caption{Summary of negation in Nakoda}
\label{tab:nak:3}
\end{table}



\section*{Abbreviations}

% \aut{the glossing system was deviced to be short and sweet, now it becomes longer for nothing really

% VI2 > VB.INTR.2

% If you decide to do that, please send me the text again so I can double check.}
% \hc{We [M+L+H] suggest to delete the numbers in vb.intr2 and similar cases, because according to the general glossing principles adopted in this collective volume, numbers are used in glossing for semantic categories like person and noun class. This distinction between classes of intransitive verbs does not seem to be relevant to negation (since you do not mention it in the text), so glossing all intransitive verbs with \textsc{vb.intr} seems to be sufficient information in the context of this volume (and the same for \textsc{vb.tr}. These only occur in parentheses in the text and are marked with comments, on p8,15,23,24,30)}
% \aut{Ok, you're right. Verb classes has no effect on negation, but since it's a split intransitive lg. it is useful to know of the existence of classes in order to figure out why 'I' and 'you' have 3 different allomorphs each.}

\begin{tabularx}{.45\textwidth}{lQ}
1 & first person \\
2 & second person \\
3 & third person \\
\scshape a  & agent  \\
\scshape abstr  & abstract \\ 
\scshape adv  & adverb \\
\scshape advs  & adversative \\
\scshape ben & benefactive  \\
\scshape caus  & causative \\
\scshape comp  & complementizer \\
\scshape cont  & continuative  \\
\scshape decl  & declarative  \\
\scshape dem  & demonstrative  \\
\scshape dim  & diminutive  \\
\scshape dst & discontinous stem \\
\scshape dur  & durative  \\
\scshape emph  & emphatic  \\
\scshape hab  & habitual  \\
\scshape imp  & imperative \\
\scshape indf  & indefinite \\ 
\scshape ins  & instrumental \\
\scshape intr  & intransitive \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape ints  & intensifier \\
\scshape loc  & locative \\
\scshape male  & male speaker \\
\scshape n  & noun \\
\scshape neg & negation \\
\scshape nmlz  & nominalizer \\
\scshape p  & patient \\
\scshape pass & passive  \\ 
\scshape pl  & plural \\
\scshape poss  & possessive  \\
\scshape pot  & potential   \\
\scshape pro  & pronoun \\
\scshape q  & interrogative  \\ 
\scshape recp  & reciprocal \\ 
\scshape refl  & reflexive \\
\scshape rln  & relational \\
\scshape sbj  & subject  \\ 
\scshape sg  & singular \\
\scshape spec  & specific  \\
\scshape stat  & stative   \\
\scshape tr  & transitive  \\
\scshape vb  & verb \\
\end{tabularx}

\printbibliography[heading=subbibliography,notkeyword=this]

\end{document}
