\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233983}
\author{Ekaterina Gruzdeva\affiliation{University of Helsinki} and Maksim Fedotov}
\title{Negation in Nivkh}
\abstract{Nivkh has a full spectrum of lexical and grammatical devices for expressing clausal negation. These devices are located at different points on the scale of grammaticalization and some of them are the result of grammatical reanalysis and areal contacts. This chapter discusses various aspects of the Nivkh system of negation both from synchronic and diachronic perspectives. A clear border is drawn between the expression of negation in declarative and non-declarative clauses. Except for this distinction, negation of all types of clauses in Nivkh follows common rules and uses basically the same devices. The discussion covers negation strategies in declarative, prohibitive and apprehensive clauses, negation in stative predications, coordinated and subordinate clauses, negative lexicalizations, negative replies, negative emphatics and indefinites, as well as the means of rendering caritive semantics and antonymic low-level negation. The material represents several different Nivkh varieties, which display a considerable variation, and comes from personal field data as well as from other extant sources, including collections of folklore texts. 

\keywords{Nivkh, Nighvng, clausal negation, low-level negation, prohibitive, apprehensive}
}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 

\section{The language}\label{sec:niv:1}

Nivkh (Gilyak, Ghilyak. ISO: \href{https://iso639-3.sil.org/code/niv}{niv}, Glottocode: \href{https://glottolog.org/resource/languoid/id/gily1242}{gily1242}, \href{https://glottolog.org/resource/languoid/id/sakh1247}{sakh1247}) is a language isolate spoken on Sakhalin Island and in the Amur region of Russia.\footnote{The first version of this article was written in 2019, when Maksim Fedotov was affiliated with the Institute for Linguistic Studies, Russian Academy of Sciences.} The language is traditionally classified as “Paleosiberian”, or “Paleoasiatic”, together with a number of other languages of Siberia and the Russian Far East. As a result of historical expansion from a homeland in central Manchuria to the Lower Amur and further to Sakhalin, Nivkh has evolved as a continuum of several distinct idioms with the major split between Amur (Nivkh proper, \href{https://glottolog.org/resource/languoid/id/gily1242}{gily1242}) and Sakhalin (Nighvng, \href{https://glottolog.org/resource/languoid/id/sakh1247}{sakh1247}) varieties, which, in fact, fulfill the criteria for separate languages and are mutually hardly comprehensible. 
The Amur Nivkh comprises Amur (A), West Sakhalin (WS) and North Sakhalin (NS) varieties, whereas the Sakhalin Nighvng consists of East Sakhalin (ES), Central Sakhalin (CS) and South Sakhalin (SS, extinct) varieties. The Amur variety is spoken along the Lower Amur on the continent, while the other varieties are used on Sakhalin Island. Two other varieties covered in this paper are represented by the ES variety, which essentially differs from the A variety, and the WS variety, which can be seen as a transitional variety displaying a mixture of features typical of other varieties. Some of the given varieties can be further divided into several local variants. For this paper, the most essential division is between the Tymovsk (ES\textsuperscript{T}) and Nyivo (ES\textsuperscript{N}) variants of the East Sakhalin variety.

For a long historical period, Nivkh has been in close contact with the Tungusic languages both in the Amur area and on Sakhalin, as well as with Ainu (isolate) on Sakhalin. One may assume that during more distant historical times, Nivkh was also influenced by other, now-extinct, local languages. More recent historical ties with Japanese and Russian date back to the mid-19\textsuperscript{th} century, when Russians began to settle in the Far East of Russia and on Sakhalin.

At the moment, Nivkh is a heavily endangered language, since the entire Nivkh population has switched to Russian. The current (2022) number of fluent Nivkh speakers, who are all bilingual, is hardly more than 50, with less than 25 for the Lower Amur, Amur Liman and West Sakhalin varieties, only 2 for the North Sakhalin variety, and less than 25 for the East Sakhalin variety.

The paper is based on personal field data collected during the period of 1989–2021 in the Amur area and on Sakhalin Island (also online),\footnote{We are extremely grateful to all our consultants for their help in collecting the data on different Nivkh varieties.} as well as on other extant sources, including collections of folklore texts in different varieties. In most cases, the examples are given in their original transcription, but with an additional morphological segmentation and glossing. Throughout the paper, field examples are marked with an ``FM" (= field materials), cf. ``(Gruzdeva FM2019: 1:11)"; ``(Fedotov FM2013)".

Typologically, Nivkh is a (poly)synthetic language with a rather complex phonology and predominant suffixation. Possessor and object (including reflexive) markers are prefixed (or rather cliticized) to nouns and transitive verbs. Nivkh is well-known for its root/morpheme-initial consonant alternations, which take place at morpheme boundaries and within noun and verb phrases.\footnote{The words entering the complexes that are subject to morphological alternations are separated in the examples by “+”. For more details on Nivkh polysynthesis see \citet{Mattissen2003}. The rules of morphophonological alternations are not discussed in this paper.}
The Nivkh phonological system has been undergoing various historical changes, the most prominent one being the loss of nasals in stem-final position.

Nivkh noun morphology distinguishes two numbers (singular vs. plural), as well as eight cases in the Amur varieties and seven cases in the Sakhalin varieties. The verbs are characterized by several aspects and moods and two tenses (non-future and future). Contemporary indicative finite forms are historically nominalizations, which explains the fact that these forms (optionally) agree with the subject only in number. The full differentiation of person and number is found only in imperative (hortative, jussive) forms. There are no adjectives that would be morphologically distinct from verbs, but practically all verbal stems irrespective of their meaning may function as adnominal modifiers (= participles) that occur in preposition to the head nominal.

At the level of the simple clause Nivkh displays some features of isolating structure, especially with respect to core arguments, which are typically unmarked. Nivkh has accusative alignment. The canonical word order (for transitive clauses) is \textsc{[(Topic)} \textsc{(Subject)} \textsc{Object} \textsc{Predicate}] (SOV). All modifiers, except for numeral / quantifier constituents, precede the head nominal. Adverbial clauses are formed by converbs (= dependent forms) of different origins and are arranged in chains concluded by the main clause. The latter, in turn, is built around the clause-final finite (= independent) form. Converbs render various interclausal relations, with the finest differentiation being made within the temporal converbs. Some converbs agree with the subject in person and number similarly to aforementioned finite forms.

Nivkh has a full spectrum of lexical and grammatical devices for expressing clausal negation. These devices are located at different points on the scale of grammaticalization and some of them are the result of grammatical reanalysis and areal contacts. There is also considerable dialectal variation, which reflects historical developments and calls for diachronic explanations. A clear border is drawn between the expression of negation in declarative and non-declarative (prohibitive and apprehensive) clauses. Besides that, negation of all types of clauses in Nivkh follows some general rules and uses more or less the same devices.

\largerpage
The description of negation in Nivkh in this chapter is based on the questionnaire provided by the editors (\citetv{chapters/questionnaire}) and mirrors its structure.


\section{Clausal negation}\label{sec:niv:2} 

This section discusses various devices for expressing clausal negation in declarative clauses (\sectref{sec:niv:2.1}) and non-declarative (prohibitive and apprehensive) clauses (\sectref{sec:niv:2.2}).

\subsection{Standard negation}\label{sec:niv:2.1} 

Standard negation, which is understood as negation in a simple declarative clause containing a verb \citep{Miestamo2005}, is rendered in Nivkh by various synthetic and analytic verb forms with different histories and degrees of productivity. We will discuss the basic synthetic negative forms in \sectref{sec:niv:2.1.1} and the basic analytic negative forms in \sectref{sec:niv:2.1.2}. As will be shown, each dialectal variety shows a clear tendency towards employing one or another type.
A special synthetic negative habitual form will be discussed in \sectref{sec:niv:2.1.3}.


\subsubsection{Synthetic forms}\label{sec:niv:2.1.1}

Synthetic forms are generated by one of the following ways: (i) compounding of the lexical verb with the negative verb \textit{ʁavr-} ‘not exist’, (ii) compounding of the lexical verb with the negative verb \textit{ləɣə{}-} ‘not have’, or (iii) suffixation.

In contemporary ES, basic negative synthetic forms are derived by compounding two verbal elements. The resulting complex verb behaves like a regular verb and may undergo subsequent inflection. The first element is the nominalized form of the negated lexical verb, which historically ends in the nominalizer \textit{-ŋ}. Note that nominalizations together with a negative lexicalized verb are very frequent in the negative existential cycle, see \citet{Veselinova2014}. 
In many cases this unstable nasal has been lost in the speech of modern speakers; however, it still determines the quality of the initial consonant of the following element, namely the root of the negative verb \textit{ʁavr-} ‘not exist’. Following the rules of morphophonological alternations, the bound form of this verb, which is used in the synthetic form, has the shape \textit{-ɢavr-}. The negative verb root is synchronically perceived as a negative suffix, though its connection with the free negative verb is rather transparent (see \sectref{sec:niv:2.3} and \sectref{sec:niv:4.6.1}), cf. \REF{ex:niv:1}.\footnote{Throughout the paper, the examples are given in (quasi)phonemic transcription, made with the International Phonetic Alphabet symbols. This transcription, however, is partly conventional. First, it does not reflect dialectal differences in the phonetic realization of cognate phonemes (such as /r̥/). Second, in case of the examples taken from other authors, especially \citet{KreinovichMs}, it is rather a transliteration from their Cyrillic notation into its rough IPA correspondence.} This type of negative construction can be seen as symmetrical to the affirmative one, since the negated verb retains all its categories, including finiteness/non-finiteness. The markers of various verbal categories follow the complex verb stem, as, for example, the future tense suffix \textit{-i} in \textit{vi-ɢavr-i-d=ra} ‘[I] won’t go’ in \REF{ex:niv:2},\footnote{The causative suffix adds the causal meaning to the narrative converb and serves as a switch-reference marker.} or the iterative suffix \textit{-tʰa} in \textit{vi-ɢavr-tʰa-d-ɣun} ‘[usually we] don’t go’ in \REF{ex:niv:3}.


\ea\label{ex:niv:1}
    \ea\label{ex:niv:1a}
        \gll Ɲi ra-d.  \\ 
             1\textsc{sg} drink-\textsc{ind}  \\
        \glt ‘I drink / drank.’
    \ex\label{ex:niv:1b}
        \gll Ɲi \textbf{ra-ɢavr-d}. \\
             1\textsc{sg} drink-\textsc{neg-ind}\\
        \glt  ‘I don’t / didn’t drink.’ [ES]
    \z
\z 

\ea\label{ex:niv:2}
    \gll Ɲ{}-əmk qo-gu-t ɲi klub-roχ \textbf{vi-ɢavr-i-d=ra}. \\ 
         \textsc{1sg}{}-mother be\_ill-\textsc{caus-cvb.narr.1pl} 1\textsc{sg} club-\textsc{dat} go-\textsc{neg-fut-ind=foc} \\
    \glt ‘Because my mother is ill, I won’t go to the club.’ (Gruzdeva FМ1991) [ES\textsuperscript{T}]
\z

\ea\label{ex:niv:3}
    \gll  Kə{}-χajnapə cʰχar+mi-roχ vi-j-d-ɣun. Ajv=at \textbf{vi-ɢavr-tʰa-d-ɣun}. \\ 
         rain-\textsc{cvb.conc} tree+inside-\textsc{dat} go-\textsc{fut-ind{}-pl} usually=\textsc{foc} go-\textsc{neg-iter.1pl-ind{}-pl}\\
    \glt ‘Despite the rain, [we]’ll go to the forest. Usually we don’t go.’ (Gruzdeva FM2000: 4:141) [ES\textsuperscript{T}]
\z 

In the Amur varieties, standard negation is normally expressed by an analytic construction. However, in the future tense, the WS variety and seemingly some other Amur varieties utilize a similar synthetic strategy as in the ES variety, viz. compounding of lexical and negative verb stems, cf. \REF{ex:niv:4} and \REF{ex:niv:5}. Note that in contrast to the ES example \REF{ex:niv:2}, where the future suffix \textit{-i} is placed after the negative verb root, in these WS examples (from \citealt{Panfilov1965}), the future suffix \textit{-nə} precedes the negative verb root. This fact also defines the use of the allomorph \textit{-ʁavr-}, which in this case follows the vowel.

\ea\label{ex:niv:4}
    \ea\label{ex:niv:4a}
        \gll Qan pʰrə{}-ɟ. \\ 
             dog come-\textsc{ind} \\
        \glt ‘The dog comes / came.’
    \ex\label{ex:niv:4b}
        \gll \textbf{Qan} \textbf{pʰrə{}-nə{}-ʁavr-ɟ}. \\
             dog come-\textsc{fut-neg-ind}\\
        \glt  ‘The dog won’t come.’ [WS]
    \z
\z 

\ea\label{ex:niv:5}
    \gll  Hə+umguola r̥atx=hagin \textbf{vi-gu-nə{}-ʁavr-ɟ}. \\ 
          that+girl where=\textsc{emph} go-\textsc{caus-fut-neg-ind} \\
    \glt ‘[He] won’t let the girl go anywhere.’ \citep[160]{Panfilov1965} [A]
\z 

Another synthetic negative form exists only in A and employs the negative possessive verb \textit{ləɣə{}-} ‘not have’ (\sectref{sec:niv:2.3.2}). This verb synthesizes with the stem of the lexical verb which is followed by the emphatic suffix \textit{-Tə}\footnote{The suffix has variants \textit{-tə}, \textit{-rə}, \textit{-də}.} (\sectref{sec:niv:4.6.2}). In this form, \textit{ləɣə{}-} is fully grammaticalized as a pure negative marker and does not bear any possessive meaning. The whole construction can be seen as symmetrical to its affirmative counterpart, cf. \REF{ex:niv:6}. This synthetic construction is much rarer than the analytic form discussed below.

\ea\label{ex:niv:6}
    \gll If \textbf{pʰrə{}-ivi-də{}-ləɣə{}-ɟ=ravej}. \\ 
         \textsc{3sg} come-\textsc{prog-emph-neg-ind=foc\_emph} \\
    \glt ‘He’s not coming all the same.’ \citep[160]{Panfilov1965} [A]
\z 

\subsubsection{Analytic forms}\label{sec:niv:2.1.2}

In various Nivkh varieties there are altogether three negative analytic constructions, all of which contain a nominalized form of a negated lexical verb and a negative verb. The corresponding lexical verb can be synchronically zero-nominalized, nominalized with the suffix ES \textit{-}(\textit{n})\textit{d} or the suffix \textit{-k}.

In WS and A, standard negation is typically expressed by an analytic construction comprising a synchronically zero-nominalized lexical verb, which is marked by the dative suffix \textit{-Toχ}.\footnote{The dative suffix has the variants \textit{-toχ}, \textit{-roχ}, \textit{-doχ}. In some works on Nivkh, the final fricative of the dative suffix is transcribed as velar: \textit{-tox}, \textit{-rox}, \textit{-dox}.} Historically, this verb form ended in a nasal, which makes \textit{-doχ} the correct allomorph in this context, according to the morphophonological rules. The resulting form is immediately followed by the negative verb \textit{qʰaw{}-} ‘not exist’, cf. (\ref{ex:niv:7}--\ref{ex:niv:8}):

\ea\label{ex:niv:7}
    \ea\label{ex:niv:7a}
        \gll If pʰrə{}-ɟ. \\ 
             \textsc{3sg} come-\textsc{ind} \\
        \glt ‘S/he came.’
    \ex\label{ex:niv:7b}
        \gll If \textbf{pʰrə-doχ} \textbf{qʰaw-ɟ}. \\
             \textsc{3sg} come-\textsc{dat} \textsc{neg-ind}\\
        \glt  ‘S/he didn’t come.’ [WS]
    \z
\z

\ea\label{ex:niv:8}
    \gll If pʰ-ətək nəmr mu-jnə{}-ɟ+xrərə imŋ-doχ \textbf{pʰrə{}-doχ} \textbf{qʰau-ɟ}. \\ 
         \textsc{3sg} \textsc{refl-}father yesterday die-\textsc{des-nmlz}+because\_of \textsc{3pl}{}-\textsc{dat} come-\textsc{dat} \textsc{neg}{}-\textsc{ind}\\
    \glt ‘He didn’t come to their place (lit. ‘to them’) because of his father’s having fallen ill the day before.’ (\citealt{NedjalkovOtaina2013}: 66) [A]
\z 

The negative form can be also derived from an ideophone. In example \REF{ex:niv:9}, it consists of the ideophone \textit{poj} describing the visual image of smoke, the nominalized form of the auxiliary verb \textit{ha{}-} ‘be / do so’, and the negative verb.

\ea\label{ex:niv:9}
    \gll Tʰuf \textbf{poj} \textbf{ha-doχ} \textbf{qʰau-ɟ}. \\ 
         smoke curl do\_so-\textsc{dat} \textsc{neg}{}-\textsc{ind} \\
    \glt ‘The smoke doesn’t curl.’ \citep[222]{Panfilov1965} [A]
\z 

As can be seen from examples (\ref{ex:niv:7}--\ref{ex:niv:9}), in declarative epistemically neutral sentences the negative verb \textit{qʰaw{}-} is typically used in the indicative mood, which is marked by the suffix A \textit{-ɟ} or \textit{-c}. In the same context, the negative verb can also be used in the nominalized form in \textit{-k} (an agentive nominalizer), cf. \REF{ex:niv:10}.
    
\ea\label{ex:niv:10}
    \gll Mu+bəvə{}-də{}-ləɣə{}-t ɲən \textbf{vi-nə{}-dox} \textbf{qʰau-k=ra}. \\ 
         boat+repair-\textsc{emph-}not\_have-\textsc{cvb.narr.1pl} 1\textsc{pl}.\textsc{excl} go-\textsc{fut-dat} \textsc{neg-nmlz.agt=foc}\\
    \glt ‘Without repairing the boat, we won’t go.’ \citep[160]{Panfilov1965} [A]
\z 

In the speech of the modern speakers of the LA and WS varieties, the negative analytic construction has undergone further changes, so that the dative suffix is nowadays pronounced as \textit{-to} or \textit{-ta} or even \textit{-tə} rather than \textit{-doχ}. The \textit{-d} > \textit{-t} change is in line with the general process of devoicing of plosives, which is attested in these varieties, whereas the loss of final \textit{χ} and the \textit{o} > \textit{a} change are connected with the reduction that takes place in word-final unstressed syllables. As a result of these processes, the connection of this marker to the dative suffix is weakening and the whole construction is becoming further grammaticalized, cf. \REF{ex:niv:11}:

\ea\label{ex:niv:11}
    \ea\label{ex:niv:11a}
        \gll Həɟ məkr-ɟ=ra. \\ 
             that be\_right-\textsc{ind=foc} \\
        \glt ‘This is right.’
    \ex\label{ex:niv:11b}
        \gll Həɟ \textbf{məkr-ta} \textbf{qʰaw-ɟ}. \\
              that be\_right-\textsc{dat} \textsc{neg-ind}\\
        \glt ‘This isn’t right.’ \citep{Jushkina2019} [AL]
    \z
\z 

The negative analytic complex displays regular inflection: aspect and modality are marked on the lexical verb, tense can be marked either on the lexical verb, as in \REF{ex:niv:10} and \REF{ex:niv:12a}, or on the negative verb, as in \REF{ex:niv:12b}, whereas mood and finiteness / non-finiteness are marked exclusively on the negative verb, cf. the finite form of the negative verb in the adversative mood in \REF{ex:niv:13}:

\ea\label{ex:niv:12}
    \ea\label{ex:niv:12a}
        \gll Tə+als-ku \textbf{ɲeɲi-nə-dox} \textbf{qʰaw-ɟ}. \\ 
             this+berry-\textsc{pl} be\_sweet-\textsc{fut-dat} \textsc{neg-ind} \\
        \glt ‘These berries won’t be sweet.’
    \ex\label{ex:niv:12b}
        \gll Tə+als-ku \textbf{ɲeɲi-dox} \textbf{qʰaw-nə{}-ɟ}. \\
             this+berry-\textsc{pl} be\_sweet-\textsc{dat} \textsc{neg-fut-ind} \\
        \glt ‘These berries won’t be sweet.’ [A]
    \z
\z 

\ea\label{ex:niv:13}
    \gll Ɲi \textbf{cʰ-sa-dox} \textbf{qʰau-bəɲevo{}-ʁar}. \\ 
         \textsc{1sg} 2\textsc{sg}{}-beat-\textsc{dat} \textsc{neg}{}-\textsc{prob-advs}\\
    \glt ‘It seems [that] I didn’t beat you.’ \citep[213]{Kreinovich1934} [A]
\z 

Synchronically, the entire analytic construction of negation is asymmetrical to the affirmative one, since the lexical verb loses its finiteness and is deverbalized (type A/Fin(/NegVerb) in \citegen{Miestamo2005} classification).

Another analytic construction seems to be used only in ES and is attested only in \citet{KreinovichMs}\footnote{We refer to the unpublished Nivkh texts that were collected by E. A. Kreinovich on Sakhalin in 1926–1928 \citep{KreinovichMs}. Copies of these documents from the E. A. Kreinovich Fund in the Sakhalin Regional Museum were kindly provided to us by A. B. Ostrovskii (part of them were discovered by the Museum and part by A. B. Ostrovskii himself). Russian translations of these texts were made in 2012–2013 by N. Ja. Tanzina, V. N. Sachgun and G. I. Paklina in collaboration with A. B. Ostrovskii and were published separately in \citet{Ostrovskii2016} and \citet{Kreinovich2016}. The authors are grateful to A. B. Ostrovskii, M. M. Prokofiev, and Sakhalin Regional Museum for the opportunity to use these data.} It is not known to the contemporary speakers. This construction comprises the lexical verb in the nominalized form with the suffix \textit{-}(\textit{n})\textit{d}, which functions as an object to the negative verb \textit{ʁavr{}-}\footnote{In the texts this verb is recorded as \textit{xovr-} or \textit{xavr-} with the initial velar voiceless fricative.}. Note that in the finite predicative form the suffix \textit{-}(\textit{n})\textit{d} was reanalyzed as an indicative mood marker.
In example \REF{ex:niv:14a}, one can see the nominalized form \textit{lu-nd} ‘singing’ and the finite form of the negative verb \textit{xovr-nd}. The contemporary synthetic variant of the same construction is given in \REF{ex:niv:14b}.

\ea\label{ex:niv:14}
    \ea\label{ex:niv:14a}
        \gll  Ni vuu-nd. Ni \textbf{lu-nd+xovr-nd}.\\ 
             \textsc{1sg} cry-\textsc{ind} \textsc{1sg} sing-\textsc{nmlz+neg-ind} \\
        \glt ‘I’m crying, I’m not singing.’ (\citealt{KreinovichMs} : D824:12v–15, «Indən. Təlgund») [ES]
    \ex\label{ex:niv:14b}
        \gll  Ɲi vuw-d. Ɲi \textbf{lu-ɢavr-d}. \\
             \textsc{1sg} cry-\textsc{ind} \textsc{1sg} sing-\textsc{neg-ind} \\
        \glt ‘I’m crying, I’m not singing.’ (an adaptation of the same text into modern ES by G. I. Paklina) [ES]
    \z
\z

Furthermore, Kreinovich’s texts contain examples of one more negative construction with the negative verb \textit{kau{}-} (= \textit{qʰaw{}-}). Here, the nominalized lexical verb in the form of \textit{-}(\textit{n})\textit{d} apparently functions as a finite verb form, whereas the negative verb, which is also nominalized, forms an additional one-word clause, cf. \textit{vi-i-nd=ra kau-k=re} ‘[I] won’t go’ in \REF{ex:niv:15}, in which one person quotes two different people.

\ea\label{ex:niv:15}
    \gll  J-aki. Cʰ-ingrə{}-i-nd=ra. \\ 
         \textsc{3sg}{}-elder\_brother \textsc{2sg}{}-go\_together-\textsc{fut-ind=foc} \\ 
    \glt ‘His elder brother [said]: I’ll go with you.’ \medskip\newline
    \gll Ugrər \textbf{vi-i-nd=ra} \textbf{kau-k=re}.\\ 
        together go-\textsc{fut-ind=foc} \textsc{neg-nmlz.agt=foc\_emph} \\ 
    \glt ‘[The other person said:] [I] won’t go together [with you].’ \medskip\newline
    \gll  Niɣəvəŋ+to-inə{}-tla agu-nd=ra[.] Park vi-inə{}-nd=ra.\\ 
         man+help-\textsc{des-asrt.1sg} not\_want-\textsc{ind=foc} only go-\textsc{des-ind=foc}\\
    \glt  ‘[I] don’t want to help the man. [I]’ll go myself.’ (\citealt{KreinovichMs} : D363:1–13, «Pal təlgund») [ES]
\z 

An alternative negation strategy, which is used by some modern speakers of ES, can be seen as an extension of the basic synthetic form. It is formed through the following steps: first, according to the general rule, the negative verb \textit{ʁavr{}-} combines with a lexical verb root. Then, the resulting negative stem is nominalized by the suffix \textit{-k}, cf. \textit{orbot-ɢavr-k} ‘a not working one’ in \REF{ex:niv:16}. The full form is ultimately created by the auxiliary verb \textit{ha{}-} ‘be / do so’, which takes all verbal markers, including non-finite ones, cf. \REF{ex:niv:16}. Synchronically, this construction is structurally not symmetrical to the positive one, but it is likely that a parallel affirmative analytic construction once existed in ES, since the verb \textit{ha{}-} ‘be / do so’ is widely used in this variety for deriving analytic forms.

\ea\label{ex:niv:16}
    \gll Ɲ{}-ətk \textbf{orbot-ɢavr-k} \textbf{ha-ŋa} eɣlŋ-gun+ɲr̥ə${\sim}$ɲr̥ə{}-d. \\ 
         \textsc{1sg-}father work-\textsc{neg-nmlz.agt} do\_so-\textsc{cvb.sim} child-\textsc{pl}+look${\sim}$\textsc{plac}{}-\textsc{ind} \\
    \glt ‘When my father didn’t work, he looked after the children.’ (Gruzdeva FM2000: 7) [ES]
\z 

\subsubsection{Negative habitual form}\label{sec:niv:2.1.3} 
Nivkh also has a specific negative habitual form that refers to a characteristic situation that does not hold at all times.\footnote{Cf. the definition of habitual aspect in \citep[27]{Comrie1976}.} Note that there are no positive habitual forms in Nivkh. This form is derived by the suffix \textit{-Ksu}\footnote{The negative habitual suffix has the variants \textit{-ksu}, \textit{-ɣsu}.}, which historically traces back to the negative verb ES \textit{j-aɣzu-}, A \textit{j-əɣzu-} ‘not know’ (\sectref{sec:niv:2.5}), cf. (\ref{ex:niv:17}--\ref{ex:niv:18}):

\ea\label{ex:niv:17}
    \gll Jaŋ arak+\textbf{ra-ɣsu-nt}. \\ 
         \textsc{3sg} vodka+drink-\textsc{hab.neg-ind} \\
    \glt ‘He (usually) doesn’t drink vodka.’ [ES]
\z 

\ea\label{ex:niv:18}
    \gll Cʰaj+χav-u-t \textbf{ra-ɣsu-ta}. Civla+cʰax=park+ta{\textasciitilde}ra-ta.\\ 
         tea+be\_hot-\textsc{tr-cvb.narr.3pl} drink-\textsc{hab.neg-coord.3pl} cold+water=only+drink{\textasciitilde}\textsc{plac}{}-\textsc{coord.3pl}\\
    \glt ‘[They] don’t drink hot tea (lit. [they] tea heating, don’t drink [it]), [they] drink only cold water.’ \citep[161]{Panfilov1965} [WS]
\z 

Like in other cases, in the A variety, the stem of the lexical verb can be followed by the emphatic suffix \textit{-Tə}, cf. \REF{ex:niv:19}:

\ea\label{ex:niv:19}
    \gll Cʰi ɲəŋ-dox \textbf{pʰrə-də-ksu-ɟ}. \\ 
         2\textsc{sg} 1\textsc{pl}.\textsc{excl-dat} come-\textsc{emph-hab.neg-ind} \\
    \glt ‘You don’t [even] come to our [place].’ \citep[161]{Panfilov1965} [A]
\z 

\subsection{Negation in non-declaratives}\label{sec:niv:2.2} 

There are two types of non-declarative sentences that express negation either directly or intrinsically: prohibitives (\sectref{sec:niv:2.2.1}) and apprehensives (\sectref{sec:niv:2.2.2}).

\subsubsection{Prohibitives}\label{sec:niv:2.2.1} 

Imperative (hortative, jussive) forms are the only forms in Nivkh that display a full differentiation of person and number. They are formed by attaching the suffixes listed in \tabref{tab:niv:1} to a verbal stem.


\begin{table}

\subtable[Amur variety]{\label{tab:mapu:2a}
    \begin{tabular}{@{~}llll}
    \lsptoprule
    &  Singular & Dual & Plural                                                 \\
    \midrule
    1 & \textit{-nəkta}, \textit{-nəxta} & \textit{-n(ə)te} & {-\textit{da}}    \\
    2 & {-\textit{ja}} & {-\textit{ve}} & {-\textit{ve}}                        \\
    3 & {-\textit{ʁazo}} & {-\textit{ʁazo}} & {-\textit{ʁazo}}                  \\
    \lspbottomrule
    \end{tabular}
}%
\subtable[East Sakhalin variety]{\label{tab:mapu:2b}
    \begin{tabular}{llll}
    \lsptoprule
     & Singular & Dual & Plural\\
    \midrule
    1 & - & \textit{-nate} & {-\textit{da}}\\
    2 & {-\textit{ja}} & {-\textit{ve}} & {-\textit{ve}} \\
    3 & \textit{-ʁaro} & {-\textit{ʁarʁ(a)ro}} & {-\textit{ʁarʁ(a)ro}}\\
    \lspbottomrule
    \end{tabular}
}
\caption{Imperative suffixes in Nivkh}
\label{tab:niv:1}
\end{table}


In all varieties, the prohibitive meaning can be expressed by a dedicated prohibitive construction with the special negator \textit{tʰa} which precedes, directly or non-directly, the corresponding positive imperative form. Prohibitive constructions are therefore completely symmetrical to the affirmative ones, cf. (\ref{ex:niv:20}--\ref{ex:niv:22}):

\protectedex{\ea\label{ex:niv:20}
    \ea\label{ex:niv:20a}
        \gll Pʰr̥ə-ja. \\ 
             come-\textsc{imp.2sg}  \\
        \glt ‘Come!’ 
    \ex\label{ex:niv:20b}
        \gll \textbf{Tʰa} \textbf{pʰr̥ə-ja}. \\
             \textsc{neg} come-\textsc{imp.2sg}\\
        \glt ‘Don’t come!’ [ES]
    \z
\z}

\ea\label{ex:niv:21}
    \gll \textbf{Tʰa} tus+pʰi-r̥ \textbf{kəpr-ja}. \\ 
         \textsc{neg} here+be\_situated-\textsc{cvb.narr.2sg} stand-\textsc{imp.2sg} \\
    \glt ‘Don’t stand here!’ [ES]
\z 

\ea\label{ex:niv:22}
    \gll Ɲi i-xez-t it-c. \textbf{Tʰa} \textbf{valc-ja}. \\ 
         \textsc{1sg} \textsc{3sg}{}-tell-\textsc{cvb.narr.1sg} say-\textsc{ind} \textsc{neg} lie-\textsc{imp.2sg} \\
    \glt ‘I told him: “Don’t lie!”’ \citep{Jushkina2019} [AL]
\z 

Furthermore, the ES variety makes use of an alternative prohibitive strategy, which is identical to standard negation. In that case the negative verb \textit{ʁavr{}-} synthesizes\textbf{ }with the lexical verb and the resulting complex verb appears in the corresponding imperative form. Again, the negative form is symmetrical to the affirmative one, cf. (\ref{ex:niv:23}--\ref{ex:niv:24}):


\ea\label{ex:niv:23}
    \ea\label{ex:niv:23a}
        \gll Pʰr̥ə-ja. \\ 
             come-\textsc{imp.2sg} \\
        \glt ‘Come!’
    \ex\label{ex:niv:23b}
        \gll \textbf{Pʰr̥ə{}-ɢavr-ja}. \\
             come-\textsc{neg-imp.2sg}\\
        \glt  ‘Don’t come!’ [ES]
    \z
\z 

\protectedex{\ea\label{ex:niv:24}
    \gll Cʰo kʰe+mi+si-n\footnotemark{} \textbf{huɟi-ɢavr-ve} qanŋ-gun i-ɲ{}-ʁar-i-d-ɣun. \\ 
         fish fishing\_net+inside+put-\textsc{cvb.narr.2pl.imp} leave-\textsc{neg-imp.2pl} dog-\textsc{pl} \textsc{3sg}{}-eat-\textsc{compl-fut-ind-pl}\\
    \glt ‘Don’t [pl.] leave fish in the fishing nets, [otherwise] the dogs will eat it.’ (Gruzdeva FM1991: 3:38:316) [ES\textsuperscript{T}]
\z }\footnotetext{The suffix \textit{-n} is a variant of the converbal suffix -\textit{r̥}, which is used when the finite verb occurs in the future or imperative forms.}

\subsubsection{Apprehensives}\label{sec:niv:2.2.2} 



Apprehensives\footnote{The corresponding forms in other languages are alternatively called ‘lest’ forms, timitives, evitatives, avoiditives, admonitives, preventives, dubitatives, precautionary imperatives, etc.} are used when a speaker attempts to prevent undesirable consequences of the situation by means of appealing to the addressee (see e.g. \citealt{Lichtenberk1995}). These forms can be seen as intrinsically negative. In Nivkh, there are special apprehensive forms, which contrast two evidential meanings (cf. \citealt{Kreinovich1979}: 316), specifying the source of fear. Based on this parameter, one can distinguish: (i) direct (visual) apprehensives and (ii) inferential apprehensives.\footnote{ This distinction seems to be typologically unusual. One example of a similar evidential contrast in apprehensive markers is found in Tariana, which has visual and non-visual apprehensive clitics \citep[384--387]{Aikhenvald2003}. We are grateful to an anonymous reviewer for this observation and for the reference.}

Visual apprehensives are issued in the situation when the speaker directly observes the situation that may lead to undesirable consequences. In the A variety, it is marked by the complex marker \textit{-nəra} that is comprised of the future tense suffix \textit{-nə} and the emphatic/direct evidential suffix \textit{-ra}, which is apparently used irrespective of the number of addressees and the person-number of the subject. Тhe resulting meaning can be literally analyzed as ‘Be aware that sb. is going to do sth.!’, cf. \textit{vi-nəra} in \REF{ex:niv:25}:

\ea\label{ex:niv:25}
    \gll Əlc-ɣo eʁgut oz-ve qʰotr \textbf{vi-nəra}. \\ 
         servant-\textsc{voc.pl} quickly.\textsc{2pl} get\_up-\textsc{imp.2pl} bear go-\textsc{appr.vis} \\
    \glt ‘Servants, get up quickly! [Otherwise] the bear will go away!’ \citep[118]{Panfilov1965} [A]
\z 

In the ES variety, there are two corresponding apprehensive markers differentiated by the number of addressees: \textit{-inəŋra} for singular and \textit{-inəŋta} for plural. Both markers contain (at least historically) the desiderative/intentional suffix \textit{-inə} and the nominalizer \textit{-ŋ}. The final segments \textit{-ra} and \textit{-ta} are the singular and plural variants of the emphatic/ direct evidential suffix that are used according to the number of the subject, as presented in \tabref{tab:niv:2}. The resulting meaning is: ‘Be aware that sb. is going to do sth.!’. For instance, when the addressee(s) is/are going along a slippery road, the speaker may warn him/her/them, cf. (\ref{ex:niv:26}--\ref{ex:niv:28}).
\ea\label{ex:niv:26}
    \ea\label{ex:niv:26a}
        \gll  Kuc-\textbf{inəŋra}.  \\ 
             fall\_down-\textsc{appr.vis.2sg} \\
        \glt ‘[Careful,] don’t fall down!’ 
    \ex\label{ex:niv:26b}
        \gll Kuc-\textbf{inəŋta}. \\
             fall\_down-\textsc{appr.vis.2pl} \\
        \glt ‘Don’t [pl.] fall down!’ [ES]
    \z
\z 

\ea\label{ex:niv:27}
    \gll Lar̥k+vaχc-u-\textbf{jnəŋra}. \\ 
         dress+burst-\textsc{tr}{}-\textsc{appr.vis.2sg} \\
    \glt ‘Don’t tear your dress!’ (Gruzdeva FM1989: 3:13) [ES\textsuperscript{T}]
\z 

\ea\label{ex:niv:28}
    \gll Pʰŋafqpʰŋafq+aɣaɣ-\textbf{inəŋta}. \\ 
         each\_other+disturb-\textsc{appr.vis.2pl} \\
    \glt ‘Don’t [pl.] disturb each other!’ (Gruzdeva FM2000: 4:36) [ES\textsuperscript{T}]
\z 

Inferential apprehensives are used when the speaker foresees undesirable consequences based not on direct evidence, but on previous experience. These forms seem to be very rare in the A variety. They are mentioned in \citep[316]{Kreinovich1979} in connection with the following example. Giving away a fragile object for a temporary use, the speaker is warning the addressee: \textit{Zosq-ijra} [break-\textsc{appr.infr}] ‘Don’t break [it]!’. The inferential apprehensive marker \textit{-ijra} or \textit{-əjra} consists of the emphatic/evidential suffix \textit{-ra} and the segment \textit{-ij} or \textit{-əj} with an unknown origin, cf. \REF{ex:niv:29}:


\ea\label{ex:niv:29}
    \gll Ola atak+vapak cʰi \textbf{zosq-əjra}. \\ 
         child.\textsc{voc} grandfather+toy 2\textsc{sg} break-\textsc{appr.infr} \\
    \glt ‘Child, don’t break grandfather’s toy!’ \citep[224]{Panfilov1965} [A]
\z

The corresponding forms also exist in the ES variety, where there are two inferential apprehensive markers opposed by number of the addressees: \textit{-jaŋra} for singular and \textit{-jaŋta} for plural. Both variants consist of the same component \textit{-ja}, which can be probably connected with the 2\textsc{sg} imperative marker \textit{-ja}, and the nominalizer \textit{-ŋ}. The final components, i.e. \textit{-ra} in the singular and \textit{-ta} in the plural, are of the same origin as the corresponding components of the visual apprehensive markers discussed above, cf. \REF{ex:niv:30}:

\ea\label{ex:niv:30}
    \ea\label{ex:niv:30a}
        \gll \textbf{Zosq-jaŋra}. \\ 
             break-\textsc{appr.infr.2sg} \\
        \glt ‘Don’t break [it]!’
    \ex\label{ex:niv:30b}
        \gll \textbf{Zosq-jaŋta}. \\
             break-\textsc{appr.infr.2pl}\\
        \glt ‘Don’t [pl.] break [it]!’ [ES]
    \z
\z

\begin{table} 
\begin{tabularx}{.6\textwidth}{llll}
    \lsptoprule
    & {Singular} & {Dual} & {Plural}\\
    \midrule
    first person & \textit{t} & \textit{t} & \textit{t}\\
    second person & \textit{\textbf{r}} & - & \textit{t}\\ 
    third person & \textit{\textbf{r}} & - & \textit{t}\\
    \lspbottomrule
\end{tabularx}
\caption{Distribution of common elements of suffixes of the apprehensive forms and the negative purposive converb}
\label{tab:niv:2}
\end{table}

\subsection{Negation in stative predications}\label{sec:niv:2.3} 

Existential (\sectref{sec:niv:2.3.1}), possessive (\sectref{sec:niv:2.3.2}) and locative (\sectref{sec:niv:2.3.3}) predications are verbal both in the affirmative and the negative and have a similar structure. Affirmative existential and locative clauses are differentiated by verbal predicates employed in the predicative position, but this differentiation is absent in the negative constructions. Due to the lack of adjectives as a separate part-of-speech category, there are no special attributive predications. Equative clauses (\sectref{sec:niv:2.3.4}) are built only with nominal predications, whereas ascriptive clauses (\sectref{sec:niv:2.3.5}) contain either nominal or nominalized predicates, following a well-known cross-linguistic tendency \citep{Veselinova2013neg.exist}. The most versatile negators for all stative predications are the negative verbs ES, WS \textit{qavr{}-} (intransitive), \textit{ʁavr{}-} (transitive), WS, A \textit{qʰaw-} (intransitive) ‘not exist’. All three verbs are diachronically connected to each other (\sectref{sec:niv:4.6.1}).

\subsubsection{Negation of existential predications}\label{sec:niv:2.3.1} 

Affirmative existential clauses are formed with the clause-final verb \textit{jiv-} ‘exist, have’ and have the structure [\textsc{existee} NP (subject, unmarked) \textsc{existential/possessive} \textsc{V} (predicate)], cf. \REF{ex:niv:31}. In case the clause contains a NP indicating location, the latter is typically case-marked with the locative (in A) or the ablative (in WS and ES) or is followed by a locative postposition and occupies the topical clause-initial position. The structure of the clause is then [\textsc{locative} NP (topic, marked) \textsc{existee} NP (subject, unmarked) \textsc{existential/possessive} V (predicate)], cf. (\ref{ex:niv:31}--\ref{ex:niv:32}). In colloquial speech, especially in dialogues, the existential/possessive verb can be omitted, cf. \REF{ex:niv:33}.

\ea\label{ex:niv:31}
    \gll Qʰo-ror tʰət oz-ŋan v-emaɲaχ ilɣ+haːs+imə{}-ɟ. Ɲa+sŋaj \textbf{jiv-ra} cʰo+sŋaj \textbf{jiv-ra}.\\ 
         sleep-\textsc{cvb.dist} morning get\_up-\textsc{cvb.sim} \textsc{3sg-}old\_woman made\_of\_deerskin+clothes+give-\textsc{ind} beast+image exist-\textsc{coord.3sg} fish+image exist-\textsc{coord.3sg}\\
    \glt ‘When, after sleeping, [he] got up, his old woman gives him clothes made of deerskin. There was an image of a beast, there was an image of a fish.’ \citep[226]{Panfilov1965} [A]
\z 

\ea\label{ex:niv:32}
    \gll Mulk+mi-x als \textbf{jiv-ɟ}. \\ 
         birch\_bark\_basket+inside-\textsc{abl} berry exist-\textsc{ind} \\
    \glt ‘There are berries in the birch bark basket.’ (Gruzdeva FM2019: 2) [WS]
\z 

\protectedex{\ea\label{ex:niv:33}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:niv:33A}
        \gll Təf+mi-x aŋ? \\ 
             house+inside-\textsc{abl} who  \\
        \glt ‘Who’s in the house?’
    \exi{A: }\label{ex:niv:33B}
        \gll Təf+mi-x ɲ{}-əmək. \\ 
             house+inside-\textsc{abl} \textsc{1sg}{}-mother \\
        \glt ‘My mother is in the house.’ (Gruzdeva FM2019: 2) [WS]
    \end{xlist}
\z }


Existential predications can be negated in several ways. Folklore texts in the various dialectal varieties contain clauses with the general negator \textit{qʰaw}, which occurs in the predicative position without any derivational or inflectional markers. The negated structure is [\textsc{existee} NP (subject, unmarked) \textsc{negator} (predicate)], which is similar to the affirmative one, except that the predicate, which completes the clause, is non-verbal, cf. (\ref{ex:niv:34}--\ref{ex:niv:35}):

\ea\label{ex:niv:34}
    \gll Ɲiɣəvəŋ \textbf{kxáuː}\footnotemark{} puiŋa \textbf{kxáu} mifŋa \textbf{kxáu} cʰo \textbf{kxáu} tol’ŋa \textbf{kxáuː}.\\ 
         man \textsc{neg} bird \textsc{neg} land\_animal \textsc{neg} fish \textsc{neg} sea\_animal \textsc{neg} \\
    \glt ‘There are no men, no birds, no land animals, no fish, no sea animals.’ (\cite{KreinovichMs}: D540:7–15+22–24v, «Nuɣə kʰen ɲaqr̥… [1]») [ES]
    \footnotetext{\textit{kxáu} corresponds to \textit{qʰau-}.}
\z

\ea\label{ex:niv:35}
    \gll Hǝ+pǝt aɲcǝj ivŋ+ŋarmu-ɉ. Napa \textbf{qʰau}. \\ 
         that+tomorrow again \textsc{3pl}+wait-\textsc{ind} still \textsc{neg} \\
    \glt ‘Tomorrow [he] waits for them again. [They’re] still not [here].’ \citep[248]{Panfilov1965} [A]
\z 

Another type of negative existential clause has a structure symmetrical to the affirmative one, i.e. [\textsc{existee} NP (subject, unmarked) \textsc{negative} \textsc{V} (predicate)], and can be further extended by the locative NP. In the ES variety, the predicate is the intransitive negative verb \textit{qavr-}, cf. \REF{ex:niv:36}, which in the A variety has the form \textit{qʰaw-} ‘not exist’. In the WS variety, the negative verb is attested in both forms, cf. \REF{ex:niv:37}. Constructions with the intransitive negative verb are also used for negation of locative predications (\sectref{sec:niv:2.3.3}).

\protectedex{\ea\label{ex:niv:36}
    \gll Ɲi cʰma-t pr̥ə{}-nd=ra. Min+ŋafk cʰ-wo+ker-i-d=lo. Ker̥ \textbf{kxavr-d=ra}… \\ 
         \textsc{1sg} visit-\textsc{cvb.narr.1sg} come-\textsc{ind}=\textsc{foc} 1\textsc{pl}.\textsc{incl}+friend \textsc{2sg}{}-settlement+tell-\textsc{fut}{}-\textsc{ind=q} news \textsc{neg-ind=foc}\\
    \glt ‘I came to visit. Our friend [says]: “Will you tell about your village?” – “There’s no news…”’ (\cite{KreinovichMs}: D536b:13–14v, «Pilan tav ɲakr…») [ES]
\z }

\ea\label{ex:niv:37}
    \gll  Kulu-r̥ tʰər-ŋan i-tʰəmr̥  \textbf{qʰau-ɣət-c}. \\ 
         open\_eyes-\textsc{cvb.narr.3sg} look-\textsc{cvb.sim} \textsc{3sg-}polecat \textsc{neg-compl-ind}\\
    \glt ‘When, opening [his] eyes, he looked, his polecat wasn’t there.’ (lit. ‘There was no his polecat.’) \citep[248]{Panfilov1965} [WS]
\z 

In the ES variety, there is an alternative way to encode negative existence with the negator \textit{ujɣe} ‘not be’, which goes back to the nominal/verbal root \textit{ujɣi-} ‘[be] a sin’. In the A variety, this item is still used almost exclusively in its original lexical meaning, cf. \REF{ex:niv:38}. The negator can be used in the predicative position as it is, cf. \REF{ex:niv:39}, or it can function as a verbal stem \textit{ujɣi-} taking various types of suffixes, cf. \REF{ex:niv:40}.

\ea\label{ex:niv:38}
    \gll Mackə+oʁla-gu huŋrə{}-ɟ+vo-ɟ \textbf{uiɣ{}-ɟ=ra}. \\ 
         be\_small+child-\textsc{pl} be\_like\_that-\textsc{nmlz}+take-\textsc{nmlz} be\_sin-\textsc{ind=foc} \\
    \glt ‘It’s a sin for small children to take [something] like that.’ \citep[225]{Panfilov1965} [A]
\z 

\ea\label{ex:niv:39}
    \gll Pax=par̥k r̥ə+kutí \textbf{uigé}. \\ 
         stone=only door+hole \textsc{neg} \\
    \glt ‘There’s no opening for a door [made] of stone.’ (Pilsudski Ms., text 3) [ES]
\z 

\ea\label{ex:niv:40}
    \gll Ɲin+du cʰo sik \textbf{ujɣi-ʁar-d}. \\ 
         1\textsc{pl}.\textsc{excl}+lake fish all \textsc{neg-compl-ind} \\
    \glt ‘There’s no fish in our lake.’ (Gruzdeva FM2000: 1:12) [ES\textsuperscript{N}]
\z 

The following text passage \REF{ex:niv:41} illustrates various devices for expressing negative existence in the ES variety:

\protectedex{\ea\label{ex:niv:41}
    \gll Əɣrǝŋǝfk mrolv+ǝrŋ-ux nud=ziŋ \textbf{qavr-ʁar-d}. Nud+mif=ciŋ nud+ɲiɣvŋ=ɟiŋ nud=ziŋ mor̥ka-k \textbf{qavr-d}. Nud+ŋa=ziŋ cʰo=ziŋ \textbf{ujɣi-d}. Cʰxul cʰaχ=pǝr̥k. Tʰas=ciŋ tʰas=ciŋ nud=ziŋ tʰmǝrmǝ=ɟiŋ \textbf{ujɣi-d}. \\ 
         long\_time\_ago be\_ancient+time-\textsc{abl} what=\textsc{emph} \textsc{neg-compl-ind} what+land=\textsc{emph} what+man=\textsc{emph} what=\textsc{emph} be\_alive-\textsc{nmlz.agt} \textsc{neg-ind} what+beast=\textsc{emph} fish=\textsc{emph} \textsc{neg-ind} completely water=only where=\textsc{emph} where=\textsc{emph} what=\textsc{emph} move=\textsc{emph} \textsc{neg-ind}\\
    \glt ‘A long time ago, at ancient times, there was nothing. No land, no man, nothing alive. There were no animals, no fish. Only water. Nowhere, nowhere, there was no one moving.’ (Gruzdeva FM2000/2013: 8) [ES\textsuperscript{N}]
\z }

In the WS variety, existentials can be also negated with the verb \textit{ləɣə{}-} ‘not have’, which is otherwise utilized for possessive negation (\sectref{sec:niv:2.3.2}) or as a basic negator within a synthetic negative form (\sectref{sec:niv:2.1}). This verb is usually transitive, but it is not completely clear how it behaves in existential predications. One of the WS speakers notes that, in contrast to the neutral negative verb \textit{qʰaw{}-}, the verb \textit{ləɣə{}-} brings up a connotation of unexpectedness (= mirativity), cf. \REF{ex:niv:42}. It may also have a rather abstract negative meaning, cf. \REF{ex:niv:43}:

\ea\label{ex:niv:42}
    \gll Ɲi məɣ{}-t j-ama-ba cʰ-mu+\textbf{ləɣə{}-ta}.\\ 
         \textsc{1sg} come\_down-\textsc{cvb.narr.1sg} \textsc{3sg-}look-\textsc{cvb.imm} \textsc{2sg-}boat+not\_have-\textsc{evid.dir.1sg}\\
    \glt ‘When I came down [to the shore], when [I] looked, your boat wasn’t there (unexpectedly).’ (lit. “there was no your boat”) (Gruzdeva FM2019: 11) [WS]
\z 

\ea\label{ex:niv:43}
    \gll Ɲi morqa-nə{}-ɟ=lu \textbf{ləɣə{}-nə{}-ɟ=lu}? \\ 
         \textsc{1sg} be\_alive-\textsc{fut-ind=q} not\_have-\textsc{fut-ind=q} \\
    \glt ‘Will I be alive or not?’ (\citealt{Panfilov1965}: 160–161) [A]
\z 

\subsubsection{Negation of possessive predication}\label{sec:niv:2.3.2} 

Possessive clauses (which involve indefinite possessees) are also formed by the existential/possessive verb \textit{jiv-} ‘exist, have’, but it is additionally required to make a reference to the possessor. The structure of a possessive clause depends on the dialectal variety.

In the ES variety, possessive predication is built on the model of a transitive clause, so that the verb \textit{j-iv-} takes the possessee NP as an object: [\textsc{possessor} \textsc{NP} (subject, unmarked) \textsc{possessee} \textsc{NP} (object, unmarked) + \textsc{existential}/\textsc{possessive} \textsc{V} (predicate, transitive)]. As a result, the verb drops its initial \textit{j-}, which is typical of Nivkh transitive verbs that bear the \textsc{3sg} object marker, cf. \REF{ex:niv:44a} and \REF{ex:niv:49a}. In colloquial speech, the existential/possessive verb can also be omitted, cf. \REF{ex:niv:45}.


\ea\label{ex:niv:44}
    \ea\label{ex:niv:44a}
        \gll Ɲi cʰχa+iv-d. \\ 
             \textsc{1sg} money+exist-\textsc{ind} \\
        \glt ‘I have money.’
    \ex\label{ex:niv:44b}
        \gll Ɲi cʰχa+\textbf{iv-ɢavr-d}. \\
             \textsc{1sg} money+exist-\textsc{neg-ind}\\
        \glt ‘I don’t have money.’ (Gruzdeva FM2000: 1:42) [ES\textsuperscript{N}]
    \z
\z 

\ea\label{ex:niv:45}
    \gll Ɲi eɣlŋ+me-nŋ. \\ 
         \textsc{1sg} child+two-\textsc{clf.hum} \\
    \glt ‘I have two children.’ (Gruzdeva FM2000: 1:48) [ES\textsuperscript{N}]
\z 

In the WS and A varieties, the possessive clause employs the same verb \textit{jiv-} ‘exist, have’, but the structure of the clause is different. It is intransitive, maintains the structure of the existential clause and forms a construction of the “Topic Possessive” type in Stassen's (\citeyear[58]{Stassen2009}) terms: [\textsc{possessor} \textsc{NP} (topic, unmarked) \textsc{possessee} \textsc{NP} (subject, unmarked) \textsc{existential/possessive} \textsc{V} (predicate, intransitive)]. The possessee NP functions as a subject, whereas the possessor NP functions as a topical clause-initial element. The corresponding clause can be literally translated as ‘As for \textsc{possessor} \textsc{NP,} \textsc{possessee} \textsc{NP} exists’. The intransitivity of the verb is confirmed by the fact that the verb \textit{jiv-} does not drop its initial \textit{j}, which is illustrated by examples \REF{ex:niv:46} and \REF{ex:niv:47}. Alternatively, especially in the modern language, the possessor can be expressed by the locative case, cf. \REF{ex:niv:48}, resulting in a construction belonging to Stassen’s “Locational Possessive” type.

\ea\label{ex:niv:46}
    \gll If mu+me-kr \textbf{jiv-ra}. \\ 
         \textsc{3sg} boat+two-\textsc{clf.gnr} exist-\textsc{evid.dir.3sg} \\
    \glt ‘He has two boats.’ (Gruzdeva FM2019: 10) [WS]
\z 

\ea\label{ex:niv:47}
    \gll Tʰaɣrpʰik oʁla+ŋe-n=pak \textbf{jiv-ra}… \\ 
         middle\_one child+one-\textsc{clf.hum}=only exist-\textsc{coord.3sg} \\
    \glt ‘The middle [brother] has one child…’ \citep[160]{Panfilov1965} [A]
\z 

\ea\label{ex:niv:48}
    \gll Cʰ-ujn cʰo jiv-bara. \\ 
         2\textsc{sg-loc} fish exist-\textsc{advs} \\
    \glt ‘You certainly have fish.’ \citep[121]{Panfilov1965} [A]
\z 

In the ES variety, the possessive clause is often negated in adherence to the standard negation strategy by synthesizing the negative verb \textit{ʁavr-} with the existential/possessive verb \textit{j-iv-}, cf. \REF{ex:niv:44b} and \REF{ex:niv:49b}. The resulting construction is symmetrical to the corresponding affirmative one. Note that this strategy is not used for negating existential clauses.

\ea\label{ex:niv:49}
    \ea\label{ex:niv:49a}
        \gll Xevgun mam+iv-d. \\ 
             Xevgun old\_woman+exist-\textsc{ind} \\
        \glt ‘Xevgun has a wife.’
    \ex\label{ex:niv:49b}
        \gll Xevgun mam+\textbf{iv-ɢavr-d}. \\
             Xevgun old\_woman+exist-\textsc{neg-ind} \\
        \glt ‘Xevgun doesn’t have a wife.’
    \ex\label{ex:niv:49c}
        \gll Xevgun mam+\textbf{ʁavr-d}. \\
             Xevgun old\_woman+\textsc{neg-ind}\\
        \glt ‘Xevgun doesn’t have a wife.’ (Gruzdeva FM2000: 1:44) [ES\textsuperscript{N}]
    \z
\z 

In addition, possessive negation can be rendered by placing the standard negative verb in the predicative position. In the ES variety and occasionally in the WS variety the transitive verb \textit{ʁavr-}takes the possessee NP as an object, so that the structure of the clause is symmetrical to the affirmative one [\textsc{possessor} \textsc{NP} (subject, unmarked) \textsc{possessee} \textsc{NP} (object, unmarked) + \textsc{negative} \textsc{V} (predicate, transitive)], cf. \REF{ex:niv:49c}, \REF{ex:niv:50}, and \REF{ex:niv:51}:

\ea\label{ex:niv:50}
    \gll Jaŋ nud=ziŋ+\textbf{ɢavr-ʁar-d}. \\ 
          \textsc{3sg} what=\textsc{emph}+\textsc{neg-compl-ind} \\
    \glt ‘He doesn’t have anything.’ [ES]
\z 

\ea\label{ex:niv:51}
    \gll Hə+haʁs cʰup-xir̥ cʰŋaj+\textbf{ʁavr-ɟ}. \\ 
         that+clothing at\_all-\textsc{ins} image+\textsc{neg}{}-\textsc{ind} \\
    \glt ‘These clothes don’t have any images at all.’ (Gruzdeva FM2019: 1) [WS]
\z 

In the A and WS varieties, it is the intransitive negative verb \textit{qʰaw-} that takes the predicative position in a negative possessive clause, so that the resulting structure is also symmetrical to the affirmative one: [\textsc{possessor} NP (topic, unmarked) \textsc{possessee} NP (subject, unmarked) \textsc{negative} V (predicate, intransitive)], cf. (\ref{ex:niv:52}--\ref{ex:niv:53}). In such constructions, the topical element can also be case-marked with the locative, cf. \REF{ex:niv:54}.

\ea\label{ex:niv:52}
    \gll Əɣrəkon+ɲivɣ{}-gu pila+vəɲ \textbf{qʰau-ɟ}. \\ 
          ancient+man-\textsc{pl} be\_big+cauldron \textsc{neg}{}-\textsc{ind} \\
    \glt ‘Ancient people didn’t have big cauldrons.’ \citep[160]{Panfilov1965} [A]
\z 

\ea\label{ex:niv:53}
    \gll  Aŋ=hakisk \textbf{qʰaw-ɟ}. \\ 
         who=\textsc{emph} \textsc{neg-ind} \\
    \glt ‘There’s nobody.’ (Gruzdeva FM2019: 2) [WS]
\z 

\ea\label{ex:niv:54}
    \gll Ɲ{}-ujn cʰo qʰaw-ɟ=ra. \\ 
         1\textsc{sg}{}-\textsc{loc} fish \textsc{neg-ind{}=foc} \\
    \glt ‘I don’t have fish.’ \citep[121]{Panfilov1965} [A]
\z 

Alternatively, possessive negation can be rendered by inherently negative possessive verbs. In the ES variety, it is the transitive verb \textit{avli-} ‘not have’, cf. \REF{ex:niv:55}:


\ea\label{ex:niv:55}
    \gll Jaŋ nanq-χainapə ŋawr̥k+\textbf{avli-d}. \\ 
         \textsc{3sg} elder\_sister-\textsc{cvb.conc} brain+not\_have-\textsc{ind} \\
    \glt ‘Though she’s an elder sister, [she] lacks the brain.’ (Gruzdeva FM1991: 3:32) [ES\textsuperscript{T}]
\z 

For the same purpose the WS and A varieties use another transitive verb \textit{ləɣə{}-} ‘not have’, cf. \REF{ex:niv:56}. The subject of \textit{ləɣə{}-} is often followed by the emphatic clitic \textit{=Tə}:


\ea\label{ex:niv:56}
    \gll Hoʁar̥ əkəɟ umgu jiv-ra. Haŋgr toɟ umgu=də+\textbf{ləɣə{}-ra}. \\
         then older\_brother woman exist-\textsc{coord.3sg} then middle\_brother woman=\textsc{emph}+not\_have-\textsc{coord.3sg}\\
    \glt ‘Then the older brother had a wife. Then the middle brother didn’t have a wife.’ \citep[160]{Panfilov1965} [A]
\z 


\subsubsection{Negation of locative predications}\label{sec:niv:2.3.3} 



There are several constructions with locative predications that differ in their structure and in the verbs used in the predicative position.
The first construction is built on the model [\textsc{locatee} NP (subject, unmarked) \textsc{locative} NP (marked) \textsc{locative} V (predicate)] and consists of the intransitive verb ES \textit{hunv-}, WS/A \textit{hum-} ‘be, live’. The topical clause-initial position is taken by an NP indicating the locatee, whereas the referent of locative NP is conceived of as a new information. This is the typical pragmatic structure of locative clauses that usually differentiates them from existential ones. However, in Nivkh, this model works rather straightforwardly only for elicited examples, whereas in narratives, the locative clause may have a structure similar to that of the existential one, with the locative NP in the clause-initial position: [\textsc{locative} NP (marked) \textsc{locatee} NP (subject, unmarked) \textsc{locative} V]. Furthermore, in an appropriate context reference neither to the location nor to the located entity is obligatory. In the ES and WS varieties, the verb \textit{hunv-}, \textit{hum-} can be used only with animate referents, cf. (\ref{ex:niv:57}--\ref{ex:niv:58}), whereas in the A variety, \textit{hum-} occurs with any referent irrespective of its animacy/inanimacy, cf. \REF{ex:niv:59}. As a predicate of locative clauses, the verb retains its locative meaning, but it has also been completely grammaticalized as a marker of continuative in other contexts, cf. the second verb \textit{hum-} in \REF{ex:niv:58}.

\ea\label{ex:niv:57}
    \gll Tuŋ+hajmŋar̥ ɲe-nɲ=bǝr̥k tav-ux \textbf{hunv-d}. \\ 
         this+old\_man one-\textsc{clf.hum}=only house-\textsc{abl} be-\textsc{ind} \\
    \glt ‘This old man is/lives in the house alone.’ (Gruzdeva FM2000: 1:76) [ES\textsuperscript{N}]
\z 

\protectedex{\ea\label{ex:niv:58}
    \gll Ɲi təf-toχ təvʁa-ŋan aŋ=hakisk ʁavr-ɣər̥-c. Umgu-oːla=bǝːk \textbf{hum-r̥} \textbf{hum-ɟ}. \\ 
         \textsc{1sg} house-\textsc{dat} come\_in-\textsc{cvb.sim} who=\textsc{emph} \textsc{neg-compl-ind} woman-child=only be-\textsc{cvb.narr.3sg} be-\textsc{ind}\\
    \glt ‘When I came into the house, there was nobody [there], only a girl was there.’ (Gruzdeva FM2019: 3) [WS]
\z}

\ea\label{ex:niv:59}
    \gll Siɟ=agin qʰau{}-ra paχ+ɲa{}-qr pʰal{}-uin \textbf{hum-ra}.\\ 
         what=\textsc{emph} \textsc{neg-coord.3sg} stone+one-\textsc{clf.gnr} floor-\textsc{loc} be{}-\textsc{coord.3sg}\\
    \glt ‘There’s nothing, there’s only a stone on the floor.’ \citep[226]{Panfilov1965} [A]
\z 

Another locative verb \textit{Xu-} ‘be located [swh.]’ is used in all varieties only with inanimate referents and obligatorily requires a reference to the location. In all varieties, it functions as a transitive verb and takes a locative NP as an object, so that the clause structure is [\textsc{locatee} NP (subject, unmarked) \textsc{locative} NP (object, unmarked) + \textsc{locative} V], cf. (\ref{ex:niv:60}--\ref{ex:niv:61}).

\ea\label{ex:niv:60}
    \gll Pitɣaŋ ostol+vaj+\textbf{xu-d}. \\ 
         book table+under+be\_located-\textsc{ind} \\
    \glt ‘The book is under the table.’ (Gruzdeva FM2000: 1:18) [ES\textsuperscript{N}]
\z 


\ea\label{ex:niv:61}
    \gll Umgu zu+ŋir-gu tǝr+tʰxǝ+\textbf{xu-ɟ}.\\ 
         woman wash+bowl-\textsc{pl} table+on+be\_located-\textsc{ind} \\
    \glt ‘The bowls (which) the woman had washed are on the table.’ (\citealt{NedjalkovOtaina2013}: 280) [A]
\z 

One more type of locative predication can be formed by the transitive verb \linebreak\textit{Fi-}\footnote{The
    verb \textit{fi-} has the bound variant \textit{-pʰi-}.
} ‘be situated (swh.)’, which in all varieties takes a locative NP as an object. The construction with this verb has the structure [\textsc{locatee} NP (subject, unmarked) \textsc{locative} NP (object, unmarked) + \textsc{locative} V], cf. (\ref{ex:niv:62}--\ref{ex:niv:63}).

\ea\label{ex:niv:62}
    \gll Liɣr̥-kun tuŋ+pʰxi+\textbf{fi-d-ɣun=la}? \\ 
         wolf-\textsc{pl} this+forest+be\_situated{}-\textsc{pl=q} \\
    \glt ‘Are there wolves in this forest?’ (Gruzdeva FM2000: 1:16) [ES\textsuperscript{N}]
\z 

\ea\label{ex:niv:63}
    \gll Naf ɲ{}-umgu tʰama-r təf+\textbf{pʰi-r} o\textbf{ː}la-gu+tʰof-c. \\ 
         now \textsc{1sg-}woman not\_do-\textsc{cvb.narr.3sg} house+be\_situated\textsc{{}-cvb.narr.3sg} child-\textsc{pl}+take\_care-\textsc{ind}\\
    \glt ‘Now my wife, not doing [anything], staying at home, takes care of children.’ (Gruzdeva FM2019: 3) [WS]
\z 

Locative predications are negated in the ES variety with the transitive negative verb \textit{ʁavr-}, so that the negative construction has the structure symmetrical to the affirmative one [\textsc{locatee} NP (subject, unmarked) \textsc{locative} NP (object, unmarked) + \textsc{negative} V], cf. \REF{ex:niv:64}. This construction is different from the intransitive negative existential one (\sectref{sec:niv:2.3.1}).


\ea\label{ex:niv:64}
    \gll \textbf{Mu} \textbf{muspi}+\textbf{ʁavr-d}. Mu taf+lef+kʰu-d. \\ 
         boat shore+\textsc{neg-ind} boat house+near+be\_located{}-\textsc{ind} \\
    \glt ‘The boat is not on the shore. The boat is near the house.’ (Gruzdeva FM2021: Mess.) [ES\textsuperscript{T}]
\z 

In the WS and A varieties, negation of locative predications is built according to the same model as negative existence and is expressed by the intransitive negative verb \textit{qʰaw-} ‘not exist’, cf. (\ref{ex:niv:65}--\ref{ex:niv:66}):


\ea\label{ex:niv:65}
    \gll  Mu co-wx \textbf{qʰaw-ɟ}. Mu təf+laɣa hunr̥um-ɟ.\\ 
         boat shore-\textsc{abl} \textsc{neg-ind} boat house+near be-\textsc{ind} \\
    \glt ‘The boat is not on the shore. The boat is near the house.’ (Gruzdeva FM2020: Mess.) [WS]
\z 


\ea\label{ex:niv:66}
    \gll Vi-ke pʰ-r̥əf-tox vi-ra nanak \textbf{qʰau-ra}. \\ 
         go-\textsc{cvb.dur} \textsc{refl}{}-house-\textsc{dat} go-\textsc{coord.3sg} older\_sister \textsc{neg-coord.3sg} \\
    \glt ‘Going for a long time, [he] came to his house, [his] older sister wasn’t there [in the house].’ (lit. “there was no [his] older sister”) \citep[225]{Panfilov1965} [A]
\z 

\subsubsection{Equative negation}\label{sec:niv:2.3.4} 

Equative (or identificational) clauses equate the referents of the subject and the nominal predicate. The predicate of an equative clause is typically a regular noun phrase, which in all varieties takes the focus clitic \textit{=Ta},\footnote{The focus clitic has the variants =\textit{ta}, =\textit{ra}, =\textit{da}.} cf. \REF{ex:niv:67a} and other examples below. No copula is needed in the non-future indicative, though in the future the verb \textit{mu-} ‘become’ must be used, cf. \REF{ex:niv:70b}. Equative clauses are negated in one of the following ways.

In the ES variety, the predicative position of the negative equative clause is typically occupied by the nominalized form of the negative verb \textit{ʁavr-}, in which case the negated NP seems to function as its attribute, cf. \REF{ex:niv:67b}.


\ea\label{ex:niv:67}
    \ea\label{ex:niv:67a}
        \gll Tu+ɲiɣvŋ ɲ-χalŋ=da. \\ 
             this+man \textsc{1sg}{}-clan=\textsc{foc} \\
        \glt ‘This person is my relative.’
    \ex\label{ex:niv:67b}
        \gll Tu+ɲiɣvŋ \textbf{ɲ-χalŋ+ɢavr̥-k}. \\
             this+man \textsc{1sg}{}-clan+\textsc{neg-nmlz.agt} \\
        \glt ‘This person is not my relative.’ [Gruzdeva FM2000: 1:23] [ES]
    \z
\z 

In the A and WS varieties, equative negative predicates are formed by three elements: negated NP, the dative form of the nominalized auxiliary verb \textit{ha-} ‘be / do so’ and the negative verb, which in \REF{ex:niv:68b} occurs in the finite indicative form, which is historically also a nominalization. In example \REF{ex:niv:69}, which illustrates contrastive negation, the negative verb is used in a nominalized form:


\ea\label{ex:niv:68}
    \ea\label{ex:niv:68a}
        \gll Təɟ ɲ-axmalk=ra. \\ 
             this\_one \textsc{1sg}{}-father\_in\_law=\textsc{foc} \\
        \glt ‘This one’s my father-in-law.’
    \ex\label{ex:niv:68b}
        \gll Təɟ \textbf{ɲ-axmalk} \textbf{ha-to} \textbf{qʰaw-ɟ=ra}. \\
             this\_one \textsc{1sg}{}-father\_in\_law do\_so{}-\textsc{dat} \textsc{neg-ind=foc}\\
        \glt ‘This one’s not my father-in-law.’ (Gruzdeva FM2019: 1:11) [WS]
    \z
\z 

\ea\label{ex:niv:69}
    \gll Tə+ɲivx umgu=da \textbf{utku} \textbf{ha-dox} \textbf{qʰau-k=ra}. \\ 
         this+man woman=\textsc{foc} man do\_so{}-\textsc{dat} \textsc{neg-nmlz.agt=foc} \\
    \glt ‘This person is a woman, not a man.’ \citep[158]{Panfilov1965} [A]
\z 

As can be seen from the following WS examples, in the future, the negative verb \textit{qʰaw-} can synthesize with the verb \textit{mu-} ‘become’. Note that in the resulting form the place of the future marker may vary, cf. \REF{ex:niv:70c} and \REF{ex:niv:70d}. Equative clauses seem to be the only type of clause in which the negative verb \textit{qʰaw-} can be used as a part of a synthetic form in a similar way to the verb \textit{ʁavr-}.


\ea\label{ex:niv:70}
    \ea\label{ex:niv:70a}
        \gll Tə+ɲivx ɲ-axmalk=ra. \\ 
             this+man \textsc{1sg}{}-father\_in\_law=\textsc{foc} \\
        \glt ‘This man is my son-in-law.’
    \ex\label{ex:niv:70b}
        \gll Tə+ɲivx ɲ-axmalk mu-nə-ɟ=ra. \\
             this+man \textsc{1sg}{}-father\_in\_law become-\textsc{fut-ind=foc} \\
        \glt ‘This man will be my father-in-law.’
    \ex\label{ex:niv:70c}
        \gll Tə+ɲivx \textbf{ɲ-axmalk} \textbf{mu-nə-qʰaw-ɟ=ra}. \\ 
             this+man \textsc{1sg}{}-father\_in\_law become-\textsc{fut-neg-ind=foc} \\
        \glt ‘This man won’t be my father-in-law.’
    \ex\label{ex:niv:70d}
        \gll Tə+ɲivx \textbf{ɲ-axmalk} \textbf{mu-qʰaw-nə-ɟ=ra}. \\
             this+man \textsc{1sg}{}-father\_in\_law become-\textsc{neg-fut-ind=foc}\\
        \glt ‘This man won’t be my father-in-law.’ (Gruzdeva FM2019: 1:11) [WS]
    \z
\z 

\subsubsection{Ascriptive negation}\label{sec:niv:2.3.5} 

Ascriptive clauses assign a certain property to the referents of their subjects. The predicate is either a full noun phrase, cf. \REF{ex:niv:71}, or a nominalized verb form, cf. \REF{ex:niv:72a}. The latter is derived from a verbal stem with the nominalizer \textit{-k}. As with equative clauses, no copula is needed in the non-future. Negation of ascriptive clauses is done in the standard way. In case the predicate of an affirmative equative clause is itself a nominalized verb form, the negative verb ES/WS \textit{ʁavr{}-} synthesizes with it, cf. \REF{ex:niv:72b}. WS and A make use of the standard analytic negative construction, cf. \REF{ex:niv:73}.

\ea\label{ex:niv:71}
    \gll Liɣs umla+ŋa. \\ 
         wolf angry+beast \\
    \glt ‘A wolf is an angry beast.’ (Gruzdeva FM2000: 1:26) [ES\textsuperscript{N}]
\z 

\ea\label{ex:niv:72}
    \ea\label{ex:niv:72a}
        \gll Ɲ-eɣlŋ mraŋi-k. \\ 
             \textsc{1sg}{}-child be\_lazy-\textsc{nmlz.agt} \\
        \glt ‘My child is a lazy one.’
    \ex\label{ex:niv:72b}
        \gll Ɲ-eɣlŋ \textbf{mraŋi-ɢavr̥-k}. \\
            \textsc{1sg-}child be\_lazy-\textsc{neg-nmlz.agt} \\
        \glt  ‘My child is not a lazy one.’ (Gruzdeva FM2000: 1:22) [ES\textsuperscript{N}]
    \z
\z

\ea\label{ex:niv:73}
    \gll Ɲ-oɣla \textbf{mraŋi-to} \textbf{qʰaw-k}. If mu-jvi-ɟ. \\ 
         \textsc{1sg}{}-child be\_lazy-\textsc{dat} \textsc{neg-nmlz.agt} \textsc{3sg} die-\textsc{prog-ind} \\
    \glt ‘My child is not lazy, he’s ill.’ (Gruzdeva FM2020: Mess.) [WS]
\z 

\subsection{Negation in non-main clauses}\label{sec:niv:2.4}
In a complex sentence with subordinated clause(s), the main clause typically occurs sentence-finally and is built around a clause-final finite (= independent) form. Subordinate clauses are represented by complement, relative or adverbial clauses.

Complement clauses function as sentential complements to main clauses containing verbs of perception. The predicate of the complement clause is marked by the nominalizer ES \textit{-}(\textit{n})\textit{d/-t}, WS/A \textit{-ɟ/-c}, which synchronically also functions as the indicative marker in the finite verb form. Both clauses have the same structure and can be negated in the same way, cf. \REF{ex:niv:152}:

\ea\label{ex:niv:152}
    \gll Ɲ-aki-roχ jaŋ jaŋr̥ \textbf{pʰ-rau-ɢavr-d}+otot-ba  jaŋ pʰe-ʁan caftrer-vur̥ it-t.\\ 
         \textsc{1sg}{}-older\_brother{}-\textsc{dat} \textsc{3sg} why:\textsc{3sg} \textsc{refl-}study-\textsc{neg-ind}+ask-\textsc{cvb.imm} \textsc{3sg} \textsc{refl-rep} be\_busy-\textsc{cvb.rep.1sg} say-\textsc{ind} \\
    \glt ‘When [someone] asks my older brother, why he doesn’t study, he says [that] he’s busy.’ (Gruzdeva FM1991: 3:18) [ES\textsuperscript{N}]
\z 

Relative clauses are headed by noun-modifying participles and embedded into the main clause. They can also be negated with the standard negators without any restrictions, cf. (\ref{ex:niv:153}--\ref{ex:niv:154}):

\ea\label{ex:niv:153}
    \gll Iosif mu+ɲivɣ{}-gu+ŋav+pos cʰʁa-ɣir̥ ɣe-ɟ. Hoqoror̥ krest-ux Iisus+ʁoc-r̥ pos-kir̥  ŋav-ror̥ aŋ=hakisk napa huŋk \textbf{vəks-ta} \textbf{qʰau}+hur̥kif+si-ɟ.  \\ 
         Joseph boat+man-\textsc{pl}+wrap+cloth money-\textsc{ins} take-\textsc{ind} then cross-\textsc{abl} Jesus+take\_off-\textsc{cvb.narr.3sg} cloth-\textsc{ins} wrap-\textsc{cvb.dist.3sg} who=\textsc{emph} yet there bury-\textsc{dat} \textsc{neg}+cave+put\textsc{{}-ind} \\
    \glt ‘Joseph bought the cloth in which dead bodies are wrapped, then [he] took Jesus’s body off the cross, wrapped [it] in the cloth and put it into the cave, where nobody was buried yet.’ \citep[156]{Histories2017} [WS]
\z 

\ea\label{ex:niv:154}
    \gll \textbf{Pil-də{}-ləɣə}+dəf pal-uin hum-ɟ. \\ 
         be\_big-\textsc{emph}{}-not\_have+house forest\_hill-\textsc{loc} be-\textsc{ind} \\
    \glt ‘A small house is situated on the forest hill.’ \citep[160]{Panfilov1965} [A]
\z 

Adverbial clauses are formed by converbs (= dependent forms) of different origin and are arranged in chains which end with the main clause. Converbs render various interclausal relations, with the finest differentiation within the temporal ones.

Nivkh has a special negative purposive converb in ES \textit{-lakrr̥/-lakrt}, WS/A \textit{-ləkrr/-ləkrt} with a complex marker which consists of the bound form of the lexical verb ES \textit{lakr-}, WS/A \textit{ləkr-} ‘fear’ and the suffix of the narrative converb ES \textit{-r̥ }: \textit{-t}, WS/A \textit{-r }: \textit{-t}. The converb agrees in person and number not with its own subject, but rather with the subject of the finite (= independent) verb, following the pattern in \tabref{tab:niv:2}, cf. \REF{ex:niv:155} and \REF{ex:niv:156}:

\ea\label{ex:niv:155}
    \gll Qan-gu \textbf{vi-jnə{}-ləkrt} ɲi tʰu+zeqaw-ɟ. \\ 
         dog-\textsc{pl} go-\textsc{des-cvb.neg\_purp.1sg} \textsc{1sg} sledge+fasten-\textsc{ind} \\
    \glt ‘I fastened the sledge, so that the dogs wouldn’t go away.’ (Gruzdeva FM2019: 1:21) [WS]
\z 

\ea\label{ex:niv:156}
    \ea\label{ex:niv:156a}
        \gll Ɲi e-f-t r̥orvi-ɟ=ra oːla-gu \textbf{ra-i-ləkrt}. \\ 
             \textsc{1sg} \textsc{3sg-}take\textsc{{}-cvb.narr.1sg} carry\_away\textsc{{}-ind=foc} child-\textsc{pl} drink-\textsc{ep}{}-\textsc{cvb.neg\_purp.1sg} \\
        \glt ‘Taking it, I carried [the medicine] away so that the children wouldn’t drink [it].’
    \ex\label{ex:niv:156b}
        \gll If e-v-r r̥orvi-ɟ=ra oːla-gu \textbf{ra-i-ləkrr}. \\
             \textsc{3sg} \textsc{3sg-}take\textsc{{}-cvb.narr.3sg} carry\_away\textsc{{}-ind=foc} child-\textsc{pl} drink-\textsc{ep-cvb.neg.purp.3sg} \\
        \glt ‘Taking it, he carried [the medicine] away so that the children wouldn’t drink [it].’ \citep[321]{Kreinovich1979} [A]
    \z
\z 

All other converbs are negated according to the standard negation strategies used in one or another variety, cf. narrative converbs in (\ref{ex:niv:157}--\ref{ex:niv:160}), a simultaneous/anterior converb in \REF{ex:niv:161}, and conditional converbs in (\ref{ex:niv:162}--\ref{ex:niv:163}):

\protectedex{\ea\label{ex:niv:157}
    \gll Jaŋ pʰr̥ə-ŋa jaŋr̥pər̥klu \textbf{j-uʁ-ɢavr-r̥} r̥ə+or̥+pʰi-r̥ kəp-r̥ hunv-nd. \\ 
         \textsc{3sg} come-\textsc{cvb.sim} for\_some\_reason \textsc{3sg}{}-come\_in-\textsc{neg-cvb.narr.3sg} door+edge+be\_situated-\textsc{cvb.narr.3sg} stand-\textsc{cvb.narr.3sg} be{}-\textsc{ind} \\
    \glt ‘When he came, for some reason not coming in, [he] stands at the threshold.’ (Gruzdeva FM1991: 3:13), [ES\textsuperscript{T}]
\z}

\ea\label{ex:niv:158}
    \gll Slovar-ux mer+dif-ku \textbf{məkr-da} \textbf{qʰaw-gu-t} raju-ɟ{}-ɣu.\footnotemark{} \\ 
         dictionary{}-\textsc{abl} 1\textsc{pl}.\textsc{incl}+word-\textsc{pl} be\_right-\textsc{dat} \textsc{neg-caus-cvb.narr.3pl} write-\textsc{ind-pl} \\
    \glt ‘In the dictionary, [they] wrote our words wrong.’ \citep{Jushkina2019} [A]
\z\footnotetext{The sentence can be literally translated as ‘In the dictionary, our words making wrong, [they] write.’}

\ea\label{ex:niv:159}
    \gll Məx-ke \textbf{i-ndə-ləɣə-r}  məɣ-r i-coŋr-tox pol-ra  mulk cosq-xət-ra. \\ 
         come\_down-\textsc{cvb.dur} \textsc{3sg}{}-look-not\_have-\textsc{cvb.narr.3sg} come\_down-\textsc{cvb.narr.3sg} \textsc{3sg-}head-\textsc{dat} fall\_down-\textsc{coord.3sg} birch\_bark\_basket break-\textsc{compl-coord.3sg} \\
    \glt ‘When [he] was coming down, [he] didn’t look [around], [he] fell down on his head, the birch basket broke.’ \citep[227]{Panfilov1965} [A]
\z

\ea\label{ex:niv:160}
    \gll \textbf{To-jrucez-r̥} pʰer-r qʰo-ɟ. \\ 
         cry-be\_unable-\textsc{cvb.narr.3sg} be\_tired-\textsc{cvb.narr.3sg} sleep-\textsc{ind} \\
    \glt ‘Not being able to cry [anymore], being tired, [he] fell asleep.’ \citep[248]{Panfilov1965} [WS]
\z

\ea\label{ex:niv:161}
    \gll Ɲ-ətk cik \textbf{pʰr̥ə-ŋ-ɢavr-ŋa} sikm xurŋuftu-d-ɣun. \\ 
         \textsc{1sg}{}-father for\_a\_long\_time come-\textsc{nmlz-neg-cvb.sim} all worry-\textsc{ind-pl} \\
    \glt ‘When my father didn’t come for a long time, everyone worried.’ (Gruzdeva FM1991: 3:14) [ES\textsuperscript{N}]
\z

\ea\label{ex:niv:162}
    \gll Cʰi tə+ɟus+\textbf{ɲi-də-ləɣə-ʁaj} cʰi-ɣ{}-nə-ɟ=ra. \\ 
         2\textsc{sg} this+meat+eat-\textsc{emph}{}-not\_have-\textsc{cvb.cond} 2\textsc{sg}{}-kill-\textsc{fut-ind=foc} \\
    \glt ‘If you don’t eat this meat, [I]’ll kill you.’ \citep[229]{Panfilov1965} [A]
\z

\ea\label{ex:niv:163}
    \gll Məkrgut ɲivɣɟif+\textbf{raju-da} \textbf{qʰaw-ʁa} məkr-la+ɲivɣɟif+bəkz-nə{}-ɟ=ra. \\ 
         correctly:\textsc{3pl} Nivkh\_language+write-\textsc{dat} \textsc{neg-cvb.cond} be\_correct-\textsc{qual}+Nivkh\_language+lose-\textsc{fut-ind=foc} \\
    \glt ‘If [they] doesn’t write correctly in Nivkh, [they]’ll lose Nivkh. \citep[227]{Jushkina2019} [A]
\z



\subsection{Negative lexicalizations}\label{sec:niv:2.5} 

There is a whole inventory of lexical/lexicalized negative verbs, some of which have been already discussed in connection with stative predications (\sectref{sec:niv:2.3}). Here we will examine negative verbs and some other lexemes that denote lack of desire, knowledge, attraction, and physical ability, as well as some other negative meanings. Most of them fall in line with what \citet{Zeshan2004} and \citet{Veselinova2013lexicalization} identified for sign and spoken languages, respectively. The inherently negative verbs often (but not always) form semantic pairs with their positive counterparts. They do not require any explicit negators but in some exceptional cases can combine with them. In general, using a standard negator with an inherently negative lexeme is perceived by the speakers as ungrammatical. Some of the lexicalized negative verbs can be used both independently and as a part of the synthetic negative form, whereas the others are synchronically attested only in synthetic negative forms. In the latter case, the root of a lexicalized modal negative verb synthesizes with the root of a lexical verb. The resulting forms behave as a single verb and can be used both in finite and non-finite functions. Synthetic modal negative forms express two meanings: non-desire and non-ability.

\subsubsection{Non-desire}

Desire vs. non-desire is indicated by the pair of verbs ES \textit{j-aɣɲi}, cf. \REF{ex:niv:74},\textbf{ }WS/A \textit{j-aɣɲ-}, cf. \REF{ex:niv:75}, ‘want’ vs.\textbf{ }ES \textit{aɣu-}, WS/A \textit{əɣi-} ‘not want’. In ES, the negative verb \textit{aɣu-} displays variation with respect to transitivity. This can be illustrated by example \REF{ex:niv:76}, where it occurs as a predicate of the main clause with a dislocated complement clause, and example \REF{ex:niv:77}, where it takes an oblique verbal complement formed by the locative nominalizer \textit{-f} and the dative suffix. These negative verbs are often used in the beginning of imperative (hortative, jussive) sentences in the form of a bare stem ES \textit{aɣo}, cf. \REF{ex:niv:78}, and WS/A \textit{əɣi}, cf. \REF{ex:niv:79}, or in the imperative form, cf. \REF{ex:niv:80} and \textit{əɣ-ja} [not\_want-\textsc{imp.2sg}] with the meaning ‘Don’t do it!’, which is close to the prohibitive meaning (2.2.1). In the WS and A varieties, the corresponding verb \textit{əɣi-} is always transitive and takes a verbal complement derived by the nominalizer \textit{-ɟ/-c}, cf. \REF{ex:niv:81}. In all attested WS examples, the verbal complement contains the future suffix. \citet[163]{Panfilov1965} notes that the verb \textit{əɣi-} is used rather rarely.


\ea\label{ex:niv:74}
    \gll Cʰi pʰ-əmk-aχ qo-ŋ-ɢavr-gu-jn\textbf{{}-aɣɲi-ʁaj} jaŋ-aχ oχt+ra-gu-ja. \\ 
         \textsc{2sg} \textsc{refl-}mother-\textsc{acc} be\_sick-\textsc{nmlz-neg-caus-des}{}-want-\textsc{cvb.cond} \textsc{3sg}{}-\textsc{acc} medicine+drink-\textsc{caus-imp.2sg} \\
    \glt ‘If you want your mother not to be sick, make her drink the medicine.’ (Gruzdeva FM2000: 4:152) [ES\textsuperscript{T}]
\z 

\ea\label{ex:niv:75}
    \gll Cʰi siɟ+\textbf{aɣɲ{}-ɟ=ŋa}? \\ 
         2\textsc{sg} what+want-\textsc{ind=q} \\
    \glt ‘What do you want?’ (\citealt{SaveljevaTaksami1970}: 494) [A]
\z 

\ea\label{ex:niv:76}
    \gll Ɲi \textbf{aɣu-d=ra} ɲin nud+xeraj-d-ɣun jaŋ-aχ j-ajm-ɢavr-gu-jnə-d.\\ 
         \textsc{1sg} not\_want-\textsc{ind=foc} 1\textsc{pl.excl} what+say-\textsc{ind-pl} \textsc{3sg}{}-\textsc{acc} \textsc{3sg}{}-know-\textsc{neg-caus-des-nmlz}\\
    \glt ‘I don’t want him to know what we said.’ (Gruzdeva FM1991: 3:10a) [ES\textsuperscript{T}]
\z 

\ea\label{ex:niv:77}
    \gll Ɲi vi-jnə{}-f-toχ \textbf{aɣu-d}. \\ 
         \textsc{1sg} go-\textsc{des-nmlz.loc-dat} not\_want-\textsc{ind} \\
    \glt ‘I don’t want to go.’ (Gruzdeva FM1991: 3:23, 193) [ES\textsuperscript{T}]
\z 

\ea\label{ex:niv:78}
    \gll Huŋ+unir̥k it-t. \textbf{aɣo} jan ətk+χa-jnə{}-d… \\
         that+cannibal say-\textsc{ind} not\_want why.\textsc{2sg} father+shoot-\textsc{des-ind}\\
    \glt ‘This cannibal says: “Don’t do [that]! Why are you going to shoot at the father...”’ (\cite{KreinovichMs}: D824:12v–15, «Indən. Tʰəlgur̥») (an adaptation of an old text into modern ES by G. I. Paklina) [ES]
\z 

\ea\label{ex:niv:79}
    \gll \textbf{Əɣi} huk-toχ vi-to qʰaw-ja. \\ 
         not\_want there-\textsc{dat} go-\textsc{dat} \textsc{neg}{}-\textsc{imp.2sg} \\
    \glt ‘Don’t do that! Don’t go there!’ (Gruzdeva FM2019: 1:24) [WS]
\z 

\ea\label{ex:niv:80}
    \gll Nirǝvun-kun polax-ŋa. \textbf{axgu-ja} izʲn+aɲrej-kun+var-kov-r morxov-ja. \\ 
         man-\textsc{pl} cry-\textsc{cvb.sim} not\_want-\textsc{imp.2sg} \textsc{3pl}+wife-\textsc{pl}+rot-\textsc{neg-cvb.narr.2sg} be\_alive-\textsc{imp.2sg}\\
    \glt ‘People crying: “Don’t do [that]! Leave their wives alive!”’ (\cite{KreinovichMs}: D824:12v–15, «Indən. Təlgund») [ES]
\z 

\ea\label{ex:niv:81}
    \gll Hoʁoŋan j-acik i-ɲ{}-nə{}-ɟ=ri\textbf{+əɣi-ɟ=ra}. \\ 
         then \textsc{3sg}{}-younger\_brother \textsc{3sg}{}-eat-\textsc{fut-nmlz=emph}+not\_want-\textsc{ind=foc} \\
    \glt ‘Then her younger brother didn’t want to eat.’ \citep[163]{Panfilov1965} [WS]
\z 

In all varieties, there is also another lexicalized intransitive negative verb \textit{ker-} with a slightly different meaning ‘not want, refuse’. Due to scarcity of examples of its use, the syntactic behavior of \textit{ker-} is not completely clear. Thus, in ES example \REF{ex:niv:82}, \textit{ker-d=ra} behaves as the predicate of a separate clause, whereas in A example \REF{ex:niv:83}, \textit{ker-ɟ} takes oblique complements in the dative case:


\ea\label{ex:niv:82}
    \gll Ɲi cʰi+lu-d \textbf{ker-d=ra}. \\ 
         \textsc{1sg} 2\textsc{sg}+sing-\textsc{nmlz/ind} not\_want-\textsc{ind=foc} \\
    \glt ‘I don’t want you to sing.’ (\cite{KreinovichMs}: D824:12v–15, «Indən. Tʰəlgur̥») (an adaptation of an old text into modern ES by G. I. Paklina) [ES]
\z 

\ea\label{ex:niv:83}
    \gll Ɲi cʰax+ta-inə{}-dox \textbf{ker-ɟ}. \\ 
         \textsc{1sg} water+drink-\textsc{des-dat} not\_want-\textsc{ind} \\
    \glt ‘I don’t want to drink water.’ \citep[162]{Panfilov1965} [A]
\z 

The corresponding transitive form of \textit{ker-} would be \textit{ɣer-}, which, however, is not attested in its free form, but which must be the source of the bound root \textit{-ger-} with a similar meaning. The latter is used in a synthetic modal negative form in which \textit{-ger-} is synthesized with a nominalized lexical verb, as in \textit{it-ŋ-ger-d} ‘refused to say’, cf. \REF{ex:niv:84}. This construction forms a parallel to the basic synthetic forms with the verb \textit{ʁavr-} (see 2.1.1). In the A variety, the root of the negative verb is often preceded by the epenthetic vowel /i/, cf. \REF{ex:niv:85}, or by an emphatic element \textit{-Ci}\footnote{This emphatic element has the variants \textit{-ci}, \textit{-zi}, \textit{-ɟi}.}or \textit{-Tə}, cf. \REF{ex:niv:86}.

\ea\label{ex:niv:84}
    \gll Ja-mam j-otot-χajnapə jaŋ nud=ziŋ+\textbf{it-ŋ-ger-d}. \\ 
         \textsc{3sg-}old\_woman \textsc{3sg}{}-ask-\textsc{cvb.conc} \textsc{3sg} what=\textsc{emph}+say-\textsc{nmlz}{}-not\_want-\textsc{ind} \\
    \glt ‘Though his wife asked [him], he refused to say anything.’ (Gruzdeva FM1991: 3:33:274) [ES\textsuperscript{T}]
\z 

\ea\label{ex:niv:85}
    \gll Kʰe \textbf{vərk-ku-i-ger-t} se-u-tʰa-ɟ. \\ 
         fishing\_net be\_spoiled-\textsc{caus-ep}{}-not\_want-\textsc{cvb.narr.3pl} dry-\textsc{tr-iter.3pl-ind} \\
    \glt ‘(They) dry fishing nets (because they) don’t want to spoil them.’ \citep[162]{Panfilov1965} [A]
\z 

\ea\label{ex:niv:86}
    \gll Hemaɲax ɲ{}-ax \textbf{kʰrəu-gu-ɟi-ger-ɟ=ra}. \\ 
         old\_woman \textsc{1sg}{}-\textsc{acc} rest-\textsc{caus-emph}{}-not\_want-\textsc{ind=foc} \\
    \glt ‘The old woman doesn’t want to let me rest.’ \citep[162]{Panfilov1965} [A]
\z 

A similar meaning of non-desire is expressed in the A variety by the synchronically bound verb -\textit{mol}(\textit{o})-, cf. \REF{ex:niv:87}:

\ea\label{ex:niv:87}
    \gll V-aɲχ j-oc-ʁar \textbf{ici-mol-ra}. \\ 
         \textsc{3sg-}wife \textsc{3sg-}ask-\textsc{cvb.sbjv} answer-not\_want-\textsc{evid.dir} \\
    \glt ‘Though his wife asked, (he) didn’t want to answer.’ \citep[162]{Panfilov1965} [A]
\z 

\subsubsection{Non-knowledgeability}

Knowledgeability versus non-knowledgeability is most commonly expressed by the pair of transitive verbs ES \textit{j-ajm-}, cf. \REF{ex:niv:88}, WS/A \textit{j-əjm-/-həjm-} ‘know [sb./sth.]’, cf. \REF{ex:niv:89}, vs. ES \textit{j-aɣzu-}, cf. \REF{ex:niv:90}, WS \textit{j-əɣzə-}, A \textit{j-əɣzu-}, cf. \REF{ex:niv:91}, ‘not know [sb./sth.]’. The negative verb obviously gave rise to the negative habitual suffix \textit{-Ksu} (see \sectref{sec:niv:2.1.3}). The verbs take either a nominal, cf. \REF{ex:niv:91} or a clausal complement, cf. \REF{ex:niv:92}. The WS folklore texts demonstrate that the corresponding lexical meaning can be expressed by a bare verb stem, which retains its transitivity feature, cf. \textit{əɣzu} in \REF{ex:niv:92}.

\ea\label{ex:niv:88}
    \gll Ɲiɣvŋ-gun \textbf{j-ajm-d-ɣun} tʰəlv-ux u-xmu-d-ɣun. \\ 
         man-\textsc{pl} \textsc{3sg}{}-know-\textsc{ind-pl} far\_away-\textsc{abl} \textsc{recp}{}-fight-\textsc{nmlz-pl} \\
    \glt ‘People knew [that] [they] fought far away.’ (Gruzdeva FM1991: 3:2) [ES\textsuperscript{N}]
\z 

\ea\label{ex:niv:89}
    \gll Nuknuko cʰ-oːla paɲd-r pil-ŋan cʰi+\textbf{həjm-ɟ=ra}.\\ 
          stop \textsc{2sg}{}-child grow\_up-\textsc{cvb.narr.3sg} be\_big-\textsc{cvb.sim} \textsc{2sg}+know-\textsc{ind=foc}\\
    \glt ‘Wait, when your child growing up will become big, [he]’ll get to know you.’ \citep[228]{Panfilov1965} [A]
\z 

\ea\label{ex:niv:90}
    \gll Ɲi \textbf{j-aɣzu-d=ra} jaŋ hotoŋ-ux nud+r̥xop-d=lo. \\ 
         \textsc{1sg} \textsc{3sg-}not\_know-\textsc{ind=foc} \textsc{3sg} city-\textsc{abl} what+do-\textsc{nmlz}=\textsc{q} \\
    \glt ‘I don’t know what he is doing in the city.’ (Gruzdeva FM1991: 3:2a) [ES\textsuperscript{T}]
\z 

\ea\label{ex:niv:91}
    \gll Hə+ɲivx pʰ-oʁla+\textbf{əɣzu-ɣət-ra} oʁla pʰ-ətək+\textbf{əɣzu-ɣət-ra}. \\ 
         that+man \textsc{refl-}child+not\_know-\textsc{compl-coord.3sg} child \textsc{refl-}father+not\_know-\textsc{compl-coord.3sg} \\
    \glt ‘This man doesn’t know his son at all. The child doesn’t know his father at all.’ \citep[161]{Panfilov1965} [WS]
\z 

\ea\label{ex:niv:92}
    \gll Lumr̥+cʰəur̥ i-z-nə{}-ɟ+\textbf{əɣzu}. \\ 
         sable+take\_out \textsc{3sg}{}-skin-\textsc{fut-nmlz}+not\_know \\
    \glt ‘Taking out a sable [from a trap], [he] doesn’t know how to skin it.’ \citep[248]{Panfilov1965} [WS]
\z 

The ES variety has yet another negative verb \textit{qoʁr-} ‘not know’, which appears rarely, though it is attested in several elicited examples, cf. \REF{ex:niv:93}:


\ea\label{ex:niv:93}
    \gll Ɲ=at \textbf{qoʁr-d=ra} naf tʰak-toχ vi-j-d=lo. \\ 
         \textsc{1sg}=\textsc{foc} not\_know-\textsc{ind=foc} now where-\textsc{dat} go-\textsc{fut-nmlz=q} \\
    \glt ‘I don’t know where to go now.’ (Gruzdeva FM2000: 4:118) [ES\textsuperscript{T}]
\z 

The verb WS \textit{laja-}, A \textit{laji-}\footnote{The negative verb A \textit{laji-} ‘not know’ is mentioned only in \citet[161]{Panfilov1965} without any examples.}, whose meaning also refers to the sphere of non-knowledgeability, implies less categorical negation and rather infers doubt than a complete lack of knowledge, cf. \REF{ex:niv:94}:

\ea\label{ex:niv:94}
    \gll \textbf{Tʰajʁo-ra} \textbf{laja-ɟ=ra}. \\ 
         not\_know-\textsc{evid.dir} doubt-\textsc{ind=foc} \\
    \glt ‘[I] don’t know, [I] doubt.’ (Gruzdeva FM2019: 4) [WS]
\z 

Additionally, there is a negative verb ES, A \textit{tʰajʁo-} ‘not know’ (which can appear in the form \textit{tʰaj}). It is typically used as a one-word clause either in a sentence-final, cf. \REF{ex:niv:95}, or sentence-initial position, cf. \REF{ex:niv:94} and \REF{ex:niv:96}. In the latter case it is typically used in a direct evidential form and is followed by another more explicit negative clause, or alone as an answer to a question.

\ea\label{ex:niv:95}
    \gll Pʰi tʰanx vəkz-d=ziŋ \textbf{tʰajʁo-ʁar-d=la?} \\ 
         \textsc{refl} where lose\textsc{{}-nmlz=emph} not\_know-\textsc{compl-ind=q} \\
    \glt ‘Don’t you know at all where you lost [it]?’ (FM2000: 5:60) [ES\textsuperscript{N}]
\z 

\ea\label{ex:niv:96}
    \gll \textbf{Tʰajʁo-ŋra} jaŋ pʰ-ətk+voci-d=lu. \\ 
          not\_know-\textsc{evid.dir} \textsc{3sg} \textsc{refl}{}-father+resemble-\textsc{ind=q}\\
    \glt ‘I don’t know [if] he resembles his father.’ (Gruzdeva FM1991: 3:10) [ES\textsuperscript{N}]
\z 

The word A \textit{pʰanpara} is a bare stem of a verb, which is often translated as ‘not know’ or ‘not understand’. It seems to be impossible to find any examples of its use in the inflected form, besides a dictionary entry in \citet[282]{SaveljevaTaksami1970}, where it is represented in the indicative form \textit{pʰanpara-ɉ}. In all attested examples, the verb occurs sentence-finally and behaves as an extra-sentence one-word clause, cf. (\ref{ex:niv:97}--\ref{ex:niv:98}):

\ea\label{ex:niv:97}
    \gll Hoŋgur̥ r̥ak-tox vi-ɟ=lu \textbf{pʰanpara}. \\ 
         thus.\textsc{3sg} where-\textsc{dat} go-\textsc{ind=q} not\_know \\
    \glt ‘Thus, [he] doesn’t know where to go.’ \citep[244]{Panfilov1965} [WS]
\z 

\ea\label{ex:niv:98}
    \gll I-z-nə{}-ɟ+əɣzu jaŋgu-nə{}-ɟ \textbf{pʰanpara}. \\ 
         \textsc{3sg}{}-skin-\textsc{fut-nmlz}+not\_know how\_do-\textsc{fut}-\textsc{ind} not\_know \\
    \glt ‘[He] doesn’t know how to skin [it], doesn’t know how to do it.’ \citep[248]{Panfilov1965} [WS]
\z 

\subsubsection{Non-attraction}

Attraction vs. non-attraction is expressed in all varieties by the pair of transitive verbs \textit{e-zmu-} ‘like’, cf. \REF{ex:niv:99}, vs. \textit{e-sqa-} ‘not like’, cf. \REF{ex:niv:100}:

\ea\label{ex:niv:99}
    \gll Andχ-kun pʰrə{}-ba ɲi \textbf{e-zmu-tʰa-ɟ}. \\ 
         guest-\textsc{pl} come-\textsc{cvb.imm} \textsc{1sg} \textsc{3sg}{}-like-\textsc{iter.1sg-ind} \\
    \glt ‘I like when the guests come.’ (Gruzdeva FM2000: 4:122) [ES\textsuperscript{T}]
\z 

\ea\label{ex:niv:100}
    \gll Cʰi \textbf{e-sqa-ʁaj} ɲi=pər̥k i-ɲ{}-i-d=ra. \\ 
         \textsc{2sg} \textsc{3sg}{}-not\_like-\textsc{cvb.cond} 1\textsc{sg}=only \textsc{3sg}{}-eat-\textsc{fut-ind=foc} \\
    \glt ‘If you don’t want to eat, I’ll eat myself.’ (\cite{KreinovichMs}: D824:12v–15, «Indən. Tʰəlgur̥») (an adaptation of an old text into modern ES by G. I. Paklina) [ES]
\z 

\subsubsection{Non-ability}

Negative lexicalized verbs that covertly express the meaning of non-ability can be used independently only in the Amur varieties. The verb A \textit{j-iki-} ‘be unable’ is also used predominantly in the bound form, but there are still a few examples of its independent use, cf. \REF{ex:niv:101}. When synthesizing with the lexical verb, \textit{j-iki-} loses the initial \textit{j} which marks transitivity, cf. \REF{ex:niv:102}:\footnote{The \textit{j} seen in this example is epenthetic and not the 3\textsc{sg} marker.}


\ea\label{ex:niv:101}
    \gll Ola ɲi pʰer-ɟ=ra. Cʰe-rʁa \textbf{j-iki-ɟ=ra}. \\ 
         son.\textsc{voc} \textsc{1sg} be\_tired-\textsc{ind=foc} 2\textsc{sg-dest} \textsc{3sg}{}-be\_unable-\textsc{ind=foc} \\
    \glt ‘Son! I’m tired. I can’t cope with you.’ \citep[225]{Panfilov1965} [A]
\z 

\ea\label{ex:niv:102}
    \gll  Hə+baχ per-ʁa ɲi xici-j-\textbf{iki-nə{}-ɟ=ra}.\\ 
         that+stone be\_heavy-\textsc{cvb.cond} \textsc{1sg} lift-\textsc{ep}{}-be\_unable\textsc{{}-fut-ind=foc} \\
    \glt ‘If this stone is heavy, I won’t be able to lift it.’ \citep[162]{Panfilov1965} [A]
\z 

The verb ES/A \textit{j-ali-} with a similar meaning is, on the contrary, used only as a full free transitive verb, cf. \REF{ex:niv:103}, which can take a clausal complement, cf. \REF{ex:niv:104}.

\ea\label{ex:niv:103}
    \gll Ketv-r \textbf{tʰəŋr̥əŋa-ɟ}. Hoŋke \textbf{j-ali-ɟ}. \\ 
         stick-\textsc{cvb.narr.3sg} not\_move-\textsc{ind} so \textsc{3sg}{}-be\_unable-\textsc{ind} \\
    \glt ‘Having stuck [it] doesn’t move. So [he] wasn’t able [to move] it.’ \citep[226]{Panfilov1965} [A]
\z 

\ea\label{ex:niv:104}
    \gll If toʁa+mu+lət-c+\textbf{ali-ɟ}. \\ 
         \textsc{3sg} be\_like\_this+boat+make-\textsc{nmlz}+be\_unable-\textsc{ind} \\
    \glt ‘He wasn’t able to make such a boat.’ (\citealt{SaveljevaTaksami1970}: 497) [A]
\z 

In the ES variety, the meaning of non-ability can be expressed by synthetic negative forms with the bound form of the verb \textit{-rer-} ‘be unable’, cf. \REF{ex:niv:105}. This verb has an intransitive counterpart \textit{ter-} with the meaning ‘be weak’.

\ea\label{ex:niv:105}
    \gll Urgur qʰo-jnə{}-ŋa \textbf{qʰo-rer-d}. \\ 
         well.\textsc{3sg} sleep-\textsc{des-cvb.sim} sleep-be\_unable\textsc{{}-ind} \\
    \glt ‘When [she] wanted to sleep well, [she] couldn’t sleep.’ (Gruzdeva FM1991: 3:14a) [ES\textsuperscript{T}]
\z 

A similar meaning ‘not achieve’ is rendered by the synchronically bound element ES \textit{-}(\textit{i})\textit{roter-}, A \textit{-}(\textit{i})\textit{rucez-}. This marker goes back to the converb suffix \textit{-iro} marking an interrupted action and the modal verb ES \textit{ter-}, A \textit{cez-} ‘be weak’ \citep[163]{Panfilov1965}, cf. (\ref{ex:niv:106}--\ref{ex:niv:107}):

\ea\label{ex:niv:106}
    \gll \textbf{Tlo-jroter-t}. \\ 
         run-not\_achieve-\textsc{ind} \\
    \glt ‘[He] couldn’t run.’ (Gruzdeva FM2000: 4:13) [ES\textsuperscript{T}]
\z 

\ea\label{ex:niv:107}
    \gll Kʰlə{}-rx pʰu-r̥ vuu-ɟ. \textbf{Vuu-rucez-ɟ=ra}. \\ 
         outside-\textsc{dat} go\_out-\textsc{cvb.narr.3sg} cry-\textsc{ind} cry-not\_achieve\textsc{{}-ind=foc} \\
    \glt ‘Going out (he) shouted. (He) didn’t get the answer.’ \citep[163]{Panfilov1965} [A]
\z 

\subsubsection{Other negative lexicalizations}

The motion intransitive negative verb A \textit{tʰəŋr̥əŋa-} ‘not move’ has been exemplified by \REF{ex:niv:103}.

The verb ES \textit{amtrer-}, A \textit{amcez-} has the meaning ‘not have time, be late’ and is possibly connected, again, with the modal verb ES \textit{ter-}, A \textit{cez-} ‘be weak’ \citep[163]{Panfilov1965}, cf. \REF{ex:niv:108}:

\ea\label{ex:niv:108}
    \gll Hoʁaŋan hə+ɲivɣ+erq-ux oz-r qʰaχ-kir e-sp-inə-ke \textbf{amcez-ɟ}.\\ 
         then that+man+side-\textsc{abl} get\_up-\textsc{cvb.narr.3sg} spear-\textsc{ins} \textsc{3sg}{}-stab-\textsc{des-cvb.dur} not\_have\_time-\textsc{ind} \\
    \glt ‘When that man getting up wanted to stab him with a spear, [he] didn’t have time.’ \citep[163]{Panfilov1965} [A]
\z 

One should also mention the WS/A pair \textit{naw-} ‘be enough’, cf. \REF{ex:niv:109}, and \textit{roj-} ‘not be enough’, cf. \REF{ex:niv:110}:

\ea\label{ex:niv:109}
    \gll Ɲi cʰχa+vo-ɟ. Ɲe-rχ \textbf{naw-ɟ}. \\ 
         \textsc{1sg} money+get-\textsc{ind} 1\textsc{sg}{}-\textsc{dat} be\_enough-\textsc{ind} \\
    \glt ‘I got money. I have enough.’ (Gruzdeva FM2019: 1:7) [WS]
\z 

\ea\label{ex:niv:110}
    \gll Jeski-ŋan maɲɟu əs \textbf{roi-ɟ}. Uski \textbf{roi-ɟ}.\\ 
         sell-\textsc{cvb.sim} Chinese money not\_be\_enough-\textsc{ind} payment not\_be\_enough-\textsc{ind}\\
    \glt ‘When [he] was selling, the Chinese didn’t have enough money. [There] was not enough payment. \citep[250]{Panfilov1965} [WS]
\z 

There are also verbs ES \textit{j-akru-}, A \textit{j-ǝkru-} ‘not believe’, cf. \REF{ex:niv:111}, and ES \textit{j-axraw-} ‘not trust’, cf. \REF{ex:niv:112}:

\ea\label{ex:niv:111}
    \gll Ɲi pʰi-dǝ-ɢavr-χaj ena-gun+\textbf{akru-j-d}. \\ 
         \textsc{1sg} \textsc{refl}{}-see-\textsc{neg-cvb.cond} be\_another\textsc{{}-pl}+not\_believe-\textsc{fut-ind} \\
    \glt ‘If I hadn’t seen myself, I wouldn’t believe the others.’ (Gruzdeva FM2000: 4:18) [ES\textsuperscript{T}]
\z 

\ea\label{ex:niv:112}
    \gll Ɲi cʰ-\textbf{axraw-d}=ra. \\ 
         \textsc{1sg} \textsc{2sg}{}-not\_trust-\textsc{ind=foc} \\
    \glt ‘I don’t trust you.’ (Gruzdeva FM2000: 4:43) [ES\textsuperscript{T}]
\z 


\section{Non-clausal negation}\label{sec:niv:3} 

This section deals with negative replies (\sectref{sec:niv:3.1}), negative emphatics and indefinites (\sectref{sec:niv:3.2}), and the ways of expressing caritive semantics and antonymic low-level negation (\sectref{sec:niv:3.3}).

\subsection{Negative replies}\label{sec:niv:3.1} 

In negative replies to polar questions, \textit{qʰawk} or \textit{qʰawkra} ‘no’ is used as the negative counterpart to \textit{hǝ} or \textit{hǝŋa} ‘yes’. The same negative word is used as a reaction to some proposal or in another appropriate context, cf. (\ref{ex:niv:113}--\ref{ex:niv:115}), but it cannot be used as a reply to content questions. Interestingly, this negative reply, which represents a nominalized form of the negative verb \textit{qʰaw-}, is used in all Nivkh varieties, including those in which the negative verb is used only in the form of \textit{qavr-}. Most likely, it has been a secondary borrowing from A varieties to the Sakhalin ones. As a result of monophthongization, in the WS variety, this word is currently pronounced as \textit{qaːk}. Besides the neutral focus clitic \textit{=ra}, it can attach the emphatic focus clitic\textit{=rej}, as in \REF{ex:niv:115} in which one speaker quotes two other people. As one can see, Nivkh follows the general tendency to use a negative existential marker as a standalone negative answer \citep{Veselinova2013neg.exist}

\ea\label{ex:niv:113}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:niv:113Q}
        \gll  Cʰi pʰ-r̥omsk cʰχa+r̥o-l?  \\ 
             2\textsc{sg} \textsc{refl}{}-together money+carry-\textsc{q} \\
        \glt ‘Do you have money on you?’
    \exi{A: }\label{ex:niv:113A}
        \gll \textbf{Qʰaw-k}. \\ 
            \textsc{neg-nmlz.agt} \\
        \glt ‘No.’ (Gruzdeva FM2000: 1:43) [ES\textsuperscript{N}]
    \end{xlist}
\z 

\ea\label{ex:niv:114}
    \gll Hoʁaŋan if it-c. \textbf{Qʰau-k=ra} ɲi cʰəŋ-doχ ler-inə-t pʰrə-ɟ=ra. \\ 
         then \textsc{3sg} say-\textsc{ind} \textsc{neg}{}-\textsc{nmlz.agt}=\textsc{foc} \textsc{1sg} 2\textsc{pl-dat} play-\textsc{des-cvb.narr.1sg} come-\textsc{ind=foc} \\
    \glt ‘Then he said: “No, I came to compete with you.”’ \citep[229]{Panfilov1965} [A]
\z 


\ea\label{ex:niv:115}
    \gll Haŋa naf asʲxa ŋamgur it-nd. Hala vi-nate. \\ 
        then now younger\_brother tit say-\textsc{ind} well go-\textsc{imp.1du}\\
    \glt ‘Then the younger brother tit said: Well, let’s go!’  \medskip\newline
    \gll \textbf{Kau-k=rej}[,] ɲi r̥as-to vi-i-ra[,] tund ɲ-raf=ta[.]\\ 
        \textsc{neg-nmlz.agt=foc\_emph} \textsc{1sg} where-\textsc{dat} go-\textsc{fut-evid.dir} this\_one \textsc{1sg}{}-house=\textsc{foc}\\
    \glt ‘[The bird-woman said:] No, I won’t go anywhere. This is my house.’ \medskip\newline
    \gll  Vi-i-nd+agu-ʁaj ɲi cʰ-xe-i-nd=ra. \\ 
         go-\textsc{fut-nmlz}+not\_want-\textsc{cvb.cond} \textsc{1sg} \textsc{2sg-}take-\textsc{fut-ind=foc}\\
    \glt ‘[The younger brother said:] If [you] don’t want to go, I’ll take you.’ \medskip\newline
    \gll \textbf{Kau-k=ra}[,] təŋ+keŋ morxa-ra[,] loŋ morxa-ra[,] ɲi vi-n-tox \textbf{kau-k=ra}[.] Ɲi əziŋ+iv-nd=ra[,]\\ 
        \textsc{neg-nmlz.agt=foc} this+sun be\_alive-\textsc{coord.3sg} moon be\_alive-\textsc{coord.3sg} \textsc{1sg} go-\textsc{nmlz}{}-\textsc{dat} \textsc{neg-nmlz.agt=foc} \textsc{1sg} master+exist-\textsc{ind=foc} \\
    \glt ‘No, this sun is alive, the moon is alive, I won’t go. I have a master,’   \medskip\newline 
    \gll Ɲi cangi+iv-nd=ra[,] ajzi-ŋ+cʰevr̥kx+r̥anx it-nd. \\ 
         \textsc{1sg} boss+exist\textsc{{}-ind=foc} be\_golden-\textsc{nmlz}+bird+woman say-\textsc{ind} \\
    \glt ‘I have a boss, the golden bird-woman said.’ (\cite{KreinovichMs}: D540:7–15+22–24v, «Nuɣə kʰen ɲaqr̥… [1]») [ES]
\z


\subsection{Negative indefinites}\label{sec:niv:3.2} 

Nivkh does not have inherently negative proforms, since those proforms which are used in the scope of negation may also occur in non-negative contexts. In all varieties, they are based on interrogatives, but rely on different clitics that are attached to the derivational base.

In the ES variety, proforms which are used in negative predications are essentially emphatics, cf. \REF{ex:niv:116}, and are formed from interrogatives with the emphatic clitic \textit{=Ciŋ}, which is otherwise used for reinforcing negation (\sectref{sec:niv:4.4}), cf. \textit{nar̥} ‘who’ > \textit{nar̥=ciŋ} ‘who (exactly), nobody’, cf. \REF{ex:niv:117}, \textit{tʰawd} ‘which one’ > \textit{tʰawd=ziŋ} ‘which one (exactly), no-one’, cf. \REF{ex:niv:118}, \textit{nud} ‘what’ > \textit{nud=ziŋ} ‘what (exactly), nothing’, cf. \REF{ex:niv:119}, \textit{tʰanx} ‘where (non-precisely)’ > \textit{tʰanx=ciŋ} ‘where (non-precisely, exactly), nowhere’, cf. \REF{ex:niv:120}, \textit{tʰas} ‘where (precisely)’ > \textit{tʰas=ciŋ} ‘where (precisely, exactly), nowhere’, cf. \REF{ex:niv:41}, ES\textsuperscript{T} \textit{aɣr̥}, ES\textsuperscript{N} \textit{əɣr̥} ‘when’ > ES\textsuperscript{T} \textit{aɣr̥=ciŋ}, cf. \REF{ex:niv:121}, ES\textsuperscript{N} \textit{əɣr̥=ciŋ}, cf. \REF{ex:niv:122}, ‘when (exactly), never’, \textit{tʰans} ‘how many’ > \textit{tʰans=ciŋ} ‘how many (exactly), none’, etc. All inflectional markers that are taken by a proform come before the emphatic clitic.

\ea\label{ex:niv:116}
    \gll It-ja \textbf{aɣr̥=ciŋ} pʰr̥ə-j-d. \\ 
         say-\textsc{imp.2sg} when=\textsc{emph} come-\textsc{fut-ind} \\
    \glt ‘Tell [me], when [exactly] you’ll come.’ [ES]
\z 

\ea\label{ex:niv:117}
    \gll Tu+daf \textbf{nar̥=ciŋ}+\textbf{iv-ɢavr-d}. \\ 
         this+house who=\textsc{emph}+exist-\textsc{neg-ind} \\
    \glt ‘There’s nobody in this house.’ (Gruzdeva FM1991: 3:9) [ES\textsuperscript{T}]
\z 

\ea\label{ex:niv:118}
    \gll Toqr̥oq-ŋ-ɢavr-ve. Sikm oz-ʁar-d-ɣun=da. Cʰin  \textbf{tʰawd=ziŋ}+\textbf{oz-u-ɢavr-i-d-ɣun=da}.\\ 
         sneak\_up-\textsc{nmlz}{}-\textsc{neg-imp.2pl} all get\_up-\textsc{compl-ind-pl=foc} 2\textsc{pl} which=\textsc{emph}+get\_up-\textsc{tr-neg-fut-ind-pl=foc}\\
    \glt ‘Don’t sneak up! Everyone [has] got up. You won’t wake up anyone.’ (Gruzdeva FM1991: 3:40) [ES\textsuperscript{T}]
\z 

\ea\label{ex:niv:119}
    \gll \textbf{Nud=ziŋ} huŋ+vo-wx \textbf{ujɣi-d}. \\ 
         what=\textsc{emph} that+settlement-\textsc{abl} not\_be-\textsc{ind} \\
    \glt ‘There was nothing in that village.’ (Gruzdeva FM1991: 3:43:360) [ES\textsuperscript{T}]
\z 

\ea\label{ex:niv:120}
    \gll Huz urla-ŋ+iv-v-ux cʰo-ɣun qavr-χaj \textbf{tʰanx=ciŋ} \textbf{qavr-χar-i-d}. \\ 
         that\_place be\_good-\textsc{nmlz}+exist-\textsc{nmlz.loc-abl} fish-\textsc{pl} \textsc{neg-cvb.cond} where=\textsc{emph} \textsc{neg-compl-fut-ind} \\
    \glt ‘If there’s no fish in this good place, there’ll be no [fish] anywhere.’ (Gruzdeva FM2000: 4:98) [ES\textsuperscript{T}]
\z 

\protectedex{\ea\label{ex:niv:121}
    \gll Ɲin eχkun hoton-doχ vi-j-ʁaj ɲi \textbf{aɣr̥=ciŋ} huŋ+r̥aŋɢeʁlŋ+\textbf{ɲr̥ə-ɢavr-i-d}. \\ 
         \textsc{1pl.excl} soon city-\textsc{dat} go-\textsc{fut-cvb.cond} \textsc{1sg} when=\textsc{emph} that+girl+see-\textsc{neg-fut-ind} \\
    \glt ‘If we soon go to the city, I’ll never see that girl.’ (Gruzdeva FM1991: 3:1a) [ES\textsuperscript{T}]
\z}

\ea\label{ex:niv:122}
    \gll \textbf{Nar̥=ciŋ} \textbf{əɣr̥=ciŋ} \textbf{ɲ-zaʁo-ɢavr-i-d}. \\ 
         who=\textsc{emph} when=\textsc{emph} \textsc{1sg-}deceive-\textsc{neg-fut-ind} \\
    \glt ‘Nobody will ever deceive me.’ (Gruzdeva FM1991: 3:13) [ES\textsuperscript{N}]
\z 

In other varieties, proforms which are used in negative predications have the same form as free-choice indefinites, cf. \REF{ex:niv:123}. Free-choice indefinites are themselves derivations from interrogatives. The corresponding interrogative attaches the zero-nominalized form of the auxiliary verb \textit{ha-} in the instrumental case. The initial fricative of the auxiliary can be omitted, and the resulting form depends on the shape of the instrumental suffix, which is used in one or another variety, cf. WS/A \textit{aŋ} ‘who’ > WS \textit{aŋ=ha-kisk/-gisk/-kirk}, A \textit{aŋ=ha-gin/-girn} ‘anybody, everybody, nobody’, cf. \REF{ex:niv:124}, WS/A \textit{r̥aɟ} ‘which one’ > WS \textit{r̥aɟ=ha-kisk/-gisk/-kirk}, A \textit{r̥aɟ=ha-gin/-girn} ‘anybody, everyone, no-one’, cf. \REF{ex:niv:125}, WS/A \textit{siɟ} ‘what’ > WS \textit{siɟ=ha-kisk/-gisk/-kirk}, A \textit{siɟ=ha-gin/girn} ‘anything, everything, nothing’, cf. \REF{ex:niv:126}, WS/A \textit{r̥atx} ‘where’ > WS \textit{r̥atx=ha-kisk/-gisk/-kirk}, A \textit{r̥atx=ha-gin/-girn} ‘anywhere, everywhere, nowhere’, cf. \REF{ex:niv:127}, WS/A \textit{əɣr} ‘when’ > WS \textit{əɣr=ha-kisk/-gisk/-kirk}, A \textit{əɣr=ha-gin/-girn} ‘whenever, never’, cf. \REF{ex:niv:128}, WS/A \textit{r̥aŋs} ‘how many’ > WS \textit{r̥aŋs=ha-kisk/-gisk/-kirk}, A \textit{r̥aŋs=ha-gin/-girn} ‘however many, none’, cf. \REF{ex:niv:129}, etc. All inflectional markers come before the clitic, cf. \textit{r̥ak-tox=hagirn} ‘to nowhere’. The elements WS \textit{=ha-kisk/-gisk/-kirk}, A \textit{=ha-gin/-girn} can be attached to other nominals to reinforce negation (\sectref{sec:niv:4.4}).

\ea\label{ex:niv:123}
    \gll Cʰi \textbf{aŋ=hakisk}+oz-u-toχ qʰaw-nə-ɟ=ra. \\ 
         2\textsc{sg} who=\textsc{emph}+get\_up-\textsc{tr-dat} \textsc{neg-fut-ind=foc} \\
    \glt ‘You won’t wake up anyone.’ (Gruzdeva FM2019: 1:7) [WS]
\z 

\ea\label{ex:niv:124}
    \gll  Təɟ \textbf{aŋ=hagin} \textbf{lət-nə-ʁavr-ɟ}. \\ 
         this\_one who=\textsc{emph} do-\textsc{fut-neg-ind} \\
    \glt ‘Nobody will do this.’ \citep[165]{Panfilov1965} [A]
\z 

\ea\label{ex:niv:125}
    \gll Imŋ-ux \textbf{r̥aɟ=hagin} togut \textbf{nə{}-nə{}-ʁavr-ɟ}. \\ 
         3\textsc{pl}{}-\textsc{abl} which=\textsc{emph} in\_such\_way:\textsc{3pl} do{}-\textsc{fut-neg}{}-\textsc{ind} \\
    \glt ‘None of them will do so.’ (\citealt{SaveljevaTaksami1970}: 319) [A]
\z 

\ea\label{ex:niv:126}
    \gll \textbf{Siɟ=akirk} ivŋ+\textbf{ar-xsu-ɟ}. \\ 
         what=\textsc{emph} 3\textsc{pl}+feed-\textsc{hab\_neg{}-ind} \\
    \glt ‘They don’t feed them with anything.’ \citep[247]{Panfilov1965} [WS]
\z 

\ea\label{ex:niv:127}
    \gll Ɲi \textbf{r̥atx=hagin} \textbf{vi-nə{}-ʁavr-ɟ=ra}. \\ 
         \textsc{1sg} where=\textsc{emph} go-\textsc{fut-neg-ind=foc} \\
    \glt ‘I won’t go anywhere.’ (\citealt{SaveljevaTaksami1970}: 321) [A]
\z 

\ea\label{ex:niv:128}
    \gll Ɲi \textbf{əɣr=hakisk} \textbf{vi-nə-ɢavr-ɟ}. \\ 
         \textsc{1sg} when=\textsc{emph} go-\textsc{fut-neg-ind} \\
    \glt ‘I’ll never go [there].’ (Gruzdeva FM2019: 7) [WS]
\z 

\ea\label{ex:niv:129}
    \gll \textbf{Raŋs=hagin}+cʰχa \textbf{qʰau-ɟ}. \\ 
         how\_many=\textsc{emph}+money \textsc{neg-ind} \\
    \glt ‘There’s no money.’ (\citealt{SaveljevaTaksami1970}: 320) [A]
\z 


In the early-20\textsuperscript{th}{}-century folklore texts recorded by Kreinovich from presumably ES speakers, one can find proforms that are derived according to the WS and A model, cf. \textit{rund} ‘what’ > \textit{rund=ha-n-kir̥} [what=do\_so{}-\textsc{nmlz-ins}] ‘anything, everything, nothing’ in \REF{ex:niv:130}, \textit{əɣr} ‘when’ > \textit{əgr=a-n-kisʲk} [when=do\_so{}-\textsc{nmlz-ins}] ‘whenever, never’ in \REF{ex:niv:131}.

\ea\label{ex:niv:130}
    \gll Ɲi \textbf{rund=hankir̥} \textbf{it-n-tox} \textbf{kau-k=ra}. \\ 
         \textsc{1sg} what=\textsc{emph} say-\textsc{nmlz-dat} \textsc{neg-nmlz.agt=foc} \\
    \glt ‘I can’t say anything.’ (\cite{KreinovichMs}: D536b:13–14v, «Pilan tav ɲakr…») [ES]
\z 

\ea\label{ex:niv:131}
    \gll \textbf{Əgr=ankisʲk} \textbf{i-ndə-n+xavr-n}+ɲigəvən-kun \textbf{xavru}. \\ 
         when=\textsc{emph} \textsc{3sg}{}-see-\textsc{nmlz+neg-nmlz}+man-\textsc{pl} \textsc{neg} \\
    \glt ‘We never saw people.\footnote{This sentence can be literally translated as ‘Never people [that] we didn’t see [there] were not.’}’ (\cite{KreinovichMs}: D546:21–22, «Təlgund [6]») [ES]
\z

\subsection{Negative derivation and case marking (caritives/privatives)}\label{sec:niv:3.3}

Nivkh does not have any special nominal cases for indicating absence or non-involvement (caritive). The major strategies for expressing caritive semantics are similar to those of predicative possessive negation (\sectref{sec:niv:2.3.2}). In the ES and WS varieties, the caritive meaning is rendered by a dependent narrative clause with the converb or nominalized form of the transitive negative verb \textit{ʁavr-} as a predicate, preceded by an (unmarked object) NP expressing the absentee. Such a clause is often used either in sentence-initial topical position, as in \REF{ex:niv:132}, as an adverbial or depictive clause, as in \REF{ex:niv:133} and \REF{ex:niv:134}, or attributively, as in \REF{ex:niv:135}.


\ea\label{ex:niv:132}
    \gll \textbf{Ətk+ʁavr-r̥} ɲi=pər̥k mu+aj-ɢavr-i-d. \\ 
         father+\textsc{neg-cvb.narr.3sg} \textsc{1sg}=only boat+make-\textsc{fut-ind} \\
    \glt ‘Without [my] father, I alone won’t [be able to] build a boat.’ [ES]
\z

\ea\label{ex:niv:133}
    \gll Ɲi \textbf{ətək+ʁavr-t} pan-ɟ=ra. \\ 
         \textsc{1sg} father+\textsc{neg-cvb.narr.3sg} grow-\textsc{ind=foc} \\
    \glt ‘I grew up without a father.’ (Gruzdeva FM2019: 1:18) [WS]
\z 

\ea\label{ex:niv:134}
    \gll Jaŋgut \textbf{lums}+\textbf{qavr-t} vi-nə{}-ɟ=ŋa? \\ 
         how.\textsc{1pl} food+\textsc{neg-cvb.narr.1pl} go-\textsc{fut-ind=q} \\
    \glt ‘How will we go without food?’ (Gruzdeva FM2019: 1:19) [WS]
\z 

\ea\label{ex:niv:135}
    \gll Ɲi ŋavr̥ki+\textbf{ɢavr}-\textbf{ŋ}+ŋa+xu-d. \\ 
         \textsc{1sg} wool+\textsc{neg-nmlz}+animal+kill-\textsc{ind} \\
    \glt ‘I killed a hairless animal.’ (Fedotov FM2013) [ES]
\z 

In the same context, the WS variety also allows the use of the analytic construction in the dependent clause, cf. \REF{ex:niv:136}:

\ea\label{ex:niv:136}
    \gll Mer jaŋgut \textbf{lums} \textbf{ha-to} \textbf{qʰaw-t} vi-nə{}-ɟ=ŋa? \\ 
         \textsc{1pl}.\textsc{incl} how.\textsc{1pl} food do\_so-\textsc{dat} \textsc{neg-cvb.narr.1pl} go-\textsc{fut-ind=q}\\
    \glt ‘How will we go without food?’ (Gruzdeva FM2019: 1:20) [WS]
\z 

In the A variety, the caritive meaning is marked by a narrative clause with the inherently negative possessive verb \textit{ləɣə{}-} in predicative position and with the same clause and sentence structure as in the ES variety, cf. \REF{ex:niv:137}. This verb can be further grammaticalized into the relational noun (= postposition) \textit{ləɣə} ‘without’, which ha an affirmative counterpart \textit{Tomsk}\footnote{This relational noun has the variants \textit{tomsk}, \textit{romsk}, \textit{domsk}.} ‘with’ and can be seen as the only, relatively new, dedicated caritive marker in Nivkh, cf. \REF{ex:niv:138}.

\ea\label{ex:niv:137}
    \gll Ɲi ətək=rə+\textbf{ləɣə{}-t} pan-ɟ=ra. \\ 
         \textsc{1sg} father=\textsc{emph}+not\_have\textsc{{}-cvb.narr.1sg} grow-\textsc{ind=foc} \\
    \glt ‘I grew up without a father.’ (Gruzdeva FM2019: 1:18) [WS]
\z 

\ea\label{ex:niv:138}
    \gll Ɲəŋ vulvula+lep+\textbf{romsk} seta-də+\textbf{ləɣə}  eʁgut cʰaj+ra-ta. \\ 
         \textsc{1pl.excl} be\_black+bread+with sugar-\textsc{emph}+without quickly:\textsc{1pl} tea+drink-\textsc{coord.1pl} \\
    \glt ‘We quickly drank tea with black bread and without sugar.’ \citep[161]{Panfilov1965} [A]
\z 

Antonymic (privative) low-level negation, corresponding to English \textit{unkind} or Russian \textit{nekrasivyj} ‘not beautiful’, has no special markers in Nivkh either. It is expressed as normal negation in a subordinate clause by a negative verb in a dependent form attached to the verb expressing the quality, cf. \REF{ex:niv:139}:

\ea\label{ex:niv:139}
    \gll Vew-\textbf{ɢavr}{}-ŋ+du kəŋ-d. \\ 
         deep-\textsc{neg-nmlz}+lake freeze-\textsc{ind} \\
    \glt ‘[There are two lakes, a deep one and a shallow one.] The shallow (lit. not deep) lake has frozen’. (Fedotov FM2013) [ES\textsuperscript{T}]
\z 

\section{Other aspects of negation}\label{sec:niv:4} 

This section provides information on the scope of negation (\sectref{sec:niv:4.1}), negative polarity (\sectref{sec:niv:4.2}), marking of NPs in the scope of negation (\sectref{sec:niv:4.3}), reinforcing negation (\sectref{sec:niv:4.4}), negation in coordinated and subordinated clauses (\sectref{sec:niv:4.5}), and some diachronic observations (\sectref{sec:niv:4.6}).

\subsection{The scope of negation}\label{sec:niv:4.1} 

In Nivkh, the scope of negation is the whole clause, since negation can be marked only on the predicate. There are no negative markers that would allow narrowing of the scope to a specific constituent. All attempts to elicit the relevant examples ended in producing standard negative constructions. On the other hand, virtually any constituent can be emphasized as being in the focus of negation with the emphatic clitic, without a special intonation and without changing the word order (\sectref{sec:niv:4.4}).

\subsection{Negative polarity}\label{sec:niv:4.2} 

Basically, Nivkh does not distinguish positive and negative polarity items. The only possible exception is represented by the negative emphatics in the ES variety, which seem to be gradually shifting from the affirmative to the negative domain (\sectref{sec:niv:3.2}, \sectref{sec:niv:4.6.2}).

\subsection{Marking of NPs in the scope of negation}\label{sec:niv:4.3} 

Case marking is not affected by negation in Nivkh.

\subsection{Reinforcing negation}\label{sec:niv:4.4} 

Negation can be reinforced by the emphatic clitics, which are attached to various constituents in the clause.

The clitics WS \textit{=ha-kisk/-gisk/-kirk}, A \textit{=ha-gin/-girn} function as scalar focus markers with the meaning ‘even’, cf. \citet[158]{Haspelmath1997} and example \REF{ex:niv:140}:

\ea\label{ex:niv:140}
    \gll \textbf{If=hagin} \textbf{pʰrǝ-doχ} \textbf{qʰau-ɉ}. \\ 
         \textsc{3sg}=\textsc{emph} come-\textsc{dat} \textsc{neg-ind} \\
    \glt ‘Even he didn’t come.’ (\citealt{SaveljevaTaksami1970}: 421) [A]
\z 

Another way to emphasize a single constituent is to use the structure [interrogative + XP=\textsc{emph}], which is similar to the structure of proforms used in negative predications (\sectref{sec:niv:3.2}). In the ES variety, the resulting form has just an emphatic reading both with and without negation, whereas in the WS and A varieties in the affirmative context it has a free-choice meaning ‘every XP’, cf. \textit{siɟ + zir=hagin} ‘every tree’ in \REF{ex:niv:141}. In the scope of negation this meaning is understood as ‘no XP (whatsoever)’, cf. \textit{nud + oχt=ziŋ} [what + medicine=\textsc{emph}] ‘no medicine’ in \REF{ex:niv:142}, and \textit{siɟ + mu=hagin} ‘no boat’ in \REF{ex:niv:143}. The ES variety makes use of the emphatic clitic \textit{=Ciŋ}, whereas in the WS and A varieties, the emphatic clitic has the form WS \textit{=ha-kisk/-gisk/-kirk}, A \textit{=ha-gin/-girn}.

\ea\label{ex:niv:141}
    \gll \textbf{Siɟ+zir=hagin} tuin pan-d ur-ɟ. \\ 
         what+tree=\textsc{emph} here grow-\textsc{nmlz} be\_good-\textsc{ind} \\
    \glt ‘Every tree grows here well.’ \citep[165]{Panfilov1965} [WS]
\z 

\ea\label{ex:niv:142}
    \gll Ɲ-aki əki-qʰatn-d. \textbf{Nud+oχt=ziŋ} \textbf{ro-ŋ-ɢavr-i-d}.\\ 
         \textsc{1sg}{}-older\_brother be\_bad-\textsc{aug-ind} what+medicine=\textsc{emph} help-\textsc{nmlz}{}-\textsc{neg-fut-ind} \\
    \glt ‘My older brother [feels] very bad. No medicine will help him.’ (Gruzdeva FM1991: 3:26) [ES]
\z 

\ea\label{ex:niv:143}
    \gll \textbf{Siɟ+mu=hagin} kerq+\textbf{tozu-nə-ɢavr-ɟ}. \\ 
         what+boat=\textsc{emph} sea+cross-\textsc{fut-neg-ind} \\
    \glt ‘No boat will cross the sea.’ \citep[165]{Panfilov1965} [WS]
\z 

Coordinated constituents can be also emphasized by emphatic clitics. In case of the ES variety, it is the clitic \textit{=Ciŋ}, which can be used both in affirmative, cf. \REF{ex:niv:144}, and in negative predications, cf. (\ref{ex:niv:145}--\ref{ex:niv:146}). In the WS and A varieties, there are two possibilities. The first one is to use the clitic \textit{=Tə}, cf. \REF{ex:niv:147}, which has been discussed in other sections, especially with regard to negative synthetic forms (2.1). Alternatively, one can utilize the clitic WS \textit{=ha-kisk/-gisk/-kirk}, A \textit{=ha-gin/-girn}. This clitic can be also used both for positive, cf. \REF{ex:niv:148}, and negative, cf. (\ref{ex:niv:149}--\ref{ex:niv:150}), emphasis with the preference for the latter.


\ea\label{ex:niv:144}
    \gll Ɲi \textbf{ər̥k=ziŋ} \textbf{muɣf=ciŋ} orbot-i-d=ra pər̥k pʰ-r̥au-jnəftoχ. \\ 
         \textsc{1sg} night=\textsc{emph} day=\textsc{emph} work-\textsc{fut-ind=foc} oneself \textsc{refl-}study-\textsc{cvb.purp} \\
    \glt ‘I’ll work day and night in order to study.’ (Gruzdeva FM2000: 4:154) [ES\textsuperscript{T}]
\z 

\ea\label{ex:niv:145}
    \gll Huŋ+cʰχar+mi-x \textbf{milkmla=ziŋ} \textbf{alr̥=ciŋ} \textbf{pan-ɢavr-d}. \\ 
         that+tree+inside-\textsc{abl} mushroom=\textsc{emph} berry=\textsc{emph} grow-\textsc{neg-ind} \\
    \glt ‘Neither mushrooms, nor berries grow in that forest.’ (Gruzdeva FM1991: 3:15) [ES\textsuperscript{T}]
\z 

\ea\label{ex:niv:146}
    \gll Cʰi \textbf{taf=ciŋ} \textbf{mam=ziŋ}+\textbf{ɢavr-ʁar-d}. \\ 
         \textsc{2sg} house=\textsc{emph} wife=\textsc{emph}+\textsc{neg-compl-ind} \\
    \glt ‘You have neither a house, nor a wife.’ (Gruzdeva FM2000: 4:13) [ES\textsuperscript{T}]
\z 

\protectedex{\ea\label{ex:niv:147}
    \gll Tə+aɲ ŋəŋ \textbf{cʰo=rə}+\textbf{ŋəŋ-ɣsu-ta} \textbf{ŋa=rə}+\textbf{ŋəŋ-ɣsu-ta} ha-ɟ.\\ 
         this+year \textsc{1pl\_excl} fish=\textsc{emph}+look\_for-\textsc{hab.neg-coord.1pl} beast=\textsc{emph}+look\_for-\textsc{hab.neg-coord.1pl} do\_so\textsc{{}-ind} \\
    \glt ‘This year we didn’t fish [and] didn’t hunt.’ (Gruzdeva FM2019: 1:10) [WS]
\z }

\ea\label{ex:niv:148}
    \gll Tunx sək malɣo-ɟ. \textbf{Als=hakisk} \textbf{puk=hakisk} melʁa-ɟ. \\ 
         here all be\_a\_lot-\textsc{ind} berry=\textsc{emph} mushroom=\textsc{emph} be\_full-\textsc{ind} \\
    \glt ‘Here there’s a lot of everything. [The place] is full of berries and mushrooms.’ (Gruzdeva FM2019: 1:9) [WS]
\z 

\ea\label{ex:niv:149}
    \gll \textbf{Puk=hakisk} \textbf{als=hakisk} tunx \textbf{pan-də-xsu-ɟ}. \\ 
         mushroom=\textsc{emph} berry=\textsc{emph} here grow-\textsc{emph-hab\_neg-ind} \\
    \glt ‘Neither mushrooms nor berries grow here.’ (Gruzdeva FM2019: 1:9) [WS]
\z 

\ea\label{ex:niv:150}
    \gll If \textbf{təf=hakisk}+\textbf{ʁavr-ra} \textbf{umgu=hakisk}+\textbf{ʁavr-ra} ha-ɟ.\\
         \textsc{3sg} house=\textsc{emph}+\textsc{neg-coord.3sg} woman=\textsc{emph+neg-coord.3sg} do\_so{}-\textsc{ind}\\
    \glt ‘He has neither a house, nor a wife.’ (Gruzdeva FM2019: 1:8) [WS]
\z 

\subsection{Negation, coordination, and complex clauses}\label{sec:niv:4.5}

Nivkh has several types of complex sentences, in which clauses can be combined either by coordination or by subordination. There are no special negative coordinators or conjunctions. All clauses can be negated separately or together with the standard strategy. For negation in subordinate clauses, see  \sectref{sec:niv:2.4}.

Coordinated clauses are formed by chained finite forms with the suffix ES \textit{-ra}~: \textit{-ta }: \textit{-na}, WS/A \textit{-ra }: \textit{-ta}, optionally followed by the auxiliary verb \textit{ha-} ‘be/do so’ in the appropriate morphological form. The following example \REF{ex:niv:151} is comprised of a chain of possessive clauses, of which the first one contains the negative verb \textit{ləɣə{}-} ‘not have’.

\ea\label{ex:niv:151}
    \gll Cij qʰola-f=tə+\textbf{ləɣə{}-ta} kʰəs \textbf{jiv-ta} sək-ujn pilagut \textbf{amr-ta} \textbf{ha-ve}. \\ 
         also be\_ill-\textsc{nmlz.loc=emph}+not\_have-\textsc{coord.2pl} happiness exist-\textsc{coord.2pl} all-\textsc{loc} a\_lot\_of.\textsc{2pl} have\_time-\textsc{coord.2pl} do\_so-\textsc{imp.2pl} \\
    \glt ‘Also, don’t be ill, be happy, have a lot of success in everything.’ \citep{Jushkina2019} [A]
\z 


\subsection{Miscellaneous aspects of negation}\label{sec:niv:4.6} 
This section will focus on diachronic observations.
There are several phenomena in the realm of negation that are of interest from a diachronic perspective. This section presents a more detailed discussion of the negative verbs (\sectref{sec:niv:4.6.1}) and emphatic clitics (\sectref{sec:niv:4.6.2}).

\subsubsection{Negative verbs}\label{sec:niv:4.6.1} 

As has been noted, in the ES variety there are two negative verbs, of which \textit{qavr-} (with initial plosive) is intransitive and \textit{ʁavr-} (with initial fricative in its free form) is transitive. Without going into further diachronic explanations, let us note that this opposition of initial plosives vs. fricatives is typical of Nivkh verbs in all varieties. The verb \textit{qavr-} is used exclusively in existential predications, whereas \textit{ʁavr-} in its free form is employed in possessive and equative predications and in its bound form in most other types of predications. In the A varieties, the corresponding verb \textit{qʰaw-} is intransitive, it is used independently with existential predications and forms a part of analytic negative construction with other predications. The origin of this analytic construction should probably be searched for in the contacts with the neighboring Tungusic languages, which have similar negative constructions and have influenced mainly the A varieties of Nivkh. Interestingly, in the A varieties there are no traces of the negative transitive verb, which would have an initial fricative, in contrast to intransitive \textit{qʰaw-}, which begins with a plosive. It seems that synchronically \textit{qʰaw-} can be transitivized by the causative suffix \textit{-gu/-ku}, cf. \REF{ex:niv:158} and the following example from \citet[148]{SaveljevaTaksami1970}: \textit{Taj qʰau-gu-ɟ} <spot \textsc{neg-caus-ind}> ‘[She] eliminated the spot’. All this suggests that the rise of the transitive verb in the ES varieties may have taken place after the split of these two variants.

As for the phonological shape of the negative verb, it may be assumed that its protoform was *\textit{qawr-}, which subsequently developed into ES \textit{qavr-} and A \textit{qʰaw-}, respectively. In the ES variety, in this particular word, a non-syllabic labial glide \textit{w} underwent a development to labiodental \textit{v}, which is a different phoneme. As a result, the verb has acquired its current ES form: *\textit{qawr-} > \textit{qavr-}. As for the A variety, the negative verb has retained the glide \textit{w} in its original shape, but has lost the trill \textit{r}, which, again, is in line with the general process of loss of fricatives and trills in this variety: *\textit{qawr-} > \textit{qaw-}. According to this diachronic scenario, the initial consonant of the negative verb in the Amur variety should not be aspirated. However, since this postvelar stop is almost always spirantized, it is perceived and pronounced by the speakers as aspirated: \textit{qʰaw-}.

Apart from the historical changes affecting its phonological shape, *\textit{qawr-} has acquired verbal morphological features and developed an opposition of the intransitive (\textit{qavr-}) and transitive (\textit{ʁavr-}) forms that are now present in the ES and WS varieties. The formation of negative transitive forms led to the creation of complex synthetic negative forms and gradual grammaticalization of the negative verb into a suffixal marker. This model was later extended to the imperative (hortative, jussive) sentences. The process of compounding has spread also to inherently negative verbs, many of which followed the same grammaticalization path. The negative habitual suffix, which comes from to the full lexical verb ‘not know’, can be seen as an example of a final stage of such a process.

Due to the longstanding contacts with the Tungusic languages, the situation in the Amur variety has been developing differently. Here, *\textit{qawr-} has functioned as an intransitive verb \textit{qʰaw-}, which, besides its independent function in stative predications, is used to head an analytic negative construction with a nominalized lexical verb. In this regard, it may be noted that the structure of this construction goes against the common tendency of placing negative markers before the elements they negate, cf. \citet{Dryer2013-wals143}. This construction has undergone further changes in its phonological structure, which could potentially lead to the development of new synthetic forms – if the language had a chance to develop further.

\subsubsection{Emphatic clitics}\label{sec:niv:4.6.2}

As has been shown in \sectref{sec:niv:4.4}, there are several emphatic clitics that occur in the negative context, whose properties differ in different Nivkh varieties. Also, each of the clitics has been developing historically in an idiosyncratic way.

The clitic \textit{=Ciŋ} is attested primarily in the ES variety and can be used both with affirmative and negative predicates for general emphasis of a certain constituent and for deriving emphatic proforms, which are also used in negative predications. Lately, its use in the affirmative context has become extremely rare, which means that this emphatic clitic is following the so-called \textit{Jespersen Cycle}, see e.g. \citet{Auwera2009}, \citet{Jespersen2025}) and is on the way to being reanalyzed as a true negative marker. In this connection, let us recall the apprehensive forms, which are also built with the emphatic/evidential suffix.

The clitic WS \textit{=ha-kisk/-gisk/-kirk}, A \textit{=ha-gin/-girn} is derived from the nominalized form of the auxiliary verb \textit{ha-} ‘be / do so’ with the instrumental suffix. In negative clauses, this clitic co-occurs with the clausal negator in the same way as ES \textit{=Ciŋ}. However, in the affirmative clauses, its function is different, as it is employed for deriving free-choice indefinites and providing the free-choice reading of other constituents to which it is attached. As a result, the clitic retains its affirmative status and is not embraced by the \textit{Jespersen Cycle}.

The emphatic clitic \textit{=Tə} can be traced back to the general focus clitic \textit{=Ta}, making use of a process known as “\textit{a-}raising” for emphasis. Also, the clitic can be attached to nominalized verbal complements, which has determined its fate as a suffixalized and desemanticized element within synthetic negative forms.

\section{Summary}\label{sec:niv:5} 
As may be concluded from the previous discussion \sectref{sec:niv:4}, the Nivkh system of negation was originally formed on the basis of a fundamental opposition between declarative and imperative clauses. This opposition was apparently relevant even before the language diversified into the different dialectal varieties. Formally, the opposition was built upon two main elements: (i) the clause-final element *\textit{qawr-}, which marked negation in the declarative sentences, and (ii) the clause-initial element \textit{tʰa}, which was used to mark negation in prohibitive sentences, and which remained intact through the history of the language. The existence of special negative marking in prohibitives conforms to a general typological tendency mentioned in \citet{AuweraEtAl2013}. Except for this distinction, negation of all types of clauses in Nivkh follows common rules and uses basically the same devices. 

As we have seen, in the realm of negation, Nivkh also displays many other interesting historical processes, which involve the transition or expansion of clitics and forms from the positive sphere to the negative. Apprehensives have developed an inherently negative meaning on the basis of affirmative emphatic/evidential forms, and the same process is observed for emphatic clitics.

Among other features, one should mention the relatively large amount of negative lexicalizations and an elaborate system of proforms that co-occur with the negative predicates.
\tabref{tab:niv:3} sums up the main negation strategies in the East Sakhalin and Amur varieties of Nivkh.

\begin{table}
\small
    \begin{tabularx}{\textwidth}{lXl}
\lsptoprule
Negation strategy & Function(s) & Section(s) \\
\midrule

(-\textit{ŋ})-\textit{ɢavr}- (ES)
&standard negation
& \sectref{sec:niv:2.1.1} \\

&prohibitive (with imperative)
& \sectref{sec:niv:2.2.1} \\

&negative ascriptive  
& \sectref{sec:niv:2.3.5} \\
	
\textit{qavr}- (ES) 
& negative existential 
& \sectref{sec:niv:2.3.1}  \\

\textit{j-iv}(-\textit{ŋ})-\textit{ɢavr}- (ES) 
& negative possessive 
& \sectref{sec:niv:2.3.2}  \\

\textit{ʁavr}- (ES) 
& negative possessive
& \sectref{sec:niv:2.3.2}  \\

& negative locative 
& \sectref{sec:niv:2.3.3}  \\

\textit{ʁavr-k} (ES) 
& negative equative 
& \sectref{sec:niv:2.3.4}  \\

\textit{qʰaw}- (A) 
& negative existential
& \sectref{sec:niv:2.3.1}  \\

& negative possessive 
& \sectref{sec:niv:2.3.2}  \\

& negative locative 
& \sectref{sec:niv:2.3.3}  \\

\textit{qʰaw-k}(=\textit{ra}) (ES/A) 
& negative reply 
& \sectref{sec:niv:3.1}  \\

\textit{-Toχ qʰaw}- (A) 
& standard negation
& \sectref{sec:niv:2.1.2} \\

& negative ascriptive 
& \sectref{sec:niv:2.3.5}  \\

\textit{ha-To}(\textit{χ}) \textit{qʰaw-ɟ=ra} (A) 
& negative equative 
& \sectref{sec:niv:2.3.4}  \\

\textit{ha-To}(\textit{χ}) \textit{qʰaw-k} (A) 
& negative equative 
& \sectref{sec:niv:2.3.4}  \\

\textit{-Ksu-} (ES/A) 
& negative habitual 
& \sectref{sec:niv:2.1.3}  \\

\textit{tʰa}  -\textsc{imp} (ES/A) 
& prohibitive 
& \sectref{sec:niv:2.2.1}  \\

\textit{-inəŋra : -inəŋta} (ES) / -\textit{nəra} (A) 
& apprehensive (visual) 
& \sectref{sec:niv:2.2.2}  \\

\textit{-jaŋra : -jaŋra} (ES) / -\textit{ijra} (A) 
& apprehensive (inferential) 
& \sectref{sec:niv:2.2.2}  \\

\textit{ujɣe} / \textit{ujɣi}- (ES) 
& negative existential 
& \sectref{sec:niv:2.3.1}  \\

\textit{avli}- (ES) 
& negative possessive 
& \sectref{sec:niv:2.3.2}  \\

\textit{ləɣə}- (A) 
& negative possessive 
& \sectref{sec:niv:2.3.2}  \\

\textit{ləɣə} (A) 
& caritive 
& \sectref{sec:niv:3.3}  \\

\textit{-lakrr̥ : -lakrt} (ES) /\textit{ -ləkrr : -ləkrt} (A) 
& negative purposive converb 
& \sectref{sec:niv:2.4}  \\

interrogative=\textsc{emph} (ES) 
& negative proforms 
& \sectref{sec:niv:3.2}  \\

interrogative=\textit{ha}-\textsc{instr} (A) 
& negative proforms 
& \sectref{sec:niv:3.2}  \\

interrogative XP=\textsc{emph} (ES) 
& forms reinforcing negation 
& \sectref{sec:niv:4.4}  \\

interrogative XP=\textit{ha}-\textsc{instr} (A) 
& forms reinforcing negation 
& \sectref{sec:niv:4.4}  \\

XP=\textit{ha}-\textsc{instr} (A) 
& forms reinforcing negation 
& \sectref{sec:niv:4.4}  \\

XP=\textit{Ci}(\textit{ŋ}) (ES) 
& forms reinforcing negation 
& \sectref{sec:niv:4.4}  \\

XP=\textit{Tə} (A) 
& forms reinforcing negation 
& \sectref{sec:niv:4.4} \\
\lspbottomrule
    \end{tabularx}
    \caption{Summary of negation strategies in Nivkh}
    \label{tab:niv:3}
\end{table}

% \hc{@authors: thank you for the new table and the section links! And I have rechecked, so there should be no problems with the hyphens, let me know if something went wrong after all.}


\section*{Acknowledgements}
This article is a part of Ekaterina Gruzdeva’s project “A Grammar of Nivkh: An isolate Paleosiberian language” (Helsinki Collegium for Advanced Studies, 2019–2021) and Maksim Fedotov’s project “Grammatical periphery in the languages of the world: A typological study of caritives” (Russian Science Foundation, Grant No. 18--78--10058). We are grateful to two anonymous reviewers for their useful comments and criticism. Responsibility for the contents of this paper and for any errors or drawbacks remains our own.

\section*{Abbreviations}




\begin{tabularx}{.45\textwidth}{lQ}
1 & first person\\
2 & second person\\
3 & third person\\
A & Amur (Lower/Liman) varieties\\
\scshape abl & ablative\\
\scshape acc & accusative\\
\scshape advs & adversative\\
\scshape agt & agentive \\
\scshape appr & apprehensive\\
\scshape asrt & assertive\\
\scshape aug & augmentative (verbal)\\
\scshape caus & causative\\
\scshape clf &classifier\\
\scshape compl & completive\\
\scshape coord  & coordinating verb form\\
\scshape conc  & concessive  \\
\scshape cond  & conditional  \\
CS & Central Sakhalin variety\\
\scshape cvb  &  converb \\
\scshape dat  & dative \\
\scshape des  & desiderative \\
\scshape dest  & destinative \\
\scshape dir & direct \\
\scshape dist  & distal  \\
\scshape du  & dual\\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape dur  & durative  \\
\scshape emph & emphatic\\
\scshape ep & epenthesis\\
ES & East Sakhalin variety\\
ES\textsuperscript{N} & Nyivo variant of the East Sakhalin variety\\
ES\textsuperscript{T} & Tymovsk variant of the East Sakhalin variety\\
\scshape evid & evidentiality\\
\scshape excl & exclusive\\
\scshape foc & focus \\
FM & field materials\\
\scshape fut & future\\
\scshape gnr & generic \\
\scshape hab & habitual\\
\scshape hum & human\\
\scshape imm  & immediate  \\
\scshape imp & imperative\\
\scshape incl & inclusive\\
\scshape ind & indicative\\
\scshape infr & inferential \\
\scshape ins & instrumental\\
\scshape iter & iterative\\
\scshape loc & locative\\
\scshape neg & negation\\
\scshape narr  & narrative  \\
\scshape nmlz & nominalization\\
\end{tabularx}

\begin{tabularx}{.45\textwidth}{lQ}
NP & noun phrase\\
NS & North Sakhalin variety\\
\scshape pl & plural\\
\scshape plac  & pluractional reduplication\\
\scshape prob & probabilitive\\
\scshape prog & progressive\\
\scshape purp  & purposive  \\
\scshape q & interrogative\\
\scshape qual & qualitative\\
\scshape recp & reciprocal\\
\scshape refl & reflexive\\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape rep  & reportative  \\ 
\scshape sbjv  & subjunctive  \\
\scshape sg & singular\\
\scshape sim  & simultaneous/anterior \\
SS & South Sakhalin variety\\
\scshape tr & transitivizer\\
V & (lexical) verb\\
\scshape vis & visual \\
\scshape voc & vocative\\
WS & West Sakhalin variety\\
XP & any syntactic phrase\\
\\
\end{tabularx}
\nocite{Kreinovich1937,KreinovichMs}%HC these are fieldnotes/examples sources. What shall we do with them? If I link all of them separately, it will give many separate references (one per year). Now there are only 4 (the ones listed in this line). e.g. GruzdevaFM has 6 different years.


\sloppy\printbibliography[heading=subbibliography,notkeyword=this]

\end{document}
