\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18234079}
\author{Jean-Pierre Koenig\affiliation{University at Buffalo} and         Karin Michelson\affiliation{University at Buffalo}}
\title{Negation in Oneida}
\abstract{This chapter focuses on the expression of negation in Oneida, which involves both an independent word (usually the particle \textit{yah}) and a negative prefix on verbs or the word \textit{té·kʌ}.  Overall, negation in Oneida is uniform across phrases and clauses, and the relation between scope and negation is also uniform (the negative particle \textit{yah} must occur to the left of any operator or predicate it scopes over).  Nevertheless, the expression of negation in Oneida is remarkable for the complexity of its morphological expression, and the presence of a series of ontologically general words that are both used as negative polarity items and serve as the basis of the ``derivation'' of corresponding indefinite expressions.
\keywords{Oneida, Iroquoian, negation, morphology}
}

%move the following commands to the "local..." files of the master project when integrating this chapter
% \usepackage{tabularx}
% \usepackage{langsci-optional}
% \usepackage{langsci-gb4e}
% %\bibliography{JPK}

\IfFileExists{../localcommands.tex}{
%   \addbibresource{JPK.bib}
   \input{../localpackages}
   \input{../localcommands}
   \input{../localhyphenation}
   \togglepaper[23]%%chapternumber
}{}

%\setlength{\glossglue}{2pt plus 2pt minus 1pt} %HC wished by the authors, bigger space between the words in the example, which makes it more readable. I am not convinced and think sticking to conventions would be easier and better readable in the first place. (e.g. 1 instead of 1st.person, at.that.time instead of 'at that time', ...). For example (42) \REF{{oneida:notso 8}, it would actually help the readability 

\begin{document}
\maketitle


\section{The language}
Oneida (ISO: \href{https://iso639-3.sil.org/code/one}{one}, Glottocode: \href{https://glottolog.org/resource/languoid/id/onei1249}{onei1249}) is one of five Northern Iroquoian languages still spoken by those who learned their language as a first language, before English. The others are Mohawk, the most closely related to Oneida, Onon\-daga, Cayuga, and Seneca. The Iroquoian languages are polysynthetic, known for their pervasive noun incorporation and enormously complex prefixal inflection with sometimes funky phonological processes making it difficult to unambiguously decide on discrete morpheme boundaries.  Today, there are three Oneida territories in the United States and Canada. The Oneida Nation of New York is located on territory east of Syracuse, New York. During the 1800s groups of Oneidas moved away from New York and established settlements in southwestern Ontario and in the vicinity of Green Bay, Wisconsin. Oneida has been considered a severely endangered language. Presently Oneida is spoken by speakers who learned Oneida as a first language only in Ontario at the Oneida Nation of the Thames, where according to latest estimates there are fewer than 40 fluent first-language speakers. Fewer than 30 elders were reported to have learned Oneida before English in Wisconsin in the mid-1990s, and the last speaker, Maria Hinton, died at age 103 in 2013. There have been no first-language speakers from the Oneida Nation in NY since at least the mid-1970s or 80s. Encouraging though, is that all the Northern Iroquoian communities (including Tuscarora and Wendat in addition to the five mentioned above) have programs that are committed to the inter-generational transmission of the languages, and some of these efforts have been remarkably successful. More and more, second-language learners provide an expertise when first-language speakers cannot.

The presentation of Oneida negation patterns in this paper follows the organization in the questionnaire provided by the editors of this volume \parencitetv{chapters/questionnaire}. The data is almost exclusively from the Oneida Nation of the Thames. There are some differences between the variety of Oneida spoken in Ontario and in Wisconsin. For example, the word for `no' in Wisconsin is \textit{yáhtʌʔ}, which is related to the negative particle \textit{yah}. In Ontario, the word for `no' is \textit{táh}.
Most of the examples in this paper were gleaned from the text collections in \citet{MichelsonEtAl2016}. Extracts are reproduced exactly as in the original, including any punctuation (or lack of it) when the extract is only part of an utterance. The speaker's name is identified after the extract, followed by the page number where the extract can be found, and the utterance number. The excerpts from texts are supplemented by data from \citet{MichelsonDoxtator2002} and from collaborative work with the late Mercy Doxtator, the late Norma Kennedy, and Olive Elm. Occasionally approximations of English sentences from the questionnaire were requested from Olive Elm, and all unattributed examples were confirmed with Olive Elm.

The expression of negation in Oneida, aside from imperative and prohibitive contexts, is for the most part uniform. Only clausal negation is present in the language and the same formal expression is used in all major syntactic structures (including questions, stative predications, and indefinites) and in both main and embedded clauses. Still, slightly different constraints are operative in a number of contexts of use, and there are certain structures, for example what we call \textit{negation spreading} and different structures for negating category membership (or identification) versus negating a participant in a situation, where there is more to be said in a description that calls for completeness. Also, a few aspects of negation in Oneida stand out typologically; especially notable is that the distribution of negative formatives in Oneida offers a particularly clear example of the distinction between  \textit{morphosyntactic features}, in this case the feature \textit{negative}, and the \textit{exponents} of that feature.

\section{Clausal negation}

\subsection{Standard negation}
\label{oneida:sec:standard}

Standard negation is expressed in Oneida by a particle, which is \textit{yah} in most contexts except those noted below and in \sectref{oneida:sec:non-declarative}, plus a negative prefix on the verb. In the example of standard negation in (\ref{oneida:standard}) the negative prefix is \mbox{\textit{teʔ-}.}\footnote{\label{oneida:fn1}In the Oneida orthography the vowel \textit{u} is a high (for some speakers closer to mid) back mildly rounded nasalized vowel and \textit{ʌ} is a low-mid central nasalized vowel. A raised period indicates vowel length. Underlining indicates devoicing, one of a set of phonological processes that occur at the ends of utterances. In the interlinear glosses a bare numeral 3 abbreviates a number of third person categories: feminine singular, indefinite or non-specific, masculine dual and plural, and feminine-zoic dual and plural (see \citealt{Michelson2015} for the gender categories of Oneida). When we give excerpts from recorded texts, we generally do not cite a sentence or utterance in its entirety; any punctuation, or lack of it, is as in the original transcriptions of the recording. Also note that Oneida has over 150 uninflected particles, and speakers find some of them difficult to translate into English; these are simply glossed `\textsc{ptcl}' in the interlinear glosses.} Whereas in other Northern Iroquoian languages the particle that occurs in standard negation is the same as the word for `no,' this is not the case in Oneida as spoken at the Oneida Nation of the Thames (the variety represented in this paper); the negative particle is \textit{yah} while `no' is \textit{táh}.

\ea\label{oneida:standard}
    \ea\label{oneida:standard a}  
        \glll wakanúhteʔ \\
        wak-anuhte-ʔ \\
        \textsc{1sg.p}-know-\textsc{stat}\\
        \glt `I know'	\citep[107]{MichelsonDoxtator2002}
    \ex\label{oneida:standard b}  
        \glll yah teʔwakanúhteʔ \\
        yah teʔ-wak-anuhte-ʔ \\
        not \textsc{neg-1sg.p}-know-\textsc{stat}\\
        \glt `I don't know'	\citep[107]{MichelsonDoxtator2002}
    \z
\z 

This pattern -- particle plus verbal prefix -- is usually classified as \textit{double negation} (\citealt{Dryer2013-wals143}; \citealt{AlsenoyAuwera2014}) and it occurs in all clauses except (non-declarative) imperatives and prohibitives. 
In \citegen{Miestamo2003} typology, Oneida exhibits symmetrical negation: a negative clause involves only the addition of the particle plus prefix to the corresponding positive. However negation in Oneida is also asymmetrical because it is restricted to certain aspect--mode categories. The negative structure occurs in two of the three basic aspects, the habitual (imperfective) and stative. The third basic aspect, the punctual (or perfective) always occurs with one of three modal prefixes: future, optative, or factual. The optative mode of the punctual aspect can be negated, but the factual and future cannot. Verbs that otherwise occur in the factual mode are negated with the stative aspect (\ref{oneida:stative}), and verbs that otherwise occur in the future mode are negated with the optative mode (\ref{oneida:optative}). The example in (\ref{oneida:optative}) shows also that material can intervene between the negative particle and the negative verb form.

\ea\label{oneida:stative}
    \ea\label{oneida:stative a}
        \glll waʔukwanú·lyahkeʔ \\
        waʔ-yukwa-nuhlyaʔk-eʔ \\
        \textsc{fact-1pl.p}-get.hurt-\textsc{pnc} \\
        \glt `we got hurt'
    \ex\label{oneida:stative b}
        \glll yah teʔyukwanuhlyá·ku \\
        yah teʔ-yukwa-nuhlyaʔk-u \\
        not \textsc{neg-1sg.p}-get.hurt-\textsc{stat} \\
        \glt `we weren't hurt'	\citep[538]{MichelsonDoxtator2002}
    \z
\z

\ea\label{oneida:optative}
    \ea\label{oneida:optative a}
        \glll ʌtehsanúhtuʔ \\
        ʌ-te-hs-anuhtu-ʔ \\
        \textsc{fut-cis-2sg.a}-determine/control-\textsc{pnc} \\
        \glt `you will determine it' \citep[108]{MichelsonDoxtator2002}
    \ex\label{oneida:optative b}
        \glll nʌ kiʔ yah isé· thaesanúhtuʔ \\
        nʌ kiʔ yah isé· th-ae-s-anuhtu-ʔ \\
        now actually not you \textsc{cntr-opt-2sg.p}-determine/control-\textsc{pnc} \\
        \glt `you are not going to have your way.'	Verland Cornelius \citep[338 (256)]{MichelsonEtAl2016}
    \z 
\z 

The form \textit{th-} in (\ref{oneida:optative b}) is usually identified as the contrastive prefix. It occurs instead of the negative form \textit{teʔ-} under two conditions: (i) when the verb is in the optative mode, as in the example in (\ref{oneida:optative b}), or (ii) when either (or both) the dualic (\ref{oneida:cntr-dlc b}) or translocative (\ref{oneida:cntr-trnl b}) verbal prefixes are also present.\footnote{An additional, unusual, exponent of the negative with the translocative is discussed in \sectref{oneida:sec:diachronic}.}

\ea\label{oneida:cntr-dlc}
    \ea\label{oneida:cntr-dlc a}
        \glll tewakatuhutsyoní \\
        te-wak-atuhutsyoni \\
        \textsc{dlc-1sg.p}-want/need[\textsc{stat}] \\
        \glt `I want or need it' \citep[254]{MichelsonDoxtator2002}
    \ex\label{oneida:cntr-dlc b}
        \glll yah thaʔtewakatuhutsyoní \\
        yah th-aʔte-wak-atuhutsyoni  \\
        not \textsc{cntr-dlc-1sg.p}-want/need[\textsc{stat}] \\
        \glt `I don't want or need it'	
    \z
\z 

\ea\label{oneida:cntr-trnl}
    \ea\label{oneida:cntr-trnl a}
        \glll tho yʌhʌ́skeʔ  \\
        tho yʌh-ʌ-s-k-e-ʔ  \\
        there \textsc{trl-fut-rpt-1sg.a}-walk/go-\textsc{pnc} \\
        \glt `I will go there again' \citep[295]{MichelsonDoxtator2002}
    \ex\label{oneida:cntr-trnl b}
        \glll yah tho thyusa·ké· \\
        yah tho th-y-u-s-a-k-e-ʔ \\
        not there \textsc{cntr-trl-opt-rpt-opt-1sg.a}-go/walk-\textsc{pnc} \\
        \glt `I won't go there again'	%(Olive Elm, p.c., patterned after \citealt[145 (22)]{MichelsonEtAl2016})
    \z
\z

It is important, for the proper classification of Oneida negation, to recognize that it is not the case that in examples such as (\ref{oneida:cntr-dlc b}) and (\ref{oneida:cntr-trnl b}) the particle \textit{yah} is simply added to a positive verb form with the contrastive prefix. Outside of the negative, the contrastive is relatively unproductive, and negated contrastive forms are not accepted. \citet[12]{Abbott2000} is explicit about this point: ``the negative and contrastive are mutually exclusive semantically''. The contrastive is  the realization, or exponent, of the negative morphosyntactic \textit{feature} under the conditions stated above; it is \textit{not} just an exponent of the contrastive feature. An example of a verb form with the contrastive feature can be seen in \REF{oneida:true-cont}, which was spoken in the context of people nowadays being on welfare instead of working for their money.

\ea\label{oneida:true-cont}
    \glll nʌ seʔ kwáh ok thiyesahwista·wíheʔ \\
    nʌ seʔ kwáh ok thi-yesa-hwist-awi-heʔ \\
     then too just only \textsc{cntr-3>2sg}-money-give-\textsc{hab} \\
    \glt  `you just get money given to you.' Verland Cornelius \citep[322(100)]{MichelsonEtAl2016}
\z
 
 
In sum, clausal negation in Oneida involves a negative particle and an exponent of the negative feature, either \textit{teʔ-} or \textit{th-}. Both \textit{teʔ-} and \textit{th-} occur word-initially in the verb, i.e. in the same ``slot" in the morphological template of the Oneida verb. Two other prefixes that occur in the word-initial ``slot" are the \textit{partitive} and \textit{coincident} prefixes (using Iroquoianist terms) and thus are in complementary distribution with the negative and contrastive exponents; discussion of these in negative structures is deferred to \sectref{oneida:sec:summary}.

To end this section, we mention two additional negative particles, \textit{áhsu} `not yet', exemplified in (\ref{oneida:ahsu}), and \textit{ahsuhkʌ́} (for an example see (\ref{oneida:ahsuhkv}) in \sectref{oneida:sec:summary}). In the excerpt in (\ref{oneida:ahsu}) the narrator is telling us that the house her father was building was not even finished when they moved there. Both the particles \textit{áhsu} and \textit{ahsuhkʌ́} are attested far less often than \textit{yah}.

\ea\label{oneida:ahsu} 
    {\glll áhsu kwahotokʌ́·u teʔyotéhsuʔ \'okhnaʔ \\
    áhsu kwahotokʌ́·u teʔ-yo-atehs-uʔ \'okhnaʔ  \\
    {not.yet} {just.for.real} \textsc{neg-3z/n.sg.p}-get.finished-\textsc{stat} {and.then} \\}
    \sn{\glll yaʔakwanákla\uline{teʔ}. \\
     y-aʔ-yakw-anaklat-eʔ \\
    \textsc{trl-fact-1excl.pl.a}-move/settle.in-\textsc{pnc}\\
    \glt `it [the house] wasn't finished yet and already we moved in.' Olive Elm \citep[ 106 (3)]{MichelsonEtAl2016} }
\z%hc just.for.real, and.then

\subsection{Non-declarative negation}
\label{oneida:sec:non-declarative}

In this section we describe polar questions, and imperatives and prohibitives. Negative polar questions use standard negation and so are discussed first.

Polar questions are formed with the second-position polar question particle \textit{kʌ} (\ref{oneida:PQ}). In negative polar questions \textit{kʌ} occurs after the negative particle, and otherwise the question is like any other standard negation; (\ref{oneida:Neg PQ1}) is an example of a negative polar question. The expected reply to a negative polar question is normally \textit{táh} `no' (see also \sectref{oneida:sec:replies}). Utterances with the structure of negative polar questions, however, are attested more frequently as indirect speech acts.

The excerpt in (\ref{oneida:Neg PQ2}) was uttered by the narrator's father after he scared her mother by showing a pig head through a window; this happened after some spooky events that led the narrator's mother to believe their house was haunted, and she was not happy about being pranked by a pig head.\footnote{\citet[394]{Woodbury2018} reports that in the related language Onondaga negative polar questions occur rarely in her corpus, and when they do, they usually are interpreted as polite requests.}

\ea\label{oneida:PQ}
    \glll shekú kʌ yetshi·kʌ́heʔ thikʌ́ atyanlúhslaʔ owahá\uline{·ke}.  \\
    shekú kʌ yetshi-kʌ-heʔ thikʌ́ atyanlúhslaʔ owahá·ke  \\
    still \textsc{q} \textsc{2du$>$3}-see-\textsc{hab} that ghost {on.the.road}\\
    \glt `do you still see that ghost on the road?' Verland Cornelius \citep[67 (21)]{MichelsonEtAl2016}
\z

\ea\label{oneida:Neg PQ1}
    \glll Yah kʌ teʔsatkáthu n aké·slet. \\
    yah kʌ teʔ-s-atkath-u n ake-ʔslet \\
    not \textsc{q} \textsc{neg-2sg.p}-see-\textsc{stat} \textsc{ptcl} \textsc{1sg.poss}-car \\
    \glt `You haven't seen my car?'  Norma Kennedy \citep[133 (27)]{MichelsonEtAl2016}
    %\hc{@M\&L: in this chapter there are many weird glosses, like "particle", see also "JOINER" (i.e. epenthetic I guess), "ASSERTION" and many more, which is in my opinion not clear enough. Could you comment on that for the author if you also think it should be modified somehow?}
\z

\ea\label{oneida:Neg PQ2}
    \glll yah kʌ teʔsatshanuní k\'oskos okúhsaʔ ʌ́hseke ʌy\'olhʌ\uline{neʔ}. \\
    yah kʌ teʔ-s-atshanuni k\'oskos okúhsaʔ ʌ-hse-k-eʔ ʌy\'olhʌneʔ \\
    not \textsc{q} \textsc{neg-2sg.p}-happy[\textsc{stat}] pig face \textsc{fut-2sg.a}-eat-\textsc{pnc} tomorrow \\
    \glt `you're not happy? you can eat the pig face tomorrow?' Verland Cornelius \citep[114 (43)]{MichelsonEtAl2016}
\z 
    %\lv{To Héloïse: I agree with most of your suggestions. But I don't find particle, joiner or assertion puzzling.} 
    %\mm{but they should be abbreviated like in other chapters; in addition, there are for example cases where the dot or the underscore should be used on the gloss line instead of spaces, see e.g. "on the road" in ex8.}
    % \aut{We may have erred in some cases, but "on the road" is not a single morpheme, but simply something we did not analyze as it was irrelevant and not to overtax readers.}
    %\aut{abbreviations are not helpful to readers. we are fine with other authors not abbreviating.}
    %\hc{go through the paper and make this change everywhere (on the road to on.the.road). We abbreviate everything: ASRT for assertion, PTCL particle, JOIN joiner.}
    % \hc{I have now changed throughout ASRT for assertion, PTCL particle, JOIN joiner, also added to the list of abbreviations. I have also added dots in "on.the.road" etc throughout, since this is how LGR deals with non-one-to-one correspondences (rule 4, see also 4C)}
Polar questions can occur with modal verbs, for example \textit{-atu-} `possible, can be.' This verb, when inflected with the feminine-zoic/neuter singular prefix, expresses possibility or permission (\ref{oneida:can}) and necessity (\ref{oneida:must}). The examples in (\ref{oneida:must}) differ from those in (\ref{oneida:can}) only by the presence of the particle \textit{nok} (otherwise `only'); in other words the particle \textit{nok} has the effect of altering an expression of possibility to one of necessity. 

As expected, (\ref{oneida:can-c}), (\ref{oneida:can-d}) and (\ref{oneida:must-c}), (\ref{oneida:must-d}) adhere to the structure of standard negation. Also, as expected, while the future prefix occurs with the verbs in the positive sentences, the optative prefix occurs in the negative ones.

\ea\label{oneida:can}
    \ea\label{oneida:can-a}
        \glll\Large ʌ\normalsize wa·tú· ʌkatnutolyaʔtá\uline{·naʔ}. \\
        \Large ʌ\normalsize -w-atu-ʔ ʌ-k-atnutolyaʔtahn-aʔ \\
        \textsc{fut-3z/n.sg.a}-can-\textsc{pnc} \textsc{fut-1sg.a}-go.and.play-\textsc{pnc} \\
        \glt `I can go and play.'
    \ex\label{oneida:can-b}
        \glll\Large ʌ\normalsize wa·tú· kʌ ʌkatnutolyaʔtá\uline{·naʔ}. \\
        \Large ʌ\normalsize -w-atu-ʔ kʌ ʌ-k-atnutolyaʔtahn-aʔ \\
        \textsc{fut-3z/n.sg.a}-can-\textsc{pnc} \textsc{q} \textsc{fut-1sg.a}-go.and.play-\textsc{pnc} \\
        \glt `Can I go and play?'
    \ex\label{oneida:can-c}
        \glll Yah thau·tú· a·katnutolyaʔtá\uline{·naʔ}. \\
    yah th-aa-w-atu-ʔ aa-k-atnutolyaʔtahn-aʔ \\
    not \textsc{cntr-opt-3z/n.sg.a}-can-\textsc{pnc} \textsc{opt-1sg.a}-go.and.play-\textsc{pnc}\\
    \glt `I can't go and play.'
    \ex\label{oneida:can-d}
        \glll Yah kʌ thau·tú· a·katnutolyaʔtá\uline{·naʔ}. \\
        yah kʌ th-aa-w-atu-ʔ aa-k-atnutolyaʔtahn-aʔ \\
        not \textsc{q} \textsc{cntr-opt-3z/n.sg.a}-can-\textsc{pnc} \textsc{opt-1sg.a}-go.and.play-\textsc{pnc} \\
        \glt `I can't go and play?'
    \z
\z


\ea
\label{oneida:must}
    \ea\label{oneida:must-a}
        \glll Nok ʌwa·tú· ʌkatnutolyaʔtá\uline{·naʔ}.  \\
        nok ʌ-w-atu-ʔ  ʌ-k-atnutolyaʔtahn-aʔ  \\
        only \textsc{fut-3z/n.sg.a}-can-\textsc{pnc} \textsc{fut-1sg.a}-go.and.play-\textsc{pnc} \\
        \glt `I have to go and play.' 
    \ex\label{oneida:must-b}
        \glll Nok kʌ ʌwa·tú· ʌkatnutolyaʔtá\uline{·naʔ}.  \\
        nok kʌ ʌ-w-atu-ʔ  ʌ-k-atnutolyaʔtahn-aʔ  \\
        only \textsc{q} \textsc{fut-3z/n.sg.a}-can-\textsc{pnc} \textsc{fut-1sg.a}-go.and.play-\textsc{pnc} \\
        \glt `Do I have to go and play?'
    \ex\label{oneida:must-c}
        \glll Yah nok thau·tú· a·katnutolyaʔtá\uline{·naʔ}. \\
        yah nok th-aa-w-atu-ʔ  aa-k-atnutolyaʔtahn-aʔ \\
        not only \textsc{cntr-opt-3z/n.sg.a}-can-\textsc{pnc} \textsc{opt-1sg.a}-go.and.play-\textsc{pnc}  \\
        \glt `I don't have to go and play.'
    \ex\label{oneida:must-d}
        \glll Yah kʌ nok thau·tú· a·katnutolyaʔtá\uline{·naʔ}.   \\
        yah kʌ nok th-aa-w-atu-ʔ aa-k-atnutolyaʔtahn-aʔ   \\
        not \textsc{q} only \textsc{cntr-opt-3z/n.sg.a}-can-\textsc{pnc} \textsc{opt-1sg.a}-go.and.play-\textsc{pnc} \\
        \glt `I don't have to go and play?'
    \z
\z


Negative imperatives and prohibitives deviate from the standard negation structure in that the particle \textit{tákʌʔ} occurs instead of the negative particle \textit{yah} and the verb does not have a negative prefix. Negative imperatives require the future modal prefix on the verb, while prohibitives require the optative. An imperative form is given in (\ref{oneida:imp}), a negative imperative in (\ref{oneida:takv-fut}) and a prohibitive in (\ref{oneida:takv-opt-prohibit}). The ending of the verb form that usually occurs in the imperative (both negative and non-negative) and prohibitive is best described as based on the punctual aspect verb form, but with some substractive morphology in that the imperative lacks a final glottal stop or a final vowel \textit{e} (historically an epenthetic vowel) plus glottal stop, and if the punctual form ends in a long vowel, the imperative form ends in a short vowel.\footnote{In the punctual aspect, the verb in (\ref{oneida:imp}) would end in a long vowel, (\ref{oneida:takv-fut}) and (\ref{oneida:takv-opt-prohibit}) in \textit{ʔ}, and (\ref{oneida:takv-opt-lest b}) in \textit{-neʔ}.} Sometimes though (unpredictably) the punctual ending occurs.

\ea\label{oneida:imp}
    \glll Wá·s yaʔsenhotuk\'o ohná·kʌʔ nukwá·, \\
    wá·s yaʔ-se-nhotuko	ohná·kʌʔ nukwá· \\
    go \textsc{trl-2sg.a}-open.door[\textsc{imp}] back where \\
    \glt `Go open the back door,' Olive Elm \citep[60 (4)]{MichelsonEtAl2016} 
\z

\ea\label{oneida:takv-fut}
    \glll tákʌʔ náhteʔ ʌhsí\uline{·lu}.\\
    tákʌʔ náhteʔ ʌ-hs-ihlu\\
    don't anything \textsc{fut-2sg.a}-say[\textsc{imp}] \\
    \glt `you don't say anything.' Clifford Cornelius \citep[230 (85)]{MichelsonEtAl2016} 
\z

\ea\label{oneida:takv-opt-prohibit}
    {\glll Né· {kyuhte wí·} thikʌ́ waʔshakohlo·lí· tákʌʔ \\
    né· {kyuhte wí·} thikʌ́ waʔ-shako-hloli-ʔ tákʌʔ \\
    \textsc{asrt} supposedly that \textsc{fact-3m.sg$>$3}-tell-\textsc{pnc}  don't \\} 
    \sn{\glll  a·yakotéts\uline{hʌ}. \\
     aa-yako-atetshʌ \\
    \textsc{opt-3fi.sg.p}-become.afraid[\textsc{imp}] \\
    }
    \glt `He told her she shouldn't become afraid.' Olive Elm \citep[169 (47)]{MichelsonEtAl2016}
\z

As evident from examples (\ref{oneida:takv-opt-lest a}) and (\ref{oneida:takv-opt-lest b}), the prohibitive structure is used not just to state that something should be avoided but also to describe a reason for doing something when an alternative action would, supposedly, result in an undesirable outcome. These negative purpose clauses are perhaps the closest equivalent to contexts where the English conjunction `lest' is found.
 
\ea\label{oneida:takv-opt-lest a}
    \glll Senhotú kwí· tákʌʔ utahutáyaht tsíks. \\
    se-nhotu kwí· tákʌʔ u-t-a-hu-atayaht tsíks \\
    \textsc{2sg.a}-close.a.door[\textsc{imp}] \textsc{ptcl} don't \textsc{opt-cis-opt-3m.pl.a}-enter-[\textsc{imp}] flies \\
    \glt `Close the door so the flies don't come in!'  \citep[526]{MichelsonDoxtator2002}
\z

\ea\label{oneida:takv-opt-lest b}
    \glll tákʌʔ a·knahko·lʌ́n kyuhte wáh.\\
    tákʌʔ aa-k-nahkw-olʌʔ-n kyuhte wáh\\
    don't \textsc{opt-1sg.a}-mate-find-\textsc{imp} supposedly indeed \\
    \glt `(It's kind of like she watched over me,) so I would not get a man supposedly.' Pearl Cornelius \citep[310 (130)]{MichelsonEtAl2016}
\z

\subsection{Negation in stative predications}
\label{oneida:sec:statives}

In many languages, like English, the same negative structure is used for concepts such as attribution, possession, and locations. English neutralizes the semantic distinction between predicative phrases that describe, e.g. occupations (\textit{John is a dentist}) and membership in a category of objects (\textit{Pop is a beagle}); they are both predicative phrases. The syntax and the morphology of Oneida, however, distinguishes between stative situations and objects. The former are expressed by verbs, the latter by nouns (but see \citealt{KoenigMichelson2020}, \citeyear{KoenigMichelson2022} for what the terms \textit{verb} and \textit{noun} mean in this context), and different patterns of negation apply to these two kinds of stative predicates. In what follows we give examples that show that standard negation applies to attributes, locations, existential meaning, and possession. Then we describe a different structure for category membership and kin relations, as well as an alternative possession pattern (see \citealt{KoenigMichelson2021} for details on the morphosyntax of possession in Oneida).

Stative predications that involve verbs -- occupations and similar offices; attribution; proper inclusion; locative, existential and possessive predication -- are all negated with the clausal negation pattern described in \sectref{oneida:sec:standard}. Occupations are generally expressed with the habitual aspect of verbs (e.g. \textit{shakonawilahslu·níheʔ} `dentist,' literally, `he fixes their teeth') and attribution is generally expressed with stative verbs (indicated \textit{v.s.} in \citealt{MichelsonDoxtator2002} or \textit{state} in \citealt{AbbottEtAl1996}). These are negated as any other clause. An example of attribution and its negation is given in (\ref{oneida:attribution}). 

\ea\label{oneida:attribution}
    \ea\label{oneida:attribution a}	
        \glll yokʌháksteʔ \\
        yo-kʌh-a-kst-eʔ \\
        \textsc{3z/n.sg.p}-blanket-\textsc{join}-heavy-\textsc{stat} \\
        \glt `it's a heavy blanket'
    \ex\label{oneida:attribution b}
        \glll yah teʔyokʌháksteʔ \\
        yah teʔ-yo-kʌh-a-kst-eʔ \\
        not \textsc{neg-3z/n.sg.p}-blanket-\textsc{join}-heavy-\textsc{stat} \\
        \glt `it's not a heavy blanket' (Olive Elm, p.c.)
    \z
\z 

Locative and existential meanings, and possession, are all conveyed with verbs of position: \textit{-yʌ-/-yʌt-} `put or lie something down (on the ground),' \textit{-ot-/-hnyot-} `erect or stand something up,' \textit{-hl-\-hel-} `set or place on top of,'  and \textit{-t-} `be standing' occur most often, although there are other verbs whose meaning includes location of a particular category of entity (such as snow, in (\ref{oneida:loc b}) below). Turning first to locatives, the verb specifies the position or orientation of an entity, and the located entity is more often than not denoted by a noun stem that has been incorporated into the verb; in the examples in (\ref{oneida:loc c}) and (\ref{oneida:loc a}) the stem \textit{-nuhs-} `house' is incorporated into the verb \textit{-ot-} `stand.' The specific location (where something is located) is expressed by the particle \textit{tho} `there' plus the separate clause \textit{aʔé· naʔoháhati} `over on the other side of the road' in (\ref{oneida:loc c}), or a form that has a locative suffix, \textit{owahaʔkéshuʔ} `on the roads', in (\ref{oneida:loc b}). Negation of clauses that include a positional verb take the usual form, \textit{yah} plus negative prefix.

\ea\label{oneida:loc}
    \ea\label{oneida:loc c}
        \glll aʔé· naʔoháhati kʌ́·, tho \\
        aʔé· naʔ-yo-hah-ati  kʌ́· tho	\\
        great \textsc{par-3z/n.sg.p}-road-other.side[\textsc{stat}] {y'know} there  \\
        \sn{tkanúhsoteʔ, \\
        t-ka-nuhs-ot-eʔ \\
        \textsc{cis-3z/n.sg.a}-house-stand-\textsc{stat}}
        \glt `over on the other side of the road, there was a house,' Mercy Doxtator \citep[142--143 (5)]{MichelsonEtAl2016}
    \ex\label{oneida:loc a}
        \glll né· aolí·waʔ yah tho teʔskanúhso\uline{teʔ}. \\
                né· aolí·waʔ yah	 tho	 teʔ-s-ka-nuhs-ot-eʔ \\
                \textsc{asrt} {the.reason} not there \textsc{neg-rpt-3z/n.sg.a}-house-stand-\textsc{stat} \\
        \glt `(we had a fire,) that's why the house isn't there anymore.' Mercy Doxtator \citep[142 (4)]{MichelsonEtAl2016} 
    \ex\label{oneida:loc b}
        {\glll yah tho teʔkanye·yʌ́· \\
        yah tho teʔ-ka-nyeyʌ-ʔ  \\
        not there \textsc{neg-3z/n.sg.a}-snow.gets.on.the.ground-\textsc{stat}  \\}
        \sn{owahaʔkéshuʔ, \\
        o-wah-aʔke-shuʔ \\
        \textsc{3z/n.sg.p}-road/path-\textsc{loc-distr} }
        \glt `there wasn't snow on the roads,' Mercy Doxtator \citep[193 (13)]{MichelsonEtAl2016}
    \z
\z

Existential meaning is expressed with positional verbs, as pointed out above, and, for animate beings, with the habitual aspect of the verb \textit{-e-} `walk, go.' Examples are given in (\ref{oneida:exist}). Negative existentials often include indefinite expressions (\sectref{oneida:sec:npi}). The example in (\ref{oneida:exist c}) shows the effect of a regular phonological alternation of Oneida, whereby a glottal stop (in this case the \textit{ʔ} of the negative prefix \textit{teʔ-}) is deleted when it follows an accented vowel, and the accented vowel lengthens.

\ea\label{oneida:exist}
    \ea\label{oneida:exist a}
        \glll tahnú· kwáh kok nikú ska·yʌ́· thikʌ́ othé·tsliʔ, \\
        tahnú· kwáh kok nikú s-ka-yʌ-ʔ thikʌ́ othé·tsliʔ \\
        and just {a.little} amount \textsc{rpt-3z/n.sg.a}-put-\textsc{stat} that flour \\
        \glt `and there was just a small amount of flour left,' Clifford Cornelius \citep[224 (42)]{MichelsonEtAl2016}
    \ex\label{oneida:exist b} 
        \glll wahʌ́·luʔ thikʌ́ shakotsyapslatshʌlyá·se, wahʌ́·luʔ ``yah náhteʔ teʔka·yʌ́·'', \\
        wahʌ́·luʔ thikʌ́ shako-tsyap-sl-a-tshʌly-a-ʔse wahʌ́·luʔ yah náhteʔ teʔ-ka-yʌ-ʔ \\
        {he.said} that \textsc{3m.sg$>$3}-job-\textsc{nmlz-join}-look.for-\textsc{join-ben[stat]} {he.said} not anything \textsc{neg-3z/n.sg.a}-put-\textsc{stat}\\
        \glt `the person who finds jobs for people said, he said ``there's nothing available"' Clifford Cornelius \citep[231 (90)]{MichelsonEtAl2016}
    \ex\label{oneida:exist c} 
        \glll Wahotétshʌʔ lonúhteʔ tsiʔ yah úhkaʔ oyá· tho té·yʌhseʔ, \\
        wa-ho-atetshʌ-ʔ lo-anuhte-ʔ tsiʔ yah úhkaʔ oyá· tho teʔ-yʌ-e-ʔseʔ \\
        \textsc{fact-3m.sg.p}-get.scared-\textsc{pnc} \textsc{3m.sg.p}-know-\textsc{stat} \textsc{comp} not anyone else there \textsc{neg-3fi.a}-walk/go-\textsc{hab} \\
        \glt `He got scared knowing that there wasn't anyone else there,'	 Rose Antone \citep[100 (12)]{MichelsonEtAl2016}
    \z
\z

An example of possession is the following excerpt. It includes two positional verbs, first one where possession is asserted, and then one where possession is negated, as in \REF{oneida:poss}. 

\begin{exe}
\ex\label{oneida:poss}
    \glll kítkit ok oniʔ usayukwayʌ·tákeʔ yah oniʔ né· teʔtsyukwa·\uline{yʌ́·}. \\
    kítkit ok oniʔ u-s-a-yukwa-yʌt-ak-eʔ yah oniʔ né· teʔ-ts-yukwa-yʌ-ʔ \\
    chicken(s) only too \textsc{opt-rpt-opt-1pl.p}-put/have-\textsc{cont-pnc} not too \textsc{asrt} \textsc{neg-rpt-1pl.p}-put/have-\textsc{stat} \\
    \glt `we should have chickens too, we don't have them anymore either.' Hazel Cornelius \citep[181 (13)]{MichelsonEtAl2016}
\end{exe}

Stative predications that assert that an entity is a member of a category and involve nouns, in the simplest case, involve the so-called ``assertion'' particle \textit{né·} and either a demonstrative particle or the particle \textit{né·n}, as in the excerpt in (\ref{oneida:airplane}). Here the speaker is talking about her father telling his children about what will happen in the future, including things like airplanes, which didn't exist at the time. Membership in a category (\ref{oneida:ident 1}), as well as statements of non-identity (\ref{oneida:ident 2}), are denied with the special form \textit{té·kʌ}. Etymologically this form probably is the negative prefix \textit{teʔ-} plus a component \textit{-kʌ-} which in the contemporary language occurs most obviously in the proximal and distal demonstrative words \textit{kaʔikʌ́} `here' and \textit{thikʌ́} `there.'\footnote{\citet[214]{Woodbury2018} hypothesizes that the cognate word in Onondaga, \textit{déʔgęh} includes the verb root \textit{-i-} `to be the total of, be all of it, be only'.} (Again, by regular phonological rule, a glottal stop that follows an accented vowel deletes and the vowel lengthens, and so \textit{teʔ-kʌ} is realized as \textit{té·kʌ}.). 
 
\ea\label{oneida:airplane}
    \gll náhteʔ katiʔ né· lʌ·té· kaʔi·\uline{kʌ́}. Né·n né· kwí· né·n airplane. \\
    what then \textsc{asrt} {he.means} this \textsc{ptcl} \textsc{asrt} \textsc{ptcl} \textsc{ptcl} airplane \\
    \glt `What does he mean by this? It's an airplane.' Margaret Antone \citep[176 (27--28)]{MichelsonEtAl2016}
\z

\ea\label{oneida:ident 1}
    \glll \'okhaleʔ nʌ sʌ waʔkáttokeʔ tsiʔ yah seʔ k\'oskos té·kʌ. \\
    \'okhaleʔ nʌ sʌ waʔ-k-attok-eʔ tsiʔ yah seʔ k\'oskos té·kʌ \\
    and now too \textsc{fact-1sg.a}-notice-\textsc{pnc} \textsc{comp} not \textsc{emph} pig {it’s.not} \\
    \glt `and then I noticed that this was no pig.’ \citet[75]{Abbott1982} 
\z

\ea\label{oneida:ident 2} 
    \begin{xlist}[B: ]
        \exi{A: } 
            \glll  Isé· kʌ tho tsyá·ta\uline{leʔ}. \\
                isé· kʌ tho ts-yaʔt-a-l-eʔ \\
                you \textsc{q} there \textsc{2sg.a}-body-\textsc{join}-be.in-\textsc{stat} \\
            \glt  `Is that you in the picture?'
        \exi{B: } 
            \gll  Yah kiʔ í· té·kʌ. \\
                not {in.fact} \textsc{1} {it's.not} \\
            \glt  `It's not me.' 
    \end{xlist}
\z

Kin relations (\ref{oneida:kin tekv}) and possessed nouns (\ref{oneida:poss tekv}) can also be negated with \textit{té·kʌ}. Possession is usually conveyed by the use of the predicative structure with positional verbs described above, but subsequent references to a possessed entity can be made by means of a possessive prefix on the stem denoting the possessed entity; in this case negation is expressed with \textit{té·kʌ}.

\ea\label{oneida:kin tekv}
    \gll Yah né· laks\'otha té·kʌ, laknulhá· {nʌʔ né·}. \\
    not \textsc{asrt} {my.grandfather} {it's.not} {my.uncle} {that.one} \\
    \glt `That's not my grandfather, he's my uncle.' 
\z

\ea\label{oneida:poss tekv}
    \glll Yah í· akwá·shaleʔ té·kʌ. \\
    yah í· akw-aʔshal-eʔ té·kʌ \\
    not \textsc{1} \textsc{1sg.poss}-knife-\textsc{ns} {it's.not} \\
    \glt `It's not my knife.'  
\z

Interestingly, both kinship and possessed entities can be negated with the pattern that usually applies to clauses, i.e. with the particle \textit{yah} and the prefix \textit{teʔ-}, as in, for example, \textit{yah tehaks\'otha} `he is not my grandfather.' However, this structure is accepted by speakers only sporadically for kin relations, and for possessed entities consistently only with the stem \textit{-awʌ-} `belonging' (\ref{oneida:awv}). (Note that the negative form of \textit{-awʌ-} `belonging in (\ref{oneida:awv b}) has a verbal patient prefix, and not a possessive prefix as in the positive form in (\ref{oneida:awv a}).) Also, it should be noted that some speakers are careful to point out that, though sentences like `he is not my grandfather' do occur, the pragmatics are such that they are not often uttered (regardless of which structure is used). So the way someone might ask about the identity of a person's mother (say an older lady and a young man are standing together) is \textit{Né· uhte kʔ lonulháh?} `It's presumably his mother? i.e. I wonder/You think maybe it's his mother?' And the response then might be \textit{Táh, Lucy né· lonulháh.} `No, Lucy is his mother.'

\ea\label{oneida:awv}
    \ea\label{oneida:awv a}
        \glll Í· né· akwawʌ́ thikʌ́ (oʔwásha\uline{leʔ}). \\
        í· né· akw-awʌ thikʌ́ (oʔwáshaleʔ) \\
        \textsc{1} \textsc{asrt} \textsc{1sg.poss}-belonging that earring(s) \\
        \glt `Those are mine (my earrings).' 
    \ex\label{oneida:awv b} 
        \glll Yah né· akaulhá· teʔyakowʌ́ thikʌ́ (oʔwásha\uline{leʔ}). \\
        yah né·  aka-ulhaʔ teʔ-yako-awʌ thikʌ́ (oʔwáshaleʔ) \\
        not \textsc{asrt} \textsc{3fi.p}-self  \textsc{neg-3fi.p}-belonging that earring(s)\\
        \glt `Those aren't hers (her earrings).' 
    \z
\z
% \hc[inline,size=\tiny]{Matti and Ljuba: This is also very weird, but the authors didn't want to change their gloss \textsc{1}}
% \aut[inline,size=\tiny,color=yellow!40]{"1ST.PERS" is OK, but 1 would not be clear enough to readers.}
% \lv[inline,size=\tiny,color=blue!40]{To Héloîse: I understand why you think it's weird. I am relating this example to what they said above about the constructions encoding class membership. I think the literal translation of this sentence is [It is the case that I have] my earrings}
% \mm[inline,size=\tiny,color=green!40]{but 1st person is abbreviated as 1 in all other chapters and also in LGR, so I don't see why it should be written out full in this chapter.}
% \aut[inline,size=\tiny,color=yellow!40]{Independent pronouns are treated differently from bound pronouns specifiically to avoid confusion and ensure readability.  The difference in glossing indicates the different status of independent pronouns vs. bound pronouns in the language}
% \hc[inline,size=\tiny]{change first.person to 1 > done throughout}
% \hc{add to email: we have changed some things to follow the LGR and the conventions followed by LSP and in this book. (we yield to many things, e.g. don't and not instead of PROH and NEG).}

\subsection{Negation in non-main clauses}

\label{oneida:sec:non-main}

As noted already, standard negation is used in all clauses, and regardless of whether the clause functions as a main or dependent clause, or whether the clause expresses location, existence, membership in a category, or possession. In the excerpt in (\ref{oneida:2 neg}) both clauses are negative.

\ea\label{oneida:2 neg}
    {\glll Yah kwí· teʔwaklʌʔnhá·u yah kánikeʔ  \\
    yah kwí· teʔ-wak-lʌʔnhaʔ-u yah kánikeʔ  \\
    not \textsc{ptcl} \textsc{neg-1sg.p}-know.how-\textsc{stat} not anywhere  \\}
    \sn{tha·hné·she\uline{keʔ}. \\
    th-aa-hn-e-ʔshek-eʔ  \\
    \textsc{cntr-opt-3m.du.a}-walk/go-\textsc{cont-pnc} }
    \glt `I wasn't used to the two of them [my parents] not being there.' Norma Kennedy \citep[118 (8)]{MichelsonEtAl2016}
\z


\subsection{Negative lexicalizations}
\label{oneida:sec:lex}

Oneida has a number of stems whose meanings appear to include some negative component; several are listed in (\ref{oneida:neg stems});  inflected forms of these may be found in \citet{MichelsonDoxtator2002}. These  are not  the negative correspondents of lexical items expressing positive meanings. Only one of the verbs in (\ref{oneida:neg stems}), \mbox{\textit{-atanowʌht-}} `doubt', is clearly analyzable (it is based on the stative verb root \textit{-anowʌ-} `lie', to which the causative suffix has been added, \textit{-anowʌ-ht-} `tell lies', and then the semi-reflexive or middle prefix \textit{-at-anowʌ-ht-} `doubt, be doubtful'). Examples of three of the stems in (\ref{oneida:neg stems}) are given in (\ref{oneida:lexicalized a}--\ref{oneida:lexicalized c}). In addition to the stems in (\ref{oneida:neg stems}) there is also the particle \textit{áhsu} `not yet' mentioned in \sectref{oneida:sec:standard}.

\eabox{\label{oneida:neg stems} 
    \begin{tabular}{lp{7.5cm}}
       \textit{-ahtakw-/-ahtakwʌʔ-} & `be unable to do, fail in an attempt' \\
        \textit{-alahsalʌ-} &`hold back, not want to do something' \\
        \textit{-atanowʌht-} &`doubt' \\
        \textit{-atstahl-} &`be unable to find something' \\
        \textit{-luhyaʔtahkw-} &`be uncooperative' \\
        \textit{-nolu-} &`be unable to quite make it, have a hard time, get stuck' \\
        \textit{-oʔkt-} &`run out of, lack' \\
        \textit{-yehwa-/-yehwha-} & 'be unable to find' \\
    \end{tabular}
} 
% \begin{exe} %HC this is the same content as the table right before, as it was without table format, so it isn't too large, but doesn't look good imo. 
% \ex\label{oneida:neg stems}
% \textit{-ahtakw-/-ahtakwʌʔ-} `be unable to do, fail in an attempt' \\
% \textit{-alahsalʌ-} `hold back, not want to do something' \\
% \textit{-atanowʌht-} `doubt' \\
% \textit{-atstahl-} `be unable to find something' \\
% \textit{-luhyaʔtahkw-} `be uncooperative' \\
% \textit{-nolu-} `be unable to quite make it, have a hard time, get stuck' \\
% \textit{-oʔkt-} `run out of, lack' \\
% \textit{-yehwa-/-yehwha-} be unable to find'
% \end{exe}

\ea\label{oneida:lexicalized a} 
	\glll Né·n waʔtyutkahtúniʔ kaʔikʌ́ átste, kwáh kiʔ né· sayutatyaʔtátstaleʔ utatyʌ́ha,\\
	né·n waʔ-t-yu-atkaht-unyu-ʔ kaʔikʌ́ átste, kwáh kiʔ né· s-a-yutat-yaʔt-atstal-eʔ utatyʌ́ha\\
	\textsc{ptcl} \textsc{fact-dlc-3fi.a}-look-\textsc{distr-pnc} this outside just actually \textsc{asrt} \textsc{rpt-fact-3fi$>$3fi}-body-unable.to.find-\textsc{pnc} {her.daughter} \\
	\glt `She looked all around outside, she just couldn't find her daughter,' Norma Kennedy \citep[52 (23)]{MichelsonEtAl2016}
\z

\ea\label{oneida:lexicalized b}
	\glll swatyelʌ́ s tsiʔ naʔteyukwayenhalá·u thikʌ́ yukwatʌhni·nú· Swatyelʌ́ s waʔakwathnʌnaʔtó·ktʌʔ, \\
	swatyelʌ́ s tsiʔ n-aʔte-yukwa-yenhala{ʔ}-u thikʌ́ yukw-atʌhninu-{ʔ} swatyelʌ́ s waʔ-yakw-at-hnʌnaʔt-oʔkt-ʌʔ \\
	sometimes usually \textsc{comp} \textsc{par-dlc-1pl.p}-get.busy-\textsc{stat} that \textsc{1pl.p}-sell-\textsc{stat} sometimes usually \textsc{fact-1excl.pl.a-srefl}-potato-run.out.of-\textsc{pnc} \\
	\glt `sometimes we would be really busy selling. Sometimes we ran out of potatoes,' Georgina Nicholas \citep[216 (141--142)]{MichelsonEtAl2016}
\z

\ea\label{oneida:lexicalized c} 
	\glll Wahatano·wʌ́hteʔ náhteʔ wahetshlo·\uline{lí·}. \\
	wa-h-at-anowʌ-ht-eʔ náhteʔ wa-hets-hloli-ʔ \\
	\textsc{fact-3m.sg.a-srefl-}lie-\textsc{caus-pnc} what \textsc{fact-2sg$>$3m.sg}-tell-\textsc{pnc}  \\
	\glt `He had doubts about what you told him.' 
\z

\section{Non-clausal negation}
\label{oneida:sec:non-clausal}

\subsection{Negative replies}
\label{oneida:sec:replies}

The words for `yes' and `no' in Oneida (as spoken at the Oneida Nation of the Thames territory) are, respectively, \textit{hʌ́·} and \textit{táh}. Both can function as replies to questions. But only \textit{táh} normally answers a negative polar question.\footnote{According to \citet[180]{SadockZwicky1985} the related language Onondaga has been claimed to have only biased questions. In Oneida, positive polar questions are neutral, and only negative polar questions are biased.} The assumption by the person uttering the negative polar question is that the content is negative, and the answer \textit{táh} `no' agrees with that content. For example, \textit{táh} `no' answers the polar question in (\ref{oneida:neg reply a}), and \textit{táh} `no' or better \textit{áhsu kiʔ} `not yet in fact' is the response to the polar question in (\ref{oneida:neg reply b}).

\ea\label{oneida:neg reply a} 
    \glll Yah kʌ nok thau·tú· a·hoyo·\uline{tʌ́·}.\\
    yah kʌ nok th-aa-w-atu-ʔ aa-ho-yotʌ-ʔ\\
    not \textsc{q} only \textsc{cntr-opt-3z/n.sg.a}-can-\textsc{pnc} \textsc{opt-3m.sg.p}-work-\textsc{pnc} \\
    \glt `He doesn't have to work?' 
\z

\ea\label{oneida:neg reply b} 
    \glll Áhsu kʌ tha·yesaka·lí· otsiʔnu·\uline{wá·}. \\
    áhsu kʌ th-aa-yesa-kali-{ʔ} otsiʔnu·wá· \\
    {not.yet} \textsc{q} \textsc{cntr-opt-3>2sg}-bite-\textsc{pnc} bug(s) \\
    \glt `The bugs haven't bitten you yet?'
\z

A positive \textit{hʌ́·} `yes' response can be given to negative polar questions, but it must be followed by a confirmation, as in example (\ref{oneida:neg reply c}).

\ea\label{oneida:neg reply c} 
    \begin{xlist}[B: ]
    \exi{A: } \glll Yah kʌ tha·hakwe·ní· a·hakhu·\uline{ní·}. \\
    yah kʌ th-aa-ha-kweni-ʔ aa-ha-khuni-ʔ \\
    not \textsc{q} \textsc{cntr-opt-3m.sg.a}-able-\textsc{pnc} \textsc{opt-3m.sg.a}-cook-\textsc{pnc} \\
    \glt `He can't cook?'
    \exi{B: } \glll  Hʌ́·, ʌhakwe·ní· kih. \\
    hʌ́· ʌ-ha-kweni-ʔ kiʔ \\
     yes  \textsc{fut-3m.sg.a}-able-\textsc{pnc} actually \\
    \glt `Yes, he can actually.' 
    \end{xlist}
\z

As mentioned in \sectref{oneida:sec:non-declarative}, most attested negative polar questions in our corpus are indirect speech acts used for special effect, and often in a good-natured way. They can sometimes elicit a positive answer, without the hearer agreeing with the content, or even treating it as a question. In fact, some of these negative questions seem to invite a positive answer, but one which contradicts the intent of the person asking the question, as in the example in (\ref{oneida:neg reply d}); here a mother is admonishing her child, and the child may be thinking `yes, of course' with respect to the literal reading of the question but may not be in agreement with the intent of the question. In other cases, the negative question may be used playfully; the excerpt in (\ref{oneida:neg reply e}) was spoken at the end of a scary story about a ghost. The story-teller is uttering this question as a way of engaging her audience. (In (\ref{oneida:neg reply e}) the \textit{ʔ} of the negative prefix \textit{teʔ-} is regularly deleted before a following \textit{h}.)

\ea\label{oneida:neg reply d} 
    \glll Yah kʌ thaʔtehsahuhtaká·lu\uline{teʔ}.   \\
    yah kʌ th-aʔte-hs-ahuht-a-kahlut-eʔ   \\
    not \textsc{q} \textsc{cntr-dlc-2sg.a}-ear-\textsc{join}-be.an.opening-\textsc{stat} \\
    \glt `Don't you have holes in your ears?' \citep[83]{MichelsonDoxtator2002}
\z

\ea\label{oneida:neg reply e} 
    \glll Yah kʌ téhselheʔ a·sheyaʔtisák\uline{haʔ}.   \\
    yah kʌ teʔ-hs-elh-eʔ aa-she-yaʔt-isakh-aʔ   \\
    not \textsc{q} \textsc{neg-2sg.a}-want-\textsc{stat} \textsc{opt-2sg$>$3}-body-go.look.for-\textsc{pnc} \\
    \glt `Don't you want to go and look for her?' Verland Cornelius \citep[69 (45)]{MichelsonEtAl2016}
\z

\subsection{Negative indefinites and quantifiers}
\label{oneida:sec:indef}

There are no negative quantifiers in Oneida; negative existential quantification corresponding to English \textit{nothing}, \textit{nobody} and the like (called  \textit{negative indefinites} in \citealt[194]{Haspelmath1997}) is expressed using standard clausal negation, i.e. the negative particle \textit{yah} and the negative form of the verb together with a negative polarity item (specifically a word that otherwise functions as an interrogative pronoun) in the scope of \textit{yah}.
Negative polarity items are introduced in the section on negative polarity, \sectref{oneida:sec:npi}.
 
\subsection{Negative derivation}
\label{oneida:sec:deriv}

Semantically, notions such as the reverse of a situation may be construed as a kind of negation, and in Oneida certain pairs of  contrary meanings are conveyed morphologically: some actions (e.g. open/close) (\ref{oneida:deriv rev}a), scalar expressions (e.g. long/short) (\ref{oneida:deriv rev}b), and transfer of ownership (e.g. buy/sell) (\ref{oneida:deriv rev}c). These are conveyed, respectively, by the ``reversative'' suffix, the diminutive ending, and the ``semi-reflexive'' prefix, although it should be noted that each of these morphemes has other functions as well. There is no derivational equivalent to English \textit{un-} or \textit{-less}. 


% \eabox{\label{oneida:deriv rev} %HC This is one way I tried to make the table before look better, but it is not ideal either. See outcommented example below (without table format)
%     \begin{tabular}{lll}
%         a. & \textit{-nhotu-}  `close a door'	& \textit{-nhotukw}-	`open a door' \\
%         & \textit{-ashalut-} `tie someone up' & \textit{-ashalutakw-} `untie, unleash' \\
%         & \textit{-hweʔnuni-} `wrap into a package, bundle' & \textit{-hweʔnunyahsyu-}  `unwrap, unbundle' \\
%         b. & \textit{-es-} `be a (certain) length' & \textit{-esha-} `be short' \\
%         & \textit{-aʔslatye-} `be a (certain) width' & \textit{-aʔslatyeha-} `be narrow' \\
%         c. & \textit{-hninu-} `buy' & \textit{-atʌhninu-} `sell' \\
%         & \textit{-niha-} `lend'&	\textit{-atʌniha-} `borrow' \\
%     \end{tabular}
% }\hc{LSP do you have a solution here for example \REF{oneida:deriv rev}? see outcommented other ideas. and a direct reference link to a-b-c?}

\TabPositions{5cm}%another try, not really good either
\ea\label{oneida:deriv rev}
    \ea
        \textit{-nhotu-}  `close a door'	\tab \textit{-nhotukw}-	`open a door' \\
        \textit{-ashalut-} `tie someone up'    \tab \textit{-ashalutakw-} `untie, unleash' \\
        \textit{-hweʔnuni-} `wrap into a package, bundle' \\
                                            \tab \textit{-hweʔnunyahsyu-}  `unwrap, unbundle' \\
    \ex
        \textit{-es-} `be a (certain) length' \tab \textit{-esha-} `be short' \\
        \textit{-aʔslatye-} `be a (certain) width' \tab \textit{-aʔslatyeha-} `be narrow' \\
    \ex
        \textit{-hninu-} `buy' \tab \textit{-atʌhninu-} `sell' \\
        \textit{-niha-} `lend' \tab	\textit{-atʌniha-} `borrow' \\
    \z
\z

% \begin{exe} %HC this is the same content as the table right before, as it was without table format
%     \ex\label{oneida:deriv}
%     \begin{xlist}
%     \ex\label{oneida:deriv rev} 
%         \textit{-nhotu-}  `close a door',	\textit{-nhotukw}-	`open a door' \\
%         \textit{-ashalut-} `tie someone up',	\textit{-ashalutakw-} `untie, unleash' \\
%         \textit{-hweʔnuni-} `wrap into a package, bundle',  \textit{-hweʔnunyahsyu-}  `unwrap, unbundle' 
%     \ex\label{oneida:deriv sim} 
%         \textit{-es-}	`be a (certain) length',		\textit{-esha-}		`be short' \\
%         \textit{-aʔslatye-} `be a (certain) width', \textit{-aʔslatyeha-} `be narrow' 
%     \ex\label{oneida:deriv srefl} 
%         \textit{-hninu-}	`buy',		\textit{-atʌhninu-}	`sell' \\
%         \textit{-niha-} `lend',	\textit{-atʌniha-}  `borrow'
%     \end{xlist}
% \end{exe}

\subsection{Other negative expressions}
\label{oneida:sec:other non-clausal}

In our corpus there are examples of utterances with \textit{yah}, but no  (negative) verb. Because there is no verb they are aptly discussed in a section devoted to non-clausal negation, although we do not commit to an analysis in which all examples involve non-clausal negation. In this type of example what is denied is that a particular entity is the right one (\ref{oneida:notso 6}) and (\ref{oneida:notso 7}), or belongs to the right category (\ref{oneida:notso 5}).\footnote{We do not provide a line that segments words into morphs for examples that contain only uninflected words or where inflected words are not analyzed.} The entity is specified by a noun or other nominal element -- the particle cluster \textit{nʌʔ né·} in (\ref{oneida:notso 6}), \textit{í· lakeʔníha} in (\ref{oneida:notso 7}), \textit{é·lhal} in (\ref{oneida:notso 5}); what is being negated may be known from previous context, as in (\ref{oneida:notso 6}) and (\ref{oneida:notso 7}), or it can be mentioned later in the same utterance (\ref{oneida:notso 5}).

\ea\label{oneida:notso 6}
    \gll Ok neʔ thikʌ́, yah  {nʌʔ né·}. \\
    but \textsc{ptcl} that not {that.one} \\
    \glt `(So then you turn on a light, it illuminates the whole house; if you turn on a light.) But not that one.' Verland Cornelius \citep[112 (17)]{MichelsonEtAl2016}
\z  

\ea\label{oneida:notso 7}
    \gll Yah kiʔ né·n í· lakeʔní\uline{ha}. \\
    not {in.fact} \textsc{ptcl} \textsc{1} {my.father} \\
    \glt `(Your father is mean.) Not my father.' 
\z%\hc{Ljuba and Matti: we should write in.fact with a dot, but the auhtors are against that. I tried to argue with them, without success, even though that is quite clear in the LGR. Same problem many times in this chapter. See JPK's answer outcommented below}
    %JPK's answer: As far as we can see, the last issue concerns the addition of a dot in some glosses. We would like to argue against the dot. The Leipzig glossing rules require dots when more than a single word in the metalanguage (English in our case) corresponds to a single morph in the described language (Oneida in our case). This is not the situation here, so we cannot add dots. But your point about readability is well taken. [followed by an explanation of why the LSP's formatting isn't good enough for him and he wants to change everything]
    %HC: I do not understand why this is not the situation there and why we couldn't add dots.

\ea\label{oneida:notso 5}
    \glll yah kwí· né· shekú né·n é·lhal yah tha·hakwe·ní· aá·lakeʔ ni·y\'ot. \\
    yah kwí· né· shekú né·n é·lhal 	yah th-aa-ha-kweni-ʔ aa-hla-k-eʔ ni·y\'ot \\
    not \textsc{ptcl} \textsc{asrt} even \textsc{ptcl} dog not \textsc{cntr-opt-3m.sg.a}-able-\textsc{pnc} \textsc{opt-3m.sg.a}-eat-\textsc{pnc} {how.it.is} \\
    \glt `not even a dog could eat it, how it was.'  Clifford Cornelius \citep[225 (53)]{MichelsonEtAl2016}
\z


Sometimes it is a whole proposition that is denied, as in \REF{oneida:notso 8}; hence the English translation `it wasn't the case.'.

\ea\label{oneida:notso 8}
    \gll Ok  {neʔ th\'o·neʔ} yah  {nʌʔ né·}, \\
    but {at.that.time} not {that's.it} \\ 
    \glt `(Even to cut wood you have to have an education.) But at that time it wasn't the case,'  Clifford Cornelius \citep[236 (116)]{MichelsonEtAl2016}
\z 


The negative particle \textit{yáhtʌʔ} `not so' also functions to deny an entire proposition; but it appears additionally to be anaphoric. In other words, \textit{yáhtʌʔ} seems to function as a negative propositional anaphor (\ref{oneida:notso 1}--\ref{oneida:dependent}).\footnote{\textit{Yáhtʌʔ} is the word for `no' in the Wisconsin variety of Oneida, and is clearly related to \textit{yah}. This is  the only context where \textit{yáhtʌʔ} is attested in our corpus of the Ontario variety of Oneida.}

\begin{exe}
\ex\label{oneida:notso 1}	
    {\glll {N\'ok tsiʔ} sʌ́haʔ kiʔ yah	teʔkatt\'okhahkweʔ kwí· \\
    {n\'ok tsiʔ} sʌ́haʔ kiʔ yah	 teʔ-k-attok-hahkweʔ 	kwí·  \\
    but more actually not \textsc{neg-1sg.a}-be.wise-\textsc{hab.pst} \textsc{ptcl}  \\}
    \sn{\glll utahséhtahkweʔ ok oniʔ n yáhtʌʔ,  \\
    u-t-a-hs-ehtahkw-eʔ ok oniʔ n	 yáhtʌʔ \\
    \textsc{opt-cis-opt-2sg.a}-believe-\textsc{pnc} {and} or \textsc{ptcl} {not.so}  \\
    }
    \glt `But even more I had no sense, you can believe it or not,' Clifford Cornelius \citep[228 (74)]{MichelsonEtAl2016}
\end{exe}

\ea\label{oneida:notso 2}
    \gll {N\'ok tsiʔ}  {neʔ th\'o·neʔ} yah kwí· náhteʔ got no brain. Shekú núwaʔ né· yáhtʌʔ. \\
    but {at.that.time} not \textsc{ptcl} anything got no brain still {this.time} \textsc{asrt} {not.so} \\
    \glt `But back then I didn't have any brains. I still don't.' Verland Cornelius \citep[320 (82--83)]{MichelsonEtAl2016}
\z

\ea\label{oneida:dependent}
    \glll Tá·t yáhtʌʔ ʌhotiná·khwʌʔ kiʔ, ʌyesaná·khwahseʔ tá·t yah tha·shni·\uline{nú·}. \\
    tá·t yáhtʌʔ ʌ-hoti-na{ʔ}khwʌ-ʔ kiʔ ʌ-yesa-naʔkhw-a-hs-eʔ tá·t yah th-aa-hs-hninu-ʔ \\
    if {not.so} \textsc{fut-3m.pl.p}-get.mad-\textsc{pnc} {actually} \textsc{fut-3>2sg}-get.mad-\textsc{join-ben-pnc} if not \textsc{cntr-opt-2sg.a}-buy-\textsc{pnc} \\
    \glt `If not, they'll get mad, they will get mad at you if you won't buy it.' Verland Cornelius \citep[326--327 (146)]{MichelsonEtAl2016}
\z

\section{Other aspects of negation}
\label{oneida:sec:other aspects}

\subsection{The scope of negation}
\label{oneida:sec:scope}

Scope is generally left-to-right in Oneida. This is shown with the quantificational word \textit{akwekú} `all, the whole of' in the two contrasting examples below (\ref{oneida:not akweku}--\ref{oneida:akweku not}).

\ea\label{oneida:not akweku}
    \glll Yah akwekú teʔshakokwáhtu. \\
    yah	akwekú teʔ-shako-kwaht-u \\ 
    not all \textsc{neg-3m.sg$>$3}-invite-\textsc{stat}  \\
    \glt `He didn't invite all of them, everyone.' 
\z

\ea\label{oneida:akweku not} 
    {\glll     nʌ  {kyuniʔ} wí· né· akwekú yah kánikeʔ \\
            nʌ kiʔ-uni{ʔ} wí· né· akwekú yah kánikeʔ \\
    now	{in.fact}-too \textsc{ptcl}	\textsc{asrt} all not anywhere  \\}
    \sn{teʔshʌ·né\uline{·seʔ}. \\
    teʔ-s-hʌn-e-ʔseʔ \\
    \textsc{neg-rpt-3m.pl.a}-walk/go-\textsc{hab}
    }
    \glt `(It used to be just [names],) they're all not around anymore.' Pearl Cornelius \citep[309 (112)]{MichelsonEtAl2016}
\z  

However when eliciting examples like these, speakers generally reject the order attested in (\ref{oneida:akweku not}). In this example those who are no longer around were identified in the preceding discourse. Out of context, though, speakers prefer to use a negative indefinite expression (see \sectref{oneida:sec:npi}), as in (\ref{oneida:akweku nobody}) (`not anyone' rather than `everyone not'). Or a close (though not exact) equivalent of \textit{akwekú}, \textit{kwah tsiʔ nihatí} can be used, as in (\ref{oneida:akweku kwah tsi}).

\ea\label{oneida:akweku nobody}	
    \glll yah né· úhkaʔ tha·yesa·\uline{kʌ́·}.  \\
    yah né· úhkaʔ th-aa-yesa-kʌ-ʔ  \\
    not \textsc{asrt} anyone \textsc{cntr-opt-3$>$2sg}-see-\textsc{pnc} \\
    \glt `nobody will see you.' Norma Kennedy \citep[128 (16)]{MichelsonEtAl2016}
\z

\ea\label{oneida:akweku kwah tsi} 
    {\glll
    Kwah tsiʔ nihatí yah  \\
    kwah tsiʔ ni-hat-i 	yah  \\
    just how \textsc{par-3m.pl.a}-make.up.the.whole.of\textsc{[stat]} not  \\} %HC here we have points between the words (already in original).
    \sn{\glll tehone·ká·seʔ onuʔúse\uline{liʔ}. \\
    te-hon-ekaʔ-seʔ onuʔúseliʔ \\
    \textsc{neg-3m.pl.p}-like.the.taste-\textsc{hab} squash \\
    }
    \glt `However many there are (all of them) don't like squash.' 
\z

A contrast in scope is also attested with the modal verb \textit{-atu-} `possible, can be' (see \sectref{oneida:sec:non-declarative} for additional examples of this verb). In the excerpt in (\ref{oneida:scope 1}), the speaker is talking about something which is not possible, but in (\ref{oneida:scope 2}) the speaker is talking about the possibility of a negative situation.

\ea\label{oneida:scope 1}	
    {\glll yah kwí· thau·tú· \'oksaʔ \\
    yah kwí· th-aa-w-atu-ʔ \'oksaʔ  \\
    not \textsc{ptcl} \textsc{cntr-opt:3z/n.sg.a}-can-\textsc{pnc} {right.away}  \\}
    \sn{a·kyatekhu·ní· \\
    aa-ky-atekhuni-ʔ \\
    \textsc{opt-3fz.du.a}-eat.a.meal-\textsc{pnc}
    }
    \glt `they coudn't eat right away' Norma Kennedy \citep[50 (3)]{MichelsonEtAl2016}
\z

\ea\label{oneida:scope 2}
    {\glll Né· tsiʔ ʌwa·tú· seʔ yah \\
    né· tsiʔ ʌ-w-atu-ʔ seʔ yah  \\
     \textsc{asrt} \textsc{comp} \textsc{fut-3z/n.sg.a}-can-\textsc{pnc} \textsc{emph} not  \\ }
     \sn{\glll  tha·hotiyo·\uline{tʌ́·}. \\
    th-aa-hoti-yotʌ-ʔ \\
      \textsc{cntr-opt-3m.pl.p}-work-\textsc{pnc} \\
    \glt `Because it's possible for them not to work.' Verland Cornelius \citep[328 (156)]{MichelsonEtAl2016}
}
\z

\subsection{Negative polarity}
\label{oneida:sec:npi}

In Oneida, indefinite expressions form a paradigm with words that are used in content questions and that we analyze as negative polarity items. These are: \textit{úhkaʔ} `anyone, who?' \textit{náhteʔ} `anything, what?', \textit{kátshaʔ} `anywhere, where?,' \textit{kánhke} `anytime, when?,' and \textit{t\'o·} `how' (used in questions having to do with manner and quantity, such as `how long \ldots'). Some of these are followed in some contexts by particles which have a classificatory function, such as \textit{nú·} in \textit{kátshaʔ nú·} `when?' The particle \textit{ok} (otherwise `only, and') forms corresponding indefinite expressions, such as \textit{úhkaʔ ok} `someone,' \textit{tok náhteʔ} or \textit{thok náhteʔ} `something,' \textit{kátshaʔ ok nú·} `somewhere.' Indefinite expressions in Oneida  thus make use of both major kinds of ``derivational bases'' identified in \citet[26]{Haspelmath1997}: words used in content questions and generic ontological classifiers. Negative indefinites are expressed using standard clausal negation, i.e. the negative particle \textit{yah} and the negative form of the verb together with a negative polarity item in the scope of \textit{yah}, as in (\ref{oneida:indef a}) and (\ref{oneida:indef b}). The excerpt in (\ref{oneida:indef c}) is a negative indefinite with \textit{tákʌʔ}, used in prohibitives (see \sectref{oneida:sec:non-declarative}  on non-declarative negation.)

\ea\label{oneida:indef a} 
    \glll Yah úhkaʔ teʔyé·tluʔ,   \\
    yah úhkaʔ teʔ-ye-iʔtlu-ʔ   \\
    not anyone \textsc{neg-3fi.a}-sit/stay-\textsc{stat}  \\
    \glt `Nobody lived there,' Mercy Doxtator \citep[143 (6)]{MichelsonEtAl2016}
\z

\ea\label{oneida:indef b} 
    \glll Yah kiʔ ní· nuwʌtú náhteʔ teʔyukyatkáthu í· kháleʔ Masyha, \\
    yah kiʔ ní· nuwʌtú náhteʔ teʔ-yuky-atkath-u í· kháleʔ Masyha \\
    not {actually} \textsc{1} ever anything \textsc{neg-1du.p}-see-\textsc{stat} \textsc{1} and {dear.Mercy} \\
    \glt `But the two of us never ever saw anything, me and Mercy,' Olive Elm \citep[64 (24)]{MichelsonEtAl2016}
\z

\ea\label{oneida:indef c} 
    {\glll wahatá·listeʔ tákʌʔ úhkaʔ a·yehlotʌ́  \\
    wa-h-atahlist-eʔ tákʌʔ úhkaʔ aa-ye-hlotʌ   \\
    \textsc{fact-3m.sg.a}-forbid-\textsc{pnc} don't anyone \textsc{opt-3fi.a}-smoke[\textsc{imp}]  \\}
    \sn{\glll tho kanúsku\\
    tho kanúsku \\
    there {in.the.house}\\
    }
    \glt `he tells people no one should smoke in the house' Mercy Doxtator \citep[186 (16)]{MichelsonEtAl2016}
\z
 
The contrast between \textit{yah úhkaʔ} `not anyone' (`no one') in (\ref{oneida:not anyone}) and \textit{úhkaʔ yah} `anyone not' in (\ref{oneida:anyone not}) shows the expected scopal relation of the negative polarity item with the negation.

\ea\label{oneida:not anyone}
    \glll Yah úhkaʔ teʔyakonúh\uline{teʔ}. \\
    yah úhkaʔ teʔ-yako-anuhte-ʔ \\
    not anyone \textsc{neg-3fi.p}-know-\textsc{stat} \\
    \glt `No one knows.' Margaret Antone \citep[177 (47)]{MichelsonEtAl2016}
\z

\ea\label{oneida:anyone not}
    \glll úhkaʔ yah tha·yutawyaʔtá·naʔ, \\
    úhkaʔ yah th-aa-yu-atawyaʔtahn-aʔ \\
    anyone not \textsc{cntr-opt-3fi.a}-go.to.school-\textsc{pnc} \\
    \glt `[if] anyone wouldn't go to school,' Verland Cornelius \citep[328 (160)]{MichelsonEtAl2016}
\z

In addition to questions and clausal negations, NPIs occur with the particle \textit{ati} `no matter, never mind, although,' which does not require a negative verb, but whose meaning suggests that no one or nothing can be excluded from the situations being described. Examples are the excerpts in (\ref{oneida:ati 1}) and (\ref{oneida:ati 2}).

\ea\label{oneida:ati 1}	
    \glll Tsiʔ kwí· né· nihoyaʔtaseʔtslaká·teʔ, ati né· úhkaʔ á·neʔ kih. \\
    tsiʔ kwí· né· ni-ho-yaʔtaseʔtsl-a-kaʔte-ʔ ati né· úhkaʔ aa-hn-e-ʔ kiʔ \\
    \textsc{comp} \textsc{ptcl} \textsc{asrt} \textsc{par-3m.sg.p}-girlfriend-\textsc{join}-have.many-\textsc{stat} {no.matter} \textsc{asrt} anyone \textsc{opt-3m.du.a}-go-\textsc{pnc} actually \\ 
    \glt `He had so many girlfriends, he would go out with just anybody (nevermind who).' Georgina Nicholas \citep[32 (6)]{MichelsonEtAl2016}
\z

\ea\label{oneida:ati 2}	
    \glll tsiʔ ati úhkaʔ ʌhakwe·ní· kiʔ ʌhoyo·tʌ́· tsiʔ náhteʔ waky\'o·tʌh\uline{seʔ}. \\
    tsiʔ ati úhkaʔ ʌ-ha-kweni-ʔ kiʔ ʌ-ho-yotʌ-ʔ tsiʔ náhteʔ wak-yo{ʔ}tʌ-{ʔ}seʔ \\
    \textsc{comp} {no.matter} anyone \textsc{fut-3m.sg.a}-able-\textsc{pnc} actually \textsc{fut-3m.sg.p}-work-\textsc{pnc} \textsc{comp} anything \textsc{1sg.p}-work-\textsc{hab}\\ 
    \glt `because anybody (no matter who) could do the work that I worked at.' Clifford Cornelius \citep[233 (101)]{MichelsonEtAl2016}
\z

Two additional NPIs: \textit{nuwʌtú} `ever,' and \textit{kánikeʔ} `nowhere' have a more restricted distribution in that they cannot co-occur with the \textsc{ptcl} \textit{ok} to form indefinite expressions. The NPI \textit{nuwʌtú} `ever' occurs only in polar questions (\ref{oneida:ever}) and in negative clauses (\ref{oneida:never 1}) and (\ref{oneida:never 2}). \textit{Kánikeʔ} oocurs only in negative clauses (\ref{oneida:kanike}). Examples (\ref{oneida:never 2}) and (\ref{oneida:kanike}) have two negative polarity items in the same clause. (Another example with \textit{kánikeʔ} is (\ref{oneida:akweku not}) in \sectref{oneida:sec:scope}.)

\ea\label{oneida:ever}
    \glll Nʌ kʌ nuwʌtú Tony Roma's yesatekhu·n\uline{í}. \\
    nʌ kʌ nuwʌtú Tony Roma’s ye-s-atekhuni \\
     now \textsc{q} ever Tony Roma's \textsc{trl-2sg.p}-eat.a.meal[\textsc{stat}] \\
    \glt `Have you ever eaten at Tony Roma’s?' 
\z

\ea\label{oneida:never 1}	
    \glll Kwáh yah nuwʌtú thaʔteyukyatlihotálhu, \\
    kwáh yah nuwʌtú th-aʔte-yuky-atlihotalh-u \\
     just not ever \textsc{cntr-dlc-1du.p}-quarrel-\textsc{stat} \\ 
    \glt `We never ever quarreled,' Norma Kennedy \citep[147 (8)]{MichelsonEtAl2016}
\z

\ea\label{oneida:never 2}
    \glll Né·n yah kiʔ nuwʌtú úhkaʔ teʔshakohlo·l\uline{í}. \\
    né·n yah kiʔ nuwʌtú úhkaʔ teʔ-shako-hloli \\
    \textsc{ptcl} not actually ever anyone \textsc{neg-3m.sg$>$3}-tell[\textsc{stat}]  \\
    \glt `He never told anyone.' Norma Kennedy \citep[78 (48)]{MichelsonEtAl2016}
\z

\ea\label{oneida:kanike}
    \glll wé·ne tsiʔ yah úhkaʔ kánikeʔ té·yʌh\uline{seʔ}. \\
    wé·ne tsiʔ yah úhkaʔ kánikeʔ teʔ-yʌ-e-ʔseʔ \\
      evidently \textsc{comp} not anyone anywhere \textsc{neg-3fi.a}-walk/go-\textsc{hab}  \\
    \glt `must be that there was no one there' Norma Kennedy \citep[77 (33)]{MichelsonEtAl2016}
\z

\subsection{Marking of NPs in the scope of negation}
\label{oneida:sec:NP}

The occurrence of referring expressions (nouns or noun phrases) in the scope of negation does not affect how they are expressed.

\subsection{Reinforcing negation}
\label{oneida:sec:emph}

The so-called assertion particle \textit{né·} occurs widely in Oneida; much remains to be understood about this particle, but one of its functions is to signal (narrow) focus, as in the following excerpt:

\ea\label{oneida:focus potato}
    \begin{xlist}[B: ]
        \item[A: ] \gll Wá·lelheʔ né· kʌ n swahyo·wáneʔ, ohnaná·taʔ  {nʌʔ né·}. \\
        {he.thought} \textsc{asrt} \textsc{q} \textsc{ptcl} apple potato {that.one} \\
        \glt `He thought it was an apple, it was a potato.' \\
        \item[B: ] Ohnaná·taʔ. \\
        {} potato\\
        \glt `A potato?' \\
        \item[A: ] \gll Hʌ́·, ohnaná·taʔ né· lonʌskwʌháti\uline{ʔ}. \\
         yes potato \textsc{asrt} {he.is.going.along.stealing} \\
        \glt `Yes, he was stealing a potato .' Pearl Cornelius \citep[303 (50--52)]{MichelsonEtAl2016}
    \end{xlist}
\z

The particle \textit{né·} occurs in negative contexts, as in the second clause in the excerpt in (\ref{oneida:focus neg}).

\ea\label{oneida:focus neg}
    \glll né· kwí· tsiʔ teki·tʌ́heʔ seʔ niʔí·, yah né· tha·hatikwe·ní· ta·hati·tʌ́· thikʌ́ tho thʌ·né·seʔ,\\
    né· kwí· tsiʔ te-k-itʌ-heʔ seʔ niʔí· yah né· th-aa-hati-kweni-ʔ t-aa-hati-tʌ-ʔ thikʌ́ tho t-hʌn-e-ʔseʔ \\
    \textsc{asrt} \textsc{ptcl} \textsc{comp} \textsc{dlc-1sg.a}-fly-\textsc{hab} \textsc{emph} \textsc{1}, not \textsc{asrt} \textsc{cntr-opt-3m.pl.a}-able-\textsc{pnc} \textsc{dlc-opt-3m.pl.a}-fly-\textsc{pnc} that there \textsc{cis-3m.pl.a}-walk/go-\textsc{hab} \\
    \glt `because me too, I fly, they can't fly, the ones who are walking around over there,' Ruben Cutcut \citep[36 (8)]{MichelsonEtAl2016}
\z 

NPIs generally occur in focus position, after the negative particle \textit{yah}, and in appropriate pragmatic contexts the order of NPIs and the modal verb \textit{yah thau·tú·} `can't' (\sectref{oneida:sec:non-declarative}) can interact to encode two distinct meanings. In (\ref{oneida:scope 3}) the speaker cannot do anything because there \textit{is} nothing to do, with the NPI in narrow focus position, while in (\ref{oneida:scope 4}) the speaker cannot do anything because, say, s/he hasn't gotten permission.

\ea\label{oneida:scope 3}
    \glll Yah náhteʔ thau·tú· na·kátye\uline{leʔ}. \\
    yah náhteʔ th-aa-w-atu-ʔ n-aa-k-atyel-eʔ \\
    not anything \textsc{cntr-opt-3z/n.sg.a}-can-\textsc{pnc} \textsc{par-opt-1sg.a}-do-\textsc{pnc} \\
    \glt `I can't do anything, there's nothing I can do.' 
\z

\ea\label{oneida:scope 4}
    \glll Yah thau·tú· náhteʔ na·kátye\uline{leʔ}. \\
    yah th-aa-w-atu-ʔ náhteʔ n-aa-k-atyel-eʔ \\
    not \textsc{cntr-opt-3z/n.sg.a}-can-\textsc{pnc} anything \textsc{par-opt-1sg.a}-do-\textsc{pnc} \\
    \glt `I can't do (e.g. I'm not allowed to do) anything.' 
\z

Finally, this might be a good place to mention the emphatic word for `no' \textit{ta·ím}, although it does not appear to be used very often. The excerpt below comes from a conversation about how poor many of the Oneida were when they were growing up; one of the two speakers observes that in spite of what they now know to be true (that they were poor), that's not what they thought of their situation at the time. The other speaker responds saying \textit{ta·ím}, agreeing with the observation (that they didn't think they were poor).

\ea\label{oneida:taim}
	\gll {Nók tsiʔ} thikʌ́  {neʔ thó·neʔ} tshiwathawinútiʔ, yah kiʔ nuwʌtú aʔnyóh né· teʔtwanuhtúnyuheʔ tsiʔ niyukwʌ·tʌ́t. \\
	but that {at.that.time} {during.those.times} not actually ever {seems.like} \textsc{asrt} {we.don't.think} how {we.are.poor} \\
	\glt `But back in those times, we never seemed to think that we were that poor.' Pearl Cornelius \citep[304 (63)]{MichelsonEtAl2016}
\z

\subsection{Negation, coordination, and complex clauses}
\label{oneida:sec:except}

In this section we provide examples of clauses describing exceptions to a generalization, the equivalent of English uses of `except' and `instead of.' We also discuss here a phenomenon to which we have assigned the term \textit{negative spreading}.

We begin with coordination, and in particular with exception clauses. When the first clause includes the equivalent of universal quantification, as in (\ref{oneida:except 1}) and (\ref{oneida:except 2}), the exception clause has just the negative particle \textit{yah} or \textit{yáhtʌʔ} (see \sectref{oneida:sec:other non-clausal} for other contexts that have only the negative particle).

\ea\label{oneida:except 1}
    \gll Kwah tsyoʔk niwahy\'o·tʌhseʔ wake·ká·seʔ  {n\'ok tsiʔ} yah kiʔ né·n raspberries. \\
    just all {different.kinds.of.fruit} {I.like.the.taste} but not actually \textsc{ptcl} raspberries \\
    \glt `I like the taste of (I like to eat) every kind of fruit, but not raspberries (except for raspberries).' 
\z

\ea\label{oneida:except 2}
    \gll Akwekú wá·lakeʔ  {kwah n\'ok} ohnaná·taʔ yáhtʌʔ. \\
    all {he.ate} just potato(es) {not.(so)} \\
    \glt `He ate it all, just not the potatoes (except for the potatoes).' 
\z

In the coordination in (\ref{oneida:except 3}) the negative clause comes first and is expressed as the usual standard negation, with both negative \textsc{ptcl} and negative prefix on the verb.

\ea\label{oneida:except 3}
    \glll Yah náhteʔ oyá· tehawekú  {kwah n\'ok} ohnaná\uline{·taʔ}. \\
    yah náhteʔ oyá· te-haw-ek-u {kwah n\'ok}  ohnaná·taʔ \\
    not anything other \textsc{neg-3m.sg.p}-eat-\textsc{stat} just  potato(es) \\
    \glt `He didn't eat anything else, just the potatoes.'  
\z

A positive situation that is an alternative to a non-event is expressed as comparison (`rather than, as compared with'), as in \REF{oneida:comparison}.

\begin{exe}
\ex\label{oneida:comparison}
    \begin{xlist}
    \ex\label{oneida:comparison 1}
        \gll Sarnia yahá·leʔ tsiʔ ni·y\'ot London. \\
        Sarnia {he.went.over.there} \textsc{comp} {how.it.is} London \\
        \glt `He went to Sarnia rather than (instead of) London (Ontario).' 
    \ex\label{oneida:comparison 2}
        \gll Sʌ́haʔ lahnʌ·yés tsiʔ ní· ni·y\'ot. \\
         more {he.is.tall} \textsc{comp} \textsc{1} {how.it.is} \\
        \glt `He is taller than I am.' 
    \end{xlist}
\end{exe}

Finally, the closest equivalent to the English coordinator `lest' includes the particle \textit{tákʌʔ}; see examples in \sectref{oneida:sec:non-declarative}.

We now turn to a different phenomenon, which we call "negative "spreading". 
As we have seen so far, Oneida is a language where clausal negation involves the simultaneous expression of two morphs (particle and prefix), or \textit{double negation}. But, under certain conditions not completely understood at this time, two verbs occur with a negative prefix; yet only one instance of the particle \textit{yah}.
Some examples of this are given below.
 
It appears that negative spreading is required with certain expressions of quantity, extent, or manner (\ref{oneida:double 1}--\ref{oneida:double 4}).\footnote{Some of the forms in the examples require explanation for those interested in details that go beyond negation. In (\ref{oneida:double 1})  the verb root of \textit{té·ku} is \textit{-u-} `be  amount, amount to.' The prefix \textit{k-} is an anomalous form. In (\ref{oneida:double 2}) the idea of `every' is conveyed by the verb \textit{-ke-} `amount to' plus the translocative and dualic (and sometimes also the partitive) prefixes.}

\ea\label{oneida:double 1}
	\glll {tá·t núwaʔ} yah tho té·ku tehoti·yʌ́· atáhsliʔ, \\
	{tá·t núwaʔ} yah tho teʔ-ku te-hoti-yʌ-ʔ atáhsliʔ \\
	maybe not thus \textsc{neg}-amount \textsc{neg-3m.pl.p}-put/have-\textsc{stat} stick(s) \\
	\glt `maybe they didn't have enough sticks,' Mercy Doxtator \citep[258 (76)]{MichelsonEtAl2016}
\z


\ea\label{oneida:double 2}
	\glll Yah thyaʔtewʌhnislaké teʔwaky\'o·tʌh\uline{seʔ}.\\
	yah th-y-aʔte-w-ʌhnisl-a-ke teʔ-wak-yo{ʔ}tʌ-ʔseʔ\\
	not \textsc{cntr-trl-dlc-3z/n.sg.a}-day-\textsc{join}-amount.to[\textsc{stat}] \textsc{neg-1sg.p}-work-\textsc{hab} \\
	\glt `I don't work everyday.' 
\z

\ea\label{oneida:double 3}
	\glll Yah tho teʔyo·lé· thaʔteyakonuhsanú·yaniht tsiʔ ni·y\'ot yesaʔkʌ́\uline{ha}. \\
	yah tho teʔ-yo-le-ʔ th-aʔte-yako-nuhs-anuhyaniht tsiʔ ni·y\'ot yesaʔkʌ́ha \\
	not thus \textsc{neg-3z/n.sg.p}-how.far-\textsc{stat} \textsc{cntr-dlc-3fi.sg.p}-house-be.dirty\textsc{[stat]} \textsc{comp} {how.it.is} {your.older.sister} \\
	\glt `Her house is not as dirty as your older sister's.' 
\z

\ea\label{oneida:double 4}
	\glll yah ní· teʔwaklʌʔnhá·u a·kka·látuʔ, yah tho té·yot tsiʔ teʔwakatatlihunyʌ·n\uline{í}.	\\
	yah ní· teʔ-wak-lʌʔnhaʔ-u aa-k-kalatu-ʔ, yah tho teʔ-y-ot 	tsiʔ teʔ-wak-atat-lihunyʌni	\\
	not \textsc{1} \textsc{neg-1sg.p}-know.how-\textsc{stat} \textsc{opt-1sg.a}-tell.stories-\textsc{pnc} not thus \textsc{neg-3z/n.sg.a}-how.it.is[\textsc{stat}] \textsc{comp} \textsc{neg-1sg.p-refl}-teach[\textsc{stat}] \\
	\glt `I don't know how to tell stories, that's not the way I taught myself.' Hazel Cornelius \citep[181 (15)]{MichelsonEtAl2016}
\z

\subsection{Other aspects of negation}
\label{oneida:sec:more}

\subsubsection{Non-negative uses of negation}
\label{oneida:sec:non-neg uses}

There are two negative expressions in Oneida that have a positive import: \textit{yah teʔwé·ni} (\textit{yah teʔwé·ne} for some speakers) `it is amazing, incredible' and \textit{yah thya·ya·wʌ́· tsiʔ} `it has to be, it is necessary, must.' The first expression, in (\ref{oneida:amazing}), is  based on the verb \textit{-é·ni-/-é·ne-} `be evident', which occurs most frequently as a particle in the form \textit{wé·ni/wé·ne} `evidently, must be, I guess.' The second expression, \textit{yah thya·ya·wʌ́· tsiʔ}, in  (\ref{oneida:have to}), is based on the root \textit{-ʌ-} `happen,' which, in this meaning, occurs with the contrastive prefix (serving as the exponent of the negative feature) and translocative prefix. 

\ea\label{oneida:amazing}
    \glll Yah teʔwé·ni naʔukniká·tʌneʔ
    porridge thikʌ́ {neʔ thó·neʔ}, \\
    yah teʔ-w-eʔni n-aʔ-yukni-kaʔt-ʌʔ-neʔ
    porridge thikʌ́ {neʔ thó·neʔ}  \\
    not \textsc{neg-3z/n.sg.a}-be.evident[\textsc{stat}] \textsc{par-fact-1du.p}-have.a.lot-\textsc{inch-pnc} porridge that {at.that.time} \\
    \glt `It's incredible (lit. It's not evident) what a lot of porridge we got at that time,' Clifford Cornelius \citep[223 (37)]{MichelsonEtAl2016} %do not change the comma in the end of the translation, it is correct as it is!  > so I guess all of these ,' are OK?
\z

\ea\label{oneida:have to}
    {\glll Yah thya·ya·wʌ́· tsiʔ Muncey \\
    yah th-y-aa-yaw-ʌ-ʔ tsiʔ Muncey  \\
    not \textsc{cntr-trl-opt-3z/n.sg.p}-happen-\textsc{pnc} that Muncey  \\}
    \sn{yʌhʌ·\uline{ké·}.  \\
    yʌh-ʌ-k-e-ʔ \\
    \textsc{trl-fut-1sg.a}-walk/go-\textsc{pnc}
    }
    \glt `I have to (lit. It can't happen) go to Muncey.' Norma Kennedy \citep[92--93 (11)]{MichelsonEtAl2016}
\z

\subsubsection{A note on the distribution of the negative vs. contrastive}
\label{oneida:sec:summary}

Standard negation in Oneida is expressed with a negative particle and a prefix on the verb, either the negative or the contrastive. This structure occurs with verbs in all three verbal aspects (habitual, stative, punctual). The punctual aspect requires one of three modes -- future, factual, or optative -- and only the optative can be negated. Typically, a verb that would otherwise occur in the future mode is negated with the optative, and a verb that otherwise occurs in the factual mode is negated with the stative aspect.

It is important, for the appropriate classification of Oneida, to come back to a point that we made in \sectref{oneida:sec:standard} at the beginning of this paper -- the need to distinguish features from their exponents. The negative and contrastive prefixes both occur in the same ``slot'' or ``position class'' in the verb structure of Oneida, the word-initial slot. Prefixes that occur in the same slot cannot both have an overt realization -- complementary distribution and competition for expression are defining characteristics of a position class morphology. Since contrastive meaning is never negated, as noted in \sectref{oneida:sec:standard}, the negative and contrastive prefixes do not compete for expression -- the exponent of the contrastive simply occurs when the negative prefix cannot. There are two additional prefixes that occur in the word-initial slot, the partitive and the coincident. Here we briefly mention what happens when verb forms with the partitive or coincident are negated. When a verb that otherwise requires the partitive is negated, the negative exponent occurs at the expense of the partitive. Thus in (\ref{oneida:double 4}) \textit{yah tho té·yot} `that's not how it is' has the negative prefix \textit{teʔ-} (or \textit{té·-} under accent); the positive form \textit{tho ni·y\'ot} `that's how it is' has the partitive prefix \textit{ni-}. Likewise, the verb form \textit{té·ku}  in \textit{yah tho té·ku} `not enough` in (\ref{oneida:double 1}) includes the negative prefix when otherwise (in the positive) the same verb occurs with the partitive, \textit{tho nikú} `that's how much, that's enough.' In contrast, a verb that otherwise has the coincident keeps the coincident when negated. In the example in (\ref{oneida:ahsuhkv}) the negative particle is \textit{ahsuhkʌ́}, mentioned in \sectref{oneida:sec:standard} as one of the less common negative particles, but the verb form \textit{tshiyukwa·yʌ́·} has the exponent for the coincident feature \textit{tshi-} rather than the negative's exponent (see \citealt{DiazEtAl2019} for details). 

\ea\label{oneida:ahsuhkv}	
    \glll Kháleʔ ahsuhkʌ́ tshiyukwa·yʌ́· wí· n hydro. \\
    kháleʔ ahsuhkʌ́ tshi-yukwa-yʌ-ʔ wí· n hydro \\
    and {not.yet} \textsc{coin-1pl.p}-put/have-\textsc{stat} \textsc{ptcl} \textsc{ptcl} hydro\\
    \glt `And it was before we had hydro.' Norma Kennedy \citep[73 (5)]{MichelsonEtAl2016}
\z

\subsubsection{Diachronic and comparative notes and observations}
\label{oneida:sec:diachronic}

It is intriguing that the contrastive serves as an exponent of the negative feature. This is the case in all the Northern Iroquoian languages that have double negation (all the languages except Tuscarora, where standard negation is expressed with several different particles, without a verbal prefix). In fact, at least in Oneida, and probably in the other languages too, the contrastive prefix occurs with far greater frequency as an exponent of the negative feature than of just the contrastive feature. 

In all the languages, the contrastive prefix expounds the negative feature in the optative mood.\footnote{In Seneca, according to \citet[54]{Chafe2015}, the negative prefix, as well as the contrastive, can occur with the optative.} In Oneida (and Mohawk) the contrastive exponent occurs also with the dualic prefix; see example (\ref{oneida:cntr-dlc b}) given in \sectref{oneida:sec:standard} at the beginning of this paper. In Onondaga and Seneca, however, an allomorph \textit{t-} of the negative occurs with the dualic prefix. Thus in Oneida and Mohawk, the contrastive plus dualic is \textit{thaʔt-} but in Onondaga and Seneca it is \textit{taʔt-} (written \textit{daʔ-} in the contemporary orthography).

In all the languages except Mohawk (where the form of the negative is \textit{te-} rather than \textit{teʔ-}), outside of its co-occurrence with the optative, the exponent of the contrastive ``overlaps" with the translocative. (The term \textit{overlap} is due to \citealt[53]{Chafe2015}.) For example, in Oneida, as shown in (\ref{oneida:thye1}) and (\ref{oneida:thye2}), the contrastive plus translocative is \textit{thyeʔ-} where it is difficult to uniquely identify which portion is contrastive and which translocative. If it were not for the presence of the glottal stop in \textit{thyeʔ-}, the form would be analyzed easily as the contrastive \textit{th-} plus translocative \textit{ye-}. In Onondaga \citep[186--187]{Woodbury2018} and Seneca \citep[53]{Chafe2015} the form of the translocative is \textit{h-}, and combined with the negative there is also overlap in these languagaes: \textit{theʔ-}. It is interesting that  the translocative exponents are quite different in the two sets of languages, yet in all four languages there is overlap with the negative.

\ea\label{oneida:thye1}
    \glll Tahnú· kwí· yah thyeʔshuwahʌlú,  \\
    Tahnú· kwí· yah thyeʔ-s-huwa-hʌl-u  \\
    and \textsc{ptcl} not \textsc{neg.trl-rpt-3>3m.sg}-call/summon-\textsc{stat} \\
    \glt `And she didn't call him back,' Mercy Doxtator \citep[140 (5)]{MichelsonEtAl2016}
\z

\ea\label{oneida:thye2}
    {\glll Yah kiʔ nuwʌtú kwaʔahsuté·ke tho nú·  \\
    yah kiʔ nuwʌtú kwaʔahsuté·ke tho nú·  \\
    not actually ever nighttime that's where  \\}
    \sn{\glll  thyeʔtsyukne·n\uline{ú}. \\
     thyeʔ-ts-yukn-e-nu  \\
     \textsc{neg.trl-rpt-1du.p}-walk/go-\textsc{stat} \\
    \glt `We never went there again at night-time.' Norma Kennedy \citep[95 (39)]{MichelsonEtAl2016}}
\z

\section{Summary}

\begin{table}
\begin{tabularx}{\textwidth}{Q@{}ll}
\lsptoprule
Type of negation & Formatives & Form \\
\midrule
Standard negation & \textsc{ptcl \ldots negV} & \textit{yah \ldots teʔV} \\
% \cline{2-3}
& \textsc{ptcl \ldots contrV} & \textit{yah \ldots thV} \\
\tablevspace
Negative replies & \textsc{ptcl} & \textit{t\'ah} `no' \\ & & \textit{áhsu kiʔ} `not yet in fact' \\ & &  \textit{yáhtʌʔ} `not so'\\
\tablevspace
Negative imperatives &  \textsc{ptcl \ldots futV} & \textit{t\'akʌʔ  \ldots ʌV} \\
\tablevspace
Negative prohibitives & \textsc{ptcl \ldots pptV} & \textit{t\'akʌʔ  \ldots aaV} \\
\tablevspace
Equatives and category membership &   \textsc{ptcl \ldots negV/contrV}  & \textit{té·kʌ} \\
% \cline{2-3}
\midrule
Other stative predications & See standard negation & \\  
\lspbottomrule 
\end{tabularx}
\caption{Expressions of negation in Oneida} 
\label{oneida:tab:summary}
\end{table} 


In \tabref{oneida:tab:summary} we summarize in expressions of negation in Oneida.
As we have detailed in this chapter, Oneida's expression of negation involves both an independent word (usually the particle \textit{yah}) and a negative prefix on verbs or the word \textit{té·kʌ}. In other words, Oneida's expression of negation is an example of double clausal negation. Overall, negation in Oneida is uniform across phrases and clauses, and the relation between scope and negation is also uniform. But, the expression of negation in Oneida is remarkable for at least the following three reasons. First is the complexity of its morphological expression, including the fact that both the negative and the contrastive prefixes are exponents of the negative feature. Second are the restrictions on scopal relations that negation can enter into. In particular, it seems that the negative particle \textit{yah} must occur to the left of any operator or predicate it scopes over, and inverse scope is not allowed. The third is that Oneida includes a series of ontologically general words that are used both as negative polarity items and serve as the basis of the ``derivation'' of corresponding indefinites expressions.

\section*{Acknowledgements}
As with all of our collaborative work, the order of authors is alphabetical. We acknowledge with gratitude the late Mercy Doxtator, the late Norma Kennedy, and Olive Elm, with whom Michelson has discussed negation in Oneida (and, as it turns out, in English) and who provided some of the examples. We are also grateful to Ryan DeCaire from Wahta Nation in Ontario and Sheri Beglen from Oneida Nation in New York, who learned Mohawk and Oneida as adults, for discussions about some of the structures described in this paper.

\section*{Abbreviations}
\begin{tabularx}{.45\textwidth}{@{}lQ@{}}
1 & first person \\
2 & second person \\
3 & third person \\
\textsc{a} & agent \\
\textsc{asrt} & assertion \\
\textsc{ben} &  benefactive \\
\textsc{caus} & causative \\
\textsc{cis} & cislocative \\
\textsc{coin} & coincident \\
\textsc{comp} &  complementizer \\
\textsc{cont} & continuative \\
\textsc{cntr} & contrastive \\
\textsc{distr} & distributive\\
\textsc{dlc} & dualic (or duplicative)  \\ 
\textsc{du} & dual \\
\textsc{emph} & emphatic particle \\
\textsc{excl} & exclusive \\ 
\textsc{fact} & factual mode \\
\textsc{fi} & feminine-indefinite \\
\textsc{fut} & future \\
\textsc{fz} & feminine-zoic \\
\textsc{hab} & habitual aspect \\
\textsc{imp} & imperative \\
\textsc{inch} & inchoative \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\textsc{join} & joiner \\
\textsc{loc} & locative \\
\textsc{m} &  masculine \\
\textsc{n} & neuter \\
\textsc{neg} & negative \\
\textsc{nmlz} & nominalizer \\ 
\textsc{ns} &  noun suffix \\
\textsc{opt} & optative\\
\textsc{p} & patient \\
\textsc{par} &  partitive \\
\textsc{pl} & plural \\
\textsc{pnc} & punctual aspect \\
\textsc{poss} & possessive prefix \\
\textsc{pst} & past \\ 
\textsc{ptcl} &  particle \\
\textsc{q} & polar question particle \\
\textsc{refl} & reflexive \\
\textsc{rpt}  & repetitive \\
\textsc{sg} & singular \\
\textsc{srefl} &  semi-reflexive \\
\textsc{stat} & stative aspect \\
\textsc{trl} & translocative \\
\textsc{z/n} & (default) feminine-zoic\\
\\
\end{tabularx}


\sloppy
\largerpage
\printbibliography[heading=subbibliography,notkeyword=this]
\end{document}
