\documentclass[output=paper,colorlinks,citecolor=brown,draft]{langscibook}
\ChapterDOI{10.5281/zenodo.18234103}

\author{Bernat Bardagil\orcid{0000-0002-1190-7164}\affiliation{Ghent University}}%
\title{Negation in Panará}
\abstract{This chapter presents a detailed account of the morphosyntax of negation in Panará (Jê family, Brazil). Panará displays a wide range of strategies to express negation, with a set of negative adverbs, a negative polarity item in the polysynthetic verbal morphology, an imperative negation marker, and a negative verb. These different strategies interweave to express negation across the grammar of Panará, be it the different moods or the various types of clauses and constituents.
\keywords{negation, polysynthesis, Panará, Jê languages, Amazonian languages}
}

\IfFileExists{../localcommands.tex}{%hack to check whether this is being compiled as part of a collection or standalone
   \input{../localpackages}
   \input{../localcommands}
   \input{../localhyphenation}
    \togglepaper[23]
  }{}

\begin{document}

\maketitle

\section{The language}\label{pan:sec:PNR}

This chapter describes negation in Panará (ISO: \href{https://iso639-3.sil.org/code/kre}{kre}, Glottocode: \href{https://glottolog.org/resource/languoid/id/pana1307}{pana1307}), a Jê language spoken in southern Amazonia, in the Brazilian states of Pará and Mato Grosso. The Panará language (\textit{panärapëë}) is spoken by the Panará nation, from the autonym /panãra/ `those that are (here)'. 

The Panará people entered direct contact with neocolonial Brazilian society in 1973, as a result of the construction of highway BR-163 through their territory on the part of the Brazilian government, at the time an authoritarian military dictatorship. That led to a massive influx of settlers that turned large forested parts of Mato Grosso into areas for the production of cattle and crops such as soy and corn. The Panará lost 90\% of their population due to the spread of infectious diseases, and were forcibly relocated to the nearby Xingu Indigenous Park. In the mid-1990s they were finally able to officially demarcate a portion of their original land in which the riverine forest was still intact. The Panará population moved back to their homeland on the Iriri river basin, where they made a demographic and cultural recovery.

The Panará people today live in their own officially demarcated indigenous land, the Terra Indígena Panará, spanning half a million hectares. The current Panará population, numbering approximately 650, are almost exclusively in this area, in the villages of Kanaä, Kôtikô, Nänpôôrö, Nänsêpotiti, Sönkâräsän and Sönkwê.
In some of these villages there is also presence of Kayapó (speakers of Mẽbêngôkre; Jê, ISO : \href{https://iso639-3.sil.org/code/txu}{txu}) and Kaiabi (speakers of Kawaiwete; Tupian, ISO: \href{https://iso639-3.sil.org/code/kyz}{kyz}), through kinship connections with Panará living there. Presence of Portuguese as a second language, with different degrees of proficiency, is common in the Panará community. That is especially the case among young men who are currently in the age range between the teens to the fifties. Most women, older people and children have minimal proficiency in Portuguese. A solid passive knowledge of Mẽbêngôkre is not unusual, but very few Panará can speak it fluently.

That the Panará speak a language of the Jê family was first hypothesized by \citet{Heelas1979}, an anthropologist and the first researcher to work with the Panará. Further research by \citet{Schwartzman1988} and \citet{Dourado-Rodrigues1993} confirmed that Panará is indeed a language of the Northern Jê branch of the family \citep{Rodrigues1999}.

Panará is a polysynthetic language, in which the verb complex forms a central unit in the clause (\ref{pan:pnr-ex1}). This verb complex is made up of several proclitics, morphologically bound to the verbal root, that reference various clausal elements. These include core and peripheral participants, mood, directionality, iterativity, a causative applicative, and negation \citep{Bardagil2018}.

\ea\label{pan:pnr-ex1}
    \gll Jy=py=ra=kö=më=ra=kwyy. \\
        \Intr{}=\Iter{}=\Tpl.\Abs{}=\Com{}=\Du{}=\Fsg.\Abs{}=go \\
    \glt `The two of us went away with them.' \\
\z 

Panará case marking follows an ergative alignment, with a morphologically marked ergative argument and a morphologically unmarked absolutive argument. Panará presents two different clause-level grammatical categories corresponding to moods, realis and irrealis, and also an imperative construction. These categories are marked by modal clitics, such as intransitive realis \textit{jy=}, and even though their presence is largely obligatory, they are optionally omitted in mar\-gin\-al cases.
Panará verbal morphology presents a reality status distinction between realis and irrealis (\citealt{Dourado2001,Bardagil2015,Bardagil2018}).

The data presented throughout this paper, unless otherwise indicated, were collected by the author during fieldwork among the Panará. The language is represented using the current spelling for the language. Among the Panará, current orthographic conventions\footnote{a = /a/, â = /ə/, e = /ɛ/, ê = /e/, y = /ɨ/, o = /ɔ/, ô = /o/, u = /u/} use a diaeresis to represent nasality on vowels, and I have used this notation in the present text. 

The description of negation in Panará in this chapter is based on the questionnaire provided by the editors (\citetv{chapters/questionnaire}) and mirrors its structure.
Consequently, this paper is organized as follows. \sectref{pan:sec:clausal} provides information on clausal negation in Panará, and \sectref{pan:sec:non-clausal} deals with negation in non-clausal environments. Following these, \sectref{pan:sec:other} describes other aspects of the morphology and syntax of negation in Panará. Finally, \sectref{pan:sec:conclusion} concludes the paper and summarizes the grammar of negation in Panará.


\section{Clausal negation}\label{pan:sec:clausal}

This section reports the negation strategies used in Panará to negate clause-level predicates. 

\subsection{Standard negation}\label{pan:sec:standard}

Standard negation is the term generally used for the mechanism or mechanisms of negation that a language employs in declarative verbal main clauses \citep{Miestamo2005}. Panará has several negators that participate in standard negation: \textit{inkjown}, \textit{pjow} and \textit{rö}. There is also a verbal negation construction, \textit{jy pjow}.

As will be seen throughout the section, standard negation does not alter the syntactic properties of Panará, such as the ergative-absolutive case marking or its quite flexible verb position in the clause (\ref{pan:ex-erg1}--\ref{pan:ex-erg2}).

\ea\label{pan:ex-erg1}
    \gll Jy=ra=too inkjë. \\
    \Intr{}=\Fsg.\Abs{}=fly \Fsg{}  \\
    \glt `I went away.' \\
\z

\ea\label{pan:ex-erg2}
    \gll Jahä rê=s=anpun pakjati inkjë hë. \\
    here \Fsg.\Erg{}=\Tsg.\Abs{}=see creek \Fsg{} \Erg{}  \\
    \glt `Here I saw a creek.' \\
\z

The most frequent mechanism for standard negation is \textit{pjow} /pjoo/. It is an adverb that appears in a postverbal position or, more specifically, in a position immediately following the predicate head (\ref{pan:pjow1}--\ref{pan:pjow2}).

\ea\label{pan:pjow1}
    \gll Jy=∅=pôô pjow kri tä. \\
    \Intr{}=\Tsg.\Abs{}=arrive \Neg{} village \All{} \\
    \glt `He didn't come to the village.' \\
\z

\ea\label{pan:pjow2}
    \gll Rê=mä=issi pjow. \\
    \Fsg{}.\Erg{}=\Tsg.\Dat{}=name \Neg{} \\
    \glt `I didn't name him.' \\
\z

A second negation marker is \textit{inkjown}, often rendered as [kjowm] with a reduction of the onset and a diphthongised long vowel. \textit{Inkjown} appears in a clause-final position, taking scope over the meaning of the entire predicate (\ref{pan:inkjown1}). 

\ea\label{pan:inkjown1}
    \gll Mära hë ti=∅=pïri nänkjô {inkjown}, ti=∅=pïri ikkjyti. \\
    \Tsg{} \Erg{} \Tsg.\Erg{}=\Tsg.\Abs{}=kill peccary \Neg{}  \Tsg.\Erg{}=\Tsg.\Abs{}=kill tapir \\
    \glt `He didn't kill a peccary, he killed a tapir.' \\
\z

\textit{Inkjown} often appears in a clause that also has a postverbal negation marker, such as \textit{pjow} (\ref{pan:kjow-doble}). 

\ea\label{pan:kjow-doble}
    \gll Rê=∅=kâri pjow swa, inkjown. \\
    \Fpl.\Erg{}=\Tsg.\Abs{}=cut \Neg{} tooth \Neg{} \\
    \glt `We didn't cut its teeth, no.' \\
\z

Even though \textit{pjow} and \textit{inkjown} both appear as standard negators, they are different in two respects. First, \textit{pjow} does not stand on its own as a negative interjection, unlike \textit{inkjown} (cf. \S{}\ref{pan:sec:replies}). Second, as mentioned above, \textit{inkjown} is restricted to a clause-final position, whereas \textit{pjow} can only appear in an immediately postverbal position. Rather than a tendency, the data in my corpus shows a solid constraint governing these positional restrictions, as illustrated in (\ref{pan:pjow-kjow1}--\ref{pan:pjow-kjow2}).

\ea\label{pan:pjow-kjow1}
    \gll  Inkjë hë rê=∅=pïri  jäsy inkjown. \\
    \Fsg{} \Erg{} \Fsg.\Erg{}=\Tsg.\Abs{}=kill deer \Neg{} \\
    \glt `I didn't kill a deer.'  \\
\z

\ea\label{pan:pjow-kjow2}
    \gll  Näkää hë ti=ra=nsari {pjow} inkjë. \\
    snake \Erg{} \Tsg.\Erg{}=\Fsg.\Abs{}=bite \Neg{} \Fsg{} \\
    \glt `The snake didn't bite me.' \\
\z

The immediately postverbal position of \textit{pjow} is also a position in which other types of adverbial phrases can appear (\ref{pan:postv-adv}), which supports the hypothesis that negation in that position is also phrasal.

\ea\label{pan:postv-adv}
    \gll  Näkää hë ti=ra=nsari {pykkôômä} inkjë.\\
    snake \Erg{} \Tsg.\Erg{}=\Fsg.\Abs{}=bite morning \Fsg{}\\
    \glt `A snake bit me this morning.' \\
\z

The positional alternation of \textit{pjow} and \textit{inkjown} is summarized in (\ref{pan:pjoo-knjoo}) in relation to the position of both the verb and argumental noun phrases in the clause.

\ea\label{pan:pjoo-knjoo}
    \ea (NPs)\phantom{mint}{[}verb complex{]}\phantom{mint}\textit{pjow}\phantom{mint}(NPs)
    \ex (NPs)\phantom{mint}{[}verb complex{]}\phantom{mint}(NPs)\phantom{min}\textit{inkjown}
\z\z

A third, less common standard negation mechanism is \textit{rö}, realized as [rõ] or [nõ]. Its position in the clause mirrors that of \textit{pjoo} and other adverbials, between the verb complex and postverbal constituents. Even though \citet{Dourado2007} initially reported \textit{rö} as being a dedicated noun phrase negator, it is also used in standard negation (\ref{pan:ro1}--\ref{pan:ro2}).

\ea\label{pan:ro1}
    \gll Mämä ramä ra=pa rö inpy-ara, jy=ra=too. \\
    that \Ine{} \Tpl.\Abs{}=walk \Neg{} man-\Pl{} \Intr{}=\Tpl.\Abs{}=fly \\
    \glt `The men were not there, they had left.' \\
\z

\ea\label{pan:ro2}
    \gll Në=ri=npari rö inkj-ara, inkjown. \\
    \Fsg.\Erg{}=\Tpl.\Abs{}=hear \Neg{} woman-\Pl{} \Neg{} \\
    \glt `I didn't think about women, no.' \\
\z

Panará presents a less common standard negation strategy, namely a mood-inflected verbal \textit{pjow} that occurs following the negated clause (\ref{pan:neg-jy}). This is a different syntactic mechanism than the adverbial \textit{pjow} that we saw in earlier examples.

\ea\label{pan:neg-jy}
    \gll \ob Ka hë ka=∅=kukrë \cb{} jy=∅=pjow. \\
    \db  \Ssg{} \Erg{} \Ssg.\Erg{}=\Tsg.\Abs{}=eat \db{} \Intr{}=\Tsg.\Abs{}=\Neg{} \\
    \glt `You don't eat.' \\
\z 

In instances such as illustrated in (\ref{pan:neg-jy}), the root \textit{pjow} appears not as an adverb but as a verb. Consequently, it is accompanied by verbal morphology belonging to intransitive verbs, like the realis intransitive modal clitic \textit{jy}.

The verbal negator \textit{jy pjow} (\S\ref{pan:sec:standard}) also has an additional meaning of a state or event being finished or at an end, as in (\ref{pan:pjoo2}).

\ea\label{pan:pjoo2}
    \gll Kjëtowajï jy=∅=titi jy=∅=pjow. \\
    candle \Intr{}=\Tsg.\Abs{}=burn \Intr{}=\Tsg.\Abs{}=\Neg{} \\
    \glt `The candle went out.' \\
\z

It can be combined with roots that are typically verbal (\ref{pan:jy-verb}), or with nominal roots (\ref{pan:jy-nom}).

\ea\label{pan:jy-verb}%
    \gll Jy=ra=pëë jy=∅=pjow. \\
    \Intr{}=\Fsg.\Abs{}=speak \Intr{}=\Tsg.\Abs{}=\Neg{} \\
    \glt `I'm done speaking.'\\
\z

\ea\label{pan:jy-nom}%
    \gll Jy=∅=pjow inkô-ränkjo.  \\
    \Intr{}=\Tsg.\Abs{}=\Neg{} water-black \\
    \glt `There's no more coffee.' \\
\z

The three negation adverbs presented so far are also observed in irrealis clauses (\ref{pan:neg-irr1a}--\ref{pan:neg-irr2}).

\ea\label{pan:neg-irr1a}
    \gll Ka=ti=tä=ra=pari pjow kajapô hë. \\
    \Irr{}=\Ssg.\Irr{}=\Dir{}=\Fpl.\Abs{}=kill.\Plac{} \Neg{} Kayapó \Erg{}  \\
    \glt `The Kayapó will not kill us.' \\
\z

\ea\label{pan:neg-irr1b}
    \gll Ka=ti=pôôj kri tä inkjown. \\
    \Irr{}=\Tsg.\Irr{}=arrive.\Irr{} village \All{} \Neg{}  \\
    \glt `He won't come to the village.' \\
\z

\ea\label{pan:neg-irr2}
    \gll Inkjë ka=∅=kwy rö. \\
    \Fsg{} \Irr{}=\Fsg.\Irr{}=go \Neg{}  \\
    \glt `I won't go.' \\
\z

Besides the presence of the negative markers, negated declarative clauses are identical to standard affirmative clauses. The very free constituent order of Panará is maintained, and no additional morphological material is present besides the negative element itself. This characterizes Panará as a language with symmetrical standard negation \citep{Miestamo2005}, in which the structure of negatives coincides with the structure of affirmatives.

\subsection{Negation in non-declaratives}\label{pan:sec:non-decl}

The previous section outlined the negation mechanisms that belong to main clauses in the two moods exhibited in declaratives, realis and irrealis. This section deals with negation in non-declarative clauses, and with imperative clauses.

%\paragraph{Imperatives}
In Panará, imperatives are formed with second person, as is usual in the world's languages. Consequently, the second person absolutive clitic \textit{a} is found in intransitive clauses (\ref{pan:imper-abs}), and the second person ergative \textit{ka} in transitive ones (\ref{pan:imper-erg}). 

\ea\label{pan:imper-abs}
    \gll Py=a=kwyy.\\
    \Dir{}=\Ssg.\Abs{}=go\\
    \glt  `Go away.' \\
\z

\ea\label{pan:imper-erg}
    \gll Ka=tö=∅=krë sâti.\\
    \Ssg.\Erg{}=\Ints{}=\Tsg.\Abs{}=eat peanut\\
    \glt  `Do eat peanuts.' \\
\z
To negate an imperative clause, \textit{sä} stands in as the negative adverb (\ref{pan:imp-sa1}--\ref{pan:imp-sa3}). 

\ea\label{pan:imp-sa1}
    \gll A=sï sä. \\
    \Ssg.\Abs{}=sit \Neg{} \\
    \glt `Don't sit down.' \\
\z

\ea\label{pan:imp-sa2}
    \gll Ka=ra=jökrepajô=kwâri sä! \\
    \Ssg.\Erg{}=\Fsg.\Abs{}=throat=break \Neg{} \\
    \glt `Don't break my throat!' \\
\z

\ea\label{pan:imp-sa3}
    \gll Ka=∅=ku=krë sä sokriti.\\
    \Ssg.\Erg{}=\Tsg.\Abs{}=bite=eat \Neg{} pet \\
    \glt  `Don't eat pets.' \\
\z

Imperative negator \textit{sä} is not attested in non-imperative clauses, and speakers judge it unacceptable in those contexts. Its position in the sentence identifies it consistently as a negative adverb, as it appears in the postverbal position where both \textit{pjow} and \textit{rö} are also found in non-imperatives, preceding postverbal phrasal material, as seen in (\ref{pan:imp-sa3}).

%\paragraph{Questions}
In Panará, polar questions are introduced by the question marker \textit{a}. In content questions, the interrogative pronoun \textit{prë} is used for human referents, and \textit{pjän} or \textit{ju} for non-human referents.(\ref{pan:q-a}--\ref{pan:q-b}). 
\ea\label{pan:q-a}
    \gll A jy=a=söti inkin? \\
    \Q{} \Intr{}=\Ssg.\Abs{}=sleep good  \\
    \glt `Did you sleep well?' \\
\z

\ea\label{pan:q-b}
    \gll Pjän ka=∅=wajäri? \\
    what \Ssg.\Erg{}=\Tsg.\Abs{}=make\\
    \glt `What are you doing?' \\
\z 

When a postpositional phrase contains an interrogative pronoun, pied-piping of the postposition takes place. Questions in Panará can be negated by the three standard negators. The most common is \textit{pjow} \REF{pan:q-neg1}, with \textit{rö} being used less \REF{pan:q-neg3}. Negation of questions with \textit{inkjown} is also attested \REF{pan:q-neg2}, but also presents a lower frequency.

\ea\label{pan:q-neg1}
    \gll A ka=s=anpun pjow tititi? \\
    \Q{} \Ssg.\Erg{}=\Tsg.\Abs{}=see \Neg{} anteater  \\
    \glt `Have you not seen an anteater?' \\
\z

\ea\label{pan:q-neg2}
    \gll A ka hë ka=∅=kuri impo jï inkjown? \\
    \Q{} \Ssg{} \Erg{} \Ssg.\Erg{}=\Tsg.\Abs{}=eat  cow meat \Neg{}  \\
    \glt `You don't eat beef?' \\
\z

\ea\label{pan:q-neg3}
    \gll Pjän rahê ka=tö=∅=pyri rö? \\
    what \Caus{} \Ssg.\Erg{}=\Emph{}=\Tsg.\Abs{}=take \Neg{}  \\
    \glt `Why didn't you take it?' \\
\z

\subsection{Negation in stative predications}\label{pan:sec:stative}

Panará presents two broad syntactic mechanisms that correspond to stative predication. In most instances they involve a non-verbal predicate. However, attribution is expressed by means of a verbal predicate, and existential predication is a mixed bag. In this section I introduce each of these types of stative predication and discuss the corresponding negation strategies.

In Panará, {equation} and inclusion are expressed paratactically, with the argument usually preceding the non-verbal predicate (\ref{pan:eq-1}--\ref{pan:eq-2}).

\ea\label{pan:eq-1}
    \gll Ja inkjë jö issê. \\
    this \Fsg{} \Poss{} bow \\
    \glt `This is my bow.' \\
\z

\ea\label{pan:eq-2}
    \gll Mära toopytun. \\
    \Tsg{}  old.man \\
    \glt `He is a chief/an old man.' \\
\z

When negated, equative predicates use the negative markers \textit{pjow} (\ref{pan:eq-1-neg-a}) and \textit{rö} (\ref{pan:eq-1-neg-b}).

\ea\label{pan:eq-1-neg-a}
    \gll Ja inkjë jö issê pjow. \\
    this \Fsg{} \Poss{} bow \Neg{} \\
    \glt `This is not my bow.' \\
\z

\ea\label{pan:eq-1-neg-b}
    \gll Pârinto sö rö. \\
    shellfish food \Neg{} \\
    \glt `Shellfish are not food.' \\
\z

Locative predication follows the same general lines of Panará non-verbal predication. The one argument is followed by a locative phrase, headed by a postposition (\ref{pan:loc-1}--\ref{pan:loc-2}).

\ea\label{pan:loc-1}
    \gll Kwy puu jamä. \\
     manioc field \Ade{} \\
    \glt `There is manioc at the field.' \\
\z

\ea\label{pan:loc-2}
    \gll Mösy pjow kan ramä. \\
     corn \Neg{} basket \Ine{} \\
    \glt `There is no corn in the basket.' \\
\z

It is possible that Panará locatives would be better analyzed as existential clauses with locative adjuncts \citep[8--10]{OverallEtAl2018}. If this is indeed the case, (\ref{pan:loc-2}) would present an instance of predicate-final negation with \textit{pjow}, consistent with the distribution of \textit{pjow} seen thus far. The existential structure of locative clauses notwithstanding, there is no dedicated verbal morphology associated with it.

Attribution stands apart from other types of stative predication in that the predicate head projects a verbal complex, rather than using the juxtaposition mechanics of non-verbal predication. Attributive predicates are fully verbal, on par with other types of intransitive verbal predicates, as can be seen from the verbal morphology present in (\ref{pan:att-1}--\ref{pan:att-2}).

\ea\label{pan:att-1}
    \gll Paa, jy=ra=kjä-pepeti. \\
    yes \Intr{}=\Fsg.\Abs{}=head-soft \\
    \glt `Yes, I'm smart [soft-headed].' \\
\z

\ea\label{pan:att-2}
    \gll Mära jy=∅=toopytun. \\
    \Tsg{} \Intr{}=\Tsg.\Abs{}=old.man \\
    \glt `He is (an) old (man).' \\
\z

Contrast (\ref{pan:att-2}) with the equative, non-verbal counterpart in (\ref{pan:eq-2}). Incidentally, (\ref{pan:att-2}) cannot mean `he is a chief', suggesting that certain lexicalization processes can affect the nominal form but not the verbal form of a shared root.

The negation of attributive predicates is akin to standard negation of intransitives (\ref{pan:att-neg1}--\ref{pan:att-neg2}), as seen in the previous section (\S\ref{pan:sec:standard}).

\ea\label{pan:att-neg1}
    \gll Jy=ra=kjä-tâti pjow. \\
    \Intr{}=\Fsg.\Abs{}=head-hard \Neg{} \\
    \glt `I'm not dumb [hard-headed].' \\
\z

\ea\label{pan:att-neg2}
    \gll Mära jy=∅=toopytun rö. \\
    \Tsg{} \Intr{}=\Tsg.\Abs{}=old.man \Neg{} \\
    \glt `He is not (an) old (man).' \\
\z

Possession and belonging more generally also fall into the category of non-verbal predication. Here Panará presents a typical distinction between alienable and inalienable possession. While inalienable possession is expressed by juxtaposition, with the possessor preceding the possessum (\ref{pan:inalien}), alienable possession \citep{NicholsBickel2013-wals59} is expressed adpositionally. The possessor is encoded as a possessive postpositional phrase, headed by \textit{jö}, and is followed by the possessum (\ref{pan:alien}). It is very common for possession to be negated with \textit{rö} (\ref{pan:inalien-neg}), but \textit{pjow} is also attested (\ref{pan:alien-neg}).

\ea\label{pan:inalien}
    \gll Mära inpimpjâ. \\
    \Tsg{} husband  \\
    \glt `She has a husband.' or `her husband' \\
\z

\ea\label{pan:inalien-neg}
    \gll Mära inpimpjâ rö. \\
    \Tsg{} husband \Neg{} \\
    \glt `She is single.' (lit. `She has no husband') or `(It is) not her husband.'  \\ 
\z

\ea\label{pan:alien}
    \gll Inkjë jö issê. \\
    \Fsg{} \Poss{} bow  \\
    \glt `I have a bow.' or `my bow' \\
\z

\ea\label{pan:alien-neg}
    \gll Inkjë jö issê pjow. \\
    \Fsg{} \Poss{} bow \Neg{}  \\
    \glt `I don't have a bow.' or `(It is) not my bow.' \\
\z

In verbal possessive constructions the lexical root for the possessum appears as the verb in the clause, rather than as an argument, and possession is indicated as a proclitic morpheme (\ref{pan:verb-poss}). This is consistent with the observation by \citet[76]{Dourado2001} that, in these constructions, negative adverb \textit{pjow} appears after the possessum, and not after the morpheme \textit{sö} indicating possession in the verbal complex (\ref{pan:verb-poss-neg}).

\ea\label{pan:verb-poss}
    \gll Kjonpe jy=sö=kukre. \\
    Kjonpe \Intr{}=\Poss{}=house \\
    \glt `Kjonpe has a house.' \\
\z

\ea\label{pan:verb-poss-neg}
    \gll Kjonpe jy=sö=kukre pjow. \\
    Kjonpe \Intr{}=\Poss{}=house \Neg{} \\
    \glt `Kjonpe doesn't have a house.' \\
\z

The verb \textit{pan} `walk' is often used to indicate both existential and locative constructions, with a preference for animate arguments \REF{pan:pan-1a}. Although it exhibits the verbal morphology associated to intransitive verbs, it often appears without it \REF{pan:pan-1b}.\footnote{Having unrealized proclitics on the verb complex is not frequent, but not uncommon in Panará verbal morphology.} 

\ea\label{pan:pan-1a}
    \gll Joopy ra=pan. \\
    jaguar \Tpl.\Abs{}=walk \\
    \glt `There are jaguars.' \\
\z

\ea\label{pan:pan-1b}
    \gll Ju rï ka pan? \\
    what \Loc{} \Ssg{} walk \\
    \glt `Where are you?' \\
\z

Existential and locative use of \textit{pan} can be negated using either \textit{rö} or \textit{pjow} (\ref{pan:pan-n1}). Once again, it is unclear whether \textit{inkjown} is also licensed in these contexts besides as the negative reply, as in (\ref{pan:pan-n2}).

\ea\label{pan:pan-n1}
    \gll Ippë pan rö. \\
    non-indigenous walk \Neg{} \\
    \glt `There were no white people.' \\
\z

\ea\label{pan:pan-n2}
    \gll Inkjown, Kôtikô rï ra=pan pjow. \\
    no Kôtikô \Loc{} \Fsg.\Abs{}=walk \Neg{} \\
    \glt `No, I'm not in Kôtikô.' \\
\z

\textit{Inkjown} is present in most, if not all classes of stative predicates. It appears in its usual clause-final position (\ref{pan:stat-ink1}--\ref{pan:stat-ink2}), but with non-verbal predicates it can also appear in an initial position (\ref{pan:stat-ink-ini1}--\ref{pan:stat-ink-ini2}). This position for \textit{inkjown} is not available in verbal clausal contexts. Its distribution, the fact that it shows a broad scope over the entire predication, and its use as a standalone negative reply make \textit{inkjown} a solid candidate for a negative existential \citep{Veselinova2013neg.exist}.

\ea\label{pan:stat-ink1}  
    \gll {``}Ka=∅=kân=∅=sũn ja: ka=ra=nĩri pjow. Inkjara inkjown,'' ti=∅=jãri. \\
    {\db}\Irr{}=\Fsg.\Irr{}=\Ssg.\Dat{}=\Fsg.\Irr{}=tell this \Ssg.\Erg{}=\Tpl.\Abs{}=have.sex \Neg{} woman-\Pl{} \Neg{} \Tsg.\Erg{}=\Tsg.\Abs{}=say  \\
    \glt {`}{``}I will say this to you: don't have sex. No women,'' he said.' \\
\z

\ea\label{pan:stat-ink2}
    \gll Sõ inkjown. \\
    food \Neg{} \\
    \glt `[There is] no food,' or `It's not food.' \\
\z

\ea\label{pan:stat-ink-ini1}
    \gll Inkjown pakwa kjokjo. \\
    \Neg{} banana ripe \\
    \glt `[There are] no ripe bananas.' \\
\z

\ea\label{pan:stat-ink-ini2}
    \gll Inkjown kjorinpe. \\
    \Neg{} rice \\
    \glt `[There is] no rice.' \\
\z

Both \textit{pjow} and \textit{rö} can appear together with \textit{inkjown}. In such cases, \textit{inkjown} should be seen in its negative reply function (\S\ref{pan:sec:replies}) rather than as clause-final negation (\ref{pan:exneg-3}).

\ea\label{pan:exneg-3}
\gll Atö pjow, inkjown. \\
 gun \Neg{} no \\
\glt `There is no gun, no.' \\
\z

Panará uses two syntactic mechanisms to form the different classes of stative predicates: non-verbal predication and verbal predication. The three standard negation mechanisms are also employed in stative predicates. While \textit{pjow} is the negation marker that can occur with all subtypes of stative predication, \textit{rö} is also employed. However, these two negative adverbs are not completely interchangeable -- especially so in the case of existential and possessive predicates, where their semantics are quite different. \textit{Inkjown} is also a negative marker in non-verbal contexts, but it has a strong tendency to appear clause-initially instead of its clause-final position in main clause negation.

\subsection{Negation in non-main clauses}\label{pan:sec:non-main}

 In the previous section we saw the three syntactic mechanisms used in Panará for main clause negation: postverbal adverbs \textit{pjow} and \textit{rö}, and clause-final \textit{inkjown}, besides a dedicated imperative negator \textit{sä}.

The following examples show instances of relative and complement clauses negated using \textit{pjow} (\ref{RCneg-pjow1}--\ref{RCneg-pjow2}), \textit{rö} (\ref{RCneg-ro}) and \textit{inkjown} (\ref{RCneg-inkjown}).

\ea\label{RCneg-pjow1}
    \gll Kuupêri hë ti=∅=sün inkin pjow inkô-ränkjo. \\
    Kuupêri \Erg{} \Tsg.\Erg{}=\Tsg.\Abs{}=say good \Neg{} water-black  \\
    \glt `Kuupêri said that he doesn't like coffee.' \\
\z

\ea\label{RCneg-pjow2}
    \gll Patty hë ti=∅=pïra pjow jäsy rê=∅=pïri. \\
    Patty \Erg{} \Tsg.\Erg{}=\Tsg.\Abs{}=kill.\Rel{} \Neg{} deer \Fsg.\Erg{}=\Tsg.\Abs{}=kill   \\
    \glt `I killed the deer that Patty didn't kill.' \\
\z

\ea\label{RCneg-ro}
    \gll Mära hë ti=∅=pyri inkö s=aswâri rö amä. \\
    \Tsg{} \Erg{} \Tsg.\Erg{}=\Tsg.\Abs{}=take water \Tsg.\Abs{}=spill \Neg{} \Ine{} \\
    \glt `She brought the water without spilling it.' \\
\z

\ea\label{RCneg-inkjown}
    \gll Inpy hë ti=ra=sun rê=tö=∅=pyri pjow atö-sy. \\
    man \Erg{} \Tsg.\Erg{}=\Fsg.\Abs{}=say \Fsg.\Erg{}=\textsc{emph}=\Tsg.\Abs{}=take \Neg{} gun-grain \\
    \glt `The man said to me that I didn't bring any ammunition.' \\ %cf Dourado 393
\z
Just like in main clause negation, \textit{rö} is less frequent than \textit{pjow} but nonetheless attested. The example in (\ref{pan:kwyy-ro}) shows a final construction in which negation is indicated with \textit{rö} in both the main clause and the subordinate clause.

 
\ea\label{pan:kwyy-ro}
    \gll Ka hë ka=kjë=∅=sun rö ra=kwyy rö ahê. \\
    \Ssg{} \Erg{} \Ssg.\Erg{}=\Fsg.\Dat{}=\Tsg.\Abs{}=say \Neg{} \Fsg.\Abs{}=go \Neg{} \Fnl{}  \\
    \glt `You didn't tell me not to go.' \\
\z

Besides main clauses and phrasal constituents, the verbal negator \textit{jy pjow} can also negate non-main clauses, such as the temporal clause in (\ref{pan:temp-neg}).

\ea\label{pan:temp-neg}
    \gll Inkjë-mëra në=∅=kuri mï sö jy=∅=pjow rahä.  \\ 
    \Fsg-\Pl{} \Fpl.\Erg{}=\Tsg.\Abs{}=eat caiman food \Intr{}=\Tsg.\Abs{}=\Neg{} \Ade{} \\
    \glt `We eat caiman when we run out of food.'\\
\z

In the case of adpositional subordinate clauses, the addition of a negative marker to a final or purpose clause can result in the equivalent of something like `in order not to' in English. An example of a negative final subordinate clause is given in (\ref{pan:final-neg}).

\ea\label{pan:final-neg}
    \gll Ka=mä=∅=sun jy=∅=tëë rö ahê. \\
    \Ssg.\Erg{}=\Tsg.\Dat{}=\Tsg.\Abs{}=say \Intr{}=\Tsg.\Abs{}=leave \Neg{} \Fnl{}{} \\
    \glt `You told him not to leave.' \\
\z

\section{Non-clausal negation}\label{pan:sec:non-clausal}

This section presents the ways in which Panará exhibits non-clausal negation. The same negation markers seen in clausal negation, \textit{pjow}, \textit{rö} and \textit{inkjown}, are also found in non-clausal contexts.

\subsection{Negative replies}\label{pan:sec:replies}

In Panará, polar questions can be answered by the use of the corresponding affirmative or negative word. The affirmative reply, conveying a positive polarity answer to the question \citep{Moser2018, EnfieldEtAl2019}, is the word \textit{paa}$\sim$\textit{ngaa}$\sim$\textit{haa} (\ref{pan:paa}). The affirmative reply can also stand on its own, so that (\ref{pan:paa}A) can be just \textit{paa}. 
%     \hc{@LSP check the last reference to ex 63A, did you find a solution for labels within xlist?}

\ea\label{pan:paa}
    \begin{xlist}[Q: ]
    \exi{Q: }
       \gll A jy=a=inkin ka? \\
       \Q{} \Intr{}=\Ssg.\Abs{}=good \Ssg{}\\
       \glt `Are you good?'\\
    \exi{A: }\label{pan:paab}
        \gll Paa, inkin pytinsi.\\
        yes good very  \\
        \glt `Yes, very good.'\\
    \end{xlist}
\z

The negative counterpart \textit{inkjown} conveys a negative polarity answer (\ref{pan:no}), and coincides with a root that we already saw functioning as a standard negator in \sectref{pan:sec:standard}.

\ea\label{pan:no}
    \begin{xlist}[Q: ]
    \exi{Q: }
       \gll Ju rï pan ka? Guarantã?\\
        \Q{} \Loc{} walk \Ssg{} Guarantã \\
        \glt `Where are you? In Guarantã?'  \\
    \exi{A: }
        \gll Inkjown, Casai Peixoto rï ra=pan.\\
         no house Peixoto \Loc{} \Fsg.\Abs{}=walk \\
        \glt `No, I'm at the healthcare house\footnote{CASAI: Casa da Saúde Indígena (indigenous healthcare house). These are government houses that serve as residence for indigenous people undergoing medical treatment in the city, and where basic healthcare is administered.} in Peixoto.'  \\
    \end{xlist}
\z

It is not uncommon for the negative reply to appear following a negative clause, after a prosodic break (\ref{pan:neg-break1}--\ref{pan:neg-break2}).

\ea\label{pan:neg-break1}
    \gll Ti=ra=nsari pjow. Inkjown. \\
    \Tsg.\Erg{}=\Fsg.\Abs{}=bite \Neg{} no \\
    \glt `It did not bite me. No.' \\
\z

\ea\label{pan:neg-break2}
    \gll Në=ri=npari rö inkj-ara, inkjown. \\
    \Fsg.\Erg{}=\Tpl.\Abs{}=hear \Neg{} woman-\Pl{} \Neg{} \\
    \glt `I didn't think about women, no.' \\
\z

\subsection{Negative indefinites and quantifiers}\label{pan:sec:indef-quant}

Panará does not have lexicalized negative indefinites. The meaning of roots such as `nobody' in English is usually conveyed in Panará by saying that something was not done or did not take place, as in \REF{pan:addition}.
In some cases, the interrogative pronoun \textit{prë} appears negated with a meaning equivalent to `nobody.' This is shown in the following examples (\ref{pan:67}--\ref{pan:68}).

\ea\label{pan:addition}
    \begin{xlist}[Q: ]
    \exi{Q: }
       \gll Prë jy=∅=tëë tepi suu?\\
        \Q{} who \Intr{}=\Tsg.\Abs{}=leave fish \\
        \glt `Who went fishing?'  \\
    \exi{A: }
        \gll Jy=∅=tëë pjow.\\
         \Intr{}=\Tsg.\Abs{}=leave \Neg{}  \\
        \glt `(S)he didn't go (intended: Nobody went.)'  \\
    \end{xlist}
\z

\ea\label{pan:67}
    \gll Jy=∅=too pjow prë. \\
    \Intr{}=\Tsg.\Abs{}=fly \Neg{} who\\
    \glt `Nobody went away.' \\
\z
 
\ea\label{pan:68}
    \gll Prë hë pjow ti=∅=inkô=kujä. \\
    who \Erg{} \Neg{} \Tsg.\Erg{}=\Tsg.\Abs{}=water=carry \\
    \glt `Nobody carried water.' \\
\z

The equivalent of `nothing' is also not lexicalized in Panará. Usually it is expressed via a negation marker, such as \textit{pjow} or \textit{rö} \REF{pan:69}, and usually also the reinforcer morpheme \textit{tö} \REF{pan:70}.

\ea\label{pan:69}
    \gll Poswa jy=mä=sâpêri toopytun mä, ti=ho=∅=pjy pjow. \\
    Poswa \Intr{}=\Dat{}=work old.man \Dat{} \Tsg.\Erg{}=\Ins{}=\Tsg.\Abs{}=get \Neg{}  \\
    \glt `Poswa worked for the chief and got nothing.' \\
\z

\ea\label{pan:70}
    \gll Mära hë ti=tö=∅=swâri rö. \\
    \Tsg{} \Erg{} \Tsg.\Erg{}=\textsc{emph}=\Tsg.\Abs{}=to.fish \Neg{}  \\
    \glt `He didn't fish anything.' \\
\z

\subsection{Negative derivation and case marking}\label{pan:sec:deriv}

Panará lacks an abessive or an equivalent grammatical case to express absence or lack of something. Instead, with nouns the mechanisms used in standard negation are also used to negate existential predication (\ref{pan:absence1}--\ref{pan:absence2}), as described in \sectref{pan:sec:stative}.

\ea\label{pan:absence1}
    \gll Inkjown tepi. \\
    \Neg{} fish  \\
    \glt `There is no fish.' \\
\z

\ea\label{pan:absence2}
    \gll Mära ippëë rö. \\
    \Tsg{} speech \Neg{}  \\
    \glt `He is mute.' \\
\z

Absence or lacking of something is usually expressed with the negator \textit{rö} (see \sectref{pan:sec:scope}). Adjectives are negated using also standard negation mechanisms (\S\ref{pan:sec:standard}), such as \textit{pjow} (\ref{pan:asa-pjow}).

\ea\label{pan:asa-pjow}
    \gll Asâ pjow topjâpjâ Jakjô. \\
    fierce \Neg{} grandfather Jakjô  \\
    \glt `Grandfather Jakjô was not grumpy.' \\
\z



\section{Other aspects of negation}\label{pan:sec:other}

\subsection{The scope of negation}\label{pan:sec:scope}

Two major crosslinguistic patterns have been observed in the interaction of negation and disjunction \citep{Szabolcsi2004}: negation taking scope over disjunction, such as in English (\ref{pan:disj1}), and disjunction taking scope over negation, as in Hungarian (\ref{pan:disj2}).

\ea\label{pan:disj1}English \citep[219]{Szabolcsi-Haddican2004}
    \gll {Mary didn’t take hockey or algebra.} \\
    {`Mary took neither hockey nor algebra.'} \\
\z

\ea\label{pan:disj2}Hungarian \citep[219]{Szabolcsi-Haddican2004}\\
    \gll Mari nem járt hokira vagy algebrára. \\
    Mari not went hockey-to or algebra-to \\
    \glt `Mary didn’t take hockey or didn’t take algebra.' \\ 
\z
The way in which disjunction is realized syntactically in Panará is via juxtaposition of the disjuncts (\ref{pan:disj}). This contrasts with conjunction, where each conjunct is additionally marked with \textit{më} (\ref{pan:conj}).

\ea\label{pan:disj}
    \gll Perankô hë ti=∅=pïri ikkjyti  jäsy. \\
    Perankô \Erg{} \Tsg.\Erg{}=\Tsg.\Abs{}=kill tapir  deer  \\
    \glt `Perankô killed a tapir or a deer.' \\
\z

\ea\label{pan:conj}
    \gll Perankô hẽ ti=∅=pïri ikkjyti më jäsy më. \\
    Perankô \Erg{} \Tsg.\Erg{}=\Tsg.\Abs{}=kill tapir and deer and \\
    \glt `Perankô killed a tapir and a deer.' \\
\z
The meaning that emerges in contexts where disjunction is negated, such as in (\ref{pan:disj-neg}), suggests that Panará patterns like English in this aspect and that negation takes scope over disjunction.

\ea\label{pan:disj-neg}
    \gll Ka hë ka=∅=kuri {pjow} kôôtita jï inpo jï. \\
    \Ssg{} \Erg{} \Tsg.\Erg{}=\Tsg.\Abs{}=eat \Neg{} chicken meat cow meat \\
    \glt `You won't eat chicken meat and you won't eat cow meat.'\\
\z

It is worth noting that the negation markers \textit{pjow} and \textit{rö} exhibit a difference in the scope of negation when interacting with the existential and possessive constructions (\S\ref{pan:sec:stative}). While \textit{pjow} still expresses predicate-level negation (\ref{pan:pjow-ex}), \textit{rö} carries instead a reading of absence (\ref{pan:ro-ex}).

\ea\label{pan:pjow-ex}
    \ea
    \gll Mära inpimpjâ pjow. \\
        \Tsg{} husband \Neg{}  \\
        \glt `She has no husband.' \\
    \ex
        \gll Kukre pjow. \\
        house \Neg{} \\
        \glt `(There are) no houses.' \\
    \z
\z

\ea\label{pan:ro-ex}
    \ea
        \gll Mära inpimpjâ rö. \\
        \Tsg{} husband \Neg{}  \\
        \glt `She is husband-less,' or `She is single.' \\
    \ex
        \gll Kukre rö. \\
        house \Neg{} \\
        \glt `Without a house.' \\
    \z
\z

\subsection{Negative polarity}\label{pan:sec:polarity}

Even though some elements, such as reinforcer \textit{tö} (\sectref{pan:sec:reinforcing}), can combine their semantics with negation, no proper negative polarity items have been found so far in Panará.

\subsection{Marking of NPs in the scope of negation}\label{pan:sec:NP-marking}

In Panará, no morphological effect is observed on noun phrases that occur under the scope of negation. There is no dedicated negative determiner, just as there is no determiner more generally, and case marking remains the same as otherwise (\ref{pan:case-negA}--\ref{pan:case-negB}).

\ea\label{pan:case-negA}
    \gll Joopy hë ti=∅=pïri kôôtita. \\
    jaguar \Erg{} \Tsg.\Erg{}=\Tsg.\Abs{}=kill chicken  \\
    \glt `The jaguar killed a chicken.' \\
\z

\ea\label{pan:case-negB}
    \gll Joopy hë ti=∅=pïri pjow kôôtita. \\
    jaguar \Erg{} \Tsg.\Erg{}=\Tsg.\Abs{}=kill \Neg{} chicken  \\
    \glt `The jaguar didn't kill a chicken.' \\
\z

\subsection{Reinforcing negation}\label{pan:sec:reinforcing}

In Panará, negation can be signaled or stressed multiple times in order to achieve an effect of reinforcing a negative meaning. One very common strategy is the use of the verbal proclitic \textit{tö}. In its use in positive polarity contexts, \textit{tö} functions as a modal particle with an epistemic or mirative component. Its meaning relates to the attitude of interlocutors towards the predicate, of stressing or insisting on, as exemplified in (\ref{pan:to-pos}--\ref{pan:to-ref}).

\ea\label{pan:to-pos}
    \gll Rê=tö=∅=py kâjasâ. \\
    \Fsg.\Erg{}=\textsc{emph}=\Tsg.\Abs{}=take machete  \\
    \glt `I did take a machete.' \\
\z

\ea\label{pan:to-ref}
    \gll Ka=tö=∅=krë sâti! \\
    \Ssg.\Erg{}=\textsc{emph}=\Tsg.\Abs{}=eat peanut  \\
    \glt `Do eat peanuts!' \\
\z


In negative polarity contexts, \textit{tö} references negation in the verbal complex and contributes an emphatic element to the negative meaning (\ref{pan:npi-1}--\ref{pan:npi-2}).

\ea\label{pan:npi-1}
    \ea
        \gll Rê=k=anpun pjow inkjë hë ka. \\
        \Fsg.\Erg{}=\Ssg.\Abs{}=see \Neg{} \Fsg{} \Erg{} \Ssg{}   \\
        \glt `I didn't see you.' \\
    \ex
        \gll Rê=a=tö=npun pjow inkjë hë ka. \\
        \Fsg.\Erg{}=\Ssg.\Abs{}=\textsc{emph}=see \Neg{} \Fsg{} \Erg{} \Ssg{}   \\
        \glt `I never saw you.' \\
    \z
\z

\ea\label{pan:npi-2}
    \gll Rê=näkää=tö=pïri pjow inkjë hë. \\
    \Fsg.\Erg{}=snake=\textsc{emph}=kill \Neg{} \Fsg{} \Erg{} \\
    \glt `I absolutely did not kill a snake.' \\
\z

As several examples throughout this chapter show, it is not uncommon for negation to be indicated by more than one element in a sentence. The context to example (\ref{pan:neg-stress}) is that the consultant is recounting that an old chief called Jakjô was very \textit{swakjörä}: he advised and guided other people. Jakjô would say `no' to something ill-advised, and the person would absolutely not do it. This is conveyed in the second sentence of the example, which indicates negation with postverbal \textit{pjow}, the negative reinforcer \textit{tö} in the verbal complex, and clause-final \textit{inkjown}. 

\ea\label{pan:neg-stress} 
    \gll ``Inkjown,'' ti=∅=järi. Ti=ra=tö=kwajäri pjow, inkjown. \\ 
    {\db}no \Tsg.\Erg{}=\Tpl.\Abs{}=say \Tsg.\Erg{}=\Tpl.\Abs{}=\textsc{emph}=make \Neg{} \Neg{}  \\
    \glt {`}``No,'' [Jakjô] said. He [someone else] did not do it, no.' \\
\z


The adverb \textit{pytinsi} `very' can be used as another mechanism to reinforce negation (\ref{pan:pjow-pyt1}). It is also used with \textit{jy pjow} (\ref{pan:pjow-pyt2a}).

\ea\label{pan:pjow-pyt1}
    \gll Rê=ri=npari rö pytinsi. \\
    \Tpl.\Erg{}=\Tpl.\Abs{}=hear \Neg{} very \\
    \glt `They didn't hear them at all.' \\
\z

\ea\label{pan:pjow-pyt2a}
    \gll Inkjë-mëra jö kyypa to jy=∅=pjow pytinsi. \\
    \Fsg-\Pl{} \Poss{} land really \Intr{}=\Tsg.\Abs{}=\Neg{} very  \\
    \glt `Our land is very much gone.' \\
\z

\subsection{Negation, coordination and complex clauses}\label{pan:sec:coord}

Panará does not exhibit dedicated negative coordinators or subordinators. The equivalent of a `neither\dots{} nor' sentence in Panará is expressed by a standard negation element such as \textit{pjow} and the usual symmetric conjunction with \textit{më} mentioned in \sectref{pan:sec:scope}.


\ea\label{pan:pjow-pyt2b}
    \gll Rê=s=anpun pjow Kuupêri më Kypakjä më. \\
     \Fsg.\Erg{}=\Tsg.\Abs{}=see \Neg{} Kuupêri and Kypakjã and \\ 
    \glt `I saw neither Kuupêri nor Kypakjã.' \\
\z

\subsection{Miscellaneous aspects of negation}\label{pan:sec:other-misc}
As mentioned by \citet{Salanova2007} and \citet{Alves2010}, in other Jê languages, such as Timbira,  Mẽbêngôkre (\ref{pan:negMBG}) or Kĩsêdjê (\ref{pan:negKSJ}), negation consists of a negative verb that selects the negated element as its dependent, and which appears linearly after the negated clause or phrase \citep{Salanova:handbook, Alves-Gildea2020}. 

\ea\label{pan:negMBG}%
    \gll \ob Aje ∅=krẽn \cb{} kêt. \\
    \db \Ssg.\Erg{} \Tsg.\Abs{}=eat.\Nml{} \db{} \Neg{} \\
    \glt `You don't eat.' \\
\z

\ea\label{pan:negKSJ}%
    \gll \ob Kare ∅=khuru \cb{} khêrê. \\
    \db \Ssg.\Erg{} \Tsg.\Abs{}=eat.\Nml{} \phantom{1} \Neg{} \\
    \glt `You didn't eat.' \citep[7]{Nonato2014}\\ 
\z

Panará \textit{jy pjow } (\ref{pan:neg-pnr-ket}), discussed in \sectref{pan:sec:standard}, could be considered truly analogous to Jê negation by means of a negative verb that selects the negated clause.

\ea\label{pan:neg-pnr-ket}%
    \gll \ob Ka hë ka=∅=ku=krë \cb{} jy=∅=pjow.  \\
    \db  \Ssg{} \Erg{} \Ssg.\Erg{}=\Tsg.\Abs{}=chew=eat \db{} \Intr{}=\Tsg.\Abs{}=\Neg{} \\
    \glt `You don't eat.' \\
\z

\section{Summary}\label{pan:sec:conclusion}

Panará presents itself as a language with a robust morphosyntax of negation. Several lexical roots and several syntactic mechanisms are used to express negation, ranging from adverbs to nouns and verbs. These are summarized in \tabref{pan:pnr-table}. 

% \begin{table}
% \centering
%   \begin{tabular}{ l l l l }
%   \lsptoprule
% \multicolumn{2}{l}{\textbf{Clausal negation}} &                 &                                                     \\
%     Standard        & \textit{pjow}             & (NP) V + \textit{pjow} + (NP)                         &           \\
%     negation        & \textit{inkjown}          & (NP) V  + (NP) + \textit{inkjown}                     &           \\
%                     & \textit{rö}               & (NP) V + \textit{rö} + (NP)                           &           \\
%                     & \textit{jy pjow}          & a. clause + \textit{jy=∅=pjow}                        &           \\
%                     &                           & b.  \textit{jy=∅=pjow} + clause                       &           \\
%     Imperative      & \textit{sä}               & (NP) V + \textit{sä} + (NP)                            &           \\
%     Interrogative   & \textit{pjow, rö}         & \textsc{q} + (NP) V + \textit{pjow}/\textit{rö} + (NP) &           \\
%                     & \textit{inkjown}          & \textsc{q} + (NP) V + (NP) + \textit{inkjown}          &           \\
%     Stative         & \textit{pjow, rö}         & stative pred. + \textit{pjow}/\textit{rö}              &           \\
%                     &                           & NP + \textit{pjow}/\textit{rö} + locative pred.        &           \\
%                     &                           & (NP) + stative V + \textit{pjow/rö} + (NP)             &           \\
%                     &                           & possessive pred. + \textit{rö}/\textit{pjow}           &           \\
%                     &                           & possessive V + \textit{pjow}                           &           \\
%                     & \textit{inkjown}          & stative pred. + \textit{inkjown}                       &           \\
%                     &                           & \textit{inkjown} + existential pred.                   &           \\
%     Non-main        & \textit{pjow}             & (NP) V + \textit{pjow} + (NP)                          &           \\
%     clauses         & \textit{inkjown}          & (NP) V  + (NP) + \textit{inkjown}                      &           \\
%                     & \textit{rö}               & (NP) V + \textit{rö} + (NP)                            &           \\
% \midrule
% \multicolumn{2}{l}{\textbf{Non-clausal negation}}   &                                                   & \textbf{} \\
%     Replies         & Affirmative: \textit{paa/haa/ngaa}  & \textit{paa}/\textit{haa}/\textit{ngaa} + (XP / clause)   &           \\
%                     & Negative: \textit{inkjown}          & \textit{inkjown} + (XP / clause)                          & \\
%     Indefinites     & interr. pron. + \Neg{}              & a. nobody: \textit{prë} \dots{} \textit{pjow}             & \\
%                     &                                     & b. nothing: \textit{pjän} \dots{} \textit{pjow}           & \\
%     Adjectives      & \textit{pjow, rö}                   & NP + \textit{pjow/rö}                         & \\
%     Abessive        & \textit{rö} `lack of'             & NP + \textit{rö}                              & \\
% \midrule
% \textbf{Other}      &                           &                                                            & \textbf{} \\
%     Scope           & negation > disjunction              & (\Neg{}) + NP + NP + (\Neg{})                 & \\
%     Reinforcement   & \textit{tö}                         & (NP) + \textit{tö}=V   +\Neg{} + (NP)         & \\

% \lspbottomrule
% \end{tabular}
% \caption{Negation in Panará [OLD TABLE]}\label{pan:pnr-table}
% \end{table}

\begin{table}%HC this is the new table
\begin{tabular}{lll}
\lsptoprule
%\multicolumn{3}{l}{\textbf{Clausal negation}} \\
    \textit{pjow}       & declarative               & (NP) + V + \textit{pjow} + (NP) \\
   
                        & interrogative             & \textsc{q} + (NP) V + \textit{pjow}/\textit{rö} + (NP) \\
                        & stative                   & stative pred. + \textit{pjow} \\
                        &                           & NP + \textit{pjow} + locative pred. \\
                        &                           & (NP) + stative V + \textit{pjow} + (NP) \\
                        &                           & possessive pred. + \textit{pjow} \\
                        &                           & possessive V + \textit{pjow} \\
                        & non-main clause           & (NP) + V + \textit{pjow} + (NP)              \\
                        & adjective                 & NP(N+Adj) + \textit{pjow}  \\
                        & indefinite                & interr. pron. + \Neg{}\\
                        &                           & a. nobody: \textit{prë} \dots{} \textit{pjow} \\
                        &                           & b. nothing: \textit{pjän} \dots{} \textit{pjow}  \medskip\\
        \hline
    \textit{jy pjow}    &   declarative             & a. clause + \textit{jy=∅=pjow} \\
                        &                           & b.  \textit{jy=∅=pjow} + clause               \medskip\\
    %\medskip 
    \hline
    \textit{rö}         & declarative               & (NP) + V + \textit{rö} + (NP) \\
                        & interrogative             & \textsc{q} + (NP) + V + \textit{rö} + (NP) \\
                        & stative                   & stative pred. + \textit{rö} \\
                        &                           & NP + \textit{rö} + locative pred.  \\
                        &                           & (NP) + stative V + \textit{rö} + (NP) \\
                        & non-main clause           & (NP) + V + \textit{rö} + (NP)                \\
                        & adjective                 & NP(N+Adj) + \textit{rö} \\
                        & abessive                  & NP + \textit{rö} \medskip\\
    %\medskip 
    \hline
    \textit{inkjown}    & declarative               & (NP) + V  + (NP) + \textit{inkjown} \\
                        & interrogative             & \textsc{q} + (NP) + V + (NP) + \textit{inkjown} \\
                        & stative                   & stative pred. + \textit{inkjown} \\
                        &                           & \textit{inkjown} + existential pred. \\
                        & non-main                  & (NP) + V  + (NP) + \textit{inkjown}   \\       
                        &  negative reply           & \textit{inkjown} + (XP/clause) \medskip\\    
    %\medskip
    \hline
    \textit{sä}         & imperative                & (NP) + V + \textit{sä} + (NP)                 \medskip\\
   
%\midrule
%\multicolumn{3}{l}{\textbf{Non-clausal negation}} \\
     %\medskip
      
     %\medskip
    
    
\midrule
%\multicolumn{3}{l}{\textbf{Other}}\\
     \textit{tö}        & reinforcement             & (NP) + \textit{tö}=V + \Neg{} + (NP)  \\
\lspbottomrule
  \end{tabular}
  \caption{Negation in Panará}\label{pan:pnr-table}
\end{table}


Of these, \textit{pjow} is the least marked negator: it is compatible with all types of clause-level negation except for imperatives, and also participates in phrasal negation. It is by far the most frequent negation marker. \textit{Rö} is a less frequent counterpart of \textit{pjow}, sharing its position as a postverbal negative adverb. These two markers lean in opposite directions regarding their favoured constituents, \textit{pjow} being more often used with clausal negation and \textit{rö} with phrasal negation, although both can negate different categories of constituents.

\hspace*{-3pt}\textit{Inkjown} does double duty as the negative reply and a clause-final standard negation marker. Its canonical clause-final position suggests that perhaps it should be considered a negative predicate, in line with negation as usually employed in other Jê languages. However, it is different from \textit{jy pjow}, the verbal use of \textit{pjow}, in the lack of corresponding verbal morphology.

One characteristic of Panará negation is its imperviousness to grammatical categories. With the exception of imperatives, which exhibit a dedicated negation marker \textit{sä}, the different grammatical environments that are relevant in Panará morphosyntax are largely compatible with the standard negation mechanisms.



\section*{Acknowledgements}

I want to express my gratitude to my Panará informants and language teachers, especially Perankô, Kypakjä, Akââ, Jôôtu and Kabakra, as well as my adoptive family in the Kwakjatantêra clan and the people in the village of Nänsêpotiti. I would also like to thank the editors of this volume, Matti Miestamo and Ljuba Veselinova -- in particular Matti, who already brought the topic of negation to Amazonia in 2016 at the Amazónicas \textsc{vi} conference, in Leticia and Tabatinga. I am grateful to Héloïse Calame for her support in the editing process, and also to João Paulo Denófrio, Jack Hoeksema and Ares Llop for valuable discussions and input. Part of the research presented here was funded by an individual grant from the Endangered Languages Documentation Programme (ELDP).


\section*{Abbreviations}

\begin{tabularx}{.45\textwidth}{lQ}
\scshape 1 & first person \\
\scshape 2 & second person \\
\scshape 3 & third person \\
\scshape abs & absolutive \\
\scshape ade & adessive \\
\scshape all & allative \\
\scshape caus & causative \\
\scshape com & comitative \\
\scshape dat & dative \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape dir & directional \\
\scshape du & dual \\
\scshape emph & emphasis \\
\scshape erg & ergative \\
\scshape fnl & final \\
\scshape ine & inessive \\
\scshape ins & instrumental \\
\scshape ints & intensifier \\
\scshape intr & intransitive \\
\end{tabularx}

\begin{tabularx}{.45\textwidth}{lQ}
\scshape irr & irrealis \\
\scshape iter & iterative \\
\scshape loc & locative \\
\scshape nml & nominal form \\
\scshape neg & negative \\
\scshape pl & plural \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape plac & pluractional \\
\scshape poss & possessive \\
\scshape q & interrogative \\
\scshape rel & relative \\
\scshape sg & singular \\
\\
\end{tabularx}

\printbibliography[heading=subbibliography,notkeyword=this]
\end{document}
