\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233956}
\author{Muhammad Kamal Khan\affiliation{Allama Iqbal Open University} and Henrik Liljegren\affiliation{Stockholm University}}
\title{Negation in Pashto}
\abstract{In this paper, we explore negation in Pashto – an Eastern Iranian language spoken mainly in Pakistan and Afghanistan. Based on the Yousafzai dialect of Pashto, with the questionnaire provided by the editors as our main instrument, we investigate such phenomena as standard negation, negation in non-declaratives, stative predications, and negation in non-main clauses. Standard negation in Pashto is expressed with a single negative particle \textit{nə}, typically added pre-verbally. For imperatives, there is a dedicated negative particle \textit{mə}, also added pre-verbally, constituting the only formal contrast between imperatives and corresponding prohibitives. No structural differences between negation in main clauses and negation in non-main clauses of simplex verbs, or in non-periphrastic categories, have been observed, whereas there are instances of structural asymmetry in some periphrastically expressed verbal categories. Pashto displays paradigmatic asymmetry in that positive imperatives can make an aspectual distinction between perfective and imperfective, whereas negative imperatives are aspectually neutralized. Negation in Pashto can be further reinforced by changing the position of the negative element to occur post-verbally. Clauses with negated pronouns or adverbs always display negative concord.
\keywords{Iranian, negative particle, paradigmatic asymmetry, Pashto, structural asymmetry}
}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 
%\shorttitlerunninghead{}%%use this for an abridged title in the page headers

\section{The language}\label{sec:pas:1}

The aim of this paper is to explore negation in Pashto  (ISO: \href{https://iso639-3.sil.org/code/pbu}{pbu}, Glottocode: \href{https://glottolog.org/resource/languoid/id/nort2646}{nort2646}) in a typological perspective. Although there are several book-length grammatical descriptions of Pashto \citep{Babrakzai1999,David2014,Grjunberg1987,Lorenz1979,Penzl1955,TegeyRobson1996}, there is still no dedicated study of negation in the language. The language material used for the present study was almost exclusively elicited by means of the questionnaire provided by the editors \parencitetv{chapters/questionnaire}. The target variety is a northeastern dialect of Pashto often referred to as Yousafzai Pashto, spoken in the northern part of Khyber Pakhtunkhwa province of Pakistan. This is the native language of one of the co-authors and is also considered the basis of literary Pashto in Pakistan. Our assumption is that the characteristics of negation emerging in our description are largely common to all of the major varieties of Pashto, in Pakistan as well as in Afghanistan. In our examples, we have essentially adopted the broad phonemic transcription of Pashto used by \citet{David2014}, while the glossing reflects our own analytical choices.

Pashto is an East Iranian language spoken primarily in Pakistan and Afghani\-stan. Its exact classification in relation to other Iranian languages is in no way final, but suffice it to say that some of its closest relatives within Iranian are found among the so-called Pamir languages (spoken by comparatively small populations in northeastern Afghanistan and in adjacent areas of Pakistan and Tajikistan), whereas its relatedness to e.g. Farsi-Dari-Tajik (considered West Iranian) is of a much more distant kind (\citealt{Windfuhr2009}: 9–15; \citealt{Skjaervø1989a}: 370; \citealt{MacKenzie2009}: 443). Pashto is the second largest language of Pakistan, spoken by approximately 18 per cent of the population, surpassed only by Punjabi.\footnote{Pakistan Bureau of Statistics. 2023. Final Results of Census-2017. \url{https://www.pbs.gov.pk/content/final-results-census-2017-0} (23 August 2023).}
Most speakers are found in the west and the northwest of the country, with a main concentration in the northwestern Khyber Pakhtunkhwa province, but there are also large speaker populations in Balochistan and in all of the major cities. In Afghanistan, Pashto enjoys the status of official language, along with Dari, and is the main native language in the east, south and southwest of the country, corresponding to approximately 50 per cent of the total population \citep[721]{RobsonTegey2009}. Outside of Pakistan and Afghanistan, Pashto is also spoken by sizeable groups in India, Iran and the United Arab Emirates. Although there are no indisputable statistics when it comes to the number of speakers, there is an estimated figure of 30 to 50 million Pashto speakers in total, making it the most widely spoken Iranian language after Persian \citep[460]{MacKenzie2009}. Although Pashto exhibits significant linguistic variation, and relatively little (conscious or spontaneous) standardization, there is a high level of mutual intelligibility, even over great geographical distances \citep[721--722, 766]{RobsonTegey2009}. The latter can most likely be attributed to the fact that all major variation is phonological and lexical, rather than morphosyntactic (\citealt[231]{MacKenzie1959}; \citealt[736]{Elfenbein1997}). 

Pashto is a strongly head-final language, typically displaying an SOV clause word order. However, it also has a large inventory of second-position clitics, encoding modal, pronominal as well as adverbial meanings (\citealt{Tegey1977}; \citealt{Pate2012}). In the past tenses, nominal as well as verbal alignment is ergative-absolutive, while it is fully nominative-accusative in non-past tenses. From an Iranian perspective, Pashto has an unusually complex nominal and verbal morphology  (\citealt{MacKenzie2009}; \citealt[111]{Grjunberg1987}), with added complexities in the form of stem and stress alternations \citep[111]{Morgenstierne1942}. Nouns inflect for gender (masculine, feminine), number (singular, plural) and case (primarily contrasting direct and oblique). Numerous case relations are expressed by pre- and postpositions, or a combination of both. Verbs inflect for person, number and gender, and have two stems, one for present and another for past. Based on those stems, morphologically distinct imperfective vs. perfective verbs (both in the present and in the past) are formed. A few other verbal categories, including perfect and pluperfect, are formed periphrastically with auxiliaries derived from ‘be’ and ‘become’. Pashto verbs are a closed rather than an open class, a feature that is clearly areal in its distribution, with new predicates entering the language via productive light verb constructions (to a large extent overlapping with the class of denominal verbs) with \textit{kedəl} ‘to become’ and \textit{kol} \textit{(kawəl)} ‘to do, make’ (\citealt{Butt2010}: 50; \citealt{DavidGoodman2014}: 186; \citealt{Liljegren2010}: 51–52; \citealt{Masica1993}; \citealt{Wohlgemuth2009}: 152–155).


\section{Clausal negation}\label{sec:pas:2}

\subsection{Standard negation}\label{sec:pas:2.1}

In Pashto, standard negation, as defined and applied by e.g., \citet{Miestamo2005}, is expressed with a negative particle \textit{nə} (alternatively \textit{na,} particularly in its stressed realization) which typically precedes the finite verb of a clause. Example (\ref{ex:pas:1}) shows a declarative verbal main clause, in this case in the present imperfective, first in the affirmative, \REF{ex:pas:1a}, and then in the corresponding negative clause, \REF{ex:pas:1b}. 

\ea\label{ex:pas:1}
    \ea\label{ex:pas:1a}
        \gll Š\={a}do gaḍ{}-eg{}-i.\\
             monkey dance{}-\textsc{prs-3}\\
        \glt ‘The monkey is dancing.’
    \ex\label{ex:pas:1b}
        \gll  Š\={a}do nə gaḍ-eg{}-i.\\
             monkey \textsc{neg} dance{}-\textsc{prs-3}\\
        \glt ‘The monkey is not dancing.’
    \z
\z 

There are good reasons to treat \textit{nə} as a particle rather than an affix. On the one hand, it does form a prosodic unit with the finite verb, and it also carries the main stress of the verb phrase when occurring in this preverbal position \citep[110]{Tegey1977}. When, for instance, an \textit{a}{}-initial verb (a distinct verb class in Pashto), e.g., \textit{axli} in \REF{ex:pas:2a}, is negated, the vowel hiatus creates a long vowel, as in \textit{nāxli} in \REF{ex:pas:2b}. 

\ea\label{ex:pas:2}
    \ea\label{ex:pas:2a}
        \gll Kamāl	kitāb	axl-i. \\
             Kamal book take.\textsc{prs{}-3}\\
        \glt ‘Kamal is buying the book.’
    \ex\label{ex:pas:2b}
        \gll Kamāl kitāb nāxl{}-i. \\
             Kamal book \textsc{neg}.take.\textsc{prs-3}\\
        \glt ‘Kamal is not buying the book.’
    \z
\z 

On the other hand, the negative particle is sometimes moved to the end of the utterance, i.e. occurring after the finite verb, thus lending support to Pashto standard negation as essentially non-affixal and syntactic (\citealt{Dahl1979}: 83–84). The post-verbal, or utterance{}-final, position has a function of placing emphasis on the negation or stressing the impossibility of the proposition, as earlier noted by e.g. \citet[398]{Skjaervø1989b}. Example \REF{ex:pas:3a} shows a regular negative clause, and \REF{ex:pas:3b} shows the corresponding emphatic negative clause with a right-dislocated \textit{na}. See \sectref{sec:pas:4.4}.

\ea\label{ex:pas:3}
    \ea\label{ex:pas:3a}
        \gll Haɣa ye nə darkaw-i. \\
            \textsc{3sg.dist.dr} \textsc{3.clt} \textsc{neg} give.you.\textsc{ipfv-3} \\
        \glt ‘She is not giving it to you.’ 
    \ex\label{ex:pas:3b}
        \gll Haɣa ye darkaw-i na. \\
             \textsc{3sg.dist.dr} \textsc{3.clt} give.you.\textsc{ipfv-3} \textsc{neg}\\
        \glt ‘She is not ever giving it to you.’
    \z
\z 

Pashto standard negation is largely symmetric. Nearly every member of the verb paradigm used in the affirmative can also appear in the negative. Apart from the presence of the negative marker \textit{nə}, negative declarative clauses expressed in other non-periphrastic tense-aspect categories show no structural differences as compared to their corresponding affirmative expressions. In (\ref{ex:pas:4}), an affirmative clause in the past imperfective, another category formed exclusively by verb morphology, is shown with its corresponding negative clause. 

\ea\label{ex:pas:4}
    \ea\label{ex:pas:4a}
        \gll Š\={a}do gaḍ{}-ed{}-o.\\
            monkey dance{}-\textsc{pst-3sg.m}  \\
        \glt ‘The monkey was dancing.’
    \ex\label{ex:pas:4b}
        \gll Š\={a}do nə gaḍ-ed{}-o. \\
            monkey \textsc{neg} dance{}-\textsc{pst-3sg.m} \\
        \glt ‘The monkey was not dancing.’
    \z
\z 

The same grammatical categories are available in negative as in affirmative clauses, as can be seen in the following examples, in \REF{ex:pas:5a} a negated past imperfective clause with feminine singular agreement, and in \REF{ex:pas:5b} a negated clause with plural agreement. 

\ea\label{ex:pas:5}
    \ea\label{ex:pas:5a}
        \gll Š\={a}dogay nə gaḍ{}-ed-a.\\
             female.monkey \textsc{neg} dance-\textsc{pst-3sg} \\
        \glt ‘The female monkey was not dancing.’
    \ex\label{ex:pas:5b}
        \gll Š\={a}do{}-gān nə gaḍ-ed{}-əl. \\
            monkey{}-\textsc{pl} \textsc{neg} dance-\textsc{pst-3pl.m} \\
        \glt ‘The monkeys were not dancing.’
    \z
\z 

In addition to suffixing verb morphology, some crucial aspectual and modal differentiations are expressed with particles. That is for instance the case with the past perfective, which is (regularly) expressed with the prefixed perfective marker \textit{wə,} as in (\ref{ex:pas:6}), and with the future, which is expressed with the modal clitic \textit{ba,} as in (\ref{ex:pas:7}). It should be noted that the negative particle, provided it is placed in pre-verbal position, invariably occurs between the aspectual/modal marker(s) and the inflected verb stem, as also noted by \citet[408]{BrugmanDavid2014}. 

\ea\label{ex:pas:6}
    \ea\label{ex:pas:6a}
        \gll Š\={a}do wə{}-gaḍ{}-ed-o.\\
             monkey \textsc{pfv}{}-dance-\textsc{pst-3sg}\\
        \glt ‘The monkey danced.’
    \ex\label{ex:pas:6b}
        \gll Š\={a}do wə{}-nə{}-gaḍ{}-ed-o. \\
             monkey \textsc{pfv}{}-\textsc{neg}{}-dance-\textsc{pst-3sg} \\
        \glt ‘The monkey did not dance.’
    \z
\z 

\ea\label{ex:pas:7}
    \ea\label{ex:pas:7a}
        \gll Š\={a}do ba gaḍ{}-eg-i.\\
             monkey \textsc{fut} dance-\textsc{prs-3}\\
        \glt ‘The monkey will be dancing.’ 
    \ex\label{ex:pas:7b}
        \gll Š\={a}do ba nə gaḍ-eg{}-i. \\
             monkey \textsc{fut} \textsc{neg} dance-\textsc{prs-3}\\
        \glt  ‘The monkey will not be dancing.’
    \z
\z 

Pashto transitive clauses in the past tenses, like the ones in (\ref{ex:pas:8}), follow an ergative alignment pattern, in which the finite verb agrees with the direct object and the subject is expressed in a non-direct case form. Even in such cases a symmetrical relationship is maintained between affirmatives and their corresponding negatives. 

\ea\label{ex:pas:8}
    \ea\label{ex:pas:8a}
        \gll Saṛi mewa wə-xwaṛ-a.\\ 
              man.\textsc{obl} fruit(\textsc{f}) \textsc{pfv}{}-eat.\textsc{pst-3sg.f}\\
        \glt ‘The man ate the fruit.’
    \ex\label{ex:pas:8b}
        \gll Saṛi mewa wə-nə{}-xwaṛ-a. \\
             man.\textsc{obl} fruit(\textsc{f}) \textsc{pfv}{}-\textsc{neg}{}-eat.\textsc{pst-3sg.f}\\
        \glt ‘The man didn’t eat the fruit.’
    \z
\z 

However, in the compound (i.e. periphrastically formed) verbal categories, such as in the present and past perfect, or in the potential categories, the finite auxiliary always follows the main (participial) verb in the affirmative, as in the present perfect clause in \REF{ex:pas:9a}, whereas in the negative, the negated inflected auxiliary can occur either before the participle, as in \REF{ex:pas:9b}, or after the participle, as in \REF{ex:pas:9c}. It should be noted that it is the pre-participle order that is the unmarked one, whereas the post-participle order puts focus or emphasis on the negation (and corresponds in function to the dislocated negative particle in example \REF{ex:pas:3b} above). 

\ea\label{ex:pas:9}
    \ea\label{ex:pas:9a}
        \gll Kamāl rāɣl-ē dē.\\
             Kamal come.\textsc{pfv-ptcp.sg.m} be.\textsc{prs.3sg.m}\\
        \glt ‘Kamal has come.’
    \ex\label{ex:pas:9b}
        \gll Kamāl nə dē rāɣl-ē. \\
             Kamal \textsc{neg} be.\textsc{prs.3sg.m} come.\textsc{pfv-ptcp.sg.m}\\
        \glt ‘Kamal hasn’t come.’ 
    \ex\label{ex:pas:9c}
        \gll Kamāl rāɣl-ē nə dē. \\
             Kamal come.\textsc{pfv-ptcp.sg.m} \textsc{neg} be.\textsc{prs.3sg.m} \\
        \glt ‘Kamal has \textit{not} come.’
    \z
\z 

As was mentioned in \sectref{sec:pas:1}, numerous predicates in Pashto are formed with a non-verb element (normally an adjective or a noun) and a “light verb” (or verbalizer), of which \textit{kedəl} ‘to become’ (forming intransitive verbs) and \textit{kawəl} (or \textit{kol}) ‘to do, make’ (forming transitive verbs) are the most productive (\citealt{BrugmanDavid2014}: 401–403). This large group of Pashto verbs has variously been referred to as denominal verbs (\citealt{DavidGoodman2014}: 200{}-203) and derivative verbs (\citealt{RobsonTegey2009}: 740; \citealt{TegeyRobson1996}: 105), although these terms are not entirely overlapping in all treatments of Pashto verbs. In reality, verbs of this type occur along a continuum of lexicalization, from those in which the two components are phonologically fused in some parts of the verb paradigm to those (newer ones) that remain more phrase-like. When clauses with such complex verbs are negated in the imperfective, the negative particle \textit{nə} may occur either before the entire construction, as in \REF{ex:pas:10b}, or immediately before the inflected verb, as in \REF{ex:pas:11b}, depending on the degree of fusion/lexicalization.\footnote{Note the intransitive complex construction that is used with reference to female persons getting married, and the corresponding transitive one that is used with reference to male persons, reflecting the patrilocal marriage pattern typically observed in the Pashtun society \citep[52--57]{Grima1993}.}

\ea\label{ex:pas:10}
    \ea\label{ex:pas:10a}
        \gll D\={a}  wādə keg-i?\\
             \textsc{3sg.f.dr} marriage become.\textsc{ipfv.prs{}-3}\\
        \glt ‘Is she getting married?’
    \ex\label{ex:pas:10b}
        \gll Na, dā nə wādə keg-i.\\
             no \textsc{3sg.f.dr} \textsc{neg} marriage become.\textsc{ipfv.prs-3}\\
        \glt ‘No, she is not getting married.’
    \z
\z 

\ea\label{ex:pas:11}
    \ea\label{ex:pas:11a}
        \gll Āy\={a}  kamāl wādə kaw-i?\\
             \textsc{q} Kamal marriage do.\textsc{ipfv.prs{}-3}\\
        \glt ‘Is Kamal getting married?’
    \ex\label{ex:pas:11b}
        \gll Na, haɣa os wādə nə kaw-i.\\
             no \textsc{3sg.dist.dr} now marriage \textsc{neg} do.\textsc{ipfv.prs-3}\\
        \glt ‘No, he is not getting married now.’
    \z
\z 

In the perfective, however, the two parts of the complex are always separated by the negation, as can be seen in the examples (\ref{ex:pas:12}) and (\ref{ex:pas:13}). Note, that when the verbs ‘become’ and ‘do’ are used as verbalizers (i.e. not as simplex verbs), the perfective is not formed with the prefix \textit{wə{}-,} as in (\ref{ex:pas:6}) and (\ref{ex:pas:8}); the aspectual contrasts are instead achieved by means of stem alternations: \textit{keg-} (\textsc{ipfv.prs}), \textit{ked{}-} (\textsc{ipfv.pst}), \textit{š-} (\textsc{pfv.prs}), \textit{šw-} (\textsc{pfv.pst}) ‘become’; \textit{kaw-} (\textsc{ipfv}), \textit{kṛ{}-} (\textsc{pfv}) ‘do’.

\ea\label{ex:pas:12} 
    \gll  Də bača tror parun wādə nə šw{}-a.\\ 
         of Bacha aunt yesterday marriage \textsc{neg} become.\textsc{pfv.pst{}-sg.f}\\
    \glt ‘Bacha’s aunt didn’t get married yesterday.’
\z 

\ea\label{ex:pas:13} 
    \gll  Də bača trə parun wādə nə wə kaṛ{}-ē.\\ 
          of Bacha uncle yesterday marriage \textsc{neg} be.\textsc{pst.3sg.m} do.\textsc{pfv{}-ptcp.sg.m}\\
    \glt ‘Bacha’s uncle didn’t get married yesterday.’
\z 

One particular construction that deserves to be mentioned is \textit{l\={a} nə} (< \textit{l\={a}} ‘still, even, yet’ + the negative particle \textit{nə}), the semantic equivalent of the English phasal adverbial ‘not yet’. An example is given in (\ref{ex:pas:14}) of an affirmative stative proposition with \textit{l\={a}} ‘[is] still’ and its corresponding negative with \textit{l\={a} nə}.

\ea\label{ex:pas:14}
    \ea\label{ex:pas:14a}
        \gll Kam\={a}l l\={a}  dalta d\={e}.\\
             Kamal still here be.\textsc{prs.3sg.m}\\
        \glt ‘Kamal is still here.’
    \ex\label{ex:pas:14b}
        \gll Kam\={a}l l\={a}  dalta nə d\={e}. \\
             Kamal yet here \textsc{neg} be.\textsc{prs.3sg.m}\\
        \glt ‘Kamal is not here yet.’
    \z
\z 

As is shown in (\ref{ex:pas:15}), a clause containing \textit{l\={a}  nə} can occur as the negative answer to a question in which \textit{l\={a}wal} (< \textit{l\={a} awal} lit. ‘still first’), corresponding to the phasal adverbial ‘already’, is used.

\ea\label{ex:pas:15}
    \ea\label{ex:pas:15a}
        \gll Kam\={a}l l\={a}wal tal-\={e} d\={e}? \\
             Kamal already go.\textsc{ipfv.prs-ptcp.3sg.m} be.\textsc{prs.3sg.m} \\
        \glt ‘Has Kamal already gone?’
    \ex\label{ex:pas:15b}
        \gll Na, kam\={a}l l\={a} nə d\={e} tal-\={e}.\\
             no Kamal still \textsc{neg} be.\textsc{prs.3sg.m}  go.\textsc{ipfv.pst-ptcp.sg.m}\\
        \glt ‘No, Kamal has not yet gone.’
    \z
\z 

We leave it open to further examination, and the availability of more data, as to whether the different interpretations of the construction can be related to telicity (as suggested by one of the reviewers).

\subsection{Negation in non-declaratives}\label{sec:pas:2.2}

Pashto prohibitives have a dedicated negative construction, different in form from standard negation. This is the globally most commonly occurring type, according to \citegen{AuweraEtAl2013} cross-linguistic study of prohibitives. The particle \textit{mə} (alternatively \textit{ma}) typically occurs before the second singular or plural imperative verb form, as can be seen in example (\ref{ex:pas:16}). 

\ea\label{ex:pas:16}
    \ea\label{ex:pas:16a}
        \gll W\={a}pas ye rāwṛ{}-a!\\
             back \textsc{3.clt} bring.me{}-\textsc{imp.sg}\\
        \glt ‘Return it!’
    \ex\label{ex:pas:16b}
        \gll Wāpas ye mə rāwṛ{}-a!\\
             back \textsc{3.clt} \textsc{neg} bring.me{}-\textsc{imp.sg}\\
        \glt ‘Don’t return it!’
    \z
\z 

There is an asymmetry in the imperative. For many verbs, a perfective and an imperfective imperative are available. However, there is normally only one corresponding prohibitive construction using the imperfective, as can be seen in examples (\ref{ex:pas:17}--\ref{ex:pas:19}).\footnote{This aspectual neutralization in the negative also applies to jussives, i.e., constructions in which the deontic particle \textit{de} is used in order to express first and third person obligation (or non-obligation in the corresponding propositions with the standard negation \textit{nə}).}  This is an example of what \citet[114]{Miestamo2005} labels paradigmatic asymmetry of type A/Cat/TAM, in which neutralization between aspectual distinctions arises in the negative. Commands are typically expressed with perfective imperatives, as in \REF{ex:pas:17a}, \REF{ex:pas:18a} and \REF{ex:pas:19a}, but imperfective imperatives (\REF{ex:pas:17b}, \REF{ex:pas:18b} and \REF{ex:pas:19b}) can be described as intensifying or adding aspects of urgency, repetition or continuation (\citealt{DavidGoodman2014}: 284), and, at times, to show emphasis on the action. Similar effects of using the imperfective in imperatives have been observed, e.g., for Bulgarian and Kabyle (Ljuba Veselinova, personal communication, 25 October 2020).

\ea\label{ex:pas:17}
    \ea\label{ex:pas:17a}
        \gll Wə-xwar{}-a!\\
             \textsc{pfv}{}-eat.\textsc{prs}{}-\textsc{imp.sg}\\
        \glt ‘Eat (it)!’
    \ex\label{ex:pas:17b}
        \gll Xwar-a!\\
             eat.\textsc{prs}{}-\textsc{imp.sg}\\
        \glt ‘Go on and eat (it)!’
    \ex\label{ex:pas:17c}
        \gll Mə xwar-a!\\
             \textsc{neg} eat.\textsc{prs}{}-\textsc{imp.sg}\\
        \glt ‘Don’t eat (it)!’
    \z
\z 

\ea\label{ex:pas:18}
    \ea\label{ex:pas:18a}
        \gll Lāṛ š{}-a!\\
             go.\textsc{pfv} become\textsc{.pfv.prs{}-imp.sg}\\
        \glt ‘Go!’
    \ex\label{ex:pas:18b}
        \gll Z-a!\\
             go.\textsc{ipfv.prs-imp.sg}\\
        \glt ‘Get going!’
    \ex\label{ex:pas:18c}
        \gll Mə z-a!\\
        \textsc{neg} go.\textsc{ipfv.prs}{}-\textsc{imp.sg}\\
        \glt ‘Don’t go!’
    \z
\z 

\ea\label{ex:pas:19}
    \ea\label{ex:pas:19a}
        \gll Bača ta wə-wāy-a!\\
             Bacha to \textsc{pfv}{}-tell.\textsc{prs}{}-\textsc{imp.sg}\\
        \glt‘Tell Bacha!’ 
    \ex\label{ex:pas:19b}
        \gll Bača ta wāy-a!\\
             Bacha to tell.\textsc{prs}{}-\textsc{imp.sg}\\
        \glt ‘Keep telling Bacha!’
    \ex\label{ex:pas:19c}
        \gll Bača ta mə wāy{}-a!\\
             Bacha to \textsc{neg} tell.\textsc{prs}{}-\textsc{imp.sg}\\
        \glt ‘Don’t tell Bacha!’
    \z
\z 

Just like the particle \textit{nə} used in standard negation, the particle \textit{mə} can occur sentence-finally, too. In such cases, it is to show a special focus or put emphasis on the negation. See the contrasting examples in (\ref{ex:pas:20}).

\ea\label{ex:pas:20}
    \ea\label{ex:pas:20a}
        \gll Mə ye axl-a! \\
            \textsc{neg} \textsc{3.clt} take.\textsc{prs{}-imp.sg} \\
        \glt ‘Don’t take it!’
    \ex\label{ex:pas:20b}
        \gll Axl-a ye ma!\\
             take.\textsc{prs{}-imp.sg} \textsc{3.clt} \textsc{neg} \\
        \glt ‘\textbf{Don’t} take it!’
    \z
\z 

The alternation between a pre-verbal and non-emphatic \textit{mə} and an utterance-final and emphatic \textit{ma} is parallel to the previously mentioned (\sectref{sec:pas:2.1}) alternation between \textit{nə} and \textit{na}. 

Another somewhat special case that deserves to be mentioned is the use of \textit{mə} with second person singular inflected indicative verb forms, instead of with an imperative form, in a few set constructions with an optative meaning. These are for instance found in greetings or routine expression of well-wishing or encouragement. The expression in (\ref{ex:pas:21}) may typically be uttered by an acquaintance addressing a farmer working in his field as he passes by, thus wishing him luck and good health. Similarly, the expression in (\ref{ex:pas:22}) may be uttered encouragingly to another participant in a group-singing session to go on singing. 

\ea\label{ex:pas:21} 
    \gll  Stəṛē mə š-e! \\ 
         tired.\textsc{sg.m} \textsc{neg} become.\textsc{pfv.prs{}-2sg} \\
    \glt ‘May you not get tired!’
\z 

\ea\label{ex:pas:22} 
    \gll Xwār mə š-e!  \\ 
          poor \textsc{neg} become.\textsc{pfv.prs-2sg}\\
    \glt ‘May you not become weak!’
\z 

As exemplified by \citet[275--276]{DavidGoodman2014}, the corresponding affirmative expression \textit{xwār še} ‘may you be miserable!’ can be used to express an ill-wish, or even a curse.

\subsection{Negation in stative predications}\label{sec:pas:2.3}

In many non-verbal clauses too, the negation particle \textit{nə} is used, to negate equation (\ref{ex:pas:23}), proper inclusion (\ref{ex:pas:24}), attribution (including temporary attribution) (\ref{ex:pas:25}), and locative predication (\ref{ex:pas:26}), just as is applied in standard negation. 

\ea\label{ex:pas:23} Equation
    \ea\label{ex:pas:23a}
        \gll Bača zəmā ust\={a}z d\={e}.  \\
             Bacha \textsc{1sg.poss} teacher be.\textsc{prs.3sg.m} \\
        \glt ‘Bacha is my teacher.’
    \ex\label{ex:pas:23b}
        \gll Bača zəmā ust\={a}z nə d\={e}. \\
             Bacha \textsc{1sg.poss} teacher \textsc{neg} be.\textsc{prs.3sg.m} \\
        \glt ‘Bacha is not my teacher.’ 
    \z
\z 

\ea\label{ex:pas:24} Proper inclusion
    \ea\label{ex:pas:24a}
        \gll Malang yaw tark\={a}ṇ d\={e}.  \\
              Malang one carpenter be.\textsc{prs.3sg.m}\\
        \glt ‘Malang is a carpenter.’
    \ex\label{ex:pas:24b}
        \gll Malang yaw tarkāṇ nə d\={e}.  \\
            Malang one carpenter \textsc{neg} be.\textsc{prs.3sg.m} \\
        \glt ‘Malang is not a carpenter.’
    \z
\z 

\ea\label{ex:pas:25} Attribution
    \ea\label{ex:pas:25a}
        \gll Adam lewan\={e}  d\={e}. \\
             Adam insane.\textsc{sg.m} be.\textsc{prs.3sg.m} \\
        \glt  ‘Adam is insane.’ 
    \ex\label{ex:pas:25b}
        \gll Adam lewan\={e}  nə d\={e}.  \\
             Adam insane.\textsc{sg.m} \textsc{neg} be.\textsc{prs.3sg.m}\\
        \glt ‘Adam is not insane.’
     \ex\label{ex:pas:25c}
        \gll Adam os xapa nə d\={e}.  \\
             Adam now sad \textsc{neg} be.\textsc{prs.3sg.m} \\
        \glt ‘Adam is not sad now.’ [temporal attribution] 
    \z
\z 

\ea\label{ex:pas:26} Locative predication
    \ea\label{ex:pas:26a}
        \gll Bača pa kor dē. \\
             Bacha at house be.\textsc{prs.3sg.m}\\
        \glt ‘Bacha is at home.’
    \ex\label{ex:pas:26b}
        \gll Bača pa kor nə dē. \\
             Bacha at house \textsc{neg} be.\textsc{prs.3sg.m}\\
        \glt ‘Bacha is not at home.’
    \z
\z 

However, existential (\ref{ex:pas:27}) and, at least some, possessive predications (\ref{ex:pas:28}) make use of a non-inflected existential negator (NegEx), the particle \textit{nište} ‘doesn’t exist, isn’t there, isn’t available’ (< \textit{nə šte}) for negating. This reflects a general cross-linguistic tendency (\citealt{Veselinova2013neg.exist}: 123–124). The corresponding affirmative predications use a dedicated existential copula \textit{šte} ‘exists, is there, is available’.

\ea\label{ex:pas:27} Existential predication
    \ea\label{ex:pas:27a}
        \gll Pa zangal ki mzari šte. \\
              in forest in lion.\textsc{pl} \textsc{ex}\\
        \glt ‘There are lions in the forest.’
    \ex\label{ex:pas:27b}
        \gll Pa zangal ki mzari nište.  \\
            in forest in lion.\textsc{pl} \textsc{negex}\\
        \glt ‘There are no lions in the forest.’ 
    \z
\z 

\ea\label{ex:pas:28} Possessive predication
    \ea\label{ex:pas:28a}
        \gll Adam sara topak šte. \\
              Adam with gun \textsc{ex}\\
        \glt ‘Adam has a gun.’
    \ex\label{ex:pas:28b}
        \gll Adam sara topak nište.  \\
            Adam with gun \textsc{negex}\\
        \glt ‘Adam does not have a gun.’
    \z
\z 

NegEx can also be used in contrastive negation, when the presence of somebody or something in one place is contrasted with its absence in another place. In this case, however, the affirmative part is expressed with the copula used in locative predication, as in (\ref{ex:pas:29}).

\ea\label{ex:pas:29} 
    \gll  Kamal hujra ki d\={e} kor ki nište. \\ 
          Kamal hujra in be\textsc{.prs.sg.m} home in \textsc{negex}\\
    \glt ‘Kamal is at the \textit{hujra} (village common house), not at home.’ 
\z 

Additionally, NegEx can carry a connotation of being gone for good, i.e., having died, as in (\ref{ex:pas:30}).

\ea\label{ex:pas:30} 
    \gll  Zəmung də stargo ašn\={a}  os nište.\\ 
          \textsc{1pl.poss} of eye.\textsc{pl.obl} known now \textsc{negex}\\
    \glt ‘(People) known to our eyes are no more (alive).’
\z 

Occasionally, it can be used adjectivally, in a contracted form, \textit{ništ}, again with the sense ‘gone’, ‘is no more’, as in \REF{ex:pas:31}, in this case along with the inflected form of ‘become’.

\ea\label{ex:pas:31} 
    \gll  Zəmung xwag{}-ə y\={a}r{}-\={a}n os ništ š-u.\\ 
          \textsc{1pl.poss} sweet{}-\textsc{pl.m} friend{}-\textsc{pl} now gone become.\textsc{pfv.prs-1pl}\\
    \glt ‘Our sweet friends are now gone.’ 
\z 

Certain locative predications may alternatively use the construction in (\ref{ex:pas:26}) and the one normally associated with existential predication, as in (\ref{ex:pas:27}). This is particularly the case if the presence of a third person is queried over the phone (\citealt{TegeyRobson1996}: 93--94), as in \REF{ex:pas:32a}. The reply may use either construction, but it is more common for the negative reply to make use of the negative existential, as in \REF{ex:pas:32b}, and the affirmative to use the inflected copula \textit{d\={e} }, etc., as in \REF{ex:pas:32c}. 

\ea\label{ex:pas:32}
    \ea\label{ex:pas:32a}
        \gll Kamāl pa kor ki šte? \\
             Kamal in house in \textsc{ex} \\
        \glt  ‘Is Kamal at home (i.e. available)?’
    \ex\label{ex:pas:32b}
        \gll Na, kamāl pa kor ki nište. \\
             no Kamal in house in \textsc{negex}\\
        \glt ‘No, Kamal is not at home.’
     \ex\label{ex:pas:32c}
        \gll Ao, kamāl pa kor ki d\={e}. \\
             yes Kamal in house in be.\textsc{prs.3sg.m}\\
        \glt \textsc{‘Y}es, Kamal is at home.’
    \z
\z 

\subsection{Negation in non-main clauses}\label{sec:pas:2.4}

Negation in Pashto dependent and subordinate clauses is expressed with the negation particle \textit{nə,} just as in standard negation. The following examples illustrate its use in both a finite, relative clause (\ref{ex:pas:33}), and in a non-finite, noun complement, clause (\ref{ex:pas:34}).

\ea\label{ex:pas:33} 
     \gll  Haɣa halək [če imtih\={a}n ye nə šo p\={a}s kol{}-e] skul prexud.\\ 
     that.\textsc{dr} boy \textsc{comp} test \textsc{3.clt} \textsc{neg} become.\textsc{pfv.pst.3sg.m} passing do.\textsc{pfv.pst{}-opt} school abandon.\textsc{pst.pfv.3sg.m} \\
    \glt ‘The boy [who could not pass the examination] left school.’
\z 

\ea\label{ex:pas:34} 
     \gll [Ṭem nə z\={a}ya kol] yaw xa \={a}dat d\={e}.  \\ 
          time \textsc{neg} wastage do.\textsc{inf} one good habit be.\textsc{prs.3sg.m}\\
    \glt ‘[Wasting no time] is a good habit.’
\z 

The main verb of the subordinate clause in (\ref{ex:pas:33}) is part of a complex verb construction, \textit{pās kol} ‘to pass, succeed’, and occurs here in the past potential, in which it is the auxiliary \textit{šo} that is inflected and is immediately preceded by the negation particle. The non-finite verb in the noun complement in (\ref{ex:pas:34}) is also part of a complex verb, \textit{z\={a}ya kol} ‘to waste’; in this case, the negation particle precedes the complex verb construction as a whole.

In Youzafzai Pashto, negative purpose could be expressed in two different ways. Either the verb in the subordinate clause that is introduced with the complementizer \textit{če} is negated, as in (\ref{ex:pas:35}), also mentioned by \citet[85]{Pate2012}, or the complementizer itself is negated, \textit{hasi na če} ‘thus not’, as in (\ref{ex:pas:36}). Note that in the latter case, the verb in the purpose clause is not negated in concord.

\ea\label{ex:pas:35} 
    \gll  Zar kaw-a če dar b\={a}ndi n\={a}waxt\={a}  nə š-i. \\ 
          quick do.\textsc{ipfv{}-imp.sg} \textsc{comp} \textsc{2} on late \textsc{neg} become.\textsc{pfv.prs{}-3}\\
    \glt ‘Hurry up, lest you get late.’
\z 

\ea\label{ex:pas:36} 
    \gll  Zar kaw-a hasi na če dar b\={a}ndi n\={a}waxt\={a}  š-i.\\ 
          quick do.\textsc{ipfv{}-imp.sg} thus \textsc{neg} \textsc{comp} \textsc{2} on late become.\textsc{pfv.prs{}-3} \\
    \glt ‘Hurry up, lest you should get late.’
\z 
 
\subsection{Negative lexicalizations}\label{sec:pas:2.5}

No clear examples of negative lexicalizations have been found in Pashto. 

\section{Non-clausal negation}\label{sec:pas:3}
\subsection{Negative replies}\label{sec:pas:3.1}

In Pashto, negative replies to polar questions are often expressed by using a one-word negative reply \textit{na}, similar in use to English ‘no’ (see further examples above: \REF{ex:pas:10b}, \REF{ex:pas:11b}, \REF{ex:pas:32b}). Note the following two polar questions, (\ref{ex:pas:37Q}Q) and (\ref{ex:pas:38Q}Q), and their answers, (\ref{ex:pas:37A}A) and (\ref{ex:pas:38A}A), respectively.

\ea\label{ex:key:37}
    \begin{xlist}
    \exi{Q: }\label{ex:pas:37Q}
        \gll  D\={a}  sp\={e}  ɣap{}-i?\\ 
              this.\textsc{dr} dog bark{}-3\\ 
        \glt  ‘Is the dog barking?’\footnote{Polar questions with an utterance-initial question particle \textit{āyā}, as reported by \citet[417]{BrugmanDavid2014}, are also observed in Youzafzai Pashto. In everyday speech, however, polar questions are normally identical to statements apart from an utterance-final rising pitch in questions, whereas the \textit{āyā{}-}construction is mostly used in formal speech.} 
    \exi{A: }\label{ex:pas:37A}
        \gll  Na!\\ 
                no\\
        \glt ‘No (it is not barking)!’
    \end{xlist}
\z
\ea\label{ex:key:38}
    \begin{xlist}
    \exi{Q: }\label{ex:pas:38Q}
        \gll  D\={a}  sp\={e}  nə ɣap{}-i?\\ 
              this.\textsc{dr} dog \textsc{neg} bark-3\\ 
        \glt ‘Isn’t the dog barking?’
    \exi{A: }\label{ex:pas:38A}
        \gll  Na!\\ 
            no\\
        \glt ‘No (it is not barking)!’
    \end{xlist}
\z

As these examples show, the polarity particle \textit{na} expresses in both cases a negative answer, in the case of a positive question, a non-confirmation of the positive polarity of the question, and in the case of a negative question, a confirmation of its negative polarity. It is therefore an example of a so-called yes-no answer system \citep[320–321]{KönigSiemund2007}. 

\subsection{Negative indefinites and adverbs}\label{sec:pas:3.2}

The negation of indefinite pronouns and adverbs (corresponding to English ‘nobody’, ‘no-one’, ‘nowhere’, ‘never’) is obtained by prefixing \textit{h\={i}ts} ‘none’ to an indefinite pronoun or adverb. Some examples are given in (\ref{ex:pas:39}--\ref{ex:pas:42}). Note the negative concord, i.e. that also the verb in those clauses must be negated, with the standard negation \textit{nə} as in (\ref{ex:pas:39}), (\ref{ex:pas:41}) and (\ref{ex:pas:43}), the negative existential \textit{nište}, as in (\ref{ex:pas:40}), or the negative prohibitive particle \textit{mə}, as in (\ref{ex:pas:42}). Note that \textit{h\={i}ts} can be left out in (\ref{ex:pas:43}), or be entirely replaced by \textit{bəl}, as in (\ref{ex:pas:44}), to express a negative exclusivity, ‘nobody else’.

\ea\label{ex:pas:39} 
    \gll  Haɣa h\={i}ts kala nə š{}-i r\={a}tl-e. \\ 
         \textsc{3sg.dist.dr} none when \textsc{neg} become.\textsc{pfv.prs{}-3} 
        come.\textsc{ipfv.pst{}-opt} \\
    \glt ‘She can never come.’
\z 

\ea\label{ex:pas:40} 
    \gll  Ap\={i}m was h\={i}ts čarta nište.\\ 
          opium now none where \textsc{negex}\\
    \glt ‘Opium is now found nowhere.’
\z 

\ea\label{ex:pas:41} 
    \gll  Zəmā h\={i}ts tsok mrasta nə kaw{}-i.\\ 
          \textsc{1sg.poss} none who.\textsc{dr} help \textsc{neg} do.\textsc{ipfv.prs-3}\\
    \glt ‘Nobody helps me.’
\z 

\ea\label{ex:pas:42} 
    \gll  Xpala jəba h\={i}ts kala mə pregd-a.\\ 
          own language none when \textsc{neg} abandon.\textsc{prs-imp.sg}\\
    \glt ‘Never give up your own language!’
\z 

\ea\label{ex:pas:43}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:pas:42Q}
        \glt ‘What are you buying?’
    \exi{A: }\label{ex:pas:42A}
        \gll  Zə (h\={i}ts) tsə n\={a}xl{}-əm.\\ 
              \textsc{1sg.nom} none what \textsc{neg.}buy.\textsc{ipfv.prs-1sg}\\
        \glt ‘I’m buying nothing.’
    \end{xlist}
\z 

\ea\label{ex:pas:44} 
    \gll  Xesta ḍer turzən dē; bəl tsok dāse nə š{}-i kaw-e.\\ 
        Khesta very brave be.\textsc{prs.3sg.m} other who.\textsc{dr}
        like.that \textsc{neg} become.\textsc{pfv.prs{}-3} do\textsc{.ipfv.prs-opt}\\
    \glt ‘Khesta is so brave. Nobody else can do it.’
\z 

\subsection{Negative derivation and case marking}\label{sec:pas:3.3}

The Pashto lexicon includes numerous, particularly adjectival, derivations that encode absence, opposition or antonymy. A number of different negating derivational prefixes occur, approximately corresponding to English ‘\nobreakdash-less’, ‘un\nobreakdash-’, ‘non\nobreakdash-’, ‘in\nobreakdash-’, etc. The most common such negating prefixes are \textit{n\={a}-}, \textit{be-} and \textit{la-}. The prefix \textit{n\={a}-} is the most productive, while \textit{be-} (of Persian origin) mostly occurs with roots of Persian (or Perso-Arabic) origin, and \textit{l\={a}-} (of Arabic origin) primarily occurs with roots of Arabic origin. Some examples are displayed in \tabref{tab:pas:1}.

%\hc{@matti: I changed \textsc{n} to N for Noun as we talked about, just making sure it's OK here in the table, compare with \textsc{adj}. You can delete the comment if it's ok.}
\begin{table}
\fittable{
\begin{tabular}{lll}
\lsptoprule
{\bfseries Root} & {\bfseries Neg. prefix} & {\bfseries Derivation}\\
\midrule 
\textit{joṛ} ‘healthy, well’ \textsc{adj} & \textit{n\={a}-} & \textit{nājoṛa} ‘unwell, sick’ \textsc{adj}\\
\textit{roɣ} ‘complete, healthy, well’ \textsc{adj} & & \textit{nāroɣa} ‘incomplete, unwell, sick’ \textsc{adj} \\
\textit{poha} ‘intelligent, learned’ \textsc{adj} &  & \textit{nāpoha} ‘ignorant, foolish’ \textsc{adj}\\
\textit{āšnā} ‘acquaintance’ \textsc{N} &  & \textit{nāšnā} ‘unknown, strange’ \textsc{adj}\\
\textit{hak} ‘just’ \textsc{adj,} \textsc{N} &  & \textit{nāhak} ‘unjust’ \textsc{adj}\\
\textit{rawā} ‘lawful, permitted’ \textsc{adj} &  & \textit{nārawā} ‘illegal, unlawful’ \textsc{adj}\\
\textit{tars} ‘fear’ \textsc{N} &  & \textit{nātars} ‘pitiless, fearless’ \textsc{adj}\\
\textit{pedā} ‘born, created’ \textsc{adj} &  & \textit{nāped} ‘rare, unavailable’ \textsc{adj}\\
\textit{lew} ‘plastering’ \textsc{N} &  & \textit{nālewa} ‘not plastered’ \textsc{adj}\\
\textit{waxt} ‘time’ \textsc{N} &  & \textit{nāwaxta} ‘late, out of time’ \textsc{adj}\\
\textit{gumān} ‘expectation, thought’ \textsc{N} &  & \textit{nāgumāna} ‘unexpected’ \textsc{adj}\\
\midrule
\textit{saṛitob} ‘human nature’ \textsc{N} & {\itshape be-} & \textit{besaṛitoba} ‘immodest’ \textsc{adj}\\
\textit{k\={a}r} ‘work’ \textsc{N} &  & \textit{bekāra} ‘useless’ \textsc{adj}\\
\textit{darak} ‘place; trace’ \textsc{N} &  & \textit{bedaraka} ‘lost, without direction’ \textsc{adj}\\
\textit{akal} ‘wisdom, reason’ \textsc{N} &  & \textit{beakla} ‘unwise, stupid’ \textsc{adj}\\
\textit{waxt} ‘time’ \textsc{N} &  & \textit{bewaxta} ‘untimely’ \textsc{adj}\\
\midrule 
\textit{hāsil} ‘produce, profit’ \textsc{N} & {\itshape lā-} & \textit{lāhāsil} ‘fruitless’ \textsc{adj}\\
\textit{parwā} ‘care, concern’ \textsc{N} &  & \textit{lāparwā} ‘careless’ \textsc{adj}\\
\lspbottomrule
\end{tabular}
}
\caption{Negative derivation in Pashto}
\label{tab:pas:1}
\end{table}

\section{Other aspects of negation}\label{sec:pas:4}

\subsection{The scope of negation}\label{sec:pas:4.1}

Pashto clausal negation, expressed with a pre-verbal negative particle, typically takes scope over the proposition as a whole, as in \REF{ex:pas:45a}. However, the scope can also be narrowed down to a specific lower-level constituent by means of prosodic stress (while the sentence structure remains unaffected). In \REF{ex:pas:45b}, emphatic stress on the subject pronoun \textit{haɣo} shifts the scope of negation from the proposition as a whole to the subject argument of the clause. The sentence now means that ‘they’, i.e., a few specific people, are not coming to the village, while some others might be coming. Similarly, in \REF{ex:pas:45c}, emphatic stress on \textit{kəli} shifts the scope of negation to the goal, meaning that they are coming to the village rather than to some other place, such as the city, which could be uttered with reference to some expected visitors coming from abroad. 

\ea\label{ex:pas:45}
    \ea\label{ex:pas:45a}
        \gll Haɣo kəli ta nə rāz-i.\\
             \textsc{3pl.dist.dr} village.\textsc{obl} to \textsc{neg} come.\textsc{ipfv.prs-3}\\
        \glt ‘They are not coming to the village.’ 
    \ex\label{ex:pas:45b}
        \gll \uline{Haɣo} kəli ta nə rāz-i.\\
             \textsc{3pl.dist.dr} village.\textsc{obl} to \textsc{neg} come.\textsc{ipfv.prs-3}\\
        \glt ‘\textit{They} are not coming to the village.’

    \ex\label{ex:pas:45c}
        \gll Haɣo \uline{kəli} ta nə rāz-i.\\
             \textsc{3pl.dist.dr} village.\textsc{obl} to \textsc{neg} come.\textsc{ipfv.prs-3}\\
        \glt ‘They are not coming \textit{to the village}.’ 
    \z
\z 

Another way of narrowing down the scope of negation is by negating indefinite pronouns or adverbs with \textit{h\={i}ts} ‘none’ in negative concord with a clause-level negative particle such as \textit{nə}, as previously described in \sectref{sec:pas:3.2}.

\subsection{Negative polarity}\label{sec:pas:4.2}

The following is by no means intended as an exhaustive list (\tabref{tab:pas:2}) but covers some of the most frequent and prominent negative polarity items in Pashto. Note that many of them include \textit{hīts} as a component, co-lexicalized, and sometimes phonologically fused, with an indefinite (or interrogative) pronoun/adverb (see \sectref{sec:pas:3.2}), or occurring as a nominal modifier (e.g. with \textit{kār} ‘work’, \textit{waxt} ‘time’, \textit{sūrat} ‘image’). In each case this negative polarity item displays negative concord with a clause-level negative particle. It is probably more correct to describe these as negative polarity constructions along with the clause-level negative particle rather than as negative polarity items in isolation (especially since \textit{hīts} in some cases can be left out, while the negative meaning of the sentence remains).

\begin{table}
\begin{tabularx}{\textwidth}{llQ}
\lsptoprule
{Form} & {Meaning} & {Use (licencing conditions)}\\
\midrule 
\textit{hīts{}-tsok} & ‘anybody, nobody’ & negated nominative NP with human reference\\
\textit{bəl-tsok} & ‘anybody else, nobody else’ & negated nominative NP with human reference\\
\textit{hīts-čā}, \textit{hī{}-čā} & ‘anybody, nobody’ & negated non-nominative NP with human reference\\
\textit{bəl-čā} & ‘anybody else, nobody else’ & negated non-nominative NP with human reference\\
\textit{(hīts{}-)tsə} & ‘anything, nothing’ & negated NP with inanimate reference\\
\textit{hīts-kār} & ‘anything, nothing’ & negated NP with reference to activity\\
\textit{hīts-šē} & ‘anything, nothing’ & negated NP with reference to inanimate physical entity\\
\textit{hīts-čerta}, \textit{hī-čerta} & ‘anywhere, nowhere’ & negated locational adverbial\\
\textit{hīts-čere}, \textit{hī-čere} & ‘never’ & negated time adverbial\\
\textit{hīts-kala} & ‘never’ & negated time adverbial\\
\textit{hīts-waxt} & ‘never’ & negated time adverbial\\
\textit{hīts-tor} & ‘by no means, in no way’ & negated manner adverbial\\
\textit{hīts-sūrat} & ‘by no means, in no way’ & negated manner adverbial\\
\lspbottomrule
\end{tabularx}
    \caption{Negative Polarity Items (NPIs) in Pashto}
    \label{tab:pas:2}
\end{table}

\subsection{Marking of NPs in the scope of negation}\label{sec:pas:4.3}

As shown in many other examples, clausal negation is largely symmetric in Pashto. The addition of a negative particle is for most of the TAM categories the only formal difference between a negative and a corresponding affirmative clause. Neither is case marking in any way affected under negation. As can be seen in the present tense utterances in (\ref{ex:pas:46}), the direct case marking of the transitive subject and the oblique case marking of the pronominal direct object remain the same, whether affirmative or negative.

\ea\label{ex:pas:46}
    \ea\label{ex:pas:46a}
        \gll Spē  mā xwar-i.\\
             dog.\textsc{dr} \textsc{1sg.obl} eat.\textsc{prs-3} \\
        \glt ‘The dog is biting me.’
    \ex\label{ex:pas:46b}
        \gll Spē mā nə xwar-i.\\
            dog.\textsc{dr} \textsc{1sg.obl} \textsc{neg} eat.\textsc{prs-3} \\
        \glt ‘The dog is not biting me.’
    \z
\z 

Similarly, in (\ref{ex:pas:47}), when the ergative alignment of the corresponding past tense utterances assigns oblique case to the transitive subject and direct case to the pronominal direct object, no changes in case marking takes place between the affirmative expression and the corresponding negative one.

\ea\label{ex:pas:47}
    \ea\label{ex:pas:47a}
        \gll Spi zə wə-xwaṛ-əm.\\
             dog.\textsc{obl} \textsc{1sg.dr} \textsc{pfv}{}-eat.\textsc{pst-1sg}  \\
        \glt ‘The dog bit me.’
    \ex\label{ex:pas:47b}
        \gll Spi zə wə-nə-xwaṛ-əm.\\
          dog.\textsc{obl} \textsc{1sg.dr} \textsc{pfv}{}-\textsc{neg}{}-eat.\textsc{pst-1sg}\\
        \glt ‘The dog did not bite me.’
    \z
\z 

\subsection{Reinforcing negation}\label{sec:pas:4.4}

As already noted, and exemplified in \REF{ex:pas:3b}, movement of the negative particle from its unmarked pre-verbal position to the post-verbal, or utterance-final, position has a function of reinforcing the negation. In periphrastically formed verbal tenses, it is the negative particle along with the following auxiliary that is moved to the utterance-final position, as exemplified in \REF{ex:pas:9c}. The same movement of the negative particle is applied when expressing non-optionality in prohibitive utterances, as seen in \REF{ex:pas:20b}. 

An adverb \textit{bīxi} that can, although rarely, be used affirmatively as an intensifier of e.g., human-propensity adjectives, as in (\ref{ex:pas:48}), with the approximate meaning ‘completely, very’, in order to emphasize the negative truth-value of a clause-level negation, as in (\ref{ex:pas:49}), translatable as ‘not at all’. 

\ea\label{ex:pas:48} 
    \gll Zə bīxi warxatā yəm.\\ 
          \textsc{1sg.nom} completely confused be.\textsc{prs.1sg} \\
    \glt ‘I’m completely confused.’
\z 

\ea\label{ex:pas:49} 
    \gll Xesta bīxi kəli ta wāpas nə rāz-i.\\ 
          Khesta at.all village\textsc{.obl} to back \textsc{neg} come.\textsc{ipfv.prs-3}\\
    \glt ‘Khesta is not coming back to the village at all.’ 
\z 

\subsection{Negation, coordination and complex clauses}\label{sec:pas:4.5}

Negative coordination (connective negation) in Pashto is expressed bisyndetically with a \textit{na … na} ‘neither … nor’ construction, as exemplified with verbal predicates in (\ref{ex:pas:50}), and with non-verbal in (\ref{ex:pas:51}). The negator is in each case placed before the constituent or proposition that is being negated.

\ea\label{ex:pas:50} 
    \gll Pa roja ki saṛē na ḍoḍəy xwar{}-i aw na obə tsk{}-i.\\ 
          in fasting in man \textsc{neg} food eat.\textsc{prs-3} and \textsc{neg} water drink.\textsc{prs-3}\\
    \glt ‘During Ramazan, we neither eat food nor drink water.’
\z 

\ea\label{ex:pas:51} 
    \gll Haɣa na ɣal d\={e}  aw na də ɣal mal d\={e}.  \\ 
    \textsc{3sg.dist.dr} \textsc{neg} thief be.\textsc{prs.3sg} and \textsc{neg} of thief friend be.\textsc{prs.3sg.m}  \\
    \glt ‘He is neither a thief nor the friend of a thief.’
\z 

The coordinating conjunction \textit{aw,} occurring before the second \textit{na} in (\ref{ex:pas:50}) and (\ref{ex:pas:51}), can also be left out, as in (\ref{ex:pas:52}).

\ea\label{ex:pas:52} 
    \gll Haɣa na uxyār d\={e} na nər.\\ 
          \textsc{3sg.dist.dr} \textsc{neg} wise be.\textsc{prs.3sg.m} \textsc{neg} brave\\
    \glt ‘He is neither wise nor brave.’
\z 

In another bisyndetic construction \textit{na yaw\={a} ze … balki ham} ‘not only … but also’, it is only the first element that is expressed with a negative particle, as in (\ref{ex:pas:53}).

\ea\label{ex:pas:53} 
    \gll Haɣa na yaw\={a}ze yaw xə ust\={a}z d\={e} balki yaw xə ins\={a}n ham d\={e}. \\ 
          \textsc{3sg.dist.dr} \textsc{neg} only one good.\textsc{sg.m} teacher be.\textsc{prs.3sg.m} but one good.\textsc{sg.m} human also be.\textsc{prs.sg.m}\\
    \glt ‘He is not only a good teacher but also a nice person.’
\z 

\subsection{Miscellaneaous aspects of negation}\label{sec:pas:4.6}
Pashto has a frequently used affirmation particle or interrogative tag \textit{kana} ‘isn’t it?’, ‘don’t you?’, ‘won’t we?’ etc. (< \textit{ka na} ‘or not’), as exemplified in (\ref{ex:pas:54}) and \REF[Q]{ex:pas:55Q}. Often the affirmative answer \textit{ao} ‘yes’, is followed by the same particle, as in \REF[A]{ex:pas:55A}.
Intonation further differentiates a request (whether actual or rhetorical) for affirmation and an affirmative answer, corresponding to the intonational difference between statements and questions in general in Pashto.

\ea\label{ex:pas:54} 
    \gll Xob darz-i kana? \\ 
          sleep come.to.you.\textsc{ipfv.prs-3} \textsc{tag}\\
    \glt ‘You feel sleepy, don’t you?’
\z 

\ea\label{ex:pas:55}
    \begin{xlist}[Q: ]\label{ex:pas:55Q}
    \exi{Q: }\label{ex:pas:54Q}
        \gll Kəli ta ba z-u kana?\\
             village.\textsc{obl} to \textsc{fut} go.\textsc{ipfv.prs-1pl} \textsc{tag}\\
        \glt ‘We will go to the village, won’t we?’
    \exi{A: }\label{ex:pas:55A}
        \gll Ao kana.\\
             yes \textsc{tag}\\
        \glt ‘Yes, of course.’
    \end{xlist}
\z 

Pashto can in various ways express negative meanings as understatements, i.e., by means of clauses that are formally affirmative but include phrases indicating “impossibility” or “difficulty”, thus in principle negating the proposition. It may in some cases be a way of politely declining a request without literally – and bluntly – expressing it with a standard negator. In example \REF{ex:pas:56}, it is virtually out of question that the proposition would be realized. 

\ea\label{ex:pas:56} 
    \gll  Haɣio ɡrān{}-a da če rāš{}-i. \\ 
         \textsc{3pl.dist.obl} difficult-\textsc{f} be.\textsc{prs.3sg.f} \textsc{comp} come.\textsc{pfv.prs{}-3} \\
    \glt ‘They might not come.’ / ‘They aren’t coming.’ (lit. ‘It is difficult for them to come’)
\z 

In yet other cases, it is the larger (cultural) context, along with intonation, that triggers a negative reading. In example (\ref{ex:pas:57}), the extremely negative connotation of \textit{xīrən} as ritually unclean and in every way objectionable, implicitly/contextually attributed to a particular person, effectively excludes an affirmative meaning.

\ea\label{ex:pas:57} 
    \gll Haɣa x\={i}rən ba rāš{}-i.\\ 
    \textsc{3sg.dist.dr} dirty \textsc{fut} come.\textsc{pfv.prs{}-3} \\
    \glt ‘That scumbag is \textit{not} coming here.’ (lit. ‘That dirty will come here.’)
\z 

\section{Summary}\label{sec:pas:5}


\begin{table}[t]
\begin{tabularx}{\textwidth}{lQl}
\lsptoprule
Form & Formal characteristics & Function\\
\midrule 
  {\itshape nə=} & Pre-verbal (finite) particle & Standard negation (SN)\\
\tablevspace
{\itshape mə=} & Pre-verbal negative particle of imperatives &  Prohibitive negation\\
\tablevspace
 \textit{nište} & Replacing clause-final existential copula \textit{šte} & Negative existential \\
\tablevspace
 {\itshape na} & Confirming/non-confirming (yes-no system) & Negative reply\\
\tablevspace
 \textit{hīts{}-...nə} & Typical prefixed element of indefinite pronoun or adverb; negative concord with clausal negation & Indefinite pronouns/adverbs\\
\tablevspace
  \textit{na…na…} & Negative elements bisyndetically preceding negated constituents & Negative coordination\\
  \tablevspace
& The default clausal scope narrowed down by prosodic stress & Scope of negation \\
\tablevspace
  & Right-dislocation of clausal negation & Reinforcement\\
\tablevspace
 \textit{nā-, be-, lā} & Numerous, largely idiosyncratic, derivations by means of prefixes & Negative derivation\\
\lspbottomrule
\end{tabularx}
\parbox{\textwidth}{\raggedright\footnotesize
 {The asymmetries observed in the domain of negation are of the  A/Cat/TAM type:} \\
 {a) Constructional asymmetry in periphrastic verbal categories;} \\
 {b) Paradigmatic asymmetry in prohibitives (perfective-imperfective neutralization)}´
 }
\caption{Main properties of Pashto negation}
\label{tab:pas:3}
\end{table}


\tabref{tab:pas:3} is a summary of the most important observations regarding negation in Pashto.
Standard negation is expressed with a single negative particle \textit{nə}, typically added pre-verbally. When negating clauses in non-periphrastic verbal categories there are no accompanying structural differences between affirmatives and negatives. All such clauses can be negated solely by adding \textit{nə}, regardless of TAM or nominal grammatical categories. Periphrastically formed categories, however, show constructional asymmetry in applying the order \textsc{verb+aux} in affirmative clauses, while having the order \textsc{neg+aux+verb} in negative clauses. For imperatives, there is a dedicated negative particle \textit{mə}, normally added pre-verbally, constituting the only formal contrast between imperatives and corresponding prohibitives. Pashto shows paradigmatic asymmetry in that positive imperatives can make an aspectual distinction between perfective and imperfective, whereas negative imperatives (normally) only occur in the imperfective. Most stative predications are negated by means of standard negation too. The inflected copula used in the affirmative is prefixed by \textit{nə} in the corresponding negative predications. Existential, most possessive and a few locative predications, instead make use of a non-inflected existential copula, \textit{šte.} In the corresponding negated predications, the existential copula is replaced with an existential negator, \textit{nište}.
No structural differences between negation in main clauses and negation in non-main clauses have been observed. Negative replies to polar questions can be given by using a one-word negative particle \textit{na}. Its use, as non-confirming and confirming answers to questions with positive and negative polarity, respectively, suggests that Pashto applies a yes-no answer system. Indefinite/interrogative pronouns and adverbs are typically negated by a preceding (or prefixed) \textit{h\={i}ts}. Clauses with such negated pronouns or adverbs always display negative concord. Negation in Pashto is normally reinforced by changes in the position of the negative element. In main clauses in the non-periphrastic categories, this is obtained by placing the negative particle, \textit{nə} or \textit{mə}, in utterance-final position. In periphrastically formed categories, reinforcement is instead exercised by the occurrence of \textsc{neg+aux} post-verbally vis-à-vis its typical pre-verbal position. Pashto clausal negation typically takes scope over the proposition as a whole, but can be narrowed down to a lower-level constituent by means of prosodic stress.

Pashto gives evidence of a few other specific negative constructions, including a bisyndetic negative coordination, and a small number of negative – primarily adjectival – derivations are productively applied, all of which are prefixal.

\section*{Acknowledgements}

This work is part of the project \textit{Language contact and relatedness in the Hindukush region}, supported by Vetenskapsrådet, the Swedish Research Council, grant 421-2014-631.

\section*{Abbreviations}

\begin{tabularx}{.45\textwidth}{lQ}
{\scshape 1} & first person\\
{\scshape 2} & second person\\
{\scshape 3} & third person\\
\textsc{adj} & adjective \\
{\scshape aux} & auxiliary \\
{\scshape clt} & clitical element \\
{\scshape comp} & complementizer \\
{\scshape dr} & direct case \\
{\scshape dist} & distal \\
{\scshape ex} & existential copula \\
{\scshape f} & feminine \\
{\scshape fut} & future \\
{\scshape imp} & imperative \\
{\scshape inf} & infinitive \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
{\scshape ipfv} & imperfective \\
{\scshape m} & masculine \\
N & noun \\
\textsc{neg} & negation\\
\textsc{negex} & negative existential \\
{\scshape obl} & oblique case\\
{\scshape opt} & optative\\
{\scshape pfv} & perfective\\
{\scshape pl} & plural\\
{\scshape poss} & possessive\\
{\scshape prs} & present\\
{\scshape pst} & past\\
{\scshape ptcp} & participle\\
{\scshape sg} & singular\\
{\scshape tag} & tag question\\
\end{tabularx}

\sloppy\printbibliography[heading=subbibliography,notkeyword=this]

\end{document} 
