\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.17788121}
\author{Matti Miestamo\affiliation{University of Helsinki}}
\title{Questionnaire for describing the negation system of a language}
% \abstract{\noabstract}
\abstract{[revised August 2016, small corrections August 2018]\newline
[February 2019, revisions done together with Ljuba Veselinova]}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 
%\shorttitlerunninghead{Negation questionnaire}% abridged title in page headers%HC see with M&L later on if we want that
\label{chapter:questionnaire}
\section*{General remarks and instructions}
This questionnaire\footnote{As explained in the introduction to this work, the questionnaire is reproduced here in the version in which it was distributed to the authors in early 2019. The only changes are the formatting following the LSP style sheet and some very small corrections. These are mainly limited to correcting typographic errors, and changes to the content are limited to a couple of instances of replacing a word with a more fitting one and making a necessary clarification to clause structure. The history and the underlying logic of the questionnaire is discussed in the introduction. This 2019 version results from revisions done together with Ljuba Veselinova and her input is gratefully recognized.} explicates the different aspects of negation that should be covered in the description of the domain of negation in a language. To keep the length of the questionnaire in reasonable proportions, the questionnaire itself does not explain very deeply the typology of the different subdomains of negation, or give examples, and typological literature is meant to be consulted while working with the questionnaire. Some pointers to relevant literature are given in the questionnaire, but this is not done systematically, and the user of the questionnaire is advised to make use of typological overviews such as \citealt{Miestamo2007,Miestamo2017}, and further typological literature cited in these.

The questionnaire starts from function and asks what the formal constructions expressing these functions are in each language. %HC in my opinion it would look better to have one bigger paragraph instead of these short sentences

Note that depending on the language, some forms and constructions may fall under more than one subsections. 

In Sections \sectref{sec:quest:2} Clausal negation and \sectref{sec:quest:3} Non-clausal negation, describe all the different constructions used to express negation in the language, paying attention to:

\begin{itemize}
    \item Negative marker(s) (see \citealt{Dahl1979}; \citealt{Payne1985}; \citealt{Dryer2005-wals112}, \citeyear{Dryer2011-wals143,Dryer2011-wals144}): 
    \begin{itemize}
        \item type: particle, clitic, affix, verb, noun, … 
        \item position: preverbal, postverbal, clause-initial, clause-final…
        \item number: one marker, two markers (discontinous), or even more…
    \end{itemize}
    \item Structural differences between positives vs. negatives (see \citealt[51ff]{Miestamo2005}, \citeyear{Miestamo2007,Miestamo2017})
    \begin{itemize}
        \item Are negative markers simply added to a corresponding positive, or does the structure of the clause differ from the affirmative in other ways, too (constructional asymmetry)? Describe the structural differences.
        \item Are the same grammatical categories available in the negative as well, or are some distinctions that are made in the affirmative lost in the negative (paradigmatic asymmetry)? Describe the differences in the paradigms available in the negative vs. affirmative.
    \end{itemize}
    \item What are the more specific functions of these negative constructions –which environments are they used in, i.e. what do they negate?
    \begin{itemize}
        \item Note specifically which categories/environments use the same negative construction.
    \end{itemize}
\end{itemize}

Languages often have different negative constructions for negation in different environments (clausal negation, different clause types, constituent negation, negative indefinites, etc.). In Sections \sectref{sec:quest:2} and \sectref{sec:quest:3}, the different negative constructions should be described: clausal negation in \sectref{sec:quest:2} and non-clausal negation in section \sectref{sec:quest:3}. The subsections deal with different clause types and environments, in which negation may show dedicated constructions, different from the negation of other clause types / environments. 

In the description of each negative construction, please take into account the points discussed above (negative markers, structural (a)symmetry, functions). Please also give illustrative examples (both the negative and its non-negative counterpart whenever possible). 

\section{The language}\label{sec:quest:1}

Give a general characterization of the language in terms of geography, genealogy, contacts, sociolinguistic status, dialectal variation etc. Please, provide also the ISO and Glottolog codes identifying the language.

%[For the papers proceeding from the SWL8 Negation workshop in Paris in September 2018, the length of this section should be 1--1.5 pages, i.e. up to 750 words.]

\section{Clausal negation}\label{sec:quest:2}
\subsection{Standard negation}\label{sec:quest:2.1}

Standard negation (SN) refers to the (basic) way(s) a language has for negating declarative verbal main clauses (see especially \citealt{Payne1985}; \citealt{Miestamo2005}). For example, in Spanish \REF{ex:quest:1}, SN is expressed by a construction in which the negative marker \textit{no} precedes the verb. In Finnish \REF{ex:quest:2}, standard negation is expressed by a construction in which the negative auxiliary \textit{e-} appears as the finite element of the sentence, carrying the verbal person-number markers, and the lexical verb is in a non-finite form (uninflected present connegative in the present and past participle in the past tense).

\ea\label{ex:quest:1}Spanish (constructed examples)
    \ea\label{ex:quest:1a}
        \gll El perro está ladrando. \\ 
             \textsc{def} dog be.\textsc{3sg} bark.\textsc{ptcp} \\
        \glt ‘The dog is barking.’ 
    \ex\label{ex:quest:1b}
        \gll El perro no está ladrando. \\
             \textsc{def} dog \textsc{neg} be.\textsc{3sg} bark.\textsc{ptcp}\\
        \glt ‘The dog is not barking.’
    \ex\label{ex:quest:1c}
        \gll El perro ladró. \\ 
             \textsc{def} dog bark.\textsc{pst.3sg} \\
        \glt ‘The dog barked.’
    \ex\label{ex:quest:1d}
        \gll El perro no ladró. \\
             \textsc{def} dog \textsc{neg} bark.\textsc{pst.3sg} \\
        \glt ‘The dog did not bark.’
    \z
\z 

\ea\label{ex:quest:2}Finnish (constructed examples)
    \ea\label{ex:quest:2a}
        \gll Koira haukku-u. \\ 
             dog bark-3\textsc{sg} \\
        \glt ‘The dog is barking.’
    \ex\label{ex:quest:2b}
        \gll Koira ei hauku. \\
             dog \textsc{neg.3sg} bark.\textsc{cng.prs} \\
        \glt ‘The dog is not barking.’
    \ex\label{ex:quest:2c}
        \gll Koira haukku-i. \\ 
             dog bark-\textsc{pst.}3\textsc{sg} \\
        \glt ‘The dog barked.’
    \ex\label{ex:quest:2d}
        \gll Koira ei haukku-nut. \\
             dog \textsc{neg.3sg} bark-\textsc{pst.ptcp.sg} \\
        \glt ‘The dog did not bark.’
    \z
\z 

As explained above, pay attention to the type and position of the negative marker(s) as well as to any structural differences between the negatives and the corresponding affirmatives. Are negative markers simply added to a corresponding positive, or does the structure of the clause differ from the affirmative in other ways, too (constructional asymmetry)? Are the same grammatical categories available in the negative as well, or are some distinctions made in the affirmative lost in the negative (paradigmatic asymmetry)? Which types of asymmetry identified by \citet{Miestamo2005} do these differences instantiate? For the relationship between aspectual categories and negation, see also \citet{MiestamoAuwera2011}.

Languages may have different standard negation constructions, e.g., in different TAM categories, in different person/number/gender categories, etc. For example in Komi, the present and the past use negative verb constructions with a different stem of the negative verb (\ref{ex:quest:3a}--\ref{ex:quest:3d}), and the perfect and the pluperfect use a completely different construction with a negative particle  (\ref{ex:quest:3e}--\ref{ex:quest:3h}). 

\ea\label{ex:quest:3} Komi-Zyrian \citep[105--109]{Rédei1978}
    \ea\label{ex:quest:3a}
        \gll Śet-e̮.\\ 
             give-\textsc{3sg.prs} \\
        \glt ‘(S)he gives.’ 
    \ex\label{ex:quest:3b}
        \gll O-z śet. \\
             \textsc{neg}{}-3 give \\
        \glt ‘(S)he does not give.’
     \ex\label{ex:quest:3c}
        \gll Śet-i-s. \\ 
             give-\textsc{pret-3sg} \\
        \glt ‘(S)he gave.’ 
    \ex\label{ex:quest:3d}
        \gll E-z śet. \\
             \textsc{neg.pret}{}-3 give\\
        \glt ‘(S)he did not give.’
    \ex\label{ex:quest:3e}
        \gll Śet-e̮m-a. \\ 
             give-\textsc{perf-3sg} \\
        \glt ‘(S)he has given.’ 
    \ex\label{ex:quest:3f}
        \gll Abu śet-e̮m-a. \\
             \textsc{neg} give-\textsc{perf-3sg}\\
        \glt ‘(S)he has not given.’
    \ex\label{ex:quest:3g}
        \gll Śet-e̮m-a ve̮li.\\ 
             give-\textsc{perf-3sg} be.\textsc{pret.3sg} \\
        \glt ‘(S)he had given.’ 
    \ex\label{ex:quest:3h}
        \gll Abu śet-e̮m-a ve̮li. \\
             \textsc{neg} give-\textsc{perf-3sg} be.\textsc{pret.3sg}\\
        \glt ‘(S)he had not given.’
    \z
\z  

Describe all the different constructions used in different standard negation environments (declarative verbal main clauses). One temporal meaning that very often has its own dedicated negative construction are the so-called ‘not yet’  or \textit{nondum} tenses (\citealt[54]{Comrie1985}; \citealt{Veselinova2017}), and special attention should be paid to the existence of such negative tenses. 

Languages may have clausal negation constructions differing from standard negation in different clause types. There may be a dedicated negative construction for imperatives, non-verbal clauses, etc. Sections \sectref{sec:quest:2.2}--\ref{sec:quest:2.4} should describe negation in each of these clause types. In case a clause type does not have a dedicated negative construction, it should be briefly noted which negative construction, already described, is used to negate it, giving also an illustrative example.

\subsection{Negation in non-declaratives}\label{sec:quest:2.2}

Do imperatives have a dedicated negative construction different from standard negation (or show special behaviour with respect to standard negation in some way at least)? Describe the dedicated construction (special behaviour) according to the instruction given above (neg-markers, (a)symmetry, functions) and give examples. You may also characterize negative imperatives in terms of the typological classification proposed by \citet{AuweraEtAl2013}; for differences between negative and positive imperatives, see also \citealt{MiestamoAuwera2007} and \citealt{Aikhenvald2010}. 


What about other non-declaratives (questions, (other) non-declarative mood categories)? If there are any special constructions, describe them here, too. For negative questions, it is especially interesting to comment on their function/use – e.g., is there an expectation of a positive or negative answer or are they neutral in this respect? 

\subsection{Negation in stative predications}\label{sec:quest:2.3}

Stative predications (aka non-verbal predications) can be divided into different types following \citet[111ff]{Payne1997}, see also \citet{Veselinova2014}:

\begin{itemize}
    \item equation, e.g., \textit{She is my mother. – She is not my mother.}
    \item proper inclusion, e.g., \textit{Kurumaku is a hunter. – Kurumaku is not a hunter.}
    \item attribution, e.g, \textit{She is intelligent}. – \textit{She is not intelligent}.
    \item locative predication, e.g., \textit{The cat is on the mat. – The cat is not on the mat.}
    \item existential predication, e.g., \textit{There are wild cats. – There are no wild cats.}
    \item possessive predication, e.g., \textit{Tom has a car}. – \textit{Tom doesn’t have a car}.
\end{itemize}

As \citet{Veselinova2015} notes, attribution can be further divided into permanent and temporary property assignment: \textit{Luna is tall} vs. \textit{Luna is happy}, and under existential predication, we can further identify locative-presentative constructions such as \textit{There are wild cats in Africa}. 


Note that the distinction between existentials and locatives lies in that locatives have a definite subject and they do not introduce new participants in the discourse, whereas existentials do introduce new participants. The subjects of existential predications are generic; they typically appear in bare form or if marked, their marking will be different from that of subjects in other kinds of predications, cf. also \citet{Dryer2007}, \citet{Veselinova2013lexicalization}, \citet{McNally2016}. 


How are these clause types negated? Do any of these clause types have a dedicated negative construction different from standard negation or from the other non-verbal clause types (or show special behaviour with respect to standard negation or the other non-verbal clause types in some way at least)? Which types are negated similarily and which ones differently? Is there a difference between temporary and permanent property assignment as regards negation? Do generic vs. specific predicates behave differently as regards negation? Describe the dedicated construction (special behaviour) according to the instruction given above (neg-markers, (a)symmetry, functions) and give examples.

\subsection{Negation in non-main clauses}\label{sec:quest:2.4}

How is negation in dependent/subordinate clauses expressed – standard negation or dedicated constructions? Describe the constructions according to the above instructions. Pay attention to both finite and non-finite dependent clauses. Can non-finite clauses be negated? Are they negated with the standard negator, a special negator or are there special negative non-finite forms?


In Finnish, for example \REF{ex:quest:4}, finite subordinate clauses are negated by standard negation but non-finite dependent clauses such as \REF{ex:quest:4c} cannot be negated. In English, on the other hand, the equivalent of \REF{ex:quest:4d} ‘I saw her/him not come’ would be grammatical. %HC I deleted the dot in the end of the example sentence here, it looked a bit weird

\ea\label{ex:quest:4}Finnish (constructed examples)
    \ea[]{\label{ex:quest:4a}
        \gll Näin että hän tule-e. \\ 
             see.\textsc{pst.1sg} that \textsc{3sg} come-\textsc{3sg} \\
        \glt ‘I saw that (s)he’s coming.’}
    \ex[]{\label{ex:quest:4b}
        \gll Näin että hän ei tule. \\
             see.\textsc{pst.1sg} that \textsc{3sg} \textsc{neg.3sg} come.\textsc{cng.prs} \\
        \glt ‘I saw that (s)he’s not coming.’}
    \ex[]{\label{ex:quest:4c}
        \gll Näin häne-n tule-va-n. \\ 
             see.\textsc{pst.1sg} \textsc{3sg-gen} come-\textsc{ptcp.prs-gen} \\
        \glt ‘I saw her/him come.’}
    \ex[*]{\label{ex:quest:4d}
        \gll Näin häne-n ei tule-va-n. \\
             see.\textsc{pst.1sg} \textsc{3sg-gen} \textsc{neg} come-\textsc{ptcp.prs-gen}\\
        \glt ‘I saw her/him not come.’}
    \z
\z 

\subsection{Negative lexicalizations}\label{sec:quest:2.5}

Lexical meanings may combine with negation to form lexically idiosyncratic negatives. Cross-linguistically common lexicalizations include ‘not yet’, ‘not exist’, ‘not be of identity', ‘not know', ‘cannot, not be able to', ‘not want', ‘not talk', ‘not need', ‘not enough', ‘not like', and ‘not see' \citep[cf.][]{Veselinova2013neg.exist}. Describe the form and behaviour of any such items in the language. 


Note that some of these may be better treated in other sections, e.g. ‘not exist’ and ‘not be of identity’ under stative predications, and ‘not yet’ under standard negation in case it can be seen as part of the tense system rather than as a more independent lexical item.

\subsection{Other clausal negation constructions}\label{sec:quest:2.6}

If there are other clausal negation constructions, not covered in sections \sectref{sec:quest:2.1}--\sectref{sec:quest:2.5}, section \sectref{sec:quest:2.6} can be added to discuss them. One possible example would be clausal negation constructions appearing in special pragmatic contexts or showing special pragmatic effects. 


\section{Non-clausal negation}\label{sec:quest:3}

This section deals with constructions/elements expressing negation other than clausal negation.

\subsection{Negative replies}\label{sec:quest:3.1}

How are negative replies to polar questions expressed? Are there one-word negative replies like English \textit{no} ? Relate them to the corresponding affirmative replies.

What is the semantics of negative replies – do they disagree with the content or with the polarity of the question, i.e. do both of the replies in (\ref{ex:quest:5}--\ref{ex:quest:6}) mean that the dog is not barking or does the latter mean that the dog is barking?

% \medskip\hc{should that be formatted as an example? See my proposition right below}
% \textit{Is the dog barking? – No!}

% \textit{Isn’t the dog barking – No!}
% \medskip

\ea\label{ex:quest:5}
    \begin{xlist}
    \exi{A: }
        \glt Is the dog barking?
    \exi{B: }
        \glt No!
    \end{xlist}
\z 
\ea\label{ex:quest:6}
    \begin{xlist}
    \exi{A: }
        \glt Isn't the dog barking?
    \exi{B: }
        \glt No!
    \end{xlist}
\z 

Particles that disagree with the polarity of the question are called polarity-reversing particles by \citet{Holmberg2015}. The Guaraní particle \textit{tove} is a polarity-reversing particle that can occur as reply to both positive and negative questions, while French \textit{si} or German \textit{doch} can only function as positive answers to negative questions. Please comment also on (the semantics of) affirmative replies to negative questions. 

\subsection{Negative indefinites and quantifiers}\label{sec:quest:3.2}

Describe the negation of indefinite pronouns and adverbs in the language, e.g. \textit{nobody, no-one, nowhere, never, none, no; anybody, anyone, anywhere, ever, one, any}. 

\begin{itemize}
    \item How are these related to indefinites in non-negatives? What is the range of use of these indefinites in non-negative contexts (e.g., English \textit{nobody} is negative only but \textit{anybody} has non-negative uses as well); these functions can be described in terms of the semantic map proposed by \citet{Haspelmath1997}, cf. also \citet{Alsenoy2014}.
    \item How are they used in clauses: Are they used together with clausal negation or not (English \textit{I saw nobody} vs. \textit{I didn’t see anybody})? \citep[cf.][]{Haspelmath2005}. In which positions can they occur in the clause?
    \item In case the language does not have indefinites or cannot use them in negatives, how are the equivalent meanings expressed, e.g., ‘I didn’t see anybody’, ‘Nobody came’, ‘The dog never barked’, ‘You didn’t go anywhere’? 
    \item Note that this subsection is related to negative polarity which is also a topic in \sectref{sec:quest:4} below. Try to find a balance between what is treated here and what is treated in \sectref{sec:quest:4}.
\end{itemize}

\subsection{Negative derivation and case marking (abessives/caritives/privatives)}\label{sec:quest:3.3}

How are meanings of absence, opposition and antonymy, such as ‘without’, ‘\nobreakdash-less’, or ‘un-', expressed on the lexical/phrasal level, e.g., \textit{without a book, bookless, unread, unreadable}?

\begin{itemize}
    \item Do verbs, nouns and adjectives behave similarly or do they have different markers?
    \item Are the markers adpositions, inflectional case markers, or derivation; if it is a primarily nominal marker, how does it combine with verbs?
    \item If the language has several of these, what is their division of labour, i.e. which functions does each marker express and what is its distribution?
    \item NB! If these markers are used in clausal negation constructions, these functions should be described in \sectref{sec:quest:2}.
\end{itemize}

\subsection{Other negative constructions/expressions}\label{sec:quest:3.4}

Describe and illustrate any other negative constructions/expressions that are not covered above. E.g. negative coordinators such as English \textit{neither... nor.}

\section{Other aspects of negation}\label{sec:quest:4}

This section pays attention to various morphosyntactic, semantic and pragmatic phenomena that are not negative constructions/expressions themselves, but arise in connection with negation.

Note that some of the topics overlap with each other or with points raised in Sections \sectref{sec:quest:2}--\ref{sec:quest:3}. Please give careful thought to how to relate the topics in different sections to each other so that the same information is not repeated but that the readers can easily see the connections.

\subsection{The scope of negation}\label{sec:quest:4.1}
How is the scope of negation narrowed to a specific constituent (e.g., Foc Neg-Verb vs. Neg-Foc Verb)? What is the role of intonation and stress in coding the scope of negation? Prosodic prominence may be indicated, e.g., by underlining.

Scope-related questions more generally? Note that examples discussed under different topics of the questionnaire may be discussed here in terms of their scope properties.

\subsection{Negative polarity}\label{sec:quest:4.2}

List negative polarity items, their form/meaning/use (licensing conditions). Note that this overlaps to some extent with section \sectref{sec:quest:3.2}.

\subsection{Marking of NPs in the scope of negation}\label{sec:quest:4.3}

Is case marking affected under negation (e.g., partitive/genitive objects or subjects)? Any other effects negation might have on the marking of NPs, such as change or loss of determiners, effects on the marking of focus, etc? Cf. \citet{Miestamo2014}.

\subsection{Reinforcing negation}\label{sec:quest:4.4}

Describe and illustrate the means used for reinforcing negation. Prosody should also be paid attention to here. To the extent that emphatic negatives involve elements forming separate negative constructions/expressions, they can also be treated or at least mentioned at relevant points in Sections \sectref{sec:quest:2} or \sectref{sec:quest:3}.

\subsection{Negation, coordination and complex clauses}\label{sec:quest:4.5}

How is the coordination of positive+negative clause or two negative clauses expressed? Are there special negative coordinators, such as \textit{neither, nor} ? Are there special constructions for contrastive negation (e.g. \textit{This is a dog, not a cat}; \textit{This is not a cat but a dog}); this is also related to the scope of negation and can be taken up in \sectref{sec:quest:4.1} as well?

What about subordination? Are there negative conjunctions, such as \textit{lest}? Note that this is related to section \sectref{sec:quest:2.4}.

To the extent that the negative coordinators and conjunctions are elements forming separate negative constructions/expressions, they can also be treated or at least mentioned at relevant points in Sections \sectref{sec:quest:2} or \sectref{sec:quest:3}.

\subsection{Miscellaneous aspects of negation}\label{sec:quest:4.6}

Here you can take up further aspects of negation to the extent that you have been able to find data on them. It is advisable to try to cover as many as possible of the phenomena that have been given a separate subsection below. Any other aspects of negation can also be addressed in case they are of special interest in the language described. 

\subsubsection{Negative transport}\label{sec:quest:4.6.1}

Negative transport (neg-raising) means that a higher-clause negative is interpreted as negating a lower-clause predicate, e.g. \textit{I don’t think they’re coming} meaning \textit{I think they’re not coming} (see \citealt{Horn1978,Horn1989}). Does neg-transport occur? Which predicates allow it and which ones do not?


\subsubsection{Metalinguistic negation}\label{sec:quest:4.6.2}

Metalinguistic negation means that what is negated is not the content of the proposition but rather the way it is expressed – “a device for objecting to a previous utterance on any grounds whatever, including the conventional or conversational implicata it potentially induces, its morphology, its style or register, or its phonetic realization.” \citep[363]{Horn1989} 

\begin{itemize}
    \item Some examples of metalinguistic negation (from \citealt{Horn1989}):
        \begin{itemize}
            \item \itshape He doesn’t have three children, he has four.
            \item \itshape Around here we don’t like coffee – we love it.
            \item \itshape He didn’t call the [pólis], he called the [polís]
            \item \itshape Phydeaux didn’t shit the rug, he soiled the carpet.
        \end{itemize}
    \item Metalinguistic negation may lead to different behaviour of negative polarity items:
        \begin{itemize}
            \item \textit{John didn’t manage to solve \{some/*any\} problems – they were quite easy for him to do.} \citep[130]{Horn1985}
        \end{itemize}
\end{itemize}

How does the language treat metalinguistic negation? Does it show special behaviour different from ordinary (“descriptive”) negation? How does metalinguistic negation relate to contrastive negation, which may or may not be metalinguistic?

\subsubsection{Non-negative uses of negatives and expletive negation}\label{sec:quest:4.6.3}
Are there any interesting non-negative uses of negative constructions or constructions resembling negative constructions in the language? One particular phenomenon to pay attention to here is the so-called expletive negation in complements to predicates such as ‘fear’ or ‘doubt’, e.g. in French \textit{Je crains qu’elle} \textbf{\textit{ne}} \textit{vienne.} ‘I’m afraid she will come’, or in subordinate clauses introduced by conjunctions meaning ‘until’ or before’, e.g. in French \textit{...avant qu’il} \textbf{\textit{ne}} \textit{vienne.} ‘...before he comes.’ If this phenomenon is found in the language, which contexts trigger it?

\subsubsection{Summary of what can be negated in the language}\label{sec:quest:4.6.4}

According to the instructions at the beginning of this questionnaire, the description of each negative construction should pay attention to whether some grammatical categories are lost in the negative. Here, you can still come back to the question of what is negatable in the language? Can, e.g., quantifiers be negated – which ones can and which ones cannot? What about clauses with indefinite subjects?

\subsubsection{Diachronic notes and observations}\label{sec:quest:4.6.5}

The questionnaire aims at a synchronic description of negation in a language. Notes on diachronic issues, such as Jespersen Cycles \citep{Auwera2009,Auwera2010} or the Negative Existential Cycle (\citealt{Croft1991}; \citealt{Veselinova2014,Veselinova2016}), may be made here in this final section or in the relevant sections that they concern.

\subsubsection{Any other phenomena}\label{sec:quest:4.6.6}

Any other phenomena that should be taken into account in describing the system of negation in your language but that has not been covered above? Anything that is of special interest in the negation system of the language but does not fit under any of the headings above.


\section*{Abbreviations}
\begin{tabularx}{.45\textwidth}{lQ}
1 & first person \\
3 & third person \\
\scshape cng & connegative \\
\scshape def & definite \\
\scshape gen & genitive \\
\scshape neg & negative \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape perf & perfect \\
\scshape prs & present \\
\scshape pret & preterite \\
\scshape pst& past \\
\scshape ptcp & participle \\
\scshape sg & singular \\
\end{tabularx}

\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
\end{document} 


