\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233981}
\author{Rachel Weymuth\affiliation{}}
\title{Negation in Rumai Palaung}
\abstract{In Rumai, a variety of Palaung (Palaungic, Austroasiatic), negation is dependent on the clause type. While the negative morphemes \textit{ʔaːw-} ‘\textsc{neg}’, \textit{buː-} ‘\textsc{neg}’, \textit{{}-maʔ} ‘\textsc{neg}’ and \textit{ɲjʌm} ‘not yet’ can be used in independent clauses, dependent clauses are restricted to \textit{buː-} and \textit{ɲjʌm}. This is probably due to the constituent order change from VAP to AVP in the Rumai independent clauses. Additionally, there are the prohibitives \textit{kʰuː{}-} and \textit{maj-}. The aim of this chapter is to describe and to analyze the means of negation in Rumai. It follows the “Questionnaire for describing the negation system of a language” by \textcitetv{chapters/questionnaire}.
\keywords{Austroasiatic, Palaung, Rumai, negation, Jespersen Cycle}
}
\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 

\section{The language} \label{sec:rum:1}

Rumai Palaung (ISO: \href{https://iso639-3.sil.org/code/rbb}{rbb}, Glottocode: \href{https://glottolog.org/resource/languoid/id/ruma1248}{ruma1248}), an Austroasiatic language of the western Palaungic branch, is one of three varieties of Palaung (Ta’ang) so far mentioned in the linguistic literature, the other two being Shwe Palaung and Ruching Palaung. The three varieties (called in the following simply Rumai, Shwe and Ruching) differ to a considerable extent in the grammatical morphemes, and, somewhat less, in the lexicon.

Rumai, with a population of approximately 140,000, is mainly spoken in Myanmar in the townships Nam Hkam and Man Ton in Northern Shan State and Mogok in Mandalay Region, but also in adjacent regions of the Chinese province Yunnan. For a long time, Shan (Tai-Kadai) has been the main contact language, but with the introduction of the public school system after the independence of Myanmar in 1948, Burmese (Tibeto-Burman) gained more influence. This means that, today the sociolinguistic hierarchy in the region is Burmese > Shan > Rumai. In Northern Shan State, the Rumai live in villages scattered over mountain ridges and slopes. The Rumai speakers in the villages in the same valley have relatively easy contact with each other, but this is not the case for villages in different valleys. This results in several dialects, which slightly diverge in phonology, lexicon, and syntax. The present study is based on fieldwork by the author mainly conducted in Mandalay, between 2013 and 2020, with consultants from Nam Hkam and Man Ton townships, in a village in Nam Hkam township and in Lashio, the informal capital of Northern Shan State. The corpus includes elicited sentences from several questionnaires, interviews, conversations, picture stories, and published texts from a phrasebook, journals, and a booklet with short stories. The references to the sources are explained in the appendix at the end of the chapter.

The Palaung have their own script, which was invented in 1972 by the Ta’ang Literature and Culture Committee, and in 2012 a dictionary Burmese-Shwe-Rumai-Ruching appeared. There is also a dictionary Rumai-English-Burmese from 2011. The script is based on the Burmese script with some additions from the Shan script. However, as teaching of (and even speaking) ethnic languages in public schools was forbidden for a long time, literacy in Palaung is still low and a real standardization of the script has not taken place. Moreover, Ruching communities in Southern Shan State use Khün script, which is mainly used for Tai languages in this region.

Rumai constituent order is mainly SV/AVP in main clauses while dependent clauses have predominantly VS/VAP order. Multi-verbal predicates are common, and although the language is mainly analytic, there are verbal affixes expressing aspect, modality, and negation. The multi-verbal predicates include preverbal and postverbal secondary verbs expressing several grammatical categories which partly coincide with the functions of the verbal prefixes.\footnote{For the affixes and secondary verbs in Rumai see \citealt{Weymuth2018} and \citeyear{Weymuth2023}.}

The description of negation in Rumai in this chapter is based on the questionnaire provided by the editors (\citetv{chapters/questionnaire}) and mirrors its structure.


\section{Clausal negation} \label{sec:rum:2}

The way clausal negation is expressed in Rumai is partly connected to constituent order. 
Negation is predominantly preverbal, but there is also the postverbal negator \textit{{}-maʔ}, which occurs only in independent declarative (SV/AVP) clauses. 
This conforms to the findings of \citet[97]{Dryer1988} that postverbal negation in verb initial languages is highly unusual.
It was originally a reinforcement of negation, but today it also occurs as the sole negator of a verb.\footnote{See \sectref{sec:rum:4.6.2} for Jespersen Cycle.} 

\subsection{Standard negation} \label{sec:rum:2.1}

The negators used in Rumai standard negation are the prefixes \textit{ʔaːw{}-} and \textit{buː{}-}, the suffix \textit{{}-maʔ}, and the preverbal negative particle \textit{ɲjʌm} ‘not yet’. \textit{ɲjʌm} is not an affix like the other negators, but a free morpheme as it can be used alone in the negative answer to a polar question (\sectref{sec:rum:3.1}).\footnote{Due to the lack of data, \textit{ɲjʌm} was analyzed as a prefix in \citet{Weymuth2018}.}
\textit{ʔaːw-} and \textit{{}-maʔ} only occur in independent clauses while \textit{buː-} and \textit{ɲjʌm} are also used in dependent ones.

In independent clauses, the preverbal affixes usually occur together with the suffix \textit{{}-maʔ} (\ref{ex:rum:1b}\footnote{The transcription \textit{eə} in words like \textit{təkjeət} ‘tidy up’ in example \REF{ex:rum:1} does not represent a diphthong in Rumai, but a vowel between /e/ and /ɛ/.}, \ref{ex:rum:3b}), but there are also examples where the prefixes are the only negator (\ref{ex:rum:2b}\footnote{In the paper about secondary verbs in Rumai \citep{Weymuth2023}, \textit{pʌŋ} is glossed as a nominalizer. However, to analyze \textit{pʌŋ} as a relativizer is more compatible with the structure of the language.}, \ref{ex:rum:4b}).

\ea\label{ex:rum:1}
    \ea\label{ex:rum:1a}
        \gll pi	ʔʌ̂w	təkjeət	pʌʔ	sɔŋfâj \\ 
             elder.sibling 1\textsc{sg} tidy.up \textsc{loc} kitchen \\
        \glt ‘My sister cleans up in the kitchen.’ (ENEG\_13\_M\_1/2\_115)
    \ex\label{ex:rum:1b}
        \gll pi	ʔʌ̂w	\textbf{ʔaːw}-təkjeət-\textbf{maʔ}	pʌʔ	sɔŋfâj \\
             elder.sibling \textsc{1sg} \textsc{neg}{}-tidy.up-\textsc{neg} \textsc{loc} kitchen \\
        \glt ‘My sister does not clean up in the kitchen.’ (ENEG\_13\_M\_1/2\_117)
    \z
\z 

\ea\label{ex:rum:2}
    \ea\label{ex:rum:2a}
        \gll ʔʌ̂w	jêəw	təɲâːw	pʌŋ	dʌh	mâj \\ 
             1\textsc{sg} see listen \textsc{rel} say 2\textsc{sg} \\
        \glt  ‘I hear what you say.’ (pc\_3)
    \ex\label{ex:rum:2b}
        \gll ʔʌ̂w	\textbf{ʔaːw}-jêəw	təɲâːw	pʌŋ	dʌh	mâj \\
             1\textsc{sg} \textsc{neg}{}-see listen \textsc{rel} say 2\textsc{sg} \\
        \glt  ‘I don’t hear what you say.’ (EQB\_17\_M\_3\_023)
    \z
\z 

\textit{buː{}-} is the main negator in dependent clauses (\sectref{sec:rum:2.4}), but it is also used in independent ones \REF{ex:rum:3} and \REF{ex:rum:4}.

\ea\label{ex:rum:3}
    \ea\label{ex:rum:3a}
        \gll paːt	bʌ̂ːn	ʔʌ̂w	hâːw	jangûːn \\ 
              week back 1\textsc{sg} go Yangon \\
        \glt  ‘Next week I will go to Yangon.’ (ENEG\_13\_M\_1/2\_040)
    \ex\label{ex:rum:3b}
        \gll paːt	bʌ̂ːn	ʔʌ̂w	\textbf{buː}-hâːw-\textbf{maʔ}	jangûːn \\
              week back 1\textsc{sg} \textsc{neg}{}-go{}-\textsc{neg} Yangon \\
        \glt  ‘Next week I will not go to Yangon.’ (ENEG\_13\_M\_1/2\_041)
    \z
\z 

\ea\label{ex:rum:4}
    \ea\label{ex:rum:4a}
        \gll ʔɛ̂	nʌ̀	ɟɯ	kɛ̂	dih \\ 
              \textsc{1pl.incl} know name 3\textsc{pl} \textsc{disc} \\
        \glt ‘We know their names.’ (pc\_3)
    \ex\label{ex:rum:4b}
        \gll ʔɛ̂	\textbf{buː}-nʌ̀	ɟɯ	kɛ̂	dih \\
              \textsc{1pl.incl} \textsc{neg}{}-know name 3\textsc{pl} \textsc{disc} \\
        \glt  ‘We don’t know their names.’ (C3\_17\_MS\_9\_107)
    \z
\z 

Besides its function as a neutral negation marker in independent clauses, \textit{buː{}-}, contrary to \textit{ʔaːw-}, may also mean ‘no longer, no more, not anymore’, depending on the context or the intention of the speaker. Here, the negative suffix \textit{{}-maʔ} is always present \REF{ex:rum:5}.

\ea\label{ex:rum:5}
    \gll ʔân	\textbf{buː}-deəh-\textbf{maʔ}	dih\\
         3\textsc{sg} \textsc{neg}{}-come-\textsc{neg} \textsc{disc} \\
    \glt ‘He won’t come anymore.’ (EQB\_17\_M\_3\_006)
\z    

\textit{ʔaːw-} and \textit{buː{}-} (as a neutral negation marker) are sometimes freely interchangeable. However, depending on the intention of the speaker, there may also be the connotation of ‘not anymore, no longer’ in \REF{ex:rum:6b}, but not in \REF{ex:rum:6a}.

\ea\label{ex:rum:6}
    \ea\label{ex:rum:6a}
        \gll mâj	\textbf{ʔaːw}-bɤ̂n-maʔ	ʔʌk \\ 
             2\textsc{sg} \textsc{neg}{}-get-\textsc{neg} sit \\
        \glt ‘You may not sit here.’ (EGGD\_16\_L\_6\_142)
    \ex\label{ex:rum:6b}
        \gll mâj	\textbf{buː}-bɤ̂n-maʔ	ʔʌk \\
             2\textsc{sg} \textsc{neg}{}-get-\textsc{neg} sit \\
        \glt ‘You may not sit here (anymore).’ (EGGD\_16\_M\_1\_141)
    \z
\z 

In combination with the negative prefixes, it is possible that the suffix \textit{{}-maʔ} originally had a reinforcing function, but today it is very frequent in negative declarative clauses, so that it is now rather a neutral negation marker. Sometimes it also occurs as the sole negator in a clause \REF{ex:rum:7b}.\footnote{See \sectref{sec:rum:4.6.2} for Jespersen Cycle.}

\ea\label{ex:rum:7}
    \ea\label{ex:rum:7a}
        \gll ʔʌ̂w	hâːw	pʌʔ	jangûːn \\ 
             1\textsc{sg} go \textsc{loc} Yangon \\
        \glt ‘I go to Yangon.’ (ENEG\_13\_M\_1/2\_034)
    \ex\label{ex:rum:7b}
        \gll ʔʌ̂w	hâːw-\textbf{maʔ} pʌʔ	jangûːn \\
              1\textsc{sg} go-\textsc{neg} \textsc{loc} Yangon \\
        \glt ‘I don’t go to Yangon.’ (ENEG\_13\_M\_1/2\_035)
    \z
\z 

The three Palaung varieties have a common preverbal negative marker with the meaning ‘not yet’, namely Rumai \textit{ɲjʌm}, Shwe \textit{hɲɐm} \citep[176]{Milne1921} and Ruching \textit{hɲam} (\citealt{DeepadungEtAl2015}: 1078). In the Burmese-Palaung dictionary \citep[131]{Unknown2012}, the Shwe and the Ruching expressions are cited as verbs with the meaning ‘dilatory, stiff’ (the corresponding Rumai verb in the dictionary is \textit{moj}). This suggests that a former common verb of the three Palaung varieties is the source of the Rumai negation marker \textit{ɲjʌm}.
Clauses containing the prefix \textit{hɔ̰j}- ‘\textsc{nsit}’  (\ref{ex:rum:8a}, \ref{ex:rum:9b}), the secondary verb \textit{hɔ̀ːj} ‘finish’ (\ref{ex:rum:9a}), or both (\ref{ex:rum:10a}) are usually negated by \textit{ɲjʌm}.\footnote{For more details about the verbal affixes see \citet{Weymuth2018}, and for the secondary verbs see \citet{Weymuth2023}.}

The prefix and the secondary verb are aspectual markers which denote a new (and usually expected) situation after the transgression of a limit \citep[125]{Jenny2001}. Both may be translated by ‘already’ (\ref{ex:rum:8a}, \ref{ex:rum:10a}), but they do not have to be translated at all (\ref{ex:rum:9}). In negative clauses, they are always omitted.

Furthermore, as illustrated in the following negative clauses, \textit{ɲjʌm} is always used together with the negative suffix \textit{{}-maʔ} in declarative verbal main clauses.

\ea\label{ex:rum:8}
    \ea\label{ex:rum:8a}
        \gll ʔân	\textbf{hɔ̰j}-tʰîw	cəʔuʔ \\ 
             3\textsc{sg} \textsc{nsit}{}-give book \\
        \glt ‘He has already given back the book.’ (ENEG\_13\_M\_1/2\_019)
    \ex\label{ex:rum:8b}
        \gll ʔân	\textbf{ɲjʌm}	tʰîw-\textbf{maʔ}	cəʔuʔ \\
             3\textsc{sg} not.yet give-\textsc{neg} book \\
        \glt ‘He has not yet given back the book.’ (ENEG\_13\_M\_1/2\_020)
    \z
\z 

The secondary verb \textit{hɔ̀ːj} \REF{ex:rum:9a} and the prefix \textit{hɔ̰j-} \REF{ex:rum:9b} are freely interchangeable, the negated clause \REF{ex:rum:9c} is the same for both \REF{ex:rum:9a} and \REF{ex:rum:9b}.

\ea\label{ex:rum:9}
    \ea\label{ex:rum:9a}
        \gll ʔʌ̂w	ɟɤh	\textbf{hɔ̀ːj} \\ 
             1\textsc{sg} buy finish \\
        \glt ‘I have bought it.’ (pc\_3)
    \ex\label{ex:rum:9b}
        \gll ʔʌ̂w	\textbf{hɔ̰j}-ɟɤh \\
             1\textsc{sg} \textsc{nsit}{}-buy \\
        \glt ‘I have bought it.’ (pc\_3)
    \ex\label{ex:rum:9c}
        \gll ʔʌ̂w	\textbf{ɲjʌm}	ɟɤh-\textbf{maʔ} \\
              1\textsc{sg} not.yet buy-\textsc{neg} \\
        \glt ‘I haven’t bought it yet.’ (pc\_3)
    \z
\z 

In \REF{ex:rum:10a}, the prefix as well as the secondary verb are used while in the negative clause \REF{ex:rum:10b} both are omitted. When both grammatical markers are used in a positive clause, one probably emphasizes the other, but it is not yet clear which is the emphasizing one.

\ea\label{ex:rum:10}
    \ea\label{ex:rum:10a}
        \gll ʔʌ̂w	\textbf{hɔ̰j}-ɟɤh	\textbf{hɔ̀ːj}	tɛ̂w.hlâ \\ 
              1\textsc{sg} \textsc{nsit}{}-buy finish vegetable \\
        \glt ‘I have already bought vegetables.’ (ENEG\_13\_M\_1/2\_059)
    \ex\label{ex:rum:10b}
        \gll ʔʌ̂w	\textbf{ɲjʌm}	ɟɤh-\textbf{maʔ}	tɛ̂w.hlâ \\
              1\textsc{sg} not.yet buy-\textsc{neg} vegetable \\
        \glt ‘I have not yet bought vegetables.’ (ENEG\_13\_M\_1/2\_060)
    \z
\z 

The replacement of the new situation markers by the negator \textit{ɲjʌm} ‘not yet’ indicates a constructional asymmetry \citep[52]{Miestamo2005}. As the asymmetry concerns a grammatical category of the aspectual domain, it belongs to Miestamo’s asymmetry type A/Cat/TAM\footnote{Asymmetry that affects the grammatical categories tense, aspect and/or mood.} \citep[116]{Miestamo2005}.

In Rumai, the lexical verbs in a clause belong to one of two functional categories: semantically primary verbs (main verbs) or semantically secondary verbs (secondary verbs). The secondary verbs (V2) can be divided into two sub-categories, namely preverbal and postverbal secondary verbs. The preverbal secondary verbs always have syntactic priority, which means that they take the verbal affixes and the postverbal secondary verbs. The postverbal secondary verbs never have syntactic priority. In \tabref{tab:rum:1}, the possible combinations of main verbs and secondary verbs are given.

\begin{table}\small 
\begin{tabularx}{\textwidth}{llXllXl}
\lsptoprule
& preverbal V2 &+& main verb & main verb & (+& postverbal V2)\\
\midrule 
semantic priority  & – && + & + && –\\
syntactic priority & + && – & + && –\\
\lspbottomrule
\end{tabularx}
\caption{Semantic and syntactic priority of the Rumai lexical verbs}
\label{tab:rum:1}
\end{table}

This categorization is important for negation because it is always the syntactic primary verb in a clause which takes the negation markers (as well as the other verbal prefixes and the postverbal secondary verbs) and is therefore in the scope of negation. As it is shown in \tabref{tab:rum:1}, in absence of a preverbal secondary verb, the syntactic primary verb is the main verb, otherwise it is the preverbal secondary verb. The most common preverbal secondary verbs are listed in \tabref{tab:rum:2}.\footnote{The analysis of Rumai is a work in progress, and there may be more marginally used preverbal secondary verbs. The same holds for the postverbal secondary verbs and the verbal affixes.}

\begin{table}
\begin{tabularx}{.9\textwidth}{llll}
\lsptoprule
{Verb} & {Lexical meaning} & {Function} & {Domain}\\
\midrule 
{\itshape lo} & {‘need’} & {obligation} & {modality}\\
{\itshape bɛ} & {‘can’} & {possibility, inherent ability} & {modality}\\
{\itshape səmâ} & {‘know’} & {acquired ability} & {modality}\\
{\textit{bɤ̂n}} & {‘get’} & {possibility, permission} & {modality}\\
{\textit{dɛh / tʰîw}} & {‘give’} & {causative} & {valency}\\
\lspbottomrule
\end{tabularx}
\caption{The most common preverbal secondary verbs in Rumai}
\label{tab:rum:2}
\end{table}

Examples (\ref{ex:rum:11}--\ref{ex:rum:15}) show that the means of negation with preverbal secondary verbs are the same as with main verbs.

\ea\label{ex:rum:11}
    \ea\label{ex:rum:11a}
        \gll pi	jê	\textbf{lo}	rêən	sɔːŋfâj \\ 
             elder.sibling \textsc{1pl.excl} need do kitchen \\
        \glt ‘Our elder sister has to clean the kitchen.’ (ENEG\_13\_M\_1/2\_114)
    \ex\label{ex:rum:11b}
        \gll pi	jê	\textbf{ʔaːw-lo}	rêən	sɔːŋfâj \\
             elder.sibling \textsc{1pl.excl} \textsc{neg}{}-need do kitchen \\
        \glt ‘Our elder sister does not have to clean the kitchen.’ (ENEG\_13\_M\_1/2\_116)
    \z
\z 

\ea\label{ex:rum:12}
    \gll ma	ʔʌ̂w	\textbf{buː-lo-maʔ}	rêən	ɟɯ̂ŋ	kjâːm \\
         mother 1\textsc{sg} \textsc{neg{}-}need\textsc{{}-neg} do work rough \\
    \glt ‘My mother doesn’t have to do rough work anymore.’ (WA\_15\_M\_3\_038)
\z 

\ea\label{ex:rum:13}
    \gll kɛ̂	\textbf{ɲjʌm}	\textbf{bɛ-maʔ}	lʌʔ	vɤh	dih\\
         3\textsc{pl} not.yet can\textsc{{}-neg} \textsc{comp} open \textsc{disc} \\
    \glt ‘They cannot yet open (school).’ (C3\_17\_MS\_9\_022)
\z 

\ea\label{ex:rum:14}
    \gll mâj	\textbf{ʔaːw-bɤ̂n-maʔ}	deəh	hla	pʌʔ	ɟɯ̂ŋ\\
         2\textsc{sg} \textsc{neg{}-}get\textsc{{}-neg} come late \textsc{loc} work \\
    \glt ‘You shouldn’t come late to work.’ (WPKT\_14\_L\_097)
\z 

\ea\label{ex:rum:15}
    \gll ma	ʔʌ̂w	siŋ-\textbf{dɛh-maʔ}	kɔ̂ːn	ʔân	hâːw	pʌʔ	raŋguŋ \\
         mother 1\textsc{sg} \textsc{des{}-}give\textsc{{}-neg} child 3\textsc{sg} go \textsc{loc} Yangon \\
    \glt ‘My mother does not want to let her children go to Yangon.’ (EMR\_19\_M\_3\_090)
\z 

Except for the negative constructions with \textit{ɲjʌm} (A/Cat/TAM), the examples provided so far suggest that negation in declarative verbal main clauses is symmetric. However, there is a certain paradigmatic asymmetry \citep[52]{Miestamo2005} regarding some verbal prefixes and postverbal secondary verbs. These grammatical markers are not generally incompatible with negation, but they are often omitted in negative clauses. Although \citet[58]{Miestamo2005} includes only fully grammaticalized, obligatory asymmetries in his study, this partial asymmetry is mentioned here as the tendency to omit certain grammatical markers is very strong. Like the asymmetry with \textit{ɲjʌm}, this asymmetry can be classified as type A/Cat/TAM. Of the verbal affixes, the imperfective marker \textit{giːj-} and the irrealis marker \textit{nʌŋ-} are most often affected. The feature is shown in example \REF{ex:rum:16}.

\ea\label{ex:rum:16}
    \ea\label{ex:rum:16a}
        \gll ʔân	dih	cəʔuʔ \\ 
             3\textsc{sg} read book \\
        \glt  ‘He reads a book.’
    \ex\label{ex:rum:16b}
        \gll ʔân	\textbf{nʌŋ}-dih	cəʔuʔ \\
             3\textsc{sg} \textsc{irr-}read book \\
        \glt ‘He will read a book.’
    \ex\label{ex:rum:16c}
        \gll ʔân	\textbf{giːj}-dih	cəʔuʔ \\
             3\textsc{sg} \textsc{ipfv-}read book \\
        \glt ‘He is reading a book.’
    \ex\label{ex:rum:16d}
        \gll ʔân	(\textbf{ʔaːw}-)dih-\textbf{maʔ}	cəʔuʔ \\
             3\textsc{sg} \textsc{neg-}read\textsc{{}-neg} book \\
        \glt ‘He doesn’t/will not read/isn’t reading a book.’ (pc\_3)
    \z
\z 

While in the affirmative clauses (\ref{ex:rum:16b}, \ref{ex:rum:16c}) the affixes are used to indicate modality and aspect, the negative clause \REF{ex:rum:16d} is underspecified and may therefore be read in different ways. However, when the function of the prefix is emphasized, it is also maintained under negation \REF{ex:rum:17}.

\ea\label{ex:rum:17}
    \ea\label{ex:rum:17a}
        \gll ʔân	\textbf{giːj}-dih	cəʔuʔ	dɤ̂ \\ 
             3\textsc{sg} \textsc{ipfv-}read book \textsc{ana} \\
        \glt ‘She is still reading that book.’
    \ex\label{ex:rum:17b}
        \gll ʔân	\textbf{ʔaːw-giːj}-dih-\textbf{maʔ}	cəʔuʔ	dɤ̂ \\
             3\textsc{sg} \textsc{neg-ipfv-}read\textsc{{}-neg} book \textsc{ana} \\
        \glt ‘She is still not reading that book.’ (pc\_3)
    \z
\z 

Of the post-verbal secondary verbs, \textit{ʔûːn} ‘keep’ is the one that is most often omitted in negative clauses. The basic function of this aspectual marker is to express duration without emphasizing initial or final boundaries, although they may be present implicitly. This includes ongoing actions or situations, but also general truths. General truths, in particular, convey certainty of the situation or event, and therefore the durative marker may additionally have developed into an emphasis marker which conveys more certainty. This is probably the case in \REF{ex:rum:18a}. In the negative clause \REF{ex:rum:18b}, the secondary verb is omitted, but the negative suffix \textit{{}-maʔ} is used which, in combination with the preverbal negative marker \textit{ʔaːw-}, may have an emphasizing function. As there are only a few examples of this pattern in the data, it is not clear at this point whether the correspondence is regular.

\ea\label{ex:rum:18}
    \ea\label{ex:rum:18a}
        \gll ʔân	jêəw	\textbf{ʔûːn}	bʌw.pʰɯ̂ŋ	ʔân \\ 
             3\textsc{sg} see keep friend 3\textsc{sg} \\
        \glt ‘He has (definitely) met his friend.’ (ENEG\_14\_M\_028)
    \ex\label{ex:rum:18b}
        \gll ʔân	\textbf{ʔaːw-}jêəw-\textbf{maʔ}	bʌw.pʰɯ̂ŋ	ʔân \\
             3\textsc{sg} \textsc{neg}{}-see-\textsc{neg} friend 3\textsc{sg} \\
        \glt ‘He did not meet his friend.’ (ENEG\_14\_M\_029)
    \z
\z 

Other secondary verbs like \textit{deəh} ‘come’ are usually retained in negative clauses \REF{ex:rum:19}. The function of \textit{deəh} is to express a motion towards the center of interest \citep[173]{Jenny2005}.

\ea\label{ex:rum:19}
    \gll səŋâj	təniŋla	paːt	bʌ̂ːn	mʌ	kɛ̂	\textbf{ɲjʌm}	vî	deəh-maʔ \\
         sun Monday week back \textsc{restr} 3\textsc{pl} not.yet return come{}-\textsc{neg} \\
    \glt ‘By next Monday they won’t have returned yet.’ (EGGD\_16\_M\_3\_158)
\z 

Some pronouns merge with the negative prefix \textit{ʔaːw-}, like \textit{ʔɛ̂w-} ‘\textsc{1pl.incl.neg}’ \REF{ex:rum:20} from \textit{ʔɛ̂} ‘\textsc{1pl.incl}’ and \textit{kɛ̂w-} ‘\textsc{3pl.neg}’ \REF{ex:rum:21} from \textit{kɛ̂} ‘\textsc{3pl}’.

\ea\label{ex:rum:20}
    \gll \textbf{ʔɛ̂w}-nʌ̀	pʌŋ	nʌŋ-rêən	tɛ	hôm \\
         1\textsc{pl.incl.neg}{}-know \textsc{rel} \textsc{irr}{}-do begin eat \\
    \glt  ‘We don't know what we should do for eating.’ (N2\_17\_MS\_7\_012)
\z 

\ea\label{ex:rum:21}
    \gll kɛ̂	nî	mʌ	\textbf{kɛ̂w}-meəh-maʔ	lʌʔ	jʌ̀ːj	rêən	hôm	siʔ	pʌŋ	hâj \\
         3\textsc{pl} \textsc{prox} \textsc{restr} \textsc{3pl.neg}{}-exist{}-\textsc{neg} \textsc{comp} rise do eat something \textsc{rel} other \\
    \glt ‘They don't have to do something else for living.’ (N2\_17\_MS\_7\_007)
\z
 
These are the only two mergers between a pronoun and \textit{ʔaːw-} that occur in the corpus. \tabref{tab:rum:3} contains the personal pronouns of Rumai.

\begin{table}
\begin{tabularx}{.8\textwidth}{lXXXl}
\lsptoprule
&  & {Singular} & {Dual} & {Plural}\\
\midrule
{1} & {Inclusive} & \multirow{2}{*}{\textit{ʔ\^{ʌ}w}} & {\textit{ʔ\^{a}ːj}} & {\itshape ʔ\^{ɛ}}\\
& {Exclusive} &  & {\textit{j\^{a}ːj}} & {\textit{j\^{e}}}\\
{2} &  & {\textit{m\^{a}j}} & {\textit{p\^{a}ːj}} & {\textit{p\^{ɛ}}}\\
{3} &  & {\textit{ʔân}} & {\textit{kâːj}} & {\textit{k\^{ɛ}}}\\
\lspbottomrule
\end{tabularx}
\caption{The Rumai personal pronouns}
\label{tab:rum:3}
\end{table}

The phonological shape of the pronouns seems to be a factor in whether a merger occurs. While a merger with pronouns ending in a vowel is possible, for phonotactic reasons it is unlikely with those ending in a consonant.

\subsection{Negation in non-declaratives} \label{sec:rum:2.2}

Non-declaratives in Rumai are imperative and interrogative clauses. Both have several means of expressing negation, which will be described in the following sections.

\subsubsection{Imperative clauses} \label{sec:rum:2.2.1}

Imperative clauses in Rumai are usually negated by the prohibitive marker \textit{kʰuː{}-} \REF{ex:rum:22}.

\ea\label{ex:rum:22}
    \ea\label{ex:rum:22a}
        \gll hâːw	pʌʔ	kəlah \\ 
             go \textsc{loc} market \\
        \glt ‘Go to the market!’ (ENEG\_13\_M\_1/2\_249)
    \ex\label{ex:rum:22b}
        \gll \textbf{kʰuː}-hâːw	pʌʔ	kəlah \\
              \textsc{proh}{}-go \textsc{loc} market \\
        \glt ‘Don’t go to the market!’ (ENEG\_13\_M\_1/2\_234)
    \z
\z 

Sometimes the prohibitive marker \textit{maj{}-} is used. This seems to be only possible if a corresponding positive imperative is semantically implausible as in \REF{ex:rum:23}. So, it may rather express a well-meaning advice.

\ea\label{ex:rum:23}
    \gll \textbf{maj}-bî	teəh	cɔːŋ \\
         \textsc{proh}{}-forget take umbrella \\
    \glt ‘Don’t forget to take an umbrella!’ (WPKT\_14\_L\_194)
\z 

So far, negation of imperatives is symmetric, but there are two features which can trigger an asymmetry. On the one hand, the second person pronoun may be used in prohibitive clauses for emphasis like \textit{mâj} ‘2\textsc{sg}’ in \REF{ex:rum:24}. According to the data, this seems impossible in positive imperatives.

\ea\label{ex:rum:24}
    \gll pʌŋtâːj	ciʔ	dʌh	\textbf{mâj}	kʰuː-leəp \\
         rabbit \textsc{tcl} say 2\textsc{sg} \textsc{proh}{}-enter \\
    \glt ‘The rabbit said: “Don’t enter!”’ (PCHT\_16\_L\_6\_035)
\z 

On the other hand, for more politeness, the polite marker \textit{kʰʌŋ} can be used in positive imperatives \REF{ex:rum:25}. This is not possible in negative ones \REF{ex:rum:26}.

\ea\label{ex:rum:25}
    \gll vɤh	\textbf{kʰʌŋ}	mòj	ploh	dî \\
         open \textsc{pol} mouth door \textsc{med} \\
    \glt  ‘Please open the door!’ (WPKT\_14\_L\_148)
\z 

\ea\label{ex:rum:26}
    \gll kʰuː-vɤh	\textbf{(*kʰʌŋ)}	mòj	ploh	dî\\
         \textsc{proh}{}-open \textsc{pol} mouth door \textsc{med}\\
    \glt ‘Don’t open the door!’ (pc\_3)
\z 

In their analysis of the prohibitive, \citet{AuweraEtAl2013-wals70} propose a four-way distinction which pays attention to the similarity between the verbal constructions of the prohibitive and the imperative, and to the similarity between the negative strategy used for prohibitives and for declaratives. 

\subsubsection{Interrogative clauses} \label{sec:rum:2.2.2}

Rumai interrogative clauses include polar questions and content questions. The negators which are used in standard negation are also used in negative interrogatives.

The polar questions mostly contain an interrogative particle, which is \textit{lʌh} in positive questions (\ref{ex:rum:27a}, \ref{ex:rum:28a}), while \textit{hɯh} is usually used in negative ones (\ref{ex:rum:27b}, \ref{ex:rum:28b}). All the interrogative particles directly follow the verb complex. In example \REF{ex:rum:27b}, the negator \textit{maʔ} is used, and in \REF{ex:rum:28b} it is \textit{ʔaːw-}. This choice is probably due to the intention of the speakers.

\ea\label{ex:rum:27}
    \ea\label{ex:rum:27a}
        \gll mâj	hâːw	\textbf{lʌh}	pʌʔ	ɟɯ̂ŋ \\ 
             2\textsc{sg} go \textsc{q} \textsc{loc} work \\
        \glt ‘Do you go to work?’ (ENEG\_13\_M\_1/2\_235)
    \ex\label{ex:rum:27b}
        \gll mâj	hâːw-\textbf{maʔ}	\textbf{hɯh}	pʌʔ	ɟɯ̂ŋ \\
             \textsc{2sg} go-\textsc{neg} \textsc{q} \textsc{loc} work \\
        \glt ‘Don’t you go to work?’ (ENEG\_13\_M\_1/2\_238)
    \z
\z 

\ea\label{ex:rum:28}
    \ea\label{ex:rum:28a}
        \gll mâj	dih	\textbf{lʌh}	bèəh	tɔ̂ːn	səŋâj \\ 
             2\textsc{sg} read \textsc{q} text every sun \\
        \glt ‘Do you read the newspaper every day?’ (ENEG\_13\_M\_1/2\_245)
    \ex\label{ex:rum:28b}
        \gll mâj	\textbf{ʔaːw}-dih	\textbf{hɯh}	bèəh	tɔ̂ːn	səŋâj \\
             2\textsc{sg} \textsc{neg-}read \textsc{q} text every sun \\
        \glt ‘Don’t you read the newspaper every day?’ (ENEG\_13\_M\_1/2\_246)
    \z
\z 

Neither the positive nor the negative question in \REF{ex:rum:29} contains a question marker, and the clauses are identical to the corresponding declarative clauses. Here, only the intonation allows the hearer to distinguish between the clause types. In declarative clauses and in questions including a question particle, the intonation is falling towards the end; in interrogative clauses without a question marker, it is rising.
Example \REF{ex:rum:29b} contains the common negative markers \textit{buː-} and \textit{{}-maʔ} to express the meaning ‘not anymore’.


\ea\label{ex:rum:29}
    \ea\label{ex:rum:29a}
        \gll ʔân	nʌŋ-hâːw	tʰɛ́ŋ	bô	pʌʔ	ranguŋ \\ 
             3\textsc{sg} \textsc{irr}{}-go repeat fill \textsc{loc} Yangon \\
        \glt ‘He will go to Yangon again.’
        \glt ‘Will he go to Yangon again?’
    \ex\label{ex:rum:29b}
        \gll ʔân	nʌŋ-\textbf{buː}-hâːw	tʰɛ́ŋ-\textbf{maʔ}	pʌʔ	ranguŋ \\
             3\textsc{sg} \textsc{irr-neg}{}-go repeat{}-\textsc{neg} \textsc{loc} Yangon \\
        \glt ‘He won’t go to Yangon anymore.’
        \glt ‘Won’t he go to Yangon anymore?’ (pc\_3)
    \z
\z 


Additionally, the particle \textit{hʌʔ} can be used as a question marker. Example \REF{ex:rum:30} shows an affirmative clause that includes the secondary verb \textit{hɔ̀ːj}, which is usually translated as ‘already’.

\ea\label{ex:rum:30}
    \gll ʔân	ɟɤh	\textbf{hɔ̀ːj}	kâ \\
         3\textsc{sg} buy finish car \\
    \glt ‘He already bought the car.’ (pc\_3)
\z 

In the corresponding question \REF{ex:rum:31}, the secondary verb \textit{hɔ̀ːj} is replaced by the particle \textit{hʌʔ}, but it can also be maintained to emphasize the meaning ‘already’ \REF{ex:rum:31a}. In the negative question \REF{ex:rum:31b}, \textit{ɲjʌm} ‘not yet’ replaces \textit{hɔ̀ːj}, and \textit{jêw} ‘perceive’, a secondary verb with a function that is not yet clear, follows the main verb.

\ea\label{ex:rum:31}
    \ea\label{ex:rum:31a}
        \gll ʔân	ɟɤh	(hɔ̀ːj)	\textbf{hʌʔ}	kâ \\ 
             3\textsc{sg} buy finish \textsc{ptcl} car \\
        \glt ‘Did he already buy the car?’
    \ex\label{ex:rum:31b}
        \gll ʔân	\textbf{ɲjʌm}	ɟɤh	jêw	kâ \\
             3\textsc{sg} not.yet buy perceive car \\
        \glt ‘Hasn’t he bought the car yet?’ (pc\_3)
    \z
\z 

According to native speakers, \textit{hʌʔ} means ‘to stop’ or ‘be all’. This and the form of the particle are hints that \textit{hʌʔ} may be derived from the verb \textit{hɔ̀ːj} ‘finish’. However, the postverbal position of \textit{hʌʔ} in interrogative clauses implies a derivation from the postverbal secondary verb \textit{hɔ̀ːj}. This would be unusual as \textit{hʌʔ} is an independent morpheme while \textit{hɔ̀ːj} in this function is not.

There do not seem to be any means to express an expectation for a positive or negative answer in polar questions.

Content questions can be expressed with interrogative particles (\ref{ex:rum:32}, \ref{ex:rum:33b}) or without (\ref{ex:rum:33a}, \ref{ex:rum:34}).

\ea\label{ex:rum:32}
    \ea\label{ex:rum:32a}
        \gll sim.baːj	mâj	sʌm	giːj-dih	\textbf{lʌh}	bèə \\ 
             why 2\textsc{sg} \textsc{conc} \textsc{ipfv}{}-read \textsc{q} text \\
        \glt ‘Why are you reading?’ (ENEG\_13\_M\_1/2\_208)
    \ex\label{ex:rum:32b}
        \gll sim.baːj	mâj	sʌm	\textbf{buː}-dih	\textbf{lʌh}	bèə \\
             why 2\textsc{sg} \textsc{conc} \textsc{neg}{}-read \textsc{q} text \\
        \glt ‘Why are you not reading?’ (ENEG\_13\_M\_1/2\_209)
    \z
\z 

\ea\label{ex:rum:33}
    \ea\label{ex:rum:33a}
        \gll cʰi	gɔ̂ːj	deəh	pʌʔ	dî \\ 
             who stay come \textsc{loc} \textsc{med} \\
        \glt ‘Who was there?’ (ENEG\_13\_M\_1/2\_222)
    \ex\label{ex:rum:33b}
        \gll cʰi	\textbf{ʔaːw}-gɔ̂ːj	deəh	\textbf{hɯh}	pʌʔ	dî \\
             who \textsc{neg}{}-stay come \textsc{q} \textsc{loc} \textsc{med} \\
        \glt ‘Who was not there?’ (ENEG\_13\_M\_1/2\_223)
    \z
\z 

\ea\label{ex:rum:34}
    \ea\label{ex:rum:34a}
        \gll si	mâj	ɟɤh	deəh \\ 
             what 2\textsc{sg} buy come \\
        \glt ‘What did you buy?’ (ENEG\_13\_M\_1/2\_219)
    \ex\label{ex:rum:34b}
        \gll si	mâj	\textbf{buː}-ɟɤh	deəh \\
             what 2\textsc{sg} \textsc{neg}{}-buy come \\
        \glt ‘What did you not buy?’ (ENEG\_13\_M\_1/2\_220)
    \z
\z  

Negative questions introduced by \textit{sim.baːj} ‘why’ \REF{ex:rum:32b} are always negated by the prefix \textit{buː{}-} and the question particle \textit{lʌh} follows the verb instead of \textit{hɯh}, which is usually used in negative questions. The use of \textit{buː-} here and in \REF{ex:rum:34b} may be due to a possible reading as ‘not anymore’, while in \REF{ex:rum:33b} this is less plausible. However, the use of the negative markers is flexible and different consultants may give different translations of the sentences.

The examples in this section show that negation in questions is usually asymmetric, only \REF{ex:rum:34b} is symmetric. But here also, translations given by other consultants may differ.

\subsection{Negation in stative predications} \label{sec:rum:2.3}

In Rumai, stative predications are verbal, and negation is expressed by using standard negation strategies. These are \textit{ʔaːw-}V\textit{{}-maʔ} ‘not’, \textit{ɲjʌm} V-\textit{maʔ} ‘not yet’ and \textit{buː{}-}V\textit{{}-maʔ} ‘not anymore’. The type of negation that seems impossible is negation by a simple preverbal or postverbal negator. 

Negation in stative predications is symmetric unless aspectual markers are involved. While clauses expressing “already” include the prefix \textit{hɔ̰j-} ‘\textsc{nsit}’ and/or the postverbal secondary verb \textit{hɔ̀ːj} ‘finish’, these aspectual markers are replaced by \textit{ɲjʌm} ‘not.yet’ in the corresponding negative clauses, as illustrated for \textit{hɔ̀ːj} in \REF{ex:rum:36d}. Other aspectual markers like \textit{giːj-} ‘\textsc{ipfv}’ are mostly omitted in negative clauses (36f, 38e). For \textit{ʔɯN-} ‘\textsc{dur}’ see \sectref{sec:rum:2.3.5}.

\subsubsection{Equation} \label{sec:rum:2.3.1}

Equation uses the verb \textit{moh} ‘be so’ that is negated by the standard strategies.\REF{ex:rum:35}.\footnote{The Rumai verbs are so far divided into primary and secondary verbs depending on their function in a clause (\sectref{sec:rum:2.1}, \tabref{tab:rum:1}). Whether \textit{moh} ‘be so’ can be assigned to one of these categories, or whether it is appropriate to introduce an additional category like “auxiliary” or “copula” is not yet clear. The research for the syntactic constructions in which \textit{moh} is used will give more insight.}

\ea\label{ex:rum:35}
    \ea\label{ex:rum:35a}
        \gll ʔân	moh	ma	ʔʌ̂w \\ 
             3\textsc{sg} be.so mother 1\textsc{sg} \\
        \glt ‘She is my mother.’
    \ex\label{ex:rum:35b}
        \gll ʔân	\textbf{ʔaːw}-moh-\textbf{maʔ}	ma	ʔʌ̂w \\
             3\textsc{sg} \textsc{neg}{}-be.so-\textsc{neg} mother 1\textsc{sg} \\
        \glt ‘She is not my mother.’ (pc\_3)
    \z
\z 

\subsubsection{Proper inclusion} \label{sec:rum:2.3.2}

In proper inclusion, the verb \textit{moh} 'be so' is used and is negated by the standard strategies as shown in \REF{ex:rum:36}.

\ea\label{ex:rum:36}
    \ea\label{ex:rum:36a}
        \gll ʔân	moh	kəɲjâ	nâ \\ 
             3\textsc{sg} be.so owner field \\
        \glt  ‘He is a farmer.’
    \ex\label{ex:rum:36b}
        \gll ʔân	\textbf{ʔaːw}-moh-\textbf{maʔ}	kəɲjâ	nâ \\
             3\textsc{sg} \textsc{neg}{}-be.so-\textsc{neg} owner field \\
        \glt ‘He isn’t a farmer.’
    \ex\label{ex:rum:36c}
        \gll ʔân	moh	\textbf{hɔ̀ːj}	kəɲjâ	nâ \\
             3\textsc{sg} be.so finish owner field \\
        \glt ‘He is already a farmer.’
    \ex\label{ex:rum:36d}
        \gll ʔân	\textbf{ɲjʌm}	moh-\textbf{maʔ}	kəɲjâ	nâ \\
             3\textsc{sg} not.yet be.so-\textsc{neg} owner field \\
        \glt ‘He is not yet a farmer.’
    \ex\label{ex:rum:36e}
        \gll ʔân	\textbf{giːj}-moh	kəɲjâ	nâ \\
             3\textsc{sg} \textsc{ipfv}{}-be.so owner field \\
        \glt ‘He is still a farmer.’
    \ex\label{ex:rum:36f}
        \gll ʔân	\textbf{buː}-moh-\textbf{maʔ}	kəɲjâ	nâ \\
              3\textsc{sg} \textsc{neg}{}-be.so-\textsc{neg} owner field \\
        \glt  ‘He isn’t a farmer anymore.’ (pc\_3)
    \z
\z 

In proper inclusion, sentences with a generic subject \REF{ex:rum:37} do not differ from sentences with specific subjects \REF{ex:rum:36}.

\ea\label{ex:rum:37}
    \ea\label{ex:rum:37a}
        \gll ʔəɲɛw	moh	to.sʌt \\ 
             cat be.so animal \\
        \glt  ‘Cats are animals.’
    \ex\label{ex:rum:37b}
        \gll ʔəɲɛw	\textbf{ʔaːw}-moh-\textbf{maʔ}	to.sʌt \\
             cat \textsc{neg}{}-be.so-\textsc{neg} animal \\
        \glt ‘Cats are not animals.’ (pc\_3)
    \z
\z 

\subsubsection{Attribution} \label{sec:rum:2.3.3}

Attribution is expressed by stative verbs \REF{ex:rum:38} that behave like intransitive active verbs \REF{ex:rum:39}, which means that they directly follow the subject and are within the scope of negation. Here also, the regular means of negation are used.

\ea\label{ex:rum:38}
    \ea\label{ex:rum:38a}
        \gll ʔân	pjɔ \\ 
             3\textsc{sg} happy \\
        \glt ‘She is happy.’
    \ex\label{ex:rum:38b}
        \gll ʔân	\textbf{ʔaːw}-pjɔ-\textbf{maʔ} \\
             3\textsc{sg} \textsc{neg}{}-happy-\textsc{neg} \\
        \glt  ‘She is not happy.’
    \ex\label{ex:rum:38c}
        \gll ʔân	\textbf{ɲjʌm}	pjɔ-\textbf{maʔ} \\
             3\textsc{sg} not.yet happy{}-\textsc{neg} \\
        \glt ‘She isn’t happy yet.’
    \ex\label{ex:rum:38d}
        \gll ʔân	\textbf{giːj}-pjɔ \\
             3\textsc{sg} \textsc{ipfv}{}-happy \\
        \glt ‘She is still happy.’
    \ex\label{ex:rum:38e}
        \gll ʔân	\textbf{buː}-pjɔ-\textbf{maʔ} \\
             3\textsc{sg} \textsc{neg}{}-happy-\textsc{neg} \\
        \glt ‘She isn’t happy anymore.’ (pc\_3)
    \z
\z 

\ea\label{ex:rum:39}
    \ea\label{ex:rum:39a}
        \gll ʔân	kəɲjah \\ 
              3\textsc{sg} laugh\\
        \glt ‘He laughs.’
    \ex\label{ex:rum:39b}
        \gll ʔân	\textbf{ʔaːw}-kəɲjah-\textbf{maʔ} \\
             3\textsc{sg} \textsc{neg}{}-laugh-\textsc{neg} \\
        \glt ‘He doesn’t laugh.’
    \ex\label{ex:rum:39c}
        \gll ʔân	\textbf{ɲjʌm}	kəɲjah-\textbf{maʔ} \\
             3\textsc{sg} not.yet laugh-\textsc{neg} \\
        \glt ‘He doesn’t laugh yet.’
    \ex\label{ex:rum:39d}
        \gll ʔân	\textbf{buː}-kəɲjah-\textbf{maʔ} \\
             3\textsc{sg} \textsc{neg}{}-laugh-\textsc{neg} \\
        \glt  ‘He doesn’t laugh anymore.’ (pc\_3)
    \z
\z 

There is no difference in negation between temporary \REF{ex:rum:38} and permanent \REF{ex:rum:40} property assignment.

\ea\label{ex:rum:40}
    \ea\label{ex:rum:40a}
        \gll kʰəleəp	nî	dâːŋ \\ 
             house \textsc{prox} big \\
        \glt  ‘This house is big.’ (ENEG\_13\_M\_1/2\_188)
    \ex\label{ex:rum:40b}
        \gll kʰəleəp	nî	\textbf{ʔaːw}-dâːŋ-\textbf{maʔ} \\
              house \textsc{prox} \textsc{neg}{}-big-\textsc{neg} \\
        \glt ‘This house is not big.’ (ENEG\_13\_M\_1/2\_189)
    \z
\z 

\subsubsection{Locative predication} \label{sec:rum:2.3.4}

Locative predication is expressed by the verb \textit{gɔ̂ːj} ‘stay’ \REF{ex:rum:41} which also means ‘live’ \REF{ex:rum:42}. Negation is regular.

\ea\label{ex:rum:41}
    \ea\label{ex:rum:41a}
        \gll ʔəɲɛw	gɔ̂ːj	têəŋ	kuː.ʔiʔ \\ 
             cat stay on bed \\
        \glt  ‘The cat is on the bed.’
    \ex\label{ex:rum:41b}
        \gll ʔəɲɛw	\textbf{ʔaːw}-gɔ̂ːj-\textbf{maʔ}	têəŋ	kuː.ʔiʔ \\
             cat \textsc{neg}{}-stay-\textsc{neg} on bed \\
        \glt ‘The cat is not on the bed.’
    \ex\label{ex:rum:41c}
        \gll ʔəɲɛw	\textbf{ɲjʌm}	gɔ̂ːj-\textbf{maʔ}	têəŋ	kuː.ʔiʔ \\
             cat not.yet stay-\textsc{neg} on bed \\
        \glt ‘The cat isn’t on the bed yet.’
    \ex\label{ex:rum:41d}
        \gll ʔəɲɛw	\textbf{buː}-gɔ̂ːj-\textbf{maʔ}	têəŋ	kuː.ʔiʔ \\
             cat \textsc{neg}{}-stay-\textsc{neg} on bed \\
        \glt ‘The cat isn’t on the bed anymore.’ (pc\_3)
    \z
\z 

\ea\label{ex:rum:42}
    \gll ʔʌ̂w	\textbf{gɔ̂ːj}	pʌʔ	ʔɯn.ɟuh	pʰədɛn \\
         1\textsc{sg} stay \textsc{loc} village \textsc{pn} \\
    \glt ‘I live in Pha Den village.’ (WPKT\_14\_L\_008)
\z 

\subsubsection{Existential predication} \label{sec:rum:2.3.5}

For existential predications, the verb \textit{meəh} ‘exist’ is used \REF{ex:rum:43}. Existential clauses are always verb initial. For locative-presentative constructions \citep[108]{Veselinova2013neg.exist}, a locative adverbial can be added to the existential clause without any other changes. Existential predications show asymmetry. This is due to the durative prefix \textit{ʔɯN-}, which describes a situation in which the boundaries are not known or are not of any importance. The negative clauses \REF{ex:rum:43b} and \REF{ex:rum:43d} both show this feature, but only in \REF{ex:rum:43d} is \textit{ʔɯN-} used. The reason for its presence in this clause is not yet clear, and more research is needed. In \REF{ex:rum:43c}, there is a boundary as \textit{ɲjʌm} ‘not.yet' also means “not until now”.

\ea\label{ex:rum:43} 
    \ea\label{ex:rum:43a}
        \gll (kʰêəŋ	bjâj)	\textbf{ʔɯm}-meəh	sâːŋ \\ 
             in forest \textsc{dur}{}-exist elephant \\
        \glt ‘There are elephants (in the forest).’
    \ex\label{ex:rum:43b}
        \gll (kʰêəŋ	bjâj)	\textbf{ʔaːw}-meəh-\textbf{maʔ}	sâːŋ \\
             in forest \textsc{neg}{}-exist-\textsc{neg} elephant \\
        \glt ‘There are no elephants (in the forest).’
    \ex\label{ex:rum:43c}
        \gll (kʰêəŋ	bjâj)	\textbf{ɲjʌm}	meəh-\textbf{maʔ}	sâːŋ \\
             in forest not.yet exist-\textsc{neg} elephant \\
        \glt ‘There aren't elephants (in the forest) yet.’
    \ex\label{ex:rum:43d}
        \gll (kʰêəŋ	bjâj)	\textbf{ʔɯm-buː}-meəh-\textbf{maʔ}	sâːŋ \\
             in forest \textsc{dur}{}-\textsc{neg}{}-exist-\textsc{neg} elephant \\
        \glt ‘There aren't elephants (in the forest) anymore.’ (pc\_3)
    \z
\z 

\subsubsection{Possessive predication} \label{sec:rum:2.3.6}

The verb \textit{meəh} ‘exist’ also has the sense ‘have’ in Rumai and is therefore used in possessive predication \REF{ex:rum:44}. Negation is regular.

\ea\label{ex:rum:44}
    \ea\label{ex:rum:44a}
        \gll ʔân 	meəh	kʰəleəp \\ 
             3\textsc{sg} exist house \\
        \glt ‘He has a house.’
    \ex\label{ex:rum:44b}
        \gll ʔân 	\textbf{ʔaːw}-meəh-\textbf{maʔ}	kʰəleəp \\
             3\textsc{sg} \textsc{neg}{}-exist-\textsc{neg} house \\
        \glt ‘He doesn’t have a house.’
    \ex\label{ex:rum:44c}
        \gll ʔân 	\textbf{ɲjʌm}	meəh-\textbf{maʔ}	kʰəleəp \\
             3\textsc{sg} not.yet exist-\textsc{neg} house \\
        \glt ‘He doesn’t have a house yet.’
    \ex\label{ex:rum:44d}
        \gll ʔân 	\textbf{buː}-meəh-\textbf{maʔ}	kʰəleəp \\
             3\textsc{sg} \textsc{neg}{}-exist-\textsc{neg} house \\
        \glt ‘He doesn’t have a house anymore.’ (pc\_3)
    \z
\z 

\subsection{Negation in non-main clauses} \label{sec:rum:2.4}

Rumai dependent clauses are mainly verb-initial. They are negated only by \textit{buː{}-} ‘\textsc{neg}’ or \textit{ɲjʌm} ‘not yet’, which means that the negators \textit{ʔaːw-} and \textit{{}-maʔ} are never used in dependent clauses. Concerning the postverbal \textit{{}-maʔ}, its absence in dependent clauses is probably due to the tendency to avoid postverbal negators in verb-initial languages as observed by \citet{Dryer1988} in his study about negative position \citep[97]{Dryer1988}. Although the current basic word order in Rumai is verb-medial, one can assume that it was generally verb-initial as dependent clauses are believed to be more conservative than independent ones \citep{Bybee2001}.
The following examples include relative clauses, temporal clauses, complement clauses and conditional clauses. These constructions are also representative of the other dependent clauses like causal and concessive clauses.

Relative clauses are usually introduced by the relativizers \textit{ʔî} for persons \REF{ex:rum:45} and \textit{pʌŋ} for things \REF{ex:rum:46}.\footnote{This is only a general distinction. Animals, for example, can be relativized by \textit{ʔî} or \textit{pʌŋ}, depending on the intention of the speaker.}

\ea\label{ex:rum:45}
    \ea\label{ex:rum:45a}
        \gll ʔî.mâːj	dɛʔ	cʰeh	[ʔî	dih	bèəh]	nʌŋ-ʔɔːŋ	bèəh \\ 
             man small \textsc{clf} {\db}\textsc{rel} read text \textsc{irr}{}-pass text \\
        \glt ‘The boy who learns the text will pass the exam.’ (ENEG\_14\_M\_007)
    \ex\label{ex:rum:45b}
        \gll ʔî.mâːj	dɛʔ	cʰeh	[ʔî	\textbf{buː}-dih	bèəh]	nʌŋ-ʔaːw-ʔɔːŋ-maʔ	bèəh \\
             man small \textsc{clf} {\db}\textsc{rel} \textsc{neg}{}-read text \textsc{irr-neg}{}-pass-\textsc{neg} text \\
        \glt ‘The boy who does not learn the text will not pass the exam.’ (ENEG\_14\_M\_008)
    \z
\z 

\ea\label{ex:rum:46}
    \gll kiʔ		nî	moh	doʔ	[pʌŋ	\textbf{ɲjʌm}	dih	ʔʌ̂w] \\
         \textsc{nmlz} \textsc{prox} be.so exhausted {\db}\textsc{rel} not.yet read 1\textsc{sg}\\
    \glt ‘These are the ones I have not read yet.’ (EMSR\_19\_M\_3\_059)
\z 

Relative clauses can include verbal affixes like the irrealis \textit{nʌŋ-} in \REF{ex:rum:47}. Here, the irrealis marker is maintained in the negative clause.

\ea\label{ex:rum:47}
    \ea\label{ex:rum:47a}
        \gll ʔɯŋ.hmêəw	ʔʌ̂w	jêəw	bʌw.pʰɯ̂ŋ	cʰeh	[ʔî	\textbf{nʌŋ}-ɟoh	kʰəlêəp] \\ 
             yesterday 1\textsc{sg} see friend \textsc{clf} {\db}\textsc{rel} \textsc{irr}{}-buy house \\
        \glt ‘Yesterday, I met the friend who will buy a house.’
    \ex\label{ex:rum:47b}
        \gll ʔɯŋ.hmêəw	ʔʌ̂w	jêəw	bʌw.pʰɯ̂ŋ	cʰeh	[ʔî	\textbf{nʌŋ-buː}-ɟoh	kʰəlêəp] \\
             yesterday 1\textsc{sg} see friend \textsc{clf} {\db}\textsc{rel} \textsc{irr}{}-\textsc{neg}{}-buy house \\
        \glt ‘Yesterday, I met the friend who will not buy a house.’ (pc\_3)
    \z
\z 

Temporal clauses are introduced by \textit{ʔəkʰjɯŋ} ‘time’ \REF{ex:rum:48} which can be interpreted as the relative construction ‘the time when’, or by a conjunction like \textit{ʔàːj} ‘before’ \REF{ex:rum:49}.

\ea\label{ex:rum:48}
    \gll [\textbf{ʔəkʰjɯŋ}	\textbf{buː}-deəh	pi	ʔʌ̂w	pʌʔ	kʰəleəp]	ʔʌ̂w	nɔh.hɲjap-maʔ \\
         {\db}time \textsc{neg}{}-come elder.sibling 1\textsc{sg} \textsc{loc} house 1\textsc{sg} worried{}-\textsc{neg} \\
    \glt ‘When my brother did not come home, I was not worried.’ (ENEG\_13\_M\_1/2\_167)
\z 

\ea\label{ex:rum:49}
    \ea\label{ex:rum:49a}
        \gll [\textbf{ʔàːj}	teəh	mâj	tʌʔ	ɟɯ̂ŋ	təmâːj	dɤ̂]	hmɔ̀	mâj	kûːn	mâj \\ 
             {\db}before take 2\textsc{sg} \textsc{obl} work new \textsc{ana} ask 2\textsc{sg} father 2\textsc{sg} \\
        \glt ‘Before you accept that new job, ask your father.’ 
    \ex\label{ex:rum:49b}
        \gll [\textbf{ʔàːj}	\textbf{ɲjʌm}	teəh	mâj	tʌʔ	ɟɯ̂ŋ	təmâːj	dɤ̂]	hmɔ̀	mâj	kûːn	mâj \\
             {\db}before not.yet take 2\textsc{sg} \textsc{obl} work new \textsc{ana} ask 2\textsc{sg} father 2\textsc{sg} \\
        \glt ‘Before you don’t accept that new job, ask your father.’ (pc\_3)
    \z
\z 

In \REF{ex:rum:49b}, the negator \textit{ɲjʌm} ‘not yet’ is used, but neither the prefix \textit{hɔ̰j-} ‘\textsc{nsit’} nor the secondary verb \textit{hɔ̀ːj} occur in the positive clause \REF{ex:rum:49a}. That means, \textit{ɲjʌm} is not triggered by the aspectual markers, but its use depends on the intention of the speaker.

For the complements in \REF{ex:rum:50}, an English translation with a finite verb as well as a non-finite one is possible. 

\ea\label{ex:rum:50}
    \ea\label{ex:rum:50a}
        \gll ʔʌ̂w	jêəw	ʔûːn	[hôm	ʔân] \\ 
             1\textsc{sg} see keep {\db}eat 3\textsc{sg} \\
        \glt ‘I saw that she was eating.’
        \glt ‘I saw her eat.’
    \ex\label{ex:rum:50b}
        \gll ʔʌ̂w	jêəw	ʔûːn	[\textbf{buː}-hôm	ʔân] \\
             1\textsc{sg} see keep {\db}\textsc{neg}{}-eat 3\textsc{sg} \\
        \glt ‘I saw that she was not eating.’
        \glt ‘I saw her not eat.’ (pc\_3)
    \z
\z 

This brings up the question about finiteness in Rumai. \citet{Bisang2001} discusses this feature from different viewpoints. These are finiteness as a scalar phenomenon and finiteness due to morphological or syntactical reasons. Rumai lexical verbs are invariable in their form, but there are verbal affixes and the postverbal secondary verbs that have aspectual and modality functions. It is currently unclear whether these affixes are excluded in specific clause types. More work is required on this issue.

%Rumai lexical verbs are invariable in their form, but there are the aspectual affixes and postverbal secondary verbs. If there are clause types, where some of these markers are excluded, or at least very uncommon, one could say that these clauses are less finite. However, this needs more research.

Complement clauses following verbs of thinking \REF{ex:rum:51} and saying \REF{ex:rum:52} are constructed like main clauses. This means that they are verb medial, and they use the negators of main clauses. The suffix {}-\textit{maʔ} is possible, but not frequent in these complement clauses.

\ea\label{ex:rum:51}
    \ea\label{ex:rum:51a}
        \gll ʔʌ̂w	tʰaŋ	[ʔân	nʌŋ-deəh] \\ 
             1\textsc{sg} think {\db}3\textsc{sg} \textsc{irr}{}-come \\
        \glt ‘I think he will come.’
    \ex\label{ex:rum:51b}
        \gll ʔʌ̂w	tʰaŋ	[ʔân	nʌŋ-\textbf{ʔaːw}-deəh(-maʔ)] \\
             1\textsc{sg} think {\db}3\textsc{sg} \textsc{irr}{}-\textsc{neg}{}-come-\textsc{neg} \\
        \glt ‘I think he will not come.’
    \ex\label{ex:rum:51c}
        \gll ʔʌ̂w	\textbf{ʔaːw}-tʰaŋ-\textbf{maʔ}	[ʔân	nʌŋ-deəh] \\
             1\textsc{sg} \textsc{neg}{}-think-\textsc{neg} {\db}3\textsc{sg} \textsc{irr}{}-come \\
        \glt ‘I don’t think he will come.’
    \ex\label{ex:rum:51d}
        \gll ʔʌ̂w	\textbf{ʔaːw}-tʰaŋ-\textbf{maʔ}	[ʔân	nʌŋ-\textbf{ʔaːw}-deəh(-maʔ)] \\
             1\textsc{sg} \textsc{neg}{}-think-\textsc{neg} {\db}3\textsc{sg} \textsc{irr}{}-\textsc{neg}{}-come-\textsc{neg} \\
        \glt ‘I don’t think he will not come.’ (pc\_3)
    \z
\z 

\ea\label{ex:rum:52}
    \ea\label{ex:rum:52a}
        \gll ʔân	dʌh	[ʔân	nʌŋ-deəh] \\ 
             \textsc{3sg} say {\db}3\textsc{sg} \textsc{irr}{}-come \\
        \glt ‘He said that he would come.’
    \ex\label{ex:rum:52b}
        \gll ʔân	dʌh	[ʔân	nʌŋ-\textbf{ʔaːw}-deəh(-maʔ)] \\
             \textsc{3sg} say {\db}\textsc{3sg} \textsc{irr}{}-\textsc{neg}{}-come-\textsc{neg} \\
        \glt ‘He said that he would not come.’
    \ex\label{ex:rum:52c}
        \gll ʔân	\textbf{ʔaːw}-dʌh-\textbf{maʔ}	[ʔân	nʌŋ-deəh] \\
             \textsc{3sg} \textsc{neg}{}-say-\textsc{neg} {\db}\textsc{3sg} \textsc{irr}{}-come \\
        \glt ‘He didn’t say that he would come.’
    \ex\label{ex:rum:52d}
        \gll ʔân	\textbf{ʔaːw}-dʌh-\textbf{maʔ}	[ʔân	nʌŋ-\textbf{ʔaːw}-deəh(-maʔ)] \\
             \textsc{3sg} \textsc{neg}{}-say-\textsc{neg} {\db}\textsc{3sg} \textsc{irr}{}-\textsc{neg}{}-come-\textsc{neg} \\
        \glt ‘He didn’t say that he would not come.’ (pc\_3)
    \z
\z 

Conditional clauses are verb-medial; nevertheless, like other dependent clauses, the negators are limited to \textit{buː{}-} (\ref{ex:rum:53}) and \textit{ɲjʌm} \REF{ex:rum:54}.

\ea\label{ex:rum:53}
    \gll [sʌ	mâj	\textbf{buː}-təɲâːw]	mʌŋ-jɔh	ŋərâːj \\
         {\db}\textsc{cond} 2\textsc{sg} \textsc{neg}{}-listen 2\textsc{sg.irr}{}-fall hell \\
    \glt ‘If you don’t listen, you will fall to hell.’ (WPKT\_14\_L\_082)
\z 

\ea\label{ex:rum:54}
    \ea\label{ex:rum:54a}
        \gll [kʰʌ	mâj	hɔ̰j-teəh	hʌ̂ːj]	pù	ŋâj \\ 
             {\db}\textsc{cond} 2\textsc{sg} \textsc{nsit}{}-take wood blow fire \\
        \glt ‘If you have brought the wood already, make a fire.’
    \ex\label{ex:rum:54b}
        \gll [kʰʌ	mâj	\textbf{ɲjʌm}	teəh	hʌ̂ːj]	teəh	cʌŋ.naj \\
             {\db}\textsc{cond} 2\textsc{sg} not.yet take wood take now \\
        \glt ‘If you haven’t brought the wood yet, bring it now.’ (pc\_3)
    \z
\z 

\subsection{Negative lexicalizations} \label{sec:rum:2.5}

In Rumai, there are no negative lexicalizations. The only negative marker that comes close to them ist \textit{ɲjʌm} ‘not yet’. It is probably derived from a former verb with the meaning ‘dilatory’ or ‘stiff’. However, as it usually replaces the new situation prefix \textit{hɔ̰j-} and/or the secondary verb \textit{hɔ̀ːj} ‘finish’ of the positive clause, it is part of the Rumai aspect system, and is therefore treated in the section about standard negation (\sectref{sec:rum:2.1}).

\section{Non-clausal negation} \label{sec:rum:3}

Non-clausal negation is rare in Rumai. The only examples which have been found so far are answers to questions that include ‘already’ in the translation. The features of the other structures included in this section will be described in as much detail as possible.

\subsection{Negative replies} \label{sec:rum:3.1}

Example \REF{ex:rum:55} shows a positive and a negative question, and \REF{ex:rum:56} shows the corresponding answers.

\ea\label{ex:rum:55}
    \ea\label{ex:rum:55a}
        \gll sʌw	bjɔh	\textbf{lʌh} \\ 
             dog bark \textsc{q} \\
        \glt ‘Is the dog barking?’
    \ex\label{ex:rum:55b}
        \gll sʌw	\textbf{ʔaːw}-bjɔh	\textbf{hɯh} \\
             dog \textsc{neg}{}-bark \textsc{q} \\
        \glt ‘Is the dog not barking?’ (pc\_3)
    \z
\z 

The answers to both questions in \REF{ex:rum:55} are clausal, and they are the same both for the positive and the negative question \REF{ex:rum:56}. This means that the positive answer \textit{bjɔh} ‘bark’ does not agree with the polarity of the negative question and the negative answer (‘not bark’) does not agree with the polarity of the positive question.\footnote{Several studies propose different categories of polar answers. However, the sample of Rumai answers is too small to draw a conclusion in this respect. For more information about polar answers see for example \citet{Moser2018} and \citet{EnfieldEtAl2019}.}

\ea\label{ex:rum:56}
    \ea\label{ex:rum:56a}
        \gll bjɔh	(ʔɤ) \\ 
             bark \textsc{asrt} \\
        \glt ‘Yes (the dog is barking).'
    \ex\label{ex:rum:56b}
        \gll \textbf{ʔaːw}-bjɔh-\textbf{maʔ} \\
             \textsc{neg}{}-bark-\textsc{neg} \\
        \glt ‘No (the dog is not barking) (pc\_3).'
    \z
\z 

The assertive marker \textit{ʔɤ} is not obligatory, and so negation in the answers is symmetric.

Questions including \textit{hʌʔ} \REF{ex:rum:57} which is usually translated by ‘already’ (see also \sectref{sec:rum:2.2.2}) can be answered in two ways. The first answers are clausal (\ref{ex:rum:57b} and \ref{ex:rum:57c}), and the second ones are not (\ref{ex:rum:57d} and \ref{ex:rum:57e}). 

\ea\label{ex:rum:57}
    \ea\label{ex:rum:57a}
        \gll mâj	ɟɤh	\textbf{hʌʔ}	kâ \\ 
             \textsc{2sg} buy \textsc{ptcl} car \\
        \glt ‘Did you already buy the car?’
    \ex\label{ex:rum:57b}
        \gll ɟɤh	hɔ̀ːj \\
             buy finish \\
        \glt ‘Yes.’
    \ex\label{ex:rum:57c}
        \gll \textbf{ɲjʌm}	ɟɤh-\textbf{maʔ} \\
             not.yet buy-\textsc{neg} \\
        \glt ‘No.’
    \ex\label{ex:rum:57d}
        \gll hʌʔ \\
             \textsc{ptcl} \\
        \glt ‘Yes.’ 
    \ex\label{ex:rum:57e}
        \gll ɲjʌm \\
             not.yet \\
        \glt ‘No.’ (pc\_3)
    \z
\z 
 
These answers, which are a simple \textit{hʌʔ} for the positive one and \textit{ɲjʌm} ‘not yet’ for the negative one, are the only non-clausal answers so far detected in Rumai.

\subsection{Negative indefinites and quantifiers} \label{sec:rum:3.2}

There are five indefinites in Rumai that are also used in negative clauses. Three of them are derived from interrogatives, namely \textit{hmeəʔ} ‘any’ from \textit{hmeə} ‘which, where’, \textit{cʰiʔ} ‘anyone’ from \textit{cʰi} ‘who’ and \textit{siʔ} ‘something, anything’ from \textit{si} ‘what’. Additionally, there are the indefinite pronoun \textit{ʔî} ‘others’ and the experiential prefix \textit{tʌm-} ‘ever’. All the clauses that include indefinites are negated by means of standard negation.

The use of the indefinite \textit{hmeəʔ} ‘any’ in negative clauses is shown in \REF{ex:rum:58} and \REF{ex:rum:59}.

\ea\label{ex:rum:58}
    \gll cəʔuʔ	\textbf{ʔaːw}-meəh	pʌʔ	\textbf{hmeəʔ}\\
         book \textsc{neg}{}-exist \textsc{loc} any \\
    \glt ‘The book is nowhere.’ (pc\_3)
\z 

\ea\label{ex:rum:59}
    \gll bɯn	\textbf{hmeəʔ}	mʌ	ʔʌ̂w	\textbf{buː}-bî	mâj \\
         time any \textsc{restr} 1\textsc{sg} \textsc{neg}{}-forget 2\textsc{sg} \\
    \glt ‘I will never forget you.’ (WPKT\_14\_L\_077)
\z 
 
Examples \REF{ex:rum:60} and \REF{ex:rum:61} illustrate \textit{cʰiʔ} ‘anyone’ in a positive and a negative clause.

\ea\label{ex:rum:60}
    \gll \textbf{cʰiʔ}	sʌ	jʌŋ	hɲja	lɯ̂jləbùj	nʌŋ-kəɲâ	tʌm	jʌ̂ːm \\
         anyone \textsc{cond} sell tobacco white \textsc{irr}{}-come.across punishment death \\
    \glt ‘Anyone, if they sell heroin, will be punished with the death penalty.’ (WPKT\_14\_L\_096)
\z 

\ea\label{ex:rum:61}
    \gll pʌŋ	səmâ	jê	\textbf{cʰiʔ}	mʌ	\textbf{ʔaːw}-bɛ-\textbf{maʔ}	lʌʔ	deəh	bjâ	bɤ̂n \\
         \textsc{rel} know \textsc{1pl.excl} anyone \textsc{restr} \textsc{neg}{}-can{}-\textsc{neg} \textsc{comp} come steal get \\
    \glt ‘Nobody can steal what we know.’ (W1\_025)
\z 

\textit{siʔ} ‘something, anything’ is also used in positive \REF{ex:rum:62} and in negative \REF{ex:rum:63} clauses.

\ea\label{ex:rum:62}
    \gll kʌp	lo	cʌm	rêən	\textbf{siʔ}	mʌ	jʌ̂w	ʔɯn.sətà	dih \\
         tight need \textsc{incl} do anything \textsc{restr} fear heat \textsc{disc} \\
    \glt ‘Although there was something to do, we feared the heat.’ (C3\_17\_MS\_9\_057)
\z 

\ea\label{ex:rum:63}
    \gll ʔân	\textbf{ʔaːw}-kəsòj	dʌh-\textbf{maʔ}	\textbf{siʔ} \\
         3\textsc{sg} \textsc{neg}{}-respond say-\textsc{neg} anything \\
    \glt ‘He doesn’t respond to anything.’ (WPKT\_14\_L\_271)
\z

The indefinite pronoun \textit{ʔî} ‘others’ is shown in \REF{ex:rum:64} and \REF{ex:rum:65}. The semantics of this pronoun suggest that the speaker is excluded, but this is still to be clarified.

\ea\label{ex:rum:64}
    \gll pjatəna	nî	\textbf{ʔî}	bɛ	lʌʔ	pʰje\\
         problem \textsc{prox} others can \textsc{comp} answer \\
    \glt ‘Anyone can solve this problem.’ (pc\_3)
\z 

\ea\label{ex:rum:65}
    \gll \textbf{ʔaːw}-meəh-\textbf{maʔ}	\textbf{ʔî}	nʌ̀	ʔəpʰje \\
         \textsc{neg}{}-exist-\textsc{neg} others know answer \\
    \glt ‘Nobody knows the answer.’ (pc\_3)
\z 

Examples \REF{ex:rum:66} and \REF{ex:rum:67} show the use of the experiential prefix \textit{tʌm-}.

\ea\label{ex:rum:66}
    \gll ʔàːj	nî	mâj	\textbf{tʌm}-cʌ̂ːm	lʌh	kɯ̂ŋ	korijâ \\
         before \textsc{prox} 2\textsc{sg} \textsc{exp}{}-reach \textsc{q} state Korea \\
    \glt ‘Have you ever been in Korea before?’ (WPKT\_14\_L\_044)
\z 

\ea\label{ex:rum:67}
    \gll ʔʌ̂w	\textbf{ʔaːw}-\textbf{tʌm}-cʌ̂ːm-\textbf{maʔ} \\
         1\textsc{sg} \textsc{neg}{}-\textsc{exp}{}-reach-\textsc{neg} \\
    \glt ‘I have never been there.’ (WPKT\_14\_L\_045)
\z 

\subsection{Negative derivation and case marking} \label{sec:rum:3.3}

There are no special expressions for meanings like absence, opposition or antonymy in Rumai. Instead, these meanings are expressed by common lexical and syntactic means. Examples (\ref{ex:rum:68}--\ref{ex:rum:70}) show some of these constructions.

\ea\label{ex:rum:68}
    \ea\label{ex:rum:68a}
        \gll pen	dâ \\ 
             good use \\
        \glt  ‘useful’ 
    \ex\label{ex:rum:68b}
        \gll \textbf{buː}-pen	dâ \\
             \textsc{neg}{}-good use \\
        \glt ‘useless’ (pc\_3)
    \z
\z 

\ea\label{ex:rum:69}
    \gll cʰiʔ	ɲjʌm	dih-\textbf{maʔ}	bèəh	nî \\
         anyone not.yet read-\textsc{neg} text \textsc{prox} \\
    \glt ‘This text is unread.’ (lit. ‘No one has read this text so far.’) (pc\_3)
\z 

\ea\label{ex:rum:70}
    \gll cʰiʔ	\textbf{buː}-bɛ-\textbf{maʔ}	lʌʔ	dih	bèəh	nî \\
         anyone \textsc{neg}{}-can-\textsc{neg} \textsc{comp} read text \textsc{prox} \\
    \glt ‘This text is unreadable.’ (lit. ‘No one can read this text.’) (pc\_3)
\z 

\section{Other aspects of negation} \label{sec:rum:4}

The phenomena discussed in this section are not negative constructions or expressions themselves but occur in connection with negation.

\subsection{The scope of negation} \label{sec:rum:4.1}

In Rumai, the scope of negation is dependent on the context. This means, the negators keep their usual position and there are no scope-indicating morphemes. In \REF{ex:rum:71}, the scope is on the object and in \REF{ex:rum:72} it is on the whole clause.

\ea\label{ex:rum:71}
    \gll ʔʌ̂w	dih-maʔ	cəʔuʔ	ʔʌ̂w	dih	bèə \\
         1\textsc{sg} read-\textsc{neg} book 1\textsc{sg} read text \\
    \glt  ‘I am not reading a book but a letter.’ (pc\_1)
\z 

\ea\label{ex:rum:72}
    \gll bɯn	hmeəʔ	mʌ	ʔʌ̂w	tʌm-gɔ̂ːj-maʔ	mandale \\
         time any \textsc{restr} 1\textsc{sg} \textsc{exp}{}-stay-\textsc{neg} Mandalay \\
    \glt ‘I have never lived in Mandalay.’ (ENEG\_13\_M\_1/2\_051)
\z 

Not much about the scope of negation in Rumai is known so far, and more research is needed.

\subsection{Negative polarity} \label{sec:rum:4.2}

No negative polarity items have been found in research on the language so far.

\subsection{Marking of NPs in the scope of negation.} \label{sec:rum:4.3}

Negation does not affect the marking of noun phrases in Rumai.

\subsection{Reinforcing negation} \label{sec:rum:4.4}

The negative suffix \textit{{}-maʔ} that was originally the means of reinforcing negation is pragmatically bleached today as it is used very frequently, and it may also be the sole negator in a clause \REF{ex:rum:73}.\footnote{See also \sectref{sec:rum:2.1}., especially the comment after example \REF{ex:rum:6}, and \sectref{sec:rum:4.6.2} for Jespersen Cycle.}

\ea\label{ex:rum:73}
    \gll kɛ̂	nʌ̀-\textbf{maʔ}	ʔɯn.dʌ̂ːn	hméə	kɛ̂	nʌŋ-hâːw \\
         3\textsc{pl} know\textsc{{}-neg} way which 3\textsc{pl} \textsc{irr-}go \\
    \glt ‘They don’t know which way they have to go.’ (EMR\_19\_M\_3\_026)
\z 

Whether prosody plays a role in reinforcing negation is not yet clear.

\subsection{Negation, coordination and complex clauses} \label{sec:rum:4.5}

According to the present database, there are no special coordinators, constructions for contrastive negation, or negative conjunctions in dependent clauses in Rumai.

For example, “neither…nor” constructions are expressed by a simple juxtaposition of two clauses \REF{ex:rum:74}, or the coordination of objects \REF{ex:rum:75}, showing standard negation. Whether the choice of the negators, namely \textit{{}-maʔ} in \REF{ex:rum:74}, and \textit{ʔaːw-…-maʔ} in \REF{ex:rum:75} is mandatory in these constructions is still to be clarified.

\ea\label{ex:rum:74}
    \gll ʔʌ̂w	dih-\textbf{maʔ}	cəʔuʔ	ʔʌ̂w	dih-\textbf{maʔ}	bèəh \\
         \textsc{1sg} read-\textsc{neg} book \textsc{1sg} read-\textsc{neg} text\\
    \glt ‘I am neither reading a book nor a letter.’ (ENEG\_13\_M\_1/2\_004)
\z 

\ea\label{ex:rum:75}
    \gll ʔân	\textbf{ʔaːw}-ɟɤh-\textbf{maʔ}	səkjɤ̂	tu	klɔ̂ːj \\
         \textsc{3sg} \textsc{neg}{}-buy-\textsc{neg} cucumber and banana\\
    \glt ‘He buys neither a cucumber nor a banana.’ (pc\_3)
\z 

\textit{tu} ‘and’ is not very frequent in the corpus, but it is also used in positive clauses, so that the meaning ‘nor’ can be excluded.

\subsection{Miscellaneous aspects of negation} \label{sec:rum:4.6}

In this section, negative transport (\sectref{sec:rum:4.6.1}) and diachronic observations (\sectref{sec:rum:4.6.2}) are discussed.

\subsubsection{Negative transport} \label{sec:rum:4.6.1}

By negative transport, the negation of an independent clause is interpreted as the negation of the predicate of a dependent clause \citep[223]{Miestamo2009}. 
In such a reading, in sentences like \REF{ex:rum:76c} and \REF{ex:rum:77c}, not the thinking would be negated but the coming, and the sentences would be nearly synonymous with \REF{ex:rum:76b} and \REF{ex:rum:77b}, respectively. However, in Rumai there is no evidence for this phenomenon. The two clauses in the corresponding Rumai sentences (\ref{ex:rum:76}, here repeated from \ref{ex:rum:51}) and \REF{ex:rum:77} have both AVP word order as it is usual in independent clauses, and as in other independent clauses, they can be negated by \textit{ʔaːw-} and \textit{{}-maʔ}, the negators that never occur in dependent ones. So, in these Rumai sentences there is rather coordination than subordination. Furthermore, the sentences can be translated by using different negators, as it is shown in \REF{ex:rum:76} and \REF{ex:rum:77}. The translations are from the same consultant but elicited independently at different times.

\ea\label{ex:rum:76}
    \ea\label{ex:rum:76a}
        \gll ʔʌ̂w	tʰaŋ	ʔân	nʌŋ-deəh \\ 
             1\textsc{sg} think 3\textsc{sg} \textsc{irr}{}-come \\
        \glt ‘I think he will come.’
    \ex\label{ex:rum:76b}
        \gll ʔʌ̂w	tʰaŋ	ʔân	nʌŋ-\textbf{ʔaːw}-deəh(-maʔ) \\
             1\textsc{sg} think 3\textsc{sg} \textsc{irr}{}-\textsc{neg}{}-come-\textsc{neg} \\
        \glt ‘I think he will not come.’
    \ex\label{ex:rum:76c}
        \gll ʔʌ̂w	\textbf{ʔaːw}-tʰaŋ-\textbf{maʔ}	ʔân	nʌŋ-deəh \\
             1\textsc{sg} \textsc{neg}{}-think-\textsc{neg} 3\textsc{sg} \textsc{irr}{}-come \\
        \glt ‘I don’t think he will come.’
    \ex\label{ex:rum:76d}
        \gll ʔʌ̂w	\textbf{ʔaːw}-tʰaŋ-\textbf{maʔ}	ʔân	nʌŋ-\textbf{ʔaːw}-deəh(-maʔ) \\
             1\textsc{sg} \textsc{neg}{}-think-\textsc{neg} 3\textsc{sg} \textsc{irr}{}-\textsc{neg}{}-come-\textsc{neg} \\
        \glt ‘I don’t think he will not come.’ (pc\_3)
    \z
\z 

\ea\label{ex:rum:77}
    \ea\label{ex:rum:77a}
        \gll ʔʌ̂w	tʰaŋ	ʔân	nʌŋ-deəh \\ 
             \textsc{1sg} think 3\textsc{sg} \textsc{irr}{}-come \\
        \glt ‘I think that he will come.’
    \ex\label{ex:rum:77b}
        \gll ʔʌ̂w	tʰaŋ	ʔân	nʌŋ-\textbf{buː}-deəh-\textbf{maʔ} \\
             \textsc{1sg} think \textsc{3sg} \textsc{irr}{}-\textsc{neg}{}-come-\textsc{neg} \\
        \glt ‘I think that he will not come (anymore).’
    \ex\label{ex:rum:77c}
        \gll ʔʌ̂w	tʰaŋ-\textbf{maʔ}	ʔân	nʌŋ-deəh \\
             \textsc{1sg} think-\textsc{neg} \textsc{3sg} \textsc{irr}{}-come \\
        \glt ‘I don’t think that he will come.’
    \ex\label{ex:rum:77d}
        \gll ʔʌ̂w	tʰaŋ-\textbf{maʔ}	ʔân	nʌŋ-\textbf{buː}-deəh \\
             \textsc{1sg} think-\textsc{neg} \textsc{3sg} \textsc{irr}{}-\textsc{neg}{}-come \\
        \glt ‘I don’t think that he will not come.’ (pc\_3)
    \z
\z 

\subsubsection{Diachronic notes and observations} \label{sec:rum:4.6.2}

The introduction of \textit{{}-maʔ} was probably due to reinforcement of the negation and there might be a Jespersen Cycle, as it is known in other languages and shown for French by \citet[78--79]{Auwera2010}. The possible stages for Palaung are shown in \tabref{tab:rum:5}, which follows \citet[79]{Auwera2010}\footnote{For Jespersen Cycle see also \citet{Jespersen2025}}

\begin{table}
\begin{tabularx}{\textwidth}{Xllllllll}
\lsptoprule
{Stage 1}       & {→} & {Stage 2}               & {→} & {Stage 3}             & {→} & {Stage 4}              & {→}   & {Stage 5}\\
{\itshape buː-} &     & {\itshape buː-}         &     & {\itshape buː-}       &     & {\itshape buː- {}-maʔ} &    & {\itshape {}-maʔ} \\
                &     & {\itshape buː- {}-maʔ}  &     & {\itshape buː- {}-maʔ}&     & {\itshape {}-maʔ}      &       &\\
                &     &                         &     & {\itshape {}-maʔ}     &     &                        &       &\\
\lspbottomrule
\end{tabularx}
\caption{Stages of negation in Rumai}
\label{tab:rum:5}
\end{table}


However, there is a difference between dependent and independent clauses. While the dependent clauses still stay on Stage 1, the independent ones are on Stage 3. As already mentioned, most of the dependent clauses are verb initial, and a postverbal negator in verb-initial languages is highly unusual, as \citegen{Dryer1988} study shows.

Dependent clauses are, as described by \citet{Bybee2001}, more conservative than independent ones. Therefore, the original general constituent order in Palaung was probably VS/VAP. Then, one can assume that the general negator in Rumai was originally \textit{buː{}-}. Only after the constituent order in independent clauses completely changed to AVP, was the postverbal \textit{{}-mɑʔ} introduced. \citep{Weymuth2014}

\section{Summary} \label{sec:rum:5}

In Rumai, the expression of negation is dependent on the clause type. While the negative morphemes \textit{ʔaːw-} ‘\textsc{neg}’, \textit{buː-} ‘\textsc{neg}’, \textit{{}-maʔ} ‘\textsc{neg}’ and \textit{ɲjʌm} ‘not yet’ can all be used in independent clauses, the use is restricted to \textit{buː-} and \textit{ɲjʌm} in dependent ones. The constituent order change from VAP to AVP in the Rumai independent clauses is probably the reason for this feature, because postverbal negators are very unusual in verb initial languages. Additionally, there are the prohibitives \textit{kʰuː{}-} and \textit{maj-}. The latter seems only to be used, when a corresponding positive imperative clause does not make sense and therefore, it may rather express a well-meaning advice. The negative suffix \textit{{}-maʔ} was originally a means of reinforcing the negation but today it is widespread, and it can also be used alone, so that it is rather a neutral negator now. There might be a Jespersen Cycle with only \textit{buː-} on stage 1, resulting in \textit{{}-mɑʔ} as the only negator on stage 5. However, the dependent clauses are still on stage 1, and the independent ones are on stage 3, where the two negators can either be used alone or together. Furthermore, the construction \textit{buː-}V\textit{{}-mɑʔ} often expresses the meaning of ‘not anymore’ in independent clauses. Several aspects of negation in Rumai, like the scope of negation, are still unclear and need more research. \tabref{tab:rum:6} summarizes the Rumai morphemes connected to negation and their functions.

\begin{table}
\begin{tabularx}{\textwidth}{lX}
\lsptoprule
{Morpheme} & {Function}\\
\midrule
{\itshape buː-} & {Main negator in dependent clauses}

{In independent clauses, in combination with \textit{{}-maʔ}, implicitly or explicitly ‘not anymore’}\\
\tablevspace
{\itshape ʔaːw-} & {Main negator in independent clauses}\\
\tablevspace
{\itshape {}-maʔ} & {In combination with \textit{buː-}, \textit{ʔaːw-} and \textit{ɲjʌm}, sometimes reinforcement of negation}

{Negator in absence of another negator}\\
\tablevspace
{\itshape ɲjʌm} & {‘not yet’}\\
\tablevspace
{\textit{kʰ}\textit{uː-}} & {Prohibitive}\\
\tablevspace
{\itshape maj-} & {Prohibitive}\\
\tablevspace
{\itshape hɯh} &  {Question particle used in negative questions, but not itself negative}\\
\lspbottomrule
\end{tabularx}
\caption{Rumai morphemes connected to negation}
\label{tab:rum:6}
\end{table}
\largerpage[2]

\section*{Abbreviations}
\begin{tabularx}{.45\textwidth}{lQ}
1 & first person \\
2 & second person \\
3 & third person \\
\textsc{ana} & anaphoric demonstrative \\
\textsc{asrt} & assertive \\
\textsc{clf} & classifier \\
\textsc{comp} & complementizer \\
\textsc{conc} & concessive \\
\textsc{cond} & conditional \\
\textsc{des} & desiderative \\
\textsc{disc} & discourse marker \\
\textsc{dur} & durative \\
\textsc{excl} & exclusive \\
\textsc{exp} & experiential \\
\textsc{incl} & inclusive \\
\textsc{ipfv} & imperfective \\
\textsc{irr} & irrealis \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\textsc{loc} & locative \\
\textsc{med} & medial \\
\textsc{neg} & negative \\
\textsc{nmlz} & nominalizer \\
\textsc{nsit} & new situation \\
\textsc{obl} & oblique \\
\textsc{pl} & plural \\
\textsc{pn} & proper name \\
\textsc{pol} & polite \\
\textsc{proh} & prohibitive \\
\textsc{prox} & proximal \\
\textsc{ptcl} & particle \\
\textsc{q} & question marker \\
\textsc{rel} & relativizer \\
\textsc{restr} & restrictive \\
\textsc{sg} & singular \\
\textsc{tcl} & topic-comment linker \\
\end{tabularx}


\begin{paperappendix}
\section*{Appendix}\label{sec:rum:Appendix}
%     \hc[noline,size=\tiny]{@LSP: we would like to have an appendix here that we can refer to in the text, but ideally unnumbered here, and referenced in the text as "Appendix". Shall we use begin paperappendix instead? Could you find a solution for that? Thanks. The appendix goes until right before the References.}
\begin{table}
\caption{Structure of the data labels}
\begin{tabularx}{\textwidth}{llll}
\lsptoprule
\multicolumn{2}{l}{First letter of the shortcuts}  & \multicolumn{2}{l}{Places} \\
\midrule
C&conversation                  & M &Mandalay \\
E&elicitation                   & MS& Man Sat (Nam Hkam township) \\
N&narrative                     & L &Lashio \\
P&picture story                 & \\
W&written text                  & \\
\lspbottomrule
\end{tabularx}
\parbox{\textwidth}{\raggedright\footnotesize
1. Short name: shortcut\_year\_place of recording/edition\\
2. \_consultant(s)/author\_example in toolbox\\
3. pc\_X: personal conversation with consultant X}
\end{table}

\begin{table}
\caption{Sources of the examples}

\begin{tabularx}{\textwidth}{>{\ttfamily}lQ}
\lsptoprule
{\rmfamily Short name} & {Description}\\
\midrule
{C3\_17\_MS} & {Conversation about the life in a village}\\
{EGGD\_16\_M} & {Questionnaire about the use of “give” and “get” and about ditransitive constructions, English}\\
{ENEG\_13\_M} & {Questionnaire about negation, Burmese/English}\\
{ENEG\_14\_M} & {Questionnaire about negation, Burmese}\\
{EQB\_17\_M} & {Questionnaire about Burmese constructions, Burmese/English}\\
{EMR\_19\_M} & {Questionnaire resumptive pronoun, Burmese}\\
{EMSR\_19\_M} & {Questionnaire subordinate clauses, Burmese/English}\\
{N2\_17\_MS} & {About green tea cultivation in a village}\\
{PCHT\_16\_L} & {Picture story: The chicken thief.} {Béatrice Rodriguez. 2008. \textit{Der Hühnerdieb}. Wuppertal: Peter Hammer Verlag.}\nocite{Rodriguez2008}\\
{WA\_15\_M} & Sentences containing aspectual markers from: Sa Pe. 2015. \textit{bjat gɔ̂ːj kʰêəŋ bèə} [Being clever with texts]. Mandalayː Self-publishing.\nocite{SaPe2015}\\
{WPKT\_14\_L} & {Phrase book Rumai – Shwe – Burmese – English. 2014. Lashio: Ta’ang Students and Youth Union.}\nocite{Unknown2014}\\
{W1} & {Story in \textit{Our Ta’ang magazine}}\\
\lspbottomrule
\end{tabularx}
\end{table}

\begin{table}
\caption{Consultants}
\begin{tabularx}{\textwidth}{lllQQQQ}
\lsptoprule
{Nr.} & {Gender} & {Y.o.b.} & {Origin} & {Education} & {Occupation} & {Languages}\\
\midrule 
{1} & {f} & {1996} & {Na Aw Gyi village, Man Ton township}  &                       & {bachelor student} & {Rumai, Burmese, English}\\
\tablevspace
{2} & {m} & {1994} & {Man Ton township}                     &                       & {bachelor student, novice} & {Rumai, Burmese, English}\\
\tablevspace
{3} & {m} & {1988} & {Man Sat village, Nam Hkam township}   & {Bachelor diploma}    & {monk} & {Rumai, Burmese, English, Pali}\\
\tablevspace
{6} & {m} & {1993} & {Sar Lu village, Nam Hkam township}    &                       & {student} & {Rumai, Burmese}\\
\tablevspace
{7} & {f} & {1959} & {Man Sat village, Nam Hkam township}   & {Public school in Shan} & {housewife, farmer} & {Rumai, Shan, (Burmese)}\\
\tablevspace
{9} & {f} & {1998} & {Pha Daen village, Nam Hkam township}  & {Public school in Burmese} & {housewife, farmer} & {Rumai, Burmese, (English)}\\
\lspbottomrule
\end{tabularx}
\end{table}
\end{paperappendix}

\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
\end{document} 
