\documentclass[output=paper,colorlinks,citecolor=brown,draftmode]{langscibook}
\ChapterDOI{10.5281/zenodo.18233766}
\author{Eloisa Ruppert\affiliation{Catholic University of Leuven} %\orcid{orcid.org/0000-0002-9514-0641}
and Saudah Namyalo\affiliation{Makerere University} and Alena Witzlack-Makarevich\affiliation{Hebrew University of Jerusalem}%\orcid{https://orcid.org/0000-0003-0138-4635}
}
\title{Negation in Ruuli}
\abstract{This paper explores the different aspects of negation in the previously undescribed Great Lakes Bantu language Ruuli (JE.103). 
It first considers clausal negation. 
All tense-aspect combinations are negated by means of the prefix \textit{ti-}. 
Except for the near future tense, the constructions are symmetric.
In the negative imperative, the negation prefix \textit{ta-} is used for various constructions which are distinguished in the positive. 
Thus, we classify Ruuli as a language that has a special negative imperative construction and a special negator in the negative imperative which is not used in standard negation. 
In non-verbal positive clauses, a range of copulas is used. 
The negative construction of non-verbal predication depends on the type of predication, as well as on tense-aspect. 
In addition to the invariable negative copula \textit{ti}, the copulas \textit{li} and \textit{ta}, as well as the semi-copula \textit{bba} are negated by means of the standard negation marker \textit{ti-} in non-present. 
Furthermore, there is a dedicated negative existential \textit{ndoo}. 
The article further provides an overview of non-clausal negation and negation in non-main clauses, as well as discusses negative replies, phasal polarity items, and negative indefinites.

\keywords{negation, prohibitive, non-verbal predication, Bantu}
}
 
\IfFileExists{../localcommands.tex}{%hack to check whether this is being compiled as part of a collection or standalone
    \input{../localpackages}
    \input{../localcommands}
%    \input{../localhyphenation}
%     \bibliography{localbibliography}
    \togglepaper[23]
}{}

\begin{document}
\maketitle

\section{The language}\label{sec-intro}

This paper deals with negation in Ruuli. 
It closely follows the questionnaire provided by the editors \parencitetv{chapters/questionnaire} and mirrors its structure.
Ruuli (ISO: \href{https://iso639-3.sil.org/code/ruc}{ruc}, Glottocode: \href{https://glottolog.org/resource/languoid/id/ruul1235}{ruul1235}, also known as Ruruuli-Lunyala) is a Great Lakes Bantu language (Niger-Congo). 
Within Great Lakes Bantu, Ruuli belongs to the Rutara group of the West Nyanza branch \citep{Schoenbrun1997}.
Apart from Ruuli, the Rutara group includes such better-described languages as Nkore-Kiga (ISO: \href{https://iso639-3.sil.org/code/cgg}{cgg} and \href{https://iso639-3.sil.org/code/nun}{nun}) and Haya (ISO: \href{https://iso639-3.sil.org/code/hay}{hay}). 
Ruuli is spoken mainly in the Nakasongola and Kayunga districts of the central region of Uganda. 
The number of ethnic members of the community is over 230,000 (190,122 indicated to be Baruuli, 47,699 indicated to be Banyala, \citealt{Uganda2016}). 
The number of Ruuli speakers is difficult to determine, we believe it to be somewhat smaller than the number of people who self-identify as Baruuli or Banyala.

Ruuli is not an official language of Uganda and the orthography has been only recently introduced by SIL. 
Most speakers are multilingual: 
They are often fluent in Ganda, the language of the majority in the area, as well as in English, the institutional language of Uganda. 
Furthermore, many speakers also speak languages of the neighboring ethnic communities, these are closely related Bantu languages, such as Soga and Nyoro.

Ruuli is a typical Bantu language: 
The dominant constituent order is SVO. 
Each noun belongs to one of the 20 noun classes. 
The noun classes are numbered from 1 to 23 corresponding to the reconstructed Proto-Bantu noun classes \citep[237–239]{VandeVelde2019}. 
Ruuli does not have the correspondences of the noun classes 19 and 21. 
Most nouns have both a singular and a plural form, which belong to two different classes. 
As in many other Bantu languages, nouns from a specific class in singular take a plural from a corresponding noun class in plural. 
The most common pairs (in descending order) are 1/2, 7/8, 9/10, and 3/4. 
The noun class prefixes on the nouns – even if they are present – are not segmented in the examples, but are indicated in the glosses, as e.g.\,\textit{Kampala} `1.Kampala’ in (\ref{ex:still}).

Ruuli nouns frequently carry an augment, also known as pre-prefix or initial vowel (see e.g.\,\citealt[247–255]{VandeVelde2019}). 
The augment precedes the noun class prefix and has the forms \textit{a-}, \textit{o-}, or \textit{e-} depending on the noun class. 
All augments are glossed as \textsc{aug}, as e.g.\,the augment on \textit{o-Kato} `\textsc{aug}-1.Kato' in~(\ref{ex:still}). 
The presence of an augment on the head noun and other elements of a noun phrase is conditioned by a combination of factors, including propositional act  (e.g.\,question vs.\,assertion), syntactic context, as well as referential properties of the respective argument \citep[cf.][249]{VandeVelde2019}. 
Also negation interacts with the presence of the augment. 
This effect is discussed in \sectref{sec:markingscoperuuli}.

The verbal inflectional morphology is primarily prefixing. 
Prefixes express such inflectional categories as negation, argument indexing, tense and aspect. 
Suffixes mark derivational categories, such as the causative and applicative, as well as inflectional categories, such as mood and further tense and aspect categories.  
The glosses of the argument indexing on the verb follow two conventions: the first and second person arguments are glossed by indicating the person and number, as in \textit{n-ku-yab-a} (1\textsc{sg.sbj}-\textsc{prog}-go-\textsc{fv}) ‘I am going’ in~(\ref{ex:presprog1}). 
The form of the third person indexes depends on the noun class and this is reflected in the glosses, which only indicate the respective class. 
That is, if the numeral 1 or 2 in the glosses is used in combination with the indication of number (\textsc{sg} or \textsc{pl}), it refers to the person, when it is used without the indication of number, the gloss indicates the noun class. 

Ruuli is a tonal language. 
The first account of its nominal and verbal tone is given in \citet{Kiriggwajjo2022}, however, other aspects of its tone, including, its interaction with negation are still being analyzed. 
In the examples in this article we indicate surface tone.\footnote{We are grateful to Anatole Kiriggwajjo for his expertise and support in marking surface tone in the examples in this article.} 
Though subsequent tone analysis might lead to the modification of some aspects of what is presented below and will definitely enrich the analysis, we are certain that most claims of this paper will remain unaffected.

The data used in the present study come from a corpus of naturalistic speech comprising over 200,000 words \citep[]{Witzlack-Makarevichetal2021Corpus}, as well as from deliberate elicitations.
The corpus was collected in 2017 and 2018 in various locations in Uganda by the team of the project \textit{A comprehensive bilingual talking Luruuli/Lunyara-English dictionary with a descriptive basic grammar for language revitalisation and enhancement of mother-tongue based education} (PI Saudah Namyalo, funded by Volkswagen Foundation).


\section{Clausal negation}\label{sec:clausalruuli}

This section presents clausal negation constructions in Ruuli. 
§\ref{sec:standardruuli} provides data on standard negation, i.e.\,negation in declarative verbal main clauses. §\ref{sec:non-declarativesruuli} describes negation of non-declarative verbal main clauses. 
In §\ref{sec:non-verbalruuli} we discuss non-verbal clauses and how they are negated. 
Finally, §\ref{sec:non-mainruuli} displays negation in non-main clauses in Ruuli.

This section does not discuss negative lexicalizations for `not speak', `not like', `not know', `cannot' or `not want', as they are not attested in Ruuli.    
The only lexically idiosyncratic negative is the lexicalized negative existential \textit{ndoo-} discussed in §\ref{sec:non-verbalruuli}. 

\subsection{Standard negation}
\label{sec:standardruuli}\label{sec:ruulietal:2.1}
The system of negation in Ruuli is similar to what has been described for other Bantu languages, which we first outline in what follows. 
The morphological structure of the Bantu finite verb has been established since \citet[108–111]{Meeussen1967} and is given in \tabref{tab:Bantuslots} following \citet[546]{Güldemann1999}. 
Across Bantu, the coding of negation is possible in three slots, viz.\,pre-initial, post-initial and post-final.

\begin{table}
\small
    \begin{tabularx}{\textwidth}{Q|l|Q|Q|Q|p{16mm}|l|Q}
%     \lsptoprule
         pre-initial & initial & post-initial & pre-radical  & radical & pre-final & final & post-final\\
         \midrule
         \scshape tam/ neg & \scshape sbj & \scshape tam/ neg & \scshape obj & verbal root & derivation/ \textsc{tam}& \textsc{tam} & clause/ object \textsc{neg}\\
%     \lspbottomrule
    \end{tabularx}
    \caption{Morphological slots of finite verbs in Bantu}
    \label{tab:Bantuslots}
\end{table}


Many Bantu languages show a regular, syntactically conditioned distribution of pre- and post-initial negatives. 
These distributional restrictions are of the following type: whereas the pre-initial negation shows a strong tendency to be confined to unmarked main clauses, the post-initial negation strategy is reported to rarely underlie distributional restrictions with regard to major clause types according to \citet[551]{Güldemann1999}. 
A functional explanation of this dichotomy is provided in \citet[305–308]{Güldemann1996} and \citet[570--576]{Güldemann1999}.


The verb in Ruuli has an extensive tense-aspect system. 
At the time of writing, 25 possible tense and aspect combinations have been recognized. 
All tense-aspect combinations are negated by means of the prefix \textit{ti-}, see~(\ref{ex:standarddrop}) for an illustration of negation in the present progressive. 
The prefix is realized as \textit{t-} before vowels, e.g.\,before the second person singular subject prefix \textit{o-} in~(\ref{ex:presprog2neg}).

\ea\label{ex:standarddrop}\label{ex:ruulietal:2}
\begin{xlist}
\ex  \label{ex:presprog1}
	\glll Nkwabâ.\\
	    n-ku-yab-a\\
		1\textsc{sg.sbj}-\textsc{prog}-go-\textsc{fv}\\
	\glt ‘I am going.’  (elicited)
\ex \label{ex:presprog1neg}
	\glll  Tínkwâba.\\
	        ti-n-ku-yab-a\\
		    \textsc{neg}-1\textsc{sg.sbj}-\textsc{prog}-go-\textsc{fv}\\
	\glt    ‘I am not going.’ 	 (elicited)
\ex \label{ex:presprog2}
	\glll   Okwabâ.\\
	        o-ku-yab-a\\
		    \textsc{2sg.sbj}-\textsc{prog}-go-\textsc{fv} \\
	\glt    `You are going.’  (elicited)
\ex \label{ex:presprog2neg}
	\glll   Tókwâba.\\
	        ti-o-ku-yab-a\\
		    \textsc{neg}-\textsc{2sg.sbj}-\textsc{prog}-go-\textsc{fv} \\
	\glt    ‘You are not going.' (elicited)
\end{xlist}
\z

Except for the near future tense, the negation constructions for all tense and aspect combinations in Ruuli are symmetric in the sense of \citet{Miestamo2005, Miestamo2017}. 
In the negative near future, the indicative suffix \textit{-a} – glossed as \textsc{fv} for final vowel, as is common in the descriptions of Bantu languages, see e.g.\,\citet[44]{Nurse2008} – is replaced by the subjunctive suffix \textit{-e}, as in~(\ref{ex:standardasyconstr}).
Thus, the negative near future construction shows A/Cat/TAM asymmetry. 
This is not the case for any of the other tense and aspect combinations. 

\ea\label{ex:standardasyconstr}\label{ex:ruulietal:3}
\begin{xlist}
\ex \label{ex:near.future}
	\glll   Nayabá.\\  
	        n-a-yab-a\\
		    1\textsc{sg.sbj}-\textsc{fut}-go-\textsc{fv}\\
	\glt ‘I will go (soon).’  (elicited)
\ex \label{ex:near.futureneg}
	\glll   Tínáyabê.\\
	        Ti-n-a-yab-e.\\
		    \textsc{neg}-1\textsc{sg.sbj}-\textsc{fut}-go-\textsc{sbjv}\\
	\glt ‘I will not go (soon).’ 	 (elicited)
\ex \label{ex:near.futureasterix}
	\gll *ti-n-a-yab-a.\\
		\textsc{neg}-1\textsc{sg.sbj}-\textsc{fut}-go-\textsc{fv} \\
\glt Intended: ‘I will not go (soon).’ (elicited)
\end{xlist}
\z

Ruuli makes use of the persistive prefix \textit{kya-} to express the phasal polarity concepts \textsc{still}, \textsc{not yet}, and \textsc{no longer} (see \citealt{Kramer2021} for a theoretical overview and  \citealt{MolochievaEtAl2021} for a description of phasal polarity items in Ruuli). 
When attached to the main verb, the prefix \textit{kya-} expresses \textsc{still}, as in~(\ref{ex:still}). 
By adding the standard negator \textit{ti-} to a verb that is marked by the persistive \mbox{\textit{kya-},} the concept \textsc{no longer} is expressed, as in~(\ref{ex:nolonger}). 
The construction denoting \textsc{not yet} is a periphrastic construction formed by the combination of a positive infinitive and the auxiliary \textit{li}\footnote{The item \textit{li} has two major functions in Ruuli. 
On the one hand, it serves as a copula for non-verbal predication, as in~(\ref{ex:near.future}) and~(\ref{ex:near.futureneg}). 
On the other hand, it surfaces in a range of analytically built tense-aspect forms, as e.g.\,in~(\ref{ex:nobody}). 
The same holds for \emph{bba} discussed in §\ref{sec:non-verbalruuli}, which serves as a copula, as well as an auxiliary, as in~(\ref{ex:complexTA}).}
which carries the prefix \textit{kya-}, as in~(\ref{ex:pers}). 
Thus, this is a semantically negative construction built without using the standard negator \textit{ti-} and lacking a direct affirmative counterpart. 
The use of a periphrastic construction is typical of many Bantu languages, however, in contrast to the majority of Bantu languages with a \textsc{not yet} periphrastic construction, this construction in Ruuli does not contain a negation marker \citep{VeselinovaDevos2021}. 

There is no dedicated construction for the expression of the corresponding affirmative semantics (i.e.\,the phasal polarity item \textsc{already}) in Ruuli.\footnote{Note that the corresponding affirmative semantics is not expressed by the negation of the construction expressing \textsc{not yet}. 
The construction of a negative persistive is formally possible and has a different meaning than the positive counterpart, but was rejected by the speakers probably because of semantic issues with the English translation.} 
Instead, one either uses the adverb \textit{(e)irai} with the primary meaning `in the past, long ago' or the periphrastic construction with the complement-taking predicate \textit{mala} ‘finish’ (see \citealt{MolochievaEtAl2021}). 

\ea\label{ex:standardpersistive}\label{ex:ruulietal:4}
\begin{xlist}
	\ex \label{ex:still}
	\glll   OKátô akyâli e Kámpálâ.\\ %it is indeed correct to keep two capital letters at the beginning here (and at a few other places too)
	        o-Kato a-kya-li e Kampala\\
		    \textsc{aug}-1.Kato 1\textsc{sbj}-\textsc{persis}-\textsc{cop} 23.\textsc{loc} 1.Kampala\\
\glt        ‘Kato is still in Kampala.’  (elicited)
	\ex \label{ex:nolonger}
	\glll    OKátô takyâlí e Kámpálâ.\\
	        o-Kato ti-a-kya-li e Kampala\\
		    \textsc{aug}-1.Kato \textsc{neg}-1\textsc{sbj}-\textsc{persis}-\textsc{cop} 23.\textsc{loc} 1.Kampala\\
\glt        ‘Kato is no longer in Kampala.’  (elicited)
	\ex \label{ex:pers}
	\glll    Nkyâli kwabâ.\\
	        n-kya-li ku-yab-a.\\
		    1\textsc{sg.sbj}-\textsc{persis}-\textsc{aux} \textsc{inf}-go-\textsc{fv}\\
\glt        ‘I have not gone yet.’ (elicited)
\end{xlist}
\z

Some tense-aspect forms consist of a combination of an auxiliary and the main verb, e.g.\,the recent past progressive. 
In these forms, the standard negation marker \textit{ti-} is prefixed to the main verb only. Compare the two examples in~(\ref{ex:complexTA}).\footnote{Note that Ruuli does not have geminate consonants, in the standard orthography <bb>, as in~\REF{ex:complexTA-pos}, represents [b], whereas <b> represents [β], see \citet[27–28]{NamyaloEtAl2021} for details.} 

\ea  \label{ex:complexTA}
	\ea \label{ex:complexTA-pos}
	\glll  Abbáíré áikalá mpáni obwômi bwámwé bwónâ.\\
	    a-bba-ire a-ikal-a mpani obwomi bu-amwe bu-ona\\
		\textsc{1sbj}-\textsc{aux}-\textsc{pfv} \textsc{1sbj}-live-\textsc{fv} here 14.life 14-1.\textsc{poss} 14-all\\
\glt ‘He has been living here for all his life.'  (elicited)

\ex \label{ex:complexTA-neg}
	\glll Emábégâ aBányálâ babbairé tibátákâ kusómâ.\\
	    e-mabega a-Banyala ba-bba-ire ti-ba-tak-a kusoma\\
		\textsc{aug}-6.past \textsc{aug}-2.Banyala 2\textsc{sbj}-\textsc{aux}-\textsc{pfv} \textsc{neg}-2\textsc{sbj}-want-\textsc{fv} 15.education\\
\glt  `In the past, the Banyala did not want education.’ 	(ll-N-KYERIMA-170218-NA-8)
\z
\z

\subsection{Negation in non-declaratives}\label{sec:non-declarativesruuli}\label{sec:ruulietal:2.2}

This section addresses the non-declarative mood categories of Ruuli viz.\,impera\-tive, hortative and jussive.

Ruuli has two imperative constructions.\footnote{Deviations from these two patterns are observed with reduplicated verbs and lexicalized reflexive verbs.} 
The first one is used for the second person singular imperative. 
It is formed by the verb stem with optional extensions followed by the final vowel \textit{-a}, as in~(\ref{ex:normalimp1}). 
It is the only independent verb form in Ruuli that lacks the subject prefix. 
The second construction is used for the second person plural imperative. 
It is formed by prefixing the regular second person plural subject prefix to the verb stem and suffixing the subjunctive suffix \textit{-e},\footnote{Both the form and the function of this suffix are similar to the cognate ones in closely related Great Lakes Bantu languages \citep{NurseMuzale1999}, as well as in many other Bantu languages \citep[44, 192]{Nurse2008}, for details, see \citet{VeselovskyEtAl2023}. 
Final *-e has been reconstructed for proto-Bantu \citep{Meeussen1967}.}
as in~(\ref{ex:normalimp2}). 
This imperative construction occurs in a range of Bantu languages, see \citet{DevosVanOlmen2013}.

\ea \label{ex:normalimp}
\begin{xlist}
	\ex \label{ex:normalimp1}
	\glll   Izâ.\\
	        iz-a\\
		    come-\textsc{fv}\\
    \glt    ‘Come (sg.)!’  (elicited)
	
	\ex \label{ex:normalimp2}
	\glll   Mwîze.\\
	        mu-iz-e\\
		    2\textsc{pl.sbj}-come-\textsc{sbjv}\\
    \glt    ‘You (pl.), come!' (elicited)
\end{xlist}
\z

Finally, imperatives with object indexing deviate from the two major patterns discussed above: 
They take the subjunctive suffix both with a second person singular and plural addressee, compare~(\ref{ex:impobj1}), which contains a nominal object, with~(\ref{ex:impobj2}--\ref{ex:impobj3}), which contain object indexes. 
The use of the final vowel \textit{-a} in the imperative construction with object indexes is ungrammatical independent of the addressee's grammatical number, as in ~(\ref{ex:impobj2wrong}--\ref{ex:impobj3wrong}).

\ea \label{ex:impobj} 
\begin{xlist}
\ex \label{ex:impobj1} 
	\glll  Kûbba enkáító zaaꜜmú!\\
	        kubb-a enkaito za-amu\\
		    polish-\textsc{fv} 10.shoe 10-2\textsc{sg.poss}\\
	\glt    ‘Polish your shoes!’  (elicited)
\ex \label{ex:impobj2}
	\glll  Zikubbê!\\
	    zi-kubb-e\\
		10\textsc{obj}-polish-\textsc{sbjv}\\
	\glt ‘Polish them (e.g.\,shoes)!’ (elicited)
	
\ex \label{ex:impobj3}
	\glll   Múzíkuꜜbbé!\\
	        mu-zi-kubb-e\\
		    2\textsc{pl.sbj}-10\textsc{obj}-polish-\textsc{sbjv}\\
	\glt ‘You (pl.), polish them (e.g.\,shoes)!’ (elicited)

\ex \label{ex:impobj2wrong}
	\gll *zi-kubb-a\\
		10\textsc{obj}-polish-\textsc{fv}\\
	\glt Intended: ‘Polish them (e.g.\,shoes)!’ (elicited)

\ex \label{ex:impobj3wrong}
	\gll *mu-zi-kubb-a\\ 
		2\textsc{pl.sbj}-10\textsc{obj}-polish-\textsc{fv}\\
	\glt Intended: ‘You (pl.), polish them (e.g.\,shoes)!’ (elicited)
\end{xlist}
\z

In the negative imperative, the post-initial negation marker \textit{ta-} and the final vowel \textit{-a} are used, as in~(\ref{ex:prohibnormal}).\footnote{This is in contrast to some other Bantu languages (cf.\,e.g. \citealt[101]{Harjula2004Ha}), where negative imperatives always take the subjunctive \textit{-e} instead of the final vowel.} 
The same construction is used with the second person plural addressee, as in~(\ref{ex:prohib2pl}), as well as with verb 
forms with object agreement prefixes, as in~ (\ref{ex:prohibobj}).  

\ea \begin{xlist}\label{ex:ruulietal:8}
\ex \label{ex:prohibnormal}
	\glll   Otákúbbâ nkáítô zaamú!\\ 
	        o-ta-kubb-a nkaito za-amu\\
		    \textsc{2sg.sbj}-\textsc{neg}-polish-\textsc{fv} 10.shoe 10-2\textsc{sg.poss}\\
	\glt ‘Don't polish your shoes!’ (elicited)

\ex \label{ex:prohib2pl}
	\glll   Mutákúbbâ nkáítô!\\
	        mu-ta-kubb-a nkaito\\
		    \textsc{2pl.sbj}-\textsc{neg}-polish-\textsc{fv} 10.shoe\\
	\glt ‘You (pl.), don't polish the shoes!’ (elicited)

\ex \label{ex:prohibobj}
	\glll   Otázíkubbâ!\\
	        o-ta-zi-kubb-a\\
		    \textsc{2sg.sbj}-\textsc{neg}-10\textsc{obj}-polish-\textsc{fv}\\
    \glt ‘Don't polish them (e.g.\,the shoes)!’ (elicited)

\end{xlist}
\z

When the imperative construction consists of two verbs, e.g.\,a verb of motion followed by another verb, the second verb takes the regular second person singular or plural subject prefix and the subjunctive suffix, as in~(\ref{ex:2verbimpsing}). 
If the addressee is the second person plural, as expected, also the first verb takes the subject prefix and the subjunctive suffix, as in~(\ref{ex:2verbimppl}). 

\ea \label{ex:2verbspos}
\begin{xlist}
\ex \label{ex:2verbimpsing}
	\glll  Izâ ozingululé ólúkoꜜsí!\\
	    iz-a o-zingulul-e o-lukosi\\
		come-\textsc{fv} \textsc{2sg.sbj}-uncoil-\textsc{sbjv} \textsc{aug}-11.thread\\
    \glt ‘Come and uncoil the thread!’ 	 (elicited)

\ex \label{ex:2verbimppl}
	\glll  Mwîzé mwéróoꜜré okú músyólî!\\
	        mu-iz-e mu-eroor-e 	o-ku musyoli\\
		    \textsc{2pl.sbj}-come-\textsc{sbjv} \textsc{2pl.sbj}-look-\textsc{sbjv}  \textsc{aug}-17.\textsc{loc} 1.magician\\ 
    \glt ‘You (pl.), come and look at the magician!’  (elicited)
\end{xlist}
\z

The corresponding prohibitive construction is slightly different from the imperative construction with one verb. 
Whereas the second verb carries a subject prefix and the subjunctive suffix in the affirmative, as in~(\ref{ex:2verbspos}), in the prohibitive the infinitive is used instead both with singular and plural addressees, as in~(\ref{ex:prohib2verb}). 


\ea \label{ex:prohib2verb}\label{ex:ruulietal:10}
\begin{xlist}
\ex \label{ex:prohib2verbsing}
	\glll  Otáyábâ kusyómá maiꜜzí.\\
	        o-ta-yab-a ku-syom-a maizi\\
		    \textsc{2sg.sbj}-\textsc{neg}-go.away-\textsc{fv} \textsc{inf}-fetch-\textsc{fv} 6.water\\
    \glt ‘Don't go and bring water.’  (elicited)
\ex \label{ex:prohib2verbpl}
	\glll   Mutáyábâ kusyómá maiꜜzí.\\
	        Mu-ta-yab-a ku-syom-a maizi\\
		    2\textsc{pl.sbj}-\textsc{neg}-go.away-\textsc{fv} \textsc{inf}-fetch-\textsc{fv} 6.water\\
    \glt ‘You (pl.), don't go and bring water.’  (elicited)
\end{xlist}
\z

Taking the different types of negative imperatives into consideration, we classify Ruuli as a language that has a special negative imperative construction and a special negator (the post-initial \emph{ta-}) in the negative imperative which is not used in standard negation. 
Ruuli thus qualifies as a type four language in the typology of \citet{AuweraEtAl2013}, i.e.\,a language in which ``the verbal construction of the second singular prohibitive is different from that of the second singular imperative, and the negative strategy is different from the sentential negative found in declaratives". 

Note, however, that in addition to the negative imperative construction discussed above, the standard negation form is sometimes used to express directive speech acts, as in~(\ref{ex:adverbial-nga-neg}). 
This seems to be the case when the statement is rather an instruction than a command.

Hortative is constructed by affixing the first person plural subject agreement marker and the subjunctive suffix \textit{-e}, which replaces the default final vowel~\textit{-a}, to the verb. 
In the negative hortative construction, the post-initial negation marker \textit{ta-} and final vowel \textit{-a} are used, as in~(\ref{ex:hortative}--\ref{ex:hortativeneg}). 
In order to form a jussive, the same construction as in hortative is used but with a non-first person plural subject, as in  (\ref{ex:jussive}--\ref{ex:jussiveneg}).

\ea \label{ex:moods}\label{ex:ruulietal:11}
\begin{xlist}
	\ex \label{ex:hortative}
	\glll   Twígále olwígê.\\
	        tu-igal-e o-lwige\\
		    1\textsc{pl.sbj}-close-\textsc{sbjv} \textsc{aug}-11.door\\
	\glt ‘Let's close the door.’  (elicited)
\ex \label{ex:hortativeneg}
	\glll   Tutáígálâ ꜜlwígê.\\
	        tu-ta-igal-a lwige\\
    		1\textsc{pl.sbj}-\textsc{neg}-close-\textsc{fv} 11.door\\
	\glt ‘Let's not close the door.’  (elicited)
\ex \label{ex:jussive}
	\glll   Báígále ꜜámádirísâ.\\
	        ba-igal-e a-madirisa\\
		    2\textsc{sbj}-close-\textsc{sbjv} \textsc{aug}-10.window\\
	\glt ‘Let them close the windows.’  (elicited)
\ex \label{ex:jussiveneg}
	\glll  Batáígálâ madirísâ.\\
	        ba-ta-igal-a madirisa\\
		    2\textsc{sbj}-\textsc{neg}-close-\textsc{fv} 10.window\\
	\glt ‘Let them not close the windows.’ (elicited)
\end{xlist}
\z

Polar questions in Ruuli do not have any morphological marking, as in (\ref{ex:interrogneg1}).\footnote{At the moment we cannot say whether polar interrogatives have a special intonation.} 
Negative polar questions are built by using the standard negator \textit{ti-}, as in (\ref{ex:interrogneg2}).

\ea \label{ex:interrogneg}
\begin{xlist}
\ex \label{ex:interrogneg1}
	\glll  We owa musoló?\\
	        we o-wa musolo\\
		    2\textsc{sg}	2\textsc{sg.sbj}-give.\textsc{fv} 3.tax\\
	\glt `You, do you pay taxes?'  (elicited)
	
\ex \label{ex:interrogneg2}
	\glll  We towá músoꜜló?\\
	        we ti-o-wa musolo\\
	    	2\textsc{sg}	\textsc{neg}-2\textsc{sg.sbj}-give.\textsc{fv} 3.tax\\
	\glt `You, don't you pay taxes?' (ll-N-KYERIMA-170220-FS-7a)
\end{xlist}
\z


\subsection{Negation in non-verbal clauses}\label{sec:non-verbalruuli}\label{sec:ruulietal:2.3}

As other Bantu languages, Ruuli employs both invariable copulas — showing no agreement with the nominal subject — and inflected copulas, which agree with the subject (see \citealt{GibsonEtAl2019}). 
There are four copulas in Ruuli: \textit{ni}, \textit{li}, \textit{ta}, %\textit{taa} 
and \textit{ti}.\footnote{Note that \textit{li} is glossed as a copula in many examples of this section, whereas in other examples it is glossed as an auxiliary, e.g.\,in (\ref{ex:pers}). 
The differing glosses are chosen to indicate the respective function of \textit{li} in Ruuli.}
Additionally, the semi-copula \textit{bba} can be used in some of the contexts where the real copulas occur, i.e.\,in locational clauses and predicative possession.
With proper inclusion, equative and attributive clauses, the usage of a copula is not required in the present tense with a non-pronominal subject, as in~(\ref{ex:nocopula1}).
In non-present tense and with pronominal subjects predication of proper inclusion, equation and attribution requires the copula \textit{li}, as in~(\ref{ex:nocopulaPST}). 

\ea \label{ex:nonverbal}
\begin{xlist}
	\ex \label{ex:nocopula1}
	\glll Omúkali musómésya / má wânge / mukóoto.\\
	    o-mukali musomesya / ma wa-ange / mu-kooto\\
		\textsc{aug}-1.woman 1.teacher / 1.mother 1-1\textsc{sg.poss} / 1-big\\
\glt ‘The woman is a teacher / my mother / big.’ (elicited)

\ex  \label{ex:nocopulaPST}
\glll Omúkali yaalí musómésya / má wânge / mukóoto.\\
	    o-mukali a-a-li musomesya / ma wa-ange / mu-kooto\\
		\textsc{aug}-1.woman \textsc{1sbj}-\textsc{pst}-\textsc{cop} 1.teacher / 1.mother 1-1\textsc{sg.poss} / 1-big\\
	\glt ‘The woman was a teacher / my mother / big.’ (elicited)
\end{xlist}
\z

The copulas \textit{li}, \textit{ta}, and the semi-copula \textit{bba} are used in locative predication, as in~(\ref{ex:locli}). 
The speakers claim that \textit{ta} in most cases is used interchangeably with \textit{li}. 
\textit{ta} is said to be an archaic form that is still in use in some dialects. 
Following \citet[6]{Pustet2003}, we classify \textit{bba} as a semi-copula since it is not semantically empty but rather adds some meaning to the predicate phrase. 
The semi-copula can be translated as `stay, remain' in examples of locational predication and as `become' with predication of proper inclusion, equation and attribution. 
It inherently conveys the aspectual continuitive value in the first case and ingressive value in the second, which is not the case for \textit{li} or \textit{ta} (cf.\,\citealt[35–36]{Hengeveld1992}, as well as \citealt[214]{GibsonEtAl2019} on Bantu specifically). 

\ea \label{ex:locli}
	\glll   Omúsáàle ꜜgúlî / gutâ / gubbâ ꜜmú kíbíra.\\
	        o-musaale gu-li / gu-ta / gu-bba mu kibira\\
	    	\textsc{aug}-3.tree \textsc{3sbj}-\textsc{cop} / \textsc{3sbj}-\textsc{cop} / \textsc{3sbj}-\textsc{cop} 18.\textsc{loc} 7.forest\\
	\glt ‘The tree is / is / stays in the forest.’ (elicited)
\z

Bantu languages often encode predicative possession by means of a copula and the comitative preposition/conjunction \textit{na} `and, with' \citep{Creissels2018}. 
In Ruuli, the most common copula \textit{li} and the preposition \textit{na} are currently being reanalyzed as the verb \textit{lina} `have', as in (\ref{ex:predicativeposs-lina}), which acquired some verbal properties, such as the ability to host locative clitics. 
Such reanalysis of a copula followed by the comitative preposition has been observed in other Bantu languages (see \citealt{Creissels2018} for an overview).

Predicative possession can also be occasionally expressed by means of the less common copula \textit{ta} followed by the preposition \textit{na}, as well as the semi-copula \textit{bba} followed by \textit{na}, as in~(\ref{ex:predicativeposs-cop}). 
The speakers claim that \textit{lina} can be used interchangeably with the copula \textit{ta} and that the preference for one or the other is dialectal, the second option is, however, very infrequent in the corpus.
There is no difference between temporary and permanent predicative possession. 


\ea \label{ex:predicativeposs}
\begin{xlist}
\ex \label{ex:predicativeposs-lina}
\glll Omúkali alina ábaana.\\
	        o-mukali a-lin-a a-baana\\
		\textsc{aug}-1.woman 1\textsc{sbj}-have-\textsc{fv}  \textsc{aug}-2.child\\
\glt ‘The woman has children.’(elicited)

\ex \label{ex:predicativeposs-cop}
\glll Omúkali atá / abbá n’ábaana.\\
	        o-mukali a-ta / a-bba na a-baana\\
		\textsc{aug}-1.woman 1\textsc{sbj}-\textsc{cop} / 1\textsc{sbj}-\textsc{cop} \textsc{com} \textsc{aug}-2.child\\
\glt ‘The woman has / stays with children.’(elicited)
\end{xlist}
\z 

Existential predication is expressed by a construction with the copula \textit{li}: The copula carries both a locative prefix and a locative enclitic of the noun classes 16, 17, 18, and 23, as in~(Table~\ref{tab:locprefixes}).  
The prefix and the enclitic typically belong to the same locative noun class, as in  (\ref{ex:16.loc}--\ref{ex:18.loc}). 
However, this is not obligatory, see e.g.\,(\ref{ex:16.23}). 
The use of \textit{waliwo} is by far the most common way to express  existential predication in the corpus, but other combinations of the copula with a locative prefix and a locative enclitic are attested as well.

\begin{table}
\caption{The locative prefixes and enclitics}
\label{tab:locprefixes}
	\begin{tabular}{lll}
\lsptoprule
	class&prefix&enclitic\\
\midrule
	16&\textit{wa-}&\textit{=wo}\\
	17&\textit{ku-}&\textit{=ku}\\
	18&\textit{mu-}&\textit{=mu}\\
	23&\textit{e-}&\textit{=yo}\\
\lspbottomrule
	\end{tabular}
\label{tab:locatives}
\end{table}

\ea \label{ex:existential}
\begin{xlist}
\ex \label{ex:16.loc}
	\glll   Wálíwo ómúzimu.\\
	        wa-li=wo o-muzumu\\
		    16\textsc{sbj}-\textsc{cop}=16.\textsc{loc} \textsc{aug}-1.devil\\
	\glt ‘There is a devil.’ (elicited)
\ex \label{ex:17.loc}
	\glll   Kúlíku omwójô.\\
	        ku-li=ku o-mwojo\\
	    	\textsc{17sbj}-\textsc{cop}=17.\textsc{loc} \textsc{aug}-1.boy\\
	\glt ‘There is a boy.’ 	(ll-R-WSKAYUNGA-170217-FS-3)
\ex \label{ex:18.loc}
	\glll   Múlímu omúntû.\\
	        mu-li=mu o-muntu\\
		    18\textsc{sbj}-\textsc{cop}=18.\textsc{loc} \textsc{aug}-1.person\\
	\glt ‘There is a person.’ (elicited)

\ex \label{ex:16.23}
	\glll   Wálíwo ekíntû ekîndi.\\
	        wa-li=yo e-kintu e-ki-ndi\\
		    16\textsc{sbj}-\textsc{cop}=23.\textsc{loc} \textsc{aug}-7.thing \textsc{aug}-7-other\\
	\glt ‘There is another thing.’ 	(elicited)
\end{xlist}
\z

The copula \textit{ni} is used when the subject of the clause is focused. 
In contrast to the other agreeing copulas, \textit{ni} takes a subject agreement suffix (cf.\,\citealt[219–202]{GibsonEtAl2019} for similar examples from other Bantu languages).
It can stand alone in predications where no copula is required, i.e.\,in proper inclusion, equative and attributive clauses in present tense with a nominal subject, as in~(\ref{ex:ni1}). 
In cases where a copula is required, i.e.\,in all other cases, \textit{ni} cannot replace another copula but can only be added to the clause, as in~(\ref{ex:ni5}--\ref{ex:ni4}).

\ea \label{ex:ni}
\begin{xlist}
	\ex \label{ex:ni1}
    	\glll   Omúkali níyê musómésyâ / má wânge / mukóoto.\\
    	        o-mukali ni-ye musomesya / ma wa-ange / mu-kooto\\
    	    	\textsc{aug}-1.woman \textsc{cop}-\textsc{1sbj} 1.teacher / 1.mother 1-1\textsc{sg.poss} / 1-big\\
        \glt ‘The woman is the one who is a teacher / my mother / big.’  (elicited)
    
    \ex \label{ex:ni5}
    	\glll Omúkali níyê alí ꜜmú Kámpálâ.\\
    	        o-mukali ni-ye a-li mu Kampala\\
    	    	\textsc{aug}-1.woman \textsc{cop}-1\textsc{sbj} 1\textsc{sbj}-\textsc{cop} 18.\textsc{loc} 1.Kampala\\
        \glt ‘It is the woman who is in Kampala.’ (elicited)
    
    \ex \label{ex:ni4}
    	\gll    *o-mukali ni-ye mu Kampala.\\
    	    	\textsc{aug}-1.woman \textsc{cop}-\textsc{1sbj} 18.\textsc{loc} 1.Kampala\\
        \glt Intended: ‘It is the woman who is in Kampala.’
\end{xlist}
\z

The Ruuli copula construction with the inflecting \textit{ni} is somewhat unusual in the typology of the Bantu languages, as it is not common for a Bantu language to show an inflecting copula based on \textit{ni} (see \citealt{GibsonEtAl2019}). 
If there is an inflecting copula, it is in most cases not based on the copula \textit{ni}. 
In Ruuli, however, there is an inflecting copula which clearly is based on \textit{ni}, as (\ref{ex:ni}) illustrate. 
In contrast to regular verbs and inflecting copulas, it shows subject agreement via enclitics (similarly to the copula \textit{ndi} in Swahili, see \citealt{GibsonEtAl2019}).

In the negation of proper inclusion, equative and attributive clauses in present tense and with a non-pronominal subject, the invariable negative copula \textit{ti} is used, as in~(\ref{ex:nonverbalneg}). 

\ea \label{ex:nonverbalneg}\label{ex:ruulietal:18}
	\glll   Ómúkaꜜlí tí músómésya / má wânge / mukóoto.\\
	        o-mukali ti musomesya / ma wa-ange / mu-kooto\\
	    	\textsc{aug}-1.woman \textsc{neg.cop} 1.teacher / 1.mother 1-1\textsc{sg.poss}  / 1-big\\
	\glt ‘The woman is not a teacher / my mother / big.’  (elicited)
\z

\noindent In non-present, as in~(\ref{ex:ti6}), as well as with pronominal subjects, as in~(\ref{ex:ti4}), however, the standard negation marker \textit{ti-} is prefixed to the copula \textit{li}. 
The same marker is prefixed to the copula \textit{ni} when the subject of the negative statement is focused, compare~(\ref{ex:ti5}) with the positive clauses in~(\ref{ex:ni}).

\ea \label{ex:nonverbalneg2}\label{ex:ruulietal:19}
\begin{xlist}
\ex \label{ex:ti6}
	\glll  Tikyálí ꜜkyángú.\\
	        ti-ki-a-li ki-angu\\
		    \textsc{neg}-\textsc{7sbj}-\textsc{pst}-\textsc{cop} 7-easy\\
\glt ‘It was not easy.’ (ll-M-WSKAYUNGA-170218-FS-3)

\ex \label{ex:ti4}
	\glll   Kubbâ tolí Múgânda.\\
	        kubba ti-o-li Muganda\\
		    because \textsc{neg}-\textsc{2sg.sbj}-\textsc{cop} 1.Muganda\\
\glt ‘Because you are not a Muganda.’  (elicited)

\ex \label{ex:ti5}
	\glll   Omúkali  tiníyê mukóoto.\\
	        o-mukali ti-ni-ye mu-kooto\\
	    	\textsc{aug}-1.woman \textsc{neg}-\textsc{cop}-1\textsc{sbj} 1-big\\
	\glt ‘It is not the woman who is big.’ 	 (elicited)
\end{xlist}
\z

As for the negation of locational predication, as in (\ref{ex:locli}) vs.\,(\ref{ex:loclineg}), and predicative possession, as in~(\ref{ex:predicativeposs}) vs.\,(\ref{ex:posslineg}), the copulas \textit{li} and \textit{ta}, as well as the semi-copula \textit{bba} are negated by means of the standard negation marker \textit{ti-}. 

\ea \label{ex:loclineg}\label{ex:ruulietal:20}
	\glll   Omúsáale tigúlî / tigútá / tigúbbá ꜜmú kíbíra.\\  
	        o-musaale ti-gu-li / ti-gu-ta / ti-gu-bba mu kibira\\
		    \textsc{aug}-3.tree \textsc{neg}-\textsc{3sbj}-\textsc{cop} / \textsc{neg}-\textsc{3sbj}-\textsc{cop} / \textsc{neg}-\textsc{3sbj}-\textsc{cop} 18.\textsc{loc} 7.forest\\
	\glt ‘The tree is not / is not / does not stay in the forest.’ 	 (elicited)
\z

\ea 
\label{ex:posslineg}\label{ex:ruulietal:21}
	\glll   Omúkali  talí / tabbá / tatá n’ ábaaꜜná.\\
	        o-mukali ti-a-li / ti-a-ta / ti-a-bba na baana\\
    		\textsc{aug}-1.woman \textsc{neg}-\textsc{1sbj}-\textsc{cop} / \textsc{neg}-\textsc{1sbj}-\textsc{cop} / \textsc{neg}-\textsc{1sbj}-\textsc{cop} \textsc{com} 2.child\\
    \glt ‘The woman does not have children.’  (elicited)
\z

Existential clauses in Ruuli may be negated by means of the standard negator \textit{ti-}. 
This strategy seems to be used only in non-present tense, as in~(\ref{ex:tiwaliwo}). 
It is infrequently found in the corpus and was never produced in elicitation sessions. 
The more common strategy includes the dedicated negative existential \textit{ndoo} with one of the locative enclitics, which are also used in the affirmative. 
This construction is used for both present and non-present, as in~(\ref{ex:ndoopres}--\ref{ex:ndoononpres}). 
The origin of the negative existential is unclear and a similar form is not available in closely related languages. 
However, the usage of a special negation construction in at least one type of non-verbal predication is common cross-linguistically \citep{Veselinova2013neg.exist, Miestamo2017}. 

\ea \label{ex:existentialneg}\label{ex:ruulietal:22}
\begin{xlist}
	\ex \label{ex:tiwaliwo}
	\glll   Eirá budí tíwáꜜálíwó sênte.\\
	        e-irai budi ti-wa-a-li=wo sente\\
		\textsc{aug}-long.ago in.the.past  \textsc{neg}-16\textsc{sbj}-\textsc{pst}-\textsc{cop}=16.\textsc{loc} 9.money\\
	\glt ‘Long ago there was no money.’ (elicited)
	
\ex \label{ex:ndoopres}\label{ex:ruulietal:22b}
	\glll   Nayê ndóówô mugásô.\\
	        naye ndoo=wo mugaso\\ 
		    but \textsc{negex}=16.\textsc{loc} 3.importance\\
	\glt ‘But there is no importance.’ (ll-N-KIBBAALE-170221-FS-7a)

\ex \label{ex:ndoononpres}\label{ex:ruulietal:22c}
	\glll   Ndóówô ꜜMúnyálâ.\\
	        ndoo=wo Munyala\\
		    \textsc{negex=16.loc} 1.Munyala\\
	\glt ‘There was no Munyala.’ (ll-N-WSKAYUNGA-170215-NA-2)
\end{xlist}
\z

Ruuli non-verbal constructions are somewhat unusual from a Bantu perspective in that they employ an inflecting copula based on \textit{ni}. 
The negation of non-verbal constructions is common for Bantu languages: 
There is a dedicated negative copula which does not show subject agreement and is formally the same as the negative copula reconstructed for Proto-Bantu \citep[115]{Meeussen1967}. 
However, \textit{ti} cannot be considered a negative counterpart to the affirmative copula \textit{ni}, as in Proto-Bantu, because its usage is not limited to negation in combination with focus. 
Finally, there is a dedicated negative existential marker \textit{ndoo} which has not been reported for related Bantu languages, though specialized seemingly unrelated forms exist in closely related languages, e.g.\,Ganda has the form \emph{mpaa}, which combines with various locative enclitics to produce \textit{mpaa=mu, mpaa=wo}, and \textit{mpaa=yo} `there is no' \citep{AshtonEtAl1954}, \citep[341]{Murphy1972}; see also \citet{Veselinova2016} on other language families and cross-linguistic commonalities of new forms of negative existentials


\subsection{Negation in non-main clauses} \label{sec:non-mainruuli}\label{sec:ruulietal:2.4}

In this section, we will discuss negation strategies in non-main clauses. 
We focus on three types of non-main clauses viz.\,clausal complementation in §\ref{sec:non-mainruulicompl}, relative clauses in §\ref{sec:non-mainrelativeclauses}, as well as temporal and conditional adverbial clauses in  §\ref{sec:nonmaintemporal}. 
The negation patterns found in other types of non-main clauses do not deviate from the patterns described for main clauses in §\ref{sec:standardruuli}.


\subsubsection{Clausal complementation}\label{sec:non-mainruulicompl}\label{sec:ruulietal:2.4.1}

Ruuli distinguishes three major types of complement clauses depending on the complement verb form: indicative complements with the final vowel suffix \textit{-a} on the verb, 
subjunctive complements with the verb suffix \textit{-e}, 
and infinitive complements, which are marked by the class 15 prefix \textit{ku-} glossed here as \textsc{inf}. 
The most frequent and versatile complement type in Ruuli are infinitive complements (see \citealt{SorensenWitzlack-Makarevich2020} for details). 

In indicative complements, the verb form of the complement clause does not differ from the verb form used in independent declarative clauses: 
There is an obligatory subject agreement and optional object agreement on the verb. 
Furthermore, the verb can occur with the same TA-categories as verbs in  independent declarative clauses. 
When the verb in indicative complement clauses is negated, the standard pre-initial negator \textit{ti-} is used, as in~(\ref{ex:compindicative}).

\ea \label{ex:compindicative} \label{ex:ruulietal:23}
	\glll   Baabóíné ↓tékyâli yà mugaso.\\
	        ba-bon-ire ti-e-kya-li ya mugaso\\
		    2\textsc{sbj}-see-\textsc{pfv} \textsc{neg}-9\textsc{sbj}-\textsc{persis}-\textsc{cop} 9.\textsc{gen} 3.importance\\
    \glt ‘They have seen that they are no longer of importance.’ (ll-R-NAKASONGOLA-170225-FS-1a)
\z

The Ruuli subjunctive shows subject agreement but neither marking for tense nor aspect, as in~(\ref{ex:subjunctivecompl}). 
As we discussed in §\ref{sec:non-declarativesruuli}, the Ruuli subjunctive is used in independent clauses to express jussive and hortative, as in~(\ref{ex:subjunctiveindependent}), as well as optative and modal meanings. 
The subjunctive can also be used in certain types of complement clauses, as in (\ref{ex:subjunctivedependent}) (see \citealt{SorensenWitzlack-Makarevich2020} and \citealt{VeselovskyEtAl2023} for details).
The negation of a subjunctive is morphologically possible and occur in other contexts, however, negative subjunctive complements do not occur in the corpus (apart from cases of reported discourse). 
Instead of negating the subjunctive complement, either the negative infinitive (see (\ref{ex:bu}) below) or a periphrastic construction with the morphologically positive form of one of the inherently negative verbs, such as \text{leka} ‘leave, stop (doing smth.)’, as in (\ref{ex:leka}), or \textit{lema} ‘refuse', is used (see also \citet{Güldemann1999} for examples of periphrastic negation in other Bantu languages). 
 
\ea \label{ex:subjunctivecompl}
\begin{xlist}
\ex \label{ex:subjunctiveindependent}
	\glll   Túsómeꜜsyé ábáana.\\
	        tu-somesy-e a-baana\\
	    	1\textsc{pl.sbj}-teach-\textsc{sbjv} \textsc{aug}-2.child\\
	\glt ‘Let us educate children.’ 	(ll-N-BBALE-170220-FS-2)
\ex \label{ex:subjunctivedependent}
	\glll Otáká áté óꜜtééwô olúꜜwónkô.\\
	        o-tak-a ate o-ta-e=wo o-lukonko\\ 
		    \textsc{2sg.sbj}-want-\textsc{fv} \textsc{foc} \textsc{2sg.sbj}-put-\textsc{sbjv}=16.\textsc{loc} \textsc{aug}-11.rift\\
	\glt ‘You want to cause a rift.' (ll-N-BBALE-170220-FS-2)
\ex \label{ex:leka}
	\gll        a-a-ba-kob-a-nga             ba-lek-e        ku-tiin-a\\ 
		    \textsc{1sbj}-\textsc{pst}-\textsc{2pl.obj}-say-\textsc{fv}-\textsc{hab} \textsc{2sbj}-let-\textsc{sbjv} \textsc{inf}-fear-\textsc{fv}\\
	\glt ‘He used to tell them not to fear.' (ll-X-AniinireAiguru-nd-W-1)
\end{xlist}
\z

In order to form the infinitive, class 15 prefix \textit{ku-} is added to the verb stem that is followed by the final vowel \textit{-a}. 
The class prefix is often preceded by the corresponding augment \textit{o-}. 
The distribution of the augment is conditioned by a range of syntactic and semantic factors and its description goes beyond the scope of the present study. 
Suffice it to say that the augment \textit{o-} cannot be used on the infinitive when the complement-taking predicate is negated, as in~(\ref{ex:ku}) (cf.\,\citealt{SorensenWitzlack-Makarevich2020}).

\ea \label{ex:ku} 
	\glll   Tintáká kúꜜnywâ isárâ.\\
	        ti-n-tak-a ku-nyw-a isara\\
	    	\textsc{neg}-1\textsc{sg.sbj}-like-\textsc{fv} \textsc{inf}-drink-\textsc{fv} 5.stale.brew\\
    \glt ‘I don’t like drinking stale brew.’ (\citealt[296]{NamyaloEtAl2021})
\z

Infinitives themselves are negated by means of the post-initial negation marker \textit{ta-}. 
When negated, the infinitive carries the class 14 prefix \textit{bu-} instead of the regular class 15 prefix \textit{ku-}. 
The class prefix can be preceded by the respective augment \textit{o-}, as in~(\ref{ex:bu}). 
It is not yet clear under which circumstances the speakers prefer to negate the infinitive directly instead of negating the matrix verb. 
In any case, the pattern with  \textit{bu-} is used to negate infinitives also when they do not form a complement.

\ea \label{ex:bu}\label{ex:ruulietal:26}
	\glll    Ómúsaiza ayínzâ óbúꜜtáléétáwô mukalí.\\
	        o-musaiza a-yinz-a o-bu-ta-leet-a=wo mukali\\
	    	\textsc{aug}-1.man 1\textsc{sbj}-be.able-\textsc{fv} \textsc{aug}-14-\textsc{neg}-bring-\textsc{fv}=16.\textsc{loc} 1.wife\\
\glt ‘The man may not bring there a wife.’ (ll-M-WSKAYUNGA-170218-FS-2b)
\z

\subsubsection{Relative clauses} \label{sec:non-mainrelativeclauses}

There are two main relativizing strategies in Ruuli. 
The construction used to relativize on subjects employs relative prefixes that are formally identical to the augment of the relativized noun, viz.\,\textit{a-, e-} or \textit{o-}, as in~(\ref{ex:relativesbj}). 
The construction for relativization of non-subjects includes the relative pronouns, as in~(\ref{ex:relativeobj}). 
With both strategies, the relative clause is negated by means of the post-initial negative prefix \textit{ta-}, as in~(\ref{ex:relativesbjneg}--\ref{ex:relativeobjneg}), and not with the standard negator \textit{ti-}. 
Other than the use of the negator the structure of positive and negative relative clauses is identical.

\ea \label{ex:relative}\label{ex:ruulietal:27}
\begin{xlist}
\ex \label{ex:relativesbj}\label{ex:ruulietal:27a}
	\glll   Ekíbírâ ekíndíráine kírímu enkémbo.\\
	        e-kibira e-ki-n-liraine ki-li=mu e-nkembo\\
	    	\textsc{aug}-7.forest \textsc{rel}-\textsc{7sbj}-\textsc{1sg.obj}-be\_close\_to.\textsc{pfv} \textsc{7sbj}-\textsc{cop}=18.\textsc{loc} \textsc{aug}-10.monkey\\
    \glt ‘There are monkeys in the forest which is close to me.’ (ll-M-KIBBALE-170221-FS-4b)

\ex \label{ex:relativeobj}\label{ex:ruulietal:27b}
	\glll    Túlína onyónyî gwé ꜜbéeta okísyo.\\
	        tu-lin-a o-nyonyi gwe ba-yet-a o-kisyo\\
	    	1\textsc{pl.sbj}-have-\textsc{fv} \textsc{aug}-1.bird 1.\textsc{rel} 2\textsc{sbj}-call-\textsc{fv} \textsc{aug}-1.weaver.bird\\
    \glt `We have a bird called weaver bird.' (ll-M-KIBBALE-170221-FS-4b)

\ex \label{ex:relativesbjneg}\label{ex:ruulietal:27c}
	\glll  Nayê ábátaabbaꜜngá ná kifô  nibáyába.\\
	        naye a-ba-ta-a-bba-a-nga na kifo ni-ba-yab-a\\
	    	but \textsc{rel}-2\textsc{sbj}-\textsc{neg}-\textsc{pst}-\textsc{cop}-\textsc{fv}-\textsc{hab} \textsc{com} 7.place \textsc{narr}-2\textsc{sbj}-go-\textsc{fv}\\
    \glt ‘But those who didn't have a place went.’ (\citealt[100]{NamyaloEtAl2021})

\ex \label{ex:relativeobjneg}\label{ex:ruulietal:27d}
	\glll  Ámáani gá Káncâ gwé óꜜtásóbolá kúꜜbónáku.\\
	        a-maani ga kanca gwe o-ta-sobol-a  ku-bon-a=ku\\ 
	    	\textsc{aug}-6.power 6.\textsc{gen} 1.God 1.\textsc{rel} \textsc{2sg.sbj}-\textsc{neg}-can-\textsc{fv} \textsc{inf}-see-\textsc{fv}=17.\textsc{loc}\\
    \glt `The power of God whom you cannot see.' (ll-N-WSKAYUNGA-170218-FS-1a)
\end{xlist}
\z

\subsubsection{Temporal and conditional adverbial clauses}
\label{sec:nonmaintemporal}

The most frequent type of adverbial clauses are temporal and conditional clauses marked by the conjunction \textit{nga} `while, when', as in~(\ref{ex:adverbial-nga}). 
The conjunction \textit{nga} is always the first item in the respective adverbial clause.\footnote{Ruuli has further less common types of adverbial clauses, which are currently being investigated, e.g.\,the concessive adverbial clause introduced by the conjunction \emph{wade} `though, even if, although'. For most of these clause types the same pattern of negation as the one observed with \emph{nga}-clauses holds.} 
These clauses are negated by means of the pre-initial standard negator \emph{ti-} (see §~\ref{sec:standardruuli}), as in~(\ref{ex:adverbial-nga-neg}).

\ea \label{ex:adverbialclause}\label{ex:ruulietal:28}
\begin{xlist}
\ex \label{ex:adverbial-nga}\label{ex:ruulietal:28a}
    \glll   Nkumanyisyémú nga twiréyo okú bbáíbbúli?\\
        n-ku-many-isy-e=mu			nga	tu-ire=yo			o-ku		bbaibbuli\\
        1\textsc{sg.sbj}-2\textsc{sg.obj}-know-\textsc{caus}-\textsc{sbjv}=18.\textsc{loc}	when	1\textsc{pl.sbj}-return.\textsc{pvf}=23.\textsc{loc}	\textsc{aug}-17.\textsc{loc}	9.Bible\\
    \glt `Should I let you know when we return to the Bible?' (ll-R-NAKASONGOLA-190801-FS-2)

\ex \label{ex:adverbial-nga-neg}\label{ex:ruulietal:28b}
    \glll Toyámbángá ꜜmú kísokô nga tokwáite mwígo.\\
         ti-o-yab-a-nga mu kisiko nga ti-o-kwaite 	        	mwigo\\
	   \textsc{neg}-2\textsc{sg.sbj}-go-\textsc{fv-hab} 18.\textsc{loc}	7.bush	while \textsc{neg}-2\textsc{sg.sbj}-hold.\textsc{pfv} 3.stick\\
	\glt ‘Never go to the bush without a stick.’ (lit.\,‘You never go to the bush while not holding a stick.’) (\citealt[62]{NamyaloEtAl2021})
\end{xlist}
\z

A separate clause often introduced by the conjunction \textit{nga} `while, when' can also be used to express the caritive meaning of \textit{without + noun} and \textit{without + verb}.
For the meaning of \textit{without + noun}, as in (\ref{ex:withoutnoun2}), the predicative possession construction including the copula \textit{li} merged with the comitative preposition \textit{na} (see §\ref{sec:non-verbalruuli}) is used in the \textit{nga}-clause, as in~(\ref{ex:withoutnoun2}--\ref{ex:withoutnoun}).

\ea \label{ex:ruulietal:29}
\begin{xlist}
\ex \label{ex:withoutnoun2}\label{ex:ruulietal:29a}
    \glll  Yaizíréyô eyô nga talínâ ꜜmótóka.\\
	        a-a-iz-ire=yo eyo nga ti-a-lin-a motoka\\
	    	\textsc{1sbj}-\textsc{pst}-come-\textsc{pfv}=23.\textsc{loc} 23.\textsc{loc} while \textsc{neg}-\textsc{1sbj}-have-\textsc{fv} 9.car\\
	\glt ‘He came there without a car.' (lit.\,`He came there while he does not have a car.') (\citealt[129]{NamyaloEtAl2021})

\ex \label{ex:withoutnoun}\label{ex:ruulietal:29b}
	\glll   Yayabiré omú írwáaro nga talínâ mugendyá.\\
	        a-a-yab-ire		omu	irwaliro nga	ti-a-lin-a		mugendya\\ 
	    	\textsc{1sbj}-\textsc{pst}-go-\textsc{pfv} 	18\textsc{.loc}	5.hospital while	\textsc{neg}-1\textsc{sbj}-have-\textsc{fv}	1.caretaker\\
    \glt ‘He went to the hospital without a caretaker.’ (Lit. ‘He went to the hospital while he did not have a caretaker.’) (\citealt[129]{NamyaloEtAl2021})

\ex \label{ex:withoutndoo}\label{ex:ruulietal:29c}
    \glll Balyangá ómúkuꜜbí ndóómû muunyú.\\
        ba-ly-a-nga	omukubi	ndoo=mu	muunyu\\
	        2\textsc{sbj}-eat-\textsc{fv-hab}	3.sauce \textsc{negex}=18.\textsc{loc}	3.salt\\
	\glt ‘They used to eat sauce without salt.’ (ll-N-WSKAYUNGA-170216-NA-1)

\ex \label{ex:withoutngandoo}\label{ex:ruulietal:29d}
    \glll Baamubbweeré nga ndóówô bujúurwa.\\
    ba-a-mu-bbw-ere nga ndoo=wo bujuurwa\\
	2\textsc{sbj}-\textsc{pst}-1\textsc{obj}-detain-\textsc{pfv}	while	\textsc{negex}=16.\textsc{loc}	14.evidence\\
	\glt ‘He was imprisoned without any evidence.’ (\citealt[172]{NamyaloEtAl2021})
\end{xlist} 
\z

In order to express \textit{without + verb}, the \textit{nga}-clause has the negated form of the respective verb, as in~(\ref{ex:withoutverb}).

\ea \label{ex:withoutverb}
	\glll   Nga talíísíꜜryé nkóko yatambwirê mbe.\\
	        nga ti-a-liisirye nkoko a-a-tambwire mbe\\
	    	while \textsc{neg}-\textsc{1sbj}-feed.\textsc{pfv} 10.chicken \textsc{1sbj}-\textsc{pst}-walk.\textsc{pfv} \textsc{foc}\\
	\glt ‘He just walked away without feeding the chicken.' (lit.\,`While he had not fed the chicken, he just walked away.') (\citealt[130]{NamyaloEtAl2021})
\z


\section{Non-clausal negation}
\label{sec:non-clausalruuli}

This section describes negative constructions that do not operate on clause-level, i.e.\,negative constructions that do not turn affirmative clauses into negative claus\-es but express negation on another type of constituent \citep[424]{Miestamo2017}.
The following sections are devoted to properties of negative replies to polar questions (§\ref{sec:repliesruuli}), the construction of negative indefinites (§\ref{sec:indefinitesruuli}), and negative derivation and case marking (§\ref{sec:derivationruuli}) in Ruuli.

\subsection{Negative replies}\label{sec:ruulietal:3.1}
\label{sec:repliesruuli}
Negative replies in Ruuli are either expressed by means of a one-word negative reply \textit{bbe} `no', by the repetition of the verb of the question in its negative form or a combination of the two strategies. \textit{Bbe} is a dedicated marker used only to give a negative reply.

In Ruuli, giving a negative reply to a negative question indicates that the speaker disagrees with the polarity but not the propositional content of the question. 
Thus, answering \textit{bbe} to the negative question in (\ref{ex:question}Q) is understood as `it (the child) is sleeping' (\ref{ex:question-answer1}A1).
Other grammatical responses indicating disagreement are given in~(\ref{ex:combination}A2) and~(\ref{ex:combinationbbe}A3). 
A combination of \textit{ee} `yes' and the verb in the affirmative is not possible as an answer to a negative question, as in~(\ref{ex:combinationwrong}A4). 
\textit{Ee} is the common word for `yes' and not a polarity reversing particle, like German \textit{doch}. 
The combination of \textit{bbe} `no' and the verb in the negative in the given context is not possible either as in~(\ref{ex:combinationbbewrong}A5). 
However, (\ref{ex:combinationwrong}A4) and (\ref{ex:combinationbbewrong}A5) are acceptable answers to a positive question.

\ea \label{ex:negreplies}\label{ex:ruulietal:31}
\begin{xlist}[A5: ]
\exi{Q.}[]{ \label{ex:question}
	\glll   Ómwána tabbásyá? \\
	        o-mwana ti-a-bbasy-a \\
	    	\textsc{aug}-1.child \textsc{neg}-\textsc{1sbj}-sleep-\textsc{fv} \\
	\glt ‘Doesn't the child sleep?' (elicited)} 
\exi{A1.}[]{\label{ex:question-answer1}
	\glll  Bbé.\\
	        bbe\\
	    	 no \\
	\glt `No (the child sleeps).' (elicited)} 
\exi{A2.}[]{ \label{ex:combination}
	\glll   ée, tabbásyá.\\
	        ee ti-a-bbasy-a\\
	    	yes \textsc{neg}-\textsc{1sbj}-sleep-\textsc{fv}\\
	\glt ‘Yes, it does not sleep.' (elicited)} 
\exi{A3.}[]{ \label{ex:combinationbbe}
	\glll   Bbé, abbásyâ.\\
	        bbe a-bbasy-a\\
            no \textsc{1sbj}-sleep-\textsc{fv} 
           \\
	\glt ‘No, it sleeps.'  (elicited)}
\exi{A4.}[*]{ \label{ex:combinationwrong}
	\gll  ee a-bbasy-a\\
		    yes \textsc{1sbj}-sleep-\textsc{fv} 
		   \\
	\glt Intended: `Yes, it sleeps.'}
\exi{A5.}[*]{\label{ex:combinationbbewrong}
	\gll  bbe t-a-bbasy-a\\
	    	no \textsc{neg}-\textsc{1sbj}-sleep-\textsc{fv}\\
	\glt Intended: `No, it does not sleep.'}
\end{xlist}
\z



\subsection{Negative indefinites} \label{sec:indefinitesruuli}

In Ruuli, the semantic equivalent of the indefinite pronouns `something' and `somebody' are expressed by the use of the generic nouns \textit{kintu} `7.thing', as in~(\ref{ex:something}), and \textit{muntu} `1.person', as in~(\ref{ex:somebody}). 

\ea \label{ex:negindefinites}
\begin{xlist}
	\ex \label{ex:something}
	\glll  Asóbólâ okúꜜkóláwo ekintû.\\
	        a-sobol-a o-ku-kol-a=wo e-kintu\\
	    	\textsc{1sbj}-can-\textsc{fv} \textsc{aug}-\textsc{inf}-do-\textsc{fv}=16.\textsc{loc} \textsc{aug}-7.thing\\
	\glt ‘He can do something there.'  (elicited)
\ex \label{ex:somebody}
	\glll    Omúntû n’ áꜜkólá ékíbbi omukobakú.\\
	        o-muntu ni a-kol-a e-ki-bbi o-mu-kobaku\\ 
	    	\textsc{aug}-1.man if \textsc{1sbj}-do-\textsc{fv} \textsc{rel}-7-wrong \textsc{2sg.sbj}-1\textsc{obj}-counsel\\
	\glt ‘If someone does wrong, you counsel him.' (\citealt[412]{NamyaloEtAl2021})
\end{xlist}
\z

\citet{Haspelmath2013-wals115} states that languages with generic-noun-based indefinites typically also have generic-noun-based negative indefinites. This is not the case in Ruuli. 
Instead, a negative existential construction is used to express `nobody' and `nothing', as in~(\ref{ex:nothing}--\ref{ex:nobody}).
Note that (\ref{ex:nothing}) is actually a relative clause in which the relativizer \textit{e-} is dropped due to the presence of negation, as will be discussed in~§\ref{sec:markingscoperuuli}. More research on negative indefinites in other Bantu language is needed to assess how typical or atypical this way of expression is.

\ea \label{ex:negindefinites2}\label{ex:ruulietal:33}
\begin{xlist}
\ex \label{ex:nothing}
	\glll   Ndóówô ꜜkíréma.\\
	        ndoo=wo ki-lem-a\\
		    \textsc{negex}=16.\textsc{loc} \textsc{7sbj}-fail-\textsc{fv}\\
	\glt    ‘Nothing fails.' (lit.\,`There is nothing that fails.') (ll-R-NAKASONGOLA-170225-FS-1d)

\ex \label{ex:nothingdoppelt}
	\glll   Olina kí? Ndóówô.\\
	        o-lin-a ki ndoo=wo\\ 
		    \textsc{2sg.sbj}-have-\textsc{fv} \textsc{com} what \textsc{negex}=16.\textsc{loc}\\
	\glt    ‘What do you have? Nothing.'  (elicited)

\ex \label{ex:nobody}
	\glll   Ndóówô yaalí ánkúbbiꜜré.\\
	        ndoo=wo a-a-li a-n-kubb-ire\\ 
	    	\textsc{negex}=16.\textsc{loc} \textsc{1sbj}-\textsc{pst}-\textsc{aux} \textsc{1sbj}-\textsc{1sg.obj}-call-\textsc{pfv}\\
	\glt ‘Nobody has ever called me.' (ll-M-WSKAYUNGA-170218-FS-2b)
\end{xlist}
\z

\subsection{Negative derivation and case marking}
\label{sec:derivationruuli}

The preposition \textit{nambula} `without' can be compounded with nouns to indicate the lack of the compounded noun. 
For example, \textit{buculeeri} `14.peace' is compound\-ed with the  preposition \textit{nambula}, which results in the adjective \textit{nambulabuculeeri} `restless, insecure', as in~(\ref{ex:without}).
The noun \textit{kinambulamugaso} `7.useless thing' consists of the noun \textit{mugaso} `3.importance' and the preposition \textit{nambula} `without'. 
This noun is assigned to class 7 and receives the prefix \textit{ki-}.  This noun class also includes the generic noun \textit{kintu} `7.thing'.
At present, it is difficult to assess the productivity of this process, however, only a  small number of negative derivations of the above type is registered in the dictionary and no examples are found in the corpus.

\ea \label{ex:without}
	\glll   Ensí egínambulabuculéeri tekulaakulaná.\\
	        e-nsi e-gi-nambulabuculeeri ti-e-kulaakulan-a\\
	    	\textsc{aug}-9.country \textsc{aug}-9-insecure \textsc{neg}-9\textsc{sbj}-develop-\textsc{fv}\\
    \glt ‘An insecure country does not develop.' (\citealt[81]{NamyaloEtAl2021})
\z


\section{Other aspects of negation}\label{sec:otheraspectsruuli}

This section discusses those aspects of negation that operate on neither clausal nor non-clausal level.
We provide a short overview of the scope of negation~(§\ref{sec:scoperuuli}) and negative polarity items (§\ref{sec:polarityruuli}) in Ruuli. 
§\ref{sec:markingscoperuuli} analyses the marking of noun phrases in the scope of negation. §\ref{sec:reinforcingruuli} presents the strategies used to reinforce Ruuli negation. 
Negation in sentence coordination is addressed in~§\ref{sec:complexruuli}.


\subsection{The scope of negation} \label{sec:scoperuuli}

Ruuli is a language that shows verb-dependent negative placement, i.e.\,it places the negator in relation to the verb instead of the element in focus. 
Speakers translate the sentence \textit{I don't read the newspaper on Saturday} always as shown in (\ref{ex:scope}) even when different elements, i.e.\,\textit{read}, \textit{newspaper} or \textit{Saturday}, were put in focus. 
Probably, prosody is used to narrow the scope of negation to a specific constituent.

\ea \label{ex:scope}
	\glll  Tinsómá ámákuru okú Lwámúkáaga.\\
	        ti-n-som-a makuru o-ku Lwamukaaga\\
	    	\textsc{neg}-1\textsc{sg.sbj}-read-\textsc{fv} 6.newspaper \textsc{aug}-17.\textsc{loc} 10.Saturday\\
    \glt ‘I don't read the newspaper on Saturday.’  (elicited)
\z


\subsection{Negative polarity items}\label{sec:polarityruuli}

At the current state of language description, we have not been able to identify either positive or negative polarity items. 
The expression of phasal polarity (the concepts such as \textsc{yet} and \textsc{already}) 
is discussed in detail in \citet{MolochievaEtAl2021} and does not involve dedicated lexical items (see \sectref{sec:standardruuli} for a brief overview).

\subsection{Marking of NPs in the scope of negation} \label{sec:markingscoperuuli}
In Ruuli and other Bantu languages \citep[80]{Miestamo2014}, noun phrases can be effected by the scope of negation: 
When nouns are preceded by a negative verb form, the augment is typically omitted, as in~(\ref{ex:withaug}--\ref{ex:withoutaug}). 
The absence of the augment is not only observed on objects, which regularly follow the verb, but also on adjuncts introduced e.g.\,by the locative noun class markers \textit{ku} `17.\textsc{loc}' and \textit{mu} `18.\textsc{loc}'.\footnote{Class 16 locative has the invariable form \textit{e} and does not take an augment.}
When the negative verb is not directly followed  by a noun or when it is followed by a headless noun phrase, the augment is present on either the noun (in the first case) or other modifiers (in the second case), as in~(\ref{ex:negwithaug}). 
Note that even in complement clauses with infinitive verb forms, the complementation verb never carries the augment when the matrix predicated is negated, as described in §\ref{sec:non-mainruuli}.  

\ea \label{ex:markingnps}
\begin{xlist}
\ex \label{ex:withaug}
	\glll  Ibbâ enkáítô!\\
	        ibb-a e-nkaito\\
	    	steal-\textsc{fv} \textsc{aug}-10.shoe\\
	\glt ‘Steal the shoes!’  (elicited)
\ex \label{ex:withoutaug}
	\glll   Otáíbbâ nkáítô!\\
	        o-ta-ibb-a nkaito\\
	    	\textsc{2sg.sbj}-\textsc{neg}-steal-\textsc{fv} 10.shoe\\
    \glt `Don't steal the shoes!’ (elicited)
\ex \label{ex:negwithaug}
	\glll   Nayé lwâki titúsómesya olwáíswê?\\
	        naye lwaki ti-tu-somesy-a o-lu-aiswe\\
	    	but why \textsc{neg}-1\textsc{pl.sbj}-teach-\textsc{fv} \textsc{aug}-11-1\textsc{pl.poss}\\
    \glt `But why don't we teach ours?’ (ll-N-BBALE-170220-FS-2)
\end{xlist}
\z

In Ruuli, the class 17 locative enclitic \textit{=ku} can be used to express partitive meaning (see \citealt{PersohnDevos2017} for an overview of the use of locative enclitic to express partitive in Bantu). 
However, in contrast to the Circum-Baltic languages (see \citealt{Miestamo2014}), the partitive does not play a role in the marking of NPs in the scope of negation.

\subsection{Reinforcing negation}
\label{sec:reinforcingruuli}

Several strategies can be used to reinforce negation. 
First, there are three adverbs that are used to reinforce negation: \textit{kimwei} `totally, completely', \textit{kaakimwei} `thoroughly' and \textit{kakyarumwei} `totally, completely'. 
When used with negative constructions, they are often translated as `at all'. 
However, they express different degrees of reinforcement: \textit{kimwei} adds the least fortification to a negative statement, as in~(\ref{ex:inflittle}). 
The reinforcement is strengthened when \textit{kaakimwei} is used, as in~(\ref{ex:infmore}). 
The strongest intensifier of negation is achieved by the use of \textit{ka\-kyarumwei}, as in~(\ref{ex:infstrongest}). 
With these adverbs, the verb takes an applicative suffix both in negative and in positive forms.

\ea \label{ex:reinforcing}
\begin{xlist}

\ex \label{ex:noinf}
	\glll   Tindyá múꜜcéere.\\
	        ti-n-ly-a muceere\\
		    \textsc{neg}-1\textsc{sg.sbj}-eat-\textsc{fv} 3.rice\\
	\glt ‘I don't eat rice.’  (elicited)
\ex \label{ex:inflittle}
	\glll   Tindíírá kíꜜmwéi mucéere.\\
	        ti-n-lya-ir-a kimwei muceere\\
	    	\textsc{neg}-1\textsc{sg.sbj}-eat-\textsc{appl}-\textsc{fv} totally 3.rice\\
	\glt ‘I don't eat rice at all.’  (elicited)
\ex \label{ex:infmore}
	\glll    Tindíírá kááꜜkímwéi mucéere.\\
	        ti-n-lya-ir-a kaakimwei muceere\\
	    	\textsc{neg}-1\textsc{sg.sbj}-eat-\textsc{appl}-\textsc{fv} thoroughly 3.rice\\
	\glt ‘I don't eat rice at all.’  (elicited)
\ex \label{ex:infstrongest}
	\glll   Tindíírá káꜜkyáárúmwê mucéere.\\
	        ti-n-lya-ir-a kakyarumwei muceere.\\
		    \textsc{neg}-1\textsc{sg.sbj}-eat-\textsc{appl}-\textsc{fv} completely 3.rice\\
	\glt ‘I don't eat rice at all.’   (elicited)
\end{xlist}
\z

Note, however, that there is no sign of bleaching of the emphatic meaning of the adverbs. 
Furthermore, the use of the adverbs in the negative contexts is not conventionalized and they appear more frequently with positive than with negative verb forms in the corpus and are overall relatively 
We can thus rule out the possibility that the adverbs in combination with the standard negation marker \textit{ti-} form a double negation construction. 

Second, similarly to what has been reported for some other Bantu languages (see \citealt{DevosAuwera2013}, \citealt{ PersohnDevos2017}), 
Ruuli also uses a locative enclitic to reinforce negation, e.g.\,the class 17 \textit{=ku}, as in~(\ref{ex:encliticreinforcing}). 
Note that it is the locative enclitic with the partitive function, see \sectref{sec:markingscoperuuli}.

\ea \label{ex:encliticreinforcing}
    \ea \label{ex:enclitic1.a}
    	\gll    iswe ti-tu-a-zwal-a-nga=ku mbugo nayenga a-baala\\ %no punctuation here and in the next examples because this is the "second line" and the first (phonetic with tones) isn't available for these few sentences.  
    	we \textsc{neg}-1\textsc{pl.sbj}-\textsc{pst}-wear-\textsc{fv}-\textsc{hab} 10.bark\_cloth but \textsc{aug}-2.girl\\
    	\glt ‘As for us, apart for the girls, we have never worn bark cloth.’ (ll-N-KIDERA-170221-FS-5a)
    	
    \ex \label{ex:enclitic1.b}%
    \gll	        ti-n-bon-a-nga=ku e-ngeri gye ba-ba-kaly-a=mu o-mu Kinyala\\ 
    	    	\textsc{neg}-1\textsc{sg.sbj}-see-\textsc{fv}-\textsc{hab}=\textsc{17.loc} \textsc{aug}-9.way \textsc{9.rel} \textsc{2sbj}-\textsc{2obj}-handle-\textsc{fv}=\textsc{18.loc} \textsc{aug}-\textsc{18.loc} 7.Banyala\_culture\\
    	\glt ‘I have never seen how they are handled in the culture of the Banyala.’ (ll-N-KYERIMA-170218-NA-4)
    \z 
\z

Finally, the regular focus marker \textit{mbe}, as e.g.\,in (\ref{ex:withoutverb}), can be used immediately following negated verbs, as in~(\ref{ex:mbe}). 
It can be combined with one of the adverbs discussed above, as in~(\ref{ex:mbe2}).

\ea \label{ex:mbe}
\begin{xlist}
\ex \label{ex:mbe1}
\gll	        naye a-bantu ti-ba-ki-kwaite mbe\\
	    	but \textsc{aug}-2.person \textsc{neg}-\textsc{2sbj}-\textsc{7obj}-understand.\textsc{pfv} \textsc{foc} \\
	\glt ‘[He taught that thing,] but people did understand it.’ (ll-R-WSKAYUNGA-170217-FS-3)
	
\ex \label{ex:mbe2}
	       \gll ti-mu-ly-a mbe kibuulo kimwei adi\\
		    \textsc{neg}-1\textsc{pl.sbj}-eat-\textsc{fv} \textsc{foc} 7.meal totally there\\
	\glt ‘You just don't eat the same meal there.' (ll-R-NAKASONGOLA-170224-FS-3b)
\end{xlist}
\z

\subsection{Negation and coordination}
\label{sec:complexruuli}

Negative coordination of nominals is achieved by means of the particle \textit{gonze} `not even', as in~(\ref{ex:nominalcoord}). 
Note that the augment is only omitted on the nominal that directly follows the negated verb, whereas the nominal following \textit{gonze} can take the augment. 
Two negative clauses can be coordinated with the same marker, as in~(\ref{ex:clausecoord}). 
When a negative and a positive clause are coordinated, the coordinators \textit{ate} `but' or \textit{baitu} `but' can be used, as in~(\ref{ex:but}). 

\ea \label{ex:complex}
\begin{xlist}
\ex \label{ex:nominalcoord}
	\glll   Tintáká náíbúbyái gonzé enswâ.\\
	        ti-n-tak-a naibubyai gonze e-nswa\\
	    	\textsc{neg}-1\textsc{sg.sbj}-like-\textsc{fv} 10.spider not.even \textsc{aug}-9.flying.termite\\
	\glt ‘I like neither spiders nor flying termites.’ (elicited)
\ex \label{ex:clausecoord}
	\glll   Enkómyô tí gíkóoto gonzé ókúꜜbbáamu  abányágî.\\
	        e-nkomyo ti gi-kooto gonze o-ku-bba=mu a-banyagi\\
	    	\textsc{aug}-9.prison \textsc{neg} 9-big not.even \textsc{aug}-\textsc{inf}-\textsc{cop}=18.\textsc{loc} \textsc{aug}-2.prisoner\\
	\glt ‘The prison is neither big nor are there many prisoners.’  (elicited)
\ex \label{ex:but}
	\glll   Atáká áꜜbántû baitú oibá wáámwê tatáká ꜜbántû.\\
	        a-tak-a a-bantu baitu o-iba wa-amwe ti-a-tak-a bantu\\
	    	\textsc{1sbj}-like-\textsc{fv} \textsc{aug}-2.person but \textsc{aug}-1.husband 1-1\textsc{poss} \textsc{neg}-\textsc{1sbj}-like-\textsc{fv} 2.person\\
	\glt ‘She is friendly, but her husband is unfriendly.’ (lit.\,`She likes people, but her husband does not like people.')  (elicited)
\end{xlist}
\z

\section{Summary} 
We started this article by considering the clausal negation. 
All tense-aspect combinations are negated by means of the prefix \textit{ti-}. 
Except for the near feature tense, the negation constructions are symmetric with respect to segmental marking. 
There might well be tonal differences between the tenses, the area of Ruuli grammar still under investigation.

As discussed in \sectref{sec:standardruuli}, many Bantu languages show a regular, syntactically conditioned distribution of pre- and post-initial negatives: whereas the pre-initial negation tends to be used in unmarked main clauses, the post-initial negation strategy is used in more “marked” clause types, such as various subordinate clauses and prohibitives (\citealt[261–279, 305–308]{Güldemann1996}, \citeyear[551–564]{Güldemann1999}). 
The distribution of the pre-initial \emph{ti-} and the post-initial \emph{ta-} in Ruuli follows this trend.
Thus, we classify Ruuli as a language that has a special negative imperative construction and a special negator in the negative imperative which is not used in standard negation (see \citealt{DevosVanOlmen2013} on similar patterns in other Bantu languages). 

In non-verbal clauses there are three copulas: \textit{ni}, \textit{li}, and \textit{ta}, as well as the semi-copula \textit{bba}, which can be used in some of the contexts where the real copulas occur, i.e.\,in locational clauses and predicative possession.
In the negation of proper inclusion, equative and attributive clauses in the present tense, as well as with a non-pronominal subject, the invariable negative copula \emph{ti} is used. 
In other non-verbal clauses, the copulas \textit{li}, \textit{ta}, as well as the semi-copula \textit{bba} are negated by means of the standard negation marker \textit{ti-}. 
Furthermore, there is a dedicated negative existential \textit{ndoo}, which to our knowledge has not been reported for closely-related Bantu languages. 

The article further provides an overview of other aspects of clausal and non-clausal negation, as well as discusses i.a.\,the scope of negation, phasal polarity items, and reinforcing negation. 
A short summary of the different aspects of negation that are discussed in this paper is provided in Table~\ref{tab:ruulietal:3}.

\begin{table}
\small
\begin{tabularx}{\textwidth}{p{23mm}Qp{23mm}l}
\lsptoprule
{\bfseries Negation strategy} & {\bfseries Functions} & {\bfseries Examples} & \textbf{Section}\\
\midrule
Prefix \textit{ti-} & Standard negation (all tenses and aspects) & (\ref{ex:ruulietal:2}--\ref{ex:ruulietal:3}) & \sectref{sec:ruulietal:2.1}\\
& Various non-main clauses & \REF{ex:ruulietal:23}, (\ref{ex:ruulietal:27c}--\ref{ex:ruulietal:27d}), \REF{ex:ruulietal:28b}, (\ref{ex:ruulietal:29a}--\ref{ex:ruulietal:29b}) & \sectref{sec:ruulietal:2.4}\\
& Negation of non-verbal locational predication & \REF{ex:ruulietal:20} & \sectref{sec:ruulietal:2.3}\\
& Negation of non-verbal predication of proper inclusion, equative and attributive clauses (non-present and with pronominal subjects) & \REF{ex:ruulietal:19} & \sectref{sec:ruulietal:2.3}\\
\midrule 
Copula \textit{ti}  & Negation of non-verbal predication of proper inclusion, equative and attributive clauses (present with non-pronominal subjects) & \REF{ex:ruulietal:18} & \sectref{sec:ruulietal:2.3}\\
\midrule 
Prefix \textit{ta-} & Negative imperative, negative hortative & \REF{ex:ruulietal:8}, (\ref{ex:ruulietal:10}--\ref{ex:ruulietal:11}) & \sectref{sec:ruulietal:2.2}\\
\midrule 
Particle \textit{ndoo} & Negative existential & (\ref{ex:ruulietal:22b}--\ref{ex:ruulietal:22c}), \REF{ex:ruulietal:33} & \sectref{sec:ruulietal:2.3}\\
\midrule 
Prefix \textit{bu-} & Negation of infinitives & \REF{ex:ruulietal:26} & \sectref{sec:ruulietal:2.4.1}\\
\midrule 
Prefix \textit{ti- +} prefix \textit{kya-} ‘PERS’ & Phasal polarity & \REF{ex:ruulietal:4} & \sectref{sec:ruulietal:2.1}\\
\midrule 
Interjection \textit{bbe} & Negative reply & (\ref{ex:ruulietal:31}A1--\ref{ex:ruulietal:31}A3) & \sectref{sec:ruulietal:3.1}\\
\lspbottomrule
\end{tabularx}
\caption{Overview of negation strategies in Ruuli}
\label{tab:ruulietal:3}
\end{table}


% \begin{table}%old table
% \caption{Summary of the different aspects of negation in Ruuli}
% \label{tab:summaryRuuli}
% \begin{tabular}{ll}
% \toprule
% Context & Construction\\
% \midrule
% Standard negation    & \textit{ti-}\\
% Non-declarative clauses      & \textit{ta-}\\
% Non-verbal clauses  & \textit{ti}/\textit{ti-}/negative existential\\
% Non-main clauses    & \textit{t(a)-}/\textit{ti-}\\
% Lexicalized negatives     & negative existential \textit{ndoo} \\
% Negative replies     & \textit{bbe}, disagreement with polarity\\
% Negative indefinites  & negative existential construction\\
% Negative derivation  & \textit{nambula}\\
% Scope of negation   & verb-dependent\\
% Marking of NPs & no augment\\
% Reinforcing negation & postverbal adverbials\\
% Complex clauses & conjunctions \textit{gonze, ate, baitu} \\
% \bottomrule                                         
% \end{tabular}
% \end{table}
% \lv{This table summarizes functions and their possible expressions. We would rather you started from the negation strategies available in Ruuli and listed the functions relevant for each one of them. We are appending examples of such tables in the mail.}

\section*{Authors' contribution}
ER wrote her MA thesis under supervision of AWM on the topic of this paper, collected additional data to supplement the available corpus, and lead the preparation of the first manuscript of this paper. 
All authors contributed to the first manuscript and first review round. 
SN and AWM revised the paper in the second review the round and finalized the article. 
SN served as the PI of the project within which the data collection for this article was carried out.

\section*{Acknowledgements}
We thank the Volkswagen Foundation for funding the project \textit{A comprehensive bilingual talking Ruruuli/Lunyala-English dictionary with a descriptive basic grammar for language revitalisation and enhancement of mother-tongue based education} (2017--2019), which allowed the data collection for this article.
We are grateful to the editors of this volume for the comments and suggestions in the first review round and to the two anonymous reviewers in the second review round for the many insights on negation in Bantu.

\newpage
\section*{Abbreviations}
\begin{tabularx}{.45\textwidth}{lQ}
1, 2, 3, etc.& noun class\\
1\textsc{sg} & first person singular\\ 
2\textsc{sg} & second person singular\\
1\textsc{pl} & first person plural\\
2\textsc{pl} & second person plural\\
\textsc{appl} & applicative\\
\textsc{aug} & augment\\
\textsc{aux} & auxiliary\\
\textsc{caus} & causative\\
\textsc{com} & comitative\\
\textsc{cop} & copula\\
\textsc{fv} & final vowel\\
\textsc{foc} & focus\\
\textsc{fut} & future\\
\textsc{gen} & genitive\\
\textsc{hab} & habitual\\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\textsc{inf} & infinitive\\
\textsc{loc} & locative\\
\textsc{narr} & narrative\\
\textsc{neg} & negative\\
\textsc{negex} & negative existential\\
\textsc{obj} & object\\
\textsc{persis} & persistive\\
\textsc{pfv} & perfective\\
\textsc{pl} & plural\\
\textsc{poss} & possessive\\
\textsc{prog} & progressive\\
\textsc{pst} & past\\
\textsc{rel} & relativizer\\
\textsc{sbj} & subject\\
\textsc{sbjv} & subjunctive\\
\textsc{sg} & singular\\
\textsc{tam} & tense, aspect, mood\\
\\
\end{tabularx}



\printbibliography[heading=subbibliography,notkeyword=this]

\end{document}
