\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233756}
\author{Valentina Schiattarella\affiliation{University of Naples, L'Orientale}}%vale.schiattarella@gmail.com
\title{Negation in Siwi} 
\abstract{The paper focuses on the description of negation in Siwi, a Berber (Afroasiatic) language spoken in Siwa (Egypt). The Siwi negative system in declarative main clauses is strikingly symmetric, if compared to other Berber languages. Symmetry is also found between declaratives and non-declaratives, as they share the same negative marker. The paper explores many aspects of the system, including negation in non-main clauses and non-verbal predications. The final part presents some diachronic notes and a short comparison with some other Berber languages.

\keywords{Siwi Berber, Egypt, Afroasiatic, negation, typology}}

\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 

\section{The language}\label{sec:siw:1}
Siwi (ISO: \href{https://iso639-3.sil.org/code/siz}{siz}, Glottocode: \href{https://glottolog.org/resource/languoid/id/siwi1239}{siwi1239}) belongs to the Berber language family (Afroasiatic phylum). Berber languages are spoken in Morocco, Algeria, Tunisia, Libya and Egypt, as well as in Mauritania, Mali, Niger and Burkina Faso. Siwi is the only Berber language spoken in Egypt. Its speakers are concentrated in two oases: Siwa and El Gaṛa. These two oases are located in the Western Desert and are very close to the Libyan border. The main oasis, Siwa, is inhabited by over 30,000 people, including Siwi, Bedouins and Egyptians who have come from other parts of the country, and settled in the oasis mainly for work. Siwi people are almost all bilingual, as the vast majority of the population speaks Bedouin and/or Egyptian Arabic as well as Siwi. 

There is almost no dialectal variation between the different parts of the oasis. Small differences can be found between the Siwi spoken in Siwa and the one in El Gaṛa, though data on the latter are scarce \citep[102--109]{Schiattarella2017}. The language is influenced by different strata of Arabic in many parts of its grammar \citep{Souag2013}. 

Internal classification is still a matter of debate among Berberologists, but Siwi is oftentimes considered a language which shares many aspects with Sokna and El Fogaha (Libya), as well as with the Zenatic group (\citealt{Souag2013}; \citealt{Kossmann2020}). A description of some aspects of Siwi’s negation system can be found in the first grammar of Siwi by \citet[48--49]{Laoust1931}, but also in \citet{Leguil1986}, \citet[224, 230, 244]{Vycichl2005}, \citet[209--214]{Souag2013} and \citet{Schiattarella2015}.

Some grammatical features of Siwi are particularly important – as unique within Berber – and others are also typologically rare. Among them, the stress shift\footnote{Stress [  \'{}  ] will only be marked in examples taken from spontaneous discourse. A stress that falls on a long vowel will be marked only by [\={}].} on nouns as a means to code – together with other functions – definiteness and indefiniteness, and the presence of a special series of suffixes for demonstratives that agree in gender and number with the addressee.
Unlike most Berber languages, there is no state distinction on nouns in Siwi. 
The language has a distinct class of adjectives. Its verbal system is based mainly on three stems: perfective, imperfective and aorist. The perfect of result, based on the perfective, is absent in most Berber languages. More details are given in \sectref{sec:siw:2.1}. Subject affixes are obligatory on verbs, and lexical subjects co-referent with those affixes are rarely present. 
No postpositions are synchronically attested. In order to modify a noun, the genitive preposition \textit{n} ‘of’ can be used before a noun (possessor), an adjective and a relative clause in adnominal, adjectival and relative clause constructions, respectively. Relative clauses are postnominal, and the presence of a relativizer depends on the nature of the relative clause (descriptives require no relativizer, restrictives do). Word order is syntactically and pragmatically determined when only one noun is present. When two nouns are lexically expressed in constructions with transitive verbs, the only attested order is AVO (\citealt{MettouchiSchiattarella2018}). 

There are two ways negation can be marked in Siwi: the prefix \textit{la-} negates verbs and members of several other grammatical categories and is also found in the negative nominal coordination (\textit{la}... \textit{la}), while the negative marker \textit{ačči /} \textit{qačči} (\citealt{Laoust1931}; \citealt{Vycichl2005}: 244; \citealt{Souag2013}: 210) is used for some non-verbal clauses and for contrastive/metalinguistic negation. \textit{ula} is the pro-sentential negation marker, but it can also be found in other contexts. The functions of these markers in different domains will be analysed in detail in the following sections. Diachronic and comparative notes will be given in \sectref{sec:siw:4.6.3} and concluding remarks in \sectref{sec:siw:5}. 

The data for this article were collected by the author over the course of several fieldwork trips and through online sessions between 2011 and 2022; collection methods mainly included recordings of spontaneous data (monologues and dialogues of variable length) and elicitation sessions with both male and female speakers of different ages. 

The paper is based on the negation questionnaire provided by the editors (\citetv{chapters/questionnaire}).
All examples come from my corpus.

\section{Clausal negation}\label{sec:siw:2}

\subsection{Standard negation}\label{sec:siw:2.1}

Aspectual forms relevant for verbs in Siwi (as in all other Berber languages) include the perfective, the imperfective and the aorist. Two additional verb forms based on the perfective and the imperfective stems, but featuring a suffix -\textit{a} that follows all other suffixes (subject, object and dative indexes when present), are also attested. 

Without going into many details, it can be useful to recall here that, irrespective of the time frame (it can have present and past reference), the perfective is mainly used in narratives, for processes that only take place once or are seen as completed, and with stative verbs. The imperfective is used for habitual, iterative and frequentative actions, while the aorist is mainly found for the hypothetical, future, irrealis, potential and optative, but also for actions repeated in block (see example \REF{ex:siw:8} below). The perfect of result (based on the perfective) focuses on the result and establishes a pragmatic link with what precedes it. The imperfective -\textit{a} in the affirmative is rare and is used when the event or process expressed by the verb is concomitant with another (see \citealt[197--199]{Souag2013}; \citealt{Schiattarella2019}).

Negation is conveyed by the addition of the negative prefix \textit{la}{}- only, see \REF{ex:siw:1} for perfective, \REF{ex:siw:2} for imperfective, \REF{ex:siw:3} for perfect of result and \REF{ex:siw:6} for aorist. Negation of the imperfective -\textit{a} \REF{ex:siw:4} is not attested.

When \textit{la}{}- is prefixed to the verb and it precedes a vowel, some phonetic changes take place: it is reduced to \textit{l-} or it triggers an assimilation between the \textit{a{}-} of the negative prefix and the following vowel or semi-vowel (\textit{i}, \textit{u} and \textit{y} only). Both possibilities are exemplified with the perfective stem in \REF{ex:siw:1}.

\eabox{\label{ex:siw:1}\label{ex:siw:1a}\label{ex:siw:1b}\label{ex:siw:1c} Perfective\medskip

\begin{tabularx}{0.85\textwidth}{ll  llll}
        & Affirmative & \multicolumn{4}{c}{{Negative}} \\
    \cmidrule(l){2-3}\cmidrule(r){5-6}
        &           &               &   &                   & \hfill a + u > o \\
    a. & usi-x     & *la-usi-x     & > & l-usi-x / lo-si-x \\
        & come{\textbackslash}\textsc{pfv-1sg} &  &  & \multicolumn{2}{l}{{ \textsc{neg}{}-come{\textbackslash}\textsc{pfv-1sg}}} \\
        &‘I came.’ &&& \multicolumn{2}{l}{‘I did not come.’}\\

        &           &               &   &                   &  \hfill a + i > e \\
    b.  & if-ax     & *la-if-ax     & > & l-if-ax / le-f-ax \\
        & find{\textbackslash}\textsc{pfv-1sg} &  &  & \multicolumn{2}{l}{ \textsc{neg}{}-find{\textbackslash}\textsc{pfv-1sg}}\\
    & ‘I found.’ &  &  & \multicolumn{2}{l}{‘I did not find.’}\\
        &           &               &   &                   &  \hfill a + y > ey \\
    c. & yə-čča    & *la-yə-čča    & > & l-yə-čča / le-yə-čča \\
    & \textsc{3sg.m}{}-eat{\textbackslash}\textsc{pfv} &  &  & \multicolumn{2}{l}{\textsc{neg-3sg.m}{}-eat{\textbackslash}\textsc{pfv}}\\
    & ‘He ate.’ &  &  & \multicolumn{2}{l}{‘He did not eat.’}\\
\end{tabularx}
}
% \mm{@LSP: fix table so that "a + u > o", "a + i > e" and "a + y > ey" appear on one line as in the proofreading version on paperhive.}

% \newpage
% \ea\label{ex:siw:1} Perfective
%     \ea  a + u > o
%         %came
%         \ea Affirmative\\ \label{ex:siw:1a}
%             usi-x\\
%             come{\textbackslash}\textsc{pfv-1sg}\\
%             \glt `I came.'
%         \ex Negative\\
%         \gll *la-usi-x     >  {l-usi-x / lo-si-x}\\
%              {}           {} \textsc{neg}{}-come{\textbackslash}\textsc{pfv-1sg}\\
%              \glt `I did not come.'
%         \z
%         % found
%     \ex a + i > e
%         \ea Affirmative\\ \label{ex:siw:1b}
%         if-ax\\
%         find{\textbackslash}\textsc{pfv-1sg}\\
%         \glt `I found.'
%
%         \ex Negative\\
%         \gll *la-if-ax > {l-if-ax / le-f-ax}\\
%         {} {} \textsc{neg}{}-find{\textbackslash}\textsc{pfv-1sg}\\
%         \glt `I did not find.'
%         \z
%         %ate
%     \ex a + y > ey
%         \ea Affirmative\\ \label{ex:siw:1c}
%         yə-čča\\
%         \textsc{3sg.m}{}-eat{\textbackslash}\textsc{pfv}\\
%         \glt `He ate.'
%
%         \ex Negative\\
%         \gll *la-yə-čča  > {l-yə-čča / le-yə-čča} \\
%              {} {} \textsc{neg-3sg.m}{}-eat{\textbackslash}\textsc{pfv}\\
%         \glt `He did not eat.'
%         \z
%     \z
% \z
%
% \newpage

\ea\label{ex:siw:2} Imperfective
    \ea
        \gll əttasi-x. \\
            come{\textbackslash}\textsc{ipfv-1sg} \\
        \glt ‘I am coming.’ \\
    \ex
        \gll l-əttasi-x. \\ 
             \textsc{neg}{}-come{\textbackslash}\textsc{ipfv-1sg}\\
        \glt ‘I am not coming.’
    \z
\z

\ea\label{ex:siw:3} Perfect of result 
    \ea
        \gll usi-x-a. \\
            come{\textbackslash}\textsc{pfv-1sg-prag}  \\ 
        \glt ‘I have come.’
    \ex 
        \gll l-usi-x-a.\\
         \textsc{neg-}come{\textbackslash}\textsc{pfv-1sg-prag}\\
        \glt ‘I have not come.’
    \z
\z

\ea\label{ex:siw:4}Imperfective -\textit{a}\newline
    \gll əttasi-x-a...\\
         come{\textbackslash}\textsc{ipfv-1sg-prag}\\
    \glt ‘(While) I was coming...’
\z


Negative imperfective-\textit{a} is not found in my corpus and rejected by the speakers in elicitations. 

The aorist stem is always preceded by the irrealis marker \textit{ga}{}-. \citet[187]{Souag2013} mentions the occasional variant \textit{ʔa}{}-, which “may be a retention of \textit{a(d)},\footnote{The prefix \textit{a(d)} is the most widespread preverb for the aorist in Berber.} or just a shortening of \textit{ga-}” . This form is not attested in my corpus, probably because of its low frequency. The prefixation of \textit{ga}{}- involves the same phonological changes seen before with \textit{la}{}-, see \REF{ex:siw:5}. 
Both forms are attested in my corpus. 

\eabox{\label{ex:siw:5}
    \begin{tabular}{lllll}
        a. & *ga-usi-x &>& g-usi-x / go-si-x \\ 
        & & & \multicolumn{2}{l}{\textsc{irr}{}-come{\textbackslash}\textsc{aor-1sg}} \\
        & & & \multicolumn{2}{l}{‘I will come.’} \\
        
        b. & *ga-if-ax &>& g-if-ax / ge-f-ax \\
        & & & \multicolumn{2}{l}{\textsc{irr}{}-find{\textbackslash}\textsc{aor-1sg}} \\
        & & & \multicolumn{2}{l}{‘I will find.’}\\
        
        c. & *ga-yə-čč &>& g-yə-čč / ge-yə-čč \\
        & & & \multicolumn{2}{l}{\textsc{irr-3sg.m}{}-eat{\textbackslash}\textsc{aor}}\\
        & & & \multicolumn{2}{l}{‘He will eat.’}\\
    \end{tabular}
} 


The aorist with \textit{ga}{}- in the affirmative corresponds to two negative forms, both featuring the prefixation of \textit{la}{}-. Examples in \REF{ex:siw:6} show them both: the first example in the negative is where \textit{la}{}- precedes \textit{ga}{}-, the second where \textit{la}{}- precedes the prefix \textit{a}{}- only (dropping of \textit{g}). In contrast to those seen before, the latter form entails more phonetic changes (vowel lengthening).

\eabox{\label{ex:siw:6} \small
    \begin{tabularx}{\textwidth}{p{1mm}l ll}
    & {Affirmative} & \multicolumn{2}{c}{{Negative}}\\
    \cmidrule(r){2-2}\cmidrule(l){3-4}
    &  &  & la- + a- + C = lā \\
    a. & ga-t-usəd & \textbf{la}{}-ga-t-usəd & \textbf{lā}{}-t-usəd\\
    & \textsc{irr}{}-3\textsc{sg.f}{}-come{\textbackslash}\textsc{aor} & \textsc{neg}{}-\textsc{irr}{}-3\textsc{sg.f}{}-come{\textbackslash}\textsc{aor} & \textsc{neg{\textbackslash}irr}{}-3\textsc{sg.f}{}-come{\textbackslash}\textsc{aor}\\
    & ‘She will come.’ & \multicolumn{2}{c}{‘She will not come.’}\\
    \tablevspace

    &  &  &  la- + a- + u = lō \\
    b. &  g-usi-x & \textbf{la}{}-g-usi-x / la-go-si-x & \textbf{lō}{}-si-x\\
    & \textsc{irr}{}-come{\textbackslash}\textsc{aor}{}-1\textsc{sg} & \textsc{neg-irr}{}-come{\textbackslash}\textsc{aor}{}-1\textsc{sg} & \textsc{neg{\textbackslash}irr}{}-come{\textbackslash}\textsc{aor}{}-1\textsc{sg}\\
    &  ‘I will come.’ & \multicolumn{2}{c}{ ‘I will not come.’}\\
\tablevspace

    &  &  & la- + a- + i = lē\\
    c. & g-if-ax & \textbf{la}{}-g-if-ax / la-ge-f-ax & \textbf{lē}{}-f-ax\\
    & \textsc{irr}{}-find{\textbackslash}\textsc{aor-1sg} & \textsc{neg}{}- \textsc{irr}{}-find{\textbackslash}\textsc{aor-1sg} & \textsc{neg{\textbackslash}irr}{}-find{\textbackslash}\textsc{aor-1sg}\\
    & ‘I will find.’ & \multicolumn{2}{c}{‘I will not find.’}\\
\tablevspace

    &  &  & la- + a- + y = l\={e}y\\
    d. &  g-yə-čč [ge-yə-čč] & \textbf{la}{}-g-yə{}-čč / \textbf{la}{}-ge-yə{}-čč & \textbf{l\={e}}-yə{}-čč \\
    & \textsc{irr}{}-3\textsc{sg.m}{}-eat{\textbackslash}\textsc{aor} & \textsc{neg}{}-\textsc{irr}{}-\textsc{3sg.m}{}-eat{\textbackslash}\textsc{aor} & \textsc{neg}{\textbackslash}\textsc{irr}{}-\textsc{3sg.m}{}-eat{\textbackslash}\textsc{aor}\\
    & ‘He will eat.’ & \multicolumn{2}{c}{‘He will not eat.’}\\
    \end{tabularx}
}

% \largerpage
% \ea\label{ex:siw:6}
%     \ea la- + a- + C = lā %used
%         \ea Affirmative\\
%         ga-t-usəd\\
%         \textsc{irr}{}-3\textsc{sg.f}{}-come{\textbackslash}\textsc{aor}\\
%         \glt `She will come.'
%
%         \ex Negative I\\
%         \textbf{la}{}-ga-t-usəd\\
%         \textsc{neg}{}-\textsc{irr}{}-3\textsc{sg.f}{}-come{\textbackslash}\textsc{aor}\\
%
%         \ex Negative II\\
%         \textbf{lā}{}-t-usəd\\
%         \textsc{neg{\textbackslash}irr}{}-3\textsc{sg.f}{}-come{\textbackslash}\textsc{aor}\\
%         \glt `She will not come.'
%         \z
%
%     \ex la- + a- + u = lō%usi
%         \ea Affirmative\\
%          g-usi-x\\
%         \textsc{irr}{}-come{\textbackslash}\textsc{aor}{}-1\textsc{sg}\\
%         \glt `I will come.'
%         \ex Negative I\\
%         \textbf{la}{}-g-usi-x / la-go-si-x\\
%         \textsc{neg-irr}{}-come{\textbackslash}\textsc{aor}{}-1\textsc{sg}\\
%         \ex Negative II\\
%         \textbf{lō}{}-si-x\\
%         \textsc{neg{\textbackslash}irr}{}-come{\textbackslash}\textsc{aor}{}-1\textsc{sg}\\
%         \glt `I will not come.'
%         \z
%     \ex la- + a- + i = lē%if
%         \ea Affirmative\\
%         g-if-ax \\
%         \textsc{irr}{}-find{\textbackslash}\textsc{aor-1sg}
%         \glt `I will find.'
%         \ex Negative I\\
%         \textbf{la}{}-g-if-ax / la-ge-f-ax\\
%         \textsc{neg}- \textsc{irr}{}-find{\textbackslash}\textsc{aor-1sg}
%         \ex Negative II\\
%         \textbf{lē}{}-f-ax\\
%         \textsc{neg{\textbackslash}irr}{}-find{\textbackslash}\textsc{aor-1sg}
%         \glt `I will not find.'
%         \z
%      \ex la- + a- + y = l\={e}y%ye
%         \ea Affirmative \\
%          g-yə-čč [ge-yə-čč]\\
%         \textsc{irr}{}-3\textsc{sg.m}{}-eat{\textbackslash}\textsc{aor}\\
%         \glt `He will eat.'
%         \ex Negative I\\
%         \textbf{l\={e}}-yə{}-čč \\
%         \textsc{neg}{}-\textsc{irr}{}-\textsc{3sg.m}{}-eat{\textbackslash}\textsc{aor}
%         \ex Negative II \\
%          \textbf{la}{}-g-yə{}-čč / \textbf{la}{}-ge-yə{}-čč\\
%         \textsc{neg}{\textbackslash}\textsc{irr}{}-\textsc{3sg.m}{}-eat{\textbackslash}\textsc{aor}
%         \glt `He will not eat.'
%         \z
%     \z
% \z


Perfective and aorist stems are, synchronically, not distinguished morphologically in both the affirmative and negative domains – except for a restricted class of verbs, such as CəC (\textit{fəl} ‘leave’, \textit{wən} ‘mount’), əCC (\textit{əčč} ‘eat’) and VC (\textit{uš} 'give'). When the irrealis marker \textit{ga-} is dropped in the negative, the vowel lengthening is the only way of distinguishing the perfective form from the aorist, compare \REF{ex:siw:7a} and \REF{ex:siw:7b}.


\ea\label{ex:siw:7}
    \ea\label{ex:siw:7a}
        \gll \textbf{la}{}-t-usəd. \\ 
             \textsc{neg-3sg.f}{}-come{\textbackslash}\textsc{pfv} \\
        \glt ‘She did not come.’
    \ex\label{ex:siw:7b}
        \gll \textbf{lā}{}-t-usəd. \\
             \textsc{neg}{\textbackslash}\textsc{irr-3sg.f}{}-come{\textbackslash}\textsc{aor} \\
        \glt ‘She will not come.’
    \z
\z 

Following \citegen{Miestamo2005} terminology, it can be said that the Siwi negative system in declarative main clauses only shows a slight paradigmatic asymmetry: while most verbal clauses form their negative by prefixing the negative marker only, there is no one-to-one correspondence as far as the imperfective -\textit{a} and the aorist are concerned. The former is indeed not attested in the negative and the latter corresponds to two different forms in the negative. 

Other paradigmatic asymmetries are also found in terms of distribution. Although the negation of the \textit{ga}{}-aorist is attested, its semantic functions in the negative domain are not the same as in the affirmative. As mentioned above, the aorist can be used in the affirmative in the domain of the irrealis (future, possibility, probability, hypothetical, etc.), but also for a series of actions that repeat in blocks \citep[100]{Leguil1986} (i.e. in recipes, where each step takes place once, but the whole procedure needs to be repeated in its entirety every time one wants to prepare the same dish again). Often, the referent is generic: see example \REF{ex:siw:8} in which the speaker explains what everybody does every year during the Eid celebration.

\ea\label{ex:siw:8}
    \gll kull әǧǧə́n g-yə́-ɣṛəṣ iɣéd / g-yə́-čč-әn / g-yə́-sw-әn.\\
         every one{\textbackslash}\textsc{m} \textsc{irr-3sg.m}{}-slaughter{\textbackslash}\textsc{aor} male\_goat{\textbackslash}\textsc{sg.m} / \textsc{irr}{}-3-eat{\textbackslash}\textsc{aor-pl} / \textsc{irr}{}-3-drink{\textbackslash}\textsc{aor-pl}  \\ 
    \glt ‘Everyone slaughters a male goat, (then) they eat and they drink.’
\z

When the aorist is used in the negative, it never expresses this latter function (the imperfective is used instead). The \textit{ga}{}-aorist in the negative domain is only used when the speaker wants to express his/her evaluation that something hypothetical could be realized, or for the future, as in \REF{ex:siw:9}.



\ea\label{ex:siw:9}
    \gll kan t-om t-uṭ\'{a} / əlmuškl\'{ə}t / aħəǧǧáǧ g-y\'{ə}-ṃṃut / múmken \textbf{lō-səd} / aʕtáqəd yáʕni / aħ\'{ə}ǧǧaǧ daw-óm hánta / g-y\'{ə}-ṃṃut.\\
         if \textsc{dem{\textbackslash}f-2sg.f} \textsc{3sg.f}{}-fall{\textbackslash}\textsc{pfv} / problem{\textbackslash}\textsc{sg.f} / pilgrim{\textbackslash}\textsc{sg.m} \textsc{irr-3sg.m}-die{\textbackslash}\textsc{aor} / maybe \textbf{\textsc{neg{\textbackslash}irr.(3sg.m)}-come{\textbackslash}\textsc{aor}} / belief{\textbackslash}\textsc{sg.m} I\_mean / pilgrim{\textbackslash}\textsc{sg.m} \textsc{dem{\textbackslash}m-2sg.f} what / \textsc{irr-3sg.m}{}-die{\textbackslash}\textsc{aor}\\
    \glt‘If it [the stick] falls, it is a problem (i.e. bad luck). The pilgrim might die, he may not come back. I mean, it is a belief. What [happens] to the pilgrim? He might die.’
\z

Nevertheless, the \textit{ga}{}-aorist is obligatory, in both the negative and affirmative domain, when it follows the complementizer \textit{anni} introducing a purpose clause (see \sectref{sec:siw:2.4}). 

Contrary to many other Berber languages (see \sectref{sec:siw:4.6.3}), in Siwi, object and dative indexes always follow the verb (and its subject affixes), even in the presence of the negative marker \textit{la}{}-, as in \REF{ex:siw:10}.

\ea\label{ex:siw:10}
    \gll \textbf{la}-t-uš-\textbf{as-t}.\\
         \textsc{\textbf{neg}-3sg.f}-give\textbackslash\textsc{pfv-\textbf{dat\textbackslash3sg-do\textbackslash3sg.m}}\\
    \glt ‘She did not give it to him/her.’
\z

As far as frequency is concerned, there is a much higher percentage of verbs in the affirmative domain than in the negative, which is not at all surprising, as the negative is not merely the opposite counterpart of the affirmative, but differs from it in a number of ways. Affirmatives are usually pragmatically different from negatives: affirmatives are considered to be present as background knowledge and negatives take longer to process \citep{Givon1978,Horn1989,Miestamo2005}.


\subsection{Negation in non-declaratives}\label{sec:siw:2.2}

As in declarative clauses, \textit{la}{}- is used also to mark negation in non-declarative ones. Imperative uses the aorist \REF{ex:siw:11a}, while prohibitive uses the imperfective stem \REF{ex:siw:11b}.


\ea\label{ex:siw:11}
    \ea\label{ex:siw:11a}
        \gll ə{\fdot}{\fdot}aɣ!\\
             go\_out{\textbackslash}\textsc{aor.imp} \\
        \glt ‘Go out!’ 
    \ex\label{ex:siw:11b}
        \gll \textbf{la-tə{\fdot}{\fdot}aɣ}! \\
             \textbf{\textsc{neg}-go\_out{\textbackslash}\textsc{ipfv.imp}}\\
        \glt ‘Do not go out!’
    \z
\z 

In this respect, there is symmetry between declaratives and imperatives in marking negation, because they both use the negative marker \textit{la}{}-, but there is asymmetry between affirmative and negative imperatives, because, besides \textit{la}{}-, the prohibitive uses a different aspectual stem (see \citealt{MiestamoAuwera2007} for different types of (a)symmetries between declarative and imperative clauses in other languages).

The function of the prohibitive is to prevent something from happening or to ask for something that is already happening to stop, see \REF{ex:siw:12}.

\ea\label{ex:siw:12}
    \gll tə-bdú aglás. / yə-ṃṃ{}-ás \textbf{lá-gəlləs}. \\
         \textsc{3sg.f-}start{\textbackslash}\textsc{pfv} cry{\textbackslash}\textsc{vn} / \textsc{3sg.m}-say{\textbackslash}\textsc{pfv-dat\textbackslash3sg} \textbf{\textsc{neg}-cry\textbackslash\textsc{ipfv.imp}} \\
    \glt ‘[The woman] started to cry. [The Christian] said to her: “Don’t cry.”’
\z

The hortative uses the first person plural form of the \textit{ga}{}-aorist as in \REF{ex:siw:13}.


\ea\label{ex:siw:13}
    \gll g\'{a}-n-fəl! \\
         \textsc{irr-1pl}-leave{\textbackslash}\textsc{aor}\\
    \glt  ‘Let’s leave!’
\z

The negation of the hortative is made through the prefixation of \textit{la}- only, see \REF{ex:siw:14}.

% \ea\label{ex:siw:14}
%     \gll gá-n-fəl! // \textbf{lā-fl-əm}! \\
%          \textsc{irr-1pl}-leave{\textbackslash}\textsc{aor} // \textbf{\textsc{neg{\textbackslash}irr}-leave{\textbackslash}\textsc{aor-2pl}}\\
%     \glt‘Let’s leave!’  ‘You will not go!’
% \z
\ea\label{ex:siw:14}
        \gll lá-ga-n-fəl!  \\
             \textsc{neg-irr-1pl}-leave{\textbackslash}\textsc{aor} \\
        \glt‘Let’s not leave!’  

\z


The jussive is a rare form which uses the aorist stem and the prefix \textit{ədda}{}-, see \REF{ex:siw:15}.\footnote{This form is called by \citet[192--193]{Souag2013} “suggestative”, because it is also found in contexts where the speaker suggests something, but at the same time asserts their indifference to the action being accomplished. See also \citet[13]{Prasse1973} on the Tuareg jussive (\textit{injonctif} in French) and references to forms with similar functions in other Berber languages.} It is used mainly with first and third persons.
This form cannot be negated.


\ea\label{ex:siw:15}
    \gll ədda-t-úsəd ssra b\'{a}dri / áčči dəgyáṭ.\\
         \textsc{juss-3sg.f-}come{\textbackslash}\textsc{aor} morning early / \textsc{neg} evening \\
    \glt ‘Let her come in the morning, not in the evening.’
    \z


There are no special constructions for negative interrogatives, as shown by \REF{ex:siw:16a} and \REF{ex:siw:16b}.

\ea\label{ex:siw:16}
    \ea\label{ex:siw:16a}
        \gll xsi-ṭ?  \\ 
             want{\textbackslash}\textsc{pfv-2sg}? \\
        \glt ‘Do you want?’
    
    \ex\label{ex:siw:16b}
        \gll \textbf{l\'{a}-xsi-ṭ}? \\
             \textbf{\textsc{neg-}want{\textbackslash}\textsc{pfv-2sg}?}\\
        \glt ‘Don’t you want?’
    \z
\z 

Oftentimes, negative interrogatives are used when the speakers rather aim at an affirmative interpretation, as in \REF{ex:siw:17}.

\ea\label{ex:siw:17}
    \gll \textbf{lá-ẓṛi-ṭ} aggʷíd / yə-flá s əgdá. \\
         \textbf{\textsc{neg}{}-see{\textbackslash}\textsc{pfv-2sg}} man{\textbackslash}\textsc{sg.m} /  \textsc{3sg.m-}leave{\textbackslash}\textsc{pfv} from here \\
    \glt ‘Didn’t you see a man, he left from here…?’ (meaning ‘Did you see a man…’)
\z

In cases where the question is a way to ask for the addressee’s attention, the question tag \textit{nola}, which is a merger of \textit{na} ‘or’ and \textit{ula} ‘no’, is used.

In example \REF{ex:siw:18}, the speaker, after listing different kinds of food in Siwa, asks me if I am happy with what she is telling me and then continues her narration.

\ea\label{ex:siw:18}
    \gll ssn-\'{a}ṭ-a aksúm yə-qqúṛ-a? / i-ħə́llay-am \textbf{nóla}? / g-yə-ʕṃáṛ-ən-t yə-qqúṛ-a. \\
         know{\textbackslash}\textsc{pfv-2sg-prag} meat{\textbackslash}\textsc{sg.m} \textsc{3sg.m-}be\_dry{\textbackslash}\textsc{pfv-prag} / \textsc{3sg.m}{}-be\_happy{\textbackslash}\textsc{ipfv-dat\textbackslash2sg.f} \textbf{or\_not} / \textsc{irr-3-}do{\textbackslash}\textsc{aor-pl-do\textbackslash3sg.m} \textsc{3sg.m}{}-be\_dry{\textbackslash}\textsc{pfv-prag}\\
    \glt ‘Do you know the dried meat? Are you happy or not? They make it dry.’
\z


\subsection{Negation in stative predications}\label{sec:siw:2.3}

Affirmative equative and proper inclusion predications in Siwi simply feature juxtaposition of the subject and the predicate. Their negative counterparts (examples \REF{ex:siw:19} and \REF{ex:siw:20}) feature the insertion of the negative marker \textit{ačči/qačči}, which is used also for negative contrastive focus with nouns, adjectives and verbs (see \sectref{sec:siw:4.1}).


\ea\label{ex:siw:19}
    \gll šəm \textbf{áčči} wəltmá. \\
         \textsc{idp.2sg.f} \textbf{\textsc{neg}} sister{\textbackslash}\textsc{sg.f} \\
    \glt ‘You are not my sister.’
    \z

\ea\label{ex:siw:20}
    \gll nə́tta \textbf{áčči} limám. \\
         \textsc{idp.3sg.m} \textbf{\textsc{neg}} imam{\textbackslash}\textsc{sg.m}\\
    \glt ‘He is not an imam.’
    \z

 Similarly to equative and proper inclusion predication, attributive predication also features juxtaposition of the subject and the adjective. The adjective can be in its basic form \REF{ex:siw:21a} or followed by the suffix -\textit{a} \REF{ex:siw:22a}. Negative attributive negation is achieved in both cases through the prefixation of \textit{la-}, as in \REF{ex:siw:21b} and \REF{ex:siw:22b}.

\ea\label{ex:siw:21}
    \ea\label{ex:siw:21a}
        \gll bass ənnh\'{a}ṛdin əddbáš kwayysə́n. \\ 
             but in\_the\_past thing{\textbackslash}\textsc{sg.m} good{\textbackslash}\textsc{pl.m} \\
        \glt ‘But in the past things were good.’
    
    \ex\label{ex:siw:21b}
        \gll əlǧaw \textbf{lá-kwayyəs}. \\
             weather{\textbackslash}\textsc{sg.m} \textbf{\textsc{neg-}good}{\textbackslash}\textsc{sg.m}\\
        \glt ‘The weather is not good.’
    \z
\z 

\ea\label{ex:siw:22}
    \ea\label{ex:siw:22a}
        \gll ənnháṛdin / isíwan taħk\'{i}kt-a. \\ 
             in\_the\_past / Siwa small{\textbackslash}\textsc{sg.f-prag} \\
        \glt ‘In the past, Siwa was small.’
    
    \ex\label{ex:siw:22b}
        \gll asúni n ṣɣáṛən \textbf{la-sahíl-a}. \\
             lift.up.\textsc{vn} of wood{\textbackslash}\textsc{pl.m} \textbf{\textsc{neg}{}-easy{\textbackslash}\textsc{sg.m-prag}} \\
        \glt ‘Lifting up the wood is not easy.’
    \z
\z 

A third possibility is for adjectives to be combined with the genitive preposition \textit{n} ‘of’ (see \citealt{Schiattarella2020} for the different functions of these three possibilities). In those cases, the referent having the property expressed by the adjective is in contrast with referents which do not have that property (or have a different one) or the property is considered as being exceptional or unexpected. In example \REF{ex:siw:23}, the speaker talks about different types of houses, pointing at the big one and mentioning, afterwards, that smaller ones exist as well.

\ea\label{ex:siw:23}
    \gll agbən daw-om n azəwwar. \\
         house{\textbackslash}\textsc{sg.m} \textsc{dem-2sg.f} of big{\textbackslash}\textsc{sg.m} \\
    \glt ‘This house is big.’
\z

In both attribution and predication, adjectives combined with \textit{n} ‘of’ cannot be negated. 

Locative predication features the juxtaposition of the subject and the locative phrase. The negative counterpart inserts the negative marker \textit{ačči/qačči}, as in \REF{ex:siw:24}. 

\ea\label{ex:siw:24}
    \gll šəm \textbf{áčči} g ágbən. \\
         \textsc{idp.2sg.f} \textbf{\textsc{neg}} in house{\textbackslash}\textsc{sg.m} \\
    \glt ‘You are not in the house.’
\z

Existential predication is expressed through the invariable particle \textit{di/di-}(\textit{y})\textit{a} ‘there is’. The suffix \textit{{}-}(\textit{y})\textit{a }is used when the speaker wants to intensify the actuality of the predication \citep[173]{Schiattarella2019}, as shown in \REF{ex:siw:25}.


\ea\label{ex:siw:25}
    \gll di inuwə́n / al ámra dí-ya. \\
         \textsc{ex} well{\textbackslash}\textsc{pl.m} / until now \textsc{ex-prag}\\
    \glt ‘There are some wells... Until now there are, indeed (two which still have water).’
\z

Possessive predication uses the preposition \textit{ɣuṛ} ‘at’ to which a pronoun indicating the possessor is suffixed, as in \REF{ex:siw:26}.


\ea\label{ex:siw:26}
    \gll agbə́n-sən n azə́wwar ɣúṛ-əs sə́tti n tiɣə́ṛ{\fdot}a.\\
         house{\textbackslash}\textsc{sg.m-poss{\textbackslash}3pl} of big{\textbackslash}\textsc{sg.m} at-\textsc{3sg} six of room{\textbackslash}\textsc{pl.f} \\
    \glt ‘Their big house has six rooms.’
\z

Negative existential \REF{ex:siw:27} and possessive predication \REF{ex:siw:28} use \textit{la}{}- (\textit{la}{}-\textit{di}; \textit{la}{}-\textit{ɣuṛ}{}-), when negated:


\ea\label{ex:siw:27}
    \gll \textbf{lá-di} šal l-í-ṛaħ. \\
         \textsc{neg-ex} country{\textbackslash}\textsc{sg.m} \textsc{neg-3sg.m-}go{\textbackslash}\textsc{pfv} \\
    \glt ‘There was no place he did not go.’
\z

\ea\label{ex:siw:28} 
    \gll ənšní g ədd\'{a}ffa  / \textbf{la-ɣúṛ-i} ankán. \\
         \textsc{idp.1pl} in desert{\textbackslash}\textsc{sg.m}  / \textbf{\textsc{neg}-at-1\textsc{sg}} place{\textbackslash}\textsc{sg.m} \\
    \glt ‘(Talking to someone in search for a place to hide) “We are in the desert, I don’t have any place...”’%why single+double quotation marks? explanation of the author: it is a storyteller who is quoting two different people having a conversation  > keep as it is now with both single and double quotation marks
\z

Negative existential predication is used mainly to express the absence of an entity (of something \textit{la-di šra}, or someone \textit{la-di ħədd}) \citep{Veselinova2013neg.exist}, as in \REF{ex:siw:29}.


\ea\label{ex:siw:29}
    \gll tɣáṛa nnúba i-rəḅḅáṭ-ən / \textbf{lá-di} tɣáṛa yə-qquṛ-ín-a. \\
         bread{\textbackslash}\textsc{pl.f} all 3-be\_wet{\textbackslash}\textsc{pfv-pl} / \textbf{\textsc{neg-ex}} bread{\textbackslash}\textsc{pl.f} 3-be\_dry{\textbackslash}\textsc{pfv-pl-prag}\\
    \glt ‘The pieces of bread were all wet, there were no dry pieces of bread.’
\z

\textit{la-di-}(\textit{y})\textit{a} is not attested. The function of the negative existential predication (absence) is incompatible with the function of -(\textit{y})\textit{a} after \textit{di}. This suffix, which can be attached to several grammatical categories, including verbs and adjectives, as seen in examples \REF{ex:siw:3} and \REF{ex:siw:22}, is used to mark pragmatic relevance \citep{Schiattarella2019}: in the case of existential predications, it marks the actuality of the predication. It is indeed often used in replies, when someone asks if something is present or when the fact of something being present is questioned and needs confirmation. By contrast, negative existential predications are used to assert the absence of an entity, with no presupposition of their existence in mind. 

Reinforcing negative existential and possessive predication will be dealt with in \sectref{sec:siw:4.4}.


\subsection{Negation in non-main clauses}\label{sec:siw:2.4}

In Siwi, subject and non-subject relative clauses use the same verbal form as independent clauses (compare \REF{ex:siw:30a} and \REF{ex:siw:30b}), and no change is involved in the negation, apart from the addition of the negative marker. This differs from most Berber languages, see \sectref{sec:siw:4.6.3}.


\ea\label{ex:siw:30}
    \ea\label{ex:siw:30a}
        \gll tálti tən t-sáwal \\ 
             woman{\textbackslash}\textsc{sg.f} \textsc{rel} \textsc{3sg.f-}speak{\textbackslash}\textsc{ipfv} \\
        \glt ‘the woman who speaks’
    
    \ex\label{ex:siw:30b}
        \gll tálti tən \textbf{lá-t-sawal} \\
             woman{\textbackslash}\textsc{sg.f} \textsc{rel} \textbf{\textsc{neg-3sg.f-}speak{\textbackslash}\textsc{ipfv}} \\
        \glt ‘the woman who does not speak’
    \z
\z


The complementizer \textit{anni} introduces complement clauses. Some verbs allow non-finite complement clauses, but only when the subject of the main and the complement clauses is the same (for instance, \textit{xəs} ‘want’ allows finite and non-finite complement clauses, while \textit{bdu} ‘start’ allows non-finite complement clauses, preceded or not by the preposition \textit{i} ‘to’).

Negation of non-finite complement clauses is not attested in my corpus, and the finite form of the verb is used instead \REF{ex:siw:31c}. Both finite \REF{ex:siw:31b} and non-finite \REF{ex:siw:31a} complement clauses are allowed in the affirmative, but only finite ones in the negative \REF{ex:siw:31c}.

\protectedex{\ea\label{ex:siw:31}
    \ea\label{ex:siw:31a}
        \gll xsi-x ačč\'{u}.  \\ 
             want{\textbackslash}\textsc{pfv}{}-1\textsc{sg} eat{\textbackslash}\textsc{vn} \\
        \glt ‘I want to eat.’
    \ex\label{ex:siw:31b} 
        \gll xsi-x gá-čč-ax. \\
             want{\textbackslash}\textsc{pfv-1sg} \textsc{irr}{}-eat{\textbackslash}\textsc{aor-1sg} \\
        \glt ‘I want to eat.’
    \ex\label{ex:siw:31c}
        \gll xsi-x \textbf{lá-ga-čč-ax}. \\
            want{\textbackslash}\textsc{pfv-1sg} \textbf{\textsc{neg}-\textsc{irr}-eat{\textbackslash}\textsc{aor}-\textsc{1sg}} \\
        \glt ‘I don’t want to eat.’
    \z
\z}

\noindent It is nevertheless more common to negate the verb in the main clause, as in \REF{ex:siw:32}.


\ea\label{ex:siw:32}
    \gll \textbf{l\'{a}-xsi-x} gá-čč-ax. \\
         \textbf{\textsc{neg}-want{\textbackslash}\textsc{pfv}-\textsc{1sg}} \textsc{irr}{}-eat{\textbackslash}\textsc{aor-1sg}\\
    \glt ‘I don’t want to eat.’
\z

When the complementizer \textit{anni} introduces purpose clauses, the \textit{ga}{}-aorist form is obligatory, in both the affirmative and the negative \REF{ex:siw:33}.

\ea\label{ex:siw:33}
    \gll i-ħaṭṭ\'{u}-n-as tisə́nt dabb ánni \textbf{lá-g-yə-ʕfən}. \\
         3-put{\textbackslash}\textsc{ipfv-pl-dat\textbackslash3sg} salt{\textbackslash}\textsc{sg.f} a\_lot \textsc{comp} \textbf{\textsc{neg-irr-3sg.m}-putrify{\textbackslash}\textsc{aor}} \\
    \glt ‘They will add a lot of salt, so that (the meat) will not putrify.’
\z

The negative form of the aorist preceded by \textit{a}{}- only (dropping of \textit{g}) is often attested in the negative in the context of purpose clauses, as in \REF{ex:siw:34}.
\largerpage
{\ea\label{ex:siw:34}
    \gll ga-ħáṭṭ-ax ssukkár kóm-a msabb šahín n ttáni / ánni \textbf{lē-yə-ʕṃaṛ} aẓáy. \\
         \textsc{irr}{}-put{\textbackslash}\textsc{aor-1sg} sugar{\textbackslash}\textsc{sg.m} a\_lot-\textsc{prag} because tea{\textbackslash}\textsc{sg.m} of second / \textsc{comp} \textbf{\textsc{neg{\textbackslash}irr{}-3sg.m}-do{\textbackslash}\textsc{aor}} bitter{\textbackslash}\textsc{sg.m}\\
    \glt ‘I will add a lot of sugar, because it is the second tea, so that it does not become bitter.’
\z}

Example \REF{ex:siw:33} is one of the very few cases in my corpus where the preverb \textit{ga}{}- is used after \textit{la}, when the complementizer \textit{anni} introduces purpose clauses. Nevertheless, since both forms are attested in this environment, it is not possible to talk about asymmetry.


\subsection{Negative lexicalization}\label{sec:siw:2.5}

Examples of negative lexicalizations are treated in different sections of this chapter, especially \sectref{sec:siw:3.2}, \sectref{sec:siw:3.3} and \sectref{sec:siw:3.4}.

The verb \textit{ugay} is translated as ‘refuse, forbid’, see \REF{ex:siw:35}.

\ea\label{ex:siw:35}
    \gll f\'{ə}tk-i a bénti / \textbf{t-ugáy}. \\
         open{\textbackslash}\textsc{pfv.imp-dat\textbackslash1sg} \textsc{voc} my\_girl / \textbf{\textsc{3sg.f}-refuse{\textbackslash}\textsc{pfv}} \\
    \glt ‘“Open up for me, girl.” She refused.’ 
\z

The verb \textit{ɣləb} is translated as ‘cannot, not be able to’, see \REF{ex:siw:36}.

\ea\label{ex:siw:36}
    \gll \textbf{nə-ɣlə́b} ga-nə́-ħħəl ləmšakəl-ə́nnax.\\
         \textbf{\textsc{1pl-}cannot{\textbackslash}\textsc{pfv}} \textsc{irr-1pl-}solve{\textbackslash}\textsc{aor} problem{\textbackslash}\textsc{pl.m-poss{\textbackslash}1pl}\\
    \glt ‘We could not solve our problem.'
\z

\textit{Lubah} ‘it is unlikely, probably not’ is formally a negative third person singular perfective (\textit{l-(y)-ubah}), and it is invariable in its adverbial function, see \REF{ex:siw:37}.


\ea\label{ex:siw:37}
\begin{xlist}[A: ]
    \exi{A:}
    \gll di teni g aṭil \\
         \textsc{ex} date{\textbackslash}\textsc{sg.f} in garden{\textbackslash}\textsc{sg.m} \\
    \glt ‘Are there dates in the garden?’
    \exi{B:}
    \gll \textbf{lubah} yə-ṣṣy-ən-tit-a. \\
          \textbf{probably\_not} 3-take{\textbackslash}\textsc{pfv-pl-do\textbackslash3sg.f-prag} \\
    \glt ‘Probably not, they took them.’
\end{xlist}
\z

\section{Non-clausal negation}\label{sec:siw:3}

\subsection{Negative replies}\label{sec:siw:3.1}


Polar questions in Siwi feature final vowel lengthening and higher pitch.  The negative reply uses the pro-sentential marker \textit{ula}, while the positive uses \textit{ah}, or \textit{aywa} (Arabic) or the lateral click ⟨ǁ⟩. Negative replies are also conveyed by \textit{aʔa}, also used to ask someone to stop doing something and by the dental click ⟨ǀ⟩, often repeated twice.

A negative reply can confirm a negative question or reverse the content of an affirmative question, see \REF{ex:siw:38} and \REF{ex:siw:39}.

\ea\label{ex:siw:38}
\begin{xlist}[A: ]
    \exi{Q:}
        \gll \textbf{l\'{a}-sawal-aṭ} is\'{i}wan? \\
             \textbf{\textsc{neg}-speak{\textbackslash}\textsc{ipfv-2sg}} Siwi \\
        \glt ‘Don’t you speak Siwi?’
    \exi{A:}
        \gll \textbf{ulá}. \\
              \textbf{\textsc{neg}} \\
        \glt ‘No, (I don’t)!’
\end{xlist}
\z

\ea\label{ex:siw:39}
\begin{xlist}[A: ]
    \exi{Q:}
        \gll sáwal-aṭ is\'{i}wan?  \\
             speak{\textbackslash}\textsc{ipfv-2sg} Siwi \\
        \glt ‘Do you speak Siwi?’
    \exi{A:}
        \gll \textbf{ulá}. \\
             \textbf{\textsc{neg}} \\
        \glt ‘No, (I don’t)!’
\end{xlist}
\z

There is no need to confirm through the verb a positive reply to an affirmative question \REF{ex:siw:40} or a negative reply to a negative or affirmative question (\ref{ex:siw:38}--\ref{ex:siw:39}).

\ea\label{ex:siw:40}
\begin{xlist}[A: ]
    \exi{Q:}
        \gll sáwal-aṭ iṭáli?  \\
             speak{\textbackslash}\textsc{ipfv}{}-2\textsc{sg} Italian  \\
        \glt ‘Do you speak Italian?’
    \exi{A:}
        \gll ah. \\
             yes \\
        \glt ‘Yes.’
\end{xlist}
\z

In order to reverse the content of a negative question, one can reply with both the affirmative or negative marker, but confirmation must be added through the affirmative form of the verb (\ref{ex:siw:41}--\ref{ex:siw:42}).

\ea\label{ex:siw:41}
\begin{xlist}[A: ]
    \exi{Q:}
        \gll \textbf{l\'{a}-ħħi-ṭ} i təṭ? \\
             \textbf{\textsc{neg}-go{\textbackslash}\textsc{pfv}-\textsc{2sg}} to source{\textbackslash}\textsc{sg.f} \\
        \glt ‘Didn’t you go to the source?’
    \exi{A:}
        \gll ah ħħí-x-a.\\
             yes go{\textbackslash}\textsc{pfv-1sg-prag}\\
        \glt ‘Yes, I went!’
\end{xlist}
\z

\ea\label{ex:siw:42}
\begin{xlist}[A: ]
    \exi{Q:}
        \gll \textbf{l\'{a}-ħħi-ṭ} i təṭ? \\
             \textbf{\textsc{neg}-go{\textbackslash}\textsc{pfv}-2\textsc{sg}} to source{\textbackslash}\textsc{sg.f} \\
        \glt ‘Didn’t you go to the source?’
    \exi{A:}
        \gll \textbf{ulá} ħħí-x-a.\\
             \textbf{\textsc{neg}} go{\textbackslash}\textsc{pfv-1sg-prag}\\
        \glt ‘No, I went!’
\end{xlist}
\z

If the speaker confirms the negative content of the question, the affirmative \textit{ah} ‘yes’, rather than \textit{ula}, can be used, but it needs to be followed by a negative form of the verb, as in \REF{ex:siw:43}.

\ea\label{ex:siw:43}
\begin{xlist}[A: ]
    \exi{Q:}
        \gll \textbf{lá-sawal-aṭ} iṭáli? \\
             \textbf{\textsc{neg}-speak{\textbackslash}\textsc{ipfv}-\textsc{2sg}} Italian \\
        \glt ‘Don’t you speak Italian?’
    \exi{A:}
        \gll  ah \textbf{lá-sawal-ax}.\\
              yes \textbf{\textsc{neg}-speak{\textbackslash}\textsc{ipfv}-\textsc{1sg}}\\
        \glt ‘(Yes), I don’t speak (it).’
\end{xlist}
\z

Contrary to \REF{ex:siw:41}, when reversing the content of an affirmative question, only \textit{ula} (as in \REF{ex:siw:39}) can be used and/or the verb in its negative form (as in \REF{ex:siw:44} and \REF{ex:siw:45}).

\ea\label{ex:siw:44}
\begin{xlist}[A: ]
    \exi{Q:}
        \gll ħħi-ṭ i təṭ? \\
             go{\textbackslash}\textsc{pfv}{}-2\textsc{sg} to source{\textbackslash}\textsc{sg.f} \\
        \glt ‘Did you go to the source?’
    \exi{A:}
        \gll  \textbf{ul\'{a}} \textbf{la-ħħ\'{i}-x-a}.\\
             \textbf{\textsc{neg}} \textbf{\textsc{neg}-go{\textbackslash}\textsc{pfv}-1\textsc{sg-prag}} \\
        \glt ‘No, I did not go.’
\end{xlist}
\z

\ea\label{ex:siw:45}
    \gll ħħ\'{i}-ṭ-a yən \'{a}ggʷid dawá / n w\'{i}hin / wən i-zə́nzu ləqmáš / aggʷíd la-kwayy\'{ə}s? / tə́-ṃṃ-as \textbf{la-ħħí-x-a}. /\\
          go{\textbackslash}\textsc{pfv-2sg-prag} to man{\textbackslash}\textsc{sg.m} \textsc{dem{\textbackslash}sg.m} / of whatsit / \textsc{rel} \textsc{3sg.m}-sell{\textbackslash}\textsc{ipfv} fabric{\textbackslash}\textsc{pl.m} /  man{\textbackslash}\textsc{sg.m} \textsc{neg}-good{\textbackslash}\textsc{sg.m} / \textsc{3sg.f}-say{\textbackslash}\textsc{pfv-dat\textbackslash3sg} \textbf{\textsc{neg}-go{\textbackslash}\textsc{pfv}-\textsc{1sg}-\textsc{prag}} / \\
    \glt ‘“Did you go to the man, whatsit, the one who sells the fabric, the untrustworthy man?” She said to him: “I did not go.”’
\z

When the negative marker is repeated several times, \textit{la} is used instead of \textit{ula}, as shown in example \REF{ex:siw:46}.


\ea\label{ex:siw:46}
    \gll yə́-ṃṃ{}-as \textbf{la} \textbf{la} \textbf{la} ħħí-ṭ{}-a. / tə́-ṃṃ{}-as \textbf{la-ħħí-x-a}. // \\
         \textsc{3sg.m}-say{\textbackslash}\textsc{pfv-dat\textbackslash3sg}  \textbf{\textsc{neg}} \textbf{\textsc{neg}} \textbf{\textsc{neg}} go{\textbackslash}\textsc{pfv-2sg-prag} / \textsc{3sg.f}-say{\textbackslash}\textsc{pfv-dat\textbackslash3sg} \textbf{\textsc{neg}-go{\textbackslash}\textsc{pfv}-\textsc{1sg-prag}} // \\
    \glt ‘He said to her: “No, no, no, you went (there)...” She replied: “I did not go.”’ 
\z

\subsection{Negative indefinites, adverbs and quantifiers}\label{sec:siw:3.2}

The numeral \textit{əǧǧən} ‘one’ functions as an indefinite pronoun in Siwi. It can also be used in the negative, preceded by the morpheme \textit{la}{}- (apart from my corpus, an example of this is also found in \citealt[80]{Walker1921}). In example \REF{ex:siw:47}, both the indefinite pronoun and the verb are negated by \textit{l(a)}{}-.

\ea\label{ex:siw:47}
    \gll \textbf{l-ə́ǧǧən} \textbf{l-yə́-ssən} hánta i-tṣáṛa. \\
         \textbf{\textsc{neg}-one{\textbackslash}\textsc{m}} \textbf{\textsc{neg}-\textsc{3sg.m}-know{\textbackslash}\textsc{pfv}} what \textsc{3sg.m}-happen{\textbackslash}\textsc{ipfv} \\
    \glt ‘No one knew what happened.’ (lit. ‘No one did not know what happened.’)
\z

Otherwise, in the preverbal position, the negative existential construction is used (\citealt{Haspelmath1997}; \citealt{Veselinova2013neg.exist}), followed by \textit{ħədd}, as in \REF{ex:siw:48}. This is a borrowing from Arabic ‘one, person’ widely used in both Berber and Arabic in negative contexts (\citealt{ChakerCaubet1996}).

\ea\label{ex:siw:48}
    \gll \textbf{lá-di} ħədd g-í-ṛaħ i šáli / \textbf{lá-di} ħədd g-í-ṛaħ i mmraqíyət. \\
         \textbf{\textsc{neg-ex}} person{\textbackslash}\textsc{sg.m} \textsc{irr-3sg.m}-go{\textbackslash}\textsc{aor} to citadel{\textbackslash}\textsc{sg.m} / \textbf{\textsc{neg-ex}} person{\textbackslash}\textsc{sg.m} \textsc{irr-3sg.m}-go{\textbackslash}\textsc{aor} to Maraqi \\
    \glt ‘Nobody would go to the citadel, nobody would go to Maraqi.’ (lit. ‘There is no one who would go...’)
\z

Alone (without the existential predication), it cannot appear in preverbal position, meaning ‘nobody’: if the indefinite pronoun follows the verb \REF{ex:siw:49}, the negative marker \textit{la}{}- on the verb is required, and \textit{ħədd} appears without any negative exponent (in line with the NegEarly principle, see \citealt{AuweraAlsenoy2016}, which is also the case of example \REF{ex:siw:48}, where \textit{la}{}- is on the existential \textit{di} ‘there is’, rather than on the noun).

\ea\label{ex:siw:49}
    \gll \textbf{lā-fətk-ʕ-as} albáb i \textbf{ħədd}.  \\
         \textbf{\textsc{neg{\textbackslash}irr}-open{\textbackslash}\textsc{aor}-\textsc{1sg}-\textsc{dat\textbackslash3sg}} door{\textbackslash}\textsc{sg.m} to \textbf{person{\textbackslash}\textsc{sg.m}} \\
    \glt ‘I will not open the door to anyone.’
\z

The element \textit{ħədd} can also be used in affirmative contexts (see example \REF{ex:siw:50} and \sectref{sec:siw:4.2}). In the affirmative, the indefinite pronoun \textit{əǧǧən} can be used as well, and in some cases it can be referential \REF{ex:siw:51}.



\ea\label{ex:siw:50}
    \gll ənnh\'{a}ṛdin / makənni yə-ṃṃút \textbf{ħədd} / g isíwan / təltáwen i-ṣə́rrx-ən / d i-ħáṭṭu-n iɣə́d g əṣṣḅaħ-ə́nsən. \\
         in\_the\_past / when \textsc{3sg.m}{}-die{\textbackslash}\textsc{pfv} \textbf{person{\textbackslash}\textsc{sg.m}} / in Siwa / woman{\textbackslash}\textsc{pl.f} \textsc{3-}yell{\textbackslash}\textsc{ipfv-pl} / and 3-put{\textbackslash}\textsc{ipfv-pl} ash{\textbackslash}\textsc{sg.m} in face{\textbackslash}\textsc{sg.m-poss{\textbackslash}3pl}  \\
    \glt ‘In the past, when/if somebody died in Siwa, women yelled and used to put ash on their faces.’
\z

\ea\label{ex:siw:51}
    \gll afə́nni tə-zwə́r / tə-nǧə́f \textbf{ə́ǧǧən} səg líbya.\\
         when \textsc{3sg.f}{}-grow{\textbackslash}\textsc{pfv} / \textsc{3sg.f-}marry{\textbackslash}\textsc{pfv} \textbf{one{\textbackslash}\textsc{m}} from Libya\\
    \glt ‘When she grew up, she married a guy from Libya.’
    \z

When the speaker wants to emphasize the negation, \textit{ħidda} is used instead of \textit{ħədd}, as in \REF{ex:siw:52}.

\ea\label{ex:siw:52}
    \gll \textbf{l\={a}-fətk-ʕ-as} i \textbf{ħídda}!  \\
         \textbf{\textsc{neg{\textbackslash}irr}-open{\textbackslash}\textsc{aor}-\textsc{1sg}-\textsc{dat\textbackslash3sg}} to nobody{\textbackslash}\textsc{sg.m} \\
    \glt ‘I will not open (the door) to anyone!’
\z 

Differently from \textit{ħədd} which can be used also in the affirmative, \textit{ħídda} is only used in negative contexts (after a negative marker or in elliptical answers) (minimal-unit expression, \citealt[227--228]{Haspelmath1997}), see \REF{ex:siw:53}.

\ea\label{ex:siw:53}
    \begin{xlist}[B: ]
    \exi{Q: }
        \gll  bə́ttin yə-ččá tɣáṛa?\\ 
              who \textsc{3sg.m}{}-eat{\textbackslash}\textsc{pfv} bread{\textbackslash}\textsc{pl.f}\\ 
        \glt ‘Who ate the bread?’
    \exi{A: }
        \gll  \textbf{ħídda}.\\ 
              \textbf{nobody{\textbackslash}\textsc{sg.m}} \\
        \glt ‘Nobody.’
    \end{xlist}
\z 


The elements \textit{ħədd} and \textit{ħídda} are both negative indefinites, as pointed out by \citealt{AuweraAlsenoy2016}, who consider this kind of pronouns as negative despite the fact that 1) the negation is conveyed also by other negative exponents, and 2) \textit{ħədd} can appear (and actually is more frequent) in affirmative contexts. Elliptic answers, like \REF{ex:siw:53}, have special features that make it possible to express a negative meaning, even if the pronoun is not \textit{per se} negative, and there is no other negative exponent (\citealt[476--478]{AuweraAlsenoy2016}). The only difference, in Siwi, is that in the case of elliptical answers, only the emphatic negative indefinite \textit{ħídda} can be used.

‘Nothing’ is formed through the negative marker \textit{la}{}- and the word \textit{šra} or, rarely *\textit{ħħila} ‘thing’: \textit{la-šra} / \textit{la-ħħila}. Contrary to \textit{šra}, *\textit{ħħila} is not attested anymore and only appears with the prefix \textit{la}{}- (or \textit{ula}{}-, see \sectref{sec:siw:3.4}). 

When \textit{šra} follows a negative clause (verbal or non-verbal), it may (\ref{ex:siw:54b} and \ref{ex:siw:54d}) or may not (\ref{ex:siw:54a} and \ref{ex:siw:54c}) be preceded by the negative marker \textit{la}{}-. In the former case, \textit{la}{}- translates as ‘not even’.



\ea\label{ex:siw:54}
    \ea\label{ex:siw:54a}
        \gll \textbf{lá-di} šra. \\ 
             \textbf{\textsc{neg-ex}} thing{\textbackslash}\textsc{sg.m} \\
        \glt ‘There is nothing.’
    \ex\label{ex:siw:54b}
        \gll \textbf{lá-di} \textbf{lá-šra}. \\
             \textbf{\textsc{neg}-\textsc{ex}} \textbf{\textsc{neg}-thing{\textbackslash}\textsc{sg.m}} \\
        \glt ‘There is not even a single thing.’
    \ex\label{ex:siw:54c}
        \gll \textbf{lá-xsi-x} šra. \\
             \textbf{\textsc{neg}-want{\textbackslash}\textsc{pfv}-\textsc{1sg}} thing{\textbackslash}\textsc{sg.m} \\
        \glt ‘I don’t want anything.’
    \ex\label{ex:siw:54d}
        \gll \textbf{lá-xsi-x} \textbf{lá-šra}. \\
             \textbf{\textsc{neg}-want{\textbackslash}\textsc{pfv}-\textsc{1sg}} \textbf{\textsc{neg}-thing{\textbackslash}\textsc{sg.m}} \\
        \glt ‘I don’t even want a single thing.’
    \z
\z

The same applies to ‘nowhere’ which uses the word \textit{ankan} ‘place’ after another negative exponent (in example \REF{ex:siw:55a}, the negative form of the verb). In parallel to what happens to \textit{šra}, if \textit{ankan} is preceded by \textit{la}{}-, it translates as ‘not even’. In example \REF{ex:siw:55b}, the allative preposition \textit{i} ‘to’ is also omitted.

\ea\label{ex:siw:55}
    \ea\label{ex:siw:55a}
        \gll \textbf{lá-ħħi-x} i ankán. \\ 
             \textbf{\textsc{neg}-go{\textbackslash}\textsc{pfv}-\textsc{1sg}} to place{\textbackslash}\textsc{sg.m} \\
        \glt ‘I did not go anywhere.’
    \ex\label{ex:siw:55b}
        \gll \textbf{lá-ħħi-x} \textbf{l-ankán}. \\
             \textbf{\textsc{neg}-go{\textbackslash}\textsc{pfv}-\textsc{1sg}} \textbf{\textsc{neg}-place{\textbackslash}\textsc{sg}.\textsc{m}} \\
        \glt ‘I did not even go to a single place.’
    \z
\z 


In general, the same distinction is found with nouns, as we will see in \sectref{sec:siw:4.4}.
The negative marker is nevertheless obligatory in elliptic answers, see \REF{ex:siw:56}.

\ea\label{ex:siw:56}
\begin{xlist}[A: ]
    \exi{Q:}
        \gll ta ɣúṛ-əm? \\
             what at-\textsc{2sg.f} \\
        \glt ‘What do you have?’ (i.e. ‘What is wrong with you?’)
    \exi{A:}
        \gll \textbf{lá-šra}! \\
             \textbf{\textsc{neg}-thing{\textbackslash}\textsc{sg.m}} \\
        \glt ‘Nothing!’
\end{xlist}
\z

‘Never’ is expressed through the Arabic loan \textit{abdan} or with the verb in its negative form and \textit{əxʷ}\textit{ṛa} ‘again’ and/or \textit{xaḷəṣ} ‘extremely’, see \REF{ex:siw:57}, which can also be used in affirmative contexts.

\ea\label{ex:siw:57}
    \gll \textbf{l\={a}-ẓṛ{}-a-šəm} xáḷəṣ əxʷṛá. \\
         \textbf{\textsc{neg}{\textbackslash}\textsc{irr}-see{\textbackslash}\textsc{aor-1sg-do\textbackslash2sg.f}} extremely again\\
    \glt ‘I will never see you again.’
\z

‘Never’ can also be expressed by the Arabic loanword \textit{maṛṛawaħda} ‘ever’ \citep[127]{Souag2013}. 

The Arabic loanword \textit{mazal} is used in both negative \REF{ex:siw:58} and affirmative contexts \REF{ex:siw:59}. This is cross-linguistically common (\citealt{VeselinovaDevos2021}; \citealt{VeselinovaAsplund2020}), as they are both continuative of an existent (‘still’, positive polarity) or non-existent (‘not yet’, negative polarity) situation (\citealt{vanBaar1997}; \citealt{Auwera1998}).

\ea\label{ex:siw:58}
    \gll n wən yə-nṭṛ-\'{i}n-a d n wən \textbf{maz\'{a}l} l-yə-nṭṛ{}-\'{i}n-a. \\
         of \textsc{rel} 3-bake{\textbackslash}\textsc{pfv-pl-prag} and of \textsc{rel}  \textbf{yet} \textsc{neg}-3-bake{\textbackslash}\textsc{pfv-pl-prag} \\
    \glt ‘those which were baked and those which were not baked yet’
\z 

\ea\label{ex:siw:59}
    \gll šəm \textbf{mazál} / tləččá n sláṭṭin. \\
         \textsc{idp.2sg.f} \textbf{yet} / girl{\textbackslash}\textsc{sg.f} of yesterday \\
    \glt ‘You are still a young girl (lit. a girl of yesterday).’
\z

The prefix \textit{la}{}- is also used to negate the quantifiers \textit{kom} ‘a lot’, \textit{dabb} ‘many’ and \textit{ħəbba} ‘a bit’. It must be noted that \textit{la-dabb} and \textit{la}{}-\textit{ħəbba} are present in my corpus only in elicitations.


\subsection{Privative marking}\label{sec:siw:3.3}

Privative marking uses the preposition \textit{bla} ‘without’, see \REF{ex:siw:60}. It is widely considered to be Arabic in origin: Classical Arabic \textit{bi-lā} = Egyptian Arabic \textit{bala} \citep[170]{Souag2013}. 

\ea\label{ex:siw:60}
    \gll tə-ṃṃə́l ə́ngr-ax ə́ǧǧət n ssənt \textbf{bla} tarwá.\\
         \textsc{3sg.f-}say{\textbackslash}\textsc{pfv} live{\textbackslash}\textsc{pfv-1sg} one{\textbackslash}\textsc{f} of year{\textbackslash}\textsc{sg.f} \textbf{without} give\_birth{\textbackslash}\textsc{vn} \\
    \glt ‘She said: “I stayed one year without giving birth."’
\z

\subsection{Other negative constructions/expressions} \label{sec:siw:3.4}

Two expressions use the negative marker \textit{ula} instead of \textit{la}, when negating a noun: \textit{af ula-ħħila} ‘for no reason’ \citep[210]{Souag2013} and \textit{af ula-šra} ‘for nothing’ \citep{Schiattarella2015}, as in \REF{ex:siw:61}.

\ea\label{ex:siw:61}
\begin{xlist}[A: ]
    \exi{Q:}
        \gll tádi fáṭma t-gə́lləs?\\
             why fatma \textsc{3sg.f}{}-cry{\textbackslash}\textsc{ipfv}?  \\
        \glt ‘Why is Fatma crying?’
    \exi{A:}
        \gll  af \textbf{ulá-šra}. \\
             on \textbf{\textsc{neg}-thing{\textbackslash}\textsc{sg.m}} \\
        \glt ‘For no reason.’
\end{xlist}
\z

\section{Other aspects of negation}\label{sec:siw:4}

\subsection{The scope of negation}\label{sec:siw:4.1}

When the scope of negation has to be narrowed to a specific constituent, the negative marker \textit{ačči/qačči} is used, for both contrastive focus or metalinguistic negation. The latter will be dealt with in \sectref{sec:siw:4.6.1}. This negative marker can be placed before verbal \REF{ex:siw:62}, nominal \REF{ex:siw:63}, pronominal, adverbial or adjectival \REF{ex:siw:64} predicates: in all such cases, its function is to oppose the presuppositions that may be held by the speaker and/or the addressee.

In example \REF{ex:siw:62}, the speaker talks about the role of old women, who also act as teachers to small children.

\ea\label{ex:siw:62}
    \gll \textbf{qáčči} ət-təṃṃ-ásən tan{\fdot}ás w xḷaṣ.\\
         \textbf{\textsc{neg}} \textsc{3sg.f}{}-say{\textbackslash}\textsc{ipfv-dat\textbackslash3pl} tale{\textbackslash}\textsc{pl.f} and stop\\
    \glt ‘It was not only about telling them stories (she also transmitted a pedagogical message).’
\z

The same notion of contrast can be found when this negative marker precedes a noun or an adjective, see \REF{ex:siw:63} and \REF{ex:siw:64}.

\ea\label{ex:siw:63}
    \gll w-om əlfaṭṭ\'{ə}t / \textbf{ačči} ənnħás. \\
         \textsc{dem-2sg.f} silver{\textbackslash}\textsc{sg.f} / \textsc{\textbf{neg}} copper{\textbackslash}\textsc{sg.m} \\
    \glt ‘This is silver, not copper.’
\z

\ea\label{ex:siw:64}
    \gll tálti tat-óm \textbf{áčči} tašmált. \\
         woman{\textbackslash}\textsc{sg.f} \textsc{dem{\textbackslash}f-2sg.f} \textbf{\textsc{neg}} bad{\textbackslash}\textsc{sg.f}\\
    \glt ‘This woman is not bad.’ (Replying to someone who thinks she is)
\z

Example \REF{ex:siw:65} is taken from a recording in which the speaker is talking about the role of women in Siwa and claims that Siwi women do not need to work, according to their customs. At the same time, she comments that it would be better to reconsider this assumption, given that Siwi women actually do work, even if their occupation concerns house chores and childcare.

\ea\label{ex:siw:65}
    \gll t\'{a}lti t-xə́ddəm / \textbf{qáčči} l\'{a}-t-xəddəm / t-xə́ddəm g ágbən. \\
         woman{\textbackslash}\textsc{sg.f} \textsc{3sg.f}-work{\textbackslash}\textsc{ipfv} / \textbf{\textsc{neg}} \textsc{neg-3sg.f}-work{\textbackslash}\textsc{ipfv} / \textsc{3sg.f}-work{\textbackslash}\textsc{ipfv} in house{\textbackslash}\textsc{sg.m}\\
    \glt ‘The woman works, it is not that she does not work, (because) she works at home (taking care of the house, of the children...).’
\z

\subsection{Negative polarity}\label{sec:siw:4.2}

The indefinite pronoun \textit{ħədd} ‘one, person’, see \sectref{sec:siw:3.2}, can be identified as a negative polarity item as it is found mostly in negative contexts or in conditional clauses introduced by \textit{kan} or \textit{mak}(-\textit{ənni}) ‘if’, as in \REF{ex:siw:66}. The variant \textit{ħidda} is only used in negative contexts. 


\ea\label{ex:siw:66}
    \gll kan ħədd  yə-xs\'{a} aħáṭṭi mági / g-yə́-ħħaṭṭ.\\
         if person{\textbackslash}\textsc{sg.m} \textsc{3sg.m}{}-want{\textbackslash}\textsc{pfv} put{\textbackslash}\textsc{vn} stock{\textbackslash}\textsc{sg.m} / \textsc{irr-3sg.m}{}-put{\textbackslash}\textsc{aor} \\
    \glt ‘If anyone wants to put in a stock cube, he will.’
\z

For the negative polarity item \textit{qaṭṭ}/\textit{qəṭṭ} ‘ever, at all’ of Arabic origin \citep{Souag2009}, see \sectref{sec:siw:4.4}.


\subsection{Marking of NPs in the scope of negation}\label{sec:siw:4.3}

No special marking of NPs in the scope of negation is attested in my data.


\subsection{Reinforcing negation}\label{sec:siw:4.4}

The verb \textit{ʕṃaṛ} ‘do’ has several grammatical functions in Siwi, among which the possibility of encoding emphasis on what follows: a verb, an existential or a possessive predication. In this case, the verb is a third person singular masculine imperfective. 

In order to reinforce the existential or possessive predication, the negative form of this verb is used, see \REF{ex:siw:67} and \REF{ex:siw:68}.

\ea\label{ex:siw:67}
    \gll kan fəkkr-áṭ-a / \textbf{l-i-ʕə́ṃṃaṛ} di ššɣul g ə́ṣṣef. \\
         if remember{\textbackslash}\textsc{pfv-2sg-prag} / \textbf{\textsc{neg-3sg.m-}do{\textbackslash}\textsc{ipfv}} \textsc{ex} job{\textbackslash}\textsc{sg.m} in summer{\textbackslash}\textsc{sg.m} \\
    \glt ‘If you remember, there is no work (at all) during summer.’
\z

The elicited example in \REF{ex:siw:68} shows the difference between the reinforced possessive predication and the simple negative possessive predication. Indeed, the speaker points out that greengrocers do not sell olive oil and that regular grocery shops do not sell meat. The difference between the two is that in the first case, the impossibility of buying olive oil is categorical, while in the second case, one might be able to buy some sort of meat, such as minced meat (though meat is regularly sold at the butcher's shop).

\ea\label{ex:siw:68}
    \gll əddəkkan n ləxḍaṛ \textbf{l-i-ʕəṃṃaṛ} \textbf{ɣuṛ-əs} əddhan / kəllamənni əddəkkan \textbf{la-ɣuṛ-əs} aksum. \\
         shop{\textbackslash}\textsc{sg.m} of fruit\_and\_vegetable{\textbackslash}\textsc{pl.m} \textbf{\textsc{neg-3sg.m}-do{\textbackslash}\textsc{ipfv}} \textbf{at-\textsc{3sg}} oil{\textbackslash}\textsc{sg.m} / on\_the\_contrary shop{\textbackslash}\textsc{sg.m} \textbf{\textsc{neg-}at\textsc{{}-3sg}} meat{\textbackslash}\textsc{sg.m} \\
    \glt ‘The greengrocer definitely does not have oil, while the grocery shop does not have meat.’
\z

The negative polarity item \textit{qaṭṭ}/\textit{qəṭṭ} ‘ever, at all’ is used to reinforce negative statements, see \REF{ex:siw:69}.

\ea\label{ex:siw:69}
    \gll l-yə-xsa asiwəl did-i \textbf{qaṭṭ}. \\
         \textsc{neg-3sg.m}-want{\textbackslash}\textsc{pfv} speak{\textbackslash}\textsc{vn} with-\textsc{1sg} \textbf{at\_all} \\
    \glt ‘He does not want to talk to me at all.’
\z

As seen in \sectref{sec:siw:3.2}, the indefinite pronoun \textit{ħidda} is used instead of \textit{ħədd} as an emphatic negative pronoun and differently from the latter, it can only be used in negative contexts. Example \REF{ex:siw:52} is repeated in \REF{ex:siw:70} for practical reasons.

\ea\label{ex:siw:70}
    \gll \textbf{l\={a}-fətk-ʕ-as} i \textbf{ħídda}! \\
         \textbf{\textsc{neg{\textbackslash}irr}-open{\textbackslash}\textsc{aor-1sg-dat\textbackslash3sg}} to \textbf{nobody{\textbackslash}\textsc{sg.m}}  \\
    \glt ‘I will not open (the door) to anyone!’
\z

A noun can be preceded by \textit{la}, mainly with negative existential predications (as in \REF{ex:siw:71}), but also after a negative verb. In this case, \textit{la} functions as a scalar particle ‘not even’. Sometimes, \textit{ħətta} ‘even’ from Arabic, also used in affirmative contexts (\citealt[214, 220]{Souag2013}) is used instead.

\ea\label{ex:siw:71}
    \gll \textbf{lá-di} \textbf{la} əlmə́ṣṛəb / ánni g-yə́-ħħ{}-ən \'{i}  zdat. \\
          \textbf{\textsc{neg-ex}} \textsc{\textbf{neg}} road{\textbackslash}\textsc{sg.f} / \textsc{comp} \textsc{irr-3}{}-go{\textbackslash}\textsc{aor-pl} to in\_front\_of \\
    \glt ‘There was not even a road in order for them to go in front (of it).’
\z

If more than one noun is present, \textit{la} is used for negative coordination, see \sectref{sec:siw:4.5}. 


\subsection{Negation, coordination and complex clauses}\label{sec:siw:4.5}

Negative verb coordination uses the morpheme \textit{ul(a)}{}- \citep[54]{Schiattarella2017}, see example \REF{ex:siw:72}.

\ea\label{ex:siw:72}
    \gll ħħi-x i áʕṛuṣ / \textbf{ul-ə́čči-x} d \textbf{ul-ə́swi-x}. \\
         go{\textbackslash}\textsc{pfv}{}-1\textsc{sg} to wedding{\textbackslash}\textsc{sg.m} / \textbf{\textsc{neg}-eat{\textbackslash}\textsc{pfv}-\textsc{1sg}} and \textbf{\textsc{neg}-drink{\textbackslash}\textsc{pfv}-\textsc{1sg}} \\
    \glt ‘I went to the wedding, I did not eat, nor did I drink.’
\z

Negative nominal coordination (as in example \ref{ex:siw:73}) uses \textit{la}.


\ea\label{ex:siw:73}
    \gll yə-ħlay-anáx-a / \textbf{lá-di} \textbf{la} aṭṭáw / \textbf{la} ənternét / \textbf{la} əddbáš n w-i-(y)ók. \\ 
         \textsc{3sg.m}-be\_happy.\textsc{pfv}-\textsc{dat\textbackslash1pl-prag} / \textbf{\textsc{neg-ex}} \textbf{\textsc{neg}} light\textbackslash\textsc{sg.m} / \textbf{\textsc{neg}} internet\textbackslash\textsc{sg}.\textsc{m} /  \textbf{\textsc{neg}} thing\textbackslash\textsc{sg}.\textsc{m} of \textsc{dem-pl-2sg.m} \\
    \glt ‘We were happy (because) there wasn’t any electricity, or internet, or any other such things.’
\z 


While \textit{la} is the most common marker of negative coordination \REF{ex:siw:74a}, \textit{ula} is not excluded. It can, indeed, precede the second nominal item of the coordination \REF{ex:siw:74b} or both of them \REF{ex:siw:74c}.
All three sentences in \REF{ex:siw:74} have the same reading.

\ea\label{ex:siw:74}
    \ea\label{ex:siw:74a}
        \gll \textbf{la-təčč-ax} \textbf{la} aksum \textbf{la} ǧǧəbnət. \\ 
             \textbf{\textsc{neg}-eat{\textbackslash}\textsc{ipfv-1sg}} \textbf{\textsc{neg}} meat{\textbackslash}\textsc{sg.m} \textbf{\textsc{neg}} cheese{\textbackslash}\textsc{sg.f} \\
    \ex\label{ex:siw:74b}
        \gll \textbf{la-təčč-ax} \textbf{la} aksum \textbf{ula} ǧǧəbnət. \\
             \textbf{\textsc{neg}-eat{\textbackslash}\textsc{ipfv-1sg}} \textbf{\textsc{neg}} meat{\textbackslash}\textsc{sg.m} \textbf{\textsc{neg}} cheese{\textbackslash}\textsc{sg.f} \\
    \ex\label{ex:siw:74c}
        \gll \textbf{la-təčč-ax} \textbf{ula} aksum \textbf{ula} ǧǧəbnət. \\
             \textbf{\textsc{neg}-eat{\textbackslash}\textsc{ipfv-1sg}} \textbf{\textsc{neg}} meat{\textbackslash}\textsc{sg.m} \textbf{\textsc{neg}} cheese{\textbackslash}\textsc{sg.f} \\
        \glt ‘I eat neither meat nor cheese.’
    \z
\z 

Alternatively, \textit{(la)... wala} (\textit{wa} ‘and’ in Arabic + the negative marker \textit{la)}, can be used. More details about the distribution of \textit{la} and \textit{ula}, not only in the context of negative coordination, in both Berber and Arabic is given in \sectref{sec:siw:4.6.3}.

\subsection{Miscellaneous aspects of negation}\label{sec:siw:4.6}
\subsubsection{Metalinguistic negation}\label{sec:siw:4.6.1}

Metalinguistic negation uses the negative marker \textit{ačči/qačči} to correct a wrong statement or the way it is expressed by the speaker or by his/her addressee.

In example \REF{ex:siw:75} the speaker clarifies what his addressee has just said, objecting to how he has presented the situation.

\ea\label{ex:siw:75}
\begin{xlist}[A: ]
    \exi{A:}
        \gll ámra dáwa kan əṃṃí-ṭ-as i ħədd asíli agəǧǧí n sɣáṛən / l-yə́-qḍaṛ.  \\
             now \textsc{dem{\textbackslash}sg.m} if say{\textbackslash}\textsc{pfv-2sg-dat{\textbackslash}3sg} to person{\textbackslash}\textsc{sg.m} load{\textbackslash}\textsc{vn} bundle{\textbackslash}\textsc{sg.m} of wood{\textbackslash}\textsc{pl.m} / \textsc{neg}-3\textsc{sg.m}-be\_able{\textbackslash}\textsc{pfv} \\
        \glt ‘Nowadays, if you ask somebody (a young boy) to load a bundle of wood, he is not able (to do it).’
    \exi{B:}
        \gll \textbf{áčči} l-yə́-qḍaṛ / l-yə́-ssən mámək. \\
             \textbf{not} \textsc{neg-3sg.m}{}-be\_able{\textbackslash}\textsc{pfv} / \textsc{neg}{}-3\textsc{sg.m}{}-know{\textbackslash}\textsc{pfv} how\\
        \glt ‘It is not that he can’t, he doesn’t know how.’
\end{xlist}
\z

\subsubsection{Non-negative uses of negatives}\label{sec:siw:4.6.2}

The negative marker \textit{ačči}/\textit{qačči} found in most stative predications and in contrastive/metalinguistic negation occurs in one context where it takes a non-negative meaning. It can, in fact, be used when the speaker wants to confirm or agree with what the addressee is saying. It roughly translates ‘Indeed!’, ‘Right!’, ‘Exactly!’.
% \mm{@LSP: for some reason this paragraphis not right-justified.}

With this function, it is always followed by the sentence-final discourse particle \textit{ya} ‘you know’, see example \REF{ex:siw:76}.

\ea\label{ex:siw:76}
\begin{xlist}[A: ]
    \exi{A:}
        \gll ásfa dáw-om ləħm\'{u} kom!  \\
             today \textsc{dem{\textbackslash}sg.m-2sg.f} heat{\textbackslash}\textsc{sg.m} a\_lot \\
        \glt ‘It is hot today!’
    \exi{B:}
        \gll \textbf{áčči} \textbf{ya}! \\
             \textbf{\textsc{neg}} \textbf{you\_know} \\
        \glt ‘Indeed!’
\end{xlist}
\z

\subsubsection{Diachronic and comparative notes}\label{sec:siw:4.6.3}

\citet[265]{Laoust1931} argues that the negative marker \textit{la}{}- comes from Arabic.
In several Arabic varieties, \textit{la} is the negative pro-sentential marker, but can be used also for the prohibitive (with or without the post-verbal negation, depending on the variety), for oaths, and for negative coordination with nouns or verbs (\citealt[88--90]{Caubet1996}, \citealt{TaineCheikh2000}). Sometimes, the second term of the coordination is \textit{u la} or \textit{wa} \textit{la}, where \textit{u}/\textit{wa} is the Arabic conjunction ‘and’.

\citet{Souag2009,Souag2013} defends the hypothesis of the Arabic origin of the marker \textit{la}, but he further adds that it is possible to imagine a double etymology (Berber and Arabic), at least in the nominal context. The author motivates this hypothesis showing that in one fixed expression \textit{af ulaħħila} (\textit{ula-ħħila} ‘no-thing’, \textit{af} \textit{ulaħħila} ‘for no reason’), \textit{ula}, instead of \textit{la}, is used to negate a noun. \textit{ula} ‘not even’ is used to negate nouns in other Berber languages too, like the variety spoken in Mzab (Algeria) and in Tamashek, the latter being very little influenced by Arabic \citep[209--210]{Souag2013}.

The marker \textit{ula} is also found in the Berber variety of Wargla (Algeria), in affirmative contexts, for ‘even’: \textit{ula šək} ‘even you’ (\citealt[20]{Biarnay1908}, transcription and translation mine). Other varieties (Kabyle, Algeria, At Jennad) make a difference between ‘even’ \textit{la} in affirmative contexts and ‘not even’ \textit{ula} in negative contexts, with some exceptions of \textit{la} being used in the negative too (after a negative verb) or \textit{ula} in affirmative contexts (mainly in hypothetical clauses). \textit{la} is also a coordinative marker in Kabyle, used in both affirmative (‘either... or’) or negative (‘neither... nor’) contexts (the latter only after a negative exponent \citep[403]{Brugnatelli2010}). 

In other varieties, both \textit{la} and \textit{ula} can be used in affirmative contexts, with the same meaning ‘even’ (as in the Beni Snous (Algeria) variety, \citealt[217]{Destaing1914}). Moreover, in Kabyle, \textit{ula} can occur in the composition of the existential negative predicative marker \textit{ulaš} ‘there is not’, as well as other compounds, i.e. \textit{ulaħədd} ‘there is no one’ (\citealt[401]{Brugnatelli2010}; \citealt[52]{Mettouchi2021}). 

\tabref{tab:siw:1} summarizes the use of \textit{la} and \textit{ula} in the nominal domain in some Berber varieties. The list is not exhaustive.


\begin{table}\small
\begin{tabularx}{\textwidth}{llXcc}
\lsptoprule
& Function & Language & \textsc{neg} & \textsc{aff} \\
\midrule
{\itshape la (d)} & scalar focus particle & Beni Snous, Kabyle (At Jennad) &  &  +\\
{\itshape la... la} & coordinative marker & Kabyle &  + & +\\
\midrule
\multirow{2}{*}{\itshape ula (d)} & \multirow{2}{*}{scalar focus particle} & Kabyle (At Jennad), Mzab, Tamashek & + & \\
&  & Wargla, Beni Snous &  & +\\
\lspbottomrule
\end{tabularx}
\caption{Functions of \textit{la} and \textit{ula} in the nominal domain}
\label{tab:siw:1}
\end{table}

\newpage

As far as the verb is concerned, \textit{ula} can be used as a postverbal negation, in Eastern Kabyle (as reported by \citealt[403--404]{Brugnatelli2010}, citing \citealt{Rahmani1939}, \citealt{Rabhi1992}) and in some Western varieties of Tasaħlit (Algeria, among other markers, \citealt[28]{Garaoun2019}), for example. It can also be a preverbal negative marker, see Western Rif varieties (Morocco, \citealt[234--235]{Lafkioui2007}, together with \textit{la}{}-), Wargla (Algeria, \citealt[126--132]{Biarnay1908}) and the Western varieties of Tasaħlit \citep[28]{Garaoun2019}, as shown in \tabref{tab:siw:2}.

\begin{table}
\begin{tabularx}{.95\textwidth}{lll}
    \lsptoprule
    preverbal negation & \textit{la{}- -š}, \textit{la{}- -šay} & Western Rif \\
    & {\itshape ula} & Wargla, Western Tasaħlit \\
    & {\itshape ula{}- -š} & Western Rif \\
    \midrule
    postverbal negation & {\itshape ula} & Eastern Kabyle, Western Tasaħlit \\
    \lspbottomrule
\end{tabularx}
\caption{Functions of \textit{la} and \textit{ula} in the verbal domain}
\label{tab:siw:2}
\end{table}


Brugnatelli recognizes the interference that could arise between Arabic and Berber uses of this marker, also because \textit{la}…\textit{la} is used as an affirmative or negative coordinative marker (‘either… or’; ‘neither… nor’) in some varieties of Berber, as mentioned before, while \textit{la}… \textit{la}… \textit{wala} is only used for Arabic negative coordination \citep[403]{Brugnatelli2010}. 

He then proposes to analyse \textit{ula} as \textit{u} (negation element) and \textit{la} (fixed form of the verb ‘be’ that, alone, has the meaning of ‘also, even’ and is used in affirmative contexts in some varieties, as we have mentioned before) \citep[404]{Brugnatelli2010}.

Our corpus shows that \textit{ula} is used for negative verb coordination (‘neither... nor’). In negative nominal coordination both \textit{la} and \textit{ula} are only possible if they follow a negative verb or a negative non-verbal predicator. In the nominal domain, \textit{ula} is used only for negative coordination and in some fixed expressions, while \textit{la} is also used as a scalar particle. Contrary to other Berber varieties, \textit{la} (or \textit{ula}) cannot be used in the affirmative domain, probably because of the interference of Arabic, where \textit{la} is only used in negative contexts.

As suggested by Souag, it is likely that Berber and Arabic have both contributed to the establishment of the current system of Siwi negative markers, where Berber \textit{ula} is restricted to some specific contexts and fixed expressions and Arabic \textit{la} is only used in negative contexts.

The negative marker \textit{ačči/qačči} is also considered to have an Arabic origin by Souag. More precisely, he analyses it as \textit{qaṭṭ} ‘not at all, never’ + \textit{šay} ‘thing’ \citep{Souag2009}.

Just to mention some aspects relevant to the comparison with some Berber varieties, we can notice that Siwi lacks specific negative stems such as the negative perfective and the negative imperfective found elsewhere. The negative perfective is indeed attested in many Berber languages, except for Siwi and for some varieties of Tashelhiyt in Morocco, while the negative imperfective is no longer found in Kabyle, Siwi, Tashelhiyt, some varieties in Central Morocco, and others \citep[41]{Kossmann2012}. Another feature of other Berber languages, which is lacking in Siwi, is the change of position of clitics when there is a negative marker: they are shifted from a post-verbal to a pre-verbal position, right after the negative marker. This change of position is not restricted to negation, but also occurs in other environments, such as relative clauses, after the aorist preverbal particle and after some conjunctions.
In the majority of the Berber languages, the negative markers are \textit{wər}, \textit{wəl} and \textit{ul}. Some varieties (Kabyle, Aures, Nefusi, among others) have a post-verbal negation marker in some contexts (as in Kabyle, \citealt{Mettouchi2009,Mettouchi2021}).

The asymmetries between affirmative and negative domains, in terms of aspect, vary from one language to the other (for an overview of negation in Berber, see \citealt{Chaker1996}; \citealt{Mettouchi2009,Mettouchi2010,Mettouchi2012,Mettouchi2021}).

Subject relative clauses in most Berber languages use a special verbal form, traditionally called participle in Berber studies, characterized by a special series of prefixes and/or suffixes. Depending on the varieties, this form can be a unique form or agree in gender, number and person with the head noun. Analogously, in some varieties, the participle undergoes morphological changes when the negative marker is added (apart from the addition of the negative marker) (see \citealt{Mettouchi2017}). In Siwi no such form is attested, as mentioned in \sectref{sec:siw:2.4}.

In Berber, there is also a high degree of variation as far as imperative and prohibitive are concerned. The use of the aorist in the former and the imperfective in the latter is also attested in other Berber languages such as Kabyle, Tarifit and Zenaga (as is shown in the survey by \citealt[68--71]{Mettouchi2021}; see also \citealt{Lafkioui1996}; \citealt{TaineCheikh2011}). 

Many features present in Berber languages are absent in Siwi because of the Arabic influence, which is particularly evident in this domain. For example, the use of the indefinite pronouns or the absence of dedicated negative stems were probably lost due to the strong influence of Arabic \citep[212]{Souag2013}. 

\newpage
\section{Summary}\label{sec:siw:5}


This study has provided a synchronic analysis of negation in Siwi, with additional notes on diachronic and comparative aspects. The standard negation system is characterized by a relatively symmetric structure, with the primary distinction between affirmative and negative contexts being the presence of the preverbal marker \textit{la}- in the latter. The same negative marker is used for non-declarative clauses and certain stative predications, such as attributional, possessive, and existential ones. Other nonverbal predications employ a different negative marker, \textit{ačči}/\textit{qačči}, which is also used for contrastive/metalinguistic negation. There appears to be no significant asymmetry between negation in main and non-main clauses. Negative replies involve the use of \textit{ula}. The presence of both \textit{la} and \textit{ula} in verbal and nominal contexts is noteworthy, especially considering their implications in terms of historical development and contact with Arabic.

\tabref{tab:siw:3} summarizes the negation strategies found in Siwi and their functions.

\begin{table}[t]
    \centering\small 
    \begin{tabularx}{\textwidth}{lXl}
\lsptoprule
Marker & Function & Section \\ 
\midrule
\textit{la} 
& Negative declarative clauses & \sectref{sec:siw:2.1} \\
& Negative non-declarative clauses &  \sectref{sec:siw:2.2}  \\
& Negative attributive, existential and possessive predication & \sectref{sec:siw:2.3} \\
& Negative non-main clauses & \sectref{sec:siw:2.4} \\
% & Preverbal particle in negative declarative clauses,non-declarative clauses, in main and non-main clauses & \sectref{sec:siw:2.1}; \sectref{sec:siw:2.2}; \sectref{sec:siw:2.4} \\
&Negative pro-sentential marker, only when repeated &  \sectref{sec:siw:3.1}  \\
&Negative indefinites and quantifiers &  \sectref{sec:siw:3.2}  \\
&Before nouns, for emphatic negation &  \sectref{sec:siw:4.4}  \\
&Negative coordination (\textit{la… la}) & \sectref{sec:siw:4.5}  \\

\textit{ačči} / \textit{qačči}%: negative particle 
&Negative equative, proper attribution and locative predication & \sectref{sec:siw:2.3} \\
&Contrastive focus &\sectref{sec:siw:4.1}  \\
&Metalinguistic negation & \sectref{sec:siw:4.6.1} \\

\textit{ačči ya} 
& ‘indeed’ &  \sectref{sec:siw:4.6.2}  \\

\textit{nola} 
& Question tag &  \sectref{sec:siw:2.2}  \\

\textit{lubah}  
& ‘probably not’ &  \sectref{sec:siw:2.5}  \\

\textit{ula} 
& Pro-sentential marker &  \sectref{sec:siw:3.1} \\
& Negative coordination (\textit{ula… ula}) &  \sectref{sec:siw:4.5}  \\

\textit{mazal} 
& ‘not yet’ &  \sectref{sec:siw:3.2}  \\

\textit{abdan}, \textit{maṛṛawaħda} 
& ‘never’ &  \sectref{sec:siw:3.2}  \\

\textit{ħədd}, \textit{ħidda} 
& Negative polarity item, ‘one, person’ &  \sectref{sec:siw:3.2}; \ref{sec:siw:4.2}; \ref{sec:siw:4.4}  \\

\textit{bla} 
&  ‘without’ &  \sectref{sec:siw:3.3} \\

\textit{af ula-ħħila}, \textit{af ula-šra} 
& ‘for no reason’ &  \sectref{sec:siw:3.4}  \\

\textit{qaṭṭ}/\textit{qəṭṭ} 
& Negative polarity item, ‘at all’ &  \sectref{sec:siw:4.4}  \\

\textit{liʕəṃṃaṛ} 
& Negative emphatic existential and possessive predication &  \sectref{sec:siw:4.4} \\ 
\lspbottomrule
 \end{tabularx}
    \caption{Summary of negation in Siwi}
    \label{tab:siw:3}
\end{table}



\section*{Acknowledgments}

All examples were transcribed and translated into English with the help of Siwi speakers, who I wish to thank here. I am also grateful for the comments and suggestions I received from the editors of the volume, the anonymous reviewers and the colleagues who read earlier versions of this paper. All errors are my own.

% \clearpage

\section*{Abbreviations}

\begin{tabularx}{.5\textwidth}{ll}
1 & first person \\
2 & second person \\
3 & third person \\
\scshape adj  &  adjective \\
\scshape aor  &  aorist \\
\scshape comp  &  complementizer \\
\scshape dat &   dative \\
% \scshape dem  &  demonstrative \\
% \scshape do   & direct object \\
% \scshape ex &   existential \\
% \scshape f   & feminine \\
% \scshape idp  &  independent pronoun \\
% \scshape imp  &  imperative \\
% \scshape ipfv &   imperfective \\
% \end{tabularx}
% \begin{tabularx}{.54\textwidth}{lQ}% a bit larger to save lines
\scshape irr  &  irrealis \\
\scshape juss &   jussive \\
\scshape m   & masculine \\
\scshape pfv &   perfective \\
\scshape pl  &  plural \\
\scshape poss &   possessive \\
\scshape prag &   pragmatic relevance marker \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{ll}
% \scshape neg   & negative \\
% \scshape rel &   relative marker \\
% \scshape sg &   singular \\
% \scshape voc  &  vocative \\
% \scshape vn &   verbal noun \\
% /  &  end of a minor prosodic unit \\
% // &   end of a major prosodic unit \\
% 1 & first person \\
% 2 & second person \\
% 3 & third person \\
% \scshape adj  &  adjective \\
% \scshape aor  &  aorist \\
% \scshape comp  &  complementizer \\
% \scshape dat &   dative \\
\scshape dem  &  demonstrative \\
\scshape do   & direct object \\
\scshape ex &   existential \\
\scshape f   & feminine \\
\scshape idp  &  independent pronoun \\
\scshape imp  &  imperative \\
\scshape ipfv &   imperfective \\
% \end{tabularx}
% \begin{tabularx}{.54\textwidth}{lQ}% a bit larger to save lines
% \scshape irr  &  irrealis \\
% \scshape juss &   jussive \\
% \scshape m   & masculine \\
% \scshape pfv &   perfective \\
% \scshape pl  &  plural \\
% \scshape poss &   possessive \\
% \scshape prag &   pragmatic relevance marker \\
\scshape neg   & negative \\
\scshape rel &   relative marker \\
\scshape sg &   singular \\
\scshape voc  &  vocative \\
\scshape vn &   verbal noun \\
/  &  end of a minor prosodic unit \\
// &   end of a major prosodic unit \\
\end{tabularx}

\newpage
\sloppy\printbibliography[heading=subbibliography,notkeyword=this]

\end{document} 
