\documentclass[output=paper]{langscibook}
\ChapterDOI{10.5281/zenodo.18234065}
\title{Negation in Sundanese}
\author{Marielle Butters\affiliation{University of Colorado at Boulder}}
\abstract{The Malayo-Polynesian language Sundanese is characterized by a rich system for the encoding of negation. While the position and number of negators in clausal negation follow the cross-linguistic majority, some elements of Sundanese negation are less common, including the distinction made in negators used in verbal versus nominal predicates and the productive use of negative lexicalizations in certain areas of cognition and volition. Negation in Sundanese also includes distinct negators in certain tense and aspectual categories; alternative negation strategies to achieve greater levels of politeness or indirectness; prohibitives with uses beyond imperative mood; and polarity phenomena relying on reduplication and scalar additive particles. This chapter comprises a detailed description of Sundanese negation following use of \textcitetv{chapters/questionnaire} in a field environment in West Java.
\keywords{Malayo-Polynesian, negation, negative lexicalizations, scalar semantics, existential predication}
}
\IfFileExists{../localcommands.tex}{
 \addbibresource{../localbibliography.bib}
 \input{../localpackages}
 \input{../localcommands} 
 \input{../localhyphenation} 
 \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle

\section{The language}\label{sec:ngo:1}

Sundanese, known by its speakers as \textit{Basa Sunda,} (lit. ‘Language of Sunda’, ISO: \href{https://iso639-3.sil.org/code/sun}{sun}, Glottocode: \href{https://glottolog.org/resource/languoid/id/sund1252}{sund1252}), is a Malayo-Polynesian, Austronesian language spoken primarily in West Java and Banten, Indonesia. Centuries-old migration and more recent governmental transmigration have also resulted in pockets of Sundanese speakers elsewhere in Indonesia, such as southern Sumatra, West Kalimantan, and Riau. With around 34 million speakers, Sundanese is the third most widely spoken language in Indonesia, after \textit{Bahasa Indonesia} (Indonesian) and Javanese. Though its speakers account for approximately 15\% of the population, Sundanese, like other \textit{bahasa daerah} (regional languages), is in decline as speakers increasingly shift to greater use of Indonesian, the official and national language of the country. In contemporary times, monolingual Sundanese speakers are found almost exclusively in rural areas among older populations. 

Like other languages of Indonesia, Sundanese has been considerably influenced by languages outside the Austronesian family. External influence on the Austronesian world began 2,000 years ago and occurred in the following order: Indian (Sanskrit), Chinese (Hokkien, Hakka, Cantonese, etc.), Islamic (Arabic), and European (Portuguese, Spanish, Dutch, English) \citep{Blust2013}. The fall of the Sundanese Pajajaran Kingdom in the 16\textsuperscript{th} century resulted in the loss of a Sundanese cultural center as power became concentrated in the \textit{kabupaten} (the seat of the local leader) with no strong single point of influence. Today the center of Sundanese culture is widely viewed to be Priangan, in the highlands around the city of Bandung in West Java province. While there are numerous dialects of Sundanese, it is the Priangan dialect that is taken to be the standard and taught from elementary to high school in West Java and Banten provinces. 

Sundanese is characterized by basic SVO word order, mostly isolating and agglutinating morphology, nasal harmony, and reduplication. Also notable is the elaborate coding of speech levels in most dialects. \citeauthor{Lezer1931} (\citeyear{Lezer1931}, cited in \citealt{Wessing1974}) identifies four basic levels: i) \textit{lemes pisan} ‘very polite’, ii) \textit{lemes} ‘polite/deferential’, iii) \textit{kasar} ‘ordinary/ colloquial’, iv) \textit{kasar pisan} ‘vulgar’. These speech levels, which consist of nearly fully distinct lexicons at each level, have long been recognized as a borrowing from Javanese. This is supported by the fact that speech levels are absent in Baduy, the region of the Sundanese world considered to be the most archaic (\citealt{Gonda1948} and oral tradition). 

One particularly rich area of Sundanese grammar is negation, the focus of this chapter. The structure of the chapter mirrors that of the questionnaire provided by the editors \parencitetv{chapters/questionnaire}.
A large portion of the data presented herein was collected by the author from June 2019 to August 2019 in the village of Cipta Gelar, one of the larger villages of Kasepuhan Banten Kidul, a community in the southern region of Mount Halimun Salak National Park in West Java. 
The work is supplemented by data gathered by the author in Bogor, West Java from January 2019 to June 2019, and in Bandung in February 2020. The data includes utterances from different speech levels, though the majority are from the \textit{lemes} or high speech level. The author follows Sundanese orthographic conventions throughout the chapter wherein ‘é’ is used for the mid front vowel /ɛ/\textsf{} and ‘e’ represents mid central /ə/. Some of the glossed examples are also included in the analysis presented in \citet{Butters2021}. 

For a full table of the negators and negative constructions discussed throughout the chapter, the reader can refer to \tabref{tab:sun:2} at the end of the chapter. 

\section{Clausal negation}\label{sec:sun:2}

\subsection{Standard negation}\label{sec:sun:2.1}

Standard negation (\citealt{Payne1985}, \citealt{Miestamo2005}) refers to the basic ways languages have for negating declarative main clauses with a verbal predicate. The standard negator in Sundanese is the uninflected particle \textit{henteu,} frequently in its clipped form \textit{teu}. This negator occurs directly before the verbal predicate, the most common position for negation in Austronesian languages \citep[61]{VossenAuwera2014} and cross-linguistically (\citealt{Dahl1979}, \citealt{Dryer2013}, \citealt[411]{AuweraDumon2015}). An affirmative utterance is presented in \REF{ex:sun:1a} along with its negative counterpart in \REF{ex:sun:1b}. 

\ea\label{ex:sun:1}
    \ea\label{ex:sun:1a}
        \gll  Urang nga-guna-keun éta émber.\\ 
              person.1\textsc{pl} \textsc{act}{}-use-\textsc{appl} \textsc{dem} bucket\\
        \glt ‘We use that bucket.’
    \ex\label{ex:sun:1b}
        \gll  Urang (\textbf{hen)teu} nga-guna-keun éta émber.\\
             person.1\textsc{pl} \textsc{neg} \textsc{act}{}-use-\textsc{appl} \textsc{dem} bucket\\
        \glt ‘We don’t use that bucket.’ 
    \z
\z 

Evidence that the negator is a particle rather than a verb is provided by its inability to take the plural infix \textit{{}-ar-} (or its allophone \textit{{}-al-}) which typically affixes to nouns and verbs, as shown in \REF{ex:sun:2}.
 
\ea\label{ex:sun:2}
    \ea[]{\label{ex:sun:2a}
        \gll \textbf{Ar}anjeunna teu t\textbf{ar}uang tos sabab\textbf{ar}aha dinten.  \\ 
              \textsc{3pl} \textsc{neg} \textsc{pl}-\textsc{eat} \textsc{iam} several day\\
        \glt ‘They hadn’t eaten for several days.’ \citep[43]{Butters2021}}
    \ex[*]{\label{ex:sun:2b}
        \gll \textbf{Ar}anjeunna t\textbf{ar}eu t\textbf{ar}uang tos sabab\textbf{ar}aha dinten. \\
              {\db}\textsc{3pl} \textsc{neg.pl} \textsc{pl}-eat \textsc{iam} several day\\
        \glt  ‘They hadn’t eaten for several days.’}
    \z
\z 

Sundanese has a number of special negators that are used in environments where \textit{(hen)teu} is dispreferred. Many of these will be presented in later sections. For the moment, I discuss clausal negators that are used in particular tenses or aspects. While Sundanese does not possess overt tense morphology, there are temporal adverbials (e.g. \textit{ayeuna} ‘now’, \textit{dinten ieu} ‘today’, \textit{énjing} ‘tomorrow’) and temporal/aspectual auxiliaries (e.g. \textit{nuju} for the progressive, \textit{atos} ‘already’ for the iamitive) whose occurrence and position in the clause are quite flexible. I turn now to address a few areas of negation related to tense and aspectual distinctions. These include: a special negative future marker, ‘not anymore’, ‘not yet’, and ‘never’. 

Negative future is encoded by a special marker, \textit{moal,} as demonstrated in \REF{ex:sun:3a}. The future counterpart in affirmative utterances is the high form \textit{badé} ‘want’ or the middle-to-low form \textit{(a)rék} ‘want’, shown in \REF{ex:sun:3b}, both of which have grammaticalized as future markers, though the lexical meaning is still transparent to speakers.\footnote{Other verbs, e.g., \textit{hoyong} ‘wish to’, only code desire and do not have the same function of encoding future.} \textit{Badé} and \textit{(a)rék} cannot be negated by the standard negator. \textit{Moal} thus constitutes an example of asymmetric negation, while most negation in Sundanese is generally symmetric, as in \REF{ex:sun:1a} and \REF{ex:sun:1b} (following the classification in \citealt{Miestamo2000, Miestamo2003, Miestamo2005}). 

\ea\label{ex:sun:3} 
    \ea\label{ex:sun:3a}
        \gll  Abdi \textbf{moal} ngiring ka ditu kumargi aya peryogi di kampus.\\ 
               1\textsc{sg} \textsc{neg}.\textsc{fut} \textsc{act}.follow to there because \textsc{ex} matter \textsc{loc} campus\\
        \glt ‘I will not join because I have something to do on campus.’
    \ex\label{ex:sun:3b}
        \gll  Abdi \textbf{badé} ngiring ka ditu.\\
             1\textsc{sg} want \textsc{act}.follow to there\\
        \glt ‘I will join there.’ \citep[76]{Butters2021} 
    \z
\z 

As demonstrated in \REF{ex:sun:4}, \textit{moal} is often followed by the adverbial \textit{waka} ‘too early, before its time’, which tends to collocate with negatives. Consultants report the perception that leaving out \textit{waka} ‘too early, before its time’ results in a statement that is too definite. Adverbials like \textit{waka} ‘too early, before its time’ and \textit{heula} ‘first’ are frequently affixed to verbs in Sundanese to encode politeness. When one leaves a group, for instance, one should say \textit{indit heula} ‘(I’m) leaving first’, even when one is not the first to go and when one does intend to return that day. Consultants note that this leaves open the possibility that one might return. 

\ea\label{ex:sun:4} 
    \gll  \textbf{Moal} \textbf{waka} angkat heula.\\ 
          \textsc{neg}.\textsc{fut} too.early go first\\
    \glt ‘(I) will not go just yet.’ \citep[77]{Butters2021}
\z 

\textsc{not} \textsc{anymore} is encoded by the co-occurrence of the standard negator, \textit{(hen)teu,} and \textit{deui} ‘again’, as shown in \REF{ex:sun:5}. 

\ea\label{ex:sun:5}
    \ea\label{ex:sun:5a}
        \gll  Aki atos \textbf{teu} damel di dieu \textbf{deui}.\\ 
              grandfather \textsc{iam} \textsc{neg} work \textsc{loc} here again\\
        \glt  ‘The older man (lit. grandfather) does not work here anymore.’ 
    \ex\label{ex:sun:5b}
        \gll  Kopi éta atos \textbf{teu} panas \textbf{deui}.\\
             coffee \textsc{dem} \textsc{iam} \textsc{neg} hot again\\
        \glt ‘That coffee is not hot anymore.’ 
    \z
\z 



\textsc{not} \textsc{yet} can be represented by a number of formal means. The first is the co-occurrence of the standard negator with \textit{heula} ‘first’, as demonstrated in \REF{ex:sun:6}. Such a strategy is also attested in Bantu languages \citep{VeselinovaDevos2021}. 

\ea\label{ex:sun:6} 
    \gll  Abdi mah \textbf{teu} \textbf{heula} mésér tas.\\ 
          1\textsc{sg} \textsc{dp} \textsc{neg} first buy bag\\
    \glt ‘I’m not going to buy the bag just yet.’ \citep[88]{Butters2021}
\z 

\textsc{not} \textsc{yet} can also be expressed by the verbal negator plus the polarity sensitive item \textit{acan} or its clipped form \textit{can}, as shown in \REF{ex:sun:7a}. In spoken Sundanese, the fused form \textit{tacan} is also common, as in \REF{ex:sun:7b}. Meanwhile, the iamitives\footnote{I follow \citet{DahlWälchli2016} in using the term \textit{iamitive} rather than \textit{perfect}, as these markers in Sundanese can combine with stative predicates like ‘be here’, as in \textit{atos di dieu} ‘already here’ and ‘be sick’, as in \textit{atos gering} ‘already sick’.}  include the polite form \textit{parantos/ atos/ tos} ‘already’ and the colloquial \textit{(ng)geus} ‘already’, as demonstrated in \REF{ex:sun:7c}; the iamitives cannot be negated, as shown in \REF{ex:sun:7d}. 



\ea\label{ex:sun:7}
    \ea[]{\label{ex:sun:7a}
        \gll  Manéhna \textbf{teu} \textbf{acan} nikah.\\ 
              3\textsc{sg} \textsc{neg} \textsc{phpc} marry\\
        \glt ‘S/he is not married yet.’
        }
    \ex[]{\label{ex:sun:7b}
        \gll  Abdi \textbf{tacan} pa-pendak sa-saha.\\
             1\textsc{sg} not.yet \textsc{red}{}-meet \textsc{red}{}-who\\
        \glt‘I haven’t met anyone yet.’
        }
    \ex[]{\label{ex:sun:7c}
        \gll  Manéhna \textbf{tos} nikah.\\ 
        3\textsc{sg} \textsc{iam} marry\\
        \glt ‘S/he is already married.’ \citep[92]{Butters2021}
        }
    \ex[*]{\label{ex:sun:7d}
        \gll  Manéhna \textbf{teu} \textbf{tos} nikah.\\
              3\textsc{sg} \textsc{neg} \textsc{iam} marry\\
        \glt  ‘S/he is not married yet.’
        }
    \z
\z 
    

Interestingly, in colloquial speech the particle \textit{(a)can} frequently appears alone yet retains its negative meaning, as demonstrated in a declarative utterance in \REF{ex:sun:8a} and a short conversation in \REF{ex:sun:8b}. 

\ea\label{ex:sun:8a}
    \gll  Ti isuk \textbf{can} ibak. \\ 
          from morning \textsc{phpc} bathe\\
    \glt ‘From morning, (I) haven’t bathed yet.’ 
\z 

\ea\label{ex:sun:8b}
    \begin{xlist}[Q: ]
        \exi{Q: }\label{ex:sun:8bA}
         \gll  Geus dahar \textbf{acan}?\\ 
              \textsc{iam} eat \textsc{phpc}\\ 
         \glt ‘Have you eaten yet, or not?’ 
        \exi{A: }\label{ex:sun:8bB}
         \gll  \textbf{Acan}.\\ 
               \textsc{phpc}\\
         \glt ‘Not yet.’ \citep[92]{Butters2021}
    \end{xlist}
\z 


The particle \textit{acan} will appear again in \sectref{sec:sun:4.2} on negative polarity. I reserve the analysis of ‘{not yet}’ for  \sectref{sec:sun:4.6.3} which handles diachronic issues. 

The last expression to be covered in this section is the coding of ‘{never}’, which is done through the negation of the perfect particle \textit{pernah}, as in \REF{ex:sun:9}, which encodes either negative present perfect or negative habitual. \textit{Acan} is optional but preferred in the former instance, according to consultants. 

\ea\label{ex:sun:9}
    \ea\label{ex:sun:9a}
        \gll Anjeuna \textbf{teu} (acan) \textbf{pernah} indit ka sekolah.  \\ 
              3\textsc{sg} \textsc{neg} \textsc{phpc} \textsc{prf} go to school\\
        \glt ‘He has never gone to school’ / ‘He never goes to school.’
    \ex\label{ex:sun:9b}
        \gll  Anjeuna \textbf{pernah} linggih di Ciamis.\\
        3\textsc{sg} \textsc{prf} live \textsc{loc} Ciamis \\
        \glt ‘He once lived in Ciamis’ / ‘He has lived in Ciamis.’ \citep[95]{Butters2021}
    \z
\z 

Alternatively, ‘{never}’ is encoded by a special particle \textit{tara,} as in \REF{ex:sun:10}. Speakers suggest that \textit{tara} means both ‘never’ and ‘rarely’, which again may have to do with a tendency to avoid definite negation. 

\ea\label{ex:sun:10} 
    \gll  Manéhna \textbf{tara} sare dina kasur.\\ 
          3\textsc{sg} never sleep \textsc{loc} mattress\\
    \glt ‘S/he never sleeps on a mattress.’ (when noting a girl’s preference to sleep on a mat) \citep[92]{Butters2021}
\z 
\subsection{Negation in non-declaratives}\label{sec:sun:2.2}

This section focuses on negation in imperative mood and interrogative mood. In imperative mood in affirmative contexts, the verb occurs in clause initial position and the subject is dropped, as in \REF{ex:sun:11a}, or postponed to the end of the utterance, as in \REF{ex:sun:11b}. A less direct command is encoded by passive voice, as in \REF{ex:sun:11c}. 

\ea\label{ex:sun:11}
    \ea\label{ex:sun:11a}
        \gll  Leueut kopi anjeun.\\ 
              drink coffee 2\textsc{sg}.\textsc{poss}\\
        \glt ‘Drink your coffee.’
    \ex\label{ex:sun:11b}
        \gll  Bobo di dieu anjeun.\\
              sleep \textsc{loc} here 2\textsc{sg}\\
        \glt ‘Sleep here, you.’
    \ex\label{ex:sun:11c}
        \gll  Mangga di-leueut kopi-na. \\
              \textsc{hort} \textsc{pass}{}-drink coffee-\textsc{lig}\\
        \glt ‘Go ahead and drink your coffee.’ \citep[137]{Butters2021}
    \z
\z 

The prohibitives, \textit{ulah} in higher speech level and \textit{tong} in lower speech level, are separate uninflected particles that occurs in phrase initial position. Sundanese joins the majority in the 495-language sample described by \citet{AuweraEtAl2013}, in having prohibitives that differ in form and position from the standard negator. The prohibitives can be used with both activities and attributes, as indicated in \REF{ex:sun:13}.

\ea\label{ex:sun:12}
    \ea\label{ex:sun:12a}
        \gll  \textbf{Ulah} saré kapetingan supados henteu hudang kaberangan.\\ 
              \textsc{proh} sleep late.night in.order.to \textsc{neg} wake late.day \\
        \glt ‘Don’t sleep late at night in order to not wake up too late.’
    \ex\label{ex:sun:12b}
        \gll  \textbf{Ulah} osok langlang-lingling.\\
              \textsc{proh} always wander-energetic \\
        \glt ‘Don’t always be aimless.’
    \ex\label{ex:sun:12c}
        \gll  \textbf{Ulah} gélo. \\
              \textsc{proh} crazy\\
        \glt ‘Don’t be crazy.’ \citep[137]{Butters2021}
    \z
\z 

A negative command is softened by the use of passive voice, as in \REF{ex:sun:13a}, or the addition of the existential predicator, \textit{aya,} as in \REF{ex:sun:13b}. 

\ea\label{ex:sun:13}
    \ea\label{ex:sun:13a}
        \gll  \textbf{Ulah} di-inum kopi manéh.\\ 
              \textsc{proh} \textsc{pass}{}-drink coffee 2\textsc{sg}\\
        \glt ‘Do not drink your coffee.’ (if, for instance, there is a fly in the coffee)
    \ex\label{ex:sun:13b}
        \gll  \textbf{Ulah} aya nu tinggal-eun.\\
              \textsc{proh} \textsc{ex} \textsc{rel} stay-\textsc{pass}\\
        \glt ‘Don’t leave anything behind.’ \citep[149]{Butters2021}
    \z
\z 

When the prohibitive is followed by the first person inclusive plural pronoun, a suggestion against collective action is encoded, as in \REF{ex:sun:14}. 

\ea\label{ex:sun:14} 
    \gll  \textbf{Ulah} arurang nyalah-keun manéhna.\\ 
          \textsc{proh} person.\textsc{pl} \textsc{act}.wrong-\textsc{appl} 3\textsc{sg}\\
    \glt ‘Let us not blame him.’ \citep[148]{Butters2021}
\z 
The hortative \textit{hayu,} which occurs in phrase initial position, as in \REF{ex:sun:15a}, can also be used with the prohibitive, as in \REF{ex:sun:15b}. The prohibitive is frequently used on its own for urging, even in absence of a verb, as in \REF{ex:sun:15c}. 
 
\ea\label{ex:sun:15}
    \ea\label{ex:sun:15a}
        \gll  \textbf{Hayu} urang ka sawah!\\ 
              \textsc{hort} person.1\textsc{pl} to paddy field\\
        \glt ‘Let’s go to the paddy field!’
    \ex\label{ex:sun:15b}
        \gll  \textbf{Hayu} urang \textbf{ulah} ka lapangan heula poé ieu.\\
              \textsc{hort} person.1\textsc{pl} \textsc{proh} to field first day \textsc{dem}\\
        \glt ‘Let’s not go to the field yet today.’
    \ex\label{ex:sun:15c}
        \gll  \textbf{Ulah} ka lapangan poé ieu!\\
              \textsc{proh} to field day \textsc{dem}\\
        \glt ‘Let’s not go to the field today!’ 
    \z
\z 

The prohibitive form has functions beyond those discussed thus far, but these are reserved for \sectref{sec:sun:4.6.2}. In the meantime, I turn to another category of negation in non-declarative utterances; namely, the behavior of negation in interrogative clauses.

Polar questions in Sundanese are typically expressed simply through the marking of a declarative utterance with rising intonation, as in \REF{ex:sun:16a} and with the negator in \REF{ex:sun:16b}. An additional strategy is fronting the verb and optionally postposing the subject, as in \REF{ex:sun:16c}, and with the negator in \REF{ex:sun:16d}. 

 
\ea\label{ex:sun:16}
    \ea\label{ex:sun:16a}
        \gll  Meri ngartos?\\ 
              Meri \textsc{act}.understand\\
        \glt ‘Do you (Meri) understand?’
    \ex\label{ex:sun:16b}
        \gll  Meri \textbf{teu} ngartos?\\
              Meri \textsc{neg} understand\\
        \glt ‘You (Meri) don't understand?’
    \ex\label{ex:sun:16c}
        \gll  Ngartos (anjeun) Meri?\\
             \textsc{act}.understand \textsc{2sg} Meri\\
        \glt ‘Do you understand, Meri?’
    \ex\label{ex:sun:16d}
        \gll  \textbf{Teu} ngartos (anjeun) Meri?\\
              \textsc{neg} \textsc{act}.understand \textsc{2sg} Meri\\
        \glt ‘Don't you understand, Meri?’
    \z
\z 

The verbal negator \textit{(hen)teu} also functions as a negative tag, as in \REF{ex:sun:17}. It is fairly flexible in its position in an utterance and its scope is always leftward. It is a neutral tag, used to seek information without prior expectation. 
 
\ea\label{ex:sun:17}
    \ea\label{ex:sun:17a}
        \gll  Jarak mal sareng stasiun tebih \textbf{henteu}?\\ 
              distance mall \textsc{com} station far \textsc{neg}\\
        \glt ‘Is the distance from the mall to the station far or not?’
    \ex\label{ex:sun:17b}
        \gll  Tebih \textbf{henteu} sareng stasiun jarak-na?\\
              far \textsc{neg} \textsc{com} station distance-\textsc{lig}\\
        \glt ‘Is it far or not from the station?’
    \z
\z 

The productive nominal negator \textit{lain,} presented in the next section, can also be used as a tag, though it is far more restricted in its position than \textit{(hen)teu}. It occurs at the end of the utterance and has wide sentential scope. It can be paraphrased, ‘isn’t it?/is it not the case?’, as demonstrated in \REF{ex:sun:18}. 

\ea\label{ex:sun:18}
    \ea\label{ex:sun:18a}
        \gll  Éta anjing ngagonggong \textbf{lain}?\\ 
              \textsc{dem} dog bark \textsc{neg} \\
        \glt ‘That’s a dog barking, isn’t it?’
    \ex\label{ex:sun:18b}
        \gll  Manéhna nu datang téh \textbf{lain}?\\
              3\textsc{sg} \textsc{rel} come \textsc{dp} \textsc{neg}\\
        \glt ‘He was the one who came, wasn’t he?’
    \z
\z  

Unlike the verbal negator, the nominal negator when used as a tag encodes prior expectation on the part of the speaker and expects an affirmative response. 

\subsection{Negation in stative predications}\label{sec:sun:2.3}

\citet[111]{Payne1997} divides stative predication into the following types: equation, proper inclusion, attribution, locative predication, existential predication, and possessive predication. These types are addressed in this section. 

Sundanese, like many non-Oceanic Austronesian languages, employs an entirely different negator for nominal predications than for verbal or adjectival predications. This negator, \textit{sanés} in the high speech level or \textit{lain} in middle speech level, occurs in the same position of the utterance as the verbal negator. \REF{ex:sun:19} demonstrates the use of \textit{lain} in an equative predication, where there is no change in word order between the affirmative and negative utterances. The use of the standard negator \textit{(hen)teu} here would result in an ungrammatical sentence. 

\ea\label{ex:sun:19}
    \ea\label{ex:sun:19a}
        \gll  Manéhna budak leutik.\\ 
              3\textsc{sg} child small\\
        \glt ‘S/he is a small child.’
    \ex\label{ex:sun:19b}
        \gll  Manéhna \textbf{lain} budak leutik.\\
              3\textsc{sg} \textsc{neg} child small\\
        \glt ‘S/he is not a small child.’
    \z
\z  

 The same negators are used to encode proper inclusion, as shown in \REF{ex:sun:20}. Once again, the standard negator is ungrammatical here. 

 
\ea\label{ex:sun:20}
    \ea\label{ex:sun:20a}
        \gll  Kasép tukang moro.\\ 
              Kasep craftsman hunt\\
        \glt ‘Kasep is a hunter.’
    \ex\label{ex:sun:20b}
        \gll  Kasép \textbf{lain} tukang moro. \\
              Kasep \textsc{neg} craftsman hunt\\
        \glt ‘Kasep is not a hunter.’ \citep[44]{Butters2021}
    \z
\z 

 The negation of attribution, on the other hand, is most typically expressed with the standard negator \textit{(hen)teu}, as in \REF{ex:sun:21b}. However, in contrastive negation like \REF{ex:sun:21c}, the nominal negator is preferred. 

 
\ea\label{ex:sun:21}
    \ea\label{ex:sun:21a}
        \gll  Awak manéhna tiis. \\ 
              body 3\textsc{sg} cold\\
        \glt ‘S/he is very cold.’
    \ex\label{ex:sun:21b}
        \gll  Awak manéhna \textbf{teu} tiis.\\
              body 3\textsc{sg} \textsc{neg} cold \\
        \glt ‘S/he is not very cold.’
    \ex\label{ex:sun:21c}
        \gll  Manéhna \textbf{lain} males tapi tunduh.\\
             3\textsc{sg} \textsc{neg} lazy but sleepy\\
        \glt ‘S/he’s not lazy, but sleepy.’ (when defending a child not helping with a task)
    \z
\z 

In Sundanese, affirmative existential predicates are used to code locative predication, existential predication, locative-presentative constructions (identified in \citealt{Veselinova2013neg.exist}), and possessive predication.\footnote{The literature on existential predications is vast and is beyond the scope of the present paper.} In locative predications, as in \REF{ex:sun:22}, the use of the existential predicator, \textit{aya,} is optional, while locative markers are obligatory; the standard negator precedes the affirmative existential. 

\ea\label{ex:sun:22}
    \ea\label{ex:sun:22a}
        \gll  Ucing \textbf{(aya)} \textbf{di} samak.\\ 
              cat \textsc{ex} on mat\\
        \glt ‘The cat is on the mat.’ 
    \ex\label{ex:sun:22b}
        \gll  Ucing \textbf{teu} \textbf{(aya)} \textbf{di} samak.\\
              cat \textsc{neg} \textsc{ex} on mat\\
        \glt ‘The cat is not on the mat.’
    \z
\z 

Locative-presentative constructions, presented in \REF{ex:sun:23}, contain a non-specific subject preceded by an existential or negative existential. 

 
\ea\label{ex:sun:23}
    \ea\label{ex:sun:23a}
        \gll  \textbf{Aya} jalmi \textbf{di} kampung ieu nu gaduh imah ageung.\\ 
              \textsc{ex} person \textsc{loc} village \textsc{dem} \textsc{rel} have house big\\
        \glt ‘There are people in this village who have big houses.’
    \ex\label{ex:sun:23b}
        \gll  \textbf{Teu} \textbf{aya} jurig \textbf{di} leuweung.\\
              \textsc{neg} \textsc{ex} ghost \textsc{loc} forest\\
        \glt ‘There are no ghosts in the forest.’
    \z
\z 

 Sundanese existential predications and negative existential predications are shown in \REF{ex:sun:24}. The affirmative existential is \textit{aya}, which can be negated by the standard negator, as in \textit{teu aya} or the fused form \textit{taya}. Additionally, there is a special negative existential predicator – that is to say, a variant of the type discussed in \citet{Veselinova2013neg.exist} that is distinct from the negative existential formed by the standard negator. In Sundanese, this form is \textit{euweuh}, shown in \REF{ex:sun:24d}. 

 \ea\label{ex:sun:24}
    \ea\label{ex:sun:24a}
        \gll  \textbf{Aya} loba béas.\\ 
              \textsc{ex} many rice\_grains\\
        \glt ‘There is a lot of rice.’
    \ex\label{ex:sun:24b}
        \gll  \textbf{Teu} \textbf{aya} loba béas.
\\
              \textsc{neg} \textsc{ex} many rice\_grains\\
        \glt ‘There is a lot of rice.’

    \ex\label{ex:sun:24c}
        \gll  \textbf{Taya} loba béas!\\
              \textsc{neg}.\textsc{ex} many rice\_grains\\
        \glt ‘There is not much rice!’

    \ex\label{ex:sun:24d}
        \gll  \textbf{Euweuh} anjing anu ngagonggong di luar.\\
              \textsc{neg}.\textsc{ex} dog \textsc{rel} \textsc{act}.bark \textsc{loc} outside\\
        \glt ‘There is no dog barking outside.’ \citep[109]{Butters2021}
    \z
\z 

The negative existentials can be used interchangeably, though \textit{euweuh} is considered very coarse and is probably the most recent predicator, following the principle that forms associated with innovative sociolinguistic forces are likely to be newer \citep{Greenberg1966synchronic}.\footnote{I suspect \textit{euweuh} might have developed from a lexeme meaning ‘lost’, but more evidence is needed to support this analysis} The selection of negative existentials in Sundanese appears to have more to do with attending to the correct speech levels than with any distinct functions performed by the existentials. 

Possessive predication can be encoded through use of the high form transitive verb \textit{gaduh} ‘have’ or the lower form \textit{boga} ‘have’, as in \REF{ex:sun:25a}, which like other verbs are negated by \textit{(hen)teu,} as in \REF{ex:sun:25b}. Additionally, the existential predicator can be used to code possession, as in \REF{ex:sun:25c}, and the negative existential can negate possession, but only when there is a definite subject in the initial slot of the utterance, as in \REF{ex:sun:25d}. Each of the existential predicators discussed thus far are compatible with coding possession. 

  \ea\label{ex:sun:25}
    \ea\label{ex:sun:25a}
        \gll  Abdi \textbf{gaduh} seueur waktu. \\ 
              1\textsc{sg} have many time \\
        \glt ‘I have lots of time.’
    \ex\label{ex:sun:25b}
        \gll  Abdi \textbf{teu} \textbf{gaduh} seueur waktu. \\
              1\textsc{sg} \textsc{neg} have many time \\
        \glt ‘I don’t have a lot of time.’

    \ex\label{ex:sun:25c}
        \gll  Faheem \textbf{aya} karanjang nu hampang.\\
              Faheem \textsc{ex} basket \textsc{rel} lightweight\\
        \glt ‘Faheem has a lightweight basket.’

    \ex\label{ex:sun:25d}
        \gll Faheem t\textbf{aya} karanjang nu hampang.\\
              Faheem \textsc{neg}.\textsc{ex} basket \textsc{rel} lightweight\\
        \glt ‘Faheem does not have a lightweight basket.’
    \z 
\z

As shown in \REF{ex:sun:26}, when the possessed entity is in sentence initial position, the comitative \textit{sareng} or the locative \textit{di} is obligatory, while the existential predicate is optional. 

 \ea\label{ex:sun:26}
    \ea\label{ex:sun:26a}
        \gll  Poci-na (\textbf{aya)} \textbf{sareng} abdi.\\ 
              teapot-\textsc{def} \textsc{ex} \textsc{com} 1\textsc{sg}\\
        \glt ‘I have the teapot.’
    \ex\label{ex:sun:26b}
        \gll  Poci-na \textbf{teu} \textbf{(aya)} \textbf{sareng} abdi.\\
              teapot-\textsc{def} \textsc{neg} \textsc{ex} \textsc{com} 1\textsc{sg}\\
        \glt ‘I do not have the teapot.’
    \ex\label{ex:sun:26c}
        \gll  Konci-na (\textbf{aya)} \textbf{di} abdi.\\
              key-\textsc{def} \textsc{ex} \textsc{loc} 1\textsc{sg}\\
        \glt ‘I have the key.’
    \ex\label{ex:sun:26d}
        \gll  Konci-na \textbf{teu} \textbf{(aya)} \textbf{di} abdi.\\
              key-\textsc{def} \textsc{neg} \textsc{ex} \textsc{loc} 1\textsc{sg}\\
        \glt ‘I don’t have the key.’
    \z
\z 

The use of the comitative corresponds to a Companion schema while the use of the locative corresponds to a Location schema, in the sense of \citet{Heine1997}.

\subsection{Negation in non-main clauses}\label{sec:sun:2.4}

Negation in dependent and subordinate clauses is expressed through standard negation. Complementizers are optional and rarely used in spoken language. The variants in \REF{ex:sun:27} are all possible. 

\ea\label{ex:sun:27}
    \ea\label{ex:sun:27a}
        \gll  Urang ningali (yén) anjeuna datang. \\ 
              person.1\textsc{sg} \textsc{act}.see \textsc{comp} 3\textsc{sg} come.\\
        \glt ‘I saw (that) he’s coming.’ 
    \ex\label{ex:sun:27b}
        \gll  Urang ningali (yén) anjeuna teu datang. \\
              person.1\textsc{sg} \textsc{act}.see \textsc{comp} 3\textsc{sg} \textsc{neg} come.\\
        \glt ‘I saw (that) he’s not coming.’ 
    \ex\label{ex:sun:27c}
        \gll  Urang teu ningali (yén) anjeuna datang. \\
              person.1\textsc{sg} \textsc{neg} \textsc{act}.see \textsc{comp} 3\textsc{sg} come\\
        \glt ‘I didn’t see (that) he came.’
    \ex\label{ex:sun:27d}
        \gll  Urang teu ningali (yén) anjeuna teu datang. \\
              person.1\textsc{sg} \textsc{neg} \textsc{act}.see \textsc{comp} 3\textsc{sg} \textsc{neg} come \\
        \glt ‘I didn’t see (that) he did not come.’
    \z
\z 

\subsection{Negative lexicalizations}\label{sec:sun:2.5}

Earlier in this chapter, I described special negators like the future negator, \textit{moal,} and the ‘{not yet}’ forms, \textit{teu acan, tacan,} or \textit{can.} The aim of this section is to present two other irregular negative verbs that fall within the realm of volition and cognition: ‘{not want}’ and ‘{not know}’. The existence of special negators for not wanting and not knowing is not particularly surprising given what is known of domains that tend to exhibit special negative lexicalization. \citet{Veselinova2013lexicalization}, for instance, notes that after negative existentials and ‘not yet’ expressions, the next most common negative lexicalizations include ‘not know’, ‘not be of identity’, ‘cannot’, ‘not want’, ‘not talk’ and ‘need not’.

In Sundanese, ‘{not want}’ can be encoded by the irregular low form negative verb, \textit{embung}, or the high form \textit{alim}.\footnote{\citet[219]{Hardjadibrata2003} also lists the forms \textit{narah} and \textit{teu kersa} for \textsc{not want}, which are not attested in my data.} These negators occur in the same slot in the utterance as standard negation, as demonstrated in \REF{ex:sun:28a} and \REF{ex:sun:28b}. \textit{Alim} can also take the third person suffix -\textit{eun} as shown in \REF{ex:sun:28c}. 

 \ea\label{ex:sun:28}
    \ea\label{ex:sun:28a}
        \gll  Abdi \textbf{alim} kopi atanapi téh. \\ 
              1\textsc{sg} \textsc{neg}.want coffee or tea\\
        \glt ‘I don’t want coffee nor tea.’
    \ex\label{ex:sun:28b}
        \gll  Hena \textbf{alim} emam buah ngora éta da haseum.\\
              Hena \textsc{neg}.want eat fruit young \textsc{dem} because sour\\
        \glt ‘Hena doesn’t want to eat that unripe fruit because (it’s) sour.’
    \ex\label{ex:sun:28c}
        \gll  Anjeunna mah \textbf{alim-eun}. \\
              3\textsc{sg} \textsc{dp} unwilling-3 \\
        \glt ‘He doesn’t want to.’ / ‘He is unwilling.’ \citep[73]{Butters2021}
    \z
\z 

 Alternatively, the lexeme \textit{hoyong} ‘want’ (or the colloquial \textit{hayang}) can be negated through ordinary means, as in \REF{ex:sun:29}. 

\ea\label{ex:sun:29} 
    \gll  Hena \textbf{teu} hoyong tuang buah éta.\\ 
          Hena \textsc{neg} desire eat fruit \textsc{dem}\\
    \glt ‘Hena doesn’t want to eat that fruit.’ \citep[73]{Butters2021}
\z 

In colloquial speech, \textit{alim} is often used in replies, as in \REF{ex:sun:30a}. It is frequently accompanied by the discourse particle \textit{ah} indicating frustration. 

\ea\label{ex:sun:30a}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:sun:30aQ}
        \gll  Badé ngiring sareng bapa?\\
            want \textsc{act}.follow \textsc{com} father\\
        \glt ‘Do you want to go with me (lit. father)?’
    \exi{A: }\label{ex:sun:30aA}
        \gll  \textbf{Alim} da badé ngiring sareng ibu.\\
            \textsc{neg}.want because want \textsc{act}.follow \textsc{com} mother\\
        \glt ‘(I) don’t want to because (I’m) going with mother.’ \citep[74]{Butters2021}
    \end{xlist}
\z 

Example \REF{ex:sun:30b} suggests that \textit{alim} may be broadening into the domain of ability modality. 

\ea\label{ex:sun:30b}    
    \begin{xlist}[B: ]
    \exi{A: }\label{ex:sun:30bA}
        \gll   Kang hayu urang ka sawah!\\
            big.brother \textsc{hort} person.1\textsc{pl} to paddy.fields\\
        \glt ‘Come on, man, let’s go to the paddy fields!’
    \exi{B: }\label{ex:sun:30bB}
        \gll  \textbf{Alim} ah abdi badé ngala suluh! \\
            \textsc{neg}.want \textsc{dp} 1\textsc{sg} want \textsc{act}.gather firewood\\
        \glt ‘I don’t want to because I have to collect firewood!’ \citep[74]{Butters2021}
    \end{xlist}
\z

A second negative lexicalization is \textit{duka} meaning ‘{don't know}’. \textit{Duka} occurs most often in sentence initial position or in answer to questions, as in \REF{ex:sun:31}. 

 
\ea\label{ex:sun:31}
    \begin{xlist}
    \exi{A: }\label{ex:sun:31A}
        \gll  Manéh kunaon batuk waé? \\ 
              2\textsc{sg} why cough just\\ 
        \glt ‘Why are you coughing?’
    \exi{B: }\label{ex:sun:31B}
        \gll  \textbf{Duka} kana kebul meureun. \\ 
               \textsc{neg}.know to dust maybe\\
        \glt ‘I don’t know, maybe from the dust.’ \citep[82]{Butters2021}
    \end{xlist}
\z 

The polite \textit{terang} ‘know’ (or the colloquial \textit{nyaho} ‘know’) can also be negated by the regular verbal negator, as in \REF{ex:sun:32a} and \REF{ex:sun:32c}. Speakers do not accept the substitution of \textit{duka} ‘not know’ here. 

 \ea\label{ex:sun:32}
    \ea[]{\label{ex:sun:32a}
        \gll  Hena \textbf{teu} \textbf{terang} kunaon manéhna ngambek.\\ 
              Hena \textsc{neg} know why 3\textsc{sg} \textsc{act}.angry\\
        \glt ‘Hena doesn’t know why s/he is angry.’
        }
    \ex[*]{\label{ex:sun:32b}
        \gll  Hena \textbf{duka} kunaon manéhna ngambek.\\
              Hena \textsc{neg}.know why 3\textsc{sg} \textsc{act}.angry\\
        \glt ‘Hena doesn’t know why s/he is angry.’
        }
    \ex[]{\label{ex:sun:32c}
        \gll  Upami \textbf{teu} \textbf{terang} jalan-na saé-na ulah angkat nyalira.\\
              if \textsc{neg} know road-\textsc{def} good-\textsc{lig} \textsc{proh} go alone\\
        \glt ‘If you don’t know the right road, it’s better not to go alone.’ 
        }
    \ex[*]{\label{ex:sun:32d}
        \gll  Upami \textbf{duka} jalan-na saé-na ulah angkat nyalira.\\
              if \textsc{neg}.know road-\textsc{def} good-\textsc{lig} \textsc{proh} go alone\\
        \glt ‘If you don’t know the right road, it’s better not to go alone.’
        }
    \z
\z 

\textit{Duka} has a broader function of appealing to general ignorance, as in \REF{ex:sun:33a}, corresponding to the Indonesian \textit{entah} ‘who knows’. This broader function is particularly evident in expressions where \textit{duka} occurs in the same utterance as \textit{teu terang,} as in \REF{ex:sun:33b}. 

 \ea\label{ex:sun:33}
    \ea\label{ex:sun:33a}
        \gll  \textbf{Duka} kunaon manéhna ngambek deui!\\ 
              \textsc{neg}.know why 3\textsc{sg} angry again\\
        \glt ‘Who knows why she is angry again!’
    \ex\label{ex:sun:33b}
        \gll  \textbf{Duka} abdi \textbf{teu} \textbf{terang}. \\
              \textsc{neg}.know 1\textsc{sg} \textsc{neg} know\\
        \glt ‘Who knows, I don’t know.’ 
    \z
\z 

In Sundanese and in Indonesian, \textit{duka} can also mean ‘sorrow, sadness’, appearing in the Sundanese \textit{duduka} ‘anger, displeasure, disgrace’ and \textit{kadukaan} ‘grief, sorrow’ \citep[210]{Hardjadibrata2003}, as well as in the Indonesian \textit{berduka cita} ‘to mourn, to grieve’, among other lexemes. It is possible that the ignorative meaning grammaticalized from an expression of sorrow or apology. 

\subsection{Other clausal negation constructions}\label{sec:sun:2.6}

This chapter has already included several means to soften negative force. An additional strategy is the substitution of the lexeme \textit{kirang} ‘less’ for the standard negator, as in \REF{ex:sun:34}. \textit{Kirang} can occur directly before verbs and attributes. 

\ea\label{ex:sun:34}
    \ea\label{ex:sun:34a}
        \gll  Abdi \textbf{kirang} percanten kana jawab-an-na. \\ 
              1\textsc{sg} less believe-3 to answer-\textsc{nmlz}-3\textsc{sg}\\
        \glt ‘I don’t really believe his answer.’
    \ex\label{ex:sun:34b}
        \gll  Cuaca dinten ieu \textbf{kirang} saé pisan.\\
              weather day \textsc{dem} less good very\\
        \glt ‘The weather today was really not so good.’ \citep[83]{Butters2021}
    \z
\z 

The form \textit{kurang} ‘less’ encodes the same function in Standard Indonesian. Given the high level of bilingualism between the two languages, it is likely that the construction was borrowed, though the direction of that borrowing is unclear. 

\section{Non-clausal negation}\label{sec:sun:3}

\subsection{Negative replies}\label{sec:sun:3.1}

Replies expressing ‘no’ make use of any of the negators discussed thus far, depending on the predicate type. Example \REF{ex:sun:35a} demonstrates a standard negative reply. \REF{ex:sun:35b} shows the use of \textit{alim} ‘not want’ in a reply. There are no changes to these forms in replies, though there is perhaps a general preference to use a full form rather than a clipped form, especially if the reply is emphatic.

\ea\label{ex:sun:35a}
    \begin{xlist}
    \exi{A: }\label{ex:sun:35a1}
        \gll  Linggih di dieu?\\ 
              live \textsc{loc} here\\ 
        \glt ‘Do you live here?’
    \exi{B: }\label{ex:sun:35a2}
        \gll  \textbf{Henteu}.\\ 
               \textsc{neg}\\
        \glt  ‘No (I don’t).’
    \end{xlist}
\z
\ea\label{ex:sun:35b}
    \begin{xlist}
    \exi{A: }\label{ex:sun:35b1}
        \gll  Badé ngiring ka ditu?\\ 
              want \textsc{act}.follow to there\\ 
        \glt‘Do you want to join?’ 
    \exi{B: }\label{ex:sun:35b2}
        \gll  \textbf{Alim}.\\ 
                \textsc{neg}.want\\
        \glt ‘No, I don’t want to.’
    \end{xlist}
\z  

A negative reply agrees with the polarity of a negative question, such that in \REF{ex:sun:36} the answer, \textit{moal,} confirms the supposition of the questioner and in the process negates the propositional content. The affirmative future marker, \textit{badé,} disagrees with the negative question, making it polarity reversing. 

 
\ea\label{ex:sun:36}
    \begin{xlist}[A2: ]
    \exi{Q: }\label{ex:sun:36A}
        \gll  Asep \textbf{moal} bobo di dieu?\\ 
              Asep \textsc{neg}.\textsc{fut} sleep \textsc{loc} here\\ 
        \glt ‘Asep is not going to sleep here?’
    \exi{A1: }\label{ex:sun:36B1}
        \gll  \textbf{Moal}.\\ 
               \textsc{neg}.\textsc{fut}\\
        \glt ‘Yes, that’s right, (Asep) is not going to (sleep here).’ 
    \exi{A2: }\label{ex:sun:36B2}
        \gll  \textbf{Badé}.\\ 
              \textsc{fut}\\
        \glt ‘No, that’s not right, (Asep) is going to (sleep here).’
    \end{xlist}
\z

\subsection{Negative indefinites and quantifiers}\label{sec:sun:3.2}

Sundanese does not have special negative indefinite pronouns like the English \textit{nothing} nor the Latin \textit{nemo.} Rather, like many Austronesian languages, indefiniteness in Sundanese is most often encoded via existential predication, reduplication of interrogatives, and co-occurrence of interrogatives and special particles with additive and downward entailing functions. Example \REF{ex:sun:37} demonstrates indefinite pronouns formulated through the existential \textit{aya} and the relativizer \textit{(a)nu}. The tendency for existentials to code indefinite pronouns has also been observed in Tagalog, Cebuano, and other languages of the Philippines, as well as in Oceanic languages \citep{Haspelmath1997}. 

\ea\label{ex:sun:37}
    \ea\label{ex:sun:37a}
        \gll  \textbf{Aya} nu dongkap.\\ 
              \textsc{ex} \textsc{rel} come\\
        \glt ‘Someone came.’ (lit. ‘There are that came.’)
    \ex\label{ex:sun:37b}
        \gll  \textbf{Teu} \textbf{aya} nu dongkap.\\
              \textsc{neg} \textsc{ex} \textsc{rel} come\\
        \glt ‘No one came.’ (lit. ‘There was not that came.’) 
    \ex\label{ex:sun:37c}
        \gll  \textbf{Aya} nu bisa di-béré-keun. \\
               \textsc{ex} \textsc{rel} can \textsc{pass}{}-give{}-\textsc{appl}\\
        \glt ‘Something can be given.’ (lit. ‘There is that can be given.’)
    \ex\label{ex:sun:37d}
        \gll  \textbf{Teu} \textbf{aya} nu bisa di-béré-keun.\\
              \textsc{neg} \textsc{ex} \textsc{rel} can \textsc{pass}{}-give-\textsc{appl}\\
        \glt ‘Nothing can be given. (lit. ‘There is naught that can be given.’)
    \z
\z 

Quantification can additionally be encoded through both full reduplication and partial reduplication of interrogatives. The latter is more common, probably due to phonological reduction. Where partial reduplication is found, it is the first syllable that repeats, as shown in \tabref{tab:sun:1} below.

\begin{table}
    \centering
    \begin{tabular}{llll}
        \itshape Saha & ‘who’ & \itshape saha-saha/ sasaha & ‘everyone’\\        
        \itshape Naon & ‘what’ & \itshape naon-naon/ nanaon & ‘everything’\\        
        \itshape Iraha & ‘when’ & \itshape iraha-iraha/*ir-iraha & ‘every time’\\        
        \itshape Ka mana & ‘where to’ &\itshape  ka mana mana/ ka mamana & ‘to everywhere’\\        
        \itshape Di mana & ‘where at’ & \itshape di mana mana/ di mamana & ‘(at) everywhere’\\
    \end{tabular}
    \caption{Simplex interrogatives and reduplicated interrogatives}
    \label{tab:sun:1}
\end{table}

The reduplicated interrogatives most frequently occur in negative declarative utterances, given that reduplication is one means that is sensitive to negation and other polarity phenomena. This is demonstrated in \REF{ex:sun:38} below, where each utterance is used in emphatic, newsworthy contexts. 

 \ea\label{ex:sun:38}
    \ea\label{ex:sun:38a}
        \gll  Manéh \textbf{teu} terang \textbf{na-naon} ngeunaan bisnis!\\ 
              2\textsc{sg} \textsc{neg} know \textsc{red}{}-what about business\\
        \glt ‘You don’t know anything about business!’
    \ex\label{ex:sun:38b}
        \gll  Manéhna \textbf{teu} resep ka \textbf{sa-saha}!\\
              3\textsc{sg} \textsc{neg} like to \textsc{red}{}-who\\
        \glt ‘She doesn’t like anyone!’
    \ex\label{ex:sun:38c}
        \gll  Tata \textbf{teu} boga \textbf{na-naon}!\\
              Tata \textsc{neg} have \textsc{red}{}-what\\
        \glt ‘Tata doesn’t have anything!’
    \ex\label{ex:sun:38d}
        \gll  Galuh \textbf{teu} ka \textbf{ma-mana}!\\
              Galuh \textsc{neg} to \textsc{red}{}-where\\
        \glt ‘Galuh didn’t go anywhere!’ \citep[172]{Butters2021}
    \z
\z 

 The reduplicated interrogatives can also appear in affirmative declaratives, as in \REF{ex:sun:39}. 

 \ea\label{ex:sun:39} 
    \ea\label{ex:sun:39a}
        \gll  Anjeun terang \textbf{na-naon} tentang bisnis.\\ 
              2\textsc{sg} know \textsc{red}{}-what about business\\
        \glt ‘You know many things about business.’
    \ex\label{ex:sun:39b}
        \gll  Manéhna resep ka \textbf{sa-saha}.\\
              3\textsc{sg} like to \textsc{red}{}-who\\
        \glt ‘She likes everyone.’
    \ex\label{ex:sun:39c}
        \gll  Tata boga \textbf{na-naon}.\\
              Tata have \textsc{red}{}-what\\
        \glt ‘Tata has everything.’
    \ex\label{ex:sun:39d}
        \gll  Galuh ka \textbf{ma-mana}.\\
              Galuh to \textsc{red}{}-where\\
        \glt ‘Galuh goes everywhere.’ \citep[172]{Butters2021}
    \z
\z 

The reduplicated interrogatives are used fairly infrequently in questions, except when they occur in phrase initial position, as demonstrated by the reduplication of \textit{saha} ‘who’ in \REF{ex:sun:40}. There appear to be slightly different functions encoded in the examples below; in \REF{ex:sun:40a} plurality of subjects is encoded, while in \REF{ex:sun:40b} specificity is encoded. 

 
\ea\label{ex:sun:40}
    \ea\label{ex:sun:40a}
        \gll  \textbf{Saha-saha} nu ulin di buruan?\\ 
              who-\textsc{red} \textsc{rel} play \textsc{loc} yard\\
        \glt ‘Who all is playing in the yard?’
    \ex\label{ex:sun:40b}
        \gll  \textbf{Saha-saha} nu meupeues-keun kaca?\\
              who-\textsc{red} \textsc{rel} \textsc{act}.break-\textsc{appl} glass\\
        \glt ‘Who exactly was it that broke the glass?’ \citep[173]{Butters2021}
    \z
\z 

 More commonly, interrogatives are followed by the minimizing particle \textit{wé/ waé/ baé} ‘just, merely, only’, as in \REF{ex:sun:41}. 

 
\ea\label{ex:sun:41}
    \ea\label{ex:sun:41a}
        \gll  Emangna manéh terang \textbf{naon} \textbf{waé} tentang bisnis?\\ 
              indeed 2\textsc{sg} know what just about business\\
        \glt ‘Truly, what all do you know about business?’
    \ex\label{ex:sun:41b}
        \gll  Manéhna resep ka \textbf{saha} \textbf{waé}?\\
              3\textsc{sg} like to who just\\
        \glt ‘Who all does she like?’ (\citealt{Butters2021}: 178, 179)
    \z
\z 
The same construction encodes free-choice in declarative utterances, as in \REF{ex:sun:42}. 

 \ea\label{ex:sun:42}
    \ea\label{ex:sun:42a}
        \gll  \textbf{Saha} \textbf{wé} bisa nga-lapor ka kantor polisi.\\ 
              who just can \textsc{act}{}-report to office police\\
        \glt ‘Anybody at all can report to the police office.’
    \ex\label{ex:sun:42b}
        \gll  Manéh tiasa dongkap ka bumi abdi \textbf{iraha} \textbf{waé}.\\
              2\textsc{sg} can come to house 1\textsc{sg} when just\\
        \glt ‘You can come to my house anytime.’ \citep[178]{Butters2021}
    \z
\z 

In the negative, expressions with \textit{waé} appear to be strongly dispreferred while reduplicated interrogatives commonly appear with negators, as in \REF{ex:sun:43a}. The interrogative can be used with \textit{waé} under the scope of negation if more information follows, as in \REF{ex:sun:43b}. However, the use of \textit{waé} with interrogatives is most felicitous in concessive utterances, as in \REF{ex:sun:43c}, where the interrogative + \textit{waé} occur outside the scope of negation. 

\ea\label{ex:sun:43}
    \ea\label{ex:sun:43a}
        \gll  Abdi mah \textbf{teu} tiasa ningali \textbf{na-naon}!\\ 
              1\textsc{sg} \textsc{dp} \textsc{neg} can see \textsc{red}-what\\
        \glt ‘I can’t see anything!’
    \ex\label{ex:sun:43b}
        \gll  Abdi mah \textbf{teu} tiasa ningali \textbf{naon} \textbf{waé} nu aya di dieu. \\
              1\textsc{sg} \textsc{dp} \textsc{neg} can see what just \textsc{rel} \textsc{ex} \textsc{loc} here\\
        \glt ‘I can’t see anything that is here.’
    \ex\label{ex:sun:43c}
        \gll  \textbf{Naon} \textbf{waé} nu aya di dieu \textbf{teu} ka-tingali ku abdi. \\
              what just \textsc{rel} \textsc{ex} \textsc{loc} here \textsc{neg} \textsc{pass}{}-see by 1\textsc{sg}\\
        \glt ‘Whatever was here was not seen by me.’ \citep[179]{Butters2021}
    \z
\z 

It appears that expressions with \textit{waé} are less compatible with negation given that \textit{waé} is already downward entailing. I return to the importance of reduplication in polarity readings in \sectref{sec:sun:4.2}. 

\subsection{Negative derivation and case marking (abessives/caritives/privatives)} \label{sec:sun:3.3}

Most frequently, ‘without’ is expressed through a periphrastic expression where\-in the negator precedes the polite \textit{ngagunakeun} ‘use’ (or the colloquial \textit{make} ‘use’), as in \REF{ex:sun:44b}.

\ea\label{ex:sun:44}
    \ea\label{ex:sun:44a}
        \gll Anjeuna tuang \textbf{nga-guna-keun} sendok. \\ 
                3\textsc{sg} eat \textsc{act}{}-use-\textsc{appl} spoon\\
        \glt ‘S/he eats with a spoon.’
    \ex\label{ex:sun:44b}
        \gll  Anjeuna tuang \textbf{teu} \textbf{nga-guna-keun} sendok.\\
             3\textsc{sg} eat \textsc{neg} \textsc{act}{}-use-\textsc{appl} spoon\\
        \glt ‘S/he eats without a spoon.’
    \z
\z 

It is also possible to negate the comitative \textit{sareng}, as in \REF{ex:sun:45}. This strategy can be employed for both animate and inanimate entities.

 \ea\label{ex:sun:45}
    \ea\label{ex:sun:45a}
        \gll  Abdi nga-leueut kopi \textbf{teu} \textbf{sareng} gula.\\ 
              1\textsc{sg} \textsc{act}{}-drink coffee \textsc{neg} \textsc{com} sugar\\
        \glt ‘I drink coffee without sugar.’
    \ex\label{ex:sun:45b}
        \gll  Abdi maehan rusa \textbf{teu} \textbf{sareng} amang abdi.\\
              1\textsc{sg} hunt deer \textsc{neg} \textsc{com} uncle 1\textsc{sg}\\
        \glt ‘I hunt deer without my uncle.’
    \ex\label{ex:sun:45c}
        \gll  Abdi bakal nyeri haté pisan lamun hirup \textbf{teu} \textbf{sareng} pun bapak. \\
              1\textsc{sg} \textsc{fut} sick liver very if live \textsc{neg} \textsc{com} \textsc{hon} father\\
        \glt ‘I would be so very sick to my heart if I had to live without my father.’
    \z
\z

\ea\label{ex:sun:46} 
    \gll  Barudak sadayana ningali abdi téh meni \textbf{teu} leupas-leupas. \\ 
          child.\textsc{pl} all \textsc{act}.look 1\textsc{sg} \textsc{dp} very \textsc{neg} release{}-\textsc{red}  \\
    \glt ‘All the children looked at me ceaselessly.’
\z

At the lexical level, the verbal negator is used to mark absence, as demonstrated in \REF{ex:sun:46}. 

Alternatively, as shown in \REF{ex:sun:47}, the negative existential predicators can be used to encode absence when preceding a noun, a cross-linguistically common expression described in \citet{Veselinova2013neg.exist}. In \REF{ex:sun:47b} it is ungrammatical to leave \textit{aya} out.

 \ea\label{ex:sun:47}
    \ea\label{ex:sun:47a}
        \gll  Manéhna gawé \textbf{euweuh} ka-capé nga-jelang acara pernikahan. \\ 
              3\textsc{sg} work \textsc{neg}.\textsc{ex} \textsc{nmlz}{}-tired \textsc{act}{}-approach event wedding\\
        \glt ‘She worked tirelessly leading up to the wedding.’
    \ex\label{ex:sun:47b}
        \gll  Pun bapak \textbf{teu} \textbf{aya} ka-capé nga-damel karanjang siang wengi.\\
              \textsc{hon} father \textsc{neg} \textsc{ex} \textsc{nmlz}{}-tired \textsc{act}{}-work basket day night\\
        \glt ‘My dad doesn’t feel tired making a basket all day.’ \citep[110]{Butters2021}
    \z
\z 

Finally, the form \textit{béak} ‘finished, used up, elapsed’ encodes absence, e.g. \textit{béak} \textit{tanaga} ‘powerless’, \textit{béak napas} ‘breathless’, \textit{béak déngkak} ‘at wit’s end.’

\section{Other aspects of negation}\label{sec:sun:4}

\subsection{The scope of negation}\label{sec:sun:4.1}

As indicated in \sectref{sec:sun:2.1}, the standard negator \textit{(hen)teu} most commonly has scope over verbs and adjectives; however, it also has scope over adverbs (e.g., \textit{teu énjing} ‘not tomorrow), numerals (e.g., \textit{teu hiji} ‘not one’), and interrogatives like \textit{naon} ‘what’ (e.g., \textit{teu naon} ‘no problem’). In some situations, the scope of the negator cannot be determined syntactically, as analysis depends on context. For instance, the interpretation of \REF{ex:sun:48} is ambiguous, given the sensitivity of reduplicated interrogatives to negation. 

\ea\label{ex:sun:48} 
    \gll  Lami \textbf{teu} \textbf{aya} \textbf{na-naon} urang tiasa papendak deui énjing.\\ 
          if \textsc{neg} \textsc{ex} \textsc{red}{}-what person.1\textsc{pl} can meet again tomorrow\\
    \glt ‘If it is no problem (with you), we can meet again tomorrow.’ / ‘If there is nothing going on, we can meet again tomorrow.’
\z 

Negative tags can also result in ambiguity given that they are of the same form as the standard negator. In \REF{ex:sun:49a} the negator can either be interpreted as a tag or as a standard negator with scope only over the predicate that follows it. Meanwhile, \REF{ex:sun:49b} unambiguously has sentential scope. 

\ea\label{ex:sun:49}
    \ea\label{ex:sun:49a}
        \gll  Manéhna resep ngi-num téh \textbf{henteu} \textbf{maké} gula. \\ 
              3\textsc{sg} like \textsc{act}{}-drink tea \textsc{neg} use sugar\\
        \glt ‘He likes to drink tea, does he not, with sugar.’ / ‘He likes to drink tea without sugar.’
    \ex\label{ex:sun:49b}
        \gll  Manéhna resep nginum téh maké gula \textbf{henteu}?\\
              3\textsc{sg} like \textsc{act}.drink tea use sugar \textsc{neg}\\
        \glt ‘He likes to drink tea with sugar, doesn’t he?’
    \z
\z 

The ambiguity of \REF{ex:sun:49a} is resolved through prosody, including a pause before the negator accompanied by upward intonation if it is a tag.

\subsection{Negative polarity}\label{sec:sun:4.2}

This section examines the contribution of two particles in particular to the polarity of an utterance. The first particle, \textit{ogé/ gé} ‘too, also’, functions as both an additive particle and a focus particle, in the sense of \citet{König1991}. The interpretation of the particle is entirely pragmatically dependent, as demonstrated in \REF{ex:sun:50}. Indeed, it is rarely possible to determine the precise meaning of the particle without knowledge of preceding utterances. 

\ea\label{ex:sun:50} 
    \gll  Anjeuna \textbf{ogé} hoyong masihan kabar ka urang!\\ 
          3\textsc{sg} too want \textsc{act}.give news to person.1\textsc{pl}\\
    \glt ‘He also wanted to tell us something!’ / ‘\textsc{he} wanted to tell us something!’ \citep[184]{Butters2021}
\z 

\textit{Ogé} is sensitive to scalar semantics, where a scale is a set of contrastive expressions of the same category that can be arranged in linear order according to semantic strength (\citealt{Horn1972}, \citealt{Gazdar1979}). The examples in \REF{ex:sun:51} include elicited utterances with \textit{ogé} in both affirmative and negative contexts. While in the affirmative, \textit{ogé} serves as an ordinary additive/focus particle, in the negative it combines with elements that are quantitatively low on a scale like \textit{hiji} ‘one’, \textit{sakeudik} ‘little’, or the indefinite suffix \textit{sa-} to create a highly newsworthy reading. 

\ea\label{ex:sun:51}
    \ea\label{ex:sun:51a}
        \gll  Anjeunna \textbf{ogé} luka saeutik. \\ 
              3\textsc{sg} too hurt little \\
        \glt ‘\textsc{he} was a little hurt.’ / ‘He too was a little hurt.’
    \ex\label{ex:sun:51b}
        \gll  Anjeunna \textbf{henteu} luka \textbf{saeutik} \textbf{ogé}!\\
              3\textsc{sg} \textsc{neg} hurt little too\\
        \glt ‘He wasn’t even a bit hurt!’
    \ex\label{ex:sun:51c}
        \gll  Abdi \textbf{ogé} nyandak \textbf{hiji} pulpén. \\
             1\textsc{sg} too \textsc{act}.bring one pen  \\
        \glt ‘I brought a pen.’ / ‘I too brought a pen.’
    \ex\label{ex:sun:51d}
        \gll  Aduh poho pisan abdi \textbf{teu} nyandak \textbf{hiji} \textbf{pulpen} \textbf{ogé}!\\
              \textsc{dp} forget very 1\textsc{sg} \textsc{neg} bring one pen too\\
        \glt ‘Oh I forgot! I didn’t bring a single pen!’ \citep[184]{Butters2021}
    \z
\z  


 An even more pragmatically marked reading is achieved if the minimal unit expression is fronted, as in \REF{ex:sun:52}. The particle \textit{deui} ‘again’ can precede \textit{ogé} in such instances for greater emphasis. 

\ea\label{ex:sun:52} 
    \gll  \textbf{Sa-pérak} deui \textbf{ogé} manéhna \textbf{teu} boga!\\ 
          \textsc{indf}{}-silver more too 3\textsc{sg} \textsc{neg} have\\
    \glt ‘He doesn’t have a single coin (lit. piece of silver) more!’ \citep[184]{Butters2021}
\z 
A second polarity sensitive item is the additive scalar particle \textit{(a)can,} which, as demonstrated in \sectref{sec:sun:2.1}, is also used to code \textsc{not} \textsc{yet}. \textit{Acan} is productive in negative expressions that co-occur with reduplication. The meaning of examples \REF{ex:sun:53} are the same as the negative expressions that used \textit{ogé} above. 

\ea\label{ex:sun:53}
    \ea\label{ex:sun:53a}
        \gll  Anjeuna teu luka saeutik-eutik acan! \\ 
              3\textsc{sg} \textsc{neg} hurt \textsc{indf}{}-little-\textsc{red} \textsc{phpc}\\
        \glt ‘He wasn’t even a bit hurt!’
    \ex\label{ex:sun:53b}
        \gll  Aduh abdi poho pisan teu nga-bawa pulpen hiji-hiji acan!\\
              \textsc{dp} 1\textsc{sg} forget very \textsc{neg} \textsc{act}{}-bring pen one-\textsc{red} \textsc{phpc}\\
        \glt ‘Oh I forgot! I didn’t bring a single pen!’ \citep[185]{Butters2021}
    \z
\z 

As evident in examples \REF{ex:sun:53} as well as the reduplication of interrogatives in \sectref{sec:sun:3.2}, reduplication is an important means in Sundanese to specify scalar readings. The inherently augmenting semantics of reduplication serve to specify an extreme end on a scale of possibility.

I return to a discussion of the role played by \textit{acan} in ‘{not yet}’ in \sectref{sec:sun:4.6.3}.  


\subsection{Marking of NPs in the scope of negation}\label{sec:sun:4.3}


There do not appear to be any effects on the marking of NPs in the scope of negation. 


\subsection{Reinforcing negation}\label{sec:sun:4.4}


There are several means through which negation is reinforced in Sundanese. The use of polarity sensitive particles was already discussed in \sectref{sec:sun:4.2}. Negative reinforcement can also be achieved through the addition of \textit{pisan} ‘very’ or \textit{meni} ‘very’ which can be used in both affirmative and negative contexts. Examples (\ref{ex:sun:54a}--\ref{ex:sun:54b}) demonstrate \textit{pisan} in negative contexts, and \REF{ex:sun:54c} shows both \textit{pisan} and \textit{meni} used in the same utterance. 

\ea\label{ex:sun:54}
    \ea\label{ex:sun:54a}
        \gll  Abdi \textbf{teu} resep \textbf{pisan} ka manéhna. \\ 
              1\textsc{sg} \textsc{neg} like very to 3\textsc{sg} \\
        \glt ‘I don’t like him at all/much.’
    \ex\label{ex:sun:54b}
        \gll  Abdi \textbf{teu} tiasa hilap \textbf{pisan}.\\
              1\textsc{sg} \textsc{neg} can forget very\\
        \glt ‘I can’t forget at all.’
    \ex\label{ex:sun:54c}
        \gll  \textbf{Meni} \textbf{teu} \textbf{aya} sa-potong omong-an \textbf{pisan} deui dosén éta nyandak murid-murid-na indit. \\
              very \textsc{neg} \textsc{ex} \textsc{indf}{}-slice talk-\textsc{nmlz} very more instructor \textsc{dem} bring student-\textsc{red}-\textsc{def} go \\
        \glt ‘Without saying a single word more, that instructor took his students away.’
    \z
\z 

Negation can also be reinforced through the emphatic marker \textit{teuing} ‘too, overly’, which directly follows verbs or negative lexemes like \textit{tara} ‘never’, demonstrated in \REF{ex:sun:55}. It can be used in both affirmative and negative contexts. 
\ea\label{ex:sun:55}
    \ea\label{ex:sun:55a}
        \gll  Urang \textbf{teu} kudu hariwang \textbf{teuing}.\\ 
              person.1\textsc{pl} \textsc{neg} must worry too\\
        \glt ‘We need not worry too much.’
    \ex\label{ex:sun:55b}
        \gll  Teu nyaho \textbf{teuing}!\\
              \textsc{neg} know too \\
        \glt ‘I know absolutely nothing!’
    \ex\label{ex:sun:55c}
        \gll  Tara \textbf{teuing}!\\
              never too\\
        \glt ‘Never ever!’
    \z
\z 

Another construction is the postponement of the verbal negator \textit{(hen)teu} to the final position in the utterance, as demonstrated in \REF{ex:sun:56a}. The scalar additive particle \textit{ogé} ‘too’ focuses the proposition under consideration, \textit{geulis} ‘beautiful’, which is then rejected resoundingly. This is in contrast with the ordinary negation of attribution, as shown in \REF{ex:sun:56b}, which does not carry the same pragmatic force.

\ea\label{ex:sun:56}
    \ea\label{ex:sun:56a}
        \gll  Manéhna geulis ogé \textbf{henteu}!\\ 
              3\textsc{sg} beautiful too \textsc{neg}\\
        \glt ‘She’s not even beautiful!’
    \ex\label{ex:sun:56b}
        \gll Manéhna \textbf{henteu} geulis. \\
              3\textsc{sg} \textsc{neg} beautiful\\
        \glt ‘She is not beautiful.’
    \z
\z 

 Finally, negation is reinforced through the form \textit{sama sakali} ‘at all’, which appears only in negative or downward entailing environments. The form can occur before the negator, as in \REF{ex:sun:57}, but also in sentence initial position, sentence final position, or after the verb. 

\ea\label{ex:sun:57} 
    \gll  Abdi \textbf{sama} \textbf{sa-kali} \textbf{teu} nga-gantung ka manéhna.\\ 
          1\textsc{sg} same \textsc{indf}{}-time \textsc{neg} \textsc{act}{}-hang to 3\textsc{sg}\\
    \glt ‘I don’t depend on her at all.’
\z 

The form \textit{sama sekali} exists in Standard Indonesian as well. As with the softening device, \textit{kirang} ‘less’ mentioned in \sectref{sec:sun:2.6}, the direction of influence is unclear. 

\subsection{Negation, coordination and complex clauses}\label{sec:sun:4.5}

Two nominal phrases are typically coordinated simply by occurring side-by-side, as in \REF{ex:sun:58}. 

\ea\label{ex:sun:58}
    \ea\label{ex:sun:58a}
        \gll  Manéhna \textbf{sanés} lalaki manéhna awéwé.\\ 
              3\textsc{sg} \textsc{neg} man 3\textsc{sg} woman\\
        \glt ‘He is not a man, he is a woman.’
    \ex\label{ex:sun:58b}
        \gll Manéhna lalaki \textbf{sanés} awéwé. \\
              3\textsc{sg} woman \textsc{neg} man\\
        \glt ‘He is a man, not a woman.’
    \z
\z 

In \sectref{sec:sun:2.3}, I demonstrated that the use of nominal negation with attributes encodes contrastive negation. A second strategy for marking contrastive negation employs the verbal negator in the first proposition and \textit{malah} ‘even’ in the second, as in \REF{ex:sun:59}. 

\ea\label{ex:sun:59}
    \ea\label{ex:sun:59a}
        \gll  Manéhna \textbf{teu} manggil polisi manéhna \textbf{malah} manggil kepala désa.\\ 
              3\textsc{sg} \textsc{neg} call police 3\textsc{sg} even \textsc{act}.call head village\\
        \glt ‘S/he didn’t call the police, but rather the village head.’
    \ex\label{ex:sun:59b}
        \gll  Putri \textbf{henteu} se-seurian \textbf{malah} jadi pucet. \\
              Putri \textsc{neg} \textsc{red}{}-laugh even become pale\\
        \glt ‘Putri didn’t laugh, but rather turned pale.’
    \z
\z 


 Expressions of the type \textit{neither … nor} are formulated with the co-occurrence of the negator and the comitative \textit{sareng} ‘with, and’, as in \REF{ex:sun:60a}, which is contrasted with the use of \textit{sareng} in an affirmative utterance, as in \REF{ex:sun:60b}. 

\ea\label{ex:sun:60}
    \ea\label{ex:sun:60a}
        \gll  Abdi mah \textbf{teu} gaduh ketan bodas \textbf{sareng} ketan hideung.\\ 
              1\textsc{sg} \textsc{dp} \textsc{neg} have sticky.rice white \textsc{com} sticky.rice black\\
        \glt ‘I have neither white sticky rice, nor black sticky rice.’
    \ex\label{ex:sun:60b}
        \gll  Abdi mah gaduh ketan bodas \textbf{sareng} ketan hideung.\\
              1\textsc{sg} \textsc{dp} have sticky.rice white \textsc{com} sticky.rice black\\
        \glt ‘I have white sticky rice and black sticky rice.’
    \z
\z 

\subsection{Miscellaneous aspects of negation}\label{sec:sun:4.6}

\subsubsection{Negative transport}\label{sec:sun:4.6.1}

Negative transport (that is, negative-raising) occurs with verbs of desire like \textit{hayang} ‘want, wish to’ (or the polite form, \textit{hoyong}), as demonstrated in \REF{ex:sun:61}. 

\ea\label{ex:sun:61}
    \ea\label{ex:sun:61a}
        \gll  Urang teu hayang manéhna melak paré.\\ 
              person.1\textsc{sg} \textsc{neg} want 3\textsc{sg} plant paddy\\
        \glt ‘I don’t want him to plant paddy.’
    \ex\label{ex:sun:61b}
        \gll  Urang hayang manéhna teu melak paré.\\
              person.1\textsc{sg} want 3\textsc{sg} \textsc{neg} plant paddy\\
        \glt ‘I want him to not plant paddy.’ 
    \z
\z 

\subsubsection{Additional functions of the prohibitive}\label{sec:sun:4.6.2}


The prohibitives \textit{ulah} and \textit{tong} perform functions beyond simply marking the negative imperative described in \sectref{sec:sun:2.2}. One such function is to reject an option from a series of possible choices. For instance, \REF{ex:sun:62a} was uttered when a speaker rejected the author’s proposal of a meeting time, while \REF{ex:sun:62b} was said during a conversation between two women when a yellow sarong (chosen from a collection of sarongs) was rejected for purchase. 

\ea\label{ex:sun:62}
    \ea\label{ex:sun:62a}
        \gll  \textbf{Ulah} wengi ieu. \\ 
              \textsc{proh} night \textsc{dem}\\
        \glt ‘Not tonight.’
    \ex\label{ex:sun:62b}
        \gll  \textbf{Ulah} (anu) konéng moronyoy teuing. \\
              \textsc{proh} \textsc{rel} yellow bright too \\
        \glt ‘Not the yellow (one), it’s too bright.’ \citep[152]{Butters2021}
    \z
\z 

Additionally, the prohibitive \textit{ulah} can be embedded in an utterance where it behaves more like a standard negator. In \REF{ex:sun:63a} it is fronted by a proposition expressing deontic modality, and in \REF{ex:sun:63b} it follows the coordinator \textit{asal} ‘as long as’. 

\ea\label{ex:sun:63}
    \ea\label{ex:sun:63a}
        \gll  Langkung saé manéhna \textbf{ulah} ng-ereun-keun kécap batur.\\ 
              more good 3\textsc{sg} \textsc{proh} \textsc{act}{}-interrupt-\textsc{appl} words others\\
        \glt ‘It’s better that he doesn’t interrupt others.’ \citep[155]{Butters2021}
    \ex\label{ex:sun:63b}
        \gll  Manéhna natap ka mana waé asal \textbf{ulah} ka arah Udin. \\
              3\textsc{sg} look to where just as.long.as \textsc{proh} to direction Udin\\
        \glt ‘He looked anywhere except at Udin.’ / ‘He looked anywhere but at Udin.’ \citep[156]{Butters2021}
    \z
\z 

A final special construction is the use of the prohibitive \textit{ulah} with the particle \textit{ogé} to mean ‘never mind X’, as demonstrated in \REF{ex:sun:64}. Note that in rural areas of West Java where the expression was uttered, cassava is much more widely available than water spinach. 

\ea\label{ex:sun:64} 
    \gll  \textbf{Ulah} \textbf{gé} sayur kangkong singkong \textbf{ogé} abdi teu boga!\\ 
          \textsc{proh} too vegetable spinach casava too 1\textsc{sg} \textsc{neg} have\\
    \glt ‘Never mind water spinach, I don’t even have cassava!’ \citep[159]{Butters2021}
\z 

\subsubsection{Diachronic notes and observations}\label{sec:sun:4.6.3}

 This chapter has demonstrated the presence of a diverse set of negators with multivariant functions. In the interest of space, the focus of this diachronic section will be on two areas of Sundanese negation discussed thus far: 1) the productive distinction between the verbal and nominal negator, and 2) the development of ‘{not yet}’. 

The distinction between the Sundanese verbal negator, \textit{(hen)teu,} and the nominal negators, \textit{lain} or \textit{sanés}, was described earlier in the chapter. The distinction is presented again here for clarity.  

\ea\label{ex:sun:65}
    \ea\label{ex:sun:65a}
        \gll  Manéhna \textbf{(hen)teu} leutik.\\ 
              3\textsc{sg} \textsc{neg} small\\
        \glt ‘S/he is not small.’
    \ex\label{ex:sun:65b}
        \gll  Manéhna \textbf{lain} lalaki leutik.\\
              3\textsc{sg} \textsc{neg} man small \\
        \glt ‘He is not a small man.’
    \z
\z 

The existence of a distinction between verbal and nominal negators is extremely common in Indonesian languages. In Sundanese, the nominal negators have the additional transparent lexical meaning of ‘other, different’ – a use that is retained in contemporary speech, as demonstrated in \REF{ex:sun:66}. It is also attested in verbs like \textit{nga}\textbf{\textit{lain}}\textit{keun} ‘to other, to give the cold shoulder to’. 

\ea\label{ex:sun:66} 
    \gll  Éta mah \textbf{lain} deui.\\ 
           \textsc{dem} \textsc{dp} different again\\
    \glt ‘That’s entirely different.’ \citep[61]{Butters2021}
\z 
The grammaticalization of the lexeme ‘other’ to a negator of nominal predicates is widely attested in other Western Malayo-Polynesian languages – an understandable development given that to be ‘other than X’ is equal to ‘not be X’. The use of \textit{lain} as a negator was already present in the Old Sundanese manuscript, \textit{The sons of Rama and Rawana}, which probably dates from around the 16\textsuperscript{th} century. An example from this manuscript is shown in \REF{ex:sun:67}, taken from the transliteration of this work. 

\ea\label{ex:sun:67} 
    \gll  Lun aing \textbf{lain} na seuweu.\\ 
          if 1\textsc{sg} \textsc{neg} \textsc{lig} son\\
    \glt ‘If I’m not your son.’ (\citealt{NoorduynTeeuw2006}: 206).
\z 

It is difficult to determine precisely how long the distinction between the nominal and the verbal negator has existed, but its presence can be accounted for by the different functions performed by the two sets of negators. The nominal negator possesses the special function of encoding contrastive negation, which is less effectively encoded by the verbal negator. Moreover, the use of \textit{lain} and \textit{sanés} as a rhetorical negative tag entails prior expectation on the part of the speaker. There are, therefore, important semantic and pragmatic distinctions between these negators to maintain this distinction. 

 A second area worthy of greater discussion is the development of ‘{not yet}’ in Sundanese, which, as presented in \sectref{sec:sun:2.1}, can be encoded by \textit{teu (a)can, tacan,} or \textit{can} alone, as demonstrated again for clarity in \REF{ex:sun:68}. 

\ea\label{ex:sun:68} 
    \gll  Manéhna \textbf{tacan} saré.\\ 
          3\textsc{sg} not.yet sleep\\
    \glt ‘He has not slept yet.’
\z

In \sectref{sec:sun:4.2}, I presented a function of \textit{(a)can} beyond its use in \textsc{not} \textsc{yet}; like \textit{ogé, (a)can} has additive functions and can be used to reinforce negation in highly newsworthy contexts. Unlike \textit{ogé, (a)can} is sensitive to reduplication, as demonstrated again in \REF{ex:sun:69}. 

\ea\label{ex:sun:69}
    \ea\label{ex:sun:69a}
        \gll  Teu bisa nulis-nulis \textbf{acan}!\\ 
              \textsc{neg} can \textsc{act}.write-\textsc{red} \textsc{phpc}\\
        \glt ‘(I) can’t even write!’
    \ex\label{ex:sun:69b}
        \gll  Teu apal laptop-laptop \textbf{acan}!\\
              \textsc{neg} know laptop-\textsc{red} \textsc{phpc}\\
        \glt ‘(I) don’t even know what laptops are!’
    \z
\z 

 When the preceding element is not reduplicated, as in \REF{ex:sun:70b}, the reinforcing particle \textit{pisan} is used. 

\ea\label{ex:sun:70}
    \ea\label{ex:sun:70a}
        \gll  Abdi teu bisa dahar-dahar acan.\\ 
              1\textsc{sg} \textsc{neg} can eat{}-\textsc{red} \textsc{phpc}\\
        \glt ‘I couldn’t even eat.’
    \ex\label{ex:sun:70b}
        \gll  Abdi teu bisa dahar \textbf{pisan}.\\
             1\textsc{sg} \textsc{neg} can eat very \\
        \glt ‘I couldn’t eat at all.’ \citep[93]{Butters2021}
    \z
\z 

\citet[2]{Hardjadibrata2003} suggests that when \textit{acan} occurs in affirmative contexts following reduplication, the interpretation is that a situation still holds true, as in \REF{ex:sun:71},\footnote{It should be noted that \citet{Hardjadibrata2003} is a translation of \citet{Eringa1984}, written in Dutch. The expression is thus older than would be suggested by the publication date.} with the author’s own glossing. 

\ea\label{ex:sun:71} 
    \gll Manéhna di sawah-sawah \textbf{acan}. \\ 
          3\textsc{sg} \textsc{loc} paddy.field-\textsc{red} \textsc{phpc} \\
    \glt ‘He is still in the rice fields.’ \citep[2]{Hardjadibrata2003}
\z 

Consultants do not accept \REF{ex:sun:71} but are fine with the same expression occurring in the negative, as in \REF{ex:sun:72}, where the understanding is that someone has broken their daily expected habit of going to the paddy fields. 

\ea\label{ex:sun:72} 
    \gll  Manéhna \textbf{teu} di sawah-sawah \textbf{acan}.\\ 
          3\textsc{sg} \textsc{neg} \textsc{loc} paddy.field-\textsc{red} \textsc{phpc} \\
    \glt  ‘He is not even in the paddy fields.’ (He can’t be seen there at all) \citep[93]{Butters2021}
\z 

\citet[170]{KönigTraugott1982} described the close historical relationship between ‘still’ and ‘yet’ in English, defining ‘still’ as “continuation without boundary” and ‘yet’ as “continuation up to an imminent boundary/change”. The authors suggest that the shared quality of continuation has resulted in frequent overlap in meanings in English but not true convergence. In a study of phasal polarity in Malayo-Polynesian languages, \citet[15]{VeselinovaEtAlFC} observe the frequent presence of periphrastic expressions of ‘not yet’ consisting of internally negated ‘still’ expressions, noting that such constructions “express the continuation of a non-realized situation and as such demonstrate the direct relation between the concepts of ‘{still}’ and ‘{not yet}’ as well as their encoding. In many languages what had initially been a periphrastic expression became fused into univerbations with various degrees of transparency.”

The examples presented in this chapter for the Sundanese \textit{acan} also appear to suggest the direct relationship between ‘{still}’ and ‘{not yet}’. An additional clue to the early meaning of \textit{acan} is evident in an archaic entry in Kamus Besar Bahasa Indonesia (KBBI) for the Indonesian \textit{mengacan} meaning ‘to hope for’ or ‘to aspire to’. The appearance of \textit{acan} in a closely related language reflects the early use of the word for an unrealized state of affairs, further elucidating the relationship between ‘still not’ and ‘not yet’. 

\section{Summary}\label{sec:sun:5}


\begin{table}%this is not the best-looking code but it works as it should so I'll leave it as it is
\caption{Sundanese negators and functions}
\label{tab:sun:2}
\begin{tabularx}{\textwidth}{lX}
\lsptoprule
{Negator}  & {Function} \\
\midrule
{(\textit{hen})\textit{teu}} & {standard negation}\newline
    {‘not anymore’ [\textsc{neg} ‘again’]}\newline     
    {‘not yet’ [\textsc{neg} scalar additive particle]}\newline
    {‘never’ [\textsc{neg} perfect]}\newline
    {neutral negative tag}\newline
    {negator of attributes}\newline
    {‘no’}\newline
    {negative quantifier [\textsc{neg} \textsc{inter}-\textsc{red}]}\newline
    {‘without’ [\textsc{neg} \textsc{com}] or [\textsc{neg} ‘use’]}\\
{\textit{lain}, \textit{sanés}} & {negator of equative predications}\newline
    {negator of proper inclusion}\newline
    {rhetorical negative tag}\newline
    {negative reply “no”} \\
{\textit{moal}}  & {negative future}\\
{\textit{tara}}  & {‘never’}\\
{\textit{teu acan}, \textit{tacan}, \textit{can}} & {‘not yet’}\newline
    {negative polarity item} \\
{\textit{alim}} & {‘not want’}\\
{\textit{duka}}  & {‘not know, who knows’}\\
{\textit{ulah}, \textit{tong}} & {prohibitive} \newline
    {rejection of a choice} \newline
    {standard negation in embedded clauses}\\
{\textit{teu aya}, \textit{taya}, \textit{euweuh}}  & {negative existential predication}\newline
    {locative predication (optional)}\newline
    {locative presentative predication}\newline
    {possessive predication} \newline
    {negative quantifier} \newline
    {negative derivation} \\
{\textit{kirang}}  & {standard negation (polite)}\\
\lspbottomrule
\end{tabularx}
\end{table}

The negative forms and functions discussed in this chapter are summarized in \tabref{tab:sun:2}.
Key points of this chapter include the demonstration that Sundanese negation makes a productive distinction for verbal vs. nominal predicates, for certain areas of tense and aspect, and for certain realms of cognition and volition. Additionally, Sundanese possesses special prohibitive forms which participate in environments beyond imperative mood. Existential, locative, and possessive predication all make use of the same existential and negative existential predicators, though such a use is only obligatory in the former category, while it is optional in the latter categories. Polarity phenomena in Sundanese rely on reduplication, specifically the reduplication of interrogatives, as well as the use of scalar additive particles and downward entailing particles. Future work might devote more attention to the precise source of some of the negators discussed herein, drawing from a larger body of Old Sundanese manuscripts. 

\section*{Acknowledgments}
This research was funded by a Boren Fellowship (2018) and a Luce Fellowship from AIFIS, The American Institute for Indonesian Studies (Spring 2020). I am extremely thankful to all who patiently assisted me during the course of this research, especially Muchlis Julianda Situmorang. The chapter also benefited from the helpful comments of anonymous reviewers.

\section*{Abbreviations}
\begin{tabularx}{.45\textwidth}{lQ}
{\scshape 1} & first person\\
{\scshape 2} & second person\\
{\scshape 3} & third person\\
{\scshape appl} & applicative\\
{\scshape act} &  active \\
{\scshape comp} & complementizer\\
{\scshape com} & comitative\\
{\scshape def} & definite\\
{\scshape dem} & demonstrative\\
{\scshape dp} & discourse particle\\
{\scshape ex} & extistential \\
{\scshape fut} & future\\
{\scshape hon} & honorific\\
{\scshape hort} & hortative\\
{\scshape iam} & iamitive\\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
{\scshape indf} & indefinite\\
{\scshape lig} & ligature\\
{\scshape loc} & locative\\
{\scshape neg} & negative\\
{\scshape nmlz} & nominalizer\\
\scshape phpc & phasal poparity with continuity \\
{\scshape pl} & plural\\
{\scshape poss} & possessive\\
{\scshape prf} & perfective\\
{\scshape proh} & prohibitive\\
{\scshape pass} & passive \\
{\scshape red} & reduplication\\
{\scshape rel} & relative\\
{\scshape sg} & singular\\
\end{tabularx}



\sloppy\printbibliography[heading=subbibliography,notkeyword=this]

\end{document}
