\documentclass[output=paper,colorlinks=true, citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18233962}
\author{Riho Grünthal\affiliation{University of Helsinki}}
\title{Negation in Veps}
\abstract{This chapter presents an overview of different negation strategies and diachronic changes in Veps, an endangered northwestern Uralic language belonging to the Finnic branch. Basic negation patterns in Veps share many characteristics with other northern Finnic languages. However, Veps displays a negation system with many innovations and manifests several language-specific morphosyntactic patterns, a lot of syntactic variation, and secondary lexicalization in negative phrases. Several features and syntactic changes specific to Veps, such as number agreement between the negative auxiliary and the verb in plural, alternating word order, and negative existentials increase the divergence of Veps negation in comparison with neighboring and related languages. Language contacts play a significant role as well, as lexical negation and indefinites involve a strong Russian influence, whereas in clausal negation it is not as significant. 

\keywords{Veps, Finnic, negation, language contact, Russian, scope of negation}
}

\IfFileExists{../localcommands.tex}{
  \addbibresource{../localbibliography.bib}
  \input{../localpackages}
  \input{../localcommands} 
  \input{../localhyphenation} 
  \togglepaper[1]%%chapternumber
}{}

\begin{document}
\maketitle 


\section{The language} \label{sec:vep:1}

Veps (ISO: \href{https://iso639-3.sil.org/code/vep}{vep}, Glottocode: \href{https://glottolog.org/resource/languoid/id/veps1250}{veps1250}) is the easternmost Finnic language spoken in Northwestern Russia in the Republic of Karelia, Leningrad oblast, and Vologda oblast. The historical core area is located between the lakes Ladoga, Onega, Beloe, and along the shores of river Oyat’. Veps is more heavily influenced by Russian than other Finnic languages and long-term bilingualism involves both intensive lexical and grammatical borrowing. The population size has constantly decreased since 1930s, which is caused by the lack of sustainable development, insufficient political rights and an unfavorable position of minority languages in the Soviet Union and Russia. The estimated current number of speakers is less than 2500\footnote{The last population census of Russia was carried out in 2021. The data were published at the end of 2022. The official statistics show a decrease in the number of ethnic Veps from over 5900 in 2010 to over 4300 in 2021 \citep{Prozes2023}. The number of Veps speakers is critically assessed in literature (\citealt{Grünthal2011}; \citealt[23--32]{PuuraEtAl2013}; \citealt{Strogal’shchikova1989, Strogal’shchikova2006}), which shows that it is considerably smaller than the number of ethnicity.} and they mainly belong to elder generations with a rural background (\citealt{Grünthal2011}; \citealt[23--32]{PuuraEtAl2013}; \citealt{Strogal’shchikova1989, Strogal’shchikova2006}). In the beginning of 1990s a new Veps standardized language was introduced. It is regularly used in a monthly newspaper and more randomly in media and other printed publications. 

A number of linguistic works describing the Veps language and the history of the Veps people have been published since the 19\textsuperscript{th} century, the first linguistic descriptions dating back to 1840s and 1850s  (\citealt{Branch1973}: 83–92; \citealt{Grünthal2015}: 18–24; \citealt{Karjalainen2016}; \citealt{Korhonen1986}: 43–45, 76–80; \citealt{Puura2019}; \citealt{PuuraEtAl2013}: 2–8; \citealt{Strogal’shchikova2008}; \citealt{Zaitseva1981, Zaitseva2001}). The dialects are divided into Northern, Central and Southern Veps, which are mutually easily intelligible, the Southern dialect having more distinctive features than the two other ones.

Typologically, the Veps language displays abundant suffixal morphology and has clearly distinct nominal and verbal paradigms. It has 17–18 inflectional cases and distinguishes between two conjugation types as it displays reflexive conjugation that diverges from the basic paradigm (\citealt{Grünthal2015, Grünthal2017}: 133–135; \citealt{MSFOu86}: 421–434; \citealt{Zaitseva1981, Zaitseva2001}). It manifests much less morphophonological alternation than other Finnic and Saamic languages and lacks consonant gradation typical of these two Uralic branches. Language contact with Russian has influenced word order, resulting in variation between the Finnic default SVO pattern and the inherited Uralic SOV order, which is secondary in Veps and used for pragmatic purposes. Word order variation is a language-specific characteristic of Veps negation in declarative clauses (see \sectref{sec:vep:2.1.1} and \sectref{sec:vep:2.1.2}). 

The considerably high amount of variation and divergence in Veps negation shows that negative phrases are liable to morphosyntactic changes. Like other northern Finnic languages, Veps does not distinguish between present and past tense negative auxiliary, which is characteristic of more eastern Uralic languages, such as Mordvin and Mari (\citealt{HamariAasmäe2015}; \citealt{Saarinen2015}) and has been preserved as a recessive archaic feature in South Estonian, Insular dialects of Estonian, and Livonian. Language-specific changes have secondarily increased the means of negation in Veps, which displays strongly divergent features in plural negation and the fusional negative existential. (For negation in other Finnic languages, cf. \citealt{LindströmEtAl2021}; \citealt{MetslangEtAl2015}; \citealt{RozhanskiMarkus2017}; \citealt{Vilkuna2015, Vilkuna2022}.)

In this paper, data illustrating Veps negation are drawn from published text collections (\citealt{Kettunen1920,Kettunen1925,MSFOu86}; \citealt{MSFOu70}) representing spoken language from the first half of the 20\textsuperscript{th} century, and the author’s fieldtrips between 2005--2014.\footnote{Data from these fieldtrips are marked with the village and year in which the data were collected in parentheses after the translation of the example.} Examples from vernacular speech are presented in a phonologically standardized form lacking diacritics.  

The description of negation in Veps in this chapter is based on the questionnaire provided by the editors \parencitetv{chapters/questionnaire} and the structure of the chapter follows the model given in the questionnaire.

\section{Clausal negation}\label{sec:vep:2}

\subsection{Standard negation}\label{sec:vep:2.1}
\subsubsection{Basic declarative sentence}\label{sec:vep:2.1.1}

Verb conjugation distinguishes between present and simple past tenses, and, most notably in the standardized language, two compound past tenses (\citealt{Zaitseva2000, Zaitseva2001}: 107–120). In colloquial speech and dialects, however, the system often operates with a simplified contrast between two tenses, namely present and simple past (\citealt{Grünthal2015}: 120–149; \citealt{MSFOu86}: 62–63, 76–77). The same negative auxiliary is used both in present and past tense inflecting for person as in most other Finnic languages. In present tense, the main verb displays an unmarked connegative form (\ref{ex:vep:1}--\ref{ex:vep:2}), whereas past tense is indicated with the past tense participle \REF{ex:vep:3}. The copula \textit{lii-da} be.\textsc{fut-inf} ‘be (fut.)’ is the only verb that indicates future, whereas the declarative negative clause is identical with present tense (\ref{ex:vep:4}--\ref{ex:vep:5}). The default word order is \textsc{neg} + verb.

\ea
    \label{ex:vep:1}
    \gll  Mams  se  sanu{}-i:  \textbf{e{}-n} \textbf{rohti} libu-da. \\
         mother  it say-\textsc{pst.3sg} \textsc{neg-1sg} dare\textsc{.cng} arise-\textsc{inf}\\
    \glt ‘The mother said: I daren’t arise.’ (VeC, \citealt[32]{MSFOu70})
\z    

\ea
    \label{ex:vep:2}
    \gll \textbf{E-taa}   \textbf{mahta} išt-taaze. \\
         \textsc{neg{}-}2\textsc{pl}   can.\textsc{cng} sit-\textsc{inf.refl} \\
    \glt ‘You can’t sit down.’ (VeS, \citealt[116]{Kettunen1920})
\z

\ea
    \label{ex:vep:3}
    \gll Kana-š-t, sanu-b, \textbf{e-d} \textbf{to-nd}. \\
         chicken-\textsc{dim{}-par} say-\textsc{3sg}  \textsc{neg{}-2sg} bring-\textsc{pst.ptcp}\\
    \glt ‘You didn’t bring a chicken, (s)he says.’ (VeC, \citealt[92]{MSFOu70})
\z

\ea
    \label{ex:vep:4}
    \gll Ma-d  jaugoo-d’en aa \textbf{ii} \textbf{muja-nd}. \\
         soil-\textsc{par} foot{}-\textsc{pl{}-gen} under \textsc{neg.3} feel-\textsc{pst.ptcp} \\
    \glt ‘(S)he didn’t feel the soil under her/his feet.’ (VeS, \citealt[83]{Kettunen1920})
\z

\ea
    \label{ex:vep:5}
    \gll Nece-n maksa-n taga \textbf{ii} \textbf{liine} annie  mehe-l. \\
         this\textsc{{}-gen} friend\textsc{{}-gen} behind \textsc{neg.3} \textsc{fut}.\textsc{cng} Annie man-\textsc{ade}\\
    \glt ‘Annie will not marry this friend.’ (VeC, \citealt[166]{MSFOu70})
\z 

    
In affirmative clauses the main verb inflects for both person \REF{ex:vep:6} and tense \REF{ex:vep:7}. In the negative phrase person marking is transferred to the auxiliary as indicated above, whereas in past tense the main verb has a past tense participle affix (\ref{ex:vep:3}--\ref{ex:vep:4}) instead of the primary past tense marker.

\ea\label{ex:vep:6}
    \gll Mina tabada-n tii-d da riko-n. \\
          \textsc{1sg} catch-\textsc{1sg} \textsc{2pl}-\textsc{par} and kill-\textsc{1sg}\\
    \glt ‘I will catch and kill you!’ (VeC, \citealt[39]{MSFOu70})
\z 

\ea
    \label{ex:vep:7}
    \gll El-i-maa  nel’ vo-t hän-mu.\\
          live-\textsc{pst-1pl} four year-\textsc{par} \textsc{3sg-com}\\
    \glt ‘We lived four years with him.’ (VeS, \citealt[5]{MSFOu70})
\z 

Structurally a basic affirmative declarative clause and its negative pair are asymmetric. The morphosyntactic pattern of the negative VP diverges from the affirmative one as the person marker shifts to the inflecting negative auxiliary (\textsc{neg}.\textsc{aux}) in negative phrases. This happens both in the subjective (basic) and the reflexive paradigm (\tabref{tab:vep:1})\footnote{Historically, present tense \textsc{3pl} forms originate from an impersonal form labeled as “passive” in the grammatical descriptions of various Finnic languages. However, vernacular Veps does not make the distinction between \textsc{3pl} and passive form because both can be used in impersonal functions. Paradigmatically, this has caused the merger of \textsc{3pl} and passive paradigms. In this article, historical passive forms that mostly represent Northern and Central Veps are encoded as \textsc{3pl}, whereas in Ludic the glossing \textsc{pa}\textsc{ss} is applied.}.  However, reflexive verbs display a morphologically split strategy in negation as the auxiliary adopts the regular person marker, whereas the lexical verb manifests a distinct reflexive stem for connegative. In most dialects connegative forms index plural. In contrast to declarative clauses, the affirmative and negative imperative clauses are, however, symmetric (see \sectref{sec:vep:2.2}).

\begin{table}
\centering\small 
\begin{tabularx}{\textwidth}{lllll}
\lsptoprule
 & \multicolumn{2}{l}{subjective conjugation}  &  \multicolumn{2}{l}{reflexive conjugation}    \\
& \multicolumn{2}{l}{teh-ta (do\textsc{{}-inf})}  & \multicolumn{2}{l}{pes-tas (wash\_oneself\textsc{{}-inf})}   \\
\midrule 
\textsc{1sg} & tege-n & e-n tege & peze-moi & {e-n peste}\\
\textsc{2sg} & tege-d & e-d tege & peze-toi & {e-d peste}\\
\textsc{3sg} & tege-b & ii /ei tege & peze-soi / peze-se & {ii /ei peste}\\
\textsc{1pl} & tege-m & e-m / e-mei teh-koi & peze-moiš & {e-m peste / e-mei pes-koiš} \\
\textsc{2pl} & tege-t & e-d / e-tei teh-koi & peze-toiš & {e-d peste / e-tei pes-koiš} \\
\textsc{3pl} & teh-tas & ii / ei teh-koi & peze-soiš & {ii peste / ei pes{}-koiš} \\
\lspbottomrule
\end{tabularx}
    \caption{Affirmative and negative paradigm of the verbs \textit{tehta} ‘do’ and \textit{pestas} ‘wash (oneself)’ in Central Veps \citep[254, 267--268]{Zaitseva2001}}
    \label{tab:vep:1}
\end{table}

A similar negation pattern is used in simple past tense clauses. Given that there is no past tense \textsc{neg}.\textsc{aux}, tense is marked on the lexical verb both in affirmative \mbox{(\textit{{}-i-}}) and negative clauses that have a past tense participle (\textit{{}-nu}(\textit{d}) {\textasciitilde} \textit{{}-nd}) like other Finnic languages (\tabref{tab:vep:2}). The morpheme order in past tense negative forms is \textsc{neg-person} V{}-\textsc{pst}{}-\textsc{pl} as in \REF{ex:vep:8additional} and \REF{ex:vep:9additional}.
Connegative forms agree in number (plural) with inflecting plural \textsc{neg}.\textsc{aux} (see \sectref{sec:vep:2.1.3}).

\ea\label{ex:vep:8additional}
    \gll E-mei teh-nu-goi.  \\
        \textsc{neg{}-1pl} do-\textsc{pst}{}-\textsc{pl}\\
    \glt ‘We didn’t do.’ \\
\z 

\ea\label{ex:vep:9additional}
    \gll E-m pez-nu-ks-oi(š).\\
        \textsc{neg{}-1pl} wash-\textsc{pst-refl}{}-\textsc{pl}\\
    \glt ‘We didn’t wash ourselves.’\\
\z 



\begin{table}
\centering
\begin{tabularx}{\textwidth}{lllll}
\lsptoprule
& \multicolumn{2}{l}{subjective conjugation} &  \multicolumn{2}{l}{reflexive conjugation} \\
& \multicolumn{2}{l}{teh-ta (do\textsc{{}-inf})} &  \multicolumn{2}{l}{pes-tas (wash\_oneself\textsc{{}-inf})} \\
\midrule
\textsc{1sg} & teg-i-n & {e-n tege-n(d)} & pez-i-moi & e-n pez-nu-kse \\
&&&& e-n pez-nu-s \\
\textsc{2sg} & teg-i-d’ & {e-d tege-n(d)} & pez-i-toi &{e-d pez-nu-kse} \\
&&&& e-d pez-nu-s \\
\textsc{3sg} & teg-i & {ii tege-n(d)} & pez-i-he & {ii pez-nu-kse} \\
&&&&  ei pez-nu-s \\
\textsc{1pl} & teg-i-m & {e-m teh-nuhu} & pez-i-moiš & {e-m pez-nu-ks-oi(š)} \\
& teg-i-mei & e-mei teh-nu-goi && e-mei pez-nu-soiš \\
\textsc{2pl} & teg-i-d’& {e-d teh-nuhu} & pez-i-toiš & {e-d pez-nu-ks-oi(š)} \\
& teg-i-tei & e-tei teh-nu-goi && e-tei pez-nu-soiš \\
\textsc{3pl} & teg-i-ba & {ii teh-nuhu} & pez-i-(he-)zoiš & {ii pez-nu-ks-oi(š)} \\
&& ei teh-nu-goi && ei pez-nu-soiš \\
\lspbottomrule
\end{tabularx}
\caption{Affirmative and negative simple past tense paradigm of the verbs \textit{tehta} ‘do’ and \textit{pestas} ‘wash (oneself)’ in Eastern and Western Central Veps \citep[254--255, 268]{Zaitseva2001}}
\label{tab:vep:2}
\end{table}

Occasionally negation may trigger a specific connegative form which, contrary to affirmative declarative sentences, displays a derivative suffix {}-\textit{ške} (\ref{ex:vep:8}--\ref{ex:vep:9}) (cf. below \REF{ex:vep:11} and \REF{ex:vep:69}). This particular suffix typically occurs only in \textsc{2sg} imperative and connegative, which are morphologically unmarked categories. Here, it is labeled as the marker of perfective aspect.

\ea
    \label{ex:vep:8}
    \gll E-n  \textbf{probi-ške}, ii \textbf{ole-ške} mugoš-t. \\
         \textsc{neg-1sg}  try-\textsc{pfv} \textsc{neg.3} be-\textsc{pfv} such-\textsc{par}\\
    \glt ‘I didn’t try, there wasn’t such a thing.’ (VeC, Mäggär’v 2007)
\z 

\ea
    \label{ex:vep:9}
    \gll Ningoš{}-t igas e{}-bad \textbf{näg-i-ške} durako{}-d. \\
         such-\textsc{par} ever \textsc{neg{}-3pl} see-\textsc{pst-pfv} idiot-\textsc{par}\\
    \glt ‘They had never seen such an idiot.’ (VeS, \citealt[71]{Kettunen1920})
\z

\subsubsection{Reversed word order V+\textsc{neg}}\label{sec:vep:2.1.2}
\largerpage
In addition to the default word order \textsc{neg}+V, Veps displays a reversed word order in negative clauses V+\textsc{neg}. The former one is more frequent but the latter one is relatively frequent as well. This kind of alternation is typically associated with emphasis or topical focus in discourse. More generally, the verb characteristically precedes its dependents following the SVO pattern. A sentence-final verb may occur in affirmative clauses shifting the focus to the preceding object or adverbial, which may also occur in sentence-initial position. The sentence-final negative auxiliary shifts the person-marking unit to the last position as in affirmative clauses. 
The scope of negation of the reversed order V + \textsc{neg} is larger as the sentence-final \textsc{neg.aux} scopes over the whole clause (\ref{ex:vep:10}--\ref{ex:vep:13}).

\ea\label{ex:vep:10}
    \gll Mö \textbf{sö-htä-n}  \textbf{e-maa}. \\
         \textsc{1pl} eat\textsc{{}-mom}{}-\textsc{pst.ptcp} \textsc{neg{}-1pl}\\
    \glt ‘We didn’t eat.’ (VeS, \citealt[93]{Kettunen1920})
\z

\ea\label{ex:vep:11}
    \gll Ende bajri-d \textbf{aje-l-i-ške}  \textbf{e-bad} keloo-mu. \\
         Earlier boyar{}-\textsc{pl} ride\textsc{{}-cont-pst-mom} \textsc{neg-3pl} bell{}-\textsc{prol}\\
    \glt ‘Earlier the boyars didn’t ride around with ringing bells.’ (VeS, \citealt[17]{Kettunen1925})
\z

\ea\label{ex:vep:12}
    \gll Hän-dast minä se \textbf{tunde-nu} vuu \textbf{e-n}. \\
         \textsc{3sg.f}.\textsc{acc}  \textsc{1sg} it know-\textsc{pst.ptcp} yet \textsc{neg-1sg} \\
    \glt ‘I didn’t know her, yet.’ (VeC, Ladv 2014)
\z

\ea\label{ex:vep:13}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:vep:13A}
        \gll  A vonuka-d maht{}-tas, elge{}-tas  vepsa-ks?\\ 
               but grandchild{}-\textsc{pl} can-3\textsc{pl} understand-3\textsc{pl} Veps-\textsc{tra}\\ 
        \glt ‘But (your) grandchildren can, understand Veps?'
    \exi{A: }\label{ex:vep:13B}
        \gll  Elge{}-tas,  a \textbf{pagiš-koi} \textbf{ii}. \\ 
               understand{}-3\textsc{pl} but talk-\textsc{cng.pl} \textsc{neg.3}\\
        \glt ‘They understand but don’t speak.’  (VeC, Ladv 2007)
    \end{xlist}
\z

This kind of permutation of constituent order is not limited to the position of the lexical verb and the negative auxiliary. The most fundamental change is seen in the use of the inherited demonstrative pronoun \textit{se} ‘it’, which typically occurs in a prenominal position. However, it is very frequently used as a postnominal focus particle that has undergone both syntactic and semantic change. This postnominal use of the particle is also attested in other Finnic languages and in Northern Russian dialects. In Veps, the frequency is considerably higher than in other Finnic varieties, which correlates with secondary changes in the system of demonstrative pronouns \citep{Yurayong2020}.

The reversed constituent order \textsc{v} + \textsc{neg} instead of \textsc{neg} + \textsc{v} has not been investigated in more detail so far. A parallel instance can be shown in Ludic \REF{ex:vep:14}, an intermediate variety between Karelian and Veps, and in Eastern South Estonian (Seto), in which it is highly frequent \citep{LindströmEtAl2021}.

\ea\label{ex:vep:14} Ludic \newline %(ISO: \href{https://iso639-3.sil.org/code/lud}{lud}, Glottocode: \href{https://glottolog.org/resource/languoid/id/ludi1246}{ludi1246}, Uralic)  
    \gll Üö-kš  mindai ni ke-he \textbf{laske-da} \textbf{ei}. \\ 
        night{}-\textsc{tra} \textsc{1sg}.\textsc{acc} \textsc{neg} who-\textsc{ill} let-\textsc{pass.cng} \textsc{neg}\\
    \glt ‘I am not allowed to anywhere for the night.’ \citep[178]{MSFOu69}
\z 

\subsubsection{Number agreement with plural subjects}\label{sec:vep:2.1.3}

Compared to other neighboring and related languages, number agreement between the main verb and subject is a language-specific Veps feature (see \tabref{tab:vep:1} and \tabref{tab:vep:2} above). While an unmarked connegative form is characteristic of both the Finnic and Saamic languages, the lexical verb carries a specific plural marker in Veps (\ref{ex:vep:15}--\ref{ex:vep:16}). Thus, instead of having the verb stem as the sole connegative form, Veps has a morphologically distinct plural connegative. A similar agreement pattern is found in Komi, a more eastern Uralic language \citep[242, 246--248]{Hamari2015}.
\ea
    \label{ex:vep:15}
    \gll Bat’ da mam \textbf{ii} \textbf{roht{}-kei} külb{}-ma{}-ha päst-ta.\\
         dad and mom \textsc{neg.3} dare-\textsc{cng.pl} bath\textsc{{}-inf}{}-\textsc{ill} let\textsc{{}-inf}\\
    \glt ‘Dad and mom dare not let [her/him] go to the sauna.’ (VeN, \citealt[157]{MSFOu70})
\z 

\ea
    \label{ex:vep:16}
    \gll Papad i tütär kazaka{}-d  familiou \textbf{ii} \textbf{kuc{}-koi}. \\
         clergyman’s.wife and daughter farmhand-\textsc{par} surname.\textsc{ade} \textsc{neg.3} call-\textsc{cng.pl}\\
    \glt ‘The clergyman’s wife and daughter do not call the farm-hand by surname.’ (VeC, \citealt[101]{MSFOu70})
\z

In this case, too, Ludic \REF{ex:vep:17} shares this characteristic with Veps. The use of a plural connegative in declarative sentences has a parallel in plural imperative (prohibitive).

\ea\label{ex:vep:17} Ludic \newline %(ISO: \href{https://iso639-3.sil.org/code/lud}{lud}, Glottocode: \href{https://glottolog.org/resource/languoid/id/ludi1246}{ludi1246}, Uralic)  
    \gll \textbf{E{}-m} \textbf{ol{}-guu} d’äl’ges kävü-tud, kaik näma{}-d vahna{}-d. \\
         \textsc{neg{}-1pl} be-\textsc{cng.pl} after visit-\textsc{pst.ptcp} all \textsc{3pl-pl} old{}-\textsc{pl}\\
    \glt ‘We haven’t been there afterwards, they are all old.’ \citep[82]{Pahomov2011}
\z

\subsection{Non-declarative sentences}\label{sec:vep:2.2}
\largerpage
This section discusses negative sentences that display one of the two morphologically marked moods, i.e. the imperative (prohibitive) and conditional. Furthermore, the negative question phrase and role of copula in non-declarative sentences will be discussed. The negative imperative is an analytic construction consisting of the auxiliary \textit{ala} \textsc{proh} and the main verb. 

\begin{table}
\centering
\begin{tabularx}{.9\textwidth}{lllll}
\lsptoprule
& \multicolumn{2}{l}{pur{}-da (bite\textsc{{}-inf})} & \multicolumn{2}{l}{pes-tas (wash\_oneself\textsc{{}-inf})} \\
& {affirmative}  & {negative} & {affirmative}  & {negative}\\
\midrule 
\textsc{2sg} & pure & ala pure & peste & ala peste \\
\textsc{3sg} & pur-ga-ha & al-g-ha pur-goi & pes-ka-has & al-g-ha peste \\
\textsc{1pl} & pur-ga-m & al-ga-m pur-goi & pes-ka-moi & al-ga-m pes-koiš \\
\textsc{2pl} & pur-ga-t & al-ga-t pur-goi & pes-ka-toi & al-ga-t pes-koiš \\
\textsc{3pl} & pur-ga-ha & al-g-ha pur-goi & pes-ka-has & al-g-ha pes-koiš \\
\lspbottomrule
\end{tabularx}
    \caption{Affirmative and negative imperative paradigm of the verbs \textit{purda} ‘bite’ and \textit{pestas} ‘wash (oneself)’ in standardized Veps (\citealt{Zaitseva2000}: 212\textit{–}213, 222)}
    \label{tab:vep:3}
\end{table}

Imperative forms inflect for person except for first person singular. Second person singular imperative is unmarked both in affirmative and negative sentences. Both the prohibitive auxiliary and the connegative are unmarked (\ref{ex:vep:18}--\ref{ex:vep:19}). The imperative affix -\textit{ga}{}-, used in other persons than second singular, is attached to both the lexical verb and the auxiliary, which both inflect for person (\ref{ex:vep:20}--\ref{ex:vep:21}). The same affix occurs in affirmative plural imperative as in the Russian borrowing \textit{vzaderš-ka-t} hold-\textsc{imp}{}-2\textsc{pl} ‘(you all) hold!’. There is quite a lot of phonological variation between the Veps dialects but the morphosyntactic pattern is the same in the whole language area (\citealt{Zaitseva2001}: 121\textit{–}137, 247\textit{–}285).

\ea
    \label{ex:vep:18}
    \gll \textbf{Ala} \textbf{ambu}, Sašku, edha{}-n kondii, ii put{}-ta.\\
         \textsc{proh} shoot.\textsc{cng} Sašku distant\textsc{{}-ess} bear \textsc{neg.3} hit\textsc{{}-inf}\\
    \glt ‘Don’t shoot, Sašku, the bear is too far, one cannot hit [it].’ (VeS, \citealt[99]{Kettunen1920})
\z 

\ea
    \label{ex:vep:19}
    \gll \textbf{Ala}  \textbf{lö} mindei! \\
         \textsc{proh}  hit.\textsc{cng} \textsc{1sg.acc}\\
    \glt ‘Don’t hit me!’ (VeC, \citealt[40]{MSFOu70})
\z

\ea
    \label{ex:vep:20}
    \gll A se pagiže{}-b: \textbf{al{}-ga{}-t}  \textbf{män{}-gä{}-t}!\\
         but it talk-\textsc{3sg} \textsc{proh-imp-2pl} go-\textsc{imp-2pl}\\
    \glt ‘However, it says: don’t go!’ (VeC, \citealt[20]{MSFOu70})
\z

\ea
    \label{ex:vep:21}
    \gll \textbf{Vzaderš{}-ka{}-t} neti{}-da neidiš{}-t i ni kut \textbf{au{}-ga{}-t} \textbf{päst{}-kä{}-t}. \\
         hold-\textsc{imp}{}-2\textsc{pl} this-\textsc{par} girl-\textsc{par} and \textsc{neg} how \textsc{proh}{}-\textsc{imp}{}-2\textsc{pl} let-\textsc{imp}{}-2\textsc{pl}\\
    \glt ‘Hold this girl and don’t let her in under any circumstances.’ (VeS, \citealt[24]{Kettunen1925})
\z 
         
The other affixal mood, the conditional, inflects for person in present and past tense. Unlike the imperative, the mood affix always attaches to the main verb and never to the auxiliary. In this case, the \textsc{neg}.\textsc{aux} \textit{ei / ii} inflects for person while mood and tense are properties of the main verb, compare for instance \textit{näg-iži-n} see\textsc{{}-cond}{}-\textsc{1sg} but e-n \textit{näg-iži} \textsc{neg-1sg} see\textsc{{}-cond} (see \tabref{tab:vep:4}).

\begin{table}
\centering
\begin{tabularx}{\textwidth}{X lX ll}
\lsptoprule
& \multicolumn{2}{l}{näh{}-ta (see\textsc{-inf})} & \multicolumn{2}{l}{pes-tas (wash\_oneself\textsc{{}-inf})}\\
& {affirmative}  & {negative} & {affirmative}  & {negative}\\
\midrule 
\textsc{1sg} & näg-iži-n & e-n näg-iži & pez-iži-moi & e-n pez-iži-he \\
\textsc{2sg} & näg-iži-d & e-d näg-iži & pez-iži-toi & e-d pez-iži{}-he \\
\textsc{3sg} & näg-iži   & ei näg-iži & pez-iži-he & ei pez-iži{}-he \\
\textsc{1pl} & näg-iži{}-m & e-m näg-iži & pez-iži{}-moiš & e-m pez-iži{}-he \\
\textsc{2pl} & näg-iži{}-t & e-t näg-iži & pez-iži{}-toiš & e-t pez-iži{}-he \\
\textsc{3pl} & näg-iži{}-ba & ei näg-iži & pez-iži{}-he & ei pez-iži{}-he \\
\lspbottomrule
\end{tabularx}
    \caption{Affirmative and negative conditional paradigm of the verbs \textit{nähta} ‘see’ and \textit{pestas} ‘wash (oneself)’ in standardized Veps (\citealt{Zaitseva2000}: 213\textit{–}214, 223)}
    \label{tab:vep:4}
\end{table}

The negative conditional sentence, as a rule, correlates morphosyntactically with negation in basic declarative sentences as the auxiliary inflects for person and the main verb inflects for tense and mood (see \tabref{tab:vep:1} and \tabref{tab:vep:4} above). 
The only exception is that there is no plural agreement between the lexical verb and the subject in the conditional. In the reflexive conjugation the reflexive marker is attached to the mood affix and, hence, the morpheme order is \textsc{neg-person} \textsc{v-cond-refl}. The third person singular and plural are always distinct in Southern Veps \REF{ex:vep:22}, whereas in other varieties they often have identical forms \REF{ex:vep:23}.
    %\gloss{see comment above about P =person marker}

\ea
    \label{ex:vep:22}
    \gll Pä-n  kärii i paglaa sidoo, mise  velje-d \textbf{e{}-bad} \textbf{tedustaa{}-ž}.\\
         head\textsc{{}-gen}(-\textsc{acc}) wrap.\textsc{pst}.\textsc{3sg}  and string.\textsc{ade} tie.\textsc{pst}.\textsc{3sg}  in.order.to brother{}-\textsc{pl} \textsc{neg{}-3pl} know\textsc{{}-cond}\\
    \glt ‘He wrapped the head and tied [it] with a string so that the brothers would not recognize [him].’ (VeS, \citealt[68]{Kettunen1920})
\z 

\ea
    \label{ex:vep:23}
    \gll Peiteiči kundle-n (ukse-n taga), miise \textbf{ii} \textbf{tetä{}-iž} rahvaz perti-š.\\
         secretly listen-\textsc{1sg} (door\textsc{{}-gen} behind) in.order.to \textsc{neg.3} know\textsc{{}-cond} people house-\textsc{ine}\\
    \glt ‘I listen secretly (behind the door) so that people in the house would not know.’ (VeN, \citealt[164]{MSFOu70})
\z

The polar question phrase is formed by a question particle {}-\textit{ik}, which is attached to the verb, in the negative phrase the inflecting \textsc{neg.aux}. The main verb displays an unmarked connegative form as in declarative clauses (\ref{ex:vep:24}–\ref{ex:vep:25}).

\ea
    \label{ex:vep:24}
    \gll \textbf{E{}-n{}-ik}  mina \textbf{ole} si{}-lei mužik?\\
         \textsc{neg-1sg-q} \textsc{1sg} be.\textsc{cng} \textsc{2sg-all} man\\
    \glt ‘Am I not your husband?’ (VeN, \citealt[155]{MSFOu70})
\z
         
{\ea
    \label{ex:vep:25}
    \gll \textbf{E{}-d{}-ik} teda mi{}-dä ni, miše kazde{}-iž vaze{}-iže{}-d? \\
         \textsc{neg{}-2sg-q} know.\textsc{cng} what-\textsc{par} \textsc{neg} in.order.to grow\textsc{{}-cond} calf-\textsc{dim{}-pl}\\
    \glt ‘Don’t you know anything that would make the calves grow?’ (VeC, \citealt[100]{MSFOu70})
\z}

\subsection{Negation in stative predications}\label{sec:vep:2.3}

The Finnic languages, as a rule, always use a copula in stative relational clauses, which is also characteristic of most Saamic languages. However, it represents a fundamental syntactic difference in comparison to the more eastern Uralic languages. Veps is an exception to this rule as it often omits the copula, although there is a lot of alternation (\ref{ex:vep:26}–\ref{ex:vep:27}) between Veps varieties (\citealt{Grünthal2015}: 184–200). Dropping the copula is optional and functionally conditioned in affirmative clauses, whereas in negatives copula is syntactically obligatory and shows phonologically driven fusion. The lack of copula is most likely secondary and caused by the strong Russian influence.

\ea
    \label{ex:vep:26}
    \gll Käzi=robeh om save-sīne.\\
         hand=dish is clay-\textsc{adj}\\
    \glt ‘There is clay in the handwashing dish.’ (VeN, \citealt[72]{MSFOu86})
\z 

\ea
    \label{ex:vep:27}
    \gll Jorš se oghakas kala. \\
         ruffe it thorny fish\\
    \glt ‘The ruffe is a thorny fish.’ (VeS, \citealt[70]{MSFOu86})
\z
    
The difference between a clause with or without a copula is not seen in the paradigms. In written standardized Veps copulaless clauses do not normally occur.

\begin{table}
    \centering
\begin{tabularx}{\textwidth}{Xllll}
\lsptoprule
& \multicolumn{2}{c}{Present}   & \multicolumn{2}{c}{Past}  \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
& {affirmative}  & {negative} & {affirmative}  & {negative}\\
\midrule
\textsc{1sg} & ole-n & en ole & ol-i-n & e-n ole-n(d)\\
\textsc{2sg} & ole-d & ed ole & ol-i-d & e-d ole-n(d)\\
\textsc{3sg} & om & iile {\textasciitilde} ele {\textasciitilde} ii {\textasciitilde} ei & ol-i & iilend {\textasciitilde} ele-n\\
\textsc{1pl} & ole-m & e-m(ei) uu-goi & ol-i-m(ei) & e-m(ei) uu-nu-goi\\
\textsc{2pl} & ole-t(ai) & e{}-t(ei) uu{}-goi & ol-i-t(ei) & e-t(ei) uu-nu-goi\\
\textsc{3pl} & oma & ii uugoi {\textasciitilde} ei uu-goi & ol-i-ba & ii uu-nu-goi\\
\lspbottomrule
\end{tabularx}
\caption{Present and past tense paradigms of the verb \textit{olda} ‘be’  in Central Veps (\citealt{Zaitseva2001}: 254\textit{–}275, 222)}
\label{tab:vep:5}
\end{table}

Unlike first and second person as well as third person plural, the constituents of the negative phrase merge in the third person singular. Forms such as \textit{ii-li-ž} \textsc{neg}{}-\textsc{pst{}-cond} inflecting for more categories, occur as well \citep[266]{Zaitseva2001}.
The copula may be absent only in present tense affirmative declarative clauses. Other clause types involve more variation as there may be no copula in existential clauses as well \REF{ex:vep:28}. Tense (\ref{ex:vep:29}–\ref{ex:vep:30}), mood and negation are always marked by inflecting the copula.
Predications of identity and location use the same copula or non-copula construction as the rest of the stative predications shown here.
    % \lv[inline,size=tiny,color=blue!40]{I don't see any examples of predications of identity or locative predications with a definite subject, either positive or negative. If it is difficult to obtain such examples, please say so. Otherwise please add such examples. It would be good to follow the order suggested in the questionnaire.}
    % \aut[inline,size=tiny,color=yellow!40]{Maybe I have overlooked or misunderstood this part of the questionnaire. Would it be easiest to leave this just as it is? The question of definiteness would need a much longer discussion. In a post-nominal position the demonstrative pronoun se ‘it’ may be considered as a definiteness marker. But it is not limited to post-nominal positions but is used as a focus marker more widely in vernacular speech but not in the written language. \\ If predication of identity means sentences like I am a Veps, this would also trigger a wider discussion about deriving identity labels from topographic terms and expressions of locality.}
    % \hc[inline]{What about adding one sentence to the effect that  "predications of identity and location use the same copula as the rest of the stative predication shown here". There is no need to further discussion of these topics.}
    % \aut[inline,color=yellow!40]{Let's do it this way. The fact is that not only stative relational clauses are used in predications of identity. A frequent way is to say that "I speak in our language" indicating "I'm a Veps". However locality and topographic attributes correspond to SRC constructions. Thus, a slight modification to what HC proposes. Please, note the copula/non-copula distinction: "predications of identity and location use the same copula or non-copula construction as stative predication shown here"}

\ea
    \label{ex:vep:28}
    \gll Me-ide derevna-s vähä norišto-d. \\
         \textsc{1pl-gen} village-\textsc{ine} few youth-\textsc{par}\\
    \glt ‘There are a few young people in our village.’ (VeC, \citealt[40]{MSFOu86})
\z 

\ea
    \label{ex:vep:29}
    \gll  Pert' mugeine lačak \textbf{ol-i}. \\
         house such lousy be-\textsc{pst}.\textsc{3sg} \\
    \glt ‘The house was such a lousy one.’ (VeN, \citealt[72]{MSFOu86})
\z

\ea
    \label{ex:vep:30}
    \gll Edō mä gol'u \textbf{ole-l-ī-n} humalakaz. \\
         earlier \textsc{1sg} always be-\textsc{freq}-\textsc{pst}{}-\textsc{1sg} drunkard\\
    \glt ‘Earlier I always used to be a drunkard.’ (VeS, \citealt[74]{MSFOu86})
\z 

The basic rule is that stative relational clauses often lack the copula but marked morphosyntactic properties, such as tense and mood, trigger it. This is reflected in the emergence of a negated copula in Veps. It is based on the fusion of \textsc{neg.aux} and the copula, which may alternate with syntactically segmentable units. Thus, there is often no copula in affirmative clauses, whereas a negative copula is obligatory in existential stative relational clauses like in more eastern Uralic languages,
a feature more commonly found in the eastern Uralic languages \citep{Veselinova2015}. The fused forms in Veps are most commonly third person forms, (a) \textit{iile} be.\textsc{neg.3sg} and (b) \textit{iilen(d)} be.\textsc{neg}.\textsc{pst}.\textsc{3sg}, whereas other persons use an inflecting \textsc{neg.aux} + \textsc{v.cng} as in basic declarative sentences. Yet, the fusional negative existential is not fully lexicalized, and segmentable negative existential is frequently used besides the fusional pattern (\ref{ex:vep:31}–\ref{ex:vep:32}). Hence, the Veps negating copula alternates between two systems (cf. \citealt{Veselinova2014}).

\ea\label{ex:vep:31}%hc
    \ea\label{ex:vep:31a}
        \gll  Sinei vel’l’e-d sizare-d? \\ 
              \textsc{2sg.all} brother{}-\textsc{pl} sister{}-\textsc{pl}\\
        \glt ‘Do you have brothers or sisters?’
    \ex\label{ex:vep:31b}
        \gll  \textbf{Ei} \textbf{ole} vel’l’-i-d  ni sizar-i-d. \\
              \textsc{neg} be.\textsc{cng} brother{}-\textsc{pl}{}-\textsc{par} \textsc{neg} sister{}-\textsc{pl}{}-\textsc{par}\\
        \glt ‘There are neither brothers nor sisters.’ (VeC, Ladv 2006)
    \z
\z 

\ea
    \label{ex:vep:32}
    \gll  Mii-le kolhoza-s ka \textbf{pasporta-d}  ni \textbf{ii} \textbf{ole-nd}.\\
         \textsc{1pl-all} kolkhoz{}-\textsc{ine} yes passport-\textsc{par} \textsc{neg} \textsc{neg.3} be-\textsc{pst.ptcp}\\
    \glt ‘In the kolkhoz we did not have even a passport.’ (VeC, Ladv 2006)
\z

The fusional negating copula that occurs in both older and contemporary data (\ref{ex:vep:33}–\ref{ex:vep:34}) inflects for tense. The past participle is simply attached to the negating copula \REF{ex:vep:35} like any other verb in negative past tense forms.

\ea
    \label{ex:vep:33}
    \gll Biržuu mugažno \textbf{eole} elä-j-i-d  nügude. \\
         Birzh.\textsc{ade} also \textsc{neg}.be live-\textsc{ptcp}{}-\textsc{pl}{}-\textsc{par} now\\
    \glt ‘Nowadays there are no inhabitants at Birzh either.’ (VeC, Mäggär’v 2007)
\z

\ea
    \label{ex:vep:34}
    \gll \textbf{Iile} mü{}-i-le ni keng-ii-d’, \textbf{iile} mü-i-le ni sob-ii-d’.\\
         \textsc{neg}.be \textsc{1pl-pl-all} \textsc{neg} shoe-\textsc{pl-par} \textsc{neg}.be \textsc{1pl-pl-all} \textsc{neg.3} cloth-\textsc{pl-par}\\
    \glt ‘We don’t have shoes, we don’t have clothes.’ (VeS, \citealt[105]{Kettunen1925})
\z 

\ea
    \label{ex:vep:35}
    \gll Sö-dä ii mi-da \textbf{ol’}, ni soba-d  \textbf{iile-nd}  ni kenga{}-d \textbf{iile-nd}. \\
         eat\textsc{-inf} \textsc{neg.3} what-\textsc{par} be.\textsc{pst.3sg} \textsc{neg} cloth-\textsc{par}  \textsc{neg}.be-\textsc{pst.ptcp} \textsc{neg} shoe-\textsc{par} \textsc{neg}.be-\textsc{pst.ptcp}\\
    \glt ‘There was nothing to eat, there were no clothes and no shoes.’ (VeS, \citealt[56]{Kettunen1920})
\z

Functionally the negating copula has the role of a negative existential. We may formulate the reanalysis and grammaticalization from a segmentable construction to a fused form as in \REF{ex:vep:35b}. 

\ea\label{ex:vep:35b}
\textsc{neg}.\textsc{aux} + \textsc{cop} > \textsc{negex}
\z 

Two inflecting units merge in this change. The negative auxiliary that inflects for person and number fuses with the copula, which inflects for tense and mood. However, there is no evidence that the Veps secondary negative existential would inflect for other than third person.  A similar development is found in Erzya, in which the negative existential has clear constraints and does not inflect for all persons (\citealt{Hamari2007}: 188–207, 260–268; \citealt{HamariAasmäe2015}: 306–307). Instead of marking person, it operates as a generic negative existential. As in the case of number agreement of plural connegative forms (see \sectref{sec:vep:2.1.3} above), the secondary negative existential occurs in Ludic as a shared feature \REF{ex:vep:36} and may inflect for mood \REF{ex:vep:37}.\footnote{More research is needed on the encoding of negative existence in Veps.}

\ea\label{ex:vep:36} Ludic \newline %(ISO: \href{https://iso639-3.sil.org/code/lud}{lud}, Glottocode: \href{https://glottolog.org/resource/languoid/id/ludi1246}{ludi1246}, Uralic)  
    \gll Kunne{}-ni sua-nu aja-da, traktoruu-d ved’ \textbf{iel{}-nu}. \\
         where{}-\textsc{neg} get-\textsc{pst.ptcp} drive\textsc{{}-inf} tractor-\textsc{par} \textsc{clt} \textsc{neg}.be-\textsc{pst.ptcp} \\
    \glt ‘There was nowhere to drive, there was no tractor.’ \citep[110]{Pahomov2011}
\z


\ea\label{ex:vep:37} Ludic \newline %(ISO: \href{https://iso639-3.sil.org/code/lud}{lud}, Glottocode: \href{https://glottolog.org/resource/languoid/id/ludi1246}{ludi1246}, Uralic)  
    \gll \textbf{Iel-iiž} ni mičč{}-ii{}-d  ur’k-ii-d da, buttob \textbf{iel{}-iiž}. \\
         \textsc{neg}.be\textsc{{}-cond} \textsc{neg} any.kind.of{}-\textsc{pl}{}-\textsc{par} smell{}-\textsc{pl}{}-\textsc{par} and, if \textsc{neg}.be\textsc{{}-cond}\\
    \glt ‘If there were no smells and, if there were not.’ \citep[206]{Pahomov2011}
\z 

\subsection{Negation in non-main clauses}\label{sec:vep:2.4}

Compared to basic declarative clauses, negation in subordinate clauses does not diverge in any significant way. The constituent order, for instance, remains the same because \textsc{neg}.\textsc{aux} precedes the verb even more regularly than in affirmative clauses (cf. \sectref{sec:vep:2.1.1} on reversed word order). This is valid in subordinate clauses such as causal clauses \REF{ex:vep:38}, negative conditional clauses \REF{ex:vep:39}, temporal clauses (example \ref{ex:vep:40} showing the maintenance of the fused negative existential), relative clauses \REF{ex:vep:41}, and reversed question clauses \REF{ex:vep:42}.

\ea
    \label{ex:vep:38}
    \gll Sigä  radnik ii anda-n kirjo-i-d, mišo kut vuu škol-ha \textbf{ii} \textbf{kävele}. \\
         there employee \textsc{neg.3} give-\textsc{pst.ptcp} book{}-\textsc{pl}{}-\textsc{par} in.order.to how yet school{}-\textsc{ill} \textsc{neg.3sg} go.\textsc{cng}\\
    \glt ‘There, the employee didn’t give books, because he doesn’t go to school, yet.’ (VeC, Ladv 2014)
\z

\ea
    \label{ex:vep:39}
    \gll  Erašti  po=telefonuu  pagiže{}-mei ka po=vepsski, mišto sekreto{}-i{}-d  \textbf{ii} \textbf{teda-iž}.\\
         sometimes \textsc{clt}{}=telephone.\textsc{ade} speak-1\textsc{pl} so \textsc{clt}{}=Veps in.order.to secret{}-\textsc{pl}{}-\textsc{par} \textsc{neg.3} know\textsc{{}-cond}\\
    \glt ‘Sometimes we speak Veps on the phone, so that they wouldn’t know the secrets.’ (VeC, Mäggär’v 2007)
\z 


\ea
    \label{ex:vep:40}
    \gll [...] kuna mini seme-ta, ku ma-d \textbf{ele}.\\
        {} where \textsc{1sg.ade} plant\textsc{-inf} when land-\textsc{par} \textsc{neg}\\
    \glt ‘... where I should plant, when there is no land.’ (VeC, \citealt[40]{MSFOu70})
\z

\ea
    \label{ex:vep:41}
    \gll  Tul-i-bad pert’-he äja-d ristitu-d, kudamba-d \textbf{ii} \textbf{pagiš{}-tut} ičemoi kele{}-l.\\
         come-\textsc{pst}{}-3\textsc{pl} house{}-\textsc{ill} many{}-\textsc{pl} people{}-\textsc{pl} which{}-\textsc{pl} \textsc{neg.3} speak-\textsc{pst.ptcp} own language-\textsc{ade}\\
    \glt ‘Such people came to our house, who didn’t speak Veps.’ (VeC, Päžar’ 2005) 
\z

\ea
    \label{ex:vep:42}
    \gll Probyi-ka-t, mijak \textbf{e-t} \textbf{pagiž{}-koi}?\\
         try-\textsc{imp}{}-2\textsc{pl} why \textsc{neg{}-2pl} speak-\textsc{cng.pl}\\
    \glt ‘Try, why don’t you speak?’ (VeC, Mäggär’v 2007)
\z

In non-finite clauses, negation may involve a modal function. In (\ref{ex:vep:43}–\ref{ex:vep:44}) the negative clause consisting of \textsc{neg}.\textsc{aux} and a non-finite verb is the predicating unit that has an epistemic meaning excluding a possible action.

\ea
    \label{ex:vep:43}
    \gll Näge  mitte  dikar’,  ni  kuna  \textbf{ii} \textbf{sat{}-ta}.\\
         see what.a slacker  \textsc{neg}  where \textsc{neg.3} send-\textsc{inf}\\
    \glt ‘Look, what a slacker! You cannot send [him] anywhere.’ (VeS, \citealt[485]{MSFOu86})
\z

\ea
    \label{ex:vep:44}
    \gll Nu igä{}-n  pagiže{}-moi  sikš=ku, miše \textbf{ii} \textbf{abit{}-ta} adivo{}-d.\\
         well  age-\textsc{acc} talk-\textsc{1pl} because=of in.order.to  \textsc{neg.3} insult-\textsc{inf} guest-\textsc{par}\\
    \glt ‘We have always talked so that we wouldn’t insult the guest.’ (Päžar’ 2011)
\z

In Veps, certain non-finite verb forms may display some inflectional cases, but not all of them.
The attaching of a case affix to a verb form is characteristic of non-finites that are marked with the affix \textit{{}-ma{}-}. This is exemplified in \REF{ex:vep:45}, which displays the abessive case that functionally corresponds to a negative instrumental (see \sectref{sec:vep:3.3} below).

\ea\label{ex:vep:45}
    \gll Ii pida räkita-da, \textbf{räkita-ma-ta} pravi-ze   azd'.\\
         \textsc{neg.3} must thrash-\textsc{inf} thrash-\textsc{inf-abe} get.better-\textsc{refl.3sg} thing\\
    \glt ‘You need not thrash [him], the thing will get better without thrashing.’ (VeN, \citealt[500]{MSFOu86})
\z 
         
\subsection{Negative lexicalizations}\label{sec:vep:2.5}
The negation of the epistemic modal auxiliary \textit{voida} ‘can’ has developed in the same way as the negating copula (see \sectref{sec:vep:2.3}, above). The segmentable construction consisting of \textsc{neg}+V[\textsc{mod}] has fused into an auxiliary which, unlike the negative existential, formally inflects for person or, to put it more cautiously, actually reflects phonological assimilation in speech. The default negation of the epistemic modal auxiliary is manifested in Veps literary language (\ref{ex:vep:46}–\ref{ex:vep:47}).

\ea
    \label{ex:vep:46}
    \gll No kaik-se \textbf{ei} \textbf{voi} seižu{}-ta{}-das.\\
         but  all-\textsc{foc} \textsc{neg} can.\textsc{cng} stand-\textsc{caus{}-inf.refl}\\
    \glt ‘But everything cannot be stopped.’
\z 

\ea
    \label{ex:vep:47}
    \gll Siloi läžu-tihe  kogonaiže-l kanza-l… \textbf{ei}  \textbf{voi-nu-goi}  nous{}-ta homendese{}-l…\\
         then be.sick-\textsc{pst}.3\textsc{pl}  whole-\textsc{ade} family-\textsc{ade} \textsc{neg}  can-\textsc{pst.ptcp}{}-\textsc{pl} get.up\textsc{{}-inf} morning-\textsc{ade}\\
    \glt ‘The whole family was sick then… it was not possible to get up in the morning…’
\z

The original morpheme order is maintained in the fusion of the two segmental units. Hence, it does not involve the shift of person marking affix to the end of the fused form, which would indicate a complete reanalysis of the construction. Instead, the original morpheme order is preserved and the first syllable \textit{em} (< \textit{e-n} \textsc{neg-1sg}), alternatively \textit{ii} \textsc{neg.3} indicates person (\ref{ex:vep:48}–\ref{ex:vep:50}).

\ea
    \label{ex:vep:48}
    \gll Sousim mä \textbf{emboo} ra-ta. \\
         at.all \textsc{1sg} cannot.\textsc{1sg} work\textsc{{}-inf} \\
    \glt ‘I can’t work at all.’ (VeS, \citealt[78]{Kettunen1920})
\z

\ea
    \label{ex:vep:49}
    \gll \textbf{Emboi}  tora-ta tii-deke.\\
        cannot.\textsc{1sg} quarrel\textsc{{}-inf} \textsc{2pl-com}\\
    \glt ‘I cannot quarrel with you.’ (VeC, \citealt[92]{MSFOu70})
\z 

\ea
    \label{ex:vep:50}
    \gll Mä paigaah-ii-n’ kondja-d, put-ta \textbf{emboo-le{}-n},  pauš-gi ambuu da put-ta \textbf{iivoo-le{}-n}  put-ta.   \\
         \textsc{1sg} shoot-\textsc{pst}{}-\textsc{1sg} bear-\textsc{par} hit\textsc{{}-inf} cannot.\textsc{1sg-freq}{}-\textsc{pst.ptcp} Pauš-\textsc{foc} shoot.\textsc{pst}.\textsc{3sg} and hit\textsc{{}-inf} cannot.\textsc{3sg}-\textsc{freq}-\textsc{pst.ptcp} hit\textsc{{}-inf}  \\
    \glt ‘I shot at the bear but I could not hit (it), Pauš shot, too, and could not hit.’ (VeS, \citealt[105]{Kettunen1920})
\z

\section{Non-clausal negation}\label{sec:vep:3}

In Veps, there is a clear distinction between clausal and non-clausal negation, which is more strongly affected by Russian model. The negative focus particle \textit{ni} borrowed from Slavic (Russian) is widely used as the basic means for non-clausal negation and is not inflected. It typically precedes a pronoun or a quantifier, whereas nouns are not negated, although the negative focus particle \textit{ni} may occur as a connector \REF{ex:vep:51} in a prenominal position and occasionally shift to a postnominal position \REF{ex:vep:52}.

\ea
    \label{ex:vep:51}
    \gll Ei ole vel’l’-i-d  \textbf{ni} sizar-i-d.\\
         \textsc{neg} be.\textsc{cng} brother{}-\textsc{pl}{}-\textsc{par} \textsc{neg} sister{}-\textsc{pl}{}-\textsc{par}\\
    \glt ‘There aren’t (= I haven’t got) any brothers nor sisters.’ (VeC, Ladv 2006)
\z

\ea
    \label{ex:vep:52}
    \gll Mii-le  kolhoza-s ka pasporta-d \textbf{ni} ii ole-nd. \\
         \textsc{1pl-all} kolkhoz{}-\textsc{ine} so passport-\textsc{par} \textsc{neg} \textsc{neg.3} be-\textsc{pst.ptcp}\\
    \glt ‘We didn’t even have a passport [identity card] in the kolkhoz.’ (VeC, Ladv 2006)
\z

\subsection{Negative replies}\label{sec:vep:3.1}

There are different ways to reply to questions, proposals and statements that form polar interrogatives (cf. \citealt{Miestamo2011}). Both question particles \REF{ex:vep:53}Q and declarative clauses  that lack a question marker (\ref{ex:vep:54}Q--\ref{ex:vep:55}Q) are used to form polar interrogatives in Veps, whereas intonation, for instance, does not play any role. Negative replies are typically finite constructions either repeating the predicate and converting the person to the respondent, \textsc{2sg} > \textsc{1sg} (\ref{ex:vep:53}A--\ref{ex:vep:54}A), or simply marking the predicative with the inflecting \textsc{neg}.\textsc{aux} (\ref{ex:vep:55}A; cf. also \REF{ex:vep:31}, above). In both cases the answer exhibits a negative reply to the propositional content of the question.


\ea\label{ex:vep:53}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:vep:53A}
        \gll Mahtad-ik sinä iče kirjuta-da vepsä-ks?\\ 
              can-\textsc{q} \textsc{2sg} self write\textsc{-inf} Veps{}-\textsc{tra}\\ 
        \glt ‘Can you write in Veps yourself?’
    \exi{A: }\label{ex:vep:53B}
        \gll  \textbf{E-n} \textbf{mahta}. Nügüd ni-ken ii mahta{}-gi mii-l.  \\ 
               \textsc{neg-1sg} can.\textsc{cng} now \textsc{neg}{}-who \textsc{neg.3} can\textsc{.cng{}-foc} \textsc{1pl-ade}\\
        \glt ‘No I can’t. Nowadays none of us can.’ (VeC, Ladv 2006)
    \end{xlist}
\z

\ea\label{ex:vep:54}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:vep:54A}
        \gll Sinä mahta-d luge-da vepsä-ks? Vot nügüd oma nene kniga-d da om… \\
        \textsc{2sg} can-\textsc{2sg} read\textsc{{}-inf} Veps{}-\textsc{tra} well now are these book{}-\textsc{pl} and is \\ 
              \\ 
        \glt ‘Can you read Veps? Nowadays there are these books and…’
    \exi{A: }\label{ex:vep:54B}
        \gll  \textbf{E-n} \textbf{mahta}.\\ 
              \textsc{neg-1sg} can.\textsc{cng}\\
        \glt ‘No, I can’t.’ (VeC, Ladv 2006)
    \end{xlist}
\z

\ea\label{ex:vep:55}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:vep:55A}
        \gll  I el-i-d igä-n tägä Ladva-s?\\ 
              And live-\textsc{pst}{}-\textsc{2sg} age-\textsc{acc} here Ladv{}-\textsc{ine}\\ 
        \glt ‘And you have lived the whole life here in Ladv?’
    \exi{A: }\label{ex:vep:55B}
        \gll  \textbf{E\-}{}-\textbf{n}. Ol{}-i{}-n evakuoi{}-tud  Sverdlovskie oblasti{}-he.\\ 
               \textsc{neg-1sg} be-\textsc{pst}{}-\textsc{1sg} evacuate-\textsc{pst.ptcp} Sverdlovsk oblast{}-\textsc{ill}\\
        \glt ‘No. I was evacuated to Sverdlovsk oblast.’ (VeC, Ladv 2006)
    \end{xlist}
\z 

If the question involves a negative clause, the reply may simply repeat the \textsc{neg}.\textsc{aux}. In a focused position the fusional negative existential \textit{iile} ‘there is not’ (see \sectref{sec:vep:2.3} above) is not always used but occurs in its original form consisting of \textsc{neg}.\textsc{aux} \textsc{+} \textsc{v}. 
The grammaticalized negative existential consisting of two units is transparent in negative replies \REF{ex:vep:56}.


\ea\label{ex:vep:56}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:vep:56A}
        \gll  Sinä konz openz{}-i{}-toi  škola{}-s, vepsä{}-n kel{}-t  \textbf{ii} \textbf{ole-nd}?\\ 
              \textsc{2sg} when learn-\textsc{pst-refl.2sg} school{}-\textsc{ine} Veps\textsc{{}-gen} language-\textsc{par} \textsc{neg.3} be-\textsc{pst.ptcp}\\ 
        \glt ‘When you went to school, there was no Veps language?’
    \exi{A: }\label{ex:vep:56B}
        \gll  \textbf{Ii}.\\ 
               \textsc{neg.3}\\
        \glt ‘No.’ (VeC, Ladv 2006)
    \end{xlist}
\z 

Occasionally, the negative reply may be triggered by a negative question, which is initiated by the negative focus particle \textit{ni} ‘no’ \REF{ex:vep:57} (see (\ref{ex:vep:51}–\ref{ex:vep:52}) above and \sectref{sec:vep:3.2} below). 

\ea\label{ex:vep:57}
    \begin{xlist}[Q: ]
    \exi{Q: }\label{ex:vep:57A}
        \gll \textbf{Ni} \textbf{e-d} \textbf{mahta} kirjuta{}-da ka, mugažo? \\ 
              \textsc{neg} \textsc{neg{}-2sg} can.\textsc{cng} write\textsc{{}-inf} yes, also\\ 
        \glt ‘You cannot write either, can you?’
    \exi{A: }\label{ex:vep:57B}
        \gll  Kirjuta-da muite po=vepsski \textbf{e-n} \textbf{mahta}, muite mahta{}-n.\\ 
               write\textsc{{}-inf} otherwise \textsc{clt}{}=Veps \textsc{neg-1sg} can.\textsc{cng} otherwise can-\textsc{1sg}\\
        \glt ‘In Veps I cannot write, otherwise I can.’ (VeC, Ladv 2006)
    \end{xlist}
\z 

\subsection{Negative indefinites and quantifiers}\label{sec:vep:3.2}

Indefinite pronouns in Veps are based on an interrogative pronoun, which is typologically very widespread \citep{Haspelmath2013-wals46}. The interrogative is followed by a postnominal demonstrative pronoun \textit{se} ‘it’ as in \textit{ken-se} who-it ‘somebody’ and \textit{kus-se} where-it ‘somewhere’ (\citealt{Grünthal2015}: 278–281; \citealt{Hienonen2010, Karjalainen2016}: 51–55). A less frequent (cf. \citealt{MSFOu86}: 417–420) but possible way of forming the indefinite pronouns applies the negative particle \textit{ni} instead of the postinterrogative demonstrative \textit{se} ‘it’ (\ref{ex:vep:58}–\ref{ex:vep:60}) (cf. \citealt{Yurayong2020}).

\ea
    \label{ex:vep:58}
    \gll  \textbf{Ken-ni}  elä{}-b sigä Mägjärve-s?\\
         who-\textsc{neg} live-\textsc{3sg} there Mäggär’v{}-\textsc{ine}\\
    \glt ‘Does anyone live there in Mäggär’v?’ (VeC, Kurba 2006)
\z

\ea\label{ex:vep:59}
    \gll Aka{}-le \textbf{mi{}-n-ni} sanu{}-b da ii tuukui.\\
         wife{}-\textsc{all} what-\textsc{acc}{}-\textsc{neg} say-\textsc{3sg} and \textsc{neg.3} understand.\textsc{cng}\\
    \glt ‘He says something to (his) wife but they don’t understand.’ (VeC, Ladv 2006)
\z

\ea
    \label{ex:vep:60}
    \gll \textbf{Ku{}-s{}-ni}  lug{}-i{}-n. \\
         where{}-\textsc{ine}{}-\textsc{neg} read-\textsc{pst}{}-\textsc{1sg}\\
    \glt ‘I have read (it) somewhere.’ (VeC, Ladv 2006)
\z

Negative indefinite pronouns are formed by using the same negative focus particle \textit{ni} ‘no’ in a pre-interrogative position, in which no affix or any other constituent can occur between the two units. This is a common pattern and any interrogatives can be converted to negative indefinites this way. The pronominal stem typically inflects for case (\ref{ex:vep:61}–\ref{ex:vep:65}). These examples represent one type of clausal negation and involve the whole verb phrase.

\ea\label{ex:vep:61}
    \gll  Minu{}-l \textbf{ni} \textbf{ken} ii küzu{}-n, ken mina ole-n.\\
         \textsc{1sg}-\textsc{ade} \textsc{neg} who \textsc{neg.3} ask-\textsc{pst.ptcp} who \textsc{1sg} be-\textsc{1sg}\\
    \glt ‘No one asked me, who I am.’ (VeC, Ladv 2005)
\z

\ea
    \label{ex:vep:62}
    \gll Ei ole{}-nd sigä enamb \textbf{ni} \textbf{ke{}-dä}. škola{}-d ei ole{}-nd i \textbf{ni} \textbf{mi{}-dä} ii ole-nd. \\
        \textsc{neg.3sg} be-\textsc{pst.ptcp}  there more \textsc{neg} who-\textsc{par} school-\textsc{par} \textsc{neg.3sg} be{}-\textsc{pst.ptcp} and \textsc{neg} what-\textsc{par} \textsc{neg.3} be-\textsc{pst.ptcp}\\
    \glt ‘There was nobody any longer. There was no school, there was nothing.’ (VeC, Ladv 2006)
\z

\ea
    \label{ex:vep:63}
    \gll Ni ühte{}-d kopeka{}-d \textbf{ni} \textbf{ku{}-spei} mii{}-lei.\\
         \textsc{neg} one-\textsc{par} kopeek-\textsc{par} \textsc{neg} where\textsc{{}-ela} \textsc{1pl-all} \\
    \glt ‘We don’t get a single kopeck from anywhere.’ (VeC, Ladv 2006)
\z 

\ea
    \label{ex:vep:64}
    \gll G’ägibaba neci-da uko-n tütar-t pagasta-ha ii pästa \textbf{ni} \textbf{kut}. \\
         witch  this-\textsc{par} man\textsc{{}-gen} daughter-\textsc{par} village{}-\textsc{ill} \textsc{neg.3} let.\textsc{cng} \textsc{neg} how\\
    \glt ‘The witch won’t let the man’s daughter to the village in any way.’ (VeC, \citealt[125]{MSFOu70})
\z 

\ea
    \label{ex:vep:65}
    \gll Ka iile mi-lei s’o-da ant-ta tii-le-tei \textbf{ni} \textbf{mi-da}. \\
         yes \textsc{neg}.be \textsc{1sg}-\textsc{all} eat\textsc{-inf} give\textsc{-inf} \textsc{2pl.all}-\textsc{2pl} \textsc{neg}  what-\textsc{par}\\
    \glt ‘Well, I don’t have anything to give you to eat.’ (VeN, \citealt[154]{MSFOu70})
\z

The same particle \textit{ni} functions as a focus particle when it precedes the quantifier \textit{üks} ‘one’ but otherwise its use in quantifier phrases is limited. The quantifier inflects for case (\ref{ex:vep:66}–\ref{ex:vep:67}).

\ea
    \label{ex:vep:66}
    \gll Eras-i-š  kül-i-š kole-b vähe-mb rahvas-t a ii  sündu \textbf{ni} \textbf{üks} laps. \\
          some{}-\textsc{pl}{}-\textsc{ine} village{}-\textsc{pl}{}-\textsc{ine} die-\textsc{3sg} less\textsc{{}-comp} people-\textsc{par} but \textsc{neg.3} be.born.\textsc{cng} \textsc{neg} one child\\
    \glt ‘In certain villages less people die but not a single child is born.’ (VeC, Päžar’ 2005)
\z 

\ea
    \label{ex:vep:67}
    \gll \textbf{Ni} \textbf{ühte{}-d} supalaš{}-t ii jä{}-nd.\\
         \textsc{neg} one-\textsc{par} morsel-\textsc{par} \textsc{neg.3} remain-\textsc{pst.ptcp}\\
    \glt ‘Not a single morsel remained.’ (VeC, \citealt[97]{MSFOu70})
\z 

Adverbs based on pronominal stems display the same negative particle \textit{ni} ‘no’ which marks negative indefinites as shown above, as well as adverbs \REF{ex:vep:68}.

\ea
    \label{ex:vep:68}
    \gll Ani hüvä{}-d ii ole{}-n mugažno. \textbf{Ni} \textbf{sinnä} \textbf{ni} \textbf{tännä}, ni huba ni hüvä.\\
         Very good-\textsc{par} \textsc{neg.3} be-\textsc{pst.ptcp} likewise \textsc{neg} there \textsc{neg} here \textsc{neg} bad \textsc{neg} good\\
    \glt ‘It wasn’t very good either. Not there nor here, not bad nor good.’ (VeC, Mäggär’v 2007)
\z

There are no exclusively negative adverbs. Certain adverbs, such as \textit{igas} ‘(n)ever’ \REF{ex:vep:69} and \textit{enamb(a)} ‘more’ can be used as negative polarity items in a negative clause but they lack any formal label of negation. 
The partitive form \textit{enamba-d} more-\textsc{par} ‘(any) more’ often occurs in a negative clause (\ref{ex:vep:70}–\ref{ex:vep:71}) and follows a salient syntactic rule in all Finnic languages, in which the negative verb phrase triggers a partitive object. 
This is a more general characteristic of the verb’s nominal dependents in the negative clause.

\ea
    \label{ex:vep:69}
    \gll  Ningoš{}-t  \textbf{igas} e{}-bad näg{}-i{}-ške durako{}-d.\\
         such-\textsc{par} ever \textsc{neg{}-3pl} see-\textsc{pst}{}-\textsc{pfv} stupid-\textsc{par}\\
    \glt ‘They had never seen such a stupid one.’ (VeS, \citealt[71]{Kettunen1920})
\z 

\ea
    \label{ex:vep:70}
    \gll \textbf{Enamba{}-d} ni kävele{}-nud e{}-n.\\
         more-\textsc{par} \textsc{neg} visit-\textsc{pst.ptcp} \textsc{neg-1sg}\\
    \glt ‘I haven’t been there since.’ (VeC, Ladv 2006)
\z 

\ea
    \label{ex:vep:71}
    \gll Uk se händäst \textbf{enamb{}-i{}-d} ii lö.\\
         man it \textsc{3sg}.\textsc{acc} more{}-\textsc{pl}{}-\textsc{par} \textsc{neg.3} hit.\textsc{cng} \\
    \glt ‘The man doesn’t hit her/him any more.’ (VeC, \citealt[40]{MSFOu70})
\z

\subsection{Negative derivation and case marking}\label{sec:vep:3.3}

The abessive case \textit{{}-ta}, functionally a negative counterpart of the comitative (instrumental) marking the lack of a certain property, is frequently used in Veps \REF{ex:vep:72}. Unlike other Finnic languages, Veps does not display a negative adposition ‘without’ (\citealt{Grünthal2003}: 81–83; \citealt{Pantermöller2007}), which increases the frequence of the abessive. In addition to abessive, nouns may display a caritive derivative affix that is likewise used in attributes. In this respect, nouns inflecting for the abessive resemble derived adjectives, which display a caritive affix \textit{{}-toi} {\textasciitilde} \textit{{}-toma-} marking the lack of a certain property \REF{ex:vep:73}. There is one significant syntactic difference between these two categories because a prenominal attribute that displays the abessive agrees in case with the head, e.g. \textit{voukta-ta hebo-ta} white-\textsc{abe} horse-\textsc{abe} ‘without a white horse’, whereas a caritive derivation does not trigger case agreement.

\ea\label{ex:vep:72}
    \gll Minä oli{}-n \textbf{rado{}-ta}.\\
         \textsc{1sg} be-\textsc{pst}{}-\textsc{1sg} work-\textsc{abe} \\
    \glt ‘I was unemployed.’ (VeC, Ladv 2014)
\z

\ea\label{ex:vep:73}
    \gll  Mö olo{}-mei \textbf{tata{}-toma{}-d} i \textbf{mama{}-toma{}-d}. \\
         \textsc{1pl} be-\textsc{1pl} father-\textsc{car-pl} and mother-\textsc{car-pl}\\
    \glt ‘We are withouth father and mother.’ (VeC, \citealt[121]{MSFOu70})
\z 

In comparison with the comitative that may occur as a coordinating affix, the abessive functions in the opposite way. While the comitative makes two constituents co-referential the abessive lacks this property (\ref{ex:vep:74}–\ref{ex:vep:75}), cf. \textit{ak mužika-nke} (woman husband-\textsc{com}) ‘woman and/with man’.

\ea\label{ex:vep:74}
    \gll  Sid hii sanu{}-i{}-ba, mišto ni kut mii \textbf{sinu-ta} elä{}-da e{}-m voi.\\
         then \textsc{3pl} say-\textsc{pst-3pl} that \textsc{neg} how \textsc{1pl} \textsc{2sg-abe} live-\textsc{inf} \textsc{neg-1pl} can.\textsc{cng} \\
    \glt ‘Then they said that we cannot live without you anyway.’ (VeC, Ladv 2006)
\z 

\ea\label{ex:vep:75}
    \gll Nece \textbf{mužika{}-ta} ak. \\
         this husband-\textsc{abe} woman\\
    \glt ‘This is an unmarried woman.’ (VeC, \citealt[354]{MSFOu86})
\z

\section{Other aspects of negation}\label{sec:vep:4}
\subsection{The scope of negation}\label{sec:vep:4.1}

The reversed word order and multifunctionality of the \textsc{neg}.\textsc{aux} has given rise to syntactically ambiguous constructions, in which the scope of negation extends beyond the verb \citep[cf.][437--439]{MSFOu86}.  In this particular phrase type \textsc{neg}.\textsc{aux} is a part of the VP and displays a standard pattern. However, as the closest constituent preceding the adverbial, it also functions as the marker of lexical negation of the adverbial despite the fact that it primarily marks the person of the verb. In these constructions both \textsc{neg.aux} and the lexical verb exceptionally mark the person. Furthermore, other constituents may occur between the verb and \textsc{neg}.\textsc{aux} (\ref{ex:vep:76}–\ref{ex:vep:78}).

\ea\label{ex:vep:76}
    \gll Mä \textbf{ajoo-n’} \textbf{e-n} \textbf{ani} \textbf{lähä-ks}.\\
         \textsc{1sg} drive.\textsc{pst}{}-\textsc{1sg} \textsc{neg-1sg} very close{}-\textsc{tra}\\
    \glt ‘I didn’t drive very close.’ (VeS, \citealt[105]{Kettunen1920})
\z

\ea\label{ex:vep:77}
    \gll \textbf{Homaiče-moiš} sinun-ked \textbf{e-n} \textbf{pigai}.\\
         notice-\textsc{refl}.1\textsc{pl} \textsc{2sg-com} \textsc{neg-1sg}/\textsc{1pl} soon\\
    \glt ‘With you we didn't notice one another very soon.’ (VeC, \citealt[81]{MSFOu70})
\z 

\ea\label{ex:vep:78}
    \gll Ven'a-ks  jo \textbf{sanu-n} da erašti \textbf{e-n} \textbf{pravil'no}. \\
         Russian{}-\textsc{tra} already say-\textsc{1sg} and sometimes \textsc{neg-1sg} right\\
    \glt ‘In Russian I sometimes do not say correctly.’ (VeC, Ladv 2007)
\z

This construction is ambiguous for two reasons as it contradicts two rules. Firstly, in a simple verb phrase, the \textsc{neg}.\textsc{aux} inflects for person, whereas the lexical verb displays a connegative form. In (\ref{ex:vep:76}–\ref{ex:vep:78}) both \textsc{neg}.\textsc{aux} and the lexical verb inflect for person manifesting a non-canonical pattern. Secondly, lexical negation should be based on an uninflecting negative particle that precedes the adverb (see \sectref{sec:vep:3.2} above). In (\ref{ex:vep:76}–\ref{ex:vep:78}) the inflection of \textsc{neg}.\textsc{aux} contradicts the rule of an uninflecting negative particle that precedes the adverb. The occurrence of post-verbal \textsc{neg}.\textsc{aux} correlates with predicates that occur in Veps in a sentence-final position more often than in other Finnic languages for topical reasons. There is no transparent reason for the emergence of this ambiguous negation pattern, which strongly deviates from two canonical morphosyntactic patterns. The reason for ambiguity lies in the location of \textsc{neg}.\textsc{aux} between the lexical verb and adverbial, and that additional verbal determinants may occur between the lexical verb and \textsc{neg}.\textsc{aux}. 

\subsection{Negative polarity}\label{sec:vep:4.2}

There are only few specialized negative polarity items in Veps. The same negative focus particle \textit{ni}, which is commonly used in negative indefinites and adverbs (see \sectref{sec:vep:3.2}), operates as a general polarity marker. Individual words repeatedly occur in negative clauses, if they share some other morphosyntactic property characteristic of negation. Adverbials such as (\textit{ni}) \textit{ühte-d} (\textsc{neg}) one-\textsc{par}, \textit{ei üh{}-t kerda{}-d} \textsc{neg} one-\textsc{par} time-\textsc{par}, (\textit{ni}) \textit{mittus{}-t} (\textsc{neg}) what.kind.of-\textsc{par}, \textit{enamba{}-d} more-\textsc{par} and \textit{enamb{}-i{}-d} more{}-\textsc{pl}{}-\textsc{par} that are compatible with quantifiers (see \sectref{sec:vep:3.2}), for instance, occur in the partitive (see \sectref{sec:vep:4.3}), a typical indication of the scope of negation in NP. The same is valid for the adverb \textit{igas} {\textasciitilde} \textit{igäs} ‘(n)ever’ \REF{ex:vep:79}, a lexicalized inessive form of \textit{igä} ‘age, time’.

\ea
    \label{ex:vep:79}
    \gll Vina{}-d  \textbf{igä{}-s} e{}-n jo{}-n.\\
        wine-\textsc{par} age{}-\textsc{ine} \textsc{neg-1sg} drink-\textsc{pst.ptcp} \\
    \glt ‘I have never drunk wine.’ (VeC, \citealt[57]{MSFOu70})
\z

None of the listed lexical units marks negation alone but functions as a dependent of a negative phrase.

\subsection{Marking of NPs in the scope of negation}\label{sec:vep:4.3}

The scope of negation is illustratively seen in those clause types and negative phrases which trigger the partitive in the dependent. Constituents that regularly display or may display the nominative or genitive(-accusative) case in an affirmative clause inflect for partitive in a corresponding negative phrase. Thus, a subject of a negative existential clause inflects for partitive \REF{ex:vep:80}, and so does the subject of a reflexive verb \REF{ex:vep:81}.

\ea
    \label{ex:vep:80}
    \gll  \textbf{Škola{}-d}  ei ole{}-nd i \textbf{ni} \textbf{mi{}-dä} ii ole{}-nd. \\
         school-\textsc{par} \textsc{neg.3sg} be-\textsc{pst.ptcp} and \textsc{neg}  what-\textsc{par} \textsc{neg.3} be-\textsc{pst.ptcp}\\
    \glt ‘There was no school and there wasn’t anything.’ (VeC, Ladv 2006)
\z 

\ea
    \label{ex:vep:81}
    \gll \textbf{Ni} \textbf{kala{}-d} \textbf{ni} \textbf{rebaš{}-t} ii teh-nu{}-nzhe. \\
         \textsc{neg}  fish-\textsc{par} \textsc{neg}  fox-\textsc{par} \textsc{neg.3} do-\textsc{pst.ptcp}{}-\textsc{refl}.\textsc{3sg}\\
    \glt ‘There wasn’t any fish or fox.’ (lit. ‘It didn’t become a fish nor a fox.’) (VeS, \citealt[440]{MSFOu86})
\z 

A similar permutation takes place in the object of a transitive clause, which regularly receives partitive marking in a negative clause \REF{ex:vep:82}.

\ea
    \label{ex:vep:82}
    \gll \textbf{Bukv-i-d} ii teda. \\
         letter{}-\textsc{pl}{}-\textsc{par} \textsc{neg.3} know.\textsc{cng}\\
    \glt ‘(S)he doesn’t know (the) letters.’ (VeC, Ladv 2006)
\z 

This characteristic is shared by all Finnic languages as the partitive is the default case of object in negative transitive clauses. The correlation between negation and partitive and the dominance of this morphosyntactic pattern is so strong that adverbials such as \textit{enamba-d} more-\textsc{par} ‘(no) more’ used as polarity items (cf. \sectref{sec:vep:4.2}) in negative clauses display the partitive as well \REF{ex:vep:83}.

\ea
    \label{ex:vep:83}
    \gll \textbf{Enamba-d} mii vepsä-n kel-t se  te-tud  e-mei.\\
         more-\textsc{par} \textsc{1pl} Veps-\textsc{gen} lang-\textsc{par} it  know-\textsc{pst.ptcp} \textsc{neg-1pl}\\
    \glt ‘We didn’t know more of the Veps language.’ (VeC, Ladv 2006)
\z 

\subsection{Reinforcing negation}\label{sec:vep:4.4}

There are several ways of reinforcing negation. Most frequently, the same negative focus particle \textit{ni}, which is widely used in non-clausal negation, occurs as a reinforcing unit in a negative clause. While the inflecting \textsc{neg}.\textsc{aux} operates as the head of the VP, the negative focus particle precedes the nominal dependent of the verb (\ref{ex:vep:84}–\ref{ex:vep:85}). 

\ea\label{ex:vep:84} 
    \gll  \textbf{Ni} \textbf{ühte{}-d} \textbf{vajeh{}-t} e{}-n ope{}-nud.\\ 
          \textsc{neg} {one-}{\textsc{par}} {word-}{\textsc{par}} \textsc{neg-1sg} {learn-}\textsc{pst.ptcp}\\
    \glt ‘I didn’t learn a single word.’ (VeC, Mäggär’v 2007)
\z 

\ea\label{ex:vep:85} 
    \gll Ii  ole \textbf{ni} \textbf{keda}.\\ 
         \textsc{neg.3} {be.}{\textsc{cng}}{} {\textsc{neg}} {who-}{\textsc{par}}\\
    \glt ‘There isn’t anybody.’ (Ladv 2006)
\z 

In certain clause types the negative focus particle parallelly functions as a coordinating conjunction \REF{ex:vep:86} (cf. \sectref{sec:vep:4.5}) and may also be reduplicated (\ref{ex:vep:87}; cf. also \ref{ex:vep:81} above). This kind of connective negation in Veps is similar to Russian, other Balto-Slavic and Romance languages (\citealt{AuweraEtAl2021}: 46–53). In both \REF{ex:vep:86} and \REF{ex:vep:87} a negative VP manifests primary negation. In \REF{ex:vep:87} the reversed order \textsc{v} + \textsc{neg} is an additional way of reinforcing clausal negation.

\ea\label{ex:vep:86} 
    \gll  Vaštke-s \textbf{ong-he} \textbf{ni} \textbf{uštke-he} ii koki.\\ 
          calm-\textsc{ine} hook{}-\textsc{ill} \textsc{neg} lure{}-\textsc{ill} \textsc{neg.3} bite.\textsc{cng}\\
    \glt ‘In calm wheather (the fish) doesn’t bite the hook nor the lure.’ (VeC, \citealt[435]{MSFOu86})
\z 

\ea\label{ex:vep:87} 
    \gll Neižne \textbf{ni} \textbf{sö-ške} \textbf{ni} \textbf{jo-ške} ii kahte-s nedali-s. \\ 
         Girl \textsc{neg} eat.\textsc{cng}-\textsc{pfv}{} \textsc{neg} drink.\textsc{cng}-\textsc{pfv} \textsc{neg.3} two{}-\textsc{ine} week{}-\textsc{ine} \\
    \glt ‘The girl hasn’t eaten or drunk for two weeks.’ (VeS, \citealt[277]{MSFOu86})
\z 

In replies to questions the interrogative \textit{kutak} ‘how’ may operate both as a polarity item characteristic of negative phrases and reinforce negation by forming a question as in \REF{ex:vep:88}. This phrase corresponds to Russian \textit{kak ne!} (how \textsc{neg}) ‘Absolutely not!’. In Veps an enclitic focus particle \textit{{}-ak} is attached to the interrogative stem \textit{kut} ‘how’. Another interrogative \textit{mijak {\textasciitilde} mejak} ‘why’ sharing the same affix may function the same way in reinforced replies.

\ea\label{ex:vep:88}
    \ea\label{ex:vep:88a}
        \gll  Tütr{}-i{}-deke  pagiže{}-d vepsa{}-ks?\\ 
               daughter{}-\textsc{pl{}-com} speak{}-\textsc{2sg} Veps{}-\textsc{tra}\\
        \glt ‘With the daughters you speak Veps?’
    \ex\label{ex:vep:88b}
        \gll  Pagiže{}-n, \textbf{kut-ak} \textbf{ii}!?\\
              speak{}-\textsc{1sg} how-\textsc{clt} \textsc{neg.3}\\
        \glt ‘I do, why shouldn’t I!?’ (Ladv 2006)
    \z
\z 

Although Veps has some focus particles, none of them is exclusively negative. In a negative clause the general focus particle \textit{{}-gi/-ki} reinforces the statement \REF{ex:vep:89}. In writing, standardized Veps separates this affix \textit{{}-ki} with a hyphen, e.g. \textit{küzund-gi} (ask.\textsc{pst.ptcp-foc}).

\ea\label{ex:vep:89}
    \gll Neci{}-l  aka{}-l ii \textbf{küzu{}-nd{}-gi}. \\
         this-\textsc{ade} woman-\textsc{ade} \textsc{neg.3} ask-\textsc{pst.ptcp}{}-\textsc{foc}\\
    \glt ‘(S)he didn’t even ask this woman.’ (VeC, \citealt[118]{MSFOu70})
\z

\subsection{Negation, coordination and complex clauses}\label{sec:vep:4.5}

A very frequent clause type is the polar clause, which operates with a multifunctional particle \textit{ka} ‘yes; so’. Syntactically, it corresponds to subordinating constructions, in which a conjunction initiates a clause that determines the main clause. However, the particle \textit{ka} actually functions as a clause connector instead of an unambiguously subordinating conjunction (\ref{ex:vep:90}–\ref{ex:vep:91}). 

\ea\label{ex:vep:90}
    \gll Tata-in  kucu-i-n  svathuzi-i-le,  ka \textbf{ii} \textbf{mäne-n}.\\
         father-\textsc{1sg.poss} invite-\textsc{pst-1sg}  courting{}-\textsc{pl}{}-\textsc{all}  yes \textsc{neg.3} go-\textsc{pst.ptcp}\\
    \glt ‘I invited my father to courting but he didn’t go.’ (VeC, \citealt[37]{MSFOu70})
\z

\ea\label{ex:vep:91}
    \gll  Om radio, ka \textbf{ii} \textbf{pagiže} pole-s aiga-d ka.\\
         is radio yes \textsc{neg.3} speak.\textsc{cng} half{}-\textsc{ine} time-\textsc{par} yes\\
    \glt ‘There is the radio, but half of the time it doesn’t speak.’ (VeC, Ladv 2006)
\z

There is a contrastive particle \textit{e(i)ka/i(i)ka} ‘or’ that originates from the fusion of the negator and \textit{ka}. It emphasizes the polar contrast between the two clauses by pointing that the statement will conditionally happen (\ref{ex:vep:92}--\ref{ex:vep:93}). The bimorphemic conjunction corresponds to English ‘if…not’.


\ea
    \label{ex:vep:92}
    \gll  Ot-ka-t  i min-daa-n’ ela-m-ha, \textbf{eka} perti-n sarv-ii-l kändä-n. \\
         take-\textsc{imp-2pl} and \textsc{1sg-par-1sg.poss} live-\textsc{inf-ill} or house-\textsc{acc} horn-\textsc{pl-ade} turn-\textsc{1sg}\\
    \glt ‘Take me to live (with you), or I’ll turn up the house with my horns.’ (VeS, \citealt[44]{Kettunen1925})
\z

\ea
    \label{ex:vep:93}
    \gll Enamba-d ii ela-nu siid, \textbf{ika} kondii sö{}-nu{}-iž  aka{}-iže{}-n. \\
         more-\textsc{par} \textsc{neg.3} live-\textsc{pst.ptcp} then or bear eat-\textsc{pst.ptcp{}-cond.3sg} woman-\textsc{dim}{}-\textsc{acc}\\
    \glt ‘(She) didn’t live any longer then, or the bear would have eaten the woman.’ (VeN, \citealt[185]{MSFOu70})
\z

In general, affirmative and negative phrases may occur as coordinated clauses that display the standard coordinating conjunction \textit{i} ‘and’ or \textit{da} ‘and’, both borrowed from Russian. Example \REF{ex:vep:94} represents the literary standardized Veps.

\ea
    \label{ex:vep:94}
    \gll Nece{}-l vede{}-l pes{}-t’he  laps{}-i{}-d,  miše  hö pare-mb  kaz-de-iž  \textbf{da} ii läžu-te-iž. \\
        this-\textsc{ade} water-\textsc{ade}  wash-\textsc{pst.3pl} child-\textsc{pl-par} in.order.to \textsc{3pl} 
        better-\textsc{comp} grow-\textsc{3pl-cond} and \textsc{neg.3} be.sick-\textsc{3pl-cond}\\
    \glt ‘They used to wash children with this water so that they would grow better and wouldn’t get sick.’ (example taken from the newspaper \textit{Kodima} 1/2008)
\z 

\subsection{Miscellaneous aspects of negation}\label{sec:vep:4.6}

A special semantic effect can be seen in the lexically conditioned modal transformation of the verb \textit{käsk-ta} order\textsc{{}-inf} ‘order, tell, ask’ and its negation \textit{ei käsk-ta} \textsc{neg} order\textsc{{}-inf} ‘forbid’ (\ref{ex:vep:95}–\ref{ex:vep:97}). Morphosyntactically, this sentence represents a basic declarative clause, whereas the semantic change based on negation is not characteristic of other verbs.
\ea\label{ex:vep:95} 
    \gll  \textbf{Käsk-i-d}, ka mida-k e-n ota-nd?\\ 
          order-\textsc{pst-2sg} yes what-Q \textsc{neg-1sg} take-\textsc{pst.ptcp}\\
    \glt ‘You told [me] to, so why didn’t I take?’ (VeC, \citealt[107]{MSFOu70})
\z 

\ea\label{ex:vep:96} 
    \gll  Näge-d, \textbf{e{}-d} \bfseries{käske{}-nd}  vö{}-da i sünd ande{}-i kümne. \\ 
          see-\textsc{2sg} \textsc{neg-2sg} order-\textsc{pst.ptcp} bring\textsc{{}-inf} and God give-\textsc{pst}.\textsc{3sg} ten\\
    \glt ‘You see, you forbid to take along and God gave ten.’ (VeC, \citealt[108]{MSFOu70})
\z

\ea
    \label{ex:vep:97}
    \gll Iče-moi  kele-l \textbf{ii} \textbf{käs-tud}  ka openz-i-n i venä-ks da i.\\
         own-1\textsc{pl}  language-\textsc{ade} \textsc{neg.3} order-\textsc{pst.ptcp} yes learn-\textsc{pst}{}-\textsc{1sg} and Russian-\textsc{tra} and and\\%it is correct that there is 'and and', do not change.
    \glt ‘They forbid to (study) in Veps, so I studied in Russian.’ (VeC, Ladv 2007)
\z

\section{Summary}\label{sec:vep:5}

This paper has discussed a number of features characteristic of Veps negation. Like in other Finnic and, more broadly, Uralic languages, certain inherent features have been preserved in Veps, whereas several language-specific innovations can also be shown. Generally speaking, Veps constitutes an illustrative example of a dynamic system in constant change. A detailed analysis of negation patterns shows a clear distinction between verbal and non-verbal negation. In declarative sentences, Veps uses a negative auxiliary like many Uralic languages do, while a negative particle is used in non-verbal negation. The negative auxiliary inflects for person but not for tense and mood, which divides the core functions of predication between the \textsc{neg}.\textsc{aux} and the main verb. This kind of ambiguity increases the number of mutually dependent features which diachronically originate from separate units and are affected by various mechanisms of change.
    % \hc{We have changed to your new sentence here. Could you still specify what you mean by 'unit' here?}\aut{Here, a unit refers to functionally and grammatically significant entities such as negative auxiliaries, modal and finite verbs. If you consider various ways of manifesting grammatical relations, affixes or free morphemes such as adpositions are another example.}
An inflecting \textsc{neg}.\textsc{aux} is an inherent Uralic feature (cf. \citealt{MiestamoEtAl2015}), whereas the particle \textit{ni} used in non-verbal constructions is borrowed from Slavic (Russian).



The most significant language-specific characteristics include the alternating constituent order of a negative clause, the emergence of a functional distinction expressed by the change of constituent order, the morphosyntactic pattern of negative plural verbs, and the emergence of a negative existential. These are all features that distinguish the Veps negation from the neighboring languages, both genetically related and unrelated.


\begin{table}
    \begin{tabularx}{\textwidth}{lQlp{25mm}}
\lsptoprule
Form & Description & Section & Examples \\
\midrule
%\multicolumn{2}{l}{Clausal negation}\\ 
\textit{e-} & inflecting auxiliary &&\\
&(1) basic declarative sentences 
& \ref{sec:vep:2.1}
 & (\ref{ex:vep:1}--\ref{ex:vep:5}, \ref{ex:vep:8additional}--\ref{ex:vep:9}, \ref{ex:vep:53}--\ref{ex:vep:55}, \ref{ex:vep:69}, \ref{ex:vep:71}, \ref{ex:vep:90}--\ref{ex:vep:91}, \ref{ex:vep:96}--\ref{ex:vep:97}) \\ 

&(2) reversed word order V+ Neg 
 & \ref{sec:vep:2.1.2}
 & (\ref{ex:vep:10}--\ref{ex:vep:13}, \ref{ex:vep:70}, \ref{ex:vep:73}--\ref{ex:vep:78}, \ref{ex:vep:87})  \\ 

&(3) conditionals 
 & \ref{sec:vep:2.2}
 & (\ref{ex:vep:22}--\ref{ex:vep:23})  \\ 

&(4) questions 
 & \ref{sec:vep:2.2}
 & (\ref{ex:vep:24}--\ref{ex:vep:25})  \\ 

&(5) stative predications 
 & \ref{sec:vep:2.3}
 & (\ref{ex:vep:31}--\ref{ex:vep:32})  \\

&(6) subordinate clauses 
&
& (\ref{ex:vep:38}--\ref{ex:vep:39}, \ref{ex:vep:41}--\ref{ex:vep:44})  \\

\textit{iile} &fusional negating copula 
& \ref{sec:vep:2.3}
& (\ref{ex:vep:33}--\ref{ex:vep:35}, \ref{ex:vep:40}, \ref{ex:vep:50})  \\

\textit{emboo}  & fusional modal negating (\textsc{1sg})  &&\\
	
\textit{ala} & imperatives
& \ref{sec:vep:2.2}
& (\ref{ex:vep:18}--\ref{ex:vep:21})  \\

\textit{e(i)ka/i(i)ka } 
& contrastive negative particle used in subordinate clauses 
& \ref{sec:vep:4.5}
& (\ref{ex:vep:92}--\ref{ex:vep:93})  \\

%\midrule \multicolumn{4}{l}{Non-clausal negation} \\

\textit{ni }
& negative particle that can be attached to all parts-of-speech &&\\

& (1) pre- or postnominal particle 
& \ref{sec:vep:3}
& (\ref{ex:vep:51}--\ref{ex:vep:52}, \ref{ex:vep:80}--\ref{ex:vep:81}, \ref{ex:vep:85}--\ref{ex:vep:86})  \\

& (2) negative focus particle preceding both affirmative and negative VP 
& \ref{sec:vep:3.1}
& (\ref{ex:vep:57}, \ref{ex:vep:87})  \\

& (3) indefinites and quantifiers 
& \ref{sec:vep:3.2}
& (\ref{ex:vep:58}--\ref{ex:vep:67}, \ref{ex:vep:84})  \\

& (4) adverbs 
& \ref{sec:vep:3.2}
& (\ref{ex:vep:68})  \\

\textit{-ta} 
& abessive case suffix (negative instrumental) 
& \ref{sec:vep:3.3}
& (\ref{ex:vep:72}, \ref{ex:vep:74}--\ref{ex:vep:75})  \\

\textit{-toi} 
& derivative suffix of caritive adjectives 
& \ref{sec:vep:3.3}
& (\ref{ex:vep:73})  \\

%\midrule \multicolumn{4}{l}{Other aspects of negation} \\

\textit{igas} 
& negative adverb reinforcing negation 
& \ref{sec:vep:3.2}, \ref{sec:vep:4.2}
& (\ref{ex:vep:69})  \\

\textit{enambad} 
& negative adverb reinforcing negation 
& \ref{sec:vep:3.2}
& (\ref{ex:vep:70}--\ref{ex:vep:71}, \ref{ex:vep:83})  \\

% [morphosyntactic rule] 
% & the mandatory use of the partitive case in object marking in negative clauses 
% & \ref{sec:vep:4.3}
% & (\ref{ex:vep:80}--\ref{ex:vep:83})  \\

\textit{-ak} 
& polar focus particle in interrogatives 
& \ref{sec:vep:4.4}
& (\ref{ex:vep:88})  \\

\textit{-gi}
& reinforcing focus particle 
& \ref{sec:vep:4.4}
& (\ref{ex:vep:89})  \\
\lspbottomrule
    \end{tabularx}
    \caption{Summary of negation in Veps}
    \label{tab:vep:summary}
\end{table}

\section*{Abbreviations}

\begin{tabularx}{.45\textwidth}{lQ}
1 & first person \\
2 & second person \\
3 & third person \\
{\scshape abe} & abessive\\
{\scshape acc} & accusative\\
{\scshape ade} & adessive\\
{\scshape adjz} & adjectivizer\\
{\scshape all} & allative\\
{\scshape aux} & auxiliary\\
{\scshape car} & caritive\\
{\scshape caus} & causative\\
{\scshape cng} & connegative\\
{\scshape com} & comitative\\
{\scshape cmpr} & comparative\\
{\scshape cond} & conditional\\
{\scshape cont} & continuative\\
{\scshape cop} & copula\\
{\scshape dim} & diminutive\\
{\scshape ela} & elative\\
{\scshape clt} & enclitic\\
{\scshape ess} & essive\\
{\scshape ex} & existential\\
{\scshape foc} & focus\\
{\scshape freq} & frequentative\\
{\scshape fut} & future\\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
{\scshape gen} & genitive\\
{\scshape ill} & illative\\
{\scshape imp} & imperative\\
{\scshape ine} & inessive\\
{\scshape inf} & infinitive\\
{\scshape mod} & modal\\
{\scshape mom} & momentative\\
{\scshape neg} & negative\\
{\scshape p} & person marker\\
{\scshape par} & partitive\\
{\scshape pass} & passive\\
{\scshape pfv} & perfective\\
{\scshape pl} & plural\\
{\scshape prol} & prolative\\
{\scshape pst} & past\\
{\scshape ptcp} & participle\\
{\scshape poss} & possessive suffix\\
{\scshape q} & question marker\\
{\scshape refl} & reflexive\\
{\scshape sg} & singular\\
{\scshape tra} & translative\\
{VeC} & Central Veps\\
{VeN} & Northern Veps\\
{VeS} & Southern Veps\\
\\
\end{tabularx}

\section*{Fieldwork data}
The fieldwork data were collected in the following villages and are archived at the Institute for the Languages of Finland (\href{https://www.kotus.fi/en}{https://www.kotus.fi/en}).
\medskip

\noindent
\begin{tabularx}{\textwidth}{lllX}
village & district / rayon & region / oblast' & year \\
\midrule
Kurba & Podporožskiy  & Leningradskaya  & 2006 \\
Ladv & Podporožskiy  & Leningradskaya  & 2005, 2006, 2007, 2014 \\
Mäggär’v & Podporožskiy  & Leningradskaya  & 2007 \\
Päžar’ & Babaevskiy  & Vologodskaya  & 2005, 2011 \\
\end{tabularx}


\sloppy\printbibliography[heading=subbibliography,notkeyword=this]

\end{document}
