\documentclass[output=paper,colorlinks,citecolor=brown]{langscibook}
\ChapterDOI{10.5281/zenodo.18234099}

\author{Magdalena Lemus Serrano\affiliation{Dynamique du Langage (CNRS/Université Lyon 2)} and Françoise Rose\affiliation{Dynamique du Langage (CNRS/Université Lyon 2)}}
\title{Negation in Yukuna}  
\abstract{This study provides an overview of negation in Yukuna, an Arawak language of Colombian Amazonia. Yukuna has two main negation strategies, Standard Negation (\textit{unká...-la}) and non-verbal negation (\textit{unká...kalé}), that perfectly map onto the verbal versus non-verbal clause distinction in the language. The same markers used in these two strategies cover most domains of negation, from polar questions, non-declaratives, non-main clauses and constituent negation. However, despite the homogeneity of markers, negation in main declarative verbal clauses is symmetric, while negation in non-main and non-finite clauses displays various types of constructional asymmetry. The most frequent type of asymmetry concerns the marking of finiteness in affirmatives versus negatives, with the particularity that in Yukuna, it is the affirmative clause that is overtly marked for nominalization, and hence, shows reduced finiteness in comparison with the corresponding negative, unmarked for nominalization. The marking of negation is thus strongly intertwined  with finiteness and the verbal versus non-verbal distinction, two central issues that crosscut the grammar of the language. 
\keywords{Arawak, non-verbal predication, finiteness, subordination}
}

\IfFileExists{../localcommands.tex}{%hack to check whether this is being compiled as part of a collection or standalone
   \input{../localpackages}
   \input{../localcommands}
   \input{../localhyphenation}
    \addbibresource{../localbibliography.bib}
%    \bibliography{localbibliographyMLS}
    \togglepaper[23]
}{}

\begin{document}
\maketitle
\section{The language}
Yukuna (ISO: \href{https://iso639-3.sil.org/code/ycn}{ycn}, Glottocode: \href{https://glottolog.org/resource/languoid/id/yucu1253}{yucu1253}) is a North-Amazonian Arawakan language spoken by less than a thousand speakers in various communities along the Mirití-Paraná river in South Eastern Colombia. The Yukuna language is spoken by the Yukuna and Matapí groups, who are in intense, long-term contact with the Tanimuka (Tukanoan) speaking groups Tanimuka and Letuama.  This study is based on ongoing work on the Yukuna language by the first author.\footnote{The Yukuna documentation and description project is led by the first author (Lemus Serrano), under the supervision of the second author (Rose). The first author carried out data collection and annotation, extracted and analyzed the data on negation, and wrote the first manuscript of this chapter. The second author provided critical feedback on the analysis of the data on negation, and contributed to the final version of the manuscript.}
Typologically, Yukuna is a nominative-accusative language with a canonical SVO constituent order. The lexical word classes are nouns, verbs, adverbs, adjectives, and postpositions. Core arguments are not case-marked, and obliques (henceforth OBL) are marked with postpositions. The nominative argument of verbs (henceforth S) is obligatorily encoded in finite clauses either with a bound person index on the verb, or by an overt S NP rigidly placed immediately before the verb. This marking strategy is identical to the encoding of possessors of nouns, and arguments of postpositions. The accusative argument of verbs (henceforth P) is neither obligatorily encoded, nor indexed on the verb, and its placement within the clause is flexible. The main encoding properties of S, P and obliques are summarized in \tabref{yukuna:tableArgEnc}.\footnote{We use the P label for the accusative argument of transitive verbs, but for practical purposes, we use O when in reference to the typology of word order.}

\begin{table}
\caption{Argument encoding features}
\label{yukuna:tableArgEnc}
 \begin{tabular}{llll} 
 \lsptoprule									
		&	Indexation	&	Role marking	&	Word order	\\
\midrule									
S	&		Yes &	No	&	[SV]	\\
P	&		No	&	No	&	[SV] P {\textasciitilde}  P [SV]	\\
OBL	&		No	&	Yes	&	[SV] OBL {\textasciitilde}  OBL [SV]\\
\lspbottomrule									
 \end{tabular}
\end{table}

All examples come from firsthand data collected in various Yukuna speaking communities between 2015 and 2018. The corpus contains roughly five hours of transcribed and translated texts, from which we extracted approximately 300 instances of negation marking of various types.
The examples are transcribed with a slightly modified version of the practical alphabet used in the Yukuna dictionary (\citealt{SchauerEtAl2005}), based on Spanish.\footnote{The following alphabetic conventions are used: <j> /h/, <ñ> /ɲ/, <y> /j/, <V'> /V̰/ (creaky vowel), <Ch> /Cʰ/ (aspirated plosive), <jC> /C̥/ (voiceless sonorant)} High tone is transcribed with an acute accent, but its surface manifestation is very variable, so the same morpheme may be transcribed with or without an accent depending on the context (e.g. past tense \textit{-cha} is found in examples both as \textit{-cha} and as \textit{-chá}). Lastly, each example contains a source given in brackets indicating the name of its audio file, and the ELAN/Flex line. Examples taken from elicitation come from the first author's field notes, and their source is indicated with the notebook and page number. The description of negation in Yukuna in this chapter is based on the questionnaire provided by the editors \parencitetv{chapters/questionnaire} and mirrors its structure.

\section{Clausal negation} 

This section is devoted to the description of various clause-level negation strategies. Here, as in the rest of the chapter, each negation strategy is described according to the guidelines and terminology used in \citet{Miestamo2005,chapters/questionnaire,Miestamo2017}.

\subsection{Standard negation}\label{yukuna:SN}

Negation in declarative main verbal clauses, or Standard Negation (SN), is encoded by a pre-verbal particle \textit{unká} and a verbal suffix \textit{-la}. 
The negation particle \textit{unká} is phonologically free and morphologically simple in main clauses, unlike its behavior in some subordinate clauses (\sectref{yukuna:negsub}). The negation suffix \textit{-la} obligatorily co-occurs with the pre-verbal particle in SN. Therefore, in terms of number of markers and their position with respect to the verb, Yukuna is classed among the type of languages  with obligatory double negation (Neg[V-Neg]) in \citet{Dryer2013-wals143}. 
Example (\ref{yukuna:ex:53.33a}) illustrates an affirmative transitive declarative clause with an overt S NP and a pronominal P NP. Example (\ref{yukuna:ex:189.177}) shows a negative transitive verbal clause, and illustrates the prototypical position of negative markers in SN. 

\ea\label{yukuna:ex:53.33a}
\gll Kája 	ri=pirá		nó-cha	ri=ikhá.\\
already \textsc{3sg.nf}=pet kill-\textsc{pst} \textsc{3sg.nf=pro}\\
\glt ‘His pet already killed him.’ [ycn0053,33]
\z

\ea\label{yukuna:ex:189.177}
\gll \textbf{Unká} wa=wáta-\textbf{la} kéelé kujnú. \\
\textbf{\textsc{neg}} \textsc{1pl}=want-\textbf{\textsc{neg}} \textsc{dem.med} cassava \\
\glt ‘We do not want that cassava.’ [ycn0189,177]
\z

The prototypical placement of the particle \textit{unká} within the clause is after sentence connectors and before the S NP, which typically immediately precedes the verb. This negator can be described as a clause-initial (before all arguments) particle, non-adjacent to the verb. In terms of the relative position of negators with respect to the verb and its arguments, Yukuna belongs to the NegS[V-Neg]O type (type 8) among languages with canonical SVO word order in the typology in \citet{Dryer2013-wals144}. The only exceptions to this word order are attested in the focalization of core arguments, where either the S or P NPs are exceptionally placed before the particle \textit{unká}, as in (\ref{yukuna:ex:189.67a}). Despite these variations in word order, there are important restrictions concerning the placement of arguments that apply in negative as well as in affirmative polarity (see \sectref{yukuna:INDF} on the placement of indefinite pro-forms).

\ea\label{yukuna:ex:189.67a}
\gll  É na=apó-chi-ya=o phiyúké, ri=ikhá \textbf{unká} apó-\textbf{la}-cha. \\
then \textsc{3pl}=wake\_up-\textsc{caus}-\textsc{pst}=\textsc{mid} all \textsc{3sg.nf}=\textsc{pro} \textbf{\textsc{neg}} wake\_up-\textbf{\textsc{neg}}-\textsc{pst}\\
\glt ‘Then they all woke up, (but) \textit{he} did not wake up.’ [ycn0189,67]
\z
The negation suffix \textit{-la} is placed after valency markers and before tense. This suffix is compatible with almost all inflectional markers of finite verbs. However, crucially, \textit{-la} is placed in the same paradigmatic position as clausal nominalizers \textit{-ka} \textsc{ev.nmlz}, and \textit{-kare} \textsc{part.nmlz}, as well as with the past imperfective suffixes \textit{-khe} and \textit{-jĩka}.\footnote{These past imperfective suffixes are both very likely diachronically related to nominalizer \textit{-ka}. However, while \textit{-khe} appears to have grammaticalized as a main clause marker, more data is needed to assess the synchronic features of \textit{-jĩka}. See \sectref{yukuna:diachrony} for a diachronic account of the negation system.} This paradigmatic position of the negative suffix \textit{-la} affects the encoding of negation in all contexts involving markers within the same slot, and leads to multiple instances of asymmetries in non-declarative (see \sectref{yukuna:ndecl}), and subordinate clauses (see \sectref{yukuna:negsub}). In main clauses, the use of the past imperfective suffix \textit{-khe} requires negation to be encoded exclusively with \textit{unká} as in (\ref{yukuna:ex:108.149}). Note that suffix \textit{-khe} is found in affirmative clauses as well (\ref{yukuna:ex:117.80}), and it is thus not a portmanteau negative marker as suggested in \citet[]{Michael2014}.\footnote{This erroneous description of the past imperfective suffix is likely due to the ambiguous phrasing in \citet[314]{SchauerEtAl2005}, where the authors suggest that the negative suffix \textit{-la} is replaced by \textit{-khe} in the past imperfective.}
\ea\label{yukuna:ex:108.149}
\gll  \textbf{Unká} na=amá-\textbf{khe} kéelé kájé itewí ri=wakajé.  \\
\textbf{\textsc{neg}} \textsc{3pl}=see-\textbf{\textsc{pst.ipfv}} \textsc{dem.med} type palm\_species \textsc{3sg.nf}=time\\ 
\glt ‘They did not usually see that type of palm trees at that time.’ [ycn0108,149]
\z

\ea\label{yukuna:ex:117.80}
\gll  na=nó-\textbf{khe} wa=jló kuwajárí,  ye'é {...} \\
\textsc{3pl}=kill-\textbf{\textsc{pst.ipfv}} \textsc{1pl}=for rodent\_species armadillo\\ 
\glt ‘they used to kill for us pacas, armadillos{...}’ [ycn0117,80]
\z

In sum, negative declarative verbal clauses have the same word order and argument encoding as affirmative clauses, and hence, they are structurally symmetric. Additionally, the presence of SN markers does not affect the marking of grammatical categories in any way. The paradigms of grammatical markers available in both SN and affirmative declarative verbal clauses are thus symmetric as well. 

\subsection{Negation in non-declaratives}\label{yukuna:ndecl}

This section provides an overview of negation in various types of non-declarative moods and speech acts. 

\subsubsection{Negative imperatives}

Yukuna has a dedicated prohibitive construction with suffix \textit{-niña}. This strategy differs from the second person imperative construction as the latter receives no overt imperative marking. However, they both require the verb to be unmarked for TAM, the S argument to be encoded with a person index, and the P argument to be postverbal. Prohibitives are then both constructionally and paradigmatically symmetric with respect to second person imperatives. Examples (\ref{yukuna:ex:68.149}--\ref{yukuna:ex:108.41a}) illustrate the affirmative imperative and prohibitive constructions respectively. 

\ea\label{yukuna:ex:68.149}
\gll Pi=ajñá ri=ikhá!\\
\textsc{2sg}=eat \textsc{3sg.nf}=\textsc{pro} \\
\glt ‘Eat that!’ [ycn0068,149]
\z

\ea\label{yukuna:ex:108.41a}
\gll  Pi=ajñá-\textbf{níña} ri=ikhá!\\
\textsc{2sg}=eat-\textbf{\textsc{proh}} \textsc{3sg.nf}=\textsc{pro}\\
\glt ‘Don't eat that!’ [ycn0108,41]
\z

The prohibitive clearly differs from the SN construction \mbox{\textit{unká}}{...}\mbox{\textit{-la}} (\sectref{yukuna:SN}), but it is symmetric with the imperative construction. Hence, Yukuna can be said to belong to the type II in the typology of \citet{AuweraEtAl2013},  among languages where the prohibitive uses the same verbal form as the second singular person imperative and a sentential negative strategy not found in indicative declaratives. 

\subsubsection{Negative interrogatives}\label{yukuna:interrog}

Affirmative polar questions in Yukuna are encoded with a construction using an interrogative particle copula \textit{é} plus a nominalized verb form with suffix \textit{-ka} \textsc{ev.nmlz} (\ref{yukuna:ex:108.139}).\footnote{The interrogative particle \textit{é} is homonymous with the postposition \textit{é} ‘at’, the sentence connector \textit{é} `then' and the temporal adverbial subordinator \textit{é}.} This construction is identical to polar questions with non-verbal predicates as in (\ref{yukuna:ex:504.51}). 

\ea\label{yukuna:ex:108.139}
\gll   É pi=jĩ'-chá-ka jíña? \\
\textsc{q} \textsc{2sg}=grab-\textsc{pst}-\textsc{ev.nmlz} fish \\
\glt ‘Did you grab fish?’ [ycn0108,139]
\z

\ea\label{yukuna:ex:504.51}
\gll   É palá-ni? \\
\textsc{q} good-\textsc{nf}\\
\glt ‘Is it good?’ [ycn0504,51]
\z

In contrast, negative polar questions are encoded by using a finite verb form carrying SN markers \textit{unká...-la} (\ref{yukuna:ex:41.52}). Negative polar questions thus radically differ from their affirmative counterparts (\ref{yukuna:ex:108.139}). The major difference between these constructions lies in the marking of finiteness, with affirmatives using an overtly nominalized verb form, and negatives using a finite verb form. 

\ea\label{yukuna:ex:41.52}
\gll  \textbf{Unká} pi=amá-\textbf{la} ná iphá-ká=no majṍ? \\
\textbf{\textsc{neg}} \textsc{2sg}=see-\textbf{\textsc{neg}} \textsc{indf} arrive-\textsc{ev.nmlz}=\textsc{hab} here \\
\glt ‘Haven't you seen who it is that came here?’ [ycn0041,52]
\z

Despite the structural difference with affirmative polar questions, negative polar questions do not show any major differences with negative declaratives, which corresponds to the most recurrent cross-linguistic scenario according to the preliminary survey in \citet{Miestamo2009neg.interr}.

\subsubsection{Negation in the potential mood}\label{yukuna:pot}

Potential modality in affirmative polarity is often encoded by combining the \textit{intentive} particle \textit{jlá} ‘try to’ with a nominalized verb form using \textit{-kare} \textsc{part.nmlz} as shown in example (\ref{yukuna:ex:53.72}). 

\ea\label{yukuna:ex:53.72}
\gll Iná i'jra-káre jlá ri=ikhá. \\
\textsc{gnr.pro} lift-\textsc{part.nmlz} \textsc{intv} \textsc{3sg.nf}=\textsc{pro}  \\
\glt ‘One can lift it up.’ [ycn0053,72]
\z

In the negative polarity, however, Yukuna uses a finite verb carrying the same markers used in SN, obligatorily preceded by an indefinite pro-adverb – most frequently \textit{méla'jẽ́} ‘any way’, but other pro-adverbs are found in this function as well – as in (\ref{yukuna:ex:63.111}). The intentive particle \textit{jlá} is optional in the negative. 

\ea\label{yukuna:ex:63.111}
\gll Unká \textbf{méla'jẽ́} pi=i'jna-lá na=jwa'té. \\ 
\textsc{neg} \textsc{indf}.manner \textsc{2sg}=go-\textsc{neg} \textsc{3sg}=with  \\
\glt ‘You cannot go with them.’ (lit. ‘You do not go with them in any way.') [ycn0063,111]
\z

Similarly to interrogatives, the potential mood shows an asymmetry in the marking of finiteness with nominalized verb forms in the affirmative versus finite verb forms using SN in the negative.

Lastly, this construction only differs structurally from SN in the obligatory use of the negative indefinite. The features of this pro-form are identical to those of other negative indefinites (\sectref{yukuna:INDF}). Like other indefinites, the marker in this construction can also be negated through constituent negation, as discussed in (\sectref{yukuna:scope}).
 
\subsection{Negation in stative predications}\label{yukuna:negnv}
In Yukuna, all word classes besides verbs can be used predicatively: nouns, adverbs, adjectives and postpositions. These predicates may be encoded through two distinct morphosyntactic strategies: a zero copula construction, and a verbal copula construction, following the terminology of \citet{Stassen1997}. The zero copula and verbal copula constructions differ on all grounds, including their strategies of negation marking. The choice of an encoding strategy depends on several criteria: time of reference, polarity and functional subcategory of non-verbal predication.

In the zero copula construction, the non-verbal predicate is placed in clause initial position, followed by its sole argument, with no finite verbal element at all (\ref{yukuna:ex:63.144}). The non-verbal predicate in the zero copula construction is negated with a strategy exclusive to non-verbal constituents, double marked with particle \textit{unká} at the left edge of the predicate, and particle \textit{kalé} at its right edge (\ref{yukuna:ex:68.253}). We refer to this construction as the non-verbal negation strategy. This is the same strategy used in constituent negation described in (\sectref{yukuna:scope}). The non-verbal negation strategy is symmetric on all levels with respect to the affirmative zero copula construction, as the only difference between them is the presence of the negation markers. 

\ea\label{yukuna:ex:63.144}
\gll A'jne-jí tá ri=ikhá.\\
food-\textsc{nposs} \textsc{emph} \textsc{3sg.nf}=\textsc{pro} \\
\glt ‘It is food.’ [ycn0063,144]
\z

\ea\label{yukuna:ex:68.253}
\gll \textbf{Unká}	inau'ké \textbf{kalé} 	nu=ikhá. \\
\textbf{\textsc{neg}}	person	\textbf{\textsc{nvb.neg}} \textsc{1sg}=\textsc{pro} \\
\glt ‘I am not a person.’ [ycn0068,253]
\z

In contrast, in the verbal copula construction, there is an inflecting copula \textit{i'ma}{\textasciitilde}\textit{i'mi}, and the single argument is encoded like the S NP in verbal clauses (\ref{yukuna:ex:151.2}). Non-verbal predicates encoded with the verbal copula construction are negated with the same markers used in SN (\textit{unká}{...}\textit{-la}), as in (\ref{yukuna:ex:108.77}). This negation strategy clearly differs from the non-verbal negation strategy in zero copula clauses, but both are double-marked and share a common marker, the negative particle \textit{unká}.

\ea\label{yukuna:ex:151.2}
\gll Inau'ké i'mi-chá.\\
person \textsc{cop-pst}\\
\glt ‘There were people.’ [ycn0151,2]
\z

\ea\label{yukuna:ex:108.77}
\gll Kája \textbf{unká} inau'ké i'ma-\textbf{lá}-cha.\\
already \textbf{\textsc{neg}} person \textsc{cop}-\textbf{\textsc{v.neg}}-\textsc{pst} \\
\glt ‘There were no people anymore.’ [ycn0108,77]
\z

The zero copula and verbal copula constructions are used in alternation to encode all six functional types of non-verbal predication, namely, equation, proper inclusion, attributive, locative, existential, and possessive predication. The choice of encoding strategy depends on an interplay between the functional type of non-verbal predication, polarity and time of reference. The preferred encoding strategy (Cop for verbal copula, and Zero for the zero copula constructions) per functional type of non-verbal predication and polarity are summarized in \tabref{yukuna:tableNVNEG2}. In short, in the affirmative, the zero copula construction is the preferred strategy in the present, while the verbal copula construction is the preferred strategy for non-present time of reference. This holds true across all six functional types of non-verbal predication. In the negative polarity, however, there is a distinction between equation, inclusion and attributive predicates on the one hand, and locative, existential and possessive predicates on the other. The former group follows the same pattern as in the affirmative polarity, with the zero copula strategy used in the present, and the verbal copula strategy elsewhere. The latter group, however, only uses the verbal copula strategy regardless of time of reference.

\begin{table}
\caption{Preferred encoding per predication type, polarity and tense}
\label{yukuna:tableNVNEG2}
\begin{tabular}{c|c|c||c|c|}
\cline{2-5}
                            & \multicolumn{2}{c||}{Affirmative\strut}         & \multicolumn{2}{c|}{Negative}            \\ \cline{2-5}
                            & Present                 & Non-present\strut               & Present                 & Non-present               \\ \hline
\multicolumn{1}{|l|}{Equation\strut}    & \multirow{6}{*}{Zero copula} & \multirow{6}{*}{Verbal copula} & \multirow{3}{*}{Zero copula} & \multirow{3}{*}{Verbal copula} \\ \cline{1-1}
\multicolumn{1}{|l|}{Inclusion\strut}    &                     &                    &                     &                    \\ \cline{1-1}
\multicolumn{1}{|l|}{Attributive\strut}  &                     &                    &                     &                    \\ \cline{1-1} \cline{4-5}
\multicolumn{1}{|l|}{Location\strut}   &                     &                    & \multicolumn{2}{c|}{\multirow{3}{*}{Verbal copula}}  \\ \cline{1-1}
\multicolumn{1}{|l|}{Existential\strut} &                     &                    & \multicolumn{2}{l|}{}                    \\ \cline{1-1}
\multicolumn{1}{|l|}{Possessive\strut}  &                     &                    & \multicolumn{2}{l|}{}                    \\ \hline
\end{tabular}
\end{table}%\hc{@LSP the lines around the table do not look good, could you make it look better?}

Indeed, the pair of affirmative and negative proper inclusion predicates in the present tense in (\ref{yukuna:ex:63.144}--\ref{yukuna:ex:68.253}) are both encoded with the zero copula strategy. In contrast, the pair of affirmative and negative existential predicates in the present tense in (\ref{yukuna:ex:108.101}--\ref{yukuna:ex:18.13}) show that each polarity is encoded with a different strategy, with the zero copula construction in the affirmative, and the verbal copula (with negative markers \textit{unká...-la}) in the negative. 

\ea\label{yukuna:ex:108.101}
\gll Ri=é=ko inau'ké.\\
\textsc{3sg.nf}=at=\textsc{emph} person \\
\glt ‘There are people there.’ [ycn0108,101]
\z

\ea\label{yukuna:ex:18.13}
\gll \textbf{Unká} kája jápá-kaje i'ma-\textbf{lá}.\\
\textbf{\textsc{neg}} \textsc{emph} work-\textsc{ev.nmlz} \textsc{cop}-\textbf{\textsc{v.neg}}\\
\glt ‘There is no work.’ [ycn0018,13]
\z

In sum, there are two distinct strategies used in the negation of non-verbal predicates: the non-verbal strategy with \textit{unká}{...}\textit{kalé}, and the verbal strategy with the copula \textit{i'ma}{\textasciitilde}\textit{i'mi} carrying the same markers as in SN (\textit{unká}{...}\textit{-la}). The distribution of these strategies depends on an interplay between tense and functional sub-type of non-verbal predication. Crucially, equation, proper inclusion and attributive predicates allow both the non-verbal negation strategy and the verbal copula with SN markers depending on tense, while location, existential and possessive predicates can only be negated with the verbal copula and SN markers.

\subsection{Negation in non-main clauses}\label{yukuna:negsub}
In this section, we describe the different negation strategies used in the negation of purpose of motion clauses (\sectref{yukuna:purp.mot}), adverbial subordinates with subordinator \textit{lojé} (\sectref{yukuna:purp.loje}), adverbial subordinates with other subordinators (\sectref{yukuna:other.advz}), and complement clauses of perception verbs (\sectref{yukuna:compl.cl}). Negation in other types of subordinate clauses is not discussed, as they are not attested in the corpus of narratives. 


\subsubsection{Purpose of motion clauses with -\textit{je}}\label{yukuna:purp.mot}

Yukuna uses the verbal suffix \textit{-je} to encode purpose of motion clauses. The verb marked with this suffix lacks all morphosyntactic features of finite main verbs. Crucially, there is no S marking nor tense in the subordinate verb, which is obligatorily understood as sharing the same subject argument and time-frame as the main clause (\ref{yukuna:ex:41.12}). The resulting construction occupies the role of an oblique argument of the main verb. 

\ea\label{yukuna:ex:41.12}
\gll Ri=i'jĩ-chá=no ri=amá-je.\\
\textsc{3sg.nf}=go-\textsc{pst}=\textsc{hab} \textsc{3sg.nf}=see-\textsc{purp.mot} \\
\glt ‘He kept going to see it.’ (lit. ‘He always went to its seeing.’) [ycn0041,12]
\z

The negation of purpose of motion clauses is encoded with the non-verbal negation strategy \textit{unká}{...}\textit{kalé} as in (\ref{yukuna:ex:58.28}). This strategy is the same one used in the negation of specific constituents in verbal clauses as in example (\ref{yukuna:ex:108.87}) (see also \sectref{yukuna:scope}). 

\ea\label{yukuna:ex:58.28}
\gll \textbf{Unká} kapichá-\textbf{je}=o \textbf{kalé} pi=i'jná, i'ma-jé i=i'jná\\
\textbf{\textsc{neg}} die-\textbf{\textsc{purp.mot}}=\textsc{mid} \textbf{\textsc{nvb.neg}} \textsc{2sg}=go live-\textsc{purp.mot} \textsc{2pl}=go\\
\glt ‘You go (there) not to die, you go (there) to live.’ [ycn0058,28]
\z

\ea\label{yukuna:ex:108.87}
\gll \textbf{Unká} ya'jnáje \textbf{kalé} kháãjĩ́-una i'jná.\\
\textbf{\textsc{neg}} far.towards \textbf{\textsc{nvb.neg}} \textsc{dem.prox}-\textsc{pl} go\\
\glt ‘These ones did not go far.’ (lit. ‘These ones went not far.') [ycn0108,87]
\z

Negative purpose of motion clauses are thus entirely symmetrical in comparison with their affirmative counterparts, and the negation strategy used is identical to the construction used in the negation of non-verbal constituents. 

\subsubsection{Purposive clauses with \textit{lojé}}\label{yukuna:purp.loje}

A second type of adverbial subordinate clause in Yukuna are purposive subordinates, encoded with event nominalizer \textit{-ka} and the subordinating particle \textit{lojé}. The event nominalizer \textit{-ka} is used to create clausal nominalizations, used in a variety of subordinating constructions (see also \sectref{yukuna:other.advz}). Nominalizations with \textit{-ka} \textsc{ev.nmlz} retain many features of finite clauses (core and oblique arguments, tense and aspect marking), while showing non-finite features typical of nominalizations in the language (loss of modality marking, possibility of combining with demonstratives and postpositions) \citep[242--243]{Lemus2020}. In purposive subordinate clauses, the verb root carrying nominalizer \textit{-ka} is followed by the purposive subordinating particle \textit{lojé} as in (\ref{yukuna:ex:108.41}), and, optionally, the entire nominalized clause is followed by the postposition \textit{penáje} `for’ (\ref{yukuna:ex:41.39}).\footnote{Subordinating particles have grammaticalized from postpositions, and are in fact synchronically homonymous with postpositions \citep[306--309]{Lemus2020}.} 

\ea\label{yukuna:ex:108.41}
\gll Ri=moto'-chá ri=jlú-wa ri=ikhá, ri=ajñá-ká lojé.\\
\textsc{3sg.nf}=cook-\textsc{pst} \textsc{3sg.nf}=for-\textsc{refl} \textsc{3sg.nf}=\textsc{pro} \textsc{3sg.nf}=eat-\textsc{ev.nmlz}  \textsc{purp}  \\
\glt ‘He cooked that for himself, in order to eat.’ [ycn0108,41]
\z

\ea\label{yukuna:ex:41.39}
\gll Ri=ikhá ri=pití-ya=nó na=jló na=ajñá-ká lojé penáje.\\
\textsc{3sg.nf}=\textsc{pro} \textsc{3sg.nf}=take-\textsc{pst}=\textsc{hab} \textsc{3pl}=for  \textsc{3pl}=eat-\textsc{ev.nmlz}  \textsc{purp} for  \\
\glt ‘He always brought that to them for them to eat.’ [ycn0041,39]
\z

Given the shared paradigmatic position between nominalizer \textit{-ka} and negative suffix \textit{-la}, most cases involving this nominalizer affect the encoding of negation. In the case of purposive clauses, negation is encoded with a portmanteau negative purposive subordinator \textit{piyá}, used instead of \textit{lojé} after the nominalized verb with \textit{-ka} as in (\ref{yukuna:ex:42.75}). There are no constructional differences between the negative purposive and its affirmative counterpart. 
%However, given that the marking of the purposive is different, we consider this strategy as showing paradigmatic\mm{I don't think this should be analysed as paradigmatic as it is not the case that some distinctions are lost or the paradigm is reorganized. Using a portmanteau marker that replaces the marker used in the positive clause is quite usual constructional asymmetry in my classification.} A/Cat asymmetry, as the asymmetry concerns the marking of a grammatical category. \aut{Noted. Please delete this final sentence.}

\ea\label{yukuna:ex:42.75}
\gll Kéchámi wa=la'á píño apú 	ri=tajná-ka=o \textbf{piyá}.\\
then \textsc{1pl}=do again other \textsc{3sg.nf}=finish-\textsc{ev.nmlz}=\textsc{mid}  \textbf{\textsc{neg.purp}} \\
\glt ‘Then we do another one again for it not to finish.’ [ycn0042,75]
\z

\subsubsection{Other adverbial subordinates}\label{yukuna:other.advz}

In addition to purposive clauses with the subordinator \textit{lojé}, Yukuna has several other adverbial subordinate clauses encoded similarly, with nominalizer -\textit{ka} plus a subordinating particle that determines the semantic link between the embedded nominalized clause and the main clause: conditional \textit{chú} (\ref{yukuna:ex:119.28}), cause markers \textit{aṹ}, \textit{lé}, and \textit{pachá} (\ref{yukuna:ex:63.207}).

\ea\label{yukuna:ex:119.28}
\gll ri=ikhá pi=a'-ká chú na=a'jné {...}\\
\textsc{3sg.nf=pro} \textsc{2sg}=give-\textsc{ev.nmlz} \textsc{cond} \textsc{3pl}=food \\
\glt ‘if you give that (to them) as their food...’ [ycn0119,28]
\z

\ea\label{yukuna:ex:63.207}
\gll {...} kawale'ké wáni ri=i'ma-ká pachá\\
{} greedy \textsc{emph} \textsc{3sg.nf=cop-ev.nmlz} \textsc{cause} \\
\glt ‘because he is very greedy’ [ycn0063,207]
\z
The negation of these clauses is encoded with the same markers used in SN \textit{unká}{...}\textit{-la}. However, the resulting construction shows some important differences with the affirmative subordinate clause. Firstly, the nominalizer \textit{-ka} used in the affirmative subordinate is absent from the negative subordinate; and secondly, the subordinator which was placed after the nominalized verb in the affirmative is placed after the negation particle \textit{unká} in the negative, as illustrated in (\ref{yukuna:ex:119.27}--\ref{yukuna:ex:108.265}).

\ea\label{yukuna:ex:119.27}
\gll eyá \textbf{unká} chú pi=matha'-\textbf{lá} ka'jnó káru\\
then \textbf{\textsc{neg}} \textsc{cond} \textsc{2sg}=cut-\textbf{\textsc{v.neg}} \textsc{dub} palm\_species  \\ 
\glt ‘if you don't cut uh{...} the palm leaves...’ [ycn0119,27]
\z

\ea\label{yukuna:ex:108.265}
\gll \textbf{unká} pachá nu=e'wé tá iphá-\textbf{lá}-cha \\
\textbf{\textsc{neg}} \textsc{cause} \textsc{1sg}=brother \textsc{emph} arrive-\textbf{\textsc{neg}}-\textsc{pst} \\
\glt ‘because my brother didn't arrive{...}’ [ycn0108,265]
\z

The asymmetry of this negation strategy with regard to its affirmative counterpart concerns the category of finiteness, with overt nominalizers in the affirmative vs. lack of nominalizing morphology in the negative. The presence of nominalizers has an impact on the degree of finiteness of the lexical verb (reduced in the affirmative with respect to the negative), and leads to structural differences in the placement of subordinating particles in the subordinate clause (verbs with \textit{-la} do not directly combine with postpositions/subordinating particles). However, despite concerning the category of finiteness, the asymmetries identified in Yukuna do not fit into the A/Fin type as defined in \citet[73]{Miestamo2005}. Indeed, this label is restricted to the cross-linguistically common cases where the \textit{lexical verb} in declarative main clauses shows reduced finiteness in the negative with respect to the corresponding affirmative. 
 
An additional negation strategy found in adverbial subordinate clauses shows a yet more ambiguous use of the negator \textit{unká}. This is the case in conditional subordinate clauses with event nominalizer \textit{-ka} and subordinator \textit{é}, as shown in (\ref{yukuna:ex:53.101}). 

\ea\label{yukuna:ex:53.101}
\gll pi=ka'-jĩ-\textbf{ka} \textbf{é}	kajrú 	ri=ikhá \\
\textsc{2sg}=throw-\textsc{fut}-\textbf{\textsc{ev.nmlz}} \textbf{\textsc{cond}} a\_lot \textsc{3sg.nf}=\textsc{pro} \\
\glt ‘if you throw it a lot{...}’ [ycn0053,101]
\z

The negation of this conditional clause is encoded exclusively with negator \textit{unká}, making this the second negation strategy where \textit{unká} is used without the negation suffix \textit{-la} (see \REF{yukuna:ex:108.149}, on the past imperfective suffix \textit{-khe}). The subordinate verb loses its overt nominalization marking, and unlike the preceding cases, it does not carry the negative suffix \textit{-la}. Lastly, and more interestingly, this construction also involves constructional asymmetry, where some of the morphology carried by the verb in the affirmative,  including the subordinating particle \textit{é}, is placed next to the negator \textit{unká} in the negative (\ref{yukuna:ex:53.33}). This change in the position of the subordinator is attested only in the context of negation, and thus, cannot be attributed to distributional features of the subordinator itself (i.e. the particle does not show features of `second position’ markers).

\ea\label{yukuna:ex:53.33}
\gll unka-\textbf{jĩ-ka} \textbf{é} 	nu=iphá \\
\textsc{neg}-\textbf{\textsc{fut}-\textsc{ev.nmlz}}  \textbf{\textsc{cond}} \textsc{1sg}=arrive \\
\glt ‘if I don't arrive {...}’ [ycn0053,33]
\z

In sum, the negative adverbial subordinate clauses described in this subsection show important asymmetries with respect to their affirmative counterparts. The asymmetries attested concern the encoding of finiteness, with overt nominalizers in the affirmative, no overt nominalizers in the negative, and a change in the placement of subordinating particles. These asymmetries highlight an important restriction in the encoding of negation, namely, that the negation suffix \textit{-la} shares the same paradigmatic slot with nominalizers like \textit{-ka} \textsc{ev.nmlz}, used in many subordinate clauses,  suggesting that it may have been itself a nominalizer in the past (see \sectref{yukuna:diachrony}).  

\subsubsection{Complement clauses}\label{yukuna:compl.cl}

The last type of subordinate clauses described here concerns complement clauses of perception verbs. Similarly to other subordinate clauses, including other complement clauses, this type of subordinates are encoded with the event nominalizer \textit{-ka} in affirmative polarity (\ref{yukuna:ex:41.25}). However, unlike other subordinates, the overt marking of nominalization is optional in these complement clauses, leading to many instances of finite subordinate clauses (\ref{yukuna:ex:151.86}).

\ea\label{yukuna:ex:41.25}
\gll Ri=amí-cha kajú na=ají-cha-ka. \\
\textsc{3sg.nf}=see-\textsc{pst} a\_lot \textsc{3pl}=fly-\textsc{pst}-\textsc{ev.nmlz} \\
\glt ‘He saw them flying a lot.’ [ycn0041,25]
\z

\ea\label{yukuna:ex:151.86}
\gll Ru\textup{\textsubscript{i}}=amí-cha kajú ru\textup{\textsubscript{j}}=apíro'-cha tá. \\
\textsc{3sg.f}=see-\textsc{pst} a\_lot \textsc{3sg.f}=suck-\textsc{pst} \textsc{emph} \\
\glt ‘She\textsubscript{i} saw that she\textsubscript{j} was eating (it) a lot.’ [ycn0151,86]
\z

These clauses are negated using the same markers as in SN \textit{unká}{..}\textit{-la}, and just like in the preceding cases, the nominalizer \textit{-ka} is absent in the negative clause (\ref{yukuna:ex:108.134}). There are no further asymmetries between the affirmative and negative complement clauses, which are, in isolation, structurally identical to main verbal negative clauses.

\ea\label{yukuna:ex:108.134}
\gll É ri=amí-cha, amá-ri \textbf{unká} líchi i'ma-\textbf{lá}-cha palá.   \\
Then \textsc{3sg.nf}=see-\textsc{pst} see-\textsc{nf} \textbf{\textsc{neg}} tobacco \textsc{cop}-\textbf{\textsc{neg}}-\textsc{pst} well \\
\glt ‘He looked and saw (that) the tobacco was not well.’ [ycn0108,134]
\z

To summarize, in \sectref{yukuna:negsub} we described four different negation strategies used in subordinate clauses. Purpose of motion clauses show no asymmetries between affirmatives and negatives (\sectref{yukuna:purp.mot}), purposive clauses (\sectref{yukuna:purp.loje}), adverbial subordinate clauses (\sectref{yukuna:other.advz}) and complement clauses (\sectref{yukuna:compl.cl}) show various types of asymmetries. The attested asymmetries involve differences in the encoding of finiteness and in the placement of subordinating particles. Crucially, we show that in the case of some adverbial subordinates, the affirmative clause is overtly nominalized, while the negative one is not, similarly to the asymmetries attested in polar questions (\sectref{yukuna:interrog}) and potential modality (\sectref{yukuna:pot}). This scenario is the opposite of the usual pattern of asymmetry in main declarative verbal clauses (A/Fin) \citep[75]{Miestamo2005}. We lack systematic typological surveys on negation in non-main clauses \citep[23--24]{Miestamo2017}, so it is unclear whether similar scenarios are attested cross-linguistically.


\subsection{Negative lexicalizations}

No negative lexicalizations used in clausal negation have been identified in Yukuna. Indeed, there are no verbal roots that have merged with negation markers, nor verbs which are negated through a specific construction. Yukuna's SN construction is extremely productive, covering all verbs in the language. The only instance of a lexicalized negative item is the interjection \textit{kaapí}, meaning `I don’t know’.

\section{Non-clausal negation}

This section describes negative constructions that do not operate at the level of the clause, namely, negative replies to polar questions (\sectref{yukuna:negrep}), negative indefinites (\sectref{yukuna:INDF}), and lastly, negative derivations (\sectref{yukuna:negder}). 

\subsection{Negative replies} \label{yukuna:negrep}

Negative replies to polar questions use either the same marking as SN with \textit{unká}{...}\textit{-la} in full negative replies, or just the negative marker \textit{unká} in one-word replies, or both (\ref{yukuna:ex:68.165-166}).

\ea\label{yukuna:ex:68.165-166}
    \begin{xlist}[Q: ]\label{yukuna:ex:68.165} 
    \exi{Q: }
        \gll Ná pi=amá=no?   é ka'jná pi=amá-ka inaá-na?   \\
        \textsc{indf} \textsc{2sg}=see=\textsc{hab} \textsc{q} \textsc{dub} \textsc{2sg}=see-\textsc{ev.nmlz} woman-\textsc{pl} \\
        \glt ‘Who did you see? Did you see any women?’ [ycn0068,165]
    \exi{A: }\label{yukuna:ex:68.166}
        \gll \textbf{Unká} nu=jiwaká, \textbf{unká} ná nu=amá-\textbf{la}. \\
        \textbf{\textsc{neg}} \textsc{1sg}=boss \textbf{\textsc{neg}} \textsc{indf} \textsc{1sg}=see-\textbf{\textsc{v.neg}} \\
        \glt ‘No, boss, I haven't seen anyone.’ [ycn0068,166]
    \end{xlist}
\z

Affirmative replies to affirmative interrogatives are introduced with \textit{ã́'a} `yes’. In the cases where the question is negative itself, affirmative and negative answers are encoded just like in answers to affirmative questions, with \textit{ã́'a} ‘yes’ followed by an affirmative clause (\ref{yukuna:ex:68.234}), and \textit{unká} `no’ followed by a negative clause (\ref{yukuna:ex:68.199}).  Answers to negative questions thus express agreement or disagreement over the propositional content of the question and not its polarity, unlike polarity reversing particles \citep{Holmberg2016,EnfieldEtAl2019}.

\ea\label{yukuna:ex:68.234}
\ea
\gll É pi=amí-cha-ka apa'wélo nu=e'we-ló? \\
\textsc{q} \textsc{2sg}=see-\textsc{pst-ev.nmlz} other.\textsc{f} \textsc{1sg}=sibling-\textsc{f} \\
\glt ‘Did you see my other sister?’ [ycn0068,234]
\ex 
\gll   “ã́'a” ké ri=ímí-cha, “kája nu=amí-cha ru=ikhá”. \\
yes like \textsc{3sg.nf}=say-\textsc{pst} already \textsc{1sg}=see-\textsc{pst} \textsc{3sg.f}=\textsc{pro} \\
\glt ‘He said “Yes, I've already seen her”.’ [ycn0068,234]
\z
\z

\ea\label{yukuna:ex:68.199}
\ea
\gll \textbf{Unká} pi=i'ma-\textbf{lá} wa=yukú-ná pi=jiwaká tá Kanumá jló? \\
\textbf{\textsc{neg}} \textsc{2sg}=tell-\textbf{\textsc{neg}} \textsc{1pl}=story-\textsc{al} \textsc{2sg}=boss \textsc{emph} K. to\\
\glt ‘Didn't you tell our story to your boss Kanumá?’ [ycn0068,199]
\ex 
\gll \textbf{Unká} nu=laké-lo, náje nu=i'má i=yukú-ná? \\
\textbf{\textsc{neg}} \textsc{1sg}=grandchild-\textsc{f} \textsc{indf}.reason \textsc{1sg}=tell \textsc{2pl}=story-\textsc{al} \\
\glt ‘No, granddaughter, why would I tell your story?’ [ycn0068,200]
\z
\z

 
\subsection{Negative indefinites and quantifiers}\label{yukuna:INDF}

Yukuna has one paradigm of indefinite pro-forms, including pronouns and pro-adverbs, that share similar functional and distributional features. These pro-forms include three different roots (\textit{ná} `nothing, nobody', \textit{náje} `no reason', \textit{mé} `no way'), and several morphologically complex forms based on pro-adverbial root \textit{mé} as summarized in \tabref{yukuna:tableINDF}.

\begin{table} 	
\caption{Indefinite pro-forms in Yukuna}				
\label{yukuna:tableINDF}				
 \begin{tabular}{ll}
\lsptoprule
Form				& Meaning \\
\midrule
\textit{ná 	}	&	nothing/nobody/who/what \\
\textit{náje 	}	&	no reason/why \\
\textit{méré	}	&	nowhere/where \\
\textit{méño'jṍ 	}	&	toward nowhere/toward where \\
\textit{méké 	}	&	no way/how \\
 \lspbottomrule				
 \end{tabular}				
\end{table}		

Functionally, these forms are used as interrogative pro-forms (\ref{yukuna:ex:63.31}) similarly to the English \textit{wh}-series, and as indefinites in negative clauses (\ref{yukuna:ex:117.97}), similarly to the English negative {indefinites} with \textit{no} (e.g. \textit{nobody}). When used in negative clauses, these pro-forms are exclusively used in combination with overt negation markers, whether SN as in (\ref{yukuna:ex:117.97}) or with non-verbal negation \textit{unká}{...}\textit{kalé} in non-verbal predicates or in constituent negation (\ref{yukuna:ex:68.124}) (see \sectref{yukuna:scope}).

\ea\label{yukuna:ex:63.31}
\gll Ná pi=la'á?   \\
\textsc{indf} \textsc{2sg}=do \\
\glt ‘What are you doing?’ [ycn0063,31]
\z

\ea\label{yukuna:ex:117.97}
\gll \textbf{Unká} ná pi=la'-\textbf{lá}.  \\
\textbf{\textsc{neg}} \textsc{indf} \textsc{2sg}=do-\textbf{\textsc{v.neg}} \\
\glt ‘You don't do anything.’ [ycn0117,97]
\z

\ea\label{yukuna:ex:68.124}
\gll \textbf{Unká} mére=ewá \textbf{kalé} nu=ata'á i=pirá. \\
\textbf{\textsc{neg}} \textsc{indf.loc}=around \textbf{\textsc{nvb.neg}} \textsc{1sg}=steal \textsc{2pl}=pet\\
\glt ‘Nowhere did I steal your pet.’ [ycn0068,43]
\z

Indefinite pro-forms show a restricted distribution in contrast to personal and demonstrative pro-forms. In negative verbal clauses, they are always pre-verbal even when encoding the P argument (\ref{yukuna:ex:117.97}) or an oblique argument (\ref{yukuna:ex:92.117}), although both of these constituents are canonically placed in post-verbal position. With respect to negation markers, indefinite pro-forms are placed after the negator \textit{unká} and before either the subject NP if there is one, or the verb carrying a S person index. Besides emphatic discourse markers, they are the only elements that can be placed in this position. 

\ea\label{yukuna:ex:92.117}
\gll \textbf{Unká} ná jló nu=yuí-\textbf{la}-je ri=ikhá. \\
\textbf{\textsc{neg}} \textsc{indf} to \textsc{1sg}=leave-\textbf{\textsc{v.neg}}-\textsc{fut} \textsc{3sg.nf}=\textsc{pro}\\
\glt ‘I will not leave it to anyone.’ [ycn0092,117]
\z

In addition to the interrogative and negative uses of these pro-forms, they are also found in contexts of hesitation (\ref{yukuna:ex:117.30}), exemplification (\ref{yukuna:ex:42.4}), or exclamation (\ref{yukuna:ex:89.69}). Crucially, however, they are not found in contexts typical of the so-called \textit{negative polarity} indefinites such as polar questions (e.g. ‘Did \textit{anyone} come?’), conditionals (e.g. ‘If \textit{anyone} comes’), the standard of comparison (e.g. ‘better than \textit{anyone}’), indirect negation (e.g. ‘I don't think \textit{anyone} will come’), nor as free choice indefinites (e.g. ‘You can do \textit{anything}’) \citep[37--53]{Haspelmath1997}.
 
\ea\label{yukuna:ex:117.30}
\gll Iná ako'ó ri=ikhá ná chojẽ́ ka'jnó? {...} galón chojẽ́ ka'jnó iná ako'ó ri=ikhá.\\
\textsc{gnr.pro} pour \textsc{3sg.nf=pro} \textsc{indf} into \textsc{dub} {} barrel into \textsc{dub} \textsc{gnr.pro} pour \textsc{3sg.nf=pro}\\
\glt ‘One pours it into what already? ... one pours it into an oil barrel.' [ycn0117,30]
\z%hc{...}

\ea\label{yukuna:ex:42.4}
\gll Wa=ajñá wa=a'jne-wá, ná kujnú ka'jnó, jíña ka'jnó {...}  \\
\textsc{1pl}=eat \textsc{1pl=}food-\textsc{refl} \textsc{indf} cassava \textsc{dub} fish \textsc{dub} \\
\glt ‘We eat our food, like cassava, or maybe fish...’ [ycn0042,4]
\z

\ea\label{yukuna:ex:89.69}
\gll Méké palá wani! \\
\textsc{indf}.manner good \textsc{emph}  \\
\glt ‘How good!’ [ycn0089,69]
\z

The only clear-cut instances of an indefinite pronoun used in declarative affirmative clauses is the complex form \textit{nákájé} ‘something’ (\ref{yukuna:ex:92.142}).\footnote{This marker is clearly related to \textit{ná}, but it is not used as an interrogative nor as a negative pronoun.} There are no instances in which the indefinite pro-forms used in negation are found in affirmative declarative clauses with a meaning similar to the \textit{some-} series in English as would be expected of so-called \textit{neutral} indefinites \citep{AuweraAlsenoy2018}. Indeed, in order to refer to indefinite humans in affirmative clauses, the generic noun \textit{inau'ké} ‘person’ is used (\ref{yukuna:ex:53.59}). There is a generic pronoun \textit{iná} ‘one’, but its meaning is collective, rather than indefinite.  

\ea\label{yukuna:ex:92.142}
\gll [...] ri=jña'á ri=jló-wa nákájé [...] \\
{} \textsc{3sg.nf}=grab \textsc{3sg.nf}=for-\textsc{refl} something \\
\glt ‘[...] he grabbed something for himself [...]' [ycn0092,142]
\z

\ea\label{yukuna:ex:53.59}
\gll É ru=amí-cha amí-cha-ri pajlúwa inau'ké.\\
then \textsc{3sg.f}=see-\textsc{pst} see-\textsc{pst-nf} one person\\
\glt ‘Then she looked and saw a person.' [ycn0053,59]
\z

In sum, the markers described here have two core functions, as interrogatives and as indefinites used in negation. Typologically, they can be classified in the type “interrogative-based negative indefinites” of Haspelmath's \citeyear{Haspelmath2013-wals115} typology, although there is absolutely no change in form between the two functions. In addition to these two core functions, these pro-forms are used in various other contexts (hesitation, exemplification, exclamation), but they are absent from declarative affirmative clauses. The functions of these pro-forms make it difficult to categorize them among the types of indefinites proposed by \citet{Kahrel1996}, namely, neutral, polar or negative indefinites. 

\subsection{Negative derivation}\label{yukuna:negder}

Yukuna has a privative marker \textit{ma-}, that derives adjectives from nouns (N-less). This prefix is well attested in many other Arawakan languages, and has been reconstructed to Proto-Arawakan \citep[377--378]{Payne1991}. Similar markers are also cross-linguistically common, as shown in \citet[]{Veselinova2013neg.exist}. The derived adjective is then marked for number with \textit{-ru} in the singular, and \textit{-runa} in the plural.\footnote{Number marking on adjectives derived with \textit{ma-} differs from gender/number marking on other loci as typically the suffix \textit{-ru} is used to encode feminine gender in the singular.} Syntactically, these adjectives function like any other adjective in Yukuna: as noun modifiers (\ref{yukuna:ex:nb5.51}), attributive predicates (\ref{yukuna:ex:41.125}), secondary predicates in verbal clauses (\ref{yukuna:ex:41.153}), and even on their own without a head noun. 

\ea\label{yukuna:ex:nb5.51}
\gll Nu=e’wé ma-ijlu-rú waícha chúwajã́.\\
\textsc{1sg}=sibling \textsc{priv}-eye-\textsc{sg} come today \\
\glt ‘My blind brother came today.’ [Elicited, NB5:51]
\z

\ea\label{yukuna:ex:41.125}
\gll Ma-chijné-rú ri=i'mi-chá.\\
\textsc{priv}-hair-\textsc{sg} \textsc{3sg.nf}=\textsc{cop}-\textsc{pst} \\
\glt ‘He was hairless.’ [ycn0041,125]
\z 

\ea\label{yukuna:ex:41.153}
\gll Yewákumi tá yuí-cha-ri=o ma-ijlu-rú, unká ri=ijlú i'ma-lá-cha.\\
Yewakúmi \textsc{emph} stay-\textsc{pst}-\textsc{nf}=\textsc{mid} \textsc{priv}-eye-\textsc{sg} \textsc{neg} \textsc{3sg.nf}=eye \textsc{cop-neg-pst}\\
\glt ‘It is Yewakúmi who became eyeless, he didn't have eyes.’ [ycn0041,153]
\z 

This marker encodes the absence of an entity, and is very similar to negative existential and possessive predicates (\ref{yukuna:ex:41.153}). The opposite meaning is encoded with attributive prefix \textit{ka-}, that also derives nouns into adjectives meaning `with N’. 

\section{Other aspects of negation}

This section explores further aspects of negation in Yukuna: the scope of negation (\sectref{yukuna:scope}), the reinforcement of negation (\sectref{yukuna:emph}), negation in complex clauses (\sectref{yukuna:complex}), and some final issues in (\sectref{yukuna:other}). 

\subsection{Scope of negation}\label{yukuna:scope}

In order to narrow the scope of negation to specific constituents, Yukuna uses the non-verbal negation strategy \textit{unká}{...}\textit{kalé} around the negated constituent, while the main verb remains in the affirmative. Very few constituents may be negated in verbal clauses, most often locative adverbs, as shown in example (\ref{yukuna:ex:108.87}) repeated here as (\ref{yukuna:ex:108.87b}) with locative adverb \textit{ya'jnájẽ} ‘far', and indefinite pro-adverbs as in (\ref{yukuna:ex:53.82}). 

\ea\label{yukuna:ex:108.87b}
\gll \textbf{Unká} ya'jnájẽ \textbf{kalé} kháãjĩ́-una i'jná.\\
\textbf{\textsc{neg}} far.towards \textbf{\textsc{nvb.neg}} \textsc{dem.prox}-\textsc{pl} go\\
\glt ‘These ones did not go far.’ (lit. ‘These ones went not far.’) [ycn0108,87]
\z

\ea\label{yukuna:ex:53.82}
\gll \textbf{Unká} náje \textbf{kalé} pi=i'jná nu=jwa'té. \\
\textbf{\textsc{neg}} \textsc{indf}.reason \textbf{\textsc{nvb.neg}} \textsc{2sg}=go \textsc{1sg}=with\\
\glt ‘You don't have to come with me.’ (lit. ‘You come with me in no way.’) [ycn0053,82]
\z

This non-verbal negation strategy used in constituent negation is also the preferred strategy used in contrastive negation, as with the negative non-verbal predicate in (\ref{yukuna:ex:91.37}). This strategy is further discussed in (\sectref{yukuna:complex}), and (\sectref{yukuna:meta}).

\ea\label{yukuna:ex:91.37}
\gll \textbf{Unká} inaá-na le'jé \textbf{kalé} kéelé mawíru, achiñá-ná le'jé, wa=le'jé ri=ikhá. \\
\textbf{\textsc{neg}} woman-\textsc{pl} possession \textbf{\textsc{nvb.neg}} \textsc{med} pineapple man-\textsc{pl} possession \textsc{1pl}=possession \textsc{3sg.nf=pro}\\
\glt ‘That pineapple is not women's possession, it is men's possession, it is ours.' [ycn0091,37]
\z

There are no attested instances of finite verbal clauses with negated core constituents in the corpus. In fact, elicited data reveals that it is agrammatical to negate S NPs with the non-verbal negation strategy in finite clauses. This is shown by the elicited grammaticality judgment example in (\ref{yukuna:ex:nb8.135}). The only way to narrow the scope of negation to this constituent is by using a pseudo-cleft containing a nominalized verb form instead (\ref{yukuna:ex:nb8.136}). More data is needed to know whether this restriction applies to both S NPs and O NPs, or to S NPs only. 

\ea[*]{\label{yukuna:ex:nb8.135}
    \gll \textbf{Unká} ru=i'rí \textbf{kalé} ií-cha. \\
    \textbf{\textsc{neg}} \textsc{3sg.f}=son \textbf{\textsc{nvb.neg}} cry-\textsc{pst}\\
    \glt (‘\textit{Her son} did not cry.') [elicitation, NB8,135]}
\z

\ea[]{\label{yukuna:ex:nb8.136}
\gll \textbf{Unká} ru=i'rí \textbf{kalé} kéelé ií-cha-ri. \\
\textbf{\textsc{neg}} \textsc{3sg.f}=son \textbf{\textsc{nvb.neg}} \textsc{dem.med} cry-\textsc{pst}-\textsc{nf} \\
\glt ‘It was not her son the one who cried.' [elicitation, NB8,136]}
\z

Although there are no attested cases of core constituents negated with the non-verbal negation strategy in finite verbal clauses, it is possible to focalize both core constituents in negative clauses marked with SN, by placing the focalized constituent before the negative particle \textit{unká}. This is shown with the focalized S NP in (\ref{yukuna:ex:189.67}). This is the only instance in which a S NP is not immediately adjacent to the finite verb in Yukuna.

\ea\label{yukuna:ex:189.67}
\gll  É na=apó-chi-ya=o phiyúké, ri=ikhá \textbf{unká} apó-\textbf{la}-cha. \\
then \textsc{3pl}=wake\_up-\textsc{caus}-\textsc{pst}=\textsc{mid} all \textsc{3sg.nf}=\textsc{pro} \textbf{\textsc{neg}} wake\_up-\textbf{\textsc{v.neg}}-\textsc{pst}\\
\glt ‘Then they all woke up, (but) \textit{he} did not wake up.’ [ycn0189,67]
\z

\subsection{Negative polarity}\label{yukuna:negpol}

The domains of negative polarity items are encoded in Yukuna with a series of negative/interrogative pro-forms, as described in \sectref{yukuna:INDF}. Note that these pro-forms also appear in contexts which are not typically associated with negative polarity. 

\subsection{Marking of NPs in the scope of negation}\label{yukuna:negnp}

No negation construction appears to affect the marking of NPs in Yukuna. 

\subsection{Reinforcing negation}\label{yukuna:emph}


Yukuna has a variety of intensifier particles such as \textit{wáni}, \textit{tá}, \textit{chí}, \textit{kája}, and others. All of these markers have subtle differences in meaning and distribution, but broadly speaking, they are found with all constituent types, in both affirmative and negative polarity. When used with SN, they are placed immediately after the negator \textit{unká}, and before the S NP and the verb (\ref{yukuna:ex:101.6}). 

\ea\label{yukuna:ex:101.6}
\gll Unká tá wa=jilá i'ma-lá.\\
\textsc{neg} \textsc{emph} \textsc{1pl}=oil \textsc{cop=neg}\\
\glt ‘We don't have any oil at all.' [ycn0101,6]
\z

\subsection{Negation, coordination and complex clauses}\label{yukuna:complex}

As shown in (\sectref{yukuna:negsub}), subordinate clauses differ on the specific negation strategy they use. Another difference concerns whether the complex clause allows negation to be marked on either the main or on the subordinate clause, or whether only one of them is marked for negation. The former case is shown by purpose of motion clauses, where negative polarity can be marked on the main verb (\ref{yukuna:ex:119.29}), or on the subordinate verb as in (\ref{yukuna:ex:58.28}) repeated here as (\ref{yukuna:ex:58.28b}). Semantically, the former implies that the motion action did not take place, whereas the latter implies that the motion action did occur. 

\ea\label{yukuna:ex:119.29}
\gll Kája \textbf{unká} iná i'jna-\textbf{lá} matha'-jẽ́.\\
Then \textbf{\textsc{neg}} \textsc{gnr.pro} go-\textbf{\textsc{neg}} cut-\textsc{purp.mot} \\
\glt ‘Then one does not go to cut.’ [ycn0119,29]
\z

\ea\label{yukuna:ex:58.28b}
\gll \textbf{Unká} kapichá-je=o \textbf{kalé} pi=i'jná, i'ma-jé i=i'jná.\\
\textbf{\textsc{neg}} die-\textsc{purp.mot}=\textsc{mid} \textbf{\textsc{nvb.neg}} \textsc{2sg}=go live-\textsc{purp.mot} \textsc{2pl}=go\\
\glt ‘You go (there) not to die, you go (there) to live.’ [ycn0058,28] 
\z

In contrast, in complex sentences with volition verbs, negation marking is only attested on the main verb, and not on the subordinate clause, as in (\ref{yukuna:ex:nb5.482}). Note that in this case, the negation of the main clause has scope over the negative indefinite \textit{ná} in the subordinate clause.  

\ea\label{yukuna:ex:nb5.482}
\gll Chúwa \textbf{unká} nu=wáta-\textbf{la} pi=a'-ká nu=jló ná tá!\\
now \textbf{\textsc{neg}} \textsc{1sg}=want-\textbf{\textsc{neg}} \textsc{2sg}=give-\textsc{ev.nmlz} \textsc{1sg}=to \textsc{indf} \textsc{emph}\\
\glt ‘Now I don't want you to give me anything!’  [elicited, NB5:482]
\z

Lastly, the marking of negation may also show irregularities in coordination. In sequences of negative verbal clauses, the first one is marked with SN \textit{unká}{...}\textit{-la}, while the following clauses omit \textit{unká} and retain only \textit{-la}, as in (\ref{yukuna:ex:545.37}).

\ea\label{yukuna:ex:545.37}
\gll Náje \textbf{unká} pi=ña'-tá-\textbf{la}-cha, pi=ka'-tá-\textbf{la}-cha kawáká ri=ikhá tá, pi=nó-\textbf{la}-cha ka'jnó ri=ikhá tá?!\\
\textsc{indf.}reason \textbf{\textsc{neg}} \textsc{2sg}=hit-\textsc{caus}-\textbf{\textsc{neg}}-\textsc{pst} \textsc{2sg}=throw-\textsc{caus}-\textbf{\textsc{neg}}-\textsc{pst} down \textsc{3sg.nf=pro} \textsc{emph} \textsc{2sg}=kill-\textbf{\textsc{neg}}-\textsc{pst} \textsc{dub} \textsc{3sg.nf=pro} \textsc{emph}\\
\glt ‘Why didn't you hit him, throw him down, kill him?!’  [ycn0545,37]
\z


\subsection{Miscellaneous aspects of negation}\label{yukuna:other}

\subsubsection{Negative transport}\label{yukuna:transport}

Negative transport or negative raising refers to the process whereby a higher-clause negative is
interpreted as negating a lower-clause predicate. In Yukuna, the negation strategy used in complex sentences varies according to each subordinate construction, as presented in \sectref{yukuna:complex}.  

\subsubsection{Metalinguistic negation}\label{yukuna:meta}

Metalinguistic negation, or the device used for objecting to a previous utterance \citep[363]{Horn1989}, is encoded in Yukuna with non-verbal negation, just like contrastive negation described in (\sectref{yukuna:scope}). This device is often used by repeating a non-verbal constituent mentioned by the interlocutor, in order to emphasize the speaker's disagreement over the interlocutor's wording. In (\ref{yukuna:ex:58.13}), speaker B quotes the indefinite pronoun \textit{ná} used by speaker A, and negates it using constituent negation, as he believes that the answer to the question is obvious. Note that in doing so, the speaker B uses the indefinite pronoun in a syntactic position typical of lexical nouns such as \textit{pána} `leaf'. 

\ea\label{yukuna:ex:58.13}
    \begin{xlist}[B: ]
    \exi{A: }\label{yukuna:ex:58.13a}
        \gll Ná chí ri=iĩ́? kéelé íju pi=kulá.\\
        \textsc{indf} \textsc{emph} \textsc{3sg.nf}=name \textsc{dem.med}  leaf \textsc{2sg}=search\\
        \glt ‘What is its name? That leaf you search.’ [ycn0058,13]
    \exi{B: }\label{yukuna:ex:58.13b}
        \gll Unká “ná” kalé, wayú pána ri=ikhá. \\
        \textsc{neg} \textsc{indf} \textsc{nvb.neg} bird\_species leaf \textsc{3sg.nf=pro} \\ 
        \glt ‘It is not a “what”, it is the leaf of the black vulture.’ [ycn0058,14]
    \end{xlist}
\z

\subsubsection{Non-negative uses of negatives}

The negation particle \textit{unká} is found in several idiomatic, synchronically unsegmentable forms used with non negative meanings: \textit{unkáchiyó} ‘before’, \textit{unkájĩchiyó}  ‘almost’, \textit{unká kéla} ‘otherwise’ (\ref{yukuna:ex:53.24}).

\ea\label{yukuna:ex:53.24}
\gll Unkájĩ́chiyó ri=nó-cha jlá ri=ikhá!\\
almost \textsc{3sg.nf}=kill-\textsc{pst} \textsc{intv} \textsc{3sg.nf=pro} \\
\glt ‘He almost killed him!’  [ycn0053,24]
\z

\subsubsection{Diachronic notes and observations}\label{yukuna:diachrony}

One of the puzzling features of negation in Yukuna concerns the diachrony of the negator \textit{unká}. This marker is used in almost all negation strategies, most notably, SN (\textit{unká}{...}\textit{-la}) and non-verbal negation (\textit{unká}{...}\textit{kalé}). 
Concerning SN, we posit that a plausible diachronic scenario is that \textit{unká} was originally a negative existential stative verb, which was used as a main clause predicate with a subordinate verb nominalized with \textit{-la}. This initially bi-clausal construction later grammaticalized as a monoclausal SN strategy through the Negative Existential Cycle \citep{Veselinova2016}. According to this diachronic hypothesis, the properties of modern Yukuna's negated verbs, and negative clauses, reflect the morphosyntax of pre-Yukuna's bi-clausal constructions. This diachronic hypothesis is based on the shared paradigmatic position of the negative suffix \textit{-la} and clausal nominalizers such as \textit{-ka} \textsc{ev.nmlz}, and provides a likely account for Yukuna's typologically unusual cases of asymmetry in finiteness in subordinate clauses, whereby the negative shows reduced finiteness with respect to the affirmative (see \sectref{yukuna:other.advz}). The former nominalizing function of \textit{-la} would also explain its incompatibility with the far past imperfective suffix \textit{-khe} in main clauses (see \sectref{yukuna:SN}). Lastly, the proposed diachronic scenario is very likely from an Arawakan perspective, as nominalizations used as a subordinating strategy are widespread in the family \citep[100]{Aikhenvald1999}.

\section{Summary}
In this chapter, we provided a detailed overview of negation in Yukuna, following the typological questionnaire by \textcitetv{chapters/questionnaire} revised with Ljuba Veselinova in February 2019. We showed that Yukuna clearly distinguishes verbal and non-verbal clauses, as reported in other Arawakan languages such as Mojeño Trinitario \citep{Rose2018}, and that the two main negation strategies in the language SN \textit{unká{...}-la} and non-verbal negation \textit{unká{...}kalé}, perfectly map onto this distinction.
We showed that the same markers of SN and non-verbal negation were employed in most other domains, from negation in non-declaratives, to subordinate clauses, and constituent negation. Some exceptional domains with distinct negation markers are the prohibitive and negative purposive subordinate clauses. The Yukuna negative markers are summarized in \tabref{yukuna:tab:summary}.

\begin{table}									
\caption{Markers of negation in Yukuna}	\label{yukuna:tab:summary}									
\begin{tabular}{lll} 									
\lsptoprule									
Domain	&	Function	&		Markers		\\
\midrule									
Clausal	&	Standard Negation	&	\textit{unká ...-la} \\
	&	Interrogative 	&	\textit{unká ...-la} \\
	&	Potential mood	&	\textit{unká ...-la} \\
	&	Non-verbal predication	&	\textit{unká ... kalé} \\
	&	Purpose of motion clauses	&	\textit{unká ... kalé} \\
	&	Constituent negation	&	\textit{unká ... kalé} \\
	&	Imperative	&	\textit{-niña}		\\
	&	Purposive clauses	&	\textit{piyá} \\
Non-clausal	&	Negative replies	&	\textit{unká} \\
	&	Negative derivation	&	\textit{ma-} \\
\lspbottomrule									
\end{tabular}									
\end{table}			

In terms of (a)symmetries, we showed that both SN and non-verbal negation are symmetrical with respect to the corresponding affirmative verbal and non-verbal predicates. However, there are several cases of constructional asymmetries in non-declarative and non-main clauses. Polar questions and potential modality clauses show an interesting alternation between nominalized verb forms in the affirmative, and finite verb forms marked with the SN markers in the negative. Similarly, certain adverbial subordinate clauses are encoded with a nominalized verb and a post-verbal subordinating particle in the affirmative, while in the negative, the verb is not nominalized, and the subordinating particle is placed after the negator \textit{unká} rather than after the subordinate verb. We  argued that these asymmetries in the marking of nominalization are interesting from a diachronic perspective, as they suggest that the origin of SN is a construction with a negative existential verb \textit{unká} and a nominalized verb form with \textit{-la}. 
Lastly, from a genealogical perspective, Yukuna shares some features with closely related languages, while showing some particularities. Indeed, although negation varies greatly in terms of both form and function within the family \citep[96]{Aikhenvald1999}, some  related languages show formally and functionally similar forms, such as the the pre-verbal particle \textit{uká} in Kawiyarí, and  \textit{noka} in Baure \citep[238]{Michael2014}. However, according to \citet[237]{Michael2014}, Arawakan languages tend to use a single pre-verbal marker in SN (21 out of 27 languages in his sample), whereas obligatory double-marking, specially involving a particle and a suffix, is quite rare. Yukuna is therefore marginal in that respect.

\section*{Acknowledgements}
\largerpage
We send our deepest thanks to all the Yukuna and Matapí collaborators who have patiently taught their language to the first author, in particular, to Virgelina Matapí Yucuna. We are also grateful to Olga Krasnoukhova for her valuable suggestions during the writing of this paper. We thank ELDP and Labex ASLAN for their financial support and lastly, we address our sincere gratitude to Matti Miestamo and Ljuba Veselinova for inviting us to be part of this book. 


\section*{Abbreviations}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape 1 & first person \\
\scshape 2 & second person \\
\scshape 3 & third person \\
\scshape al & alienable \\
\scshape caus & causative \\
\scshape cause & cause clause \\
\scshape cond & conditional  \\
\scshape cop 	& copula \\ 
\scshape dem 	& demonstrative \\ 
\scshape dub 	& dubitative \\ 
\scshape emph & emphatic particle  \\
\scshape ev 	& event \\ 
\scshape f 	& feminine \\	
\scshape fut & future \\
\scshape gnr	& generic \\
\scshape hab & habitual aspect  \\
\scshape indf 	& indefinite \\
\scshape intv & intentive \\
\scshape ipfv 	& imperfective \\
\scshape loc & locative  \\
\end{tabularx}
\begin{tabularx}{.45\textwidth}{lQ}
\scshape med	& medial \\			
\scshape mid	& middle \\
\scshape nf 	& non-feminine	\\
\scshape nvb & nonverbal \\   
\scshape neg & negative  \\
\scshape nmlz & nominalizer \\
\scshape nposs 	& unpossessed \\
%\scshape p & patient  \\   ??%to recheck what we do.
\scshape part & partitive \\  
\scshape pst & past \\
\scshape pl & plural  \\
\scshape priv 	& privative \\ 				
\scshape pro 	& pronoun \\		
\scshape proh 	& prohibitive \\	
\scshape prox 	& proximal \\ 
\scshape purp.mot & 	purpose of motion \\ 
\scshape q & polar question particle  \\
\scshape refl & reflexive \\
\scshape sg & singular  \\
\scshape v 	& verbal \\ 

\end{tabularx}


\sloppy\printbibliography[heading=subbibliography,notkeyword=this]
\end{document}
